### 2022-05-13 notebooks/Linear_algebra.Rmd
Initial commit. Includes first version of the article draft, along with several notebooks and ancillary output documents.

---
title: "Linear algebra"
output:
  html_notebook:
    number_sections: yes
---

Following this document:
https://www.math.hkust.edu.hk/~mabfchen/Math111/Week13-14.pdf

# Dot product of $\mathbb{R}^n$

All clear here.

# Inner product spaces

**Example 2.1 (adapted).**

For
$\mathbf{\theta}_1 = \begin{bmatrix}\theta_{11}\\ \theta_{12}\\ \end{bmatrix}$,
$\mathbf{\theta}_2 = \begin{bmatrix}\theta_{21}\\ \theta_{22}\\ \end{bmatrix}$
$\in \mathbb{R}^2$,
define

$$
\langle \mathbf{\theta}_1, \mathbf{\theta}_2 \rangle =
  \tfrac{4}{3} \theta_{11} \theta_{21} -
    \tfrac{2}{3} \theta_{11} \theta_{22} -
    \tfrac{2}{3} \theta_{12} \theta_{21} +
    \tfrac{4}{3} \theta_{12} \theta_{22}.
$$

Then $\langle,  \rangle$ is an inner product on $\mathbb{R}^2$.
The properties fulfill (see handwritten notebook "Linear algebra").

For each $\theta \in \Theta$, the **norm** or **length** of $\theta$ is
defined as

$$
\lVert \mathbf{\theta} \rVert :=
  \sqrt{ \langle \mathbf{\theta}, \mathbf{\theta} \rangle }
$$

# Examples of inner product spaces

**Example 3.1.**

**IMPORTANT:**
The vector space $\mathbb{R}^n$ with this special inner product (dot product)
is called the **Euclidean $n$-space**, and the dot product is called the
**standard inner product** on $\mathbb{R}^n$.

# Representation of inner product

**Theorem 4.1 (adapted).** Let $\Theta$ be an $n$-dimensional vector space with
an inner product $\langle, \rangle$, and let $\Sigma^{-1}$ be the matrix of
$\langle, \rangle$ relative to a basis $\mathcal{B}$.
Then for any vectors $\mathbf{\theta}_1', \mathbf{\theta}_2' \in \Theta$,

$$
\langle \mathbf{\theta}_1', \mathbf{\theta}_2' \rangle =
  \mathbf{\theta}_1^T \Sigma^{-1} \mathbf{\theta}_2,
$$

where $\mathbf{\theta}_1$ and $\mathbf{\theta}_2$ are the coordinate vectors
of $\mathbf{\theta}_1'$ and $\mathbf{\theta}_2'$, respectively, i.e.,
$\mathbf{\theta}_1 = [\mathbf{\theta}_1']_\mathcal{B}$ and
$\mathbf{\theta}_2 = [\mathbf{\theta}_2']_\mathcal{B}$.

**Example 4.1 (adapted).** For the inner product of $\mathbb{R}^2$ defined by

$$
\langle \mathbf{\theta}_1', \mathbf{\theta}_2' \rangle =
  \frac{4}{3} \theta_{11} \theta_{21} -
    \frac{2}{3} \theta_{11} \theta_{22} -
    \frac{2}{3} \theta_{12} \theta_{21} +
    \frac{4}{3} \theta_{12} \theta_{22},
$$

where
$\mathbf{\theta}_1 = \begin{bmatrix} \theta_{11}\\ \theta_{12}\\ \end{bmatrix}$,
$\mathbf{\theta}_2 = \begin{bmatrix} \theta_{21}\\ \theta_{22}\\ \end{bmatrix}$
$\in \mathbb{R}^2$,
its matrix relative to the standard basis
$\mathcal{E} = \{ \mathbf{e}_1, \mathbf{e}_2 \}$ is

$$
\Sigma^{-1} = \begin{bmatrix}
    \langle \mathbf{e}_1, \mathbf{e}_1 \rangle &
      \langle \mathbf{e}_1, \mathbf{e}_2 \rangle \\
    \langle \mathbf{e}_2, \mathbf{e}_1 \rangle &
      \langle \mathbf{e}_2, \mathbf{e}_2 \rangle \\
  \end{bmatrix} =
  \begin{bmatrix}
     \tfrac{4}{3} & -\tfrac{2}{3} \\
    -\tfrac{2}{3} &  \tfrac{4}{3} \\
  \end{bmatrix}
$$

(see handwritten notebook "Linear algebra").

With a change of variables:

$$
\Bigg\{
\begin{array} \\
  \theta_{11} =             \theta_{11}' \\
  \theta_{12} = \cos \alpha \theta_{11}' + \sin \alpha \theta_{12}'
\end{array}
$$

$$
\Bigg\{
\begin{array} \\
  \theta_{21} =             \theta_{21}' \\
  \theta_{22} = \cos \alpha \theta_{21}' + \sin \alpha \theta_{22}'
\end{array}
$$

we have (see handwritten notebook "Linear algebra")

$$
\langle \mathbf{\theta}_1', \mathbf{\theta}_2' \rangle =
  \theta_{11}'\theta_{21}' + \theta_{12}'\theta_{22}' =
  \mathbf{\theta}_1'^T \mathbf{\theta}_2'.
$$

This is equivalent to choosing a new basis

$$
\mathcal{B} = \left\{
  \mathbf{u}_1 = \begin{bmatrix} 1\\ \cos \alpha\\ \end{bmatrix},
  \mathbf{u}_2 = \begin{bmatrix} 0\\ \sin \alpha\\ \end{bmatrix}
\right\}
$$

so that the inner product relative to it is the identity matrix, i.e.
(see handwritten notebook "Linear algebra"),

$$
\begin{bmatrix}
  \langle \mathbf{u}_1, \mathbf{u}_1 \rangle &
    \langle \mathbf{u}_1, \mathbf{u}_2 \rangle \\
  \langle \mathbf{u}_2, \mathbf{u}_1 \rangle &
    \langle \mathbf{u}_2, \mathbf{u}_2 \rangle \\
\end{bmatrix} =
\begin{bmatrix}
  1 & 0 \\
  0 & 1 \\
\end{bmatrix}.
$$

Let $\mathbf{P}$ the transition matrix from the standard basis
$\{ \mathbf{e}_1, \mathbf{e}_2 \}$ to the basis
$\{ \mathbf{u}_1, \mathbf{u}_2 \}$, i.e.,

$$
[ \mathbf{u}_1, \mathbf{u}_2 ] = [ \mathbf{e}_1, \mathbf{e}_2 ] \mathbf{P} =
  [ \mathbf{e}_1, \mathbf{e}_2 ]
  \begin{bmatrix}
         1      &      0      \\
    \cos \alpha & \sin \alpha \\
  \end{bmatrix}
$$

Let $\mathbf{\theta}_1$ be the coordinate vector
of the vector $\mathbf{\theta}_1'$ relative to the basis $\mathcal{B}$.
(The coordinate vector of $\mathbf{\theta}_1'$ relative to the standard basis is
$\mathbf{\theta}_1'$ itself.)
Then

$$
\mathbf{\theta}_1' = [ \mathbf{e}_1, \mathbf{e}_2 ] \mathbf{\theta}_1' =
  [ \mathbf{u}_1, \mathbf{u}_2 ]                    \mathbf{\theta}_1  =
  [ \mathbf{e}_1, \mathbf{e}_2 ] \mathbf{P}         \mathbf{\theta}_1,
$$

and it follows that

$$
\mathbf{\theta}_1' = \mathbf{P} \mathbf{\theta}_1.
$$

Similarly,

$$
\mathbf{\theta}_2' = \mathbf{P} \mathbf{\theta}_2.
$$

Note that $\mathbf{\theta}_1'^T = \mathbf{\theta}_1^T \mathbf{P}^T$,
so by Theorem 4.1,

$$
\langle \mathbf{\theta}_1',         \mathbf{\theta}_2' \rangle =
  \mathbf{\theta}_1'^T \mathbf{I}_n \mathbf{\theta}_2'         =
  \mathbf{\theta}_1'^T              \mathbf{\theta}_2',
$$

on one hand, and

$$
\langle \mathbf{\theta}_1', \mathbf{\theta}_2' \rangle                       =
  \mathbf{\theta}_1^T               \Sigma^{-1}            \mathbf{\theta}_2 =
  \mathbf{\theta}_1'^T \mathbf{P}^T \Sigma^{-1} \mathbf{P} \mathbf{\theta}_2'
$$

on the other.

**Theorem 4.2.** Let $\Theta$ be a finite-dimensional inner product space.
Let $\Sigma^{-1}$, $\mathbf{B}$ be matrices of the inner product relative to
bases $\mathcal{B}$, $\mathcal{B}'$ of $\Theta$, respectively.
If $\mathbf{P}$ is the transition matrix
from $\mathcal{B}'$ to $\mathcal{B}$[^trans-matrix-basis],
then

[^trans-matrix-basis]: According to the document I'm following,
the transition matrix if from basis $\mathcal{B}$ to $\mathcal{B}'$,
but to my understanding, in the previous example this transition matrix
is from the standard basis to $\mathcal{B}$, and so it should be here
(as in my example, the standard basis $\mathcal{E}$
plays the role of $\mathcal{B}'$ in this theorem).

$$
\mathbf{B} = \mathbf{P}^T \Sigma^{-1} \mathbf{P}.
$$

**Application of Theorem 4.2. to my bersion of example 4.1.**

In my case, the standard basis $\mathcal{E}$ plays the role of $\mathcal{B}'$
in Theorem 4.2. The inner product matrix relative to it is $\mathbf{I}_n$,
and the transition matrix from $\mathcal{E}$ to $\mathcal{B}$ is

$$
\mathbf{P} = \begin{bmatrix}
         1      &      0      \\
    \cos \alpha & \sin \alpha \\
  \end{bmatrix}
$$

It is true that:

$$
\mathbf{I}_n = \mathbf{P}^T \Sigma^{-1} \mathbf{P}
$$

(see handwritten notebook for derivation).

# Cauchy-Schwarz inequality

**IMPORTANT HERE!**

The **angle $\alpha$ between $u$ and $v$** is defined by

$$
\cos \alpha = \frac{
    \langle \mathbf{u}, \mathbf{v} \rangle
  } {
    \lVert \mathbf{u} \rVert \lVert \mathbf{v} \rVert
  }.
$$

The angle exists and is unique.

(**IMPORTANT NOTE!!** $\cos \alpha$ apparently is independent of the inner
product, i.e., which basis it is computed in).

# Orthogonality

Let $\Theta$ be an inner product space.
Two vectors $\mathbf{u}$ and $\mathbf{v}$ are said to be **orthogonal** if

$$
\langle \mathbf{u}, \mathbf{v} \rangle = 0
$$

# Orthogonal sets and bases

Let $\Theta$ be an inner product space.
A subset $S = {\mathbf{u}_1, \mathbf{u}_1, \dots, \mathbf{u}_k}$
of nonzero vectors of $\Theta$ is called an **orthogonal set**
if every pair of vectors are orthogonal, i.e.,

$$
\langle \mathbf{u}_i, \mathbf{u}_j \rangle = 0, \, 1 \le i < j \le k.
$$

An orthogonal set $S = {\mathbf{u}_1, \mathbf{u}_1, \dots, \mathbf{u}_k}$
is called an **orthonormal set** if we further have

$$
\lVert \mathbf{u}_i \rVert = 1, \, 1 \le i \le k.
$$

An **orthonormal basis** or $\Theta$ is a basis which is also
an orthonormal set.

# Orthogonal projection

Let $\Theta$ be an inner product space.
Let $\mathbf{\theta}_1$ be a nonzero vector of $\Theta$.
We want to decompose an arbitrary vector $\mathbf{y}$ into the form

$$
\mathbf{y} = \alpha \mathbf{\theta}_1 + z, \,
\mathrm{where} \, \mathbf{z} \in \mathbf{\theta}_1^\perp.
$$

Since $z \perp \mathbf{\theta}_1$, we have

$$
\langle \mathbf{\theta}_1, \mathbf{y} \rangle =
  \langle \mathbf{\theta}_1, \alpha \mathbf{\theta}_1 \rangle =
  \alpha \langle \mathbf{\theta}_1, \mathbf{\theta}_1 \rangle.
$$

This implies that

$$
\alpha = \frac{
    \langle \mathbf{\theta}_1, \mathbf{y} \rangle
  }
  {
    \langle \mathbf{\theta}_1, \mathbf{\theta}_1 \rangle
  }.
$$

We define the vector

$$
\mathrm{Proj}_{\mathbf{\theta}_1}(\mathbf{y}) = \frac{
    \langle \mathbf{\theta}_1, \mathbf{y} \rangle
  }
  {
    \langle \mathbf{\theta}_1, \mathbf{\theta}_1 \rangle
  } \mathbf{\theta}_1,
$$

called the **orthogonal projection of $\mathbf{y}$ along $\mathbf{\theta}_1$**.
The linear transformation
$\mathrm{Proj}_{\mathbf{\theta}_1}: \Theta \rightarrow \Theta$
is called the
**orthogonal projection of $\Theta$ onto the direction $\mathbf{\theta}_1$**.

(Example 8.1 clear, and then generalization of the orthogonal projection from
a vector to a subspace. But note this is in an Euclidean space, i.e.,
a vector space with *standard inner product*.)

```{r, include=FALSE}
alpha <- pi / 3
rho   <- cos(alpha)

u_1 <- c(1, rho)
u_2 <- c(0, sin(alpha))

R     <- rbind(u_1, u_2) # R^T = [u_1, u_2]
R_inv <- R |> solve() |> unname()
R_inv_alt <- matrix(c(1, 0, -rho / sin(alpha), 1 / sin(alpha)), nrow = 2)
all.equal(R_inv, R_inv_alt) # Son iguales

Sigma <- t(R) %*% R

Sigma_inv <- solve(Sigma)

all.equal(Sigma_inv, R_inv %*% t(R_inv)) # Son iguales
```

# Adaptation to the case in hand {-}

Let's assume that the inner product of $\mathbb{R}^2$ relative to basis
$\mathcal{B} = \{ \mathbf{u}_1, \mathbf{u}_2 \}$,
where
$\mathbf{u}_1 = \begin{bmatrix} 1\\ -\frac{\cos \alpha}{\sin \alpha} \\ \end{bmatrix}$
and
$\mathbf{u}_2 = \begin{bmatrix} 0\\ \frac{1}{\sin \alpha} \\ \end{bmatrix}$,
is defined by

$$
\langle \mathbf{x}, \mathbf{y} \rangle = \frac{4}{3} x_1y_1 - \frac{2}{3} x_1y_2
  - \frac{2}{3} x_2y_1 + \frac{4}{3} x_2y_2,
$$

where
$\mathbf{x} = \begin{bmatrix} x_1\\ x_2\\ \end{bmatrix}$,
$\mathbf{y} = \begin{bmatrix} y_1\\ y_2\\ \end{bmatrix}$,
$\in \mathbb{R}^2$,

$$
\Sigma^{-1} = \begin{bmatrix}
    \langle \mathbf{e}_1, \mathbf{e}_1 \rangle &
      \langle \mathbf{e}_1, \mathbf{e}_2 \rangle \\
    \langle \mathbf{e}_2, \mathbf{e}_1 \rangle &
      \langle \mathbf{e}_2, \mathbf{e}_2 \rangle \\
  \end{bmatrix} =
  \begin{bmatrix}
     \tfrac{4}{3} & -\tfrac{2}{3} \\
    -\tfrac{2}{3} &  \tfrac{4}{3} \\
  \end{bmatrix}
$$

### 2022-05-13 notebooks/Search_Citation_Gecko.Rmd
Initial commit. Includes first version of the article draft, along with several notebooks and ancillary output documents.

---
title: "Search with CitationGecko"
output: html_notebook
bibliography: ../www/Multidimensional-parameters-MCLM.bib
csl: ../www/apa-old-doi-prefix.csl
---

```{r setup}
library(tidyverse)
library(bib2df)
library(writexl)
library(glue)
```

# Procedure

## Initial search

On 15 july, 2021, I use the following papers as seeds in [CitationGecko](https://citationgecko.azurewebsites.net/):

::: {#refs}
:::

<br>

Then select:

-   Papers citing seed papers

-   Filter: Year after 1984

Then click on "Export" (in the network view) and save the results as `output/citation_gecko/GeckoRecommendations.bib`[^search_citation_gecko-1].

[^search_citation_gecko-1]: A bug in `bib2pdf()` prevents from saving files in the usual resource folder `www`. (It parses the regexp `www.` as a remote URL, which is matched by `www/`; it should parse `www\\.` to interpret a literal dot.)

```{r entries-raw}
entries_raw <- bib2df("../output/citation_gecko/GeckoRecommendations.bib")

n_raw <- entries_raw %>% count() %>% pull()
```

There are `r n_raw` entries in this search.

## Manual processing

I import the bibliography to Zotero into a new collection "GeckoRecommendations". After initial curtaion (eliminate duplicates, search for available full text, and rename PDFs according to metadata), the following manual filtering is made:

-   Entries with `null` title

-   Entries with "book review" in the title

-   Entries with titles "Commentaries", "References" (content checked, non-relevant)

I export this collection as `output/citation_gecko/GeckoRecommendations_filtered.bib`.

```{r entries-filtered}
entries_filtered <-
  bib2df("../output/citation_gecko/GeckoRecommendations_filtered.bib")

n_filtered <- entries_filtered %>% count() %>% pull()
```

This filtering leaves `r n_filtered` entries.

## Exploration of the metadata

I perform an assisted exploration to find some patterns in the publications.

### Publications by journal

```{r journal-count}
entries_filtered %>% count(JOURNAL) %>% arrange(desc(n))
```

```{r}
entries_filtered %>% filter(is.na(JOURNAL))
```

One entry without journal name; corresponds to a book, the bibliographic entry is from 1985 (too old for my purposes), and barely addresses the topic of multidimensionality.

```{r}
entries_filtered <- entries_filtered %>% filter(!is.na(JOURNAL))
```

I explore several journals, that seem to be non-central to the topic, but essentially most of them seem to cite Reckase or Zhang for some relevant reason. It is difficult to filter out journals by name or keywords, as most articles seem to deal with multidimensional IRT in one way or another, even in journals that seem completely unrelated. Some of them are even theoretical reviews of multidimensional IRT and not just applications.

### Publications by title and keywords

We filter the publications that contain the keyword "multidimensional" in any field:

```{r}
entries_multidimensional <- entries_filtered %>%
  rowwise() %>%
  filter(
    c_across(where(is.character)) %>%
      tolower() %>%
      str_detect("multidimensional") %>%
      any()
  ) %>%
  ungroup()

n_entries_multidimensional <- entries_multidimensional %>% count() %>% pull()
```

This leaves `r n_entries_multidimensional` entries, which are suitable for manual classification.

## Manual classification

### Preparation

I export the selected entries to an Excel file, first reordering the fields, flattening the `AUTHOR` field (otherwise it is not written properly), and ordering by year to make the search easier.

```{r}
entries_multidimensional <- entries_multidimensional %>%
  select(AUTHOR, YEAR, TITLE, JOURNAL, DOI, everything()) %>%
  rowwise() %>%
  mutate(across(AUTHOR, glue_collapse, sep = "; ")) %>%
  ungroup() %>%
  arrange(YEAR)

entries_multidimensional %>%
  write_xlsx("../output/citation_gecko/Recommendations_wb_bak.xlsx")
```

Then I manually copy this file to `../output/citation_gecko/Recommendations_wb.xlsx`. I do this as a safeguard, in case I run again this whole notebook no information is lost in the working file if it is overwritten.

### Classification

I create a new collumn in the Excel file "CLASSIFICATION", in column "AI", where I fill in the results of the manual classification. Then, to decide whether a paper is relevant, I search for citations of @reckase_difficulty_1985, @reckase_discriminating_1991, @ackerman_creating_1994, and @zhang_conditional_1999. If the paper mentions the computation of the multidimensional difficulty (MDIFF, MID, or MDL) or discrimination (MDISC) parameter (I search for those terms and also "difficulty" and "discrimination"), then I check whether they assume an orthogonal latent space (by searching for "distribution", "orthogonal", and "correlation", if necessary).

The possible resulting categories are:

-   **Full text unavailable**

-   **Orthogonality assumed**: Assumes orthogonality among latent dimensions, and computes the MDIFF and MDISC accordingly, OR does not mention the latent space structure at all (and therefore it just assumes orthogonality, as in @reckase_difficulty_1985 and @reckase_discriminating_1991).

-   **Not relevant**: There is no mention to the multidimensional parameters or their computations, OR they are only discussed theoretically. Multidimensional traits may be simulated, dealt with, or theoretically discussed, but not in relationship with the multidimensional parameters.

-   **Oblique dimensions disregarding non-orthogonality**: Computes the MDIFF and/or MDISC parameters applying Reckase's [-@reckase_difficulty_1985] and Reckase and McKinely's [-@reckase_discriminating_1991] formulae, disregarding the fact that these only apply to orthogonal latent spaces.

### Title formatting

While doing the manual classification, I find out that at least one entry (Wang, Chang, & Boughton, 2010) has been filtered out due to the title using the term "Multi-Dimensional" (with a dash) instead of "Multidimensional". Therefore, I change the matching regular expression to "`multi-?dimensional"`, which matches both terms. However, the BibTeX entries are "[case protected](https://tex.stackexchange.com/a/10775/171606)" with curly braces, which may prevent entries like this one from being detected ("{Multi}-{Dimensional}" instead of "{Multi-Dimensional}". In order to avoid this, the curly braces are removed from the titles of the bibliographical entries first.

```{r}
entries_no_curly_braces <- entries_filtered %>%
  mutate(TITLE = TITLE %>% str_remove_all("[\\{\\}]"))

entries_multidimensional <- entries_no_curly_braces %>%
  rowwise() %>%
  filter(
    c_across(where(is.character)) %>%
      tolower() %>%
      str_detect("multi-?dimensional") %>%
      any()
  ) %>%
  ungroup() %>%
  select(AUTHOR, YEAR, TITLE, JOURNAL, DOI, everything()) %>%
  rowwise() %>%
  mutate(across(AUTHOR, glue_collapse, sep = "; ")) %>%
  ungroup() %>%
  arrange(YEAR)

n_entries_multidimensional_hyphenated <- entries_multidimensional %>%
  count() %>%
  pull()

entries_multidimensional %>%
  write_xlsx("../output/citation_gecko/Recommendations_wb_bak.xlsx")
```

The updated list has `r n_entries_multidimensional_hyphenated` (apparently, only that one entry had been missed for this reason).

### Missing entries with the acronym MIRT

While doing the manual search, I notice a few papers having the acronym "MIRT" in the title, that have not been considered in the results. As I assume they also refer to "Multidimensional Item Response Theory", and therefore should be included in the results, I re-run the filtering including these results.

```{r}
entries_with_mirt <- entries_no_curly_braces %>%
  rowwise() %>%
  filter(
    c_across(where(is.character)) %>%
      tolower() %>%
      str_detect("(multi-?dimensional|mirt)") %>%
      any()
  ) %>%
  ungroup() %>%
  select(AUTHOR, YEAR, TITLE, JOURNAL, DOI, everything()) %>%
  rowwise() %>%
  mutate(across(AUTHOR, glue_collapse, sep = "; ")) %>%
  ungroup() %>%
  arrange(YEAR)

n_entries_multidimensional_mirt <- entries_with_mirt %>%
  count() %>%
  pull()

entries_with_mirt %>%
  write_xlsx("../output/citation_gecko/Recommendations_wb_bak.xlsx")
```

The number of entries is now `r n_entries_multidimensional_mirt`. The new ones included are:

```{r}
entries_with_mirt %>% anti_join(entries_multidimensional)
```

# Known issues

There is at least one known paper that has not been included and is known to be relevant:

Fragoso, T. M., & Cúri, M. (2013). Improving psychometric assessment of the Beck Depression Inventory using Multidimensional Item Response Theory. *Biometrical Journal*, *55*(4), 527--540. <https://doi.org/10.1002/bimj.201200197>

This reference has been found because it is cited by Silva (2019), and classifies as "Oblique dimensions disregarding non-orthogonality". The problem is that it does not cite directly any of the seed papers considered. Instead, it cites Reckase (1997; A Linear Logistic Multidimensional Model for Dichotomous Item Response Data).

The problem with CitationGecko is that it does not allow to construct the whole citation network, so only direct references of the seed papers can be found.

### 2022-05-13 output/Derivations_by_Reckase/Derivation_according_to_Reckase.Rmd
Initial commit. Includes first version of the article draft, along with several notebooks and ancillary output documents.

---
title: "Multidimensional difficulty and discrimination parameters according to Reckase & McKinley"
author: "Daniel Morillo"
date: "`r Sys.Date()`"
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
    mapstyles:
      Normal: First Paragraph
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
bibliography: ../../www/Multidimensional-parameters-MCLM.bib
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)

ROOT_DIR <- "../.."
DOC_DIR <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      =   300,
  dev.args = list(type = "cairo-png"),
  fig.pos = "!H"
)

# Constants:
GRAPH_FONT   <- "serif"
LINE_WIDTH   <- .1
VECTOR_WIDTH <- 1.5
ANGLE_WIDTH  <- .5
POINT_CHAR   <- 19L

# Output configuration options:
options(digits = 4)
# theme_set(
#   theme_classic(
#     base_size      = 12,
#     base_family    = GRAPH_FONT,
#     base_line_size = LINE_WIDTH
#   ) %+replace%
#   theme(
#     axis.title.x = element_text(hjust = 1),
#     axis.title.y = element_text(vjust = 1, angle = 0)
#   )
# )
```

```{r includes, cache=FALSE}
source("R/Formulas.R")
```

# Introduction

We are going to use the method proposed by @reckase_difficulty_1985
in order to determine the difficulty of a multidimensional item
in a non-orthogonal latent space.
Then, we will determine its discrimination,
following @reckase_discriminating_1991.
In order to better understand these,
we will first review the original derivations in the orthogonal space.

# Original parameter derivations

## Multidimensional difficulty in @reckase_difficulty_1985

Equation 7 in @reckase_difficulty_1985 applies the constraint
$\sum_{k=1}^n\cos^2 \alpha_{jk} = 1$,
which implies orthogonal axes.
This constraint is necessary because the $n$ cosines are not independent;
rather, they are (non-linearly) dependent by the constraint in their Equation 7.

For a certain dimension $k$ ($k \in [1, n-1]$),
the derivative of each addend in their Equation 6 is 0
(because it is independent of $\cos \alpha_{ki}$),
except for the derivatives of the $k-\mathrm{th}$ term,
which is $\tfrac{\alpha_{ki}}{4}$,
and the $n-\mathrm{th}$ term, given by

\begin{equation}
  \begin{split}
    \frac{\delta}{\delta \cos \alpha_{jk}}
      \frac{a_{in}}{4}
      \sqrt{ 1 - \sum_{k=1}^{n-1} \cos^2 \alpha_{jk} } &=
        \frac{a_{in}}{4}
          \frac{1}{2 \sqrt{ 1 - \sum_{k=1}^{n-1} \cos^2 \alpha_{jk} } }
          \left(-2 \cos \alpha_{jk} \right) \\
      &= - \frac{a_{in}}{4}
        \frac{
            \cos \alpha_{jk}
          } {
            \sqrt{ 1 - \sum_{k=1}^{n-1} \cos^2 \alpha_{jk} }
          } \\
      &= - \frac{a_{in}}{4} \frac{\cos \alpha_{jk}}{\cos \alpha_{jn}}.
  \end{split}
\end{equation}

Therefore, the derivative of their Equation 6 is given by

\begin{equation}
  \frac{\delta}{\delta \cos \alpha_{jk}}
    \frac{1}{4}
    \sum_{k=1}^{n} a_{ik} \cos \alpha_{jk} =
  \frac{1}{4}
    \left( a_{ik} - a_{in} \frac{\cos \alpha_{jk}}{\cos \alpha_{jn}} \right).
\end{equation}

The term $\tfrac{1}{4}$ can be dropped when equating to 0,
thus resulting in the $n-1$ expressions in Equation 8 in
@reckase_difficulty_1985.
By reformulating each of these equations we can show that
$\cos \alpha_{ik}$ is proportional to $a_{ik}$ for all $k$:

\begin{equation}
  a_{ik} - a_{in} \frac{\cos \alpha_{jk}}{\cos \alpha_{jn}} = 0 \Leftrightarrow
    \frac{a_{ik}}{\cos \alpha_{jk}} = \frac{a_{in}}{\cos \alpha_{jn}} = C.
\end{equation}

Thus, the solution implies computing the proportionality constant $C$.

The system is complemented by Equation 7,
giving $n$ equations with the $n$ cosines as unknowns.
Operating on the previous equation,

\begin{equation}
  \begin{split}
    \cos \alpha_{jk} = \frac{a_{ik}}{C} &\Leftrightarrow
      \cos^2 \alpha_{jk} = \frac{a_{ik}^2}{C^2} \\
      &\Leftrightarrow \sum_{k=1}^{n} \cos^2 \alpha_{jk} =
        \sum_{k=1}^{n} \frac{a_{ik}^2}{C^2} \\
      &\Leftrightarrow \frac{1}{C^2} \sum_{k=1}^{n} a_{ik}^2 = 1 \\
      &\Leftrightarrow C^2 = \sum_{k=1}^{n} a_{ik}^2 \\
      &\Leftrightarrow C = \sqrt{ \sum_{k=1}^{n} a_{ik}^2 },
  \end{split}
\end{equation}

results in the value of $C$ which, when solving for $j = i$,
gives Equation 9 in @reckase_difficulty_1985.

The signed distance to the point of steepest slope, $D_i$,
is obtained by substituting Equation 9 for $\cos \alpha_{ik}$ in Equation 3.
The exponent of the logit function results in
(the summation index is changed from $k$ to $l$ in the denominator
to avoid collision):

\begin{equation}
  \begin{split}
    \sum_{k=1}^{n} a_{ik} `r THETA_MODULE` \cos \alpha_{jk} + d_i &=
        \sum_{k=1}^{n} a_{ik}
        `r THETA_MODULE`
        \frac{a_{ik}}{\sqrt{ \sum_{l=1}^{n} a_{il}^2} } + d_i \\
      &= \frac{ \sum_{k=1}^{n} a_{ik}^2 }{ \sqrt{ \sum_{l=1}^{n} a_{il}^2 } }
        `r THETA_MODULE`  + d_i \\
      &= \sqrt{
          \frac{ \left( \sum_{k=1}^{n} a_{ik}^2 \right)^2 }
               { \sum_{l=1}^{n} a_{il}^2 }
        }
        `r THETA_MODULE` +
        d_i \\
      &= \sqrt{ \sum_{k=1}^{n} a_{ik}^2 } `r THETA_MODULE`  + d_i.
  \end{split}
\end{equation}

Solving for $P_{ij} = .5$ implies that the exponent equals 0.
Thus solving for $D_i = `r THETA_MODULE`$,

\begin{equation}
  \begin{split}
    \sqrt{ \sum_{k=1}^{n} a_{ik}^2 } D_i + d_i = 0 &\Leftrightarrow
      \sqrt{ \sum_{k=1}^{n} a_{ik}^2 } D_i = -d_i \\
    &\Leftrightarrow D_i = - \frac{d_i}{ \sqrt{ \sum_{k=1}^{n} a_{ik}^2 } },
  \end{split}
\end{equation}

which gives Equation 10.

## Multidimensional discrimination

The multidimensional discrimination parameter ($MDISC$)
is obtained straightforwardly from the previous equations and
Equations 6 through 9 in @reckase_discriminating_1991.

# Parameters in the non-orthogonal space

## Model definition

The model is defined by Reckase's [-@reckase_difficulty_1985] Equation 1.
However, in the present case, $`r THETA_J`$ is multivariate normal:

\begin{equation}
  `r THETA_J` \sim \mathcal{N}(\mathbf{\upmu}, `r SIGMA`),
\end{equation}

where $\mathbf{\upmu}$ is an $n$-element vector of means and
$`r SIGMA` = `r S_MATRIX` `r R_MATRIX` `r S_MATRIX`'$
an $n$-dimensional positive definite covariance matrix,
with $`r R_MATRIX`$ the corresponding correlation matrix
(also positive definite) and $`r S_MATRIX`$ a scaling diagonal matrix
with $s_{ii} = \sigma_{ii}$
(i.e., the matrix of standard deviations).

When $`r THETA_J`$ is standardized, $`r S_MATRIX` = `r ID_MATRIX`$ and
$`r SIGMA` = `r R_MATRIX`$.
When $`r R_MATRIX` = `r ID_MATRIX`$, the elements of $`r THETA_J`$
are independent and $`r SIGMA` = `r S_MATRIX` `r S_MATRIX`'$,
which implies that the basis of the latent space $`r THETA`$ is orthogonal.
When both conditions occur simultaneously, $`r SIGMA` = `r ID_MATRIX`$,
and the basis of the latent space $`r THETA`$ is canonical and therefore
orthonormal.

One may assume that the latter must occur for the conditions in
@reckase_difficulty_1985 to apply.
However, Reckase's derivations apply to an orthogonal latent space,
which actually happens in the second condition.
That is, the only requirement is that the elements of $`r THETA_J`$ are
independent, being irrelevant whether they are standardized or not.
Therefore, our generalization applies to the non-orthogonal case.

## Model recasting into polar coordinates {#polar-coords}

Following the method outlined by @reckase_difficulty_1985,
the model is first transformed into polar coordinates,
as in Reckase's [-@reckase_difficulty_1985] Equation 3.
However, to apply this transformation one must consider that,
in a non-orthogonal space,
$`r THETA_J` = `r R_COS_ALPHA_THETA`$
instead of $`r THETA_MODULE` `r COS_ALPHA_J_VECTOR`$, as in Rekcase's
[-@reckase_difficulty_1985] Equation 3.

**Proof 1**.[^proof-needed]
First, let $`r THETA_ORTH` = `r B_INVERSE` `r THETA`$
be an orthogonal latent space with $`r B_MATRIX`$ a rotation matrix such that
$`r B_MATRIX` `r B_MATRIX`' = `r R_MATRIX`$.
The norm $`r THETA_J_ORTH_NORM` \equiv `r THETA_MODULE`$
of a vector $`r THETA_J_ORTH`$ in $`r THETA_ORTH`$
is defined by

\begin{equation}
  `r THETA_MODULE`^2 = `r THETA_J_ORTH`' `r THETA_J_ORTH` =
    `r THETA_J`' `r B_INVERSE`' `r B_INVERSE` `r THETA_J` =
    `r THETA_J`' (`r B_MATRIX` `r B_MATRIX`')^{-1} `r THETA_J` =
    `r THETA_J`' `r R_INVERSE` `r THETA_J`.                 (\#eq:inner-product)
\end{equation}

Therefore, we consistently define
$\langle \mathbf{x}, \mathbf{y} \rangle = \mathbf{x}' `r R_INVERSE` \mathbf{y}$ as an inner product in $`r THETA`$ with norm
$`r THETA_MODULE` = \sqrt{`r THETA_J`' `r R_INVERSE` `r THETA_J`}$.
By definition of the inner product, the cosine between
$`r THETA_J`$ and each vector $`r BASIS_VECTOR`$ of the basis of $`r THETA`$
is given by their inner product divided by the norms of the two vectors:

\begin{equation}
  `r COS_ALPHA_JK` = \frac{
      `r BASIS_VECTOR`' `r R_INVERSE` `r THETA_J`
    } {
      `r BASIS_VECTOR_NORM` `r THETA_J_NORM`
    } =
    \frac{
      `r BASIS_VECTOR`' `r R_INVERSE` `r THETA_J`
    } {
      1 `r THETA_J_NORM`
    } =
    \frac{`r BASIS_VECTOR`' `r R_INVERSE` `r THETA_J`}{ `r THETA_MODULE`}.
\end{equation}

Therefore, the vector of cosines is given by

\begin{equation}
  `r COS_ALPHA_J_VECTOR` = \frac{
      \begin{pmatrix}
        `r BASIS_VECTOR_1`' \\
        \vdots              \\
        `r BASIS_VECTOR_N`'
      \end{pmatrix}
      `r R_INVERSE` `r THETA_J`
    } {
       `r THETA_MODULE`
    } =
    \frac{`r ID_MATRIX` `r R_INVERSE` `r THETA_J`}{`r THETA_MODULE`}
\end{equation}

\begin{equation}
    `r COS_ALPHA_J_VECTOR` = \frac{`r R_INVERSE` `r THETA_J`}{`r THETA_MODULE`}.
                                                         (\#eq:director-cosines)
\end{equation}

Solving for $`r THETA_J`$,

\begin{equation}
  \begin{split}
    \frac{`r R_MATRIX` `r R_INVERSE` `r THETA_J`}{ `r THETA_MODULE`} &=
      `r R_COS_ALPHA` \\
    `r THETA_J` &= `r R_COS_ALPHA_THETA`.\blacksquare
  \end{split}
\end{equation}

[^proof-needed]: It took me quite an effort to realise that Equation 3 in
@reckase_difficulty_1985 did not apply straightforwardly,
and derivate the following proof.
However, I'm not sure if it's really necessary to proof this,
as I think it must be an already-known result of the inner product
[see e.g., @harman_modern_1970,
though probably not the most appropriate source].

Thus, the model expression in polar coordinates is given by

\begin{equation}
  `r P_MCLM_OBL` = `r LOGISTIC_OBL_POLAR`.                     (\#eq:MCLM-polar)
\end{equation}

## Multidimensional difficulty {#md-diff}

The equivalent to Reckase's [-@reckase_difficulty_1985]
Equations 4 through 6 are:

\begin{equation}
  `r IRF_DIFF2_THETA` = `r IRF_DIFF2_THETA_DEF`,
\end{equation}

\begin{equation}
  `r IRF_DIFF_THETA` = `r IRF_DIFF_THETA_DEF`,
\end{equation}

and

\begin{equation}
  \tfrac{1}{4} `r A_R_COS_ALPHA`,                               (\#eq:max-slope)
\end{equation}

To derive the direction of steepest slope we must take into account the
dependence among the components of $`r COS_ALPHA_J_VECTOR`$,
as in Reckase's [-@reckase_difficulty_1985] Equation 7.
However, this dependence is expressed here as

\begin{equation}
  `r COS_ALPHA_J_VECTOR` ' `r R_MATRIX` `r COS_ALPHA_J_VECTOR` = 1
                                                        (\#eq:cosine-constraint)
\end{equation}

**Proof 2**.[^proof-needed-2]
From Equation \@ref(eq:director-cosines), we have that

\begin{equation}
  \begin{split}
    `r R_MATRIX` `r COS_ALPHA_J_VECTOR` &=
      \frac{`r THETA_J`}{`r THETA_MODULE`},\\
    `r B_MATRIX` `r B_MATRIX`' `r COS_ALPHA_J_VECTOR` &=
      \frac{`r THETA_J`}{`r THETA_MODULE`}.
  \end{split}
\end{equation}

Also, in the orthogonal space,

\begin{equation}
  \begin{split}
    `r COS_ALPHA_J_ORTH_VECTOR` &= \frac{`r THETA_J_ORTH`}{`r THETA_MODULE`}, \\
    `r COS_ALPHA_J_ORTH_VECTOR` &=
      \frac{`r B_INVERSE` `r THETA_J`}{`r THETA_MODULE`}, \\
    `r B_MATRIX` `r COS_ALPHA_J_ORTH_VECTOR` &=
      \frac{`r THETA_J`}{`r THETA_MODULE`}.
  \end{split}
\end{equation}

Equating the left terms in both equations,

\begin{equation}
  `r B_MATRIX` `r B_MATRIX`' `r COS_ALPHA_J_VECTOR` =
    `r B_MATRIX` `r COS_ALPHA_J_ORTH_VECTOR`,
\end{equation}

\begin{equation}
  `r B_MATRIX`' `r COS_ALPHA_J_VECTOR` = `r COS_ALPHA_J_ORTH_VECTOR`
                                                      (\#eq:cosines-equivalence)
\end{equation}

and squaring,

\begin{equation}
  \begin{split}
    `r COS_ALPHA_J_VECTOR`' `r B_MATRIX` `r B_MATRIX`' `r COS_ALPHA_J_VECTOR` &=
      `r COS_ALPHA_J_ORTH_VECTOR`' `r COS_ALPHA_J_ORTH_VECTOR`; \\
    `r COS_ALPHA_J_VECTOR`' `r R_MATRIX` `r COS_ALPHA_J_VECTOR` &= 1.
      \blacksquare
  \end{split}
\end{equation}

[^proof-needed-2]: As before, I doubt this proof is strictly necessary.

(Now, I've come to a dead end here, as I don't know how to solve
\@ref(eq:cosine-constraint) for $`r COS_2_ALPHA_JN`$.
I arrive to the covariance matrix of $`r COS_ALPHA_J_VECTOR`$,
$\mathrm{cov}(`r COS_ALPHA_J_VECTOR`) = `r R_INVERSE` `r S_MATRIX` `r R_MATRIX` `r S_MATRIX`' `r R_INVERSE`'$,
but I don't see how this can be solved for a single component of
$`r COS_ALPHA_J_VECTOR`$.
A possible solution would be to define
$`r THETA_ORTH` = `r T_INVERSE` `r THETA`$ as an orthonormal latent space
with $`r T_MATRIX` = `r S_MATRIX` `r B_MATRIX`$,
such that $`r T_MATRIX` `r T_MATRIX`' = `r SIGMA`$.
This would lead to the substitution of $`r R_MATRIX`$ by $`r SIGMA`$ in all
the equations in sections \@ref(polar-coords) and \@ref(md-diff),
and therefore $\mathrm{cov}(`r COS_ALPHA_J_VECTOR`) = `r SIGMA`^{-1}$.
However, I don't know how to continue from here either.)

### Derivating the multidimensional difficulty by orthogonalization

The most straightforward way of solving it seems to be by orthogonalizing
the latent space, computing the solution in this orthogonalized space,
and then transforming the space back into oblique coordinates
(as I do in my dissertation).
Using this approach, we may find the solution in the orthogonal space
applying the method in @reckase_difficulty_1985.

First, we define $`r A_VECTOR_ORTH` = `r B_MATRIX`' `r A_VECTOR`$ such that
$`r A_VECTOR_ORTH`' `r THETA_J_ORTH` = `r A_VECTOR`' `r B_MATRIX` `r B_INVERSE` `r THETA_J` = `r A_VECTOR`' `r THETA_J`$.
Then, we can compute $`r COS_ALPHA_J_ORTH_VECTOR`$ and $`r MDIFF`$ applyting
Reckase's [-@reckase_difficulty_1985] equations 3 through 10.
This results in

\begin{equation}
  `r COS_ALPHA_IK_ORTH` =
    \frac{`r A_COMPONENT_ORTH`}{\sqrt{`r A_VECTOR_ORTH`' `r A_VECTOR_ORTH`}}
\end{equation}

and

\begin{equation}
  `r MDIFF` =
    \frac{- `r D_PARAM`}{\sqrt{`r A_VECTOR_ORTH`' `r A_VECTOR_ORTH`}}.
\end{equation}

The first one can be expressed in vector terms:

\begin{equation}
  `r COS_ALPHA_I_ORTH_VECTOR` =
    \frac{`r A_VECTOR_ORTH`}{\sqrt{`r A_VECTOR_ORTH`' `r A_VECTOR_ORTH`}},
\end{equation}

and thus both can be expressed in terms of $`r A_VECTOR`$ instead of
$`r A_VECTOR_ORTH`$, with
$`r A_VECTOR_ORTH`' `r A_VECTOR_ORTH` = `r A_VECTOR`' `r B_MATRIX` `r B_MATRIX`' `r A_VECTOR` = `r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`$:

\begin{equation}
  `r COS_ALPHA_I_ORTH_VECTOR` = \frac{
      `r B_MATRIX`' `r A_VECTOR`
    } {
      \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}
    },
\end{equation}

and

\begin{equation}
  `r MDIFF` =
    \frac{- `r D_PARAM`}{\sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}}.
\end{equation}

From Equation \@ref(eq:cosines-equivalence) we have that 
$`r COS_ALPHA_I_ORTH_VECTOR` = `r B_MATRIX`' `r COS_ALPHA_I_VECTOR`$, so
[^warning-direction]

\begin{equation}
  `r B_MATRIX`' `r COS_ALPHA_I_VECTOR` = \frac{
      `r B_MATRIX`' `r A_VECTOR`
    } {
      \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}
    },
\end{equation}

\begin{equation}
  `r COS_ALPHA_I_VECTOR` =
    \frac{`r A_VECTOR`}{\sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}}.
                                                         (\#eq:director-cosines)
\end{equation}

[^warning-direction]: Warning!
This result is not the same as what I found in the dissertation (Equation 3.7).
This one seems to be the correct one though.
I think the explanation is that, in my dissertation,
the projections should be on the orthogonal axes instead, that is,
$`r COS_ALPHA_I_ORTH_VECTOR`'$ in the left side.

### Derivating the multidimensional difficulty form the equation in polar coordinates

Using the method outlined above, one can derivate the director cosines which
describe the measurement direction.
However, the absolute value of the multidimensional difficulty can be derived
by applying this solution to the model expression in polar coordinates,
i.e. Equation \@ref(eq:MCLM-polar).

Substituting Equation \@ref(eq:director-cosines) for $`r COS_ALPHA_I_VECTOR`$ in
Equation \@ref(eq:MCLM-polar) we may find the maximum slope, solving for
$`r MDIFF` = `r THETA_MODULE`$ when $P_ij = .5$,
that is, $`r P_MCLM_OBL_MDIFF` = .5$:

\begin{equation}
  \begin{split}
    `r LOGISTIC_OBL_POLAR_MDIFF` = .5 &\Leftrightarrow
      `r EXPONENT_OBL_POLAR_MDIFF` = 0 \\
      &\Leftrightarrow \frac{
        `r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`
      } {
        \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}
      } `r MDIFF` = - `r D_PARAM`
  \end{split}
\end{equation}

\begin{equation}
  \begin{split}
      `r MDIFF` = - \frac{
        `r D_PARAM`
      } {
        \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}
      }
  \end{split}                                              (\#eq:MDIFF-solution)
\end{equation}

## Multidimensional discrimination

Analogously, we apply Equations 4 through 9 in @reckase_discriminating_1991,
arriving at the equivalent of their Equation 9 in the orthogonal space:

\begin{equation}
  `r MDISC` = \sqrt{`r A_VECTOR_ORTH`' `r A_VECTOR_ORTH`} =
    \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}.
\end{equation}

## Properties of the multidimensional parameters

### Multidimensional discrimination

Substituting $`r COS_ALPHA_J_VECTOR`$ in expression \@ref(eq:max-slope) by
$`r COS_ALPHA_I_VECTOR`$,
and the substituting Equation \@ref(eq:director-cosines) into
\@ref(eq:max-slope) results in

\begin{equation}
  \tfrac{1}{4} \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`} =
    \tfrac{1}{4} `r MDISC`.
\end{equation}

Therefore, $`r MDISC`$ is analogous
to the $a_i$ parameter in unidimensional IRT:
Its scale amounts to four times the slope and can be interpreted accordingly,
as in the orthogonal case.

When $a_{il} \neq 0$ and $a_{ik} = 0, \forall k \neq l$,
then $`r MDISC` = a_{il}$.
That is, $`r MDISC`$ reduces to the unidimensional discrimination parameter
as expected, the same way it does in the orthogonal case.

Finally, @reckase_difficulty_1985 [p. 404] states that "when $`r ALPHA_JK` = 0$
and $\alpha_{jl} = 90$ for $l \neq k$,
indicating that the slope of the surface at $P_{ij} = .5$
is determined parallel to axis $k$, $`r COS_ALPHA_JK` = 1$ and
$\cos \alpha_{jl} = 0$.
The slope parallel to axis $k$ is then given by $\tfrac{1}{4} a_{ik}$."

In the non-orthogonal space, the equality $\alpha_{jl} = 90$
does not necessarily hold,
and thus it is not generally true that $\cos \alpha_{jl} = 0$.
In fact, $`r COS_ALPHA_JK`$ equals the inner product of $`r A_VECTOR`$
with the basis vector
$`r BASIS_VECTOR`' = (e_1, \ldots, e_k, \ldots, e_n), e_k = 1, e_l = 0 \, \forall l \neq k$:

\begin{equation}
  `r COS_ALPHA_JK` = \frac{
      `r BASIS_VECTOR`' `r R_MATRIX` `r A_VECTOR`
    } {
      \left\| e \right\| \left\| `r A_VECTOR` \right\|
    } =
    \frac{
      `r BASIS_VECTOR`' `r R_MATRIX` `r A_VECTOR`
    } {
      \left\| e \right\| \left\| `r A_VECTOR` \right\|
    } =
    \frac{
      (\rho_{k1}, \ldots, 1, \ldots, \rho_{kn}) `r A_VECTOR`
    } {
      \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}
    } =
    \frac{
      a_{ik} + \sum_{l = 1, l \neq k}^n \rho_{kl} a_{il}
    } {
      \sqrt{`r A_VECTOR`' `r R_MATRIX` `r A_VECTOR`}
    }.
\end{equation}

Therefore, this condition does not generally hold in the non-orthogonal space.

### Multidimensional difficulty

$`r MDIFF`$ has the same relationship with $`r D_PARAM`$
as $b_i$ for the 2PL model, i.e.

\begin{equation}
  `r MDIFF` = - \frac{`r D_PARAM`}{`r MDISC`}.       (\#eq:relationship-MDIFF-d)
\end{equation}

# Alternative definition of the inner product

## Orthogonal interpretation of the coordinates for correlated traits

The definition of the inner product in $`r THETA`$ as
$\langle \mathbf{x}, \mathbf{y} \rangle = \mathbf{x}' `r R_INVERSE` \mathbf{y}$
is arbitrary, and it has been done in order to make it consistent with the
inner product
$\langle \mathbf{x}, \mathbf{y} \rangle = \mathbf{x}' \mathbf{y}$
in the orthogonal space $`r THETA_ORTH`$.
This is convenient in order to transform between the oblique and orthogonal
space back and forth,
and thus leverage on Reckase's [-@reckase_difficulty_1985] derivations to
solve the non-orthogonal case.
However, we could want to define an inner product in $`r THETA`$ such that
the inner product in the test space
$\langle \mathbf{x}, \mathbf{y} \rangle = \mathbf{x}' \mathbf{y}$.
This would imply an inner product
$\langle \mathbf{x}, \mathbf{y} \rangle = \mathbf{x}' \mathbf{y}$
in the latent trait space, and all the formulae in @reckase_difficulty_1985
and @reckase_discriminating_1991 would apply.
However, in applications where one would need to compute the equivalent
parameters in an orthogonal space (i.e. equating, linking, or multigroup IRT),
additional complications would arise,
as the inner product in the orthogonal space would be defined by
the inner product
$\langle \mathbf{x}, \mathbf{y} \rangle = \mathbf{x}' (`r B_MATRIX`' `r B_MATRIX`)^{-1} \mathbf{y}$.

## Orthonormalization

The orthogonal latent space $`r THETA_ORTH`$ could instead be defined as
$`r THETA_ORTH` = `r T_INVERSE` `r THETA`$,
with $`r T_MATRIX` = `r S_MATRIX` `r B_MATRIX`$,
such that $\mathrm{cov}(`r THETA_ORTH`) = `r ID_MATRIX`$.

**Proof 3.** The covariance of $`r THETA_ORTH`$ is given by
$\mathrm{cov}(`r THETA_ORTH`) = `r T_INVERSE` `r SIGMA` `r T_INVERSE`'$, and
$`r T_MATRIX` `r T_MATRIX`' = `r S_MATRIX` `r B_MATRIX` `r B_MATRIX`' `r S_MATRIX`' = `r SIGMA`$.
Thus substituting:

\begin{equation}
  \mathrm{cov}(`r THETA_ORTH`) =
    `r T_INVERSE` `r T_MATRIX` `r T_MATRIX`' `r T_INVERSE`' =
    `r ID_MATRIX`.\blacksquare
\end{equation}

Using this definition results in an inner product as defined in
@zhang_theoretical_1999; all the derivations from \@ref(eq:inner-product) to
\@ref(eq:relationship-MDIFF-d) would be equivalent,
substituting $`r B_MATRIX`$ by $`r T_MATRIX`$ and $`r R_MATRIX`$ by $`r SIGMA`$.
This has the advantage that different latent spaces can be referred to a common
metric.
However, it also has the disadvantage of changing the metric (i.e. covariances)
of the latent space, resulting in a change in variances.
This is detrimental for graphical representation purposes, on one side,
and for the property of analogy with unidimensional IRT on the other,
as $`r MDISC`$ is no longer analogous to its unidimensional counterpart $a_i$.

The original derivation, on the other side,
has a consistent metric with the original, oblique latent space
(but $`r THETA_ORTH`$ apparently does not satisfy
$\mathrm{cov}(`r THETA_ORTH`) = `r S_MATRIX` `r S_MATRIX`'$, in general).

# References

### 2022-05-13 output/Transcription_Mario_Luzardo/Transcription_of_Marios_derivations.Rmd
Initial commit. Includes first version of the article draft, along with several notebooks and ancillary output documents.

---
title: "Generalized multidimensional parameters of the MCLM"
author: "Daniel Morillo, Mario Luzardo"
date: "`r Sys.Date()`"
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
    mapstyles:
      Normal: First Paragraph
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
bibliography: ../../www/Multidimensional-parameters-MCLM.bib
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(knitr)
library(scales)
library(tidyverse)
library(latex2exp)
library(magrittr)
library(ggforce)
library(pBrackets)
library(glue)
library(flextable)
library(ftExtra)

ROOT_DIR <- "../.."
DOC_DIR <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      =   300,
  dev.args = list(type = "cairo-png"),
  fig.pos = "!H"
)

# Constants:
GRAPH_FONT   <- "serif"
LINE_WIDTH   <- .1
VECTOR_WIDTH <- 1.5
ANGLE_WIDTH  <- .5
POINT_CHAR   <- 19L

# Output configuration options:
options(digits = 4)
theme_set(
  theme_classic(
    base_size      = 12,
    base_family    = GRAPH_FONT,
    base_line_size = LINE_WIDTH
  ) %+replace%
  theme(
    axis.title.x = element_text(hjust = 1),
    axis.title.y = element_text(vjust = 1, angle = 0)
  )
)
```

```{r includes, cache=FALSE}
source("R/Formulas.R")
source("R/Graphical_example_transcriptions_Mario.R")
source("R/Output.R")
```

# Direction of the maximal slope from the origin in the MUPP-2PL model

## Original derivation by Mario Luzardo (circa 13-07-2018)

First,

\begin{equation}
  f_2(\theta_1, \theta_2) = `r LOGISTIC_RECT`                  (\#eq:model-rect)
\end{equation}

is equivalent to

\begin{equation}
  `r F_THETA_ALPHA` = `r LOGISTIC_POLAR`,                     (\#eq:model-polar)
\end{equation}

with

\begin{equation}
  \begin{split}
    \theta_1 = `r THETA_COS_ALPHA`, \\
    \theta_2 = `r THETA_SIN_ALPHA`.
  \end{split}                                             (\#eq:transform-polar)
\end{equation}

The first derivative of $`r F_THETA_ALPHA`$ (for $\theta$) is

\begin{equation}
  \begin{split}
    `r F_THETA_ALPHA_DIFF_THETA` & =
      \frac{`r EXPFUNC_POLAR`}{(`r LOG_DENOM_POLAR`)^2} `r A_COS_SIN_POLAR` \\
    & = `r LOGISTIC_POLAR` \\
      `r LOG_INV_POLAR` `r A_COS_SIN_POLAR` \\
    & = `r F_THETA_ALPHA` `r F_THETA_ALPHA_1_MINUS` `r A_COS_SIN_POLAR`.
  \end{split}                                           (\#eq:first-deriv-theta)
\end{equation}

Then, the second derivative (for $\theta$) is

\begin{equation}
  \begin{split}
    `r F_THETA_ALPHA_DIFF2_THETA` &=
      `r F_THETA_ALPHA` `r F_THETA_ALPHA_1_MINUS`^2 `r A_COS_SIN_POLAR`^2 - \\
      &`r F_THETA_ALPHA`^2 `r F_THETA_ALPHA_1_MINUS` `r A_COS_SIN_POLAR`^2 \\
      &= `r F_THETA_ALPHA` `r F_THETA_ALPHA_1_MINUS` `r A_COS_SIN_POLAR`^2
        [1 - 2 `r F_THETA_ALPHA`] \\
      &= `r A_COS_SIN_POLAR`^2 `r F_THETA_ALPHA`
        [1 - 3 `r F_THETA_ALPHA` + 2 `r F_THETA_ALPHA`^2].
  \end{split}                                          (\#eq:second-deriv-theta)
\end{equation}

The condition the multidimensional slope parameter must satisfy is

\begin{equation}
  `r F_THETA_ALPHA_DIFF2_THETA` = 0
    \Leftrightarrow
    `r F_THETA_ALPHA` = `r FRAC_1_2`.           (\#eq:condition-max-slope-theta)
\end{equation}

The slope is maximal (across $\theta$) if
$`r F_THETA_ALPHA` = \tfrac{1}{2}$,
and the maximal slope equals $\tfrac{1}{4} `r A_COS_SIN_POLAR`$.

For which $\alpha$ (i.e., in which direction) is

\begin{equation}
  `r ZETA_ALPHA` \coloneqq `r FRAC_1_4` `r A_COS_SIN_POLAR`
                                                           (\#eq:zeta-alpha-def)
\end{equation}

maximal?

\begin{equation}
  `r ZETA_ALPHA` = `r FRAC_1_4`
    \left( `r A1_COS_ALPHA` - `r A2_SIN_ALPHA_AS_COS` \right)
                                                           (\#eq:zeta-alpha-cos)
\end{equation}

\begin{equation}
  `r ZETA_ALPHA_DIFF_COS_ALPHA` =
    `r FRAC_1_4` a_1 +
      `r FRAC_1_4` a_2 \frac{`r COS_ALPHA`}{`r SIN_ALPHA_AS_COS`}
                                                         (\#eq:zeta-first-deriv)
\end{equation}

\begin{equation}
  \begin{split}
    `r ZETA_ALPHA_DIFF_COS_ALPHA` = 0
      & \Leftrightarrow
        a_1 + a_2 \frac{`r COS_ALPHA`}{`r SIN_ALPHA_AS_COS`} = 0 \\
      & \Leftrightarrow
        a_2 `r COS_ALPHA` = - a_1 `r SIN_ALPHA_AS_COS` \\
      & \Leftrightarrow
        a_2^2 `r COS_SQ_ALPHA` = (- a_1)^2 (1 - `r COS_SQ_ALPHA`) \\
      & \Leftrightarrow
        (a_1^2 + a_2^2) `r COS_SQ_ALPHA` = (- a_1)^2 \\
      & \Leftrightarrow
        `r COS_ALPHA` = `r MINUS_A1_OVER_SQRT_A_SQ`
  \end{split}                                   (\#eq:zeta-first-deriv-solution)
\end{equation}

That is, the maximal slope (in the direction from the origin) is found when

\begin{equation}
  `r COS_ALPHA` = `r MINUS_A1_OVER_SQRT_A_SQ`   (\#eq:condition-max-slope-alpha)
\end{equation}

and $\theta$ is such that $`r F_THETA_ALPHA` = \tfrac{1}{2}$.

The latter occurs when

\begin{equation}
  \begin{split}
    `r F_THETA_ALPHA` = `r LOGISTIC_MAX_THETA` = `r FRAC_1_2`
      \Leftrightarrow
      \theta = \frac {-l}{`r SQRT_A1_A2_SQUARED`}.
                                                 (\#eq:max-slope-alpha-solution)
  \end{split}
\end{equation}

## Highlights

- This derivation assumes orthogonality of the latent space, as per Equation
\@ref(eq:transform-polar).

- This document was handwritten by Mario (circa 13 July 2018),
and it is likely that the idea of generalizing to a non-orthonormal space was
still not mature by then.
The transformation of Equation \@ref(eq:zeta-alpha-def) into \@ref(eq:zeta-alpha-cos)
is the first step where the (implicit) orthogonality assumption
becomes relevant.

- Note also that these derivations apply to the MUPP-2PL model.
Clearly they are equivalent to the MCLM model derivations in
[@reckase_difficulty_1985].
However, precisely because of this equivalence,
the orthogonality assumption is equally implicit.

## Development of some steps

Some steps are rather complex so I expand the intermediate steps in order to
clarify the derivation.

### First derivative of $f(\theta, \alpha)$

The first step applies the rule
$\frac{\delta}{\delta x} \tfrac{1}{g(x)} = - \tfrac{g'(x)}{g^2(x)}$;
with

\begin{equation}
  \begin{split}
    g(\theta) &= `r LOG_DENOM_POLAR` \\
              &= 1 + \exp{[-`r A_COS_SIN_POLAR` \theta - l]},
  \end{split}                                        (\#eq:inv-rule-exponential)
\end{equation}

\begin{equation}
  g'(\theta) = - `r A_COS_SIN_POLAR` `r EXPFUNC_POLAR`
                                                   (\#eq:derivative-exponential)
\end{equation}

\begin{equation}
  \begin{split}
    `r F_THETA_ALPHA_DIFF_THETA` & = -
      \frac{ - `r A_COS_SIN_POLAR` `r EXPFUNC_POLAR` }
           {(`r LOG_DENOM_POLAR`)^2} \\
      & = \frac{`r EXPFUNC_POLAR`} {(`r LOG_DENOM_POLAR`)^2} `r A_COS_SIN_POLAR`
  \end{split}                                 (\#eq:first-deriv-theta-developed)
\end{equation}

### Second derivative of $`r F_THETA_ALPHA`$

First, the terms
$`r A_COS_SIN_POLAR`^2 `r F_THETA_ALPHA` `r F_THETA_ALPHA_1_MINUS`$
are extracted as a common factor:

\begin{equation}
  \begin{split}
    `r F_THETA_ALPHA_DIFF2_THETA` & =
      `r F_THETA_ALPHA` `r F_THETA_ALPHA_1_MINUS`^2 `r A_COS_SIN_POLAR`^2 - \\
        `r F_THETA_ALPHA`^2 `r F_THETA_ALPHA_1_MINUS` `r A_COS_SIN_POLAR`^2 \\
      & = `r DIFF2_COMMON_FACTOR` `r F_THETA_ALPHA_1_MINUS` - \\
        `r DIFF2_COMMON_FACTOR` `r F_THETA_ALPHA` \\
      & = `r DIFF2_COMMON_FACTOR` [1 - `r F_THETA_ALPHA` - `r F_THETA_ALPHA`] \\
      & = `r DIFF2_COMMON_FACTOR` [1 - 2 `r F_THETA_ALPHA`] \\
      & = `r DIFF2_COMMON_FACTOR_ALT_ORDER` [1 - 2 f(\theta, \alpha)].
  \end{split}                                (\#eq:second-deriv-theta-developed)
\end{equation}

Then, the product
$`r F_THETA_ALPHA_1_MINUS` [1 - 2 f(\theta, \alpha)]$
is expanded:

\begin{equation}
  \begin{split}
    `r F_THETA_ALPHA_DIFF2_THETA` & =
      `r DIFF2_COMMON_FACTOR_ALT_ORDER` [1 - 2 f(\theta, \alpha)] \\
      & = `r A_COS_SIN_POLAR`^2 `r F_THETA_ALPHA`
        [1 - 2`r F_THETA_ALPHA` - `r F_THETA_ALPHA` + 2f^2(\theta, \alpha) ] \\
      & = `r A_COS_SIN_POLAR`^2 `r F_THETA_ALPHA`
        [1 - 3`r F_THETA_ALPHA` + 2f^2(\theta, \alpha) ].
  \end{split}                              (\#eq:second-deriv-theta-developed-2)
\end{equation}

### Condition for the point of maximal slope

When $`r F_THETA_ALPHA_DIFF2_THETA` = 0$,
$`r A_COS_SIN_POLAR`^2 `r F_THETA_ALPHA` [1 - 3 `r F_THETA_ALPHA` + 2 `r F_THETA_ALPHA`^2] = 0$.
As this must be true for any value of $\alpha$,
the term $`r A_COS_SIN_POLAR`^2$ is irrelevant.
Thus, there is a root when $`r F_THETA_ALPHA` = 0$,
and two roots when $1 - 3 `r F_THETA_ALPHA` + 2 `r F_THETA_ALPHA`^2 = 0$.

There is one root when $`r F_THETA_ALPHA` = 1$:
$(1 - 3 · 1 + 2 · 1) = (1 - 3 + 2) = 0$.
However, $`r F_THETA_ALPHA` = 1$
when $\theta \rightarrow \infty$.
Also, $`r F_THETA_ALPHA` = 0$ is satisfied when $\theta \rightarrow \infty$
(depending on the value of $`r ALPHA`$,
$`r F_THETA_ALPHA`$ will be equal to 0 or 1).

The other root occurs when $`r F_THETA_ALPHA` = \tfrac{1}{2}$:
$(1 - 3 · 1/2 + 2 · (1/2)^2) = (1 - 1.5 + 0.5) = 0$,
as shown in Equation \@ref(eq:condition-max-slope-theta).
In such case,

\begin{equation}
  \begin{split}
    `r F_THETA_ALPHA_DIFF_THETA` &=
      \frac{1}{2} \left(1 - \frac{1}{2}\right) `r A_COS_SIN_POLAR` \\
      &= \frac{1}{4} `r A_COS_SIN_POLAR`
  \end{split}
\end{equation}

The value of $\theta$ for this maximal slope is then obtained
from \@ref(eq:max-slope-alpha-solution).
Note that, in \@ref(eq:max-slope-alpha-solution),
following the equivalent derivation in \@ref(eq:zeta-first-deriv-solution)
for $`r SIN_ALPHA`$, we get

\begin{equation}
  \begin{split}
  `r ZETA_ALPHA` &= `r FRAC_1_4`
    \left( `r A1_COS_ALPHA_AS_SIN` - `r A2_SIN_ALPHA` \right) \\
  `r ZETA_ALPHA_DIFF_SIN_ALPHA` &=
    - `r FRAC_1_4` \frac{`r SIN_ALPHA`}{`r COS_ALPHA_AS_SIN`} a_1 -
      `r FRAC_1_4` a_2 \\
    `r ZETA_ALPHA_DIFF_SIN_ALPHA` = 0
      & \Leftrightarrow
        - a_1 \frac{`r SIN_ALPHA`}{`r COS_ALPHA_AS_SIN`} - a_2 = 0 \\
      & \Leftrightarrow
        a_1 `r SIN_ALPHA` = - a_2 `r COS_ALPHA_AS_SIN` \\
      & \Leftrightarrow
        a_1^2 `r SIN_SQ_ALPHA` = (- a_2)^2 (1 - `r SIN_SQ_ALPHA`) \\
      & \Leftrightarrow
        (a_1^2 + a_2^2) `r SIN_SQ_ALPHA` = (- a_2)^2 \\
      & \Leftrightarrow
        `r SIN_ALPHA` = `r MINUS_A2_OVER_SQRT_A_SQ`
  \end{split}
\end{equation}

To get to the right part of this expression, the following intermediate steps
have been omitted[^sol_theta_max_slope]:

\begin{equation}
  \begin{split}
    `r LOG_DENOM_MAX_THETA` &= 2 \\
    `r EXP_MAX_THETA` &= 1 \\
    `r LOG_EXPONENT_MAX_THETA` &= 0 \\
    - \frac{a_1_2 + a_2^2}{ `r SQRT_A1_A2_SQUARED` } \theta &= -l \\
    \frac{ \left(`r SQRT_A1_A2_SQUARED` \right)^2 }{ `r SQRT_A1_A2_SQUARED` } \theta &= l \\
    `r SQRT_A1_A2_SQUARED` \theta &= l \\
    \theta &= \frac {l}{`r SQRT_A1_A2_SQUARED`}
  \end{split}
\end{equation}

[^sol_theta_max_slope]: It seems like there could be a mistake,
being the numerator in the left part of \@ref(eq:max-slope-alpha-solution)
$l$ instead of $-l$.
Also, as the exponent in \@ref(eq:max-slope-alpha-solution)
(i.e. inside the parenthesis) seems to be
$`r LOG_EXPONENT_MAX_THETA_CORR`$,
due to the minus sign of $a_2$ in Equation \@ref(eq:model-polar).
If so, the correct result in \@ref(eq:max-slope-alpha-solution) must be
$\tfrac {l`r SQRT_A1_A2_SQUARED`}{(a_1^2 - a_2^2)}$.

\newpage

# About the axis transformation and graphical representation in the MUPP-2PL model

## Original derivation by Mario Luzardo (13-07-2019)

### Transformation of the axes

There are two ways of treating it: geometrically, or as a change of base.

#### Geometrical interpretation of the axis transformation

Following the notation in my dissertation, let

\begin{equation}
  `r THETA_ORTH` = `r T_INVERSE` `r THETA`       (\#eq:orthogonalized-theta-def)
\end{equation}

with $`r THETA` \in V \subset R^D$,
$`r THETA_ORTH`$ the trait vector in the orthogonal space,
and $`r THETA`$ the trait vector in the oblique space.

Regarding $`r SIGMA` = `r T_MATRIX` `r T_MATRIX`'$: This is not correct.
If we are in the context of factor analysis
(where the variables and factors are standardized),
the correlation matrix fulfills the previous relationship,
as can be seen in Harman [-@harman_modern_1970].
`r insert_anchor("assertion-T-root-Sigma")`

**Justification:** Given Equation \@ref(eq:orthogonalized-theta-def),
$`r THETA` = `r T_MATRIX` `r THETA_ORTH`$.
If all the variables are standardized, then

\begin{equation}
  `r R_MATRIX` = `r COV`(`r THETA`, `r THETA`) =
    `r COV`(`r T_MATRIX``r THETA_ORTH`, `r T_MATRIX``r THETA_ORTH`) =
    `r T_MATRIX` `r COV`(`r THETA_ORTH`, `r THETA_ORTH`) `r T_MATRIX`' =
    `r T_MATRIX` `r T_MATRIX`',                         (\#eq:r-matrix-equality)
\end{equation}

because  $`r COV`(`r THETA_ORTH`, `r THETA_ORTH`) = `r ID_MATRIX`$,
as they are orthogonal and have variance 1.

If they are not standardized, but are just orthogonal instead,

\begin{equation}
  `r COV`(`r THETA`, `r THETA`) =
    `r T_MATRIX` `r D_MATRIX` `r T_MATRIX`', \qquad
    `r D_MATRIX` = \begin{pmatrix} 
      s_1^2  & \cdots &      0 \\
      \vdots & \ddots & \vdots \\
      0      & \cdots &  s_d^2 \\
    \end{pmatrix},                           (\#eq:sigma-matrix-equality-non-sd)
\end{equation}

that is, $`r SIGMA` = `r T_MATRIX` `r D_MATRIX` `r T_MATRIX`'$.

Say we are in the context where the orthogonal variables are standardized
(e.g., factor analysis or IRT estimation); then $`r T_MATRIX`$ is a rotation
and, because the distances are invariant to rotations, it is true that
$`r SIGMA` = `r T_MATRIX` `r T_MATRIX`'$.

We see that this is not the case,
because $`r T_MATRIX`$ is a transformation that is *not strictly* a rotation.

Let's see how the transformed parameters are. Considering the MCLM model,

\begin{equation}
  `r P_THETA` = `r LOGISTIC_MCLM` = `r LOGISTIC_MCLM_ORTH`     (\#eq:MCLM-model)
\end{equation}

that is, $(`r A_VECTOR_ORTH`)' = `r A_VECTOR`' `r T_MATRIX`$, or
$`r A_VECTOR_ORTH` = `r T_MATRIX`' `r A_VECTOR`$.

\begin{equation}
  `r MBL` = `r MBL_DEF`                                               (\#eq:MBL)
\end{equation}

(but the expression with $`r SIGMA`$ **instead of** $`r T_MATRIX``r T_MATRIX`'$,
i.e. in equation 3.6 in my dissertation, is not correct).

Now let's see the change of coordinates.
Let $`r COORD_SYSTEM`$ be the original, oblique coordinate system
that will be transformed into the orthogonal coordinate system,
$`r COORD_SYSTEM_ORTH`$[^orth_coords].
In order to this, we consider $`r THETA_1` \equiv `r THETA_1`^\mathrm{o}$
(i.e., it is not transformed),
and we transform $`r THETA_2`$ in $`r THETA_2`^\mathrm{o}$ such that
$`r THETA_2`^\mathrm{o} \perp `r THETA_1`^\mathrm{o}$. Graphically:

[^orth_coords]: Although Mario calls the orthogonal axes
$`r THETA_1`'$ and $`r THETA_2`'$, I think it is more appropriate to use the
notation $`r THETA_1`^\mathrm{o}$ and $`r THETA_2`^\mathrm{o}$,
as it is consistent with the notation used previously.

```{r coord-transform, fig.id="coord-transform", warning=FALSE, fig.cap="Axis transformation."}
GUIDE_SEGMENTS <- tibble(
  x    = c(0, 0, 2, 1.5),
  y    = c(0, 1, 0, 0),
  xend = c(2, 2, 2, 2),
  yend = c(4, 1, 1, 1),
  line = c("solid", "longdash" %>% rep(3))
)

point_labels <- tribble(
  ~x,  ~y, ~show,   ~label,
   0,   0,  FALSE,  '0', ## TODO: = in italics in the equation
   1.5, 0,  FALSE,  'A',
   2,   0,  FALSE,  'B',
   0.5, 1,  FALSE,  'C',
   0,   1,  FALSE,  'D',
   2,   1,  TRUE,   'P'
) %>%
  add_column(
    nudgex = c( 0  %>% rep(3), -.12 %>% rep(3)),
    nudgey = c(-.1 %>% rep(3),  .12 %>% rep(3))
  )

ggplot() +
  geom_segment(
    mapping = aes(x, y, xend = xend, yend = yend, linetype = I(line)),
    data    = GUIDE_SEGMENTS,
    size    = LINE_WIDTH
  ) +
  geom_point(x = 2, y = 1, aes(shape = I(POINT_CHAR))) +
  # geom_text(
  #   data     = point_labels,
  #   mapping  = aes(x, y, label = label),
  #   position = point_labels %$% position_nudge(nudgex, nudgey),
  #   family   = GRAPH_FONT
  # ) +
  annotate(
    geom     = "text",
    x        = point_labels$x + point_labels$nudgex,
    y        = point_labels$y + point_labels$nudgey,
    label    = point_labels$label %>% TeX(italic = TRUE),
    family   = GRAPH_FONT
  ) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = .5, start = pi / 2, end = acos(2 / sqrt(5))),
    size   = ANGLE_WIDTH
  ) +
  annotate(
    geom   = "text",
    label  = c("$\\theta_1$", glue("${ALPHA}$")) %>% TeX(),
    x      = c(1.75, .5),
    y      = c(3.88, .35),
    family = GRAPH_FONT
  ) +
  xlab("$\\theta_1 \\equiv \\theta_1^o$" %>% TeX()) +
  ylab("$\\theta_2^o$" %>% TeX()) +
  scale_x_continuous(limits = c(0, 4), breaks = NULL, oob = oob_keep) +
  scale_y_continuous(limits = c(0, 4), breaks = NULL, oob = oob_keep) +
  coord_fixed(clip = "off", expand = FALSE)
```

We know that $`r COS_ALPHA` = `r RHO_COORD_SYSTEM`$.
Let $P$ be a point
with coordinates $(x, y)$ in the system $`r COORD_SYSTEM`$
and coordinates $(x', y')$ in the system $`r COORD_SYSTEM_ORTH`$.

We have that[^point_notation]

\begin{equation}
  \begin{split}
    &0B = x' \qquad &0D = y', \\
    &0A = x  \qquad &0C = y,  \\
    &0B = 0A + AB.
  \end{split}
\end{equation}

[^point_notation]: Here Mario uses $`r THETA_1`$ and $`r THETA_2`$
instead of $x$ and $y$,
and $`r THETA_1`'$ and $`r THETA_2`'$ instead of $x'$ and $y'$;
however, he has just defined $(x, y)$ and $(x', y')$ as the coordinates of
point $P$ in the oblique and orthogonal reference system respectively.
Therefore, my understanding in that $`r THETA_1`$ should be considered as an
*axis*, and $x$ as a coordinate on that axis and so on,
so e.g., $x$ and $`r THETA_1`$ are semantically different symbols.
Therefore I'm using here $x$, $y$, $x'$ and $y'$ to represent the coordinates.

$x' = x + AP \, `r COS_ALPHA`$,
because $\measuredangle PAB = \alpha$;  
$AP = y$, because $AP = 0C$;  
therefore $\boxed{x' = x + y \, `r COS_ALPHA`}$.

Also, $0D = BP$, i.e.,  
$y' = AP \, `r SIN_ALPHA`$, $\boxed{y' = y \, `r SIN_ALPHA`}$

But $`r COS_ALPHA` = `r RHO_COORD_SYSTEM`$,
and $`r SIN_ALPHA` = `r SIN_ALPHA_COORD_SYSTEM`$.

\begin{equation}
  \begin{split}
    x' &= x + y \, `r RHO_COORD_SYSTEM` \\
    y' &=     y \, `r SIN_ALPHA_COORD_SYSTEM`
  \end{split}
\end{equation}

or, equivalently,

\begin{equation}
  \begin{pmatrix} 
    x' \\
    y' \\
  \end{pmatrix}
    =
    \begin{pmatrix} 
      1 & `r RHO_COORD_SYSTEM` \\
      0 & `r SIN_ALPHA_COORD_SYSTEM` \\
    \end{pmatrix}
    \begin{pmatrix} 
      x \\
      y \\
    \end{pmatrix}.                                        (\#eq:coord-transform)
\end{equation}

This is Equation \@ref(eq:orthogonalized-theta-def)
($`r THETA_ORTH` = `r T_INVERSE` `r THETA`$),
in the notation of my dissertation (Equation 3.3 there).

The previous equation in horizontal vector form is

\begin{equation}
  (x',  y') = (x, y)
    \begin{pmatrix} 
      1                    & 0                    \\
      `r RHO_COORD_SYSTEM` & `r SIN_ALPHA_COORD_SYSTEM` \\
    \end{pmatrix},                             (\#eq:coord-transform-horizontal)
\end{equation}

$`r THETA_ORTH`' = `r THETA`' (`r T_INVERSE`)'$.

#### Change-of-base interpretation of the axis transformation

Another way of putting this is by making a change of base.
The base of the orthogonal system $`r COORD_SYSTEM_ORTH`$
is $`r ORTHONORMAL_BASE`$, i.e., the canonical base.

The base of the oblique system $`r COORD_SYSTEM`$ will be given by the vectors

```{r oblique-base, fig.id="oblique-base", warning=FALSE, fig.cap="Oblique base."}
SLOPE_V2 <- 2

base_vectors <- tibble(
  x     = c(1, 1),
  y     = c(0, SLOPE_V2),
  label = c("$(1, 0)$" %>% TeX(), expression(bold(v)[2])),
  width = c(VECTOR_WIDTH %>% rep(2))
) %>% 
  mutate(
    norm = sqrt(x^2 + y^2),
    across(x:y,  ~ . / norm, .names = "{.col}_norm")
  )

base_vectors %>% ggplot() +
  geom_segment(x = 0, y = 0, xend = 1, yend = SLOPE_V2, size = LINE_WIDTH) +
  geom_segment(
    aes(xend = x_norm, yend = y_norm, size = I(width)),
    x        = 0,
    y        = 0,
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = .5, start = pi / 2, end = x_norm[2]),
    data = base_vectors,
    size = ANGLE_WIDTH
  ) +
  annotate(
    geom   = "text",
    label  = base_vectors$label %>% c(c("$\\theta_2$", "$\\alpha$") %>% TeX()),
    x      = c(.8, .27,  .85, .5),
    y      = c(.1, .75, 1.88, .27),
    family = GRAPH_FONT
  ) +
  xlab(TeX("$\\theta_1 \\equiv \\theta_1'$")) +
  ylab(NULL) +
  scale_x_continuous(limits = c(0, 2), breaks = NULL) +
  scale_y_continuous(limits = c(0, 2), breaks = NULL) +
  coord_fixed(clip = "off", expand = FALSE) +
  theme(axis.line.y = element_blank())
```

$`r V_2_VECTOR`$ is given by the normalization of the vector

\begin{equation}
  (1, \, \tan \alpha) = `r DIRECTOR_VECTOR`.              (\#eq:director-vector)
\end{equation}

Its norm is

\begin{equation}
  \left\| `r DIRECTOR_VECTOR` \right\|^2 = 1 + \frac{1 - \rho^2}{\rho^2} =
    \frac{1}{\rho^2},
\end{equation}

\begin{equation}
  \left\| `r DIRECTOR_VECTOR` \right\| = \frac{1}{\rho};
                                                     (\#eq:director-vector-norm)
\end{equation}

therefore

\begin{equation}
  `r V_2_VECTOR` = \frac{ \left( 1, \tfrac{`r SIN_ALPHA_AS_RHO`}{\rho} \right) }
                        { \tfrac{1}{\rho} }
                 = `r NORM_DIR_VECTOR`.     (\#eq:director-vector-normalization)
\end{equation}

The oblique base is

\begin{equation}
  `r OBLIQUE_BASE`.                                          (\#eq:oblique-base)
\end{equation}

The change of base matrix that we need contains the coordinates of the base
$`r OBLIQUE_BASE`$ in the base $`r ORTHONORMAL_BASE`$, that is,

\begin{equation}
  \begin{split}
    (1, \, 0)                                  &= 1 (1, \, 0) + 0 (0, \, 1) \\
    \left(\rho, \, `r SIN_ALPHA_AS_RHO`\right) &=
      \rho (1, \, 0) + `r SIN_ALPHA_AS_RHO` (0, \, 1).
  \end{split}                            (\#eq:change-of-base-matrix-derivation)
\end{equation}

The resulting matrix is

\begin{equation}
  `r T_INVERSE` = `r T_INVERSE_DEF`.                (\#eq:change-of-base-matrix)
\end{equation}

### Representation of the multidimensional parameters in an oblique space

In order to represent the *MBL* and *MBS* in the oblique axes
we should compute them in the orthogonal space,
as the formulae found by Reckase
[-@reckase_difficulty_1985, -@reckase_discriminating_1991]
are for orthogonal axes.

\begin{equation}
  `r MBL` = `r MBL_DEF_ORTH` \qquad `r MBS` = `r MBS_DEF_ORTH`
                                       (\#eq:multidimensional-params-orthogonal)
\end{equation}

If $`r GAMMA_COMPONENT`$ is the angle in axis $k$
(we do not call it $\alpha$,
as in @reckase_difficulty_1985 and @reckase_discriminating_1991
to avoid confusing it with the angle between
$`r THETA_1`$ and $`r THETA_2`$),

\begin{equation}
  `r DIRECTOR_COSINE` = \frac{`r A_COMPONENT_ORTH`}{`r MBS_DEF_ORTH`}
    \Rightarrow
    `r DIR_COS_VECTOR` = \frac{`r A_VECTOR_ORTH`}{`r MBS_DEF_ORTH`}.
                                                           (\#eq:dir-cos-vector)
\end{equation}

(`r DIR_COS_VECTOR` is the vector of director cosines.)

Given $`r A_VECTOR_ORTH`' = `r A_VECTOR`' `r T_MATRIX`$,

\begin{equation}
  \begin{split}
    `r MBL`            &= `r MBL_DEF` \\
    `r MBS`            &= `r MBS_DEF` \\
    `r DIR_COS_VECTOR` &=
      \frac{`r T_TRANSP_A`}{`r MBS_DEF`} =
      `r DIR_COS_VECTOR_DEF`
  \end{split}                             (\#eq:multidimensional-params-oblique)
\end{equation}

(with $`r GAMMA_VECTOR`$ the cosines expressed as a vertical vector).

```{r multidimensional-components-orthogonal, fig.id="multidimensional-components-orthogonal", warning=FALSE, fig.cap="Multidimensional parameters."}
SLOPE <- .8

POINTS <- tribble(
  ~x, ~y,   ~name,
   1,   .8,  'P',
   2,  1.6,  'Q'
)

angle <- atan(SLOPE)

POINTS %>% ggplot() +
  geom_abline(slope = SLOPE, size = LINE_WIDTH, linetype = "11") +
  geom_point(aes(x, y)) +
  # geom_text(
  #   # aes(x, y, label = substitute(expression(italics(name)), POINTS)),
  #   aes(x, y, label = name %>% TeX(italic = TRUE)),
  #   position = position_nudge(x = .05, y = -.05),
  #   family   = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    x      = POINTS$x + .05,
    y      = POINTS$y - .05,
    label  = POINTS$name %>% TeX(italic = TRUE),
    family = GRAPH_FONT
  ) +
  geom_arc(
    aes(
      x0 = 0,
      y0 = 0,
      r = c(.5, .6),
      start = c(pi / 2, 0),
      end = pi / 2 - angle
    ),
    size   = ANGLE_WIDTH
  ) +
  annotate(
    geom   = "text",
    label  = c("$\\gamma_1'$", "$\\gamma_2'$", "$(MBL_i)$", "$MBS_i$") %>% TeX(),
    x      = c(.55, .29, 1.05, 1.25),
    y      = c(.2, .65,  .65, 1.37),
    family = GRAPH_FONT
  ) +
  xlab(TeX("$\\theta_1^o$")) +
  ylab(TeX("$\\theta_2^o$")) +
  scale_x_continuous(limits = c(0, 2.2), breaks = NULL) +
  scale_y_continuous(limits = c(0, 2),   breaks = NULL) +
  coord_fixed(expand = FALSE)

grid.brackets(576, 656, 1062, 270, lwd = 1, col = "black")
```

We need to find the trasformed parameters in order to represent this
in the oblique axes.
Given $`r THETA` = `r T_MATRIX` `r THETA_ORTH`$[^theta_orth],
by Equation \@ref(eq:orthogonalized-theta-def),
we must transform points $P$ and $Q$ [^point_Q].

[^theta_orth]: Note that Mario uses here the notation
$`r THETA` = `r T_MATRIX` `r THETA`'$,
but he does not mean $`r THETA`$ transposed but a point in the coordinate system
$`r COORD_SYSTEM_ORTH`$, i.e., the orthogonal coordinate system.
This equality comes from Equation \@ref(eq:orthogonalized-theta-def),
$`r THETA_ORTH` = `r T_INVERSE` `r THETA` \Rightarrow `r T_MATRIX` `r THETA_ORTH` = `r T_MATRIX` `r T_INVERSE` `r THETA` \Rightarrow `r T_MATRIX` `r THETA_ORTH` = `r ID_MATRIX` `r THETA` \Rightarrow `r T_MATRIX` `r THETA_ORTH` = `r THETA`$

[^point_Q]: Here Mario calls the points $P$ and $T$,
but I think he later notices that the orthogonalizing matrix is named
$`r T_MATRIX`$ and changes the name of the point to $Q$ in order to avoid
confusion with it.

```{r multidimensional-components-oblique, fig.id="multidimensional-components-oblique", warning=FALSE, fig.cap="Point coordinates in the orthogonal space."}
guide_segments <- POINTS %>%
  slice(1:n() %>% rep(each = 2)) %>%
  mutate(across(x:y, .fns =  ~., .names = "{.col}_0")) %>%
  mutate(x_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(0, x_0)) %>% 
  mutate(y_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(y_0, 0))

POINTS %>% ggplot() +
  geom_abline(slope = SLOPE, size = LINE_WIDTH, linetype = "11") +
  geom_point(aes(x, y)) +
  # geom_text(
  #   aes(x, y, label = name),
  #   position = position_nudge(x = .05, y = -.05),
  #   family   = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    x      = POINTS$x + .05,
    y      = POINTS$y - .05,
    label  = POINTS$name %>% TeX(italic = TRUE),
    family = GRAPH_FONT
  ) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = .5, start = pi / 2, end = pi / 2 - angle),
    size   = ANGLE_WIDTH
  ) +
  annotate(
    geom   = "text",
    label  = c("$\\beta$", 'x', 'y',  'l') %>% TeX(italic = TRUE),
    # label  = expression(beta, italic(x), italic(y), italic(l)),
    x      = c( .55,        1,   -.04, 1.5),
    y      = c( .2,         -.04, .8,  1.25),
    family = GRAPH_FONT
  ) +
  xlab(TeX("$\\theta_1^o$")) +
  ylab(TeX("$\\theta_2^o$")) +
  scale_x_continuous(limits = c(0, 2.2), breaks = NULL, oob = oob_keep) +
  scale_y_continuous(limits = c(0, 2),   breaks = NULL, oob = oob_keep) +
  coord_fixed(expand = FALSE, clip = "off")
```

$P \equiv (x, y) \qquad Q \equiv (x', y')$[^quote_mark]

[^quote_mark]: These $(x', y')$ mean something different than before.
In the previous section, they made reference to the coordinates $(x, y)$
transformed to the orthonormal space.
Here though, both $(x, y)$ and $(x', y')$
are coordinates in the orthonormal space,
but $(x, y)$ are the coordinates of point $P$
and $(x', y')$ are the coordinates of point $Q$.

$(x, y)$, $\beta$[^beta_gamma1_equiv], and $l$[^l_clarification] are known.
A vector in the direction of the line with norm 1 is $`r DIR_VECTOR_NORM`$.

[^beta_gamma1_equiv]: $\beta$ may be called $\gamma_1$ directly;
this equivalence is stated later.

[^l_clarification]: $l$ is the norm of the segment $\overline{PQ}$,
NOT to be mistaken with $l_i$,
the intersection parameter of item $i$.

This comes from the normalization of

\begin{equation}
  `r DIR_VECTOR` = \left(1, `r TAN_BETA_DEF` \right).
\end{equation}

It's (squared) norm is

\begin{equation}
  1 + \frac{\sin^2 \beta}{\cos^2 \beta} = \frac{1}{\cos^2 \beta};
\end{equation}

thus normalizing,

\begin{equation}
  \frac{`r DIR_VECTOR`}{\tfrac{1}{\cos \beta}} = `r DIR_VECTOR_NORM`.
\end{equation}

$Q = P + l \, `r DIR_VECTOR_NORM` = (`r X_COORD_Q`, `r Y_COORD_Q`)$.

Then, the transformed coordinates are

\begin{equation}
  `r T_MATRIX` \begin{pmatrix} x \\ y \\ \end{pmatrix}
                                               (\#eq:point-P-coords-transformed)
\end{equation}

and

\begin{equation}
  `r T_MATRIX` \begin{pmatrix} `r X_COORD_Q` \\ `r Y_COORD_Q` \\ \end{pmatrix}.
                                               (\#eq:point-Q-coords-transformed)
\end{equation}

We know that $P \equiv `r MBL` `r DIR_VECTOR_NORM`$, $l = `r MBS`$,
and $\cos \beta$ is the component of $`r DIR_COS_VECTOR`$
with respect to axis $`r THETA_1`'$.

\begin{equation}
  `r T_MATRIX` P = `r MBL_DEF` `r T_MATRIX` `r DIR_VECTOR_NORM`
                                                      (\#eq:point-P-transformed)
\end{equation}

Let $e' = (0, \dots, 0, 1, 0, \dots, 0)$ be the vector with a 1 in the
component that represents $`r THETA_1`$.
Given the Equations in \@ref(eq:multidimensional-params-oblique),

\begin{equation}
  \begin{split}
    \cos \beta &= \frac{e' `r T_TRANSP_A`}{`r MBS_DEF`} \\
    \sin \beta &= `r SIN_BETA_AS_COS`
  \end{split},                                           (\#eq:cos-sin-beta-def)
\end{equation}

\begin{equation}
  Q \equiv \left[ `r MBL_DEF` - sgn(l_i) `r MBS_DEF` \right]
    `r DIR_VECTOR_NORM_T`,                                        (\#eq:point-Q)
\end{equation}

and[^sign_l_i]`r insert_anchor("fn-sign_l_i")`

\begin{equation}
  `r T_MATRIX` Q = \left[ `r MBL_DEF` - sgn(l_i) `r MBS_DEF` \right]
    `r T_MATRIX` `r DIR_VECTOR_NORM_T`.               (\#eq:point-Q-transformed)
\end{equation}

[^sign_l_i]: I don't understand where $sgn(l_i)$ comes from in
Equations \@ref(eq:point-Q) and \@ref(eq:point-Q-transformed).
As I understand it, the direction should be given, in any case,
by the sign of the component on the axis, that is, $sgn(a_{i1})$.
However, $\sin \beta$ may (or should?) already imply the sign here.

From Equation \@ref(eq:change-of-base-matrix), we have

\begin{equation}
  `r T_INVERSE` = `r T_INVERSE_DEF`;
\end{equation}

\begin{equation}
  \det(`r T_INVERSE`) = `r SIN_ALPHA_AS_RHO`,
\end{equation}

\begin{equation}
  `r T_MATRIX` = \frac{
    \begin{pmatrix}
      `r SIN_ALPHA_AS_RHO` & 0 \\
      -\rho                & 1
    \end{pmatrix}'
  }
  {`r SIN_ALPHA_AS_RHO`} =
    \frac{
      \begin{pmatrix}
        `r SIN_ALPHA_AS_RHO` & -\rho \\
        0                    & 1
      \end{pmatrix}
    }
    {`r SIN_ALPHA_AS_RHO`} = `r T_MATRIX_DEF`.        (\#eq:T-matrix-definition)
\end{equation}

We check that $`r T_MATRIX` `r T_INVERSE` = `r ID_MATRIX`$:

\begin{equation}
  `r T_MATRIX` `r T_INVERSE` = `r T_MATRIX_DEF` `r T_INVERSE_DEF` =
    \begin{pmatrix}
      1 & 0 \\
      0 & 1
    \end{pmatrix}.
\end{equation}

However, $`r T_MATRIX` `r T_MATRIX`'$ is not equal to $`r SIGMA`$:
[^T_squared_clarification]

[^T_squared_clarification]: But $`r T_MATRIX` `r T_MATRIX`' = `r SIGMA`^{-1}$,
as we will see later on.

\begin{equation}
  \begin{split}
    `r T_MATRIX` `r T_MATRIX`' &= `r T_MATRIX_DEF`
      \begin{pmatrix}
        1                                   & 0 \\
        -\tfrac{\rho}{`r SIN_ALPHA_AS_RHO`} &
          \tfrac{1}{`r SIN_ALPHA_AS_RHO`}
      \end{pmatrix} =
      \begin{pmatrix}
        1 + \frac{\rho^2}{1 - \rho^2} & \frac{-\rho}{1 - \rho^2} \\
            \frac{-\rho}{1 - \rho^2}  &     \frac{1}{1 - \rho^2}
      \end{pmatrix} \\
      &=
      \begin{pmatrix}
            \frac{1}{1 - \rho^2} & \frac{-\rho}{1 - \rho^2} \\
        \frac{-\rho}{1 - \rho^2} &     \frac{1}{1 - \rho^2}
      \end{pmatrix}                         (\#eq:T-T-transposed-matrix-product)
  \end{split}
\end{equation}

#### Example of graphical representation

Let's assume an item $a_1$ in a bidimensional latent space where
$\rho = `r RHO`$, with parameters $a_{1_1} = `r A_1_1`$, $a_{1_2} = `r A_1_2`$,
and $l_1 = `r L_1`$:

\begin{equation}
  `r A_1_VECTOR`' = `r A_1_VECTOR_DEF_T`
\end{equation}

\begin{equation}
  `r T_MATRIX` = `r T_MATRIX_DEF_EXAMPLE`
\end{equation}

\begin{equation}
  `r A_1_VECTOR_ORTH`' = `r A_1_VECTOR_DEF_T` \begin{pmatrix}
    1 & `r T_VALUE[1, 2]` \\
    0 & `r T_VALUE[2, 2]`
  \end{pmatrix}
\end{equation}

\begin{equation}
  `r T_MATRIX` `r T_MATRIX`' = `r T_SQ`
\end{equation}

\begin{equation}
  \begin{split}
    `r PROD_SQ_A_1_T` &= `r A_1_VECTOR_DEF_T` `r T_SQ` `r A_1_VEC_DEF_VERT` \\
                      &= \left(
                        \tfrac{8}{3} + \tfrac{2}{6}, \,
                        -\tfrac{4}{3} - \tfrac{4}{6}
                      \right)
                           `r A_1_VEC_DEF_FRAC` \\
                      &= (3, \, -2) `r A_1_VEC_DEF_FRAC` =
                        `r A_1_ORTH_SQ_VALUE`
  \end{split}
\end{equation}

\begin{equation}
  `r MBL_1` = \frac{`r L_1`}{\sqrt{`r PROD_SQ_A_1_T`}} =
    -\frac{`r L_1`}{\sqrt{`r A_1_ORTH_SQ_VALUE`}} = `r MBL_1_VALUE`.
\end{equation}

We need to find the direction

\begin{equation}
  \cos \beta = \frac{e' `r PROD_T_A_1`}{\|`r PROD_T_A_1`\|} =
    \frac{(1, \, 0) `r PROD_T_A_1`}{\sqrt{`r A_1_ORTH_SQ_VALUE`}}.
\end{equation}

($e'$ is the direction given by $\theta_1 \equiv \theta_1'$.)

In our case,

\begin{equation}
  `r T_MATRIX`' =
    `r T_MATRIX_TRANSP_DEF` =
    \begin{pmatrix}
      1                                             & 0 \\
      \frac{-\tfrac{1}{2}}{\sqrt{1 - \tfrac{1}{4}}} &
        \frac{1}{\sqrt{1 - \tfrac{1}{4}}}
    \end{pmatrix} =
    `r T_MATRIX_TRANSP_FRAC_DEF`.
\end{equation}

Thus,

\begin{equation}
  `r PROD_T_A_1` = `r T_MATRIX_TRANSP_FRAC_DEF` `r A_1_VEC_DEF_FRAC` =
    \begin{pmatrix}
      2 \\
      - \frac{2}{\sqrt{3}} - \frac{1}{\sqrt{3}}
    \end{pmatrix} =
    \begin{pmatrix}
      2 \\
      - \frac{3}{\sqrt{3}}
    \end{pmatrix} =
    \begin{pmatrix}
      2 \\
      - \sqrt{3}
    \end{pmatrix},
\end{equation}

\begin{equation}
  e' `r PROD_T_A_1` = (1, \, 0)
    \begin{pmatrix}
      2 \\
      - \sqrt{3}
    \end{pmatrix} =
    2,
\end{equation}

\begin{equation}
  \cos \beta = \frac{2}{\sqrt{7}} = `r COS_BETA_VALUE` \Rightarrow
    \beta \approx `r number(BETA_VALUE_DEG, 1e-2)`º.
\end{equation}

```{r graphical-representation-example-angles, fig.id="graphical-representation-example-angles", fig.cap="Director angles."}
angles <- tibble(init = c(0, BETA_VALUE, pi / 2, BETA_VALUE + pi, 3/2 * pi)) %>%
  mutate(end = init %>% lead(1)) %>%
  slice(-n()) %>%
  mutate(across(.fns = ~pi / 2 - ., .names = "{.col}_alt")) %>%
  add_column(rad = c(.5, .9, .7, .5)) %>%
  rowwise() %>%
  mutate(
    angle       = (end - init) / pi * 180,
    angle_label = angle %>% number(1e-2, suffix = "º"),
    label_angle = mean(c_across(init:end)),
    label_rad   = rad + .3,
    label_x     = cos(label_angle) * label_rad,
    label_y     = sin(label_angle) * label_rad
  ) %>%
  ungroup()

angles %>% ggplot() +
  geom_abline(slope = BETA_VALUE %>% tan()) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = rad, start = init_alt, end = end_alt),
    size   = ANGLE_WIDTH
  ) +
  geom_text(
    aes(label_x, label_y, label = angle_label),
    family   = GRAPH_FONT
  ) +
  annotate(
    geom   = "text",
    label  = "direction",
    x      = 1.8,
    y      = 1.2,
    family = GRAPH_FONT
  ) +
  scale_x_continuous(
    limits = c(-2, 2),
    breaks = 0,
    labels = NULL,
    name   = NULL
  ) +
  scale_y_continuous(
    limits = c(-2, 2),
    breaks = 0,
    labels = NULL,
    name   = NULL
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme(
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid.major = element_line(color = "black", size = LINE_WIDTH)
  )
```

$\sin \beta = \sqrt{1 - \tfrac{4}{7}} = \sqrt{\tfrac{3}{7}} \approx `r sqrt(3/7)`$
[^sine_neg]`r insert_anchor("fn-sine_neg")`

[^sine_neg]: I think there is a mistake here, as the sine function should
be negative here, i.e., $-`r SIN_ALPHA_AS_RHO`$ in this case.
I think this because the coordinate $a_{1_2}$ is negative
($a_{1_2} = `r A_1_2`$), and therefore the vector should be pointing towards
a direction in the fourth quadrant (down-right) but,
as we can see in Figure
\@ref(fig:graphical-representation-example-angles), this is not the case.
This makes me think that the sign of $a_{1_2}$ must be considered somehow in
computing the coordinates in this case.

Let's plot $P$:

```{r graphical-representation-example-point-P, fig.id="graphical-representation-example-point-P", fig.cap="Origin of the item vector."}
angle_dir <- angles %>% slice(1) %>%
  mutate(
    label_rad = .6,
    label_x   = cos(label_angle) * label_rad,
    label_y   = sin(label_angle) * label_rad
  )
point_P <- tibble(coord = c('x', 'y'), value = A_1_START_COORDS) %>%
  pivot_wider(names_from = coord)

guide_segments <- point_P %>%
  slice(1 %>% rep(2)) %>%
  mutate(across(x:y, .fns =  ~., .names = "{.col}_0")) %>%
  mutate(x_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(0, x_0)) %>% 
  mutate(y_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(y_0, 0))

angle_dir %>% ggplot() +
  geom_abline(slope = tan(BETA_VALUE)) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = rad, start = init_alt, end = end_alt),
    size = ANGLE_WIDTH
  ) +
  geom_text(
    aes(label_x, label_y, label = angle_label),
    family   = GRAPH_FONT
  ) +
  geom_point(aes(x, y), data = point_P) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  annotate(
    geom   = "text",
    label  = MBL_1_VALUE %>% number(1e-4),
    x      = -.18,
    y      = -.07,
    family = GRAPH_FONT
  ) +
  scale_x_continuous(
    limits = c(-.65, .65),
    breaks = 0,
    labels = NULL,
    name   = NULL
  ) +
  scale_y_continuous(
    limits = c(-.65, .65),
    breaks = 0,
    labels = NULL,
    name = NULL
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme(
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid.major = element_line(color = "black", size = LINE_WIDTH)
  )
```

\begin{equation}
  P' \equiv `r MBL_1_VALUE`
    \left( \tfrac{2}{\sqrt{7}}, \, \sqrt{\tfrac{3}{7}} \right) =
    (`r A_1_START_COORDS`)
\end{equation}

$Q$ has the form[^point_transposed]

\begin{equation}
  Q' \equiv \left[ `r MBL_DEF` - sgn(l_i) `r MBS_DEF` \right]
    `r DIR_VECTOR_NORM`
\end{equation}

as given by Equation \@ref(eq:point-Q).

\begin{equation}
  Q' \equiv \left[ -\frac{`r L_1`}{\sqrt{`r A_1_ORTH_SQ_VALUE`}} -
    `r A_1_ORTH_SQ_VALUE` \right]
    \left(
      \tfrac{2}{\sqrt{`r A_1_ORTH_SQ_VALUE`}},
      \sqrt{\tfrac{3}{`r A_1_ORTH_SQ_VALUE`}}
    \right) =
    (`r A_1_END_COORDS`)
\end{equation}

[^point_transposed]: I use the notation $P'$ and $Q'$ here,
as we are representing all points as vertical vectors,
but it makes no difference otherwise with Mario's notation ($P$ and $Q$),
given the term in square brackets is a scalar
(so transposing these vectors makes no difference in the result).

```{r graphical-representation-example-vector, fig.id="graphical-representation-example-vector", fig.cap="Complete item vector in the orthogonal space.", warning=FALSE}
angle_dir <- angles %>% slice(1) %>%
  mutate(
    rad       = .3,
    label_rad = .6,
    label_x   = cos(label_angle) * label_rad,
    label_y   = sin(label_angle) * label_rad
  )

points <- tibble(
  coord = c('x', 'y') %>% rep(2),
  value = c(A_1_START_COORDS, A_1_END_COORDS),
  point = c('P', 'Q') %>% rep(each = 2)
) %>%
  pivot_wider(id_cols = point, names_from = coord)

guide_segments <- points %>%
  slice(2 %>% rep(2)) %>%
  mutate(across(x:y, .fns =  ~., .names = "{.col}_0")) %>%
  mutate(x_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(0, x_0)) %>% 
  mutate(y_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(y_0, 0))

item_vector <- points %>% pivot_wider(names_from = point, values_from = x:y)

angle_dir %>% ggplot() +
  geom_abline(slope = tan(BETA_VALUE)) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = rad, start = init_alt, end = end_alt),
    size = ANGLE_WIDTH
  ) +
  geom_text(
    aes(label_x, label_y, label = angle_label),
    family   = GRAPH_FONT
  ) +
  geom_point(aes(x, y), data = points) +
  # geom_text(
  #   aes(x, y, label = point),
  #   data   = points,
  #   position = position_nudge(y = -.2),
  #   family = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    x      = points$x,
    y      = points$y - .2,
    label  = points$point %>% TeX(italic = TRUE),
    family = GRAPH_FONT
  ) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  geom_segment(
    aes(x_P, y_P, xend = x_Q, yend = y_Q, size = I(VECTOR_WIDTH)),
    data     = item_vector,
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  scale_x_continuous(
    limits = c(-3, 1),
    breaks = -3:-1,
    minor_breaks = 0,
    name   = NULL
  ) +
  scale_y_continuous(
    limits = c(-3, 1),
    breaks = -3:-1,
    minor_breaks = 0,
    name   = NULL
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme(
    axis.line        = element_blank(),
    panel.grid.major = element_line(
      color    = "black",
      size     = LINE_WIDTH,
      linetype = "17"
    ),
    panel.grid.minor = element_line(color = "black", size = LINE_WIDTH)
  )
```

What we are interested in is the plot in the oblique plane:

\begin{equation}
  `r T_MATRIX` P = \begin{pmatrix}
    1 & `r T_VALUE[1, 2]` \\
    0 & `r T_VALUE[2, 2]`
  \end{pmatrix}
  \begin{pmatrix}
    `r A_1_START_COORDS[1]` \\
    `r A_1_START_COORDS[2]`
  \end{pmatrix} =
  \begin{pmatrix}
    `r A_1_START_COORDS_OBLIQUE[1]` \\
    `r A_1_START_COORDS_OBLIQUE[2]`
  \end{pmatrix}
\end{equation}

\begin{equation}
  `r T_MATRIX` Q = \begin{pmatrix}
    1 & `r T_VALUE[1, 2]` \\
    0 & `r T_VALUE[2, 2]`
  \end{pmatrix}
  \begin{pmatrix}
    `r A_1_END_COORDS[1]` \\
    `r A_1_END_COORDS[2]`
  \end{pmatrix} =
  \begin{pmatrix}
    `r A_1_END_COORDS_OBLIQUE[1]` \\
    `r A_1_END_COORDS_OBLIQUE[2]`
  \end{pmatrix}
\end{equation}

Given $\rho = `r RHO`$, the angle in $`r THETA_1`$ and $`r THETA_2`$ is
$`r RHO_ARC`$º[^rot_matrix].

[^rot_matrix]: As we will see later on, following the procedure outlined by
Mario gives an incorrect result, where the vector is actually "rotated"
counterclockwise instead of clockwise,
as would be expected and shown in this plot.
The reason, as we will see, is that the transformation must be done by
pre-multiplying by $`r T_MATRIX`'$.
In this plot however, for reasons I do not fully understand yet,
I had to pre-multiply by $`r T_INVERSE`$ instead of $`r T_MATRIX`'$
in order to obtain Mario's plot.

```{r graphical-representation-example-vector-oblique, include=FALSE}
## This chunk follows exactly the method explained by Mario, but it
##   does not work as expected.

points_obl <- tibble(
  coord = c('x', 'y') %>% rep(2),
  value = c(A_1_START_COORDS_OBLIQUE, A_1_END_COORDS_OBLIQUE),
  point = c('P', 'Q') %>% rep(each = 2)
) %>%
  pivot_wider(id_cols = point, names_from = coord) %>%
  mutate(point_label = glue("{T_MATRIX}{point} = {point}_{{oblique}}"))

guide_segments_obl <- guide_segments %>%
  add_column(name = c("H_spike", "V_spike")) %>%
  pivot_longer(cols = where(is.numeric), names_to = "coord") %>%
  mutate(pt = coord %>% str_detect(., "_0") %>% if_else("Q", "axis")) %>%
  group_by(name, pt) %>%
  mutate(value_obl = drop(T_VALUE %*% value)) %>%
  ungroup() %>%
  pivot_wider(id_cols = name, names_from = coord, values_from = value_obl)

item_vector_obl <- points_obl %>%
  select(-point_label) %>% 
  pivot_wider(names_from = point, values_from = x:y)

ggplot() +
  geom_abline(slope = tan(RHO_ARC_RAD), size = LINE_WIDTH) +
  geom_hline(yintercept = 0, size = LINE_WIDTH) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = .3, start = pi / 2, end = pi / 2 - RHO_ARC_RAD),
    size = ANGLE_WIDTH
  ) +
  geom_text(
    aes(x = .4, y = .2),
    label  = RHO_ARC %>% number(suffix = 'º'),
    family = GRAPH_FONT
  ) +
  geom_point(aes(x, y), data = points_obl) +
  # geom_text(
  #   aes(x, y, label = point_label),
  #   data     = points_obl,
  #   position = position_nudge(y = -.2),
  #   family   = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    label  = points_obl$point_label %>% TeX(italic = TRUE),
    x      = points_obl$x + .5,
    y      = points_obl$y - .2,
    family = GRAPH_FONT
  ) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments_obl,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  geom_segment(
    aes(x_P, y_P, xend = x_Q, yend = y_Q, size = I(VECTOR_WIDTH)),
    data     = item_vector_obl,
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  scale_x_continuous(
    limits = c(-3, 1),
    breaks = 0,
    labels = NULL,
    name   = NULL,
    oob = oob_keep
  ) +
  scale_y_continuous(
    limits = c(-3, 1),
    breaks = 0,
    labels = NULL,
    name   = NULL,
    oob = oob_keep
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme(
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid.major = element_blank()
  )
```

```{r graphical-representation-example-vector-oblique-alt, fig.id="graphical-representation-example-vector-oblique-alt", warning=FALSE, fig.cap="Complete item vector in the oblique space."}
## This chunk replicates the previous one, but uses the inverse of the T
##   matrix (as defined by Mario) as the coordinate transformation.

A_1_START_COORDS_OBLIQUE_ALT <- T_INV_VAL %*% A_1_START_COORDS
A_1_END_COORDS_OBLIQUE_ALT   <- T_INV_VAL %*% A_1_END_COORDS

points_obl <- tibble(
  coord = c('x', 'y') %>% rep(2),
  value = c(A_1_START_COORDS_OBLIQUE_ALT, A_1_END_COORDS_OBLIQUE_ALT),
  point = c('P', 'Q') %>% rep(each = 2)
) %>%
  pivot_wider(id_cols = point, names_from = coord) %>%
  mutate(point_label = glue("{T_MATRIX}{point} = {point}_{{oblique}}"))

guide_segments_obl <- guide_segments %>%
  add_column(name = c("H_spike", "V_spike")) %>%
  pivot_longer(cols = where(is.numeric), names_to = "coord") %>%
  mutate(pt = coord %>% str_detect(., "_0") %>% if_else("Q", "axis")) %>%
  group_by(name, pt) %>%
  mutate(value_obl = drop(T_INV_VAL %*% value)) %>%
  ungroup() %>%
  pivot_wider(id_cols = name, names_from = coord, values_from = value_obl)

item_vector_obl <- points_obl %>%
  select(-point_label) %>%
  pivot_wider(names_from = point, values_from = x:y)

ggplot() +
  geom_abline(slope = tan(RHO_ARC_RAD), size = LINE_WIDTH) +
  geom_abline(
    slope     = tan(RHO_ARC_RAD),
    intercept = -tan(RHO_ARC_RAD) * (-5:2),
    size      = LINE_WIDTH,
    linetype  = "17"
  ) +
  geom_hline(yintercept = 0, size = LINE_WIDTH) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = .3, start = pi / 2, end = pi / 2 - RHO_ARC_RAD),
    size = ANGLE_WIDTH
  ) +
  geom_text(
    aes(x = .45, y = .22),
    label  = RHO_ARC %>% number(suffix = 'º'),
    family = GRAPH_FONT
  ) +
  geom_point(aes(x, y), data = points_obl) +
  # geom_text(
  #   aes(x, y, label = point_label),
  #   data     = points_obl,
  #   position = position_nudge(y = -.2),
  #   family   = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    label  = points_obl$point_label %>% TeX(italic = TRUE),
    x      = points_obl$x + .5,
    y      = points_obl$y - .2,
    family = GRAPH_FONT
  ) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments_obl,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  geom_segment(
    aes(x_P, y_P, xend = x_Q, yend = y_Q, size = I(VECTOR_WIDTH)),
    data     = item_vector_obl,
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  scale_x_continuous(
    limits       = c(-4, 1),
    breaks       = (-2:2) - 3 / tan(RHO_ARC_RAD),
    labels       = -2:2,
    minor_breaks = NULL,
    name         = NULL,
    oob          = oob_keep
  ) +
  scale_y_continuous(
    limits       = c(-3, 1),
    breaks       = -3:1 * cos(RHO),
    labels       = function(x) x / cos(RHO),
    minor_breaks = 0,
    name         = NULL,
    oob          = oob_keep
  ) +
  coord_fixed(expand = FALSE, clip = "on") +
  theme(
    axis.line          = element_blank(),
    panel.grid.major.y = element_line(
      color    = "black",
      size     = LINE_WIDTH,
      linetype = "17"
    )
  )
```

## Highlights and comments

### About the transformation matrix $`r T_MATRIX`$

Mario
[states that my assertion that $`r SIGMA` = `r T_MATRIX` `r T_MATRIX`'$ is not correct](#assertion-T-root-Sigma)
(see Equations \@ref(eq:r-matrix-equality) and
\@ref(eq:sigma-matrix-equality-non-sd)).
This formula is the definition of a square root matrix [@harman_modern_1970];
that is, $`r T_MATRIX`$ is a square root matrix of $`r SIGMA`$ if they fulfill
that equality (which is implicit in Equation 3.6 in my dissertation).
However, a transformation matrix obtained by the Gram-Schmidt method
*orthogonalizes* the axes, but does not *normalize* (i.e., standardize) them.
Therefore, if
$`r T_MATRIX`$ is a square root matrix of $`r SIGMA`$ then it should be called
an *orthonormalizing* matrix instead, but this is not the same as the
transformation matrix obtained by the Gram-Schmidt method
(or any other orthogonalizing method).
Conversely, if $`r T_MATRIX`$ is an orthogonalizing matrix,
then it is not necessarily a square root matrix of $`r SIGMA`$, as Mario says.
(And, as he adds, this would only be true if $`r SIGMA` = `r R_MATRIX`$.)

In order to generalize these assertions, we may define the following:

Let $\mathbf{N}$ be an orthogonalizing matrix, obtained by the Gram-Schmidt
(or any other) method.
Then,
$`r THETA`^\mathbf{p} = \mathbf{N}^{-1} `r THETA`$ is the trait vector in an
orthogonal space such that
$Cov(`r THETA`^\mathbf{p}, `r THETA`^\mathbf{p}) = \mathbf{D}$.
Let also $\mathbf{S}$ be a scaling matrix,
such that $\mathbf{S} \mathbf{S}' = \mathbf{D}$,
with $\mathbf{D}$ a diagonal matrix.
Then, $`r THETA_ORTH` = \mathbf{S}^{-1} `r THETA`^\mathbf{p}$
is the trait vector in the *orthonormal* space, such that
$Cov(`r THETA_ORTH`, `r THETA_ORTH`) = `r ID_MATRIX`$.
Then, we can define the transformation matrix
$`r T_MATRIX`  = \mathbf{N} \mathbf{S}$ such that
$`r THETA_ORTH` = \mathbf{S}^{-1} `r THETA`^\mathbf{p} = \mathbf{S}^{-1} \mathbf{N}^{-1} `r THETA` = (\mathbf{N} \mathbf{S})^{-1} `r THETA` =  `r T_INVERSE` `r THETA`$.
Given this, $`r T_MATRIX`$ would be a square root matrix of
$`r SIGMA`$, i.e., $`r T_MATRIX` `r T_MATRIX`' = `r SIGMA`$.

### About the product $`r T_MATRIX` `r T_MATRIX`$' {#matrix-squared}

Equation \@ref(eq:T-T-transposed-matrix-product) seems to contradict Equation
\@ref(eq:r-matrix-equality). How is this possible?
Note that if $`r T_MATRIX`$ is an orthogonalizing matrix,
we have seen that it is *not necessarily* a square root matrix of $`r SIGMA`$.
It will be if $`r T_MATRIX`$ is an *orthonormalizing* matrix, as seen before.
However, it should be *necessarily* a square root matrix of $`r R_MATRIX`$.

Equation \@ref(eq:r-matrix-equality) though shows that $`r T_MATRIX`$ is a
square root matrix of $`r R_INVERSE`$ instead,
as $`r T_MATRIX` `r T_MATRIX`' = `r R_INVERSE`$.
We show this by computing the inverse of $`r T_MATRIX` `r T_MATRIX`'$.

\begin{equation}
  \det(`r T_MATRIX` `r T_MATRIX`') =
    \frac{1}{(1 - \rho^2)^2} - \frac{\rho^2}{(1 - \rho^2)^2} =
    \frac{1 - \rho^2}{(1 - \rho^2)^2} =
    \frac{1}{1 - \rho^2}
\end{equation}

\begin{equation}
  (`r T_MATRIX` `r T_MATRIX`')^{-1} = \frac{
    \begin{pmatrix}
         \tfrac{1}{1 - \rho^2} & \tfrac{\rho}{1 - \rho^2} \\
      \tfrac{\rho}{1 - \rho^2} &    \tfrac{1}{1 - \rho^2}
    \end{pmatrix}
  }{\tfrac{1}{1 - \rho^2}} = \begin{pmatrix}
       1 & \rho \\
    \rho &    1
  \end{pmatrix} = `r R_MATRIX`
\end{equation}

Why this result?
It is not clear to me by now, but the explanation may be that the item
parameters do not belong to the same space as the trait parameters and should
not be represented in that same space.
This result appears in Zhang & Stout [-@zhang_theoretical_1999]:
"Note that although both the latent space and the test space
are *d*-dimensional Euclidean spaces,
mathematically they are different spaces because the inner products
in those two spaces are different" (p. 221).
It is worth considering this reference before proceeding.
However, that seems to contradict Reckase and McKinley
[-@reckase_difficulty_1985; -@reckase_discriminating_1991],
who (apparently) define the multidimensional parameters
in the latent trait space.
(This seems to make sense as well,
as the item features give a sense of how the item "behaves"
in relationship to the variation of the measured latent traits.)
Up to now, I have not found the correct solution to this issue.

### About the graphical representation in an oblique space

For what I can say, including the sign of $l_i$ in Equations
\@ref(eq:point-Q) and \@ref(eq:point-Q-transformed) must be wrong.
Point $Q$ is the sum of two vectors: One representing the *difficulty*
parameter (given by point $P$),
and another one representing the *discrimination* parameter,
represented by the distance
$l \equiv \overline{PQ}$
and the direction given by $`r GAMMA_VECTOR`$.
The two terms in the square brackets in
\@ref(eq:point-Q) and \@ref(eq:point-Q-transformed) correspond to the point
$Q$ and the distance $l$, respectively;
$l_i$ is only involved in computing the *diffculty*,
and thus in representing $P$.
Therefore, the sign of $l_i$ should not be involved in computing
the second term in the square brackets.

Also, there is the issue that applying the $`r T_MATRIX`$ transformation
leads to an incorrect graphical representation in the oblique plane.
Following Mario's example,
when we compute the coordinates in the oblique plane by
pre-multiplying by $`r T_MATRIX`$, the following plot results:

```{r graphical-representation-example-vector-oblique, fig.id="graphical-representation-example-vector-oblique", warning=FALSE, fig.cap=glue("Item vector in the oblique space, applying the ${T_MATRIX}$ matrix transformation.")}
```

Here, the axis line and angle annotation have been added by hand
to represent the correlated axes.
However, the vector and the spikes of the $Q$ point
have been computed by applying the transformation given in Equation
\@ref(eq:point-Q-transformed).
This shows how this transformation rotates the angle in the opposite direction
as expected.
That is, the transformation must correspond to a correlation of `r -RHO`
instead of `r RHO`.
On the other hand,
by computing the inverse of this $`r T_MATRIX`$ and pre-multiplying by it
I arrived at the graphical representation in Figure
\@ref(fig:graphical-representation-example-vector-oblique-alt),
which corresponds to Mario's example.

Given the above, my initial guess woud be that these two equations
should actually be

\begin{equation}
  Q \equiv \left[ `r MBL_DEF` + `r MBS_DEF` \right]
    `r DIR_VECTOR_NORM_T`,                              (\#eq:point-Q-corrected)
\end{equation}

(i.e., dropping $- sgn(l_i)$ altogether), as follows from
\@ref(eq:point-Q-coords-transformed)
(see also [this footnote](#fn-sign_l_i)),
and

\begin{equation}
  `r T_INVERSE` Q = \left[ `r MBL_DEF` + `r MBS_DEF` \right]
    `r T_INVERSE` `r DIR_VECTOR_NORM_T`.    (\#eq:point-Q-transformed-corrected)
\end{equation}

(i.e., changing $`r T_MATRIX`$ by $`r T_INVERSE`$).
There is the problem though of whether $`r T_MATRIX`$ is properly defined,
or we are defining in Equations \@ref(eq:coord-transform-horizontal) and
\@ref(eq:change-of-base-matrix) what should actually be called $`r T_MATRIX`$,
as discussed in section \@ref(matrix-squared) above.
If the latter is true, then $`r T_INVERSE`$ should actually be defined
by Equation \@ref(eq:T-matrix-definition).
I think this has yet to be resolved, but in fact the representation according to
Mario's derivations is obtained by pre-multiplying by $`r T_MATRIX`'$
instead of $`r T_INVERSE`$, as we will see later on.

### About the graphical representation example

Regarding the example, [we have already seen in this footnote](#fn-sine_neg)
that, intuitively, it does not seem to be correct:
The graphical representation of the vector should be in the fourth
quadrant, given the signs of the scale parameters.
Also, the oblique axes should have the effect of "shortening" the vector norm
(i.e. length), instead of making it longer.
However, this latter effect is probably due to
being represented in the third quadrant
(because the transformation in the fourth quadrant,
i.e., direction "south-east" or "down-right",
will most likely have this shortening effect).

To my understanding, when we compute the direction given by the angles betweeen
the item vector and the axes, it is important to take into account the
sign of these angles.
Thus, in the example, it should be noted that

\begin{equation}
  \cos \beta = \frac{2}{\sqrt{7}} = `r COS_BETA_VALUE` \Rightarrow
    \beta \approx \pm `r number(BETA_VALUE_DEG, 1e-2)`º,
\end{equation}

i.e., both positive and negative angles with the same absolute value are
valid solutions.
Then, when we make $\sin \beta = `r SIN_BETA_AS_COS`$, we are also disregarding
the sign of $a_{1_2}$ in the computation.
Starting from the definition of $`r DIR_COS_VECTOR`$
in Equations \@ref(eq:multidimensional-params-oblique)
(and noting that these director cosines are given in the orthogonal space)

\begin{equation}
  \sin \beta = \cos \gamma_{i2} = \frac{e' `r PROD_T_A_1`}{\|`r PROD_T_A_1`\|}
\end{equation}

with $e' = (0, \, 1)$, that is

\begin{equation}
    \sin \beta = \frac{(0, \, 1) `r PROD_T_A_1`}{\sqrt{`r A_1_ORTH_SQ_VALUE`}}
\end{equation}

\begin{equation}
  e' `r PROD_T_A_1` = (0, \, 1)
    \begin{pmatrix}
      2 \\
      - \sqrt{3}
    \end{pmatrix} =
      - \sqrt{3},
\end{equation}

\begin{equation}
  \sin \beta = - \sqrt{\frac{3}{7}} = `r SIN_BETA_VALUE_CORR` \Rightarrow
    \beta \approx
    \{
      - `r number(BETA_VALUE_DEG, 1e-2)`º,
      180º + `r number(BETA_VALUE_DEG, 1e-2)`º
    \}=
    \{
      - `r number(BETA_VALUE_DEG, 1e-2)`º,
      `r number(180 + BETA_VALUE_DEG, 1e-2)`º
    \}.
\end{equation}

Therefore, the common (valid) solution is -`r number(BETA_VALUE_DEG, 1e-2)`
and not `r number(BETA_VALUE_DEG, 1e-2)`.
This also converges with the intuitive idea about the direction of the item
I expressed in the first paragraph.

As a result, the coordinates of $P$ and $Q$ would be given by

\begin{equation}
  P' \equiv `r MBL_1_VALUE`
    \left( \tfrac{2}{\sqrt{7}}, \, -\sqrt{\tfrac{3}{7}} \right) =
    (`r A_1_START_COORDS_CORR`)
\end{equation}

and

\begin{equation}
  Q' \equiv \left[ -\frac{`r L_1`}{\sqrt{`r A_1_ORTH_SQ_VALUE`}} -
    `r A_1_ORTH_SQ_VALUE` \right]
    \left(
      \tfrac{2}{\sqrt{`r A_1_ORTH_SQ_VALUE`}},
      - \sqrt{\tfrac{3}{`r A_1_ORTH_SQ_VALUE`}}
    \right) =
    (`r A_1_END_COORDS_CORR`).
\end{equation}

```{r graphical-representation-example-vector-correct, fig.id="graphical-representation-example-vector-correct", fig.cap=glue("Item vector in the orthogonal space, with corrected director angles."), warning=FALSE}
angle <- tibble(init = c(0, -BETA_VALUE)) %>%
  mutate(end = init %>% lead(1)) %>%
  slice(-n()) %>%
  mutate(across(.fns = ~pi / 2 - ., .names = "{.col}_alt")) %>%
  add_column(rad = .5) %>%
  rowwise() %>%
  mutate(
    angle       = (end - init) / pi * 180,
    angle_label = angle %>% number(1e-2, suffix = "º"),
    label_angle = mean(c_across(init:end)),
    label_rad   = rad + .3,
    label_x     = cos(label_angle) * label_rad,
    label_y     = sin(label_angle) * label_rad
  ) %>%
  ungroup()

angle_dir <- angle %>%
  mutate(
    rad       = .3,
    label_rad = .6,
    label_x   = cos(label_angle) * label_rad,
    label_y   = sin(label_angle) * label_rad
  )

points <- tibble(
  coord = c('x', 'y') %>% rep(2),
  value = c(A_1_START_COORDS_CORR, A_1_END_COORDS_CORR),
  point = c('P', 'Q') %>% rep(each = 2)
) %>%
  pivot_wider(id_cols = point, names_from = coord)

guide_segments <- points %>%
  slice(2 %>% rep(2)) %>%
  mutate(across(x:y, .fns =  ~., .names = "{.col}_0")) %>%
  mutate(x_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(0, x_0)) %>% 
  mutate(y_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(y_0, 0))

item_vector <- points %>% pivot_wider(names_from = point, values_from = x:y)

angle_dir %>% ggplot() +
  geom_abline(slope = tan(-BETA_VALUE)) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = rad, start = init_alt, end = end_alt),
    size = ANGLE_WIDTH
  ) +
  geom_text(
    aes(label_x, label_y, label = angle_label),
    family   = GRAPH_FONT
  ) +
  geom_point(aes(x, y), data = points) +
  # geom_text(
  #   aes(x, y, label = point),
  #   data   = points,
  #   position = position_nudge(y = -.2),
  #   family = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    label  = points$point %>% TeX(italic = TRUE),
    x      = points$x,
    y      = points$y - .15,
    family = GRAPH_FONT
  ) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  geom_segment(
    aes(x_P, y_P, xend = x_Q, yend = y_Q, size = I(VECTOR_WIDTH)),
    data     = item_vector,
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  scale_x_continuous(
    limits = c(-1, 2),
    breaks = 0,
    labels = NULL,
    name   = NULL
  ) +
  scale_y_continuous(
    limits = c(-2, 1),
    breaks = 0,
    labels = NULL,
    name   = NULL
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme(
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid.major = element_line(color = "black", size = LINE_WIDTH)
  )
```

Finally, in the oblique plane (but pre-multiplying by $`r T_INVERSE`$ instead of
$`r T_MATRIX`$:

\begin{equation}
  `r T_INVERSE` P = \begin{pmatrix}
    1 & `r T_VALUE[1, 2]` \\
    0 & `r T_VALUE[2, 2]`
  \end{pmatrix}
  \begin{pmatrix}
    `r A_1_START_COORDS_CORR[1]` \\
    `r A_1_START_COORDS_CORR[2]`
  \end{pmatrix} =
  \begin{pmatrix}
    `r A_1_START_COORDS_OBLIQUE_CORR[1]` \\
    `r A_1_START_COORDS_OBLIQUE_CORR[2]`
  \end{pmatrix}
\end{equation}

\begin{equation}
  `r T_INVERSE` Q = \begin{pmatrix}
    1 & `r T_VALUE[1, 2]` \\
    0 & `r T_VALUE[2, 2]`
  \end{pmatrix}
  \begin{pmatrix}
    `r A_1_END_COORDS_CORR[1]` \\
    `r A_1_END_COORDS_CORR[2]`
  \end{pmatrix} =
  \begin{pmatrix}
    `r A_1_END_COORDS_OBLIQUE_CORR[1]` \\
    `r A_1_END_COORDS_OBLIQUE_CORR[2]`
  \end{pmatrix}
\end{equation}

```{r graphical-representation-example-vector-oblique-correct, fig.id="graphical-representation-example-vector-oblique-correct", warning=FALSE, fig.cap=glue("Item vector in the oblique space, with corrected director angles and applying the ${T_INVERSE}$ transformation.")}
points_obl <- tibble(
  coord = c('x', 'y') %>% rep(2),
  value = c(A_1_START_COORDS_OBLIQUE_CORR, A_1_END_COORDS_OBLIQUE_CORR),
  point = c('P', 'Q') %>% rep(each = 2)
) %>%
  pivot_wider(id_cols = point, names_from = coord)# %>%
  # mutate(point_label = glue("{T_INVERSE}{point} = {point}_{{oblique}}"))

guide_segments_obl <- guide_segments %>%
  add_column(name = c("H_spike", "V_spike")) %>%
  pivot_longer(cols = where(is.numeric), names_to = "coord") %>%
  mutate(pt = coord %>% str_detect(., "_0") %>% if_else("Q", "axis")) %>%
  group_by(name, pt) %>%
  mutate(value_obl = drop(T_INV_VAL %*% value)) %>%
  ungroup() %>%
  pivot_wider(id_cols = name, names_from = coord, values_from = value_obl)

item_vector_obl <- points_obl %>%
  # select(-point_label) %>% 
  pivot_wider(names_from = point, values_from = x:y)

ggplot() +
  geom_abline(slope = tan(RHO_ARC_RAD), size = LINE_WIDTH) +
  geom_hline(yintercept = 0, size = LINE_WIDTH) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = .3, start = pi / 2, end = pi / 2 - RHO_ARC_RAD),
    size = ANGLE_WIDTH
  ) +
  geom_text(
    aes(x = .4, y = .2),
    label  = RHO_ARC %>% number(suffix = 'º'),
    family = GRAPH_FONT
  ) +
  geom_point(aes(x, y), data = points_obl) +
  # geom_text(
  #   aes(x, y, label = point_label),
  #   data     = points_obl,
  #   position = position_nudge(y = -.2),
  #   family   = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    # label  = points_obl$point_label %>% TeX(italic = TRUE),
    label  = expression(
      paste(
        bold(paste("T")),
        phantom()^paste("", "-1"),
        italic("P = P", phantom()[paste("oblique")])
      ),
      paste(
        bold(paste("T")),
        phantom()^paste("", "-1"),
        italic("Q = Q", phantom()[paste("oblique")])
      )
    ),
    x      = points_obl$x - .4,
    y      = points_obl$y - .1,
    family = GRAPH_FONT
  ) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments_obl,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  geom_segment(
    aes(x_P, y_P, xend = x_Q, yend = y_Q, size = I(VECTOR_WIDTH)),
    data     = item_vector_obl,
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  scale_x_continuous(
    limits = c(-1, 2),
    breaks = 0,
    labels = NULL,
    name   = NULL,
    oob = oob_keep
  ) +
  scale_y_continuous(
    limits = c(-2, 1),
    breaks = 0,
    labels = NULL,
    name   = NULL,
    oob = oob_keep
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme(
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid.major = element_blank()
  )
```

This representation seems to be closer to what would be expected
from the signs of the parameters.
However, there is still an issue:
Given the values of the parameters,
`r A_1_1` and `r A_1_2` for $a_{1_1}$ and $a_{1_2}$, respectively,
one would expect the projection of the vector on the horizontal
axis to be four times as big in magnitude as
the projection on the oblique axis.
However, both projections seem to be roughly equal.

The problem seems to be due to the computation of the coordinates
in the oblique space in the first place.
The matrix of the inner product of the test space (i.e. the item vectors),
according to Zhang & Stout [-@zhang_theoretical_1999],
should be $`r SIGMA`$. However,
as Equation \@ref(eq:T-T-transposed-matrix-product) shows,
$`r T_MATRIX` `r T_MATRIX`'$ is not equal to $`r SIGMA`$.
In fact, $`r T_MATRIX` `r T_MATRIX`' = `r SIGMA`^{-1}$ ($= `r R_MATRIX`^{-1}$
if we assume that all latent trait dimensions are standardized).
Applying Mario's procedure to compute the coordinates, we have that
$`r MBS_DEF` = `r A_TRANSP_T` `r T_TRANSP_A` = `r A_VECTOR` `r SIGMA`^{-1} `r A_VECTOR`'$.
That is, the matrix of the inner product of the test space would be
$`r SIGMA`^{-1}$ instead of $`r SIGMA`$.

Also, the axis projections in this oblique space
should not be orthogonal but oblique.
Therefore, instead of computing the \beta angle and then using its cosine
and sine to compute the horizontal and vertical coordinates, respectively,
the two director cosines need to be computed as [@harman_modern_1970]
$`r DIR_COS_VECTOR` = `r A_VECTOR` `r SIGMA` / `r MBS`$
(note that each of these director cosines already has its proper sign).
Finally, as these coordinates are computed in an oblique space,
we need to transform them to the orthogonal space for representing them
on a two-dimensional display.

We have previously transformed them by pre-multiplying by $`r T_INVERSE`$.
However, this leads to an incorrect representation,
while pre-multiplying by $`r T_MATRIX`'$ gives the proper representation.
The reason for this is still unclear to me, but I think it has to do with
how $`r T_MATRIX`$ and $`r T_INVERSE`$ have been defined.

```{r graphical-representation-example-vector-oblique-corrected-coordinates, fig.id="graphical-representation-example-vector-oblique-corrected-coordinates", warning=FALSE, fig.cap="Item vector in the oblique space after properly computing the multidimensional discrimination paramater."}
points <- tibble(
  coord     = c('x', 'y') %>% rep(2),
  value     = c(A_1_START_COORDS_ORTH_CORRECT, A_1_END_COORDS_ORTH_CORRECT),
  value_obl = c(A_1_START_COORDS_CORRECT, A_1_END_COORDS_CORRECT),
  point     = c('P', 'Q') %>% rep(each = 2)
) %>%
  pivot_wider(
    id_cols     = point,
    names_from  = coord,
    values_from = starts_with("value")
  ) %>%
  rename_with(str_remove, pattern = "^value_") %>%
  mutate(point_label = glue("{T_MATRIX}'{point} = {point}_{{oblique}}"))

guide_segments <- points %>%
  slice(2 %>% rep(2)) %>%
  select(x, y) %>%
  mutate(across(x:y, .fns =  ~., .names = "{.col}_0")) %>%
  mutate(x_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(0, x_0)) %>% 
  mutate(y_0 = 1:n() %>% mod(2) %>% as.logical() %>% if_else(y_0, 0)) %>%
  add_column(name = c("H_spike", "V_spike"))

## TODO: Not sure why this graphical representation (REVIEW)
guide_segments_orth <- guide_segments %>%
  mutate(
    x_0 = if_else(
      name == "V_spike",
      x_0 - y * tan(asin(RHO)),
      x_0 + y * tan(asin(RHO))
    )
  )

item_vector_obl <- points %>%
  select(point:y) %>% 
  pivot_wider(names_from = point, values_from = x:y)

ggplot() +
  geom_abline(slope = tan(RHO_ARC_RAD), size = LINE_WIDTH) +
  geom_hline(yintercept = 0, size = LINE_WIDTH) +
  geom_arc(
    aes(x0 = 0, y0 = 0, r = .3, start = pi / 2, end = pi / 2 - RHO_ARC_RAD),
    size = ANGLE_WIDTH
  ) +
  geom_text(
    aes(x = .4, y = .2),
    label  = RHO_ARC %>% number(suffix = 'º'),
    family = GRAPH_FONT
  ) +
  geom_point(aes(x, y), data = points) +
  # geom_text(
  #   aes(x, y, label = point_label),
  #   data     = points_obl,
  #   position = position_nudge(y = -.2),
  #   family   = GRAPH_FONT
  # ) +
  annotate(
    geom   = "text",
    label  = points$point_label %>% TeX(italic = TRUE),
    x      = points$x + .2,
    y      = points$y + .1,
    family = GRAPH_FONT
  ) +
  geom_segment(
    aes(x_0, y_0, xend = x, yend = y),
    data     = guide_segments_orth,
    linetype = "longdash",
    size     = LINE_WIDTH
  ) +
  geom_segment(
    aes(x_P, y_P, xend = x_Q, yend = y_Q, size = I(VECTOR_WIDTH)),
    data     = item_vector_obl,
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  scale_x_continuous(
    limits = c(-.65, 1.65),
    breaks = 0,
    labels = NULL,
    name   = NULL,
    oob    = oob_keep
  ) +
  scale_y_continuous(
    # limits = c(-.5, .5),
    breaks = 0,
    labels = NULL,
    name   = NULL,
    oob    = oob_keep
  ) +
  coord_fixed(expand = FALSE, clip = "off") +
  theme(
    axis.line        = element_blank(),
    axis.ticks       = element_blank(),
    panel.grid.major = element_blank()
  )
```

### Complete procedure for the correct graphical representation

The following compares the two procedures for generating
the graphical representation:
The original one, following Mario's derivation,
and the one that gives the correct representation.

The summary of the procedure according to Mario's derivation implies
the following steps:

1. Compute the item discrimination parameters in the orthogonal space
   $`r A_1_VECTOR_ORTH`$ ($= `r T_MATRIX` `r A_1_VECTOR`$)

1. Compute $`r MBS`$
   ($= \sqrt{`r A_1_VECTOR_ORTH`' `r A_1_VECTOR_ORTH`}$)

1. Compute $`r MBL`$ ($\tfrac{-l}{`r MBS`}$)

1. Compute the director cosine $\cos \beta$
   ($= \cos \gamma_{i1} = \tfrac{`r A_1_VECTOR_ORTH`}{`r MBS`}$)
   and its complementary sine $\sin \beta$
   ($= \sqrt{1 - \cos^2 \beta}$)

1. Compute the vector origin coordinates $\mathbf{P}$
   ($= `r MBL` (\cos \beta, \sin \beta)$)

1. Compute the vector end coordinates $\mathbf{Q}$
   ($= \mathbf{P} - sgn (l_i) `r MBS` (\cos \beta, \sin \beta)$)

1. Transform the coordinates $\mathbf{P}$ and $\mathbf{Q}$
   to their equivalent $\mathbf{P}'$ and $\mathbf{Q}'$
   in the oblique space
   ($\mathbf{P}' = `r T_MATRIX` \mathbf{P}; \mathbf{Q}' = `r T_MATRIX` \mathbf{Q}$)

The procedure, according to my corrections, would be as follows:

1. Compute the item discrimination parameters in the orthogonal space
   $`r A_1_VECTOR_ORTH`$ ($= `r T_INVERSE` `r A_1_VECTOR`$)

1. Compute $`r MBS`$
   ($= \sqrt{`r A_1_VECTOR_ORTH`' `r A_1_VECTOR_ORTH`} = \sqrt{`r A_1_VECTOR`' `r T_INVERSE`' `r T_INVERSE` `r A_1_VECTOR`}$)

1. Compute $`r MBL`$ ($\tfrac{-l}{`r MBS`}$)

1. Compute the director cosine vector
   $`r DIR_COS_VECTOR`$ ($= \tfrac{`r A_VECTOR``r SIGMA`}{`r MBS`}$)

1. Compute the origin coordinates $\mathbf{P}$
   ($= `r MBL` `r DIR_COS_VECTOR`$)

1. Compute the end coordinates $\mathbf{Q}$
   ($= \mathbf{P} + `r MBS` `r DIR_COS_VECTOR`$)

1. Transform the coordinates $\mathbf{P}$ and $\mathbf{Q}$ to their equivalent
   $\mathbf{P}'$ and $\mathbf{Q}'$ in the orthogonal space
   ($\mathbf{P}' = `r T_MATRIX`' \mathbf{P}$;
    $\mathbf{Q}' = `r T_MATRIX`' \mathbf{Q}$)

Therefore, the differences can be summed up in the following four:

1. using $`r T_INVERSE`$ instead of $`r T_MATRIX`$ to transform the coordinates,

1. dropping the coefficient $- sgn(l_i)$ from the second term
   in the computation of $\mathbf{Q}$,

1. computing and using the whole director cosine vector $`r DIR_COS_VECTOR`$
   using the original, oblique discrimination parameter $`r A_VECTOR`$
   multiplied by the covariance matrix $`r SIGMA`$, and

1. pre-multiplying the coordinates by $`r T_MATRIX`'$ instead of $`r T_MATRIX`$
   to orthogonalize them.

### Example of graphical representation from my dissertation (Figure 3.1)

Starting from the example in Figure 3.1 in my dissertation,
let's compute and plot the item vectors using the original procedure,
and the one with the corrections.
We need to take into account two things:

- The representation in the orthogonal space does not pose any problem;
  thus, I'll only replicate the example in the oblique space, with $\rho = .5$.

- The parameters in the figure footnote are given in the MUPP-2PL metric,
  so they need to be transformed to the MCLM metric.

The equivalence between the MUPP-2PL and the MCLM parameters
is given elsewhere, e.g. @morillo_dominance_2016.
\@ref(tab:item-params-example-table) gives the parameters in both metrics,
along with the MCLM multidimensional parameters.

```{r set-flextable-wd, cache=FALSE}
# Necessary for officedown to find the template file:
opts_knit$set(root.dir = DOC_DIR)
```

```{r item-params-example-table, tab.id="item-params-example-table", tab.cap="Item parameters for the graphical example"}
MUPP_2PL_params <- tribble(
   ~a_i_1, ~a_i_2, ~l_i, ~dim_i_1, ~dim_i_2,
      2  ,    2  ,  0  ,  1,        2,
      2  ,   -0.5,  1  ,  1,        2,
     -1.5,    1.5, -0.8,  1,        2,
      1.8,    1.3,  1  ,  1,        1,
      1  ,   -1.2, -0.4,  2,        2
)

item_params <- MUPP_2PL_params %>%
  rownames_to_column(var = "item") %>%
  rowwise() %>%
  mutate(
    a_1i         = a_i_1 * (dim_i_1 == 1) - a_i_2 * (dim_i_2 == 1),
    a_2i         = a_i_1 * (dim_i_1 == 2) - a_i_2 * (dim_i_2 == 2),
    l_i_         = l_i,
    a_orth_1i    = T_INV_VAL %*% c_across(a_1i:a_2i) %>% extract(1),
    a_orth_2i    = T_INV_VAL %*% c_across(a_1i:a_2i) %>% extract(2),
    MBS_i        = c_across(a_orth_1i:a_orth_2i) %>%
      { t(.) %*% . } %>%
      drop() %>%
      sqrt(),
    MBL_i        = -l_i_ / MBS_i,
    cos_gamma    = c_across(a_1i:a_2i) %*% CORR_MATRIX / MBS_i,
    cos_gamma_1i = cos_gamma[, 1],
    cos_gamma_2i = cos_gamma[, 2]
  ) %>%
  select(-starts_with("a_orth"), -cos_gamma)

## TODO: Revisar notación de los parámetros (subíndices)
item_params_output <- item_params %>%
  flextable() %>%
  set_header_df(
    tibble(
      col_keys = item_params %>% colnames(),
      metric   = c(
        "Item",
        "MUPP-2PL"    %>% rep(3),
        "Dim."        %>% rep(2),
        "MCLM"        %>% rep(3),
        "Md. params." %>% rep(4)
      ),
      param    = c(
        "Item",
        "a_{i_{1}}", "a_{i_{2}}", "l_i",
        "i_1",       "i_2",
        "a_{1i}",    "a_{2i}",    "l_i",
        "MBS_i",     "MBL_i",     "\\cos \\gamma_{1i}", "\\cos \\gamma_{2i}"
      )
    )
  ) %>%
  mk_par(
    i = 2, j = -1,
    part = "header",
    value = as_paragraph(as_equation(.)),
    use_dot = TRUE
  ) %>%
  merge_v(part = "header") %>%
  merge_h(part = "header", i = 1) %>%
  colformat_double(j = 10:13, digits = 3) %>%
  theme_vanilla() %>%
  add_footer_lines(
    values = paste(
      "Note: Dim. = Dimension;",
      "Md. params. = Multidiemnsional parameters"
    )
  ) %>%
  align(align = "center", part = "header") %>%
  set_table_properties(layout = "autofit")

item_params_output
```

Using these parameters, let's compute the graphical
representation according to Mario's original proposal.
The following R code implements the computation of the coordinates,
with the resulting plot below.
The MCLM parameters $a_{1i}$, $a_{2i}$, and $l_i$
are represented in variables `a_1i`, `a_2i`, and `l_i`, respectively.

```{r item-params-Marios-version, cache=FALSE, echo=TRUE}
item_coords <- item_params %>%
  rowwise() %>%
  mutate(
    a_orth_1i = T_VALUE %*% c_across(a_1i:a_2i) %>% extract(1),
    a_orth_2i = T_VALUE %*% c_across(a_1i:a_2i) %>% extract(2),
    MBS_i     = c_across(a_orth_1i:a_orth_2i) %>%
      { t(.) %*% . } %>%
      drop() %>%
      sqrt(),
    MBL_i     = -l_i_ / MBS_i,
    cos_beta  = a_orth_1i / MBS_i,
    sin_beta  = sqrt(1 - cos_beta^2),
    P_1       = MBL_i * cos_beta,
    P_2       = MBL_i * sin_beta,
    Q_1       = P_1 - sign(l_i) * MBS_i * cos_beta,
    Q_2       = P_2 - sign(l_i) * MBS_i * sin_beta,
    P_1_orth = T_VALUE %*% c_across(P_1:P_2) %>% extract(1),
    P_2_orth = T_VALUE %*% c_across(P_1:P_2) %>% extract(2),
    Q_1_orth = T_VALUE %*% c_across(Q_1:Q_2) %>% extract(1),
    Q_2_orth = T_VALUE %*% c_across(Q_1:Q_2) %>% extract(2)
  )
```

```{r item-params-Marios-version-plot, fig.width=6}
palette <- c("darkred", "darkgoldenrod3", "green3", "cyan3", "blue3")

item_coords %>%
  ggplot(
    aes(
      P_1_orth, P_2_orth,
      xend  = Q_1_orth, yend = Q_2_orth,
      size  = I(VECTOR_WIDTH),
      color = item, fill = item
    ),
  ) +
  geom_segment(
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  geom_abline(slope = tan(RHO_ARC_RAD), size = LINE_WIDTH) +
  geom_abline(
    slope     = tan(RHO_ARC_RAD),
    intercept = -tan(RHO_ARC_RAD) * (-5:4),
    size      = LINE_WIDTH,
    linetype  = "17"
  ) +
  geom_hline(yintercept = 0, size = LINE_WIDTH) +
  scale_x_continuous(
    limits       = c(-3, 3),
    breaks       = (-1:4) - 2 / tan(RHO_ARC_RAD),
    labels       = -1:4,
    minor_breaks = NULL,
    name         = NULL,
    oob          = oob_keep
  ) +
  scale_y_continuous(
    limits       = c(-2, 3.6),
    breaks       = -3:4 * cos(RHO),
    labels       = function(x) x / cos(RHO),
    minor_breaks = 0,
    name         = NULL,
    oob          = oob_keep
  ) +
  scale_color_manual(values = palette) +
  coord_fixed(expand = FALSE, clip = "on") +
  theme(
    axis.line          = element_blank(),
    panel.grid.major.y = element_line(
      color    = "black",
      size     = LINE_WIDTH,
      linetype = "17"
    )
  )
```

As we can see, the item lengths and directions do not correspond to what would
be expected by looking at \@ref(tab:item-params-example-table).
Particularly, item 1 has length 0 due to the fact that $l_1 = 0$ according to
Mario's procedure, and thus $sgn(l_1) = 0$;
therefore, the coefficient $sgn(l_i)$ in Equation \@ref(eq:point-Q)
yields coordinate values of 0 for point $Q$.

Using the corrected procedure I outlined above,
the code for computing the coordinates and resulting plot are:

```{r item-params-corrected-version, cache=FALSE, echo=TRUE}
item_coords <- item_params %>%
  rowwise() %>%
  mutate(
    a_orth_1i    = T_INV_VAL %*% c_across(a_1i:a_2i) %>% extract(1),
    a_orth_2i    = T_INV_VAL %*% c_across(a_1i:a_2i) %>% extract(2),
    MBS_i        = c_across(a_orth_1i:a_orth_2i) %>%
      { t(.) %*% . } %>%
      drop() %>%
      sqrt(),
    MBL_i        = -l_i / MBS_i,
    cos_gamma = c_across(a_1i:a_2i) %*% CORR_MATRIX / MBS_i,
    P_1       = MBL_i * cos_gamma[, 1],
    P_2       = MBL_i * cos_gamma[, 2],
    Q_1       = P_1 + MBS_i * cos_gamma[, 1],
    Q_2       = P_2 + MBS_i * cos_gamma[, 2],
    P_1_orth = t(T_VALUE) %*% c_across(P_1:P_2) %>% extract(1),
    P_2_orth = t(T_VALUE) %*% c_across(P_1:P_2) %>% extract(2),
    Q_1_orth = t(T_VALUE) %*% c_across(Q_1:Q_2) %>% extract(1),
    Q_2_orth = t(T_VALUE) %*% c_across(Q_1:Q_2) %>% extract(2)
  )
```

```{r item-params-corrected-version-plot, fig.width=6}
item_coords %>%
  ggplot(
    aes(
      P_1_orth, P_2_orth,
      xend  = Q_1_orth, yend = Q_2_orth,
      size  = I(VECTOR_WIDTH),
      color = item, fill = item
    ),
  ) +
  geom_segment(
    arrow    = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
    linejoin = "mitre"
  ) +
  geom_abline(slope = tan(RHO_ARC_RAD), size = LINE_WIDTH) +
  geom_abline(
    slope     = tan(RHO_ARC_RAD),
    intercept = -tan(RHO_ARC_RAD) * (-4:4),
    size      = LINE_WIDTH,
    linetype  = "17"
  ) +
  geom_hline(yintercept = 0, size = LINE_WIDTH) +
  scale_x_continuous(
    limits       = c(-3, 3),
    breaks       = (-1:4) - 2 / tan(RHO_ARC_RAD),
    labels       = -1:4,
    minor_breaks = NULL,
    name         = NULL,
    oob          = oob_keep
  ) +
  scale_y_continuous(
    limits       = c(-2, 2.8),
    breaks       = -3:3 * cos(RHO),
    labels       = function(x) x / cos(RHO),
    minor_breaks = 0,
    name         = NULL,
    oob          = oob_keep
  ) +
  scale_color_manual(values = palette) +
  coord_fixed(expand = FALSE, clip = "on") +
  theme(
    axis.line          = element_blank(),
    panel.grid.major.y = element_line(
      color    = "black",
      size     = LINE_WIDTH,
      linetype = "17"
    )
  )
```

The vectors in this representation seem to match the expected lengths and
directions of the vectors given the multidimensional parameters in
\@ref(tab:item-params-example-table).
Also, the representation is coincident with the right panel in Figure 3.1
of my dissertation.

# Next steps

- Clarifying whether $`r T_MATRIX`$ and $`r T_INVERSE`$ are interchanged;
  note that $`r T_MATRIX`$ is defined as the square root matrix of $`r SIGMA`$
  (when all the variables are standardized),
  and therefore this equivalence should hold for that case.
  Remember to see also Zhang & Stout [-@zhang_theoretical_1999, p. 221].

- There is also another pending fact, which is the relationship between
  $`r T_MATRIX`$, $`r SIGMA`$, and $`r R_MATRIX`$.
  If $`r T_MATRIX`$ is defined as the squared root of $`r SIGMA`$,
  then it is an *orthonormalizing* instead of orthogonalizing matrix.
  However, for the graphical representation, one needs to *orthogonalize*
  without standardizing.
  This issue comes from the ambiguity used (by me) between $`r SIGMA`$ and
  $`r T_MATRIX`$, as I (incorrectly) assume that $`r SIGMA`$ is standardized.
  I think this can be resolved by being more strict with the definitions,
  as in section \@ref(matrix-squared).

- These issues are important for clarifying the notation and the graphical
  representation.
  However, for computing the multidimensional parameters,
  I think we should strictly follow Reckase & McKinley's
  [-@reckase_difficulty_1985, -@reckase_discriminating_1991]
  procedure in the first place.

# References

### 2022-05-13 output/paper_draft/paper_M2PLM_parameters.Rmd
Initial commit. Includes first version of the article draft, along with several notebooks and ancillary output documents.

---
title: "A generalized definition of multidimensional item response theory parameters"
author: "Daniel Morillo"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
    mapstyles:
      Normal: First Paragraph
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Includes:

library(knitr)
library(patchwork)

ROOT_DIR <- "../.."
DOC_DIR <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      =   300,
  dev.args = list(type = "cairo-png"),
  fig.pos = "!H"
)

# Output configuration options:
options(digits = 3)
```

```{r write-bib, cache=TRUE}
# automatically create a bib database for R packages
write_bib(
  c(.packages(), 'bookdown', 'knitr', 'rmarkdown', 'lavaan'),
  'www/packages.bib'
)
```

```{r sources, cache=FALSE, include=FALSE}
source("R/Formulae.R",                encoding = 'UTF-8')
```

```{r read-chunks}
read_chunk("src/Graphical_example_paper.R")
```

```{r libraries, message=FALSE}
```

```{r sources, message=FALSE}
```

```{r constants}
```

```{r graphical-output-conf}
```

# Introduction

Multidimensional item response theory (MIRT) models have been widely used since
their inception.<!--# Ref. needed? -->
The so called *Multidimensional 2-Parameter Logistic*
[M2PL, @mckinley_extension_1983] model is one of the most important and
widely used multidimensional models,
partly due to its close relationship to the common factor model
[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
The M2PL model is a generalization of the 2-parameter logistic model
[@birnbaum_latent_1968] to more than one latent trait
[@mckinley_extension_1983].
Its usefulness lies in describing the variability of the population when
the correct response does not depend on a single trait;
more specifically, when a deficit in a certain trait
and a surplus in another one cancel out their effects --thus it is sometimes
referred to as a *compensatory* model.

@reckase_difficulty_1985 argues in favor of a need to characterize
the statistical properties of items taking dimensionality into account.
As such, he outlines a general procedure for determining
the difficulty of multidimensional items.
Following this procedure, he also derives
a *multidimensional difficulty* parameter (MID) for M2PL items.
This parameter consists of two parts:
The first one is a scalar value representing the distance from
the origin of the multidimensional latent space to
the maximally discriminating point of the item.
The second one, the direction of the vector from the origin to that same point.
Following a similar rationale, @reckase_discriminating_1991 define
a *multidimensional discrimination* (MDISC) parameter definition for that model.

A crucial assumption leading to those formulations is that the axes of the
latent space are orthogonal
[see @reckase_difficulty_1985, p. 404 and Equation 7].
Of course, orthogonality is a property of a geometrical space instead of a
latent trait space.
Thus, nothing prevents a researcher from interpreting an abstraction such as
a latent trait space, no matter its structure, as an orthogonal space
--exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
the R package *psych* <!--# Reference needed? -->graphically represent
factor loadings on orthogonal axes, even after an oblique rotation.
Nevertheless, the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
This assumption is not so easily found in the MIRT literature though,
which mostly deals with uncorrelated dimensions
[see @reckase_multidimensional_2009 for a comprehensive treatment on the topic].
This fact is probably due to the different spotlights
the two theoretical programs put the focus on;
Factor analysis aims at a clear interpretation of
the latent covariance structure, and thus focuses on finding
a structure matrix as simple as possible. MIRT literature, on the other hand,
aims at modeling responses to certain stimuli and estimating person scores,
being the model interpretation secondary in relevance.<!--# Reference needed? -->
Nevertheless, the confluence of the two theoretical approaches into a common
modern measurement theory [@mcdonald_test_1999] imposes
the relaxation of traditional constrains on both approaches.
It is not uncommon to find recent examples of MIRT models where the
correlation matrix of the latent dimensions is freely estimated
[see, e.g., @reckase_multidimensional_2009, pp. 224-228]. Despite this, the
correspondence between correlated latent traits and orthogonal coordinate axes
is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
[see @zhang_theoretical_1999 for a notable exception].
What is more, disregarding this geometrical interpretation may lead to
misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->

A latent space where the traits are correlated is more readily interpretable as
a geometrical space with non-orthogonal axes,
often referred to as *general Cartesian coordinates* [@harman_modern_1970].
Such a space has the advantage that the cosines of the angles between two axes
can be interpreted as the correlation of the corresponding latent dimensions.
<!--# Reference needed? -->
<!-- Problem here: The latent space is not defined like that, according to
the results of section "Linear algebra of non-orthogonal coordinates". -->
The general case of MIRT model estimation implies estimating possibly
non-standard latent dimensions and non-null covariances among them,
a more realistic case than an identity matrix.
Assuming orthogonality among traits is usually unrealistic, be it in the
cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
Moreover, certain applications impose the estimation of
non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed?. -->
interpreting the latent dimensions as substantive constructs is not a goal.
As we will argue, the usual definitions of the
MID and MDISC parameters do not generalize to the non-orthogonal geometry used
to represent correlated latent spaces.

In the non-cognitive domain, the use of so-called negatively-keyed items
--items that tap the negative pole of a certain trait, and thus
are usually reverse scored-- is also widespread.<!--# Reference and example? -->
It is common to find instruments that combine
positively- and negatively-keyed items.
Moreover, a complex item may have discrimination parameters with opposite signs
for different latent dimensions.<!--# Examples? -->
Take the multidimensional version of the nominal categories model
[@thissen_nominal_2010] as an example,
and consider two response categories tapping a different construct each:
A positive change in the dimension tapped by one category will inevitably
lead to a decrease in the probability of response to the other category.
One can easily note that a monotonically decreasing probability for a certain
dimension is given by a non-positive discrimination parameter
(and an unchanging probability by a null discrimination parameter),
contrary to the constrains imposed by @reckase_difficulty_1985 [p. 411].
However, the multidimensional item parameters can also be defined for
items with monotonically non-increasing probabilities, as we will see.

The main objective of this paper is thus to provide a generalization of
the MIRT parameters in two senses: on one hand, in an oblique coordinate system
that can be soundly applied to the correlated trait case; on the other, to
items that yield monotonous response probabilities, but in the broader sense of
unchanging monotony. Additionally, we aim at providing a method for
graphically representing these items in the fashion of @ackerman_graphical_2015,
but taking into account the non-orthogonality of the coordinate axes.
In order to pursue these objectives, we first define the generalization of
the M2PL model and describe its scope of application.
Then, we follow the method outlined by @reckase_difficulty_1985 and
@reckase_discriminating_1991 to obtain
the formulation of the multidimensional parameters.
We will introduce two versions of these indices,
showing that either the original or one of the new ones is obtained
as a function of how we define and interpret the latent space.
Next we discuss the properties of the indices,
propose a graphical representation method,
and present an example of computation<!-- # TODO: Pending -->
and representation of the items based on these parameters.
We round up with a discussion about
the applicability and generalization of our results.

# Method

## Model formulation

Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$,
defined by vector space $V$ and basis $`r LS_BASIS`$,
being $`r LS_BASIS_EQ`$ an arbitrary set of linearly independent vectors.
Person $j$ is represented in $`r LATENT_SPACE`$ by vector
$`r TRAIT_VC_PERSON_J`$,
being $`r TRAIT_VECTOR`$ a multivariate normal variable

\begin{equation}
  `r TRAIT_MV_DEF`,
\end{equation}

where $`r MEAN_VECTOR`$ is an $n$-element vector of means and
$`r COV_MATRIX` = `r SD_MATRIX` `r CORR_MATRIX` `r SD_MATRIX`$
an $n$-dimensional positive definite covariance matrix,
with $`r CORR_MATRIX`$ the corresponding correlation matrix
(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
with $`r VAR_COV_EQ`$ (i.e., the variances).

As in @reckase_difficulty_1985, the characteristic function of item
$`r ITEM_INDEX`$ according to the M2PL model is formulated as follows:

\begin{equation}
  `r M2PL_FORMULATION`,                                  (\#eq:M2PL-formulation)
\end{equation}

being  
$`r IRF_M2PL`$ ($= `r IRF_ABBR`$ henceforth, for simplicity) the probability of
a positive response to item $`r ITEM_INDEX`$ by person $`r PERSON_INDEX`$,  
$`r DISCR_VECTOR`$ a $`r N_DIMS`$-dimensional vector of
discrimination parameters, and  
$`r INTERCEPT_PARAM`$ an intercept parameter related to the location of the item
in the latent space.

We must make a few remarks here. First, as we are not constraining ourselves to
cognitive traits, we speak of _positive_ instead of _correct response_.
This may imply a correct response, if we are talking about a cognitive test, or
endorsing an item (or a certain response option) in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in Reckse
[-@reckase_difficulty_1985, a decrease in the probability of a correct response with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is *inverse* for (i.e., taps the negative pole of)
latent dimension $`r DIM_INDEX`$.
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
instead of the item difficulty.

Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
$`r LATENT_SPACE`$. As in @reckase_difficulty_1985, we need to
reparameterize Equation \@ref(eq:M2PL-formulation) in polar coordinates,
because the slope is dependent on the direction considered.
However, Reckase's Equation 3 implicitly assumes orthogonality.
In order to contemplate the general case,
we need to introduce a few linear algebra results before proceeding futher.

## Linear algebra of non-orthogonal coordinates
<!-- # TODO: To appendix, OR citation about these? -->

### Inner product

Let $`r INNER_PROD_EQ`$ be the inner product in $`r LATENT_SPACE`$, with
the positive-definite matrix $`r INNER_PROD_MATRIX`$ its inner product matrix.
We may apply a change of basis in $`r LATENT_SPACE`$ that results in a space
$`r LATENT_SPACE_STD`$ with the standard basis $`r LS_STD_BASIS_EQ`$.
Such a change of basis can be defined as

\begin{equation}
  `r BASIS_CHANGE`,                                          (\#eq:change-basis)
\end{equation}

with $`r TRAIT_VECTOR_STD`$ the coordinates
of $`r TRAIT_VECTOR`$ in $`r LATENT_SPACE_STD`$ and $`r TRANSFORM_MATRIX_EQ`$
(being $`r BASIS_VECTOR_ANY_TRANSF`$ the coordinates of
$`r BASIS_VECTOR_ANY`$ in $`r LATENT_SPACE_STD`$
for any dimension $`r DIM_INDEX`$).
This new space can be defined as a Euclidean space with standard inner product
(i.e. dot product), such that the norm $`r NORM_EQ`$ of $`r TRAIT_VECTOR`$
in $`r LATENT_SPACE_STD`$ is given by<!-- # TODO: Reference needed? -->

\begin{equation}
  `r TRAIT_MODULE`^2 = `r TRAIT_VECTOR_STD_SQ` =
    `r TRANSF_TRAIT_SQUARED` =
    `r TRAIT_TRANSF_TRAIT`.                                 (\#eq:inner-product)
\end{equation}

Thus, the inner product matrix $`r INNER_PROD_MATRIX`$ in $`r LATENT_SPACE`$
is consistently defined as

\begin{equation}
  `r INNER_PROD_MAT_EQ`.                            (\#eq:inner-prod-matrix-def)
\end{equation}

### Direction cosines

The cosines between two vectors follow the usual definition,
<!-- # TODO: Is this true?? See Harman 1970 (eq. 4.43, p. 60 to check) -->

\begin{equation}
  `r COS_VECTORS_EQ`,                                         (\#eq:cos-vectors)
\end{equation}

independently of whether the inner product is standard or not.
Therefore, to compute the direction cosine $`r DIR_COS_ANY`$
of vector $`r TRAIT_VECTOR`$ along dimension $`r DIM_INDEX`$,
we plug it into Equation \@ref(eq:cos-vectors) along with
the corresponding basis vector $`r BASIS_VECTOR_ANY`$:

\begin{equation}
  `r DIR_COSINE_EQ`.                                     (\#eq:direction-cosine)
\end{equation}

By Equation \@ref(eq:inner-prod-matrix-def) it is clear that
$`r DIAG_INNER_PROD_MATRIX_EQ`$
(but note that the basis vectors are not necessarily unitary).
By defining $`r DIAG_MATRIX_INNER_PROD`$ as
the diagonal matrix with the same diagonal as $`r INNER_PROD_MATRIX`$,
Equation \@ref(eq:direction-cosine) in matrix notation is<!-- # TODO:
Not sure if it’s necessary to explain this further.
There may be a certain “inferential leap” assuming that D^(-1/2) is
the matrix form equivalent to 1/u_k,
(but I “know” intuitively that the generalization holds) -->

\begin{equation}
  `r DIR_COS_VEC_EQ`.                             (\#eq:direction-cosine-vector)
\end{equation}

## Model recasting into polar coordinates

From Equation \@ref(eq:direction-cosine-vector),
any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as

\begin{equation}
  `r TRAIT_VECTOR_POLAR_EQ`.                           (\#eq:trait-vector-polar)
\end{equation}

Thus, the M2PL model is expressed in polar coordinates as

\begin{equation}
  `r IRF_POLAR_EQ`.                                    (\#eq:M2PL-formula-polar)
\end{equation}

## Point of maximum slope

Following the procedure outlined by @reckase_difficulty_1985
[@reckase_discriminating_1991],
we first compute the point of maximum slope by finding the root(s) of
the second derivative with respect to $`r TRAIT_J_MODULE`$.

\begin{equation}
  `r IRF_2ND_DIFF_EQ`,                                      (\#eq:M2PL-2nd-diff)
\end{equation}

The root we are interested in is found for $`r IRF_MAX_SLOPE_EQ`$
(the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
which result in improper values of $`r TRAIT_J_MODULE`$).
The slope in direction $`r DIR_ANGLE_J_VEC`$ is given by

\begin{equation}
  `r IRF_1ST_DIFF_EQ`.                                      (\#eq:M2PL-1st-diff)
\end{equation}

When $`r IRF_MAX_SLOPE_EQ`$,

\begin{equation}
  `r SLOPE_MAX_EQ`.                                             (\#eq:slope-max)
\end{equation}

Notation aside, these equations differ from
Reckase's [-@reckase_difficulty_1985] Equations 3 through 6
only in the term $`r INNER_PROD_INV_DIAG_SQ_PROD`$.
When $`r LS_BASIS`$ is the standard basis,
this term simplifies to an identity matrix
(which results in Equation \@ref(eq:M2PL-formula-polar) through
\@ref(eq:slope-max) simplyfing to rectangular Cartesian coordinates,
i.e. the equations outlined by Reckase).
To compute the direction of the maximum slope,
Reckase leverages on the property that the squared direction cosines sum up to 1
(see Equation 7 therein).
However, this assertion does not apply to the general case,
as it implicitly assumes orthogonal coordinates.<!-- # TODO: To appendix the following, or maybe give for granted? -->
Instead, we may consider the direction cosine vector in $`r LATENT_SPACE_STD`$:

\begin{equation}
  `r DIR_COS_VEC_STD_EQ`.                     (\#eq:direction-cosine-vector-std)
\end{equation}

Operating on Equation \@ref(eq:trait-vector-polar),

\begin{equation}
  \begin{split}
    `r DIR_COS_AS_DIR_COS_STD_EQ` \\
    `r TRAIT_POLAR_COEFF_COS_STD_EQ`
  \end{split}                                 (\#eq:direction-cosine-vectors-eq)
\end{equation}

and, substituting in \@ref(eq:slope-max),

\begin{equation}
  `r SLOPE_MAX_STD_COSINES_EQ`.                     (\#eq:slope-max-std-cosines)
\end{equation}

We can define

\begin{equation}
  `r DISCR_VECTOR_STD_EQ`.                               (\#eq:discr-vector-std)
\end{equation}

such that Equation \@ref(eq:slope-max-std-cosines) is similar
(in vectorial form) to Reckase's [-@reckase_difficulty_1985] Equation 6,
and thus can be solved by the same procedure
applying their Equations 7 through 9.
The resulting equivalent to their Equation 9, in vectorial form
and solving for $`r DIR_ANGLE_STD_SUBSTITUTE`$ is

\begin{equation}
  `r DIR_COS_STD_ITEM_VEC_EQ`.                            (\#eq:dir-cos-vec-std)
\end{equation}

Substituting $`r DIR_COS_STD_ITEM_VEC`$ and $`r DISCR_VECTOR_STD`$
by their respective expressions results in the expression for
the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,

\begin{equation}
  `r DIR_COS_ITEM_VEC_EQ`,                                    (\#eq:dir-cos-vec)
\end{equation}

which gives the direction from the origin to the point where
the slope is maximum.
Finally, to determine the signed distance $`r DISTANCE_PARAM`$
from the origin to that point,
we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$,
$`r DIR_COS_SUBSTITUTE`$ and
$`r DISTANCE_SUBSTITUTE`$,
and then substitute Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$.
The result, solving for $`r DISTANCE_PARAM`$, is

\begin{equation}
  `r DISTANCE_PARAM_EQ`,                                   (\#eq:distance-param)
\end{equation}

The slope $`r SLOPE_MAX_PARAM`$ at the point defined by Equations
\@ref(eq:dir-cos-vec) and \@ref(eq:distance-param)
is given by Equation \@ref(eq:slope-max).
Solving for $`r DIR_COS_SUBSTITUTE`$
and substituting Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$
[@reckase_discriminating_1991],

\begin{equation}
  `r SLOPE_MAX_PARAM_EQ`.                                     (\#eq:slope-param)
\end{equation}

# Results

## Test space and latent space

Before defining the multidimensional parameters, we need to make a clarification
regarding the representation of the items in the latent space.
Equation \@ref(eq:slope-param) makes it clear that a space $`r TEST_SPACE`$
with inner product matrix $`r INNER_PROD_MATRIX_INV`$ can be defined,
such that $`r INNER_PROD_DISCR_EQ`$ and, subsequently,
$`r DISCR_VECTOR_MODULE_EQ`$.
We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of the
discrimination vectors.
This space is where the item vectors can be properly represented.
It is in general different from the latent space $`r LATENT_SPACE`$,
although there is a bijective relationship between the two sets of spaces,
as their inner product matrices are inverse of one another.

The basis $`r ITEM_SPACE_BASIS_EQ`$ of $`r TEST_SPACE`$ is such that
$`r ITEM_TRANSF_MATRIX_EQ`$ (see Equation \@ref(eq:discr-vector-std)).
Thus, similarly to equation \@ref(eq:dir-cos-vec),
the vector of direction cosines of a vector in $`r TEST_SPACE`$
can be defined as

\begin{equation}
  `r DIR_COS_ITEM_VEC_TS_EQ`,                      (\#eq:dir-cos-vec-test-space)
\end{equation}

with $`r DIAG_MATRIX_INNER_PROD_INV`$ the diagonal matrix with
the same diagonal as $`r INNER_PROD_MATRIX_INV`$.

## Generalized multidimensional parameters

The item location $`r MIL_PARAM`$ is defined as the distance and direction
from the origin to the point of maximum slope [@reckase_difficulty_1985].
By analogy with the unidimensional case, $`r MDISC_PARAM`$ is defined as
$4 `r SLOPE_MAX_PARAM`$ [@reckase_discriminating_1991].
Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_PARAM`$ using
equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
However, as they depend on the inner product matrix $`r INNER_PROD_MATRIX`$,
the choice we make of this matrix will lead us to different definitions of
the parameters.

Up to this point,
we have derived the maximum slope and its location in the latent space
$`r LATENT_SPACE`$ independently of its basis $`r LS_BASIS`$.
If we assume $`r BASIS_EQ`$ or, more generally,
that $`r LS_BASIS`$ is orthonormal, $`r TRANSFORM_MATRIX`$ will be
also orthonormal and thus $`r INNER_PROD_MAT_STD_EQ`$.
In such a case the multidimensional item location and discrimination simplify
to the expressions derived by @reckase_difficulty_1985
and @reckase_discriminating_1991.
These expresssions disregard the covariance structure of $`r TRAIT_VECTOR`$ by
implicitly assuming the orthonormality of $`r LS_BASIS`$.
Therefore, we shall refer to these as the *agnostic* version of the indices:

\begin{equation}
  \begin{split}
    `r MDISC_AG_PARAM_EQ` \\
    `r MIL_AG_PARAM_EQ`.
  \end{split}
\end{equation}

On the other hand, we can account for the covariance structure of
the latent space letting $`r LATENT_SPACE`$ be its geometrical representation.
Thus, we must define $`r LS_BASIS`$ in such a way that

\begin{equation}
  `r TRAIT_MV_STD_EQ`,
\end{equation}

where $`r COV_MATRIX_STD`$ is a diagonal matrix.
We may consider now two different possibilities:
We may define $`r TRANSFORM_MATRIX`$ such that $`r INNER_PROD_CORR_COND`$.
This results in a *correlation-based* version of the indices
(note that, in this case, $`r DIAG_MATRIX_INNER_PROD_INV`$ is
an identity matrix):

\begin{equation}
  \begin{split}
    `r MDISC_CORR_PARAM_EQ` \\
    `r MIL_CORR_PARAM_EQ`.
  \end{split}
\end{equation}

However, we may also define $`r TRANSFORM_MATRIX`$ such that
$`r INNER_PROD_COV_COND`$, so

\begin{equation}
  `r TRAIT_MV_STD_NORM_EQ`.
\end{equation}

In this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in the
following *covariance-based* version of the indices:

\begin{equation}
  \begin{split}
    `r MDISC_COV_PARAM_EQ` \\
    `r MIL_COV_PARAM_EQ`.
  \end{split}
\end{equation}

## Properties of the parameters

There are three conditions to meet for regarding these parameters as
a valid generalization of the (unidimensional) IRT parameters
[@reckase_discriminating_1991]:

1. If an item measures only dimension $`r DIM_INDEX`$,
   then $`r MDISC_UNIDIM_EQ`$.

2. The distance $`r DISTANCE_PARAM`$ has the same relationship with the
   intercept as $b_i$ in the unidimensional case,
   i.e. $`r DIST_INTERCEPT_REL`$.

3. $`r MDISC_PARAM`$ is four times the maximum slope $`r SLOPE_MAX_PARAM`$.

The second property derives straightforwardly from
Equation \@ref(eq:distance-param),
and the third one is implicit in the definition of the $`r MDISC_PARAM`$.
Simple arithmetic can show that the first holds true whenever the
$`r DIAG_KTH_ELEMENT`$-th element of $`r INNER_PROD_MATRIX_INV`$ equals 1.
Therefore, both the agnostic and the correlation-based version fulfill it.
$`r MDISC_COV_PARAM`$ does not though,
as $`r COV_ELEMENT`$ is not generally equal to 1.
Nevertheless, it can be interpreted as a *scaled version* of the index,
with the standard deviation of the corresponding dimension as scaling factor.
This transformation may be made in the unidimensional case as well,
thus finding a *standard-metric* discrimination parameter
(i.e. referred to a unitary variance space).

The (signed) distance to the origin,
given by the second element of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_PARAM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.

The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
also due to the strictly positiveness of $`r MDISC_PARAM`$,
are equal to the signs of $`r DISCR_VECTOR`$.
This property leads directly to generalizing the indices to the monotonically
non-increasing case: The measurement direction relative to dimension
$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
When $`r IRF_ABBR`$ is constant with respect to
variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
However, note that this does not necessarily imply that the vector is
parallel to any other axis or strictly contained in the hyperplane formed
by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
orthogonality assumption, but will depend on the correlation matrix
$`r CORR_MATRIX`$ for the other two versions of the index.

## Graphical representation

The aforementioned properties are easier to apprehend with
a visual representation of the items.
A vector plot [@ackerman_creating_1994] is the most appropriate way of
representing the items graphically;
it allows one to plot several items altogether and analyze them
in terms of their multidimensional parameters [@ackerman_graphical_1996]:
An item vector will be applied at a direction and (signed) distance
from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
and will have a length equal to its $`r MDISC_PARAM`$ parameter.

As plotting more than two dimensions is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we will limit ourselves to the bidimensional case here.
Plotting on a two-dimensional display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we will need to compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
However, this implies pre-multiplying by $`r TRANSFORM_MATRIX`$,
instead of $`r TRANSFORM_MATRIX_TRANSP_INV`$ as would be expected from
Equation \@ref(eq:discr-vector-std).
This is because computing the equivalent rectangular coordinates
of the oblique ones is different from computing the item parameters
transformed to an orthogonal space.
While the latter implies rotating the axes to an orthogonal configuration and
computing the new coordinates in these axes,
the former implies transforming the space altogether along with the
axes transformation.

Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
that allows us to represent the test space structure
and the items by transforming their coordinates to
a rectangular Cartesian system.
However, we want to avoid changing the scaling of the parameters
in order to facilitate the interpretation,
and thus we will prefer the correlation-based over the covariance-based indices.
Thus, we define $`r TRANSFORM_MATRIX`$ such that
$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.

The general procedure for plotting an item $`r ITEM_INDEX`$ comprises
the following steps:

1. Define $`r TRANSFORM_MATRIX`$ such that $`r TRANSF_MATRIX_SQ_CORR_INV_EQ`$.

1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
   respectively.

1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.

1. Compute the end coordinates $`r END_ITEM_EQ`$.

1. Compute the rectangular coordinates
   $`r ORIGIN_ITEM_TRANSF`$ and $`r END_ITEM_TRANSF`$ by pre-multiplying
   $`r ORIGIN_ITEM`$ and $`r END_ITEM`$, respectively, by
   the transformation matrix $`r TRANSFORM_MATRIX`$
   ($`r ORIGIN_ITEM_TRANSF_EQ`$; $`r END_ITEM_TRANSF_EQ`$).

### Graphical representation example

```{r compute-example-items}
```

The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
In the first case, the two latent dimensions are independent,
with $`r TRANSFORM_MATRIX`$ being an identity matrix.
In the second one, the correlation between the dimensions is $`r corr_out`$.
In order to keep the horizontal axis invariant,
we define $`r TRANSFORM_MATRIX`$ such that it only rotates the vertical axis
[@harman_modern_1970]:

\begin{equation}
  `r TRANSFORM_MATRIX_EXAMPLE_EQ`.
\end{equation}

The sample items have the same M2PL parameters in both latent spaces.
However, their multidimensional parameters differ due to the correlation among
the latent dimensions.
The parameters in both spaces,
along with their corresponding M2PL parameters,
are shown in Table \@ref(tab:example-items-table-out).

```{r compose-example-items-table}
```

```{r set-flextable-wd, cache=FALSE}
# Necessary for officedown to find the template file:
opts_knit$set(root.dir = DOC_DIR)
```

```{r example-items-table-out, cache=FALSE, tab.id="example-items-table-out", tab.cap="Item parameters for the graphical example"}
item_params_output
```

<br>

```{r compose-oblique-plot}
```

```{r compose-orthogonal-plot}
```

```{r item-plot-out, cache=FALSE, fig.cap="Item plots in an orthogonal (left) and oblique (right) test space.", fig.height=3.1, fig.width=6.65}
plot_orth + plot_oblique
```

Figure \@ref(fig:item-plot-out) illustrates the effect of
a positive correlation on the $`r MDISC_CORR_PARAM`$
(and consequently the $`r MIL_CORR_PARAM`$) value:
the components of the discrimination parameters tend to
*sum up* in the same direction as they get *more aligned*
with a positive increase in the correlation.
Thus, if their discrimination parameters have the same sign, like items 1 and 5,
their $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
on the contrary, when the discrimination parameters have opposite sign
they tend to cancel each other out, as happens with items 2 and 3.
This effect can be especially noticed by comparing
the $`r MDISC_CORR_PARAM_ITEM`$ of items 3 and 5:
Their discrimination parameters are equal in absolute value,
so their $`r MDISC_CORR_PARAM_ITEM`$ values are equal in the orthogonal space;
however, the effect of the correlation shrinks the former and
stretches the latter.
Finally, note that item 4 has one discrimination parameter equal to 0,
so its $`r MDISC_CORR_PARAM`$ is unaffected by the correlation.

We can also see how the sign of $`r INTERCEPT_PARAM`$ affects the item distance:
The item is applied at the origin when its value is null (item 1),
and tends to be displaced
*against* the item direction when its value is positive (items 2 and 4) and
*towards* the item direction when it is negative (items 3 and 5).
Finally, we can see how the item direction is determined by the discrimination
parameters *and* the correlation; the sign of the discrimination parameters
determine the quadrant where the item will point at;
thus, an item with two positive (negative) discrimination parameters will be
in the first (third) quadrant,
while an item with opposite-sign discrimination parameters will be in the
second or fourth quadrant.
The direction relative to the axes will be given by the relative absolute value
of the two parameters, but also by the correlation:
Items 2 and 3 especially illustrate this effect,
as each of them is orthogonal to one of the axes in the oblique space,
even when they are not parallel to the other axis.

## Example with empirical data

<!-- # TODO: Complete? -->

# Discussion

@reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
general enough to "be used with any model that yields probabilities that
increase monotonically with an increase in ability on any dimension" (p. 411).
However, we have discussed that this definition implicitly assumes
the orthonormality of the latent space.
Applying the general procedure outlined by @reckase_difficulty_1985 and later
extended by @reckase_discriminating_1991
we have obtained a set of multidimensional parameters
that generalize the original results in two aspects:
(1) to a non-orthonormal space that accounts for the covariance structure of
the multivariate latent variable, and (2) to any case of unchanging monotony.
It is worth noting that @zhang_theoretical_1999 already proposed a formulation
of the $`r MDISC_PARAM`$ equivalent to our covariance-based version,
but for a more general, semi-parametric formulation
(although constrained to non-decreasing monotony).
However, they provided no formal proof for this result.
Our results provide this formal derivation, although only for the specific
case of the M2PL model.

As we have seen, taking the latent space structure into account may have a
substantial effect in the resulting values and interpretation of the parameters,
something not contemplated by the original
(hereby referred to as agnostic) version.
Moreover, in applications where the latent space structure is critical
(e.g. multi-group IRT, equating, and linking),
the agnostic version may overlook relevant differences in
the values of the multidimensional parameters.
Thus, the covariance structure of the latent space should be taken into account
when computing the multidimensional parameters of a MIRT model.

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based index has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
Once again, this is important in applications where
potentially different latent spaces are involved.

Given these considerations, we propose to adopt the covariance-based version of
the multidimensional parameters, and refer to them simply as
the $`r MDISC_PARAM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of preserving the metric of the parameters,
which is convenient for fidelity when representing the items.
Therefore, we recommend using the correlation-based version of the parameters
for graphical representation purposes only.

One may consider that the formal derivations presented here are overly
complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the indices in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 april 2015).
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.

Strictly following @reckase_difficulty_1985's method in the original,
oblique space has also provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, in order to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although models such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the M2PL model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying measurement theory, its implications and its interpretation.

\newpage

# References

### 2022-05-13 R/Formulae.R

@@ -1,0 +1,902 @@
+# ==============================================================================
+# 
+# FILE NAME:   Formulas.R
+# DESCRIPTION: Constants for facilitating the writing of LaTeX equations
+# 
+# AUTHOR:      Mori (danivmorillo@gmail.com)
+# 
+# DATE:        22/01/2021
+# 
+# ==============================================================================
+
+
+## ---- GLOBAL OPTIONS: --------------------------------------------------------
+
+
+
+## ---- INCLUDES: --------------------------------------------------------------
+
+# p_load(tidyverse, magrittr)
+
+source("R/LaTeX_math.R")
+
+## ---- CONSTANTS: -------------------------------------------------------------
+
+### Mathematical constants ----
+
+FRAC_1_2 <- latex_frac(1, 2, .sep = NO_SEP)
+FRAC_1_4 <- latex_frac(1, 4, .sep = NO_SEP)
+
+### Latent space and variable definition ----
+
+# Latent space definition:
+LATENT_SPACE       <- latex_bf("\\upTheta")
+N_DIMS             <- latex('n')
+BASIS_VECTOR_MOD   <- latex('u')
+BASIS_VECTOR       <- latex_bf(BASIS_VECTOR_MOD)
+BASIS_VECTOR_FIRST <- latex("$BASIS_VECTOR$_1")
+BASIS_VECTOR_LAST  <- latex("$BASIS_VECTOR$_$N_DIMS$")
+LS_BASIS           <- latex_cal("B")
+LS_BASIS_ELEMENTS  <- latex("$BASIS_VECTOR_FIRST$, ... , $BASIS_VECTOR_LAST$")
+LS_BASIS_SET       <- latex_curlybraces(LS_BASIS_ELEMENTS)
[…]

### 2022-05-13 R/Formulas.R

@@ -1,0 +1,440 @@
+# ==============================================================================
+# 
+# FILE NAME:   Formulas.R
+# DESCRIPTION: Constants for facilitating the writing of LaTeX equations
+# 
+# AUTHOR:      Mori (danivmorillo@gmail.com)
+# 
+# DATE:        22/01/2021
+# 
+# ==============================================================================
+
+
+## ---- GLOBAL OPTIONS: --------------------------------------------------------
+
+
+
+## ---- INCLUDES: --------------------------------------------------------------
+
+# p_load(tidyverse, magrittr)
+
+
+source("R/LaTeX_math.R")
+
+
+## ---- CONSTANTS: -------------------------------------------------------------
+
+### Mathematical constants ----
+
+FRAC_1_2 <- latex_frac(1, 2)
+FRAC_1_4 <- latex_frac(1, 4)
+
+
+### MUPP-2PL model maximal slope ----
+
+#### Rectangular coordinates ----
+
+F_THETA_RECT <- "f_2(\\theta_1, \\theta_2)"
+
+EXPONENT_RECT <- "(a_1 \\theta_1 - a_2 \\theta_2 + l)"
+LOGISTIC_RECT <- latex_logistic("$EXPONENT_RECT$")
+
+
+#### Polar coordinates ----
+
+ALPHA <- "\\alpha"
+
+COS_ALPHA <- latex("\\cos $ALPHA$")
+SIN_ALPHA <- latex("\\sin $ALPHA$")
+
+COS_SQ_ALPHA <- latex("\\cos^2 $ALPHA$")
+
+THETA_COS_ALPHA <- latex("\\theta $COS_ALPHA$")
+THETA_SIN_ALPHA <- latex("\\theta $SIN_ALPHA$")
+
+A1_COS_ALPHA <- latex("a_1 $COS_ALPHA$")
[…]

### 2022-05-13 R/Graphical_example_transcriptions_Mario.R

@@ -1,0 +1,84 @@
+# ==============================================================================
+# 
+# FILE NAME:   Graphical_example_transcriptions_Mario.R
+# DESCRIPTION: Constants for using with Mario's graphical representation example
+# 
+# AUTHOR:      Mori (danivmorillo@gmail.com)
+# 
+# DATE:        19/02/2021
+# 
+# ==============================================================================
+
+
+## ---- GLOBAL OPTIONS: --------------------------------------------------------
+
+
+
+## ---- INCLUDES: --------------------------------------------------------------
+
+
+
+## ---- CONSTANTS: -------------------------------------------------------------
+
+A_1_1 <-  2
+A_1_2 <- 0
+L_1   <-  1
+RHO   <-   .5
+
+RHO_ARC_RAD <- acos(RHO)
+RHO_ARC     <- RHO_ARC_RAD / pi * 180
+
+A_1_VALUE <- as.matrix(c(A_1_1, A_1_2))
+T_VALUE   <- matrix(c(1, 0, -RHO/sqrt(1 - RHO^2), 1/sqrt(1 - RHO^2)), nrow = 2)
+T_INV_VAL <- solve(T_VALUE)
+
+
+A_1_ORTH_TRANSP_VALUE <- t(A_1_VALUE) %*% T_VALUE
+
+T_T_TRANSP_VALUE <- T_VALUE %*% t(T_VALUE)
+
+A_1_ORTH_SQ_VALUE <- drop(A_1_ORTH_TRANSP_VALUE %*% t(A_1_ORTH_TRANSP_VALUE))
+
+MBS_1_VALUE <- sqrt(A_1_ORTH_SQ_VALUE)
+MBL_1_VALUE <- - L_1 / MBS_1_VALUE
+
+COS_BETA_VALUE <- A_1_ORTH_TRANSP_VALUE %*% c(1, 0) / MBS_1_VALUE
+SIN_BETA_VALUE <- sqrt(1 - COS_BETA_VALUE^2)
+BETA_VALUE     <- acos(COS_BETA_VALUE)
+BETA_VALUE_DEG <- BETA_VALUE / pi * 180
+
[…]

### 2022-05-13 R/LaTeX_math.R

@@ -1,0 +1,207 @@
+# ==============================================================================
+# 
+# FILE NAME:   LaTeX_math.R
+# DESCRIPTION: Functions for outputting LaTeX equations
+# 
+# AUTHOR:      Mori (danivmorillo@gmail.com)
+# 
+# DATE:        22/01/2021
+# 
+# ==============================================================================
+
+
+## ---- GLOBAL OPTIONS: --------------------------------------------------------
+
+## ---- INCLUDES: --------------------------------------------------------------
+
+library(glue)
+
+
+# source("R/<File_name>.R")
+
+
+## ---- CONSTANTS: -------------------------------------------------------------
+
+DELIMITER  <- '$'
+SEP        <- ' '
+EQ_SIGN    <- ' = '
+EQUIV_SIGN <- ' \\equiv '
+DEF_SIGN   <- ' \\coloneq'
+SIM_SIGN   <- ' \\sim '
+NO_SEP     <- ''
+SEP_COMMA  <- ', '
+UNDERSCORE <- '_'
+ELLIPSIS   <- '...'
+
+TRANSPOSED_SUFFIX <- '^T'
+PRIME_SUFFIX      <- "'"
+
+INNER_PROD_LEFT  <- "\\langle"
+INNER_PROD_RIGHT <- "\\rangle"
+INNER_PROD_SEP   <- ",\\,"
+
+LATEX_CLASS <- "laTeR"
+
+
+## ---- FUNCTIONS: -------------------------------------------------------------
+
+### Engine: ----
+
+latex <- function(..., open = DELIMITER, close = DELIMITER, .sep = SEP) {
+  
+  result <- glue(
+    ...,
+    .open  = open,
+    .close = close,
+    .envir = parent.frame(),
+    .sep   = .sep
+  )
+  
+  class(result) <- c(LATEX_CLASS, class(result))
+  
+  result
+}
+
+
+### Notation: ----
+
+latex_bf  <- function(...) latex("\\mathbf{",  latex(...), "}", .sep = NO_SEP)
+latex_cal <- function(...) latex("\\mathcal{", latex(...), "}", .sep = NO_SEP)
+latex_rm  <- function(...) latex("\\mathrm{",  latex(...), "}", .sep = NO_SEP)
+
+latex_prime <- function(x) latex(x, PRIME_SUFFIX, .sep = NO_SEP)
[…]

### 2022-05-13 R/Output.R

@@ -1,0 +1,61 @@
+# ==============================================================================
+# 
+# FILE NAME:   Output.R
+# DESCRIPTION: Functions for output formatting
+# 
+# AUTHOR:      Mori (danivmorillo@gmail.com)
+# 
+# DATE:        03/06/2021
+# 
+# ==============================================================================
+
+
+## ---- GLOBAL OPTIONS: --------------------------------------------------------
+
+
+## ---- INCLUDES: --------------------------------------------------------------
+
+library(officer)
+library(assertive)
+
+# source("R/<file>")
+
+
+## ---- CONSTANTS: -------------------------------------------------------------
+
+### Mathematical constants ----
+
+
+## ---- FUNCTIONS: -------------------------------------------------------------
+
+insert_anchor <- function(anchor_id, bg_color = "white") {
+  
+  assert_is_a_string(anchor_id)
+  
+  ftext(
+    paste0("{#", anchor_id, "}"), prop = fp_text(color = bg_color))
+}
+
+format_prop_like <- function(values, sig = 3, drop_0 = TRUE) {
+  
+  ## Constants: ----
+  LEADING_ZERO_PATTERN <- "0(?=\\.)"
+  
+  ## Argument checking and formatting: ----
+  
+  assert_is_numeric(values)
+  assert_all_are_in_closed_range(values, -1L, 1L)
+  assert_is_scalar(sig)
+  assert_all_are_whole_numbers(sig)
+  assert_is_a_bool(drop_0)
+  
+  
+  ## Main: ----
+  
+  result <- values %>% round(digits = sig) %>%
+    format(digits = sig, nsmall = sig)
+  
+  if (drop_0) result <- result %>% str_remove(LEADING_ZERO_PATTERN)
+  
+  result
+}

### 2022-05-13 src/Graphical_example_paper.R

@@ -1,0 +1,292 @@
+# ==============================================================================
+# 
+# FILE NAME:   Graphical_example_paper.R
+# DESCRIPTION: Script for computing and plotting the graphical representation
+#              example in the paper.
+# 
+# AUTHOR:      Mori (danivmorillo@gmail.com)
+# 
+# DATE:        18/02/2022
+# 
+# ==============================================================================
+
+
+## ---- GLOBAL OPTIONS: --------------------------------------------------------
+
+## ---- INCLUDES: --------------------------------------------------------------
+
+## ----libraries----
+library(tidyverse)
+library(flextable)
+library(scales)
+
+## ----sources----
+source("R/Output.R")
+source("R/Formulae.R")
+
+## ---- CONSTANTS: -------------------------------------------------------------
+
+## ----constants----
+
+# Graphical representation parameters:
+GRAPH_FONT   <- "serif"
+LINE_WIDTH   <- .1
+VECTOR_WIDTH <- .3
+# POINT_CHAR   <- 19L
+PALETTE      <- c("darkred", "darkgoldenrod3", "green3", "cyan3", "blue3")
+
+# Item parameters:
+items_MCLM <- tribble(
+  ~a_1, ~a_2, ~l  ,
+   2  ,  2  ,  0  ,
+   2  , -1  ,  1  ,
+  -0.5,  1  , -0.5,
+   0  ,  1.2,  0.5,
+  -0.5, -1  , -0.5
+)
+
+# Oblique latent space correlation:
+CORR_VALUE   <- .5
+CORR_ARC     <- acos(CORR_VALUE)
+CORR_ARC_SIN <- sin(CORR_ARC)
+CORR_ARC_TAN <- tan(CORR_ARC)
+
+corr_out <- latex_eq(CORR, CORR_VALUE |> format_prop_like(sig = 1))
+
+
+## ---- CONFIGURATION: ---------------------------------------------------------
+
+## ----graphical-output-conf----
+theme_set( # `ggplot` output configuration
+  theme_classic(
+    base_size      = 12,
+    base_family    = GRAPH_FONT,
+    base_line_size = LINE_WIDTH
+  ) %+replace%
+    theme(
+      axis.title.x = element_text(hjust = 1),
+      axis.title.y = element_text(vjust = 1, angle = 0)
+    )
+)
+
+## ---- MAIN: ------------------------------------------------------------------
+
+## ----compute-example-items----
+
[…]

### 2022-05-13 output/paper_draft/paper_M2PLM_parameters.Rmd
Expressions clarified after reviewing and updating mind map

@@
 
-A latent space where the traits are correlated is more readily interpretable as
+A latent space where the traits are correlated is easier to represent in
 a geometrical space with non-orthogonal axes,

### 2022-05-20 output/paper_draft/paper_M2PLM_parameters.Rmd
lavaan package is not being used so it's not written to the bibliography

--- 2022-05-13 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 write_bib(
-  c(.packages(), 'bookdown', 'knitr', 'rmarkdown', 'lavaan'),
+  c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
   'www/packages.bib'

### 2022-05-31 output/paper_draft/paper_M2PLM_parameters.Rmd
Minor corrections after review, up to "Properties"

@@
 latent space are orthogonal
-[see @reckase_difficulty_1985, p. 404 and Equation 7].
+[see @reckase_difficulty_1985, p.
+404 and Equation 7].<!--# It may be worth mentioning that the orthogonality assumption is also implicit in the expression of the M2PL model in polar coordinates (p.
+403 and Eq.
+3 in Reckase 1985) -->
 Of course, orthogonality is a property of a geometrical space instead of a
@@
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed?. -->
+non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
 interpreting the latent dimensions as substantive constructs is not a goal.
@@
 dimension is given by a non-positive discrimination parameter
-(and an unchanging probability by a null discrimination parameter),
+(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
 contrary to the constrains imposed by @reckase_difficulty_1985 [p. 411].
@@
 The main objective of this paper is thus to provide a generalization of
-the MIRT parameters in two senses: on one hand, in an oblique coordinate system
+the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
 that can be soundly applied to the correlated trait case; on the other, to
@@
 (see Equation 7 therein).
-However, this assertion does not apply to the general case,
-as it implicitly assumes orthogonal coordinates.<!-- # TODO: To appendix the following, or maybe give for granted? -->
+However, this assertion implicitly assumes orthogonal coordinates,
+so it does not apply in general to `r DIR_COS_VEC`.<!-- # TODO: To appendix the following, or maybe give for granted? -->
 Instead, we may consider the direction cosine vector in $`r LATENT_SPACE_STD`$:
@@
 The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
-also due to the strictly positiveness of $`r MDISC_PARAM`$,
+also due to the strict positivity of $`r MDISC_PARAM`$,
 are equal to the signs of $`r DISCR_VECTOR`$.
@@
 by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
-orthogonality assumption, but will depend on the correlation matrix
+orthogonality assumption, but it will depend on the correlation matrix
 $`r CORR_MATRIX`$ for the other two versions of the index.

### 2022-06-10 output/paper_draft/paper_M2PLM_parameters.Rmd
Minor changes after creating mind map of the draft structure

@@
 latent space are orthogonal
-[see @reckase_difficulty_1985, p.
-404 and Equation 7].<!--# It may be worth mentioning that the orthogonality assumption is also implicit in the expression of the M2PL model in polar coordinates (p.
-403 and Eq.
-3 in Reckase 1985) -->
+[see @reckase_difficulty_1985, p. 404 and Equation 7].
 Of course, orthogonality is a property of a geometrical space instead of a
@@
 unchanging monotony. Additionally, we aim at providing a method for
-graphically representing these items in the fashion of @ackerman_graphical_2015,
+graphically representing these items in the fashion of @ackerman_creating_1994,
 but taking into account the non-orthogonality of the coordinate axes.
@@
 \end{equation}
+
+As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
+involved in the expression of the model in polar coordinates.
+Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
+already assumes the orthogonality of the latent space.
 
@@
 (and consequently the $`r MIL_CORR_PARAM`$) value:
-the components of the discrimination parameters tend to
-*sum up* in the same direction as they get *more aligned*
-with a positive increase in the correlation.
-Thus, if their discrimination parameters have the same sign, like items 1 and 5,
-their $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
-on the contrary, when the discrimination parameters have opposite sign
-they tend to cancel each other out, as happens with items 2 and 3.
+The components of the discrimination parameters tend to
+*sum up* in the same direction as they get *more aligned*.
+Thus, if the discrimination parameters have the same sign, as in items 1 and 5,
+the $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
+on the contrary, when the discrimination parameters have opposite signs
+they tend to cancel each other out and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
+as happens with items 2 and 3.
 This effect can be especially noticed by comparing
@@
 Finally, note that item 4 has one discrimination parameter equal to 0,
-so its $`r MDISC_CORR_PARAM`$ is unaffected by the correlation.
+so its $`r MDISC_CORR_PARAM_ITEM`$ is unaffected by the correlation.
 
-We can also see how the sign of $`r INTERCEPT_PARAM`$ affects the item distance:
-The item is applied at the origin when its value is null (item 1),
-and tends to be displaced
+We can also see how the sign of $`r INTERCEPT_PARAM`$ affects
+the distance to the origin:
+The item vector is applied at the origin when its value is null (item 1),
+and tends to be shifted
 *against* the item direction when its value is positive (items 2 and 4) and
 *towards* the item direction when it is negative (items 3 and 5).
-Finally, we can see how the item direction is determined by the discrimination
+Finally, we can see how the direction is determined by the discrimination
 parameters *and* the correlation; the sign of the discrimination parameters
-determine the quadrant where the item will point at;
-thus, an item with two positive (negative) discrimination parameters will be
-in the first (third) quadrant,
+determine where the item will point at;
+thus, an item with two positive (negative) discrimination parameters will point
+at the first (third) quadrant,
 while an item with opposite-sign discrimination parameters will be in the
@@
 potentially different latent spaces are involved.
-
 Given these considerations, we propose to adopt the covariance-based version of
@@
 with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
-then compute the indices in the transformed space using
+then compute the parameters in the transformed space using
 @reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
@@
 which may give relevant information if these are actual, substantive traits.
-
-Strictly following @reckase_difficulty_1985's method in the original,
+Moreover, strictly following @reckase_difficulty_1985's method in the original,
 oblique space has also provided us with an important insight:

### 2022-07-19 output/paper_draft/paper_M2PLM_parameters.Rmd
Includes empirical example table in paper draft

@@
 source("R/Formulae.R",                encoding = 'UTF-8')
-```
-
-```{r read-chunks}
-read_chunk("src/Graphical_example_paper.R")
-```
-
-```{r libraries, message=FALSE}
-```
-
-```{r sources, message=FALSE}
-```
-
-```{r constants}
-```
-
-```{r graphical-output-conf}
 ```
@@
 
+```{r read-chunks-graphical-example}
+read_chunk("src/Graphical_example_paper.R")
+```
+
+```{r libraries, message=FALSE}
+```
+
+```{r sources, message=FALSE}
+```
+
+```{r constants}
+```
+
+```{r graphical-output-conf}
+```
+
 ```{r compute-example-items}
@@
 ## Example with empirical data
+
+```{r reset-root-wd}
+# Necessary for knitr to find the source file:
+opts_knit$set(root.dir = ROOT_DIR)
+```
+
+```{r read-chunks-empirical-example}
+read_chunk("src/Empirical_example.R")
+```
+
+```{r libraries, message=FALSE}
+```
+
+```{r sources, message=FALSE}
+```
+
+```{r read-items, message=FALSE}
+```
+
+```{r orthogonal-params}
+```
+
+```{r oblique-params}
+```
+
+```{r params-table}
+```
+
+```{r compose-empirical-items-table}
+```
+
+```{r set-flextable-wd, cache=FALSE}
+```
+
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r empirical-items-table-out, cache=FALSE, tab.id="empirical-items-table-out", tab.cap="Item parameters for the empirical example"}
+item_params_output
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
 

### 2022-07-19 src/Empirical_example.R

@@ -17,6 +17,7 @@
 ## ----libraries----
 library(tidyverse)
 library(flextable)
+library(assertive.numbers)
 
 ## ----sources----
 source("R/Mirt_toolbox.R")
@@ -38,11 +39,10 @@ source("R/Mirt_toolbox.R")
 ## ----read-items----
 
 # Item parameters:
-items <- read_csv2("dat/Table_5_Tezza_et_al_2018.csv", )
+items <- read_csv2("dat/Table_5_Tezza_et_al_2018.csv")
 
 ## ----orthogonal-params----
-items_orth <- items_mirt_params |>
-  compute_mirt_params(d, starts_with('a'), dir_out = "deg")
+items_orth <- items |> compute_mirt_params(d, starts_with('a'), dir_out = "deg")
 
 ## ----compare-orthogonal-params----
 
@@ -88,7 +88,7 @@ item_params <- item_params |> select(
   starts_with('deg_' |> paste0(1:4))                 # direction angles
 )
 
-## ----compose-example-items-table----
+## ----compose-empirical-items-table----
 
 # Add blank columns to format table (separate different types of parameters)
 item_params <- item_params                   |>
@@ -98,33 +98,42 @@ item_params <- item_params                   |>
 
 # Table header (for flextable object):
 item_headers <- tibble(
-  col_keys = item_params %>% colnames(),
+  col_keys = item_params |> colnames(),
+  metric   = col_keys %>% {
+    case_when(
+      . == "item"            ~ "Item",
+      str_detect(., "a\\d+") ~ "M2PL",
+      . == 'l'               ~ "M2PL",
+      str_detect(., "sep1")  ~ ' ',
+      TRUE                   ~ "Multidimensional parameters"
+    )
+  },
   param    = col_keys %>% {
     case_when(
       . == "item"              ~ "Item",
-      str_detect(., "a\\d+")   ~ "Discrimination",
-      . == 'l'                 ~ "l",
+      str_detect(., "a\\d+")   ~ "a_i",
+      . == 'l'                 ~ "l_i",
       str_detect(., "sep\\d+") ~ ' ',
-      str_detect(., "MDISC_")  ~ "MDISC",
-      str_detect(., "D_")      ~ "D",
-      str_detect(., "deg_")    ~ "\\alpha",
+      str_detect(., "MDISC_")  ~ "MDISC_i",
+      str_detect(., "D_")      ~ "D_i",
[…]

### 2022-07-19 output/paper_draft/paper_M2PLM_parameters.Rmd
Drops person index "j" from the whole notation

@@
 being $`r LS_BASIS_EQ`$ an arbitrary set of linearly independent vectors.
-Person $j$ is represented in $`r LATENT_SPACE`$ by vector
-$`r TRAIT_VC_PERSON_J`$,
-being $`r TRAIT_VECTOR`$ a multivariate normal variable
+A respondent is represented in $`r LATENT_SPACE`$ by
+coordinate vector $`r TRAIT_VECTOR`$, assumed to be multivariate normal:
 
@@
 $`r IRF_M2PL`$ ($= `r IRF_ABBR`$ henceforth, for simplicity) the probability of
-a positive response to item $`r ITEM_INDEX`$ by person $`r PERSON_INDEX`$,  
+a positive response to item $`r ITEM_INDEX`$ for a given $`r TRAIT_VECTOR`$,  
 $`r DISCR_VECTOR`$ a $`r N_DIMS`$-dimensional vector of
@@
 we first compute the point of maximum slope by finding the root(s) of
-the second derivative with respect to $`r TRAIT_J_MODULE`$.
+the second derivative with respect to $`r TRAIT_VECTOR`$.
 
@@
 (the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
-which result in improper values of $`r TRAIT_J_MODULE`$).
-The slope in direction $`r DIR_ANGLE_J_VEC`$ is given by
+which result in improper values of $`r TRAIT_MODULE`$).
+The slope in direction $`r DIR_ANGLE_VEC`$ is given by
 

### 2022-07-27 output/paper_draft/paper_M2PLM_parameters.Rmd
Terminology correction, due to the ambiguity in the use of "vector" both as a geometric vector and an algebraic vector (i.e. column matrix) of the coordinates of such vector in a vector space.

@@
 being $`r LS_BASIS_EQ`$ an arbitrary set of linearly independent vectors.
-A respondent is represented in $`r LATENT_SPACE`$ by
-coordinate vector $`r TRAIT_VECTOR`$, assumed to be multivariate normal:
+A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
+with coordinates $`r TRAIT_VECTOR`$ in $`r LATENT_SPACE`$,
+which is assumed to be a multivariate normal random variable:
 

### 2022-07-27 R/Formulae.R

@@ -32,8 +32,8 @@ FRAC_1_4 <- latex_frac(1, 4, .sep = NO_SEP)
 # Latent space definition:
 LATENT_SPACE       <- latex_bf("\\upTheta")
 N_DIMS             <- latex('n')
-BASIS_VECTOR_MOD   <- latex('u')
-BASIS_VECTOR       <- latex_bf(BASIS_VECTOR_MOD)
+BASIS_VECTOR_EL    <- latex('u')
+BASIS_VECTOR       <- latex_bf(BASIS_VECTOR_EL)
 BASIS_VECTOR_FIRST <- latex("$BASIS_VECTOR$_1")
 BASIS_VECTOR_LAST  <- latex("$BASIS_VECTOR$_$N_DIMS$")
 LS_BASIS           <- latex_cal("B")
@@ -42,11 +42,12 @@ LS_BASIS_SET       <- latex_curlybraces(LS_BASIS_ELEMENTS)
 LS_BASIS_EQ        <- latex_eq(LS_BASIS, LS_BASIS_SET)
 
 # Latent vector definition:
-TRAIT_VECTOR      <- latex_bf("\\uptheta")
-MEAN_VECTOR       <- latex_bf("\\upmu")
-COV_MATRIX        <- latex_bf("\\upSigma")
-NORMAL_DISTR      <- latex_cal('N')
-MV_DISTRIBUTION   <- latex(
+TRAIT_VECTOR    <- latex_bf("\\uptheta")
+TRAIT_VEC_IN_LS <- latex_in(TRAIT_VECTOR, LATENT_SPACE)
+MEAN_VECTOR     <- latex_bf("\\upmu")
+COV_MATRIX      <- latex_bf("\\upSigma")
+NORMAL_DISTR    <- latex_cal('N')
+MV_DISTRIBUTION <- latex(
   NORMAL_DISTR,
   latex_parentheses("$MEAN_VECTOR$, $COV_MATRIX$")
 )
@@ -92,7 +93,8 @@ M2PL_FORMULATION <- latex_eq(IRF_M2PL, LOGISTIC_M2PL)
 
 ### Change of basis in the latent space ----
 
-STD_BASIS_VECTOR      <- latex_bf('e')
+STD_BASIS_VECTOR_EL   <- latex('e')
[…]

### 2022-07-27 R/LaTeX_math.R

@@ -26,8 +26,9 @@ DELIMITER  <- '$'
 SEP        <- ' '
 EQ_SIGN    <- ' = '
 EQUIV_SIGN <- ' \\equiv '
-DEF_SIGN   <- ' \\coloneq'
+DEF_SIGN   <- ' \\coloneq '
 SIM_SIGN   <- ' \\sim '
+IN_SIGN    <- ' \\in '
 NO_SEP     <- ''
 SEP_COMMA  <- ', '
 UNDERSCORE <- '_'
@@ -196,6 +197,11 @@ latex_innerprod <- function(x, y) {
   latex(INNER_PROD_LEFT, x, INNER_PROD_SEP, y, INNER_PROD_RIGHT)
 }
 
+
+
+### Sets: ----
+
+latex_in <- function(element, set) latex(element, set, .sep = IN_SIGN)
 
 
 ### Enclosing functions: ----

### 2022-07-27 notebooks/Space_transformation_theorem.qmd
Cosine director lemmas, initial version [WIP]

---
title: "Space transformation in Multidimensional IRT (Theorem)"
author: "Mario Luzardo, Daniel Morillo"
format:
  docx:
    number-sections: true
    highlight-style: github  
  html: default
callout-collapse: true
editor_options: 
  chunk_output_type: console
---

# Introduction

Transcription of Mario's notes

```{r setup, include=FALSE}
# Includes:

library(knitr)

ROOT_DIR <- ".."
DOC_DIR <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
# opts_chunk$set(
#   echo     = FALSE,
#   cache    = FALSE,
#   dev      = "png",
#   dpi      =   300,
#   dev.args = list(type = "cairo-png"),
#   fig.pos = "!H"
# )
```

```{r sources, cache=FALSE, include=FALSE}
source("R/Formulae.R", encoding = 'UTF-8')
```

# Main

Let's consider a $`r N_DIMS`$ latent space $`r LATENT_SPACE`$ with Euclidean
vector structure, and let $`r LS_BASIS_EQ`$ be a basis of $`r LATENT_SPACE`$.

::: callout-note
Stating that $`r LS_BASIS`$ is a basis implies that its vectors are linearly
independent and are a generator of $`r LATENT_SPACE`$; i.e., any vector
$`r TRAIT_VEC_IN_LS`$ can be expressed as a linear combination of
$`r LS_BASIS`$.
:::

Let's consider a dichotomous item $`r ITEM_INDEX`$ such that the probability
that respondent with trait $`r LS_BASIS_EQ`$ gives a correct[^1] response is
given by

$$
  `r M2PL_FORMULATION`
$$ {#eq-M2PL-formulation}

Let's consider in $`r LATENT_SPACE`$ the standard basis $`r LS_STD_BASIS_EQ`$ as
well.

For all $`r TRAIT_VEC_IN_LS`$ we have both representations[^2]:

$`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and $`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$

Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix.

[^1]: I prefer to refer to the responses as "positive/negative", instead of
    "correct/incorrect", as this terminology contemplates the possibility of
    non-cognitive as well as cognitive domains.

[^2]: I'm preferring here the notation I use in the draft, that is,
    $`r DIM_INDEX`$ for indexing the dimensions (instead of $`r ITEM_INDEX`$,
    which is also used to index the items), $`r TRAIT_MODULE`$ for the
    coordinates of $`r TRAIT_VECTOR`$ in $`r LS_BASIS`$,
    and $`r TRAIT_VEC_STD_EL`$ for the coordinates in $`r LS_STD_BASIS`$.
    Nevertheless, as Mario reminded me, it is worth noting that this notation
    for the vector coordinates is ambiguous, as we are using the same symbol
    ($`r TRAIT_VECTOR`$) for the "geometric vector" and
    the "algebraic vector" (i.e. column matrix) of its coordinates in
    $`r LATENT_SPACE`$ (this notation is usually accepted though).

### 2022-07-27 R/LaTeX_math.R

@@ -41,6 +41,8 @@ INNER_PROD_LEFT  <- "\\langle"
 INNER_PROD_RIGHT <- "\\rangle"
 INNER_PROD_SEP   <- ",\\,"
 
+SUMMATION_OP <- "\\sum"
+
 LATEX_CLASS <- "laTeR"
 
 
@@ -107,6 +109,31 @@ latex_sign <- function(..., .par = NA) {
   if (.par) argument <- latex_parentheses(argument)
   
   latex(latex_rm("sign"), argument)
+}
+
+### Iterated operations: ----
+
+latex_summation <- function(index, ..., from = 1, to = NULL, .par = NA) {
+  
+  if (is.na(.par)) .par <- length(list(...)) != 1
+  
+  if (is.null(to)) {
+    
+    sub <- latex_in(index, from)
+    
+  } else {
+    
+    sub <- latex_eq(index, from)
+  }
+  
+  operator <- latex_sub(SUMMATION_OP, sub)
+  
+  if (!is.null(to)) operator <- latex_raised_to(operator, exp = to)
+  
+  result <- latex(...)
+  if (.par) result <- latex_parentheses(result)
+  
+  latex(operator, result)
 }
 
 ### Powers and roots: ----

### 2022-07-27 R/Formulae.R
Cosine director lemmas, initial version [WIP]

@@ -633,6 +633,34 @@ TRANSFORM_MATRIX_EXAMPLE_EQ <- latex_def(
   TRANSFORM_MATRIX_EXAMPLE
 )
 
+### Mario's Theorem: ----
+
+TRAIT_VEC_STD_MOD    <- latex_prime(TRAIT_MODULE)
+TRAIT_VEC_STD_COMP   <- latex_sub(TRAIT_VEC_STD_MOD, DIM_INDEX)
+STD_BASIS_VECTOR_ANY <- latex_sub(STD_BASIS_VECTOR, DIM_INDEX)
+
+TRAIT_VECTOR_DEF_BASIS        <- latex_summation(
+  TRAIT_COMPONENT, BASIS_VECTOR_ANY,
+  index = DIM_INDEX, from = 1, to = N_DIMS,
+  .par = FALSE
+)
+TRAIT_VECTOR_DEF_BASIS_EQ     <- latex_eq(TRAIT_VECTOR, TRAIT_VECTOR_DEF_BASIS)
+TRAIT_VECTOR_DEF_STD_BASIS    <- latex_summation(
+  TRAIT_VEC_STD_COMP, BASIS_VECTOR_ANY_TRANSF,
+  index = DIM_INDEX, from = 1, to = N_DIMS,
+  .par = FALSE
+)
+TRAIT_VECTOR_DEF_STD_BASIS_EQ <- latex_eq(
+  TRAIT_VECTOR,
+  TRAIT_VECTOR_DEF_STD_BASIS
+)
+
+BASIS_CHANGE_SYM <- latex(
+  "{}_{$LS_STD_BASIS$}",
+  latex_sub(latex_parentheses('I'), LS_BASIS)
+)
+BASIS_CHANGE_DEF <- latex_eq(BASIS_CHANGE_SYM, TRANSFORM_MATRIX)
+
 ## ---- FUNCTIONS: -------------------------------------------------------------
 
 ### <Section name>: ----

### 2022-07-27 R/Formulae.R
Cosine director lemmas, initial version [WIP]

@@ -630,6 +630,34 @@ TRANSFORM_MATRIX_EXAMPLE_EQ <- latex_def(
   TRANSFORM_MATRIX_EXAMPLE
 )
 
+### Mario's Theorem: ----
+
+TRAIT_VEC_STD_MOD    <- latex_prime(TRAIT_MODULE)
+TRAIT_VEC_STD_COMP   <- latex_sub(TRAIT_VEC_STD_MOD, DIM_INDEX)
+STD_BASIS_VECTOR_ANY <- latex_sub(STD_BASIS_VECTOR, DIM_INDEX)
+
+TRAIT_VECTOR_DEF_BASIS        <- latex_summation(
+  TRAIT_COMPONENT, BASIS_VECTOR_ANY,
+  index = DIM_INDEX, from = 1, to = N_DIMS,
+  .par = FALSE
+)
+TRAIT_VECTOR_DEF_BASIS_EQ     <- latex_eq(TRAIT_VECTOR, TRAIT_VECTOR_DEF_BASIS)
+TRAIT_VECTOR_DEF_STD_BASIS    <- latex_summation(
+  TRAIT_VEC_STD_COMP, BASIS_VECTOR_ANY_TRANSF,
+  index = DIM_INDEX, from = 1, to = N_DIMS,
+  .par = FALSE
+)
+TRAIT_VECTOR_DEF_STD_BASIS_EQ <- latex_eq(
+  TRAIT_VECTOR,
+  TRAIT_VECTOR_DEF_STD_BASIS
+)
+
+BASIS_CHANGE_SYM <- latex(
+  "{}_{$LS_STD_BASIS$}",
+  latex_sub(latex_parentheses('I'), LS_BASIS)
+)
+BASIS_CHANGE_DEF <- latex_eq(BASIS_CHANGE_SYM, TRANSFORM_MATRIX)
+
 ## ---- FUNCTIONS: -------------------------------------------------------------
 
 ### <Section name>: ----

### 2022-07-27 R/Formulae.R
Cosine director lemmas continued [WIP]

@@ -658,6 +658,34 @@ BASIS_CHANGE_SYM <- latex(
 )
 BASIS_CHANGE_DEF <- latex_eq(BASIS_CHANGE_SYM, TRANSFORM_MATRIX)
 
+TRAIT_COORDS_DEF <- latex_eq(
+  TRAIT_VECTOR,
+  latex(
+    "coord" |> latex_rm() |> latex_sub(LS_BASIS, .abbr = TRUE),
+    latex_parentheses(TRAIT_VECTOR, .sep = NO_SEP)
+  )
+)
+
+TRAIT_STD_COORDS_DEF <- latex_eq(
+  TRAIT_VECTOR_STD,
+  latex(
+    "coord" |> latex_rm() |> latex_sub(LS_STD_BASIS, .abbr = TRUE),
+    latex_parentheses(TRAIT_VECTOR, .sep = NO_SEP)
+  )
+)
+
+BASIS_CHANGE_EQ <- latex_eq(TRAIT_VECTOR_STD, TRAIT_TRANSFORM)
+
+BASIS_VEC_STD_COORDS_DEF <- latex_eq(
+  BASIS_VECTOR_ANY_TRANSF,
+  latex(
+    "coord" |> latex_rm() |> latex_sub(LS_STD_BASIS, .abbr = TRUE),
+    latex_parentheses(BASIS_VECTOR_ANY, .sep = NO_SEP)
+  )
+)
+
+INNER_PROD_BASIS <- latex_innerprod('\\,', '') |> latex_sub(LS_BASIS)
+
 ## ---- FUNCTIONS: -------------------------------------------------------------
 
 ### <Section name>: ----

### 2022-07-27 notebooks/Space_transformation_theorem.qmd
Cosine director lemmas continued [WIP]

@@
 Let's consider a dichotomous item $`r ITEM_INDEX`$ such that the probability
-that respondent with trait $`r LS_BASIS_EQ`$ gives a correct[^1] response is
+that respondent with trait $`r LS_BASIS_EQ`$ gives a positive[^1] response is
 given by
@@
 
-Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix.
+Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix; then, if we denote
+$`r TRAIT_COORDS_DEF`$ and $`r TRAIT_STD_COORDS_DEF`$ we have that
+
+$$
+  `r BASIS_CHANGE_EQ`
+$$ {#eq-change-basis}
+
+The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$. If
+$`r BASIS_VEC_STD_COORDS_DEF`$ then[^3]
+
+$$
+  `r TRANSFORM_MATRIX_EQ`
+$$ {#eq-transform-matrix-def}
+
+In $`r LATENT_SPACE`$ with the standard basis we have the usual scalar product.
+In $`r LATENT_SPACE`$ with basis $`r LS_BASIS`$ we define an inner product
+$`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant.
 
@@
     "correct/incorrect", as this terminology contemplates the possibility of
-    non-cognitive as well as cognitive domains.
+    non-cognitive (as well as cognitive) domains.
 
@@
     which is also used to index the items), $`r TRAIT_MODULE`$ for the
-    coordinates of $`r TRAIT_VECTOR`$ in $`r LS_BASIS`$,
-    and $`r TRAIT_VEC_STD_EL`$ for the coordinates in $`r LS_STD_BASIS`$.
+    coordinates of $`r TRAIT_VECTOR`$ in $`r LS_BASIS`$, and
+    $`r TRAIT_VEC_STD_COMP`$ for the coordinates in $`r LS_STD_BASIS`$.
     Nevertheless, as Mario reminded me, it is worth noting that this notation
     for the vector coordinates is ambiguous, as we are using the same symbol
-    ($`r TRAIT_VECTOR`$) for the "geometric vector" and
-    the "algebraic vector" (i.e. column matrix) of its coordinates in
-    $`r LATENT_SPACE`$ (this notation is usually accepted though).
+    ($`r TRAIT_VECTOR`$) for the "geometric vector" and the "algebraic vector"
+    (i.e. column matrix) of its coordinates in $`r LATENT_SPACE`$ (this notation
+    is usually accepted though).
+
+[^3]: It is ok to use square brackets here? I'm using it instead of parentheses
+    because I had already written this (with a function), but if it's not
+    correct I change it to parentheses.

### 2022-07-27 R/Formulae.R
Cosine director lemmas continued [WIP]

@@ -661,6 +661,34 @@ BASIS_CHANGE_SYM <- latex(
 )
 BASIS_CHANGE_DEF <- latex_eq(BASIS_CHANGE_SYM, TRANSFORM_MATRIX)
 
+TRAIT_COORDS_DEF <- latex_eq(
+  TRAIT_VECTOR,
+  latex(
+    "coord" |> latex_rm() |> latex_sub(LS_BASIS, .abbr = TRUE),
+    latex_parentheses(TRAIT_VECTOR, .sep = NO_SEP)
+  )
+)
+
+TRAIT_STD_COORDS_DEF <- latex_eq(
+  TRAIT_VECTOR_STD,
+  latex(
+    "coord" |> latex_rm() |> latex_sub(LS_STD_BASIS, .abbr = TRUE),
+    latex_parentheses(TRAIT_VECTOR, .sep = NO_SEP)
+  )
+)
+
+BASIS_CHANGE_EQ <- latex_eq(TRAIT_VECTOR_STD, TRAIT_TRANSFORM)
+
+BASIS_VEC_STD_COORDS_DEF <- latex_eq(
+  BASIS_VECTOR_ANY_TRANSF,
+  latex(
+    "coord" |> latex_rm() |> latex_sub(LS_STD_BASIS, .abbr = TRUE),
+    latex_parentheses(BASIS_VECTOR_ANY, .sep = NO_SEP)
+  )
+)
+
+INNER_PROD_BASIS <- latex_innerprod('\\,', '') |> latex_sub(LS_BASIS)
+
 ## ---- FUNCTIONS: -------------------------------------------------------------
 
 ### <Section name>: ----

### 2022-08-05 R/LaTeX_math.R
Direction cosines complete

@@ -22,17 +22,20 @@ library(glue)
 
 ## ---- CONSTANTS: -------------------------------------------------------------
 
-DELIMITER  <- '$'
-SEP        <- ' '
-EQ_SIGN    <- ' = '
-EQUIV_SIGN <- ' \\equiv '
-DEF_SIGN   <- ' \\coloneq '
-SIM_SIGN   <- ' \\sim '
-IN_SIGN    <- ' \\in '
-NO_SEP     <- ''
-SEP_COMMA  <- ', '
-UNDERSCORE <- '_'
-ELLIPSIS   <- '...'
+DELIMITER     <- '$'
+SEP           <- ' '
+EQ_SIGN       <- ' = '
+EQUIV_SIGN    <- ' \\equiv '
+DEF_SIGN      <- ' \\coloneq '
+SIM_SIGN      <- ' \\sim '
+IN_SIGN       <- ' \\in '
+NO_SEP        <- ''
+SEP_COMMA     <- ', '
+UNDERSCORE    <- '_'
+NEWLINE       <- '\\\\'
+ALIGN         <- '&'
+ELLIPSIS      <- '...'
+DIAG_ELLIPSIS <- '\\ddots'
 
 TRANSPOSED_SUFFIX <- '^T'
 PRIME_SUFFIX      <- "'"

### 2022-08-05 notebooks/Space_transformation_theorem.qmd
Direction cosines complete
Rebase merge conflict addressed. Adds footnote nº 9

@@
 ---
-title: "Space transformation in Multidimensional IRT (Theorem)"
+title: "Space transformation in Multidimensional IRT (Lemmas)"
 author: "Mario Luzardo, Daniel Morillo"
@@
     highlight-style: github  
-  html: default
 callout-collapse: true
@@
 In $`r LATENT_SPACE`$ with basis $`r LS_BASIS`$ we define an inner product
-$`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant.
+$`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant,
+i.e., $`r TRAIT_NORMS_EQ`$. Thus,[^4]
+
+$$
+  `r TRAIT_NORM_STD_EQ`.
+$$
+
+Therefore we must define the inner product $`r INNER_PROD_BASIS`$ with Gram
+matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$.[^5]
+
+The Gram matrix has as element $`r INNER_PROD_MAT_EL_EQ`$. As
+$`r TRANSFORM_MATRIX`$ (and consequently, $`r TRANSFORM_MATRIX_TRANSP`$) is
+invertible so $`r INNER_PROD_MATRIX`$ is, and
+$`r INNER_PROD_MATRIX_INV_EQ`$.[^6]
+
+Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
+up by $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
+vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.[^7]
+
+::: {#lem-dircos}
+Under the previously defined conditions, for all $`r TRAIT_VEC_IN_LS`$,[^8]
+
+$$
+  `r DIR_COS_VEC_EQ`
+$$
+
+with
+
+$$
+  `r DIAG_MATRIX_INNER_PROD_INV_SQ_EQ`.
+$$
+
+Therefore,
+
+$$
+  `r TRAIT_VECTOR_POLAR_EQ2`.
+$$
+
+::: proof
+For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,
+
+$$
+  `r DIR_COSINE_PROOF`,
+$$
+
+with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
+$`r INNER_PROD_MATRIX`$.
+
+Therefore, $`r DIR_COS_VEC_EQ2`$ and $`r TRAIT_VECTOR_POLAR_EQ2`$.[^9]
+$\blacksquare$
+:::
+:::
+
+::: {#lem-convert-spaces}
+Let's consider another basis $`r NEW_BASIS_EQ`$ in $`r LATENT_SPACE`$ and the
+inner product $`r INNER_PROD_NEWBS`$ that keeps the norm invariant.
+
+Let $`r DIR_COS_VEC_NEWBS`$ be the vector of direction cosines of
+$`r TRAIT_VECTOR`$ (i.e., the cosines with the elements of $`r NEW_BASIS`$).
+
+Then,
+
+$$
+  `r LEMMA_2_EQ`,
+$$ {#eq-lemma-2}
+
+where
+
+$$
+  `r DIAG_MATRIX_INNER_PROD_NEWBS_INV_SQ_EQ`
+$$
+
+and $`r BASIS_CHANGE_NEWBS_DEF`$.
+
+::: proof
+Let's call $`r TRAIT_NEWBS_COORDS_DEF`$ and let $`r BASIS_CHANGE_NEWBS_DEF`$ be
+the change of basis matrix such that $`r BASIS_CHANGE_NEWBS`$.
+
+We have that $`r BASES_CHANGE_EQ`$.
+
+The inner product $`r INNER_PROD_NEWBS`$ that keeps the norm invariant is
+defined by[^10] $`r INNER_PROD_NEWBS_EQ`$, and we define
+$`r INNER_PROD_MAT_NEWBS_EQ`$.
+
+By @lem-dircos, for each case we have:
+
+1.  $`r TRAIT_VECTOR_POLAR_STD_EQ`$
+
+2.  $`r TRAIT_VECTOR_POLAR_EQ2`$
+
+3.  $`r TRAIT_VEC_NEWBS_POLAR_EQ`$
+
+Given (1),
+
+$$
+  `r TRAIT_COORDS_EQUIV`
+$$
+
+$$
+  `r TRAIT_COORDS_EQUIV_NEWBS`.
+$$
+
+Using (2) and (3) we get[^11]
+
+$$
+  `r TRAIT_COORDS_EQUIV_EXPANDED`
+$$
+
+$$
+  `r TRAIT_COORDS_EQUIV_NEWBS_EXPANDED`.
+$$
+
+Therefore,
+
+$$
+  `r TRAIT_COORDS_EQUIV_EXPANDED_EQ`,
+$$
+
+and thus $`r LEMMA_2_EQ`$.$\blacksquare$
+:::
+:::
 
@@
 
-[^3]: It is ok to use square brackets here? I'm using it instead of parentheses
+[^3]: Is it ok to use square brackets here? I'm using it instead of parentheses
     because I had already written this (with a function), but if it's not
     correct I change it to parentheses.
+
+[^4]: Mario wrote in the leftmost therm of the equality "the norm of
+    $`r TRAIT_VECTOR`$", but it should be "the *squared* norm of
+    $`r TRAIT_VECTOR`$".
+
+[^5]: Having already defined $`r TRAIT_COORDS_DEF`$, I have found this notation
+    more convenient (consistent with what I had already written, for which I had
+    it already coded).
+
+[^6]: I haven't considered it necessary to add the observation, as I think is a
+    rather known theorem in matrix algebra (see e.g. [this response in Math
+    Stackexchange](https://math.stackexchange.com/a/340234/1082326 "Response to 'Transpose of inverse vs inverse of transpose'")).
+
+[^7]: I'm omitting here (for the moment) the mathematical expressions defining
+    the direction cosine vectors.
+
+[^8]: Mario is using $V$ here but I understand he refers to the latent (vector)
+    space, which I have denoted it by $`r LATENT_SPACE`$. Is this correct?
+
+[^9]: I don't quite understand the whole proof, particularly, the equivalence
+    $`r latex_eq(latex(latex_transp(STD_BASIS_VECTOR_ANY), INNER_PROD_MATRIX), INNER_PROD_MATRIX_ROW)`$,
+    and the step from $`r DIR_COS_ANY`$ to $`r DIR_COS_VEC`$. This step is
+    similar to my step from Equation 6 to Equation 7 in the draft, which I had
+    the feeling was missing some intermediate steps.
+
+[^10]: Again I'm using here the notation I have used in the draft and before
+    (see also footnote 5).
+
+[^11]: I have skipped some basic matrix algebra steps here.

### 2022-08-05 R/Formulae.R
Direction cosines complete

@@ -630,7 +630,7 @@ TRANSFORM_MATRIX_EXAMPLE_EQ <- latex_def(
   TRANSFORM_MATRIX_EXAMPLE
 )
 
-### Mario's Theorem: ----
+### Mario's lemmas: ----
 
 TRAIT_VEC_STD_MOD    <- latex_prime(TRAIT_MODULE)
 TRAIT_VEC_STD_COMP   <- latex_sub(TRAIT_VEC_STD_MOD, DIM_INDEX)
@@ -685,6 +685,258 @@ BASIS_VEC_STD_COORDS_DEF <- latex_eq(
 )
 
 INNER_PROD_BASIS <- latex_innerprod('\\,', '') |> latex_sub(LS_BASIS)
+
+TRAIT_NORM_BASIS     <- TRAIT_NORM |> latex_sub(LS_BASIS)
+TRAIT_NORM_STD_BASIS <- TRAIT_NORM |> latex_sub(LS_STD_BASIS)
+TRAIT_NORMS_EQ       <- latex_eq(TRAIT_NORM_BASIS, TRAIT_NORM_STD_BASIS)
+
+TRAIT_NORM_STD_BASIS_SQ <- latex_squared(TRAIT_NORM_STD_BASIS)
+
+TRAIT_NORM_STD_EQ <- latex_eq(
+  TRAIT_NORM_STD_BASIS_SQ,
+  TRAIT_VECTOR_STD_SQ,
+  TRAIT_TRANSF_TRAIT
+)
+
+INNER_PROD_MAT_EL    <- latex_sub('m', "$AUX_INDEX$$DIM_INDEX$")
+BASIS_VECTOR_AUX     <- latex_sub(BASIS_VECTOR, AUX_INDEX)
+BASIS_VEC_INNER_PROD <- latex_innerprod(BASIS_VECTOR_AUX, BASIS_VECTOR_ANY)
+INNER_PROD_MAT_EL_EQ <- latex_eq(INNER_PROD_MAT_EL, BASIS_VEC_INNER_PROD)
+
+TRANSFORM_MATRIX_INV_TRANSP <- TRANSFORM_MATRIX_INV |> latex_transp()
+INNER_PROD_MATRIX_INV_EQ    <- latex_eq(
+  INNER_PROD_MATRIX_INV,
+  "$TRANSFORM_MATRIX_INV$ $TRANSFORM_MATRIX_TRANSP_INV$",
[…]

### 2022-08-05 notebooks/Space_transformation_theorem.qmd
The intermediate steps are basic algebra results and can be left implicit without further consideration.

@@
 
-Therefore, $`r DIR_COS_VEC_EQ2`$ and $`r TRAIT_VECTOR_POLAR_EQ2`$.[^9]
+Therefore, $`r DIR_COS_VEC_EQ2`$ and $`r TRAIT_VECTOR_POLAR_EQ2`$.
 $\blacksquare$
@@
 
-[^9]: I don't quite understand the whole proof, particularly, the equivalence
-    $`r latex_eq(latex(latex_transp(STD_BASIS_VECTOR_ANY), INNER_PROD_MATRIX), INNER_PROD_MATRIX_ROW)`$,
-    and the step from $`r DIR_COS_ANY`$ to $`r DIR_COS_VEC`$. This step is
-    similar to my step from Equation 6 to Equation 7 in the draft, which I had
-    the feeling was missing some intermediate steps.
-
 [^10]: Again I'm using here the notation I have used in the draft and before

### 2022-08-11 notebooks/Space_transformation_theorem.qmd
Direction cosines lemmas revised for errata

@@
 
-# Main
+# Direction consines transformations
 
-Let's consider a $`r N_DIMS`$ latent space $`r LATENT_SPACE`$ with Euclidean
-vector structure, and let $`r LS_BASIS_EQ`$ be a basis of $`r LATENT_SPACE`$.
+Let's consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$ with
+Euclidean vector structure, and let $`r LS_BASIS_EQ`$ be a basis of
+$`r LATENT_SPACE`$.
 
@@
 Let's consider a dichotomous item $`r ITEM_INDEX`$ such that the probability
-that respondent with trait $`r LS_BASIS_EQ`$ gives a positive[^1] response is
-given by
+that a respondent with trait $`r TRAIT_VEC_IN_LS`$ gives a positive[^1] response
+is given by
 
@@
 
-Let's consider in $`r LATENT_SPACE`$ the standard basis $`r LS_STD_BASIS_EQ`$ as
-well.
+Let's consider as well the standard basis $`r LS_STD_BASIS_EQ`$ in
+$`r LATENT_SPACE`$.
 
@@
 $$
-  `r BASIS_CHANGE_EQ`
+  `r BASIS_CHANGE_EQ`.
 $$ {#eq-change-basis}
@@
 
-In $`r LATENT_SPACE`$ with the standard basis we have the usual scalar product.
-In $`r LATENT_SPACE`$ with basis $`r LS_BASIS`$ we define an inner product
+With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product.
+With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
 $`r INNER_PROD_BASIS`$ such that the norm of $`r TRAIT_VECTOR`$ is invariant,
@@
 Therefore we must define the inner product $`r INNER_PROD_BASIS`$ with Gram
-matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$.[^5]
+matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$,[^5] which keeps
+the norm invariant.
 
-The Gram matrix has as element $`r INNER_PROD_MAT_EL_EQ`$. As
+The Gram matrix has as element $`r INNER_PROD_MAT_EL_EQ`$.[^6] As
 $`r TRANSFORM_MATRIX`$ (and consequently, $`r TRANSFORM_MATRIX_TRANSP`$) is
 invertible so $`r INNER_PROD_MATRIX`$ is, and
-$`r INNER_PROD_MATRIX_INV_EQ`$.[^6]
+$`r INNER_PROD_MATRIX_INV_EQ`$.[^7]
 
 Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
-up by $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
-vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.[^7]
+up by the $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
+vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.[^8]
 
 ::: {#lem-dircos}
-Under the previously defined conditions, for all $`r TRAIT_VEC_IN_LS`$,[^8]
+Under the previously defined conditions, for all $`r TRAIT_VEC_IN_LS`$,[^9]
 
@@
 Let's consider another basis $`r NEW_BASIS_EQ`$ in $`r LATENT_SPACE`$ and the
-inner product $`r INNER_PROD_NEWBS`$ that keeps the norm invariant.
+inner product $`r INNER_PROD_NEWBS`$ that keeps the norm of the Euclidean space
+invariant.
 
 Let $`r DIR_COS_VEC_NEWBS`$ be the vector of direction cosines of
-$`r TRAIT_VECTOR`$ (i.e., the cosines with the elements of $`r NEW_BASIS`$).
+$`r TRAIT_VECTOR`$ in $`r NEW_BASIS`$ (i.e., the cosines with the elements of
+$`r NEW_BASIS`$).
 
@@
 
+# Point of maximum slope
+
 [^1]: I prefer to refer to the responses as "positive/negative", instead of
-    "correct/incorrect", as this terminology contemplates the possibility of
-    non-cognitive (as well as cognitive) domains.
+    "correct/incorrect", as this terminology allows for non-cognitive (as well
+    as cognitive) domains.
 
@@
     $`r DIM_INDEX`$ for indexing the dimensions (instead of $`r ITEM_INDEX`$,
-    which is also used to index the items), $`r TRAIT_MODULE`$ for the
+    which is also used to index the items), $`r TRAIT_COMPONENT`$ for the
     coordinates of $`r TRAIT_VECTOR`$ in $`r LS_BASIS`$, and
@@
     for the vector coordinates is ambiguous, as we are using the same symbol
-    ($`r TRAIT_VECTOR`$) for the "geometric vector" and the "algebraic vector"
-    (i.e. column matrix) of its coordinates in $`r LATENT_SPACE`$ (this notation
-    is usually accepted though).
+    ($`r TRAIT_VECTOR`$) for the "geometric vector" (i.e. "arrow" in vector
+    space) and the "algebraic vector" (i.e. column matrix) of its coordinates in
+    $`r LATENT_SPACE`$ (although this notation is usually accepted).
 
@@
     because I had already written this (with a function), but if it's not
-    correct I change it to parentheses.
+    correct I change it to parentheses. **Also:** Is this a common algebra
+    result? It feels like the this step is left implicit without justification;
+    in my case, in the draft I needed to go all the way by Equations 2 through 4
+    (section 2.2.1)
 
@@
 
-[^6]: I haven't considered it necessary to add the observation, as I think is a
+[^6]: There is an erratum here in Mario's notes, I think. It should be the inner
+    product of $`r BASIS_VECTOR_AUX_TRANSF`$ and $`r BASIS_VECTOR_ANY_TRANSF`$
+    as I have written, instead of $`r BASIS_VECTOR_AUX`$ and
+    $`r BASIS_VECTOR_ANY`$, as Mario wrote.
+
+[^7]: I haven't considered it necessary to add the observation, as I think is a
     rather known theorem in matrix algebra (see e.g. [this response in Math
@@
 
-[^7]: I'm omitting here (for the moment) the mathematical expressions defining
+[^8]: I'm omitting here (for the moment) the mathematical expressions defining
     the direction cosine vectors.
 
-[^8]: Mario is using $V$ here but I understand he refers to the latent (vector)
+[^9]: Mario is using $V$ here but I understand he refers to the latent (vector)
     space, which I have denoted it by $`r LATENT_SPACE`$. Is this correct?

### 2022-08-11 R/Formulae.R

@@ -646,7 +646,7 @@ TRAIT_VECTOR_DEF_BASIS        <- latex_summation(
 )
 TRAIT_VECTOR_DEF_BASIS_EQ     <- latex_eq(TRAIT_VECTOR, TRAIT_VECTOR_DEF_BASIS)
 TRAIT_VECTOR_DEF_STD_BASIS    <- latex_summation(
-  TRAIT_VEC_STD_COMP, BASIS_VECTOR_ANY_TRANSF,
+  TRAIT_VEC_STD_COMP, STD_BASIS_VECTOR_ANY,
   index = DIM_INDEX, from = 1, to = N_DIMS,
   .par = FALSE
 )
@@ -701,10 +701,14 @@ TRAIT_NORM_STD_EQ <- latex_eq(
   TRAIT_TRANSF_TRAIT
 )
 
-INNER_PROD_MAT_EL    <- latex_sub('m', "$AUX_INDEX$$DIM_INDEX$")
-BASIS_VECTOR_AUX     <- latex_sub(BASIS_VECTOR, AUX_INDEX)
-BASIS_VEC_INNER_PROD <- latex_innerprod(BASIS_VECTOR_AUX, BASIS_VECTOR_ANY)
-INNER_PROD_MAT_EL_EQ <- latex_eq(INNER_PROD_MAT_EL, BASIS_VEC_INNER_PROD)
+INNER_PROD_MAT_EL       <- latex_sub('m', "$AUX_INDEX$$DIM_INDEX$")
+BASIS_VECTOR_AUX        <- latex_sub(BASIS_VECTOR, AUX_INDEX)
+BASIS_VECTOR_AUX_TRANSF <- latex_prime(BASIS_VECTOR_AUX)
+BASIS_VEC_INNER_PROD    <- latex_innerprod(
+  BASIS_VECTOR_AUX_TRANSF,
+  BASIS_VECTOR_ANY_TRANSF
+)
+INNER_PROD_MAT_EL_EQ    <- latex_eq(INNER_PROD_MAT_EL, BASIS_VEC_INNER_PROD)
 
 TRANSFORM_MATRIX_INV_TRANSP <- TRANSFORM_MATRIX_INV |> latex_transp()
 INNER_PROD_MATRIX_INV_EQ    <- latex_eq(

### 2022-08-11 output/paper_draft/paper_M2PLM_parameters.Rmd
Minor errata corrected in the manuscript draft

@@
 we first compute the point of maximum slope by finding the root(s) of
-the second derivative with respect to $`r TRAIT_VECTOR`$.
+the second derivative with respect to $`r TRAIT_MODULE`$.
 
@@
 However, this assertion implicitly assumes orthogonal coordinates,
-so it does not apply in general to `r DIR_COS_VEC`.<!-- # TODO: To appendix the following, or maybe give for granted? -->
+so it does not apply in general to $`r DIR_COS_VEC`$.<!-- # TODO: To appendix the following, or maybe give for granted? -->
 Instead, we may consider the direction cosine vector in $`r LATENT_SPACE_STD`$:

### 2022-08-11 notebooks/Space_transformation_theorem.qmd
Final transcriptions of Mario's lemmas about the direction cosines

@@
 
+There is little difference (notation and a few steps) from my own computataions,
+so I think its ok not to transcribe it (Mario and I are "on the same page").
+
+# Test space and latent space
+
+I think this part details the derivation of my Equation 21 (which I have
+omitted, but is equivalent to Equations 13 through 15), so it is ok not to
+transcribe this part for now.
+
 [^1]: I prefer to refer to the responses as "positive/negative", instead of

### 2022-09-14 notebooks/Search_Citation_Gecko.Rmd
Formats the "Citation Gecko notebook" free text with maximum column width

@@
 title: "Search with CitationGecko"
-output: html_notebook
-bibliography: ../www/Multidimensional-parameters-MCLM.bib
-csl: ../www/apa-old-doi-prefix.csl
+output:
+  html_document:
+    df_print: paged
+    toc: yes
+    toc_float: yes
+  html_notebook: default
+bibliography: "../www/Multidimensional-parameters-MCLM.bib"
+csl: "../www/apa-old-doi-prefix.csl"
 ---
@@
 
-On 15 july, 2021, I use the following papers as seeds in [CitationGecko](https://citationgecko.azurewebsites.net/):
+On 15 july, 2021, I use the following papers as seeds in
+[CitationGecko](https://citationgecko.azurewebsites.net/):
 
@@
 
-Then click on "Export" (in the network view) and save the results as `output/citation_gecko/GeckoRecommendations.bib`[^search_citation_gecko-1].
+Then click on "Export" (in the network view) and save the results as
+`output/citation_gecko/GeckoRecommendations.bib`[^search_citation_gecko-1].
 
-[^search_citation_gecko-1]: A bug in `bib2pdf()` prevents from saving files in the usual resource folder `www`.
-(It parses the regexp `www.` as a remote URL, which is matched by `www/`; it should parse `www\\.` to interpret a literal dot.)
+[^search_citation_gecko-1]: A bug in `bib2pdf()` prevents from saving files in
+the usual resource folder `www`. (It parses the regexp `www.` as a remote URL,
+which is matched by `www/`; it should parse `www\\.`
+to interpret a literal dot.)
 
@@
 
-I import the bibliography to Zotero into a new collection "GeckoRecommendations".
-After initial curtaion (eliminate duplicates, search for available full text, and rename PDFs according to metadata), the following manual filtering is made:
+I import the bibliography to Zotero into a new collection
+"GeckoRecommendations". After initial curtaion (eliminate duplicates, search for
+available full text, and rename PDFs according to metadata), the following
+manual filtering is made:
 
@@
 
--   Entries with titles "Commentaries", "References" (content checked, non-relevant)
+-   Entries with titles "Commentaries", "References"
+    (content checked, non-relevant)
 
-I export this collection as `output/citation_gecko/GeckoRecommendations_filtered.bib`.
+I export this collection as
+`output/citation_gecko/GeckoRecommendations_filtered.bib`.
 
@@
 
-One entry without journal name; corresponds to a book, the bibliographic entry is from 1985 (too old for my purposes), and barely addresses the topic of multidimensionality.
+One entry without journal name; corresponds to a book, the bibliographic entry
+is from 1985 (too old for my purposes),
+and barely addresses the topic of multidimensionality.
 
@@
 
-I explore several journals, that seem to be non-central to the topic, but essentially most of them seem to cite Reckase or Zhang for some relevant reason.
-It is difficult to filter out journals by name or keywords, as most articles seem to deal with multidimensional IRT in one way or another, even in journals that seem completely unrelated.
-Some of them are even theoretical reviews of multidimensional IRT and not just applications.
+I explore several journals, that seem to be non-central to the topic, but
+essentially most of them seem to cite Reckase or Zhang for some relevant reason.
+It is difficult to filter out journals by name or keywords, as most articles
+seem to deal with multidimensional IRT in one way or another, even in journals
+that seem completely unrelated. Some of them are even theoretical reviews of
+multidimensional IRT and not just applications.
 
@@
 
-We filter the publications that contain the keyword "multidimensional" in any field:
+We filter the publications that contain the keyword "multidimensional"
+in any field:
 
@@
 
-This leaves `r n_entries_multidimensional` entries, which are suitable for manual classification.
+This leaves `r n_entries_multidimensional` entries,
+which are suitable for manual classification.
 
@@
 
-I export the selected entries to an Excel file, first reordering the fields, flattening the `AUTHOR` field (otherwise it is not written properly), and ordering by year to make the search easier.
+I export the selected entries to an Excel file, first reordering the fields,
+flattening the `AUTHOR` field (otherwise it is not written properly),
+and ordering by year to make the search easier.
 
@@
 
-Then I manually copy this file to `../output/citation_gecko/Recommendations_wb.xlsx`.
-I do this as a safeguard, in case I run again this whole notebook no information is lost in the working file if it is overwritten.
+Then I manually copy this file to
+`../output/citation_gecko/Recommendations_wb.xlsx`. I do this as a safeguard,
+in case I run again this whole notebook no information is lost in
+the working file if it is overwritten.
 
@@
 
-I create a new collumn in the Excel file "CLASSIFICATION", in column "AI", where I fill in the results of the manual classification.
-Then, to decide whether a paper is relevant, I search for citations of @reckase_difficulty_1985, @reckase_discriminating_1991, @ackerman_creating_1994, and @zhang_conditional_1999.
-If the paper mentions the computation of the multidimensional difficulty (MDIFF, MID, or MDL) or discrimination (MDISC) parameter (I search for those terms and also "difficulty" and "discrimination"), then I check whether they assume an orthogonal latent space (by searching for "distribution", "orthogonal", and "correlation", if necessary).
+I create a new column in the Excel file "CLASSIFICATION", in column "AI",
+where I fill in the results of the manual classification.
+Then, to decide whether a paper is relevant, I search for citations of
+@reckase_difficulty_1985, @reckase_discriminating_1991, @ackerman_creating_1994,
+and @zhang_conditional_1999. If the paper mentions the computation of
+the multidimensional difficulty (MDIFF, MID, or MDL) or discrimination (MDISC)
+parameter (I search for those terms and also "difficulty" and "discrimination"),
+then I check whether they assume an orthogonal latent space (by searching for
+"distribution", "orthogonal", and "correlation", if necessary).
 
@@
 
--   **Orthogonality assumed**: Assumes orthogonality among latent dimensions, and computes the MDIFF and MDISC accordingly, OR does not mention the latent space structure at all (and therefore it just assumes orthogonality, as in @reckase_difficulty_1985 and @reckase_discriminating_1991).
+-   **Orthogonality assumed**: Assumes orthogonality among latent dimensions,
+    and computes the MDIFF and MDISC accordingly, OR does not mention the latent
+    space structure at all (and therefore it just assumes orthogonality,
+    as in @reckase_difficulty_1985 and @reckase_discriminating_1991).
 
--   **Not relevant**: There is no mention to the multidimensional parameters or their computations, OR they are only discussed theoretically.
-Multidimensional traits may be simulated, dealt with, or theoretically discussed, but not in relationship with the multidimensional parameters.
+-   **Not relevant**: There is no mention to the multidimensional parameters or
+    their computations, OR they are only discussed theoretically.
+    Multidimensional traits may be simulated, dealt with, or theoretically
+    discussed, but not in relationship with the multidimensional parameters.
 
--   **Oblique dimensions disregarding non-orthogonality**: Computes the MDIFF and/or MDISC parameters applying Reckase's [-@reckase_difficulty_1985] and Reckase and McKinely's [-@reckase_discriminating_1991] formulae, disregarding the fact that these only apply to orthogonal latent spaces.
+-   **Oblique dimensions disregarding non-orthogonality**: Computes the MDIFF
+    and/or MDISC parameters applying Reckase's [-@reckase_difficulty_1985] and
+    Reckase and McKinely's [-@reckase_discriminating_1991] formulae,
+    disregarding the fact that these only apply to orthogonal latent spaces.
 
@@
 
-While doing the manual classification, I find out that at least one entry (Wang, Chang, & Boughton, 2010) has been filtered out due to the title using the term "Multi-Dimensional" (with a dash) instead of "Multidimensional".
-Therefore, I change the matching regular expression to "`multi-?dimensional"`, which matches both terms.
-However, the BibTeX entries are "[case protected](https://tex.stackexchange.com/a/10775/171606)" with curly braces, which may prevent entries like this one from being detected ("{Multi}-{Dimensional}" instead of "{Multi-Dimensional}".
-In order to avoid this, the curly braces are removed from the titles of the bibliographical entries first.
+While doing the manual classification, I find out that at least one entry
+(Wang, Chang, & Boughton, 2010) has been filtered out due to the title using
+the term "Multi-Dimensional" (with a dash) instead of "Multidimensional".
+Therefore, I change the matching regular expression to "`multi-?dimensional"`,
+which matches both terms. However, the BibTeX entries are
+"[case protected](https://tex.stackexchange.com/a/10775/171606)" with
+curly braces, which may prevent entries like this one from being detected
+("{Multi}-{Dimensional}" instead of "{Multi-Dimensional}".
+In order to avoid this, the curly braces are removed from the titles of
+the bibliographical entries first.
 
@@
 
-The updated list has `r n_entries_multidimensional_hyphenated` (apparently, only that one entry had been missed for this reason).
+The updated list has `r n_entries_multidimensional_hyphenated`
+(apparently, only that one entry had been missed for this reason).
 
@@
 
-While doing the manual search, I notice a few papers having the acronym "MIRT" in the title, that have not been considered in the results.
-As I assume they also refer to "Multidimensional Item Response Theory", and therefore should be included in the results, I re-run the filtering including these results.
+While doing the manual search, I notice a few papers having the acronym "MIRT"
+in the title, that have not been considered in the results.
+As I assume they also refer to "Multidimensional Item Response Theory", and
+therefore should be included in the results,
+I re-run the filtering including these results.
 
@@
 
-The number of entries is now `r n_entries_multidimensional_mirt`. The new ones included are:
+The number of entries is now `r n_entries_multidimensional_mirt`.
+The new ones included are:
 
@@
 
-There is at least one known paper that has not been included and is known to be relevant:
+There is at least one known paper that has not been included and
+is known to be relevant:
 
-Fragoso, T.
-M., & Cúri, M.
-(2013).
-Improving psychometric assessment of the Beck Depression Inventory using Multidimensional Item Response Theory.
+Fragoso, T. M., & Cúri, M. (2013). Improving psychometric assessment of the
+Beck Depression Inventory using Multidimensional Item Response Theory.
 *Biometrical Journal*, *55*(4), 527--540.
@@
 
-This reference has been found because it is cited by Silva (2019), and classifies as "Oblique dimensions disregarding non-orthogonality".
+This reference has been found because it is cited by Silva (2019),
+and classifies as "Oblique dimensions disregarding non-orthogonality".
 The problem is that it does not cite directly any of the seed papers considered.
-Instead, it cites Reckase (1997; A Linear Logistic Multidimensional Model for Dichotomous Item Response Data).
+Instead, it cites Reckase (1997; A Linear Logistic Multidimensional Model for
+Dichotomous Item Response Data).
 
-The problem with CitationGecko is that it does not allow to construct the whole citation network, so only direct references of the seed papers can be found.
+The problem with CitationGecko is that it does not allow to construct
+the whole citation network, so only direct references of
+the seed papers can be found.

### 2022-09-14 output/paper_draft/paper_M2PLM_parameters.Rmd
Credits Ackerman (1994, 2005) for acknowledging that Reckase (1985) and Reckase & McKinley (1991) imply orthogonal axes

@@
 latent space are orthogonal
-[see @reckase_difficulty_1985, p. 404 and Equation 7].
+[@ackerman_using_1994; @ackerman_multidimensional_2005; see also @reckase_difficulty_1985, p. 404 and Equation 7].
 Of course, orthogonality is a property of a geometrical space instead of a
@@
 Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
-non-negative real numbers in Reckse
+non-negative real numbers in Reckase
 [-@reckase_difficulty_1985, a decrease in the probability of a correct response with an increase in ability makes little sense].

### 2022-09-16 notebooks/Multidimensional_information.qmd
Multidimensional information notebook versioned

---
title: "Multidimensional information"
author: "Daniel Morillo"
format: html
bibliography: ../www/Multidimensional-parameters-MCLM.bib
editor: visual
---

# Introduction

Hereby I am going to derive the multidimensional information, according to
@ackerman_creating_1994.

# Derivations

@ackerman_creating_1994's Equations 1 through 7 and 13 through 16 for
bidimensional latent spaces apply directly to the 3-or-more dimensional case,
regardless of the latent covariance structure, as this is not involved at all in
those expressions. The only caveat is that he uses the normal parameterization,
and thus for the M2PL as we have been formulating it the 1.7 factor in Equation
1 does not apply (and subsequently, the 1.7\^2 = 2.89 factor in Equations 7 and
14).

The first deviation from his work comes in the paragraph right before Equation
13, where he states that

> the unidimensional trait level $\theta$ is equal to the linear composite,
> $\theta_c$, defined by $\theta_c = \beta_1 \theta_1 + \beta_2 \theta_2$, in
> which $\theta_c$ represents a linear composite in the angular direction of
> $\alpha$ and $\beta_1 = \cos \alpha$ and $\beta_2 = \sin \alpha$.

In our case, $\mathbf{\beta}$ must be a $D$-dimensional vector, being its $k$-th
component $\beta_k = \cos \alpha_k$, $k = [1, \dots, n]$.

## Components of $\mathbf{\beta}$

The main problem with determining the composite direction given by
$\mathbf{\beta}$ is that its components must be the cosines of the angles with
the $\theta_k$ axes. At this point, my main doubt is which space to use for
computing the angles: the latent trait space or the test space?

I will first try with a correlated, bidimensional latent space as an example to
explore the solution. In this space, the information of simple-structure items
in the direction parallel to the axes of its tapped trait should be equal to the
information of the unidimensional case. In the direction of the other trait
axis, the information should be more positive the more positive the correlation
is (constrained by the information of the former axis as maximum value).

I first define the latent space, by considering a bivariate standardize normal
distribution with correlated dimensions:

```{r}
rho     <- .5
sigma   <- c(1, rho, rho, 1) |> matrix(2)
sig_inv <- sigma |> solve()
```

I define the item discrimination parameters:

```{r}
a_item <- c(1, 0) |> matrix()
d_item <- 0
```

Then, the composite direction must be given by any arbitrary vector that is then
multiplied (by the inner product) with the basis vectors to be transformed into
a vector of cosines:

```{r}
comp_vec <- c(0, 1) |> matrix()
```

Now, computing the module of the basis vectors and the composite vector requires
the inner product matrix for the first time. I will try with the inverse of the
covariance matrix first.

```{r}
library(magrittr)

ip_matrix <- sig_inv

comp_vec_mod <- t(comp_vec) %*% ip_matrix %*% comp_vec |>
  drop() |>
  sqrt()

u_1 <- c(1, 0) |> matrix()
u_2 <- c(0, 1) |> matrix()

basis_matrix <- cbind(u_1, u_2)
basis_vec_mods <- t(basis_matrix) %*% ip_matrix %*% basis_matrix |>
  diag() |>
  sqrt()

comp_cos_vec <- t(comp_vec) %*% ip_matrix %*% basis_matrix |>
  divide_by(comp_vec_mod * basis_vec_mods) |>
  matrix()

comp_cos_vec
```

These are the cosines, the components of the vector $\beta$. The covariance
error matrix, Equation 13 in @ackerman_creating_1994, is given by the inverse of
the information matrix. The latter can be expressed as
$\mathbf{A}^T\mathbf{A}PQ$, being $\mathbf{A}$ the matrix of all item
parameters; the covariance matrix can be subsequently expressed as
$(\mathbf{A}^T\mathbf{A})^{-1}PQ^{-1}$. In our case, for a single item,
$\mathbf{A} = a^T$.

```{r}
A_matrix <- a_item |> t()

A_sq     <- t(A_matrix) %*% A_matrix
# A_sq_inv <- A_sq |> solve() ## Not run
```

As @ackerman_creating_1994 states (and I already knew) "for a single item or for
$n$ items that measure in exactly the same direction \[...\], Equation 7 (i.e.
the information matrix) is singular" (p. 259). This implies that I need at least
two items, and I want the information to be mostly aligned with one of the axes,
but they must not measure in the same direction. Therefore, I create a second
item, with near simple structure on the trait of the first axis.

```{r}
a_item_1 <- a_item
a_item_2 <- c(.95, .1) |> matrix()

A_matrix <- cbind(a_item_1, a_item_2) |> t()

A_sq     <- t(A_matrix) %*% A_matrix
A_sq_inv <- A_sq |> solve()
```

Once we have the error covariance matrix (except for the $PQ^{-1}$ factor), we
can pre- and post- multiply by (transposed) $\mathbf{\beta}$ to compute first
the error variance in that direction, and then its inverse to obtain the
information.

```{r}
error_var <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
comp_info <- 1 / error_var
```

The resulting multidimensional information coefficient for the direction
parallel to the second axis is `r comp_info`. Let's try to compute it in the
direction of the first axis:

```{r}
comp_vec     <- c(1, 0) |> matrix()
comp_vec_mod <- t(comp_vec) %*% ip_matrix %*% comp_vec |>
  drop() |>
  sqrt()
comp_cos_vec <- t(comp_vec) %*% ip_matrix %*% basis_matrix |>
  divide_by(comp_vec_mod * basis_vec_mods) |>
  matrix()

error_var <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
comp_info <- 1 / error_var
```

In the first axis the coefficient is `r comp_info`. This seems rather small when
one should expect a much higher information than in the second axis.

### 2022-09-28 notebooks/Multidimensional_information.qmd
Minor corrections to multidimensional infromation notebook

@@
 and thus for the M2PL as we have been formulating it the 1.7 factor in Equation
-1 does not apply (and subsequently, the 1.7\^2 = 2.89 factor in Equations 7 and
+1 does not apply (and subsequently, the 1.7^2^ = 2.89 factor in Equations 7 and
 14).
@@
 information of the unidimensional case. In the direction of the other trait
-axis, the information should be more positive the more positive the correlation
-is (constrained by the information of the former axis as maximum value).
+axis, the information should be higher the more positive the correlation is
+(constrained by the information of the former axis as maximum value).
 
-I first define the latent space, by considering a bivariate standardize normal
+I first define the latent space, by considering a bivariate standardized normal
 distribution with correlated dimensions:
@@
 
-These are the cosines, the components of the vector $\beta$. The covariance
-error matrix, Equation 13 in @ackerman_creating_1994, is given by the inverse of
-the information matrix. The latter can be expressed as
+These are the cosines, the components of the vector $\beta$. The error
+covariance matrix, Equation 13 in @ackerman_creating_1994, is given by the
+inverse of the information matrix. The latter can be expressed as
 $\mathbf{A}^T\mathbf{A}PQ$, being $\mathbf{A}$ the matrix of all item
-parameters; the covariance matrix can be subsequently expressed as
+parameters; the error covariance matrix can be subsequently expressed as
 $(\mathbf{A}^T\mathbf{A})^{-1}PQ^{-1}$. In our case, for a single item,

### 2022-09-28 notebooks/Multidimensional_information.qmd
Update to the multidimensional information notebook

@@
 author: "Daniel Morillo"
-format: html
+format:
+  html:
+    toc: true
+    toc-location: left
+    toc-depth: 4
 bibliography: ../www/Multidimensional-parameters-MCLM.bib
@@
 
+```{r setup}
+#| include: false
+options(digits = 5)
+```
+
 # Introduction
 
-Hereby I am going to derive the multidimensional information, according to
-@ackerman_creating_1994.
+Hereby I am going to try to derive the multidimensional information, according
+to @ackerman_creating_1994.
 
 # Derivations
+
+## Multidimensional information
 
@@
 14).
+
+## Linear composite $\mathbf{\beta}$
 
@@
 
-## Components of $\mathbf{\beta}$
+### Components of $\mathbf{\beta}$
 
@@
 computing the angles: the latent trait space or the test space?
+
+## Example with a correlated bidimensional space
 
@@
 
-```{r}
+```{r def-corr}
 rho <- .5
+```
+
+```{r def-covariance}
 sigma   <- c(1, rho, rho, 1) |> matrix(2)
@@
 
-```{r}
+```{r def-item}
 a_item <- c(1, 0) |> matrix()
@@
 
-```{r}
+```{r def-composite}
 comp_vec <- c(0, 1) |> matrix()
 ```
+
+### Computation in the latent space {#latent-space}
 
@@
 
-```{r}
+```{r def-inner-product}
+ip_matrix <- sig_inv
+```
+
+```{r compute-composite-components}
 library(magrittr)
-
-ip_matrix <- sig_inv
 
@@
 
-```{r}
+```{r compute-discr-matrix-attemt}
 A_matrix <- a_item |> t()
@@
 
-As @ackerman_creating_1994 states (and I already knew) "for a single item or for
-$n$ items that measure in exactly the same direction \[...\], Equation 7 (i.e.
-the information matrix) is singular" (p. 259). This implies that I need at least
-two items, and I want the information to be mostly aligned with one of the axes,
-but they must not measure in the same direction. Therefore, I create a second
-item, with near simple structure on the trait of the first axis.
+As @ackerman_creating_1994 states "for a single item or for $n$ items that
+measure in exactly the same direction \[...\], Equation 7 (i.e. the information
+matrix) is singular" (p. 259). This I already knew, but hadn't realized it was
+the case; therefore, my previous assertion that "in the direction of the other
+trait axis, the information should be higher the more positive the correlation
+is" is actually false (for one item). Therefore, I need at least two items, and
+I want the information to be mostly aligned with one of the axes, but they must
+not measure in the same direction. Thus I create a second item, with near simple
+structure on the trait of the first axis.
 
-```{r}
+```{r def-items}
 a_item_1 <- a_item
-a_item_2 <- c(.95, .1) |> matrix()
+a_item_2 <- c(1.5, .1) |> matrix()
 
@@
 
-```{r}
+```{r compute-info-axis2}
 error_var  <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
-comp_info <- 1 / error_var
+comp_info2 <- 1 / error_var
 ```
@@
 The resulting multidimensional information coefficient for the direction
-parallel to the second axis is `r comp_info`. Let's try to compute it in the
+parallel to the second axis is `r comp_info2`. Let's try to compute it in the
 direction of the first axis:
 
-```{r}
+```{r compute-info-axis1}
 comp_vec     <- c(1, 0) |> matrix()
@@
 error_var  <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
-comp_info <- 1 / error_var
+comp_info1 <- 1 / error_var
+
+ratio <- comp_info1 / comp_info2
 ```
 
-In the first axis the coefficient is `r comp_info`. This seems rather small when
-one should expect a much higher information than in the second axis.
+In the first axis the coefficient is `r comp_info1`. This seems rather small
+when one should expect a much higher information than in the second axis; the
+ratio in the two directions is `r ratio`.
+
+#### Test with higher correlation
+
+Common sense tells us that the more aligned the axes are, the more similar the
+information will be on the two dimensions parallel with the two axes. To test
+this commonsense assumption, we try the previous computations again with a
+higher correlation.
+
+```{r test-higher-corr}
+# Correlation and covariance matrix
+rho       <- .93
+sigma     <- c(1, rho, rho, 1) |> matrix(2)
+sig_inv   <- sigma |> solve()
+ip_matrix <- sig_inv
+
+# Basis vectors
+basis_vec_mods <- t(basis_matrix) %*% ip_matrix %*% basis_matrix |>
+  diag() |>
+  sqrt()
+
+# Composite cosine vectors
+comp_vecs     <- diag(2)
+comp_vecs_mod <- t(comp_vecs) %*% ip_matrix %*% comp_vecs |>
+  diag() |>
+  sqrt()
+comp_cos_vecs <- t(comp_vecs) %*% ip_matrix %*% basis_matrix |>
+  divide_by(comp_vecs_mod * basis_vec_mods)
+
+# Error variance and information in the composite directions
+error_var <- t(comp_cos_vecs) %*% A_sq_inv %*% comp_cos_vecs |> diag()
+comp_info <- 1 / error_var
+
+ratio <- comp_info[1] / comp_info[2]
+```
+
+The information in the two axes is now `r comp_info[1]` and `r comp_info[2]`,
+being their ratio `r ratio`. This fulfills the assumption we wanted to test.
+However, it doesn't make much sense that the information in the direction of the
+first axis has decreased.
+
+### Computation in the test space
+
+We are going to test the same assumption, but this time in the test space. This
+implies using the covariance matrix as inner product matrix, instead of its
+inverse.
+
+```{r}
+rho <- .5
+```
+
+We try first with a correlation of `r rho`.
+
+```{r compute-test-space}
+# Covariance matrix
+sigma     <- c(1, rho, rho, 1) |> matrix(2)
+ip_matrix <- sigma
+
+# Basis vectors
+basis_vec_mods <- t(basis_matrix) %*% ip_matrix %*% basis_matrix |>
+  diag() |>
+  sqrt()
+
+# Composite cosine vectors
+comp_vecs     <- diag(2)
+comp_vecs_mod <- t(comp_vecs) %*% ip_matrix %*% comp_vecs |>
+  diag() |>
+  sqrt()
+comp_cos_vecs <- t(comp_vecs) %*% ip_matrix %*% basis_matrix |>
+  divide_by(comp_vecs_mod * basis_vec_mods)
+
+# Error variance and information in the composite directions
+error_var <- t(comp_cos_vecs) %*% A_sq_inv %*% comp_cos_vecs |> diag()
+comp_info <- 1 / error_var
+
+ratio <- comp_info[1] / comp_info[2]
+```
+
+The information in both axes directions would be `r comp_info[1]` and
+`r comp_info[2]`, and their ratio `r ratio`.
+
+```{r}
+rho <- .93
+```
+
+Now we compute the same result, but with a correlation of `r rho` as we did
+before.
+
+```{r compute-test-space-high-corr}
+# Covariance matrix
+sigma     <- c(1, rho, rho, 1) |> matrix(2)
+ip_matrix <- sigma
+
+# Basis vectors
+basis_vec_mods <- t(basis_matrix) %*% ip_matrix %*% basis_matrix |>
+  diag() |>
+  sqrt()
+
+# Composite cosine vectors
+comp_vecs     <- diag(2)
+comp_vecs_mod <- t(comp_vecs) %*% ip_matrix %*% comp_vecs |>
+  diag() |>
+  sqrt()
+comp_cos_vecs <- t(comp_vecs) %*% ip_matrix %*% basis_matrix |>
+  divide_by(comp_vecs_mod * basis_vec_mods)
+
+# Error variance and information in the composite directions
+error_var <- t(comp_cos_vecs) %*% A_sq_inv %*% comp_cos_vecs |> diag()
+comp_info <- 1 / error_var
+
+ratio <- comp_info[1] / comp_info[2]
+```
+
+The information would be now `r comp_info[1]` and `r comp_info[2]` for the first
+and second axis direction, respectively, and their ratio `r ratio`.
+
+### Interpretation of the differences in the results in both spaces
+
+The information in the direction of the first axis decreases also, as happened
+when computing the directions in the latent space. However, the information in
+the direction of the second axis increases, while when computing it in the
+latent space it also decreased. It makes more sense to assume that the direction
+would increase in the second axis direction with an increase in the correlation,
+as it gets more aligned with the first axis.
+
+On the other hand, exploring the cosines that determine the composite direction
+one can see that, when using the latent space, they get more aligned **in the
+opposite direction** when the correlation increases. This makes me think that,
+in this case, the information is more evenly distributed across all direction in
+the latent space. However, what I would expect is that the information get more
+"concentrated" in a single direction.
+
+### Correction about the computation of the information matrix
+
+There is also one last thing when computing the information with two (or more)
+items: I have assumed that the $PQ^{-1}$ factor can be factored out of the
+matrix (see section [Computation in the latent space](#latent-space) above).
+However, this is wrong. Its value is different for each item, and thus
+it must be taken into account when computing the sum of the information for all
+the items. This implies that the information matrix cannot be computed as
+$\mathbf{A}^T\mathbf{A}PQ$ as I assumed before, for more than one item.

### 2022-10-21 notebooks/Distributions_Carlson_2017.qmd
Versions notebook about the issuess in Carlson's (2017) report

---
title: "Distributions in Carlson (2017)"
format: html
editor: visual
bibliography: ../www/Multidimensional-parameters-MCLM.bib
---

# Introduction

These distributions are taken from @carlson_unidimensional_2017's Table 2.

```{r table-2}
covs <- c(
  0.496, 0.138, 0.678,
  0.496, 0.223, 0.870,
  0.507, 0.328, 0.902,
  0.295, 0.428, 0.928,
  0.085, 0.383, 0.808,
  0.069, 0.546, 0.380
) |>
  matrix(ncol = 3, byrow = TRUE)

get_corr_mat <- function(vals) {

  cov <- vals[3] |> matrix(2, 2)
  diag(cov) <- vals[1:2]

  cov2cor(cov)
}

get_corr_mat(covs[1, ])
get_corr_mat(covs[2, ])
get_corr_mat(covs[3, ])
get_corr_mat(covs[4, ])
get_corr_mat(covs[5, ])
get_corr_mat(covs[6, ])
```

### 2022-10-21 output/paper_draft/paper_M2PLM_parameters.Rmd
Update with Mario's review (WIP)

@@
 Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$,
-defined by vector space $V$ and basis $`r LS_BASIS`$,
-being $`r LS_BASIS_EQ`$ an arbitrary set of linearly independent vectors.
+defined by vector space $V$ and basis $`r LS_BASIS_EQ`$.
 A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
-with coordinates $`r TRAIT_VECTOR`$ in $`r LATENT_SPACE`$,
 which is assumed to be a multivariate normal random variable:
@@
 
-As in @reckase_difficulty_1985, the characteristic function of item
-$`r ITEM_INDEX`$ according to the M2PL model is formulated as follows:
+According to the M2PL model, the probability of a positive response to item
+$`r ITEM_INDEX`$ is
 
@@
 
-being  
-$`r IRF_M2PL`$ ($= `r IRF_ABBR`$ henceforth, for simplicity) the probability of
-a positive response to item $`r ITEM_INDEX`$ for a given $`r TRAIT_VECTOR`$,  
-$`r DISCR_VECTOR`$ a $`r N_DIMS`$-dimensional vector of
-discrimination parameters, and  
-$`r INTERCEPT_PARAM`$ an intercept parameter related to the location of the item
-in the latent space.
+where $`r DISCR_VECTOR`$ is a $`r N_DIMS`$-dimensional vector of discrimination
+parameters, and $`r INTERCEPT_PARAM`$ an intercept parameter related to the
+location of the item in the latent space.
 
@@
 ## Linear algebra of non-orthogonal coordinates
-<!-- # TODO: To appendix, OR citation about these? -->
 
-### Inner product
+Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
+TODO: Incomplete in Mario's version, is it correct to say "in the latent space"?
+We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
+Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
+$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$, where $`r TRAIT_COORDS_EQ`$ and
+$`r TRAIT_COORDS_STD_EQ`$.
 
-Let $`r INNER_PROD_EQ`$ be the inner product in $`r LATENT_SPACE`$, with
-the positive-definite matrix $`r INNER_PROD_MATRIX`$ its inner product matrix.
-We may apply a change of basis in $`r LATENT_SPACE`$ that results in a space
-$`r LATENT_SPACE_STD`$ with the standard basis $`r LS_STD_BASIS_EQ`$.
-Such a change of basis can be defined as
+Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix; then, we have that
 
 \begin{equation}
-  `r BASIS_CHANGE`,                                          (\#eq:change-basis)
+  `r BASIS_CHANGE`.                                          (\#eq:change-basis)
 \end{equation}
+
+The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$
 
@@
 ```{r read-chunks-empirical-example}
-read_chunk("src/Empirical_example.R")
+read_chunk("src/Empirical_example_Tezza2018.R")
 ```

### 2022-10-21 R/Formulae.R

@@ -93,7 +93,7 @@ EXPFUNC_M2PL  <- latex_exp(latex_sqbrackets("-$LOGIT_M2PL$"))
 LOGISTIC_M2PL <- latex_logistic(LOGIT_M2PL)
 LOG_INV_M2PL  <- latex_inv_log(LOGIT_M2PL)
 
-M2PL_FORMULATION <- latex_eq(IRF_M2PL, LOGISTIC_M2PL)
+M2PL_FORMULATION <- latex_eq(IRF_M2PL, IRF_ABBR, LOGISTIC_M2PL)
 
 ### Change of basis in the latent space ----
 
@@ -110,11 +110,43 @@ LS_STD_BASIS_ELEMENTS <- latex_enum(
 LS_STD_BASIS_SET      <- latex_curlybraces(LS_STD_BASIS_ELEMENTS)
 LS_STD_BASIS_EQ       <- latex_eq(LS_STD_BASIS, LS_STD_BASIS_SET)
 
+BASIS_VECTOR_ANY          <- latex_sub(BASIS_VECTOR, DIM_INDEX)
+BASIS_VECTOR_ANY_TRANSF   <- latex_prime(BASIS_VECTOR_ANY)
+TRAIT_MODULE              <- latex("\\theta")
+TRAIT_COMPONENT           <- latex_sub(TRAIT_MODULE, DIM_INDEX)
+TRAIT_VECTOR_DEF_BASIS    <- latex_summation(
+  TRAIT_COMPONENT, BASIS_VECTOR_ANY,
+  index = DIM_INDEX, from = 1, to = N_DIMS,
+  .par = FALSE
+)
+TRAIT_VECTOR_DEF_BASIS_EQ <- latex_eq(TRAIT_VECTOR, TRAIT_VECTOR_DEF_BASIS)
+
+TRAIT_VEC_STD_MOD             <- latex_prime(TRAIT_MODULE)
+TRAIT_VEC_STD_COMP            <- latex_sub(TRAIT_VEC_STD_MOD, DIM_INDEX)
+STD_BASIS_VECTOR_ANY          <- latex_sub(STD_BASIS_VECTOR, DIM_INDEX)
+TRAIT_VECTOR_DEF_STD_BASIS    <- latex_summation(
+  TRAIT_VEC_STD_COMP, STD_BASIS_VECTOR_ANY,
[…]

### 2022-10-27 output/paper_draft/paper_M2PLM_parameters.Rmd
Update with Mario's review, until start of Proposition 1 (WIP)

@@
 In order to contemplate the general case,
-we need to introduce a few linear algebra results before proceeding futher.
+we need to introduce a few linear algebra results before proceeding further.
 
@@
 Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
-TODO: Incomplete in Mario's version, is it correct to say "in the latent space"?
+<!-- TODO: Incomplete in Mario's version, is it correct to say "in the latent space"? -->
 We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
@@
 
-The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$
-
-with $`r TRAIT_VECTOR_STD`$ the coordinates
-of $`r TRAIT_VECTOR`$ in $`r LATENT_SPACE_STD`$ and $`r TRANSFORM_MATRIX_EQ`$
-(being $`r BASIS_VECTOR_ANY_TRANSF`$ the coordinates of
-$`r BASIS_VECTOR_ANY`$ in $`r LATENT_SPACE_STD`$
-for any dimension $`r DIM_INDEX`$).
-This new space can be defined as a Euclidean space with standard inner product
-(i.e. dot product), such that the norm $`r NORM_EQ`$ of $`r TRAIT_VECTOR`$
-in $`r LATENT_SPACE_STD`$ is given by<!-- # TODO: Reference needed? -->
+The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.
+<!-- # TODO: For all k? -->
+If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
 
 \begin{equation}
-  `r TRAIT_MODULE`^2 = `r TRAIT_VECTOR_STD_SQ` =
-    `r TRANSF_TRAIT_SQUARED` =
-    `r TRAIT_TRANSF_TRAIT`.                                 (\#eq:inner-product)
+  `r TRANSFORM_MATRIX_EQ`.                           (\#eq:transform-matrix-def)
 \end{equation}
 
-Thus, the inner product matrix $`r INNER_PROD_MATRIX`$ in $`r LATENT_SPACE`$
-is consistently defined as
+With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product<!-- (i.e. dot product) -->.
+With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
+$`r LS_BASIS_INNER_PROD`$, such that the norm of $`r TRAIT_VECTOR`$ is
+invariant, i.e., $`r TRAIT_NORM_EQ`$. Therefore,
 
 \begin{equation}
-  `r INNER_PROD_MAT_EQ`.                            (\#eq:inner-prod-matrix-def)
+  `r TRAIT_NORM_STD_SQ_EQ`.                             (\#eq:inner-product-def)
 \end{equation}
 
-### Direction cosines
+Thus, we must define the inner product $`r LS_BASIS_INNER_PROD`$ with Gram
+matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$, which keeps the
+norm invariant.
+
+The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
+$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
+invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
 
@@
 
+Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
+up by the $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
+vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.
+
+::: {#prp-cosines-equivalence}
+
+**Proposition 1.** Under the previously defined conditions, for all
+$`r TRAIT_VEC_IN_LS`$,
+
+\begin{equation}
+  `r DIR_COS_VEC_EQ`.                             (\#eq:direction-cosine-vector)
+\end{equation}
+
+with
+
+:::
+
 By Equation \@ref(eq:inner-prod-matrix-def) it is clear that
@@
 (but I “know” intuitively that the generalization holds) -->
-
-\begin{equation}
-  `r DIR_COS_VEC_EQ`.                             (\#eq:direction-cosine-vector)
-\end{equation}
 

### 2022-10-27 R/Formulae.R

@@ -68,14 +68,15 @@ VAR_COV_EQ  <- latex_eq(VAR_ELEMENT, COV_ELEMENT)
 
 ### M2PL model formulation ----
 
-ITEM_INDEX       <- latex('i')
-INTERCEPT_PARAM  <- latex_sub('d', ITEM_INDEX)
-DISCR_VECTOR_ANY <- latex_bf('a')
-DISCR_VECTOR     <- latex_sub(DISCR_VECTOR_ANY, ITEM_INDEX)
-DISCR_PARAM      <- latex_sub('a', "$ITEM_INDEX$$DIM_INDEX$")
-
+# Model parameters:
+ITEM_INDEX          <- latex('i')
+INTERCEPT_PARAM     <- latex_sub('d', ITEM_INDEX)
+DISCR_VECTOR_ANY    <- latex_bf('a')
+DISCR_VECTOR        <- latex_sub(DISCR_VECTOR_ANY, ITEM_INDEX)
+DISCR_PARAM         <- latex_sub('a', "$ITEM_INDEX$$DIM_INDEX$")
 DISCR_VECTOR_TRANSP <- latex_transp(DISCR_VECTOR)
 
+# IRF denotation:
 IRF_M2PL <- latex(
   'P',
   latex_parentheses(
@@ -86,17 +87,18 @@ IRF_M2PL <- latex(
 )
 IRF_ABBR <- latex_sub('P', "$ITEM_INDEX$")
 
+# Model formula:
 LOGIT_M2PL    <- latex_parentheses(
   DISCR_VECTOR_TRANSP, "$TRAIT_VECTOR$ + $INTERCEPT_PARAM$"
 )
-EXPFUNC_M2PL  <- latex_exp(latex_sqbrackets("-$LOGIT_M2PL$"))
-LOGISTIC_M2PL <- latex_logistic(LOGIT_M2PL)
-LOG_INV_M2PL  <- latex_inv_log(LOGIT_M2PL)
-
+EXPFUNC_M2PL     <- latex_exp(latex_sqbrackets("-$LOGIT_M2PL$"))
+LOGISTIC_M2PL    <- latex_logistic(LOGIT_M2PL)
+LOG_INV_M2PL     <- latex_inv_log(LOGIT_M2PL)
 M2PL_FORMULATION <- latex_eq(IRF_M2PL, IRF_ABBR, LOGISTIC_M2PL)
 
[…]

### 2022-10-27 R/LaTeX_math.R

@@ -22,6 +22,12 @@ library(glue)
 
 ## ---- CONSTANTS: -------------------------------------------------------------
 
+EMPTY_STRING      <- ''
+
+TRANSPOSED_SUFFIX <- '^T'
+PRIME_SUFFIX      <- "'"
+EMPTY_OPERAND     <- EMPTY_STRING
+
 DELIMITER  <- '$'
 SEP        <- ' '
 EQ_SIGN    <- ' = '
@@ -29,13 +35,10 @@ EQUIV_SIGN <- ' \\equiv '
 DEF_SIGN   <- ' \\coloneq '
 SIM_SIGN   <- ' \\sim '
 IN_SIGN    <- ' \\in '
-NO_SEP     <- ''
+NO_SEP     <- EMPTY_STRING
 SEP_COMMA  <- ', '
 UNDERSCORE <- '_'
 ELLIPSIS   <- '...'
-
-TRANSPOSED_SUFFIX <- '^T'
-PRIME_SUFFIX      <- "'"
 
 INNER_PROD_LEFT  <- "\\langle"
 INNER_PROD_RIGHT <- "\\rangle"
@@ -184,8 +187,14 @@ latex_seconddiff <- function(fun, par) {
 ### Exponentials and logistics: ----
 
 latex_exp       <- function(x) latex("\\exp{$latex(x)$}")
-latex_log_den   <- function(x) latex("1 + ", latex_exp(latex("\\left[-$x$\\right]")))
+
+latex_log_den   <- function(x) {
+  
+  latex("1 + ", latex_exp(latex("\\left[-$x$\\right]")))
+}
+
 latex_logistic  <- function(x) latex_frac(1, latex_log_den(x))
+
 latex_inv_log   <- function(x) latex_frac(1, latex("1 + ", latex_exp(x)))
 
 
@@ -214,6 +223,7 @@ latex_basis_change <- function(init_basis, end_basis) {
     latex_sub(latex_parentheses('I'), init_basis)
   )
 }
+
 latex_coords <- function(vector, basis) {
   
   COORDINATES_OP <- latex_rm(COORDINATES_OP)
@@ -241,9 +251,15 @@ latex_transp <- function(x) {
 #   latex(x, '^', transp_sym, .sep = NO_SEP)
 # }
 
-latex_innerprod <- function(x, y) {
+latex_innerprod <- function(x     = EMPTY_OPERAND,
+                            y     = EMPTY_OPERAND,
+                            basis = NULL) {
   
[…]

### 2022-11-04 R/Formulae.R
WIP: Inner product matrix diagonal

@@ -247,6 +247,24 @@ INNER_PROD_MATRIX_INV_EQ    <- latex_eq(
   INNER_PROD_MATRIX_INV_DEF
 )
 
+### Proposition 1 ----
+
+BASIS_VECTOR_FIRST_NORM <- latex_norm(BASIS_VECTOR_FIRST)
+BASIS_VECTOR_LAST_NORM  <- latex_norm(BASIS_VECTOR_LAST)
+
+BASIS_VECTOR_FIRST_NORM_SQ <- latex_squared(BASIS_VECTOR_FIRST_NORM)
+BASIS_VECTOR_LAST_NORM_SQ  <- latex_squared(BASIS_VECTOR_LAST_NORM)
+
+DIAG_MATRIX_INNER_PROD     <- latex_bf('D')
+DIAG_MATRIX_INNER_PROD_DEF <- latex_enum_diagmatrix(
+  BASIS_VECTOR_FIRST_NORM_SQ,
+  BASIS_VECTOR_LAST_NORM_SQ
+)
+DIAG_MATRIX_INNER_PROD_EQ <- latex_eq(
+  DIAG_MATRIX_INNER_PROD,
+  DIAG_MATRIX_INNER_PROD_DEF
+)
+
 ### Direction cosines ----
 
 # Cosine of the angle between two verctors:
@@ -290,7 +308,6 @@ DIR_ANGLE_VEC_STD <- latex_prime(DIR_ANGLE_VEC)
 DIR_COS_VEC_STD   <- latex_cos(DIR_ANGLE_VEC_STD)
 
 # Direction cosine vector in original space (proposition):
-DIAG_MATRIX_INNER_PROD        <- latex_bf('D')
 DIAG_MATRIX_INNER_PROD_INV_SQ <- latex_raised_to(
   DIAG_MATRIX_INNER_PROD,
   exp = latex("-$FRAC_1_2$")

### 2022-11-10 output/paper_draft/paper_M2PLM_parameters.Rmd
WIP: Inner product matrix diagonal
Finishes Proposition 1 and proof

@@
 
-::: {#prp-cosines-equivalence}
+::: {#prp-dir-cosines-generalized-basis}
 
@@
 \begin{equation}
-  `r DIR_COS_VEC_EQ`.                             (\#eq:direction-cosine-vector)
+  `r DIR_COS_VEC_EQ`                              (\#eq:direction-cosine-vector)
 \end{equation}
@@
 
+\begin{equation}
+  `r DIAG_MATRIX_INNER_PROD_EQ`.
+\end{equation}
+
+Therefore,
+
+\begin{equation}
+  `r TRAIT_VECTOR_POLAR_EQ`.                           (\#eq:trait-vector-polar)
+\end{equation}
+
+::: {.proof}
+*Proof.*
+
+For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,
+
+\begin{equation}
+  `r DIR_COSINE_PROOF_EQ`,                        (\#eq:dir-cosine-proof-prop-1)
+\end{equation}
+
+with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
+$`r INNER_PROD_MATRIX`$.
+
+Therefore, $`r DIR_COS_VEC_EQ`$ and $`r TRAIT_VECTOR_POLAR_EQ`$.
+$`r END_OF_PROOF`$
 :::
 
-By Equation \@ref(eq:inner-prod-matrix-def) it is clear that
-$`r DIAG_INNER_PROD_MATRIX_EQ`$
-(but note that the basis vectors are not necessarily unitary).
-By defining $`r DIAG_MATRIX_INNER_PROD`$ as
-the diagonal matrix with the same diagonal as $`r INNER_PROD_MATRIX`$,
-Equation \@ref(eq:direction-cosine) in matrix notation is<!-- # TODO:
-Not sure if it’s necessary to explain this further.
-There may be a certain “inferential leap” assuming that D^(-1/2) is
-the matrix form equivalent to 1/u_k,
-(but I “know” intuitively that the generalization holds) -->
+:::
 
@@
 any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
-
-\begin{equation}
-  `r TRAIT_VECTOR_POLAR_EQ`.                           (\#eq:trait-vector-polar)
-\end{equation}
 

### 2022-11-10 R/Formulae.R

@@ -247,24 +247,6 @@ INNER_PROD_MATRIX_INV_EQ    <- latex_eq(
   INNER_PROD_MATRIX_INV_DEF
 )
 
-### Proposition 1 ----
-
-BASIS_VECTOR_FIRST_NORM <- latex_norm(BASIS_VECTOR_FIRST)
-BASIS_VECTOR_LAST_NORM  <- latex_norm(BASIS_VECTOR_LAST)
-
-BASIS_VECTOR_FIRST_NORM_SQ <- latex_squared(BASIS_VECTOR_FIRST_NORM)
-BASIS_VECTOR_LAST_NORM_SQ  <- latex_squared(BASIS_VECTOR_LAST_NORM)
-
-DIAG_MATRIX_INNER_PROD     <- latex_bf('D')
-DIAG_MATRIX_INNER_PROD_DEF <- latex_enum_diagmatrix(
-  BASIS_VECTOR_FIRST_NORM_SQ,
-  BASIS_VECTOR_LAST_NORM_SQ
-)
-DIAG_MATRIX_INNER_PROD_EQ <- latex_eq(
-  DIAG_MATRIX_INNER_PROD,
-  DIAG_MATRIX_INNER_PROD_DEF
-)
-
 ### Direction cosines ----
 
 # Cosine of the angle between two verctors:
@@ -307,7 +289,37 @@ DIR_COS_VEC       <- latex_cos(DIR_ANGLE_VEC)
 DIR_ANGLE_VEC_STD <- latex_prime(DIR_ANGLE_VEC)
 DIR_COS_VEC_STD   <- latex_cos(DIR_ANGLE_VEC_STD)
 
-# Direction cosine vector in original space (proposition):
+## TODO: Delete this block?
+# DIAG_INNER_PROD_MATRIX     <- latex(
+#   latex_rm("diag"),
+#   latex_parentheses(INNER_PROD_MATRIX)
+# )
+# BASIS_VECTOR_FIRST_MOD     <- latex_sub(BASIS_VECTOR_EL, 1)
+# BASIS_VECTOR_LAST_MOD      <- latex_sub(BASIS_VECTOR_EL, N_DIMS)
+# DIAG_INNER_PROD_MATRIX_DEF <- latex_transp(
+#   latex_sqbrackets(
+#     latex_enum(
+#       latex_squared(BASIS_VECTOR_FIRST_MOD),
+#       ELLIPSIS,
[…]

### 2022-11-10 R/LaTeX_math.R

@@ -27,7 +27,6 @@ EMPTY_STRING      <- ''
 
 TRANSPOSED_SUFFIX <- '^T'
 PRIME_SUFFIX      <- "'"
-EMPTY_OPERAND     <- EMPTY_STRING
 
 DELIMITER  <- '$'
 SEP        <- ' '
@@ -61,9 +60,13 @@ ENCLOSING_VALUES <- c(
   "doublepipes"
 )
 
-COLUMN_SEP <- " & "
-LINE_BREAK <- "\\\\\\n"
+COLUMN_SEP    <- " & "
+LINE_BREAK    <- "\\\\"
+EMPTY_OPERAND <- SEP
 
+END_OF_PROOF <- "\\blacksquare"
+# TODO: Maybe change by `\rule{2mm}{2mm}`
+#       (see https://tex.stackexchange.com/a/302260 )
 
 ## ---- FUNCTIONS: -------------------------------------------------------------
 
@@ -239,14 +242,15 @@ latex_basis_change <- function(init_basis, end_basis) {
   )
 }
 
-latex_coords <- function(vector, basis) {
+latex_coords <- function(vector, basis, transpose = FALSE) {
   
   COORDINATES_OP <- latex_rm(COORDINATES_OP)
-  latex(
-    latex_sub(COORDINATES_OP, LS_BASIS, .abbr = TRUE),
-    latex_parentheses(vector, .sep = NO_SEP),
-    .sep = NO_SEP
-  )
+  
+  result <- latex_sub(COORDINATES_OP, LS_BASIS, .abbr = TRUE)
+  
+  if (transpose) result <- latex_transp(result)
+  
+  latex(result, latex_parentheses(vector, .sep = NO_SEP), .sep = NO_SEP)
 }
 
 latex_norm <- function(..., basis = NULL) {
@@ -302,7 +306,7 @@ latex_diagmatrix <- function(elements,
   if (is.list(elements)) {
     
     matrix <- as.list(matrix)
-    matrix[matrix == ''] <- SEP_SPACE
+    matrix[matrix == ''] <- EMPTY_OPERAND
   }
   
   dim(matrix) <- rep(n, 2)

### 2022-11-10 output/paper_draft/paper_M2PLM_parameters.Rmd
Drops the use of `TRAIT_MODULE` (substituted by `TRAIT_NORM`)

@@
 we first compute the point of maximum slope by finding the root(s) of
-the second derivative with respect to $`r TRAIT_MODULE`$.
+the second derivative with respect to $`r TRAIT_NORM`$.
 
@@
 (the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
-which result in improper values of $`r TRAIT_MODULE`$).
+which result in improper values of $`r TRAIT_NORM`$).
 The slope in direction $`r DIR_ANGLE_VEC`$ is given by

### 2022-11-10 R/Formulae.R

@@ -388,8 +388,7 @@ INNER_PROD_INV_DIAG_SQ_PROD <- latex(
   DIAG_MATRIX_INNER_PROD_SQ
 )
 TRAIT_VECTOR_POLAR_COEFF    <- latex(INNER_PROD_INV_DIAG_SQ_PROD, DIR_COS_VEC)
-TRAIT_MODULE                <- TRAIT_SYMBOL
-TRAIT_VECTOR_POLAR_DEF      <- latex(TRAIT_VECTOR_POLAR_COEFF, TRAIT_MODULE)
+TRAIT_VECTOR_POLAR_DEF      <- latex(TRAIT_VECTOR_POLAR_COEFF, TRAIT_NORM)
 TRAIT_VECTOR_POLAR_EQ       <- latex_eq(TRAIT_VECTOR, TRAIT_VECTOR_POLAR_DEF)
 
 LOGIT_M2PL_POLAR     <- latex_parentheses(
@@ -407,7 +406,7 @@ IRF_SQUARED           <- latex_squared(IRF_ABBR)
 DISCR_VEC_TRAIT_COEFF <- latex(DISCR_VECTOR_TRANSP, TRAIT_VECTOR_POLAR_COEFF)
 
 # Second derivative:
-IRF_2ND_DIFF     <- latex_seconddiff(IRF_ABBR, TRAIT_MODULE)
+IRF_2ND_DIFF     <- latex_seconddiff(IRF_ABBR, TRAIT_SYMBOL)
 IRF_2ND_DIFF_DEF <- latex(
   latex_squared(DISCR_VEC_TRAIT_COEFF, .par = TRUE),
   IRF_ABBR,
@@ -418,7 +417,7 @@ IRF_MAX_SLOPE    <- '.5' ## TODO: Format prop-like!!
 IRF_MAX_SLOPE_EQ <- latex_eq(IRF_ABBR, IRF_MAX_SLOPE)
 
 # First derivative:
-IRF_1ST_DIFF     <- latex_firstdiff(IRF_ABBR, TRAIT_MODULE)
+IRF_1ST_DIFF     <- latex_firstdiff(IRF_ABBR, TRAIT_SYMBOL)
 IRF_1ST_DIFF_DEF <- latex(
   IRF_ABBR,
   latex_parentheses("1 - $IRF_ABBR$"),
[…]

### 2022-11-11 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds Proposition 2 and proof

@@
 
-**Proposition 1.** Under the previously defined conditions, for all
+**Proposition 1** Under the previously defined conditions, for all
 $`r TRAIT_VEC_IN_LS`$,
@@
 \end{equation}
+<!-- # TODO: I have changed here $D^{-1/2}$ by $D$; discuss with Mario -->
 
@@
 $`r END_OF_PROOF`$
+:::
+
+:::
+
+We can also obtain a result that relates the direction cosines of any two bases.
+
+::: {#prp-dir-cosines-bases-relationship}
+
+**Proposition 2** Let's consider another basis $`r ALT_BASIS_EQ`$ in
+$`r LATENT_SPACE`$ and the inner product $`r ALT_BASIS_INNER_PROD`$ that keeps
+the norm of the Euclidean space invariant.
+
+Let $`r DIR_COS_VEC_ALT`$ be the vector of direction cosines of
+$`r TRAIT_VECTOR`$ in $`r ALT_BASIS`$ (i.e., the cosines with the elements of
+$`r ALT_BASIS`$).
+
+Then,
+
+\begin{equation}
+  `r DIR_COS_VEC_ALT_EQ`,                      (\#eq:dir-cosines-transformation)
+\end{equation}
+
+where
+
+\begin{equation}
+  `r DIAG_MATRIX_INNER_PROD_ALT_EQ`
+\end{equation}
+
+and $`r TRANSFORM_MATRIX_ALT_EQ`$.
+<!-- TODO: Comment with Mario (order of terms inverted here) -->
+
+::: {.proof}
+*Proof.*
+
+Let's call $`r TRAIT_COORDS_ALT_EQ`$ and let $`r TRANSFORM_MATRIX_ALT_EQ`$
+be the change of basis matrix, such that $`r TRAIT_TRANSFORM_ALT_EQ`$.
+
+Then, we have that $`r BASIS_CHANGE_ORG_ALT_EQ`$.
+
+The inner product $`r ALT_BASIS_INNER_PROD`$ that keeps the norm invariant is
+defined by $`r INNER_PROD_TRAIT_ALT_EQ`$, and we define
+$`r INNER_PROD_MATRIX_ALT_EQ`$.
+
+By proposition 1, for each case we have:
+
+1. $`r TRAIT_VEC_STD_POLAR_EQ`$
+
+1. $`r TRAIT_VECTOR_POLAR_EQ`$
+
+1. $`r TRAIT_VEC_ALT_POLAR_EQ`$
+
+Given (1),
+
+\begin{equation}
+  `r TRAIT_VEC_STD_POLAR_EQ_XPAND`
+\end{equation}
+
+and
+
+\begin{equation}
+  `r TRAIT_VEC_ALT_POLAR_ALT_EQ`.
+\end{equation}
+
+Using (2) and (3) we get that
+
+\begin{equation}
+  `r TRAIT_VEC_STD_POLAR_ORG_EQ`
+\end{equation}
+
+and
+
+\begin{equation}
+  `r TRAIT_VEC_STD_POLAR_ALT_EQ`.
+\end{equation}
+
+Therefore,
+
+\begin{equation}
+  `r TRAIT_VEC_POLAR_ORG_ALT_COEFF_EQ`,
+\end{equation}
+
+and thus $`r DIR_COS_VEC_ALT_EQ`$.$`r END_OF_PROOF`$
+
 :::

### 2022-11-11 R/Formulae.R

@@ -365,7 +365,6 @@ DIR_COS_DEF_COORDS <- latex_frac(
   latex(TRAIT_NORM, BASIS_VECTOR_ANY_NORM)                    # Denominator
 )
 
-
 # Inner product (defined with standard basis vector):
 STD_BASIS_VECTOR_ANY_TRANSP <- latex_transp(STD_BASIS_VECTOR_ANY)
 DIR_COS_DEF_BASIS_VEC       <- latex_frac(
@@ -389,19 +388,171 @@ DIR_COSINE_PROOF_EQ <- latex_eq(
   DIR_COS_DEF_IPM_ROW
 )
 
-### Polar coordinates ----
+### Proposition 2 ----
 
-DIAG_MATRIX_INNER_PROD_SQ   <- latex_raised_to(
-  DIAG_MATRIX_INNER_PROD,
+# New latent space basis:
+ALT_BASIS              <- latex_cal('A')
+ALT_BASIS_VECTOR_EL    <- latex('v')
+ALT_BASIS_VECTOR       <- latex_bf(ALT_BASIS_VECTOR_EL)
+ALT_BASIS_VECTOR_FIRST <- latex_sub(ALT_BASIS_VECTOR, 1)
+ALT_BASIS_VECTOR_LAST  <- latex_sub(ALT_BASIS_VECTOR, N_DIMS)
+ALT_BASIS_ELEMENTS     <- latex_enum(
+  ALT_BASIS_VECTOR_FIRST,
+  ELLIPSIS,
+  ALT_BASIS_VECTOR_LAST
+)
+ALT_BASIS_SET       <- latex_curlybraces(ALT_BASIS_ELEMENTS)
+ALT_BASIS_EQ        <- latex_eq(ALT_BASIS, ALT_BASIS_SET)
+
+# Inner product in new basis :
+ALT_BASIS_INNER_PROD <- latex_innerprod(basis = ALT_BASIS)
+
+# Direction cosine vectors equivalence (new and original bases):
+DIR_ANGLE_VEC_ALT                 <- latex_prime(latex_prime(DIR_ANGLE_VEC))
+DIR_COS_VEC_ALT                   <- latex_cos(DIR_ANGLE_VEC_ALT)
[…]

### 2022-11-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Completes section "Linear algebra of non-orthogonal coordinates"

@@
 \begin{equation}
-  `r DIR_COSINE_PROOF_EQ`,                        (\#eq:dir-cosine-proof-prop-1)
+  `r DIR_COSINE_PROOF_EQ`,
 \end{equation}
@@
 
-From Equation \@ref(eq:direction-cosine-vector),
+From Proposition 1,
 any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
 
+\begin{equation}
+  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
+\end{equation}
+
 Thus, the M2PL model is expressed in polar coordinates as
+<!-- # TODO: Use the "module" symbol instead of the "norm"? -->
 
@@
 the second derivative with respect to $`r TRAIT_NORM`$.
+<!-- # TODO: Use the "module" symbol instead of the "norm"? -->
+
+The slope in direction $`r DIR_ANGLE_VEC`$ is given by
+
+\begin{equation}
+  `r IRF_1ST_DIFF_EQ`,                                      (\#eq:M2PL-1st-diff)
+\end{equation}
+
+and
 
@@
 which result in improper values of $`r TRAIT_NORM`$).
-The slope in direction $`r DIR_ANGLE_VEC`$ is given by
-
-\begin{equation}
-  `r IRF_1ST_DIFF_EQ`.                                      (\#eq:M2PL-1st-diff)
-\end{equation}
-
-When $`r IRF_MAX_SLOPE_EQ`$,
+
+The slope in direction $`r DIR_ANGLE_VEC`$ when $`r IRF_MAX_SLOPE_EQ`$ is given
+by
 
@@
 only in the term $`r INNER_PROD_INV_DIAG_SQ_PROD`$.
-When $`r LS_BASIS`$ is the standard basis,
-this term simplifies to an identity matrix
-(which results in Equation \@ref(eq:M2PL-formula-polar) through
-\@ref(eq:slope-max) simplyfing to rectangular Cartesian coordinates,
-i.e. the equations outlined by Reckase).
+
 To compute the direction of the maximum slope,
@@
 However, this assertion implicitly assumes orthogonal coordinates,
-so it does not apply in general to $`r DIR_COS_VEC`$.<!-- # TODO: To appendix the following, or maybe give for granted? -->
-Instead, we may consider the direction cosine vector in $`r LATENT_SPACE_STD`$:
-
-\begin{equation}
-  `r DIR_COS_VEC_STD_EQ`.                     (\#eq:direction-cosine-vector-std)
-\end{equation}
-
-Operating on Equation \@ref(eq:trait-vector-polar),
+so it does not apply in general to $`r DIR_COS_VEC`$.
+Instead, we may consider the direction cosine vector with the standard basis
+$`r LS_STD_BASIS`$.<!-- # TODO: Mario wrote "\mathcal{B}", I think he refers to "\mathcal{E}", the standard basis -->
+Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain
 
@@
     `r DIR_COS_AS_DIR_COS_STD_EQ` \\
-    `r TRAIT_POLAR_COEFF_COS_STD_EQ`
+    `r TRAIT_POLAR_COEFF_COS_STD_EQ`.
   \end{split}                                 (\#eq:direction-cosine-vectors-eq)
@@
 
-and, substituting in \@ref(eq:slope-max),
+Therefore, substituting in \@ref(eq:slope-max),
 
@@
 
-We can define
-
-\begin{equation}
-  `r DISCR_VECTOR_STD_EQ`.                               (\#eq:discr-vector-std)
-\end{equation}
-
-such that Equation \@ref(eq:slope-max-std-cosines) is similar
-(in vectorial form) to Reckase's [-@reckase_difficulty_1985] Equation 6,
-and thus can be solved by the same procedure
-applying their Equations 7 through 9.
-The resulting equivalent to their Equation 9, in vectorial form
-and solving for $`r DIR_ANGLE_STD_SUBSTITUTE`$ is
-
-\begin{equation}
-  `r DIR_COS_STD_ITEM_VEC_EQ`.                            (\#eq:dir-cos-vec-std)
-\end{equation}
-
-Substituting $`r DIR_COS_STD_ITEM_VEC`$ and $`r DISCR_VECTOR_STD`$
-by their respective expressions results in the expression for
-the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
+We can then define
+
+\begin{equation}
+  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
+\end{equation}
+
+so
+
+\begin{equation}
+  `r SLOPE_MAX_STD_EQ`.                                     (\#eq:slope-max-std)
+\end{equation}
+
+We are thus considering the direction cosines with respect to the orthogonal
+axes, so we can just apply @reckase_difficulty_1985's results, obtaining
+
+\begin{equation}
+  `r DIR_COS_VEC_STD_EQ`.                                 (\#eq:dir-cos-vec-std)
+\end{equation}
+
+Substituting $`r DIR_COS_VEC_STD`$
+(Equation \@ref(eq:direction-cosine-vectors-eq)) and
+$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
+by their respective expressions results in the
+expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
+<!-- # TODO: with basis \mathcal{B}? -->
 
@@
 from the origin to that point,
-we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$,
-$`r DIR_COS_SUBSTITUTE`$ and
-$`r DISTANCE_SUBSTITUTE`$,
-and then substitute Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$.
-The result, solving for $`r DISTANCE_PARAM`$, is
+we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$ to get
+$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
+<!-- # TODO: Verify that it's equation 20, not 18 -->
 
@@
 is given by Equation \@ref(eq:slope-max).
-Solving for $`r DIR_COS_SUBSTITUTE`$
-and substituting Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$
+Substituting Equation \@ref(eq:dir-cos-vec) in Equation \@ref(eq:slope-max),
 [@reckase_discriminating_1991],

### 2022-11-18 R/Formulae.R

@@ -565,19 +565,7 @@ IRF_POLAR_EQ         <- latex_eq(IRF_ABBR, LOGISTIC_M2PL_POLAR)
 ### Point and direction of maximum slope ----
 
 # Terms:
-IRF_SQUARED           <- latex_squared(IRF_ABBR)
 DISCR_VEC_TRAIT_COEFF <- latex(DISCR_VECTOR_TRANSP, TRAIT_VECTOR_POLAR_COEFF)
-
-# Second derivative:
-IRF_2ND_DIFF     <- latex_seconddiff(IRF_ABBR, TRAIT_SYMBOL)
-IRF_2ND_DIFF_DEF <- latex(
-  latex_squared(DISCR_VEC_TRAIT_COEFF, .par = TRUE),
-  IRF_ABBR,
-  latex_parentheses("1 - 3$IRF_ABBR$ + 2$IRF_SQUARED$")
-)
-IRF_2ND_DIFF_EQ  <- latex_eq(IRF_2ND_DIFF, IRF_2ND_DIFF_DEF)
-IRF_MAX_SLOPE    <- '.5' ## TODO: Format prop-like!!
-IRF_MAX_SLOPE_EQ <- latex_eq(IRF_ABBR, IRF_MAX_SLOPE)
 
 # First derivative:
 IRF_1ST_DIFF     <- latex_firstdiff(IRF_ABBR, TRAIT_SYMBOL)
@@ -588,18 +576,24 @@ IRF_1ST_DIFF_DEF <- latex(
 )
 IRF_1ST_DIFF_EQ  <- latex_eq(IRF_1ST_DIFF, IRF_1ST_DIFF_DEF)
 
-# Maximum slope:
-SLOPE_MAX     <- latex_sub("\\left. $IRF_1ST_DIFF$ \\right|", IRF_MAX_SLOPE_EQ)
-SLOPE_MAX_DEF <- latex(FRAC_1_4, DISCR_VEC_TRAIT_COEFF)
-SLOPE_MAX_EQ  <- latex_eq(SLOPE_MAX, SLOPE_MAX_DEF)
+# Second derivative:
+IRF_2ND_DIFF     <- latex_seconddiff(IRF_ABBR, TRAIT_SYMBOL)
+IRF_2ND_DIFF_DEF <- latex(
+  latex_squared(DISCR_VEC_TRAIT_COEFF, .par = TRUE),
+  IRF_ABBR,
+  latex_parentheses("1 -  $IRF_ABBR$"),
[…]

### 2022-11-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Starts test space section (WIP) until definition of the inner product matrix elements.

@@
 Equation \@ref(eq:slope-param) makes it clear that a space $`r TEST_SPACE`$
-with inner product matrix $`r INNER_PROD_MATRIX_INV`$ can be defined,
+with basis $`r ITEM_SPACE_BASIS_EQ`$ and inner product matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined,
 such that $`r INNER_PROD_DISCR_EQ`$ and, subsequently,
@@
 as their inner product matrices are inverse of one another.
+
+With this definition we have that
+
+\begin{equation}
+  \begin{split}
+    `r DIR_COS_ITEM_VEC_NORM_EQ` \\
+    `r DISTANCE_PARAM_NORM_EQ`   \\
+    `r SLOPE_MAX_PARAM_NORM_EQ`.
+  \end{split}                             (\#eq:item-params-discrimination-norm)
+\end{equation}
+<!-- # TODO: Three equations collapsed into one altogether -->
+
+It is clear that $`r INNER_PROD_INV_MAT_ELEMENT_EQ`$ were
+$`r INNER_PROD_INV_MAT_ELEMENT`$ is the $`r INNER_PROD_INV_MAT_SUBINDEX`$-th
+element of $`r INNER_PROD_MATRIX_INV`$.
+<!-- # TODO: Error here? Mario means product of the basis elements? -->
 

### 2022-11-18 R/Formulae.R

@@ -288,27 +288,6 @@ DIR_COS_VEC       <- latex_cos(DIR_ANGLE_VEC)
 DIR_ANGLE_VEC_STD <- latex_prime(DIR_ANGLE_VEC)
 DIR_COS_VEC_STD   <- latex_cos(DIR_ANGLE_VEC_STD)
 
-## TODO: Delete this block?
-# DIAG_INNER_PROD_MATRIX     <- latex(
-#   latex_rm("diag"),
-#   latex_parentheses(INNER_PROD_MATRIX)
-# )
-# BASIS_VECTOR_FIRST_MOD     <- latex_sub(BASIS_VECTOR_EL, 1)
-# BASIS_VECTOR_LAST_MOD      <- latex_sub(BASIS_VECTOR_EL, N_DIMS)
-# DIAG_INNER_PROD_MATRIX_DEF <- latex_transp(
-#   latex_sqbrackets(
-#     latex_enum(
-#       latex_squared(BASIS_VECTOR_FIRST_MOD),
-#       ELLIPSIS,
-#       latex_squared(BASIS_VECTOR_LAST_MOD)
-#     )
-#   )
-# )
-# DIAG_INNER_PROD_MATRIX_EQ  <- latex_eq(
-#   DIAG_INNER_PROD_MATRIX,
-#   DIAG_INNER_PROD_MATRIX_DEF
-# )
-
 ### Proposition 1 ----
 
 # Basis vectors:
@@ -628,16 +607,6 @@ DISCR_VEC_TRAIT_STD_COEFF <- latex(DISCR_VECTOR_STD_TRANSP, DIR_COS_VEC_STD)
 SLOPE_MAX_STD_DEF         <- latex(FRAC_1_4, DISCR_VEC_TRAIT_STD_COEFF)
 SLOPE_MAX_STD_EQ          <- latex_eq(SLOPE_MAX, SLOPE_MAX_STD_DEF)
 
-# Standardized direction cosines:
-## TODO: Review (and possibly delete)
-# DIR_COS_VEC_STD_DEFF     <- latex_frac(TRAIT_VECTOR_STD, TRAIT_NORM)
-# DIR_COS_VEC_STD_DEFF_INV <- latex_frac(TRAIT_TRANSFORM, TRAIT_NORM)
-# DIR_COS_VEC_STD_EQ       <- latex_eq(
-#   DIR_COS_VEC_STD,
[…]

### 2022-11-21 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds change of basis matrix in test space (keeping norm invariant)

@@
 
-The basis $`r ITEM_SPACE_BASIS_EQ`$ of $`r TEST_SPACE`$ is such that
-$`r ITEM_TRANSF_MATRIX_EQ`$ (see Equation \@ref(eq:discr-vector-std)).
-Thus, similarly to equation \@ref(eq:dir-cos-vec),
-the vector of direction cosines of a vector in $`r TEST_SPACE`$
-can be defined as
+Let $`r LS_STD_BASIS`$ be the standard basis of $`r TEST_SPACE`$.
+<!-- # TODO: Can the standard basis of \Alpha be also \mathcal{E}? Otherwise, may I use \mathcal{E} instead? -->
+Then, $`r BASIS_CHANGE_TS_EQ`$ is the change of basis matrix and
+<!-- # TODO: Result pending of demonstration? -->
+$`r DISCR_VECTOR_STD_MOD_EQ`$.
+
+Applying propositon 1,
 

### 2022-11-21 R/Formulae.R
Changes definition of direction angles in test space, using the "norm" of the discrimination vector (instead of the equivalent square root of the inner product)

@@ -757,7 +757,7 @@ DIAG_M_DISCR_PROD                 <- latex(
 )
 DIR_COS_ITEM_VEC_TS_DEF           <- latex_frac(
   DIAG_M_DISCR_PROD,
-  DISCR_VECTOR_MODULE_DEF
+  DISCR_VECTOR_MODULE
 )
 DIR_COS_ITEM_VEC_TS_EQ            <- latex_eq(
   DIR_COS_ITEM_VEC_TS,

### 2022-11-25 output/paper_draft/paper_M2PLM_parameters.Rmd
Changes definition of direction angles in test space, using the "norm" of the discrimination vector (instead of the equivalent square root of the inner product)
Deletes the TODO's in Mario's review (they are intended to be included in the Github review of the PR as comments.

@@
 Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
-<!-- TODO: Incomplete in Mario's version, is it correct to say "in the latent space"? -->
 We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
@@
 coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.
-<!-- # TODO: For all k? -->
 If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
@@
 
-With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product<!-- (i.e. dot product) -->.
+With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product.
 With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
@@
 The cosines between two vectors follow the usual definition,
-<!-- # TODO: Is this true?? See Harman 1970 (eq. 4.43, p. 60 to check) -->
 
@@
 \end{equation}
-<!-- # TODO: I have changed here $D^{-1/2}$ by $D$; discuss with Mario -->
 
@@
 and $`r TRANSFORM_MATRIX_ALT_EQ`$.
-<!-- TODO: Comment with Mario (order of terms inverted here) -->
 
@@
 Thus, the M2PL model is expressed in polar coordinates as
-<!-- # TODO: Use the "module" symbol instead of the "norm"? -->
 
@@
 the second derivative with respect to $`r TRAIT_NORM`$.
-<!-- # TODO: Use the "module" symbol instead of the "norm"? -->
 
@@
 Instead, we may consider the direction cosine vector with the standard basis
-$`r LS_STD_BASIS`$.<!-- # TODO: Mario wrote "\mathcal{B}", I think he refers to "\mathcal{E}", the standard basis -->
+$`r LS_STD_BASIS`$.
 Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain
@@
 expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
-<!-- # TODO: with basis \mathcal{B}? -->
 
@@
 $`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
-<!-- # TODO: Verify that it's equation 20, not 18 -->
 
@@
 \end{equation}
-<!-- # TODO: Three equations collapsed into one altogether -->
 
@@
 element of $`r INNER_PROD_MATRIX_INV`$.
-<!-- # TODO: Error here? Mario means product of the basis elements? -->
 
 Let $`r LS_STD_BASIS`$ be the standard basis of $`r TEST_SPACE`$.
-<!-- # TODO: Can the standard basis of \Alpha be also \mathcal{E}? Otherwise, may I use \mathcal{E} instead? -->
 Then, $`r BASIS_CHANGE_TS_EQ`$ is the change of basis matrix and
-<!-- # TODO: Result pending of demonstration? -->
 $`r DISCR_VECTOR_STD_MOD_EQ`$.

### 2022-12-22 notebooks/Distributions_Carlson_2017.qmd
Updates the loading of the items and starts the simulations

@@
 
-# Introduction
+```{r setup}
+#| message: false
+library(tidyverse)
+library(waldo)
+library(mnormt)
+```
+
+```{r functions}
+get_cov_mat <- function(vals) {
+
+  vals <- vals                 |>
+    pivot_longer(everything()) |>
+    deframe()
+
+  corr_matrix <- diag(2)
+  corr_matrix[1, 2] <- corr_matrix[2, 1] <- vals["r"]
+
+  vars <- vals['sd_' |> paste0(1:2)]
+
+  diag(vars) %*% corr_matrix %*% diag(vars)
+}
+
+get_mean_vec <- function(vals) {
+
+  vals                               |>
+    pivot_longer(everything())       |>
+    filter(name |> str_detect("^m_")) |>
+    deframe()
+}
+
+simulate <- function(params, N = 2000) {
+
+  covs  <- params |> get_cov_mat()
+  means <- params |> get_mean_vec()
+
+  rmnorm(N, mean = means, varcov = covs)
+}
+```
+
+# Item parameters
+
+The item parameters are taken from the Appendix, and are as follows:
+
+```{r table-appendix}
+items_table <- read_delim(
+  "../dat/Table_Appendix_Carlson_2017.csv", 
+  delim         = ";",
+  escape_double = FALSE,
+  trim_ws       = TRUE,
+  col_types     = cols(
+    Item           = col_character(),
+    `Level admin.` = col_character(),
+    .default       = col_double()
+  )
+)
+
+# Set apart items and their descriptives:
+suppressWarnings(
+  items_table <- items_table |> mutate(
+    is_descriptive = Item |>
+      as.numeric()        |>
+      is.na()
+  )
+)
+
+item_parameters <- items_table |>
+  filter(!is_descriptive)      |>
+  select(-is_descriptive)      |>
+  mutate(
+    `Level admin.` = `Level admin.` |>
+      str_split(", ")               |>
+      map(parse_integer)
+  )
+```
+
+We check that the items follow the specifications stated:
+
+```{r check-items}
+item_parameters          |>
+  count(`Level admin.`)  |>
+  unnest(`Level admin.`) |>
+  count(`Level admin.`, wt = n)
+
+item_parameters |>
+  ggplot(aes(b_1, b_2, color = `Level admin.`)) +
+  geom_point()
+```
+
+The number of items per level are as specified (but they are not coincident
+with Table 1; there seems to be a mistake in that table). Also, the graphical
+representation seems to be equivalent to @carlson_unidimensional_2017's Figure
+5.
+
+Now, we check whether the rectangular parameters are properly computed from the
+polar parameters provided. In order to that, we apply Equations 23, 24, and 25
+in @carlson_unidimensional_2017 to compute the rectangular parameters
+(`d`, `b_m` `a_m`, respectively, for $m = {1, 2}$).
+
+```{r check-item-parameters-rectangular}
+computed_item_params <- item_parameters |> transmute(
+  Item,
+  b_1 =  `b*` * cospi(omega_1 / 180),
+  b_2 =  `b*` * sinpi(omega_1 / 180),
+  d   = -`b*` / `a*`,
+  a_1 =  `a*` * cospi(omega_1 / 180),
+  a_2 =  `a*` * sinpi(omega_1 / 180)
+)
+
+item_parameters |> select(Item, b_1:a_2) |>
+  compare(computed_item_params, tolerance = 5e-3)
+```
+
+The values of `d` seem to be miscalculated.
+
+We extract the descriptive statistics of the parameters and compare them with
+the ones provided in the appendix table:
+
+```{r check-item-descriptives}
+items_descriptives <- items_table                                     |>
+  filter(is_descriptive)                                              |>
+  select(stat = Item, everything(), -is_descriptive, -`Level admin.`) |>
+  pivot_longer(-stat, names_to = "parameter")                         |>
+  mutate(across(-value, parse_factor))
+
+# Check item descriptives:
+computed_item_desc <- item_parameters                                      |>
+  summarize(
+    across(omega_1:a_2, .fns = lst(mean, sd, min, max))
+  )                                                                        |>
+  pivot_longer(everything())                                               |>
+  separate(col = name, into = c("parameter", "stat"), sep = "_(?=[^_]+$)") |>
+  mutate(
+    across(-value, parse_factor),
+    stat = stat                    |>
+      fct_relabel(str_to_sentence) |>
+      fct_recode(SD = "Sd")
+  )                                                                        |>
+  select(stat, parameter, value)                                           |>
+  arrange(stat)
+
+compare(items_descriptives, computed_item_desc, tolerance = 5e-3)
+compare(items_descriptives, computed_item_desc, tolerance = 1e-3)
+```
+
+Coincidence is exact (up to the second decimal).
+
+We will compute now the multidimensional parameters using these distribution
+parameters:
+
+```{r simulate}
+params |> slice(1) |> get_cov_mat()
+```
+
+# Latent trait distributions
 
 These distributions are taken from @carlson_unidimensional_2017's Table 2.
+They are all bivariate normal with the following parameters:
 
 ```{r table-2}
-covs <- c(
-  0.496, 0.138, 0.678,
-  0.496, 0.223, 0.870,
-  0.507, 0.328, 0.902,
-  0.295, 0.428, 0.928,
-  0.085, 0.383, 0.808,
-  0.069, 0.546, 0.380
+params <- tribble(
+  ~m_1,   ~m_2,   ~sd_1,  ~sd_2,  ~r,
+  -3.228, -1.093,  0.496,  0.138,  0.678,
+  -1.528, -0.779,  0.496,  0.223,  0.870,
+  -0.265, -0.101,  0.507,  0.328,  0.902,
+   0.598,  0.801,  0.295,  0.428,  0.928,
+   1.100,  1.893,  0.085,  0.383,  0.808,
+   1.330,  3.233,  0.069,  0.546,  0.380
+)
+
+params
+```
+
+# Simulations
+
+2,000 latent trait values are drawn from bivariate normal distributions with the
+parameters given above.
+
+```{r simulate-traits}
+set.seed(1325498132)
+
+sim_traits <- params |>
+  nrow()             |>
+  seq_len()          |>
+  map_df(
+    ~params      |>
+      slice(.x)  |>
+      simulate() |>
+      as_tibble(),
+    .id = "level"
   )                  |>
-  matrix(ncol = 3, byrow = TRUE)
+  mutate(level = level |> parse_integer())
 
-get_corr_mat <- function(vals) {
+sim_traits                                |>
+  mutate(level = level |> as.character()) |>
+  ggplot(aes(V1, V2, color = level)) +
+  geom_point(alpha = .1)
+```
 
-  cov <- vals[3] |> matrix(2, 2)
-  diag(cov) <- vals[1:2]
+The resulting distributions quite resemble @carlson_unidimensional_2017's Figure
+6.
 
-  cov2cor(cov)
-}
+With these values, item responses are first simulated to the items that
+correspond to each level.
 
-get_corr_mat(covs[1, ])
-get_corr_mat(covs[2, ])
-get_corr_mat(covs[3, ])
-get_corr_mat(covs[4, ])
-get_corr_mat(covs[5, ])
-get_corr_mat(covs[6, ])
+```{r simulate-responses}
+sim_responses <- sim_traits |>
+  full_join(
+    item_parameters          |>
+      unnest(`Level admin.`) |>
+      select(Item, `Level admin.`, d:a_2),
+    by = c(level = "Level admin.")
+  )        |>
+  mutate(
+    prob     = V1 * a_1 + V2 * a_2 + d,
+    response = prob |> rbinom(n = n(), size = 1)
+  )
 ```

### 2022-12-22 notebooks/Distributions_Carlson_2017.qmd
Corrects the computation of the probability for the simulations

@@
 
-```{r simulate}
-params |> slice(1) |> get_cov_mat()
-```
-
 # Latent trait distributions
@@
   mutate(
-    prob     = V1 * a_1 + V2 * a_2 + d,
+    prob     = (1 + exp(- (V1 * a_1 + V2 * a_2 + d)))^(-1),
     response = prob |> rbinom(n = n(), size = 1)

### 2023-01-13 notebooks/Distributions_Carlson_2017.qmd
Updates use of pipe to comply with style guide and make the response simulation easier to understand

@@
 ```{r simulate-responses}
+item_params_by_level <- item_parameters |>
+      unnest(`Level admin.`)            |>
+      select(Item, level = `Level admin.`, d:a_2)
+
 sim_responses <- sim_traits                     |>
-  full_join(
-    item_parameters          |>
-      unnest(`Level admin.`) |>
-      select(Item, `Level admin.`, d:a_2),
-    by = c(level = "Level admin.")
-  )        |>
+  full_join(item_params_by_level, by = "level") |>
   mutate(

### 2023-01-13 output/paper_draft/paper_M2PLM_parameters.Rmd
Changes knit hooks to new version compatible with Quarto

@@
 
-```{r setup, include=FALSE}
+```{r setup}
+#| include: false
+
 # Includes:
@@
 
-```{r write-bib, cache=TRUE}
+```{r write-bib}
+#| cache: true
+
 # automatically create a bib database for R packages
@@
 
-```{r sources, cache=FALSE, include=FALSE}
+```{r sources}
+#| cache:   false
+#| include: false
 source("R/Formulae.R", encoding = 'UTF-8')
@@
 Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$,
-defined by vector space $V$ and basis $`r LS_BASIS_EQ`$.
+defined by vector space $V$ and basis $`r LS_BASIS`$,
+being $`r LS_BASIS_EQ`$ an arbitrary set of linearly independent vectors.
 A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
+with coordinates $`r TRAIT_VECTOR`$ in $`r LATENT_SPACE`$,
 which is assumed to be a multivariate normal random variable:
@@
 
-According to the M2PL model, the probability of a positive response to item
-$`r ITEM_INDEX`$ is
+As in @reckase_difficulty_1985, the characteristic function of item
+$`r ITEM_INDEX`$ according to the M2PL model is formulated as follows:
 
@@
 
-where $`r DISCR_VECTOR`$ is a $`r N_DIMS`$-dimensional vector of discrimination
-parameters, and $`r INTERCEPT_PARAM`$ an intercept parameter related to the
-location of the item in the latent space.
+being  
+$`r IRF_M2PL`$ ($= `r IRF_ABBR`$ henceforth, for simplicity) the probability of
+a positive response to item $`r ITEM_INDEX`$ for a given $`r TRAIT_VECTOR`$,  
+$`r DISCR_VECTOR`$ a $`r N_DIMS`$-dimensional vector of
+discrimination parameters, and  
+$`r INTERCEPT_PARAM`$ an intercept parameter related to the location of the item
+in the latent space.
 
@@
 In order to contemplate the general case,
-we need to introduce a few linear algebra results before proceeding further.
+we need to introduce a few linear algebra results before proceeding futher.
 
 ## Linear algebra of non-orthogonal coordinates
+<!-- # TODO: To appendix, OR citation about these? -->
 
-Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
-We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
-Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
-$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$, where $`r TRAIT_COORDS_EQ`$ and
-$`r TRAIT_COORDS_STD_EQ`$.
+### Inner product
 
-Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix; then, we have that
+Let $`r INNER_PROD_EQ`$ be the inner product in $`r LATENT_SPACE`$, with
+the positive-definite matrix $`r INNER_PROD_MATRIX`$ its inner product matrix.
+We may apply a change of basis in $`r LATENT_SPACE`$ that results in a space
+$`r LATENT_SPACE_STD`$ with the standard basis $`r LS_STD_BASIS_EQ`$.
+Such a change of basis can be defined as
 
 \begin{equation}
-  `r BASIS_CHANGE`.                                          (\#eq:change-basis)
+  `r BASIS_CHANGE`,                                          (\#eq:change-basis)
 \end{equation}
 
-The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
-coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.
-If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
+with $`r TRAIT_VECTOR_STD`$ the coordinates
+of $`r TRAIT_VECTOR`$ in $`r LATENT_SPACE_STD`$ and $`r TRANSFORM_MATRIX_EQ`$
+(being $`r BASIS_VECTOR_ANY_TRANSF`$ the coordinates of
+$`r BASIS_VECTOR_ANY`$ in $`r LATENT_SPACE_STD`$
+for any dimension $`r DIM_INDEX`$).
+This new space can be defined as a Euclidean space with standard inner product
+(i.e. dot product), such that the norm $`r NORM_EQ`$ of $`r TRAIT_VECTOR`$
+in $`r LATENT_SPACE_STD`$ is given by<!-- # TODO: Reference needed? -->
 
 \begin{equation}
-  `r TRANSFORM_MATRIX_EQ`.                           (\#eq:transform-matrix-def)
+  `r TRAIT_MODULE`^2 = `r TRAIT_VECTOR_STD_SQ` =
+    `r TRANSF_TRAIT_SQUARED` =
+    `r TRAIT_TRANSF_TRAIT`.                                 (\#eq:inner-product)
 \end{equation}
 
-With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product.
-With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
-$`r LS_BASIS_INNER_PROD`$, such that the norm of $`r TRAIT_VECTOR`$ is
-invariant, i.e., $`r TRAIT_NORM_EQ`$. Therefore,
+Thus, the inner product matrix $`r INNER_PROD_MATRIX`$ in $`r LATENT_SPACE`$
+is consistently defined as
 
 \begin{equation}
-  `r TRAIT_NORM_STD_SQ_EQ`.                             (\#eq:inner-product-def)
+  `r INNER_PROD_MAT_EQ`.                            (\#eq:inner-prod-matrix-def)
 \end{equation}
 
-Thus, we must define the inner product $`r LS_BASIS_INNER_PROD`$ with Gram
-matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$, which keeps the
-norm invariant.
-
-The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
-$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
-invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
+### Direction cosines
 
 The cosines between two vectors follow the usual definition,
+<!-- # TODO: Is this true?? See Harman 1970 (eq. 4.43, p. 60 to check) -->
 
@@
 
-Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
-up by the $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
-vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.
-
-::: {#prp-dir-cosines-generalized-basis}
-
-**Proposition 1** Under the previously defined conditions, for all
-$`r TRAIT_VEC_IN_LS`$,
+By Equation \@ref(eq:inner-prod-matrix-def) it is clear that
+$`r DIAG_INNER_PROD_MATRIX_EQ`$
+(but note that the basis vectors are not necessarily unitary).
+By defining $`r DIAG_MATRIX_INNER_PROD`$ as
+the diagonal matrix with the same diagonal as $`r INNER_PROD_MATRIX`$,
+Equation \@ref(eq:direction-cosine) in matrix notation is<!-- # TODO:
+Not sure if it’s necessary to explain this further.
+There may be a certain “inferential leap” assuming that D^(-1/2) is
+the matrix form equivalent to 1/u_k,
+(but I “know” intuitively that the generalization holds) -->
 
 \begin{equation}
-  `r DIR_COS_VEC_EQ`                              (\#eq:direction-cosine-vector)
+  `r DIR_COS_VEC_EQ`.                             (\#eq:direction-cosine-vector)
 \end{equation}
 
-with
+## Model recasting into polar coordinates
 
-\begin{equation}
-  `r DIAG_MATRIX_INNER_PROD_EQ`.
-\end{equation}
-
-Therefore,
+From Equation \@ref(eq:direction-cosine-vector),
+any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
 
@@
   `r TRAIT_VECTOR_POLAR_EQ`.                           (\#eq:trait-vector-polar)
-\end{equation}
-
-::: {.proof}
-*Proof.*
-
-For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,
-
-\begin{equation}
-  `r DIR_COSINE_PROOF_EQ`,
-\end{equation}
-
-with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
-$`r INNER_PROD_MATRIX`$.
-
-Therefore, $`r DIR_COS_VEC_EQ`$ and $`r TRAIT_VECTOR_POLAR_EQ`$.
-$`r END_OF_PROOF`$
-:::
-
-:::
-
-We can also obtain a result that relates the direction cosines of any two bases.
-
-::: {#prp-dir-cosines-bases-relationship}
-
-**Proposition 2** Let's consider another basis $`r ALT_BASIS_EQ`$ in
-$`r LATENT_SPACE`$ and the inner product $`r ALT_BASIS_INNER_PROD`$ that keeps
-the norm of the Euclidean space invariant.
-
-Let $`r DIR_COS_VEC_ALT`$ be the vector of direction cosines of
-$`r TRAIT_VECTOR`$ in $`r ALT_BASIS`$ (i.e., the cosines with the elements of
-$`r ALT_BASIS`$).
-
-Then,
-
-\begin{equation}
-  `r DIR_COS_VEC_ALT_EQ`,                      (\#eq:dir-cosines-transformation)
-\end{equation}
-
-where
-
-\begin{equation}
-  `r DIAG_MATRIX_INNER_PROD_ALT_EQ`
-\end{equation}
-
-and $`r TRANSFORM_MATRIX_ALT_EQ`$.
-
-::: {.proof}
-*Proof.*
-
-Let's call $`r TRAIT_COORDS_ALT_EQ`$ and let $`r TRANSFORM_MATRIX_ALT_EQ`$
-be the change of basis matrix, such that $`r TRAIT_TRANSFORM_ALT_EQ`$.
-
-Then, we have that $`r BASIS_CHANGE_ORG_ALT_EQ`$.
-
-The inner product $`r ALT_BASIS_INNER_PROD`$ that keeps the norm invariant is
-defined by $`r INNER_PROD_TRAIT_ALT_EQ`$, and we define
-$`r INNER_PROD_MATRIX_ALT_EQ`$.
-
-By proposition 1, for each case we have:
-
-1. $`r TRAIT_VEC_STD_POLAR_EQ`$
-
-1. $`r TRAIT_VECTOR_POLAR_EQ`$
-
-1. $`r TRAIT_VEC_ALT_POLAR_EQ`$
-
-Given (1),
-
-\begin{equation}
-  `r TRAIT_VEC_STD_POLAR_EQ_XPAND`
-\end{equation}
-
-and
-
-\begin{equation}
-  `r TRAIT_VEC_ALT_POLAR_ALT_EQ`.
-\end{equation}
-
-Using (2) and (3) we get that
-
-\begin{equation}
-  `r TRAIT_VEC_STD_POLAR_ORG_EQ`
-\end{equation}
-
-and
-
-\begin{equation}
-  `r TRAIT_VEC_STD_POLAR_ALT_EQ`.
-\end{equation}
-
-Therefore,
-
-\begin{equation}
-  `r TRAIT_VEC_POLAR_ORG_ALT_COEFF_EQ`,
-\end{equation}
-
-and thus $`r DIR_COS_VEC_ALT_EQ`$.$`r END_OF_PROOF`$
-
-:::
-
-:::
-
-## Model recasting into polar coordinates
-
-From Proposition 1,
-any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
-
-\begin{equation}
-  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
 \end{equation}
@@
 we first compute the point of maximum slope by finding the root(s) of
-the second derivative with respect to $`r TRAIT_NORM`$.
-
-The slope in direction $`r DIR_ANGLE_VEC`$ is given by
-
-\begin{equation}
-  `r IRF_1ST_DIFF_EQ`,                                      (\#eq:M2PL-1st-diff)
-\end{equation}
-
-and
+the second derivative with respect to $`r TRAIT_MODULE`$.
 
@@
 (the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
-which result in improper values of $`r TRAIT_NORM`$).
+which result in improper values of $`r TRAIT_MODULE`$).
+The slope in direction $`r DIR_ANGLE_VEC`$ is given by
 
-The slope in direction $`r DIR_ANGLE_VEC`$ when $`r IRF_MAX_SLOPE_EQ`$ is given
-by
+\begin{equation}
+  `r IRF_1ST_DIFF_EQ`.                                      (\#eq:M2PL-1st-diff)
+\end{equation}
+
+When $`r IRF_MAX_SLOPE_EQ`$,
 
@@
 only in the term $`r INNER_PROD_INV_DIAG_SQ_PROD`$.
-
+When $`r LS_BASIS`$ is the standard basis,
+this term simplifies to an identity matrix
+(which results in Equation \@ref(eq:M2PL-formula-polar) through
+\@ref(eq:slope-max) simplyfing to rectangular Cartesian coordinates,
+i.e. the equations outlined by Reckase).
 To compute the direction of the maximum slope,
@@
 However, this assertion implicitly assumes orthogonal coordinates,
-so it does not apply in general to $`r DIR_COS_VEC`$.
-Instead, we may consider the direction cosine vector with the standard basis
-$`r LS_STD_BASIS`$.
-Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain
+so it does not apply in general to $`r DIR_COS_VEC`$.<!-- # TODO: To appendix the following, or maybe give for granted? -->
+Instead, we may consider the direction cosine vector in $`r LATENT_SPACE_STD`$:
+
+\begin{equation}
+  `r DIR_COS_VEC_STD_EQ`.                     (\#eq:direction-cosine-vector-std)
+\end{equation}
+
+Operating on Equation \@ref(eq:trait-vector-polar),
 
@@
     `r DIR_COS_AS_DIR_COS_STD_EQ` \\
-    `r TRAIT_POLAR_COEFF_COS_STD_EQ`.
+    `r TRAIT_POLAR_COEFF_COS_STD_EQ`
   \end{split}                                 (\#eq:direction-cosine-vectors-eq)
@@
 
-Therefore, substituting in \@ref(eq:slope-max),
+and, substituting in \@ref(eq:slope-max),
 
@@
 
-We can then define
+We can define
 
 \begin{equation}
-  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
+  `r DISCR_VECTOR_STD_EQ`.                               (\#eq:discr-vector-std)
 \end{equation}
 
-so
+such that Equation \@ref(eq:slope-max-std-cosines) is similar
+(in vectorial form) to Reckase's [-@reckase_difficulty_1985] Equation 6,
+and thus can be solved by the same procedure
+applying their Equations 7 through 9.
+The resulting equivalent to their Equation 9, in vectorial form
+and solving for $`r DIR_ANGLE_STD_SUBSTITUTE`$ is
 
 \begin{equation}
-  `r SLOPE_MAX_STD_EQ`.                                     (\#eq:slope-max-std)
+  `r DIR_COS_STD_ITEM_VEC_EQ`.                            (\#eq:dir-cos-vec-std)
 \end{equation}
 
-We are thus considering the direction cosines with respect to the orthogonal
-axes, so we can just apply @reckase_difficulty_1985's results, obtaining
-
-\begin{equation}
-  `r DIR_COS_VEC_STD_EQ`.                                 (\#eq:dir-cos-vec-std)
-\end{equation}
-
-Substituting $`r DIR_COS_VEC_STD`$
-(Equation \@ref(eq:direction-cosine-vectors-eq)) and
-$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
-by their respective expressions results in the
-expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
+Substituting $`r DIR_COS_STD_ITEM_VEC`$ and $`r DISCR_VECTOR_STD`$
+by their respective expressions results in the expression for
+the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
 
@@
 from the origin to that point,
-we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$ to get
-$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
+we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$,
+$`r DIR_COS_SUBSTITUTE`$ and
+$`r DISTANCE_SUBSTITUTE`$,
+and then substitute Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$.
+The result, solving for $`r DISTANCE_PARAM`$, is
 
@@
 is given by Equation \@ref(eq:slope-max).
-Substituting Equation \@ref(eq:dir-cos-vec) in Equation \@ref(eq:slope-max),
+Solving for $`r DIR_COS_SUBSTITUTE`$
+and substituting Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$
 [@reckase_discriminating_1991],
@@
 Equation \@ref(eq:slope-param) makes it clear that a space $`r TEST_SPACE`$
-with basis $`r ITEM_SPACE_BASIS_EQ`$ and inner product matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined,
+with inner product matrix $`r INNER_PROD_MATRIX_INV`$ can be defined,
 such that $`r INNER_PROD_DISCR_EQ`$ and, subsequently,
@@
 
-With this definition we have that
-
-\begin{equation}
-  \begin{split}
-    `r DIR_COS_ITEM_VEC_NORM_EQ` \\
-    `r DISTANCE_PARAM_NORM_EQ`   \\
-    `r SLOPE_MAX_PARAM_NORM_EQ`.
-  \end{split}                             (\#eq:item-params-discrimination-norm)
-\end{equation}
-
-It is clear that $`r INNER_PROD_INV_MAT_ELEMENT_EQ`$ were
-$`r INNER_PROD_INV_MAT_ELEMENT`$ is the $`r INNER_PROD_INV_MAT_SUBINDEX`$-th
-element of $`r INNER_PROD_MATRIX_INV`$.
-
-Let $`r LS_STD_BASIS`$ be the standard basis of $`r TEST_SPACE`$.
-Then, $`r BASIS_CHANGE_TS_EQ`$ is the change of basis matrix and
-$`r DISCR_VECTOR_STD_MOD_EQ`$.
-
-Applying propositon 1,
+The basis $`r ITEM_SPACE_BASIS_EQ`$ of $`r TEST_SPACE`$ is such that
+$`r ITEM_TRANSF_MATRIX_EQ`$ (see Equation \@ref(eq:discr-vector-std)).
+Thus, similarly to equation \@ref(eq:dir-cos-vec),
+the vector of direction cosines of a vector in $`r TEST_SPACE`$
+can be defined as
 
@@
 
-```{r libraries, message=FALSE}
+```{r libraries}
+#| message: false
 ```
 
-```{r sources, message=FALSE}
+```{r sources}
+#| message: false
 ```
@@
 
-```{r set-flextable-wd, cache=FALSE}
+```{r set-flextable-wd}
+#| cache: false
+
 # Necessary for officedown to find the template file:
@@
 
-```{r example-items-table-out, cache=FALSE, tab.id="example-items-table-out", tab.cap="Item parameters for the graphical example"}
+```{r example-items-table-out}
+#| cache:   false
+#| tab_id:  example-items-table-out
+#| tab.cap: Item parameters for the graphical example
 item_params_output
@@
 
-```{r item-plot-out, cache=FALSE, fig.cap="Item plots in an orthogonal (left) and oblique (right) test space.", fig.height=3.1, fig.width=6.65}
+```{r item-plot-out}
+#| cache:      false
+#| fig.cap:   Item plots in an orthogonal (left) and oblique (right) test space.
+#| fig.height: 3.1
+#| fig.width:  6.65
 plot_orth + plot_oblique
@@
 ```{r read-chunks-empirical-example}
-read_chunk("src/Empirical_example_Tezza2018.R")
+read_chunk("src/Empirical_example.R")
 ```
@@
 ```{r set-flextable-wd, cache=FALSE}
+```{r set-flextable-wd}
+#| cache: false
 ```
@@
 
-```{r empirical-items-table-out, cache=FALSE, tab.id="empirical-items-table-out", tab.cap="Item parameters for the empirical example"}
+```{r empirical-items-table-out}
+#| cache:   false
+#| tab.id:  empirical-items-table-out
+#| tab.cap: Item parameters for the empirical example
 item_params_output

### 2023-01-13 src/Empirical_example_Reckase2009.R
Includes new version of the first empirical example table

@@ -28,7 +28,15 @@ source("R/Mirt_toolbox.R")
 ## ---- CONSTANTS: -------------------------------------------------------------
 
 ## ----constants----
-
+COV_MATRIX <- matrix(
+  c(
+    1.210, .297, 1.232,
+     .297, .810,  .252,
+    1.232, .252, 1.960
+  ),
+  nrow = 3
+)
+# Version in Reckase, 2009, p. 153 (see also p. 183 for the close to 1D case)
 
 ## ---- CONFIGURATION: ---------------------------------------------------------
 
@@ -60,32 +68,10 @@ items                                            |>
 
 ## ----oblique-params----
 
-# ## Compute the covariance matrix:
-
-# cov_matrix <- matrix(
-#   c(
-#     .1670, .2361, .2892,
-#     .2361, .3340, .4090,
-#     .2892, .4090, .5010
-#   ),
-#   nrow = 3
-# )
-# # See Reckase, 2009, p. 183
-cov_matrix <- matrix(
-  c(
-    1.210, .297, 1.232,
-     .297, .810,  .252,
-    1.232, .252, 1.960
-  ),
-  nrow = 3
-)
-# See Reckase, 2009, p. 183
-
-
 ## Compute the parameters:
 items_oblique <- items |> compute_mirt_params(
   d, matches('^a\\d$'),
-  cov_matrix = cov_matrix,
+  cov_matrix = COV_MATRIX,
   dir_out = "deg"
 )
 

### 2023-01-13 output/paper_draft/paper_M2PLM_parameters.Rmd
Includes new version of the first empirical example table
Ignores messages in chunk outputs

@@
   echo     = FALSE,
+  message  = FALSE,
   cache    = FALSE,
@@
 ```{r libraries}
-#| message: false
 ```
@@
 ```{r sources}
-#| message: false
 ```
@@
 ```{r read-chunks-empirical-example}
-read_chunk("src/Empirical_example.R")
+source("src/Empirical_example_Reckase2009.R", encoding = 'UTF-8')
 ```
 
-```{r libraries, message=FALSE}
-```
-
-```{r sources, message=FALSE}
-```
-
-```{r read-items, message=FALSE}
-```
-
-```{r orthogonal-params}
-```
-
-```{r oblique-params}
-```
-
-```{r params-table}
-```
-
-```{r compose-empirical-items-table}
-```
-
-```{r set-flextable-wd, cache=FALSE}
 ```{r set-flextable-wd}

### 2023-01-13 src/Empirical_example_Reckase2009.R
Reformats tables according to APA format

@@ -17,6 +17,7 @@
 ## ----libraries----
 library(tidyverse)
 library(flextable)
+library(officer)
 library(assertive.numbers)
 
 ## ----sources----
@@ -67,7 +68,8 @@ items                                            |>
   pull(MDIFF)                                    |>
   is_equal_to(items_orth |> pull(D), tol = 5e-3) |>
   all()
-
+# Both multidimensional parameters are equal for all items, up to the tolerance
+#   level.
 
 ## ----oblique-params----
 
@@ -103,46 +105,41 @@ item_params <- item_params                   |>
 
 # Table header (for flextable object):
 item_headers <- tibble(
-  col_keys = item_params |> colnames(),
-  metric   = col_keys %>% {
+  col_keys     = item_params |> colnames(),
+  metric       = col_keys %>% {
     case_when(
-      . == "item"                         ~ "Item",
+                 . == "item"              ~ "Item",
+                 . == 'l'                 ~ "M2PL",
       str_detect(., DISCR_PARAMS_PATTERN) ~ "M2PL",
-      . == 'l'                            ~ "M2PL",
       str_detect(., "sep1")               ~ ' ',
       TRUE                                ~ "Multidimensional parameters"
     )
   },
-  param    = col_keys %>% {
+  metric_param = col_keys %>% {
     case_when(
-      . == "item"                         ~ "Item",
-      str_detect(., DISCR_PARAMS_PATTERN) ~ "a_i",
-      . == 'l'                            ~ "l_i",
-      str_detect(., "sep\\d+")            ~ ' ',
-      str_detect(., "MDISC_")             ~ "MDISC_i",
-      str_detect(., "D_")                 ~ "D_i",
-      str_detect(., "deg_")               ~ "\\alpha_i",
+      str_detect(., "MDISC_") ~ "MDISC_i",
+      str_detect(., "D_")     ~ "D_i",
+      str_detect(., "deg_")   ~ paste0(
+                                  "\\alpha_{i",
+                                  str_extract(., "\\d+"),
+                                  '}'
+                                ),
+      str_detect(., "sep")    ~ ' ',
+      TRUE                    ~ metric
     )
   },
-  dim      = col_keys %>% {
+  param_space  = col_keys %>% {
     case_when(
-      . == "item"                         ~ "Item",
-      . == 'l'                            ~ "l_i",
-      str_detect(., DISCR_PARAMS_PATTERN) ~ "(dimension)",
[…]

### 2023-01-20 output/paper_draft/paper_M2PLM_parameters.Rmd
Reformats tables according to APA format
Corrects erratum
Updates table outputs of the graphical and the empirical examples, in order to make the code more generic, necessary for both empirical example outputs (WIP)

@@
 library(patchwork)
+library(rstudioapi)
 
-ROOT_DIR <- "../.."
+ROOT_DIR <- getActiveProject()
 DOC_DIR  <- getwd()
@@
   dpi      =   300,
-  dev.args = list(type = "cairo-png"),
-  fig.pos  = "!H"
+  dev.args = list(type = "cairo-png")
 )
@@
 source("R/Formulae.R",  encoding = 'UTF-8')
+source("R/Constants.R", encoding = 'UTF-8')
 ```
@@
 The so called *Multidimensional 2-Parameter Logistic*
-[M2PL, @mckinley_extension_1983] model is one of the most important and
-widely used multidimensional models,
+[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is one of the most important
+and widely used multidimensional models,
 partly due to its close relationship to the common factor model
 [@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
-The M2PL model is a generalization of the 2-parameter logistic model
-[@birnbaum_latent_1968] to more than one latent trait
-[@mckinley_extension_1983].
+The `r MODEL_ACRONYM` model is a generalization of
+the 2-parameter logistic model [@birnbaum_latent_1968]
+to more than one latent trait [@mckinley_extension_1983].
 Its usefulness lies in describing the variability of the population when
@@
 Following this procedure, he also derives
-a *multidimensional difficulty* parameter (MID) for M2PL items.
+a *multidimensional difficulty* parameter (MID) for `r MODEL_ACRONYM` items.
 This parameter consists of two parts:
@@
 In order to pursue these objectives, we first define the generalization of
-the M2PL model and describe its scope of application.
+the `r MODEL_ACRONYM` model and describe its scope of application.
 Then, we follow the method outlined by @reckase_difficulty_1985 and
@@
 As in @reckase_difficulty_1985, the characteristic function of item
-$`r ITEM_INDEX`$ according to the M2PL model is formulated as follows:
+$`r ITEM_INDEX`$ according to the `r MODEL_ACRONYM` model is formulated
+as follows:
 
@@
 
-Thus, the M2PL model is expressed in polar coordinates as
+Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
 
@@
 \begin{equation}
-  `r DISCR_VECTOR_STD_EQ`.                               (\#eq:discr-vector-std)
+  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
 \end{equation}
@@
 from the origin to the point of maximum slope [@reckase_difficulty_1985].
-By analogy with the unidimensional case, $`r MDISC_PARAM`$ is defined as
+By analogy with the unidimensional case, $`r MDISC_SYM`$ is defined as
 $4 `r SLOPE_MAX_PARAM`$ [@reckase_discriminating_1991].
-Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_PARAM`$ using
+Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
 equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
@@
 
-3. $`r MDISC_PARAM`$ is four times the maximum slope $`r SLOPE_MAX_PARAM`$.
+3. $`r MDISC_SYM`$ is four times the maximum slope $`r SLOPE_MAX_PARAM`$.
 
@@
 Equation \@ref(eq:distance-param),
-and the third one is implicit in the definition of the $`r MDISC_PARAM`$.
+and the third one is implicit in the definition of the $`r MDISC_SYM`$.
 Simple arithmetic can show that the first holds true whenever the
@@
 By the definition of the inner product,
-$`r MDISC_PARAM`$ is strictly positive for non-null vectors.
+$`r MDISC_SYM`$ is strictly positive for non-null vectors.
 Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
@@
 The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
-also due to the strict positivity of $`r MDISC_PARAM`$,
+also due to the strict positivity of $`r MDISC_SYM`$,
 are equal to the signs of $`r DISCR_VECTOR`$.
@@
 from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
-and will have a length equal to its $`r MDISC_PARAM`$ parameter.
+and will have a length equal to its $`r MDISC_SYM`$ parameter.
 
@@
 with $`r TRANSFORM_MATRIX`$ being an identity matrix.
-In the second one, the correlation between the dimensions is $`r corr_out`$.
+In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
 In order to keep the horizontal axis invariant,
@@
 
-The sample items have the same M2PL parameters in both latent spaces.
+The sample items have the same `r MODEL_ACRONYM` parameters
+in both latent spaces.
 However, their multidimensional parameters differ due to the correlation among
@@
 The parameters in both spaces,
-along with their corresponding M2PL parameters,
+along with their corresponding `r MODEL_ACRONYM` parameters,
 are shown in Table \@ref(tab:example-items-table-out).
@@
 
-## Example with empirical data
+## Examples with empirical data
 
 ```{r reset-root-wd}
+#| cache: false
+
 # Necessary for knitr to find the source file:
@@
 ```{r read-chunks-empirical-example}
-source("src/Empirical_example_Reckase2009.R", encoding = 'UTF-8')
+read_chunk("src/Empirical_example.R")
+```
+
+```{r libraries}
+```
+
+```{r sources}
+```
+
+```{r general-constants}
+```
+
+```{r set-cov-matrix-Reckase2009}
+# Version in Reckase, 2009, p. 153; this represents a usual 3D case
+COV_MATRIX <- matrix(
+  c(
+    1.210, .297, 1.232,
+     .297, .810,  .252,
+    1.232, .252, 1.960
+  ),
+  nrow = 3
+)
+# (see also Reckase, 2009, p. 183 for a "close-to-1D" case).
+```
+
+```{r read-items}
+```
+
+```{r compute-orthogonal-params}
+```
+
+```{r compute-oblique-params}
+```
+
+```{r create-params-table}
+```
+
+```{r compose-output-table}
+```
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Reckase2009-out}
+#| cache:   false
+#| tab.id:  empirical-table-Reckase2009-out
+#| tab.cap: Item parameters for the empirical example
+item_params_output
+```
+
+```{r set-cov-matrix-Tezza2018}
+# See Tezza et al., 2018, p. 927:
+# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
+# > foi de aproximadamente 0,4.
+COV_MATRIX <- diag(4)
+COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
+```
+
+```{r reset-root-wd}
+#| cache: false
+```
+
+```{r read-items-Tezza2018}
+items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
+```
+
+```{r compute-orthogonal-params}
+```
+
+```{r compare-orthogonal-params}
+```
+
+<!-- Computed item parameters in the orthonormal space match the ones in given
+in Tezza 2018, table 5 -->
+
+```{r compute-oblique-params}
+```
+
+```{r create-params-table}
+```
+
+```{r compose-output-table}
 ```
@@
 
-```{r empirical-items-table-out}
+```{r empirical-table-Tezza2018-out}
 #| cache:   false
-#| tab.id:  empirical-items-table-out
+#| tab.id:  empirical-table-Tezza2018-out
 #| tab.cap: Item parameters for the empirical example
@@
 <!---BLOCK_LANDSCAPE_STOP--->
-
-<!-- # TODO: Complete? -->
 
@@
 It is worth noting that @zhang_theoretical_1999 already proposed a formulation
-of the $`r MDISC_PARAM`$ equivalent to our covariance-based version,
+of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
 but for a more general, semi-parametric formulation
@@
 Our results provide this formal derivation, although only for the specific
-case of the M2PL model.
+case of the `r MODEL_ACRONYM` model.
 
@@
 the multidimensional parameters, and refer to them simply as
-the $`r MDISC_PARAM`$ and $`r MIL_PARAM`$ parameters.
+the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
@@
 The same can be said from the multidimensional parameters of other models
-for dichotomous items different from the M2PL model.
+for dichotomous items different from the `r MODEL_ACRONYM` model.
 We expect this work to help set the foundations for investigating those

### 2023-01-20 R/Formulae.R

@@ -428,6 +428,10 @@ ITEM_TRANSF_MATRIX_EQ  <- latex_eq(
 
 # Item direction cosines (in test space):
 ANGLE_TS                          <- latex("\\alpha")
+ANGLE_TS_ITEM                     <- latex_sub(
+  ANGLE_TS,
+  "$ITEM_INDEX$$DIM_INDEX$"
+)
 DIR_ANGLE_VEC_TS                  <- latex_bf(ANGLE_TS)
 DIR_COS_VEC_TS                    <- latex_cos(DIR_ANGLE_VEC_TS)
 DIR_ANGLE_ITEM_VEC_TS             <- latex_sub(DIR_ANGLE_VEC_TS, ITEM_INDEX)
@@ -455,8 +459,9 @@ DIR_COS_ITEM_VEC_TS_EQ            <- latex_eq(
 ### Generalized multidimensional parameters ----
 
 # Parameters:
-MDISC_PARAM <- latex("MDISC")
-MIL_PARAM   <- latex("MIL")
+MDISC_SYM  <- latex("MDISC")
+MDISC_ITEM <- latex_sub(MDISC_SYM, ITEM_INDEX)
+MIL_PARAM  <- latex("MIL")
 
 #### Agnostic version of the indices: ----
 
@@ -467,7 +472,7 @@ INNER_PROD_MAT_STD_EQ <- latex_eq(INNER_PROD_MAT_EQ, ID_MATRIX)
 
 # MDISC:
 AGNOSTIC_SUBINDEX          <- latex("ag")
-MDISC_AG_PARAM             <- latex_sub(MDISC_PARAM, AGNOSTIC_SUBINDEX)
+MDISC_AG_PARAM             <- latex_sub(MDISC_SYM, AGNOSTIC_SUBINDEX)
 DISCR_VECTOR_AG_INNER_PROD <- latex(DISCR_VECTOR_TRANSP, DISCR_VECTOR)
 DISCR_VECTOR_AG_MODULE     <- latex_sqrt(DISCR_VECTOR_AG_INNER_PROD)
 MDISC_AG_PARAM_EQ          <- latex_def(MDISC_AG_PARAM, DISCR_VECTOR_AG_MODULE)
@@ -510,7 +515,7 @@ DISCR_VECTOR_CORR_INNER_PROD <- latex(
   DISCR_VECTOR
 )
[…]

### 2023-01-20 R/Mirt_toolbox.R

@@ -80,12 +80,14 @@ compute_mirt_params <- function(items,
   items                                                                      |>
     tidyr::pivot_longer(!!discrimination, names_to = "par", values_to = "a") |>
     dplyr::group_by(!!item_id)                                               |>
+    # This renaming is necessary for accessing the variable in `transmute`:
+    dplyr::rename(d = !!intercept)                                           |>
     dplyr::transmute(
       !!item_id,
       ## Compute the multidimensional parameters:
       dim   = par |> stringr::str_extract("\\d+"),
       MDISC = (t(a) %*% cov_matrix %*% a) |> drop() |> sqrt(),
-      D     = - !!intercept / MDISC,
+      D     = - d / MDISC,
       cos   = (inv_sd_matrix %*% cov_matrix %*% a / MDISC)  |> drop(),
       rad   = acos(cos),
       deg   = rad * RAD_DEG_FACTOR

### 2023-01-20 src/Empirical_example.R

@@ -1,0 +1,209 @@
+# ==============================================================================
+# 
+# FILE NAME:   Empirical_example.R
+# DESCRIPTION: Script for computing the empirical example for the paper.
+# 
+# AUTHOR:      Mori (danivmorillo@gmail.com)
+# 
+# DATE:        18/07/2022
+# 
+# ==============================================================================
+
+
+## ---- GLOBAL OPTIONS: --------------------------------------------------------
+
+## ---- INCLUDES: --------------------------------------------------------------
+
+## ----libraries----
+library(tidyverse)
+library(flextable)
+library(officer)
+library(assertive.numbers)
+
+## ----sources----
+source("R/Mirt_toolbox.R", encoding = 'UTF-8')
+source("R/Constants.R",    encoding = 'UTF-8')
+source("R/Formulae.R")
+# source("R/Output.R")
+
+## ---- CONSTANTS: -------------------------------------------------------------
+
+## ----general-constants----
+
+# Table manipulation constants:
+DISCR_PARAMS_SELECTION <- quo(matches(DISCR_PARAMS_PATTERN))
+
+# Numerical constants for computations:
+TOL_LEVEL <- 5e-3 # for comparing computed and given multidimensional parameters
+
+## ----set-cov-matrix----
+# Version in Reckase, 2009, p. 183; this represents a "close-to-1D case";
+#   dimension 2 covariance was changed by +.0001 (.3341 instead of .3340) to
+#   avoid negative definiteness of the matrix, which would cause errors
+#   computing the multidimensional parameters.
+COV_MATRIX <- matrix(
+  c(
+    .1670, .2362, .2892,
+    .2362, .3341, .4091,
+    .2892, .4091, .5010
+  ),
+  nrow = 3
+)
+# (see also Reckase, 2009, p. 153 for the more common 3D case)
+
+# # TODO: Delete
+# COV_MATRIX <- diag(4)
+# COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
+
+## ---- CONFIGURATION: ---------------------------------------------------------
+
+## ----graphical-output-conf----
+
+## ---- MAIN: ------------------------------------------------------------------
+
+## ----read-items----
+
+# Item parameters:
+items <- read_csv2(ITEM_PARAMS_RECKASE_2009_FILEPATH)
+
+# # TODO: Delete
+# items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
+
+## ----compute-orthogonal-params----
+items_orth <- items |>
[…]

### 2023-01-20 src/Empirical_example_Reckase2009.R

@@ -1,190 +1,0 @@
-# ==============================================================================
-# 
-# FILE NAME:   Empirical_example.R
-# DESCRIPTION: Script for computing the empirical example for the paper.
-# 
-# AUTHOR:      Mori (danivmorillo@gmail.com)
-# 
-# DATE:        18/07/2022
-# 
-# ==============================================================================
-
-
-## ---- GLOBAL OPTIONS: --------------------------------------------------------
-
-## ---- INCLUDES: --------------------------------------------------------------
-
-## ----libraries----
-library(tidyverse)
-library(flextable)
-library(officer)
-library(assertive.numbers)
-
-## ----sources----
-source("R/Mirt_toolbox.R")
-# source("R/Formulae.R")
-# source("R/LaTeX_math.R")
-# source("R/Output.R")
-
-## ---- CONSTANTS: -------------------------------------------------------------
-
-## ----constants----
-COV_MATRIX <- matrix(
-  c(
-    1.210, .297, 1.232,
-     .297, .810,  .252,
-    1.232, .252, 1.960
-  ),
-  nrow = 3
-)
-# Version in Reckase, 2009, p. 153 (see also p. 183 for the close to 1D case)
-
-DISCR_PARAMS_PATTERN   <- '^a\\d+$'
-DISCR_PARAMS_SELECTION <- quo(matches(DISCR_PARAMS_PATTERN))
-
-## ---- CONFIGURATION: ---------------------------------------------------------
-
-## ----graphical-output-conf----
-
-## ---- MAIN: ------------------------------------------------------------------
-
-## ----read-items----
-
-# Item parameters:
-items <- read_csv2("dat/Table_6.1_Reckase_2009.csv")
-
-## ----orthogonal-params----
-items_orth <- items |>
-  compute_mirt_params(d, !!DISCR_PARAMS_SELECTION, dir_out = "deg")
-
-## ----compare-orthogonal-params----
-
-items                                                |>
-  pull(MDISC)                                        |>
-  is_equal_to(items_orth |> pull(MDISC), tol = 5e-3) |>
-  all()
-
-items                                            |>
-  pull(MDIFF)                                    |>
-  is_equal_to(items_orth |> pull(D), tol = 5e-3) |>
-  all()
-# Both multidimensional parameters are equal for all items, up to the tolerance
-#   level.
-
-## ----oblique-params----
-
-## Compute the parameters:
-items_oblique <- items |> compute_mirt_params(
-  d, !!DISCR_PARAMS_SELECTION,
[…]

### 2023-01-20 src/Graphical_example_paper.R

@@ -26,6 +26,7 @@ source("R/Mirt_toolbox.R")
 source("R/Formulae.R")
 source("R/LaTeX_math.R")
 source("R/Output.R")
+source("R/Constants.R")
 
 ## ---- CONSTANTS: -------------------------------------------------------------
 
@@ -40,7 +41,7 @@ PALETTE      <- c("darkred", "darkgoldenrod3", "green3", "cyan3", "blue3")
 
 # Item parameters:
 items_M2PL <- tribble(
-  ~a_1, ~a_2, ~l  ,
+  ~a_1, ~a_2, ~d  ,
    2  ,  2  ,  0  ,
    2  , -1  ,  1  ,
   -0.5,  1  , -0.5,
@@ -54,7 +55,10 @@ CORR_ARC     <- acos(CORR_VALUE)
 CORR_ARC_SIN <- sin(CORR_ARC)
 CORR_ARC_TAN <- tan(CORR_ARC)
 
-corr_out <- latex_eq(CORR, CORR_VALUE |> format_prop_like(sig = 1))
+# Table output objects:
+CORR_VALUE_OUT <- CORR_VALUE |> format_prop_like(sig = 1)
+CORR_OBL_OUT   <- latex_eq(CORR, CORR_VALUE_OUT) |> as.character()
+CORR_ORTH_OUT  <- latex_eq(CORR, 0)              |> as.character()
 
 
 ## ---- CONFIGURATION: ---------------------------------------------------------
@@ -77,19 +81,19 @@ theme_set( # `ggplot` output configuration
 ## ----compute-example-items----
 
 # Item parameters:
-items_M2PL <- items_M2PL |> rownames_to_column(var = "item")
+items_M2PL <- items_M2PL |> rownames_to_column(var = ITEM_COLKEY)
 
 # Orthogonal case:
-items_orth <- items_M2PL |> compute_mirt_params(l, starts_with('a_'))
-items_orth <- items_orth |>
-  select(-(rad_1:rad_2)) |>
-  full_join(
-    items_orth |> compute_mirt_coords(
-      D, MDISC, starts_with('cos_'),
-      original_coords = FALSE
-    ),
-    by = "item"
-  )
+items_orth_params <- items_M2PL |> compute_mirt_params(
+  all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
[…]

### 2023-01-27 output/paper_draft/paper_M2PLM_parameters.Rmd
Omits output from the comparison of the computed and given orthogonal parameters
`COV_MATRIX` collapsed with the corresponding symbol

@@
 # Version in Reckase, 2009, p. 153; this represents a usual 3D case
-COV_MATRIX <- matrix(
+COV_MATRIX_VALUE <- matrix(
   c(
@@
 # > foi de aproximadamente 0,4.
-COV_MATRIX <- diag(4)
-COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
+COV_MATRIX_VALUE <- diag(4)
+COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- .4
 ```
@@
 ```{r compare-orthogonal-params}
+#| results:  hide
 ```

### 2023-01-27 src/Empirical_example.R

@@ -36,12 +36,15 @@ DISCR_PARAMS_SELECTION <- quo(matches(DISCR_PARAMS_PATTERN))
 # Numerical constants for computations:
 TOL_LEVEL <- 5e-3 # for comparing computed and given multidimensional parameters
 
+# The original "\\mathbf{\\upSigma}" raises an error in flextable (KaTeX):
+COV_MATRIX <- "\\mathbf{\\Sigma}"
+
 ## ----set-cov-matrix----
 # Version in Reckase, 2009, p. 183; this represents a "close-to-1D case";
 #   dimension 2 covariance was changed by +.0001 (.3341 instead of .3340) to
 #   avoid negative definiteness of the matrix, which would cause errors
 #   computing the multidimensional parameters.
-COV_MATRIX <- matrix(
+COV_MATRIX_VALUE <- matrix(
   c(
     .1670, .2362, .2892,
     .2362, .3341, .4091,
@@ -50,10 +53,6 @@ COV_MATRIX <- matrix(
   nrow = 3
 )
 # (see also Reckase, 2009, p. 153 for the more common 3D case)
-
-# # TODO: Delete
-# COV_MATRIX <- diag(4)
-# COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
 
 ## ---- CONFIGURATION: ---------------------------------------------------------
 
@@ -96,7 +95,7 @@ items                                                 |>
 ## Compute the parameters:
 items_oblique <- items |> compute_mirt_params(
   d, !!DISCR_PARAMS_SELECTION,
-  cov_matrix = COV_MATRIX,
+  cov_matrix = COV_MATRIX_VALUE,
   dir_out    = DEGREE_DIRTYPE
 )
 

### 2023-01-27 R/Constants.R
Specifies UFT-8 encoding in all sourced files

@@ -14,6 +14,7 @@
 
 library(glue)
 
+source("R/Formulae.R", encoding = 'UTF-8')
 
 ## ---- CONSTANTS: -------------------------------------------------------------
 

### 2023-01-27 R/Output.R

@@ -18,7 +18,7 @@
 library(officer)
 library(assertive)
 
-# source("R/<file>")
+# source("R/<file>", encoding = 'UTF-8')
 
 
 ## ---- CONSTANTS: -------------------------------------------------------------

### 2023-01-27 src/Graphical_example_paper.R

@@ -22,11 +22,11 @@ library(officer)
 library(scales)
 
 ## ----sources----
-source("R/Mirt_toolbox.R")
-source("R/Formulae.R")
-source("R/LaTeX_math.R")
-source("R/Output.R")
-source("R/Constants.R")
+source("R/Mirt_toolbox.R", encoding = 'UTF-8')
+source("R/Formulae.R",     encoding = 'UTF-8')
+source("R/LaTeX_math.R",   encoding = 'UTF-8')
+source("R/Output.R",       encoding = 'UTF-8')
+source("R/Constants.R",    encoding = 'UTF-8')
 
 ## ---- CONSTANTS: -------------------------------------------------------------
 

### 2023-02-03 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 
-As in @reckase_difficulty_1985, the characteristic function of item
-$`r ITEM_INDEX`$ according to the `r MODEL_ACRONYM` model is formulated
-as follows:
+According to the M2PL model, the probability of a positive response to item
+$`r ITEM_INDEX`$ is
 
@@
 In order to contemplate the general case,
-we need to introduce a few linear algebra results before proceeding futher.
+we need to introduce a few linear algebra results before proceeding further.
 
 ## Linear algebra of non-orthogonal coordinates
-<!-- # TODO: To appendix, OR citation about these? -->
-
-### Inner product
-
-Let $`r INNER_PROD_EQ`$ be the inner product in $`r LATENT_SPACE`$, with
-the positive-definite matrix $`r INNER_PROD_MATRIX`$ its inner product matrix.
-We may apply a change of basis in $`r LATENT_SPACE`$ that results in a space
-$`r LATENT_SPACE_STD`$ with the standard basis $`r LS_STD_BASIS_EQ`$.
-Such a change of basis can be defined as
-
-\begin{equation}
-  `r BASIS_CHANGE`,                                          (\#eq:change-basis)
-\end{equation}
-
-with $`r TRAIT_VECTOR_STD`$ the coordinates
-of $`r TRAIT_VECTOR`$ in $`r LATENT_SPACE_STD`$ and $`r TRANSFORM_MATRIX_EQ`$
-(being $`r BASIS_VECTOR_ANY_TRANSF`$ the coordinates of
-$`r BASIS_VECTOR_ANY`$ in $`r LATENT_SPACE_STD`$
-for any dimension $`r DIM_INDEX`$).
-This new space can be defined as a Euclidean space with standard inner product
-(i.e. dot product), such that the norm $`r NORM_EQ`$ of $`r TRAIT_VECTOR`$
-in $`r LATENT_SPACE_STD`$ is given by<!-- # TODO: Reference needed? -->
-
-\begin{equation}
-  `r TRAIT_MODULE`^2 = `r TRAIT_VECTOR_STD_SQ` =
-    `r TRANSF_TRAIT_SQUARED` =
-    `r TRAIT_TRANSF_TRAIT`.                                 (\#eq:inner-product)
-\end{equation}
-
-Thus, the inner product matrix $`r INNER_PROD_MATRIX`$ in $`r LATENT_SPACE`$
-is consistently defined as
-
-\begin{equation}
-  `r INNER_PROD_MAT_EQ`.                            (\#eq:inner-prod-matrix-def)
-\end{equation}
-
-### Direction cosines
+
+Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
+<!-- TODO: Incomplete in Mario's version, is it correct to say "in the latent space"? -->
+We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
+Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
+$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$, where $`r TRAIT_COORDS_EQ`$ and
+$`r TRAIT_COORDS_STD_EQ`$.
+
+Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix; then, we have that
+
+\begin{equation}
+  `r BASIS_CHANGE`.                                          (\#eq:change-basis)
+\end{equation}
+
+The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.
+<!-- # TODO: For all k? -->
+If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
+
+\begin{equation}
+  `r TRANSFORM_MATRIX_EQ`.                           (\#eq:transform-matrix-def)
+\end{equation}
+
+With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product<!-- (i.e. dot product) -->.
+With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
+$`r LS_BASIS_INNER_PROD`$, such that the norm of $`r TRAIT_VECTOR`$ is
+invariant, i.e., $`r TRAIT_NORM_EQ`$. Therefore,
+
+\begin{equation}
+  `r TRAIT_NORM_STD_SQ_EQ`.                             (\#eq:inner-product-def)
+\end{equation}
+
+Thus, we must define the inner product $`r LS_BASIS_INNER_PROD`$ with Gram
+matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$, which keeps the
+norm invariant.
+
+The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
+$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
+invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
 
@@
 \end{equation}
+
+Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
+up by the $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
+vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.
+
+::: {#prp-cosines-equivalence}
+
+**Proposition 1.** Under the previously defined conditions, for all
+$`r TRAIT_VEC_IN_LS`$,
+
+\begin{equation}
+  `r DIR_COS_VEC_EQ`.                             (\#eq:direction-cosine-vector)
+\end{equation}
+
+with
+
+:::
 

### 2023-02-03 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 
-\begin{equation}
-  `r TRAIT_VECTOR_POLAR_EQ`.                           (\#eq:trait-vector-polar)
-\end{equation}
-
-Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+Thus, the M2PL model is expressed in polar coordinates as
 

### 2023-02-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Completes example in Reckase, 2009

@@
 
+### @reckase_multidimensional_2009
+
 ```{r reset-root-wd}
@@
 
+```{r format-covariance-matrix}
+```
+
 ```{r create-params-table}
@@
 ```
+
+Table \@ref(tab:empirical-table-Reckase2009-out) draws from a classical example
+by @reckase_multidimensional_2009.
+In this 30-test item, there are three ten-item blocks assumed to approximate
+a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
+and items 21-30 dimension 2 (see Table 6.1, p. 153 therein).
+In his example, Reckase tests the estimation of these items
+on two simulated datasets: In the first one, the latent trait distribution
+is standard with null correlations; however, the second one has
+(apart from a non-null vector of means) as covariance matrix
+
+\begin{equation}
+  `r cov_matrix_value_out`,
+\end{equation}
+
+which corresponds to a correlation matrix of
+
+\begin{equation}
+  `r cor_matrix_value_out`.
+\end{equation}
+
+As we can see in Table \@ref(tab:empirical-table-Reckase2009-out), the agnostic
+versions of the multidimensional parameters
+(columns labelled _`r AGNOSTIC_ABBR`_)
+coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
+However, when we take into account the latent space structure
+(columns labelled $`r COV_MATRIX`$), we notice several differences.
+First, the $`r MDISC_ITEM`$ parameters tend to be larger
+than their agnostic counterparts.
+This happens because all the discrimination parameters as and the latent space
+correlations are positive; in this respect, these items are similar to item 1
+in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
+However, for the items mostly aligned with dimension 2 (items 21-30), as the
+$`r MDISC_ITEM`$ parameters are also scaled by a standard deviation smaller
+than 1, their values decrease, being some of them smaller than the agnostic
+versions. The $`r DISTANCE_PARAM`$ parametes decrease or increase, consequently,
+(see Equation \@ref(eq:M2PL-formula-polar)), thus being most of them smaller
+than the agnostic ones (again, with the exceptions only happening
+in the items aligned with dimension 2).
+Finally, we see that the direction angles are not as clearly separated as one
+might expect from the agnostic versions. This effect is more pronounced among
+the items aligned with dimensions 1 and 3, which are strongly correlated.
+In this situation, we notice that the angles tend to be much smaller, as the
+high correlation induces an alignment between the two dimensions
+(compare item 1 in the two test spaces in
+Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
 
@@
 ```
+
+The latter effect can be better noticed in another case, in which Reckase
+uses the same item parameters to demonstrate the over-specification of
+dimensions (see p. 183).
+In this example, he uses a covariance matrix that practically has rank 1,
+implying that the (alleged) three dimensions collapse into a single one, being
+thus the test unidimensional. In such a case, the covariance-based version of
+the direction angles are extremely close to 0 for all items, as one would expect
+from a unidimensional test, where all items are strictly aligned with the one
+dimension they measure. The reader interested can check out the parameters
+computed in such case in the Supplementary Table 1.
+<!-- #TODO: Include Supplementary Table -->
 

### 2023-02-03 R/Constants.R

@@ -65,6 +65,7 @@ MULTIDIM_PARAMS_TITLE <- "Multidimensional parameters"
 MODEL_ACRONYM         <- "M2PL"
 ITEM_TABLE_TITLE      <- "Item"
 SPACE_SEP             <- ' '
+AGNOSTIC_ABBR         <- glue("{AGNOSTIC_SUBINDEX}{DOT}") |> as.character()
 
 ## Footer elements:
 FOOTER_PREFFIX        <- as_i("Note. ")
@@ -99,7 +100,7 @@ ANGLE_EXPLANATION     <- list(
   " = Direction component of the multidimensional item location parameter",
   " (in dimension ", as_equation(as.character(DIM_INDEX)), ")"
 )
-AG_VER_EXPLANATION    <- glue("{AGNOSTIC_SUBINDEX}{DOT} = Agnostic version; ")
+AG_VER_EXPLANATION    <- glue("{AGNOSTIC_ABBR} = Agnostic version; ")
 COV_VER_EXPLANATION   <- list(
   as_equation(as.character(COV_MATRIX)),
   " = Covariance-based version."

### 2023-02-03 src/Empirical_example.R

@@ -16,6 +16,7 @@
 
 ## ----libraries----
 library(tidyverse)
+library(scales)
 library(flextable)
 library(officer)
 library(assertive.numbers)
@@ -85,6 +86,19 @@ items                                                 |>
   cat()
 # Both multidimensional parameters are equal for all items, up to the tolerance
 #   level.
+
+## ----format-covariance-matrix----
+
+cov_matrix_value_out      <- COV_MATRIX_VALUE |> number(1e-3)
+dim(cov_matrix_value_out) <- dim(COV_MATRIX_VALUE)
+cov_matrix_value_out      <- latex_matrix(cov_matrix_value_out)
+
+cor_matrix_value_out      <- COV_MATRIX_VALUE |>
+  cov2cor() |>
+  latex_matrix()
+# cor_matrix_value_out      <- COV_MATRIX_VALUE |> cov2cor() |> number(1e-3)
+# dim(cor_matrix_value_out) <- dim(COV_MATRIX_VALUE)
+# cor_matrix_value_out      <- latex_matrix(cor_matrix_value_out)
 
 ## ----compute-oblique-params----
 
@@ -159,8 +173,7 @@ item_headers <- tibble(
                                               DIM_INDEX,
                                               str_extract(., DIGIT_PATTERN)
                                             ),
-      str_detect(., ORTH_SUFFIX)          ~ glue("{AGNOSTIC_SUBINDEX}{DOT}") |>
-                                              as.character(),
+      str_detect(., ORTH_SUFFIX)          ~ AGNOSTIC_ABBR,
       str_detect(., OBL_SUFFIX)           ~ COV_MATRIX |> as.character(),
       TRUE                                ~ metric_param
     )

### 2023-02-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Finishes including example in Tezza et al., 2018

@@
 
+### @tezza_modelo_2018
+
 ```{r set-cov-matrix-Tezza2018}
@@
 COV_MATRIX_VALUE <- diag(4)
-COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- .4
+DIMS_CORR        <- .4
+COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
 ```
@@
 ```
+
+@tezza_modelo_2018 provide us with another example that allows showcasing
+the effect of correlated dimensions on negative discrimination parameters.
+In their model, dimensions 1 and 4 are estimated to have a correlation of
+`r DIMS_CORR`.
+With such latent space structure, items that strongly discriminate in any or
+both of these two dimensions change drastically their multidimensional
+parameters from the agnostic to the covariance-based version
+(see Table \@ref(tab:empirical-table-Tezza2018-out)) ---this happens,
+for example, with items 57 and 60. These items also have different-sign
+discrimination parameters in dimensions 1 and 4, so their situation is
+similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
+Figure \@ref(fig:item-plot-out),
+thus the decrease in their $`r MDISC_ITEM`$ parameter.
+The effect on the direction is also patent in the angles these items make with
+these dimensions, which change up to almost 20º, compared with at most 7º in the
+other two dimensions, and around 1º in most cases.
+
+On the other hand, in cases such as items 12 or 59, which have
+relatively low discrimination parameters on those two dimensions
+(compared with the ones in dimension 2 and 3, respectively) are barely affected.
+In these cases, we also notice that the angles are barely affected.
+The most extreme change in these two examples occurs in
+the direction of item 12 with respect to dimension 4. We see how the large
+discrimination in dimension 1, when combined with the correlation
+with dimension 4, induces a decrease in the angular direction
+with that dimension.
+In this case, being the discrimination in dimension 4 so low, the situation is
+very similar to item 4 in Table \@ref(tab:example-items-table-out)
+and Figure \@ref(fig:item-plot-out).
+
+<!---BLOCK_LANDSCAPE_START--->
 
@@
 ```
-
-<!---BLOCK_LANDSCAPE_START--->
 

### 2023-02-10 output/paper_draft/paper_M2PLM_parameters.Rmd
Makes minor style changes Corrects some errata Properly refers to "orthonormality" instead of "orthogonality" when needed

--- 2023-02-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-```{r setup}
-#| include: false
-
+```{r setup, include=FALSE}
 # Includes:
@@
 library(patchwork)
-library(rstudioapi)
-
-ROOT_DIR <- getActiveProject()
+
+ROOT_DIR <- "../.."
 DOC_DIR <- getwd()
@@
   echo     = FALSE,
-  message  = FALSE,
   cache    = FALSE,
@@
   dpi      =   300,
-  dev.args = list(type = "cairo-png")
+  dev.args = list(type = "cairo-png"),
+  fig.pos = "!H"
 )
@@
 
-```{r write-bib}
-#| cache: true
-
+```{r write-bib, cache=TRUE}
 # automatically create a bib database for R packages
@@
 
-```{r sources}
-#| cache:   false
-#| include: false
+```{r sources, cache=FALSE, include=FALSE}
 source("R/Formulae.R", encoding = 'UTF-8')
-source("R/Constants.R", encoding = 'UTF-8')
 ```
@@
 The so called *Multidimensional 2-Parameter Logistic*
-[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is one of the most important
-and widely used multidimensional models,
-partly due to its close relationship to the common factor model
+[M2PL, @mckinley_extension_1983] model is among the most important
+and widely used ones.
+This is partly due to its close relationship to the common factor model
 [@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
-The `r MODEL_ACRONYM` model is a generalization of
-the 2-parameter logistic model [@birnbaum_latent_1968]
-to more than one latent trait [@mckinley_extension_1983].
+It consists of a generalization of the 2-parameter logistic model
+[@birnbaum_latent_1968] to more than one latent trait
+[@mckinley_extension_1983].
 Its usefulness lies in describing the variability of the population when
@@
 more specifically, when a deficit in a certain trait
-and a surplus in another one cancel out their effects --thus it is sometimes
-referred to as a *compensatory* model.
-
-@reckase_difficulty_1985 argues in favor of a need to characterize
-the statistical properties of items taking dimensionality into account.
-As such, he outlines a general procedure for determining
-the difficulty of multidimensional items.
+and a surplus in another one cancel out their effects. Because of this property,
+it is sometimes referred to as a *compensatory* model.
+
+@reckase_difficulty_1985 argues for a need to take dimensionality into account
+when characterizing the statistical properties of multidimensional items.
+As such, he outlines a general procedure for determining their difficulty.
 Following this procedure, he also derives
-a *multidimensional difficulty* parameter (MID) for `r MODEL_ACRONYM` items.
+a *multidimensional difficulty* parameter for M2PL items.
 This parameter consists of two parts:
@@
 the origin of the multidimensional latent space to
-the maximally discriminating point of the item.
-The second one, the direction of the vector from the origin to that same point.
-Following a similar rationale, @reckase_discriminating_1991 define
-a *multidimensional discrimination* (MDISC) parameter definition for that model.
+the maximally discriminating locus of the item.
+The second one, the direction of the vector from the origin to that same locus.
+Following a similar rationale, @reckase_discriminating_1991 derive a
+*multidimensional discrimination* (`r MDISC_PARAM`) parameter definition
+for that model.
 
@@
 latent space are orthogonal
-[@ackerman_using_1994; @ackerman_multidimensional_2005; see also @reckase_difficulty_1985, p. 404 and Equation 7].
+[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
+@reckase_difficulty_1985, p. 404 and Equation 7].
 Of course, orthogonality is a property of a geometrical space instead of a
@@
 the R package *psych* <!--# Reference needed? -->graphically represent
-factor loadings on orthogonal axes, even after an oblique rotation.
+factor loadings on orthogonal axes even after an oblique rotation.
 Nevertheless, the assimilation of correlated latent traits to non-orthogonal
@@
 This fact is probably due to the different spotlights
-the two theoretical programs put the focus on;
-Factor analysis aims at a clear interpretation of
+these two theoretical programs put the focus on;
+factor analysis aims at a clear interpretation of
 the latent covariance structure, and thus focuses on finding
@@
 Nevertheless, the confluence of the two theoretical approaches into a common
-modern measurement theory [@mcdonald_test_1999] imposes
+modern measurement theory [@mcdonald_test_1999] requires
 the relaxation of traditional constrains on both approaches.
@@
 [see @zhang_theoretical_1999 for a notable exception].
-What is more, disregarding this geometrical interpretation may lead to
+Moreover, disregarding this geometrical interpretation may lead to
 misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->
 
-A latent space where the traits are correlated is easier to represent in
-a geometrical space with non-orthogonal axes,
+A latent space where the traits are correlated is easier to understand as a
+geometrical space with non-orthogonal axes,
 often referred to as *general Cartesian coordinates* [@harman_modern_1970].
@@
 The general case of MIRT model estimation implies estimating possibly
-non-standard latent dimensions and non-null covariances among them,
-a more realistic case than an identity matrix.
-Assuming orthogonality among traits is usually unrealistic, be it in the
+non-standard latent dimensions and non-null covariances among them
+(a more realistic case than an identity matrix).
+Assuming orthonormality among traits is usually unrealistic, be it in the
 cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
@@
 interpreting the latent dimensions as substantive constructs is not a goal.
-As we will argue, the usual definitions of the
-MID and MDISC parameters do not generalize to the non-orthogonal geometry used
-to represent correlated latent spaces.
+As we will argue, the usual definitions of the multidimensional difficulty and
+discrimination parameters do not generalize to the non-orthonormal geometry
+necessary to accurately represent correlated latent spaces.
 
@@
 graphically representing these items in the fashion of @ackerman_creating_1994,
-but taking into account the non-orthogonality of the coordinate axes.
+but taking the non-orthogonality of the coordinate axes into account.
 In order to pursue these objectives, we first define the generalization of
-the `r MODEL_ACRONYM` model and describe its scope of application.
+the M2PL model and describe its scope of application.
 Then, we follow the method outlined by @reckase_difficulty_1985 and
@@
 propose a graphical representation method,
-and present an example of computation<!-- # TODO: Pending -->
-and representation of the items based on these parameters.
+and present two examples of application of our results.
 We round up with a discussion about
@@
 
-where $`r DISCR_VECTOR`$ is a $`r N_DIMS`$-dimensional vector of discrimination
+where $`r DISCR_VECTOR`$ is an $`r N_DIMS`$-dimensional vector of discrimination
 parameters, and $`r INTERCEPT_PARAM`$ an intercept parameter related to the
@@
 non-negative real numbers in Reckase
-[-@reckase_difficulty_1985, a decrease in the probability of a correct response with an increase in ability makes little sense].
+[-@reckase_difficulty_1985, a decrease in the probability of a correct response
+with an increase in ability makes little sense].
 In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
@@
 Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
-<!-- TODO: Incomplete in Mario's version, is it correct to say "in the latent space"? -->
-We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
+We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each basis,
+$`r LS_BASIS`$ or $`r LS_STD_BASIS`$.
 Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
@@
 
-Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix; then, we have that
+Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix between the two bases;
+then, we have that
 
@@
 The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
-coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.
-<!-- # TODO: For all k? -->
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.<!--# Does this need to be justified? -->
 If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
@@
 
-With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product<!-- (i.e. dot product) -->.
+With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product
+(i.e., dot product).
 With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
@@
 norm invariant.
-
 The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
@@
 
+We can also obtain a result that relates the direction cosines of any two bases.
+
+::: {#prp-dir-cosines-bases-relationship}
+
+**Proposition 2** Let's consider another basis $`r ALT_BASIS_EQ`$ in
+$`r LATENT_SPACE`$ and the inner product $`r ALT_BASIS_INNER_PROD`$ that keeps
+the norm of the Euclidean space invariant.
+
+Let $`r DIR_COS_VEC_ALT`$ be the vector of direction cosines of
+$`r TRAIT_VECTOR`$ in $`r ALT_BASIS`$ (i.e., the cosines with the elements of
+$`r ALT_BASIS`$).
+
+Then,
+
+\begin{equation}
+  `r DIR_COS_VEC_ALT_EQ`,                      (\#eq:dir-cosines-transformation)
+\end{equation}
+
+where
+
+\begin{equation}
+  `r DIAG_MATRIX_INNER_PROD_ALT_EQ`
+\end{equation}
+
+and $`r TRANSFORM_MATRIX_ALT_EQ`$.
+
+::: {.proof}
+*Proof.*
+
+Let's call $`r TRAIT_COORDS_ALT_EQ`$ and let $`r TRANSFORM_MATRIX_ALT_EQ`$
+be the change of basis matrix, such that $`r TRAIT_TRANSFORM_ALT_EQ`$.
+
+Then, we have that $`r BASIS_CHANGE_ORG_ALT_EQ`$.
+
+The inner product $`r ALT_BASIS_INNER_PROD`$ that keeps the norm invariant is
+defined by $`r INNER_PROD_TRAIT_ALT_EQ`$, and we define
+$`r INNER_PROD_MATRIX_ALT_EQ`$.
+
+By proposition 1, for each case we have:
+
+1. $`r TRAIT_VEC_STD_POLAR_EQ`$
+
+1. $`r TRAIT_VECTOR_POLAR_EQ`$
+
+1. $`r TRAIT_VEC_ALT_POLAR_EQ`$
+
+Given (1),
+
+\begin{equation}
+  `r TRAIT_VEC_STD_POLAR_EQ_XPAND`
+\end{equation}
+
+and
+
+\begin{equation}
+  `r TRAIT_VEC_ALT_POLAR_ALT_EQ`.
+\end{equation}
+
+Using (2) and (3) we get that
+
+\begin{equation}
+  `r TRAIT_VEC_STD_POLAR_ORG_EQ`
+\end{equation}
+
+and
+
+\begin{equation}
+  `r TRAIT_VEC_STD_POLAR_ALT_EQ`.
+\end{equation}
+
+Therefore,
+
+\begin{equation}
+  `r TRAIT_VEC_POLAR_ORG_ALT_COEFF_EQ`,
+\end{equation}
+
+and thus $`r DIR_COS_VEC_ALT_EQ`$.$`r END_OF_PROOF`$
+
+:::
+
+:::
+
 ## Model recasting into polar coordinates
 
-From Equation \@ref(eq:direction-cosine-vector),
+From Proposition 1,
 any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
+
+\begin{equation}
+  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
+\end{equation}
 
@@
 Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
-already assumes the orthogonality of the latent space.
+already assumes the orthonormality of the latent space.
 
@@
 we first compute the point of maximum slope by finding the root(s) of
-the second derivative with respect to $`r TRAIT_MODULE`$.
+the second derivative with respect to $`r TRAIT_NORM`$.
+
+The slope in direction $`r DIR_ANGLE_VEC`$ is given by
+
+\begin{equation}
+  `r IRF_1ST_DIFF_EQ`,                                      (\#eq:M2PL-1st-diff)
+\end{equation}
+
+and
 
@@
 (the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
-which result in improper values of $`r TRAIT_MODULE`$).
-The slope in direction $`r DIR_ANGLE_VEC`$ is given by
-
-\begin{equation}
-  `r IRF_1ST_DIFF_EQ`.                                      (\#eq:M2PL-1st-diff)
-\end{equation}
-
-When $`r IRF_MAX_SLOPE_EQ`$,
+which result in improper values of $`r TRAIT_NORM`$).
+
+The slope in direction $`r DIR_ANGLE_VEC`$ when $`r IRF_MAX_SLOPE_EQ`$ is given
+by
 
@@
 However, this assertion implicitly assumes orthogonal coordinates,
-so it does not apply in general to $`r DIR_COS_VEC`$.<!-- # TODO: To appendix the following, or maybe give for granted? -->
-Instead, we may consider the direction cosine vector in $`r LATENT_SPACE_STD`$:
-
-\begin{equation}
-  `r DIR_COS_VEC_STD_EQ`.                     (\#eq:direction-cosine-vector-std)
-\end{equation}
-
-Operating on Equation \@ref(eq:trait-vector-polar),
+so it does not apply in general to $`r DIR_COS_VEC`$.
+Instead, we may consider the direction cosine vector with the standard basis
+$`r LS_STD_BASIS`$.
+Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain
 
@@
 
-We can define
+We can then define
 
@@
 
-such that Equation \@ref(eq:slope-max-std-cosines) is similar
-(in vectorial form) to Reckase's [-@reckase_difficulty_1985] Equation 6,
-and thus can be solved by the same procedure
-applying their Equations 7 through 9.
-The resulting equivalent to their Equation 9, in vectorial form
-and solving for $`r DIR_ANGLE_STD_SUBSTITUTE`$ is
-
-\begin{equation}
-  `r DIR_COS_STD_ITEM_VEC_EQ`.                            (\#eq:dir-cos-vec-std)
-\end{equation}
-
-Substituting $`r DIR_COS_STD_ITEM_VEC`$ and $`r DISCR_VECTOR_STD`$
-by their respective expressions results in the expression for
-the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
+so
+
+\begin{equation}
+  `r SLOPE_MAX_STD_EQ`.                                     (\#eq:slope-max-std)
+\end{equation}
+
+We are thus considering the direction cosines with respect to the orthonormal
+axes, so we can just apply @reckase_difficulty_1985's results, obtaining
+
+\begin{equation}
+  `r DIR_COS_VEC_STD_EQ`.                                 (\#eq:dir-cos-vec-std)
+\end{equation}
+
+Substituting $`r DIR_COS_VEC_STD`$
+(Equation \@ref(eq:direction-cosine-vectors-eq)) and
+$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
+by their respective expressions results in the
+expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
 
@@
 from the origin to that point,
-we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$,
-$`r DIR_COS_SUBSTITUTE`$ and
-$`r DISTANCE_SUBSTITUTE`$,
-and then substitute Equation \@ref(eq:dir-cos-vec) for $`r DIR_COS_ITEM_VEC`$.
-The result, solving for $`r DISTANCE_PARAM`$, is
+we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$ to get
+$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
 
@@
 
-The basis $`r ITEM_SPACE_BASIS_EQ`$ of $`r TEST_SPACE`$ is such that
-$`r ITEM_TRANSF_MATRIX_EQ`$ (see Equation \@ref(eq:discr-vector-std)).
-Thus, similarly to equation \@ref(eq:dir-cos-vec),
-the vector of direction cosines of a vector in $`r TEST_SPACE`$
-can be defined as
+With this definition we have that
+
+\begin{equation}
+  \begin{split}
+    `r DIR_COS_ITEM_VEC_NORM_EQ` \\
+    `r DISTANCE_PARAM_NORM_EQ`   \\
+    `r SLOPE_MAX_PARAM_NORM_EQ`.
+  \end{split}                             (\#eq:item-params-discrimination-norm)
+\end{equation}
+
+It is clear that $`r INNER_PROD_INV_MAT_ELEMENT_EQ`$ were
+$`r INNER_PROD_INV_MAT_ELEMENT`$ is the $`r INNER_PROD_INV_MAT_SUBINDEX`$-th
+element of $`r INNER_PROD_MATRIX_INV`$.
+
+Let $`r LS_STD_BASIS`$ be the standard basis of $`r TEST_SPACE`$.
+Then, $`r BASIS_CHANGE_TS_EQ`$ is the change of basis matrix and
+$`r DISCR_VECTOR_STD_MOD_EQ`$.
+
+Applying Proposition 1,
 
@@
 
-with $`r DIAG_MATRIX_INNER_PROD_INV`$ the diagonal matrix with
-the same diagonal as $`r INNER_PROD_MATRIX_INV`$.
+with $`r DIAG_MATRIX_INNER_PROD_INV`$ a diagonal matrix where
+$`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
 
@@
 from the origin to the point of maximum slope [@reckase_difficulty_1985].
-By analogy with the unidimensional case, $`r MDISC_SYM`$ is defined as
+By analogy with the unidimensional case, $`r MDISC_PARAM`$ is defined as
 $4 `r SLOPE_MAX_PARAM`$ [@reckase_discriminating_1991].
-Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
+Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_PARAM`$ using
 equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
@@
 
-3. $`r MDISC_SYM`$ is four times the maximum slope $`r SLOPE_MAX_PARAM`$.
+3. $`r MDISC_PARAM`$ is four times the maximum slope $`r SLOPE_MAX_PARAM`$.
 
@@
 Equation \@ref(eq:distance-param),
-and the third one is implicit in the definition of the $`r MDISC_SYM`$.
+and the third one is implicit in the definition of the $`r MDISC_PARAM`$.
 Simple arithmetic can show that the first holds true whenever the
@@
 By the definition of the inner product,
-$`r MDISC_SYM`$ is strictly positive for non-null vectors.
+$`r MDISC_PARAM`$ is strictly positive for non-null vectors.
 Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
@@
 The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
-also due to the strict positivity of $`r MDISC_SYM`$,
+also due to the strict positivity of $`r MDISC_PARAM`$,
 are equal to the signs of $`r DISCR_VECTOR`$.
@@
 from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
-and will have a length equal to its $`r MDISC_SYM`$ parameter.
+and will have a length equal to its $`r MDISC_PARAM`$ parameter.
 
@@
 
-```{r libraries}
-```
-
-```{r sources}
+```{r libraries, message=FALSE}
+```
+
+```{r sources, message=FALSE}
 ```
@@
 with $`r TRANSFORM_MATRIX`$ being an identity matrix.
-In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
+In the second one, the correlation between the dimensions is $`r corr_out`$.
 In order to keep the horizontal axis invariant,
@@
 
-The sample items have the same `r MODEL_ACRONYM` parameters
-in both latent spaces.
+The sample items have the same M2PL parameters in both latent spaces.
 However, their multidimensional parameters differ due to the correlation among
@@
 The parameters in both spaces,
-along with their corresponding `r MODEL_ACRONYM` parameters,
+along with their corresponding M2PL parameters,
 are shown in Table \@ref(tab:example-items-table-out).
@@
 
-```{r set-flextable-wd}
-#| cache: false
-
+```{r set-flextable-wd, cache=FALSE}
 # Necessary for officedown to find the template file:
@@
 
-```{r example-items-table-out}
-#| cache:   false
-#| tab_id:  example-items-table-out
-#| tab.cap: Item parameters for the graphical example
+```{r example-items-table-out, cache=FALSE, tab.id="example-items-table-out", tab.cap="Item parameters for the graphical example"}
 item_params_output
@@
 
-```{r item-plot-out}
-#| cache:      false
-#| fig.cap:   Item plots in an orthogonal (left) and oblique (right) test space.
-#| fig.height: 3.1
-#| fig.width:  6.65
+```{r item-plot-out, cache=FALSE, fig.cap="Item plots in an orthogonal (left) and oblique (right) test space.", fig.height=3.1, fig.width=6.65}
 plot_orth + plot_oblique
@@
 
-## Examples with empirical data
+## Example with empirical data
 
 ```{r reset-root-wd}
-#| cache: false
-
 # Necessary for knitr to find the source file:
@@
 
-```{r libraries}
-```
-
-```{r sources}
-```
-
-```{r general-constants}
-```
-
-```{r set-cov-matrix-Reckase2009}
-# Version in Reckase, 2009, p. 153; this represents a usual 3D case
-COV_MATRIX_VALUE <- matrix(
-  c(
-    1.210, .297, 1.232,
-     .297, .810,  .252,
-    1.232, .252, 1.960
-  ),
-  nrow = 3
-)
-# (see also Reckase, 2009, p. 183 for a "close-to-1D" case).
-```
-
-```{r read-items}
-```
-
-```{r compute-orthogonal-params}
-```
-
-```{r compute-oblique-params}
-```
-
-```{r create-params-table}
-```
-
-```{r compose-output-table}
-```
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Reckase2009-out}
-#| cache:   false
-#| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: Item parameters for the empirical example
+```{r libraries, message=FALSE}
+```
+
+```{r sources, message=FALSE}
+```
+
+```{r read-items, message=FALSE}
+```
+
+```{r orthogonal-params}
+```
+
+```{r oblique-params}
+```
+
+```{r params-table}
+```
+
+```{r compose-empirical-items-table}
+```
+
+```{r set-flextable-wd, cache=FALSE}
+```
+
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r empirical-items-table-out, cache=FALSE, tab.id="empirical-items-table-out", tab.cap="Item parameters for the empirical example"}
 item_params_output
@@
 
-```{r set-cov-matrix-Tezza2018}
-# See Tezza et al., 2018, p. 927:
-# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
-# > foi de aproximadamente 0,4.
-COV_MATRIX_VALUE <- diag(4)
-COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- .4
-```
-
-```{r reset-root-wd}
-#| cache: false
-```
-
-```{r read-items-Tezza2018}
-items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
-```
-
-```{r compute-orthogonal-params}
-```
-
-```{r compare-orthogonal-params}
-#| results:  hide
-```
-
-<!-- Computed item parameters in the orthonormal space match the ones in given
-in Tezza 2018, table 5 -->
-
-```{r compute-oblique-params}
-```
-
-```{r create-params-table}
-```
-
-```{r compose-output-table}
-```
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-<!---BLOCK_LANDSCAPE_START--->
-
-```{r empirical-table-Tezza2018-out}
-#| cache:   false
-#| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: Item parameters for the empirical example
-item_params_output
-```
-
 <!---BLOCK_LANDSCAPE_STOP--->
+
+<!-- # TODO: Complete? -->
 
@@
 It is worth noting that @zhang_theoretical_1999 already proposed a formulation
-of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
+of the $`r MDISC_PARAM`$ equivalent to our covariance-based version,
 but for a more general, semi-parametric formulation
@@
 Our results provide this formal derivation, although only for the specific
-case of the `r MODEL_ACRONYM` model.
+case of the M2PL model.
 
@@
 the multidimensional parameters, and refer to them simply as
-the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
+the $`r MDISC_PARAM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
@@
 The same can be said from the multidimensional parameters of other models
-for dichotomous items different from the `r MODEL_ACRONYM` model.
+for dichotomous items different from the M2PL model.
 We expect this work to help set the foundations for investigating those

### 2023-02-10 R/Formulae.R

@@ -273,7 +273,7 @@ DIR_COSINE_DEF             <- latex_frac(
 TRAIT_NORM_INV             <- latex_frac(1, TRAIT_NORM, .sep = NO_SEP)
 BASIS_VECTOR_ANY_NORM_INV <- latex_frac(1, BASIS_VECTOR_ANY_NORM, .sep = NO_SEP)
 DIR_COSINE_DEF_EXPANDED    <- latex(
-  TRAIT_NORM_INV, BASIS_VECTOR_ANY_NORM_INV,         # Denominators
+  BASIS_VECTOR_ANY_NORM_INV, TRAIT_NORM_INV,         # Denominators
   BASIS_VECTOR_ANY, INNER_PROD_MATRIX, TRAIT_VECTOR, # Numerators
 )
 DIR_COSINE_EQ              <- latex_eq(
@@ -763,6 +763,27 @@ DIR_COS_ITEM_VEC_TS_EQ            <- latex_eq(
   DIR_COS_ITEM_VEC_TS,
   DIR_COS_ITEM_VEC_TS_DEF
 )
+
+# Elements of the diagonal matrix of the inverse inner product matrix:
+DIAG_MATRIX_INNER_PROD_INV_EL     <- latex_prime('d')
+DIAG_MATRIX_INNER_PROD_INV_EL_ANY <- latex_sub(
+  DIAG_MATRIX_INNER_PROD_INV_EL,
+  DIM_INDEX
+)
+INNER_PROD_INV_MAT_DIAG_EL        <- latex_sub('m', '$DIM_INDEX$$DIM_INDEX$') |>
+  latex_inverse()
+ITEM_BASIS_VECTOR_ANY_NORM        <- latex_norm(ITEM_BASIS_VECTOR_AUX)
+ITEM_BASIS_VECTOR_ANY_NORM_SQ     <- latex_squared(ITEM_BASIS_VECTOR_ANY_NORM)
+MATRIX_INNER_PROD_INV_EL_DIAG     <- latex_sub(
+  DIAG_MATRIX_INNER_PROD_INV_EL,
+  DIM_INDEX
+)
+MATRIX_INNER_PROD_INV_EL_DIAG_EQ  <- latex_eq(
+  DIAG_MATRIX_INNER_PROD_INV_EL_ANY,
+  INNER_PROD_INV_MAT_DIAG_EL,
+  ITEM_BASIS_VECTOR_ANY_NORM_SQ
+)
+
 
 ### Generalized multidimensional parameters ----
 

### 2023-02-15 output/paper_draft/paper_M2PLM_parameters.Rmd
pedido git

@@
 The so called *Multidimensional 2-Parameter Logistic*
-[M2PL, @mckinley_extension_1983] model is among the most important
-and widely used ones.
-This is partly due to its close relationship to the common factor model
+[M2PL, @mckinley_extension_1983] model is one of the most important and
+widely used multidimensional models,
+partly due to its close relationship to the common factor model
 [@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
-It consists of a generalization of the 2-parameter logistic model
+The M2PL model is a generalization of the 2-parameter logistic model
 [@birnbaum_latent_1968] to more than one latent trait
@@
 more specifically, when a deficit in a certain trait
-and a surplus in another one cancel out their effects. Because of this property,
-it is sometimes referred to as a *compensatory* model.
-
-@reckase_difficulty_1985 argues for a need to take dimensionality into account
-when characterizing the statistical properties of multidimensional items.
-As such, he outlines a general procedure for determining their difficulty.
+and a surplus in another one cancel out their effects --thus it is sometimes
+referred to as a *compensatory* model.
+
+@reckase_difficulty_1985 argues in favor of a need to characterize
+the statistical properties of items taking dimensionality into account.
+As such, he outlines a general procedure for determining
+the difficulty of multidimensional items.
 Following this procedure, he also derives
-a *multidimensional difficulty* parameter for M2PL items.
+a *multidimensional difficulty* parameter (MID) for M2PL items.
 This parameter consists of two parts:
@@
 the origin of the multidimensional latent space to
-the maximally discriminating locus of the item.
-The second one, the direction of the vector from the origin to that same locus.
-Following a similar rationale, @reckase_discriminating_1991 derive a
-*multidimensional discrimination* (`r MDISC_PARAM`) parameter definition
-for that model.
+the maximally discriminating point of the item.
+The second one, the direction of the vector from the origin to that same point.
+Following a similar rationale, @reckase_discriminating_1991 define
+a *multidimensional discrimination* (MDISC) parameter definition for that model.
 
@@
 latent space are orthogonal
-[@ackerman_using_1994; @ackerman_multidimensional_2005; see also
-@reckase_difficulty_1985, p. 404 and Equation 7].
+[@ackerman_using_1994; @ackerman_multidimensional_2005; see also @reckase_difficulty_1985, p. 404 and Equation 7].
 Of course, orthogonality is a property of a geometrical space instead of a
@@
 the R package *psych* <!--# Reference needed? -->graphically represent
-factor loadings on orthogonal axes even after an oblique rotation.
+factor loadings on orthogonal axes, even after an oblique rotation.
 Nevertheless, the assimilation of correlated latent traits to non-orthogonal
@@
 This fact is probably due to the different spotlights
-these two theoretical programs put the focus on;
-factor analysis aims at a clear interpretation of
+the two theoretical programs put the focus on;
+Factor analysis aims at a clear interpretation of
 the latent covariance structure, and thus focuses on finding
@@
 Nevertheless, the confluence of the two theoretical approaches into a common
-modern measurement theory [@mcdonald_test_1999] requires
+modern measurement theory [@mcdonald_test_1999] imposes
 the relaxation of traditional constrains on both approaches.
@@
 [see @zhang_theoretical_1999 for a notable exception].
-Moreover, disregarding this geometrical interpretation may lead to
+What is more, disregarding this geometrical interpretation may lead to
 misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->
 
-A latent space where the traits are correlated is easier to understand as a
-geometrical space with non-orthogonal axes,
+A latent space where the traits are correlated is easier to represent in
+a geometrical space with non-orthogonal axes,
 often referred to as *general Cartesian coordinates* [@harman_modern_1970].
@@
 The general case of MIRT model estimation implies estimating possibly
-non-standard latent dimensions and non-null covariances among them
-(a more realistic case than an identity matrix).
-Assuming orthonormality among traits is usually unrealistic, be it in the
+non-standard latent dimensions and non-null covariances among them,
+a more realistic case than an identity matrix.
+Assuming orthogonality among traits is usually unrealistic, be it in the
 cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
@@
 interpreting the latent dimensions as substantive constructs is not a goal.
-As we will argue, the usual definitions of the multidimensional difficulty and
-discrimination parameters do not generalize to the non-orthonormal geometry
-necessary to accurately represent correlated latent spaces.
+As we will argue, the usual definitions of the
+MID and MDISC parameters do not generalize to the non-orthogonal geometry used
+to represent correlated latent spaces.
 
@@
 graphically representing these items in the fashion of @ackerman_creating_1994,
-but taking the non-orthogonality of the coordinate axes into account.
+but taking into account the non-orthogonality of the coordinate axes.
 In order to pursue these objectives, we first define the generalization of
@@
 propose a graphical representation method,
-and present two examples of application of our results.
+and present an example of computation<!-- # TODO: Pending -->
+and representation of the items based on these parameters.
 We round up with a discussion about
@@
 
-where $`r DISCR_VECTOR`$ is an $`r N_DIMS`$-dimensional vector of discrimination
+where $`r DISCR_VECTOR`$ is a $`r N_DIMS`$-dimensional vector of discrimination
 parameters, and $`r INTERCEPT_PARAM`$ an intercept parameter related to the
@@
 non-negative real numbers in Reckase
-[-@reckase_difficulty_1985, a decrease in the probability of a correct response
-with an increase in ability makes little sense].
+[-@reckase_difficulty_1985, a decrease in the probability of a correct response with an increase in ability makes little sense].
 In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
@@
 Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
-We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each basis,
-$`r LS_BASIS`$ or $`r LS_STD_BASIS`$.
+We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each baseis.
 Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
@@
 
-Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix between the two bases;
-then, we have that
+Let $`r BASIS_CHANGE_DEF`$ be the change of basis matrix; then, we have that
 
@@
 The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
-coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.<!--# Does this need to be justified? -->
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.
 If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
@@
 
-With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product
-(i.e., dot product).
+With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product.
 With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
@@
 norm invariant.
+
 The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
@@
 
-Then, we have that $`r BASIS_CHANGE_ORG_ALT_EQ`$.
+Then, we have that $`r BASIS_CHANGE_ORG_ALT_EQ`$. (ERROR aca)
 
@@
 Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
-already assumes the orthonormality of the latent space.
+already assumes the orthogonality of the latent space.
 
@@
 
-We are thus considering the direction cosines with respect to the orthonormal
+We are thus considering the direction cosines with respect to the orthogonal
 axes, so we can just apply @reckase_difficulty_1985's results, obtaining
@@
 
-Applying Proposition 1,
+Applying propositon 1,
 
@@
 
-with $`r DIAG_MATRIX_INNER_PROD_INV`$ a diagonal matrix where
-$`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
+with $`r DIAG_MATRIX_INNER_PROD_INV`$ the diagonal matrix with
+the same diagonal as $`r INNER_PROD_MATRIX_INV`$.
 

### 2023-02-15 output/paper_draft/paper_M2PLM_parameters.Rmd
agregados y coemntarios

--- 2023-02-10 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
+OBS.
+En la ecuación debe estar el producto inteno con el subindice que indique que es en la base B.
+No se ponen en las normas debido a la invarianza de estas.
+Como $\theta$ son las coordanadas en B se dejan igual.
+
 \begin{equation}
@@
 For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,
+
+OBS en esta primera ecuación recordar poner el subindice B en el producto interno
 
@@
 ## Test space and latent space
+
+AGREGADO
+Aca pongo redacciön alternativa dejo igual la origunal debajo
+
+Before defining the multidimensional parameters, we need to make a clarification
+regarding the representation of the items in the latent space.
+
+We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of the discrimination vectors.
+This space is where the item vectors can be properly represented.
+It is in general different from the latent space $`r LATENT_SPACE`$,
+although there is a bijective relationship between the two sets of spaces
+Let us consider the model given by the equation (1 poner la referencia) where the latent traits and  the parameters $a_i$ are expressed in the same base B in their respective spaces.
+
+We have  $\theta^{'}=P \theta$ so since the probability is invariant, the parameters $a_i$ must also be transformed in $a^{'}_i$.
+
+Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
+This shows that first a base change to the standard base is performed and then a transformation through $M^{-1}$, so the $a^{'}_i$ are expressed in the standard base.
+
+Then,
+
+\begin{equation}
+\|a_i^{'} \|=a_i^T P^{-1}P^{-1^T}a_i=a_i^TM^{-1}a_i
+(\#eq:slope-param)
+\end{equation}
+
+Therefore, in the test space, a inner product matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined,
+such that $`r INNER_PROD_DISCR_EQ`$ to keep the norm invariant
+
+With this definition we have that
+
+\begin{equation}
+  \begin{split}
+    `r DIR_COS_ITEM_VEC_NORM_EQ` \\
+    `r DISTANCE_PARAM_NORM_EQ`   \\
+    `r SLOPE_MAX_PARAM_NORM_EQ`.
+  \end{split}                             (\#eq:item-params-discrimination-norm)
+\end{equation}
+
+Applying Proposition 1,
+
+\begin{equation}
+  `r DIR_COS_ITEM_VEC_TS_EQ`,                      (\#eq:dir-cos-vec-test-space)
+\end{equation}
+
+with $`r DIAG_MATRIX_INNER_PROD_INV`$ a diagonal matrix where
+$`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
+
+Las vk no deben ir
+
+Esta es la original
 

### 2023-02-16 output/paper_draft/paper_M2PLM_parameters.Rmd
notacion

@@
 Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
-$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$, where $`r TRAIT_COORDS_EQ`$ and
+$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$. From now on we  will note  $`r TRAIT_COORDS_EQ`$ and
 $`r TRAIT_COORDS_STD_EQ`$.

### 2023-02-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Puts the product of norms (of the unitary vector and the theta vector) always in the same order, and explicitly states that the inner product is in basis B

@@
 Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
-$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$. From now on we  will note  $`r TRAIT_COORDS_EQ`$ and
-$`r TRAIT_COORDS_STD_EQ`$.
+$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$.
+From now on, we denote $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_STD_EQ`$.
 
@@
 
-OBS.
-En la ecuación debe estar el producto inteno con el subindice que indique que es en la base B.
-No se ponen en las normas debido a la invarianza de estas.
-Como $\theta$ son las coordanadas en B se dejan igual.
-
 \begin{equation}
@@
 For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,
-
-OBS en esta primera ecuación recordar poner el subindice B en el producto interno
 

### 2023-02-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Reorders and rewrites (WIP) the "test space" section

@@
 
-## Model recasting into polar coordinates
-
-From Proposition 1,
-any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
-
-\begin{equation}
-  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
-\end{equation}
-
-Thus, the M2PL model is expressed in polar coordinates as
-
-\begin{equation}
-  `r IRF_POLAR_EQ`.                                    (\#eq:M2PL-formula-polar)
-\end{equation}
-
-As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
-involved in the expression of the model in polar coordinates.
-Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
-already assumes the orthonormality of the latent space.
-
-## Point of maximum slope
-
-Following the procedure outlined by @reckase_difficulty_1985
-[@reckase_discriminating_1991],
-we first compute the point of maximum slope by finding the root(s) of
-the second derivative with respect to $`r TRAIT_NORM`$.
-
-The slope in direction $`r DIR_ANGLE_VEC`$ is given by
-
-\begin{equation}
-  `r IRF_1ST_DIFF_EQ`,                                      (\#eq:M2PL-1st-diff)
-\end{equation}
-
-and
-
-\begin{equation}
-  `r IRF_2ND_DIFF_EQ`,                                      (\#eq:M2PL-2nd-diff)
-\end{equation}
-
-The root we are interested in is found for $`r IRF_MAX_SLOPE_EQ`$
-(the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
-which result in improper values of $`r TRAIT_NORM`$).
-
-The slope in direction $`r DIR_ANGLE_VEC`$ when $`r IRF_MAX_SLOPE_EQ`$ is given
-by
-
-\begin{equation}
-  `r SLOPE_MAX_EQ`.                                             (\#eq:slope-max)
-\end{equation}
-
-Notation aside, these equations differ from
-Reckase's [-@reckase_difficulty_1985] Equations 3 through 6
-only in the term $`r INNER_PROD_INV_DIAG_SQ_PROD`$.
-
-To compute the direction of the maximum slope,
-Reckase leverages on the property that the squared direction cosines sum up to 1
-(see Equation 7 therein).
-However, this assertion implicitly assumes orthogonal coordinates,
-so it does not apply in general to $`r DIR_COS_VEC`$.
-Instead, we may consider the direction cosine vector with the standard basis
-$`r LS_STD_BASIS`$.
-Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain
-
-\begin{equation}
-  \begin{split}
-    `r DIR_COS_AS_DIR_COS_STD_EQ` \\
-    `r TRAIT_POLAR_COEFF_COS_STD_EQ`.
-  \end{split}                                 (\#eq:direction-cosine-vectors-eq)
-\end{equation}
-
-Therefore, substituting in \@ref(eq:slope-max),
-
-\begin{equation}
-  `r SLOPE_MAX_STD_COSINES_EQ`.                     (\#eq:slope-max-std-cosines)
-\end{equation}
-
-We can then define
-
-\begin{equation}
-  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
-\end{equation}
-
-so
-
-\begin{equation}
-  `r SLOPE_MAX_STD_EQ`.                                     (\#eq:slope-max-std)
-\end{equation}
-
-We are thus considering the direction cosines with respect to the orthonormal
-axes, so we can just apply @reckase_difficulty_1985's results, obtaining
-
-\begin{equation}
-  `r DIR_COS_VEC_STD_EQ`.                                 (\#eq:dir-cos-vec-std)
-\end{equation}
-
-Substituting $`r DIR_COS_VEC_STD`$
-(Equation \@ref(eq:direction-cosine-vectors-eq)) and
-$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
-by their respective expressions results in the
-expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
-
-\begin{equation}
-  `r DIR_COS_ITEM_VEC_EQ`,                                    (\#eq:dir-cos-vec)
-\end{equation}
-
-which gives the direction from the origin to the point where
-the slope is maximum.
-Finally, to determine the signed distance $`r DISTANCE_PARAM`$
-from the origin to that point,
-we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$ to get
-$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
-
-\begin{equation}
-  `r DISTANCE_PARAM_EQ`,                                   (\#eq:distance-param)
-\end{equation}
-
-The slope $`r SLOPE_MAX_PARAM`$ at the point defined by Equations
-\@ref(eq:dir-cos-vec) and \@ref(eq:distance-param)
-is given by Equation \@ref(eq:slope-max).
-Substituting Equation \@ref(eq:dir-cos-vec) in Equation \@ref(eq:slope-max),
-[@reckase_discriminating_1991],
-
-\begin{equation}
-  `r SLOPE_MAX_PARAM_EQ`.                                     (\#eq:slope-param)
-\end{equation}
-
-# Results
-
 ## Test space and latent space
 
-AGREGADO
-Aca pongo redacciön alternativa dejo igual la origunal debajo
+********************************************************************************
 
@@
 
-We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of the discrimination vectors.
+We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of the
+discrimination vectors.
 This space is where the item vectors can be properly represented.
 It is in general different from the latent space $`r LATENT_SPACE`$,
-although there is a bijective relationship between the two sets of spaces
-Let us consider the model given by the equation (1 poner la referencia) where the latent traits and  the parameters $a_i$ are expressed in the same base B in their respective spaces.
-
-We have  $\theta^{'}=P \theta$ so since the probability is invariant, the parameters $a_i$ must also be transformed in $a^{'}_i$.
+although there is a bijective relationship between the two sets of spaces.
+Let us consider the model given by Equation \@ref(eq:M2PL-formulation),
+where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
+in the same basis $`r LS_BASIS`$, in their respective spaces.
+<!-- # TODO: Discuss; don't think this is actually true -->
+
+We have  $\theta^{'}=P \theta$ so since the probability is invariant,
+the parameters $a_i$ must also be transformed in $a^{'}_i$.
 
@@
 
-Esta es la original
+********************************************************************************
 
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
+
+## Model recasting into polar coordinates
+
+From Proposition 1,
+any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
+
+\begin{equation}
+  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
+\end{equation}
+
+Thus, the M2PL model is expressed in polar coordinates as
+
+\begin{equation}
+  `r IRF_POLAR_EQ`.                                    (\#eq:M2PL-formula-polar)
+\end{equation}
+
+As Equation \@ref(eq:M2PL-formula-polar) shows, the inner product matrix is
+involved in the expression of the model in polar coordinates.
+Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
+already assumes the orthonormality of the latent space.
+
+## Point of maximum slope
+
+Following the procedure outlined by @reckase_difficulty_1985
+[@reckase_discriminating_1991],
+we first compute the point of maximum slope by finding the root(s) of
+the second derivative with respect to $`r TRAIT_NORM`$.
+
+The slope in direction $`r DIR_ANGLE_VEC`$ is given by
+
+\begin{equation}
+  `r IRF_1ST_DIFF_EQ`,                                      (\#eq:M2PL-1st-diff)
+\end{equation}
+
+and
+
+\begin{equation}
+  `r IRF_2ND_DIFF_EQ`,                                      (\#eq:M2PL-2nd-diff)
+\end{equation}
+
+The root we are interested in is found for $`r IRF_MAX_SLOPE_EQ`$
+(the other ones occur when $`r IRF_ABBR`$ equals 0 and 1,
+which result in improper values of $`r TRAIT_NORM`$).
+
+The slope in direction $`r DIR_ANGLE_VEC`$ when $`r IRF_MAX_SLOPE_EQ`$ is given
+by
+
+\begin{equation}
+  `r SLOPE_MAX_EQ`.                                             (\#eq:slope-max)
+\end{equation}
+
+Notation aside, these equations differ from
+Reckase's [-@reckase_difficulty_1985] Equations 3 through 6
+only in the term $`r INNER_PROD_INV_DIAG_SQ_PROD`$.
+
+To compute the direction of the maximum slope,
+Reckase leverages on the property that the squared direction cosines sum up to 1
+(see Equation 7 therein).
+However, this assertion implicitly assumes orthogonal coordinates,
+so it does not apply in general to $`r DIR_COS_VEC`$.
+Instead, we may consider the direction cosine vector with the standard basis
+$`r LS_STD_BASIS`$.
+Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain
+
+\begin{equation}
+  \begin{split}
+    `r DIR_COS_AS_DIR_COS_STD_EQ` \\
+    `r TRAIT_POLAR_COEFF_COS_STD_EQ`.
+  \end{split}                                 (\#eq:direction-cosine-vectors-eq)
+\end{equation}
+
+Therefore, substituting in \@ref(eq:slope-max),
+
+\begin{equation}
+  `r SLOPE_MAX_STD_COSINES_EQ`.                     (\#eq:slope-max-std-cosines)
+\end{equation}
+
+We can then define
+
+\begin{equation}
+  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
+\end{equation}
+
+so
+
+\begin{equation}
+  `r SLOPE_MAX_STD_EQ`.                                     (\#eq:slope-max-std)
+\end{equation}
+
+We are thus considering the direction cosines with respect to the orthonormal
+axes, so we can just apply @reckase_difficulty_1985's results, obtaining
+
+\begin{equation}
+  `r DIR_COS_VEC_STD_EQ`.                                 (\#eq:dir-cos-vec-std)
+\end{equation}
+
+Substituting $`r DIR_COS_VEC_STD`$
+(Equation \@ref(eq:direction-cosine-vectors-eq)) and
+$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
+by their respective expressions results in the
+expression for the direction cosine vector of item $i$ in $`r LATENT_SPACE`$,
+
+\begin{equation}
+  `r DIR_COS_ITEM_VEC_EQ`,                                    (\#eq:dir-cos-vec)
+\end{equation}
+
+which gives the direction from the origin to the point where
+the slope is maximum.
+Finally, to determine the signed distance $`r DISTANCE_PARAM`$
+from the origin to that point,
+we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$ to get
+$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
+
+\begin{equation}
+  `r DISTANCE_PARAM_EQ`,                                   (\#eq:distance-param)
+\end{equation}
+
+The slope $`r SLOPE_MAX_PARAM`$ at the point defined by Equations
+\@ref(eq:dir-cos-vec) and \@ref(eq:distance-param)
+is given by Equation \@ref(eq:slope-max).
+Substituting Equation \@ref(eq:dir-cos-vec) in Equation \@ref(eq:slope-max),
+[@reckase_discriminating_1991],
+
+\begin{equation}
+  `r SLOPE_MAX_PARAM_EQ`.                                     (\#eq:slope-param)
+\end{equation}
+
+# Results
 

### 2023-02-21 output/paper_draft/paper_M2PLM_parameters.Rmd
Moves the definition of the cosine vectors in the standard basis where they are needed for the first time.

--- 2023-02-17 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-Let's consider now the vectors $`r DIR_COS_VEC`$ and $`r DIR_COS_VEC_STD`$, made
-up by the $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with the basis
-vectors of $`r LS_BASIS`$ and $`r LS_STD_BASIS`$, respectively.
+Let's consider now the vector $`r DIR_COS_VEC`$, made up by the $`r N_DIMS`$
+direction cosines of $`r TRAIT_VECTOR`$ with
+the basis vectors of $`r LS_BASIS`$.
 
@@
 
-Let $`r DIR_COS_VEC_ALT`$ be the vector of direction cosines of
-$`r TRAIT_VECTOR`$ in $`r ALT_BASIS`$ (i.e., the cosines with the elements of
-$`r ALT_BASIS`$).
+Let $`r DIR_COS_VEC_STD`$ and $`r DIR_COS_VEC_ALT`$ be the vectors of
+direction cosines of $`r TRAIT_VECTOR`$ in $`r LS_STD_BASIS`$ and
+$`r ALT_BASIS`$, respectively.
 

### 2023-02-21 output/paper_draft/paper_M2PLM_parameters.Rmd
Generalizes the standard basis to "any orthonormal basis"
Changes the notation to distinguish a vector and its coordinates in a certain basis (WIP)

@@
   echo     = FALSE,
-  cache    = FALSE,
+  cache    = TRUE,
   dev      = "png",
@@
 
-Let's consider the standard basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$.
-We can write all $`r TRAIT_VEC_IN_LS`$ as a linear combination of each basis,
-$`r LS_BASIS`$ or $`r LS_STD_BASIS`$.
-Therefore, we have both representations: $`r TRAIT_VECTOR_DEF_BASIS_EQ`$, and
-$`r TRAIT_VECTOR_DEF_STD_BASIS_EQ`$.
-From now on, we denote $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_STD_EQ`$.
+Let's consider an orthonormal basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$
+with the dot product as inner product.
+Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_STD_EQ`$ be the coordinates of
+any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_STD_BASIS`$,
+respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
 
@@
 The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
-coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$.<!--# Does this need to be justified? -->
-If $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$;
+formally, if $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
 
@@
 
-With the standard basis in $`r LATENT_SPACE`$ we have the usual scalar product
-(i.e., dot product).
-With basis $`r LS_BASIS`$ in $`r LATENT_SPACE`$ we define an inner product
-$`r LS_BASIS_INNER_PROD`$, such that the norm of $`r TRAIT_VECTOR`$ is
-invariant, i.e., $`r TRAIT_NORM_EQ`$. Therefore,
-
-\begin{equation}
-  `r TRAIT_NORM_STD_SQ_EQ`.                             (\#eq:inner-product-def)
-\end{equation}
-
-Thus, we must define the inner product $`r LS_BASIS_INNER_PROD`$ with Gram
-matrix $`r INNER_PROD_MAT_EQ`$. That is, $`r INNER_PROD_EQ`$, which keeps the
-norm invariant.
-The Gram matrix has $`r INNER_PROD_MAT_ELEMENT_EQ`$ as element. As
+As the norm of $`r TRAIT_VECTOR`$ must be invariant (i.e., $`r TRAIT_NORM_EQ`$),
+we need to define $`r LS_BASIS`$ in $`r LATENT_SPACE`$ with inner product
+$`r LS_BASIS_INNER_PROD`$ as
+
+\begin{equation}
+  `r TRAIT_NORM_SQ_EQ`.                                 (\#eq:inner-product-def)
+\end{equation}
+
+Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
+Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
+the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
 $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
@@
 
-The cosines between two vectors follow the usual definition,
+The cosine between two vectors
 
@@
 
-independently of whether the inner product is standard or not.
+is independent of whether the inner product is standard or not.
 Therefore, to compute the direction cosine $`r DIR_COS_ANY`$

### 2023-02-21 R/Formulae.R

@@ -122,7 +122,6 @@ TRAIT_VECTOR_DEF_BASIS <- latex_summation(
   index = DIM_INDEX, from = 1, to = N_DIMS,
   .par = FALSE
 )
-TRAIT_VECTOR_DEF_BASIS_EQ <- latex_eq(TRAIT_VECTOR, TRAIT_VECTOR_DEF_BASIS)
 
 # Vector definition in standard basis:
 TRAIT_VEC_STD_COMP            <- latex_raised_to(
@@ -135,16 +134,20 @@ TRAIT_VECTOR_DEF_STD_BASIS    <- latex_summation(
   index = DIM_INDEX, from = 1, to = N_DIMS,
   .par = FALSE
 )
-TRAIT_VECTOR_DEF_STD_BASIS_EQ <- latex_eq(
+
+TRAIT_VECTOR_DEF_EQ <- latex_eq(
   TRAIT_VECTOR,
+  TRAIT_VECTOR_DEF_BASIS,
   TRAIT_VECTOR_DEF_STD_BASIS
 )
 
 # Trait coordinates in both original and standard bases:
+TRAIT_COORDS         <- latex_raised_to(TRAIT_VECTOR, exp = LS_BASIS)
 TRAIT_COORDS_DEF     <- latex_coords(TRAIT_VECTOR, LS_BASIS)
-TRAIT_COORDS_EQ      <- latex_eq(TRAIT_VECTOR, TRAIT_COORDS_DEF)
+TRAIT_COORDS_EQ      <- latex_eq(TRAIT_COORDS, TRAIT_COORDS_DEF)
+TRAIT_STD_COORDS     <- latex_raised_to(TRAIT_VECTOR, exp = LS_STD_BASIS)
 TRAIT_COORDS_STD_DEF <- latex_coords(TRAIT_VECTOR, LS_STD_BASIS)
-TRAIT_COORDS_STD_EQ  <- latex_eq(TRAIT_VECTOR_STD, TRAIT_COORDS_STD_DEF)
+TRAIT_COORDS_STD_EQ  <- latex_eq(TRAIT_STD_COORDS, TRAIT_COORDS_STD_DEF)
 
 # Change of basis matrix:
 TRANSFORM_MATRIX <- latex_bf('P')
@@ -152,22 +155,30 @@ BASIS_CHANGE_SYM <- latex_basis_change(LS_BASIS, LS_STD_BASIS)
[…]

### 2023-02-21 output/paper_draft/paper_M2PLM_parameters.Rmd
Drops the use of the term "Euclidean" (not used in the whole paper)

@@
 $`r LATENT_SPACE`$ and the inner product $`r ALT_BASIS_INNER_PROD`$ that keeps
-the norm of the Euclidean space invariant.
+the norm of the space invariant.
 

### 2023-02-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Updates the notation to consistently refer to the coordinates of the vectors in the latent space referred to a concrete basis

@@
   echo     = FALSE,
-  cache    = TRUE,
+  cache    = FALSE,
   dev      = "png",
@@
 is independent of whether the inner product is standard or not.
-Therefore, to compute the direction cosine $`r DIR_COS_ANY`$
+Therefore, to compute the direction cosine $`r DIR_COS_ANY_ORG`$
 of vector $`r TRAIT_VECTOR`$ along dimension $`r DIM_INDEX`$,
@@
 
-Let's consider now the vector $`r DIR_COS_VEC`$, made up by the $`r N_DIMS`$
+Let's consider now the vector $`r DIR_COS_VEC_ORG`$, made up by the $`r N_DIMS`$
 direction cosines of $`r TRAIT_VECTOR`$ with
@@
 However, this assertion implicitly assumes orthogonal coordinates,
-so it does not apply in general to $`r DIR_COS_VEC`$.
+so it does not apply in general to $`r DIR_COS_VEC_ORG`$.
 Instead, we may consider the direction cosine vector with the standard basis

### 2023-02-24 R/Formulae.R

@@ -102,7 +102,6 @@ M2PL_FORMULATION <- latex_eq(IRF_M2PL, IRF_ABBR, LOGISTIC_M2PL)
 STD_BASIS_VECTOR_EL   <- latex('e')
 STD_BASIS_VECTOR      <- latex_bf(STD_BASIS_VECTOR_EL)
 LATENT_SPACE_STD      <- latex_prime(LATENT_SPACE)
-TRAIT_VECTOR_STD      <- latex_prime(TRAIT_VECTOR)
 LS_STD_BASIS          <- latex_cal('E')
 LS_STD_BASIS_ELEMENTS <- latex_enum(
   latex_sub(STD_BASIS_VECTOR, 1),
@@ -199,8 +198,11 @@ TRAIT_NORM_EQ           <- latex_eq(
 # Latent vector invariance condition:
 TRAIT_NORM              <- latex_norm(TRAIT_VECTOR)
 TRAIT_NORM_SQ           <- latex_squared(TRAIT_NORM)
-TRAIT_VECTOR_STD_TRANSP <- latex_transp(TRAIT_VECTOR_STD)
-TRAIT_VECTOR_STD_SQ     <- latex(TRAIT_VECTOR_STD_TRANSP, TRAIT_VECTOR_STD)
+TRAIT_VECTOR_STD_TRANSP <- latex_raised_to(
+  TRAIT_VECTOR, # TODO: Transposing a "vector in basis" may typeset wrongly
+  exp = "{$LS_STD_BASIS$T}"
+)
+TRAIT_VECTOR_STD_SQ     <- latex(TRAIT_VECTOR_STD_TRANSP, TRAIT_STD_COORDS)
 TRAIT_VECTOR_TRANSP     <- latex_transp(TRAIT_VECTOR)
 INNER_PROD_TRANSF_DEF   <- latex(TRANSFORM_MATRIX_TRANSP, TRANSFORM_MATRIX)
 TRAIT_TRANSF_TRAIT      <- latex(
@@ -282,7 +284,8 @@ COS_VECTORS_EQ  <- latex_eq(COS_VECTORS, COS_VECTORS_DEF)
 
 # Generic direction cosine of a vector:
 DIR_ANGLE_ANY              <- latex_sub(ANGLE, DIM_INDEX)
[…]

### 2023-02-27 output/IMPS-2023_proposal/IMPS-2023_abstract.Rmd
Versions the abstract for the IMPS 2023 proposal

---
title: A generalization of multidimensional item response theory parameters
author: "Daniel Morillo, Mario Luzardo"
date: "2023-02-27"
output: word_document
bibliography: ../../www/Multidimensional-parameters-MCLM.bib
csl:          ../../www/apa-old-doi-prefix.csl
---

# Abstract

```{r setup}
#| include: false

# Includes:

library(knitr)
library(rstudioapi)

ROOT_DIR <- getActiveProject()
DOC_DIR  <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE
)
```

```{r sources}
#| cache:   false
#| include: false
source("R/Constants.R", encoding = 'UTF-8')
```

The Multidimensional 2-Parameter Logistic
[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is one of the most important
and widely used models in Multidimensional Item Response Theory (MIRT).
@reckase_difficulty_1985 and @reckase_discriminating_1991 gave a precise
definition of a Multidimensional Item Difficulty (MID) and Multidimensional
Discrimination (MDISC) parameter, respectively, and derived their expressions
for the M2PL model.
However, we argue here that (1) their derivations implicitly assume the
orthonormality of the latent space, and (2) it is straightforwardly extensible
to applications where the items have marginally unchanging monotony, instead of
just to monotonously increasing items (which is important for non-cognitive
applications).

We derive generalized expressions of those parameters, that apply to any item
with unchanging monotony in any non-orthonormal space, thus allowing to properly
represent disparate covariance structures.
We then use these definitions in two applications: to represent M2PL items in
non-orthogonal spaces, and to compute the multidimensional parameters in various
applications, showing how they provide better information compared with their
"covariance-agnostic" counterparts.

These results become increasingly relevant as the modern measurement theory in
social and behavioral sciences experiences as a convergence between MIRT and the
common factor model. Nevertheless, we warn against generalizing our results to
other MIRT models, beyond the M2PL, without carefully considering each
particular case.

# References

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Specializes the cosine direction angles for the item (adding the item subindex) where appropriate

@@
 
-The slope in direction $`r DIR_ANGLE_VEC`$ when $`r IRF_MAX_SLOPE_EQ`$ is given
-by
+The slope in direction $`r DIR_ANGLE_ITEM_VEC_EQ`$ when $`r IRF_MAX_SLOPE_EQ`$
+is given by
 
@@
 
-Substituting $`r DIR_COS_VEC_STD`$
+Substituting $`r DIR_COS_ITEM_VEC_STD`$
 (Equation \@ref(eq:direction-cosine-vectors-eq)) and

### 2023-03-03 R/Formulae.R

@@ -631,9 +631,19 @@ TRAIT_POLAR_COEFF_COS_STD_EQ   <- latex_eq(
 )
 
 # Maximum slope:
-SLOPE_MAX     <- latex_sub("\\left. $IRF_1ST_DIFF$ \\right|", IRF_MAX_SLOPE_EQ)
-SLOPE_MAX_DEF <- latex(FRAC_1_4, DISCR_VEC_TRAIT_COEFF)
-SLOPE_MAX_EQ  <- latex_eq(SLOPE_MAX, SLOPE_MAX_DEF)
+SLOPE_MAX            <- latex_sub(
+  "\\left. $IRF_1ST_DIFF$ \\right|",
+  IRF_MAX_SLOPE_EQ
+)
+DIR_ANGLE_ITEM_VEC   <- latex_sub(DIR_ANGLE_VEC_ORG, ITEM_INDEX)
+DIR_COS_ITEM_VEC     <- latex_cos(DIR_ANGLE_ITEM_VEC)
+ITEM_POLAR_COEFF     <- latex(
+  INNER_PROD_INV_DIAG_SQ_PROD,
+  DIR_COS_ITEM_VEC
+)
+DISCR_VEC_ITEM_COEFF <- latex(DISCR_VECTOR_TRANSP, ITEM_POLAR_COEFF)
+SLOPE_MAX_DEF        <- latex(FRAC_1_4, DISCR_VEC_ITEM_COEFF)
+SLOPE_MAX_EQ         <- latex_eq(SLOPE_MAX, SLOPE_MAX_DEF)
 
 # Standardized discrimination vector:
 DISCR_VECTOR_STD     <- latex_prime(DISCR_VECTOR)
@@ -641,30 +651,36 @@ DISCR_VECTOR_STD_DEF <- latex(TRANSFORM_MATRIX_TRANSP_INV, DISCR_VECTOR)
 DISCR_VECTOR_STD_EQ  <- latex_def(DISCR_VECTOR_STD, DISCR_VECTOR_STD_DEF)
 
 # Maximum slope as a function of standardized discrimination vectors:
-DISCR_VECTOR_STD_TRANSP   <- latex_transp(DISCR_VECTOR_STD)
-DISCR_VEC_TRAIT_STD_COEFF <- latex(DISCR_VECTOR_STD_TRANSP, DIR_COS_VEC_STD)
[…]

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Reverts aggregation by Mario in section "Test space and latent space" in 6ac4c2a5

@@
 
-AGREGADO
-Aca pongo redacciön alternativa dejo igual la origunal debajo
-
-Before defining the multidimensional parameters, we need to make a clarification
-regarding the representation of the items in the latent space.
-
-We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of the discrimination vectors.
-This space is where the item vectors can be properly represented.
-It is in general different from the latent space $`r LATENT_SPACE`$,
-although there is a bijective relationship between the two sets of spaces
-Let us consider the model given by the equation (1 poner la referencia) where the latent traits and  the parameters $a_i$ are expressed in the same base B in their respective spaces.
-
-We have  $\theta^{'}=P \theta$ so since the probability is invariant, the parameters $a_i$ must also be transformed in $a^{'}_i$.
-
-Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
-This shows that first a base change to the standard base is performed and then a transformation through $M^{-1}$, so the $a^{'}_i$ are expressed in the standard base.
-
-Then,
-
-\begin{equation}
-\|a_i^{'} \|=a_i^T P^{-1}P^{-1^T}a_i=a_i^TM^{-1}a_i
-(\#eq:slope-param)
-\end{equation}
-
-Therefore, in the test space, a inner product matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined,
-such that $`r INNER_PROD_DISCR_EQ`$ to keep the norm invariant
-
-With this definition we have that
-
-\begin{equation}
-  \begin{split}
-    `r DIR_COS_ITEM_VEC_NORM_EQ` \\
-    `r DISTANCE_PARAM_NORM_EQ`   \\
-    `r SLOPE_MAX_PARAM_NORM_EQ`.
-  \end{split}                             (\#eq:item-params-discrimination-norm)
-\end{equation}
-
-Applying Proposition 1,
-
-\begin{equation}
-  `r DIR_COS_ITEM_VEC_TS_EQ`,                      (\#eq:dir-cos-vec-test-space)
-\end{equation}
-
-with $`r DIAG_MATRIX_INNER_PROD_INV`$ a diagonal matrix where
-$`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
-
-Las vk no deben ir
-
-Esta es la original
-
 Before defining the multidimensional parameters, we need to make a clarification

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds again Mario's changes to section "Test space and latent space"

@@
 regarding the representation of the items in the latent space.
-Equation \@ref(eq:slope-param) makes it clear that a space $`r TEST_SPACE`$
-with basis $`r ITEM_SPACE_BASIS_EQ`$ and inner product matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined,
-such that $`r INNER_PROD_DISCR_EQ`$ and, subsequently,
-$`r DISCR_VECTOR_MODULE_EQ`$.
-We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of the
-discrimination vectors.
+We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of
+the discrimination vectors.
 This space is where the item vectors can be properly represented.
 It is in general different from the latent space $`r LATENT_SPACE`$,
-although there is a bijective relationship between the two sets of spaces,
-as their inner product matrices are inverse of one another.
-
+although there is a bijective relationship between the two sets of spaces.
+Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
+where the latent traits and the parameters $a_i$ are expressed in the same basis
+$`r LS_BASIS`$ in their respective spaces.
+
+Since the response probability must be invariant, because of Equation
+\@ref(eq:change-basis), parameter $`r DISCR_VECTOR`$ must also be transformed
+in $`r DISCR_VECTOR_STD`$.
+Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
+This shows that first a basis change to the standard base is performed and then
+a transformation through $`r INNER_PROD_MATRIX_INV`$,
+so the $`r DISCR_VECTOR_STD`$ are expressed in the standard basis.
+
+\begin{equation}
+  `r DISCR_VECTOR_MODULE_EQ`                     (\#eq:discr-param-squared-norm)
+\end{equation}
+
+Therefore, in the test space, an inner product matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined such that $`r INNER_PROD_DISCR_EQ`$,
+to keep the norm invariant.
 With this definition we have that
@@
 \end{equation}
-
-It is clear that $`r INNER_PROD_INV_MAT_ELEMENT_EQ`$ were
-$`r INNER_PROD_INV_MAT_ELEMENT`$ is the $`r INNER_PROD_INV_MAT_SUBINDEX`$-th
-element of $`r INNER_PROD_MATRIX_INV`$.
-
-Let $`r LS_STD_BASIS`$ be the standard basis of $`r TEST_SPACE`$.
-Then, $`r BASIS_CHANGE_TS_EQ`$ is the change of basis matrix and
-$`r DISCR_VECTOR_STD_MOD_EQ`$.
 

### 2023-03-03 R/Formulae.R

@@ -742,10 +742,17 @@ INNER_PROD_DISCR_DEF <- latex(
 INNER_PROD_DISCR_EQ  <- latex_eq(INNER_PROD_DISCR, INNER_PROD_DISCR_DEF)
 
 # Discrimination vector module:
-DISCR_VECTOR_MODULE    <- latex_norm(DISCR_VECTOR)
-DISCR_VECTOR_MODULE_SQ <- latex_squared(DISCR_VECTOR_MODULE)
-DISCR_VECTOR_MODULE_EQ <- latex_eq(
+DISCR_VECTOR_MODULE      <- latex_norm(DISCR_VECTOR)
+DISCR_VECTOR_MODULE_SQ   <- latex_squared(DISCR_VECTOR_MODULE)
+DISCR_VECTOR_TRANSF_PROD <- latex(
+  DISCR_VECTOR_TRANSP,
+  TRANSFORM_MATRIX_INV,
+  TRANSFORM_MATRIX_TRANSP_INV,
+  DISCR_VECTOR
+)
+DISCR_VECTOR_MODULE_EQ   <- latex_eq(
   DISCR_VECTOR_MODULE_SQ,
+  DISCR_VECTOR_TRANSF_PROD,
   DISCR_VECTOR_INNER_PROD
 )
 

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Generalizes the transformation of the direction cosines to any orthonormal basis, instead of the standard basis

@@
 so it does not apply in general to $`r DIR_COS_VEC_ORG`$.
-Instead, we may consider the direction cosine vector with the standard basis
+Instead, we may consider the direction cosine vector with an orthonormal basis
 $`r LS_STD_BASIS`$.
-Now, applying Proposition 2 (with $`r ALT_MATRIX_AS_ID`$) we obtain
+Now, applying Proposition 2 (with $`r TRANSFORM_MATRIX_ALT`$
+an orthonormal matrix and $`r ALT_MATRIX_AS_ID`$) we obtain
 

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 library(patchwork)
-
-ROOT_DIR <- "../.."
+library(rstudioapi)
+
+ROOT_DIR <- getActiveProject()
 DOC_DIR  <- getwd()
@@
   dpi      =   300,
-  dev.args = list(type = "cairo-png"),
-  fig.pos = "!H"
+  dev.args = list(type = "cairo-png")
 )
@@
 
-```{r sources, cache=FALSE, include=FALSE}
+```{r sources}
+#| cache:   false
+#| include: false
 source("R/Formulae.R",  encoding = 'UTF-8')
+source("R/Constants.R", encoding = 'UTF-8')
 ```
@@
 Following this procedure, he also derives
-a *multidimensional difficulty* parameter for M2PL items.
+a *multidimensional difficulty* parameter (MID) for `r MODEL_ACRONYM` items.
 This parameter consists of two parts:
@@
 
-```{r libraries, message=FALSE}
-```
-
-```{r sources, message=FALSE}
+```{r libraries}
+```
+
+```{r sources}
 ```
@@
 
-```{r libraries, message=FALSE}
-```
-
-```{r sources, message=FALSE}
-```
-
-```{r read-items, message=FALSE}
-```
-
-```{r orthogonal-params}
-```
-
-```{r oblique-params}
-```
-
-```{r params-table}
-```
-
-```{r compose-empirical-items-table}
-```
-
-```{r set-flextable-wd, cache=FALSE}
+```{r libraries}
+```
+
+```{r sources}
+```
+
+```{r general-constants}
+```
+
+```{r set-cov-matrix-Reckase2009}
+# Version in Reckase, 2009, p. 153; this represents a usual 3D case
+COV_MATRIX <- matrix(
+  c(
+    1.210, .297, 1.232,
+     .297, .810,  .252,
+    1.232, .252, 1.960
+  ),
+  nrow = 3
+)
+# (see also Reckase, 2009, p. 183 for a "close-to-1D" case).
+```
+
+```{r read-items}
+```
+
+```{r compute-orthogonal-params}
+```
+
+```{r compute-oblique-params}
+```
+
+```{r create-params-table}
+```
+
+```{r compose-output-table}
+```
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Reckase2009-out}
+#| cache:   false
+#| tab.id:  empirical-table-Reckase2009-out
+#| tab.cap: Item parameters for the empirical example
+item_params_output
+```
+
+```{r set-cov-matrix-Tezza2018}
+# See Tezza et al., 2018, p. 927:
+# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
+# > foi de aproximadamente 0,4.
+COV_MATRIX <- diag(4)
+COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
+```
+
+```{r reset-root-wd}
+#| cache: false
+```
+
+```{r read-items-Tezza2018}
+items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
+```
+
+```{r compute-orthogonal-params}
+```
+
+```{r compare-orthogonal-params}
+```
+
+<!-- Computed item parameters in the orthonormal space match the ones in given
+in Tezza 2018, table 5 -->
+
+```{r compute-oblique-params}
+```
+
+```{r create-params-table}
+```
+
+```{r compose-output-table}
+```
+
+```{r set-flextable-wd}
+#| cache: false
 ```
@@
 
-```{r empirical-items-table-out, cache=FALSE, tab.id="empirical-items-table-out", tab.cap="Item parameters for the empirical example"}
+```{r empirical-table-Tezza2018-out}
+#| cache:   false
+#| tab.id:  empirical-table-Tezza2018-out
+#| tab.cap: Item parameters for the empirical example
 item_params_output

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Adapts draft to work with the last changes merged from the main branch

@@
 The so called *Multidimensional 2-Parameter Logistic*
-[M2PL, @mckinley_extension_1983] model is among the most important
+[`r MODEL_ACRONYM`, @mckinley_extension_1983] model is among the most important
 and widely used ones.
@@
 Following a similar rationale, @reckase_discriminating_1991 derive a
-*multidimensional discrimination* (`r MDISC_PARAM`) parameter definition
+*multidimensional discrimination* (`r MDISC_SYM`) parameter definition
 for that model.
@@
 
-According to the M2PL model, the probability of a positive response to item
-$`r ITEM_INDEX`$ is
+According to the `r MODEL_ACRONYM` model, the probability of
+a positive response to item $`r ITEM_INDEX`$ is
 

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Changes the reference to the discrimination vector by its symbol

@@
 Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
-where the latent traits and the parameters $a_i$ are expressed in the same basis
-$`r LS_BASIS`$ in their respective spaces.
+where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
+in the same basis $`r LS_BASIS`$ in their respective spaces.
 

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
References Ackerman 1994 where it is mentioned the "independence of the geometrical representation from the latent space correlation"

@@
 a latent trait space, no matter its structure, as an orthogonal space
---exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
+[@ackerman_creating_1994]
+---exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
 the R package *psych* <!--# Reference needed? -->graphically represent

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
References Ackerman 1994 where it is mentioned the "independence of the geometrical representation from the latent space correlation"

@@
 
-Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$,
-defined by vector space $V$ and basis $`r LS_BASIS_EQ`$.
+Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
+in basis $`r LS_BASIS_EQ`$.
 A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
@@
 
-Let's consider an orthonormal basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$
+Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
 with the dot product as inner product.
-Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_STD_EQ`$ be the coordinates of
-any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_STD_BASIS`$,
+Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_ORTH_EQ`$ be the coordinates of
+any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_ORTH_BASIS`$,
 respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
@@
 The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
-coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$;
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_ORTH_BASIS`$;
 formally, if $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
@@
 
-with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
+with $`r STD_BASIS_VECTOR_ANY`$ the $`r DIM_INDEX`$-th standard unitary vector,
+and $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
 $`r INNER_PROD_MATRIX`$.
@@
 Let $`r DIR_COS_VEC_STD`$ and $`r DIR_COS_VEC_ALT`$ be the vectors of
-direction cosines of $`r TRAIT_VECTOR`$ in $`r LS_STD_BASIS`$ and
+direction cosines of $`r TRAIT_VECTOR`$ in $`r LS_ORTH_BASIS`$ and
 $`r ALT_BASIS`$, respectively.
@@
 
-1. $`r TRAIT_VEC_STD_POLAR_EQ`$
+1. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
 
@@
 \begin{equation}
-  `r TRAIT_VEC_STD_POLAR_EQ_XPAND`
+  `r TRAIT_VEC_ORTH_POLAR_EQ_XPAND`
 \end{equation}
@@
 \begin{equation}
-  `r TRAIT_VEC_STD_POLAR_ORG_EQ`
+  `r TRAIT_VEC_ORTH_POLAR_ORG_EQ`
 \end{equation}
@@
 \begin{equation}
-  `r TRAIT_VEC_STD_POLAR_ALT_EQ`.
+  `r TRAIT_VEC_ORTH_POLAR_ALT_EQ`.
 \end{equation}
@@
 Instead, we may consider the direction cosine vector with an orthonormal basis
-$`r LS_STD_BASIS`$.
+$`r LS_ORTH_BASIS`$.
 Now, applying Proposition 2 (with $`r TRANSFORM_MATRIX_ALT`$
@@
   \begin{split}
-    `r DIR_COS_AS_DIR_COS_STD_EQ` \\
-    `r TRAIT_POLAR_COEFF_COS_STD_EQ`.
+    `r DIR_COS_AS_DIR_COS_ORTH_EQ` \\
+    `r TRAIT_POLAR_COEFF_COS_ORTH_EQ`.
   \end{split}                                 (\#eq:direction-cosine-vectors-eq)
@@
 ```{r compare-orthogonal-params}
+#| results: hide
 ```

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds minor style changes

@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
+is usually implicit in the MIRT literature without further consideration<!--# Examples? -->
 [see @zhang_theoretical_1999 for a notable exception].
@@
 The general case of MIRT model estimation implies estimating possibly
-non-standard latent dimensions and non-null covariances among them
-(a more realistic case than an identity matrix).
+non-null covariances among latent dimensions, as well as non-standard scales.
 Assuming orthonormality among traits is usually unrealistic, be it in the
@@
 interpreting the latent dimensions as substantive constructs is not a goal.
-As we will argue, the usual definitions of the multidimensional difficulty and
-discrimination parameters do not generalize to the non-orthonormal geometry
+As we will argue, the usual definitions of the MID and MDISC parameters
+do not generalize to the non-orthonormal geometry
 necessary to accurately represent correlated latent spaces.
@@
 In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait, and thus
-are usually reverse scored-- is also widespread.<!--# Reference and example? -->
+--items that tap the negative pole of a certain trait,
+usually reverse scored-- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine
@@
 One can easily note that a monotonically decreasing probability for a certain
-dimension is given by a non-positive discrimination parameter
+dimension is given by a negative discrimination parameter
 (and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
@@
 The main objective of this paper is thus to provide a generalization of
-the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
+the `r MODEL_ACRONYM` multidimensional parameters in two aspects:
+On one hand, in an oblique coordinate system
 that can be soundly applied to the correlated trait case; on the other, to
@@
 as a function of how we define and interpret the latent space.
-Next we discuss the properties of the indices,
+Next we discuss the properties of these indices,
 propose a graphical representation method,
-and present two examples of application of our results.
+and present two practical examples applying them.
 We round up with a discussion about
@@
 
-We must make a few remarks here. First, as we are not constraining ourselves to
-cognitive traits, we speak of _positive_ instead of _correct response_.
-This may imply a correct response, if we are talking about a cognitive test, or
-endorsing an item (or a certain response option) in a non-cognitive domain.
+We must stop here and make a few remarks. First, as we are not constraining
+ourselves to cognitive traits,
+we speak of _positive_ instead of _correct response_. This may imply giving
+a correct response indeed, if we are talking about a cognitive test.
+But it may mean endorsing an item (or a certain response option)
+in a non-cognitive domain.
 Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$.
+latent dimension $`r DIM_INDEX`$ (e.g. a negatively-keyed item).
 Finally, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
 instead of the item difficulty.
@@
 because the slope is dependent on the direction considered.
-However, Reckase's Equation 3 implicitly assumes orthogonality.
+However, Reckase's Equation 3 implicitly assumes orthogonality, as we will see.
 In order to contemplate the general case,
@@
 Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
-Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
+Gram matrix, i.e. $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
 the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
 $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
-invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
+invertible, so is $`r INNER_PROD_MATRIX`$, and $`r INNER_PROD_MATRIX_INV_EQ`$.
 

### 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds minor style changes

@@
 
-Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
-in basis $`r LS_BASIS_EQ`$.
+Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$,
+defined by vector space $V$ and basis $`r LS_BASIS_EQ`$.
 A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
@@
 
-Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
+Let's consider an orthonormal basis $`r LS_STD_BASIS_EQ`$ in $`r LATENT_SPACE`$
 with the dot product as inner product.
-Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_ORTH_EQ`$ be the coordinates of
-any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_ORTH_BASIS`$,
+Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_STD_EQ`$ be the coordinates of
+any $`r TRAIT_VEC_IN_LS`$ in $`r LS_BASIS`$ and $`r LS_STD_BASIS`$,
 respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
@@
 The $`r DIM_INDEX`$-th column of $`r TRANSFORM_MATRIX`$ is given by the
-coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_ORTH_BASIS`$;
+coordinates of $`r BASIS_VECTOR_ANY`$ in basis $`r LS_STD_BASIS`$;
 formally, if $`r BASIS_VECTOR_ANY_TRANSF_EQ`$, then
@@
 
-with $`r STD_BASIS_VECTOR_ANY`$ the $`r DIM_INDEX`$-th standard unitary vector,
-and $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
+with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
 $`r INNER_PROD_MATRIX`$.
@@
 Let $`r DIR_COS_VEC_STD`$ and $`r DIR_COS_VEC_ALT`$ be the vectors of
-direction cosines of $`r TRAIT_VECTOR`$ in $`r LS_ORTH_BASIS`$ and
+direction cosines of $`r TRAIT_VECTOR`$ in $`r LS_STD_BASIS`$ and
 $`r ALT_BASIS`$, respectively.
@@
 
-1. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
+1. $`r TRAIT_VEC_STD_POLAR_EQ`$
 
@@
 \begin{equation}
-  `r TRAIT_VEC_ORTH_POLAR_EQ_XPAND`
+  `r TRAIT_VEC_STD_POLAR_EQ_XPAND`
 \end{equation}
@@
 \begin{equation}
-  `r TRAIT_VEC_ORTH_POLAR_ORG_EQ`
+  `r TRAIT_VEC_STD_POLAR_ORG_EQ`
 \end{equation}
@@
 \begin{equation}
-  `r TRAIT_VEC_ORTH_POLAR_ALT_EQ`.
+  `r TRAIT_VEC_STD_POLAR_ALT_EQ`.
 \end{equation}
@@
 Instead, we may consider the direction cosine vector with an orthonormal basis
-$`r LS_ORTH_BASIS`$.
+$`r LS_STD_BASIS`$.
 Now, applying Proposition 2 (with $`r TRANSFORM_MATRIX_ALT`$
@@
   \begin{split}
-    `r DIR_COS_AS_DIR_COS_ORTH_EQ` \\
-    `r TRAIT_POLAR_COEFF_COS_ORTH_EQ`.
+    `r DIR_COS_AS_DIR_COS_STD_EQ` \\
+    `r TRAIT_POLAR_COEFF_COS_STD_EQ`.
   \end{split}                                 (\#eq:direction-cosine-vectors-eq)
@@
 ```{r compare-orthogonal-params}
-#| results: hide
 ```

### 2023-03-10 output/paper_draft/paper_M2PLM_parameters.Rmd
Moves the "Test space" section before (temptatively)
Adds minor relevant cross-mention
Changes the "test space" section according to its precedence and the derivation based on the invariance of the M2PL model.

--- 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
+### Test space and latent space
+
+Before proceeding further, we need to make a clarification
+regarding the representation of the items in the latent space.
+Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
+where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
+in the same basis $`r LS_BASIS`$ in their respective spaces.
+The coordinates of $`r DISCR_VECTOR`$ are then $`r DISCR_COORDS`$.
+Since the response probability must be invariant with respect to
+a change of basis, $`r DISCR_COORDS`$ must also be transformed
+into $`r DISCR_STD_COORDS`$, such that $`r INVARIANCE_EQ`$;
+therefore, $`r DISCR_STD_COORDS_EQ`$.
+This shows that first a transformation through $`r INNER_PROD_MATRIX_INV`$
+is performed and then a change to the standard basis,
+so the $`r DISCR_STD_COORDS`$ are expressed in the standard basis.
+The norm of $`r DISCR_VECTOR`$ must then fulfill
+
+\begin{equation}
+  `r DISCR_VECTOR_MODULE_EQ`.                    (\#eq:discr-param-squared-norm)
+\end{equation}
+
+Therefore, an inner product $`r ITEM_SPACE_INNER_PROD`$ with Gram matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
+keeps the norm of $`r DISCR_VECTOR`$ invariant.
+The discrimination vectors are thus represented in their own space
+$`r TEST_SPACE`$ with basis $`r ITEM_SPACE_BASIS`$,
+which we refer to as the *test space* [@zhang_theoretical_1999].
+The test space is in general different from the latent space $`r LATENT_SPACE`$,
+although there is a bijective relationship between the two sets of spaces.
+
+### Direction cosines in the latent space
+
 The cosine between two vectors
@@
 Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
-already assumes the orthonormality of the latent space.
+already assumes the orthonormality of the latent space, as we stated before.
 
@@
 \begin{equation}
-  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
+  `r DISCR_STD_COORDS_EQ`                                (\#eq:discr-vector-std)
 \end{equation}
@@
 (Equation \@ref(eq:direction-cosine-vectors-eq)) and
-$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
+$`r DISCR_STD_COORDS`$ (Equation \@ref(eq:discr-vector-std))
 by their respective expressions results in the
@@
 
-# Results
-
-## Test space and latent space
-
-Before defining the multidimensional parameters, we need to make a clarification
-regarding the representation of the items in the latent space.
-We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of
-the discrimination vectors.
-This space is where the item vectors can be properly represented.
-It is in general different from the latent space $`r LATENT_SPACE`$,
-although there is a bijective relationship between the two sets of spaces.
-Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
-where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
-in the same basis $`r LS_BASIS`$ in their respective spaces.
-
-Since the response probability must be invariant, because of Equation
-\@ref(eq:change-basis), parameter $`r DISCR_VECTOR`$ must also be transformed
-in $`r DISCR_VECTOR_STD`$.
-Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
-This shows that first a basis change to the standard base is performed and then
-a transformation through $`r INNER_PROD_MATRIX_INV`$,
-so the $`r DISCR_VECTOR_STD`$ are expressed in the standard basis.
-
-\begin{equation}
-  `r DISCR_VECTOR_MODULE_EQ`                     (\#eq:discr-param-squared-norm)
-\end{equation}
-
-Therefore, in the test space, an inner product matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined such that $`r INNER_PROD_DISCR_EQ`$,
-to keep the norm invariant.
-With this definition we have that
+In the test space we have that
 
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
+
+# Results
 

### 2023-03-10 R/Formulae.R
Changes the "test space" section according to its precedence and the derivation based on the invariance of the M2PL model.

@@ -88,8 +88,9 @@ IRF_M2PL <- latex(
 IRF_ABBR <- latex_sub('P', "$ITEM_INDEX$")
 
 # Model formula:
-LOGIT_M2PL    <- latex_parentheses(
-  DISCR_VECTOR_TRANSP, "$TRAIT_VECTOR$ + $INTERCEPT_PARAM$"
+DISCR_TRAIT_PROD <- latex(DISCR_VECTOR_TRANSP, TRAIT_VECTOR)
+LOGIT_M2PL       <- latex_parentheses(
+  "$DISCR_TRAIT_PROD$ + $INTERCEPT_PARAM$"
 )
 EXPFUNC_M2PL     <- latex_exp(latex_sqbrackets("-$LOGIT_M2PL$"))
 LOGISTIC_M2PL    <- latex_logistic(LOGIT_M2PL)
@@ -154,7 +155,7 @@ BASIS_CHANGE_SYM <- latex_basis_change(LS_BASIS, LS_STD_BASIS)
 BASIS_CHANGE_DEF <- latex_eq(TRANSFORM_MATRIX, BASIS_CHANGE_SYM)
 
 # Basis vector coordinates in standard basis:
-BASIS_VEC_ANY_STD_COORDS     <- latex_raised_to(
+BASIS_VEC_ANY_STD_COORDS    <- latex_raised_to(
   BASIS_VECTOR_ANY,
   exp = LS_STD_BASIS
 )
@@ -266,6 +267,84 @@ INNER_PROD_MATRIX_INV_DEF   <- latex(
 INNER_PROD_MATRIX_INV_EQ    <- latex_eq(
   INNER_PROD_MATRIX_INV,
   INNER_PROD_MATRIX_INV_DEF
+)
+
+### Test space ----
+
+# Discrimination vector coordinates:
+ITEM_SPACE_BASIS     <- latex_cal("B^*")
+DISCR_COORDS         <- latex_raised_to(DISCR_VECTOR, exp = ITEM_SPACE_BASIS)
+ITEM_SPACE_STD_BASIS <- latex_cal("E^*")
+DISCR_STD_COORDS    <- latex_raised_to(DISCR_VECTOR, exp = ITEM_SPACE_STD_BASIS)
+
+# Invariance assumption:
[…]

### 2023-04-20 output/paper_draft/paper_M2PLM_parameters.Rmd
prueba de cambio

--- 2023-03-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 a latent trait space, no matter its structure, as an orthogonal space
-[@ackerman_creating_1994]
----exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
+--exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
 the R package *psych* <!--# Reference needed? -->graphically represent
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration<!--# Examples? -->
+is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
 [see @zhang_theoretical_1999 for a notable exception].
@@
 The general case of MIRT model estimation implies estimating possibly
-non-null covariances among latent dimensions, as well as non-standard scales.
+non-standard latent dimensions and non-null covariances among them
+(a more realistic case than an identity matrix).
 Assuming orthonormality among traits is usually unrealistic, be it in the
@@
 interpreting the latent dimensions as substantive constructs is not a goal.
-As we will argue, the usual definitions of the MID and MDISC parameters
-do not generalize to the non-orthonormal geometry
+As we will argue, the usual definitions of the multidimensional difficulty and
+discrimination parameters do not generalize to the non-orthonormal geometry
 necessary to accurately represent correlated latent spaces.
@@
 In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait,
-usually reverse scored-- is also widespread.<!--# Reference and example? -->
+--items that tap the negative pole of a certain trait, and thus
+are usually reverse scored-- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine
@@
 One can easily note that a monotonically decreasing probability for a certain
-dimension is given by a negative discrimination parameter
+dimension is given by a non-positive discrimination parameter
 (and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
@@
 The main objective of this paper is thus to provide a generalization of
-the `r MODEL_ACRONYM` multidimensional parameters in two aspects:
-On one hand, in an oblique coordinate system
+the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
 that can be soundly applied to the correlated trait case; on the other, to
@@
 as a function of how we define and interpret the latent space.
-Next we discuss the properties of these indices,
+Next we discuss the properties of the indices,
 propose a graphical representation method,
-and present two practical examples applying them.
+and present two examples of application of our results.
 We round up with a discussion about
@@
 \end{equation}
+
+NOT NECESSARY
 
@@
 
-We must stop here and make a few remarks. First, as we are not constraining
-ourselves to cognitive traits,
-we speak of _positive_ instead of _correct response_. This may imply giving
-a correct response indeed, if we are talking about a cognitive test.
-But it may mean endorsing an item (or a certain response option)
-in a non-cognitive domain.
+We must make a few remarks here. First, as we are not constraining ourselves to
+cognitive traits, we speak of _positive_ instead of _correct response_.
+This may imply a correct response, if we are talking about a cognitive test, or
+endorsing an item (or a certain response option) in a non-cognitive domain.
 Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$ (e.g. a negatively-keyed item).
+latent dimension $`r DIM_INDEX`$.
 Finally, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
 instead of the item difficulty.
@@
 because the slope is dependent on the direction considered.
-However, Reckase's Equation 3 implicitly assumes orthogonality, as we will see.
+However, Reckase's Equation 3 implicitly assumes orthogonality.
 In order to contemplate the general case,
@@
 Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
-Gram matrix, i.e. $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
+Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
 the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
 $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
-invertible, so is $`r INNER_PROD_MATRIX`$, and $`r INNER_PROD_MATRIX_INV_EQ`$.
+invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
 

### 2023-04-20 output/paper_draft/paper_M2PLM_parameters.Rmd
changes 1

@@
 Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
-Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
+Gram matrix, i.e.$\langle \theta_j, \theta_k \rangle_B=\theta_j^{BT}M\theta_k^{B}$ with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
 the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
@@
 \end{equation}
+
+BE CAREFUL with $e_k$
 
@@
 in $`r DISCR_VECTOR_STD`$.
-Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
-This shows that first a basis change to the standard base is performed and then
-a transformation through $`r INNER_PROD_MATRIX_INV`$,
-so the $`r DISCR_VECTOR_STD`$ are expressed in the standard basis.
+Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i$ .
+
+JUSTIFY
 

### 2023-07-27 notebooks/example_script.Rmd
Versions initial script with Mario's derivations (sent on 2023-05-17)

---
title: "Ejemplificando art"
author: "Mario Luzardo"
date: "2023-05-16"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls()) 
library(pspline)
library(mvtnorm)
library(sn)
library(mnormt)
library(scatterplot3d)
library(copula)
library(ggplot2)
library(fungible)
library(pracma)
library(matlib)
library(rgl)
set.seed(123)
```
En esta simulacion queremos ejemplificar lo habllado hasta la fecha
Prumeramente se generar'a la base canónica y una base no ortogonal $\mathfrak{B}$
Para simplifcar la base no ortogonal $\mathfrak{B}$  la normalizamos aunque esto no es necesario

```{r base canonica, include=FALSE}
c_v1=c(1,0)
c_v2=c(0,1)
Base_C=matrix(cbind(c_v1,c_v2),ncol=2)
```

```{r base no ortogonal, include=FALSE}
b1_v1=c(1,4)
b1_v2=c(3,1)
b1_v1=b1_v1/len(b1_v1)
b1_v2=b1_v2/len(b1_v2)
Base_1=matrix(cbind(b1_v1,b1_v2),ncol=2)
```

La base no ortogonal  es:

$\mathfrak{B}$=\{$v_1$=(`r b1_v1[1] `, `r b1_v1[2] `) ;$v_2$=(`r b1_v2[1] `, `r b1_v2[2] `) \}

Estos vectores tienen norma 1 y un ángulo entre ellos de `r round(matlib::angle(Base_1[,1],Base_1[,2]),2)` grados

Gráficamente los vectores de las bases canónica y B quedan

```{r grafico bases, echo=FALSE, include=TRUE}
plot(NA, xlim=c(-2,2), ylim=c(-2,2), xlab="X", ylab="Y")
vecs <- data.frame(vname=c("a","b"),x0=c(0,0,0,0),y0=c(0,0,0,0), x1=c(1,0,Base_1[1,1],Base_1[1,2]) ,y1=c(0,1,Base_1[2,1],Base_1[2,2]),col=1:4) 
with( vecs, mapply("arrows", x0, y0, x1,y1,col=col,lwd=3) )
with(vecs, mapply('text', x=x1[1:2]-.1, y=y1[1:2]+.1,labels=expression(list(a), list(b) ) ))
```

```{r angulos, include=FALSE}
alfa_1_1=matlib::angle(Base_1[,1],Base_C[,1])
alfa_1_2=matlib::angle(Base_1[,1],Base_C[,2])
alfa_2_1=matlib::angle(Base_1[,2],Base_C[,1])
alfa_2_2=matlib::angle(Base_1[,2],Base_C[,2])
```

Los ángulos con los ejes coordenados son:

Entre $v_1$ y $e_1$ `r round(alfa_1_1,2) ` grados

Entre $v_1$ y $e_2$ `r round(alfa_1_2,2) ` grados

Entre $v_2$ y $e_1$ `r round(alfa_2_1,2) ` grados

Entre $v_2$ y $e_2$ `r round(alfa_2_2,2) ` grados

Vamos a silular 2000 coordenadas en la base $\mathfrak{B}$. Tomaremos habilidades normales bivariadas correlacionadas 0.5 y con marginales normales típicas.

```{r coordenadas en B, include=FALSE}
CopulaNormal<-normalCopula(0.5, dim = 2, dispstr = "un")
MvdcNorm1<-mvdc(copula=CopulaNormal,margins=c("norm","norm"),paramMargins=list(list(mean=0,sd=1),list(mean=0,sd=1)))
coord_B<-rMvdc(2000,MvdcNorm1)
```

```{r coordenadas en C, include=FALSE}
P=Base_1
P_inv=solve(P)
P_invT=t(P_inv)
M=t(P)%*%P
coord_C <- t(apply(coord_B, 1, function(x) P %*% as.matrix(x)))
```

Los puntos quedan gráficamente

```{r coordenadas, echo=FALSE, include=TRUE}
plot(coord_C,xlim=c(-4,4),ylim=c(-4,4))
arrows(0, 0, 4*Base_1[1,1],4*Base_1[2,1] , lwd=3, col=2,code=2)
arrows(0, 0, 4*Base_1[1,2], 4*Base_1[2,2], lwd=3, col=2,code=2)
arrows(0, 0, -3*Base_1[1,1],-3*Base_1[2,1] , lwd=3, col=2,code=0)
arrows(0, 0, -3*Base_1[1,2], -3*Base_1[2,2], lwd=3, col=2,code=0)
grid()
```

En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la matriz de cambio de base P

$$\theta^{\mathfrak{E}}=P\theta^{\mathfrak{B}}$$

```{r }
print(P)
```

Debemos observar que los puntos quedan fijos en el espacio lo que cambia son las coordenadas dado que P es la transformación identidad expresada en ambas bases.

Para mantener invariante la norma debemos definir en el espacio de la habilidad con la base B un producto interno con matriz de Gramm M
$$ M= P^T P$$
```{r }
print(M)
```

Por lo tanto la norma permanece invariante esto se puede comprobar mediante los comandos que siguen.

Consderamos un caso particular de habilidad

```{r norma, echo=TRUE, include=TRUE}
theta_C=coord_C[17,]
theta_C
theta_B=solve(Base_1,theta_C)
norma_B=as.numeric(sqrt(t(theta_B)%*%M%*%theta_B))
norma_C=as.numeric(sqrt(t(theta_C)%*%theta_C))
norma_B
norma_C
```

Es claro que al existir una transformación lineal sobre las coordenadas esto afectará a las varianzas (no a la distribución ya que partimos de una normal) y también las correlaciones.

Veremos como son estas relaciones en el siguiente código

```{r varianzas,echo=TRUE}
var(coord_B)
var(coord_C)
#Relacione entre las varianzas a partir de matrices
P%*%var(coord_B)%*%t(P)
cor(coord_B)
cor(coord_C)
#Relacione entre las correlaciones a partir de matrices

U=diag(diag(sqrt(var(coord_C))^(-1)))

U%*%P%*%var(coord_B)%*%t(P)%*%U
```

Veamos  como funciona la proposición 1 que dice que

$$ cos\gamma^B=\frac{D^{-\frac{1}{2}}M \theta^B}{\| \theta \|}$$

```{r prop1,echo=TRUE}
D=diag(c(len(b1_v1),len(b1_v2)))
D_inv=solve(D)
normaC=apply(coord_C,1,len)
cosdir_B=NULL
for(i in 1: length(normaC)){
  aux=(D_inv^(0.5)%*%M%*%coord_B[i,])/normaC[i]  
  cosdir_B=rbind(cosdir_B,aux)
}
cosdir_B[1:5]
angulos=r2d(acos(cosdir_B))
angulos[1:5]
```
Observemos que la norma que tomamos en la funci'on es la norma respecto a la base canonica. Si queremos considerar la base no ortogonal debe calcularse la norma con el producto interno con matriz M

#Modelo

```{r modelo,include=FALSE}
a=c(0.5,1.5)
d=0.2
P_theta=NULL
for(i in 1:dim(coord_B)[1]){
  x=coord_B[i,]
  aux=1/(1+exp(-(t(a)%*%x+d)))
  P_theta=rbind(P_theta,aux)
}

a_p=P_invT%*%a
P_theta_C=NULL
for(i in 1:dim(coord_C)[1]){
  x=coord_C[i,]
  aux=1/(1+exp(-(t(a_p)%*%x+d)))
  P_theta_C=rbind(P_theta_C,aux)
}

```

Consideremos el modelo expresado en la base no ortogonal B con discriminación `r a` y d=`r d`

Con el cambio

$$a_i^{'}=(P^{-1})^T a_i$$
se puede modelizar utilizando las coordenadas en la base can'onica

```{r comprueba}
P_theta[1:5]
P_theta_C[1:5]
```

#Cálculo de punto de máxima pendiente y pendiente

Los cosenos directores respecto a los ejes de la base canónica de la dirección del punto de máxima pendiente son

$$cos \gamma_i^E=\frac{a_i^{'}}{\sqrt{a_i^{'T}a_i^{'}}}$$
o con respecto a la base B

$$cos \gamma_i^B=\frac{D^{-\frac{1}{2}} a_i}{\sqrt{a_i^{T}M^{-1}a_i}}$$

La distencia al origen en esta dirección esta dada por

$$D_i=\frac{-d_i}{\sqrt{a_i^{T}M^{-1}a_i}}$$
y la pendiente
$$ S_i=\frac{\sqrt{a_i^{T}M^{-1}a_i}}{4}$$

```{r slope, include=FALSE}
cos_gamma_C=a_p/len(a_p)
M_inv=solve(M)
cos_gamma_B=(D_inv^(0.5)%*%a)/as.numeric(sqrt(t(a)%*%M_inv%*%a))
dista=-d/as.numeric(sqrt(t(a)%*%M_inv%*%a))
slope=as.numeric(sqrt(t(a)%*%M_inv%*%a))/4
```

Obtenemos

$cos \gamma_i^E$= `r cos_gamma_C`

$cos \gamma_i^B$= `r cos_gamma_B`

$D_i$=`r dista`

$S_i$= `r slope`

El ángulo con los ejes de la base canónica es `r r2d(acos(cos_gamma_C))`

Graficamos un vector cuya norma es $S_i$ aplicado sobre el punto donde se da el m'aximo slope

```{r graslope}
m=cos_gamma_C[2]/cos_gamma_C[1]
punto_x=(dista/abs(dista))*sqrt(dista^2/(1+m^2))
punto_y=m*punto_x

plot(coord_C,xlim=c(-4,4),ylim=c(-4,4))
arrows(0, 0, 4*Base_1[1,1],4*Base_1[2,1] , lwd=3, col=2,code=2)
arrows(0, 0, 4*Base_1[1,2], 4*Base_1[2,2], lwd=3, col=2,code=2)
arrows(0, 0, -3*Base_1[1,1],-3*Base_1[2,1] , lwd=3, col=2,code=0)
arrows(0, 0, -3*Base_1[1,2], -3*Base_1[2,2], lwd=3, col=2,code=0)
points(punto_x, punto_y,lwd=4, col=5)
grid()

```

En un grafico de perspectiva tenemos
```{r perspe}
x=seq(-3,3,length=100)
y=seq(-3,3,length=100)

ICC=function(x,y){1/(1+exp(-(a_p[1]*x+a_p[2]*y+d)))}

z=outer(x, y, ICC)
persp(x,y,z,phi=40,border=NA,col="blue")

image(x,y,z)
contour(x,y,z,add=T)
points(punto_x,punto_y,lwd=4, col=5)

```

### 2023-07-27 notebooks/example_script.Rmd
Updates Mario's script to version sent on 2023-07-14

@@
 author: "Mario Luzardo"
-date: "2023-05-16"
+date: "2023-07-13"
 output: word_document
@@
 ```
-En esta simulacion queremos ejemplificar lo habllado hasta la fecha
-Prumeramente se generar'a la base canónica y una base no ortogonal $\mathfrak{B}$
+En esta simulacion queremos ejemplificar lo hallado hasta la fecha
+
+Primeramente se generará a la base canónica (puede no ser la canónica, cualquier base ortonormal sirve) y una base no ortogonal $\mathfrak{B}$
 Para simplifcar la base no ortogonal $\mathfrak{B}$  la normalizamos aunque esto no es necesario
@@
 ```{r }
-print(P)
+print("La matriz de cambio de base P es")
+P
 ```
@@
 ```{r }
-print(M)
+print("M resulta")
+M
 ```
 
-Por lo tanto la norma permanece invariante esto se puede comprobar mediante los comandos que siguen.
-
-Consderamos un caso particular de habilidad
-
-```{r norma, echo=TRUE, include=TRUE}
+```{r norma, echo=TRUE, include=FALSE}
 theta_C=coord_C[17,]
-theta_C
+#theta_C
 theta_B=solve(Base_1,theta_C)
@@
 norma_C=as.numeric(sqrt(t(theta_C)%*%theta_C))
-norma_B
-norma_C
+#norma_B
+#norma_C
 ```
+
+Por lo tanto la norma permanece invariante esto se puede comprobar mediante un ejemplo particular.
+
+Consderamos un caso particular de habilidad dada sus coordenadas en la base canónica por `r round(theta_C,4) ` y su norma usando el producto punto es `r round(norma_C,4) `
+
+Las coordenadas de este punto expresados en la base B son `r round(theta_B,4) ` y su norma utilizando el producto internos con matriz de Gramm M es  `r round(norma_B,4) `
 
@@
 
+La relación viene dada por 
+$$Var(coord_C(\theta))=P Var(coord_B(\theta)) P^T$$
+
+$$Cor(coord_C(\theta))=U P Var(coord_B(\theta)) P^T U=UVar(coord_C(\theta))U$$
+
+siendo
+
+$$U=diag((Var(coord_C(\theta)))^{-1/2})$$
 Veremos como son estas relaciones en el siguiente código
@@
 
+## Proposición 1
+
 Veamos  como funciona la proposición 1 que dice que
@@
 for(i in 1: length(normaC)){
-  aux=(D_inv^(0.5)%*%M%*%coord_B[i,])/normaC[i]  
+  aux=t((D_inv^(0.5)%*%M%*%coord_B[i,])/normaC[i])  
   cosdir_B=rbind(cosdir_B,aux)
 }
-cosdir_B[1:5]
+cosdir_B[1:5,]
 angulos=r2d(acos(cosdir_B))
-angulos[1:5]
+angulos[1:5,]
 ```
-Observemos que la norma que tomamos en la funci'on es la norma respecto a la base canonica. Si queremos considerar la base no ortogonal debe calcularse la norma con el producto interno con matriz M
+Observemos que la norma que tomamos en la función es la norma respecto a la base canonica. Si queremos considerar la base no ortogonal debe calcularse la norma con el producto interno con matriz M
 
-#Modelo
+## Modelo
 
@@
 $$a_i^{'}=(P^{-1})^T a_i$$
-se puede modelizar utilizando las coordenadas en la base can'onica
+se puede modelizar utilizando las coordenadas en la base canónica, es decir la ecuación anterior da la transformación de los a al transformar los $\theta$.
+Esto imlica que si utilizamos las coordeadas de las habilidades en la base can'onica debemos utilizar la $a'$.
+
+Veamos que coinciden ambas modelizaciones
 
@@
 
-#Cálculo de punto de máxima pendiente y pendiente
+## Cálculo de punto de máxima pendiente y pendiente
 
@@
 
-La distencia al origen en esta dirección esta dada por
+La distancia al origen en esta dirección esta dada por
 
@@
 ```
+
+## Función de Representación
+Lo anteriore lo podemos representar para cualesquiera bases mediante la función representa
+
+```{r representa}
+# BaseO es la base oblicua u original
+#BaseC base ortogonal cualesquiera
+#K matriz de cambio de base de O a C
+# a y d parámetros del modelo original
+#coord_O coordenadas del rasgo enla base original
+
+representa=function(BaseO,BaseC,K,a,d,coord_O){
+  K_inv=solve(K)
+  M=t(K)%*%K
+  M_inv=solve(M)
+  D=diag(c(len(K[,1]),len(K[,2])))
+  D_inv=solve(D)
+  coord_C=t(apply(coord_O, 1, function(x) K %*% as.matrix(x)))
+  a_p=t(K_inv)%*%a
+  #cosenos directores entre la base C y a_p
+  cos_gamma_C=a_p/len(a_p)
+  angulo_C=r2d(acos(cos_gamma_C))
+  #cosenos directores entre la base O y a_p
+  cos_gamma_B=(D_inv^(0.5)%*%a)/as.numeric(sqrt(t(a)%*%M_inv%*%a))
+  angulo_B=r2d(acos(cos_gamma_B))
+  dista=-d/as.numeric(sqrt(t(a)%*%M_inv%*%a))
+  slope=as.numeric(sqrt(t(a)%*%M_inv%*%a))/4
+  m=cos_gamma_C[2]/cos_gamma_C[1]
+    if(d==0){
+    punto_x=0
+  }else{
+  punto_x=(dista/abs(dista))*sqrt(dista^2/(1+m^2))}
+  punto_y=m*punto_x
+  #Coordenadas de la BaseO orignal en la BaseC
+  v1= K[,1] # viene de K*c(1,0) pues en la BaseO estas son las coordenadas
+  v2= K[,2] #K * (0,1)
+  plot(coord_C,xlim=c(-4,4),ylim=c(-4,4))
+  arrows(0, 0, v1[1],v1[2] , lwd=3, col=2,code=2)
+  arrows(0, 0, v2[1], v2[2], lwd=3, col=2,code=2)
+  arrows(0, 0, -v1[1],-v1[2] , lwd=3, col=2,code=0)
+  arrows(0, 0, -v2[1], -v2[2], lwd=3, col=2,code=0)
+  arrows(-2, m*(-2), 2, m*2, lwd=3, col=4,code=0)
+  points(punto_x, punto_y,lwd=4, col=5)
+  grid()
+  return(list(direccion=m,D=dista,coord_X=punto_x,coord_Y=punto_y,Slope=slope))  
+}
+
+```
+
+Podemos presentar un gráfico de nibel mediante la función repnivel
+
+```{r reprenivel}
+# x, y grilla
+#K matriz de cambio de base de O a C
+# a y d parámetros del modelo original
+#punto_X ,punto_y punto de máximo slope
+
+repnivel<-function(x=seq(-3,3,length=100),y=seq(-3,3,length=100),K,a,d,punto_x,punto_y){
+  a_p=t(solve(K))%*%a
+  ICC=function(x,y){1/(1+exp(-(a_p[1]*x+a_p[2]*y+d)))}  
+  z=outer(x, y, ICC)  
+  image(x,y,z)
+  contour(x,y,z,add=T)
+  points(punto_x,punto_y,lwd=4, col=5)
+}
+```
+
+Por ejemplo si usamos nuevamente la base $\mathfrak{B}$ y la canónica pero ahora el ítem tiene a=(0.5,2) y d=1.1
+
+```{r }
+Ej1=representa(Base_1,Base_C,P,c(0.5,2),1.1,coord_B)
+repnivel(seq(-3,3,length=100),seq(-3,3,length=100),P,c(0.5,2),1.1,Ej1$coord_X,Ej1$coord_Y)
+
+```
+
+## Test Space
+
+Observemos que el parámetro a ¨vive¨en otro espacio vectorial que $\theta$.
+
+Llamemos a ese espacio vectorial $\mathcal A$.
+
+El modelo de TRI expresa una relación en el espacio de coordenadas de la habilidad y la discriminación y en ese espacio de coordenadas $R^n$ se considera el producto punto.
+En principio el rasgo $\theta$ tiene sus coordenadas expresadas en la base $\mathfrak{B}$ que representa a las reales dimensiones que tiene el rasgo.
+También las discriminaciones estan expresadas en alguna base que al no tener restricciones previas podemos suponer cualquier base que fijará las transformaciones del espacio.
+En particular supondremos que la base de origen en el espacio $\mathcal A$ tiene como vectores las columnas de $(P^{-1})^T$. La llamaremos $\mathfrak{A}$
+
+Al realizar un cambio de base en el espacio $\Theta$ donde se transforman las coordenadas  de la base $\mathfrak{B}$ a la base $\mathfrak{E}$ (en nuestro caso la canónica) mediante la matriz P a partir de la ecuación de ligadura dada por el modelo observamos que las coordenadas de las dicriminaciones ($a_i$) se trasforman en $a^'_i$ mediante la ecuación
+
+$$a_i^{'}=(P^{-1})^T a_i$$
+
+Al ser $(P^{-1})^T$ invertible representa una matriz de cambio de base que lleva las coordenasas de $a_i$ expresadas en la base $\mathfrak{A}$ a la base canónica
+
+Además al tratarse de una transformación identidad la norma de $a_i$ no cambia al expresarse en las diferentes bases.
+Para mantener esta condición se induce un producto interno en el espacio original con matriz de Gramm $M^{-1}$
+
+$$\parallel a_i^{'} \parallel=a_i^T P^{-1} (P^{-1})^T a_i =a_i^T M^{-1}a_i$$
+
+En nuestro ejemplo la base $\mathfrak{A}$ se puede dibujar
+
+```{r coord3, echo=FALSE, include=TRUE}
+plot(NA,xlim=c(-4,4),ylim=c(-4,4))
+arrows(0, 0, 2*P_invT[1,1],2*P_invT[2,1] , lwd=3, col=2,code=2)
+arrows(0, 0, 2*P_invT[1,2], 2*P_invT[2,2], lwd=3, col=2,code=2)
+grid()
+```
+
+## Consideramdo la estructura probabilistica de $\Theta$
+
+Si ahora consideramos la estructura probabilistica de los rasgos por ejemplo
+
+$$\theta \sim N(0, \Sigma)$$
+
+El hecho de que la media sea 0 no hace perder generalidad pies se puede centrar el rasgo.
+
+Suponemos además que la base natural de los rasgos es  $\mathfrak{B}$ que en nuestro ejemplo es
+
+```{r b1, include=FALSE}
+print(Base_1)
+```
+
+Como ya vimos graficamente tenemos
+
+```{r coord, echo=FALSE, include=TRUE}
+plot(coord_C,xlim=c(-4,4),ylim=c(-4,4))
+arrows(0, 0, 4*Base_1[1,1],4*Base_1[2,1] , lwd=3, col=2,code=2)
+arrows(0, 0, 4*Base_1[1,2], 4*Base_1[2,2], lwd=3, col=2,code=2)
+arrows(0, 0, -3*Base_1[1,1],-3*Base_1[2,1] , lwd=3, col=2,code=0)
+arrows(0, 0, -3*Base_1[1,2], -3*Base_1[2,2], lwd=3, col=2,code=0)
+grid()
+```
+
+La pregunta es cual es la base ortonormal conveniente para tomarnos
+
+```{r DEB, echo=FALSE, include=TRUE}
+aux=eigen(var(coord_B))
+Q=aux$vectors
+Qinv=solve(Q)
+lam=diag(aux$values)
+auxi=Q%*%lam%*%t(Q)
+```
+
+```{r coord2, echo=FALSE, include=TRUE}
+plot(coord_C,xlim=c(-4,4),ylim=c(-4,4))
+arrows(0, 0, 4*Base_1[1,1],4*Base_1[2,1] , lwd=3, col=2,code=2)
+arrows(0, 0, 4*Base_1[1,2], 4*Base_1[2,2], lwd=3, col=2,code=2)
+arrows(0, 0, -3*Base_1[1,1],-3*Base_1[2,1] , lwd=3, col=2,code=0)
+arrows(0, 0, -3*Base_1[1,2], -3*Base_1[2,2], lwd=3, col=2,code=0)
+arrows(0, 0, 4*Q[1,1],4*Q[2,1] , lwd=3, col=3,code=2)
+arrows(0, 0, 4*Q[1,2],4*Q[2,2] , lwd=3, col=3,code=2)
+grid()
+
+```
+
+El ángulo entre los vectores propios es 1
+
+```{r , echo=FALSE, include=TRUE}
+print(matlib::angle(Q[,1],Q[,2]))
+```
+
+Debemos entonces desarrollar como queda la matriz de cambio de base entre la base $\mathfrak{B}$ y la base de vectores propios $\mathfrak{L}$ de la matriz $\Sigma$
+
+Tenemos al descomponer espectralmente a $\Sigma$ que
+
+$$\Sigma=Q \Lambda Q^T$$
+
+con $Q^T=Q^{-1}$
+
+La base $\mathfrak{L}$  tiene como vectores a las columnas de Q
+
+```{r , echo=FALSE, include=TRUE}
+print(Q)
+```
+
+La matriz de cambio de base entre $\mathfrak{L}$ y $\mathfrak{E}$ es Q por lo tanto la matriz de cambio de base entre $\mathfrak{B}$ y $\mathfrak{L}$ es $$Q^{-1}P$$
+
+Veamos un ejemplo emplendo como base ortogonal los vectores propios de $\Sigma$. El par'ametro a=(0.5,2) y d=1.1
+
+```{r , echo=FALSE, include=TRUE}
+a=c(0.5,2) 
+d=1.1
+Kform1=solve(Q)%*%P
+Ej2=representa(Base_1,Q,Kform1,a,d,coord_B)
+repnivel(seq(-3,3,length=100),seq(-3,3,length=100),Kform1,a,d,Ej2$coord_X,Ej2$coord_Y)
+print(Ej2)
+```

### 2023-07-27 notebooks/example_script.Rmd
Turns markdown text to "canonical"

@@
 output: word_document
+editor_options:
+  markdown:
+    canonical: true
 ---
@@
 ```
+
 En esta simulacion queremos ejemplificar lo hallado hasta la fecha
 
-Primeramente se generará a la base canónica (puede no ser la canónica, cualquier base ortonormal sirve) y una base no ortogonal $\mathfrak{B}$
-Para simplifcar la base no ortogonal $\mathfrak{B}$  la normalizamos aunque esto no es necesario
+Primeramente se generará a la base canónica (puede no ser la canónica, cualquier
+base ortonormal sirve) y una base no ortogonal $\mathfrak{B}$
+
+Para simplifcar la base no ortogonal $\mathfrak{B}$ la normalizamos aunque esto
+no es necesario
 
@@
 
-$\mathfrak{B}$=\{$v_1$=(`r b1_v1[1] `, `r b1_v1[2] `) ;$v_2$=(`r b1_v2[1] `, `r b1_v2[2] `) \}
+$\mathfrak{B}$={$v_1$=(`r b1_v1[1]`, `r b1_v1[2]`) ;$v_2$=(`r b1_v2[1]`,
+`r b1_v2[2]`) }
 
-Estos vectores tienen norma 1 y un ángulo entre ellos de `r round(matlib::angle(Base_1[,1],Base_1[,2]),2)` grados
+Estos vectores tienen norma 1 y un ángulo entre ellos de
+`r round(matlib::angle(Base_1[,1],Base_1[,2]),2)` grados
 
@@
 
-Entre $v_1$ y $e_1$ `r round(alfa_1_1,2) ` grados
+Entre $v_1$ y $e_1$ `r round(alfa_1_1,2)` grados
 
-Entre $v_1$ y $e_2$ `r round(alfa_1_2,2) ` grados
+Entre $v_1$ y $e_2$ `r round(alfa_1_2,2)` grados
 
-Entre $v_2$ y $e_1$ `r round(alfa_2_1,2) ` grados
+Entre $v_2$ y $e_1$ `r round(alfa_2_1,2)` grados
 
-Entre $v_2$ y $e_2$ `r round(alfa_2_2,2) ` grados
+Entre $v_2$ y $e_2$ `r round(alfa_2_2,2)` grados
 
-Vamos a silular 2000 coordenadas en la base $\mathfrak{B}$. Tomaremos habilidades normales bivariadas correlacionadas 0.5 y con marginales normales típicas.
+Vamos a silular 2000 coordenadas en la base $\mathfrak{B}$. Tomaremos
+habilidades normales bivariadas correlacionadas 0.5 y con marginales normales
+típicas.
 
@@
 
-En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la matriz de cambio de base P
+En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la
+matriz de cambio de base P
 
@@
 
-Debemos observar que los puntos quedan fijos en el espacio lo que cambia son las coordenadas dado que P es la transformación identidad expresada en ambas bases.
+Debemos observar que los puntos quedan fijos en el espacio lo que cambia son las
+coordenadas dado que P es la transformación identidad expresada en ambas bases.
 
-Para mantener invariante la norma debemos definir en el espacio de la habilidad con la base B un producto interno con matriz de Gramm M
-$$ M= P^T P$$
+Para mantener invariante la norma debemos definir en el espacio de la habilidad
+con la base B un producto interno con matriz de Gramm M $$ M= P^T P$$
+
 ```{r }
@@
 
-Por lo tanto la norma permanece invariante esto se puede comprobar mediante un ejemplo particular.
+Por lo tanto la norma permanece invariante esto se puede comprobar mediante un
+ejemplo particular.
 
-Consderamos un caso particular de habilidad dada sus coordenadas en la base canónica por `r round(theta_C,4) ` y su norma usando el producto punto es `r round(norma_C,4) `
+Consderamos un caso particular de habilidad dada sus coordenadas en la base
+canónica por `r round(theta_C,4)` y su norma usando el producto punto es
+`r round(norma_C,4)`
 
-Las coordenadas de este punto expresados en la base B son `r round(theta_B,4) ` y su norma utilizando el producto internos con matriz de Gramm M es  `r round(norma_B,4) `
+Las coordenadas de este punto expresados en la base B son `r round(theta_B,4)` y
+su norma utilizando el producto internos con matriz de Gramm M es
+`r round(norma_B,4)`
 
-Es claro que al existir una transformación lineal sobre las coordenadas esto afectará a las varianzas (no a la distribución ya que partimos de una normal) y también las correlaciones.
+Es claro que al existir una transformación lineal sobre las coordenadas esto
+afectará a las varianzas (no a la distribución ya que partimos de una normal) y
+también las correlaciones.
 
-La relación viene dada por 
-$$Var(coord_C(\theta))=P Var(coord_B(\theta)) P^T$$
+La relación viene dada por $$Var(coord_C(\theta))=P Var(coord_B(\theta)) P^T$$
 
@@
 
-$$U=diag((Var(coord_C(\theta)))^{-1/2})$$
-Veremos como son estas relaciones en el siguiente código
+$$U=diag((Var(coord_C(\theta)))^{-1/2})$$ Veremos como son estas relaciones en
+el siguiente código
 
@@
 var(coord_C)
-#Relacione entre las varianzas a partir de matrices
+
+#Relaciones entre las varianzas a partir de matrices
 P%*%var(coord_B)%*%t(P)
@@
 cor(coord_C)
-#Relacione entre las correlaciones a partir de matrices
 
+#Relaciones entre las correlaciones a partir de matrices
 U=diag(diag(sqrt(var(coord_C))^(-1)))
@@
 ```
-Observemos que la norma que tomamos en la función es la norma respecto a la base canonica. Si queremos considerar la base no ortogonal debe calcularse la norma con el producto interno con matriz M
+
+Observemos que la norma que tomamos en la función es la norma respecto a la base
+canonica. Si queremos considerar la base no ortogonal debe calcularse la norma
+con el producto interno con matriz M
 
@@
 
-Consideremos el modelo expresado en la base no ortogonal B con discriminación `r a` y d=`r d`
+Consideremos el modelo expresado en la base no ortogonal B con discriminación
+`r a` y d=`r d`
 
@@
 
-$$a_i^{'}=(P^{-1})^T a_i$$
-se puede modelizar utilizando las coordenadas en la base canónica, es decir la ecuación anterior da la transformación de los a al transformar los $\theta$.
-Esto imlica que si utilizamos las coordeadas de las habilidades en la base can'onica debemos utilizar la $a'$.
+$$a_i^{'}=(P^{-1})^T a_i$$ se puede modelizar utilizando las coordenadas en la
+base canónica, es decir la ecuación anterior da la transformación de los a al
+transformar los $\theta$. Esto imlica que si utilizamos las coordeadas de las
+habilidades en la base canónica debemos utilizar la $a'$.
 
@@
 
-Los cosenos directores respecto a los ejes de la base canónica de la dirección del punto de máxima pendiente son
+Los cosenos directores respecto a los ejes de la base canónica de la dirección
+del punto de máxima pendiente son
 
-$$cos \gamma_i^E=\frac{a_i^{'}}{\sqrt{a_i^{'T}a_i^{'}}}$$
-o con respecto a la base B
+$$cos \gamma_i^E=\frac{a_i^{'}}{\sqrt{a_i^{'T}a_i^{'}}}$$ o con respecto a la
+base B
 
@@
 
-$$D_i=\frac{-d_i}{\sqrt{a_i^{T}M^{-1}a_i}}$$
-y la pendiente
+$$D_i=\frac{-d_i}{\sqrt{a_i^{T}M^{-1}a_i}}$$ y la pendiente
 $$ S_i=\frac{\sqrt{a_i^{T}M^{-1}a_i}}{4}$$
@@
 
-Graficamos un vector cuya norma es $S_i$ aplicado sobre el punto donde se da el m'aximo slope
+Graficamos un vector cuya norma es $S_i$ aplicado sobre el punto donde se da el
+máximo slope
 
@@
 En un grafico de perspectiva tenemos
+
 ```{r perspe}
@@
 ## Función de Representación
-Lo anteriore lo podemos representar para cualesquiera bases mediante la función representa
+
+Lo anteriore lo podemos representar para cualesquiera bases mediante la función
+representa
 
@@
 
-Podemos presentar un gráfico de nibel mediante la función repnivel
+Podemos presentar un gráfico de nivel mediante la función repnivel
 
@@
 
-Por ejemplo si usamos nuevamente la base $\mathfrak{B}$ y la canónica pero ahora el ítem tiene a=(0.5,2) y d=1.1
+Por ejemplo si usamos nuevamente la base $\mathfrak{B}$ y la canónica pero ahora
+el ítem tiene a=(0.5,2) y d=1.1
 
@@
 
-Observemos que el parámetro a ¨vive¨en otro espacio vectorial que $\theta$.
+Observemos que el parámetro a "vive" en otro espacio vectorial que $\theta$.
 
@@
 
-El modelo de TRI expresa una relación en el espacio de coordenadas de la habilidad y la discriminación y en ese espacio de coordenadas $R^n$ se considera el producto punto.
-En principio el rasgo $\theta$ tiene sus coordenadas expresadas en la base $\mathfrak{B}$ que representa a las reales dimensiones que tiene el rasgo.
-También las discriminaciones estan expresadas en alguna base que al no tener restricciones previas podemos suponer cualquier base que fijará las transformaciones del espacio.
-En particular supondremos que la base de origen en el espacio $\mathcal A$ tiene como vectores las columnas de $(P^{-1})^T$. La llamaremos $\mathfrak{A}$
+El modelo de TRI expresa una relación en el espacio de coordenadas de la
+habilidad y la discriminación y en ese espacio de coordenadas $R^n$ se considera
+el producto punto. En principio el rasgo $\theta$ tiene sus coordenadas
+expresadas en la base $\mathfrak{B}$ que representa a las reales dimensiones que
+tiene el rasgo. También las discriminaciones estan expresadas en alguna base que
+al no tener restricciones previas podemos suponer cualquier base que fijará las
+transformaciones del espacio. En particular supondremos que la base de origen en
+el espacio $\mathcal A$ tiene como vectores las columnas de $(P^{-1})^T$. La
+llamaremos $\mathfrak{A}$
 
-Al realizar un cambio de base en el espacio $\Theta$ donde se transforman las coordenadas  de la base $\mathfrak{B}$ a la base $\mathfrak{E}$ (en nuestro caso la canónica) mediante la matriz P a partir de la ecuación de ligadura dada por el modelo observamos que las coordenadas de las dicriminaciones ($a_i$) se trasforman en $a^'_i$ mediante la ecuación
+Al realizar un cambio de base en el espacio $\Theta$ donde se transforman las
+coordenadas de la base $\mathfrak{B}$ a la base $\mathfrak{E}$ (en nuestro caso
+la canónica) mediante la matriz P a partir de la ecuación de ligadura dada por
+el modelo observamos que las coordenadas de las dicriminaciones ($a_i$) se
+trasforman en $a^'_i$ mediante la ecuación
 
@@
 
-Al ser $(P^{-1})^T$ invertible representa una matriz de cambio de base que lleva las coordenasas de $a_i$ expresadas en la base $\mathfrak{A}$ a la base canónica
+Al ser $(P^{-1})^T$ invertible representa una matriz de cambio de base que lleva
+las coordenasas de $a_i$ expresadas en la base $\mathfrak{A}$ a la base canónica
 
-Además al tratarse de una transformación identidad la norma de $a_i$ no cambia al expresarse en las diferentes bases.
-Para mantener esta condición se induce un producto interno en el espacio original con matriz de Gramm $M^{-1}$
+Además al tratarse de una transformación identidad la norma de $a_i$ no cambia
+al expresarse en las diferentes bases. Para mantener esta condición se induce un
+producto interno en el espacio original con matriz de Gramm $M^{-1}$
 
@@
 
-El hecho de que la media sea 0 no hace perder generalidad pies se puede centrar el rasgo.
+El hecho de que la media sea 0 no hace perder generalidad pies se puede centrar
+el rasgo.
 
-Suponemos además que la base natural de los rasgos es  $\mathfrak{B}$ que en nuestro ejemplo es
+Suponemos además que la base natural de los rasgos es $\mathfrak{B}$ que en
+nuestro ejemplo es
 
@@
 
-Debemos entonces desarrollar como queda la matriz de cambio de base entre la base $\mathfrak{B}$ y la base de vectores propios $\mathfrak{L}$ de la matriz $\Sigma$
+Debemos entonces desarrollar como queda la matriz de cambio de base entre la
+base $\mathfrak{B}$ y la base de vectores propios $\mathfrak{L}$ de la matriz
+$\Sigma$
 
@@
 
-La matriz de cambio de base entre $\mathfrak{L}$ y $\mathfrak{E}$ es Q por lo tanto la matriz de cambio de base entre $\mathfrak{B}$ y $\mathfrak{L}$ es $$Q^{-1}P$$
+La matriz de cambio de base entre $\mathfrak{L}$ y $\mathfrak{E}$ es Q por lo
+tanto la matriz de cambio de base entre $\mathfrak{B}$ y $\mathfrak{L}$ es
+$$Q^{-1}P$$
 
-Veamos un ejemplo emplendo como base ortogonal los vectores propios de $\Sigma$. El par'ametro a=(0.5,2) y d=1.1
+Veamos un ejemplo emplendo como base ortogonal los vectores propios de $\Sigma$.
+El parámetro a=(0.5,2) y d=1.1
 

### 2023-07-27 notebooks/example_script.Rmd
Makes the figure outputs "square" so the directions of the vectors can be better understood

@@
 date: "2023-07-13"
-output: word_document
+output:
+  word_document:
+    fig_width:  6
+    fig_height: 6
 editor_options:

### 2023-07-27 notebooks/example_script.Rmd
Adds some clarifying comments (for myself)

@@
 b1_v2=c(3,1)
-b1_v1=b1_v1/len(b1_v1)
-b1_v2=b1_v2/len(b1_v2)
+b1_v1=b1_v1/len(b1_v1) # Normalized, but not necessary
+b1_v2=b1_v2/len(b1_v2) # Normalized, but not necessary
 Base_1=matrix(cbind(b1_v1,b1_v2),ncol=2)

### 2023-08-08 notebooks/example_script.Rmd
Versions update from Mario sent by mail on 2023-07-13) superseding version sent on 2023-07-14, which apparently lacked the latest advances, that were sent only in Word format)

@@
 
-La base $\mathfrak{L}$ tiene como vectores a las columnas de Q
+La descomposicion espectral se expresa en la base canónica tanto para los datos
+con coordenadas en la base $\mathfrak{B}$ como en la base $\mathfrak{E}$.
 
-```{r , echo=FALSE, include=TRUE}
-print(Q)
-```
+Si descomponemos $Sigma_E$ (con los datos en la base canónica) entonces
 
-La matriz de cambio de base entre $\mathfrak{L}$ y $\mathfrak{E}$ es Q por lo
-tanto la matriz de cambio de base entre $\mathfrak{B}$ y $\mathfrak{L}$ es
-$$Q^{-1}P$$
+$$
+\Sigma_E = Q_E \Lambda_E Q_E^T
+$$
 
-Veamos un ejemplo emplendo como base ortogonal los vectores propios de $\Sigma$.
-El parámetro a=(0.5,2) y d=1.1
+y lo propio con $\Sigma$ (con las coordenadas oblicuas)
 
-```{r , echo=FALSE, include=TRUE}
-a=c(0.5,2) 
-d=1.1
-Kform1=solve(Q)%*%P
-Ej2=representa(Base_1,Q,Kform1,a,d,coord_B)
-repnivel(seq(-3,3,length=100),seq(-3,3,length=100),Kform1,a,d,Ej2$coord_X,Ej2$coord_Y)
-print(Ej2)
-```
+$$
+\Sigma = Q \Lambda Q^T
+$$
+
+Además
+
+$$
+\Sigma_E=P \Sigma P^T
+$$
+
+de donde
+
+$$
+Q_E \Lambda_E Q_E^T = P Q \Lambda Q^T P^T
+$$
+
+Si tomamos la matriz $L$ definida como la matriz diagonal que contiene el
+cociente de los valores propios de $\Lambda$ entre $\Lambda_E$
+
+Entonces
+
+$$
+Q_E \Lambda_E Q_E^T = P Q L^{1/2} \Lambda_E (P Q L^{1/2} )^T
+$$
+
+o sea se cumple que
+
+$$
+Q_E = P Q L^{1/2}
+$$
+
+Por otro lado la matriz de cambio de base entre $\mathfrak{B}$ y $\mathfrak{L}$
+es $PQ_E^{-1}$ ya que $Q_E$ es la matriz de cambio de base entre $\mathfrak{B}$
+y la base canónica
+
+Luego la matriz de cambio de base queda
+
+$$
+PQ_E^{-1} = P( PQL^{1/2} )^{-1} = L^{-1/2} PQ^{-1} P^{-1}
+$$
+
+Se puede aplicar con esta matriz de cambio de base los resultados anteriores
+hallados.

### 2023-09-07 output/paper_draft/paper_M2PLM_parameters.Rmd

--- 2023-04-20 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 \end{equation}
-
-NOT NECESSARY
 

### 2023-09-15 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 
-Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+::: {.proof}
+*Proof.*
+
+For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,
+
+\begin{equation}
+  `r DIR_COSINE_PROOF_EQ`,                        (\#eq:dir-cosine-proof-prop-1)
+\end{equation}
+
+with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
+$`r INNER_PROD_MATRIX`$.
+
+Therefore, $`r DIR_COS_VEC_EQ`$ and $`r TRAIT_VECTOR_POLAR_EQ`$.
+$`r END_OF_PROOF`$
+:::
+
+:::
+
+## Model recasting into polar coordinates
+
+From Equation \@ref(eq:direction-cosine-vector),
+any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
+
+Thus, the M2PL model is expressed in polar coordinates as
 
@@
 
+The latter effect can be better noticed in another case, in which Reckase
+uses the same item parameters to demonstrate the over-specification of
+dimensions (see p. 183).
+In this example, he uses a covariance matrix that practically has rank 1,
+implying that the (alleged) three dimensions collapse into a single one, being
+thus the test unidimensional. In such a case, the covariance-based version of
+the direction angles are extremely close to 0 for all items, as one would expect
+from a unidimensional test, where all items are strictly aligned with the one
+dimension they measure. The reader interested can check out the parameters
+computed in such case in the Supplementary Table 1.
+<!-- #TODO: Include Supplementary Table -->
+
+### @tezza_modelo_2018
+
 ```{r set-cov-matrix-Tezza2018}
@@
 # > foi de aproximadamente 0,4.
-COV_MATRIX <- diag(4)
-COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
+COV_MATRIX_VALUE <- diag(4)
+DIMS_CORR        <- .4
+COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
 ```

### 2023-10-31 notebooks/example_script.Rmd
cambio 30 de octubre 2023
otro del 30

@@
 
-Primeramente se generará a la base canónica (puede no ser la canónica, cualquier
-base ortonormal sirve) y una base no ortogonal $\mathfrak{B}$
+Primeramente se generará a la base canónica (puede no ser la canónica, cualquier base ortonormal sirve) y una base no ortogonal $\mathfrak{B}$
 
-Para simplifcar la base no ortogonal $\mathfrak{B}$ la normalizamos aunque esto
-no es necesario
+Para simplifcar la base no ortogonal $\mathfrak{B}$ la normalizamos aunque esto no es necesario
 
@@
 
-$\mathfrak{B}$={$v_1$=(`r b1_v1[1]`, `r b1_v1[2]`) ;$v_2$=(`r b1_v2[1]`,
-`r b1_v2[2]`) }
+$\mathfrak{B}$={$v_1$=(`r b1_v1[1]`, `r b1_v1[2]`) ;$v_2$=(`r b1_v2[1]`, `r b1_v2[2]`) }
 
-Estos vectores tienen norma 1 y un ángulo entre ellos de
-`r round(matlib::angle(Base_1[,1],Base_1[,2]),2)` grados
+Estos vectores tienen norma 1 y un ángulo entre ellos de `r round(matlib::angle(Base_1[,1],Base_1[,2]),2)` grados
 
@@
 
-Vamos a silular 2000 coordenadas en la base $\mathfrak{B}$. Tomaremos
-habilidades normales bivariadas correlacionadas 0.5 y con marginales normales
-típicas.
+Vamos a simular 2000 coordenadas en la base $\mathfrak{B}$. Tomaremos habilidades normales bivariadas correlacionadas 0.5 y con marginales normales típicas.
 
@@
 
-En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la
-matriz de cambio de base P
+En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la matriz de cambio de base P
 
@@
 
-Debemos observar que los puntos quedan fijos en el espacio lo que cambia son las
-coordenadas dado que P es la transformación identidad expresada en ambas bases.
+Debemos observar que los puntos quedan fijos en el espacio lo que cambia son las coordenadas dado que P es la transformación identidad expresada en ambas bases.
 
-Para mantener invariante la norma debemos definir en el espacio de la habilidad
-con la base B un producto interno con matriz de Gramm M $$ M= P^T P$$
+Para mantener invariante la norma debemos definir en el espacio de la habilidad con la base B un producto interno con matriz de Gramm M $$ M= P^T P$$
 
@@
 
-Por lo tanto la norma permanece invariante esto se puede comprobar mediante un
-ejemplo particular.
+Por lo tanto la norma permanece invariante esto se puede comprobar mediante un ejemplo particular.
 
-Consderamos un caso particular de habilidad dada sus coordenadas en la base
-canónica por `r round(theta_C,4)` y su norma usando el producto punto es
-`r round(norma_C,4)`
+Consderamos un caso particular de habilidad dada sus coordenadas en la base canónica por `r round(theta_C,4)` y su norma usando el producto punto es `r round(norma_C,4)`
 
-Las coordenadas de este punto expresados en la base B son `r round(theta_B,4)` y
-su norma utilizando el producto internos con matriz de Gramm M es
-`r round(norma_B,4)`
+Las coordenadas de este punto expresados en la base B son `r round(theta_B,4)` y su norma utilizando el producto internos con matriz de Gramm M es `r round(norma_B,4)`
 
-Es claro que al existir una transformación lineal sobre las coordenadas esto
-afectará a las varianzas (no a la distribución ya que partimos de una normal) y
-también las correlaciones.
+Es claro que al existir una transformación lineal sobre las coordenadas esto afectará a las varianzas (no a la distribución ya que partimos de una normal) y también las correlaciones.
 
@@
 
-$$U=diag((Var(coord_C(\theta)))^{-1/2})$$ Veremos como son estas relaciones en
-el siguiente código
+$$U=diag((Var(coord_C(\theta)))^{-1/2})$$ Veremos como son estas relaciones en el siguiente código
 
@@
 
-Observemos que la norma que tomamos en la función es la norma respecto a la base
-canonica. Si queremos considerar la base no ortogonal debe calcularse la norma
-con el producto interno con matriz M
+Observemos que la norma que tomamos en la función es la norma respecto a la base canonica. Si queremos considerar la base no ortogonal debe calcularse la norma con el producto interno con matriz M
 
@@
 
-Consideremos el modelo expresado en la base no ortogonal B con discriminación
-`r a` y d=`r d`
+Consideremos el modelo expresado en la base no ortogonal B con discriminación `r a` y d=`r d`
 
@@
 
-$$a_i^{'}=(P^{-1})^T a_i$$ se puede modelizar utilizando las coordenadas en la
-base canónica, es decir la ecuación anterior da la transformación de los a al
-transformar los $\theta$. Esto imlica que si utilizamos las coordeadas de las
-habilidades en la base canónica debemos utilizar la $a'$.
+$$a_i^{'}=(P^{-1})^T a_i$$ se puede modelizar utilizando las coordenadas en la base canónica, es decir la ecuación anterior da la transformación de los a al transformar los $\theta$.
+Esto imlica que si utilizamos las coordeadas de las habilidades en la base canónica debemos utilizar la $a'$.
 
@@
 
-Los cosenos directores respecto a los ejes de la base canónica de la dirección
-del punto de máxima pendiente son
+Los cosenos directores respecto a los ejes de la base canónica de la dirección del punto de máxima pendiente son
 
-$$cos \gamma_i^E=\frac{a_i^{'}}{\sqrt{a_i^{'T}a_i^{'}}}$$ o con respecto a la
-base B
+$$cos \gamma_i^E=\frac{a_i^{'}}{\sqrt{a_i^{'T}a_i^{'}}}$$ o con respecto a la base B
 
@@
 
-$$D_i=\frac{-d_i}{\sqrt{a_i^{T}M^{-1}a_i}}$$ y la pendiente
-$$ S_i=\frac{\sqrt{a_i^{T}M^{-1}a_i}}{4}$$
+$$D_i=\frac{-d_i}{\sqrt{a_i^{T}M^{-1}a_i}}$$ y la pendiente $$ S_i=\frac{\sqrt{a_i^{T}M^{-1}a_i}}{4}$$
 
@@
 
-Graficamos un vector cuya norma es $S_i$ aplicado sobre el punto donde se da el
-máximo slope
+Graficamos un vector cuya norma es $S_i$ aplicado sobre el punto donde se da el máximo slope
 
@@
 
-Lo anteriore lo podemos representar para cualesquiera bases mediante la función
-representa
+Lo anteriore lo podemos representar para cualesquiera bases mediante la función representa
 
@@
 
-Por ejemplo si usamos nuevamente la base $\mathfrak{B}$ y la canónica pero ahora
-el ítem tiene a=(0.5,2) y d=1.1
+Por ejemplo si usamos nuevamente la base $\mathfrak{B}$ y la canónica pero ahora el ítem tiene a=(0.5,2) y d=1.1
 
@@
 
-El modelo de TRI expresa una relación en el espacio de coordenadas de la
-habilidad y la discriminación y en ese espacio de coordenadas $R^n$ se considera
-el producto punto. En principio el rasgo $\theta$ tiene sus coordenadas
-expresadas en la base $\mathfrak{B}$ que representa a las reales dimensiones que
-tiene el rasgo. También las discriminaciones estan expresadas en alguna base que
-al no tener restricciones previas podemos suponer cualquier base que fijará las
-transformaciones del espacio. En particular supondremos que la base de origen en
-el espacio $\mathcal A$ tiene como vectores las columnas de $(P^{-1})^T$. La
-llamaremos $\mathfrak{A}$
+El modelo de TRI expresa una relación en el espacio de coordenadas de la habilidad y la discriminación y en ese espacio de coordenadas $R^n$ se considera el producto punto.
+En principio el rasgo $\theta$ tiene sus coordenadas expresadas en la base $\mathfrak{B}$ que representa a las reales dimensiones que tiene el rasgo.
+También las discriminaciones estan expresadas en alguna base que al no tener restricciones previas podemos suponer cualquier base que fijará las transformaciones del espacio.
+En particular supondremos que la base de origen en el espacio $\mathcal A$ tiene como vectores las columnas de $(P^{-1})^T$.
+La llamaremos $\mathfrak{A}$
 
-Al realizar un cambio de base en el espacio $\Theta$ donde se transforman las
-coordenadas de la base $\mathfrak{B}$ a la base $\mathfrak{E}$ (en nuestro caso
-la canónica) mediante la matriz P a partir de la ecuación de ligadura dada por
-el modelo observamos que las coordenadas de las dicriminaciones ($a_i$) se
-trasforman en $a^'_i$ mediante la ecuación
+Al realizar un cambio de base en el espacio $\Theta$ donde se transforman las coordenadas de la base $\mathfrak{B}$ a la base $\mathfrak{E}$ (en nuestro caso la canónica) mediante la matriz P a partir de la ecuación de ligadura dada por el modelo observamos que las coordenadas de las dicriminaciones ($a_i$) se trasforman en $a^'_i$ mediante la ecuación
 
@@
 
-Al ser $(P^{-1})^T$ invertible representa una matriz de cambio de base que lleva
-las coordenasas de $a_i$ expresadas en la base $\mathfrak{A}$ a la base canónica
+Al ser $(P^{-1})^T$ invertible representa una matriz de cambio de base que lleva las coordenasas de $a_i$ expresadas en la base $\mathfrak{A}$ a la base canónica
 
-Además al tratarse de una transformación identidad la norma de $a_i$ no cambia
-al expresarse en las diferentes bases. Para mantener esta condición se induce un
-producto interno en el espacio original con matriz de Gramm $M^{-1}$
+Además al tratarse de una transformación identidad la norma de $a_i$ no cambia al expresarse en las diferentes bases.
+Para mantener esta condición se induce un producto interno en el espacio original con matriz de Gramm $M^{-1}$
 
@@
 
-El hecho de que la media sea 0 no hace perder generalidad pies se puede centrar
-el rasgo.
+El hecho de que la media sea 0 no hace perder generalidad pies se puede centrar el rasgo.
 
-Suponemos además que la base natural de los rasgos es $\mathfrak{B}$ que en
-nuestro ejemplo es
+Suponemos además que la base natural de los rasgos es $\mathfrak{B}$ que en nuestro ejemplo es
 
@@
 
-Debemos entonces desarrollar como queda la matriz de cambio de base entre la
-base $\mathfrak{B}$ y la base de vectores propios $\mathfrak{L}$ de la matriz
-$\Sigma$
+Debemos entonces desarrollar como queda la matriz de cambio de base entre la base $\mathfrak{B}$ y la base de vectores propios $\mathfrak{L}$ de la matriz $\Sigma$
 
@@
 
-La descomposicion espectral se expresa en la base canónica tanto para los datos
-con coordenadas en la base $\mathfrak{B}$ como en la base $\mathfrak{E}$.
+La descomposicion espectral se expresa en la base canónica tanto para los datos con coordenadas en la base $\mathfrak{B}$ como en la base $\mathfrak{E}$.
 
@@
 
-y lo propio con $\Sigma$ (con las coordenadas oblicuas)
+y lo propio con $\Sigma_B$ (con las coordenadas oblicuas)
 
 $$
-\Sigma = Q \Lambda Q^T
+\Sigma_B = Q_B \Lambda_B Q_B^T
 $$
 
-Además
+Además como ya vimos
 
 $$
-\Sigma_E=P \Sigma P^T
+\Sigma_E=P \Sigma_B P^T
 $$
@@
 $$
-Q_E \Lambda_E Q_E^T = P Q \Lambda Q^T P^T
+Q_E \Lambda_E Q_E^T = P Q_B \Lambda_B Q_B^T P^T
 $$
 
-Si tomamos la matriz $L$ definida como la matriz diagonal que contiene el
-cociente de los valores propios de $\Lambda$ entre $\Lambda_E$
+Si tomamos la matriz $L$ definida como la matriz diagonal que contiene el cociente de los valores propios de $\Lambda_B$ entre $\Lambda_E$
 
@@
 $$
-Q_E \Lambda_E Q_E^T = P Q L^{1/2} \Lambda_E (P Q L^{1/2} )^T
+Q_E \Lambda_E Q_E^T = P Q_B L^{1/2} \Lambda_E (P Q_B L^{1/2} )^T
 $$
@@
 $$
-Q_E = P Q L^{1/2}
+Q_E = P Q_B L^{1/2}
 $$
 
-Por otro lado la matriz de cambio de base entre $\mathfrak{B}$ y $\mathfrak{L}$
-es $PQ_E^{-1}$ ya que $Q_E$ es la matriz de cambio de base entre $\mathfrak{B}$
-y la base canónica
+Como $Q_E$ tiene los vectores propios como columnas, es la matriz de cambio de base entre la base $\mathfrak{L}$ y la base canónica o sea
 
-Luego la matriz de cambio de base queda
+$$\theta^{\mathfrak{E}}=Q_E\theta^{\mathfrak{L}}$$
+
+Como $$\theta^{\mathfrak{E}}=P\theta^{\mathfrak{B}}$$
+
+Luego la matriz de cambio de base entre $\mathfrak{B}$ y $\mathfrak{L}$ es $Q_E^{-1}P$
+
+Una expresion de la matriz de cambio de base es
 
 $$
-PQ_E^{-1} = P( PQL^{1/2} )^{-1} = L^{-1/2} PQ^{-1} P^{-1}
+Q_E^{-1}P = ( PQ_BL^{1/2} )^{-1} P= L^{-1/2} Q_B^{-1}
 $$
 
-Se puede aplicar con esta matriz de cambio de base los resultados anteriores
-hallados.
+Veamos como queda
+
+```{r DEBC, echo=FALSE, include=TRUE}
+sigma_E=var(coord_C)
+aux_E=eigen(sigma_E)
+Q_E=aux_E$vectors
+Qinv_E=solve(Q_E)
+lam_E=diag(aux_E$values)
+auxi_E=Q_E%*%lam_E%*%t(Q_E)
+
+sigma_B=var(coord_B)
+aux_B=eigen(sigma_B)
+Q_B=aux_B$vectors
+Qinv_B=solve(Q_B)
+lam_B=diag(aux_B$values)
+auxi_B=Q_B%*%lam_B%*%t(Q_B)
+
+KK=Qinv_E%*%P
+llam_E=diag(lam_E)
+llam_B=diag(lam_B)
+L=diag(llam_B/llam_E)
+L_raiz=diag(diag(L)^(-1/2))
+L_raiz%*%Qinv
+
+coord_L <- t(apply(coord_B, 1, function(x) KK %*% as.matrix(x)))
+
+```
+
+$Q_B$ es
+
+```{r , echo=FALSE, include=TRUE}
+print(Q_B)
+```
+
+$Q_E$ es
+
+```{r , echo=FALSE, include=TRUE}
+print(Q_E)
+```
+
+La matriz de cambio de base $Q_E^{-1}P$ es
+
+```{r , echo=FALSE, include=TRUE}
+print(KK)
+print(len(KK[,1]))
+print(len(KK[,2]))
+```
+
+Se puede aplicar con esta matriz de cambio de base los resultados anteriores hallados.
+
+Como queda la nube ahora
+
+```{r coord4, echo=FALSE, include=TRUE}
+PP2=KK
+coord_VP <- t(apply(coord_B, 1, function(x) PP2 %*% as.matrix(x)))
+plot(coord_VP,xlim=c(-4,4),ylim=c(-4,4))
+arrows(0, 0, 4*PP2[1,1],4*PP2[2,1] , lwd=3, col=2,code=2)
+arrows(0, 0, 4*PP2[1,2], 4*PP2[2,2], lwd=3, col=2,code=2)
+arrows(0, 0, -3*PP2[1,1],-3*PP2[2,1] , lwd=3, col=2,code=0)
+arrows(0, 0, -3*PP2[1,2], -3*PP2[2,2], lwd=3, col=2,code=0)
+grid()
+
+```
+
+La matriz de varianza covarianza es
+
+```{r , echo=FALSE, include=TRUE}
+print(var(coord_VP))
+```
+
+Otra manera de proceder para encontrar la base y su matriz es de nuevo considerar
+
+$$
+\Sigma_B = Q_B \Lambda_B^{1/2} \Lambda_B^{1/2} Q_B^T
+$$ Si consideramos como mariz de cambio de base
+
+$$P=\Lambda_B^{1/2} Q_B^T$$ entonces $$\Sigma_B =P^T P=M$$ La matriz de cambio de be queda
+
+```{r , echo=FALSE, include=TRUE}
+lam_medio=diag(aux_B$values^(1/2))
+KK2=lam_medio%*%t(Q)
+```
+
+```{r , echo=FALSE, include=TRUE}
+print(KK2)
+print(t(KK2)%*%KK2)
+```
+
+```{r coord5, echo=FALSE, include=TRUE}
+PP2=KK2
+coord_sigma <- t(apply(coord_B, 1, function(x) PP2 %*% as.matrix(x)))
+plot(coord_sigma,xlim=c(-4,4),ylim=c(-4,4))
+arrows(0, 0, 4*PP2[1,1],4*PP2[2,1] , lwd=3, col=2,code=2)
+arrows(0, 0, 4*PP2[1,2], 4*PP2[2,2], lwd=3, col=2,code=2)
+arrows(0, 0, -3*PP2[1,1],-3*PP2[2,1] , lwd=3, col=2,code=0)
+arrows(0, 0, -3*PP2[1,2], -3*PP2[2,2], lwd=3, col=2,code=0)
+grid()
+
+```
+
+También recordando que \$ Q_B^T=Q_B^{-1}\$ tenemos
+
+$$
+\Sigma_B^{-1} = Q_B \Lambda_B^{-1/2} \Lambda_B^{-1/2} Q_B^T
+$$ Si consideramos como matriz de cambio de base
+
+$$P=\Lambda_B^{-1/2} Q_B^T$$ entonces $$\Sigma_B^{-1} =P^T P=M$$ La matriz de cambio de be queda
+
+```{r , echo=FALSE, include=TRUE}
+lam_mediomenos=diag(aux_B$values^(-1/2))
+KK3=lam_mediomenos%*%t(Q)
+```
+
+```{r , echo=FALSE, include=TRUE}
+print(KK3)
+print(t(KK3)%*%KK3)
+```
+
+```{r coord6, echo=FALSE, include=TRUE}
+PP2=KK3
+coord_menossigma <- t(apply(coord_B, 1, function(x) PP2 %*% as.matrix(x)))
+plot(coord_menossigma,xlim=c(-4,4),ylim=c(-4,4))
+arrows(0, 0, 4*PP2[1,1],4*PP2[2,1] , lwd=3, col=2,code=2)
+arrows(0, 0, 4*PP2[1,2], 4*PP2[2,2], lwd=3, col=2,code=2)
+arrows(0, 0, -3*PP2[1,1],-3*PP2[2,1] , lwd=3, col=2,code=0)
+arrows(0, 0, -3*PP2[1,2], -3*PP2[2,2], lwd=3, col=2,code=0)
+grid()
+
+```
+
+Varianza de la nube
+
+```{r , echo=FALSE, include=TRUE}
+print(var(coord_menossigma))
+```

### 2023-11-02 notebooks/example_script.Rmd
Discards writing canonical markdown to avoid wrapping lines

@@
   markdown:
-    canonical: true
+    canonical: false
 ---

### 2023-11-13 notebooks/Graphical_representation_alternatives.Rmd
Versions notebook of alternatives to the graphical representation

---
title: 'Graphical representation example: Alternative interpretations'
output:
  html_document:
    df_print: paged
    code_folding: hide
---

```{r setup}
library(knitr)
library(patchwork)
library(rstudioapi)

ROOT_DIR <- getActiveProject()
DOC_DIR  <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      =   300,
  dev.args = list(type = "cairo-png")
)

# Output configuration options:
options(digits = 3)
```

# Introduction

After the 2023/11/02 discussion, there are two issues that remain unclear
regarding the graphical representation of the items.

1. Whether the correlation-based or the covariance-based matrix should be used
in the graphical representation, and whether it has an effect on its interpretation.

2. The need to represent the items in the latent vs. the test space.

We have a look at the example changing these two issues to see the effect.

# Preface

```{r read-chunks-graphical-example}
#| echo: false
read_chunk("src/Graphical_example_paper.R")
```

Load libaries:

```{r libraries}
```

Source R scripts:

```{r sources}
```

Define constants:

```{r constants}
```

Configure graphical output:

```{r graphical-output-conf}
```

# Covariance-based parameters and representation

As I say in the draft, my understanding is that the correlation-based version of
the parameters should be used for the graphical representation, as these are
scale preserving, and therefore can be interpreted straightforwardly.

The graphical representation, as computed in the draft, is:

```{r compute-example-items}
```

```{r compose-oblique-plot}
```

```{r compose-orthogonal-plot}
```

```{r item-plot-out}
#| fig.cap:   Item plots in an orthogonal (left) and oblique (right) test space.
#| fig.height: 3.1
#| fig.width:  6.65
plot_orth + plot_oblique
```

And the correlation-based parameters:

```{r example-items-table-out}
#| cache: false
#| echo:  false
item_params |> select(-ends_with("_orth"))
```

Let's now compute a covariance-based version of the parameters. In this case,
we defined the variances of the two latent dimensions as 1, so the
covariance-based and the correlation-based versions of the indices are the same.
In order to solve this, we define a new covariance matrix where the variance are
different from 1. We set them to $[2, 1.5]$ instead:

```{r}
variances  <- c(2, 1.5)
sd_matrix  <- variances |> sqrt() |> diag()
cov_matrix <- sd_matrix %*% corr_matrix %*% sd_matrix
```

With this covariance matrix, we compute the item parameters again:

```{r}
# Oblique case:

## Compute parameters and coordinates:
items_oblique_params <- items_M2PL |> compute_mirt_params(
  all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
  cov_matrix = cov_matrix,                              # Input covariances
  dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
)
items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
  D, MDISC, starts_with(COSINE_DIRTYPE),
  transform       = transform_matrix,
  original_coords = FALSE
)
items_oblique        <- full_join(
  items_oblique_params, items_oblique_coords,
  by = ITEM_COLKEY
)

# Create the parameter table again:
item_params <- items_M2PL |>
  full_join(
    items_orth |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
  )                       |>
  full_join(
    items_oblique |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
    by     = ITEM_COLKEY,
    suffix = paste0('_', c(ORTH_SUFFIX, OBL_SUFFIX))
  )
```

```{r example-items-table-out}
```

### 2023-11-13 notebooks/Graphical_representation_alternatives.Rmd
Separates the tables with the output parameters of the orthogonal and oblique representation

@@
 
-The graphical representation, as computed in the draft, is:
+We first compute the original item parameters and coordinates:
 
@@
 ```
+
+For reference, the MCLM version of the parameters, along with the "orthogonal"
+(i.e. agnostic) ones, are:
+
+```{r example-items-table-out}
+item_params |> select(-ends_with("_obl"))
+```
+
+The graphical representation (only the oblique version is shown), as computed in
+the draft, is:
 
@@
 ```{r item-plot-out}
-#| fig.cap:   Item plots in an orthogonal (left) and oblique (right) test space.
-#| fig.height: 3.1
-#| fig.width:  6.65
-plot_orth + plot_oblique
+#| fig.cap: Item plots in an oblique test space.
+plot_oblique
 ```
 
-And the correlation-based parameters:
+And the correlation-based parameters and coordinates:
 
-```{r example-items-table-out}
-#| cache: false
-#| echo:  false
-item_params |> select(-ends_with("_orth"))
+```{r example-items-obl-out}
+items_oblique
 ```
@@
 # Create the parameter table again:
-item_params <- items_M2PL |>
-  full_join(
-    items_orth |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
-  )                       |>
-  full_join(
+item_params <- items_M2PL |> full_join(
   items_oblique |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
-    by     = ITEM_COLKEY,
-    suffix = paste0('_', c(ORTH_SUFFIX, OBL_SUFFIX))
+  by = ITEM_COLKEY
 )
@@
 
-```{r example-items-table-out}
+```{r example-items-obl-out}
 ```

### 2023-11-15 notebooks/Graphical_representation_alternatives.Rmd
Formats output with table of contents and section headers

@@
     code_folding:    hide
+    toc:             yes
+    toc_float:       yes
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
+## MCLM parameters
+
 For reference, the MCLM version of the parameters, along with the "orthogonal"
@@
 ```
+
+## Parameters and representation in the original oblique space
 
@@
 ```{r item-plot-out}
-#| fig.cap: Item plots in an oblique test space.
-plot_oblique
+#| fig.cap: |
+#|   Item plots in an oblique test space, with original (i.e., standard)
+#|   covariance matrix
+plot_oblique + scale_color_manual(values = PALETTE)
 ```
@@
 
+## Parameters and representation in oblique space with different variances
+
 Let's now compute a covariance-based version of the parameters. In this case,
@@
 
-```{r}
-variances  <- c(2, 1.5)
-sd_matrix  <- variances |> sqrt() |> diag()
+```{r define-cov-matrix}
+VARIANCES  <- c(2, 1.5)
+sd_matrix  <- VARIANCES |> sqrt() |> diag()
 cov_matrix <- sd_matrix %*% corr_matrix %*% sd_matrix
@@
 
-```{r}
+```{r recompute-oblique-items}
 # Oblique case:

### 2023-11-15 notebooks/Graphical_representation_alternatives.Rmd
versions output and interpretation

@@
 ```
+
+As can be seen, the directions change, due to the difference in the scale of the
+two dimensions induced by the different variances. The resulting graphical
+representation is:
+
+```{r compose-oblique-plot}
+```
+
+```{r item-plot-cov-out}
+#| fig.cap: |
+#|   Item plots in an oblique test space, with variances [2, 1.5]
+plot_oblique +
+  scale_x_continuous(
+    limits       = c(-2, 4.2),
+    breaks       = (0:4) - 2 / CORR_ARC_TAN,
+    labels       = 0:4,
+    minor_breaks = NULL,
+    name         = NULL,
+    oob          = oob_keep
+  )          +
+  scale_color_manual(values = PALETTE)
+```
+
+### Interpretation
+
+The vectors are "stretched" in both directions, due to the different variances.
+The "length" by which they are scaled depends on the standard deviation of each
+dimension. In this case, we are using variances of 2 and 1.5. Therefore, the
+vectors are stretched more in the first (horizontal) direction than in the
+second (oblique) one.

### 2023-11-15 notebooks/Graphical_representation_alternatives.Rmd
Versions new case with standard deviations of .5

@@
 second (oblique) one.
+
+## Case with more predictable results
+
+My expectations were that the representation if the items would be independent
+of the covariance matrix, but this has not been the case. I had not predicted
+that the item vectors would be "stretched" in space, as has happened, although I
+should have expected this because I mention it in the draft, in section 3.3
+(Properties of the parameters). Therefore, I acknowledge this representation is
+the correct one.
+
+In order to test the prediction that the items are scaled by the standard
+deviation, let's test a case that is easier to predict. In order to it, I will
+use variances of 0.25, which result in standard deviations of 0.5. Then, if my
+understanding of the representation of the vectors is correct, all the vectors
+should be scaled by $tfrac{1}{2}$; that is, they should have half the original
+length. Their direction however should not change.
+
+```{r define-cov-matrix-prediction}
+VARIANCES  <- c(.25, .25)
+sd_matrix  <- VARIANCES |> sqrt() |> diag()
+cov_matrix <- sd_matrix %*% corr_matrix %*% sd_matrix
+```
+
+```{r recompute-oblique-items}
+```
+
+```{r example-items-obl-out}
+```
+
+```{r compose-oblique-plot}
+```
+
+```{r item-plot-cov-prediction-out}
+#| fig.cap: |
+#|   Item plots in an oblique test space, with variances [0.25, 0.25]
+plot_oblique + scale_color_manual(values = PALETTE)
+```
+
+### Interpretation
+
+The resulting MIDSC parameters are effectively half the length, as predicted.
+On the other hand, as the MIL is affected by the inverse of the MDICS, the MIL
+values are the double of the original ones. This result in vectors that are not
+only half the length of the originals. Also, their origin is twice as far from
+the origin of the latent space.
+
+## Conclusions
+
+The graphical representation reflects correctly the values of the parameters.
+However, the coordinates of the parameters in the latent space do not reflect
+the values of the original MCLM parameters.
+
+This is the reason why I understand that a correlation-based representation is
+more appropriate.

### 2023-11-15 notebooks/Graphical_representation_alternatives.Rmd
Versions section testing the effect of using the "covariance-based" transformation matrix

@@
 ```{r recompute-oblique-items}
-# Oblique case:
-
 ## Compute parameters and coordinates:
@@
 more appropriate.
+
+## Using the "covariance-based" transformation matrix
+
+After pondering the results in the previous sections, I have realized the
+following: Although the parameters and the coordinates in the original space are
+correctly computed, when I transform the coordinates into the rectangular
+coordinate system the transformation (i.e. mapping?) matrix I am using is not a
+square root of the inverse covariance matrix, as I state in the algorithm for
+computing the graphical representation (section 3.4 in the draft). Instead, what
+I was using was the square root fo the inverse correlation matrix, therefore
+applying a transformation that intends to be "scale invariant". As a result, the
+graphical representation shows the parameters scaled by the standard deviation
+of each dimension.
+
+However, I foresee that the following will happen: If I use the _correct_ square
+root matrix, that is, the square root matrix of the inverse of the covariance
+matrix, the graphical representation will be always the same, independently of
+the variances of the dimensions (it will only depend on the correlations).
+
+Let's try with the previous two examples, computing first the square root
+matrix. Given that $\mathbf{\Sigma} = \mathbf{SRS}$:
+
+$$\mathbf{\Sigma}^{-1} = (\mathbf{SRS})^{-1};$$
+
+$$\mathbf{\Sigma}^{-1} = \mathbf{S}^{-1}\mathbf{R}^{-1}\mathbf{S}^{-1};$$
+
+$$\mathbf{P^TP} = \mathbf{S}^{-1}\mathbf{P^T_RP_R}\mathbf{S}^{-1},$$
+
+where $\mathbf{P_R}$ refers to the (previously used) correlation-based
+transformation matrix. Therefore, we can equate[^mat-eq]
+$\mathbf{P} = \mathbf{P_R}\mathbf{S}^{-1}$.
+
+[^mat-eq]: They are not necessarily the same, but we can define them to be for
+our convenience, as defining it that way allegedly keeps the desired property of
+leaving the horizontal axis unrotated.
+
+With this formula, we compute the new (covariance-based) transformation matrix:
+
+```{r recompute-transform-matrix}
+cb_transform_matrix <- transform_matrix %*% solve(sd_matrix)
+```
+
+Let's verify that this matrix is a square root of the inverse covariance matrix
+
+```{r check-transform-matrix}
+t(cb_transform_matrix) %*%
+  cb_transform_matrix  |>
+  solve()              |>
+  all.equal(cov_matrix)
+```
+
+Both matrices are the same (with double numerical precision).
+
+Let's now try the latter example with this new transform matrix.
+
+```{r recompute-oblique-coords}
+## Compute coordinates:
+items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
+  D, MDISC, starts_with(COSINE_DIRTYPE),
+  transform       = cb_transform_matrix,
+  original_coords = FALSE
+)
+items_oblique        <- full_join(
+  items_oblique_params, items_oblique_coords,
+  by = ITEM_COLKEY
+)
+
+# Create the parameter table again:
+item_params <- items_M2PL |> full_join(
+  items_oblique |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
+  by = ITEM_COLKEY
+)
+```
+
+```{r example-items-obl-out}
+```
+
+```{r compose-oblique-plot}
+```
+
+```{r item-plot-cov-based-transform-out}
+#| fig.cap: |
+#|   Item plots in an oblique test space, with variances [0.25, 0.25], and using
+#|   the "covariance-based transformation matrix.
+plot_oblique + scale_color_manual(values = PALETTE)
+```

### 2023-11-22 notebooks/Graphical_representation_alternatives.Rmd
Versions preliminary interpretation of using the "covariance-based" transformation matrix
Versions conclusions of the "covariance-based representation" section

@@
 The resulting MIDSC parameters are effectively half the length, as predicted.
-On the other hand, as the MIL is affected by the inverse of the MDICS, the MIL
+On the other hand, as the MIL is affected by the inverse of the MDISC, the MIL
 values are the double of the original ones. This result in vectors that are not
-only half the length of the originals. Also, their origin is twice as far from
+only half the length of the originals; their origin is twice as far from
 the origin of the latent space.
@@
 ```
+
+### Interpretation
+
+The discrimination parameters are "rescaled" back to their original value.
+However, the shift of the location is even more exaggerated due to being
+scaled _again_ by a factor dependent on the inverse of the variances. Because of
+this, the new location in the graphic is thus meaningless, in terms of the
+interpretation of the multidimensional parameters.
+
+## Conclusion
+
+There are two ways to represent the item vectors in the bidimensional space:
+
+1. Using the correlation-based definition of the multidimensional parameters.
+
+2. Using the covariance-based definition of the multidimensional parameters.
+
+The first one has the advantage that the items are represented in the original
+scale of the test space, assuming a distribution with the given variances; the
+original MCLM parameters are therefore kept invariant. The second one implies
+that _the items are normalized to a standard distribution_; _the original MCLM
+parameters are thus scaled accordingly_ (see [Section 3.3][gh-section-ref] of
+the current draft), and the representation of the distribution in this case
+would be made with a unitary variance.
+
+[gh-section-ref]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/output/paper_draft/paper_M2PLM_parameters.Rmd#L676
+
+In each of these two cases though, the oblique coordinates must be transformed
+back to rectangular coordinates _using a transformation matrix that is a
+square-root matrix of the correlation and **not the covariance matrix**_.

### 2023-11-22 notebooks/Graphical_representation_alternatives.Rmd
Versions introduction of the "Representation int the latent space" section

@@
 1. Whether the correlation-based or the covariance-based matrix should be used
-in the graphical representation, and whether it has an effect on its interpretation.
+in the graphical representation, and whether it has an effect on its
+interpretation.
 
@@
 square-root matrix of the correlation and **not the covariance matrix**_.
+
+# Representation in the latent space instead of the test space
+
+It is a concern that the multidimensional parameters should be defined
+(according to Reckase's, -@reckase_difficulty_1985 definition), in the latent
+space. All the computations (both the ones here and in the draft) assume that
+the representation of the vectors are in the test space, after arguing that this
+should be the correct representation instead of the one in the latent space.
+
+It should be argued (probably more clearly in the draft) that the fact that
+Reckase proposes the representation in the latent space may be due to the
+confusion resulting from considering that both constitute a single entity. This
+is because in the orthogonal case they overlap, making them indistinguishable in
+practice.
+
+Let's try to represent these sample items in the latent space, in order to
+explore this representation.

### 2023-11-22 notebooks/Graphical_representation_alternatives.Rmd
Versions start of "Parameters and representation" section, including the function computing the multidimensional parameters

@@
 explore this representation.
+
+## Parameters and representation in the oblique space
+
+The MDISC and the distance component of the MIL are the same in both spaces.
+However, in order to represent the item in the latent space, instead of
+[Equation 25][gh-eq-25], we must apply [Equation 20][gh-eq-20] to compute the
+angles with the axes of the latent space. In order to that, we first need to 
+redefine the function [`compute_mirt_params()`][gh-par-fun] to compute the
+latent space angles instead of the test space ones:
+
+[gh-eq-25]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/output/paper_draft/paper_M2PLM_parameters.Rmd#L599
+
+[gh-eq-20]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/output/paper_draft/paper_M2PLM_parameters.Rmd#L532
+
+[gh-par-fun]: https://github.com/DaniMori/mirt-parameters/blob/8fa41540ff077f07db4b87840f0acc86eb4be2ea/R/Mirt_toolbox.R#L30
+
+```{r redefine-compute-mirt-params}
+compute_mirt_params <- function(items,
+                                intercept,
+                                discrimination,
+                                cov_matrix,
+                                item_id = item,
+                                dir_out = c("cos", "rad", "deg")) {
+  ## Argument checking and formatting: ----
+
+  # Enquote column selections:
+  item_id        <- rlang::enquo(item_id)
+  intercept      <- rlang::enquo(intercept)
+  discrimination <- rlang::enquo(discrimination)
+
+  # Check number of columns selected:
+
+  id_col     <- items |> dplyr::select(!!item_id)
+  intr_col   <- items |> dplyr::select(!!intercept)
+  discr_cols <- items |> dplyr::select(!!discrimination)
+
+  n_discr_cols <- discr_cols |> ncol()
+
+  assertive.properties::assert_is_of_length(id_col,   1)
+  assertive.properties::assert_is_of_length(intr_col, 1)
+
+  assertive.extra::assert_is_greater_than_or_equal_to(n_discr_cols, 1)
+
+  # Check/format covariance matrix:
+  if (missing(cov_matrix)) {
+
+    cov_matrix <- n_discr_cols |> diag()
+
+  } else {
+
+    dim_cov <- cov_matrix |> ncol()
+
+    assertive.base::assert_are_identical(dim_cov, n_discr_cols)
+    assertive.extra::assert_is_covariance_matrix(cov_matrix)
+  }
+
+  # Direction options:
+  dir_out <- dir_out |> match.arg(several.ok = TRUE)
+  one_dir <- dir_out                      |>
+    assertive.properties::is_of_length(1) |>
+    as.logical()
+
+  ## Main: ----
+
+  # Compute inverse s.d. matrix:
+  var_matrix    <- cov_matrix |> diag() |> diag()
+  inv_sd_matrix <- var_matrix |> sqrt() |> solve()
+
+  items                                                                      |>
+    tidyr::pivot_longer(!!discrimination, names_to = "par", values_to = "a") |>
+    dplyr::group_by(!!item_id)                                               |>
+    # This renaming is necessary for accessing the variable in `transmute`:
+    dplyr::rename(d = !!intercept)                                           |>
+    dplyr::transmute(
+      !!item_id,
+      ## Compute the multidimensional parameters:
+      dim   = par |> stringr::str_extract("\\d+"),
+      MDISC = (t(a) %*% cov_matrix %*% a) |> drop() |> sqrt(),
+      D     = - d / MDISC,
+      cos   = (inv_sd_matrix %*% cov_matrix %*% a / MDISC)  |> drop(),
+      rad   = acos(cos),
+      deg   = rad * RAD_DEG_FACTOR
+    )                                                                        |>
+    dplyr::ungroup()                                                         |>
+    select(!!item_id, dim, MDISC, D, all_of(dir_out))                        |>
+    tidyr::pivot_wider(
+      names_from  = dim,
+      values_from = all_of(dir_out),
+      names_glue  = one_dir |> if_else('{.value}_{.name}', '{.name}')
+    )
+}
+```

### 2023-11-22 notebooks/Graphical_representation_alternatives.Rmd
Update parameter matrix to give the direction parameters in the latent space

@@
 
-  # Compute inverse s.d. matrix:
-  var_matrix    <- cov_matrix |> diag() |> diag()
-  inv_sd_matrix <- var_matrix |> sqrt() |> solve()
+  # Compute Gram (i.e. inner product) matrix and "scaling" (D) matrix:
+  gram_matrix  <- cov_matrix |> solve()
+  scale_matrix <- gram_matrix |> diag() |> diag() |> sqrt() |> solve()
 
@@
       D     = - d / MDISC,
-      cos   = (inv_sd_matrix %*% cov_matrix %*% a / MDISC)  |> drop(),
+      cos   = (scale_matrix %*% a / MDISC) |> drop(),
       rad   = acos(cos),

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Add control for computing the arc-cosine of the "computational 1" cosine exception

@@
       cos   = (scale_matrix %*% a / MDISC) |> drop(),
-      rad   = acos(cos),
+      rad   = dplyr::if_else(
+        assertive.numbers::is_equal_to(cos, 1), # If cos is a "computational 1",
+        0,                                  # then fill in its arc-cos with a 0;
+        suppressWarnings(acos(cos))   # otherwise, compute its arc-cos normally.
+      ),
       deg   = rad * RAD_DEG_FACTOR

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Adds (original) chunk that computes the coordinates

@@
 ```
+
+```{r compute-oblique-items-latent-space}
+## Compute parameters and coordinates:
+items_oblique_params <- items_M2PL |> compute_mirt_params(
+  all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
+  cov_matrix = corr_matrix,                             # Input covariances
+  dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
+)
+items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
+  D, MDISC, starts_with(COSINE_DIRTYPE),
+  transform       = transform_matrix,
+  original_coords = FALSE
+)
+items_oblique        <- full_join(
+  items_oblique_params, items_oblique_coords,
+  by = ITEM_COLKEY
+)
+
+# Create the parameter table again:
+item_params <- items_M2PL |> full_join(
+  items_oblique |> select(item:deg_2, -starts_with(COSINE_DIRTYPE)),
+  by = ITEM_COLKEY
+)
+```

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Explains how the function is adapted to compute the directions in the latent space

@@
 
+In the previous function, we have inverted the roles of the covariance matrix
+(i.e., the Gram matrix in the test space) and the "Gram matrix" of the latent
+space, which is the inverse of the covariance matrix. Also, we have computed a
+"scale matrix" that is formed by the square root elements of the diagonal of the
+Gram matrix, inverted (i.e. matrix $\mathbf{D}$ in the manuscript). Note that,
+contrary to what happens in the test space (where the "scale" matrix is the
+inverse of the standard deviation matrix), this scale matrix in the latent space
+is "meaningless" (in terms of interpretation). Finally, in the test pace, the
+covariance matrix was used in the computation of the cosines, but here the Gram
+matrix is not; this makes sense since, as we have seen before, transforming from
+the test to the latent space implies multiplying by the inverse of the
+covariance matrix.
+
 ```{r compute-oblique-items-latent-space}

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Explains how the chunk is adapted to compute the coordinates in the latent space by calling the new function and transforming the coordinates accordingly.

@@
 
+We now compute the coordinates in the latent space. In order to compute the
+transformed coordinates, instead of the transformation matrix, I understand we
+need its "complementary" with the same role in the latent space. That would be
+the inverse transposed of that transformation matrix.
+
+We return in this case the original coordinates (in the latent space) along with
+the transformed ones (to rectangular coordinates).
+
 ```{r compute-oblique-items-latent-space}
@@
   D, MDISC, starts_with(COSINE_DIRTYPE),
-  transform       = transform_matrix,
-  original_coords = FALSE
+  transform       = transform_matrix_inv_transp,
+  original_coords = TRUE
 )

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Adds item table output selection
Versions chunks with original plot output in the test space

@@
 ```
+
+```{r example-items-obl-out}
+```
+```{r compose-oblique-plot-latent-space}
+plot_oblique <- items_oblique |>
+  arrange(desc(item))         |> # To plot them in reverse order
+  ggplot(
+    aes(
+      origin_transf_1, origin_transf_2,
+      xend  = end_transf_1, yend = end_transf_2,
+      color = item, fill = item
+    ),
+  )                                                         +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  )                                                         +
+  scale_x_continuous(
+    limits       = c(-2, 3),
+    breaks       = (0:4) - 2 / CORR_ARC_TAN,
+    labels       = 0:4,
+    minor_breaks = NULL,
+    name         = NULL,
+    oob          = oob_keep
+  )                                                         +
+  scale_y_continuous(
+    limits       = c(-2, 2.8),
+    breaks       = -3:3 * cos(CORR_VALUE),
+    labels       = function(x) x / cos(CORR_VALUE),
+    minor_breaks = 0,
+    name         = NULL,
+    oob          = oob_keep
+  )                                                         +
+  geom_abline(slope = CORR_ARC_TAN, linewidth = LINE_WIDTH) +
+  coord_fixed(expand = FALSE, clip = "on")                  +
+  scale_color_manual(values = PALETTE)                +
+  geom_abline(
+    slope     =  CORR_ARC_TAN,
+    intercept = -CORR_ARC_TAN * (-4:4),
+    linewidth = LINE_WIDTH,
+    linetype  = "17"
+  )                                                         +
+  geom_hline(yintercept = 0, linewidth = LINE_WIDTH)        +
+  theme(
+    axis.line          = element_blank(),
+    panel.grid.major.y = element_line(
+      color     = "black",
+      linewidth = LINE_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+
+```{r item-plot-latent-space-out}
+#| fig.cap: |
+#|   Item plots in an oblique, orthonormal latent space (with correlation = .5)
+plot_oblique
+```

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Versions changes to create the latent space grid

--- 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
@@
 ```
+
+```{r example-items-obl-out}
+```
+
+We now compose the plot in the latent space. Apart from the item representation,
+we need to plot the grid that corresponds to the latent space instead of the
+test space.
+
+```{r compose-oblique-plot-latent-space}
+TAN_LS <- -CORR_VALUE / CORR_ARC_SIN
+
+plot_oblique <- items_oblique |>
+  arrange(desc(item))         |> # To plot them in reverse order
+  ggplot(
+    aes(
+      origin_transf_1, origin_transf_2,
+      xend  = end_transf_1, yend = end_transf_2,
+      color = item, fill = item
+    ),
+  )                                                   +
+  scale_x_continuous(
+    limits       = c(-2, 3),
+    minor_breaks = NULL,
+    name         = NULL,
+    oob          = oob_keep
+  )                                                   +
+  scale_y_continuous(
+    limits       = c(-2, 2.8),
+    breaks       = -3:3 * 2 * TAN_LS,
+    labels       = function(y) 1 + y / (2 * TAN_LS),
+    minor_breaks = 0,
+    name         = NULL,
+    oob          = oob_keep
+  )                                                   +
+  coord_fixed(expand = FALSE, clip = "on")            +
+  scale_color_manual(values = PALETTE)                +
+  geom_abline(slope = TAN_LS, linewidth = LINE_WIDTH) +
+  geom_abline(
+    slope     =  TAN_LS,
+    intercept = -3:3 * 2 * TAN_LS,
+    linewidth = LINE_WIDTH,
+    linetype  = "17"
+  )                                                         +
+  geom_vline(xintercept = 0, linewidth = LINE_WIDTH)  +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  )                                                         +
+  theme(
+    axis.line          = element_blank(),
+    panel.grid.major.x = element_line(
+      color     = "black",
+      linewidth = LINE_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+
+```{r item-plot-latent-space-out}
+#| fig.cap: |
+#|   Item plots in an oblique, orthonormal latent space (with correlation = .5)
+plot_oblique
+```

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Versions explanation of the plot in the latent space

@@
 ```
+
+Unless there is a mistake, we can see that the representation is the same. The
+angles seem to be correct, and the computed coordinates seem to be given
+"perpendicular" to the latent space axes[^ls-coords]
+
+[^ls-coords]: This seems odd, there may be a mistake in the computation of the
+latent space grid. Also, it seems like the coordinates are not correct, even if
+they are (apparently) computed perpendicularly to the latent space axes.

### 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
Simplifies the use of the scale limits in the x axis

@@
 plot_oblique    +
-  scale_x_continuous(
-    limits       = c(-2, 4.2),
-    breaks       = (0:4) - 2 / CORR_ARC_TAN,
-    labels       = 0:4,
-    minor_breaks = NULL,
-    name         = NULL,
-    oob          = oob_keep
-  )          +
+  xlim(-2, 4.2) +
   scale_color_manual(values = PALETTE)

### 2023-11-24 notebooks/Graphical_representation_alternatives.Rmd
Computes the latent space grid (with detailed explanation)

--- 2023-11-23 notebooks/Graphical_representation_alternatives.Rmd
@@
 we need to plot the grid that corresponds to the latent space instead of the
-test space.
+test space. I made a previous version of the grid, but then I realised that I
+didn't quite understand how the grid should be built, apart from the axes.
+
+Therefore, I decide to try to build it following this procedure:
+
+1. First, I try to find the points where the two lines of the grid (one for each
+axis) instersects in the coordinate "1" of the other axis.
+
+2. Then I compute the transformed coordinates of point (1, 1).
+
+3. Tracing a line from this point to each of the intersections of the two axes
+gives the "base" grid lines. Then I only need to translate its coordinates by
+an integer number to get any other grid line.
+
+The points in step 1 must be the end of each of the two basis vectors.
+Therefore, this step implies simply obtaining the coordinates of those basis
+vectors in the orthonormal basis.
+
+Then, step 2 is:
+
+$$
+\begin{pmatrix}
+  1                          & 0 \\
+  - \rho / \sqrt{1 - \rho^2} & 1 / \sqrt{1 - \rho^2}
+\end{pmatrix}
+\begin{pmatrix}
+  1 \\ 1
+\end{pmatrix} = 
+\begin{pmatrix}
+  1 \\ (1 - \rho) / \sqrt{1 - \rho^2}
+\end{pmatrix}
+$$
+
+Finally, step 3:
+
+- The first grid line through axis 2, in point/slope form:
+
+$$
+\begin{pmatrix}
+  1 \\ (1 - \rho) / \sqrt{1 - \rho^2}
+\end{pmatrix} -
+\begin{pmatrix}
+  0 \\ 1 / \sqrt{1 - \rho^2}
+\end{pmatrix} =
+\begin{pmatrix}
+  1 \\ - \rho / \sqrt{1 - \rho^2}
+\end{pmatrix}
+$$
+
+$$
+slope = \frac{- \rho / \sqrt{1 - \rho^2}}{1} =
+  - \frac{\rho}{\sqrt{1 - \rho^2}}
+$$
+
+$$
+intercept = \frac{1}{\sqrt{1 - \rho^2}}
+$$
+
+- The first grid line through axis 1:
+
+$$
+\begin{pmatrix}
+  1 \\ (1 - \rho) / \sqrt{1 - \rho^2}
+\end{pmatrix} -
+\begin{pmatrix}
+  1 \\ - \rho / \sqrt{1 - \rho^2}
+\end{pmatrix} =
+\begin{pmatrix}
+  0 \\ 1 / \sqrt{1 - \rho^2}
+\end{pmatrix}
+$$
+
+It is thus a vertical line through (rectangular) coordinate
+$(1, (1 - \rho) / \sqrt{1 - \rho^2})$.
 
@@
     limits       = c(-2, 3),
+    breaks       = -2:3,
+    # labels       = function(x) x * CORR_ARC_COSEC,
     minor_breaks = NULL,
@@
     limits       = c(-2, 2.8),
-    breaks       = -3:3 * 2 * TAN_LS,
-    labels       = function(y) 1 + y / (2 * TAN_LS),
+    breaks       = -3:3 / CORR_ARC_SIN,
+    labels       = function(y) y * CORR_ARC_SIN - 2 * CORR_VALUE,
     minor_breaks = 0,
@@
     slope     = TAN_LS,
-    intercept = -3:3 * 2 * TAN_LS,
+    intercept = -3:3 / CORR_ARC_SIN,
     linewidth = LINE_WIDTH,
@@
 latent space grid. Also, it seems like the coordinates are not correct, even if
-they are (apparently) computed perpendicularly to the latent space axes.
+they were computed perpendicularly to the latent space axes.

### 2023-11-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Omits output from the check of the parameter computation with the original ones (in the empirical example)
Renames "ITEM_SPACE" objects more properly as "TEST_SPACE"

--- 2023-08-04 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
-invertible, so is $`r INNER_PROD_MATRIX`$, and $`r INNER_PROD_MATRIX_INV_EQ`$.
+invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
 
@@
 
-Therefore, an inner product $`r ITEM_SPACE_INNER_PROD`$ with Gram matrix
+Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
 $`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
@@
 The discrimination vectors are thus represented in their own space
-$`r TEST_SPACE`$ with basis $`r ITEM_SPACE_BASIS`$,
+$`r TEST_SPACE`$ with basis $`r TEST_SPACE_BASIS`$,
 which we refer to as the *test space* [@zhang_theoretical_1999].

### 2023-11-28 R/Formulae.R

@@ -277,16 +277,16 @@ INNER_PROD_MATRIX_INV_EQ    <- latex_eq(
 ### Test space ----
 
 # Discrimination vector coordinates:
-ITEM_SPACE_BASIS     <- latex_cal("B^*")
-DISCR_COORDS         <- latex_raised_to(DISCR_VECTOR, exp = ITEM_SPACE_BASIS)
-ITEM_SPACE_STD_BASIS <- latex_cal("E^*")
-DISCR_STD_COORDS    <- latex_raised_to(DISCR_VECTOR, exp = ITEM_SPACE_STD_BASIS)
+TEST_SPACE_BASIS     <- latex_cal("B^*")
+DISCR_COORDS         <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_BASIS)
+TEST_SPACE_STD_BASIS <- latex_cal("E^*")
+DISCR_STD_COORDS    <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_STD_BASIS)
 
 # Invariance assumption:
 DISCR_COORDS_TRANSP         <- latex_transp(DISCR_COORDS)
 DISCR_TRAIT_COORDS_PROD     <- latex(DISCR_COORDS_TRANSP, TRAIT_COORDS)
 DISCR_STD_COORDS_TRANSP     <- latex_transp(DISCR_STD_COORDS)
-DISCR_TRAIT_STD_COORDS_PROD <- latex(DISCR_STD_COORDS_TRANSP, TRAIT_STD_COORDS)
+DISCR_TRAIT_STD_COORDS_PROD <- latex(DISCR_STD_COORDS_TRANSP, TRAIT_ORTH_COORDS)
 INVARIANCE_EQ               <- latex_eq(
   DISCR_TRAIT_PROD,
   DISCR_TRAIT_COORDS_PROD,
@@ -312,18 +312,18 @@ ITEM_BASIS_VECTOR_EL      <- latex('v')
 ITEM_BASIS_VECTOR         <- latex_bf(ITEM_BASIS_VECTOR_EL)
 ITEM_BASIS_VECTOR_FIRST   <- latex_sub(ITEM_BASIS_VECTOR, 1)
 ITEM_BASIS_VECTOR_LAST    <- latex_sub(ITEM_BASIS_VECTOR, N_DIMS)
[…]

### 2023-11-29 output/paper_draft/paper_M2PLM_parameters.Rmd
Rewrites "Test space and laten space" section content to match its new order in the draft

@@
 Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
-Gram matrix, i.e. $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
+Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
 the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
@@
 
-Before proceeding further, we need to make a clarification
-regarding the representation of the items in the latent space.
-Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
-where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
-in the same basis $`r LS_BASIS`$ in their respective spaces.
-The coordinates of $`r DISCR_VECTOR`$ are then $`r DISCR_COORDS`$.
+Consider now the M2PL model formulation given by Equation
+\@ref(eq:M2PL-formulation), on one hand, and Equation \@ref(eq:change-basis) on
+the other.
 Since the response probability must be invariant with respect to
-a change of basis, $`r DISCR_COORDS`$ must also be transformed
-into $`r DISCR_STD_COORDS`$, such that $`r INVARIANCE_EQ`$;
-therefore, $`r DISCR_STD_COORDS_EQ`$.
-This shows that first a transformation through $`r INNER_PROD_MATRIX_INV`$
-is performed and then a change to the standard basis,
-so the $`r DISCR_STD_COORDS`$ are expressed in the standard basis.
-The norm of $`r DISCR_VECTOR`$ must then fulfill
-
-\begin{equation}
-  `r DISCR_VECTOR_MODULE_EQ`.                    (\#eq:discr-param-squared-norm)
+a change of basis,
+we must define the coordinates $`r DISCR_STD_COORDS`$ of the parameter vector
+$`r DISCR_VECTOR`$ in an orthonormal basis $`r TEST_SPACE_STD_BASIS`$,
+such that $`r INVARIANCE_EQ`$.
+If we assume that $`r DISCR_VECTOR`$ exists in vector space $`r TEST_SPACE`$,
+with coordinates $`r DISCR_COORDS`$ in basis $`r TEST_SPACE_BASIS`$,
+then we must define $`r DISCR_COORDS`$ such that
+
+\begin{equation}
+  `r DISCR_STD_COORDS_EQ`                                (\#eq:discr-vector-std)
+\end{equation}
+
+We want @reckase_difficulty_1985's definition of the multidimensional parameters
+in rectangular coordinates to be a particular case of our more general
+definition, so we can conveniently assume that the inner product in
+$`r TEST_SPACE_STD_BASIS`$ is the dot product,
+as we did with $`r LS_ORTH_BASIS`$.
+The norm of $`r DISCR_VECTOR`$ thus fulfills
+
+\begin{equation}
+  `r DISCR_VECTOR_MODULE_EQ`.
 \end{equation}
@@
 keeps the norm of $`r DISCR_VECTOR`$ invariant.
-The discrimination vectors are thus represented in their own space
-$`r TEST_SPACE`$ with basis $`r TEST_SPACE_BASIS`$,
-which we refer to as the *test space* [@zhang_theoretical_1999].
-The test space is in general different from the latent space $`r LATENT_SPACE`$,
-although there is a bijective relationship between the two sets of spaces.
+
+This shows that discrimination vectors are indeed represented in their own
+vector space $`r TEST_SPACE`$ with basis $`r TEST_SPACE_BASIS`$.
+We shall refer to this new space as the *test space* [@zhang_theoretical_1999].
+Although the inverse relationship between their respective inner product
+Gram matrices induces a bijective relationship between the two sets of spaces,
+it is worth highlighting that test space $`r TEST_SPACE`$
+and latent space $`r LATENT_SPACE`$ are different entities. Only in the
+orthonormal case the two spaces share a common inner product
+(i.e., the dot product), and thus can be mistaken.
 
@@
 
-The slope in direction $`r DIR_ANGLE_VEC`$ is given by
+The slope in direction $`r DIR_ANGLE_VEC_ORG`$ is given by
 
@@
   \begin{split}
-    `r DIR_COS_AS_DIR_COS_ORTH_EQ` \\
+    `r DIR_COS_AS_DIR_COS_ORTH_EQ`, \\
     `r TRAIT_POLAR_COEFF_COS_ORTH_EQ`.
@@
 
-We can then define
-
-\begin{equation}
-  `r DISCR_STD_COORDS_EQ`                                (\#eq:discr-vector-std)
-\end{equation}
-
-so
+As $`r DISCR_VECTOR`$ is defined in test space $`r TEST_SPACE`$ with basis
+$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
+coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
+\@ref(eq:discr-vector-std),
 

### 2023-11-29 R/Formulae.R

@@ -284,26 +284,20 @@ DISCR_STD_COORDS    <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_STD_BASIS)
 
 # Invariance assumption:
 DISCR_COORDS_TRANSP         <- latex_transp(DISCR_COORDS)
-DISCR_TRAIT_COORDS_PROD     <- latex(DISCR_COORDS_TRANSP, TRAIT_COORDS)
 DISCR_STD_COORDS_TRANSP     <- latex_transp(DISCR_STD_COORDS)
 DISCR_TRAIT_STD_COORDS_PROD <- latex(DISCR_STD_COORDS_TRANSP, TRAIT_ORTH_COORDS)
+DISCR_STD_TRANSF_TRAIT_PROD <- latex(DISCR_STD_COORDS_TRANSP, TRAIT_TRANSFORM)
 INVARIANCE_EQ               <- latex_eq(
   DISCR_TRAIT_PROD,
-  DISCR_TRAIT_COORDS_PROD,
-  DISCR_TRAIT_STD_COORDS_PROD
+  DISCR_TRAIT_STD_COORDS_PROD,
+  DISCR_STD_TRANSF_TRAIT_PROD
 )
 
 # Standardized discrimination vector:
-DISCR_STD_COORDS_DEF        <- latex(TRANSFORM_MATRIX_TRANSP_INV, DISCR_COORDS)
-DISCR_STD_COORDS_MATRIX_DEF <- latex(
-  TRANSFORM_MATRIX,
-  INNER_PROD_MATRIX_INV,
-  DISCR_COORDS
-)
-DISCR_STD_COORDS_EQ         <- latex_eq(
+DISCR_STD_COORDS_DEF <- latex(TRANSFORM_MATRIX_TRANSP_INV, DISCR_COORDS)
+DISCR_STD_COORDS_EQ  <- latex_eq(
   DISCR_STD_COORDS,
-  DISCR_STD_COORDS_DEF,
-  DISCR_STD_COORDS_MATRIX_DEF
+  DISCR_STD_COORDS_DEF
 )
 
 # Test space and test space basis:
@@ -335,6 +329,7 @@ INNER_PROD_DISCR_EQ  <- latex_eq(INNER_PROD_DISCR, INNER_PROD_DISCR_DEF)
 # Discrimination vector module:
[…]

### 2023-11-29 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects some punctuation signs in equations

--- 2023-09-15 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-::: {.proof}
-*Proof.*
-
-For every $`r DIM_INDEX`$; $`r DIM_ENUM_EQ`$,
-
-\begin{equation}
-  `r DIR_COSINE_PROOF_EQ`,                        (\#eq:dir-cosine-proof-prop-1)
-\end{equation}
-
-with $`r INNER_PROD_MATRIX_ROW`$ the $`r DIM_INDEX`$-th row of
-$`r INNER_PROD_MATRIX`$.
-
-Therefore, $`r DIR_COS_VEC_EQ`$ and $`r TRAIT_VECTOR_POLAR_EQ`$.
-$`r END_OF_PROOF`$
-:::
-
-:::
-
-## Model recasting into polar coordinates
-
-From Equation \@ref(eq:direction-cosine-vector),
-any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
-
-Thus, the M2PL model is expressed in polar coordinates as
+Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
 
@@
 we solve Equation \@ref(eq:M2PL-formula-polar) for $`r IRF_MAX_SLOPE_EQ`$ to get
-$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec), then
-
-\begin{equation}
-  `r DISTANCE_PARAM_EQ`,                                   (\#eq:distance-param)
+$`r TRAIT_NORM_SOLVED_EQ`$. Using equation \@ref(eq:dir-cos-vec),
+
+\begin{equation}
+  `r DISTANCE_PARAM_EQ`.                                   (\#eq:distance-param)
 \end{equation}
@@
    intercept as $b_i$ in the unidimensional case,
-   i.e. $`r DIST_INTERCEPT_REL`$.
+   i.e., $`r DIST_INTERCEPT_REL`$.
 
@@
 thus finding a *standard-metric* discrimination parameter
-(i.e. referred to a unitary variance space).
+(i.e., referred to a unitary variance space).
 
@@
 
-### @reckase_multidimensional_2009
-
 ```{r reset-root-wd}
@@
 # Version in Reckase, 2009, p. 153; this represents a usual 3D case
-COV_MATRIX_VALUE <- matrix(
+COV_MATRIX <- matrix(
   c(
@@
 
-```{r format-covariance-matrix}
-```
-
 ```{r create-params-table}
@@
 ```
-
-Table \@ref(tab:empirical-table-Reckase2009-out) draws from a classical example
-by @reckase_multidimensional_2009.
-In this 30-test item, there are three ten-item blocks assumed to approximate
-a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
-and items 21-30 dimension 2 (see Table 6.1, p. 153 therein).
-In his example, Reckase tests the estimation of these items
-on two simulated datasets: In the first one, the latent trait distribution
-is standard with null correlations; however, the second one has
-(apart from a non-null vector of means) as covariance matrix
-
-\begin{equation}
-  `r cov_matrix_value_out`,
-\end{equation}
-
-which corresponds to a correlation matrix of
-
-\begin{equation}
-  `r cor_matrix_value_out`.
-\end{equation}
-
-As we can see in Table \@ref(tab:empirical-table-Reckase2009-out), the agnostic
-versions of the multidimensional parameters
-(columns labelled _`r AGNOSTIC_ABBR`_)
-coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
-However, when we take into account the latent space structure
-(columns labelled $`r COV_MATRIX`$), we notice several differences.
-First, the $`r MDISC_ITEM`$ parameters tend to be larger
-than their agnostic counterparts.
-This happens because all the discrimination parameters as and the latent space
-correlations are positive; in this respect, these items are similar to item 1
-in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
-However, for the items mostly aligned with dimension 2 (items 21-30), as the
-$`r MDISC_ITEM`$ parameters are also scaled by a standard deviation smaller
-than 1, their values decrease, being some of them smaller than the agnostic
-versions. The $`r DISTANCE_PARAM`$ parametes decrease or increase, consequently,
-(see Equation \@ref(eq:M2PL-formula-polar)), thus being most of them smaller
-than the agnostic ones (again, with the exceptions only happening
-in the items aligned with dimension 2).
-Finally, we see that the direction angles are not as clearly separated as one
-might expect from the agnostic versions. This effect is more pronounced among
-the items aligned with dimensions 1 and 3, which are strongly correlated.
-In this situation, we notice that the angles tend to be much smaller, as the
-high correlation induces an alignment between the two dimensions
-(compare item 1 in the two test spaces in
-Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
 
@@
 
-The latter effect can be better noticed in another case, in which Reckase
-uses the same item parameters to demonstrate the over-specification of
-dimensions (see p. 183).
-In this example, he uses a covariance matrix that practically has rank 1,
-implying that the (alleged) three dimensions collapse into a single one, being
-thus the test unidimensional. In such a case, the covariance-based version of
-the direction angles are extremely close to 0 for all items, as one would expect
-from a unidimensional test, where all items are strictly aligned with the one
-dimension they measure. The reader interested can check out the parameters
-computed in such case in the Supplementary Table 1.
-<!-- #TODO: Include Supplementary Table -->
-
-### @tezza_modelo_2018
-
 ```{r set-cov-matrix-Tezza2018}
@@
 # > foi de aproximadamente 0,4.
-COV_MATRIX_VALUE <- diag(4)
-DIMS_CORR        <- .4
-COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
+COV_MATRIX <- diag(4)
+COV_MATRIX[1, 4] <- COV_MATRIX[4, 1] <- .4
 ```
@@
 ```
-
-@tezza_modelo_2018 provide us with another example that allows showcasing
-the effect of correlated dimensions on negative discrimination parameters.
-In their model, dimensions 1 and 4 are estimated to have a correlation of
-`r DIMS_CORR`.
-With such latent space structure, items that strongly discriminate in any or
-both of these two dimensions change drastically their multidimensional
-parameters from the agnostic to the covariance-based version
-(see Table \@ref(tab:empirical-table-Tezza2018-out)) ---this happens,
-for example, with items 57 and 60. These items also have different-sign
-discrimination parameters in dimensions 1 and 4, so their situation is
-similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
-Figure \@ref(fig:item-plot-out),
-thus the decrease in their $`r MDISC_ITEM`$ parameter.
-The effect on the direction is also patent in the angles these items make with
-these dimensions, which change up to almost 20º, compared with at most 7º in the
-other two dimensions, and around 1º in most cases.
-
-On the other hand, in cases such as items 12 or 59, which have
-relatively low discrimination parameters on those two dimensions
-(compared with the ones in dimension 2 and 3, respectively) are barely affected.
-In these cases, we also notice that the angles are barely affected.
-The most extreme change in these two examples occurs in
-the direction of item 12 with respect to dimension 4. We see how the large
-discrimination in dimension 1, when combined with the correlation
-with dimension 4, induces a decrease in the angular direction
-with that dimension.
-In this case, being the discrimination in dimension 4 so low, the situation is
-very similar to item 4 in Table \@ref(tab:example-items-table-out)
-and Figure \@ref(fig:item-plot-out).
-
-<!---BLOCK_LANDSCAPE_START--->
 
@@
 ```
+
+<!---BLOCK_LANDSCAPE_START--->
 

### 2023-11-29 output/paper_draft/paper_M2PLM_parameters.Rmd
Applies minor grammar and style corrections

@@
 their inception.<!--# Ref. needed? -->
-The so called *Multidimensional 2-Parameter Logistic*
+The so-called *Multidimensional 2-Parameter Logistic*
 [`r MODEL_ACRONYM`, @mckinley_extension_1983] model is among the most important
@@
 a latent trait space, no matter its structure, as an orthogonal space
---exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
-the R package *psych* <!--# Reference needed? -->graphically represent
+---exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
+the R package *psych*, <!--# Reference needed? -->graphically represents
 factor loadings on orthogonal axes even after an oblique rotation.
@@
 which mostly deals with uncorrelated dimensions
-[see @reckase_multidimensional_2009 for a comprehensive treatment on the topic].
+[see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
 This fact is probably due to the different spotlights
@@
 modern measurement theory [@mcdonald_test_1999] requires
-the relaxation of traditional constrains on both approaches.
+the relaxation of traditional constraints on both approaches.
 It is not uncommon to find recent examples of MIRT models where the
@@
 In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait, and thus
-are usually reverse scored-- is also widespread.<!--# Reference and example? -->
+---items that tap the negative pole of a certain trait, and thus
+usually reverse scored--- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine
@@
 (and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
-contrary to the constrains imposed by @reckase_difficulty_1985 [p. 411].
+contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
 However, the multidimensional item parameters can also be defined for
@@
 unchanging monotony. Additionally, we aim at providing a method for
-graphically representing these items in the fashion of @ackerman_creating_1994,
+graphically representing these items in the fashion of @ackerman_creating_1994
 but taking the non-orthogonality of the coordinate axes into account.
@@
 as a function of how we define and interpret the latent space.
-Next we discuss the properties of the indices,
+Next, we discuss the properties of the indices,
 propose a graphical representation method,
@@
 Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
-$`r LATENT_SPACE`$. As in @reckase_difficulty_1985, we need to
-reparameterize Equation \@ref(eq:M2PL-formulation) in polar coordinates,
-because the slope is dependent on the direction considered.
+$`r LATENT_SPACE`$.
+Because the slope is dependent on the direction considered,
+we need to reparameterize Equation \@ref(eq:M2PL-formulation)
+in polar coordinates, as in @reckase_difficulty_1985.
+
 However, Reckase's Equation 3 implicitly assumes orthogonality.
@@
 where $`r COV_MATRIX_STD`$ is a diagonal matrix.
-We may consider now two different possibilities:
+Now we may consider two different possibilities:
 We may define $`r TRANSFORM_MATRIX`$ such that $`r INNER_PROD_CORR_COND`$.
@@
 a rectangular Cartesian system.
-However, we want to avoid changing the scaling of the parameters
-in order to facilitate the interpretation,
-and thus we will prefer the correlation-based over the covariance-based indices.
+However, to facilitate the interpretation,
+we want to avoid changing the scaling of the parameters.
+We will thus prefer the correlation-based over the covariance-based indices.
 Thus, we define $`r TRANSFORM_MATRIX`$ such that
@@
 
-The general procedure for plotting an item $`r ITEM_INDEX`$ comprises
+The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
 the following steps:
@@
 parameters *and* the correlation; the sign of the discrimination parameters
-determine where the item will point at;
+determines where the item will point at;
 thus, an item with two positive (negative) discrimination parameters will point
@@
 the values of the multidimensional parameters.
-Thus, the covariance structure of the latent space should be taken into account
+Thus, the covariance structure of the latent space should be considered
 when computing the multidimensional parameters of a MIRT model.
@@
 Given these considerations, we propose to adopt the covariance-based version of
-the multidimensional parameters, and refer to them simply as
+the multidimensional parameters and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
@@
 the target of the instrument"
-(Reckase, M.D., personal communication, 28 april 2015).
+(Reckase, M.D., personal communication, 28 April 2015).
 However, the computations in the oblique space allow us to obtain

### 2023-11-29 output/paper_draft/paper_M2PLM_parameters.Rmd

--- 2023-11-29 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 [@ackerman_creating_1994]
----exploratory factor analysis software such as SPSS <!--# Reference needed? -->or
-the R package *psych* <!--# Reference needed? -->graphically represent
+---exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
+the R package *psych*, <!--# Reference needed? -->graphically represents
 factor loadings on orthogonal axes even after an oblique rotation.
@@
 as a function of how we define and interpret the latent space.
-Next we discuss the properties of these indices,
+Next, we discuss the properties of the indices,
 propose a graphical representation method,
@@
 Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
-$`r LATENT_SPACE`$. As in @reckase_difficulty_1985, we need to
-reparameterize Equation \@ref(eq:M2PL-formulation) in polar coordinates,
-because the slope is dependent on the direction considered.
-However, Reckase's Equation 3 implicitly assumes orthogonality, as we will see.
+$`r LATENT_SPACE`$.
+Because the slope is dependent on the direction considered,
+we need to reparameterize Equation \@ref(eq:M2PL-formulation)
+in polar coordinates, as in @reckase_difficulty_1985.
+However, Reckase's Equation 3 implicitly assumes orthogonality.
 In order to contemplate the general case,

### 2024-01-12 notebooks/Mirt_solution_exploration.Rmd
Creates initial version of the notebook

---
title: 'Exploration of the MIRT solutions'
output:
  html_document: 
    df_print:        paged
    code_folding:    hide
    toc:             yes
    toc_float:       yes
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(knitr)
library(mirt)

ROOT_DIR <- rstudioapi::getActiveProject()
DOC_DIR  <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  cache    = FALSE
)

# Output configuration options:
options(digits = 3)
```

# Introduction

We want to explore the solutions of package `{mirt}` when computing latent trait
estimates, given that, apparently, it computes a rotated solution. Rotated
solutions typically try to find the simplest structure possible, thus performing
an oblique rotation, if necessary, to find the highest number of null
discrimination parameters.

However, the MDISC and MID parameters are possibly estimated from the original,
unrotated solution.

### 2024-01-16 notebooks/Mirt_solution_exploration.Rmd
Checks `{mirt}` solution, including factor scores and multidimensional parameters

@@
 library(mirt)
+
+source('R/Mirt_toolbox.R', encoding = 'UTF-8')
 
@@
 However, the MDISC and MID parameters are possibly estimated from the original,
-unrotated solution.
+unrotated solution. Note that the values of these multidimensional parameters
+is invariant to the rotation.
+
+# Model estimation
+
+We start by fitting a model with sample data bundled with the `{mirt}` package
+itself. Let's start by using the first multidimensional example:
+
+```{r fit-model}
+data <- expand.table(LSAT7)
+itemstats(data)
+
+mod2 <- mirt(data, 2)
+coef(mod2)
+summary(mod2, rotate = 'oblimin') #oblimin rotation
+```
+
+## Parameter computation
+
+The default model is a 2-PL, so the items only have two discrimination
+parameters (one for each dimension), and an intercept parameter. Parameters `g`
+and `u` are restricted to `0` and `1`, respectively.
+
+After applying the `oblimin` rotation, the item parameters and covariances are:
+
+```{r rotated-irt-params}
+coefs <- coef(mod2, rotate = 'oblimin') |> map(as_tibble)
+
+item_pars <- coefs |>
+  bind_rows()      |>
+  slice(1:5)       |>
+  select(a1:d)     |>
+  rownames_to_column(var = "item")
+
+cov_pars <- coefs[['GroupPars']] |>
+  pivot_longer(everything())     |>
+  deframe()
+
+cov_matrix <- cov_pars[c('COV_11', 'COV_21', 'COV_21', 'COV_22')] |>
+  matrix(nrow = 2)
+```
+
+In order to check the result expressed in factor-analytic metric, we turn the
+factor-analytic parameters into IRT metric to see whether they are coincident:
+
+```{r param-equivalence}
+sum <- summary(mod2, rotate = 'oblimin') #oblimin rotation
+sum$rotF |> as_tibble() |> mutate(
+  across(everything(), ~1.702 * . / sqrt(1-as.numeric(sum$h2)))
+)
+```
+
+The equivalence holds.
+
+## Factor scores
+
+Factor scores are computed by means of the `fscores()` function. By default,
+`{mirt}` applies an `oblimin` rotation.
+
+We use the original LSAT7 response patterns to compute the factor scores:
+
+```{r compute-fscores}
+scores <- fscores(mod2, response.pattern = LSAT7[, 1:5])
+```
+
+## Multidimensional IRT parameters
+
+We compute the multidimensional IRT parameters using the `{mirt}` functions, and
+also the IRT parameters of the rotated solution, and compare them.
+
+```{r multidim-params}
+mod2 |> MDISC()
+mod2 |> MDIFF()
+
+item_pars |>
+  compute_mirt_params(d, a1:a2, cov_matrix = cov_matrix, dir_out = 'cos')
+```
+
+We see that they have the same values. Also, we can use the original, unrotated
+solution:
+
+```{r multidim-params-unrotated}
+coefs_unrot <- coef(mod2) |> map(as_tibble)
+
+item_pars_unrot <- coefs_unrot |>
+  bind_rows()                  |>
+  slice(1:5)                   |>
+  select(a1:d)                 |>
+  rownames_to_column(var = "item")
+
+item_pars_unrot |>
+  compute_mirt_params(d, a1:a2, dir_out = 'cos')
+```
+
+As expected, the MDISC and the location component of the MDIFF are coincident,
+but the direction cosines change.
+
+A visual inspection of the `MDISC()` and `MDIFF()` functions shows that the
+parameters are computed applying Reckase's formulae to the unrotated solution.

### 2024-01-16 notebooks/Mirt_solution_exploration.Rmd
Adds necessary packages
Corrects notebook output

@@
     df_print:        paged
-    code_folding:    hide
+    code_folding:    show
     toc:             yes
@@
 library(mirt)
+library(tidyverse)
 
@@
   matrix(nrow = 2)
+
+item_pars
 ```
@@
 ```{r compute-fscores}
-scores <- fscores(mod2, response.pattern = LSAT7[, 1:5])
+fscores(mod2, response.pattern = LSAT7[, 1:5])
 ```

### 2024-02-02 notebooks/Mirt_solution_exploration.Rmd
Changes the sourcing of R file so that it works independently of knitting directory
Improves notebook output to make it clearer

@@
 editor_options: 
-  chunk_output_type: console
+  chunk_output_type: inline
 ---
@@
 ```{r setup}
+#| message: false
 library(knitr)
@@
 library(tidyverse)
-
-source('R/Mirt_toolbox.R', encoding = 'UTF-8')
 
@@
   message  = FALSE,
-  cache    = FALSE
+  cache    = FALSE,
+  results  = 'hold'
 )
@@
 options(digits = 3)
+```
+
+```{r sources}
+source('R/Mirt_toolbox.R', encoding = 'UTF-8')
 ```
@@
 ```{r fit-model}
+#| message: false
 data <- expand.table(LSAT7)
@@
 
-mod2 <- mirt(data, 2)
+mod2 <- mirt(data, 2, verbose = FALSE)
 coef(mod2)
@@
 sum <- summary(mod2, rotate = 'oblimin') #oblimin rotation
-sum$rotF |> as_tibble() |> mutate(
+sum$rotF      |>
+  as_tibble() |>
+  mutate(
     across(everything(), ~1.702 * . / sqrt(1-as.numeric(sum$h2)))
-)
+  )           |>
+  # Update the column names so they refer correctly to the discrimination
+  #   parameters instead of the factor loadings.
+  rename_with(str_replace, everything(), pattern = 'F', replacement = 'a')
 ```

### 2024-02-15 output/Transcription_Mario_Luzardo/Transcription_of_Marios_derivations.Rmd

@@
 ```{r includes, cache=FALSE}
-source("R/Formulas.R")
-source("R/Graphical_example_transcriptions_Mario.R")
-source("R/Output.R")
+source("R/Formulas.R", encoding = 'UTF-8')
+source("R/Graphical_example_transcriptions_Mario.R", encoding = 'UTF-8')
+source("R/Output.R", encoding = 'UTF-8')
 ```

### 2024-02-24 notebooks/example_script.Rmd
Adds html output to Mario's notebook

--- 2023-10-31 notebooks/example_script.Rmd
@@
     fig_height:   6
+  html_document:
+    fig_width:    6
+    fig_height:   6
+    code_folding: hide
+    toc:          yes
+    toc_float:    yes
 editor_options:
   markdown:
-    canonical: true
+    canonical:    false
 ---

### 2024-02-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Deletes commnets with discarded suggestions (to add references and/or examples)

--- 2024-02-15 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 Multidimensional item response theory (MIRT) models have been widely used since
-their inception.<!--# Ref. needed? -->
+their inception.
 The so-called *Multidimensional 2-Parameter Logistic*
@@
 axes is prevalent in the common factor literature
-[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
+[see, e.g., @harman_modern_1970].
 This assumption is not so easily found in the MIRT literature though,
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
+is usually implicit in the MIRT literature without further consideration
 [see @zhang_theoretical_1999 for a notable exception].

### 2024-02-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Drops definition of the latent trait vector as a multivariate random normal variable in the model formulation

@@
 Multidimensional item response theory (MIRT) models have been widely used since
-their inception.
+their inception.<!--# Ref. needed? -->
 The so-called *Multidimensional 2-Parameter Logistic*
@@
 axes is prevalent in the common factor literature
-[see, e.g., @harman_modern_1970].
+[see, e.g., @harman_modern_1970].<!--# Thurstone's Multiple Factor Analysis, 1947? -->
 This assumption is not so easily found in the MIRT literature though,
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration
+is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
 [see @zhang_theoretical_1999 for a notable exception].
@@
 Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
-in basis $`r LS_BASIS_EQ`$.
-A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
-which is assumed to be a multivariate normal random variable:
-
-\begin{equation}
-  `r TRAIT_MV_DEF`,
-\end{equation}
-
-where $`r MEAN_VECTOR`$ is an $n$-element vector of means and
-$`r COV_MATRIX` = `r SD_MATRIX` `r CORR_MATRIX` `r SD_MATRIX`$
-an $n$-dimensional positive definite covariance matrix,
-with $`r CORR_MATRIX`$ the corresponding correlation matrix
-(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
-with $`r VAR_COV_EQ`$ (i.e., the variances).
-
+in basis $`r LS_BASIS_EQ`$,
+where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
 According to the `r MODEL_ACRONYM` model, the probability of

### 2024-02-28 R/Formulae.R

@@ -48,6 +48,8 @@ LS_BASIS_EQ        <- latex_eq(LS_BASIS, LS_BASIS_SET)
 # Latent vector definition:
 TRAIT_VECTOR    <- latex_bf("\\uptheta")
 TRAIT_VEC_IN_LS <- latex_in(TRAIT_VECTOR, LATENT_SPACE)
+
+# TODO: Move forward this block:
 MEAN_VECTOR     <- latex_bf("\\upmu")
 COV_MATRIX      <- latex_bf("\\upSigma")
 NORMAL_DISTR    <- latex_cal('N')

### 2024-02-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Solves merging error dropping repeated content in "Proposition 1" proof and section "Model recasting into polar coordinates"
Corrects reference to the MIL components

@@
 The (signed) distance to the origin,
-given by the second element of $`r MIL_PARAM`$, has an interesting property:
+given by the first component of $`r MIL_PARAM`$, has an interesting property:
 By the definition of the inner product,

### 2024-02-29 output/paper_draft/paper_M2PLM_parameters.Rmd
Solves merging error dropping repeated content in "Proposition 1" proof and section "Model recasting into polar coordinates"

--- 2024-02-28 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
-in basis $`r LS_BASIS_EQ`$,
-where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
+in basis $`r LS_BASIS_EQ`$.
+A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
+which is assumed to be a multivariate normal random variable:
+
+\begin{equation}
+  `r TRAIT_MV_DEF`,
+\end{equation}
+
+where $`r MEAN_VECTOR`$ is an $n$-element vector of means and
+$`r COV_MATRIX` = `r SD_MATRIX` `r CORR_MATRIX` `r SD_MATRIX`$
+an $n$-dimensional positive definite covariance matrix,
+with $`r CORR_MATRIX`$ the corresponding correlation matrix
+(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
+with $`r VAR_COV_EQ`$ (i.e., the variances).
+
 According to the `r MODEL_ACRONYM` model, the probability of
@@
 
-Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+Thus, the M2PL model is expressed in polar coordinates as
 

### 2024-03-01 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects minor erratum

--- 2024-02-28 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 ---items that tap the negative pole of a certain trait, and thus
-usually reverse scored--- is also widespread.<!--# Reference and example? -->
+usually reverse-scored--- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine

### 2024-03-12 notebooks/ObservacionesMario.Rmd
nueva rama

---
title: "ObservacionesMario"
author: "Mario Luzardo"
date: "2024-03-12"
output: html_document
---

  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls()) 
library(pspline)
library(mvtnorm)
library(sn)
library(mnormt)
library(scatterplot3d)
library(copula)
library(ggplot2)
library(fungible)
library(pracma)
library(matlib)
library(rgl)
library(MASS)
set.seed(123)
```

# Observación 1

Tenemos que $M=P^TP$ es decir viene a partir de la matriz de cambio de base. 
En nuestro problema tenemos conocidos  los ejes oblicuos dado que ellos representan los rasgos latentes del fenómeno de estudio y por lo tanto la matriz P.

Pero como saso de estudio consideremos que tenemos $M$ y queremos determinar P y por lo tanto la base oblicua. 
A partir de $M$ no se determina univocamente la matriz $P$ (y por lo tanto la base). Para ver la clase de las matrices $P$ o de cambio de base que son compatibles observemos que si $P_1$ es una matriz de cambio de base que verifica $M=P_1^TP_1$ entonces $P_2=AP_1$ con A matriz ortogonal ($A^T=A^{-1}$) determina la misma matriz de producto interno pues

$$P_2^TP_2=P_1^T A^TAP_1=P_1^T=P_1=M$$
Las rotaciones cumplen con esta propiedad

Esto implica que cualquiera de estas matrices (finalmente bases) determinan la misma matriz de Gramm sin embargo cualitativamente no tienen porque dar cuenta del fenómeno.

## Caso $M=R^{-1}$

Veamos ahora algunas observaciones respecto a la seccion Generalized multidimensional parameters cuando $M=R^{-1}$

  Además $M=P^tP$ donde P es una matriz de cambio de base.
Entonces cabe hallar P tal que $R^{-1}=P^tP$

  Estamos psrtiendo del hecho que no conocemos la base B pues si la conocemos, tenemos P y por ende M y esta puede o no coincidir con $R^{-1}$ (en general no).

Pondré dos posible formas de hallar P que no son la únicas pero sirven para ejemplificar.

a) via descompisición de Cholesky
b) via descomposición espectral

## Halla P mediante descompisición de Cholesky

Como R es simétrica y supongamos definida positiva entonces por Cholesky podemos escribir

$$R=L L^t$$
  donde L es una matrz triangular inferior,

$$M=R^{-1}=(L^{-1})^T L^{-1}$$

  luego

o sea $P=L^{-1}$

## Halla P mediante descompisición de espectral

  Idem como R es simétrica  y definida positiva por el teorema de la descomposición espectral

$$R=QDQ^T$$
  con Q matriz de vectores propios y D matriz diagonal de valores propios

luego

$$R=(U \sqrt(D))(\sqrt(D) U)^T$$

  por lo que $$P=(U\sqrt(D))^{-1}$$

  Ejemplo

Simulemos datos con matriz de varianza covarianzas

```{r  include=FALSE, echo=FALSE}
S1 <- matrix(c(4.4, 2.5, 2.5, 8), nrow = 2, ncol = 2)
print(S1)
```

```{r  include=FALSE, echo=FALSE}
S1 <- matrix(c(4.4, 2.5, 2.5, 8), nrow = 2, ncol = 2)
# Generar datos normales bivariados
n <- 30000  # Número de observaciones
mu <- c(0, 0)  # Medias de las variables
set.seed(1234567)
data <- mvrnorm(n = n, mu = mu, Sigma = S1)

# Convertir a data frame para visualización
data_df <- as.data.frame(data)
```

Los datos simulados tiene esta matriz $\Sigma$

```{r   echo=FALSE}
S1_E=var(data_df)
print(S1_E)
```

y matriz $R$
  ```{r   echo=FALSE}
R1=cor(data_df)
print(R1)
```
```{r   echo=FALSE}
plot(data_df$V1,data_df$V2)
```

Vamos a encontrar dos bases que pueden servir para estos datos y que generan la misma matriz de producto interno.

### Base usando Cholesky queda

```{r   echo=FALSE}
L=t(chol(R1))
P=solve(L)
print(P)
```

Angulo entre los ejes y la base canónica

```{r }
matlib::angle(P[,1],P[,2])
matlib::angle(P[,1],c(1,0))
matlib::angle(P[,2],c(1,0))
matlib::angle(P[,1],c(0,1))
matlib::angle(P[,2],c(0,1))
```

Usando que $$\theta^E=P\theta^B$$ (con E la base canónica) obtenemos las coordenadas en E

```{r   echo=FALSE}
thetaE1=as.matrix(data_df)%*%t(P)
```
Con varianza covarianza y correlaciones

```{r }
var(thetaE1)
cor(thetaE1)
```

Vemos que $\Sigma'$ no es diagonal

### Base usando descomposición espectral
```{r   echo=FALSE}
DE=eigen(R1)
raizD=diag(sqrt(DE$values))
K=DE$vectors%*%raizD
P2=solve(K)
print(P2)
```
Angulo entre los ejes y la base canónica

```{r }
matlib::angle(P2[,1],P2[,2])
matlib::angle(P2[,1],c(1,0))
matlib::angle(P2[,2],c(1,0))
matlib::angle(P2[,1],c(0,1))
matlib::angle(P2[,2],c(0,1))
```
Igual que antes se transforman los datos a la base canónica y se halla la matriz de varianza de los datos transformados

```{r   echo=FALSE}
thetaE2=as.matrix(data_df)%*%t(P2)
print(cov(thetaE2))
print(cor(thetaE2))
```

Igual que antes la matriz Vemos que $\Sigma'$ no es diagonal

## Caso $M=\Sigma ^{-1}$

En este caso $\Sigma ^{-1}=P^TP$ por lo tanto la matriz de varianzas y covarianzas del transformado es

$$Var(\theta^E)=Pvar(\theta^B) P^T=P \Sigma P^T=P P^{-1}(P^T)^{-1}P^T=I$$
Vemos que sto se verifica con los datos simulados

### Base usando Cholesky queda

```{r   echo=FALSE}
LS=t(chol(S1_E))
PS=solve(LS)
print(PS)
```

Angulo entre los ejes y la base canónica

```{r }
matlib::angle(PS[,1],PS[,2])
matlib::angle(PS[,1],c(1,0))
matlib::angle(PS[,2],c(1,0))
matlib::angle(PS[,1],c(0,1))
matlib::angle(PS[,2],c(0,1))
```

Coordenadas en E

```{r   echo=FALSE}
thetaES=as.matrix(data_df)%*%t(PS)
```
Con varianza covarianza y correlaciones

```{r }
var(thetaES)
cor(thetaES)
```

### Base usando descomposición espectral

```{r   echo=FALSE}
DE_S=eigen(S1_E)
raizD_S=diag(sqrt(DE_S$values))
K_S=DE_S$vectors%*%raizD_S
P2_S=solve(K_S)
print(P2_S)
```
Angulo entre los ejes y la base canónica

```{r }
matlib::angle(P2_S[,1],P2_S[,2])
matlib::angle(P2_S[,1],c(1,0))
matlib::angle(P2_S[,2],c(1,0))
matlib::angle(P2_S[,1],c(0,1))
matlib::angle(P2_S[,2],c(0,1))
```
Igual que antes se transforman los datos a la base canónica y se halla la matriz de varianza de los datos transformados

```{r   echo=FALSE}
thetaE2_S=as.matrix(data_df)%*%t(P2_S)
print(cov(thetaE2_S))
print(cor(thetaE2_S))
```

# Observación 2
Este punto para mi es importante pues las secciones anteriores se formulan en base a que conocemos la base oblicua donde el fenómeno es interpretable y tiene las dimensiones "verdaderas".

Entonces si uno tiene las respuestas por ej dicotomicas de los ítems hay una directa relación entre los parámetros y los loadings y comunalidades del análisis factorial

Por ejemplo en mirt se realiza un analisis factorial (con nfact) donde las variables son los ítems y se calculan los parámetros como

$a_{ij}=\frac{loading_{ij}}{\sqrt(1-h^2)}$

  Esto esta computado en el test space

Si usamos una rotación (por ej oblimin) podemos calcular la base en que estan los parametros comparandola con la solucion sin rotar. (PONER FORMULAS)

### 2024-03-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds reference to software in the introduction

@@
 a latent trait space, no matter its structure, as an orthogonal space
----exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
-the R package *psych*, <!--# Reference needed? -->graphically represents
+---exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
+the R package *psych* [@revelle_psych_2024], graphically represents
 factor loadings on orthogonal axes even after an oblique rotation.

### 2024-03-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Applies approved changes in "Model Formulation"
Changes approved by me and @mluzardov in the Google Docs version

--- 2024-03-01 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 ---items that tap the negative pole of a certain trait, and thus
-usually reverse-scored--- is also widespread.<!--# Reference and example? -->
+usually reverse scored--- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine
@@
 Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
-in basis $`r LS_BASIS_EQ`$.
-A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
-which is assumed to be a multivariate normal random variable:
-
-\begin{equation}
-  `r TRAIT_MV_DEF`,
-\end{equation}
-
-where $`r MEAN_VECTOR`$ is an $n$-element vector of means and
-$`r COV_MATRIX` = `r SD_MATRIX` `r CORR_MATRIX` `r SD_MATRIX`$
-an $n$-dimensional positive definite covariance matrix,
-with $`r CORR_MATRIX`$ the corresponding correlation matrix
-(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
-with $`r VAR_COV_EQ`$ (i.e., the variances).
-
+in basis $`r LS_BASIS_EQ`$, being $`r LS_BASIS`$ not necessarily orthonormal,
+where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
 According to the `r MODEL_ACRONYM` model, the probability of
@@
 
-Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
-$`r LATENT_SPACE`$.
+Our aim is to find in the first place
+the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
 Because the slope is dependent on the direction considered,

### 2024-03-28 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 a latent trait space, no matter its structure, as an orthogonal space
+[@ackerman_creating_1994]
 ---exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
@@
 In the non-cognitive domain, the use of so-called negatively-keyed items
----items that tap the negative pole of a certain trait, and thus
-usually reverse scored--- is also widespread.<!--# Reference and example? -->
+--items that tap the negative pole of a certain trait,
+usually reverse scored-- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine
@@
 propose a graphical representation method,
-and present two examples of application of our results.
+and present two practical examples applying them.
 We round up with a discussion about
@@
 in polar coordinates, as in @reckase_difficulty_1985.
-
 However, Reckase's Equation 3 implicitly assumes orthogonality.
@@
 
+### Test space and latent space
+
+Consider now the M2PL model formulation given by Equation
+\@ref(eq:M2PL-formulation), on one hand, and Equation \@ref(eq:change-basis) on
+the other.
+Since the response probability must be invariant with respect to
+a change of basis,
+we must define the coordinates $`r DISCR_STD_COORDS`$ of the parameter vector
+$`r DISCR_VECTOR`$ in an orthonormal basis $`r TEST_SPACE_STD_BASIS`$,
+such that $`r INVARIANCE_EQ`$.
+If we assume that $`r DISCR_VECTOR`$ exists in vector space $`r TEST_SPACE`$,
+with coordinates $`r DISCR_COORDS`$ in basis $`r TEST_SPACE_BASIS`$,
+then we must define $`r DISCR_COORDS`$ such that
+
+\begin{equation}
+  `r DISCR_STD_COORDS_EQ`                                (\#eq:discr-vector-std)
+\end{equation}
+
+We want @reckase_difficulty_1985's definition of the multidimensional parameters
+in rectangular coordinates to be a particular case of our more general
+definition, so we can conveniently assume that the inner product in
+$`r TEST_SPACE_STD_BASIS`$ is the dot product,
+as we did with $`r LS_ORTH_BASIS`$.
+The norm of $`r DISCR_VECTOR`$ thus fulfills
+
+\begin{equation}
+  `r DISCR_VECTOR_MODULE_EQ`.
+\end{equation}
+
+Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
+keeps the norm of $`r DISCR_VECTOR`$ invariant.
+
+This shows that discrimination vectors are indeed represented in their own
+vector space $`r TEST_SPACE`$ with basis $`r TEST_SPACE_BASIS`$.
+We shall refer to this new space as the *test space* [@zhang_theoretical_1999].
+Although the inverse relationship between their respective inner product
+Gram matrices induces a bijective relationship between the two sets of spaces,
+it is worth highlighting that test space $`r TEST_SPACE`$
+and latent space $`r LATENT_SPACE`$ are different entities. Only in the
+orthonormal case the two spaces share a common inner product
+(i.e., the dot product), and thus can be mistaken.
+
+### Direction cosines in the latent space
+
 The cosine between two vectors
@@
 
-We can then define
-
-\begin{equation}
-  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
-\end{equation}
-
-so
+As $`r DISCR_VECTOR`$ is defined in test space $`r TEST_SPACE`$ with basis
+$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
+coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
+\@ref(eq:discr-vector-std),
 

### 2024-03-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Expresses "Test space" section start more formally

@@
 
-Consider now the M2PL model formulation given by Equation
-\@ref(eq:M2PL-formulation), on one hand, and Equation \@ref(eq:change-basis) on
-the other.
+Let $`r TEST_SPACE`$ be the vector space made up by
+the set of $`r DISCR_VECTOR`$ vectors,
+with the sum and product by a scalar operations as in $`r REAL_N_SPACE`$.
+Consider also an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
+in $`r TEST_SPACE`$, such that $`r DISCR_STD_COORDS_ALT_EQ`$.
+<!-- #TODO: Continue here -->
 Since the response probability must be invariant with respect to

### 2024-03-28 R/Formulae.R

@@ -278,11 +278,17 @@ INNER_PROD_MATRIX_INV_EQ    <- latex_eq(
 
 ### Test space ----
 
+# Test space definition:
+TEST_SPACE   <- latex_bf("A")
+REAL_N_SPACE <- latex_raised_to("\\mathbb{R}", exp = N_DIMS)
+
 # Discrimination vector coordinates:
-TEST_SPACE_BASIS     <- latex_cal("B^*")
-DISCR_COORDS         <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_BASIS)
-TEST_SPACE_STD_BASIS <- latex_cal("E^*")
+TEST_SPACE_BASIS        <- latex_cal("B^*")
+DISCR_COORDS            <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_BASIS)
+TEST_SPACE_STD_BASIS    <- latex_cal("E^*")
 DISCR_STD_COORDS    <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_STD_BASIS)
+DISCR_STD_COORDS_ALT <- latex_coords(DISCR_VECTOR, basis = TEST_SPACE_STD_BASIS)
+DISCR_STD_COORDS_ALT_EQ <- latex_eq(DISCR_STD_COORDS, DISCR_STD_COORDS_ALT)
 
 # Invariance assumption:
 DISCR_COORDS_TRANSP         <- latex_transp(DISCR_COORDS)
@@ -302,8 +308,7 @@ DISCR_STD_COORDS_EQ  <- latex_eq(
   DISCR_STD_COORDS_DEF
 )
 
-# Test space and test space basis:
-TEST_SPACE                <- latex_bf("A")
+# Test space basis:
 ITEM_BASIS_VECTOR_EL      <- latex('v')
 ITEM_BASIS_VECTOR         <- latex_bf(ITEM_BASIS_VECTOR_EL)
 ITEM_BASIS_VECTOR_FIRST   <- latex_sub(ITEM_BASIS_VECTOR, 1)

### 2024-03-29 output/paper_draft/paper_M2PLM_parameters.Rmd
Makes preliminary updates to "Test space" section
There are still pending changes to approve, but these are already approved. The remaining ones will be implemented later on, when they are definitively approved.

@@
 in $`r TEST_SPACE`$, such that $`r DISCR_STD_COORDS_ALT_EQ`$.
-<!-- #TODO: Continue here -->
-Since the response probability must be invariant with respect to
-a change of basis,
+Since the response probability to the `r MODEL_ACRONYM`,
+given by \@ref(eq:M2PL-formulation), must be invariant with respect to
+the change of basis expressed in \@ref(eq:change-basis),
 we must define the coordinates $`r DISCR_STD_COORDS`$ of the parameter vector
-$`r DISCR_VECTOR`$ in an orthonormal basis $`r TEST_SPACE_STD_BASIS`$,
+$`r DISCR_VECTOR`$ in the orthonormal basis $`r TEST_SPACE_STD_BASIS`$,
 such that $`r INVARIANCE_EQ`$.
@@
 \begin{equation}
-  `r DISCR_STD_COORDS_EQ`                                (\#eq:discr-vector-std)
+  `r DISCR_STD_COORDS_EQ`,                               (\#eq:discr-vector-std)
 \end{equation}
@@
 
-This shows that discrimination vectors are indeed represented in their own
-vector space $`r TEST_SPACE`$ with basis $`r TEST_SPACE_BASIS`$.
 We shall refer to this new space as the *test space* [@zhang_theoretical_1999].

### 2024-03-29 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects "Polar coordinatess" section
The deleted content was redundant with Proposition 1

@@
 
-From Proposition 1,
-any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
-
-\begin{equation}
-  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
-\end{equation}
-
-Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+Applying Proposition 1 to the model formulation in \@ref(eq:M2PL-formulation),
+the `r MODEL_ACRONYM` model is expressed in polar coordinates as
 

### 2024-04-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Makes initial changes to "Maximum slope" section
Implements approved corrections (some corrections are still pending approval)

@@
 
-In the test space we have that
+We have that
 
@@
 
-Applying Proposition 1,
+Applying Proposition 1, in the test space $`r TEST_SPACE`$, we get
 

### 2024-04-03 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 [@ackerman_creating_1994]
----exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
-the R package *psych*, <!--# Reference needed? -->graphically represents
+---exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
+the R package *psych* [@revelle_psych_2024], graphically represents
 factor loadings on orthogonal axes even after an oblique rotation.
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration<!--# Examples? -->
+is usually implicit in the MIRT literature without further consideration
 [see @zhang_theoretical_1999 for a notable exception].
@@
 In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait,
-usually reverse scored-- is also widespread.<!--# Reference and example? -->
+---items that tap the negative pole of a certain trait, and thus
+usually reverse-scored--- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine

### 2024-04-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects minor erratum

--- 2024-03-29 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 the set of $`r DISCR_VECTOR`$ vectors,
-with the sum and product by a scalar operations as in $`r REAL_N_SPACE`$.
+with the sum and product by a scalar operation as in $`r REAL_N_SPACE`$.
 Consider also an orthonormal basis $`r TEST_SPACE_STD_BASIS`$

### 2024-04-09 output/paper_draft/paper_M2PLM_parameters.Rmd
Changes grammar (minor suggestion by Grammarly)

--- 2024-04-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 a latent trait space, no matter its structure, as an orthogonal space
-[@ackerman_creating_1994]
----exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
-the R package *psych*, <!--# Reference needed? -->graphically represents
+---exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
+the R package *psych* [@revelle_psych_2024], graphically represents
 factor loadings on orthogonal axes even after an oblique rotation.
@@
 The main objective of this paper is thus to provide a generalization of
-the `r MODEL_ACRONYM` multidimensional parameters in two aspects:
-On one hand, in an oblique coordinate system
+the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
 that can be soundly applied to the correlated trait case; on the other, to
 items that yield monotonous response probabilities, but in the broader sense of
-unchanging monotony. Additionally, we aim at providing a method for
+unchanging monotony. Additionally, we aim to provide a method for
 graphically representing these items in the fashion of @ackerman_creating_1994
@@
 propose a graphical representation method,
-and present two practical examples applying them.
+and present two examples of application of our results.
 We round up with a discussion about
@@
 Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
-in basis $`r LS_BASIS_EQ`$, being $`r LS_BASIS`$ not necessarily orthonormal,
-where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
+in basis $`r LS_BASIS_EQ`$.
+A respondent is represented by vector $`r TRAIT_VEC_IN_LS`$,
+which is assumed to be a multivariate normal random variable:
+
+\begin{equation}
+  `r TRAIT_MV_DEF`,
+\end{equation}
+
+where $`r MEAN_VECTOR`$ is an $n$-element vector of means and
+$`r COV_MATRIX` = `r SD_MATRIX` `r CORR_MATRIX` `r SD_MATRIX`$
+an $n$-dimensional positive definite covariance matrix,
+with $`r CORR_MATRIX`$ the corresponding correlation matrix
+(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
+with $`r VAR_COV_EQ`$ (i.e., the variances).
+
 According to the `r MODEL_ACRONYM` model, the probability of
@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$ (e.g. a negatively-keyed item).
+latent dimension $`r DIM_INDEX`$.
 Finally, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
 instead of the item difficulty.
 
-Our aim is to find in the first place
-the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
+Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
+$`r LATENT_SPACE`$.
 Because the slope is dependent on the direction considered,
@@
 in polar coordinates, as in @reckase_difficulty_1985.
+
 However, Reckase's Equation 3 implicitly assumes orthogonality.
@@
 
-### Test space and latent space
-
-Let $`r TEST_SPACE`$ be the vector space made up by
-the set of $`r DISCR_VECTOR`$ vectors,
-with the sum and product by a scalar operation as in $`r REAL_N_SPACE`$.
-Consider also an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
-in $`r TEST_SPACE`$, such that $`r DISCR_STD_COORDS_ALT_EQ`$.
-Since the response probability to the `r MODEL_ACRONYM`,
-given by \@ref(eq:M2PL-formulation), must be invariant with respect to
-the change of basis expressed in \@ref(eq:change-basis),
-we must define the coordinates $`r DISCR_STD_COORDS`$ of the parameter vector
-$`r DISCR_VECTOR`$ in the orthonormal basis $`r TEST_SPACE_STD_BASIS`$,
-such that $`r INVARIANCE_EQ`$.
-If we assume that $`r DISCR_VECTOR`$ exists in vector space $`r TEST_SPACE`$,
-with coordinates $`r DISCR_COORDS`$ in basis $`r TEST_SPACE_BASIS`$,
-then we must define $`r DISCR_COORDS`$ such that
-
-\begin{equation}
-  `r DISCR_STD_COORDS_EQ`,                               (\#eq:discr-vector-std)
-\end{equation}
-
-We want @reckase_difficulty_1985's definition of the multidimensional parameters
-in rectangular coordinates to be a particular case of our more general
-definition, so we can conveniently assume that the inner product in
-$`r TEST_SPACE_STD_BASIS`$ is the dot product,
-as we did with $`r LS_ORTH_BASIS`$.
-The norm of $`r DISCR_VECTOR`$ thus fulfills
-
-\begin{equation}
-  `r DISCR_VECTOR_MODULE_EQ`.
-\end{equation}
-
-Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
-keeps the norm of $`r DISCR_VECTOR`$ invariant.
-
-We shall refer to this new space as the *test space* [@zhang_theoretical_1999].
-Although the inverse relationship between their respective inner product
-Gram matrices induces a bijective relationship between the two sets of spaces,
-it is worth highlighting that test space $`r TEST_SPACE`$
-and latent space $`r LATENT_SPACE`$ are different entities. Only in the
-orthonormal case the two spaces share a common inner product
-(i.e., the dot product), and thus can be mistaken.
-
-### Direction cosines in the latent space
-
 The cosine between two vectors
@@
 
-Applying Proposition 1 to the model formulation in \@ref(eq:M2PL-formulation),
-the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+From Proposition 1,
+any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
+
+\begin{equation}
+  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
+\end{equation}
+
+Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
 
@@
 Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
-already assumes the orthonormality of the latent space, as we stated before.
+already assumes the orthonormality of the latent space.
 
@@
 
-The slope in direction $`r DIR_ANGLE_VEC_ORG`$ is given by
+The slope in direction $`r DIR_ANGLE_VEC`$ is given by
 
@@
   \begin{split}
-    `r DIR_COS_AS_DIR_COS_ORTH_EQ`, \\
+    `r DIR_COS_AS_DIR_COS_ORTH_EQ` \\
     `r TRAIT_POLAR_COEFF_COS_ORTH_EQ`.
@@
 
-As $`r DISCR_VECTOR`$ is defined in test space $`r TEST_SPACE`$ with basis
-$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
-coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
-\@ref(eq:discr-vector-std),
+We can then define
+
+\begin{equation}
+  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
+\end{equation}
+
+so
 
@@
 (Equation \@ref(eq:direction-cosine-vectors-eq)) and
-$`r DISCR_STD_COORDS`$ (Equation \@ref(eq:discr-vector-std))
+$`r DISCR_VECTOR_STD`$ (Equation \@ref(eq:discr-vector-std))
 by their respective expressions results in the
@@
 
-We have that
+# Results
+
+## Test space and latent space
+
+Before defining the multidimensional parameters, we need to make a clarification
+regarding the representation of the items in the latent space.
+We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of
+the discrimination vectors.
+This space is where the item vectors can be properly represented.
+It is in general different from the latent space $`r LATENT_SPACE`$,
+although there is a bijective relationship between the two sets of spaces.
+Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
+where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
+in the same basis $`r LS_BASIS`$ in their respective spaces.
+
+Since the response probability must be invariant, because of Equation
+\@ref(eq:change-basis), parameter $`r DISCR_VECTOR`$ must also be transformed
+in $`r DISCR_VECTOR_STD`$.
+Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
+This shows that first a basis change to the standard base is performed and then
+a transformation through $`r INNER_PROD_MATRIX_INV`$,
+so the $`r DISCR_VECTOR_STD`$ are expressed in the standard basis.
+
+\begin{equation}
+  `r DISCR_VECTOR_MODULE_EQ`                     (\#eq:discr-param-squared-norm)
+\end{equation}
+
+Therefore, in the test space, an inner product matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined such that $`r INNER_PROD_DISCR_EQ`$,
+to keep the norm invariant.
+With this definition we have that
 
@@
 
-Applying Proposition 1, in the test space $`r TEST_SPACE`$, we get
+Applying Proposition 1,
 
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
-
-# Results
 
@@
 The (signed) distance to the origin,
-given by the first component of $`r MIL_PARAM`$, has an interesting property:
+given by the second element of $`r MIL_PARAM`$, has an interesting property:
 By the definition of the inner product,

### 2024-04-09 output/paper_draft/paper_M2PLM_parameters.Rmd
Implements changes approved in Google Docs

--- 2024-04-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-Our aim is to find in the first place
+Following @reckase_difficulty_1985’s procedure,
+we must find in the first place
 the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
@@
 with the sum and product by a scalar operation as in $`r REAL_N_SPACE`$.
-Consider also an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
-in $`r TEST_SPACE`$, such that $`r DISCR_STD_COORDS_ALT_EQ`$.
+We want @reckase_difficulty_1985's definition of
+the multidimensional parameters in rectangular coordinates to be
+a particular case of our more general definition;
+we can thus consider an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
+in $`r TEST_SPACE`$ such that $`r DISCR_STD_COORDS_ALT_EQ`$
+corresponds in the model to $`r TRAIT_ORTH_COORDS`$.
 Since the response probability to the `r MODEL_ACRONYM`,
@@
 the change of basis expressed in \@ref(eq:change-basis),
-we must define the coordinates $`r DISCR_STD_COORDS`$ of the parameter vector
-$`r DISCR_VECTOR`$ in the orthonormal basis $`r TEST_SPACE_STD_BASIS`$,
-such that $`r INVARIANCE_EQ`$.
-If we assume that $`r DISCR_VECTOR`$ exists in vector space $`r TEST_SPACE`$,
-with coordinates $`r DISCR_COORDS`$ in basis $`r TEST_SPACE_BASIS`$,
-then we must define $`r DISCR_COORDS`$ such that
+Therefore, we find a basis $`r TEST_SPACE_BASIS`$ in $`r TEST_SPACE`$
+defined by
 
@@
 
-We want @reckase_difficulty_1985's definition of the multidimensional parameters
-in rectangular coordinates to be a particular case of our more general
-definition, so we can conveniently assume that the inner product in
+with $`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
+Note that, as $`r TEST_SPACE_STD_BASIS`$ is orthonormal,
+$`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
+
+To make @reckase_difficulty_1985's definition a particular case of ours
+we can conveniently assume that the inner product in
 $`r TEST_SPACE_STD_BASIS`$ is the dot product,
@@
 
-We shall refer to this new space as the *test space* [@zhang_theoretical_1999].
+We shall refer to this new space $`r TEST_SPACE`$ as
+the *test space* [@zhang_theoretical_1999].
 Although the inverse relationship between their respective inner product

### 2024-04-09 R/Formulae.R

@@ -285,7 +285,7 @@ REAL_N_SPACE <- latex_raised_to("\\mathbb{R}", exp = N_DIMS)
 # Discrimination vector coordinates:
 TEST_SPACE_BASIS        <- latex_cal("B^*")
 DISCR_COORDS            <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_BASIS)
-TEST_SPACE_STD_BASIS    <- latex_cal("E^*")
+TEST_SPACE_STD_BASIS    <- latex_cal("U^*")
 DISCR_STD_COORDS    <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_STD_BASIS)
 DISCR_STD_COORDS_ALT <- latex_coords(DISCR_VECTOR, basis = TEST_SPACE_STD_BASIS)
 DISCR_STD_COORDS_ALT_EQ <- latex_eq(DISCR_STD_COORDS, DISCR_STD_COORDS_ALT)

### 2024-04-09 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds (some) pending references and sample item

--- 2024-04-09 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
+cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
+[e.g., @thielmann_comparability_2022].
 Moreover, certain applications impose the estimation of
@@
 ---items that tap the negative pole of a certain trait, and thus
-usually reverse-scored--- is also widespread.<!--# Reference and example? -->
+usually reverse-scored--- is also widespread
+[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
+of the Big-Five Factor Markers, @noauthor_international_nodate].
 It is common to find instruments that combine

### 2024-04-29 output/paper_draft/paper_M2PLM_parameters.Rmd
Implements changes approved in Google Docs
Implements changes approved in MLV's review on 2024-04-28 in section "Test space and latent space"

--- 2024-04-09 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 the set of $`r DISCR_VECTOR`$ vectors,
-with the sum and product by a scalar operation as in $`r REAL_N_SPACE`$.
+with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
 We want @reckase_difficulty_1985's definition of
@@
 Since the response probability to the `r MODEL_ACRONYM`,
-given by \@ref(eq:M2PL-formulation), must be invariant with respect to
-the change of basis expressed in \@ref(eq:change-basis),
-Therefore, we find a basis $`r TEST_SPACE_BASIS`$ in $`r TEST_SPACE`$
+given by Equation \@ref(eq:M2PL-formulation), must be invariant with respect to
+the change of basis expressed in Equation \@ref(eq:change-basis),
+$`r INVARIANCE_EQ`$.
+Therefore, we find a basis $`r TEST_SPACE_BASIS_EQ`$ in $`r TEST_SPACE`$
 defined by
@@
 with $`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
-Note that, as $`r TEST_SPACE_STD_BASIS`$ is orthonormal,
-$`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
+Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
 

### 2024-05-10 apps/Graphical_representation_interactive.qmd
Versison shiny app inputs
Creates the inputs and their rendering logic

---
title: "MCLM parameters, interactive example"
author: "Daniel Morillo"
format: html
server: shiny
---

```{r setup}
#| warning: false
library(shinyjs)
library(tibble)
library(mnormt)
library(dplyr)
library(stringr)
```

# Description

This app show an interactive graphical representation of the
Multidimensional parameters (location and discrimination) of the Multidimensional
Compensatory Logistic IRT Model (MCLM).

# Input

## MCLM item parameters

```{r item-params-input}
#| layout: [[1], [1], [1]]
numericInput('a_1', label = 'a1', value = 1, step = .1, min = -3, max = 3)
numericInput('a_2', label = 'a2', value = 1, step = .1, min = -3, max = 3)
numericInput('d',   label = 'd',  value = 0, step = .1, min = -3, max = 3)
```

## Latent space structure

### Covariance matrix

```{r cov-input}
#| layout: [[1, 1], [1, 1]]
numericInput('var_1', label = 'var1', value = 1, step = .1, min = .1, max = 3)
textOutput('cov1')
textOutput('cov2')
numericInput('var_2', label = 'var2', value = 1, step = .1, min = .1, max = 3)
```

### Correlation matrix

```{r corr-input}
#| layout: [[1, 1], [1, 1]]
h5("1")
numericInput(
  'corr',
  label = 'corr',
  value = 0,
  step  =  .05,
  min   = -.95,
  max   =  .95
)
textOutput('corr_lower')
h5("1")
```

```{r compute-ls}
#| context: server

# Logic for computing & rendering the latent space structure:

## Values and matrices:
covariance  <- reactiveVal()
cov_matrix  <- reactiveVal()
corr_matrix <- reactiveVal()

## Compute covariance:
observe(
  covariance(sqrt(input$var_1 * input$var_2) * input$corr)
)

## Compute covariance matrix:
observe(
  cov_matrix(
    matrix(c(input$var_1, covariance(), covariance(), input$var_2), nrow = 2)
  )
)

## Compute correlation matrix:
observe(
  corr_matrix(matrix(c(1, input$corr, input$corr, 1), nrow = 2))
)

## Output correlation and covariance values:
output$corr_lower <- renderText(input$corr)
output$cov1       <- renderText(covariance() |> round(2))
output$cov2       <- renderText(covariance() |> round(2))
```

## Sample

```{r sample-choice}
radioButtons(
  'sample_type',
  label   = "Plot:",
  choices = c("unique value", "sample"),
  inline  = TRUE
)
```

```{r single-value}
#| panel: sidebar
numericInput(
  'theta_1',
  label = 'theta_1',
  value =  1,
  step  =   .1,
  min   = -3,
  max   =  3
)
numericInput(
  'theta_2',
  label = 'theta_2',
  value =  1,
  step  =   .1,
  min   = -3,
  max   =  3
)
```

```{r sample-size}
#| panel: sidebar
numericInput('N', label = 'N', value = 10, min = 10, max = 1000)
```

```{r compute-sample}
#| context: server

# Logic for computing & rendering the sample values:

# Sample values:
single_lt       <- reactiveVal() # Single latent trait vector
complete_sample <- reactiveVal() # A N = 1000 sample is pre-computed to save
                                 #   time (it only changes if the covariance
                                 #   matrix changes)
n_subsample     <- reactiveVal() # Subsets the first N cases

# Compute single vector:
observe(
  single_lt(tibble(theta_1 = input$theta_1, theta_2 = input$theta_2))
)

# Compute complete sample:
observeEvent(
  cov_matrix(),
  complete_sample(
    rmnorm(n = 1000, varcov = cov_matrix()) |>
    as_tibble(.name_repair = "unique") |>
    rename_with(str_replace, pattern = "\\.{3}", replacement = "theta_")
  )
)

# Compute N subset sample:
observeEvent(
  list(input$N, complete_sample()),
  n_subsample(complete_sample() |> slice_head(n = input$N))
)

useShinyjs()

observeEvent(
  input$sample_type == "unique value",
  disable('N')
)
observeEvent(
  input$sample_type == "sample",
  {
    disable('theta_1')
    disable('theta_2')
  }
)
```

### 2024-05-10 apps/Graphical_representation_interactive.qmd
Outputs item multidimensional parameters table

@@
 
-This app show an interactive graphical representation of the
-Multidimensional parameters (location and discrimination) of the Multidimensional
-Compensatory Logistic IRT Model (MCLM).
+This app show an interactive graphical representation of the Multidimensional
+parameters (location and discrimination) of the Multidimensional Compensatory
+Logistic IRT Model (MCLM).
 
@@
 numericInput('d',   label = 'd',  value = 0, step = .1, min = -3, max = 3)
+```
+
+```{r compute-item}
+#| context: server
+
+# Logic for computing the item object:
+item_M2PL <- reactiveVal()
+
+observeEvent(
+  list(input$a_1, input$a_2, input$d),
+  item_M2PL(tibble(item = 1, a_1 = input$a_1, a_2 = input$a_2, d = input$d))
+)
 ```
@@
 ```
+
+# Output
+
+# Item multidimensional parameters
+
+```{r item-output}
+tableOutput('item')
+```
+
+```{r compute-multidim-params}
+#| context: server
+#| warning: false
+
+# Multidimensional item parameters logic:
+
+source("../R/Mirt_toolbox.R", encoding = 'UTF-8')
+source("../R/Constants.R",    encoding = 'UTF-8')
+
+item_oblique <- reactiveVal()
+
+observeEvent(
+  list(input$a_1, input$a_2, input$d, cov_matrix()),
+  item_oblique({
+    oblique_params <- item_M2PL() |> compute_mirt_params(
+      all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
+      cov_matrix = corr_matrix(),                           # Input correlations
+      dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
+    )
+    oblique_coords <- oblique_params |> compute_mirt_coords(
+      D, MDISC, starts_with(COSINE_DIRTYPE),
+      transform       = cov_matrix(),
+      original_coords = TRUE
+    )
+
+    full_join(oblique_params, oblique_coords, by = 'item')
+  })
+)
+
+output$item <- renderTable(
+  item_oblique() |> select(MDISC, D, starts_with('deg'))
+)
+```

### 2024-05-10 apps/Graphical_representation_interactive.qmd
Includes latent space "original" plot
This first plot plots the latent trait sample in the "original basis", meaning that this algebraic basis is "mapped" to the geometric standard basis, so the latent trait values are represented in "rectangular Cartesian coordinates"

@@
 author: "Daniel Morillo"
-format: html
+format:
+  html:
+    fig-asp: 1
 server: shiny
@@
 library(stringr)
+library(ggplot2)
+library(scales)
+```
+
+```{r constants}
+#| context: server
+# Graphical representation parameters:
+GRAPH_FONT   <- "serif"
+LINE_WIDTH   <- .1
+VECTOR_WIDTH <- .5
+PALETTE      <- c("darkred", "darkgoldenrod3", "green3", "cyan3", "blue3")
+```
+
+```{r graphical-output-conf}
+#| context: server
+theme_set( # `ggplot` output configuration
+  theme_classic(
+    base_size      = 12,
+    base_family    = GRAPH_FONT,
+    base_line_size = LINE_WIDTH
+  ) %+replace%
+    theme(
+      axis.title.x = element_text(hjust = 1),
+      axis.title.y = element_text(vjust = 1, angle = 0)
+    )
+)
 ```
@@
 ```
+
+## Plots
+
+```{r plot-outputs}
+#| layout: [[1, 1], [1, 1]]
+plotOutput('ls_original')
+```
+
+```{r render-plots}
+#| context: server
+output$ls_original <- renderPlot({
+
+  n_subsample() |> ggplot(aes(x = theta_1, y = theta_2)) +
+    geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
+    geom_hline(yintercept = 0, linewidth = LINE_WIDTH) +
+    geom_point() +
+    scale_x_continuous(
+      limits       = c(-3, 3),
+      breaks       = -2:2,
+      minor_breaks = NULL,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    scale_y_continuous(
+      limits       = c(-3, 3),
+      breaks       = -3:3,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    coord_fixed(expand = FALSE, clip = "on") +
+    theme(
+      axis.line        = element_blank(),
+      panel.grid.major = element_line(
+        color     = "black",
+        linewidth = LINE_WIDTH,
+        linetype  = "17"
+      )
+    )
+})
+```

### 2024-05-10 apps/Graphical_representation_interactive.qmd
Creates plot structure

@@
 plotOutput('ls_original')
+plotOutput('ts_original')
+plotOutput('ls_transformed')
+plotOutput('ts_transformed')
 ```
@@
 })
+output$ts_original <- renderPlot({
+})
+output$ls_transformed <- renderPlot({
+})
+output$ts_transformed <- renderPlot({
+})
 ```

### 2024-05-14 apps/Graphical_representation_interactive.qmd
Computes transformed latent trait coordinates

@@
 library(dplyr)
+library(tidyr)
 library(stringr)
@@
 ## Values and matrices:
-covariance  <- reactiveVal()
-cov_matrix  <- reactiveVal()
-corr_matrix <- reactiveVal()
+covariance        <- reactiveVal() # Covariance value
+cov_matrix        <- reactiveVal() # Covariance matrix
+corr_matrix       <- reactiveVal() # Correlation matrix
+transf_inv_transp <- reactiveVal() # Transpose inverse of the latent space
+                                   #   transform matrix (== Square root of
+                                   #   the correlation matrix scaled by the
+                                   #   standard deviations).
+transf_matrix     <- reactiveVal() # Latent space transform matrix
 
 ## Compute covariance:
-observe(
+observeEvent(
+  list(input$var_1, input$var_2, input$corr),
   covariance(sqrt(input$var_1 * input$var_2) * input$corr)
@@
 ## Compute covariance matrix:
-observe(
+observeEvent(
+  list(input$var_1, input$var_2, covariance()),
   cov_matrix(
@@
 ## Compute correlation matrix:
-observe(
+observeEvent(
+  input$corr,
   corr_matrix(matrix(c(1, input$corr, input$corr, 1), nrow = 2))
+)
+
+## Compute (inverse transposed) transform matrix:
+observeEvent(
+  cov_matrix(),
+  transf_inv_transp({
+
+    # Square root of the correlation matrix (conveniently fixing the X axis):
+    corr_sqrt_matrix <- matrix(
+      c(1, 0, input$corr, sqrt(1 - input$corr^2)),
+      nrow = 2
+    )
+
+    # Scale matrix (i.e. standard deviations):
+    scale_matrix <- cov_matrix() |> diag() |> diag() |> sqrt()
+
+    corr_sqrt_matrix %*% scale_matrix
+  })
+)
+# Compute actual (P) transform matrix (such that $P^TP = M = \Sigma^(-1)$):
+observeEvent(
+  transf_inv_transp(),
+  transf_matrix(transf_inv_transp() |> t() |> solve())
 )
@@
                                   #   time (it only changes if the covariance
-                                 #   matrix changes)
+                                  #   matrix changes).
 n_subsample      <- reactiveVal() # Subsets the first N cases
+lt_sample        <- reactiveVal() # Sample to plot
+lt_sample_transf <- reactiveVal() # Transformed coordinates of sample to plot
 
@@
 
+# Select sample to plot:
 observeEvent(
-  input$sample_type == "unique value",
-  disable('N')
+  list(input$sample_type, single_lt(), n_subsample()),
+  lt_sample(if (input$sample_type == "sample") n_subsample() else single_lt())
 )
+
+# Transform sample coordinates to orthonormal space:
 observeEvent(
-  input$sample_type == "sample",
-  {
-    disable('theta_1')
-    disable('theta_2')
-  }
+  list(transf_inv_transp(), lt_sample()),
+  lt_sample_transf({
+
+    # Compute actual (P) transform matrix (such that $P^TP = M = \Sigma^(-1)$):
+    transf_matrix <- transf_inv_transp() |> t() |> solve()
+
+    # Compute coordinates in transformed (i.e. orthonormal) space:
+    lt_sample() |>
+      rowid_to_column() |>
+      pivot_longer(-rowid) |>
+      group_by(rowid) |>
+      mutate(value = transf_matrix %*% value |> drop()) |>
+      pivot_wider()
+  })
 )

### 2024-05-14 apps/Graphical_representation_interactive.qmd
Corrects computation of MIRT parameters
The multidimensional parameters must be computed with the covariance matrix instead of the correlation matrix

@@
       all_of(INTERCEPT_COLKEY), starts_with(DISCR_PREFFIX), # Input parameters
-      cov_matrix = corr_matrix(),                           # Input correlations
-      dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output configuration
+      cov_matrix = cov_matrix(),                            # Input covariance
+      dir_out    = c(COSINE_DIRTYPE, DEGREE_DIRTYPE)        # Output config.
     )

### 2024-05-14 apps/Graphical_representation_interactive.qmd
Updates latent trait plot in original basis

@@
 #| context: server
-output$ls_original <- renderPlot({
 
-  n_subsample() |> ggplot(aes(x = theta_1, y = theta_2)) +
+# Render scatter plot in original coordinates:
+output$ls_original <- renderPlot(
+
+  lt_sample() |> ggplot(aes(x = theta_1, y = theta_2)) +
     geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
     geom_hline(yintercept = 0, linewidth = LINE_WIDTH) +
-    geom_point() +
+    geom_point(alpha = .5) +
     scale_x_continuous(
       limits       = c(-3, 3),
-      breaks       = -2:2,
+      breaks       = -3:3,
       minor_breaks = NULL,
@@
         linewidth = LINE_WIDTH,
-        linetype  = "17"
+        linetype  = "C6"
       )
+    ) +
+    xlab("theta_1") + ## FIXME: Labels not shown
+    ylab("theta_2")
 )
-})
+
 output$ts_original <- renderPlot({

### 2024-05-14 apps/Graphical_representation_interactive.qmd
Adds latent trait plot in transformed basis

@@
 
+# Render scatter plot in transformed coordinates:
 output$ts_original <- renderPlot({
@@
 output$ls_transformed <- renderPlot({
+
+  # Constant objects:
+  BREAKS <- -3:3
+
+  # Compute grid lines:
+
+  ## Basis vectors:
+  basis_vec_1 <- transf_matrix()[, 1]
+  basis_vec_2 <- transf_matrix()[, 2]
+  basis_vec_1_norm <- basis_vec_1^2 |> sum() |> sqrt()
+  basis_vec_2_norm <- basis_vec_2^2 |> sum() |> sqrt()
+
+  ## Oblique grid lines:
+  tan_vec_1 <- basis_vec_1[2] / basis_vec_1[1] # Angle tangent
+
+  lt_sample_transf() |> ggplot(aes(x = theta_1, y = theta_2)) +
+    geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
+    geom_abline(intercept = 0, slope = tan_vec_1, linewidth = LINE_WIDTH) +
+    geom_point(alpha = .5) +
+    scale_x_continuous(
+      limits       = c(-3, 3),
+      breaks       = -3:3,
+      minor_breaks = NULL,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    scale_y_continuous(
+      limits       = c(-3, 3),
+      breaks       = -3:3,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    coord_fixed(expand = FALSE, clip = "on") +
+    theme(
+      axis.line          = element_blank(),
+      panel.grid.major.x = element_line(
+        color     = "black",
+        linewidth = LINE_WIDTH,
+        linetype  = "C6"
+     ),
+     axis.line.y = 
+    )
 })

### 2024-05-14 apps/Graphical_representation_interactive.qmd
Adds spike drawing logic (for single latent trait)
Adds input disabling logic (WiP)

@@
 
-useShinyjs()
-
 # Select sample to plot:
@@
 )
+
+## FIXME: Disabling inputs does not work
+# useShinyjs()
+# observeEvent(
+#   input$sample_type == "unique value",
+#   disable('N')
+# )
+# observeEvent(
+#   input$sample_type == "sample",
+#   {
+#     disable('theta_1')
+#     disable('theta_2')
+#   }
+# )
 ```
@@
 #| context: server
+
+draw_spikes <- reactiveVal() # Flag to indicate whether to draw spikes (for
+                             #   single value plotting only)
+observeEvent( ## TODO: Draw spikes
+  input$sample_type,
+  draw_spikes(input$sample_type == "unique value")
+)
 

### 2024-05-17 apps/Graphical_representation_interactive.qmd
Makes "include paths" relative to project path
Adds oblique grid lines in latent space plot

@@
 
-source("../R/Mirt_toolbox.R", encoding = 'UTF-8')
-source("../R/Constants.R",    encoding = 'UTF-8')
+source("R/Mirt_toolbox.R", encoding = 'UTF-8')
+source("R/Constants.R",    encoding = 'UTF-8')
 
@@
     geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
-    geom_abline(intercept = 0, slope = tan_vec_1, linewidth = LINE_WIDTH) +
+    geom_abline( # Oblique `theta_2` axis
+      intercept = 0,
+      slope     = tan_vec_1,
+      linewidth = LINE_WIDTH
+    ) +
+    geom_abline( # Oblique panel grid lines
+      intercept = -3:3 * basis_vec_2[2],
+      slope     = tan_vec_1,
+      linewidth = LINE_WIDTH,
+      linetype  = "C6"
+    ) +
     geom_point(alpha = .5) +
@@
         linetype  = "C6"
-     ),
-     axis.line.y = 
+     )
     )

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Adds y-axis breaks in transformed latent space

@@
       limits       = c(-3, 3),
-      breaks       = -3:3,
+      breaks       = -3:3 * basis_vec_2[2] -3 * basis_vec_1[2],
+      labels       = -3:3,
       name         = NULL,

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Drops redundant computation of `transf_matrix`

@@
 observeEvent(
-  list(transf_inv_transp(), lt_sample()),
+  list(transf_matrix(), lt_sample()),
   lt_sample_transf({
-
-    # Compute actual (P) transform matrix (such that $P^TP = M = \Sigma^(-1)$):
-    transf_matrix <- transf_inv_transp() |> t() |> solve()
 
@@
       group_by(rowid) |>
-      mutate(value = transf_matrix %*% value |> drop()) |>
+      mutate(value = transf_matrix() %*% value |> drop()) |>
       pivot_wider()

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Corrects computation of oblique item coordinates
The correct transformation matrix to use is the inverse transpose of the latent trait transformation matrix

@@
       D, MDISC, starts_with(COSINE_DIRTYPE),
-      transform       = cov_matrix(),
+      transform       = transf_inv_transp(),
       original_coords = TRUE

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Parameterizes breaks & limits in latent trait plots

@@
 )
+# Constant objects:
+LIM_INF <- -3L
+LIM_SUP <-  3L
+LIMITS  <- c(LIM_INF, LIM_SUP)
+BREAKS  <- LIM_INF:LIM_SUP
 
@@
     scale_x_continuous(
-      limits       = c(-3, 3),
-      breaks       = -3:3,
+      limits       = LIMITS,
+      breaks       = BREAKS,
       minor_breaks = NULL,
@@
     scale_y_continuous(
-      limits       = c(-3, 3),
-      breaks       = -3:3,
+      limits       = LIMITS,
+      breaks       = BREAKS,
       name         = NULL,
@@
 
-  # Constant objects:
-  BREAKS <- -3:3
-
   # Compute grid lines:
@@
     geom_abline( # Oblique panel grid lines
-      intercept = -3:3 * basis_vec_2[2],
+      intercept = BREAKS * basis_vec_2[2],
       slope     = tan_vec_1,
@@
     scale_y_continuous(
-      limits       = c(-3, 3),
-      breaks       = -3:3 * basis_vec_2[2] -3 * basis_vec_1[2],
-      labels       = -3:3,
+      limits       = LIMITS,
+      breaks       = BREAKS * basis_vec_2[2] + LIM_INF * basis_vec_1[2],
+      labels       = BREAKS,
       name         = NULL,

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Cleans up code of the plot renders

@@
 
-draw_spikes <- reactiveVal() # Flag to indicate whether to draw spikes (for
-                             #   single value plotting only)
-observeEvent( ## TODO: Draw spikes
-  input$sample_type,
-  draw_spikes(input$sample_type == "unique value")
-)
 # Constant objects:
@@
 
-# Render scatter plot in transformed coordinates:
 output$ts_original <- renderPlot({
 })
+
+# Render scatter plot in transformed coordinates:
 output$ls_transformed <- renderPlot({
@@
   basis_vec_2 <- transf_matrix()[, 2]
-  basis_vec_1_norm <- basis_vec_1^2 |> sum() |> sqrt()
-  basis_vec_2_norm <- basis_vec_2^2 |> sum() |> sqrt()
 
@@
 
+  # Render plot:
   lt_sample_transf() |> ggplot(aes(x = theta_1, y = theta_2)) +
     geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
-    geom_abline( # Oblique `theta_2` axis
+    geom_abline( # Oblique `theta_1` axis
       intercept = 0,

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Renders item in original coordinates

@@
 
+# Render item in original coordinates:
 output$ts_original <- renderPlot({
+
+  item_oblique() |>
+    ggplot(
+      aes(
+        origin_1, origin_2,
+        xend  = end_1, yend = end_2,
+        color = I("red"), fill = I("red")
+      ),
+    )                                                  +
+    geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
+    geom_hline(yintercept = 0, linewidth = LINE_WIDTH) +
+    geom_segment(
+      arrow     = arrow(
+        angle  = 20,
+        length = unit(10, "points"),
+        type   = "closed"
+      ),
+      linejoin  = "mitre",
+      linewidth = VECTOR_WIDTH
+    ) +
+    scale_x_continuous(
+      limits       = LIMITS,
+      breaks       = BREAKS,
+      minor_breaks = NULL,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    scale_y_continuous(
+      limits       = LIMITS,
+      breaks       = BREAKS,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    coord_fixed(expand = FALSE, clip = "on") +
+    theme(
+      axis.line        = element_blank(),
+      panel.grid.major = element_line(
+        color     = "black",
+        linewidth = LINE_WIDTH,
+        linetype  = "C6"
+      )
+    )
 })

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Corrects computation of item coordinates
It must be based on the latent space transform matrix, not the test space one

@@
 observeEvent(
-  list(input$a_1, input$a_2, input$d, cov_matrix()),
+  list(item_M2PL(), cov_matrix()),
   item_oblique({
@@
       D, MDISC, starts_with(COSINE_DIRTYPE),
-      transform       = transf_inv_transp(),
+      transform       = transf_matrix(),
       original_coords = TRUE

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Corrects reactive dependence of item coordinates
The item coordinates computation depends on the "transformation matrix" event, not the "covariance matrix" one

@@
 observeEvent(
-  list(item_M2PL(), cov_matrix()),
+  list(item_M2PL(), transf_matrix()),
   item_oblique({

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Test space rendering in transformed coordinates

@@
 })
+
+# Render item in transformed coordinates:
 output$ts_transformed <- renderPlot({
+
+  # Compute grid lines:
+
+  ## Basis vectors:
+  basis_vec_1 <- transf_inv_transp()[, 1]
+  basis_vec_2 <- transf_inv_transp()[, 2]
+
+  ## Oblique grid lines:
+  tan_vec_2 <- basis_vec_2[2] / basis_vec_2[1] # Angle tangent
+
+  # Render plot:
+  output <- item_oblique() |>
+    ggplot(
+      aes(
+        origin_transf_1, origin_transf_2,
+        xend  = end_transf_1, yend = end_transf_2,
+        color = I("red"), fill = I("red")
+      ),
+    ) +
+    geom_hline(yintercept = 0L, linewidth = LINE_WIDTH)
+
+  output <- if (basis_vec_2[1]) {
+
+    output +
+      geom_abline( # Oblique `theta_2` axis
+        intercept = 0L,
+        slope     = tan_vec_2,
+        linewidth = LINE_WIDTH
+      ) +
+      geom_abline( # Oblique panel grid lines
+        intercept = BREAKS * tan_vec_2,
+        slope     = tan_vec_2,
+        linewidth = LINE_WIDTH,
+        linetype  = "C6"
+      )
+
+  } else {
+
+    output + 
+      geom_vline(xintercept = 0L,     linewidth = LINE_WIDTH) +
+      geom_vline(xintercept = BREAKS, linewidth = LINE_WIDTH, linetype  = "C6")
+  }
+
+  output +
+    geom_segment(
+      arrow     = arrow(
+        angle  = 20,
+        length = unit(10, "points"),
+        type   = "closed"
+      ),
+      linejoin  = "mitre",
+      linewidth = VECTOR_WIDTH
+    ) +
+    scale_x_continuous(
+      limits       = LIMITS,
+      breaks       = BREAKS * basis_vec_1[1] + LIM_INF * basis_vec_2[1],
+      labels       = BREAKS,
+      minor_breaks = NULL,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    scale_y_continuous(
+      limits       = LIMITS,
+      breaks       = BREAKS * basis_vec_2[2],
+      labels       = BREAKS,
+      name         = NULL,
+      oob          = oob_keep
+    ) +
+    coord_fixed(expand = FALSE, clip = "on") +
+    theme(
+      axis.line          = element_blank(),
+      panel.grid.major.y = element_line(
+        color     = "black",
+        linewidth = LINE_WIDTH,
+        linetype  = "C6"
+      )
+    )
 })

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Drops non-working input disabling logic

@@
 #| warning: false
-library(shinyjs)
 library(tibble)
@@
 )
-
-## FIXME: Disabling inputs does not work
-# useShinyjs()
-# observeEvent(
-#   input$sample_type == "unique value",
-#   disable('N')
-# )
-# observeEvent(
-#   input$sample_type == "sample",
-#   {
-#     disable('theta_1')
-#     disable('theta_2')
-#   }
-# )
 ```

### 2024-05-20 apps/Graphical_representation_interactive.qmd
Adds plot titles

@@
     theme(
+      plot.title   = element_text(hjust = .5),
       axis.title.x = element_text(hjust = 1),
@@
     ) +
-    xlab("theta_1") + ## FIXME: Labels not shown
-    ylab("theta_2")
+    ggtitle("Original coordinates")
 )
@@
       )
-    )
+    ) +
+    ggtitle("Original coordinates")
 })
@@
      )
-    )
+    ) +
+    ggtitle("Transformed coordinates")
 })
@@
       )
-    )
+    ) +
+    ggtitle("Transformed coordinates")
 })

### 2024-05-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Clarifies interpretation of null discrimination

--- 2024-04-09 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 dimension is given by a non-positive discrimination parameter
-(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
+(and an unchanging probability by a null discrimination parameter,
+which also applies to cognitive instruments),
 contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].

### 2024-05-28 R/Formulae.R
Corrects deduction of item direction angles
Corrects deduction of item direction angles in test space

@@ -840,14 +840,14 @@ DISCR_STD_COORDS_MOD_EQ <- latex_eq(
 
 # Item direction cosines (in test space):
 ANGLE_TS                          <- latex("\\alpha")
+ANGLE_VEC_TS                      <- latex_bf(ANGLE_TS)
+ANGLE_VEC_TS_ITEM                 <- latex_sub(ANGLE_VEC_TS, ITEM_INDEX)
 ANGLE_TS_ITEM                     <- latex_sub(
   ANGLE_TS,
   "$ITEM_INDEX$$DIM_INDEX$"
 )
-DIR_ANGLE_VEC_TS                  <- latex_bf(ANGLE_TS)
-DIR_COS_VEC_TS                    <- latex_cos(DIR_ANGLE_VEC_TS)
-DIR_ANGLE_ITEM_VEC_TS             <- latex_sub(DIR_ANGLE_VEC_TS, ITEM_INDEX)
-DIR_COS_ITEM_VEC_TS               <- latex_cos(DIR_ANGLE_ITEM_VEC_TS)
+DIR_COS_VEC_TS                    <- latex_cos(ANGLE_VEC_TS)
+DIR_COS_ITEM_VEC_TS               <- latex_cos(ANGLE_VEC_TS_ITEM)
 DIAG_MATRIX_INNER_PROD_INV        <- latex_prime(DIAG_MATRIX_INNER_PROD)
 DIAG_MATRIX_INNER_PROD_INV_SQ_INV <- latex_raised_to(
   DIAG_MATRIX_INNER_PROD_INV,

### 2024-05-28 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects deduction of item direction angles
Corrects deduction of item direction angles in test space
Drops reference to covariance structure
The covariance structure of theta plays no role up to this point, so it must be obviated for now.

--- 2024-04-29 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-Applying Proposition 1, in the test space $`r TEST_SPACE`$, we get
+Applying Proposition 2, if `r ANGLE_VEC_TS_ITEM` is
+the vector of direction angles of `r DISCR_VECTOR`,
+in the test space $`r TEST_SPACE`$ we get
 
@@
 and @reckase_discriminating_1991.
-These expresssions disregard the covariance structure of $`r TRAIT_VECTOR`$ by
-implicitly assuming the orthonormality of $`r LS_BASIS`$.
-Therefore, we shall refer to these as the *agnostic* version of the indices:
+As these expressions implicitly assume the orthonormality of $`r LS_BASIS`$,
+which bears no meaning besides its pure algebraic purpose,
+we shall refer to them as the *agnostic* version of the indices:
 

### 2024-05-29 notebooks/Representacion.Rmd
Versions new Mario's notebook about representation

---
title: "Representacion"
author: "Mario Luzardo"
date: "2024-05-19"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls()) 
library(pspline)
library(mvtnorm)
library(sn)
library(mnormt)
library(scatterplot3d)
library(copula)
library(ggplot2)
library(fungible)
library(pracma)
library(matlib)
library(rgl)
library(lsa)
library(MASS)
library(mirt)
library(tidyverse)
set.seed(123)
```

Esto intenta generalizar los conceptos considerando que partimos de una base de habilidades que consideramos la que explica realmente el fenómeno.

Me baso en que en el desarrolo de Rackase no se utiliza la normalidad para obtener las f'ormulas simplemente la ortogonalidad del espacio latente.

Me basaré en la relación entre los modelos de TRI y el AF, por ello usare el paquete mirt de R

El método consistira en:

1) Considero un espacio latente oblicuo y alli una base que representa el fenómeno realmente.
2) Simulo las coordenadas de las habilidades en esa base para que su correlación sea el angulo entre los factores oblicuos. 
3) Analizo la opción de Daniel y la justifico
3) Realizo el procedimiento que llamos para calcular los parámetros.
4) Veo como queda el espacio test
5) Genero método para calcular P a partir de solucion sin rotar y rotada

Entonces veremos métodos para
a) Si conozco P
b) Si solo se el ángulo entre las habilidades (usamos solucion de Daniel)
c) Si tenemos solucion sin rotar y rotada como hallar P

Supongamos que la base que representa las habilidades verdaderas es:
```{r base no ortogonal, include=FALSE}
b1_v1=c(3,1)
b1_v2=c(1,4)
b1_v1=b1_v1/len(b1_v1)
b1_v2=b1_v2/len(b1_v2)
Base_1=matrix(cbind(b1_v1,b1_v2),ncol=2)
cosalfa=cosine(b1_v1,b1_v2)
alfa=acos(cosalfa)
```
La base no ortogonal  es:

$\mathfrak{B}$=\{$v_1$=(`r b1_v1[1] `, `r b1_v1[2] `) ;$v_2$=(`r b1_v2[1] `, `r b1_v2[2] `) \}

Estos vectores tienen norma 1 y un ángulo entre ellos de `r round(matlib::angle(Base_1[,1],Base_1[,2]),2)` grados
El coseno entre los rasgos latentes es: `r cosalfa `

Ahora simularé las coordenadas de las habilidades en esa base. Estas representan las habilidades de los sujetos respecto a esos constructos por lo tanto deben la correlacion entre los rasgos de los sujetos es la correlacion entre los ejes

Observemos que no es necesario para obtener estos datos saber la base solo el ángulo.

Es decir se puede tomar por ejemplo la base cualquier rotación de esta por ejemplo
la base

$\mathfrak{B}$=\{$v_1$= c(1,0)  ;$v_2=c(cos(\alpha),sen(\alpha))$ \} osea

$\mathfrak{B}$=\{$v_1$= c(1,0)  ;$v_2$=(`r c(cosalfa, sin(alfa))` \}

Observemos que lo que se simulan son coordenadas en una base y eso hay que tenerlo en cuenta al transformar

Nosotros seguiremo con la general

```{r simulo datos en B, include=FALSE}
cov_matrix <- matrix(c(1,cosalfa,cosalfa,1),ncol=2,byrow=T)
coord_B<- mvtnorm::rmvnorm(3000, sigma = cov_matrix)
Sigma_B=cor(coord_B)
```
La correlacion entre las habilidades es `r print(cor(coord_B)) `

En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la matriz de cambio de base P

$$\theta^{\mathfrak{E}}=P\theta^{\mathfrak{B}}$$

```{r coordenadas en C, include=FALSE}
P=Base_1
P_inv=solve(P)
P_invT=t(P_inv)
M=t(P)%*%P
coord_C <- t(apply(coord_B, 1, function(x) P %*% as.matrix(x)))
Sigma_C=cor(coord_C)
```

Si voy a los ejes canónicos la corelación entre las habilidades es `r cor(coord_C) `

Debemos observar que los puntos quedan fijos en el espacio lo que cambia son las coordenadas dado que P es la transformación identidad expresada en ambas bases.

Para mantener invariante la norma debemos definir en el espacio de la habilidad con la base B un producto interno con matriz de Gramm M
$$ M= P^T P$$
```{r }
print("M resulta")
M
```

```{r norma, echo=TRUE, include=FALSE}
theta_C=coord_C[17,]
#theta_C
theta_B=solve(Base_1,theta_C)
norma_B=as.numeric(sqrt(t(theta_B)%*%M%*%theta_B))
norma_C=as.numeric(sqrt(t(theta_C)%*%theta_C))
#norma_B
#norma_C
```

Por lo tanto la norma permanece invariante esto se puede comprobar mediante un ejemplo particular.

Consderamos un caso particular de habilidad dada sus coordenadas en la base canónica por `r round(theta_C,4) ` y su norma usando el producto punto es `r round(norma_C,4) `

Las coordenadas de este punto expresados en la base B son `r round(theta_B,4) ` y su norma utilizando el producto internos con matriz de Gramm M es  `r round(norma_B,4) `

Es claro que al existir una transformación lineal sobre las coordenadas esto afectará a las varianzas (no a la distribución ya que partimos de una normal) y también las correlaciones.

Dependera si la correlación aumenta o disminuye entre las coordenadas en una base y la otra de la matriz P y en definitiva del ángulo entre los vectres de la base original

La relación viene dada por 
$$Var(coord_C(\theta))=P Var(coord_B(\theta)) P^T$$

$$Cor(coord_C(\theta))=U P Var(coord_B(\theta)) P^T U=UVar(coord_C(\theta))U$$

siendo

$$U=diag((Var(coord_C(\theta)))^{-1/2})$$

OBSERVACION
De los infinitas bases podemos tamar dos que llevan en su trnsformaci’on a la matriz identidad e los ejes coordenados

A partir de

$$Var(coord_C(\theta))=P Var(coord_B(\theta)) P^T$$

si consideramos $$Var(coord_C(\theta))=I$$

obtenemos dos ecuaciones que da  que

  $$I=P Var(coord_B(\theta)) P^T$$

Si las resolvemos obtenemos

 $$ P^{-1}(P^T)^{-1}=M^{-1}=Var(coord_B(\theta))$$

o sea

 $$P^TP=M=Var(coord_B(\theta))^{-1}$$

Es claro que estas P se pueden hallar pero pueden no representar el fenómeno

Igualmente avancemos con esto

Notemos por $$Var(coord_B(\theta))=\Sigma_B$$ y $$Var(coord_C(\theta))=\Sigma_C$$

Si realizamosla descomosici'on espectral de $$\Sigma_B$$ obtenemos

$$\Sigma_B=Q \Lambda Q^{-1}$$ con Q ortogonal o sea $Q^{-1}=Q^T$ la matriz de vectores propios y $\Lambda$ la matriz diagonal de valores propios
luego

$$\Sigma_B^{-1}=Q \Lambda^{-1} Q^{-1}$$

$$P^T P=M=\Sigma_B^{-1}=Q (\Lambda)^{-1/2} Q^{-1} Q  (\Lambda)^{-1/2} Q^{-1}$$

o sea encuentro la matriz raiz cuadrada $$\sqrt{\Sigma_B}^{-1}=Q (\Lambda)^{-1/2} Q^{-1}$$

Es claro que su uso $$P=P^T=Q (\Lambda)^{-1/2} Q^{-1}$$

```{r }

DE=eigen(Sigma_B)
raizD=diag(sqrt(DE$values)^{-1})
Q=DE$vectors
Qinv=solve(Q)
P_VP=Q%*%raizD%*%Qinv
coord_VP <- t(apply(coord_B, 1, function(x) P_VP %*% as.matrix(x)))
cor(coord_VP)
```

```{r }
plot(coord_VP)
```

No es necesario hallar la matriz raiz cuadrada ya que su usamos $$P^T=Q (\Lambda)^{1/2}$$
también sirve ya que 
$$\Sigma_B^{-1}=P^TP=Q (\Lambda)^{-1/2}   (\Lambda)^{-1/2} Q^{-1}$$

Observemos que estamos usando en este caso como base los vectores  propios escalados
Es decir mi base ortogonal es la de vectores propios escalados

Luego $$P=(\Lambda)^{-1/2} Q^{-1}$$

```{r }

DE=eigen(Sigma_B)
raizD=diag(sqrt(DE$values)^{-1})
Qinv=solve(DE$vectors)
P_VP2=raizD%*%Qinv
coord_VP2 <- t(apply(coord_B, 1, function(x) P_VP2 %*% as.matrix(x)))
cor(coord_VP2)
```

```{r }
plot(coord_VP2)
```

Otra forma de hallar P es usar la descomposición de Cholesqui de M o su inversa

Como $\Sigma_B$ es simétrica y supongamos definida positiva entonces por Cholesky podemos escribir

$$\Sigma_B=L L^t$$
donde L es una matrz triangular inferior, o sea $P=L^{T}$

Si trabajamos con
$$\Sigma_B^{-1}=(L^{-1})^T L^{-1}$$
o sea $P=L^{-1}

```{r   echo=FALSE}
Linv=chol(Sigma_B)
PChol=solve(Linv)
coord_Chol <- t(apply(coord_B, 1, function(x) PChol %*% as.matrix(x)))
cor(coord_VP2)
```

CASO CON MATRIZ DE VARIANZAS COVARIANZAS

```{r , include=FALSE}
cov_matrix <- matrix(c(4,3,3,9),ncol=2,byrow=T)
coord_B_2<- mvtnorm::rmvnorm(3000, sigma = cov_matrix)
Sigma_B_2=cov(coord_B_2)
cor(coord_B_2)
coord_C_2 <- t(apply(coord_B_2, 1, function(x) P %*% as.matrix(x)))
Sigma_C_2=cor(coord_C_2)
cor(coord_C_2)
```
Vemos que al utilizar la base normalizada las coordenadas en C tienen varianza 1

```{r }

DE=eigen(Sigma_B_2)
raizD=diag(sqrt(DE$values)^{-1})
Q=DE$vectors
Qinv=solve(Q)
P_VP_2=Q%*%raizD%*%Qinv
coord_VP_2 <- t(apply(coord_B_2, 1, function(x) P_VP_2 %*% as.matrix(x)))
cor(coord_VP_2)
cov(coord_VP_2)
plot(coord_VP_2)
```

Idéntico resultado da con los otros métodos
Concluimos que si se trabaja con la matriz de varianza covarianzas o la de correlacion da lo mismo

EJEMPLIFICACION

Primero ejemplifico el caso donde conozco la P luego cuando solo conozco la correlacion entre las habilidades

Los cosenos directores respecto a la base en el espacio test es

$$cos \alpha_i^{B^*}=\frac{(D')^{-\frac{1}{2}} M^{-1} a_i^{*}}{\sqrt{(a_i^{*})^{T}M^{-1}(a_i^{*})}}$$

La distancia al origen en esta dirección esta dada por

$$D_i=\frac{-d_i}{\sqrt{(a_i^{*})^{T}M^{-1}(a_i^{*})}}$$

y la pendiente

$$ S_i=\frac{\sqrt{(a_i^{*})^{T}M^{-1}(a_i^{*})}}{4}$$

```{r slope, include=FALSE}
repre=function(M,a,d){
  M_inv=solve(M)
  norm_a=(sqrt(t(a)%*%M_inv%*%a))[1,1]
  S=norm_a/4
  D=-d/norm_a
  cdiralfa=diag(diag(M_inv)^{-1/2})%*%M_inv%*%a/norm_a 
  return(list(cosdir=cdiralfa,D=D,Slope=S))
}
```
repre(M,c(0.5,1.5),0.2)

## Como determinar P cuando se puede obtener los ejes oblicuos

Este punto para mi es importante pues las secciones anteriores se formulan en base a que conocemos la base oblicua donde el fenómeno es interpretable y tiene las dimensiones "verdaderas".

Entonces si uno tiene las respuestas por ej dicotomicas de los ítems hay una directa relación entre los parámetros y los loadings y comunalidades del análisis factorial

Por ejemplo en mirt se realiza un analisis factorial (con nfact) donde las variables son los ítems y se calculan los parámetros como

$$a_{ij}=\frac{loading_{ij}}{\sqrt(1-h_j^2)}$$

y si $q_j$ es la normal inversa de la dificultad

$$d_j=\frac{q_j}{\sqrt(1-h_j^2)}$$

Esto esta computado en el test space

Si usamos una rotación (por ej oblimin) podemos calcular la base en que estan los parametros comparandola con la solucion sin rotar.

Vemos un ejemplo simple de como proceder en el caso bidimensional pero facilmente generalizable a cualquier dimensión

Consideremos el espacio test y allí se realiza el analisis factorial

Consideremos primero la solución factorial sin rotar y sea $F_1$ y $F_2$ estos factores ortogonales o sea $r_{F_1F_2}=0$ ysean $T_1$ y $T_2$ los factores oblicuos
Notemos por $\alpha$ al ángulo de $T_1$ con $F_1$ y $\beta$ al ángulo de $T_2$ también con $F_1$ y sea $\phi$ al ángulo del vector que representa a un ítem genérico (evitamos el sub índice para simplificar) con $F_1$

Sea

$$\begin{equation} L=\begin{bmatrix}
l_{11} & l_{12} \\
l_{21} & l_{22} \\ \vdots & \vdots \\
l_{n1} & l_{n2}
\end{bmatrix}
\end{equation}$$

 la matriz que contiene los pesos factoriales en los factores ortogonales $F_1$ y $F_2$

 Se cumple que $l_{j1}=h_j \cos(\phi)$ y $l_{j2}=h_j \sin(\phi)$

Sea $S$ la matriz que contiene las correlaciones de cada item con los ejere rotados o sea

$$\begin{equation} S=\begin{bmatrix}
r_{1T_1} & r_{1T_2} \\
r_{2T_1} & r_{2T_2} \\ \vdots & \vdots \\
r_{nT_1} & r_{nT_2}
\end{bmatrix}
\end{equation}$$

Es facil ver (Harman) que se cumple $$S=LT$$

con

$$\begin{equation} T=\begin{bmatrix}
\cos(\alpha) & \cos(\beta)  \\
\sin(\alpha) & \sin(\beta) 
\end{bmatrix}
\end{equation}$$

Sea ahora

$$\begin{equation} B=\begin{bmatrix}
b_{11} & b_{12} \\
b_{21} & b_{22} \\ \vdots & \vdots \\
b_{n1} & b_{n2}
\end{bmatrix}
\end{equation}$$

la matriz de coordenadas de los ítems respecto a los ejes oblicuos $T_1$ y $T_2$

Se cumple que $$S=B \Phi$$

con

$$\begin{equation} \Phi=\begin{bmatrix}
1 & r_{T_1T_2} \\
r_{T_1T_2} & 1 
\end{bmatrix}
\end{equation}$$

Sabemos que $\Phi=T^tT$ entonces a partir de lo anterior
$$B=S\Phi^{-1}=LT(T^tT)^{-1}$$
entonces

$$ B=L(T^t)^{-1}$$

Como $$S=B \Phi$$
y  $$S=LT$$

Veamos que tanto S como L son matrices  nx2 y T 2x2.
Entonces la relación debe verificarse para todos (esta sobreidentificada)
y me alcanza resolverla para dos ítems por ejemplo el j y el k.
La matriz $(L)_{jk}$ es 2x2 y si el sistema tiene soluci'o invertible

entonces $$T=(L^{-1})_{jk} (B \Phi)_{jk}$$

Nosotros consideramosen el espacio test la base $E^*$ canónica y $B^*$ que se relacionaban mediante

$$\begin{equation}
a_i^{E^*}=(P^{-1})^t a_i^{B^*}
\end{equation}$$

siendo $P$ la matriz de cambio de base en el espacio latente o sea $$\theta^E=P\theta^B$$ siendo E la base ortonormal en el espacio latente y B la base oblicua.

Recordemos que se cumple $$ B=L(T^t)^{-1}$$

de donde
$$ L=BT^t$$

Matricialmente

$$\begin{equation}
\begin{bmatrix}
l_{11} & l_{12} \\
l_{21} & l_{22} \\ \vdots & \vdots \\
l_{n1} & l_{n2}
\end{bmatrix}=
\begin{bmatrix}
b_{11} & b_{12} \\
b_{21} & b_{22} \\ \vdots & \vdots \\
b_{n1} & b_{n2}
\end{bmatrix} \begin{bmatrix}
\cos(\alpha) & \cos(\beta)  \\
\sin(\alpha) & \sin(\beta) 
\end{bmatrix}^t
\end{equation}$$

o sea

$$\begin{equation}
\begin{bmatrix}
l_{11} & l_{11} & \vdots & l_{n1} \\
l_{12} & l_{22} &  \vdots & l_{n2}\\
\end{bmatrix}=\begin{bmatrix}
\cos(\alpha) & \cos(\beta)  \\
\sin(\alpha) & \sin(\beta) 
\end{bmatrix}
\begin{bmatrix}
b_{11} & b_{11} & \vdots & b_{n1} \\
b_{12} & b_{22} &  \vdots & b_{n2}\\
\end{bmatrix} 
\end{equation}$$

O sea T es la matriz de cambio de base en el espacio test.
Esto facilmente se puede ver también geométricamente sin necesidad de operar.

Es decir $$T=(P^{-1})^t$$

por lo que $$P=(T^{-1})^t$$

Vemos un ejemplo. Simulemos datos que suponemos no sabemos ni las habilidades ni los par'ametros de los ítems.

```{r}
#generemos 3000 datos con distribucion bivariada
S1 <- matrix(c(4.4, 2.5, 2.5, 8), nrow = 2, ncol = 2)
# Generar datos normales bivariados
n <- 3000  # Número de observaciones
mu <- c(0, 0)  # Medias de las variables
set.seed(1234567)
data <- mvrnorm(n = n, mu = mu, Sigma = S1)

#generemos un banco de 10 items
a<-matrix(c(.7471, .1428,.4595, .0097, .0067, .4840,
 .0470,.5521, .0204, .1482,
1.3547, .0064, .5362,
1.3761, .0861, .4676,
.8525, .0383, .2574,0.87),ncol=2,byrow=T)

d <- matrix(c(.1826,-.1924,-.4656,.5845,-1.0403,
.6431,.0912,-.8082,-1.8398,.4139),ncol=1)*1.702

dataset1 <- simdata(a, d, itemtype = '2PL',Theta=data)

```

Primero obtenganos la solución sin rotar

```{r}
#Estimacion de la solución sin rotar

mod2=mirt(dataset1, 2, itemtype='2PL', verbose = FALSE)

coefs <- coef(mod2, rotate = 'none') |> map(as_tibble)
item_pars <- coefs |>
  bind_rows()      |>
  slice(1:10)       |>
  select(a1:d)     |>
  rownames_to_column(var = "item")
cov_pars <- coefs[['GroupPars']] |>
  pivot_longer(everything())     |>
  deframe()
cov_matrix <- cov_pars[c('COV_11', 'COV_21', 'COV_21', 'COV_22')] |>
  matrix(nrow = 2)
print(item_pars)
print(cov_matrix)
```

Si queremos ver la relación entre los factores y los parámetros de los ítems antes mencionada

```{r}
# Esto muestra la relacion de los factores con los items

sum <- summary(mod2, rotate = 'none')
sum$rotF      |>
  as_tibble() |>
  mutate(
    across(everything(), ~1.702 * . / sqrt(1-as.numeric(sum$h2)))
  )           |>
rename_with(str_replace, everything(), pattern = 'F', replacement = 'a')
```

Sea ahora la solución rotada

```{r}
#AHORA ROTADA

coefsR <- coef(mod2, rotate = 'oblimin') |> map(as_tibble)
item_parsR <- coefsR |>
  bind_rows()      |>
  slice(1:10)       |>
  select(a1:d)     |>
  rownames_to_column(var = "item")
cov_parsR <- coefsR[['GroupPars']] |>
  pivot_longer(everything())     |>
  deframe()
cov_matrixR <- cov_parsR[c('COV_11', 'COV_21', 'COV_21', 'COV_22')] |>
  matrix(nrow = 2)
print(item_parsR)
print(cov_matrixR)
```

Se da igual que hoy la relaci'on entre los parámetros de los ítems y los pesos en la solución rotada

```{r}
sumR <- summary(mod2, rotate = 'oblimin')
sumR$rotF      |>
  as_tibble() |>
  mutate(
    across(everything(), ~1.702 * . / sqrt(1-as.numeric(sum$h2)))
  )           |>
  rename_with(str_replace, everything(), pattern = 'F', replacement = 'a')

```

Calculemos entonces

```{r}
#parametros solucion ortogonal
L<-as.matrix(item_pars[,2:3])
#parametros ejes oblicuos
B<-as.matrix(item_parsR[,2:3])
# Matriz de correlacion entre los factores rotados
PHI<-cov_matrixR

```

```{r}
#Tomaremos item 1 y 2
S=B%*%PHI
T=solve(L[1:2,])%*%S[1:2,]
print(T)
```

Verifiquemos ahora que esta solución sirve para las matrices completas

```{r}
#Verificacon
round(L%*%T-S,8)
```

```{r}
P=t(solve(T))
print(P)
```

### 2024-05-29 notebooks/Basis_change_explanation.qmd
Adds initial version of "misconception" notebook

---
title: "Explicación sobre la representación gráfica y el cambio de base"
author: "Daniel Morillo"
format: html
editor: visual
---

```{r setup}
#| warning: false
library(matlib)
library(lsa)
library(mvtnorm)

set.seed(123)
```

# Introducción

En este notebook voy a intentar dar una explicación lo más detallada posible y
pormenorizada de una concepción errónea respecto a la interpretación de la base
oblicua y la distribución de los vectores de rasgo latente.

Es importante aclarar que esta interpretación incorrecta es muy difícil de
entender; a mí me costó muchísimo darme cuenta de ello y "deshacerme" de ella.
No obstante, creo que esta concepción errónea es la principal barrera que
estamos teniendo ahora mismo Mario y yo para entendernos, y por lo tanto para
hacer avances significativos. Una vez hayamos podido aclarar esto, estoy seguro
de que habremos superado una barrera muy importante y haremos progresos mayores.

Empezaré por analizar el
[notebook "Representacion" de Mario](https://github.com/DaniMori/mirt-parameters/blob/experimental/derivations-by-mario/notebooks/Representacion.Rmd),
fechado el 19 de mayo y enviado por email el 27 de mayo de 2024. Al comienzo del
mismo se crea una base oblicua en el espacio latente, y después se hace una
simulación de datos "en esa base" . Aunque dicha simulación no es incorrecta,
mostraré por qué NO responde al objetivo del artículo. Lo que sí es incorrecto
es asumir que de esta manera "su correlación [es] el angulo entre los factores
oblicuos", lo cual mostraré a continuación. Por último, explicaré con dos
ejemplos distintos cómo debería hacerse de manera esta simulación para que sí
responda a nuestro objetivo.

# Creación de la base

Mario genera en su notebook
([ll. 52-64](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L52-L64)),
de manera correcta, una base del espacio latente formada por dos vectores
normales, oblicuos entre sí. Reproduzco aquí su código (con adaptaciones de
estilo):

```{r oblique-basis}
# Crear vectores de "dirección" iniciales:
b_1 <- c(3, 1)
b_2 <- c(1, 4)

# Normalizar vectores:
b_1 <- b_1 / len(b_1)
b_2 <- b_2 / len(b_2)

```

Este código genera dos vectores, $\mathbf{b}_1$ y $\mathbf{b}_2$, los cuales
forman la base oblicua o "base original"
$\mathcal{B} = \{ \mathbf{b}_1, \mathbf{b}_2 \}$. Estos vectores están
expresados en la base canónica $\mathfrak{E}$, y por lo tanto la matriz $P$ que
transforma cualquier vector de $\mathfrak{B}$ a la base canónica $\mathfrak{E}$
es $\mathbf{P} = [ \mathbf{b}_1, \mathbf{b}_2 ]$:

```{r transform-matrix}
P <- cbind(b_1, b_2)
```

Al ser la base $\mathfrak{B}$ oblicua, los vectores forman entre sí un ángulo
dado por:

```{r basis-vectors-angle}
cos_b1_b2       <- cosine(b_1, b_2)
angle_deg_b1_b2 <- angle(b_1, b_2)
```

El ángulo es de `r angle_deg_b1_b2 |> round(2)`º, al cual corresponde un coseno
de `r cos_b1_b2 |> round(3)`.

# Simulación

A continuación Mario simula los vectores $\mathbf{\theta}$ de rasgo latente
([ll. 69-89](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69-L89)),
los cuales *deseamos representar en esta base oblicua* $\mathfrak{B}$. No
obstante, nótese que los vectores no están necesariamente "en una base" como
tal, sino que son coordenadas algebraicas. Según yo lo entiendo (y puede que
aquí sea yo mismo quien tiene una concepción errónea), **los vectores puramente
algebraicos no necesitan de ninguna base para ser generados, sino solamente para
ser representados**. Por lo tanto, **los vectores generados en esta simulación
no necesitarían de la base para ser generados**. Esto es común en TRI
multidimensional: Simulamos vectores de rasgo latente con una determinada
distribución (incluyendo su matriz de covarianzas), sin hacer referencia a
ninguna base.

Además, esto ya lo hemos visto y hecho anteriormente de hecho. En el notebook
["Ejemplificando art"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd)
del 13 de septiembre de 2023,
[Mario genera una base oblicua](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L43-L53), y después
[simula una distribución de vectores de rasgo latente que nada tienen que ver con los vectores de esa base](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L83-L89).

No obstante, lo que vemos en este notebook es que se desea que las coordenadas
de los vectores de rasgo latente (en $\mathfrak{B}$)
["representen las habilidades de los sujetos respecto a esos constructos"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69). Para ello, Mario dice que
["la correlacion entre los rasgos de los sujetos es la correlacion entre los ejes"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
**Aquí es donde está la concepción errónea que es necesario corregir**; como
digo, esto es un concepto muy contraintuitivo, y que solamente resulta después
de analizar muy en detalle las deducciones y entender adecuadamente la
representación.

::: callout-warning
La correlación entre los vectores de la base $\mathfrak{B}$ **NO representa la
correlación de la distribución de los rasgos latentes**.
:::

Voy a hacer la simulación tal y como la hace Mario, y después justifico esa
concepción errónea.

```{r simulate-coordinates}
cov_matrix <- matrix(c(1, cos_b1_b2, cos_b1_b2, 1), ncol = 2)
coords_B   <- rmvnorm(3000, sigma = cov_matrix)

cov_coords_B <- cov(coords_B)
cor_coords_B <- cor(coords_B)[2, 1]
```

La correlación muestral de los rasgos latentes simulados es
`r cor_coords_B |> round(3)`.

Ahora bien: Hay que recordar que **el objetivo de proponer una base latente
$\mathfrak{B}$ es** que las coordenadas de los vectores de rasgo latente
transformadas a una base ortogonal, tengan una distribución independiente, es
decir, que su correlación sea 0. Yendo más allá (por aclarar todos los
conceptos), queremos **que las coordenadas de los vectores de rasgo latente
transformadas a una base ortonormal** (por simplificar, a la base canónica, en
este ejemplo), **tengan una matriz de covarianzas identidad**.

::: callout-important
Las coordenadas de los vectores de rasgo latente, transformadas a la base
canónica, deben tener una matriz de covarianzas identidad.
:::

# Transformación a la base canónica

Tal y como indica Mario
([ll. 91-93](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L91-L93)),
"En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la
matriz de cambio de base $\mathbf{P}$".

Vamos a hacer el cambio de base, tal y como viene en el artículo, y
[como hace Mario en este notebook](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L96-L103). (Aunque utilizao un código distinto, el
resultado que obtengo es el mismo.)

```{r transform-coordinates}
coords_C <- P %*% t(coord_B) |> t()

cov_coords_C <- cov(coords_C)
cor_coords_C <- cor(coords_C)[2, 1]
```

La matriz de covarianzas muestral de las coordenadas en la base canónica es:

`r cov_coords_C |> round(3)`

La correlación entre las coordenadas en la base canónica es
`r cor_coords_C |> round(3)`.

Sin embargo, tal y como decíamos antes, queremos que la correlación en esta base
sea 0, pero en su lugar **es una correlación muy alta**, próxima a 1. ¿Por qué
está ocurriendo esto? Es la concepción errónea a la que me refería antes la que
lo explica.

## Matriz de covarianzas de las coordenadas de rasgo latente

Este apartado explica por qué la transformación a la base canónica no da el
resultado que esperamos.

Las coordenadas en la base original se han simulado con una matriz de
covarianzas igual a:

$$
\begin{bmatrix}
             1 & \cos{\alpha} \\
  \cos{\alpha} &            1
\end{bmatrix}
$$

 siendo $\alpha$ el ángulo entre los dos vectores de la base $\mathfrak{B}$.
 Ahora bien, esta matriz también es igual a
 $\mathbf{P}^T\mathbf{P} = \mathbf{M}$ (no pongo la justificación matemática
 porque no creo que sea necesario, pero podemos verlo en el ejemplo):

```{r compute-Gram-matrix}
M <- t(P) %*% P
M
cov_matrix

# Dan exactamente igual, salvo por la precisión numérica en el [1, 1]
cov_matrix == M
```

 Como ya sabemos, si la matriz de covarianzas de las coordenadas originales es
$\mathbf{\Sigma}$, la matriz de covarianzas de las coordenadas en la base
canónica $\mathbf{\Sigma}^\mathfrak{B}$ (tras aplicar la pre-multiplicación por
$\mathbf{P}$) responde a:

$$
\mathbf{\Sigma}^\mathfrak{E} = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T
$$

Ahora bien, si simulamos las coordenadas originales con
$\mathbf{\Sigma} = \mathbf{M}$, tenemos que:

$$
\mathbf{\Sigma}^\mathfrak{E} =
  \mathbf{P} \mathbf{M} \mathbf{P}^T =
  \mathbf{P} \mathbf{P} \mathbf{P}^T \mathbf{P}^T
$$

Lo cual (en general) NO equivale a una matriz identidad.

Es necesario darse cuenta de que hemos llegado a este punto porque **hemos
simulado los vectores de rasgo latente (originales) con una correlación igual al
coseno del ángulo entre los vectores de la base $\mathfrak{B}$**, y que hacemos
esto porque **consideramos que dicho ángulo representa la correlación entre los
constructos sustantivos que nos interesa representar**. No obstante, vemos que
**esto NO es así**.

::: callout-warning
El coseno del ángulo entre los vectores de la base $\mathfrak{B}$ NO representa
la correlación entre los constructos sustantivos que nos interesa representar.
:::

### 2024-05-30 notebooks/Basis_change_explanation.qmd
contestaciones

@@
 
-En este notebook voy a intentar dar una explicación lo más detallada posible y
-pormenorizada de una concepción errónea respecto a la interpretación de la base
-oblicua y la distribución de los vectores de rasgo latente.
+En este notebook voy a intentar dar una explicación lo más detallada posible y pormenorizada de una concepción errónea respecto a la interpretación de la base oblicua y la distribución de los vectores de rasgo latente.
 
-Es importante aclarar que esta interpretación incorrecta es muy difícil de
-entender; a mí me costó muchísimo darme cuenta de ello y "deshacerme" de ella.
-No obstante, creo que esta concepción errónea es la principal barrera que
-estamos teniendo ahora mismo Mario y yo para entendernos, y por lo tanto para
-hacer avances significativos. Una vez hayamos podido aclarar esto, estoy seguro
-de que habremos superado una barrera muy importante y haremos progresos mayores.
+Es importante aclarar que esta interpretación incorrecta es muy difícil de entender; a mí me costó muchísimo darme cuenta de ello y "deshacerme" de ella.
+No obstante, creo que esta concepción errónea es la principal barrera que estamos teniendo ahora mismo Mario y yo para entendernos, y por lo tanto para hacer avances significativos.
+Una vez hayamos podido aclarar esto, estoy seguro de que habremos superado una barrera muy importante y haremos progresos mayores.
 
-Empezaré por analizar el
-[notebook "Representacion" de Mario](https://github.com/DaniMori/mirt-parameters/blob/experimental/derivations-by-mario/notebooks/Representacion.Rmd),
-fechado el 19 de mayo y enviado por email el 27 de mayo de 2024. Al comienzo del
-mismo se crea una base oblicua en el espacio latente, y después se hace una
-simulación de datos "en esa base" . Aunque dicha simulación no es incorrecta,
-mostraré por qué NO responde al objetivo del artículo. Lo que sí es incorrecto
-es asumir que de esta manera "su correlación [es] el angulo entre los factores
-oblicuos", lo cual mostraré a continuación. Por último, explicaré con dos
-ejemplos distintos cómo debería hacerse de manera esta simulación para que sí
-responda a nuestro objetivo.
+Empezaré por analizar el [notebook "Representacion" de Mario](https://github.com/DaniMori/mirt-parameters/blob/experimental/derivations-by-mario/notebooks/Representacion.Rmd), fechado el 19 de mayo y enviado por email el 27 de mayo de 2024.
+Al comienzo del mismo se crea una base oblicua en el espacio latente, y después se hace una simulación de datos "en esa base" .
+Aunque dicha simulación no es incorrecta, mostraré por qué NO responde al objetivo del artículo.
+Lo que sí es incorrecto es asumir que de esta manera "su correlación \[es\] el angulo entre los factores oblicuos", lo cual mostraré a continuación.
+Por último, explicaré con dos ejemplos distintos cómo debería hacerse de manera esta simulación para que sí responda a nuestro objetivo.
 
@@
 
-Mario genera en su notebook
-([ll. 52-64](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L52-L64)),
-de manera correcta, una base del espacio latente formada por dos vectores
-normales, oblicuos entre sí. Reproduzco aquí su código (con adaptaciones de
-estilo):
+Mario genera en su notebook ([ll.
+52-64](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L52-L64)), de manera correcta, una base del espacio latente formada por dos vectores normales, oblicuos entre sí.
+Reproduzco aquí su código (con adaptaciones de estilo):
 
@@
 
-Este código genera dos vectores, $\mathbf{b}_1$ y $\mathbf{b}_2$, los cuales
-forman la base oblicua o "base original"
-$\mathcal{B} = \{ \mathbf{b}_1, \mathbf{b}_2 \}$. Estos vectores están
-expresados en la base canónica $\mathfrak{E}$, y por lo tanto la matriz $P$ que
-transforma cualquier vector de $\mathfrak{B}$ a la base canónica $\mathfrak{E}$
-es $\mathbf{P} = [ \mathbf{b}_1, \mathbf{b}_2 ]$:
+Este código genera dos vectores, $\mathbf{b}_1$ y $\mathbf{b}_2$, los cuales forman la base oblicua o "base original" $\mathcal{B} = \{ \mathbf{b}_1, \mathbf{b}_2 \}$.
+Estos vectores están expresados en la base canónica $\mathfrak{E}$, y por lo tanto la matriz $P$ que transforma cualquier vector de $\mathfrak{B}$ a la base canónica $\mathfrak{E}$ es $\mathbf{P} = [ \mathbf{b}_1, \mathbf{b}_2 ]$:
 
@@
 
-Al ser la base $\mathfrak{B}$ oblicua, los vectores forman entre sí un ángulo
-dado por:
+Al ser la base $\mathfrak{B}$ oblicua, los vectores forman entre sí un ángulo dado por:
 
@@
 
-El ángulo es de `r angle_deg_b1_b2 |> round(2)`º, al cual corresponde un coseno
-de `r cos_b1_b2 |> round(3)`.
+El ángulo es de `r angle_deg_b1_b2 |> round(2)`º, al cual corresponde un coseno de `r cos_b1_b2 |> round(3)`.
 
@@
 
-A continuación Mario simula los vectores $\mathbf{\theta}$ de rasgo latente
-([ll. 69-89](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69-L89)),
-los cuales *deseamos representar en esta base oblicua* $\mathfrak{B}$. No
-obstante, nótese que los vectores no están necesariamente "en una base" como
-tal, sino que son coordenadas algebraicas. Según yo lo entiendo (y puede que
-aquí sea yo mismo quien tiene una concepción errónea), **los vectores puramente
-algebraicos no necesitan de ninguna base para ser generados, sino solamente para
-ser representados**. Por lo tanto, **los vectores generados en esta simulación
-no necesitarían de la base para ser generados**. Esto es común en TRI
-multidimensional: Simulamos vectores de rasgo latente con una determinada
-distribución (incluyendo su matriz de covarianzas), sin hacer referencia a
-ninguna base.
+A continuación Mario simula los vectores $\mathbf{\theta}$ de rasgo latente ([ll.
+69-89](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69-L89)), los cuales *deseamos representar en esta base oblicua* $\mathfrak{B}$.
+No obstante, nótese que los vectores no están necesariamente "en una base" como tal, sino que son coordenadas algebraicas.
+Según yo lo entiendo (y puede que aquí sea yo mismo quien tiene una concepción errónea), **los vectores puramente algebraicos no necesitan de ninguna base para ser generados, sino solamente para ser representados**.
+Por lo tanto, **los vectores generados en esta simulación no necesitarían de la base para ser generados**.
+Esto es común en TRI multidimensional: Simulamos vectores de rasgo latente con una determinada distribución (incluyendo su matriz de covarianzas), sin hacer referencia a ninguna base.
 
-Además, esto ya lo hemos visto y hecho anteriormente de hecho. En el notebook
-["Ejemplificando art"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd)
-del 13 de septiembre de 2023,
-[Mario genera una base oblicua](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L43-L53), y después
-[simula una distribución de vectores de rasgo latente que nada tienen que ver con los vectores de esa base](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L83-L89).
+**CONTESTO A LO ANTERIOR**
 
-No obstante, lo que vemos en este notebook es que se desea que las coordenadas
-de los vectores de rasgo latente (en $\mathfrak{B}$)
-["representen las habilidades de los sujetos respecto a esos constructos"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
-Para ello, Mario dice que
-["la correlacion entre los rasgos de los sujetos es la correlacion entre los ejes"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
-**Aquí es donde está la concepción errónea que es necesario corregir**; como
-digo, esto es un concepto muy contraintuitivo, y que solamente resulta después
-de analizar muy en detalle las deducciones y entender adecuadamente la
-representación.
+**Intentare aclarar esto.
+Todo vector que al ser vector pertenece a un espacio vectorial, tiene sus coordenadas en cierta base, no se puede pensar un vector sin su base.
+Muchas veces se asume sin decirlo que la base el la canónica y es por eso no la nombramos y entonces generamos las coordenadas y asumimos que estan en la base (1,0) , (0,1).
+En mi caso yo genero coordenadas normales e impongo que sean las coordenadas en la base obicua.
+Por lo tanto si se necesita una base.**
+
+**Cuando dices que es comun simular vectores sin hacer mencion a la base es porque se presupone que las coordenadas están en la están en la canónica**
+
+Además, esto ya lo hemos visto y hecho anteriormente de hecho.
+En el notebook ["Ejemplificando art"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd) del 13 de septiembre de 2023, [Mario genera una base oblicua](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L43-L53), y después [simula una distribución de vectores de rasgo latente que nada tienen que ver con los vectores de esa base](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L83-L89).
+
+No obstante, lo que vemos en este notebook es que se desea que las coordenadas de los vectores de rasgo latente (en $\mathfrak{B}$) ["representen las habilidades de los sujetos respecto a esos constructos"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
+Para ello, Mario dice que ["la correlacion entre los rasgos de los sujetos es la correlacion entre los ejes"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
+**Aquí es donde está la concepción errónea que es necesario corregir**; como digo, esto es un concepto muy contraintuitivo, y que solamente resulta después de analizar muy en detalle las deducciones y entender adecuadamente la representación.
 
 ::: callout-warning
-La correlación entre los vectores de la base $\mathfrak{B}$ **NO representa la
-correlación de la distribución de los rasgos latentes**.
+La correlación entre los vectores de la base $\mathfrak{B}$ **NO representa la correlación de la distribución de los rasgos latentes**.
 :::
 
-Voy a hacer la simulación tal y como la hace Mario, y después justifico esa
-concepción errónea.
+**CONTESTO**
+
+**a) El notebook del 13 de setiembre tiene errores por eso lo mejor es desecharlo.**
+
+**b) Pero en el del 27 de mayo utilizo el concepto que mencionas por lo tanto es util discutirlo.**
+
+**"Ahora simularé las coordenadas de las habilidades en esa base.
+Estas representan las habilidades de los sujetos respecto a esos constructos por lo tanto deben la correlacion entre los rasgos de los sujetos es la correlacion entre los ejes"**
+
+**Lo que sucede es que los metodos de estimaciön hacen que no sean iguales pero deben serlo .Este art'iculo muestra el compromiso en la estimacion y como se da sesgo en la estimación**
+
+**The trade-off between factor score determinacy and the preservation of inter-factor correlations [André Beauducel](https://arxiv.org/search/stat?searchtype=author&query=Beauducel,+A), [Norbert Hilger](https://arxiv.org/search/stat?searchtype=author&query=Hilger,+N), [Tobias Kuhl](https://arxiv.org/search/stat?searchtype=author&query=Kuhl,+T)**
+
+**Es decir teoricamente los scores factoriales deberian tener corrrelación igual a la de los factores pero no se da por el hecho de como se estiman y esto produce un sesgo.**
+
+**Igual si no supongo eso no hay problema para el resto de las deducciones, miralo. Lo que pasa es que me parece adecuado expresar en las simulaciones lo que deberia pasar.**
+
+**Además llego a tu misma conclusión cuando quiero llevar la transformación a la corr 0 pero por motivos diferentes**
+
+Voy a hacer la simulación tal y como la hace Mario, y después justifico esa concepción errónea.
 
@@
 
-La correlación muestral de los rasgos latentes simulados es
-`r cor_coords_B |> round(3)`.
+La correlación muestral de los rasgos latentes simulados es `r cor_coords_B |> round(3)`.
 
-Ahora bien: Hay que recordar que **el objetivo de proponer una base latente
-$\mathfrak{B}$ es** que las coordenadas de los vectores de rasgo latente
-transformadas a una base ortogonal, tengan una distribución independiente, es
-decir, que su correlación sea 0. Yendo más allá (por aclarar todos los
-conceptos), queremos **que las coordenadas de los vectores de rasgo latente
-transformadas a una base ortonormal** (por simplificar, a la base canónica, en
-este ejemplo), **tengan una matriz de covarianzas identidad**.
+Ahora bien: Hay que recordar que **el objetivo de proponer una base latente** $\mathfrak{B}$ es que las coordenadas de los vectores de rasgo latente transformadas a una base ortogonal, tengan una distribución independiente, es decir, que su correlación sea 0.
+Yendo más allá (por aclarar todos los conceptos), queremos **que las coordenadas de los vectores de rasgo latente transformadas a una base ortonormal** (por simplificar, a la base canónica, en este ejemplo), **tengan una matriz de covarianzas identidad**.
 
 ::: callout-important
-Las coordenadas de los vectores de rasgo latente, transformadas a la base
-canónica, deben tener una matriz de covarianzas identidad.
+Las coordenadas de los vectores de rasgo latente, transformadas a la base canónica, deben tener una matriz de covarianzas identidad.
 :::
+
+**CONTESTO**
+
+**Para mi el objetivo de proponer una base latente** $\mathfrak{B}$ **es explicar los verdaderos constructos del test, no una base para que al transformarse los rasgos sean independientes ya que eso no se utiliza en la deducción por Reckase.**
+
+**Pero igual si ves a parrir de mi linea 149 me meto en este tema y encuentro que la P que lleva a rasgos independientes pero como caso particular.
+Mira linea 149 a 173 para ver** $$P^TP=M=Var(coord_B(\theta))^{-1}$$ **y luego como hallarlas.**
+
+**No hay contradicción pues hay infinitas bases ortiginales y una de ellas hace que las coordenadas tengan correlacion 0.**
+
+**Mira las graficas generdas en la linea 202 y 228 verás que tiene sihma Identidad**
 
@@
 
-Tal y como indica Mario
-([ll. 91-93](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L91-L93)),
-"En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la
-matriz de cambio de base $\mathbf{P}$".
+Tal y como indica Mario ([ll.
+91-93](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L91-L93)), "En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la matriz de cambio de base $\mathbf{P}$".
 
-Vamos a hacer el cambio de base, tal y como viene en el artículo, y
-[como hace Mario en este notebook](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L96-L103).
-(Aunque utilizao un código distinto, el
-resultado que obtengo es el mismo.)
+Vamos a hacer el cambio de base, tal y como viene en el artículo, y [como hace Mario en este notebook](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L96-L103).
+(Aunque utilizao un código distinto, el resultado que obtengo es el mismo.)
 
 ```{r transform-coordinates}
-coords_C <- P %*% t(coord_B) |> t()
+coords_C <- P %*% t(coords_B) |> t()
 
@@
 
-La correlación entre las coordenadas en la base canónica es
-`r cor_coords_C |> round(3)`.
+La correlación entre las coordenadas en la base canónica es `r cor_coords_C |> round(3)`.
 
-Sin embargo, tal y como decíamos antes, queremos que la correlación en esta base
-sea 0, pero en su lugar **es una correlación muy alta**, próxima a 1. ¿Por qué
-está ocurriendo esto? Es la concepción errónea a la que me refería antes la que
-lo explica.
+Sin embargo, tal y como decíamos antes, queremos que la correlación en esta base sea 0, pero en su lugar **es una correlación muy alta**, próxima a 1.
+¿Por qué está ocurriendo esto?
+Es la concepción errónea a la que me refería antes la que lo explica.
+
+**CONTESTO**
+
+**Explore diferentes bases y depende de como lo hagas a veces da correlacion alta o baja y como te comente antes hay una base ortogonal que hace que sea la correlación 0 .
+Vas a ver que hay graficas y codigo donde aparece esto.**
 
@@
 
-Este apartado explica por qué la transformación a la base canónica no da el
-resultado que esperamos.
+Este apartado explica por qué la transformación a la base canónica no da el resultado que esperamos.
 
-Las coordenadas en la base original se han simulado con una matriz de
-covarianzas igual a:
+Las coordenadas en la base original se han simulado con una matriz de covarianzas igual a:
 
@@
 siendo $\alpha$ el ángulo entre los dos vectores de la base $\mathfrak{B}$.
- Ahora bien, esta matriz también es igual a
- $\mathbf{P}^T\mathbf{P} = \mathbf{M}$ (no pongo la justificación matemática
- porque no creo que sea necesario, pero podemos verlo en el ejemplo):
+Ahora bien, esta matriz también es igual a $\mathbf{P}^T\mathbf{P} = \mathbf{M}$ (no pongo la justificación matemática porque no creo que sea necesario, pero podemos verlo en el ejemplo):
 
@@
 
- Como ya sabemos, si la matriz de covarianzas de las coordenadas originales es
-$\mathbf{\Sigma}$, la matriz de covarianzas de las coordenadas en la base
-canónica $\mathbf{\Sigma}^\mathfrak{B}$ (tras aplicar la pre-multiplicación por
-$\mathbf{P}$) responde a:
+Como ya sabemos, si la matriz de covarianzas de las coordenadas originales es $\mathbf{\Sigma}$, la matriz de covarianzas de las coordenadas en la base canónica $\mathbf{\Sigma}^\mathfrak{B}$ (tras aplicar la pre-multiplicación por $\mathbf{P}$) responde a:
 
@@
 
-Ahora bien, si simulamos las coordenadas originales con
-$\mathbf{\Sigma} = \mathbf{M}$, tenemos que:
+Ahora bien, si simulamos las coordenadas originales con $\mathbf{\Sigma} = \mathbf{M}$, tenemos que:
 
@@
 
-Es necesario darse cuenta de que hemos llegado a este punto porque **hemos
-simulado los vectores de rasgo latente (originales) con una correlación igual al
-coseno del ángulo entre los vectores de la base $\mathfrak{B}$**, y que hacemos
-esto porque **consideramos que dicho ángulo representa la correlación entre los
-constructos sustantivos que nos interesa representar**. No obstante, vemos que
-**esto NO es así**.
+Es necesario darse cuenta de que hemos llegado a este punto porque **hemos simulado los vectores de rasgo latente (originales) con una correlación igual al coseno del ángulo entre los vectores de la base** $\mathfrak{B}$, y que hacemos esto porque **consideramos que dicho ángulo representa la correlación entre los constructos sustantivos que nos interesa representar**.
+No obstante, vemos que **esto NO es así**.
 
 ::: callout-warning
-El coseno del ángulo entre los vectores de la base $\mathfrak{B}$ NO representa
-la correlación entre los constructos sustantivos que nos interesa representar.
+El coseno del ángulo entre los vectores de la base $\mathfrak{B}$ NO representa la correlación entre los constructos sustantivos que nos interesa representar.
 :::
+
+**CONTESTO**
+
+**Esto último se relaciona con lo que te digo esta en las lineas 149 en adelante.**

### 2024-05-30 notebooks/Basis_change_explanation.qmd
Revert "contestaciones"
This reverts commit 748a7a33d33ad652ef5b1f1ee31a539a4bddf9ad.
Applies automatic text wrapping

--- 2024-05-29 notebooks/Basis_change_explanation.qmd
@@
 
-Empezaré por analizar el
-[notebook "Representacion" de Mario](https://github.com/DaniMori/mirt-parameters/blob/experimental/derivations-by-mario/notebooks/Representacion.Rmd),
+Empezaré por analizar el [notebook "Representacion" de
+Mario](https://github.com/DaniMori/mirt-parameters/blob/experimental/derivations-by-mario/notebooks/Representacion.Rmd),
 fechado el 19 de mayo y enviado por email el 27 de mayo de 2024. Al comienzo del
@@
 mostraré por qué NO responde al objetivo del artículo. Lo que sí es incorrecto
-es asumir que de esta manera "su correlación [es] el angulo entre los factores
+es asumir que de esta manera "su correlación \[es\] el angulo entre los factores
 oblicuos", lo cual mostraré a continuación. Por último, explicaré con dos
@@
 
-Mario genera en su notebook
-([ll. 52-64](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L52-L64)),
+Mario genera en su notebook ([ll.
+52-64](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L52-L64)),
 de manera correcta, una base del espacio latente formada por dos vectores
@@
 A continuación Mario simula los vectores $\mathbf{\theta}$ de rasgo latente
-([ll. 69-89](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69-L89)),
+([ll.
+69-89](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69-L89)),
 los cuales *deseamos representar en esta base oblicua* $\mathfrak{B}$. No
@@
 Además, esto ya lo hemos visto y hecho anteriormente de hecho. En el notebook
-["Ejemplificando art"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd)
-del 13 de septiembre de 2023,
-[Mario genera una base oblicua](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L43-L53), y después
-[simula una distribución de vectores de rasgo latente que nada tienen que ver con los vectores de esa base](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L83-L89).
+["Ejemplificando
+art"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd)
+del 13 de septiembre de 2023, [Mario genera una base
+oblicua](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L43-L53),
+y después [simula una distribución de vectores de rasgo latente que nada tienen
+que ver con los vectores de esa
+base](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/example_script.Rmd#L83-L89).
 
 No obstante, lo que vemos en este notebook es que se desea que las coordenadas
-de los vectores de rasgo latente (en $\mathfrak{B}$)
-["representen las habilidades de los sujetos respecto a esos constructos"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
-Para ello, Mario dice que
-["la correlacion entre los rasgos de los sujetos es la correlacion entre los ejes"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
+de los vectores de rasgo latente (en $\mathfrak{B}$) ["representen las
+habilidades de los sujetos respecto a esos
+constructos"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
+Para ello, Mario dice que ["la correlacion entre los rasgos de los sujetos es la
+correlacion entre los
+ejes"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
 **Aquí es donde está la concepción errónea que es necesario corregir**; como
@@
 
-Ahora bien: Hay que recordar que **el objetivo de proponer una base latente
-$\mathfrak{B}$ es** que las coordenadas de los vectores de rasgo latente
+Ahora bien: Hay que recordar que **el objetivo de proponer una base latente**
+$\mathfrak{B}$ es que las coordenadas de los vectores de rasgo latente
 transformadas a una base ortogonal, tengan una distribución independiente, es
@@
 
-Tal y como indica Mario
-([ll. 91-93](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L91-L93)),
+Tal y como indica Mario ([ll.
+91-93](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L91-L93)),
 "En la base Canónica $\mathfrak{E}$ las coordenadas se obtienen mediante la
@@
 
-Vamos a hacer el cambio de base, tal y como viene en el artículo, y
-[como hace Mario en este notebook](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L96-L103).
-(Aunque utilizao un código distinto, el
-resultado que obtengo es el mismo.)
+Vamos a hacer el cambio de base, tal y como viene en el artículo, y [como hace
+Mario en este
+notebook](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L96-L103).
+(Aunque utilizao un código distinto, el resultado que obtengo es el mismo.)
 
 ```{r transform-coordinates}
-coords_C <- P %*% t(coord_B) |> t()
+coords_C <- P %*% t(coords_B) |> t()
 
@@
 siendo $\alpha$ el ángulo entre los dos vectores de la base $\mathfrak{B}$.
- Ahora bien, esta matriz también es igual a
- $\mathbf{P}^T\mathbf{P} = \mathbf{M}$ (no pongo la justificación matemática
- porque no creo que sea necesario, pero podemos verlo en el ejemplo):
+Ahora bien, esta matriz también es igual a $\mathbf{P}^T\mathbf{P} = \mathbf{M}$
+(no pongo la justificación matemática porque no creo que sea necesario, pero
+podemos verlo en el ejemplo):
 
@@
 simulado los vectores de rasgo latente (originales) con una correlación igual al
-coseno del ángulo entre los vectores de la base $\mathfrak{B}$**, y que hacemos
+coseno del ángulo entre los vectores de la base** $\mathfrak{B}$, y que hacemos
 esto porque **consideramos que dicho ángulo representa la correlación entre los

### 2024-05-30 notebooks/Basis_change_explanation.qmd
Clarifies trasnformation to the canonical basis

@@
 format: html
-editor: visual
+editor:
+  markdown:
+    wrap: 80
+  mode: source
 ---
@@
 
+::: callout-note
+Mario
+[comenta](https://github.com/DaniMori/mirt-parameters/pull/10/files#r1620807302),
+acertadamente, que "el objetivo de proponer una base latente $\mathfrak{B}$ es
+explicar los verdaderos constructos del test, no una base para que al
+transformarse los rasgos sean independientes".
+
+No obstante, para dar una interpretación "sustantiva" a los ejes, nos interesa
+que las $\mathbf{\theta}$s transformadas a una base ortogonal estén "no
+correlacionadas" y, más generalmente, que en una base ortonormal (la canónica,
+sin perder generalidad) tengan covarianza identidad. Esto interesa porque cuando
+se hace un procedimiento de TRI multidimensional, se estiman $\mathbf{\theta}$s
+con covarianza identidad que (de manera implícita) se asume que se representan
+en la base canónica. Una vez que se aplica a ellas una (mal llamada) "rotación
+oblicua", es cuando se obtienen las $\mathbf{\theta}$s en ejes "sustantivos",
+las cuales estarían representadas en la base oblicua a la que llamamos aquí
+$\mathfrak{B}$.
+
+En base a esto, uno pensaría que los vectores de la base $\mathfrak{B}$ son un
+fiel reflejo de la estructura de covarianzas de las $\mathbf{\theta}$s en dichos
+ejes sustantivos. Sin embargo, **esto NO es así**, y vermos por qué. De hecho,
+**esta es la concepción errónea que intento explicar en este notebook**.
+:::
+
 # Transformación a la base canónica

### 2024-05-30 notebooks/Basis_change_explanation.qmd
Corrects output error in the correlation matrix
Explains (wrong) meaning of covariance matrix

@@
 
-`r cov_coords_C |> round(3)`
+`r cov_coords_C`
 
@@
 
-Lo cual (en general) NO equivale a una matriz identidad.
+Lo cual (en general) NO equivale a una matriz identidad, y lo que **buscamos**
+es **que esto se cumpla para cualquier caso**.
+
+::: callout-important
+Si simulamos las coordenadas originales con
+$\mathbf{\Sigma} = \mathbf{M} = \mathbf{P}^T \mathbf{P}$, **las coordenadas
+transformadas NO van a tener una matriz de covarianzas identidad**.
+:::
 
@@
 ::: callout-warning
-El coseno del ángulo entre los vectores de la base $\mathfrak{B}$ NO representa
-la correlación entre los constructos sustantivos que nos interesa representar.
+El coseno del ángulo entre los vectores de la base $\mathfrak{B}$ **NO
+representa la correlación entre los constructos sustantivos que nos interesa
+representar**.
 :::

### 2024-05-30 notebooks/Basis_change_explanation.qmd
Versions "norm invariance" section

@@
 mostraré por qué NO responde al objetivo del artículo. Lo que sí es incorrecto
-es asumir que de esta manera "su correlación \[es\] el angulo entre los factores
+es asumir que de esta manera "su correlación \[es\] el ángulo entre los factores
 oblicuos", lo cual mostraré a continuación. Por último, explicaré con dos
-ejemplos distintos cómo debería hacerse de manera esta simulación para que sí
-responda a nuestro objetivo.
+ejemplos distintos cómo debería hacerse esta simulación para que sí responda a
+nuestro objetivo.
 
@@
 Las coordenadas de los vectores de rasgo latente, transformadas a la base
-canónica, deben tener una matriz de covarianzas identidad.
+canónica, **deben tener una matriz de covarianzas identidad**.
 :::
@@
 :::
+
+## Invarianza de la norma
+
+A este respecto coincido con Mario (no hay ninguna objeción). En la base
+original, $\mathbf{M}$ es la matriz de Gramm del producto interno, mientras que
+en la base canónica éste es la matriz identidad. Se cumple por tanto la
+invarianza de la norma, mostrando que la transformación a la base canónica es
+correcta.
+
+```{r norm-invariance-example}
+theta_C <- coord_C[17L, ]
+
+theta_B <- solve(P, theta_C)
+
+norm_B <- t(theta_B) %*% M %*% theta_B |> sqrt() |> drop()
+norm_C <- t(theta_C) %*%       theta_C |> sqrt() |> drop()
+norm_B == norm_C # Mismo resultado con precisión numérica exacta
+```
+
+No obstante, esto sólo muestra que la transformación es correcta. No implica
+nada respecto a la simulación de las coordenadas o su interpretación sustantiva.

### 2024-05-30 notebooks/Basis_change_explanation.qmd
Versions "deduction of covariance" section
Covariance is deduced from assuming that the transformed coordinates have identity covariance

@@
 nada respecto a la simulación de las coordenadas o su interpretación sustantiva.
+
+## Deducción de la matriz de covarianzas a partir de las coordenadas transformadas
+
+En las [líneas 149 a
+173](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L91-L93)),
+Mario deduce la matriz de covarianzas de las coordenadas originales a partir del
+supuesto de que las coordenadas transformadas a la base canónica tienen matriz
+de covarianzas identidad. A pesar de esta deducción, muestra escepticismo
+respecto de este resultado (["Es claro que estas P se pueden hallar pero pueden
+no representar el
+fenómeno"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L171))
+
+No obstante, hay un error a este respecto. Mario indica que ["De los infinitas
+bases podemos tamar dos que llevan en su trnsformaci’on a la matriz identidad e
+los ejes
+coordenados"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L150").
+Esto parece venir a decir que, "de las infinitas matrices que pueden solucionar
+$\mathbf{M} = \mathbf{P}^T \mathbf{P}$, existe una solución que hace que
+$\mathbf{\Sigma}^{-1} = \mathbf{P}^T \mathbf{P}$. Sin embargo, esto no es
+correcto. Dado que ya hemos simulado los datos con
+$\mathbf{\Sigma} = \mathbf{P}^T \mathbf{P}$, la única posibilidad de que
+$\mathbf{\Sigma}^{-1} = \mathbf{P}^T \mathbf{P}$ es si
+$\mathbf{\Sigma} = \mathbf{\Sigma}^{-1}$. Pero para esto, lo que tiene que
+ocurrir es que $\mathbf{\Sigma} = \mathbf{\Sigma}^{-1} = \mathbf{I}$, es decir,
+que $\mathbf{\Sigma}$ sea una matriz identidad.
+
+Por lo tanto, como podemos ver, en el caso general NO se cumple que, si la
+covarianza de las coordenadas transformadas es $\mathbf{I}$, la varianza de las
+coordenadas originales es $\mathbf{M}$. Sin embargo, sí se cumple siempre que,
+si la covarianza de las coordenadas transformadas es $\mathbf{I}$, la covarianza
+de las coordenadas originales es $\mathbf{M}^{-1}$.
+
+::: callout-important
+Si la covarianza de las coordenadas transformadas es $\mathbf{I}$, **la
+covarianza de las coordenadas originales necesariamente ha de ser
+$\mathbf{M}^{-1}$**.
+:::

### 2024-05-31 notebooks/Basis_change_explanation.qmd
Corrects errata and a possible misconception

@@
 b_2 <- b_2 / len(b_2)
-
 ```
@@
 ```{r norm-invariance-example}
-theta_C <- coord_C[17L, ]
+theta_C <- coords_C[17L, ]
 
@@
 En las [líneas 149 a
-173](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L91-L93)),
+173](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L149-L173)),
 Mario deduce la matriz de covarianzas de las coordenadas originales a partir del
@@
 $\mathbf{\Sigma} = \mathbf{\Sigma}^{-1}$. Pero para esto, lo que tiene que
-ocurrir es que $\mathbf{\Sigma} = \mathbf{\Sigma}^{-1} = \mathbf{I}$, es decir,
-que $\mathbf{\Sigma}$ sea una matriz identidad.
+ocurrir es que $\mathbf{\Sigma} = \mathbf{\Sigma}^{-1}$. Esto, hasta donde yo
+sé (y quizá me equivoque[1]) sólo puede ocurrir si $\mathbf{\Sigma} = \mathbf{I}$;
+es decir, que $\mathbf{\Sigma}$ sea una matriz identidad.
+
+[1]: Aunque no sea cierto esto y sólo sea un caso particular, está claro que la
+restricción $\mathbf{\Sigma} = \mathbf{\Sigma}^{-1}$ implica que estas
+deducciones no podrían aplicarse al caso general.
 
 Por lo tanto, como podemos ver, en el caso general NO se cumple que, si la
-covarianza de las coordenadas transformadas es $\mathbf{I}$, la varianza de las
-coordenadas originales es $\mathbf{M}$. Sin embargo, sí se cumple siempre que,
-si la covarianza de las coordenadas transformadas es $\mathbf{I}$, la covarianza
-de las coordenadas originales es $\mathbf{M}^{-1}$.
+covarianza de las coordenadas transformadas es igual a su inversa, la varianza
+de las coordenadas originales es $\mathbf{M}$. Sin embargo, sí se cumple siempre
+que, si la covarianza de las coordenadas transformadas es $\mathbf{I}$, la
+covarianza de las coordenadas originales es $\mathbf{M}^{-1}$. El único problema
+sería, hasta este punto, añadir la restricción de que
+$\mathbf{\Sigma} = \mathbf{M}$ y aplicar esto en la siimulación de los datos.
 
@@
 Si la covarianza de las coordenadas transformadas es $\mathbf{I}$, **la
-covarianza de las coordenadas originales necesariamente ha de ser
-$\mathbf{M}^{-1}$**.
+covarianza de las coordenadas originales necesariamente ha de ser igual a
+$\mathbf{M}^{-1}$ y no igual a $\mathbf{M}$**.
 :::

### 2024-05-31 notebooks/Basis_change_explanation.qmd
Adds "Spectral decomposition" section

@@
 :::
+
+## Descomposición espectral de la matriz de covarianzas
+
+En las [líneas 176 a
+233](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L176-L273)),
+Mario obtiene la descomposición espectral de la matriz de covarianzas en el
+espacio original (a la cual Mario llama $\Sigma_B$).
+
+Las deducciones que hace Mario de la descomposición espectral me parecen
+correctas. Además, los cálculos con el ejemplo verifican que se cumplen sus
+deducciones:
+
+```{r eigendecomposition-sample-covariance}
+eigendecomposition <- cov_coords_B |> cov2cor() |> eigen()
+
+lambda_root <- eigendecomposition$values^(-1) |> sqrt() |> diag()
+Q           <- eigendecomposition$vectors
+Q_inv       <- Q |> solve()
+
+P_ed <- Q %*% lambda_root %*% Q_inv
+
+coords_VP <- P_ed %*% t(coords_B) |> t()
+
+coords_VP |> cov()
+coords_VP |> cor()
+```
+
+Sólo tengo una objeción, y es que utiliza la matriz de correlación
+(no covarianzas, cuidado) muestral en sus deducciones en lugar de la matriz de
+covarianzas poblacional:
+
+```{r population-covariance}
+# Covarianza poblacional:
+P_ed %*% cov_matrix %*% t(P_ed)
+```
+
+Realizando los mismos cálculos con la matriz de covarianzas poblacional[2], el
+resultado es el siguiente:
+
+[2]: En este caso, matriz de covarianzas y de correlaciones son iguales, puesto
+que ambas covarianzas son iguales a 1.
+
+```{r eigendecomposition-population-covariance}
+eigendecomposition_pop <- cov_matrix |> cov2cor() |> eigen()
+
+lambda_root_pop <- eigendecomposition_pop$values^(-1) |> sqrt() |> diag()
+Q_pop           <- eigendecomposition_pop$vectors
+Q_inv_pop       <- Q |> solve()
+
+P_ed_pop <- Q_pop %*% lambda_root_pop %*% Q_inv_pop
+
+coords_VP <- P_ed_pop %*% t(coords_B) |> t()
+
+coords_VP |> cov()
+coords_VP |> cor()
+
+# Covarianza poblacional:
+P_ed_pop %*% cov_matrix %*% t(P_ed_pop)
+```
+
+Esto muestra que, efectivamente, para una matriz de transformación $\mathbf{P}$
+que cumple que $\mathbf{P}^T \mathbf{P} = \mathbf{M} = \mathbf{\Sigma}^{-1}$,
+las coordenadas transformadas $\mathbf{P} \mathbf{\theta}$ tendrán matriz de
+covarianza identidad. (Además, en este caso se cumple que
+$\mathbf{P}^T = \mathbf{P}$, pero esto es irrelevante para la demostración ya
+que cualquier matriz que cumpla la primera condición cumplirá la segunda).
+
+Sin embargo, hay algo **muy importante** que considerar en este caso: Esta
+descomposición espectral de $\mathbf{\Sigma}^{-1}$ da una matriz de
+transformación que, como en el caso anterior, sería la matriz de vectores de la
+base original (oblicua) expresados en la base transformada (ortonormal).
+Recordemos que al principio se había propuesto la existencia de dos vectores de
+una base oblicua cualquiera $\mathfrak{B}$, y que se había simulado una
+distribución bivariada con una correlación igual al ángulo que formaban los dos
+vectores de esa base. Veamos ahora cuál es el ángulo que forman los dos vectores
+de $\mathfrak{B}$, es decir, los que forman la nueva matriz $\mathbf{P}$ que
+hemos obtenido por descomposición espectral:
+
+```{r angle-eigendecomposition-vectors}
+angle_ed_pop <- angle(P_ed_pop[, 1], P_ed_pop[, 2])
+cos_ed_pop   <- cosine(P_ed_pop[, 1], P_ed_pop[, 2])
+
+angle_ed_pop
+cos_ed_pop
+```
+
+Tal y como vemos, estos dos vectores no forman un ángulo tal que ["la
+correlacion entre los rasgos de los sujetos es la correlacion entre los
+ejes"](https://github.com/DaniMori/mirt-parameters/blob/9f0549415219a90d8c45fbce674eeb6d9b0ea425/notebooks/Representacion.Rmd#L69).
+
+De hecho, pensemos por un momento qué habría pasado si hubiéramos seguido los
+mismos pasos para hacer la simulación que hemos hecho al principio. En lugar de
+simular una distribución cuya correlación es de `r cos_b1_b2 |> round(3)`,
+habríamos simulado una distribución con una correlación de
+`r cos_ed_pop |> round(3)`. (Que estas dos correlaciones sean opuestas no es
+casualidad, como veremos.)
+
+El mismo razonamiento se puede aplicar a la descomposición de Cholesky, pero no
+seguiré con ello porque lo veo innecesario lo que intento explicar aquí: Que la
+**base del espacio latente NO explica realmente el fenómeno**.

### 2024-05-31 notebooks/Basis_change_explanation.qmd
Adds section "explanation of the missconception"

@@
 es asumir que de esta manera "su correlación \[es\] el ángulo entre los factores
-oblicuos", lo cual mostraré a continuación. Por último, explicaré con dos
+oblicuos", lo cual mostraré a continuación. Por último, explicaré con tres
 ejemplos distintos cómo debería hacerse esta simulación para que sí responda a
@@
 ```{r angle-eigendecomposition-vectors}
-angle_ed_pop <- angle(P_ed_pop[, 1], P_ed_pop[, 2])
+angle_ed_pop <- angle( P_ed_pop[, 1], P_ed_pop[, 2])
 cos_ed_pop   <- cosine(P_ed_pop[, 1], P_ed_pop[, 2])
@@
 **base del espacio latente NO explica realmente el fenómeno**.
+
+# Explicación de la concepción errónea
+
+Tal y como hemos podido ver, **el problema parte de considerar que el coseno
+entre los vectores de la base del espacio latente representa la correlación de
+los constructos latentes**. Ya hemos visto, sin embargo, que **esto no se cumple
+si asumimos que la base canónica representa una distribución independiente**.
+
+No obstante, hemos visto que, aceptando este último supuesto, se cumple que
+$\mathbf{M} = \mathbf{\Sigma}^{-1}$. Veamos ahora cuál es la interpretación
+adecuada del fenómeno.
+
+Sabemos que, si los vectores de la base $\mathfrak{B}$ forman la matriz de
+transformación $\mathbf{P}$, tal que $\mathbf{P}^T \mathbf{P} = \mathbf{M}$,
+para mantener la invarianza del modelo y que los ítems se transformen también a
+un espacio ortonormal, se cumple que los vectores de $\mathfrak{B}*$ forman la
+matriz de transformación $(\mathbf{P}^{-1})^T$, siendo $\mathfrak{B}*$ la base
+original (oblicua) del espacio test.
+
+Pues bien, si esto es así, se cumple que **el ángulo entre los vectores de la
+base $\mathfrak{B}*$ sí representa la correlación entre los constructos latentes
+sustantivos**. Es decir, que no es la base del espacio latente base "la que
+explica realmente el fenómeno", sino la base del espacio test.
+
+::: callout-important
+El estructura latente de los constructos que dan lugar al fenómeno sustantivo
+**están representados por la base del espacio test.**
+:::
+
+## Ejemplo con la base obtenida por descomposición espectral
+
+Vamos a ver lo anteiror con un ejemplo, usando para ello la base obtenida
+mediante descomposición espectral:
+
+```{r eigendecomposition-test-space-basis}
+# Obtenemos la base del espacio test como la inversa traspuesta de la base del
+#   espacio latente:
+P_inv_transp <- P_ed_pop |> solve() |> t()
+
+# Ángulo entre los vectores:
+angle( P_inv_transp[, 1], P_inv_transp[, 2])
+cosine(P_inv_transp[, 1], P_inv_transp[, 2])
+
+# Cálculo de M^{-1}:
+t(P_inv_transp) %*% P_inv_transp
+cov_matrix
+
+# Iguales excepto por la precisión numérica:
+(t(P_inv_transp) %*% P_inv_transp) |> all.equal(cov_matrix)
+```
+
+## Ejemplo obtenido con la base propuesta inicialmente
+
+Veamos ahora qué ocurriría si utilizamos la base propuesta originalmente para
+llevar a cabo la simulación. Para ello, en lugar de obtener una matriz de
+covarianzas usando el coseno de los ángulos de esa base, calculamos primero la
+matriz de transformación del espacio test, y utilizamos dicha matriz para
+obtener la matriz de covarianzas.
+
+La base original da lugar a la siguiente matriz de transformación:
+
+```{r original-test-space-basis}
+# Matriz de transformación del espacio test == vectores de la base del espacio
+#   test como columnas:
+P_inv_transp <- P |> solve() |> t()
+
+cov_matrix <- t(P_inv_transp) %*% P_inv_transp
+cov_matrix # Matriz de covarianzas
+cov_matrix |> cov2cor() # Matriz de correlaciones de la distribución
+
+coords_B <- rmvnorm(3000, sigma = cov_matrix)
+
+cov_coords_B <- cov(coords_B)
+cor_coords_B <- cor(coords_B)[2, 1]
+
+cov_coords_B
+```
+
+La correlación muestral de las coordenadas simuladas ahora es
+`r cor_coords_B |> round(3)`.
+
+También podemos transformar estas coordenadas en el espacio latente, tal y como
+hacíamos antes, y veremos que tanto la covarianza poblacional como la muestral
+se aproximan a una matriz identidad. Para ello, lo que hacemos es
+**pre-multiplicar por la matriz de transformación del espacio latente, no la del
+espacio test**:
+
+```{r transform-coordinates-test-space-basis}
+coords_C <- P %*% t(coords_B) |> t()
+
+cov(coords_C) # Covarianza muestral
+cor(coords_C) # Correlación muestral
+
+# Covarianza poblacional:
+P %*% cov_matrix %*% t(P) # Matriz identidad, salvo por precisión numérica
+```
+
+## Ejemplo asumiendo la base y distribución del ejemplo inicial
+
+Si queremos asumir que la base del ejemplo original de Mario es la que
+representa el fenómeno de estudio, y que dicha base da lugar a la distribución
+de los rasgos latentes que representan dicho fenómeno, lo que tenemos que hacer
+es asumir que dicha base es la del espacio test, y no la del espacio latente.
+
+Es decir, los vectores que propusimos originalmente forman la base
+$\mathfrak{B}*$, y a partir de ellos se obtiene la matriz de transformación en
+el espacio test, a la que llamamos $(\mathbf{P}^{-1})^T$.
+
+```{r original-basis-distribution}
+P_inv_transp <- cbind(b_1, b_2)
+
+# Matriz de transformación del espacio latente:
+P <- P_inv_transp |> solve() |> t()
+P
+
+cov_matrix <- t(P_inv_transp) %*% P_inv_transp
+cov_matrix # MAtriz de covarianzas de la distribución (== correlaciones)
+
+coords_B <- rmvnorm(3000, sigma = cov_matrix)
+
+# Covarianza y correlación de la muestra:
+cov(coords_B)
+cor(coords_B)
+
+# Transformación a base ortonormal (== coordenadas "independientes"):
+coords_C <- P %*% t(coords_B) |> t()
+
+cov(coords_C) # Covarianza muestral
+cor(coords_C) # Correlación muestral
+
+# Covarianza poblacional:
+P %*% cov_matrix %*% t(P) # Matriz identidad, salvo por precisión numérica
+```

### 2024-05-31 notebooks/Basis_change_explanation.qmd
Adds section "relationship with factor analysis"

@@
 ```
+
+# Relación entre esta explicación y el análisis factorial
+
+En el análisis factorial (exploratorio), originalmente se obtiene (usualmente)
+una solución sin rotar, y posteriormente se hace una rotación (oblicua) para
+obtener los ejes que mejor representan el fenómeno sustantivo. Sin embargo,
+nótese que **lo que se obtiene son coordenadas rotadas de los pesos factoriales
+de las variables**. Esto sería **equivalente a rotar los ejes en el espacio test
+y no en el espacio latente** (en realidad, sería lo que Harman denomina factores
+primarios, en contraposición con los ejes de referencia, que representarían el
+espacio latente).

### 2024-05-31 notebooks/Basis_change_explanation.qmd
Adds section "sustantive interpretation of axes"

@@
 espacio latente).
+
+# Interpretación sustantiva de los ejes
+
+Normalmente, los ejes (en análisis factorial y, consecuentemente, en TRI
+multidimensional, tal y como hace `{mirt}`) se rotan para obtener una solución
+sustantiva que satisfaga los objetivos de la investigación. Para ello, se tiende
+a buscar la llamada "estructura simple" de los ítems. No obstante, es importante
+tener en cuenta que, tanto lo que Reckase propone como lo que nosotros
+intentamos hacer, es **aplicar estos resultados a ítems multidimensionales,
+independientemente de su complejidad**. De hecho, en mi opinión, lo más
+interesante son los ítems complejos, y no los ítems simples, ya que en éstos, al
+tener una única discriminación no nula, la solución es trivial. Por este motivo,
+creo que no hay que centrarse demasiado en las relaciones con el análisis
+factorial. Hay que asumir que nuestros resultados, por muchos y muy variados
+motivos, van más allá de esa equiparación. Mi idea es ceñirme a lo puramente
+teórico, tal y como hace Reckase, pero esto se podría debatir por supuesto.

### 2024-06-04 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds missing "math" delimiters
Omits package building wanings (due to R version)
Corrects statement about axes interpretation
The interpretation of the axes in the latent space is not that their angles represent the correlations (that happens in the test space). Therefore, that sentence in the introduction is changed to a more neutral one, that introduces this topic, which will be clarified later on in the results and discussion.

--- 2024-05-28 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-A latent space where the traits are correlated is easier to understand as a
-geometrical space with non-orthogonal axes,
-often referred to as *general Cartesian coordinates* [@harman_modern_1970].
-Such a space has the advantage that the cosines of the angles between two axes
-can be interpreted as the correlation of the corresponding latent dimensions.
-<!--# Reference needed? -->
-<!-- Problem here: The latent space is not defined like that, according to
-the results of section "Linear algebra of non-orthogonal coordinates". -->
+The representation of a latent space where the traits are correlated is more
+accurate on a geometrical coordinate system with non-orthogonal axes and
+non-standard units, often referred to as
+*general Cartesian coordinates* [@harman_modern_1970].
+Such a space has the advantage that the basis vectors and the axes they yield
+have a meaningful interpretation (albeit counterintuitive, as we will see)
+in a multivariate statistical sense.
 The general case of MIRT model estimation implies estimating possibly

### 2024-06-04 output/paper_draft/paper_M2PLM_parameters.Rmd
Updates "covariance-based" version of the indices

--- 2024-05-28 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-Applying Proposition 2, if `r ANGLE_VEC_TS_ITEM` is
-the vector of direction angles of `r DISCR_VECTOR`,
+Applying Proposition 2, if $`r ANGLE_VEC_TS_ITEM`$ is
+the vector of direction angles of $`r DISCR_VECTOR`$,
 in the test space $`r TEST_SPACE`$ we get
@@
 
-On the other hand, we can account for the covariance structure of
-the latent space letting $`r LATENT_SPACE`$ be its geometrical representation.
-Thus, we must define $`r LS_BASIS`$ in such a way that
-
-\begin{equation}
-  `r TRAIT_MV_STD_EQ`,
-\end{equation}
-
-where $`r COV_MATRIX_STD`$ is a diagonal matrix.
-Now we may consider two different possibilities:
-We may define $`r TRANSFORM_MATRIX`$ such that $`r INNER_PROD_CORR_COND`$.
-This results in a *correlation-based* version of the indices
-(note that, in this case, $`r DIAG_MATRIX_INNER_PROD_INV`$ is
-an identity matrix):
+On the other hand, we can let the geometrical representation of
+the latent space $`r LATENT_SPACE`$ account for its covariance structure,
+thus giving $`r LS_BASIS`$ a statistical meaning
+in the context of our MIRT modeling approach.
+To do that, we may assume that $`r TRAIT_COORDS`$ is
+a random vector distributed with covariance $`r COV_MATRIX_EQ`$,
+being $`r COV_MATRIX`$ an $`r N_DIMS`$-dimensional positive definite matrix,
+$`r CORR_MATRIX`$ its corresponding correlation matrix
+(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
+with $`r VAR_COV_EQ`$ (i.e., the variances).
+
+To make the representation of $`r LATENT_SPACE`$ meaningful,
+the axes in $`r LS_BASIS`$ must have
+an interpretation in terms of the latent space structure.
+Although we do not know, in principle, what
+the interpretation of a latent space basis should be, we may assume that
+it must be somewhat related to the covariance structure.
+Therefore, an orthonormal basis should represent independent coordinates;
+formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
+$`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$.
+Applying Equation \@ref(eq:change-basis), $`r COV_MATRIX_TRANSF_EQ`$, and thus,
 
@@
   \begin{split}
-    `r MDISC_CORR_PARAM_EQ` \\
-    `r MIL_CORR_PARAM_EQ`.
+    `r COV_MATRIX_TRANSF_ID_EQ` \\
+    `r COV_MATRIX_RESULT_EQ`.
   \end{split}
@@
 
-However, we may also define $`r TRANSFORM_MATRIX`$ such that
-$`r INNER_PROD_COV_COND`$, so
-
-\begin{equation}
-  `r TRAIT_MV_STD_NORM_EQ`.
-\end{equation}
-
-In this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in the
-following *covariance-based* version of the indices:
+Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
+$`r INNER_PROD_COV_COND`$;
+in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
+*covariance-based* version of the indices:
 
@@
 ```{r libraries}
+#| warning: false
 ```

### 2024-06-04 R/Formulae.R

@@ -49,29 +49,11 @@ LS_BASIS_EQ        <- latex_eq(LS_BASIS, LS_BASIS_SET)
 TRAIT_VECTOR    <- latex_bf("\\uptheta")
 TRAIT_VEC_IN_LS <- latex_in(TRAIT_VECTOR, LATENT_SPACE)
 
-# TODO: Move forward this block:
-MEAN_VECTOR     <- latex_bf("\\upmu")
-COV_MATRIX      <- latex_bf("\\upSigma")
-NORMAL_DISTR    <- latex_cal('N')
-MV_DISTRIBUTION <- latex(
-  NORMAL_DISTR,
-  latex_parentheses("$MEAN_VECTOR$, $COV_MATRIX$")
-)
-TRAIT_MV_DEF    <- latex_sim(TRAIT_VECTOR, MV_DISTRIBUTION)
-
-# Covariance matrix definition:
-CORR_MATRIX <- latex_bf('R')
-SD_MATRIX   <- latex_bf('S')
-DIM_INDEX   <- latex('k')
-SD_ELEMENT  <- latex_sub('s', "$DIM_INDEX$$DIM_INDEX$")
-VAR_ELEMENT <- latex_squared(SD_ELEMENT)
-COV_ELEMENT <- latex_squared(latex_sub("\\sigma", "$DIM_INDEX$$DIM_INDEX$"))
-VAR_COV_EQ  <- latex_eq(VAR_ELEMENT, COV_ELEMENT)
-
 ### M2PL model formulation ----
 
 # Model parameters:
 ITEM_INDEX          <- latex('i')
+DIM_INDEX           <- latex('k')
 INTERCEPT_PARAM     <- latex_sub('d', ITEM_INDEX)
 DISCR_VECTOR_ANY    <- latex_bf('a')
 DISCR_VECTOR        <- latex_sub(DISCR_VECTOR_ANY, ITEM_INDEX)
@@ -918,56 +900,40 @@ MIL_AG_PARAM_EQ         <- latex_def(
   latex_curlybraces("$DISTANCE_AG_DEF$, $DIR_COS_ITEM_VEC_AG_DEF$")
 )
 
-# Transformed latent vector as orthogonal:
-MEAN_VECTOR_STD <- latex_prime(MEAN_VECTOR)
-COV_MATRIX_STD  <- latex_prime(COV_MATRIX)
-MV_DISTR_STD    <- latex(
[…]

### 2024-06-11 notebooks/Corr_based_parameters.qmd
Versions correlation-based parameters notebook

---
title: "Parámetros basados en la matriz de correlaciones"
format: docx
editor: source
---

# Introducción

Tal y como Mario ha demostrado anteriormente, la versión "basada en la
correlación" de los parámetros multidimensionales no sería correcta, pues no
cumple la premisa de que "la matriz de covarianzas resultante" es diagonal.

No obstante, viendo las derivaciones matemáticas que ha hecho Mario también
respecto a cómo supuestamente obtener la base "verdadera" del espacio test
(en base a la descomposición espectral), creo que se puede encontrar una base
en el espacio test que cumpla esa relación.

# Deducción

## Supuestos iniciales

Como hemos hecho hasta ahora, asumimos un espacio $\mathbf{\Theta}$  con base
$\mathcal{B}$ en el que se representa el vector de rasgo latente
$\mathbf{\theta}$. Este vector es una variable aleatoria $n$-variada que se
distribuye con covarianza $\mathbf{\Sigma} = \mathbf{S} \mathbf{R} \mathbf{S}$,
siendo $\mathbf{R}$ una matriz de correlaciones, semidefinida positiva y con 1's
en la diagonal, y $\mathbf{S}$ una matriz diagonal de desviaciones típicas, tal
que $\sigma^2_{ii} = s^2_{ii}$, para todo $i$ entre 1 y $n$ (i.e., los elementos
diagonales de $\mathbf{\Sigma}$ y $\mathbf{S}^2$ son iguales).

## Premisa

Es posible encontrar una representación del espacio latente en una base
ortogonal $\mathcal{U}$ tal que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de
covarianzas de $\mathbf{\theta}$ representado en $\mathcal{U}$,
$\mathbf{\theta}^\mathcal{U}$) sea una matriz diagonal.

## Demostración

Sea $\mathbf{P}$ la matriz de transformación de la base $\mathcal{B}$ a la base
$\mathcal{U}$. Como ya sabemos,
$\mathbf{P} = [\mathbf{b}_1^\mathcal{U}, ..., \mathbf{b}_n^\mathcal{U}]$, es
decir, los vectores de la base $\mathcal{B}$ representados en la base
$\mathcal{U}$. Las coordenadas de $\mathbf{\theta}$, exprsadas en la base
$\mathcal{U}$, vienen dadas por la siguiente relación:

$$
\mathbf{\theta}^\mathcal{U} = \mathbf{P} \mathbf{\theta}
$$

La matriz de covarianzas $\mathbf{\Sigma}^\mathcal{U}$ de
$\mathbf{\theta}^\mathcal{U}$ cumple

$$
\mathbf{\Sigma}^\mathcal{U} = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T =
  \mathbf{P} \mathbf{S} \mathbf{R} \mathbf{S} \mathbf{P}^T
$$

Supongamos que $\mathbf{P}$ es una raíz cuadrada de $\mathbf{R}^{-1}$; es
decir, $\mathbf{R}^{-1} = \mathbf{P}^T \mathbf{P}$.

$$
\mathbf{\Sigma}^\mathcal{U} =
  \mathbf{P} \mathbf{S} (\mathbf{P}^T \mathbf{P})^{-1} \mathbf{S} \mathbf{P}^T =
  \mathbf{P} \mathbf{S} \mathbf{P}^{-1} (\mathbf{P}^{-1})^T \mathbf{S} \mathbf{P}^T
$$

Si $\mathbf{S}$ conmuta con $\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$, entonces
la ecuación anterior puede expresarse como

$$
\mathbf{\Sigma}^\mathcal{U} =
  \mathbf{P} \mathbf{P}^{-1} \mathbf{S} \mathbf{S} (\mathbf{P}^{-1})^T \mathbf{P}^T =
  \mathbf{S}^2
$$

### Contraejemplo

Lo anterior en general no se cumple, por lo que $\mathbf{\Sigma}^\mathcal{U}$ no
es, en general, una matriz diagonal bajo estos supuestos.

Un contraejemplo de esto es la solución con la base del ejemplo de
representación gráfica del borrador y varianzas desiguales. Supongamos

$$
\mathbf{S} = diag(\sigma_1, \sigma_2) \\
(\mathbf{P}^{-1})^T = \begin{pmatrix}
    1 & \rho \\
    0 & \sqrt{1 - \rho^2}
  \end{pmatrix}.
$$

Bajo estos supuestos,

$$
\mathbf{R} = \mathbf{P}^{-1} (\mathbf{P}^{-1})^T = \begin{pmatrix}
       1 &              0 \\
    \rho & \sqrt{1 - \rho^2}
  \end{pmatrix}
  \begin{pmatrix}
    1 & \rho \\
    0 & \sqrt{1 - \rho^2}
  \end{pmatrix} =
  \begin{pmatrix}
       1 & \rho \\
    \rho &    1
  \end{pmatrix}
$$

$$
\mathbf{\Sigma} = diag(\sigma_1, \sigma_2)
  \begin{pmatrix}
       1 & \rho \\
    \rho &    1
  \end{pmatrix}
  diag(\sigma_1, \sigma_2) =
  \begin{pmatrix}
    \sigma_1^2             & \sigma_1 \sigma_2 \rho \\
    \sigma_1 \sigma_2 \rho & \sigma_2^2
  \end{pmatrix}.
$$

Veamos que $\mathbf{S}$ No conmuta con $\mathbf{P}^{-1}$ ni con
$(\mathbf{P}^{-1})^T$:

$$
\mathbf{S} (\mathbf{P}^{-1})^T = diag(\sigma_1, \sigma_2)
  \begin{pmatrix}
    1 & \rho \\
    0 & \sqrt{1 - \rho^2}
  \end{pmatrix} =
  \begin{pmatrix}
    \sigma_1 & \sigma_1 \rho \\
    0        & \sigma_2 \sqrt{1 - \rho^2}
  \end{pmatrix} \\

(\mathbf{P}^{-1})^T \mathbf{S} = \begin{pmatrix}
    1 & \rho \\
    0 & \sqrt{1 - \rho^2}
  \end{pmatrix}
  diag(\sigma_1, \sigma_2) =
  \begin{pmatrix}
    \sigma_1 & \sigma_2 \rho \\
    0        & \sigma_2 \sqrt{1 - \rho^2}
  \end{pmatrix} \\

\mathbf{S} \mathbf{P}^{-1} = diag(\sigma_1, \sigma_2)
  \begin{pmatrix}
       1 &              0 \\
    \rho & \sqrt{1 - \rho^2}
  \end{pmatrix} =
  \begin{pmatrix}
    \sigma_1      &              0 \\
    \sigma_2 \rho & \sigma_2 \sqrt{1 - \rho^2}
  \end{pmatrix} \\

\mathbf{P}^{-1} \mathbf{S} = \begin{pmatrix}
       1 &              0 \\
    \rho & \sqrt{1 - \rho^2}
  \end{pmatrix}
  diag(\sigma_1, \sigma_2) =
  \begin{pmatrix}
    \sigma_1      &              0 \\
    \sigma_1 \rho & \sigma_2 \sqrt{1 - \rho^2}
  \end{pmatrix}
$$

Estas ecuaciones muestran que la conmutatividad de $\mathbf{S}$ con
$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$ no se cumple, en general (salvo que
$\sigma_1 = \sigma_2$).

### Condiciones adicionales

A pesar de lo anterior, $\mathbf{S}$ conmuta con cualquier otra matriz que sea
simultáneamente diagonalizable con ella. En el caso de matrices simétricas, es
decir, si $\mathbf{P}^{-1} = (\mathbf{P}^{-1})^T$, sabemos que se cumple la
conmutatividad de $\mathbf{S}$ con $\mathbf{P}^{-1}$ (y por lo tanto también con
$(\mathbf{P}^{-1})^T$), si
$\mathbf{S} \mathbf{P}^{-1} = \mathbf{P}^{-1} \mathbf{S}$ es también simétrica.

Intentamos obtener a continuación una solución que haga $\mathbf{P}^{-1}$
simétrica, y comprobamos si $\mathbf{S} \mathbf{P}^{-1}$ es también simétrica.
[^diagonalization]

[^diagonalization]: En realidad, la forma idónea de proceder sería encontrar la
diagonalización simultánea de $\mathbf{S}$ y $\mathbf{P}^{-1}$, ya que si son
simultáneamente diagonalizables entonces son conmutativas. No obstante, esto es
más complejo para mí y ahora mismo no tengo claro saber cómo hacerlo, mientras
que para obtener la $\mathbf{P}^{-1}$ simétrica puedo reaprovechar las
deducciones de Mario mediante la descomposición espectral.

### Solución mediante descomposición espectral de $\mathbf{R}$

Puesto que $\mathbf{R}$ es semidefinida positiva, existen las matrices
$\mathbf{Q}$ y $\mathbf{\Lambda}$ tal que
$\mathbf{R} = \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T$, y además:

* $\mathbf{Q}$ es una matriz ortonormal, i.e., $\mathbf{Q}^T = \mathbf{Q}^{-1}$.
* $\mathbf{\Lambda}$ es una matriz diagonal.

Dado que $\mathbf{\Lambda}$ es diagonal, aplicamos
$\mathbf{\Lambda} = \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2}$ para obtener
$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$:

$$
\mathbf{R} = \mathbf{P}^{-1} (\mathbf{P}^{-1})^T =
  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
$$

Ahora bien, para obtener una $\mathbf{P}^{-1}$ simétrica, podemos aplicar

$$
\mathbf{P}^{-1} (\mathbf{P}^{-1})^T =
  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^{-1}
    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
$$

Igualando $\mathbf{P}^{-1} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
tenemos que
$(\mathbf{P}^{-1})^T = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$
y por tanto $\mathbf{P}^{-1} = (\mathbf{P}^{-1})^T$. Luego 
$\mathbf{P}^{-1} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$ es una
solución que hace a $\mathbf{P}^{-1}$ simétrica.

### Comprobación de la simetría de $\mathbf{S} \mathbf{P}^{-1}$

Comprobamos si el producto $\mathbf{S} \mathbf{P}^{-1}$ da lugar a una matriz
simétrica.

$$
(\mathbf{S} \mathbf{P}^{-1})^T =
  (\mathbf{P}^{-1})^T \mathbf{S}^T = \mathbf{P}^{-1} \mathbf{S}
$$

Según este resultado,
$(\mathbf{S} \mathbf{P}^{-1})^T = \mathbf{S} \mathbf{P}^{-1}$, y por lo tanto
$\mathbf{S} \mathbf{P}^{-1}$ es simétrica y (tal y como muestra ese mismo
resultado) $\mathbf{S}$ y $\mathbf{P}^{-1}$ son conmutativas.

Algo no parece encajar en esta deducción, ya que según eso todos los pares de
matrices simétricas (i.e. para cualesquiera $\mathbf{S}$ y $\mathbf{P}^{-1}$)
deberían dar lugar a una matriz simétrica y cumplir la propiedad de
conmutatividad.

### 2024-06-11 notebooks/Corr_based_parameters.qmd
Deletes wrong conclusion about symmetry

@@
 $$
+\mathbf{S} \mathbf{P}^{-1} =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \\
+
 (\mathbf{S} \mathbf{P}^{-1})^T =
-  (\mathbf{P}^{-1})^T \mathbf{S}^T = \mathbf{P}^{-1} \mathbf{S}
+  (\mathbf{P}^{-1})^T \mathbf{S}^T = \mathbf{P}^{-1} \mathbf{S} \\
+
+(\mathbf{S} \mathbf{P}^{-1})^T =
+  (\mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T =
+  (\mathbf{Q}^T)^T (\mathbf{\Lambda}^{1/2})^T \mathbf{Q}^T \mathbf{S}^T =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S} \\
+
+\mathbf{P}^{-1} \mathbf{S} =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
 $$
@@
 Según este resultado,
-$(\mathbf{S} \mathbf{P}^{-1})^T = \mathbf{S} \mathbf{P}^{-1}$, y por lo tanto
-$\mathbf{S} \mathbf{P}^{-1}$ es simétrica y (tal y como muestra ese mismo
-resultado) $\mathbf{S}$ y $\mathbf{P}^{-1}$ son conmutativas.
-
-Algo no parece encajar en esta deducción, ya que según eso todos los pares de
-matrices simétricas (i.e. para cualesquiera $\mathbf{S}$ y $\mathbf{P}^{-1}$)
-deberían dar lugar a una matriz simétrica y cumplir la propiedad de
-conmutatividad.
+$(\mathbf{S} \mathbf{P}^{-1})^T = \mathbf{S} \mathbf{P}^{-1}$, pero esto no
+demuestra que $\mathbf{S} \mathbf{P}^{-1}$ sea simétrica ni que $\mathbf{S}$ y
+$\mathbf{P}^{-1}$ sean conmutativas (ver
+[aquí un contraejemplo](https://math.vanderbilt.edu/sapirmv/msapir/jan22.html)
+).

### 2024-06-12 notebooks/Corr_based_parameters.qmd
Corrects LaTeX math syntax mistakes

@@
 $$
-\mathbf{S} = diag(\sigma_1, \sigma_2) \\
+\mathbf{S} = diag(\sigma_1, \sigma_2)
+$$
+
+$$
 (\mathbf{P}^{-1})^T = \begin{pmatrix}
@@
     0        & \sigma_2 \sqrt{1 - \rho^2}
-  \end{pmatrix} \\
+  \end{pmatrix}
+$$
 
+$$
 (\mathbf{P}^{-1})^T \mathbf{S} = \begin{pmatrix}
@@
     0        & \sigma_2 \sqrt{1 - \rho^2}
-  \end{pmatrix} \\
+  \end{pmatrix}
+$$
 
+$$
 \mathbf{S} \mathbf{P}^{-1} = diag(\sigma_1, \sigma_2)
@@
     \sigma_2 \rho & \sigma_2 \sqrt{1 - \rho^2}
-  \end{pmatrix} \\
+  \end{pmatrix}
+$$
 
+$$
 \mathbf{P}^{-1} \mathbf{S} = \begin{pmatrix}
@@
 Estas ecuaciones muestran que la conmutatividad de $\mathbf{S}$ con
-$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$ no se cumple, en general (salvo que
+$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$ no se cumple en general (salvo que
 $\sigma_1 = \sigma_2$).
@@
 \mathbf{S} \mathbf{P}^{-1} =
-  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \\
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+$$
 
+$$
 (\mathbf{S} \mathbf{P}^{-1})^T =
-  (\mathbf{P}^{-1})^T \mathbf{S}^T = \mathbf{P}^{-1} \mathbf{S} \\
+  (\mathbf{P}^{-1})^T \mathbf{S}^T = \mathbf{P}^{-1} \mathbf{S}
+$$
 
+$$
 (\mathbf{S} \mathbf{P}^{-1})^T =
@@
   (\mathbf{Q}^T)^T (\mathbf{\Lambda}^{1/2})^T \mathbf{Q}^T \mathbf{S}^T =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S} \\
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
+$$
 
+$$
 \mathbf{P}^{-1} \mathbf{S} =

### 2024-06-12 notebooks/Corr_based_parameters.qmd
Completes proof
The proof needs to apply the property of positive semidefiniteness of the product $\mathbf{S} \mathbf{P}^{-1}$; this property is shown to fulfill and then applied to complete the proof.

@@
 ).
+
+Podemos utilizar las siguientes propiedades: Sabemos que $\mathbf{S}$ y
+$\mathbf{P}$ son semidefinidas positivas[^P_SDP]. Dado que $\mathbf{P}$ es
+semidefinida positiva, $\mathbf{P}^{-1}$ también lo es. Según
+[Meensakshi & Rakian, 1999](https://core.ac.uk/download/pdf/82822897.pdf)
+(referenciado en
+[esta respuesta en Math StackExchange](https://math.stackexchange.com/a/483930)
+), la condición para que $\mathbf{S} \mathbf{P}^{-1}$ sea simétrica es que
+$\mathbf{S} \mathbf{P}^{-1}$ sea "normal", es decir, que
+$\mathbf{S} \mathbf{P}^{-1} (\mathbf{S} \mathbf{P}^{-1})^T =
+  (\mathbf{S} \mathbf{P}^{-1})^T \mathbf{S} \mathbf{P}^{-1}$.
+Ahora bien,
+
+[^P_SDP]: Estoy asumiendo que $\mathbf{P}$ es semidefinida positiva porque
+$\mathbf{R}$ lo es, y $\mathbf{P}T \mathbf{P} = \mathbf{R}$. Puede que haya un
+salto inferencial aquí (i.e., que sea necesario demostrar que $\mathbf{P}$ es
+semidefinida positiva).
+
+$$
+\mathbf{S} \mathbf{P}^{-1} (\mathbf{S} \mathbf{P}^{-1})^T =
+  \mathbf{S} \mathbf{P}^{-1} (\mathbf{P}^{-1})^T \mathbf{S}^T =
+  \mathbf{S} \mathbf{P}^{-1} (\mathbf{P}^{-1})^T \mathbf{S} =
+  \mathbf{\Sigma}.
+$$
+
+Como $\mathbf{\Sigma}$ es simétrica (y semidefinida positiva),
+el producto $\mathbf{S} \mathbf{P}^{-1}$ es conmutativo. Comprobamos no obstante
+la propiedad de "normalidad" de $\mathbf{S} \mathbf{P}^{-1}$:
+
+$$
+\mathbf{S} \mathbf{P}^{-1} (\mathbf{S} \mathbf{P}^{-1})^T =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+    (\mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+    (\mathbf{Q}^T)^T \mathbf{\Lambda}^{1/2}^T \mathbf{Q}^T \mathbf{S}^T =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S} =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T \mathbf{S} =
+  \mathbf{S} \mathbf{R} \mathbf{S} =
+  \mathbf{\Sigma}
+$$
+
+$$
+(\mathbf{S} \mathbf{P}^{-1})^T \mathbf{S} \mathbf{P}^{-1} =
+  (\mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T
+    \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
+  (\mathbf{Q}^T)^T \mathbf{\Lambda}^{1/2}^T \mathbf{Q}^T \mathbf{S}^T
+    \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
+    \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
+  \mathbf{P}^{-1} \mathbf{S}^2 (\mathbf{P}^{-1})^T
+$$
+
+Esto no parece fácil de demostrar, por desgracia. La ecuación anterior, sin
+embargo, nos permite aplicar el resultado al que queríamos llegar, ya que
+obtenemos que la distribución de rasgos latentes transformada,
+$\mathbf{\theta}^\mathcal{E}$, tiene por matriz de covarianzas $\mathbf{S}^2$.
+
+$$
+\mathbf{\Sigma}^\mathcal{E} = \mathbf{P} \mathbf{S}^2 \mathbf{P}^T
+  \mathbf{P} \mathbf{P}^{-1} \mathbf{S}^2 (\mathbf{P}^{-1})^T \mathbf{P}^T =
+  \mathbf{S}^2
+$$

### 2024-06-18 notebooks/Sobre noteook Daniel corre_based_parametesr.Rmd
Versions Mario's response notebook

---
title: "Sobre notebook Daniel corre based parameters"
author: "Mario Luzardo"
date: "2024-06-17"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list=ls()) 
library(pspline)
library(mvtnorm)
library(sn)
library(mnormt)
library(scatterplot3d)
library(copula)
library(ggplot2)
library(fungible)
library(pracma)
library(matlib)
library(rgl)
library(lsa)
library(MASS)
library(mirt)
library(tidyverse)
set.seed(123)
```

## Supuestos iniciales

Como hemos hecho hasta ahora, asumimos un espacio $\mathbf{\Theta}$  con base
$\mathcal{B}$ en el que se representa el vector de rasgo latente
$\mathbf{\theta}$. Este vector es una variable aleatoria $n$-variada que se
distribuye con covarianza $\mathbf{\Sigma} = \mathbf{S} \mathbf{R} \mathbf{S}$,
siendo $\mathbf{R}$ una matriz de correlaciones, semidefinida positiva y con 1's
en la diagonal, y $\mathbf{S}$ una matriz diagonal de desviaciones típicas, tal
que $\sigma^2_{ii} = s^2_{ii}$, para todo $i$ entre 1 y $n$ (i.e., los elementos
diagonales de $\mathbf{\Sigma}$ y $\mathbf{S}^2$ son iguales).

## Premisa

Es posible encontrar una representación del espacio latente en una base
ortogonal $\mathcal{U}$ tal que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de
covarianzas de $\mathbf{\theta}$ representado en $\mathcal{U}$,
$\mathbf{\theta}^\mathcal{U}$) sea una matriz diagonal.

Sea $\mathbf{P}$ la matriz de transformación de la base $\mathcal{B}$ a la base
$\mathcal{U}$. Como ya sabemos,
$\mathbf{P} = [\mathbf{b}_1^\mathcal{U}, ..., \mathbf{b}_n^\mathcal{U}]$, es
decir, los vectores de la base $\mathcal{B}$ representados en la base
$\mathcal{U}$. Las coordenadas de $\mathbf{\theta}$, exprsadas en la base
$\mathcal{U}$, vienen dadas por la siguiente relación:

$$
\mathbf{\theta}^\mathcal{U} = \mathbf{P} \mathbf{\theta}
$$

La matriz de covarianzas $\mathbf{\Sigma}^\mathcal{U}$ de
$\mathbf{\theta}^\mathcal{U}$ cumple

$$
\mathbf{\Sigma}^\mathcal{U} = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T 
$$
Queremos que 
$$
\mathbf{\Sigma}^\mathcal{U} = S^2 
$$

Entonces igualemos

$$
S^2 = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T 
$$

luego si operamos obtenemos

 $$ 
 P^{-1} S^2(P^T)^{-1}=\mathbf{\Sigma}
 $$
 Como S es diagonal

$$ 
 (S^{-1}P)^{-1} ((S^{-1}P)^T)^{-1}=\mathbf{\Sigma}
$$
Si llamo $K=S^{-1}P$ entonces

$$ 
 K^T K=(\mathbf{\Sigma})^{-1}
$$
Si realizamosla descomosición espectral de $$\mathbf{\Sigma}$$ obtenemos

$$\mathbf{\Sigma}=Q \Lambda Q^{-1}$$ con Q ortogonal o sea $Q^{-1}=Q^T$ la matriz de vectores propios y $\Lambda$ la matriz diagonal de valores propios
luego

$$\mathbf{\Sigma}^{-1}=Q \Lambda^{-1} Q^{-1}$$
o sea

$$K^T K=\mathbf{\Sigma}^{-1}=Q (\Lambda)^{-1/2} Q^{-1} Q  (\Lambda)^{-1/2} Q^{-1}$$

o sea encuentro la matriz raiz cuadrada $$\sqrt{\Sigma^{-1}}=Q (\Lambda)^{-1/2} Q^{-1}$$

Es claro por las propiedades de Q que  
$$K=K^T=Q (\Lambda)^{-1/2} Q^{-1}$$
Luego como $K=S^{-1}P$

$$P=S Q (\Lambda)^{-1/2} Q^{-1}$$

No es necesario hallar la matriz raiz cuadrada ya que su usamos 
$$K^T=Q (\Lambda)^{1/2}$$
también sirve ya que 
$$\mathbf{\Sigma}^{-1}=K^TK=Q (\Lambda)^{-1/2}   (\Lambda)^{-1/2} Q^{-1}$$
Por lo tanto

$$K= (\Lambda)^{1/2} Q^T=(\Lambda)^{1/2} Q^{-1}$$
o sea

$$P= S (\Lambda)^{1/2} Q^{-1}$$

Podemos usar ls descomposicion de Cholesky

Como $\Sigma$ es simétrica y supongamos definida positiva entonces por Cholesky podemos escribir

$$\Sigma=L^T L$$
donde L es una matrz triangular superior, o sea $K =(L^{T})^{-1}$ de donde

$$
P=S(L^{T})^{-1}
$$

Veamos un ejemplo en $R^3$ con los tres métodos

```{r   echo=FALSE}
Sigma <- c(4,5, 4.8, 5, 25, 9, 4.8, 9, 9) |> matrix(ncol = 3)
S=diag(c(2,5,3))
R=cov_matrix <- c(1, .5, .8, .5, 1, .6, .8, .6, 1) |> matrix(ncol = 3)
print("Matriz de varianza covarianza")
print(Sigma)
print("Matriz de desviaciones")
print(S)
print("Matriz de correlaciones")
print(R)
print("Verificacion")
print(S%*%R%*%S)
```

Calculamos las matrices de cambio de base por los tres métodos

```{r   echo=FALSE}
# descomposición espectral de Sigma_U:
eigendecomposition <- Sigma |> eigen()
lambda             <- eigendecomposition$values
Q                  <- eigendecomposition$vectors

# Base calculada:
lambda_root <- eigendecomposition$values^(-1) |> sqrt() |> diag()
Q_inv       <- Q |> solve()

#Calculo de matrices de cambio de base
#Calculo de P por descomposicion espectral
P <- S%*%Q %*% lambda_root %*% Q_inv
#Calculo de P sin raiz cuadrada
P2<- S%*%lambda_root %*% Q_inv
#Calculo P cholesky
PCHOL<-S%*%solve(t(chol(Sigma)))

print("Matriz P por descomposición espectral ")
print(P)
print("Matriz P por método simplificado ")
print(P2)
print("Matriz P por Cholesky")
print(PCHOL)
```

Comprobemos que la matriz obtenida ahora es $S^2$

Comprobacion usando P

```{r   echo=FALSE}
print(round(P %*% Sigma %*% t(P),3)) 
```
Comprobacion usando P2
```{r   echo=FALSE}
print(round(P2 %*% Sigma %*% t(P2),3)) 
```
Comprobacion usando Cholesky
```{r   echo=FALSE}
print(round(PCHOL %*% Sigma %*% t(PCHOL)))
```

### 2024-06-19 notebooks/Corr_based_parameters.qmd
Specifies correctly the "correlation" condition
The condition that the inverse of the correlation must be the Gram matrix in the original space was not correctly specified as an assumption. This commit makes the necessary changes to specify it properly.

@@
 Como hemos hecho hasta ahora, asumimos un espacio $\mathbf{\Theta}$ con base
-$\mathcal{B}$ en el que se representa el vector de rasgo latente
-$\mathbf{\theta}$. Este vector es una variable aleatoria $n$-variada que se
-distribuye con covarianza $\mathbf{\Sigma} = \mathbf{S} \mathbf{R} \mathbf{S}$,
-siendo $\mathbf{R}$ una matriz de correlaciones, semidefinida positiva y con 1's
-en la diagonal, y $\mathbf{S}$ una matriz diagonal de desviaciones típicas, tal
-que $\sigma^2_{ii} = s^2_{ii}$, para todo $i$ entre 1 y $n$ (i.e., los elementos
+$\mathcal{B}$ y matriz Gramiana del producto interno $\mathbf{M}$ en el que se
+representa el vector de rasgo latente $\mathbf{\theta}$. Este vector es una
+variable aleatoria $n$-variada que se distribuye con covarianza
+$\mathbf{\Sigma} = \mathbf{S} \mathbf{R} \mathbf{S}$, siendo $\mathbf{R}$ una
+matriz de correlaciones, semidefinida positiva y con 1's en la diagonal, y
+$\mathbf{S}$ una matriz diagonal de desviaciones típicas, tal que
+$\sigma^2_{ii} = s^2_{ii}$, para todo $i$ entre 1 y $n$ (i.e., los elementos
 diagonales de $\mathbf{\Sigma}$ y $\mathbf{S}^2$ son iguales).
 
-## Premisa
-
-Es posible encontrar una representación del espacio latente en una base
-ortogonal $\mathcal{U}$ tal que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de
-covarianzas de $\mathbf{\theta}$ representado en $\mathcal{U}$,
-$\mathbf{\theta}^\mathcal{U}$) sea una matriz diagonal.
-
-## Demostración
-
 Sea $\mathbf{P}$ la matriz de transformación de la base $\mathcal{B}$ a la base
-$\mathcal{U}$. Como ya sabemos,
-$\mathbf{P} = [\mathbf{b}_1^\mathcal{U}, ..., \mathbf{b}_n^\mathcal{U}]$, es
-decir, los vectores de la base $\mathcal{B}$ representados en la base
-$\mathcal{U}$. Las coordenadas de $\mathbf{\theta}$, exprsadas en la base
+$\mathcal{U}$. Las coordenadas de $\mathbf{\theta}$, expresadas en la base
 $\mathcal{U}$, vienen dadas por la siguiente relación:
@@
 $$
+
+Como ya sabemos,
+$\mathbf{P} = [\mathbf{b}_1^\mathcal{U}, ..., \mathbf{b}_n^\mathcal{U}]$, es
+decir, los vectores de la base $\mathcal{B}$ representados en la base
+$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortogonal.
+Sabemos entonces que $\mathbf{P}^T \mathbf{P} = \mathbf{M}$.
+
+## Premisa
+
+Es posible encontrar en el espacio latente $\mathbf{\Theta}$ una matriz de
+transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal
+que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de covarianzas de
+$\mathbf{\theta}$ representado en $\mathcal{U}$, $\mathbf{\theta}^\mathcal{U}$)
+sea una matriz diagonal, siendo $\mathbf{M} = \mathbf{R}^{-1}$.
+
+## Demostración
 

### 2024-06-19 output/paper_draft/paper_M2PLM_parameters.Rmd
Makes minor punctuation correction.

--- 2024-06-04 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 the origin of the multidimensional latent space to
-the maximally discriminating locus of the item.
-The second one, the direction of the vector from the origin to that same locus.
+the maximally discriminating locus of the item;
+the second one, the direction of the vector from the origin to that same locus.
 Following a similar rationale, @reckase_discriminating_1991 derive a

### 2024-06-19 output/paper_draft/paper_M2PLM_parameters.Rmd
Moves "test space" conclusions to "Results"
This commit separates the "invariance" condition of the model from the results regarding the test space, which are moved to a new section in the results. In this new section, some properties of the test space are also explained.

--- 2024-06-04 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-### Test space and latent space
+### Item response function invariance
 
@@
 
-We shall refer to this new space $`r TEST_SPACE`$ as
-the *test space* [@zhang_theoretical_1999].
-Although the inverse relationship between their respective inner product
-Gram matrices induces a bijective relationship between the two sets of spaces,
-it is worth highlighting that test space $`r TEST_SPACE`$
-and latent space $`r LATENT_SPACE`$ are different entities. Only in the
-orthonormal case the two spaces share a common inner product
-(i.e., the dot product), and thus can be mistaken.
-
 ### Direction cosines in the latent space
@@
 \begin{equation}
-  `r SLOPE_MAX_PARAM_EQ`.                                     (\#eq:slope-param)
+  `r SLOPE_MAX_PARAM_EQ`,                                     (\#eq:slope-param)
 \end{equation}
@@
 
-Applying Proposition 2, if $`r ANGLE_VEC_TS_ITEM`$ is
+# Results
+
+## Test space vs. latent space
+
+Before presenting the results about the parameters,
+we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
+As shown before, this space is an entity by itself,
+distinct from the latent space $`r LATENT_SPACE`$.
+Despite this, the dot product in the exponent of Equation
+\@ref(eq:eq:M2PL-formulation) induces a bijective relationship between
+their two sets of respective bases.
+Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
+with inner product gram matrix $`r INNER_PROD_MATRIX`$,
+$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
+with inner product gram matrix $`r INNER_PROD_MATRIX_INV`$,
+for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
+Only when the two bases are orthonormal,
+the two spaces share a common inner product (i.e., the dot product),
+and thus can be mistaken.
+However, it is worth noting that, although their representation
+can be superimposed in this case,
+they are still two different spaces. <!-- # TODO: To discussion? -->
+Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
+the test space, as it is the space where
+the items (i.e., the test elements) are represented.
+
+We must make a few remarks here.
+First, the transformation to the orthonormal bases of the two spaces allows
+mapping the $`r MODEL_ACRONYM`$ parameters indistinctly in the two spaces,
+by projecting one space onto the other.
+This allows representing the coordinates of the discrimination vectors
+in the latent space and, conversely,
+the coordinates of the latent trait vectors in the test space.
+As we can see, the invariance of the model holds,
+no matter in which space we represent the coordinates.
+<!-- # TODO: Incluir la condición de invarianza de las coordenadas en los dos espacios (Foto pizarra)? -->
+Second, and most importantly, if the test space is
+where the item parameters naturally lie, they must be represented in this space,
+and not in the latent space.
+Therefore, we may find the direction angles of $`r DISCR_VECTOR`$
+in the test space.
+If $`r ANGLE_VEC_TS_ITEM`$ is
 the vector of direction angles of $`r DISCR_VECTOR`$,
+applying Proposition 2
 in the test space $`r TEST_SPACE`$ we get
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
-
-# Results
+<!-- # TODO: Revisar esto bien. Esto sólo aplica correctamente si efectivamente
+"equiparamos" las bases U y U*. Además, en la Proposición 2 estamos asumiendo
+que la "base de enlace" es la canónica, no una base ortonormal cualquiera. -->
+As we will see later, this representation is very convenient, as
+the alignment of the item vectors with the coordinate axes has a direct,
+meaningful interpretation.
 
@@
 Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
-equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
+Equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
 However, as they depend on the inner product matrix $`r INNER_PROD_MATRIX`$,

### 2024-06-19 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds minor corrections

@@
 
-Given (1),
+Given 1,
 
@@
 
-Using (2) and (3) we get that
+Using 2 and 3 we get that
 
@@
 
-Applying Proposition 1 to the model formulation in \@ref(eq:M2PL-formulation),
+Applying Proposition 1 to
+the model formulation in Equation \@ref(eq:M2PL-formulation),
 the `r MODEL_ACRONYM` model is expressed in polar coordinates as
@@
 
-We have that
+and thus, we have that
 
@@
 Despite this, the dot product in the exponent of Equation
-\@ref(eq:eq:M2PL-formulation) induces a bijective relationship between
+\@ref(eq:M2PL-formulation) induces a bijective relationship between
 their two sets of respective bases.
 Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
-with inner product gram matrix $`r INNER_PROD_MATRIX`$,
+with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
 $`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
-with inner product gram matrix $`r INNER_PROD_MATRIX_INV`$,
+with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
 for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
@@
 Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
-the test space, as it is the space where
+the *test space*, as it is the space where
 the items (i.e., the test elements) are represented.
 
-We must make a few remarks here.
+We must make a few observations here.
 First, the transformation to the orthonormal bases of the two spaces allows
@@
 As we will see later, this representation is very convenient, as
-the alignment of the item vectors with the coordinate axes has a direct,
-meaningful interpretation.
+the direction of the item vectors with respect to the coordinate axes
+has a direct, meaningful interpretation.
 
@@
 $`r LATENT_SPACE`$ independently of its basis $`r LS_BASIS`$.
-If we assume $`r BASIS_EQ`$ or, more generally,
-that $`r LS_BASIS`$ is orthonormal, $`r TRANSFORM_MATRIX`$ will be
+If we assume that $`r BASIS_EQ`$ or, more generally,
+that $`r LS_BASIS`$ is any orthonormal basis, $`r TRANSFORM_MATRIX`$ will be
 also orthonormal and thus $`r INNER_PROD_MAT_STD_EQ`$.
@@
 it must be somewhat related to the covariance structure.
-Therefore, an orthonormal basis should represent independent coordinates;
+Therefore, an orthonormal basis should represent independent,
+standard coordinates;
 formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
@@
 
-There are three conditions to meet for regarding these parameters as
+There are three conditions to meet for considering these parameters as
 a valid generalization of the (unidimensional) IRT parameters
@@
 and the third one is implicit in the definition of the $`r MDISC_SYM`$.
-Simple arithmetic can show that the first holds true whenever the
+Simple arithmetic can show that the first one holds whenever the
 $`r DIAG_KTH_ELEMENT`$-th element of $`r INNER_PROD_MATRIX_INV`$ equals 1.
 Therefore, both the agnostic and the correlation-based version fulfill it.
-$`r MDISC_COV_PARAM`$ does not though,
+The covariance-based version $`r MDISC_COV_PARAM`$ does not though,
 as $`r COV_ELEMENT`$ is not generally equal to 1.

### 2024-06-26 notebooks/Corr_based_parameters.qmd
Updates deduction after Mario's review

--- 2024-06-12 notebooks/Corr_based_parameters.qmd
@@
 
+Como ya sabemos,
+$\mathbf{P} = [\mathbf{b}_1^\mathcal{U}, ..., \mathbf{b}_n^\mathcal{U}]$, es
+decir, los vectores de la base $\mathcal{B}$ representados en la base
+$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortogonal,
+con producto interno estándar. Sabemos entonces que
+$\mathbf{P}^T \mathbf{P} = \mathbf{M}$[^orth_basis].
+
+[^orth_basis]: Problema: Esto puede **no ser cierto**, ya que para que ello sea
+así, la base $\mathcal{U}$ debe ser ortonormal, y no solamente ortogonal.
+
+## Premisa
+
+Es posible encontrar en el espacio latente $\mathbf{\Theta}$ una matriz de
+transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal
+que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de covarianzas de
+$\mathbf{\theta}$ representado en $\mathcal{U}$, $\mathbf{\theta}^\mathcal{U}$)
+sea una matriz diagonal, siendo $\mathbf{M} = \mathbf{R}^{-1}$.
+
+Además, esta matriz invariante en "escala"; es decir, conservará las "varianzas"
+de $\mathbf{\Theta}$: $\mathbf{\Sigma}^\mathcal{U} = \mathbf{S}^2$.
+
+## Demostración
+
 La matriz de covarianzas $\mathbf{\Sigma}^\mathcal{U}$ de
@@
 $$
-\mathbf{\Sigma}^\mathcal{U} = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T =
-  \mathbf{P} \mathbf{S} \mathbf{R} \mathbf{S} \mathbf{P}^T
+\mathbf{\Sigma}^\mathcal{U} = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T.
 $$
 
-Supongamos que $\mathbf{P}$ es una raíz cuadrada de $\mathbf{R}^{-1}$; es
-decir, $\mathbf{R}^{-1} = \mathbf{P}^T \mathbf{P}$.
+Por lo tanto, igualamos
 
 $$
-\mathbf{\Sigma}^\mathcal{U} =
-  \mathbf{P} \mathbf{S} (\mathbf{P}^T \mathbf{P})^{-1} \mathbf{S} \mathbf{P}^T =
-  \mathbf{P} \mathbf{S} \mathbf{P}^{-1} (\mathbf{P}^{-1})^T \mathbf{S} \mathbf{P}^T
+\mathbf{S}^2 = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T
 $$
 
-Si $\mathbf{S}$ conmuta con $\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$, entonces
-la ecuación anterior puede expresarse como
+y, operando, obtenemos
 
 $$
-\mathbf{\Sigma}^\mathcal{U} =
-  \mathbf{P} \mathbf{P}^{-1} \mathbf{S} \mathbf{S} (\mathbf{P}^{-1})^T \mathbf{P}^T =
-  \mathbf{S}^2
+\mathbf{P}^{-1} \mathbf{S}^2 (\mathbf{P}^{-1})^T = \mathbf{\Sigma}.
 $$
 
-### Contraejemplo
-
-Lo anterior en general no se cumple, por lo que $\mathbf{\Sigma}^\mathcal{U}$ no
-es, en general, una matriz diagonal bajo estos supuestos.
-
-Un contraejemplo de esto es la solución con la base del ejemplo de
-representación gráfica del borrador y varianzas desiguales. Supongamos
+Como $\mathbf{S}$ es diagonal,
 
 $$
-\mathbf{S} = diag(\sigma_1, \sigma_2)
+\mathbf{S} (\mathbf{P}^T)^{-1} (\mathbf{P}^T)^{-1} \mathbf{S} =
+  \mathbf{\Sigma} =
+  \mathbf{S} \mathbf{R} \mathbf{S},
 $$
 
+Dado que $\mathbf{P}$ es una raíz cuadrada de $\mathbf{R}^{-1}$,
+$\mathbf{R} = (\mathbf{P}^T \mathbf{P})^{-1}$.
+
 $$
-(\mathbf{P}^{-1})^T = \begin{pmatrix}
-    1 & \rho \\
-    0 & \sqrt{1 - \rho^2}
-  \end{pmatrix}.
+(\mathbf{P}^T \mathbf{P}^T)^{-1} = (\mathbf{P}^T \mathbf{P})^{-1}.
 $$
 
-Bajo estos supuestos,
-
-$$
-\mathbf{R} = \mathbf{P}^{-1} (\mathbf{P}^{-1})^T = \begin{pmatrix}
-       1 &              0 \\
-    \rho & \sqrt{1 - \rho^2}
-  \end{pmatrix}
-  \begin{pmatrix}
-    1 & \rho \\
-    0 & \sqrt{1 - \rho^2}
-  \end{pmatrix} =
-  \begin{pmatrix}
-       1 & \rho \\
-    \rho &    1
-  \end{pmatrix}
-$$
-
-$$
-\mathbf{\Sigma} = diag(\sigma_1, \sigma_2)
-  \begin{pmatrix}
-       1 & \rho \\
-    \rho &    1
-  \end{pmatrix}
-  diag(\sigma_1, \sigma_2) =
-  \begin{pmatrix}
-    \sigma_1^2             & \sigma_1 \sigma_2 \rho \\
-    \sigma_1 \sigma_2 \rho & \sigma_2^2
-  \end{pmatrix}.
-$$
-
-Veamos que $\mathbf{S}$ No conmuta con $\mathbf{P}^{-1}$ ni con
-$(\mathbf{P}^{-1})^T$:
-
-$$
-\mathbf{S} (\mathbf{P}^{-1})^T = diag(\sigma_1, \sigma_2)
-  \begin{pmatrix}
-    1 & \rho \\
-    0 & \sqrt{1 - \rho^2}
-  \end{pmatrix} =
-  \begin{pmatrix}
-    \sigma_1 & \sigma_1 \rho \\
-    0        & \sigma_2 \sqrt{1 - \rho^2}
-  \end{pmatrix}
-$$
-
-$$
-(\mathbf{P}^{-1})^T \mathbf{S} = \begin{pmatrix}
-    1 & \rho \\
-    0 & \sqrt{1 - \rho^2}
-  \end{pmatrix}
-  diag(\sigma_1, \sigma_2) =
-  \begin{pmatrix}
-    \sigma_1 & \sigma_2 \rho \\
-    0        & \sigma_2 \sqrt{1 - \rho^2}
-  \end{pmatrix}
-$$
-
-$$
-\mathbf{S} \mathbf{P}^{-1} = diag(\sigma_1, \sigma_2)
-  \begin{pmatrix}
-       1 &              0 \\
-    \rho & \sqrt{1 - \rho^2}
-  \end{pmatrix} =
-  \begin{pmatrix}
-    \sigma_1      &              0 \\
-    \sigma_2 \rho & \sigma_2 \sqrt{1 - \rho^2}
-  \end{pmatrix}
-$$
-
-$$
-\mathbf{P}^{-1} \mathbf{S} = \begin{pmatrix}
-       1 &              0 \\
-    \rho & \sqrt{1 - \rho^2}
-  \end{pmatrix}
-  diag(\sigma_1, \sigma_2) =
-  \begin{pmatrix}
-    \sigma_1      &              0 \\
-    \sigma_1 \rho & \sigma_2 \sqrt{1 - \rho^2}
-  \end{pmatrix}
-$$
-
-Estas ecuaciones muestran que la conmutatividad de $\mathbf{S}$ con
-$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$ no se cumple en general (salvo que
-$\sigma_1 = \sigma_2$).
-
-### Condiciones adicionales
-
-A pesar de lo anterior, $\mathbf{S}$ conmuta con cualquier otra matriz que sea
-simultáneamente diagonalizable con ella. En el caso de matrices simétricas, es
-decir, si $\mathbf{P}^{-1} = (\mathbf{P}^{-1})^T$, sabemos que se cumple la
-conmutatividad de $\mathbf{S}$ con $\mathbf{P}^{-1}$ (y por lo tanto también con
-$(\mathbf{P}^{-1})^T$), si
-$\mathbf{S} \mathbf{P}^{-1} = \mathbf{P}^{-1} \mathbf{S}$ es también simétrica.
-
-Intentamos obtener a continuación una solución que haga $\mathbf{P}^{-1}$
-simétrica, y comprobamos si $\mathbf{S} \mathbf{P}^{-1}$ es también simétrica.
-[^diagonalization]
-
-[^diagonalization]: En realidad, la forma idónea de proceder sería encontrar la
-diagonalización simultánea de $\mathbf{S}$ y $\mathbf{P}^{-1}$, ya que si son
-simultáneamente diagonalizables entonces son conmutativas. No obstante, esto es
-más complejo para mí y ahora mismo no tengo claro saber cómo hacerlo, mientras
-que para obtener la $\mathbf{P}^{-1}$ simétrica puedo reaprovechar las
-deducciones de Mario mediante la descomposición espectral.
+Es decir, la solución que buscamos es una matriz $\mathbf{P}$, raíz cuadrada de
+$\mathbf{R}^{-1}$, que
+cumpla la condición de simetría (i.e. $\mathbf{P} = \mathbf{P}^T$).
 
@@
 
-Puesto que $\mathbf{R}$ es semidefinida positiva, existen las matrices
+Puesto que $\mathbf{R}^{-1}$ es semidefinida positiva, existen las matrices
 $\mathbf{Q}$ y $\mathbf{\Lambda}$ tal que
-$\mathbf{R} = \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T$, y además:
+$\mathbf{R}^{-1} = \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T$, y además:
 
@@
 Dado que $\mathbf{\Lambda}$ es diagonal, aplicamos
-$\mathbf{\Lambda} = \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2}$ para obtener
-$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$:
+$\mathbf{\Lambda} = \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2}$:
 
 $$
-\mathbf{R} = \mathbf{P}^{-1} (\mathbf{P}^{-1})^T =
+\mathbf{R}^{-1} = \mathbf{P}^T \mathbf{P} =
   \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
@@
 
-Ahora bien, para obtener una $\mathbf{P}^{-1}$ simétrica, podemos aplicar
+Igualmente, aplicamos
 
 $$
-\mathbf{P}^{-1} (\mathbf{P}^{-1})^T =
+\mathbf{P}^T \mathbf{P} =
   \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^{-1}
@@
 
-Igualando $\mathbf{P}^{-1} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
+Igualando $\mathbf{P} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
 tenemos que
-$(\mathbf{P}^{-1})^T = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$
-y por tanto $\mathbf{P}^{-1} = (\mathbf{P}^{-1})^T$. Luego 
-$\mathbf{P}^{-1} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$ es una
-solución que hace a $\mathbf{P}^{-1}$ simétrica.
+$\mathbf{P}^T = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
+y por tanto $\mathbf{P} = \mathbf{P}^T$.
 
-### Comprobación de la simetría de $\mathbf{S} \mathbf{P}^{-1}$
+# Ejemplos
 
-Comprobamos si el producto $\mathbf{S} \mathbf{P}^{-1}$ da lugar a una matriz
-simétrica.
+## Ejemplo en $\mathbb{R}^2$
 
-$$
-\mathbf{S} \mathbf{P}^{-1} =
-  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
-$$
+Usamos el ejemplo con la correlación del borrador, y covarianzas de 2 y 3.
 
-$$
-(\mathbf{S} \mathbf{P}^{-1})^T =
-  (\mathbf{P}^{-1})^T \mathbf{S}^T = \mathbf{P}^{-1} \mathbf{S}
-$$
+```{r 2-dim-example}
+corr_matrix <- c(1, .5, .5, 1) |> matrix(ncol = 2)
+std_devs    <- c(2, 3) |> sqrt() |> diag()
+cov_matrix  <- std_devs %*% corr_matrix %*% std_devs
 
-$$
-(\mathbf{S} \mathbf{P}^{-1})^T =
-  (\mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T =
-  (\mathbf{Q}^T)^T (\mathbf{\Lambda}^{1/2})^T \mathbf{Q}^T \mathbf{S}^T =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
-$$
+## Comprobación:
+cov_matrix |> cov2cor()
+```
 
-$$
-\mathbf{P}^{-1} \mathbf{S} =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
-$$
+```{r basis-computation}
+# descomposición espectral de Sigma_U:
+eigendecomposition <- corr_matrix |> solve() |> eigen()
+lambda             <- eigendecomposition$values
+Q                  <- eigendecomposition$vectors
 
-Según este resultado,
-$(\mathbf{S} \mathbf{P}^{-1})^T = \mathbf{S} \mathbf{P}^{-1}$, pero esto no
-demuestra que $\mathbf{S} \mathbf{P}^{-1}$ sea simétrica ni que $\mathbf{S}$ y
-$\mathbf{P}^{-1}$ sean conmutativas (ver
-[aquí un contraejemplo](https://math.vanderbilt.edu/sapirmv/msapir/jan22.html)
-).
+# Base calculada:
+lambda_sqrt <- eigendecomposition$values |> sqrt() |> diag()
+Q_inv       <- Q |> t()
+
+# Calculo de P por descomposicion espectral
+P <- Q %*% lambda_sqrt %*% Q_inv
+
+# Comprobación
+
+## Raíz de la inversa de la correlación:
+solve(t(P) %*% P)
+
+## Covarianza en la base transformada:
+P %*% cov_matrix %*% t(P)
+```
+
+## Ejemplo en $\mathbb{R}^3$
+
+Usamos el ejemplo propuesto por Mario.
+
+```{r 3-dim-example}
+corr_matrix <- c(1, .5, .8, .5, 1, .6, .8, .6, 1) |> matrix(ncol = 3)
+std_devs    <- c(2, 5, 3) |> diag()
+cov_matrix  <- std_devs %*% corr_matrix %*% std_devs
+```
+
+```{r basis-computation}
+```

### 2024-06-26 notebooks/Sobre noteook Daniel corre_based_parametesr.Rmd
Leaves only valid derivations
The valid derivations are the ones that make the square root matrix of the correlation. I delete the ones that are not in line with this result.

@@
 
-No es necesario hallar la matriz raiz cuadrada ya que su usamos 
-$$K^T=Q (\Lambda)^{1/2}$$
-también sirve ya que 
-$$\mathbf{\Sigma}^{-1}=K^TK=Q (\Lambda)^{-1/2}   (\Lambda)^{-1/2} Q^{-1}$$
-Por lo tanto
-
-$$K= (\Lambda)^{1/2} Q^T=(\Lambda)^{1/2} Q^{-1}$$
-o sea
-
-$$P= S (\Lambda)^{1/2} Q^{-1}$$
-
-Podemos usar ls descomposicion de Cholesky
-
-Como $\Sigma$ es simétrica y supongamos definida positiva entonces por Cholesky podemos escribir
-
-$$\Sigma=L^T L$$
-donde L es una matrz triangular superior, o sea $K =(L^{T})^{-1}$ de donde
-
-$$
-P=S(L^{T})^{-1}
-$$
-
 Veamos un ejemplo en $R^3$ con los tres métodos
@@
 P <- S%*%Q %*% lambda_root %*% Q_inv
-#Calculo de P sin raiz cuadrada
-P2<- S%*%lambda_root %*% Q_inv
-#Calculo P cholesky
-PCHOL<-S%*%solve(t(chol(Sigma)))
 
@@
 print(P)
-print("Matriz P por método simplificado ")
-print(P2)
-print("Matriz P por Cholesky")
-print(PCHOL)
 ```
@@
 ```
-Comprobacion usando P2
-```{r   echo=FALSE}
-print(round(P2 %*% Sigma %*% t(P2),3)) 
-```
-Comprobacion usando Cholesky
-```{r   echo=FALSE}
-print(round(PCHOL %*% Sigma %*% t(PCHOL)))
-```

### 2024-06-26 notebooks/Sobre noteook Daniel corre_based_parametesr.Rmd
Corrects errors and style

@@
 ```{r   echo=FALSE}
-Sigma <- c(4,5, 4.8, 5, 25, 9, 4.8, 9, 9) |> matrix(ncol = 3)
-S=diag(c(2,5,3))
-R=cov_matrix <- c(1, .5, .8, .5, 1, .6, .8, .6, 1) |> matrix(ncol = 3)
+S     <- diag(c(2,5,3))
+R     <- c(1, .5, .8, .5, 1, .6, .8, .6, 1) |> matrix(ncol = 3)
+Sigma <- S %*% R %*% S
 print("Matriz de varianza covarianza")
@@
 print(R)
-print("Verificacion")
-print(S%*%R%*%S)
 ```
@@
 #Calculo de P por descomposicion espectral
-P <- S%*%Q %*% lambda_root %*% Q_inv
+P <- S %*% Q %*% lambda_root %*% Q_inv
 

### 2024-06-26 notebooks/Corr_based_parameters.qmd
Versions (incomplete) inline example

@@
 format: docx
+callout-icon: false
 editor: source
@@
 decir, los vectores de la base $\mathcal{B}$ representados en la base
-$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortogonal,
+$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortonormal,
 con producto interno estándar. Sabemos entonces que
-$\mathbf{P}^T \mathbf{P} = \mathbf{M}$[^orth_basis].
+$\mathbf{P}^T \mathbf{P} = \mathbf{M}$.
 
-[^orth_basis]: Problema: Esto puede **no ser cierto**, ya que para que ello sea
-así, la base $\mathcal{U}$ debe ser ortonormal, y no solamente ortogonal.
+::: note
+## Ejemplo
+
+Asumamos las siguientes matrices de desviación típica y varianzas:
+
+```{r 2d-example-matrices}
+corr_matrix <- c(1, .5, .5, 1) |> matrix(ncol = 2)
+std_devs    <- c(2, 3) |> diag()
+cov_matrix  <- std_devs %*% corr_matrix %*% std_devs
+
+corr_matrix
+std_devs
+```
+
+La matriz de covarianzas es:
+
+```{r 2d-example-cov}
+cov_matrix <- std_devs %*% corr_matrix %*% std_devs
+cov_matrix
+```
+:::
 
@@
 Es posible encontrar en el espacio latente $\mathbf{\Theta}$ una matriz de
-transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal
-que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de covarianzas de
-$\mathbf{\theta}$ representado en $\mathcal{U}$, $\mathbf{\theta}^\mathcal{U}$)
-sea una matriz diagonal, siendo $\mathbf{M} = \mathbf{R}^{-1}$.
+transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal que
+$\mathbf{\Sigma}^\mathcal{U} = \mathbf{S}^2$, siendo
+$\mathbf{M} = \mathbf{R}^{-1}$.
 
-Además, esta matriz invariante en "escala"; es decir, conservará las "varianzas"
-de $\mathbf{\Theta}$: $\mathbf{\Sigma}^\mathcal{U} = \mathbf{S}^2$.
+Es decir, que esta matriz será "invariante en escala" (conservará las varianzas
+de $\mathbf{\Theta}^\mathcal(U)$ iguales a las de $\mathbf{\Theta}$).
+
+::: note
+## Ejemplo
+
+La matriz de covarianzas de $\mathbf{\Theta}^\mathcal(U)$ que se espera hallar
+es
+
+```{r 2d-example-orthonormal-cov}
+std_devs %*% t(std_devs)
+```
+:::
 
@@
 
+::: note
+## Ejemplo
+
+Supongamos una matriz $\mathbf{P}$ hallada siguiendo estos resultados (ver más
+abajo). Vamos a comprobar que las relaciones entre las matrices se cumplen según
+se espera.
+
+```{r basis-computation}
+eigendecomposition <- cov_matrix |> eigen()
+lambda             <- eigendecomposition$values
+Q                  <- eigendecomposition$vectors
+
+# Base calculada:
+lambda_sqrt <- eigendecomposition$values^(-.5) |> diag()
+
+P <- std_devs %*% lambda_sqrt %*% solve(Q)
+```
+
+```{r 2d-example-transform-cov}
+P %*% cov_matrix %*% t(P)
+```
+:::
+
 y, operando, obtenemos
@@
 $$
-\mathbf{P}^{-1} \mathbf{S}^2 (\mathbf{P}^{-1})^T = \mathbf{\Sigma}.
+\mathbf{P}^{-1} \mathbf{S} \mathbf{S} (\mathbf{P}^{-1})^T = \mathbf{\Sigma}.
 $$
+
+::: note
+## Ejemplo
+
+```{r 2d-example-original-cov}
+P_inv <- solve(P)
+P_inv %*% std_devs %*% std_devs %*% t(P_inv)
+```
+:::
 
@@
 $$
-\mathbf{S} (\mathbf{P}^T)^{-1} (\mathbf{P}^T)^{-1} \mathbf{S} =
-  \mathbf{\Sigma} =
-  \mathbf{S} \mathbf{R} \mathbf{S},
+(\mathbf{S} (\mathbf{P}^{-1})^T)^T \mathbf{S} (\mathbf{P}^{-1})^T =
+  \mathbf{\Sigma},
 $$
 
-Dado que $\mathbf{P}$ es una raíz cuadrada de $\mathbf{R}^{-1}$,
-$\mathbf{R} = (\mathbf{P}^T \mathbf{P})^{-1}$.
+::: note
+## Ejemplo
+
+```{r 2d-example-original-transp}
+t(std_devs %*% t(P_inv)) %*% std_devs %*% t(P_inv)
+```
+:::
+
+Llamemos $\mathbf{K} = \mathbf{S} (\mathbf{P}^{-1})^T$, tal que
+$\mathbf{K}^T \mathbf{K} = \mathbf{\Sigma}$. Despejando $\mathbf{P}$, tenemos
+que
 
 $$
-(\mathbf{P}^T \mathbf{P}^T)^{-1} = (\mathbf{P}^T \mathbf{P})^{-1}.
+\mathbf{S}^{-1} \mathbf{K} = (\mathbf{P}^{-1})^T
 $$
 
-Es decir, la solución que buscamos es una matriz $\mathbf{P}$, raíz cuadrada de
-$\mathbf{R}^{-1}$, que
-cumpla la condición de simetría (i.e. $\mathbf{P} = \mathbf{P}^T$).
+$$
+\mathbf{K}^T \mathbf{S}^{-1} = \mathbf{P}^{-1}
+$$
 
-### Solución mediante descomposición espectral de $\mathbf{R}$
+$$
+\mathbf{P} = \mathbf{S} (\mathbf{K}^{-1})^T
+$$
 
-Puesto que $\mathbf{R}^{-1}$ es semidefinida positiva, existen las matrices
+::: note
+## Ejemplo
+
+```{r 2d-example-original-transp}
+K <- std_devs %*% t(P_inv)
+
+# Step 1:
+solve(std_devs) %*% K
+t(P_inv)
+
+# Step 2:
+t(K) %*% solve(std_devs)
+P_inv
+
+# Step 3:
+P
+std_devs %*% solve(t(K))
+```
+:::
+
+### Solución mediante descomposición espectral de $\mathbf{\Sigma}$
+
+Puesto que $\mathbf{\Sigma}$ es semidefinida positiva, existen las matrices
 $\mathbf{Q}$ y $\mathbf{\Lambda}$ tal que
-$\mathbf{R}^{-1} = \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T$, y además:
+$\mathbf{\Sigma} = \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T$, y además:
 
@@
 * $\mathbf{\Lambda}$ es una matriz diagonal.
+
+::: note
+## Ejemplo
+
+```{r 2d-example-eigendecomp-sigma}
+eigendecomposition <- cov_matrix |> eigen()
+lambda             <- eigendecomposition$values |> diag()
+Q                  <- eigendecomposition$vectors
+
+Q %*% lambda %*% t(Q)
+```
+:::
 
@@
 $$
-\mathbf{R}^{-1} = \mathbf{P}^T \mathbf{P} =
+\mathbf{\Sigma} = \mathbf{K}^T \mathbf{K} =
   \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
@@
 
-Igualmente, aplicamos
+::: note
+## Ejemplo
+
+```{r 2d-example-eigendecomp-sigma}
+lambda_sqsrt <- eigendecomposition$values |> diag()
+```
+:::
+
+Por lo tanto,
 
 $$
-\mathbf{P}^T \mathbf{P} =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^{-1}
-    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
-    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+\mathbf{K}^T \mathbf{K} =
+  (\mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
 $$
 
-Igualando $\mathbf{P} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
-tenemos que
-$\mathbf{P}^T = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
-y por tanto $\mathbf{P} = \mathbf{P}^T$.
+$$
+\mathbf{K} = \mathbf{\Lambda}^{1/2} \mathbf{Q}^T,
+$$
+
+y
+
+$$
+\mathbf{P} = \mathbf{S} ((\mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^{-1})^T
+$$
+
+$$
+\mathbf{P} = \mathbf{S} ((\mathbf{Q}^T)^{-1} \mathbf{\Lambda}^{-1/2})^T
+$$
+
+$$
+\mathbf{P} = \mathbf{S} \mathbf{\Lambda}^{-1/2} \mathbf{Q}^{-1}
+$$
 
@@
 ```{r basis-computation}
-# descomposición espectral de Sigma_U:
-eigendecomposition <- corr_matrix |> solve() |> eigen()
-lambda             <- eigendecomposition$values
-Q                  <- eigendecomposition$vectors
-
-# Base calculada:
-lambda_sqrt <- eigendecomposition$values |> sqrt() |> diag()
-Q_inv       <- Q |> t()
-
-# Calculo de P por descomposicion espectral
-P <- Q %*% lambda_sqrt %*% Q_inv
-
-# Comprobación
-
-## Raíz de la inversa de la correlación:
-solve(t(P) %*% P)
-
-## Covarianza en la base transformada:
-P %*% cov_matrix %*% t(P)
 ```
@@
 cov_matrix  <- std_devs %*% corr_matrix %*% std_devs
+
+## Comprobación:
+cov_matrix |> cov2cor()
 ```

### 2024-06-26 notebooks/Corr_based_parameters.qmd
Corrects notebook (WiP)
Instead of assuming that the "scale" is invariant, it assumes that the covariance in the orthonormalized space is simply a diagonal matrix. This assumption is less restrictive, and may make possible finding a "correlation-based" solution that has a solid foundation.

--- 2024-06-26 notebooks/Corr_based_parameters.qmd
@@
 decir, los vectores de la base $\mathcal{B}$ representados en la base
-$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortogonal,
+$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortonormal,
 con producto interno estándar. Sabemos entonces que
-$\mathbf{P}^T \mathbf{P} = \mathbf{M}$[^orth_basis].
-
-[^orth_basis]: Problema: Esto puede **no ser cierto**, ya que para que ello sea
-así, la base $\mathcal{U}$ debe ser ortonormal, y no solamente ortogonal.
+$\mathbf{P}^T \mathbf{P} = \mathbf{M}$.
 
@@
 Es posible encontrar en el espacio latente $\mathbf{\Theta}$ una matriz de
-transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal
-que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de covarianzas de
-$\mathbf{\theta}$ representado en $\mathcal{U}$, $\mathbf{\theta}^\mathcal{U}$)
-sea una matriz diagonal, siendo $\mathbf{M} = \mathbf{R}^{-1}$.
-
-Además, esta matriz invariante en "escala"; es decir, conservará las "varianzas"
-de $\mathbf{\Theta}$: $\mathbf{\Sigma}^\mathcal{U} = \mathbf{S}^2$.
+transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal que
+$\mathbf{\Sigma}^\mathcal{U}$ sea una matriz diagonal, siendo
+$\mathbf{M} = \mathbf{R}^{-1}$.
 
@@
 $$
-\mathbf{\Sigma}^\mathcal{U} = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T.
+(\mathbf{S}^\mathcal{U})^2 = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T,
 $$
 
-Por lo tanto, igualamos
+siendo $\mathbf{S}^\mathcal{U}$ una matriz diagonal de desviaciones típicas de
+$\mathbf{\theta}^\mathcal{U}$.
+
+Puesto que $\mathbf{P}^T \mathbf{P} = \mathbf{R}^{-1}$,
 
 $$
-\mathbf{S}^2 = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T
+(\mathbf{S}^\mathcal{U})^2 =
+  \mathbf{P} \mathbf{S} (\mathbf{P}^T \mathbf{P})^{-1} \mathbf{S} \mathbf{P}^T.
 $$
 
-y, operando, obtenemos
+Operando, tenemos que
 
 $$
-\mathbf{P}^{-1} \mathbf{S}^2 (\mathbf{P}^{-1})^T = \mathbf{\Sigma}.
+\mathbf{P}^{-1} (\mathbf{S}^\mathcal{U})^2 (\mathbf{P}^{-1})^T =
+  \mathbf{S} (\mathbf{P}^{-1})^T
+  \mathbf{P}^{-1} \mathbf{S},
 $$
 
-Como $\mathbf{S}$ es diagonal,
+$$
+\mathbf{P}^{-1} \mathbf{S}^\mathcal{U}
+  (\mathbf{P}^{-1} \mathbf{S}^\mathcal{U})^T =
+    (\mathbf{P}^{-1} \mathbf{S})^T
+    \mathbf{P}^{-1} \mathbf{S},
+$$
+
+Por lo tanto, podemos igualar
+$\mathbf{P}^{-1} \mathbf{S}^\mathcal{U} = (\mathbf{P}^{-1} \mathbf{S})^T$.
+
+Operando,
 
 $$
-\mathbf{S} (\mathbf{P}^T)^{-1} (\mathbf{P}^T)^{-1} \mathbf{S} =
-  \mathbf{\Sigma} =
-  \mathbf{S} \mathbf{R} \mathbf{S},
+\mathbf{S}^\mathcal{U} = \mathbf{P} \mathbf{S} (\mathbf{P}^{-1})^T,
 $$
-
-Dado que $\mathbf{P}$ es una raíz cuadrada de $\mathbf{R}^{-1}$,
-$\mathbf{R} = (\mathbf{P}^T \mathbf{P})^{-1}$.
-
-$$
-(\mathbf{P}^T \mathbf{P}^T)^{-1} = (\mathbf{P}^T \mathbf{P})^{-1}.
-$$
-
-Es decir, la solución que buscamos es una matriz $\mathbf{P}$, raíz cuadrada de
-$\mathbf{R}^{-1}$, que
-cumpla la condición de simetría (i.e. $\mathbf{P} = \mathbf{P}^T$).
-
-### Solución mediante descomposición espectral de $\mathbf{R}$
-
-Puesto que $\mathbf{R}^{-1}$ es semidefinida positiva, existen las matrices
-$\mathbf{Q}$ y $\mathbf{\Lambda}$ tal que
-$\mathbf{R}^{-1} = \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T$, y además:
-
-* $\mathbf{Q}$ es una matriz ortonormal, i.e., $\mathbf{Q}^T = \mathbf{Q}^{-1}$.
-* $\mathbf{\Lambda}$ es una matriz diagonal.
-
-Dado que $\mathbf{\Lambda}$ es diagonal, aplicamos
-$\mathbf{\Lambda} = \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2}$:
-
-$$
-\mathbf{R}^{-1} = \mathbf{P}^T \mathbf{P} =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
-$$
-
-Igualmente, aplicamos
-
-$$
-\mathbf{P}^T \mathbf{P} =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^{-1}
-    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
-  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
-    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
-$$
-
-Igualando $\mathbf{P} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
-tenemos que
-$\mathbf{P}^T = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
-y por tanto $\mathbf{P} = \mathbf{P}^T$.
-
-# Ejemplos
-
-## Ejemplo en $\mathbb{R}^2$
-
-Usamos el ejemplo con la correlación del borrador, y covarianzas de 2 y 3.
-
-```{r 2-dim-example}
-corr_matrix <- c(1, .5, .5, 1) |> matrix(ncol = 2)
-std_devs    <- c(2, 3) |> sqrt() |> diag()
-cov_matrix  <- std_devs %*% corr_matrix %*% std_devs
-
-## Comprobación:
-cov_matrix |> cov2cor()
-```
-
-```{r basis-computation}
-# descomposición espectral de Sigma_U:
-eigendecomposition <- corr_matrix |> solve() |> eigen()
-lambda             <- eigendecomposition$values
-Q                  <- eigendecomposition$vectors
-
-# Base calculada:
-lambda_sqrt <- eigendecomposition$values |> sqrt() |> diag()
-Q_inv       <- Q |> t()
-
-# Calculo de P por descomposicion espectral
-P <- Q %*% lambda_sqrt %*% Q_inv
-
-# Comprobación
-
-## Raíz de la inversa de la correlación:
-solve(t(P) %*% P)
-
-## Covarianza en la base transformada:
-P %*% cov_matrix %*% t(P)
-```
-
-## Ejemplo en $\mathbb{R}^3$
-
-Usamos el ejemplo propuesto por Mario.
-
-```{r 3-dim-example}
-corr_matrix <- c(1, .5, .8, .5, 1, .6, .8, .6, 1) |> matrix(ncol = 3)
-std_devs    <- c(2, 5, 3) |> diag()
-cov_matrix  <- std_devs %*% corr_matrix %*% std_devs
-```
-
-```{r basis-computation}
-```

### 2024-06-27 notebooks/Corr_based_parameters.qmd
Versions counterexample

@@
 format: docx
-callout-icon: false
 editor: source
@@
 
-# Deducción
+# Solución basada en matriz simétrica mediante descomposición espectral
 
@@
 decir, los vectores de la base $\mathcal{B}$ representados en la base
-$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortonormal,
-con producto interno estándar. Sabemos entonces que
-$\mathbf{P}^T \mathbf{P} = \mathbf{M}$.
+$\mathcal{U}$. Asumamos además que $\mathcal{U}$ es una base ortogonal.
+Sabemos entonces que $\mathbf{P}^T \mathbf{P} = \mathbf{M}$.
 
@@
 Es posible encontrar en el espacio latente $\mathbf{\Theta}$ una matriz de
-transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal que
-$\mathbf{\Sigma}^\mathcal{U}$ sea una matriz diagonal, siendo
-$\mathbf{M} = \mathbf{R}^{-1}$.
+transformación $\mathbf{P}$ de $\mathcal{B}$ en $\mathcal{U}$ tal
+que $\mathbf{\Sigma}^\mathcal{U}$ (i.e. la matriz de covarianzas de
+$\mathbf{\theta}$ representado en $\mathcal{U}$, $\mathbf{\theta}^\mathcal{U}$)
+sea una matriz diagonal, siendo $\mathbf{M} = \mathbf{R}^{-1}$.
 
@@
 $$
-(\mathbf{S}^\mathcal{U})^2 = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T,
+\mathbf{\Sigma}^\mathcal{U} = \mathbf{P} \mathbf{\Sigma} \mathbf{P}^T =
+  \mathbf{P} \mathbf{S} \mathbf{R} \mathbf{S} \mathbf{P}^T
 $$
 
-siendo $\mathbf{S}^\mathcal{U}$ una matriz diagonal de desviaciones típicas de
-$\mathbf{\theta}^\mathcal{U}$.
-
-Puesto que $\mathbf{P}^T \mathbf{P} = \mathbf{R}^{-1}$,
+Supongamos que $\mathbf{P}$ es una raíz cuadrada de $\mathbf{R}^{-1}$; es
+decir, $\mathbf{R}^{-1} = \mathbf{P}^T \mathbf{P}$.
 
 $$
-(\mathbf{S}^\mathcal{U})^2 =
-  \mathbf{P} \mathbf{S} (\mathbf{P}^T \mathbf{P})^{-1} \mathbf{S} \mathbf{P}^T.
+\mathbf{\Sigma}^\mathcal{U} =
+  \mathbf{P} \mathbf{S} (\mathbf{P}^T \mathbf{P})^{-1} \mathbf{S} \mathbf{P}^T =
+  \mathbf{P} \mathbf{S} \mathbf{P}^{-1} (\mathbf{P}^{-1})^T \mathbf{S} \mathbf{P}^T
 $$
 
-Operando, tenemos que
+Si $\mathbf{S}$ conmuta con $\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$, entonces
+la ecuación anterior puede expresarse como
 
 $$
-\mathbf{P}^{-1} (\mathbf{S}^\mathcal{U})^2 (\mathbf{P}^{-1})^T =
-  \mathbf{S} (\mathbf{P}^{-1})^T
-  \mathbf{P}^{-1} \mathbf{S},
+\mathbf{\Sigma}^\mathcal{U} =
+  \mathbf{P} \mathbf{P}^{-1} \mathbf{S} \mathbf{S} (\mathbf{P}^{-1})^T \mathbf{P}^T =
+  \mathbf{S}^2
+$$
+
+### Contraejemplo
+
+Lo anterior en general no se cumple, por lo que $\mathbf{\Sigma}^\mathcal{U}$ no
+es, en general, una matriz diagonal bajo estos supuestos.
+
+Un contraejemplo de esto es la solución con la base del ejemplo de
+representación gráfica del borrador y varianzas desiguales. Supongamos
+
+$$
+\mathbf{S} = diag(\sigma_1, \sigma_2)
 $$
@@
 $$
-\mathbf{P}^{-1} \mathbf{S}^\mathcal{U}
-  (\mathbf{P}^{-1} \mathbf{S}^\mathcal{U})^T =
-    (\mathbf{P}^{-1} \mathbf{S})^T
-    \mathbf{P}^{-1} \mathbf{S},
+(\mathbf{P}^{-1})^T = \begin{pmatrix}
+    1 & \rho \\
+    0 & \sqrt{1 - \rho^2}
+  \end{pmatrix}.
 $$
 
-Por lo tanto, podemos igualar
-$\mathbf{P}^{-1} \mathbf{S}^\mathcal{U} = (\mathbf{P}^{-1} \mathbf{S})^T$.
-
-Operando,
+Bajo estos supuestos,
 
 $$
-\mathbf{S}^\mathcal{U} = \mathbf{P} \mathbf{S} (\mathbf{P}^{-1})^T,
+\mathbf{R} = \mathbf{P}^{-1} (\mathbf{P}^{-1})^T = \begin{pmatrix}
+       1 &              0 \\
+    \rho & \sqrt{1 - \rho^2}
+  \end{pmatrix}
+  \begin{pmatrix}
+    1 & \rho \\
+    0 & \sqrt{1 - \rho^2}
+  \end{pmatrix} =
+  \begin{pmatrix}
+       1 & \rho \\
+    \rho &    1
+  \end{pmatrix}
 $$
+
+$$
+\mathbf{\Sigma} = diag(\sigma_1, \sigma_2)
+  \begin{pmatrix}
+       1 & \rho \\
+    \rho &    1
+  \end{pmatrix}
+  diag(\sigma_1, \sigma_2) =
+  \begin{pmatrix}
+    \sigma_1^2             & \sigma_1 \sigma_2 \rho \\
+    \sigma_1 \sigma_2 \rho & \sigma_2^2
+  \end{pmatrix}.
+$$
+
+Veamos que $\mathbf{S}$ No conmuta con $\mathbf{P}^{-1}$ ni con
+$(\mathbf{P}^{-1})^T$:
+
+$$
+\mathbf{S} (\mathbf{P}^{-1})^T = diag(\sigma_1, \sigma_2)
+  \begin{pmatrix}
+    1 & \rho \\
+    0 & \sqrt{1 - \rho^2}
+  \end{pmatrix} =
+  \begin{pmatrix}
+    \sigma_1 & \sigma_1 \rho \\
+    0        & \sigma_2 \sqrt{1 - \rho^2}
+  \end{pmatrix}
+$$
+
+$$
+(\mathbf{P}^{-1})^T \mathbf{S} = \begin{pmatrix}
+    1 & \rho \\
+    0 & \sqrt{1 - \rho^2}
+  \end{pmatrix}
+  diag(\sigma_1, \sigma_2) =
+  \begin{pmatrix}
+    \sigma_1 & \sigma_2 \rho \\
+    0        & \sigma_2 \sqrt{1 - \rho^2}
+  \end{pmatrix}
+$$
+
+$$
+\mathbf{S} \mathbf{P}^{-1} = diag(\sigma_1, \sigma_2)
+  \begin{pmatrix}
+       1 &              0 \\
+    \rho & \sqrt{1 - \rho^2}
+  \end{pmatrix} =
+  \begin{pmatrix}
+    \sigma_1      &              0 \\
+    \sigma_2 \rho & \sigma_2 \sqrt{1 - \rho^2}
+  \end{pmatrix}
+$$
+
+$$
+\mathbf{P}^{-1} \mathbf{S} = \begin{pmatrix}
+       1 &              0 \\
+    \rho & \sqrt{1 - \rho^2}
+  \end{pmatrix}
+  diag(\sigma_1, \sigma_2) =
+  \begin{pmatrix}
+    \sigma_1      &              0 \\
+    \sigma_1 \rho & \sigma_2 \sqrt{1 - \rho^2}
+  \end{pmatrix}
+$$
+
+Estas ecuaciones muestran que la conmutatividad de $\mathbf{S}$ con
+$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$ no se cumple en general (salvo que
+$\sigma_1 = \sigma_2$).
+
+### Condiciones adicionales
+
+A pesar de lo anterior, $\mathbf{S}$ conmuta con cualquier otra matriz que sea
+simultáneamente diagonalizable con ella. En el caso de matrices simétricas, es
+decir, si $\mathbf{P}^{-1} = (\mathbf{P}^{-1})^T$, sabemos que se cumple la
+conmutatividad de $\mathbf{S}$ con $\mathbf{P}^{-1}$ (y por lo tanto también con
+$(\mathbf{P}^{-1})^T$), si
+$\mathbf{S} \mathbf{P}^{-1} = \mathbf{P}^{-1} \mathbf{S}$ es también simétrica.
+
+Intentamos obtener a continuación una solución que haga $\mathbf{P}^{-1}$
+simétrica, y comprobamos si $\mathbf{S} \mathbf{P}^{-1}$ es también simétrica.
+[^diagonalization]
+
+[^diagonalization]: En realidad, la forma idónea de proceder sería encontrar la
+diagonalización simultánea de $\mathbf{S}$ y $\mathbf{P}^{-1}$, ya que si son
+simultáneamente diagonalizables entonces son conmutativas. No obstante, esto es
+más complejo para mí y ahora mismo no tengo claro saber cómo hacerlo, mientras
+que para obtener la $\mathbf{P}^{-1}$ simétrica puedo reaprovechar las
+deducciones de Mario mediante la descomposición espectral.
+
+### Solución mediante descomposición espectral de $\mathbf{R}$
+
+Puesto que $\mathbf{R}$ es semidefinida positiva, existen las matrices
+$\mathbf{Q}$ y $\mathbf{\Lambda}$ tal que
+$\mathbf{R} = \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T$, y además:
+
+* $\mathbf{Q}$ es una matriz ortonormal, i.e., $\mathbf{Q}^T = \mathbf{Q}^{-1}$.
+* $\mathbf{\Lambda}$ es una matriz diagonal.
+
+Dado que $\mathbf{\Lambda}$ es diagonal, aplicamos
+$\mathbf{\Lambda} = \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2}$ para obtener
+$\mathbf{P}^{-1}$ y $(\mathbf{P}^{-1})^T$:
+
+$$
+\mathbf{R} = \mathbf{P}^{-1} (\mathbf{P}^{-1})^T =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+$$
+
+Ahora bien, para obtener una $\mathbf{P}^{-1}$ simétrica, podemos aplicar
+
+$$
+\mathbf{P}^{-1} (\mathbf{P}^{-1})^T =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^{-1}
+    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+$$
+
+Igualando $\mathbf{P}^{-1} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$,
+tenemos que
+$(\mathbf{P}^{-1})^T = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$
+y por tanto $\mathbf{P}^{-1} = (\mathbf{P}^{-1})^T$. Luego 
+$\mathbf{P}^{-1} = \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T$ es una
+solución que hace a $\mathbf{P}^{-1}$ simétrica.
+
+### Comprobación de la simetría de $\mathbf{S} \mathbf{P}^{-1}$
+
+Comprobamos si el producto $\mathbf{S} \mathbf{P}^{-1}$ da lugar a una matriz
+simétrica.
+
+$$
+\mathbf{S} \mathbf{P}^{-1} =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+$$
+
+$$
+(\mathbf{S} \mathbf{P}^{-1})^T =
+  (\mathbf{P}^{-1})^T \mathbf{S}^T = \mathbf{P}^{-1} \mathbf{S}
+$$
+
+$$
+(\mathbf{S} \mathbf{P}^{-1})^T =
+  (\mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T =
+  (\mathbf{Q}^T)^T (\mathbf{\Lambda}^{1/2})^T \mathbf{Q}^T \mathbf{S}^T =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
+$$
+
+$$
+\mathbf{P}^{-1} \mathbf{S} =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
+$$
+
+Según este resultado,
+$(\mathbf{S} \mathbf{P}^{-1})^T = \mathbf{S} \mathbf{P}^{-1}$, pero esto no
+demuestra que $\mathbf{S} \mathbf{P}^{-1}$ sea simétrica ni que $\mathbf{S}$ y
+$\mathbf{P}^{-1}$ sean conmutativas (ver
+[aquí un contraejemplo](https://math.vanderbilt.edu/sapirmv/msapir/jan22.html)
+).
+
+Podemos utilizar las siguientes propiedades: Sabemos que $\mathbf{S}$ y
+$\mathbf{P}$ son semidefinidas positivas[^P_SDP]. Dado que $\mathbf{P}$ es
+semidefinida positiva, $\mathbf{P}^{-1}$ también lo es. Según
+[Meensakshi & Rakian, 1999](https://core.ac.uk/download/pdf/82822897.pdf)
+(referenciado en
+[esta respuesta en Math StackExchange](https://math.stackexchange.com/a/483930)
+), la condición para que $\mathbf{S} \mathbf{P}^{-1}$ sea simétrica es que
+$\mathbf{S} \mathbf{P}^{-1}$ sea "normal", es decir, que
+$\mathbf{S} \mathbf{P}^{-1} (\mathbf{S} \mathbf{P}^{-1})^T =
+  (\mathbf{S} \mathbf{P}^{-1})^T \mathbf{S} \mathbf{P}^{-1}$.
+Ahora bien,
+
+[^P_SDP]: Estoy asumiendo que $\mathbf{P}$ es semidefinida positiva porque
+$\mathbf{R}$ lo es, y $\mathbf{P}T \mathbf{P} = \mathbf{R}$. Puede que haya un
+salto inferencial aquí (i.e., que sea necesario demostrar que $\mathbf{P}$ es
+semidefinida positiva).
+
+$$
+\mathbf{S} \mathbf{P}^{-1} (\mathbf{S} \mathbf{P}^{-1})^T =
+  \mathbf{S} \mathbf{P}^{-1} (\mathbf{P}^{-1})^T \mathbf{S}^T =
+  \mathbf{S} \mathbf{P}^{-1} (\mathbf{P}^{-1})^T \mathbf{S} =
+  \mathbf{\Sigma}.
+$$
+
+Como $\mathbf{\Sigma}$ es simétrica (y semidefinida positiva),
+el producto $\mathbf{S} \mathbf{P}^{-1}$ es conmutativo. Comprobamos no obstante
+la propiedad de "normalidad" de $\mathbf{S} \mathbf{P}^{-1}$:
+
+$$
+\mathbf{S} \mathbf{P}^{-1} (\mathbf{S} \mathbf{P}^{-1})^T =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+    (\mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+    (\mathbf{Q}^T)^T \mathbf{\Lambda}^{1/2}^T \mathbf{Q}^T \mathbf{S}^T =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T
+    \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S} =
+  \mathbf{S} \mathbf{Q} \mathbf{\Lambda} \mathbf{Q}^T \mathbf{S} =
+  \mathbf{S} \mathbf{R} \mathbf{S} =
+  \mathbf{\Sigma}
+$$
+
+$$
+(\mathbf{S} \mathbf{P}^{-1})^T \mathbf{S} \mathbf{P}^{-1} =
+  (\mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T)^T
+    \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
+  (\mathbf{Q}^T)^T \mathbf{\Lambda}^{1/2}^T \mathbf{Q}^T \mathbf{S}^T
+    \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
+  \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T \mathbf{S}
+    \mathbf{S} \mathbf{Q} \mathbf{\Lambda}^{1/2} \mathbf{Q}^T =
+  \mathbf{P}^{-1} \mathbf{S}^2 (\mathbf{P}^{-1})^T
+$$
+
+Esto no parece fácil de demostrar, por desgracia. La ecuación anterior, sin
+embargo, nos permite aplicar el resultado al que queríamos llegar, ya que
+obtenemos que la distribución de rasgos latentes transformada,
+$\mathbf{\theta}^\mathcal{E}$, tiene por matriz de covarianzas $\mathbf{S}^2$.
+
+$$
+\mathbf{\Sigma}^\mathcal{E} = \mathbf{P} \mathbf{S}^2 \mathbf{P}^T
+  \mathbf{P} \mathbf{P}^{-1} \mathbf{S}^2 (\mathbf{P}^{-1})^T \mathbf{P}^T =
+  \mathbf{S}^2
+$$
+
+## Contraejemplo
+
+Lo anterior parece contener razonamientos circulares, por lo que no está claro
+que sea correcto. Vamos a probar con uno de los ejemplos que sabemos que pueden
+ser problemáticos, es decir, cuando las varianzas son desiguales.
+
+```{r 2d-example}
+corr_matrix <- c(1, .5, .5, 1) |> matrix(ncol = 2)
+std_devs    <- c(2, 3) |> diag()
+cov_matrix  <- std_devs %*% corr_matrix %*% std_devs
+
+corr_matrix
+std_devs
+
+cov_matrix <- std_devs %*% corr_matrix %*% std_devs
+cov_matrix
+
+eigendecomposition <- corr_matrix |> eigen()
+lambda             <- eigendecomposition$values
+Q                  <- eigendecomposition$vectors
+
+# Base calculada:
+lambda_sqrt <- eigendecomposition$values |> sqrt() |> diag()
+
+P_inv <- Q %*% lambda_sqrt %*% solve(Q)
+
+# Comprobación de la simetría y conmutatividad:
+std_devs %*% P_inv # No es simétrica
+P_inv %*% std_devs # No es conmutativa
+```
+
+Vemos que, en general, no se cumplen las condiciones de simetría y
+conmutatividad, por lo que la solución no es correct.

### 2024-06-27 notebooks/Corr_based_parameters.qmd
Versions proof of lack of solution

@@
 conmutatividad, por lo que la solución no es correct.
+
+## Demostración de la falta de solución
+
+En general, hay que considerar que estamos partiendo de la premisa de que
+$\mathbf{P} \mathbf{R} \mathbf{P}^T = \mathbf{I}$ y que
+$\mathbf{P} \mathbf{\Sigma} \mathbf{P}^T$ es una matriz diagonal.
+
+Por lo tanto, estamos asumiendo que $\mathbf{\Sigma}$ y $\mathbf{R}$ son
+simultáneamente diagonalizables. Si atendemos a esta respuesta en [Mathematics
+StackExchange](https://math.stackexchange.com/a/1580856) (y asumimos que es
+correcta), vemos que para que dos matrices sean simultáneamente diagonalizables
+han de ser conmutativas.
+
+Dado que $\mathbf{\Sigma} = \mathbf{S} \mathbf{R} \mathbf{S}$,
+
+$$
+\mathbf{\Sigma} \mathbf{R} = \mathbf{S} \mathbf{R} \mathbf{S} \mathbf{R}
+$$
+
+$$
+\mathbf{R} \mathbf{\Sigma} = \mathbf{R} \mathbf{S} \mathbf{R} \mathbf{S}
+$$
+
+Es decir, que $\mathbf{\Sigma}$ y $\mathbf{R}$ son conmutativas si $\mathbf{S}$
+y $\mathbf{R}$ lo son. Pero ya hemos visto varios contraejemplos, e.g.:
+
+```{r non-commutativity-corr-sd}
+corr_matrix %*% std_devs
+std_devs %*% corr_matrix
+```

### 2024-07-05 R/Formulae.R
Moves definition of inverse Gram matrix
Puts the definition of the inverse Gram matrix in the section where it makes more sense, right before it is used.

@@ -245,19 +245,6 @@ INNER_PROD_MAT_ELEMENT_EQ  <- latex_eq(
 )
 INNER_PROD_MATRIX_INDEX    <- latex(AUX_INDEX, DIM_INDEX, .sep = SEP_COMMA)
 
-# Inverse inner product matrix:
-INNER_PROD_MATRIX_INV       <- latex_inverse(INNER_PROD_MATRIX)
-TRANSFORM_MATRIX_INV        <- latex_inverse(TRANSFORM_MATRIX)
-TRANSFORM_MATRIX_TRANSP_INV <- latex_transp("{$TRANSFORM_MATRIX_INV$}")
-INNER_PROD_MATRIX_INV_DEF   <- latex(
-  TRANSFORM_MATRIX_INV,
-  TRANSFORM_MATRIX_TRANSP_INV
-)
-INNER_PROD_MATRIX_INV_EQ    <- latex_eq(
-  INNER_PROD_MATRIX_INV,
-  INNER_PROD_MATRIX_INV_DEF
-)
-
 ### Test space ----
 
 # Test space definition:
@@ -284,8 +271,10 @@ INVARIANCE_EQ               <- latex_eq(
 )
 
 # Standardized discrimination vector:
-DISCR_STD_COORDS_DEF <- latex(TRANSFORM_MATRIX_TRANSP_INV, DISCR_COORDS)
-DISCR_STD_COORDS_EQ  <- latex_eq(
+TRANSFORM_MATRIX_INV        <- latex_inverse(TRANSFORM_MATRIX)
+TRANSFORM_MATRIX_TRANSP_INV <- latex_transp("{$TRANSFORM_MATRIX_INV$}")
+DISCR_STD_COORDS_DEF        <- latex(TRANSFORM_MATRIX_TRANSP_INV, DISCR_COORDS)
+DISCR_STD_COORDS_EQ         <- latex_eq(
   DISCR_STD_COORDS,
   DISCR_STD_COORDS_DEF
 )
@@ -314,6 +303,17 @@ INNER_PROD_DISCR_DEF  <- latex(
   DISCR_COORDS_AUX
 )
 INNER_PROD_DISCR_EQ  <- latex_eq(INNER_PROD_DISCR, INNER_PROD_DISCR_DEF)
+
+# Inverse inner product matrix:
+INNER_PROD_MATRIX_INV       <- latex_inverse(INNER_PROD_MATRIX)
[…]

### 2024-07-05 R/Formulae.R
Defines coordinates in original test space basis

@@ -251,16 +251,13 @@ INNER_PROD_MATRIX_INDEX    <- latex(AUX_INDEX, DIM_INDEX, .sep = SEP_COMMA)
 TEST_SPACE   <- latex_bf("A")
 REAL_N_SPACE <- latex_raised_to("\\mathbb{R}", exp = N_DIMS)
 
-# Discrimination vector coordinates:
-TEST_SPACE_BASIS        <- latex_cal("B^*")
-DISCR_COORDS            <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_BASIS)
+# Discrimination vector coordinates (in transformed basis):
 TEST_SPACE_STD_BASIS    <- latex_cal("U^*")
 DISCR_STD_COORDS    <- latex_raised_to(DISCR_VECTOR, exp = TEST_SPACE_STD_BASIS)
 DISCR_STD_COORDS_ALT <- latex_coords(DISCR_VECTOR, basis = TEST_SPACE_STD_BASIS)
 DISCR_STD_COORDS_ALT_EQ <- latex_eq(DISCR_STD_COORDS, DISCR_STD_COORDS_ALT)
 
 # Invariance assumption:
-DISCR_COORDS_TRANSP         <- latex_transp(DISCR_COORDS)
 DISCR_STD_COORDS_TRANSP     <- latex_transp(DISCR_STD_COORDS)
 DISCR_TRAIT_STD_COORDS_PROD <- latex(DISCR_STD_COORDS_TRANSP, TRAIT_ORTH_COORDS)
 DISCR_STD_TRANSF_TRAIT_PROD <- latex(DISCR_STD_COORDS_TRANSP, TRAIT_TRANSFORM)
@@ -270,14 +267,11 @@ INVARIANCE_EQ               <- latex_eq(
   DISCR_STD_TRANSF_TRAIT_PROD
 )
 
-# Standardized discrimination vector:
-TRANSFORM_MATRIX_INV        <- latex_inverse(TRANSFORM_MATRIX)
-TRANSFORM_MATRIX_TRANSP_INV <- latex_transp("{$TRANSFORM_MATRIX_INV$}")
-DISCR_STD_COORDS_DEF        <- latex(TRANSFORM_MATRIX_TRANSP_INV, DISCR_COORDS)
-DISCR_STD_COORDS_EQ         <- latex_eq(
[…]

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Changes numbered list in Proposition 2 to "roman"
Instead of arabic numbers I use roman numbers, to avoid confusion with the equations when referencing these list items.
Updates "test space" section titles
Moves definition of inverse Gram matrix
Puts the definition of the inverse Gram matrix in the section where it makes more sense, right before it is used.
Defines coordinates in original test space basis
Drops (premature) mention to "test space"

@@
 Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
-the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
-$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
-invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
-
-### Item response function invariance
+the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$.
+
+### Test space definition
 
@@
 
-with $`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
+with $`r DISCR_COORDS_ALT_EQ`$, and
+$`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
 Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
@@
 as we did with $`r LS_ORTH_BASIS`$.
+As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
+invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
 The norm of $`r DISCR_VECTOR`$ thus fulfills
@@
 
-1. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
-
-1. $`r TRAIT_VECTOR_POLAR_EQ`$
-
-1. $`r TRAIT_VEC_ALT_POLAR_EQ`$
+i. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
+
+i. $`r TRAIT_VECTOR_POLAR_EQ`$
+
+i. $`r TRAIT_VEC_ALT_POLAR_EQ`$
 
@@
 
-As $`r DISCR_VECTOR`$ is defined in test space $`r TEST_SPACE`$ with basis
+As $`r DISCR_VECTOR`$ is defined in space $`r TEST_SPACE`$ with basis
 $`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
@@
 
-## Test space vs. latent space
+## Relationship between test space and latent space
 

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Justifies Proposition 2 in test space direction
The use of Proposition 2 in the derivation of the test space direction cosines must be justified by the "mapping" of the two spaces onto each other, by means of linking them through their respective orthonormal bases.

@@
 Therefore, we may find the direction angles of $`r DISCR_VECTOR`$
-in the test space.
-If $`r ANGLE_VEC_TS_ITEM`$ is
+in the test space:
+As both $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are orthonormal,
+we can map one directly onto the other. Without loss of generality, we may thus
+apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
 the vector of direction angles of $`r DISCR_VECTOR`$,
-applying Proposition 2
 in the test space $`r TEST_SPACE`$ we get

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects terminology ("index" to "parameter")

@@
 which bears no meaning besides its pure algebraic purpose,
-we shall refer to them as the *agnostic* version of the indices:
+we shall refer to them as the *agnostic* version of the parameters:
 
@@
 in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
-*covariance-based* version of the indices:
+*covariance-based* version of the parameters:
 
@@
 are equal to the signs of $`r DISCR_VECTOR`$.
-This property leads directly to generalizing the indices to the monotonically
+This property leads directly to generalizing the parameters to the monotonically
 non-increasing case: The measurement direction relative to dimension
@@
 orthogonality assumption, but it will depend on the correlation matrix
-$`r CORR_MATRIX`$ for the other two versions of the index.
+$`r CORR_MATRIX`$ for the other two versions of the parameter.
 
@@
 we want to avoid changing the scaling of the parameters.
-We will thus prefer the correlation-based over the covariance-based indices.
+We will thus prefer the correlation-based over the covariance-based parameters.
 Thus, we define $`r TRANSFORM_MATRIX`$ such that
@@
 and another one that only considers the correlations among latent dimensions.
-The covariance-based index has the drawback of not fulfilling
+The covariance-based parameter has the drawback of not fulfilling
 the first of @reckase_discriminating_1991's properties, namely,

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects terminology ("index" to "parameter")

--- 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-A latent space where the traits are correlated is easier to understand as a
-geometrical space with non-orthogonal axes,
-often referred to as *general Cartesian coordinates* [@harman_modern_1970].
-Such a space has the advantage that the cosines of the angles between two axes
-can be interpreted as the correlation of the corresponding latent dimensions.
-<!--# Reference needed? -->
-<!-- Problem here: The latent space is not defined like that, according to
-the results of section "Linear algebra of non-orthogonal coordinates". -->
+The representation of a latent space where the traits are correlated is more
+accurate on a geometrical coordinate system with non-orthogonal axes and
+non-standard units, often referred to as
+*general Cartesian coordinates* [@harman_modern_1970].
+Such a space has the advantage that the basis vectors and the axes they yield
+have a meaningful interpretation (albeit counterintuitive, as we will see)
+in a multivariate statistical sense.
 The general case of MIRT model estimation implies estimating possibly
-non-null covariances among latent dimensions, as well as non-standard scales.
+non-standard latent dimensions and non-null covariances among them
+(a more realistic case than an identity matrix).
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
+cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
+[e.g., @thielmann_comparability_2022].
 Moreover, certain applications impose the estimation of
@@
 In the non-cognitive domain, the use of so-called negatively-keyed items
---items that tap the negative pole of a certain trait,
-usually reverse scored-- is also widespread.<!--# Reference and example? -->
+---items that tap the negative pole of a certain trait, and thus
+usually reverse-scored--- is also widespread
+[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
+of the Big-Five Factor Markers, @noauthor_international_nodate].
 It is common to find instruments that combine
@@
 One can easily note that a monotonically decreasing probability for a certain
-dimension is given by a negative discrimination parameter
-(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
+dimension is given by a non-positive discrimination parameter
+(and an unchanging probability by a null discrimination parameter,
+which also applies to cognitive instruments),
 contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
@@
 the formulation of the multidimensional parameters.
-We will introduce two versions of these indices,
-showing that either the original or one of the new ones is obtained
+We will introduce three different versions,
+showing that either the original or one of one of the new ones is obtained
 as a function of how we define and interpret the latent space.
-Next, we discuss the properties of the indices,
-propose a graphical representation method,
-and present two practical examples applying them.
+Next, we discuss their properties,
+propose a graphical representation method based on the new formulation,
+and present two examples of application of our results.
 We round up with a discussion about
@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$ (e.g. a negatively-keyed item).
+latent dimension $`r DIM_INDEX`$.
 Finally, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
 instead of the item difficulty.
 
-Following @reckase_difficulty_1985’s procedure,
-we must find in the first place
-the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
+Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
+$`r LATENT_SPACE`$.
 Because the slope is dependent on the direction considered,
@@
 Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
-the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$.
-
-### Test space definition
-
-Let $`r TEST_SPACE`$ be the vector space made up by
-the set of $`r DISCR_VECTOR`$ vectors,
-with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
-We want @reckase_difficulty_1985's definition of
-the multidimensional parameters in rectangular coordinates to be
-a particular case of our more general definition;
-we can thus consider an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
-in $`r TEST_SPACE`$ such that $`r DISCR_STD_COORDS_ALT_EQ`$
-corresponds in the model to $`r TRAIT_ORTH_COORDS`$.
-Since the response probability to the `r MODEL_ACRONYM`,
-given by Equation \@ref(eq:M2PL-formulation), must be invariant with respect to
-the change of basis expressed in Equation \@ref(eq:change-basis),
-$`r INVARIANCE_EQ`$.
-Therefore, we find a basis $`r TEST_SPACE_BASIS_EQ`$ in $`r TEST_SPACE`$
-defined by
-
-\begin{equation}
-  `r DISCR_STD_COORDS_EQ`,                               (\#eq:discr-vector-std)
-\end{equation}
-
-with $`r DISCR_COORDS_ALT_EQ`$, and
-$`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
-Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
-
-To make @reckase_difficulty_1985's definition a particular case of ours
-we can conveniently assume that the inner product in
-$`r TEST_SPACE_STD_BASIS`$ is the dot product,
-as we did with $`r LS_ORTH_BASIS`$.
-As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
+the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
+$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
 invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
-The norm of $`r DISCR_VECTOR`$ thus fulfills
-
-\begin{equation}
-  `r DISCR_VECTOR_MODULE_EQ`.
-\end{equation}
-
-Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
-keeps the norm of $`r DISCR_VECTOR`$ invariant.
-
-### Direction cosines in the latent space
 
@@
 
-i. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
-
-i. $`r TRAIT_VECTOR_POLAR_EQ`$
-
-i. $`r TRAIT_VEC_ALT_POLAR_EQ`$
-
-Given 1,
+1. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
+
+1. $`r TRAIT_VECTOR_POLAR_EQ`$
+
+1. $`r TRAIT_VEC_ALT_POLAR_EQ`$
+
+Given (1),
 
@@
 
-Using 2 and 3 we get that
+Using (2) and (3) we get that
 
@@
 
-Applying Proposition 1 to
-the model formulation in Equation \@ref(eq:M2PL-formulation),
-the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+From Proposition 1,
+any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
+
+\begin{equation}
+  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
+\end{equation}
+
+Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
 
@@
 
-As $`r DISCR_VECTOR`$ is defined in space $`r TEST_SPACE`$ with basis
-$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
-coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
-\@ref(eq:discr-vector-std),
+We can then define
+
+\begin{equation}
+  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
+\end{equation}
+
+so
 
@@
 \begin{equation}
-  `r SLOPE_MAX_PARAM_EQ`,                                     (\#eq:slope-param)
-\end{equation}
-
-and thus, we have that
+  `r SLOPE_MAX_PARAM_EQ`.                                     (\#eq:slope-param)
+\end{equation}
+
+# Results
+
+## Test space and latent space
+
+Before defining the multidimensional parameters, we need to make a clarification
+regarding the representation of the items in the latent space.
+We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of
+the discrimination vectors.
+This space is where the item vectors can be properly represented.
+It is in general different from the latent space $`r LATENT_SPACE`$,
+although there is a bijective relationship between the two sets of spaces.
+Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
+where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
+in the same basis $`r LS_BASIS`$ in their respective spaces.
+
+Since the response probability must be invariant, because of Equation
+\@ref(eq:change-basis), parameter $`r DISCR_VECTOR`$ must also be transformed
+in $`r DISCR_VECTOR_STD`$.
+Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
+This shows that first a basis change to the standard base is performed and then
+a transformation through $`r INNER_PROD_MATRIX_INV`$,
+so the $`r DISCR_VECTOR_STD`$ are expressed in the standard basis.
+
+\begin{equation}
+  `r DISCR_VECTOR_MODULE_EQ`                     (\#eq:discr-param-squared-norm)
+\end{equation}
+
+Therefore, in the test space, an inner product matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined such that $`r INNER_PROD_DISCR_EQ`$,
+to keep the norm invariant.
+With this definition we have that
 
@@
 
-# Results
-
-## Relationship between test space and latent space
-
-Before presenting the results about the parameters,
-we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
-As shown before, this space is an entity by itself,
-distinct from the latent space $`r LATENT_SPACE`$.
-Despite this, the dot product in the exponent of Equation
-\@ref(eq:M2PL-formulation) induces a bijective relationship between
-their two sets of respective bases.
-Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
-with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
-$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
-with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
-for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
-Only when the two bases are orthonormal,
-the two spaces share a common inner product (i.e., the dot product),
-and thus can be mistaken.
-However, it is worth noting that, although their representation
-can be superimposed in this case,
-they are still two different spaces. <!-- # TODO: To discussion? -->
-Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
-the *test space*, as it is the space where
-the items (i.e., the test elements) are represented.
-
-We must make a few observations here.
-First, the transformation to the orthonormal bases of the two spaces allows
-mapping the $`r MODEL_ACRONYM`$ parameters indistinctly in the two spaces,
-by projecting one space onto the other.
-This allows representing the coordinates of the discrimination vectors
-in the latent space and, conversely,
-the coordinates of the latent trait vectors in the test space.
-As we can see, the invariance of the model holds,
-no matter in which space we represent the coordinates.
-<!-- # TODO: Incluir la condición de invarianza de las coordenadas en los dos espacios (Foto pizarra)? -->
-Second, and most importantly, if the test space is
-where the item parameters naturally lie, they must be represented in this space,
-and not in the latent space.
-Therefore, we may find the direction angles of $`r DISCR_VECTOR`$
-in the test space.
-If $`r ANGLE_VEC_TS_ITEM`$ is
-the vector of direction angles of $`r DISCR_VECTOR`$,
-applying Proposition 2
-in the test space $`r TEST_SPACE`$ we get
+Applying Proposition 1,
 
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
-<!-- # TODO: Revisar esto bien. Esto sólo aplica correctamente si efectivamente
-"equiparamos" las bases U y U*. Además, en la Proposición 2 estamos asumiendo
-que la "base de enlace" es la canónica, no una base ortonormal cualquiera. -->
-As we will see later, this representation is very convenient, as
-the direction of the item vectors with respect to the coordinate axes
-has a direct, meaningful interpretation.
 
@@
 Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
-Equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
+equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
 However, as they depend on the inner product matrix $`r INNER_PROD_MATRIX`$,
@@
 $`r LATENT_SPACE`$ independently of its basis $`r LS_BASIS`$.
-If we assume that $`r BASIS_EQ`$ or, more generally,
-that $`r LS_BASIS`$ is any orthonormal basis, $`r TRANSFORM_MATRIX`$ will be
+If we assume $`r BASIS_EQ`$ or, more generally,
+that $`r LS_BASIS`$ is orthonormal, $`r TRANSFORM_MATRIX`$ will be
 also orthonormal and thus $`r INNER_PROD_MAT_STD_EQ`$.
@@
 and @reckase_discriminating_1991.
-As these expressions implicitly assume the orthonormality of $`r LS_BASIS`$,
-which bears no meaning besides its pure algebraic purpose,
-we shall refer to them as the *agnostic* version of the indices:
+These expresssions disregard the covariance structure of $`r TRAIT_VECTOR`$ by
+implicitly assuming the orthonormality of $`r LS_BASIS`$.
+Therefore, we shall refer to these as the *agnostic* version of the indices:
 
@@
 
-On the other hand, we can let the geometrical representation of
-the latent space $`r LATENT_SPACE`$ account for its covariance structure,
-thus giving $`r LS_BASIS`$ a statistical meaning
-in the context of our MIRT modeling approach.
-To do that, we may assume that $`r TRAIT_COORDS`$ is
-a random vector distributed with covariance $`r COV_MATRIX_EQ`$,
-being $`r COV_MATRIX`$ an $`r N_DIMS`$-dimensional positive definite matrix,
-$`r CORR_MATRIX`$ its corresponding correlation matrix
-(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
-with $`r VAR_COV_EQ`$ (i.e., the variances).
-
-To make the representation of $`r LATENT_SPACE`$ meaningful,
-the axes in $`r LS_BASIS`$ must have
-an interpretation in terms of the latent space structure.
-Although we do not know, in principle, what
-the interpretation of a latent space basis should be, we may assume that
-it must be somewhat related to the covariance structure.
-Therefore, an orthonormal basis should represent independent,
-standard coordinates;
-formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
-$`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$.
-Applying Equation \@ref(eq:change-basis), $`r COV_MATRIX_TRANSF_EQ`$, and thus,
+On the other hand, we can account for the covariance structure of
+the latent space letting $`r LATENT_SPACE`$ be its geometrical representation.
+Thus, we must define $`r LS_BASIS`$ in such a way that
+
+\begin{equation}
+  `r TRAIT_MV_STD_EQ`,
+\end{equation}
+
+where $`r COV_MATRIX_STD`$ is a diagonal matrix.
+Now we may consider two different possibilities:
+We may define $`r TRANSFORM_MATRIX`$ such that $`r INNER_PROD_CORR_COND`$.
+This results in a *correlation-based* version of the indices
+(note that, in this case, $`r DIAG_MATRIX_INNER_PROD_INV`$ is
+an identity matrix):
 
@@
   \begin{split}
-    `r COV_MATRIX_TRANSF_ID_EQ` \\
-    `r COV_MATRIX_RESULT_EQ`.
+    `r MDISC_CORR_PARAM_EQ` \\
+    `r MIL_CORR_PARAM_EQ`.
   \end{split}
@@
 
-Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
-$`r INNER_PROD_COV_COND`$;
-in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
-*covariance-based* version of the indices:
+However, we may also define $`r TRANSFORM_MATRIX`$ such that
+$`r INNER_PROD_COV_COND`$, so
+
+\begin{equation}
+  `r TRAIT_MV_STD_NORM_EQ`.
+\end{equation}
+
+In this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in the
+following *covariance-based* version of the indices:
 
@@
 
-There are three conditions to meet for considering these parameters as
+There are three conditions to meet for regarding these parameters as
 a valid generalization of the (unidimensional) IRT parameters
@@
 and the third one is implicit in the definition of the $`r MDISC_SYM`$.
-Simple arithmetic can show that the first one holds whenever the
+Simple arithmetic can show that the first holds true whenever the
 $`r DIAG_KTH_ELEMENT`$-th element of $`r INNER_PROD_MATRIX_INV`$ equals 1.
 Therefore, both the agnostic and the correlation-based version fulfill it.
-The covariance-based version $`r MDISC_COV_PARAM`$ does not though,
+$`r MDISC_COV_PARAM`$ does not though,
 as $`r COV_ELEMENT`$ is not generally equal to 1.
@@
 ```{r libraries}
-#| warning: false
 ```

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Mentions linear algebra derivations in outline

@@
 In order to pursue these objectives, we first define the generalization of
-the `r MODEL_ACRONYM` model and describe its scope of application.
+the `r MODEL_ACRONYM` model and describe its scope of application,
+which will require introducing some linear algebra concepts and results.
 Then, we follow the method outlined by @reckase_difficulty_1985 and

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects minor errata (repetition)

@@
 We will introduce three different versions,
-showing that either the original or one of one of the new ones is obtained
+showing that either the original or one of the new ones is obtained
 as a function of how we define and interpret the latent space.

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects expression of "orthonormality" condition
Because \mathcal{U} is already considered an orthonormal basis in general, and not just the canonical one (as in previous versions).

--- 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 the origin of the multidimensional latent space to
-the maximally discriminating locus of the item;
-the second one, the direction of the vector from the origin to that same locus.
+the maximally discriminating locus of the item.
+The second one, the direction of the vector from the origin to that same locus.
 Following a similar rationale, @reckase_discriminating_1991 derive a
@@
 a latent trait space, no matter its structure, as an orthogonal space
----exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
-the R package *psych* [@revelle_psych_2024], graphically represents
+[@ackerman_creating_1994]
+---exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
+the R package *psych*, <!--# Reference needed? -->graphically represents
 factor loadings on orthogonal axes even after an oblique rotation.
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration
+is usually implicit in the MIRT literature without further consideration<!--# Examples? -->
 [see @zhang_theoretical_1999 for a notable exception].
@@
 
-The representation of a latent space where the traits are correlated is more
-accurate on a geometrical coordinate system with non-orthogonal axes and
-non-standard units, often referred to as
-*general Cartesian coordinates* [@harman_modern_1970].
-Such a space has the advantage that the basis vectors and the axes they yield
-have a meaningful interpretation (albeit counterintuitive, as we will see)
-in a multivariate statistical sense.
+A latent space where the traits are correlated is easier to understand as a
+geometrical space with non-orthogonal axes,
+often referred to as *general Cartesian coordinates* [@harman_modern_1970].
+Such a space has the advantage that the cosines of the angles between two axes
+can be interpreted as the correlation of the corresponding latent dimensions.
+<!--# Reference needed? -->
+<!-- Problem here: The latent space is not defined like that, according to
+the results of section "Linear algebra of non-orthogonal coordinates". -->
 The general case of MIRT model estimation implies estimating possibly
-non-standard latent dimensions and non-null covariances among them
-(a more realistic case than an identity matrix).
+non-null covariances among latent dimensions, as well as non-standard scales.
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
-[e.g., @thielmann_comparability_2022].
+cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
 Moreover, certain applications impose the estimation of
@@
 In the non-cognitive domain, the use of so-called negatively-keyed items
----items that tap the negative pole of a certain trait, and thus
-usually reverse-scored--- is also widespread
-[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
-of the Big-Five Factor Markers, @noauthor_international_nodate].
+--items that tap the negative pole of a certain trait,
+usually reverse scored-- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine
@@
 One can easily note that a monotonically decreasing probability for a certain
-dimension is given by a non-positive discrimination parameter
-(and an unchanging probability by a null discrimination parameter,
-which also applies to cognitive instruments),
+dimension is given by a negative discrimination parameter
+(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
 contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
@@
 The main objective of this paper is thus to provide a generalization of
-the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
+the `r MODEL_ACRONYM` multidimensional parameters in two aspects:
+On one hand, in an oblique coordinate system
 that can be soundly applied to the correlated trait case; on the other, to
 items that yield monotonous response probabilities, but in the broader sense of
-unchanging monotony. Additionally, we aim to provide a method for
+unchanging monotony. Additionally, we aim at providing a method for
 graphically representing these items in the fashion of @ackerman_creating_1994
@@
 the formulation of the multidimensional parameters.
-We will introduce three different versions,
-showing that either the original or one of one of the new ones is obtained
+We will introduce two versions of these indices,
+showing that either the original or one of the new ones is obtained
 as a function of how we define and interpret the latent space.
-Next, we discuss their properties,
-propose a graphical representation method based on the new formulation,
-and present two examples of application of our results.
+Next, we discuss the properties of the indices,
+propose a graphical representation method,
+and present two practical examples applying them.
 We round up with a discussion about
@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$.
+latent dimension $`r DIM_INDEX`$ (e.g. a negatively-keyed item).
 Finally, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
 instead of the item difficulty.
 
-Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
-$`r LATENT_SPACE`$.
+Following @reckase_difficulty_1985’s procedure,
+we must find in the first place
+the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
 Because the slope is dependent on the direction considered,
@@
 Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
-the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
-$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
+the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$.
+
+### Test space definition
+
+Let $`r TEST_SPACE`$ be the vector space made up by
+the set of $`r DISCR_VECTOR`$ vectors,
+with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
+We want @reckase_difficulty_1985's definition of
+the multidimensional parameters in rectangular coordinates to be
+a particular case of our more general definition;
+we can thus consider an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
+in $`r TEST_SPACE`$ such that $`r DISCR_STD_COORDS_ALT_EQ`$
+corresponds in the model to $`r TRAIT_ORTH_COORDS`$.
+Since the response probability to the `r MODEL_ACRONYM`,
+given by Equation \@ref(eq:M2PL-formulation), must be invariant with respect to
+the change of basis expressed in Equation \@ref(eq:change-basis),
+$`r INVARIANCE_EQ`$.
+Therefore, we find a basis $`r TEST_SPACE_BASIS_EQ`$ in $`r TEST_SPACE`$
+defined by
+
+\begin{equation}
+  `r DISCR_STD_COORDS_EQ`,                               (\#eq:discr-vector-std)
+\end{equation}
+
+with $`r DISCR_COORDS_ALT_EQ`$, and
+$`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
+Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
+
+To make @reckase_difficulty_1985's definition a particular case of ours
+we can conveniently assume that the inner product in
+$`r TEST_SPACE_STD_BASIS`$ is the dot product,
+as we did with $`r LS_ORTH_BASIS`$.
+As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
 invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
+The norm of $`r DISCR_VECTOR`$ thus fulfills
+
+\begin{equation}
+  `r DISCR_VECTOR_MODULE_EQ`.
+\end{equation}
+
+Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
+$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
+keeps the norm of $`r DISCR_VECTOR`$ invariant.
+
+### Direction cosines in the latent space
 
@@
 
-1. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
-
-1. $`r TRAIT_VECTOR_POLAR_EQ`$
-
-1. $`r TRAIT_VEC_ALT_POLAR_EQ`$
-
-Given (1),
+i. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
+
+i. $`r TRAIT_VECTOR_POLAR_EQ`$
+
+i. $`r TRAIT_VEC_ALT_POLAR_EQ`$
+
+Given 1,
 
@@
 
-From Proposition 1,
-any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
-
-\begin{equation}
-  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
-\end{equation}
-
-Thus, the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+Applying Proposition 1 to
+the model formulation in Equation \@ref(eq:M2PL-formulation),
+the `r MODEL_ACRONYM` model is expressed in polar coordinates as
 
@@
 
-We can then define
-
-\begin{equation}
-  `r DISCR_VECTOR_STD_EQ`                                (\#eq:discr-vector-std)
-\end{equation}
-
-so
+As $`r DISCR_VECTOR`$ is defined in space $`r TEST_SPACE`$ with basis
+$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
+coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
+\@ref(eq:discr-vector-std),
 
@@
 \begin{equation}
-  `r SLOPE_MAX_PARAM_EQ`.                                     (\#eq:slope-param)
-\end{equation}
-
-# Results
-
-## Test space and latent space
-
-Before defining the multidimensional parameters, we need to make a clarification
-regarding the representation of the items in the latent space.
-We call $`r TEST_SPACE`$ the *test space* [@zhang_theoretical_1999] of
-the discrimination vectors.
-This space is where the item vectors can be properly represented.
-It is in general different from the latent space $`r LATENT_SPACE`$,
-although there is a bijective relationship between the two sets of spaces.
-Let's consider the model given by Equation \@ref(eq:M2PL-formulation),
-where the latent traits and the parameters $`r DISCR_VECTOR`$ are expressed
-in the same basis $`r LS_BASIS`$ in their respective spaces.
-
-Since the response probability must be invariant, because of Equation
-\@ref(eq:change-basis), parameter $`r DISCR_VECTOR`$ must also be transformed
-in $`r DISCR_VECTOR_STD`$.
-Equating the probabilities we get $a_1^{'}=P^{-1^T}a_i=M^{-1}Pa_i$ .
-This shows that first a basis change to the standard base is performed and then
-a transformation through $`r INNER_PROD_MATRIX_INV`$,
-so the $`r DISCR_VECTOR_STD`$ are expressed in the standard basis.
-
-\begin{equation}
-  `r DISCR_VECTOR_MODULE_EQ`                     (\#eq:discr-param-squared-norm)
-\end{equation}
-
-Therefore, in the test space, an inner product matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined such that $`r INNER_PROD_DISCR_EQ`$,
-to keep the norm invariant.
-With this definition we have that
+  `r SLOPE_MAX_PARAM_EQ`,                                     (\#eq:slope-param)
+\end{equation}
+
+and thus, we have that
 
@@
 
-Applying Proposition 1,
+# Results
+
+## Relationship between test space and latent space
+
+Before presenting the results about the parameters,
+we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
+As shown before, this space is an entity by itself,
+distinct from the latent space $`r LATENT_SPACE`$.
+Despite this, the dot product in the exponent of Equation
+\@ref(eq:M2PL-formulation) induces a bijective relationship between
+their two sets of respective bases.
+Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
+with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
+$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
+with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
+for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
+Only when the two bases are orthonormal,
+the two spaces share a common inner product (i.e., the dot product),
+and thus can be mistaken.
+However, it is worth noting that, although their representation
+can be superimposed in this case,
+they are still two different spaces. <!-- # TODO: To discussion? -->
+Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
+the *test space*, as it is the space where
+the items (i.e., the test elements) are represented.
+
+We must make a few observations here.
+First, the transformation to the orthonormal bases of the two spaces allows
+mapping the $`r MODEL_ACRONYM`$ parameters indistinctly in the two spaces,
+by projecting one space onto the other.
+This allows representing the coordinates of the discrimination vectors
+in the latent space and, conversely,
+the coordinates of the latent trait vectors in the test space.
+As we can see, the invariance of the model holds,
+no matter in which space we represent the coordinates.
+<!-- # TODO: Incluir la condición de invarianza de las coordenadas en los dos espacios (Foto pizarra)? -->
+Second, and most importantly, if the test space is
+where the item parameters naturally lie, they must be represented in this space,
+and not in the latent space.
+Therefore, we may find the direction angles of $`r DISCR_VECTOR`$
+in the test space:
+As both $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are orthonormal,
+we can map one directly onto the other. Without loss of generality, we may thus
+apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
+the vector of direction angles of $`r DISCR_VECTOR`$,
+in the test space $`r TEST_SPACE`$ we get
 
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
+<!-- # TODO: Revisar esto bien. Esto sólo aplica correctamente si efectivamente
+"equiparamos" las bases U y U*. Además, en la Proposición 2 estamos asumiendo
+que la "base de enlace" es la canónica, no una base ortonormal cualquiera. -->
+As we will see later, this representation is very convenient, as
+the direction of the item vectors with respect to the coordinate axes
+has a direct, meaningful interpretation.
 
@@
 
+### Agnostic version
+
 Up to this point,
@@
 $`r LATENT_SPACE`$ independently of its basis $`r LS_BASIS`$.
-If we assume $`r BASIS_EQ`$ or, more generally,
-that $`r LS_BASIS`$ is orthonormal, $`r TRANSFORM_MATRIX`$ will be
-also orthonormal and thus $`r INNER_PROD_MAT_STD_EQ`$.
+If we assume that $`r LS_BASIS`$ is orthonormal, i.e., $`r BASIS_EQ`$,
+$`r TRANSFORM_MATRIX`$ will also be orthonormal
+and thus $`r INNER_PROD_MAT_STD_EQ`$.
 In such a case the multidimensional item location and discrimination simplify
@@
 and @reckase_discriminating_1991.
-These expresssions disregard the covariance structure of $`r TRAIT_VECTOR`$ by
-implicitly assuming the orthonormality of $`r LS_BASIS`$.
-Therefore, we shall refer to these as the *agnostic* version of the indices:
+As these expressions implicitly assume the orthonormality of $`r LS_BASIS`$,
+which bears no meaning besides its pure algebraic purpose,
+we shall refer to them as the *agnostic* version of the parameters:
 
@@
 
-On the other hand, we can account for the covariance structure of
-the latent space letting $`r LATENT_SPACE`$ be its geometrical representation.
-Thus, we must define $`r LS_BASIS`$ in such a way that
-
-\begin{equation}
-  `r TRAIT_MV_STD_EQ`,
-\end{equation}
-
-where $`r COV_MATRIX_STD`$ is a diagonal matrix.
-Now we may consider two different possibilities:
-We may define $`r TRANSFORM_MATRIX`$ such that $`r INNER_PROD_CORR_COND`$.
-This results in a *correlation-based* version of the indices
-(note that, in this case, $`r DIAG_MATRIX_INNER_PROD_INV`$ is
-an identity matrix):
+On the other hand, we can let the geometrical representation of
+the latent space $`r LATENT_SPACE`$ account for its covariance structure,
+thus giving $`r LS_BASIS`$ a statistical meaning
+in the context of our MIRT modeling approach.
+To do that, we may assume that $`r TRAIT_COORDS`$ is
+a random vector distributed with covariance $`r COV_MATRIX_EQ`$,
+being $`r COV_MATRIX`$ an $`r N_DIMS`$-dimensional positive definite matrix,
+$`r CORR_MATRIX`$ its corresponding correlation matrix
+(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
+with $`r VAR_COV_EQ`$ (i.e., the variances).
+
+To make the representation of $`r LATENT_SPACE`$ meaningful,
+the axes in $`r LS_BASIS`$ must have
+an interpretation in terms of the latent space structure.
+Although we do not know, in principle, what
+the interpretation of a latent space basis should be, we may assume that
+it must be somewhat related to the covariance structure.
+Therefore, an orthonormal basis should represent independent,
+standard coordinates;
+formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
+$`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$.
+Applying Equation \@ref(eq:change-basis), $`r COV_MATRIX_TRANSF_EQ`$, and thus,
 
@@
 
-However, we may also define $`r TRANSFORM_MATRIX`$ such that
-$`r INNER_PROD_COV_COND`$, so
-
-\begin{equation}
-  `r TRAIT_MV_STD_NORM_EQ`.
-\end{equation}
-
-In this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in the
-following *covariance-based* version of the indices:
+Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
+$`r INNER_PROD_COV_COND`$;
+in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
+*covariance-based* version of the parameters:
 

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Moves properties to covariance-based section
The properties of the multidimensional parameters need to be in the covariance-based version section, because the noncompliance of this version with the "scale invariance" property (Property 1 in Reckase & McKinley, 1991) is used to justify and introduce the correlation-based version.

@@
 
+### Covariance-based version
+
 On the other hand, we can let the geometrical representation of
@@
 
-## Properties of the parameters
-
-There are three conditions to meet for considering these parameters as
-a valid generalization of the (unidimensional) IRT parameters
-[@reckase_discriminating_1991]:
+According to @reckase_discriminating_1991, the multidimensional parameters must
+meet three conditions for being regarded as a valid generalization
+of the (unidimensional) IRT parameters:
 
@@
 and the third one is implicit in the definition of the $`r MDISC_SYM`$.
-Simple arithmetic can show that the first one holds whenever the
-$`r DIAG_KTH_ELEMENT`$-th element of $`r INNER_PROD_MATRIX_INV`$ equals 1.
-Therefore, both the agnostic and the correlation-based version fulfill it.
-The covariance-based version $`r MDISC_COV_PARAM`$ does not though,
-as $`r COV_ELEMENT`$ is not generally equal to 1.
-Nevertheless, it can be interpreted as a *scaled version* of the index,
+However, when an item measures only dimension $`r DIM_INDEX`$,
+simple arithmetic can show that $`r MDISC_UNIDIM_GENERALIZED_EQ`$.
+In the covariance-based case, this means that $`r MDISC_COV_BASED_UNIDIM_EQ`$
+and, as $`r COV_SYM_DIAG`$ is generally different from 1,
+$`r MDISC_COV_PARAM`$ does not fulfill the first property
+(we shall refer to this property as *scale invariance*, from now on).
+Nevertheless, it can be interpreted as a *scaled version* of the parameter,
 with the standard deviation of the corresponding dimension as scaling factor.
@@
 (i.e., referred to a unitary variance space).
+
+## Properties of the parameters
 

### 2024-07-05 R/Formulae.R

@@ -230,8 +230,9 @@ INNER_PROD_TRANSF <- latex(
 INNER_PROD_EQ     <- latex_eq(INNER_PROD_TRAIT, INNER_PROD_TRANSF)
 
 # Inner product matrix element:
-INNER_PROD_MAT_ELEMENT     <- latex_sub(
-  'm',
+INNER_PROD_MAT_ELEMENT     <- latex('m')
+INNER_PROD_MAT_ELEMENT_ANY <- latex_sub(
+  INNER_PROD_MAT_ELEMENT,
   latex(AUX_INDEX, DIM_INDEX, .sep = NO_SEP)
 )
 BASIS_VECTOR_AUX           <- latex_sub(BASIS_VECTOR, AUX_INDEX)
@@ -240,7 +241,7 @@ INNER_PROD_MAT_ELEMENT_DEF <- latex_innerprod(
   BASIS_VECTOR_ANY
 )
 INNER_PROD_MAT_ELEMENT_EQ  <- latex_eq(
-  INNER_PROD_MAT_ELEMENT,
+  INNER_PROD_MAT_ELEMENT_ANY,
   INNER_PROD_MAT_ELEMENT_DEF
 )
 INNER_PROD_MATRIX_INDEX    <- latex(AUX_INDEX, DIM_INDEX, .sep = SEP_COMMA)
@@ -914,7 +915,9 @@ SD_MATRIX      <- latex_bf('S')
 COV_MATRIX_DEF <- latex(SD_MATRIX, CORR_MATRIX, SD_MATRIX)
 SD_ELEMENT     <- latex_sub('s', "$DIM_INDEX$$DIM_INDEX$")
 VAR_ELEMENT    <- latex_squared(SD_ELEMENT)
-COV_ELEMENT    <- latex_squared(latex_sub("\\sigma", "$DIM_INDEX$$DIM_INDEX$"))
+COV_SYMBOL     <- latex("\\sigma")
+COV_SYM_DIAG   <- COV_SYMBOL   |> latex_sub("$DIM_INDEX$$DIM_INDEX$")
+COV_ELEMENT    <- COV_SYM_DIAG |> latex_squared()
 VAR_COV_EQ     <- latex_eq(VAR_ELEMENT, COV_ELEMENT)
 COV_MATRIX_EQ  <- latex_eq(COV_MATRIX, COV_MATRIX_LS, COV_MATRIX_DEF)
 
[…]

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds new "correlation-based section" version

@@
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
+non-identity latent covariance matrices (e.g., multigroup equating) even when<!-- # Reference needed? -->
 interpreting the latent dimensions as substantive constructs is not a goal.
@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$ (e.g. a negatively-keyed item).
+latent dimension $`r DIM_INDEX`$ (e.g., a negatively-keyed item).
 Finally, as responses are not necessarily interpreted as correct or incorrect,
@@
 
+### Correlation-based version
+
+Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
+we have argued that it is still a valid generalization of the IRT parameters.
+However, it may be convenient to find yet another generalization,
+one that accounts for the latent space structure
+while still fulfilling this property.
+To do this, let's assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
+$`r TRANSFORM_MATRIX_AUX_RESULT`$.
+Then, we have that
+
+$$
+  `r CORR_BASED_DERIVATION`.
+$$
+
+That is, defining $`r TRANSFORM_MATRIX`$ as
+$`r TRANSFORM_MATRIX_CORR_BASED_DEF`$ can be interpreted as re-scaling
+the transformed parameter back to its original metric.
+In this case, we get that $`r INNER_PROD_CORR_COND`$.
+Therefore, we can also define a *correlation-based* version of the
+multidimensional parameters as
+
+\begin{equation}
+  \begin{split}
+    `r MDISC_CORR_PARAM_EQ` \\
+    `r MIL_CORR_PARAM_EQ`.
+  \end{split}
+\end{equation}
+
+As we can see, this version also fulfills the scale invariance property,
+as the $`r INNER_PROD_MAT_INV_DIAG_ELEMENT`$ elements are
+the diagonal elements of $`r CORR_MATRIX`$, which are always equal to 1.
+It is important to note though that this transformation does not have
+an also desirable property,
+which we have used to justify the covariance-based version:
+The $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
+$`r TRAIT_COORDS`$ to the orthonormal basis $`r LS_ORTH_BASIS`$
+are not generally uncorrelated, as one would expect
+(save some exceptions, e.g., when all the variances are equal).
+Nevertheless, it is important to note how the $`r MDISC_CORR_PARAM`$ accounts
+for the correlation structure of the latent space
+while still being scale invariant.
+This will be very useful for the purpose of representing the items graphically,
+as we will see later.
+
 ## Properties of the parameters
@@
 Moreover, in applications where the latent space structure is critical
-(e.g. multi-group IRT, equating, and linking),
+(e.g., multi-group IRT, equating, and linking),
 the agnostic version may overlook relevant differences in

### 2024-07-05 R/Formulae.R

@@ -1003,11 +1003,55 @@ MDISC_COV_BASED_UNIDIM_EQ   <- latex_eq(MDISC_COV_PARAM, MDISC_COV_BASED_UNIDIM)
 
 #### Correlation-based version of the indices: ----
 
-##TODO: Review (maybe delete) this block
+# Ancillary transform matrix:
+TRANSFORM_MATRIX_AUX         <- latex_bf('Q')
+TRANSFORM_MATRIX_AUX_DEF     <- latex(TRANSFORM_MATRIX, SD_MATRIX_INV)
+TRANSFORM_MATRIX_AUX_EQ      <- latex_eq(
+  TRANSFORM_MATRIX_AUX,
+  TRANSFORM_MATRIX_AUX_DEF
+)
+TRANSFORM_MATRIX_AUX_TRANSP  <- TRANSFORM_MATRIX_AUX |> latex_transp()
+TRANSFORM_MATRIX_AUX_APPLIED <- latex(
+  TRANSFORM_MATRIX_AUX,
+  COV_MATRIX,
+  TRANSFORM_MATRIX_AUX_TRANSP
+)
+TRANSFORM_MATRIX_AUX_RESULT <- latex_eq(ID_MATRIX, TRANSFORM_MATRIX_AUX_APPLIED)
+
+# Correlation resulting as inner-product matrix:
+TRANSFORM_MATRIX_AUX_TRANSP_DEF <- latex(SD_MATRIX_INV, TRANSFORM_MATRIX_TRANSP)
+TRANSFORM_MATRIX_AUX_SUBS       <- latex(
+  TRANSFORM_MATRIX_AUX_DEF,
+  COV_MATRIX,
+  TRANSFORM_MATRIX_AUX_TRANSP_DEF
+)
+TRANSFORM_MATRIX_AUX_CORR       <- latex(
+  TRANSFORM_MATRIX_AUX_DEF,
+  SD_MATRIX,
+  CORR_MATRIX,
+  SD_MATRIX,
+  TRANSFORM_MATRIX_AUX_TRANSP_DEF
+)
+TRANSFORM_MATRIX_AUX_FINAL      <- latex(
+  TRANSFORM_MATRIX,
+  CORR_MATRIX,
+  TRANSFORM_MATRIX_TRANSP
+)
+CORR_BASED_DERIVATION           <- latex_eq(
+  TRANSFORM_MATRIX_AUX_SUBS,
[…]

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Updates "properties" section
After the addition to the "covariance-based version" section, the "properties" section must be updated to refer to those previous contents.

@@
 
+We have seen how the two new versions of the parameters fulfill
+the three conditions proposed by @reckase_discriminating_1991
+for valid generalizations of the IRT parameters to the multidimensional case,
+being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
+We will consider now two more properties,
+related to each of the two components of the $`r MIL_PARAM`$ parameter,
+namely, the distance to the origin, and the measurement direction.
+
 The (signed) distance to the origin,

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds dependency of distance on discrimination
Reflows sentences in "properties" section

@@
 its measurement direction forwards or backwards with respect to the origin.
+As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
+the distance will be inversely proportional to the discrimination of the item.
+That is, the more discriminating the item is, the further away its location will
+be from the origin.
 
 The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
-also due to the strict positivity of $`r MDISC_SYM`$,
+also due to the strict positiveness of $`r MDISC_SYM`$,
 are equal to the signs of $`r DISCR_VECTOR`$.
-This property leads directly to generalizing the parameters to the monotonically
-non-increasing case: The measurement direction relative to dimension
-$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
-monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
-When $`r IRF_ABBR`$ is constant with respect to
-variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
+This property leads directly to generalizing
+the parameters to the monotonically non-increasing case:
+The measurement direction relative to dimension $`r DIM_INDEX`$
+will be negative when $`r IRF_ABBR`$ is monotonically decreasing
+with respect to variations along $`r TRAIT_COMPONENT`$.
+When $`r IRF_ABBR`$ is constant
+with respect to variations along $`r TRAIT_COMPONENT`$,
+$`r SIGN_COS_VEC_ITEM_EQ`$,
 which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
-However, note that this does not necessarily imply that the vector is
-parallel to any other axis or strictly contained in the hyperplane formed
-by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
-orthogonality assumption, but it will depend on the correlation matrix
-$`r CORR_MATRIX`$ for the other two versions of the parameter.
+However, note that this does not necessarily imply that
+the vector is parallel to any other axis
+(or strictly contained in the hyperplane formed by other axes, for that matter);
+this will always be true for the $`r MIL_AG_PARAM`$ due to
+the orthogonality assumption,
+but it will depend on the correlation matrix $`r CORR_MATRIX`$
+for the other two versions of the parameter.
 

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects style in new "properties" addition

@@
 As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
-the distance will be inversely proportional to the discrimination of the item.
-That is, the more discriminating the item is, the further away its location will
-be from the origin.
+that displacement will be inversely proportional
+to the discrimination of the item.
+That is, the more discriminating the item is,
+the further away it will be from the origin.
 

### 2024-07-05 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects computation of oblique coordinates
Correctly states the computation of the oblique coordinates (as by a "parallellepiped" with the axes), instead of the "orthogonal projections" on the axes, as was incorrectly stated before.

@@
 and will have a length equal to its $`r MDISC_SYM`$ parameter.
+Its direction will also be given by the direction component of $`r MIL_PARAM`$,
+so its orientation will always pass through the origin.
 
@@
 of our general, possibly non-rectangular coordinates.
-However, this implies pre-multiplying by $`r TRANSFORM_MATRIX`$,
-instead of $`r TRANSFORM_MATRIX_TRANSP_INV`$ as would be expected from
-Equation \@ref(eq:discr-vector-std).
-This is because computing the equivalent rectangular coordinates
-of the oblique ones is different from computing the item parameters
-transformed to an orthogonal space.
-While the latter implies rotating the axes to an orthogonal configuration and
-computing the new coordinates in these axes,
-the former implies transforming the space altogether along with the
-axes transformation.
-
-Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
-that allows us to represent the test space structure
-and the items by transforming their coordinates to
-a rectangular Cartesian system.
+This implies pre-multiplying its coordinates in basis $`r LS_BASIS`$
+by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
+the equivalent, orthonormalized coordinates.
+
+Therefore, we need to define
+a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
+that allows us to represent the test space structure and the items
+in a rectangular Cartesian system.
 However, to facilitate the interpretation,
-we want to avoid changing the scaling of the parameters.
+we want to avoid changing the scale of the parameters.
 We will thus prefer the correlation-based over the covariance-based parameters.
-Thus, we define $`r TRANSFORM_MATRIX`$ such that
-$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.
+Thus, we define $`r TRANSFORM_MATRIX_TRANSP_INV`$
+to be a square root matrix of $`r CORR_MATRIX`$.
 
@@
 the following steps:
-
-1. Define $`r TRANSFORM_MATRIX`$ such that $`r TRANSF_MATRIX_SQ_CORR_INV_EQ`$.
 
@@
 1. Compute the end coordinates $`r END_ITEM_EQ`$.
+
+1. Define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
+   $`r TRANSF_MATRIX_INV_SQ_CORR_EQ`$.
 

### 2024-07-05 R/Formulae.R

@@ -1088,35 +1088,46 @@ SIGN_COS_VEC_ITEM_EQ <- latex_eq(SIGN_COS_VEC_ITEM, 0)
 
 ### Graphical representation: ----
 
-CORR_MATRIX_INV              <- latex_inverse(CORR_MATRIX) # TODO: Borrar (modificar rep. gráfica)?
-TRANSF_MATRIX_SQ_CORR_INV_EQ <- latex_eq(INNER_PROD_TRANSF_DEF, CORR_MATRIX_INV)
+# Item parameters:
 MDISC_CORR_PARAM_ITEM        <- latex_sub(
-  MDISC_SYM,
-  "$CORR_MATRIX$$ITEM_INDEX$"
+MDISC_SYM,
+"$CORR_MATRIX$$ITEM_INDEX$"
 )
 MIL_CORR_PARAM_ITEM          <- latex_sub(
   MIL_PARAM,
   "$CORR_MATRIX$$ITEM_INDEX$"
 )
 
+# Origin coordinates:
+CORR_MATRIX_INV     <- latex_inverse(CORR_MATRIX)
 ORIGIN              <- latex_bf('o')
 ORIGIN_ITEM         <- latex_sub(ORIGIN, ITEM_INDEX)
 DISTANCE_CORR_PARAM <- latex_sub(DISTANCE_SYM, "$CORR_MATRIX$$ITEM_INDEX$")
 DIR_CORR_PARAM      <- latex_sub(DIR_COS_VEC_TS,  "$CORR_MATRIX$$ITEM_INDEX$")
-ORIGIN_ITEM_COMP    <- latex(DISTANCE_CORR_PARAM, DIR_CORR_PARAM)
+ORIGIN_ITEM_COMP  <- latex(DISTANCE_CORR_PARAM, CORR_MATRIX_INV, DIR_CORR_PARAM)
 ORIGIN_ITEM_EQ      <- latex_eq(ORIGIN_ITEM, ORIGIN_ITEM_COMP)
 
+# End coordinates:
 END           <- latex_bf('e')
 END_ITEM      <- latex_sub(END, ITEM_INDEX)
-END_ITEM_COMP <- latex(ORIGIN_ITEM, '+', MDISC_CORR_PARAM_ITEM, DIR_CORR_PARAM)
+END_ITEM_COMP <- latex(
+  ORIGIN_ITEM,
+  '+',
[…]

### 2024-07-08 output/paper_draft/paper_M2PLM_parameters.Rmd
Computes coordinates by orthonormalization (WiP)
Uses the orthonormalization as a means for computing the coordinates in the transformed space.

@@
 a visual representation of the items.
-A vector plot [@ackerman_creating_1994] is the most appropriate way of
-representing the items graphically;
-it allows one to plot several items altogether and analyze them
+The most appropriate way of representing them graphically is
+a vector plot [@ackerman_creating_1994].
+This kind of plot allows representing several items altogether and analyze them
 in terms of their multidimensional parameters [@ackerman_graphical_1996]:
@@
 so its orientation will always pass through the origin.
+On the basis of these conditions, we will need to compute
+the origin and end coordinates of the vector,
+denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
+
+In an orthogonal basis it is easy to compute the coordinates,
+as they are simply the orthogonal projections on the respective axes.
+In the oblique case it is not so simple though, but we take advantage of the
+orthornormalization introduced above.
+In $`r TEST_SPACE_STD_BASIS`$ , the origin and end coordinates are
+$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
+and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
+according to Equation \@ref(eq:discr-vector-std),
+i.e., $`r ORIGIN_ORTH_EQ`$ and $`r END_ORTH_EQ`$.
+Given the mapping of $`r TEST_SPACE_STD_BASIS`$ basis onto $`r LS_ORTH_BASIS`$,
+Equation \@ref(eq:dir-cos-vec-std) gives
+the direction cosines in $`r TEST_SPACE_STD_BASIS`$.
+For the origin coordinates we have that
 

### 2024-07-08 R/Formulae.R

@@ -1088,6 +1088,22 @@ SIGN_COS_VEC_ITEM_EQ <- latex_eq(SIGN_COS_VEC_ITEM, 0)
 
 ### Graphical representation: ----
 
+# Item coordinates definition:
+ORIGIN            <- latex_bf('o')
+ORIGIN_ITEM       <- ORIGIN      |> latex_sub(ITEM_INDEX)
+ORIGIN_ITEM_BASIS <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
+ORIGIN_ITEM_ORTH  <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
+END              <- latex_bf('e')
+END_ITEM         <- END      |> latex_sub(ITEM_INDEX)
+END_ITEM_BASIS   <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
+END_ITEM_ORTH    <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
+
+# Coordinates equivalente:
+ORIGIN_ORTH_DEF <- latex(TRANSFORM_MATRIX_TRANSP_INV, ORIGIN_ITEM_BASIS)
+ORIGIN_ORTH_EQ  <- latex_eq(ORIGIN_ITEM_ORTH, ORIGIN_ORTH_DEF)
+END_ORTH_DEF    <- latex(TRANSFORM_MATRIX_TRANSP_INV, END_ITEM_BASIS)
+END_ORTH_EQ     <- latex_eq(END_ITEM_ORTH, ORIGIN_ORTH_DEF)
+
 # Item parameters:
 MDISC_CORR_PARAM_ITEM        <- latex_sub(
 MDISC_SYM,
@@ -1100,16 +1116,12 @@ MIL_CORR_PARAM_ITEM          <- latex_sub(
 
 # Origin coordinates:
 CORR_MATRIX_INV     <- latex_inverse(CORR_MATRIX)
-ORIGIN              <- latex_bf('o')
-ORIGIN_ITEM         <- latex_sub(ORIGIN, ITEM_INDEX)
 DISTANCE_CORR_PARAM <- latex_sub(DISTANCE_SYM, "$CORR_MATRIX$$ITEM_INDEX$")
[…]

### 2024-07-08 R/Formulae.R
Adds formulas for computing coordinates (WiP)

@@ -1088,6 +1088,42 @@ SIGN_COS_VEC_ITEM_EQ <- latex_eq(SIGN_COS_VEC_ITEM, 0)
 
 ### Graphical representation: ----
 
+# Item coordinates definition:
+ORIGIN            <- latex_bf('o')
+ORIGIN_ITEM       <- ORIGIN      |> latex_sub(ITEM_INDEX)
+ORIGIN_ITEM_BASIS <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
+ORIGIN_ITEM_ORTH  <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
+END              <- latex_bf('e')
+END_ITEM         <- END      |> latex_sub(ITEM_INDEX)
+END_ITEM_BASIS   <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
+END_ITEM_ORTH    <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
+
+# Coordinates equivalence:
+ORIGIN_ORTH_DEF <- latex(TRANSFORM_MATRIX_TRANSP_INV, ORIGIN_ITEM_BASIS)
+ORIGIN_ORTH_EQ  <- latex_eq(ORIGIN_ITEM_ORTH, ORIGIN_ORTH_DEF)
+END_ORTH_DEF    <- latex(TRANSFORM_MATRIX_TRANSP_INV, END_ITEM_BASIS)
+END_ORTH_EQ     <- latex_eq(END_ITEM_ORTH, ORIGIN_ORTH_DEF)
+
+# Origin coordinates:
+ORIGIN_ORTH_DIR_DEF   <- latex(DISTANCE_PARAM, DIR_COS_VEC_STD)
+ORIGIN_ORTH_DISCR_DEF <- latex(DISTANCE_PARAM, ) ## TODO: Complete
+
+CORR_MATRIX_INV     <- latex_inverse(CORR_MATRIX)
+DISTANCE_CORR_PARAM <- latex_sub(DISTANCE_SYM, "$CORR_MATRIX$$ITEM_INDEX$")
+DIR_CORR_PARAM      <- latex_sub(DIR_COS_VEC_TS,  "$CORR_MATRIX$$ITEM_INDEX$")
[…]

### 2024-07-09 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds formulas for computing coordinates (WiP)
Adds derivation of item coordinates

@@
 so its orientation will always pass through the origin.
-On the basis of these conditions, we will need to compute
+On the basis of these conditions, we need to compute
 the origin and end coordinates of the vector,
@@
 In an orthogonal basis it is easy to compute the coordinates,
-as they are simply the orthogonal projections on the respective axes.
+as they are simply the orthogonal projections onto the corresponding axis.
 In the oblique case it is not so simple though, but we take advantage of the
 orthornormalization introduced above.
-In $`r TEST_SPACE_STD_BASIS`$ , the origin and end coordinates are
+In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
 $`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
@@
 according to Equation \@ref(eq:discr-vector-std),
-i.e., $`r ORIGIN_ORTH_EQ`$ and $`r END_ORTH_EQ`$.
-Given the mapping of $`r TEST_SPACE_STD_BASIS`$ basis onto $`r LS_ORTH_BASIS`$,
-Equation \@ref(eq:dir-cos-vec-std) gives
-the direction cosines in $`r TEST_SPACE_STD_BASIS`$.
-For the origin coordinates we have that
+i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
+Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
+the direction cosines in $`r TEST_SPACE_STD_BASIS`$
+are $`r DIR_COS_ITEM_VEC_STD`$.
+Therefore,
+
+\begin{align}
+  `r ORIGIN_ORTH_EQ` \\
+  `r END_ORTH_EQ`.
+\end{align}
+
+Applying Equation \@ref(eq:dir-cos-vec-std),
+for the origin coordinates we have that
+
+\begin{equation}
+  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
+\end{equation}
+
+which, operating, results in
+
+\begin{align}
+  `r ORIGIN_ITEM_COORDS_EQ`    \\
+  `r ORIGIN_TRANSF_SQUARED_EQ` \\
+  `r ORIGIN_INNERPROD_EQ`      \\
+  `r ORIGIN_INNERPROD_DIAG_EQ`;
+\end{align}
+
+\begin{equation}
+  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
+\end{equation}
+
+The end coordinates can be computed by a similar procedure, which results in
+
+\begin{equation}
+  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
+\end{equation}
 

### 2024-07-09 R/Formulae.R

@@ -315,8 +315,7 @@ DISCR_COORDS_TRANSP      <- latex_transp(DISCR_COORDS)
 DISCR_VECTOR_TRANSF_PROD <- latex(
   DISCR_COORDS_TRANSP,
   TRANSFORM_MATRIX_INV,
-  TRANSFORM_MATRIX_TRANSP_INV,
-  DISCR_COORDS
+  DISCR_STD_COORDS_DEF
 )
 DISCR_VECTOR_INNER_PROD  <- latex(
   DISCR_COORDS_TRANSP,
@@ -1092,36 +1091,112 @@ ORIGIN            <- latex_bf('o')
 ORIGIN_ITEM       <- ORIGIN      |> latex_sub(ITEM_INDEX)
 ORIGIN_ITEM_BASIS <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
 ORIGIN_ITEM_ORTH  <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
-END              <- latex_bf('e')
-END_ITEM         <- END      |> latex_sub(ITEM_INDEX)
-END_ITEM_BASIS   <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
-END_ITEM_ORTH    <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
+END               <- latex_bf('e')
+END_ITEM          <- END      |> latex_sub(ITEM_INDEX)
+END_ITEM_BASIS    <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
+END_ITEM_ORTH     <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
 
-# Coordinates equivalence:
-ORIGIN_ORTH_DEF <- latex(TRANSFORM_MATRIX_TRANSP_INV, ORIGIN_ITEM_BASIS)
-ORIGIN_ORTH_EQ  <- latex_eq(ORIGIN_ITEM_ORTH, ORIGIN_ORTH_DEF)
-END_ORTH_DEF    <- latex(TRANSFORM_MATRIX_TRANSP_INV, END_ITEM_BASIS)
[…]

### 2024-07-09 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds new section "Vector representation"

@@
 
-## Properties of the parameters
+## Vector representation
+
+According to @ackerman_creating_1994, the most appropriate way of representing
+multidimensional items is in vector form.
+This allows representing several items altogether and analyzing them
+in terms of their multidimensional parameters [@ackerman_graphical_1996].
+An item vector will be applied at a direction and (signed) distance
+from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter,
+and will have a length equal to its $`r MDISC_SYM`$ parameter.
+Its direction will also be given by the direction component of $`r MIL_PARAM`$,
+so its orientation will always pass through the origin.
+On the basis of these conditions, we need to compute
+the origin and end coordinates of the vector,
+denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
+
+In an orthogonal basis it is easy to compute the coordinates,
+as they are simply the orthogonal projections onto the corresponding axis.
+In the oblique case it is not so simple though, but we can take advantage of the
+orthornormalization introduced above.
+In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
+$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
+and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
+according to Equation \@ref(eq:discr-vector-std),
+i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
+Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
+the direction cosines in $`r TEST_SPACE_STD_BASIS`$
+are $`r DIR_COS_ITEM_VEC_STD`$.
+Therefore,
+
+\begin{align}
+  `r ORIGIN_ORTH_EQ` \\
+  `r END_ORTH_EQ`.
+\end{align}
+
+Applying Equation \@ref(eq:dir-cos-vec-std),
+for the origin coordinates we have that
+
+\begin{equation}
+  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
+\end{equation}
+
+which, operating, results in
+
+\begin{align}
+  `r ORIGIN_ITEM_COORDS_EQ`    \\
+  `r ORIGIN_TRANSF_SQUARED_EQ` \\
+  `r ORIGIN_INNERPROD_EQ`      \\
+  `r ORIGIN_INNERPROD_DIAG_EQ`
+\end{align}
+
+and, finally,
+
+\begin{equation}
+  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
+\end{equation}
+
+The end coordinates can be computed by a similar procedure, which results in
+
+\begin{equation}
+  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
+\end{equation}
+
+Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
+express the items coordinates in terms of their multidimensional parameters.
+However, it is worth noting that the latent space Gram matrix
+also plays a role in computing these coordinates,
+and will therefore have an effect on the representation.
+Of course,
+from Equations \@ref(eq:origin-orth-discr) and \@ref(eq:end-coords-result)
+it is easy to express the coordinates in terms of the parameters in the original
+model formulation, namely,
+
+\begin{align}
+  `r ORIGIN_ITEM_PARAMS_EQ`                  &(\#eq:origin-from-model-params) \\
+  `r END_ITEM_PARAMS_EQ`.                    &(\#eq:end-from-model-params)
+\end{align}
+
+However, formulating them in terms of the multidimensional parameters allows us
+studying the items in terms of their multidimensional properties.
+In the following, we see how to geometrically interpret
+these multidimensional parameters.
+
+### Geometric properties of the items
 
@@
 being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
-We will consider now two more properties,
-related to each of the two components of the $`r MIL_PARAM`$ parameter,
-namely, the distance to the origin, and the measurement direction.
+From a geometrical perspective, the item $`r MDISC_PARAM`$ is simply interpreted
+as the length of its corresponding vector.
+Regarding the $`r MIL_PARAM`$ parameter,
+we will consider now two more properties, related to each of its two components:
+the distance to the origin, and the measurement direction.
 
@@
 a visual representation of the items.
-The most appropriate way of representing them graphically is
-a vector plot [@ackerman_creating_1994].
-This kind of plot allows representing several items altogether and analyze them
-in terms of their multidimensional parameters [@ackerman_graphical_1996]:
-An item vector will be applied at a direction and (signed) distance
-from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
-and will have a length equal to its $`r MDISC_SYM`$ parameter.
-Its direction will also be given by the direction component of $`r MIL_PARAM`$,
-so its orientation will always pass through the origin.
-On the basis of these conditions, we need to compute
-the origin and end coordinates of the vector,
-denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
-
-In an orthogonal basis it is easy to compute the coordinates,
-as they are simply the orthogonal projections onto the corresponding axis.
-In the oblique case it is not so simple though, but we take advantage of the
-orthornormalization introduced above.
-In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
-$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
-and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
-according to Equation \@ref(eq:discr-vector-std),
-i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
-Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
-the direction cosines in $`r TEST_SPACE_STD_BASIS`$
-are $`r DIR_COS_ITEM_VEC_STD`$.
-Therefore,
-
-\begin{align}
-  `r ORIGIN_ORTH_EQ` \\
-  `r END_ORTH_EQ`.
-\end{align}
-
-Applying Equation \@ref(eq:dir-cos-vec-std),
-for the origin coordinates we have that
-
-\begin{equation}
-  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
-\end{equation}
-
-which, operating, results in
-
-\begin{align}
-  `r ORIGIN_ITEM_COORDS_EQ`    \\
-  `r ORIGIN_TRANSF_SQUARED_EQ` \\
-  `r ORIGIN_INNERPROD_EQ`      \\
-  `r ORIGIN_INNERPROD_DIAG_EQ`;
-\end{align}
-
-\begin{equation}
-  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
-\end{equation}
-
-The end coordinates can be computed by a similar procedure, which results in
-
-\begin{equation}
-  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
-\end{equation}
+To do that, we need to chose a version of the parameters
+that yields their most faithful representation.
+We have already seen that
+the agnostic version disregards the latent space structure.
+Therefore, it will only be useful in the unlikely case that
+we have no information of such structure.
+When we have an estimate of the latent space covariances,
+we may use either the covariance-based or the correlation-based version.
+However, we have already seen how
+the distance to the origin depends on the $`r MDISC_SYM`$.
+Also, in the covariance-based,
+the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
+which, as we have argued before, implies a change of scale.
+In the case of the correlation-based version however,
+$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
+being the coordinates scale-invariant.
+Therefore, for graphical representation purposes,
+the correlation-based version will be preferred.
+
+Plotting on a display usually requires providing
+coordinates in a rectangular Cartesian system.
+Therefore, we will need to compute the equivalent rectangular coordinates
+of our general, possibly non-rectangular coordinates.
+This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
+by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
+the equivalent, orthonormalized coordinates.
+Therefore, we will need to define
+a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
+that allows us to represent the test space structure and the items
+in a rectangular Cartesian system.
+
+The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
+the following steps:
+
+1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
+   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
+   respectively.
+
+1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
+   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
+   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.
+
+1. Compute the end coordinates $`r END_ITEM_EQ`$.
+
+1. Define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
+   $`r TRANSF_MATRIX_INV_SQ_CORR_EQ`$.
+
+1. Compute the rectangular coordinates
+   $`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$ by pre-multiplying
+   $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively, by
+   the transformation matrix $`r TRANSFORM_MATRIX`$
+   ($`r ORIGIN_TRANSF_EQ`$; $`r END_TRANSF_EQ`$).
+
+Alternatively,
+one can compute $`r ORIGIN_ITEM_EQ`$ and $`r END_ITEM_EQ`$ directly from
+Equations (\#eq:origin-from-model-params) and (\#eq:end-from-model-params).
+Note though that these equations are not independent of $`r CORR_MATRIX`$,
+as it is used in the computation of $`r DISCR_VECTOR_MODULE`$,
+and thus this matrix is necessary in any case to compute the coordinates.
+
+### Graphical representation example
+
+```{r read-chunks-graphical-example}
+read_chunk("src/Graphical_example_paper.R")
+```
+
+```{r libraries}
+#| warning: false
+```
+
+```{r sources}
+```
+
+```{r constants}
+```
+
+```{r graphical-output-conf}
+```
+
+```{r compute-example-items}
+```
 
@@
 we will limit ourselves to the bidimensional case here.
-Plotting on a two-dimensional display usually requires providing
-coordinates in a rectangular Cartesian system.
-Therefore, we will need to compute the equivalent rectangular coordinates
-of our general, possibly non-rectangular coordinates.
-This implies pre-multiplying its coordinates in basis $`r LS_BASIS`$
-by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
-the equivalent, orthonormalized coordinates.
-
-Therefore, we need to define
-a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
-that allows us to represent the test space structure and the items
-in a rectangular Cartesian system.
-However, to facilitate the interpretation,
-we want to avoid changing the scale of the parameters.
-We will thus prefer the correlation-based over the covariance-based parameters.
-Thus, we define $`r TRANSFORM_MATRIX_TRANSP_INV`$
-to be a square root matrix of $`r CORR_MATRIX`$.
-
-The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
-the following steps:
-
-1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
-   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
-   respectively.
-
-1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
-   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
-   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.
-
-1. Compute the end coordinates $`r END_ITEM_EQ`$.
-
-1. Define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
-   $`r TRANSF_MATRIX_INV_SQ_CORR_EQ`$.
-
-1. Compute the rectangular coordinates
-   $`r ORIGIN_ITEM_TRANSF`$ and $`r END_ITEM_TRANSF`$ by pre-multiplying
-   $`r ORIGIN_ITEM`$ and $`r END_ITEM`$, respectively, by
-   the transformation matrix $`r TRANSFORM_MATRIX`$
-   ($`r ORIGIN_ITEM_TRANSF_EQ`$; $`r END_ITEM_TRANSF_EQ`$).
-
-### Graphical representation example
-
-```{r read-chunks-graphical-example}
-read_chunk("src/Graphical_example_paper.R")
-```
-
-```{r libraries}
-#| warning: false
-```
-
-```{r sources}
-```
-
-```{r constants}
-```
-
-```{r graphical-output-conf}
-```
-
-```{r compute-example-items}
-```
 

### 2024-07-09 R/Formulae.R

@@ -1076,15 +1076,7 @@ MIL_CORR_PARAM_EQ         <- latex_def(
   latex_curlybraces("$DISTANCE_CORR_DEF$, $DIR_COS_ITEM_VEC_CORR_DEF$")
 )
 
-### Properties of the parameters: ----
-
-# Direction cosines:
-ANGLE_VECTORS_ITEM   <- latex_sub(ANGLE, "$ITEM_INDEX$$DIM_INDEX$")
-COS_VECTORS_ITEM     <- latex_cos(ANGLE_VECTORS_ITEM)
-SIGN_COS_VEC_ITEM    <- latex_sign(COS_VECTORS_ITEM, .par = TRUE)
-SIGN_COS_VEC_ITEM_EQ <- latex_eq(SIGN_COS_VEC_ITEM, 0)
-
-### Graphical representation: ----
+### Vector representation: ----
 
 # Item coordinates definition:
 ORIGIN            <- latex_bf('o')
@@ -1105,13 +1097,8 @@ END_TRANSF_EQ     <- latex_eq(END_ITEM_ORTH, END_TRANSF_DEF)
 # Coordinates in orthonormal space:
 ORIGIN_ORTH_DEF  <- latex(DISTANCE_PARAM, DIR_COS_VEC_STD)
 ORIGIN_ORTH_EQ   <- latex_eq(ORIGIN_ITEM_ORTH, ORIGIN_ORTH_DEF, .align = TRUE)
-END_ORTH_DIR_DEF <- latex(
-  ORIGIN_ITEM_ORTH,
-  '+',
-  DISCR_VECTOR_MODULE,
-  DIR_COS_VEC_STD
-)
-END_ORTH_EQ  <- latex_eq(END_ITEM_ORTH, END_ORTH_DIR_DEF, .align = TRUE)
+END_ORTH_DIR_DEF <- latex(ORIGIN_ITEM_ORTH, '+', MDISC_ITEM, DIR_COS_VEC_STD)
+END_ORTH_EQ      <- latex_eq(END_ITEM_ORTH, END_ORTH_DIR_DEF, .align = TRUE)
 
 # Origin coordinates:
 ORIGIN_ORTH_DIR_DEF    <- latex(DISTANCE_PARAM, DIR_COS_VEC_STD)
[…]

### 2024-07-10 output/paper_draft/paper_M2PLM_parameters.Rmd
Makes minor improvements and errata corrections

@@
 being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
-From a geometrical perspective, the item $`r MDISC_PARAM`$ is simply interpreted
+From a geometrical perspective, the item $`r MDISC_SYM`$ is simply interpreted
 as the length of its corresponding vector.
@@
 to the discrimination of the item.
-That is, the more discriminating the item is,
+That is, the less discriminating the item is,
 the further away it will be from the origin.
@@
 a visual representation of the items.
-To do that, we need to chose a version of the parameters
+To do that, we need to choose a version of the parameters
 that yields their most faithful representation.
@@
 
-Alternatively,
-one can compute $`r ORIGIN_ITEM_EQ`$ and $`r END_ITEM_EQ`$ directly from
-Equations (\#eq:origin-from-model-params) and (\#eq:end-from-model-params).
+Instead of steps 1 to 3,
+one can alternatively compute $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$
+directly from Equations \@ref(eq:origin-from-model-params)
+and \@ref(eq:end-from-model-params).
 Note though that these equations are not independent of $`r CORR_MATRIX`$,

### 2024-07-12 output/paper_draft/paper_M2PLM_parameters.Rmd
Addresses "misconceptions" from orthogonality

--- 2024-07-09 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration<!--# Examples? -->
+is usually implicit in the MIRT literature without further consideration
 [see @zhang_theoretical_1999 for a notable exception].
 Moreover, disregarding this geometrical interpretation may lead to
-misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->
-
-A latent space where the traits are correlated is easier to understand as a
-geometrical space with non-orthogonal axes,
-often referred to as *general Cartesian coordinates* [@harman_modern_1970].
-Such a space has the advantage that the cosines of the angles between two axes
-can be interpreted as the correlation of the corresponding latent dimensions.
-<!--# Reference needed? -->
-<!-- Problem here: The latent space is not defined like that, according to
-the results of section "Linear algebra of non-orthogonal coordinates". -->
+critical misconceptions about the measurement model.
+
+The representation of a latent space where the traits are correlated is more
+accurate on a geometrical coordinate system with non-orthogonal axes and
+non-standard units, often referred to as
+*general Cartesian coordinates* [@harman_modern_1970].
+Such a space has the advantage that the basis vectors and the axes they yield
+have a meaningful interpretation (albeit counterintuitive, as we will see)
+in a multivariate statistical sense.
 The general case of MIRT model estimation implies estimating possibly
-non-null covariances among latent dimensions, as well as non-standard scales.
+non-standard latent dimensions and non-null covariances among them
+(a more realistic case than an identity matrix).
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
+cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
+[e.g., @thielmann_comparability_2022].
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g., multigroup equating) even when<!-- # Reference needed? -->
+non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
 interpreting the latent dimensions as substantive constructs is not a goal.
-As we will argue, the usual definitions of the MID and MDISC parameters
-do not generalize to the non-orthonormal geometry
+As we will argue, the usual definitions of the multidimensional difficulty and
+discrimination parameters do not generalize to the non-orthonormal geometry
 necessary to accurately represent correlated latent spaces.
@@
 the formulation of the multidimensional parameters.
-We will introduce two versions of these indices,
+We will introduce three different versions,
 showing that either the original or one of the new ones is obtained
 as a function of how we define and interpret the latent space.
-Next, we discuss the properties of the indices,
-propose a graphical representation method,
-and present two practical examples applying them.
+Next, we discuss their properties,
+propose a graphical representation method based on the new formulation,
+and present two examples of application of our results.
 We round up with a discussion about
@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$ (e.g., a negatively-keyed item).
+latent dimension $`r DIM_INDEX`$.
 Finally, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
 instead of the item difficulty.
 
-Following @reckase_difficulty_1985’s procedure,
-we must find in the first place
-the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
+Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
+$`r LATENT_SPACE`$.
 Because the slope is dependent on the direction considered,
@@
 
-# Results
-
-## Relationship between test space and latent space
-
-Before presenting the results about the parameters,
-we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
-As shown before, this space is an entity by itself,
-distinct from the latent space $`r LATENT_SPACE`$.
-Despite this, the dot product in the exponent of Equation
-\@ref(eq:M2PL-formulation) induces a bijective relationship between
-their two sets of respective bases.
-Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
-with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
-$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
-with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
-for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
-Only when the two bases are orthonormal,
-the two spaces share a common inner product (i.e., the dot product),
-and thus can be mistaken.
-However, it is worth noting that, although their representation
-can be superimposed in this case,
-they are still two different spaces. <!-- # TODO: To discussion? -->
-Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
-the *test space*, as it is the space where
-the items (i.e., the test elements) are represented.
-
-We must make a few observations here.
-First, the transformation to the orthonormal bases of the two spaces allows
-mapping the $`r MODEL_ACRONYM`$ parameters indistinctly in the two spaces,
-by projecting one space onto the other.
-This allows representing the coordinates of the discrimination vectors
-in the latent space and, conversely,
-the coordinates of the latent trait vectors in the test space.
-As we can see, the invariance of the model holds,
-no matter in which space we represent the coordinates.
-<!-- # TODO: Incluir la condición de invarianza de las coordenadas en los dos espacios (Foto pizarra)? -->
-Second, and most importantly, if the test space is
-where the item parameters naturally lie, they must be represented in this space,
-and not in the latent space.
-Therefore, we may find the direction angles of $`r DISCR_VECTOR`$
-in the test space:
-As both $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are orthonormal,
-we can map one directly onto the other. Without loss of generality, we may thus
-apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
-the vector of direction angles of $`r DISCR_VECTOR`$,
-in the test space $`r TEST_SPACE`$ we get
+Applying Proposition 1,
 
@@
 
-### Agnostic version
-
 Up to this point,
@@
 $`r LATENT_SPACE`$ independently of its basis $`r LS_BASIS`$.
-If we assume that $`r LS_BASIS`$ is orthonormal, i.e., $`r BASIS_EQ`$,
-$`r TRANSFORM_MATRIX`$ will also be orthonormal
-and thus $`r INNER_PROD_MAT_STD_EQ`$.
+If we assume $`r BASIS_EQ`$ or, more generally,
+that $`r LS_BASIS`$ is orthonormal, $`r TRANSFORM_MATRIX`$ will be
+also orthonormal and thus $`r INNER_PROD_MAT_STD_EQ`$.
 In such a case the multidimensional item location and discrimination simplify
@@
 and @reckase_discriminating_1991.
-As these expressions implicitly assume the orthonormality of $`r LS_BASIS`$,
-which bears no meaning besides its pure algebraic purpose,
-we shall refer to them as the *agnostic* version of the parameters:
+These expresssions disregard the covariance structure of $`r TRAIT_VECTOR`$ by
+implicitly assuming the orthonormality of $`r LS_BASIS`$.
+Therefore, we shall refer to these as the *agnostic* version of the indices:
 
@@
 
-### Covariance-based version
-
-On the other hand, we can let the geometrical representation of
-the latent space $`r LATENT_SPACE`$ account for its covariance structure,
-thus giving $`r LS_BASIS`$ a statistical meaning
-in the context of our MIRT modeling approach.
-To do that, we may assume that $`r TRAIT_COORDS`$ is
-a random vector distributed with covariance $`r COV_MATRIX_EQ`$,
-being $`r COV_MATRIX`$ an $`r N_DIMS`$-dimensional positive definite matrix,
-$`r CORR_MATRIX`$ its corresponding correlation matrix
-(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
-with $`r VAR_COV_EQ`$ (i.e., the variances).
-
-To make the representation of $`r LATENT_SPACE`$ meaningful,
-the axes in $`r LS_BASIS`$ must have
-an interpretation in terms of the latent space structure.
-Although we do not know, in principle, what
-the interpretation of a latent space basis should be, we may assume that
-it must be somewhat related to the covariance structure.
-Therefore, an orthonormal basis should represent independent,
-standard coordinates;
-formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
-$`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$.
-Applying Equation \@ref(eq:change-basis), $`r COV_MATRIX_TRANSF_EQ`$, and thus,
+On the other hand, we can account for the covariance structure of
+the latent space letting $`r LATENT_SPACE`$ be its geometrical representation.
+Thus, we must define $`r LS_BASIS`$ in such a way that
+
+\begin{equation}
+  `r TRAIT_MV_STD_EQ`,
+\end{equation}
+
+where $`r COV_MATRIX_STD`$ is a diagonal matrix.
+Now we may consider two different possibilities:
+We may define $`r TRANSFORM_MATRIX`$ such that $`r INNER_PROD_CORR_COND`$.
+This results in a *correlation-based* version of the indices
+(note that, in this case, $`r DIAG_MATRIX_INNER_PROD_INV`$ is
+an identity matrix):
 
@@
 
-Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
-$`r INNER_PROD_COV_COND`$;
-in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
-*covariance-based* version of the parameters:
+However, we may also define $`r TRANSFORM_MATRIX`$ such that
+$`r INNER_PROD_COV_COND`$, so
+
+\begin{equation}
+  `r TRAIT_MV_STD_NORM_EQ`.
+\end{equation}
+
+In this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in the
+following *covariance-based* version of the indices:
 
@@
 
-According to @reckase_discriminating_1991, the multidimensional parameters must
-meet three conditions for being regarded as a valid generalization
-of the (unidimensional) IRT parameters:
+## Properties of the parameters
+
+There are three conditions to meet for regarding these parameters as
+a valid generalization of the (unidimensional) IRT parameters
+[@reckase_discriminating_1991]:
 
@@
 and the third one is implicit in the definition of the $`r MDISC_SYM`$.
-However, when an item measures only dimension $`r DIM_INDEX`$,
-simple arithmetic can show that $`r MDISC_UNIDIM_GENERALIZED_EQ`$.
-In the covariance-based case, this means that $`r MDISC_COV_BASED_UNIDIM_EQ`$
-and, as $`r COV_SYM_DIAG`$ is generally different from 1,
-$`r MDISC_COV_PARAM`$ does not fulfill the first property
-(we shall refer to this property as *scale invariance*, from now on).
-Nevertheless, it can be interpreted as a *scaled version* of the parameter,
+Simple arithmetic can show that the first holds true whenever the
+$`r DIAG_KTH_ELEMENT`$-th element of $`r INNER_PROD_MATRIX_INV`$ equals 1.
+Therefore, both the agnostic and the correlation-based version fulfill it.
+$`r MDISC_COV_PARAM`$ does not though,
+as $`r COV_ELEMENT`$ is not generally equal to 1.
+Nevertheless, it can be interpreted as a *scaled version* of the index,
 with the standard deviation of the corresponding dimension as scaling factor.
@@
 
-### Correlation-based version
-
-Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
-we have argued that it is still a valid generalization of the IRT parameters.
-However, it may be convenient to find yet another generalization,
-one that accounts for the latent space structure
-while still fulfilling this property.
-To do this, let's assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
-$`r TRANSFORM_MATRIX_AUX_RESULT`$.
-Then, we have that
-
-$$
-  `r CORR_BASED_DERIVATION`.
-$$
-
-That is, defining $`r TRANSFORM_MATRIX`$ as
-$`r TRANSFORM_MATRIX_CORR_BASED_DEF`$ can be interpreted as re-scaling
-the transformed parameter back to its original metric.
-In this case, we get that $`r INNER_PROD_CORR_COND`$.
-Therefore, we can also define a *correlation-based* version of the
-multidimensional parameters as
-
-\begin{equation}
-  \begin{split}
-    `r MDISC_CORR_PARAM_EQ` \\
-    `r MIL_CORR_PARAM_EQ`.
-  \end{split}
-\end{equation}
-
-As we can see, this version also fulfills the scale invariance property,
-as the $`r INNER_PROD_MAT_INV_DIAG_ELEMENT`$ elements are
-the diagonal elements of $`r CORR_MATRIX`$, which are always equal to 1.
-It is important to note though that this transformation does not have
-an also desirable property,
-which we have used to justify the covariance-based version:
-The $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
-$`r TRAIT_COORDS`$ to the orthonormal basis $`r LS_ORTH_BASIS`$
-are not generally uncorrelated, as one would expect
-(save some exceptions, e.g., when all the variances are equal).
-Nevertheless, it is important to note how the $`r MDISC_CORR_PARAM`$ accounts
-for the correlation structure of the latent space
-while still being scale invariant.
-This will be very useful for the purpose of representing the items graphically,
-as we will see later.
-
-## Properties of the parameters
-
-We have seen how the two new versions of the parameters fulfill
-the three conditions proposed by @reckase_discriminating_1991
-for valid generalizations of the IRT parameters to the multidimensional case,
-being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
-We will consider now two more properties,
-related to each of the two components of the $`r MIL_PARAM`$ parameter,
-namely, the distance to the origin, and the measurement direction.
-
 The (signed) distance to the origin,
-given by the first component of $`r MIL_PARAM`$, has an interesting property:
+given by the second element of $`r MIL_PARAM`$, has an interesting property:
 By the definition of the inner product,
@@
 its measurement direction forwards or backwards with respect to the origin.
-As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
-that displacement will be inversely proportional
-to the discrimination of the item.
-That is, the more discriminating the item is,
-the further away it will be from the origin.
 
 The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
-also due to the strict positiveness of $`r MDISC_SYM`$,
+also due to the strict positivity of $`r MDISC_SYM`$,
 are equal to the signs of $`r DISCR_VECTOR`$.
-This property leads directly to generalizing
-the parameters to the monotonically non-increasing case:
-The measurement direction relative to dimension $`r DIM_INDEX`$
-will be negative when $`r IRF_ABBR`$ is monotonically decreasing
-with respect to variations along $`r TRAIT_COMPONENT`$.
-When $`r IRF_ABBR`$ is constant
-with respect to variations along $`r TRAIT_COMPONENT`$,
-$`r SIGN_COS_VEC_ITEM_EQ`$,
+This property leads directly to generalizing the indices to the monotonically
+non-increasing case: The measurement direction relative to dimension
+$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
+monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
+When $`r IRF_ABBR`$ is constant with respect to
+variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
 which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
-However, note that this does not necessarily imply that
-the vector is parallel to any other axis
-(or strictly contained in the hyperplane formed by other axes, for that matter);
-this will always be true for the $`r MIL_AG_PARAM`$ due to
-the orthogonality assumption,
-but it will depend on the correlation matrix $`r CORR_MATRIX`$
-for the other two versions of the parameter.
+However, note that this does not necessarily imply that the vector is
+parallel to any other axis or strictly contained in the hyperplane formed
+by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
+orthogonality assumption, but it will depend on the correlation matrix
+$`r CORR_MATRIX`$ for the other two versions of the index.
 
@@
 a visual representation of the items.
-The most appropriate way of representing them graphically is
-a vector plot [@ackerman_creating_1994].
-This kind of plot allows representing several items altogether and analyze them
+A vector plot [@ackerman_creating_1994] is the most appropriate way of
+representing the items graphically;
+it allows one to plot several items altogether and analyze them
 in terms of their multidimensional parameters [@ackerman_graphical_1996]:
@@
 and will have a length equal to its $`r MDISC_SYM`$ parameter.
-Its direction will also be given by the direction component of $`r MIL_PARAM`$,
-so its orientation will always pass through the origin.
-On the basis of these conditions, we need to compute
-the origin and end coordinates of the vector,
-denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
-
-In an orthogonal basis it is easy to compute the coordinates,
-as they are simply the orthogonal projections onto the corresponding axis.
-In the oblique case it is not so simple though, but we take advantage of the
-orthornormalization introduced above.
-In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
-$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
-and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
-according to Equation \@ref(eq:discr-vector-std),
-i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
-Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
-the direction cosines in $`r TEST_SPACE_STD_BASIS`$
-are $`r DIR_COS_ITEM_VEC_STD`$.
-Therefore,
-
-\begin{align}
-  `r ORIGIN_ORTH_EQ` \\
-  `r END_ORTH_EQ`.
-\end{align}
-
-Applying Equation \@ref(eq:dir-cos-vec-std),
-for the origin coordinates we have that
-
-\begin{equation}
-  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
-\end{equation}
-
-which, operating, results in
-
-\begin{align}
-  `r ORIGIN_ITEM_COORDS_EQ`    \\
-  `r ORIGIN_TRANSF_SQUARED_EQ` \\
-  `r ORIGIN_INNERPROD_EQ`      \\
-  `r ORIGIN_INNERPROD_DIAG_EQ`;
-\end{align}
-
-\begin{equation}
-  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
-\end{equation}
-
-The end coordinates can be computed by a similar procedure, which results in
-
-\begin{equation}
-  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
-\end{equation}
 
@@
 of our general, possibly non-rectangular coordinates.
-This implies pre-multiplying its coordinates in basis $`r LS_BASIS`$
-by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
-the equivalent, orthonormalized coordinates.
-
-Therefore, we need to define
-a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
-that allows us to represent the test space structure and the items
-in a rectangular Cartesian system.
+However, this implies pre-multiplying by $`r TRANSFORM_MATRIX`$,
+instead of $`r TRANSFORM_MATRIX_TRANSP_INV`$ as would be expected from
+Equation \@ref(eq:discr-vector-std).
+This is because computing the equivalent rectangular coordinates
+of the oblique ones is different from computing the item parameters
+transformed to an orthogonal space.
+While the latter implies rotating the axes to an orthogonal configuration and
+computing the new coordinates in these axes,
+the former implies transforming the space altogether along with the
+axes transformation.
+
+Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
+that allows us to represent the test space structure
+and the items by transforming their coordinates to
+a rectangular Cartesian system.
 However, to facilitate the interpretation,
-we want to avoid changing the scale of the parameters.
-We will thus prefer the correlation-based over the covariance-based parameters.
-Thus, we define $`r TRANSFORM_MATRIX_TRANSP_INV`$
-to be a square root matrix of $`r CORR_MATRIX`$.
+we want to avoid changing the scaling of the parameters.
+We will thus prefer the correlation-based over the covariance-based indices.
+Thus, we define $`r TRANSFORM_MATRIX`$ such that
+$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.
 
@@
 the following steps:
+
+1. Define $`r TRANSFORM_MATRIX`$ such that $`r TRANSF_MATRIX_SQ_CORR_INV_EQ`$.
 
@@
 1. Compute the end coordinates $`r END_ITEM_EQ`$.
-
-1. Define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
-   $`r TRANSF_MATRIX_INV_SQ_CORR_EQ`$.
 
@@
 Moreover, in applications where the latent space structure is critical
-(e.g., multi-group IRT, equating, and linking),
+(e.g. multi-group IRT, equating, and linking),
 the agnostic version may overlook relevant differences in
@@
 and another one that only considers the correlations among latent dimensions.
-The covariance-based parameter has the drawback of not fulfilling
+The covariance-based index has the drawback of not fulfilling
 the first of @reckase_discriminating_1991's properties, namely,

### 2024-07-12 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds pending references

@@
 This is partly due to its close relationship to the common factor model
-[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
+[@mcdonald_test_1999], but also to the formal tractability of its formulation.
 It consists of a generalization of the 2-parameter logistic model
@@
 The representation of a latent space where the traits are correlated is more
-accurate on a geometrical coordinate system with non-orthogonal axes and
-non-standard units, often referred to as
-*general Cartesian coordinates* [@harman_modern_1970].
+accurate on a geometrical coordinate system with non-orthogonal axes
+[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
+often referred to as *general Cartesian coordinates* [@harman_modern_1970].
 Such a space has the advantage that the basis vectors and the axes they yield
@@
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
-interpreting the latent dimensions as substantive constructs is not a goal.
+non-identity latent covariance matrices (e.g., multigroup equating) even when
+interpreting the latent dimensions as substantive constructs is not a goal
+[for for multidimensional linking and equating procedures,
+see @reckase_linking_2009].
 As we will argue, the usual definitions of the multidimensional difficulty and

### 2024-07-12 output/paper_draft/paper_M2PLM_parameters.Rmd
Improves justification of negative discrimination

@@
 
-In the non-cognitive domain, the use of so-called negatively-keyed items
+On an unrelated basis, negative loadings/discrimination parameters
+are often irrelevant in MIRT applications in the cognitive testing domain.
+It has been argued that discrimination parameters
+"are constrained to be positive" [@ackerman_graphical_1996, p. 315],
+or that “in an [independent-clusters] solution, the loadings can always
+be selected to be positive” [@mcdonald_basis_2000, p. 110].
+However, when such instances appear, they are often
+dismissed as “particularly puzzling” (p. 109),
+or implicitly assuming item malfunctioning without further discussion
+[e.g., @ackerman_multidimensional_2005-1].
+In the non-cognitive domain though, the use of so-called negatively-keyed items
 ---items that tap the negative pole of a certain trait, and thus
-usually reverse-scored--- is also widespread
+usually reverse-scored--- is widespread
 [consider e.g. the statement "Neglect my duties" in the Conscientiousness item
 of the Big-Five Factor Markers, @noauthor_international_nodate].
-It is common to find instruments that combine
-positively- and negatively-keyed items.
-Moreover, a complex item may have discrimination parameters with opposite signs
-for different latent dimensions.<!--# Examples? -->
-Take the multidimensional version of the nominal categories model
-[@thissen_nominal_2010] as an example,
-and consider two response categories tapping a different construct each:
-A positive change in the dimension tapped by one category will inevitably
-lead to a decrease in the probability of response to the other category.
-One can easily note that a monotonically decreasing probability for a certain
-dimension is given by a non-positive discrimination parameter
+Instruments that combine positively- and negatively-keyed items
+are relatively common, and
+complex indicators may have discrimination parameters
+with opposite signs for different latent dimensions
+[see, e.g. @mclarnon_be_2016, for item parcels with
+both positive and negative loadings,
+on method factors in the case of the negative ones].
+Following @ackerman_multidimensional_2005-1, one can easily note that
+a monotonically decreasing probability for a certain dimension
+is given by a non-positive discrimination parameter
 (and an unchanging probability by a null discrimination parameter,
 which also applies to cognitive instruments),
-contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
+contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411]
+for determining the multidimensional difficulty.
 However, the multidimensional item parameters can also be defined for

### 2024-07-15 output/paper_draft/paper_M2PLM_parameters.Rmd
Improves discussion after latest results

--- 2024-07-12 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 a latent trait space, no matter its structure, as an orthogonal space
----exploratory factor analysis software, such as SPSS [@ibm_corp_ibm_2022] or
-the R package *psych* [@revelle_psych_2024], graphically represents
+---exploratory factor analysis software, such as SPSS <!--# Reference needed? -->or
+the R package *psych*, <!--# Reference needed? -->graphically represents
 factor loadings on orthogonal axes even after an oblique rotation.
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration
+is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
 [see @zhang_theoretical_1999 for a notable exception].
 Moreover, disregarding this geometrical interpretation may lead to
-critical misconceptions about the measurement model.
-
-The representation of a latent space where the traits are correlated is more
-accurate on a geometrical coordinate system with non-orthogonal axes and
-non-standard units, often referred to as
-*general Cartesian coordinates* [@harman_modern_1970].
-Such a space has the advantage that the basis vectors and the axes they yield
-have a meaningful interpretation (albeit counterintuitive, as we will see)
-in a multivariate statistical sense.
+misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->
+
+A latent space where the traits are correlated is easier to understand as a
+geometrical space with non-orthogonal axes,
+often referred to as *general Cartesian coordinates* [@harman_modern_1970].
+Such a space has the advantage that the cosines of the angles between two axes
+can be interpreted as the correlation of the corresponding latent dimensions.
+<!--# Reference needed? -->
+<!-- Problem here: The latent space is not defined like that, according to
+the results of section "Linear algebra of non-orthogonal coordinates". -->
 The general case of MIRT model estimation implies estimating possibly
@@
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
-[e.g., @thielmann_comparability_2022].
+cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
 Moreover, certain applications impose the estimation of
@@
 ---items that tap the negative pole of a certain trait, and thus
-usually reverse-scored--- is also widespread
-[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
-of the Big-Five Factor Markers, @noauthor_international_nodate].
+usually reverse scored--- is also widespread.<!--# Reference and example? -->
 It is common to find instruments that combine
@@
 dimension is given by a non-positive discrimination parameter
-(and an unchanging probability by a null discrimination parameter,
-which also applies to cognitive instruments),
+(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
 contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
@@
 items that yield monotonous response probabilities, but in the broader sense of
-unchanging monotony. Additionally, we aim to provide a method for
+unchanging monotony. Additionally, we aim at providing a method for
 graphically representing these items in the fashion of @ackerman_creating_1994
@@
 In order to pursue these objectives, we first define the generalization of
-the `r MODEL_ACRONYM` model and describe its scope of application,
-which will require introducing some linear algebra concepts and results.
+the `r MODEL_ACRONYM` model and describe its scope of application.
 Then, we follow the method outlined by @reckase_difficulty_1985 and
@@
 the formulation of the multidimensional parameters.
-We will introduce three different versions,
+We will introduce two versions of these indices,
 showing that either the original or one of the new ones is obtained
 as a function of how we define and interpret the latent space.
-Next, we discuss their properties,
-propose a graphical representation method based on the new formulation,
+Next, we discuss the properties of the indices,
+propose a graphical representation method,
 and present two examples of application of our results.
@@
 the first of @reckase_discriminating_1991's properties, namely,
-that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
+the scale invariance property.
 However, we have provided a rationale for the violation of this property,
@@
 comparable among different latent spaces.
+More importantly, the covariance-based version
+of the multidimensional parameters is invariant to
+changes of latent-space basis.
 Once again, this is important in applications where
-potentially different latent spaces are involved.
-Given these considerations, we propose to adopt the covariance-based version of
-the multidimensional parameters and refer to them simply as
+potentially different bases are involved.
+Given these considerations,
+we propose to adopt the covariance-based version of
+the multidimensional parameters as the most general definition,
+and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of preserving the metric of the parameters,
+has the advantage of being scale-invariant,
 which is convenient for fidelity when representing the items.
-Therefore, we recommend using the correlation-based version of the parameters
-for graphical representation purposes only.
-
-One may consider that the formal derivations presented here are overly
-complicated,
+However, it is worth noting that it has been defined ad-hoc
+to fulfill this specific property.
+Therefore it lacks other desirable properties of the covariance-based version,
+like being invariant to a latent space scaling transformation,
+or yielding a diagonal latent covariance matrix in the corresponding
+orthogonal latent space basis.
+Therefore, we recommend its use for graphical representation purposes only.
+As for the agnostic version, we think it should only be used in
+the (unlikely) case of having no information about the latent space structure.
+
+One may consider that the formal derivations presented here are
+overly complicated,
 when one might simply transform the space to rectangular Cartesian coordinates
@@
 (Reckase, M.D., personal communication, 28 April 2015).
+Such a procedure will of course yield identical results,
+given the invariance of the parameters to a change of basis.
 However, the computations in the oblique space allow us to obtain
@@
 which may give relevant information if these are actual, substantive traits.
-Moreover, strictly following @reckase_difficulty_1985's method in the original,
-oblique space has also provided us with an important insight:
+Computing the parameters in the orthonormalized (or just orthogonalized) space,
+by contrast, would require to compute the cosines in
+the original basis afterwards.
+
+On the other hand, establishing the relationship between the original and
+the orthonormalized basis has provided us with an important insight:
 The distinction between the latent space and the test space.
@@
 its inner product matrix.
+This result is both surprising and counterintuitive;
+The MIRT literature suggests in several instances that
+"the angle between the axes [of latent ability space] represents
+the degree of correlation" [@, p. 16].
+We hope our results help shed a clearer light on
+this often missunderstood topic.
+
 It is worth noting that a similar relationship exists in
@@
 require modeling several response categories.
-Although models such as the graded-scale [@samejima_estimation_1968] and
+Although proposals such as the graded-scale [@samejima_estimation_1968] and
 the nominal response [@bock_estimating_1972] model are closely related to
@@
 Hopefully, this will contribute to a better understanding
-of the underlying measurement theory, its implications and its interpretation.
+of the underlying multidimensional measurement theory, its implications,
+and its interpretation.
 

### 2024-07-15 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds pending reference in "Discussion"
Reference to Ackerman (2005a) needed the bibliographic references updated in branch `review/intro`.

@@
 "the angle between the axes [of latent ability space] represents
-the degree of correlation" [@, p. 16].
+the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
 We hope our results help shed a clearer light on

### 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
Makes minor style improvements to "Introduction"

--- 2024-07-15 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 be selected to be positive” [@mcdonald_basis_2000, p. 110].
-However, when such instances appear, they are often
-dismissed as “particularly puzzling” (p. 109),
+However, when such instances of negative parameters appear,
+they are often dismissed as “particularly puzzling” (p. 109),
 or implicitly assuming item malfunctioning without further discussion
@@
 usually reverse-scored--- is widespread
-[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
+[consider, e.g., the statement "Neglect my duties" in the Conscientiousness item
 of the Big-Five Factor Markers, @noauthor_international_nodate].
@@
 are relatively common, and
-complex indicators may have discrimination parameters
+complex indicators may even have discrimination parameters
 with opposite signs for different latent dimensions
-[see, e.g. @mclarnon_be_2016, for item parcels with
-both positive and negative loadings,
+[see, e.g., @mclarnon_be_2016,
+for item parcels with both positive and negative loadings,
 on method factors in the case of the negative ones].
@@
 the first of @reckase_discriminating_1991's properties, namely,
-the scale invariance property.
+that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
 However, we have provided a rationale for the violation of this property,
@@
 comparable among different latent spaces.
-More importantly, the covariance-based version
-of the multidimensional parameters is invariant to
-changes of latent-space basis.
 Once again, this is important in applications where
-potentially different bases are involved.
-Given these considerations,
-we propose to adopt the covariance-based version of
-the multidimensional parameters as the most general definition,
-and refer to them simply as
+potentially different latent spaces are involved.
+Given these considerations, we propose to adopt the covariance-based version of
+the multidimensional parameters and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of being scale-invariant,
+has the advantage of preserving the metric of the parameters,
 which is convenient for fidelity when representing the items.
-However, it is worth noting that it has been defined ad-hoc
-to fulfill this specific property.
-Therefore it lacks other desirable properties of the covariance-based version,
-like being invariant to a latent space scaling transformation,
-or yielding a diagonal latent covariance matrix in the corresponding
-orthogonal latent space basis.
-Therefore, we recommend its use for graphical representation purposes only.
-As for the agnostic version, we think it should only be used in
-the (unlikely) case of having no information about the latent space structure.
-
-One may consider that the formal derivations presented here are
-overly complicated,
+Therefore, we recommend using the correlation-based version of the parameters
+for graphical representation purposes only.
+
+One may consider that the formal derivations presented here are overly
+complicated,
 when one might simply transform the space to rectangular Cartesian coordinates
@@
 (Reckase, M.D., personal communication, 28 April 2015).
-Such a procedure will of course yield identical results,
-given the invariance of the parameters to a change of basis.
 However, the computations in the oblique space allow us to obtain
@@
 which may give relevant information if these are actual, substantive traits.
-Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require to compute the cosines in
-the original basis afterwards.
-
-On the other hand, establishing the relationship between the original and
-the orthonormalized basis has provided us with an important insight:
+Moreover, strictly following @reckase_difficulty_1985's method in the original,
+oblique space has also provided us with an important insight:
 The distinction between the latent space and the test space.
@@
 its inner product matrix.
-This result is both surprising and counterintuitive;
-The MIRT literature suggests in several instances that
-"the angle between the axes [of latent ability space] represents
-the degree of correlation" [@, p. 16].
-We hope our results help shed a clearer light on
-this often missunderstood topic.
-
 It is worth noting that a similar relationship exists in
@@
 require modeling several response categories.
-Although proposals such as the graded-scale [@samejima_estimation_1968] and
+Although models such as the graded-scale [@samejima_estimation_1968] and
 the nominal response [@bock_estimating_1972] model are closely related to
@@
 Hopefully, this will contribute to a better understanding
-of the underlying multidimensional measurement theory, its implications,
-and its interpretation.
+of the underlying measurement theory, its implications and its interpretation.
 

### 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects minor errata in empirical examples

@@
 This is partly due to its close relationship to the common factor model
-[@mcdonald_test_1999], but also to the formal tractability of its formulation.
+[@mcdonald_test_1999], but also to the formal tractability of its formulation.<!--# Ref. needed? -->
 It consists of a generalization of the 2-parameter logistic model
@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration
+is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
 [see @zhang_theoretical_1999 for a notable exception].
 Moreover, disregarding this geometrical interpretation may lead to
-critical misconceptions about the measurement model.
-
-The representation of a latent space where the traits are correlated is more
-accurate on a geometrical coordinate system with non-orthogonal axes
-[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
+misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->
+
+A latent space where the traits are correlated is easier to understand as a
+geometrical space with non-orthogonal axes,
 often referred to as *general Cartesian coordinates* [@harman_modern_1970].
-Such a space has the advantage that the basis vectors and the axes they yield
-have a meaningful interpretation (albeit counterintuitive, as we will see)
-in a multivariate statistical sense.
+Such a space has the advantage that the cosines of the angles between two axes
+can be interpreted as the correlation of the corresponding latent dimensions.
+<!--# Reference needed? -->
+<!-- Problem here: The latent space is not defined like that, according to
+the results of section "Linear algebra of non-orthogonal coordinates". -->
 The general case of MIRT model estimation implies estimating possibly
@@
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
-[e.g., @thielmann_comparability_2022].
+cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g., multigroup equating) even when
-interpreting the latent dimensions as substantive constructs is not a goal
-[for for multidimensional linking and equating procedures,
-see @reckase_linking_2009].
+non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
+interpreting the latent dimensions as substantive constructs is not a goal.
 As we will argue, the usual definitions of the multidimensional difficulty and
@@
 
-On an unrelated basis, negative loadings/discrimination parameters
-are often irrelevant in MIRT applications in the cognitive testing domain.
-It has been argued that discrimination parameters
-"are constrained to be positive" [@ackerman_graphical_1996, p. 315],
-or that “in an [independent-clusters] solution, the loadings can always
-be selected to be positive” [@mcdonald_basis_2000, p. 110].
-However, when such instances of negative parameters appear,
-they are often dismissed as “particularly puzzling” (p. 109),
-or implicitly assuming item malfunctioning without further discussion
-[e.g., @ackerman_multidimensional_2005-1].
-In the non-cognitive domain though, the use of so-called negatively-keyed items
+In the non-cognitive domain, the use of so-called negatively-keyed items
 ---items that tap the negative pole of a certain trait, and thus
-usually reverse-scored--- is widespread
-[consider, e.g., the statement "Neglect my duties" in the Conscientiousness item
-of the Big-Five Factor Markers, @noauthor_international_nodate].
-Instruments that combine positively- and negatively-keyed items
-are relatively common, and
-complex indicators may even have discrimination parameters
-with opposite signs for different latent dimensions
-[see, e.g., @mclarnon_be_2016,
-for item parcels with both positive and negative loadings,
-on method factors in the case of the negative ones].
-Following @ackerman_multidimensional_2005-1, one can easily note that
-a monotonically decreasing probability for a certain dimension
-is given by a non-positive discrimination parameter
-(and an unchanging probability by a null discrimination parameter,
-which also applies to cognitive instruments),
-contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411]
-for determining the multidimensional difficulty.
+usually reverse scored--- is also widespread.<!--# Reference and example? -->
+It is common to find instruments that combine
+positively- and negatively-keyed items.
+Moreover, a complex item may have discrimination parameters with opposite signs
+for different latent dimensions.<!--# Examples? -->
+Take the multidimensional version of the nominal categories model
+[@thissen_nominal_2010] as an example,
+and consider two response categories tapping a different construct each:
+A positive change in the dimension tapped by one category will inevitably
+lead to a decrease in the probability of response to the other category.
+One can easily note that a monotonically decreasing probability for a certain
+dimension is given by a non-positive discrimination parameter
+(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
+contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
 However, the multidimensional item parameters can also be defined for
@@
 ```
+
+Table \@ref(tab:empirical-table-Reckase2009-out) draws from a classical example
+by @reckase_multidimensional_2009.
+In this 30-test item, there are three ten-item blocks assumed to approximate
+a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
+and items 21-30 dimension 2 (see Table 6.1, p. 153 therein).
+In his example, Reckase tests the estimation of these items
+on two simulated datasets: In the first one, the latent trait distribution
+is standard with null correlations; however, the second one has
+(apart from a non-null vector of means) as covariance matrix
+
+\begin{equation}
+  `r cov_matrix_value_out`,
+\end{equation}
+
+which corresponds to a correlation matrix of
+
+\begin{equation}
+  `r cor_matrix_value_out`.
+\end{equation}
+
+As we can see in Table \@ref(tab:empirical-table-Reckase2009-out), the agnostic
+versions of the multidimensional parameters
+(columns labelled _`r AGNOSTIC_ABBR`_)
+coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
+However, when we take into account the latent space structure
+(columns labelled $`r COV_MATRIX`$), we notice several differences.
+First, the $`r MDISC_ITEM`$ parameters tend to be larger
+than their agnostic counterparts.
+This happens because all the discrimination parameters and the latent space
+correlations are positive; in this respect, these items are similar to item 1
+in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
+However, for the items mostly aligned with dimension 2 (items 21-30), as the
+$`r MDISC_ITEM`$ parameters are also scaled by a standard deviation smaller
+than 1, their values decrease, being some of them smaller than the agnostic
+versions. The $`r DISTANCE_PARAM`$ parameters decrease or increase, consequently
+(see Equation \@ref(eq:M2PL-formula-polar)), thus being most of them smaller
+than the agnostic ones (again, with the exceptions only happening
+in the items aligned with dimension 2).
+Finally, we see that the direction angles are not as clearly separated as one
+might expect from the agnostic versions. This effect is more pronounced among
+the items aligned with dimensions 1 and 3, which are strongly correlated.
+In this situation, we notice that the angles tend to be much smaller, as the
+high correlation induces an alignment between the two dimensions
+(compare item 1 in the two test spaces in
+Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
 

### 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects table captions of empirical examples

@@
 #| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: Item parameters for the empirical example
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
 item_params_output
@@
 #| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: Item parameters for the empirical example
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Tezza et al. (2018).
 item_params_output

### 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
Rephrases conclusion about use of agnostic version

--- 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 be selected to be positive” [@mcdonald_basis_2000, p. 110].
-However, when such instances of negative parameters appear,
-they are often dismissed as “particularly puzzling” (p. 109),
+However, when such instances appear, they are often
+dismissed as “particularly puzzling” (p. 109),
 or implicitly assuming item malfunctioning without further discussion
@@
 usually reverse-scored--- is widespread
-[consider, e.g., the statement "Neglect my duties" in the Conscientiousness item
+[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
 of the Big-Five Factor Markers, @noauthor_international_nodate].
@@
 are relatively common, and
-complex indicators may even have discrimination parameters
+complex indicators may have discrimination parameters
 with opposite signs for different latent dimensions
-[see, e.g., @mclarnon_be_2016,
-for item parcels with both positive and negative loadings,
+[see, e.g. @mclarnon_be_2016, for item parcels with
+both positive and negative loadings,
 on method factors in the case of the negative ones].
@@
 comparable among different latent spaces.
+More importantly, the covariance-based version
+of the multidimensional parameters is invariant to
+changes of latent-space basis.
 Once again, this is important in applications where
-potentially different latent spaces are involved.
-Given these considerations, we propose to adopt the covariance-based version of
-the multidimensional parameters and refer to them simply as
+potentially different bases are involved.
+Given these considerations,
+we propose to adopt the covariance-based version of
+the multidimensional parameters as the most general definition,
+and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of preserving the metric of the parameters,
+has the advantage of being scale-invariant,
 which is convenient for fidelity when representing the items.
-Therefore, we recommend using the correlation-based version of the parameters
-for graphical representation purposes only.
-
-One may consider that the formal derivations presented here are overly
-complicated,
+However, it is worth noting that it has been defined ad-hoc
+to fulfill this specific property.
+Therefore it lacks other desirable properties of the covariance-based version,
+like being invariant to a latent space scaling transformation,
+or yielding a diagonal latent covariance matrix in the corresponding
+orthogonal latent space basis.
+Therefore, we recommend its use for graphical representation purposes only.
+As for the agnostic version, we recommend it only in
+the (unlikely) case of having no information about the latent space structure.
+
+One may consider that the formal derivations presented here are
+overly complicated,
 when one might simply transform the space to rectangular Cartesian coordinates
@@
 its inner product matrix.
+This result is both surprising and counterintuitive;
+The MIRT literature suggests in several instances that
+"the angle between the axes [of latent ability space] represents
+the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
+We hope our results help shed a clearer light on
+this often missunderstood topic.
+
 It is worth noting that a similar relationship exists in

### 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$.
+latent dimension $`r DIM_INDEX`$ (e.g., a negatively-keyed item).
 Finally, as responses are not necessarily interpreted as correct or incorrect,
@@
 
-Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
-$`r LATENT_SPACE`$.
+Following @reckase_difficulty_1985’s procedure,
+we must find in the first place
+the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
 Because the slope is dependent on the direction considered,
@@
 
-On the other hand, we can account for the covariance structure of
-the latent space letting $`r LATENT_SPACE`$ be its geometrical representation.
-Thus, we must define $`r LS_BASIS`$ in such a way that
-
-\begin{equation}
-  `r TRAIT_MV_STD_EQ`,
-\end{equation}
-
-where $`r COV_MATRIX_STD`$ is a diagonal matrix.
-Now we may consider two different possibilities:
-We may define $`r TRANSFORM_MATRIX`$ such that $`r INNER_PROD_CORR_COND`$.
-This results in a *correlation-based* version of the indices
-(note that, in this case, $`r DIAG_MATRIX_INNER_PROD_INV`$ is
-an identity matrix):
+### Covariance-based version
+
+On the other hand, we can let the geometrical representation of
+the latent space $`r LATENT_SPACE`$ account for its covariance structure,
+thus giving $`r LS_BASIS`$ a statistical meaning
+in the context of our MIRT modeling approach.
+To do that, we may assume that $`r TRAIT_COORDS`$ is
+a random vector distributed with covariance $`r COV_MATRIX_EQ`$,
+being $`r COV_MATRIX`$ an $`r N_DIMS`$-dimensional positive definite matrix,
+$`r CORR_MATRIX`$ its corresponding correlation matrix
+(also positive definite) and $`r SD_MATRIX`$ a scaling diagonal matrix
+with $`r VAR_COV_EQ`$ (i.e., the variances).
+
+To make the representation of $`r LATENT_SPACE`$ meaningful,
+the axes in $`r LS_BASIS`$ must have
+an interpretation in terms of the latent space structure.
+Although we do not know, in principle, what
+the interpretation of a latent space basis should be, we may assume that
+it must be somewhat related to the covariance structure.
+Therefore, an orthonormal basis should represent independent,
+standard coordinates;
+formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
+$`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$.
+Applying Equation \@ref(eq:change-basis), $`r COV_MATRIX_TRANSF_EQ`$, and thus,
+
+\begin{equation}
+  \begin{split}
+    `r COV_MATRIX_TRANSF_ID_EQ` \\
+    `r COV_MATRIX_RESULT_EQ`.
+  \end{split}
+\end{equation}
+
+Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
+$`r INNER_PROD_COV_COND`$;
+in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
+*covariance-based* version of the parameters:
+
+\begin{equation}
+  \begin{split}
+    `r MDISC_COV_PARAM_EQ` \\
+    `r MIL_COV_PARAM_EQ`.
+  \end{split}
+\end{equation}
+
+According to @reckase_discriminating_1991, the multidimensional parameters must
+meet three conditions for being regarded as a valid generalization
+of the (unidimensional) IRT parameters:
+
+1. If an item measures only dimension $`r DIM_INDEX`$,
+   then $`r MDISC_UNIDIM_EQ`$.
+
+2. The distance $`r DISTANCE_PARAM`$ has the same relationship with the
+   intercept as $b_i$ in the unidimensional case,
+   i.e., $`r DIST_INTERCEPT_REL`$.
+
+3. $`r MDISC_SYM`$ is four times the maximum slope $`r SLOPE_MAX_PARAM`$.
+
+The second property derives straightforwardly from
+Equation \@ref(eq:distance-param),
+and the third one is implicit in the definition of the $`r MDISC_SYM`$.
+However, when an item measures only dimension $`r DIM_INDEX`$,
+simple arithmetic can show that $`r MDISC_UNIDIM_GENERALIZED_EQ`$.
+In the covariance-based case, this means that $`r MDISC_COV_BASED_UNIDIM_EQ`$
+and, as $`r COV_SYM_DIAG`$ is generally different from 1,
+$`r MDISC_COV_PARAM`$ does not fulfill the first property
+(we shall refer to this property as *scale invariance*, from now on).
+Nevertheless, it can be interpreted as a *scaled version* of the parameter,
+with the standard deviation of the corresponding dimension as scaling factor.
+This transformation may be made in the unidimensional case as well,
+thus finding a *standard-metric* discrimination parameter
+(i.e., referred to a unitary variance space).
+
+### Correlation-based version
+
+Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
+we have argued that it is still a valid generalization of the IRT parameters.
+However, it may be convenient to find yet another generalization,
+one that accounts for the latent space structure
+while still fulfilling this property.
+To do this, let's assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
+$`r TRANSFORM_MATRIX_AUX_RESULT`$.
+Then, we have that
+
+$$
+  `r CORR_BASED_DERIVATION`.
+$$
+
+That is, defining $`r TRANSFORM_MATRIX`$ as
+$`r TRANSFORM_MATRIX_CORR_BASED_DEF`$ can be interpreted as re-scaling
+the transformed parameter back to its original metric.
+In this case, we get that $`r INNER_PROD_CORR_COND`$.
+Therefore, we can also define a *correlation-based* version of the
+multidimensional parameters as
 
@@
 
-However, we may also define $`r TRANSFORM_MATRIX`$ such that
-$`r INNER_PROD_COV_COND`$, so
-
-\begin{equation}
-  `r TRAIT_MV_STD_NORM_EQ`.
-\end{equation}
-
-In this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in the
-following *covariance-based* version of the indices:
-
-\begin{equation}
-  \begin{split}
-    `r MDISC_COV_PARAM_EQ` \\
-    `r MIL_COV_PARAM_EQ`.
-  \end{split}
-\end{equation}
-
-## Properties of the parameters
-
-There are three conditions to meet for regarding these parameters as
-a valid generalization of the (unidimensional) IRT parameters
-[@reckase_discriminating_1991]:
-
-1. If an item measures only dimension $`r DIM_INDEX`$,
-   then $`r MDISC_UNIDIM_EQ`$.
-
-2. The distance $`r DISTANCE_PARAM`$ has the same relationship with the
-   intercept as $b_i$ in the unidimensional case,
-   i.e., $`r DIST_INTERCEPT_REL`$.
-
-3. $`r MDISC_SYM`$ is four times the maximum slope $`r SLOPE_MAX_PARAM`$.
-
-The second property derives straightforwardly from
-Equation \@ref(eq:distance-param),
-and the third one is implicit in the definition of the $`r MDISC_SYM`$.
-Simple arithmetic can show that the first holds true whenever the
-$`r DIAG_KTH_ELEMENT`$-th element of $`r INNER_PROD_MATRIX_INV`$ equals 1.
-Therefore, both the agnostic and the correlation-based version fulfill it.
-$`r MDISC_COV_PARAM`$ does not though,
-as $`r COV_ELEMENT`$ is not generally equal to 1.
-Nevertheless, it can be interpreted as a *scaled version* of the index,
-with the standard deviation of the corresponding dimension as scaling factor.
-This transformation may be made in the unidimensional case as well,
-thus finding a *standard-metric* discrimination parameter
-(i.e., referred to a unitary variance space).
+As we can see, this version also fulfills the scale invariance property,
+as the $`r INNER_PROD_MAT_INV_DIAG_ELEMENT`$ elements are
+the diagonal elements of $`r CORR_MATRIX`$, which are always equal to 1.
+It is important to note though that this transformation does not have
+an also desirable property,
+which we have used to justify the covariance-based version:
+The $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
+$`r TRAIT_COORDS`$ to the orthonormal basis $`r LS_ORTH_BASIS`$
+are not generally uncorrelated, as one would expect
+(save some exceptions, e.g., when all the variances are equal).
+Nevertheless, it is important to note how the $`r MDISC_CORR_PARAM`$ accounts
+for the correlation structure of the latent space
+while still being scale invariant.
+This will be very useful for the purpose of representing the items graphically,
+as we will see later.
+
+## Vector representation
+
+According to @ackerman_creating_1994, the most appropriate way of representing
+multidimensional items is in vector form.
+This allows representing several items altogether and analyzing them
+in terms of their multidimensional parameters [@ackerman_graphical_1996].
+An item vector will be applied at a direction and (signed) distance
+from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter,
+and will have a length equal to its $`r MDISC_SYM`$ parameter.
+Its direction will also be given by the direction component of $`r MIL_PARAM`$,
+so its orientation will always pass through the origin.
+On the basis of these conditions, we need to compute
+the origin and end coordinates of the vector,
+denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
+
+In an orthogonal basis it is easy to compute the coordinates,
+as they are simply the orthogonal projections onto the corresponding axis.
+In the oblique case it is not so simple though, but we can take advantage of the
+orthornormalization introduced above.
+In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
+$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
+and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
+according to Equation \@ref(eq:discr-vector-std),
+i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
+Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
+the direction cosines in $`r TEST_SPACE_STD_BASIS`$
+are $`r DIR_COS_ITEM_VEC_STD`$.
+Therefore,
+
+\begin{align}
+  `r ORIGIN_ORTH_EQ` \\
+  `r END_ORTH_EQ`.
+\end{align}
+
+Applying Equation \@ref(eq:dir-cos-vec-std),
+for the origin coordinates we have that
+
+\begin{equation}
+  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
+\end{equation}
+
+which, operating, results in
+
+\begin{align}
+  `r ORIGIN_ITEM_COORDS_EQ`    \\
+  `r ORIGIN_TRANSF_SQUARED_EQ` \\
+  `r ORIGIN_INNERPROD_EQ`      \\
+  `r ORIGIN_INNERPROD_DIAG_EQ`
+\end{align}
+
+and, finally,
+
+\begin{equation}
+  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
+\end{equation}
+
+The end coordinates can be computed by a similar procedure, which results in
+
+\begin{equation}
+  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
+\end{equation}
+
+Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
+express the items coordinates in terms of their multidimensional parameters.
+However, it is worth noting that the latent space Gram matrix
+also plays a role in computing these coordinates,
+and will therefore have an effect on the representation.
+Of course,
+from Equations \@ref(eq:origin-orth-discr) and \@ref(eq:end-coords-result)
+it is easy to express the coordinates in terms of the parameters in the original
+model formulation, namely,
+
+\begin{align}
+  `r ORIGIN_ITEM_PARAMS_EQ`                  &(\#eq:origin-from-model-params) \\
+  `r END_ITEM_PARAMS_EQ`.                    &(\#eq:end-from-model-params)
+\end{align}
+
+However, formulating them in terms of the multidimensional parameters allows us
+studying the items in terms of their multidimensional properties.
+In the following, we see how to geometrically interpret
+these multidimensional parameters.
+
+### Geometric properties of the items
+
+We have seen how the two new versions of the parameters fulfill
+the three conditions proposed by @reckase_discriminating_1991
+for valid generalizations of the IRT parameters to the multidimensional case,
+being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
+From a geometrical perspective, the item $`r MDISC_SYM`$ is simply interpreted
+as the length of its corresponding vector.
+Regarding the $`r MIL_PARAM`$ parameter,
+we will consider now two more properties, related to each of its two components:
+the distance to the origin, and the measurement direction.
 
@@
 its measurement direction forwards or backwards with respect to the origin.
+As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
+that displacement will be inversely proportional
+to the discrimination of the item.
+That is, the less discriminating the item is,
+the further away it will be from the origin.
 
 The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
-also due to the strict positivity of $`r MDISC_SYM`$,
+also due to the strict positiveness of $`r MDISC_SYM`$,
 are equal to the signs of $`r DISCR_VECTOR`$.
-This property leads directly to generalizing the indices to the monotonically
-non-increasing case: The measurement direction relative to dimension
-$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
-monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
-When $`r IRF_ABBR`$ is constant with respect to
-variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
+This property leads directly to generalizing
+the parameters to the monotonically non-increasing case:
+The measurement direction relative to dimension $`r DIM_INDEX`$
+will be negative when $`r IRF_ABBR`$ is monotonically decreasing
+with respect to variations along $`r TRAIT_COMPONENT`$.
+When $`r IRF_ABBR`$ is constant
+with respect to variations along $`r TRAIT_COMPONENT`$,
+$`r SIGN_COS_VEC_ITEM_EQ`$,
 which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
-However, note that this does not necessarily imply that the vector is
-parallel to any other axis or strictly contained in the hyperplane formed
-by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
-orthogonality assumption, but it will depend on the correlation matrix
-$`r CORR_MATRIX`$ for the other two versions of the index.
+However, note that this does not necessarily imply that
+the vector is parallel to any other axis
+(or strictly contained in the hyperplane formed by other axes, for that matter);
+this will always be true for the $`r MIL_AG_PARAM`$ due to
+the orthogonality assumption,
+but it will depend on the correlation matrix $`r CORR_MATRIX`$
+for the other two versions of the parameter.
 
@@
 a visual representation of the items.
-A vector plot [@ackerman_creating_1994] is the most appropriate way of
-representing the items graphically;
-it allows one to plot several items altogether and analyze them
-in terms of their multidimensional parameters [@ackerman_graphical_1996]:
-An item vector will be applied at a direction and (signed) distance
-from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
-and will have a length equal to its $`r MDISC_SYM`$ parameter.
+To do that, we need to choose a version of the parameters
+that yields their most faithful representation.
+We have already seen that
+the agnostic version disregards the latent space structure.
+Therefore, it will only be useful in the unlikely case that
+we have no information of such structure.
+When we have an estimate of the latent space covariances,
+we may use either the covariance-based or the correlation-based version.
+However, we have already seen how
+the distance to the origin depends on the $`r MDISC_SYM`$.
+Also, in the covariance-based,
+the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
+which, as we have argued before, implies a change of scale.
+In the case of the correlation-based version however,
+$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
+being the coordinates scale-invariant.
+Therefore, for graphical representation purposes,
+the correlation-based version will be preferred.
+
+Plotting on a display usually requires providing
+coordinates in a rectangular Cartesian system.
+Therefore, we will need to compute the equivalent rectangular coordinates
+of our general, possibly non-rectangular coordinates.
+This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
+by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
+the equivalent, orthonormalized coordinates.
+Therefore, we will need to define
+a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
+that allows us to represent the test space structure and the items
+in a rectangular Cartesian system.
+
+The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
+the following steps:
+
+1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
+   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
+   respectively.
+
+1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
+   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
+   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.
+
+1. Compute the end coordinates $`r END_ITEM_EQ`$.
+
+1. Define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
+   $`r TRANSF_MATRIX_INV_SQ_CORR_EQ`$.
+
+1. Compute the rectangular coordinates
+   $`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$ by pre-multiplying
+   $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively, by
+   the transformation matrix $`r TRANSFORM_MATRIX`$
+   ($`r ORIGIN_TRANSF_EQ`$; $`r END_TRANSF_EQ`$).
+
+Instead of steps 1 to 3,
+one can alternatively compute $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$
+directly from Equations \@ref(eq:origin-from-model-params)
+and \@ref(eq:end-from-model-params).
+Note though that these equations are not independent of $`r CORR_MATRIX`$,
+as it is used in the computation of $`r DISCR_VECTOR_MODULE`$,
+and thus this matrix is necessary in any case to compute the coordinates.
+
+### Graphical representation example
+
+```{r read-chunks-graphical-example}
+read_chunk("src/Graphical_example_paper.R")
+```
+
+```{r libraries}
+#| warning: false
+```
+
+```{r sources}
+```
+
+```{r constants}
+```
+
+```{r graphical-output-conf}
+```
+
+```{r compute-example-items}
+```
 
@@
 we will limit ourselves to the bidimensional case here.
-Plotting on a two-dimensional display usually requires providing
-coordinates in a rectangular Cartesian system.
-Therefore, we will need to compute the equivalent rectangular coordinates
-of our general, possibly non-rectangular coordinates.
-However, this implies pre-multiplying by $`r TRANSFORM_MATRIX`$,
-instead of $`r TRANSFORM_MATRIX_TRANSP_INV`$ as would be expected from
-Equation \@ref(eq:discr-vector-std).
-This is because computing the equivalent rectangular coordinates
-of the oblique ones is different from computing the item parameters
-transformed to an orthogonal space.
-While the latter implies rotating the axes to an orthogonal configuration and
-computing the new coordinates in these axes,
-the former implies transforming the space altogether along with the
-axes transformation.
-
-Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
-that allows us to represent the test space structure
-and the items by transforming their coordinates to
-a rectangular Cartesian system.
-However, to facilitate the interpretation,
-we want to avoid changing the scaling of the parameters.
-We will thus prefer the correlation-based over the covariance-based indices.
-Thus, we define $`r TRANSFORM_MATRIX`$ such that
-$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.
-
-The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
-the following steps:
-
-1. Define $`r TRANSFORM_MATRIX`$ such that $`r TRANSF_MATRIX_SQ_CORR_INV_EQ`$.
-
-1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
-   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
-   respectively.
-
-1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
-   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
-   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.
-
-1. Compute the end coordinates $`r END_ITEM_EQ`$.
-
-1. Compute the rectangular coordinates
-   $`r ORIGIN_ITEM_TRANSF`$ and $`r END_ITEM_TRANSF`$ by pre-multiplying
-   $`r ORIGIN_ITEM`$ and $`r END_ITEM`$, respectively, by
-   the transformation matrix $`r TRANSFORM_MATRIX`$
-   ($`r ORIGIN_ITEM_TRANSF_EQ`$; $`r END_ITEM_TRANSF_EQ`$).
-
-### Graphical representation example
-
-```{r read-chunks-graphical-example}
-read_chunk("src/Graphical_example_paper.R")
-```
-
-```{r libraries}
-```
-
-```{r sources}
-```
-
-```{r constants}
-```
-
-```{r graphical-output-conf}
-```
-
-```{r compute-example-items}
-```
 
@@
 and another one that only considers the correlations among latent dimensions.
-The covariance-based index has the drawback of not fulfilling
+The covariance-based parameter has the drawback of not fulfilling
 the first of @reckase_discriminating_1991's properties, namely,
-the scale invariance property.
+that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
 However, we have provided a rationale for the violation of this property,
@@
 comparable among different latent spaces.
-More importantly, the covariance-based version
-of the multidimensional parameters is invariant to
-changes of latent-space basis.
 Once again, this is important in applications where
-potentially different bases are involved.
-Given these considerations,
-we propose to adopt the covariance-based version of
-the multidimensional parameters as the most general definition,
-and refer to them simply as
+potentially different latent spaces are involved.
+Given these considerations, we propose to adopt the covariance-based version of
+the multidimensional parameters and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of being scale-invariant,
+has the advantage of preserving the metric of the parameters,
 which is convenient for fidelity when representing the items.
-However, it is worth noting that it has been defined ad-hoc
-to fulfill this specific property.
-Therefore it lacks other desirable properties of the covariance-based version,
-like being invariant to a latent space scaling transformation,
-or yielding a diagonal latent covariance matrix in the corresponding
-orthogonal latent space basis.
-Therefore, we recommend its use for graphical representation purposes only.
-As for the agnostic version, we recommend it only in
-the (unlikely) case of having no information about the latent space structure.
-
-One may consider that the formal derivations presented here are
-overly complicated,
+Therefore, we recommend using the correlation-based version of the parameters
+for graphical representation purposes only.
+
+One may consider that the formal derivations presented here are overly
+complicated,
 when one might simply transform the space to rectangular Cartesian coordinates
@@
 its inner product matrix.
-This result is both surprising and counterintuitive;
-The MIRT literature suggests in several instances that
-"the angle between the axes [of latent ability space] represents
-the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
-We hope our results help shed a clearer light on
-this often missunderstood topic.
-
 It is worth noting that a similar relationship exists in

### 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
Reverts incorrect "method" section changes
Reverts changes inherited from merging "review/intro"

--- 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 Finally, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
 instead of the item difficulty.

### 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
Reverts missing changes from empirical examples and discussion

@@
 #| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: Item parameters for the empirical example
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
 item_params_output
 ```
+
+The latter effect can be better noticed in another case, in which Reckase
+uses the same item parameters to demonstrate the over-specification of
+dimensions (see p. 183).
+In this example, he uses a covariance matrix that practically has rank 1,
+implying that the (alleged) three dimensions collapse into a single one, being
+thus the test unidimensional. In such a case, the covariance-based version of
+the direction angles are extremely close to 0 for all items, as one would expect
+from a unidimensional test, where all items are strictly aligned with the one
+dimension they measure. The reader interested can check out the parameters
+computed in such case in the Supplementary Table 1.
+<!-- #TODO: Include Supplementary Table -->
+
+### @tezza_modelo_2018
 

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects use of test space transformation matrix
The correct transformation matrix in the test space should be "P inverse transposed" instead of "P". The use of the latter is inherited from previous versions, where the orthogonal projections on the axes (instead of the coordinates) of the oblique space were incorrectly computed.

--- 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 In the first case, the two latent dimensions are independent,
-with $`r TRANSFORM_MATRIX`$ being an identity matrix.
+with $`r TRANSFORM_MATRIX_TRANSP_INV`$ being an identity matrix.
 In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
 In order to keep the horizontal axis invariant,
-we define $`r TRANSFORM_MATRIX`$ such that it only rotates the vertical axis
-[@harman_modern_1970]:
+we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
+it only rotates the vertical axis [@harman_modern_1970]:
 

### 2024-07-17 R/Formulae.R

@@ -1253,6 +1253,6 @@ TRANSFORM_MATRIX_EXAMPLE    <- latex(
   sep = "\n"
 )
 TRANSFORM_MATRIX_EXAMPLE_EQ <- latex_def(
-  TRANSFORM_MATRIX,
+  TRANSFORM_MATRIX_TRANSP_INV,
   TRANSFORM_MATRIX_EXAMPLE
 )

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds introduction to "Literature examples" section

--- 2024-07-16 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-Applying Proposition 1 to
-the model formulation in Equation \@ref(eq:M2PL-formulation),
-the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+From Proposition 1,
+any trait vector $`r TRAIT_VECTOR`$ is expressed in polar coordinates as
+
+\begin{equation}
+  `r TRAIT_VECTOR_POLAR_EQ`.                      (\#eq:trait-equivalence-polar)
+\end{equation}
+
+Thus, the M2PL model is expressed in polar coordinates as
 
@@
 
-### Correlation-based version
-
-Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
-we have argued that it is still a valid generalization of the IRT parameters.
-However, it may be convenient to find yet another generalization,
-one that accounts for the latent space structure
-while still fulfilling this property.
-To do this, let's assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
-$`r TRANSFORM_MATRIX_AUX_RESULT`$.
-Then, we have that
-
-$$
-  `r CORR_BASED_DERIVATION`.
-$$
-
-That is, defining $`r TRANSFORM_MATRIX`$ as
-$`r TRANSFORM_MATRIX_CORR_BASED_DEF`$ can be interpreted as re-scaling
-the transformed parameter back to its original metric.
-In this case, we get that $`r INNER_PROD_CORR_COND`$.
-Therefore, we can also define a *correlation-based* version of the
-multidimensional parameters as
-
-\begin{equation}
-  \begin{split}
-    `r MDISC_CORR_PARAM_EQ` \\
-    `r MIL_CORR_PARAM_EQ`.
-  \end{split}
-\end{equation}
-
-As we can see, this version also fulfills the scale invariance property,
-as the $`r INNER_PROD_MAT_INV_DIAG_ELEMENT`$ elements are
-the diagonal elements of $`r CORR_MATRIX`$, which are always equal to 1.
-It is important to note though that this transformation does not have
-an also desirable property,
-which we have used to justify the covariance-based version:
-The $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
-$`r TRAIT_COORDS`$ to the orthonormal basis $`r LS_ORTH_BASIS`$
-are not generally uncorrelated, as one would expect
-(save some exceptions, e.g., when all the variances are equal).
-Nevertheless, it is important to note how the $`r MDISC_CORR_PARAM`$ accounts
-for the correlation structure of the latent space
-while still being scale invariant.
-This will be very useful for the purpose of representing the items graphically,
-as we will see later.
-
-## Vector representation
-
-According to @ackerman_creating_1994, the most appropriate way of representing
-multidimensional items is in vector form.
-This allows representing several items altogether and analyzing them
-in terms of their multidimensional parameters [@ackerman_graphical_1996].
-An item vector will be applied at a direction and (signed) distance
-from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter,
-and will have a length equal to its $`r MDISC_SYM`$ parameter.
-Its direction will also be given by the direction component of $`r MIL_PARAM`$,
-so its orientation will always pass through the origin.
-On the basis of these conditions, we need to compute
-the origin and end coordinates of the vector,
-denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
-
-In an orthogonal basis it is easy to compute the coordinates,
-as they are simply the orthogonal projections onto the corresponding axis.
-In the oblique case it is not so simple though, but we can take advantage of the
-orthornormalization introduced above.
-In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
-$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
-and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
-according to Equation \@ref(eq:discr-vector-std),
-i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
-Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
-the direction cosines in $`r TEST_SPACE_STD_BASIS`$
-are $`r DIR_COS_ITEM_VEC_STD`$.
-Therefore,
-
-\begin{align}
-  `r ORIGIN_ORTH_EQ` \\
-  `r END_ORTH_EQ`.
-\end{align}
-
-Applying Equation \@ref(eq:dir-cos-vec-std),
-for the origin coordinates we have that
-
-\begin{equation}
-  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
-\end{equation}
-
-which, operating, results in
-
-\begin{align}
-  `r ORIGIN_ITEM_COORDS_EQ`    \\
-  `r ORIGIN_TRANSF_SQUARED_EQ` \\
-  `r ORIGIN_INNERPROD_EQ`      \\
-  `r ORIGIN_INNERPROD_DIAG_EQ`
-\end{align}
-
-and, finally,
-
-\begin{equation}
-  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
-\end{equation}
-
-The end coordinates can be computed by a similar procedure, which results in
-
-\begin{equation}
-  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
-\end{equation}
-
-Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
-express the items coordinates in terms of their multidimensional parameters.
-However, it is worth noting that the latent space Gram matrix
-also plays a role in computing these coordinates,
-and will therefore have an effect on the representation.
-Of course,
-from Equations \@ref(eq:origin-orth-discr) and \@ref(eq:end-coords-result)
-it is easy to express the coordinates in terms of the parameters in the original
-model formulation, namely,
-
-\begin{align}
-  `r ORIGIN_ITEM_PARAMS_EQ`                  &(\#eq:origin-from-model-params) \\
-  `r END_ITEM_PARAMS_EQ`.                    &(\#eq:end-from-model-params)
-\end{align}
-
-However, formulating them in terms of the multidimensional parameters allows us
-studying the items in terms of their multidimensional properties.
-In the following, we see how to geometrically interpret
-these multidimensional parameters.
-
-### Geometric properties of the items
-
-We have seen how the two new versions of the parameters fulfill
-the three conditions proposed by @reckase_discriminating_1991
-for valid generalizations of the IRT parameters to the multidimensional case,
-being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
-From a geometrical perspective, the item $`r MDISC_SYM`$ is simply interpreted
-as the length of its corresponding vector.
-Regarding the $`r MIL_PARAM`$ parameter,
-we will consider now two more properties, related to each of its two components:
-the distance to the origin, and the measurement direction.
-
 The (signed) distance to the origin,
-given by the first component of $`r MIL_PARAM`$, has an interesting property:
+given by the second element of $`r MIL_PARAM`$, has an interesting property:
 By the definition of the inner product,
@@
 its measurement direction forwards or backwards with respect to the origin.
-As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
-that displacement will be inversely proportional
-to the discrimination of the item.
-That is, the less discriminating the item is,
-the further away it will be from the origin.
 
 The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
-also due to the strict positiveness of $`r MDISC_SYM`$,
+also due to the strict positivity of $`r MDISC_SYM`$,
 are equal to the signs of $`r DISCR_VECTOR`$.
-This property leads directly to generalizing
-the parameters to the monotonically non-increasing case:
-The measurement direction relative to dimension $`r DIM_INDEX`$
-will be negative when $`r IRF_ABBR`$ is monotonically decreasing
-with respect to variations along $`r TRAIT_COMPONENT`$.
-When $`r IRF_ABBR`$ is constant
-with respect to variations along $`r TRAIT_COMPONENT`$,
-$`r SIGN_COS_VEC_ITEM_EQ`$,
+This property leads directly to generalizing the indices to the monotonically
+non-increasing case: The measurement direction relative to dimension
+$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
+monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
+When $`r IRF_ABBR`$ is constant with respect to
+variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
 which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
-However, note that this does not necessarily imply that
-the vector is parallel to any other axis
-(or strictly contained in the hyperplane formed by other axes, for that matter);
-this will always be true for the $`r MIL_AG_PARAM`$ due to
-the orthogonality assumption,
-but it will depend on the correlation matrix $`r CORR_MATRIX`$
-for the other two versions of the parameter.
+However, note that this does not necessarily imply that the vector is
+parallel to any other axis or strictly contained in the hyperplane formed
+by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
+orthogonality assumption, but it will depend on the correlation matrix
+$`r CORR_MATRIX`$ for the other two versions of the index.
 
@@
 a visual representation of the items.
-To do that, we need to choose a version of the parameters
-that yields their most faithful representation.
-We have already seen that
-the agnostic version disregards the latent space structure.
-Therefore, it will only be useful in the unlikely case that
-we have no information of such structure.
-When we have an estimate of the latent space covariances,
-we may use either the covariance-based or the correlation-based version.
-However, we have already seen how
-the distance to the origin depends on the $`r MDISC_SYM`$.
-Also, in the covariance-based,
-the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
-which, as we have argued before, implies a change of scale.
-In the case of the correlation-based version however,
-$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
-being the coordinates scale-invariant.
-Therefore, for graphical representation purposes,
-the correlation-based version will be preferred.
-
-Plotting on a display usually requires providing
-coordinates in a rectangular Cartesian system.
-Therefore, we will need to compute the equivalent rectangular coordinates
-of our general, possibly non-rectangular coordinates.
-This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
-by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
-the equivalent, orthonormalized coordinates.
-Therefore, we will need to define
-a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
-that allows us to represent the test space structure and the items
-in a rectangular Cartesian system.
-
-The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
-the following steps:
-
-1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
-   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
-   respectively.
-
-1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
-   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
-   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.
-
-1. Compute the end coordinates $`r END_ITEM_EQ`$.
-
-1. Define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
-   $`r TRANSF_MATRIX_INV_SQ_CORR_EQ`$.
-
-1. Compute the rectangular coordinates
-   $`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$ by pre-multiplying
-   $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively, by
-   the transformation matrix $`r TRANSFORM_MATRIX`$
-   ($`r ORIGIN_TRANSF_EQ`$; $`r END_TRANSF_EQ`$).
-
-Instead of steps 1 to 3,
-one can alternatively compute $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$
-directly from Equations \@ref(eq:origin-from-model-params)
-and \@ref(eq:end-from-model-params).
-Note though that these equations are not independent of $`r CORR_MATRIX`$,
-as it is used in the computation of $`r DISCR_VECTOR_MODULE`$,
-and thus this matrix is necessary in any case to compute the coordinates.
-
-### Graphical representation example
-
-```{r read-chunks-graphical-example}
-read_chunk("src/Graphical_example_paper.R")
-```
-
-```{r libraries}
-#| warning: false
-```
-
-```{r sources}
-```
-
-```{r constants}
-```
-
-```{r graphical-output-conf}
-```
-
-```{r compute-example-items}
-```
+A vector plot [@ackerman_creating_1994] is the most appropriate way of
+representing the items graphically;
+it allows one to plot several items altogether and analyze them
+in terms of their multidimensional parameters [@ackerman_graphical_1996]:
+An item vector will be applied at a direction and (signed) distance
+from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
+and will have a length equal to its $`r MDISC_SYM`$ parameter.
 
@@
 we will limit ourselves to the bidimensional case here.
+Plotting on a two-dimensional display usually requires providing
+coordinates in a rectangular Cartesian system.
+Therefore, we will need to compute the equivalent rectangular coordinates
+of our general, possibly non-rectangular coordinates.
+However, this implies pre-multiplying by $`r TRANSFORM_MATRIX`$,
+instead of $`r TRANSFORM_MATRIX_TRANSP_INV`$ as would be expected from
+Equation \@ref(eq:discr-vector-std).
+This is because computing the equivalent rectangular coordinates
+of the oblique ones is different from computing the item parameters
+transformed to an orthogonal space.
+While the latter implies rotating the axes to an orthogonal configuration and
+computing the new coordinates in these axes,
+the former implies transforming the space altogether along with the
+axes transformation.
+
+Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
+that allows us to represent the test space structure
+and the items by transforming their coordinates to
+a rectangular Cartesian system.
+However, to facilitate the interpretation,
+we want to avoid changing the scaling of the parameters.
+We will thus prefer the correlation-based over the covariance-based indices.
+Thus, we define $`r TRANSFORM_MATRIX`$ such that
+$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.
+
+The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
+the following steps:
+
+1. Define $`r TRANSFORM_MATRIX`$ such that $`r TRANSF_MATRIX_SQ_CORR_INV_EQ`$.
+
+1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
+   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
+   respectively.
+
+1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
+   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
+   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.
+
+1. Compute the end coordinates $`r END_ITEM_EQ`$.
+
+1. Compute the rectangular coordinates
+   $`r ORIGIN_ITEM_TRANSF`$ and $`r END_ITEM_TRANSF`$ by pre-multiplying
+   $`r ORIGIN_ITEM`$ and $`r END_ITEM`$, respectively, by
+   the transformation matrix $`r TRANSFORM_MATRIX`$
+   ($`r ORIGIN_ITEM_TRANSF_EQ`$; $`r END_ITEM_TRANSF_EQ`$).
+
+### Graphical representation example
+
+```{r read-chunks-graphical-example}
+read_chunk("src/Graphical_example_paper.R")
+```
+
+```{r libraries}
+```
+
+```{r sources}
+```
+
+```{r constants}
+```
+
+```{r graphical-output-conf}
+```
+
+```{r compute-example-items}
+```
 
@@
 
-## Examples with empirical data
+## Application to examples from the literature
+
+In the following, we apply our results to two instances taken from
+the MIRT literature.
+The first one draws from a classical example by @reckase_multidimensional_2009,
+who presents a three-dimensional instrument made up by 30 items.
+It is supposed to be a cognitive test
+(it is unclear whether the item parameters are estimated from
+actual empirical data or they are a fictional example,
+created ad-hoc for simulation purposes),
+so all the discrimination parameters are positive.
+The second example, taken from @tezza_modelo_2018, is an application of
+the `r MODEL_ACRONYM` to assess the quality of e-commerce websites.
+Besides being an empirical example, applied to actual test data,
+it pertains to the non-cognitive domain and contains
+items with  negative discrimination parameters.
+This is very convenient for exemplifying our results with
+items with opposite-sign discrimination parameters.
+
+The two examples are represented in Tables
+\@ref(tab:empirical-table-Reckase2009-out)
+and \@ref(tab:empirical-table-Tezza2018-out), respectively.
+Each of these tables shows the original `r MODEL_ACRONYM` parameters
+[the estimation errors in @tezza_modelo_2018 are omitted for clarity],
+along with the agnostic version of the multidimensional parameters;
+it can be seen that the $`r MDISC_AG_PARAM`$ and the $`r DISTANCE_AG_PARAM`$
+are coincident with their respective original sources
+[see Table 6.1 in @reckase_multidimensional_2009, p. 153, and
+Table 5 in @tezza_modelo_2018, p. 926].
+In the first case, the direction angles are also coincident with the values
+provided in the original [@tezza_modelo_2018 do not provide these results].
+In order to facilitate the comparison,
+the covariance-based version of each parameter is provided in
+the column next to its corresponding agnostic version.
 
@@
 and another one that only considers the correlations among latent dimensions.
-The covariance-based parameter has the drawback of not fulfilling
+The covariance-based index has the drawback of not fulfilling
 the first of @reckase_discriminating_1991's properties, namely,
-the scale invariance property.
+that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
 However, we have provided a rationale for the violation of this property,

### 2024-07-17 R/Formulae.R

@@ -1067,3 +1067,7 @@ TRANSFORM_MATRIX_EXAMPLE_EQ <- latex_def(
   TRANSFORM_MATRIX,
   TRANSFORM_MATRIX_EXAMPLE
 )
+
+### Application to examples from the literature: ----
+
+DISTANCE_AG_PARAM <- DISTANCE_SYM |> latex_sub(AGNOSTIC_SUBINDEX)

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Improves redaction of literature examples

@@
 
-Table \@ref(tab:empirical-table-Reckase2009-out) draws from a classical example
-by @reckase_multidimensional_2009.
-In this 30-test item, there are three ten-item blocks assumed to approximate
+Table \@ref(tab:empirical-table-Reckase2009-out) shows the parameters of the
+30-item test in the @reckase_multidimensional_2009 example:
+There are three ten-item blocks assumed to approximate
 a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
-and items 21-30 dimension 2 (see Table 6.1, p. 153 therein).
+and items 21-30 dimension 2
+[see Table 6.1, p. 153, in @reckase_multidimensional_2009].
 In his example, Reckase tests the estimation of these items
@@
 is standard with null correlations; however, the second one has
-(apart from a non-null vector of means) as covariance matrix
 
@@
 
-which corresponds to a correlation matrix of
+as covariance matrix, which corresponds to a correlation matrix of
 
@@
 
-As we can see in Table \@ref(tab:empirical-table-Reckase2009-out), the agnostic
-versions of the multidimensional parameters
+As we can see in Table \@ref(tab:empirical-table-Reckase2009-out),
+the agnostic versions of the multidimensional parameters
 (columns labelled _`r AGNOSTIC_ABBR`_)
@@
 dimensions (see p. 183).
-In this example, he uses a covariance matrix that practically has rank 1,
+In that example, he uses a covariance matrix that practically has rank 1,
 implying that the (alleged) three dimensions collapse into a single one, being
-thus the test unidimensional. In such a case, the covariance-based version of
-the direction angles are extremely close to 0 for all items, as one would expect
+thus the test practically unidimensional.
+In such a case, the covariance-based version of the direction angles
+are extremely close to 0 for all items, as one would expect
 from a unidimensional test, where all items are strictly aligned with the one
 dimension they measure. The reader interested can check out the parameters
-computed in such case in the Supplementary Table 1.
-<!-- #TODO: Include Supplementary Table -->
+computed in that case in the table in the Supplementary Material.
 
@@
 (compared with the ones in dimension 2 and 3, respectively) are barely affected.
-In these cases, we also notice that the angles are barely affected.
 The most extreme change in these two examples occurs in

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects grammar as suggested by MS Word

@@
 However, Reckase's Equation 3 implicitly assumes orthogonality.
-In order to contemplate the general case,
+To contemplate the general case,
 we need to introduce a few linear algebra results before proceeding further.
@@
 so its orientation will always pass through the origin.
-On the basis of these conditions, we need to compute
+Based on these conditions, we need to compute
 the origin and end coordinates of the vector,
@@
 
-The aforementioned properties are easier to apprehend with
+The geometric properties are easier to apprehend with
 a visual representation of the items.
@@
 In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
-In order to keep the horizontal axis invariant,
+To keep the horizontal axis invariant,
 we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
@@
 
-## Application to examples from the literature
-
-In the following, we apply our results to two instances taken from
-the MIRT literature.
-The first one draws from a classical example by @reckase_multidimensional_2009,
-who presents a three-dimensional instrument made up by 30 items.
-It is supposed to be a cognitive test
-(it is unclear whether the item parameters are estimated from
-actual empirical data or they are a fictional example,
-created ad-hoc for simulation purposes),
-so all the discrimination parameters are positive.
-The second example, taken from @tezza_modelo_2018, is an application of
-the `r MODEL_ACRONYM` to assess the quality of e-commerce websites.
-Besides being an empirical example, applied to actual test data,
-it pertains to the non-cognitive domain and contains
-items with  negative discrimination parameters.
-This is very convenient for exemplifying our results with
-items with opposite-sign discrimination parameters.
-
-The two examples are represented in Tables
-\@ref(tab:empirical-table-Reckase2009-out)
-and \@ref(tab:empirical-table-Tezza2018-out), respectively.
-Each of these tables shows the original `r MODEL_ACRONYM` parameters
-[the estimation errors in @tezza_modelo_2018 are omitted for clarity],
-along with the agnostic version of the multidimensional parameters;
-it can be seen that the $`r MDISC_AG_PARAM`$ and the $`r DISTANCE_AG_PARAM`$
-are coincident with their respective original sources
-[see Table 6.1 in @reckase_multidimensional_2009, p. 153, and
-Table 5 in @tezza_modelo_2018, p. 926].
-In the first case, the direction angles are also coincident with the values
-provided in the original [@tezza_modelo_2018 do not provide these results].
-In order to facilitate the comparison,
-the covariance-based version of each parameter is provided in
-the column next to its corresponding agnostic version.
-
-### @reckase_multidimensional_2009
+## Examples with empirical data
 
@@
 ```
-
-Table \@ref(tab:empirical-table-Reckase2009-out) shows the parameters of the
-30-item test in the @reckase_multidimensional_2009 example:
-There are three ten-item blocks assumed to approximate
-a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
-and items 21-30 dimension 2
-[see Table 6.1, p. 153, in @reckase_multidimensional_2009].
-In his example, Reckase tests the estimation of these items
-on two simulated datasets: In the first one, the latent trait distribution
-is standard with null correlations; however, the second one has
-
-\begin{equation}
-  `r cov_matrix_value_out`,
-\end{equation}
-
-as covariance matrix, which corresponds to a correlation matrix of
-
-\begin{equation}
-  `r cor_matrix_value_out`.
-\end{equation}
-
-As we can see in Table \@ref(tab:empirical-table-Reckase2009-out),
-the agnostic versions of the multidimensional parameters
-(columns labelled _`r AGNOSTIC_ABBR`_)
-coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
-However, when we take into account the latent space structure
-(columns labelled $`r COV_MATRIX`$), we notice several differences.
-First, the $`r MDISC_ITEM`$ parameters tend to be larger
-than their agnostic counterparts.
-This happens because all the discrimination parameters and the latent space
-correlations are positive; in this respect, these items are similar to item 1
-in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
-However, for the items mostly aligned with dimension 2 (items 21-30), as the
-$`r MDISC_ITEM`$ parameters are also scaled by a standard deviation smaller
-than 1, their values decrease, being some of them smaller than the agnostic
-versions. The $`r DISTANCE_PARAM`$ parameters decrease or increase, consequently
-(see Equation \@ref(eq:M2PL-formula-polar)), thus being most of them smaller
-than the agnostic ones (again, with the exceptions only happening
-in the items aligned with dimension 2).
-Finally, we see that the direction angles are not as clearly separated as one
-might expect from the agnostic versions. This effect is more pronounced among
-the items aligned with dimensions 1 and 3, which are strongly correlated.
-In this situation, we notice that the angles tend to be much smaller, as the
-high correlation induces an alignment between the two dimensions
-(compare item 1 in the two test spaces in
-Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
 
@@
 #| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
+#| tab.cap: Item parameters for the empirical example
 item_params_output
 ```
-
-The latter effect can be better noticed in another case, in which Reckase
-uses the same item parameters to demonstrate the over-specification of
-dimensions (see p. 183).
-In that example, he uses a covariance matrix that practically has rank 1,
-implying that the (alleged) three dimensions collapse into a single one, being
-thus the test practically unidimensional.
-In such a case, the covariance-based version of the direction angles
-are extremely close to 0 for all items, as one would expect
-from a unidimensional test, where all items are strictly aligned with the one
-dimension they measure. The reader interested can check out the parameters
-computed in that case in the table in the Supplementary Material.
-
-### @tezza_modelo_2018
 
@@
 ```
-
-@tezza_modelo_2018 provide us with another example that allows showcasing
-the effect of correlated dimensions on negative discrimination parameters.
-In their model, dimensions 1 and 4 are estimated to have a correlation of
-`r DIMS_CORR`.
-With such latent space structure, items that strongly discriminate in any or
-both of these two dimensions change drastically their multidimensional
-parameters from the agnostic to the covariance-based version
-(see Table \@ref(tab:empirical-table-Tezza2018-out)) ---this happens,
-for example, with items 57 and 60. These items also have different-sign
-discrimination parameters in dimensions 1 and 4, so their situation is
-similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
-Figure \@ref(fig:item-plot-out),
-thus the decrease in their $`r MDISC_ITEM`$ parameter.
-The effect on the direction is also patent in the angles these items make with
-these dimensions, which change up to almost 20º, compared with at most 7º in the
-other two dimensions, and around 1º in most cases.
-
-On the other hand, in cases such as items 12 or 59, which have
-relatively low discrimination parameters on those two dimensions
-(compared with the ones in dimension 2 and 3, respectively) are barely affected.
-The most extreme change in these two examples occurs in
-the direction of item 12 with respect to dimension 4. We see how the large
-discrimination in dimension 1, when combined with the correlation
-with dimension 4, induces a decrease in the angular direction
-with that dimension.
-In this case, being the discrimination in dimension 4 so low, the situation is
-very similar to item 4 in Table \@ref(tab:example-items-table-out)
-and Figure \@ref(fig:item-plot-out).
-
-<!---BLOCK_LANDSCAPE_START--->
 
@@
 #| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Tezza et al. (2018).
+#| tab.cap: Item parameters for the empirical example
 item_params_output
@@
 which may give relevant information if these are actual, substantive traits.
-Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require to compute the cosines in
-the original basis afterwards.
-
-On the other hand, establishing the relationship between the original and
-the orthonormalized basis has provided us with an important insight:
+Moreover, strictly following @reckase_difficulty_1985's method in the original,
+oblique space has also provided us with an important insight:
 The distinction between the latent space and the test space.
-Our derivations show that, in order to properly represent the items,
+Our derivations show that, to properly represent the items,
 we need to define a vector space that has

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects grammar as suggested by MS Word

--- 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 the MIRT literature.
-The first one draws from a classical example by @reckase_multidimensional_2009,
+The first one draws from a classic example by @reckase_multidimensional_2009,
 who presents a three-dimensional instrument made up by 30 items.
@@
 (it is unclear whether the item parameters are estimated from
-actual empirical data or they are a fictional example,
+actual empirical data, or they are a fictional example,
 created ad-hoc for simulation purposes),
@@
 along with the agnostic version of the multidimensional parameters;
-it can be seen that the $`r MDISC_AG_PARAM`$ and the $`r DISTANCE_AG_PARAM`$
+the $`r MDISC_AG_PARAM`$ and the $`r DISTANCE_AG_PARAM`$
 are coincident with their respective original sources
@@
 coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
-However, when we take into account the latent space structure
+However, when we consider the latent space structure
 (columns labelled $`r COV_MATRIX`$), we notice several differences.
@@
 thus the test practically unidimensional.
-In such a case, the covariance-based version of the direction angles
+In such a case, the covariance-based direction angles
 are extremely close to 0 for all items, as one would expect

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects grammar as suggested by MS Word

@@
 correspondence between correlated latent traits and orthogonal coordinate axes
-is usually implicit in the MIRT literature without further consideration<!--# Exaples? -->
+is usually implicit in the MIRT literature without further consideration
 [see @zhang_theoretical_1999 for a notable exception].
 Moreover, disregarding this geometrical interpretation may lead to
-misconceptions that put the measurement and its conclusions into question.<!-- completar?? -->
-
-A latent space where the traits are correlated is easier to understand as a
-geometrical space with non-orthogonal axes,
+critical misconceptions about the measurement model.
+
+The representation of a latent space where the traits are correlated is more
+accurate on a geometrical coordinate system with non-orthogonal axes
+[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
 often referred to as *general Cartesian coordinates* [@harman_modern_1970].
-Such a space has the advantage that the cosines of the angles between two axes
-can be interpreted as the correlation of the corresponding latent dimensions.
-<!--# Reference needed? -->
-<!-- Problem here: The latent space is not defined like that, according to
-the results of section "Linear algebra of non-orthogonal coordinates". -->
+Such a space has the advantage that the basis vectors and the axes they yield
+have a meaningful interpretation (albeit counterintuitive, as we will see)
+in a multivariate statistical sense.
 The general case of MIRT model estimation implies estimating possibly
@@
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain<!--# Reference to example? -->.
+cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
+[e.g., @thielmann_comparability_2022].
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g. multigroup equating) even when<!-- # Reference needed? -->
-interpreting the latent dimensions as substantive constructs is not a goal.
+non-identity latent covariance matrices (e.g., multigroup equating) even when
+interpreting the latent dimensions as substantive constructs is not a goal
+[for for multidimensional linking and equating procedures,
+see @reckase_linking_2009].
 As we will argue, the usual definitions of the multidimensional difficulty and
@@
 
-In the non-cognitive domain, the use of so-called negatively-keyed items
+On an unrelated basis, negative loadings/discrimination parameters
+are often irrelevant in MIRT applications in the cognitive testing domain.
+It has been argued that discrimination parameters
+"are constrained to be positive" [@ackerman_graphical_1996, p. 315],
+or that “in an [independent-clusters] solution, the loadings can always
+be selected to be positive” [@mcdonald_basis_2000, p. 110].
+However, when such instances appear, they are often
+dismissed as “particularly puzzling” (p. 109),
+or implicitly assuming item malfunctioning without further discussion
+[e.g., @ackerman_multidimensional_2005-1].
+In the non-cognitive domain though, the use of so-called negatively-keyed items
 ---items that tap the negative pole of a certain trait, and thus
-usually reverse scored--- is also widespread.<!--# Reference and example? -->
-It is common to find instruments that combine
-positively- and negatively-keyed items.
-Moreover, a complex item may have discrimination parameters with opposite signs
-for different latent dimensions.<!--# Examples? -->
-Take the multidimensional version of the nominal categories model
-[@thissen_nominal_2010] as an example,
-and consider two response categories tapping a different construct each:
-A positive change in the dimension tapped by one category will inevitably
-lead to a decrease in the probability of response to the other category.
-One can easily note that a monotonically decreasing probability for a certain
-dimension is given by a non-positive discrimination parameter
-(and an unchanging probability by a null discrimination parameter),<!--# Actually, this also applies to the cognitive domain, which may not be very well explained. -->
-contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411].
+usually reverse-scored--- is widespread
+[consider e.g. the statement "Neglect my duties" in the Conscientiousness item
+of the Big-Five Factor Markers, @noauthor_international_nodate].
+Instruments that combine positively- and negatively-keyed items
+are relatively common, and
+complex indicators may have discrimination parameters
+with opposite signs for different latent dimensions
+[see, e.g. @mclarnon_be_2016, for item parcels with
+both positive and negative loadings,
+on method factors in the case of the negative ones].
+Following @ackerman_multidimensional_2005-1, one can easily note that
+a monotonically decreasing probability for a certain dimension
+is given by a non-positive discrimination parameter
+(and an unchanging probability by a null discrimination parameter,
+which also applies to cognitive instruments),
+contrary to the constraints imposed by @reckase_difficulty_1985 [p. 411]
+for determining the multidimensional difficulty.
 However, the multidimensional item parameters can also be defined for
@@
 items that yield monotonous response probabilities, but in the broader sense of
-unchanging monotony. Additionally, we aim at providing a method for
+unchanging monotony. Additionally, we aim to provide a method for
 graphically representing these items in the fashion of @ackerman_creating_1994
@@
 the formulation of the multidimensional parameters.
-We will introduce two versions of these indices,
+We will introduce three different versions,
 showing that either the original or one of the new ones is obtained
 as a function of how we define and interpret the latent space.
-Next, we discuss the properties of the indices,
-propose a graphical representation method,
+Next, we discuss their properties,
+propose a graphical representation method based on the new formulation,
 and present two examples of application of our results.
@@
 
-## Application to examples from the literature
-
-In the following, we apply our results to two instances taken from
-the MIRT literature.
-The first one draws from a classic example by @reckase_multidimensional_2009,
-who presents a three-dimensional instrument made up by 30 items.
-It is supposed to be a cognitive test
-(it is unclear whether the item parameters are estimated from
-actual empirical data, or they are a fictional example,
-created ad-hoc for simulation purposes),
-so all the discrimination parameters are positive.
-The second example, taken from @tezza_modelo_2018, is an application of
-the `r MODEL_ACRONYM` to assess the quality of e-commerce websites.
-Besides being an empirical example, applied to actual test data,
-it pertains to the non-cognitive domain and contains
-items with  negative discrimination parameters.
-This is very convenient for exemplifying our results with
-items with opposite-sign discrimination parameters.
-
-The two examples are represented in Tables
-\@ref(tab:empirical-table-Reckase2009-out)
-and \@ref(tab:empirical-table-Tezza2018-out), respectively.
-Each of these tables shows the original `r MODEL_ACRONYM` parameters
-[the estimation errors in @tezza_modelo_2018 are omitted for clarity],
-along with the agnostic version of the multidimensional parameters;
-the $`r MDISC_AG_PARAM`$ and the $`r DISTANCE_AG_PARAM`$
-are coincident with their respective original sources
-[see Table 6.1 in @reckase_multidimensional_2009, p. 153, and
-Table 5 in @tezza_modelo_2018, p. 926].
-In the first case, the direction angles are also coincident with the values
-provided in the original [@tezza_modelo_2018 do not provide these results].
-In order to facilitate the comparison,
-the covariance-based version of each parameter is provided in
-the column next to its corresponding agnostic version.
-
-### @reckase_multidimensional_2009
+## Examples with empirical data
 
@@
 ```
-
-Table \@ref(tab:empirical-table-Reckase2009-out) shows the parameters of the
-30-item test in the @reckase_multidimensional_2009 example:
-There are three ten-item blocks assumed to approximate
-a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
-and items 21-30 dimension 2
-[see Table 6.1, p. 153, in @reckase_multidimensional_2009].
-In his example, Reckase tests the estimation of these items
-on two simulated datasets: In the first one, the latent trait distribution
-is standard with null correlations; however, the second one has
-
-\begin{equation}
-  `r cov_matrix_value_out`,
-\end{equation}
-
-as covariance matrix, which corresponds to a correlation matrix of
-
-\begin{equation}
-  `r cor_matrix_value_out`.
-\end{equation}
-
-As we can see in Table \@ref(tab:empirical-table-Reckase2009-out),
-the agnostic versions of the multidimensional parameters
-(columns labelled _`r AGNOSTIC_ABBR`_)
-coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
-However, when we consider the latent space structure
-(columns labelled $`r COV_MATRIX`$), we notice several differences.
-First, the $`r MDISC_ITEM`$ parameters tend to be larger
-than their agnostic counterparts.
-This happens because all the discrimination parameters and the latent space
-correlations are positive; in this respect, these items are similar to item 1
-in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
-However, for the items mostly aligned with dimension 2 (items 21-30), as the
-$`r MDISC_ITEM`$ parameters are also scaled by a standard deviation smaller
-than 1, their values decrease, being some of them smaller than the agnostic
-versions. The $`r DISTANCE_PARAM`$ parameters decrease or increase, consequently
-(see Equation \@ref(eq:M2PL-formula-polar)), thus being most of them smaller
-than the agnostic ones (again, with the exceptions only happening
-in the items aligned with dimension 2).
-Finally, we see that the direction angles are not as clearly separated as one
-might expect from the agnostic versions. This effect is more pronounced among
-the items aligned with dimensions 1 and 3, which are strongly correlated.
-In this situation, we notice that the angles tend to be much smaller, as the
-high correlation induces an alignment between the two dimensions
-(compare item 1 in the two test spaces in
-Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
 
@@
 #| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
+#| tab.cap: Item parameters for the empirical example
 item_params_output
 ```
-
-The latter effect can be better noticed in another case, in which Reckase
-uses the same item parameters to demonstrate the over-specification of
-dimensions (see p. 183).
-In that example, he uses a covariance matrix that practically has rank 1,
-implying that the (alleged) three dimensions collapse into a single one, being
-thus the test practically unidimensional.
-In such a case, the covariance-based direction angles
-are extremely close to 0 for all items, as one would expect
-from a unidimensional test, where all items are strictly aligned with the one
-dimension they measure. The reader interested can check out the parameters
-computed in that case in the table in the Supplementary Material.
-
-### @tezza_modelo_2018
 
@@
 comparable among different latent spaces.
+More importantly, the covariance-based version
+of the multidimensional parameters is invariant to
+changes of latent-space basis.
 Once again, this is important in applications where
-potentially different latent spaces are involved.
-Given these considerations, we propose to adopt the covariance-based version of
-the multidimensional parameters and refer to them simply as
+potentially different bases are involved.
+Given these considerations,
+we propose to adopt the covariance-based version of
+the multidimensional parameters as the most general definition,
+and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of preserving the metric of the parameters,
+has the advantage of being scale-invariant,
 which is convenient for fidelity when representing the items.
-Therefore, we recommend using the correlation-based version of the parameters
-for graphical representation purposes only.
-
-One may consider that the formal derivations presented here are overly
-complicated,
+However, it is worth noting that it has been defined ad-hoc
+to fulfill this specific property.
+Therefore, it lacks other desirable properties of the covariance-based version,
+like being invariant to a latent space scaling transformation,
+or yielding a diagonal latent covariance matrix in the corresponding
+orthogonal latent space basis.
+Therefore, we recommend its use for graphical representation purposes only.
+As for the agnostic version, we recommend it only in
+the (unlikely) case of having no information about the latent space structure.
+
+One may consider that the formal derivations presented here are
+overly complicated,
 when one might simply transform the space to rectangular Cartesian coordinates
@@
 which may give relevant information if these are actual, substantive traits.
-Moreover, strictly following @reckase_difficulty_1985's method in the original,
-oblique space has also provided us with an important insight:
+Computing the parameters in the orthonormalized (or just orthogonalized) space,
+by contrast, would require computing the cosines in
+the original basis afterwards.
+
+On the other hand, establishing the relationship between the original and
+the orthonormalized basis has provided us with an important insight:
 The distinction between the latent space and the test space.
@@
 its inner product matrix.
+This result is both surprising and counterintuitive;
+The MIRT literature suggests in several instances that
+"the angle between the axes [of latent ability space] represents
+the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
+We hope our results help shed a clearer light on
+this often misunderstood topic.
+
 It is worth noting that a similar relationship exists in

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd

--- 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 but taking the non-orthogonality of the coordinate axes into account.
-In order to pursue these objectives, we first define the generalization of
+To pursue these objectives, we first define the generalization of
 the `r MODEL_ACRONYM` model and describe its scope of application,
@@
 However, Reckase's Equation 3 implicitly assumes orthogonality.
-To contemplate the general case,
+In order to contemplate the general case,
 we need to introduce a few linear algebra results before proceeding further.
@@
 so its orientation will always pass through the origin.
-Based on these conditions, we need to compute
+On the basis of these conditions, we need to compute
 the origin and end coordinates of the vector,
@@
 
-The geometric properties are easier to apprehend with
+The aforementioned properties are easier to apprehend with
 a visual representation of the items.
@@
 In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
-To keep the horizontal axis invariant,
+In order to keep the horizontal axis invariant,
 we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
@@
 
-## Examples with empirical data
+## Application to examples from the literature
+
+In the following, we apply our results to two instances taken from
+the MIRT literature.
+The first one draws from a classical example by @reckase_multidimensional_2009,
+who presents a three-dimensional instrument made up by 30 items.
+It is supposed to be a cognitive test
+(it is unclear whether the item parameters are estimated from
+actual empirical data or they are a fictional example,
+created ad-hoc for simulation purposes),
+so all the discrimination parameters are positive.
+The second example, taken from @tezza_modelo_2018, is an application of
+the `r MODEL_ACRONYM` to assess the quality of e-commerce websites.
+Besides being an empirical example, applied to actual test data,
+it pertains to the non-cognitive domain and contains
+items with  negative discrimination parameters.
+This is very convenient for exemplifying our results with
+items with opposite-sign discrimination parameters.
+
+The two examples are represented in Tables
+\@ref(tab:empirical-table-Reckase2009-out)
+and \@ref(tab:empirical-table-Tezza2018-out), respectively.
+Each of these tables shows the original `r MODEL_ACRONYM` parameters
+[the estimation errors in @tezza_modelo_2018 are omitted for clarity],
+along with the agnostic version of the multidimensional parameters;
+it can be seen that the $`r MDISC_AG_PARAM`$ and the $`r DISTANCE_AG_PARAM`$
+are coincident with their respective original sources
+[see Table 6.1 in @reckase_multidimensional_2009, p. 153, and
+Table 5 in @tezza_modelo_2018, p. 926].
+In the first case, the direction angles are also coincident with the values
+provided in the original [@tezza_modelo_2018 do not provide these results].
+In order to facilitate the comparison,
+the covariance-based version of each parameter is provided in
+the column next to its corresponding agnostic version.
+
+### @reckase_multidimensional_2009
 
@@
 ```
+
+Table \@ref(tab:empirical-table-Reckase2009-out) shows the parameters of the
+30-item test in the @reckase_multidimensional_2009 example:
+There are three ten-item blocks assumed to approximate
+a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
+and items 21-30 dimension 2
+[see Table 6.1, p. 153, in @reckase_multidimensional_2009].
+In his example, Reckase tests the estimation of these items
+on two simulated datasets: In the first one, the latent trait distribution
+is standard with null correlations; however, the second one has
+
+\begin{equation}
+  `r cov_matrix_value_out`,
+\end{equation}
+
+as covariance matrix, which corresponds to a correlation matrix of
+
+\begin{equation}
+  `r cor_matrix_value_out`.
+\end{equation}
+
+As we can see in Table \@ref(tab:empirical-table-Reckase2009-out),
+the agnostic versions of the multidimensional parameters
+(columns labelled _`r AGNOSTIC_ABBR`_)
+coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
+However, when we take into account the latent space structure
+(columns labelled $`r COV_MATRIX`$), we notice several differences.
+First, the $`r MDISC_ITEM`$ parameters tend to be larger
+than their agnostic counterparts.
+This happens because all the discrimination parameters and the latent space
+correlations are positive; in this respect, these items are similar to item 1
+in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
+However, for the items mostly aligned with dimension 2 (items 21-30), as the
+$`r MDISC_ITEM`$ parameters are also scaled by a standard deviation smaller
+than 1, their values decrease, being some of them smaller than the agnostic
+versions. The $`r DISTANCE_PARAM`$ parameters decrease or increase, consequently
+(see Equation \@ref(eq:M2PL-formula-polar)), thus being most of them smaller
+than the agnostic ones (again, with the exceptions only happening
+in the items aligned with dimension 2).
+Finally, we see that the direction angles are not as clearly separated as one
+might expect from the agnostic versions. This effect is more pronounced among
+the items aligned with dimensions 1 and 3, which are strongly correlated.
+In this situation, we notice that the angles tend to be much smaller, as the
+high correlation induces an alignment between the two dimensions
+(compare item 1 in the two test spaces in
+Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
 
@@
 #| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: Item parameters for the empirical example
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
 item_params_output
 ```
+
+The latter effect can be better noticed in another case, in which Reckase
+uses the same item parameters to demonstrate the over-specification of
+dimensions (see p. 183).
+In that example, he uses a covariance matrix that practically has rank 1,
+implying that the (alleged) three dimensions collapse into a single one, being
+thus the test practically unidimensional.
+In such a case, the covariance-based version of the direction angles
+are extremely close to 0 for all items, as one would expect
+from a unidimensional test, where all items are strictly aligned with the one
+dimension they measure. The reader interested can check out the parameters
+computed in that case in the table in the Supplementary Material.
+
+### @tezza_modelo_2018
 
@@
 ```
+
+@tezza_modelo_2018 provide us with another example that allows showcasing
+the effect of correlated dimensions on negative discrimination parameters.
+In their model, dimensions 1 and 4 are estimated to have a correlation of
+`r DIMS_CORR`.
+With such latent space structure, items that strongly discriminate in any or
+both of these two dimensions change drastically their multidimensional
+parameters from the agnostic to the covariance-based version
+(see Table \@ref(tab:empirical-table-Tezza2018-out)) ---this happens,
+for example, with items 57 and 60. These items also have different-sign
+discrimination parameters in dimensions 1 and 4, so their situation is
+similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
+Figure \@ref(fig:item-plot-out),
+thus the decrease in their $`r MDISC_ITEM`$ parameter.
+The effect on the direction is also patent in the angles these items make with
+these dimensions, which change up to almost 20º, compared with at most 7º in the
+other two dimensions, and around 1º in most cases.
+
+On the other hand, in cases such as items 12 or 59, which have
+relatively low discrimination parameters on those two dimensions
+(compared with the ones in dimension 2 and 3, respectively) are barely affected.
+The most extreme change in these two examples occurs in
+the direction of item 12 with respect to dimension 4. We see how the large
+discrimination in dimension 1, when combined with the correlation
+with dimension 4, induces a decrease in the angular direction
+with that dimension.
+In this case, being the discrimination in dimension 4 so low, the situation is
+very similar to item 4 in Table \@ref(tab:example-items-table-out)
+and Figure \@ref(fig:item-plot-out).
+
+<!---BLOCK_LANDSCAPE_START--->
 
@@
 which may give relevant information if these are actual, substantive traits.
-Moreover, strictly following @reckase_difficulty_1985's method in the original,
-oblique space has also provided us with an important insight:
+Computing the parameters in the orthonormalized (or just orthogonalized) space,
+by contrast, would require to compute the cosines in
+the original basis afterwards.
+
+On the other hand, establishing the relationship between the original and
+the orthonormalized basis has provided us with an important insight:
 The distinction between the latent space and the test space.
-Our derivations show that, to properly represent the items,
+Our derivations show that, in order to properly represent the items,
 we need to define a vector space that has

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 The distinction between the latent space and the test space.
-Our derivations show that, in order to properly represent the items,
+Our derivations show that, to properly represent the items,
 we need to define a vector space that has

### 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 to fulfill this specific property.
-Therefore it lacks other desirable properties of the covariance-based version,
+Therefore, it lacks other desirable properties of the covariance-based version,
 like being invariant to a latent space scaling transformation,
@@
 Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require to compute the cosines in
+by contrast, would require computing the cosines in
 the original basis afterwards.
@@
 We hope our results help shed a clearer light on
-this often missunderstood topic.
+this often misunderstood topic.
 

### 2024-07-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Defines test space before, where it should

@@
 but taking the non-orthogonality of the coordinate axes into account.
-To pursue these objectives, we first define the generalization of
+In order to pursue these objectives, we first define the generalization of
 the `r MODEL_ACRONYM` model and describe its scope of application,
@@
 keeps the norm of $`r DISCR_VECTOR`$ invariant.
+Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
+the *test space*, as it is the space where
+the items (i.e., the test elements) are represented.
 
@@
 they are still two different spaces. <!-- # TODO: To discussion? -->
-Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
-the *test space*, as it is the space where
-the items (i.e., the test elements) are represented.
 
@@
 comparable among different latent spaces.
-More importantly, the covariance-based version
-of the multidimensional parameters is invariant to
-changes of latent-space basis.
 Once again, this is important in applications where
-potentially different bases are involved.
-Given these considerations,
-we propose to adopt the covariance-based version of
-the multidimensional parameters as the most general definition,
-and refer to them simply as
+potentially different latent spaces are involved.
+Given these considerations, we propose to adopt the covariance-based version of
+the multidimensional parameters and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of being scale-invariant,
+has the advantage of preserving the metric of the parameters,
 which is convenient for fidelity when representing the items.
-However, it is worth noting that it has been defined ad-hoc
-to fulfill this specific property.
-Therefore, it lacks other desirable properties of the covariance-based version,
-like being invariant to a latent space scaling transformation,
-or yielding a diagonal latent covariance matrix in the corresponding
-orthogonal latent space basis.
-Therefore, we recommend its use for graphical representation purposes only.
-As for the agnostic version, we recommend it only in
-the (unlikely) case of having no information about the latent space structure.
-
-One may consider that the formal derivations presented here are
-overly complicated,
+Therefore, we recommend using the correlation-based version of the parameters
+for graphical representation purposes only.
+
+One may consider that the formal derivations presented here are overly
+complicated,
 when one might simply transform the space to rectangular Cartesian coordinates
@@
 which may give relevant information if these are actual, substantive traits.
-Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require computing the cosines in
-the original basis afterwards.
-
-On the other hand, establishing the relationship between the original and
-the orthonormalized basis has provided us with an important insight:
+Moreover, strictly following @reckase_difficulty_1985's method in the original,
+oblique space has also provided us with an important insight:
 The distinction between the latent space and the test space.
@@
 its inner product matrix.
-This result is both surprising and counterintuitive;
-The MIRT literature suggests in several instances that
-"the angle between the axes [of latent ability space] represents
-the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
-We hope our results help shed a clearer light on
-this often misunderstood topic.
-
 It is worth noting that a similar relationship exists in

### 2024-07-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Explains better the link between spaces
As suggested by @mluzardov in the 2024/07/18 meeting

@@
 in the test space:
-As both $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are orthonormal,
-we can map one directly onto the other. Without loss of generality, we may thus
+As $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are both
+orthonormal in $`r REAL_N_SPACE`$, without loss of generality
+we can consider that $`r BASES_EQ`$ (save for a rotation).
+We can then map the latent space directly onto the test space and
 apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
-<!-- # TODO: Revisar esto bien. Esto sólo aplica correctamente si efectivamente
-"equiparamos" las bases U y U*. Además, en la Proposición 2 estamos asumiendo
-que la "base de enlace" es la canónica, no una base ortonormal cualquiera. -->
 As we will see later, this representation is very convenient, as

### 2024-07-18 R/Formulae.R

@@ -823,6 +823,11 @@ DISCR_STD_COORDS_MOD_EQ <- latex_eq(
   DISCR_STD_COORDS_MODULE
 )
 
+### Relationship between test space and latent space: ----
+
+# Equating the orthonormal bases:
+BASES_EQ <- latex_eq(LS_ORTH_BASIS, TEST_SPACE_STD_BASIS)
+
 # Item direction cosines (in test space):
 ANGLE_TS                          <- latex("\\alpha")
 ANGLE_VEC_TS                      <- latex_bf(ANGLE_TS)

### 2024-07-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects reference to numbered items

--- 2024-07-17 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-Given 1,
+Given i,
 
@@
 
-Using 2 and 3 we get that
+Using ii and iii we get that
 

### 2024-07-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Explains misunderstanding of axes reprsentation

@@
 its inner product matrix.
-This result is both surprising and counterintuitive;
-The MIRT literature suggests in several instances that
-"the angle between the axes [of latent ability space] represents
+This implies that the axes defined by the test space basis,
+and not the latent space basis,
+are a geometrical representation of the covariance structure.
+This result is both surprising and counterintuitive
+---Several instances in the MIRT literature suggest that
+"the angle between the axes [of the latent ability space] represents
 the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].

### 2024-07-18 output/paper_draft/paper_M2PLM_parameters.Rmd
Uses Proposition 2 for derivating the coordinates

--- 2024-07-18 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 
-Applying Equation \@ref(eq:dir-cos-vec-std),
-for the origin coordinates we have that
-
-\begin{equation}
-  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
-\end{equation}
-
-which, operating, results in
-
-\begin{align}
-  `r ORIGIN_ITEM_COORDS_EQ`    \\
-  `r ORIGIN_TRANSF_SQUARED_EQ` \\
-  `r ORIGIN_INNERPROD_EQ`      \\
-  `r ORIGIN_INNERPROD_DIAG_EQ`
-\end{align}
-
-and, finally,
-
-\begin{equation}
-  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
-\end{equation}
-
-The end coordinates can be computed by a similar procedure, which results in
+Applying Proposition 2, we have that
+
+\begin{equation}
+  `r DIR_COS_ITEM_EQ`.
+\end{equation}
+
+Therefore, the coordinates result in
+
+\begin{equation}
+  `r ORIGIN_ITEM_COORDS_EQ`                          (\#eq:origin-coords-result)
+\end{equation}
+
+and
 

### 2024-07-18 R/Formulae.R

@@ -1100,32 +1100,20 @@ END_TRANSF_DEF    <- latex(TRANSFORM_MATRIX_TRANSP_INV, END_ITEM_BASIS)
 END_TRANSF_EQ     <- latex_eq(END_ITEM_ORTH, END_TRANSF_DEF)
 
 # Coordinates in orthonormal space:
-ORIGIN_ORTH_DEF  <- latex(DISTANCE_PARAM, DIR_COS_VEC_STD)
+ORIGIN_ORTH_DEF  <- latex(DISTANCE_PARAM, DIR_COS_ITEM_VEC_STD)
 ORIGIN_ORTH_EQ   <- latex_eq(ORIGIN_ITEM_ORTH, ORIGIN_ORTH_DEF, .align = TRUE)
-END_ORTH_DIR_DEF <- latex(ORIGIN_ITEM_ORTH, '+', MDISC_ITEM, DIR_COS_VEC_STD)
+END_ORTH_DIR_DEF <- latex(
+  ORIGIN_ITEM_ORTH,
+  '+',
+    MDISC_ITEM,
+    DIR_COS_ITEM_VEC_STD
+)
 END_ORTH_EQ      <- latex_eq(END_ITEM_ORTH, END_ORTH_DIR_DEF, .align = TRUE)
 
 # Origin coordinates:
-ORIGIN_ORTH_DIR_DEF    <- latex(DISTANCE_PARAM, DIR_COS_VEC_STD)
-DIR_COS_ORTH_DISCR_DEF <- latex_frac(DISCR_STD_COORDS_DEF, DISCR_VECTOR_MODULE)
-ORIGIN_ORTH_DISCR_DEF  <- latex(DISTANCE_PARAM, DIR_COS_ORTH_DISCR_DEF)
-ORIGIN_ORTH_DISCR_EQ   <- latex_eq(ORIGIN_TRANSF_DEF, ORIGIN_ORTH_DISCR_DEF)
-
-# Origin coordinates derivation:
-ORIGIN_TRANSF_DISCR_DEF       <- latex(
-  DISTANCE_PARAM,
+DIR_COS_ITEM_TRANSF           <- latex(
   TRANSFORM_MATRIX_TRANSP,
-  DIR_COS_ORTH_DISCR_DEF
-)
-ORIGIN_TRANSF_SQUARED_DEF     <- latex(
-  DISTANCE_PARAM,
-  INNER_PROD_TRANSF_DEF,
[…]

### 2024-07-24 R/Formulae.R
Expresses transform matrix correctly as identity
In deriving the test space direction cosines, the orthonormal basis equals basis C in proposition 2; therefore, basis U is equal to C, so the transform basis is not just any orthonormal matrix but the identity matrix

@@ -686,10 +686,6 @@ IRF_2ND_DIFF_EQ  <- latex_eq(IRF_2ND_DIFF, IRF_2ND_DIFF_DEF)
 IRF_MAX_SLOPE    <- '.5' ## TODO: Format prop-like!!
 IRF_MAX_SLOPE_EQ <- latex_eq(IRF_ABBR, IRF_MAX_SLOPE)
 
-# Condition for transformation to orthonormal basis:
-ID_MATRIX        <- latex_bf('I')
-ALT_MATRIX_AS_ID <- latex_eq(DIAG_MATRIX_INNER_PROD_ALT, ID_MATRIX)
-
 # Director cosines and orthonormalized direction cosines equivalence:
 DIR_COS_AS_DIR_COS_STD        <- latex(
   TRANSFORM_MATRIX_TRANSP_INV,
@@ -890,6 +886,7 @@ MIL_PARAM  <- latex("MIL")
 
 # Condition to meet:
 BASIS_EQ              <- latex_equiv(LS_BASIS, LS_ORTH_BASIS)
+ID_MATRIX             <- latex_bf('I')
 INNER_PROD_MAT_STD_EQ <- latex_eq(INNER_PROD_MAT_EQ, ID_MATRIX)
 
 # MDISC:

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Expresses transform matrix correctly as identity
In deriving the test space direction cosines, the orthonormal basis equals basis C in proposition 2; therefore, basis U is equal to C, so the transform basis is not just any orthonormal matrix but the identity matrix
Moves "equivalence of orthonormal matrices"
This assertion goes better in section "test space definition" to justify that Reckase's (1985) case is a particular case of ours.

@@
 
-To make @reckase_difficulty_1985's definition a particular case of ours
+As we did with $`r LS_ORTH_BASIS`$ before,
 we can conveniently assume that the inner product in
-$`r TEST_SPACE_STD_BASIS`$ is the dot product,
-as we did with $`r LS_ORTH_BASIS`$.
+$`r TEST_SPACE_STD_BASIS`$ is the dot product.
+Then, as both $`r TEST_SPACE_STD_BASIS`$ and $`r LS_ORTH_BASIS`$ are
+orthonormal in $`r REAL_N_SPACE`$,
+to make @reckase_difficulty_1985's definition a particular case of ours we can,
+without loss of generality  (i.e., appying an arbitrary rotation),
+consider that $`r BASES_EQ`$.
 As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
@@
 $`r LS_ORTH_BASIS`$.
-Now, applying Proposition 2 (with $`r TRANSFORM_MATRIX_ALT`$
-an orthonormal matrix and $`r ALT_MATRIX_AS_ID`$) we obtain
+Now, applying Proposition 2 (with both $`r TRANSFORM_MATRIX_ALT`$ and
+$`r DIAG_MATRIX_INNER_PROD_ALT`$ equal to the identity matrix) we obtain
 

### 2024-07-24 R/Formulae.R

@@ -246,7 +246,7 @@ INNER_PROD_MAT_ELEMENT_EQ  <- latex_eq(
 )
 INNER_PROD_MATRIX_INDEX    <- latex(AUX_INDEX, DIM_INDEX, .sep = SEP_COMMA)
 
-### Test space ----
+### Test space definition ----
 
 # Test space definition:
 TEST_SPACE   <- latex_bf("A")
@@ -295,6 +295,9 @@ DISCR_STD_COORDS_EQ         <- latex_eq(
   DISCR_STD_COORDS,
   DISCR_STD_COORDS_DEF
 )
+
+# Equating the orthonormal bases:
+BASES_EQ <- latex_eq(LS_ORTH_BASIS, TEST_SPACE_STD_BASIS)
 
 # Inverse inner product matrix:
 INNER_PROD_MATRIX_INV       <- latex_inverse(INNER_PROD_MATRIX)
@@ -820,9 +823,6 @@ DISCR_STD_COORDS_MOD_EQ <- latex_eq(
 )
 
 ### Relationship between test space and latent space: ----
-
-# Equating the orthonormal bases:
-BASES_EQ <- latex_eq(LS_ORTH_BASIS, TEST_SPACE_STD_BASIS)
 
 # Item direction cosines (in test space):
 ANGLE_TS                          <- latex("\\alpha")

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Complements previous commit
The addition in the "Test space definition" section must be complemented with the deletion of the same content in the "Relationship among the spaces" section

@@
 can be superimposed in this case,
-they are still two different spaces. <!-- # TODO: To discussion? -->
+they are still two different spaces.
 
@@
 in the test space:
-As $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are both
-orthonormal in $`r REAL_N_SPACE`$, without loss of generality
-we can consider that $`r BASES_EQ`$ (save for a rotation).
 We can then map the latent space directly onto the test space and

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Derives test space direction cosines from Prop. 1

@@
 in the test space:
-We can then map the latent space directly onto the test space and
-apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
+Applying Proposition 1, if $`r ANGLE_VEC_TS_ITEM`$ is
 the vector of direction angles of $`r DISCR_VECTOR`$,
@@
   \begin{split}
-    `r COV_MATRIX_TRANSF_ID_EQ` \\
+    `r COV_MATRIX_TRANSF_ID_EQ`, \\
     `r COV_MATRIX_RESULT_EQ`.

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects derivation of coordinates

@@
 In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
-$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
+$`r ORIGIN_ORTH_EQ`$ and $`r END_ORTH_EQ`$, respectively,
 and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
@@
 i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
-Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
-the direction cosines in $`r TEST_SPACE_STD_BASIS`$
-are $`r DIR_COS_ITEM_VEC_STD`$.
-Therefore,
-
-\begin{align}
-  `r ORIGIN_ORTH_EQ` \\
-  `r END_ORTH_EQ`.
-\end{align}
-
-Applying Proposition 2, we have that
+Applying Proposition 2 in the test space, we have that
 
@@
 Of course,
-from Equations \@ref(eq:origin-orth-discr) and \@ref(eq:end-coords-result)
+from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
 it is easy to express the coordinates in terms of the parameters in the original

### 2024-07-24 R/Formulae.R

@@ -1085,14 +1085,26 @@ MIL_CORR_PARAM_EQ         <- latex_def(
 ### Vector representation: ----
 
 # Item coordinates definition:
+
+# Origin coordinates:
 ORIGIN            <- latex_bf('o')
 ORIGIN_ITEM       <- ORIGIN      |> latex_sub(ITEM_INDEX)
 ORIGIN_ITEM_BASIS <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
 ORIGIN_ITEM_ORTH  <- ORIGIN_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
+DIR_COS_ORTH_TS   <- latex_raised_to(
+  DIR_COS_ITEM_VEC_TS,
+  exp = TEST_SPACE_STD_BASIS
+)
+ORIGIN_ORTH_DEF   <- latex(DISTANCE_PARAM, DIR_COS_ORTH_TS)
+ORIGIN_ORTH_EQ    <- latex_eq(ORIGIN_ITEM_ORTH, ORIGIN_ORTH_DEF)
+
+# End coordinates:
 END               <- latex_bf('e')
 END_ITEM          <- END      |> latex_sub(ITEM_INDEX)
 END_ITEM_BASIS    <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_BASIS)
 END_ITEM_ORTH     <- END_ITEM |> latex_raised_to(exp = TEST_SPACE_STD_BASIS)
+END_ORTH_DIR_DEF  <- latex(ORIGIN_ITEM_ORTH, '+', MDISC_ITEM, DIR_COS_ORTH_TS)
+END_ORTH_EQ       <- latex_eq(END_ITEM_ORTH, END_ORTH_DIR_DEF)
 
 # Coordinate transformation to orthonormal basis:
 ORIGIN_TRANSF_DEF <- latex(TRANSFORM_MATRIX_TRANSP_INV, ORIGIN_ITEM_BASIS)
@@ -1100,22 +1112,8 @@ ORIGIN_TRANSF_EQ  <- latex_eq(ORIGIN_ITEM_ORTH, ORIGIN_TRANSF_DEF)
 END_TRANSF_DEF    <- latex(TRANSFORM_MATRIX_TRANSP_INV, END_ITEM_BASIS)
[…]

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Moves vector and coordinate equivalence earlier

@@
 $`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
-Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
+Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$
+and, as $`r DISCR_VECTOR`$ is expressed in this basis,
+its algebraic vector values correspond to its coordinates, i.e.,
+$`r DISCR_VECTOR_COORDS_EQ`$.
 
@@
 \begin{equation}
-  `r SLOPE_MAX_STD_COSINES_EQ`.                     (\#eq:slope-max-std-cosines)
-\end{equation}
-
-As $`r DISCR_VECTOR`$ is defined in space $`r TEST_SPACE`$ with basis
-$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
-coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
-\@ref(eq:discr-vector-std),
+  `r SLOPE_MAX_STD_COSINES_EQ`                      (\#eq:slope-max-std-cosines)
+\end{equation}
+
+and, by Equation \@ref(eq:discr-vector-std),
 

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Expresses direction correctly in the test space

@@
 
-The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
+The measurement direction is given by
+the signs of $`r DIR_COS_ITEM_VEC_BASIS_TS`$ which,
 also due to the strict positiveness of $`r MDISC_SYM`$,

### 2024-07-24 R/Formulae.R

@@ -1163,7 +1163,7 @@ END_ITEM_PARAMS_EQ     <- latex_eq(
 #### Geometric properties of the parameters: ----
 
 # Direction cosines:
-ANGLE_VECTORS_ITEM   <- latex_sub(ANGLE, "$ITEM_INDEX$$DIM_INDEX$")
+ANGLE_VECTORS_ITEM   <- latex_sub(ANGLE_TS, "$ITEM_INDEX$$DIM_INDEX$")
 COS_VECTORS_ITEM     <- latex_cos(ANGLE_VECTORS_ITEM)
 SIGN_COS_VEC_ITEM    <- latex_sign(COS_VECTORS_ITEM, .par = TRUE)
 SIGN_COS_VEC_ITEM_EQ <- latex_eq(SIGN_COS_VEC_ITEM, 0)

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects style in "Graphical representation example"

@@
 currently available technology),
-we will limit ourselves to the bidimensional case here.
-
+we limit ourselves to the bidimensional case here.
 The example in Figure \@ref(fig:item-plot-out) showcases a representation of

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 
-## Examples with empirical data
+## Application to examples from the literature
+
+In the following, we apply our results to two instances taken from
+the MIRT literature.
+The first one draws from a classic example by @reckase_multidimensional_2009,
+who presents a three-dimensional instrument made up by 30 items.
+It is supposed to be a cognitive test
+(it is unclear whether the item parameters are estimated from
+actual empirical data, or they are a fictional example,
+created ad-hoc for simulation purposes),
+so all the discrimination parameters are positive.
+The second example, taken from @tezza_modelo_2018, is an application of
+the `r MODEL_ACRONYM` to assess the quality of e-commerce websites.
+Besides being an empirical example, applied to actual test data,
+it pertains to the non-cognitive domain and contains
+items with  negative discrimination parameters.
+This is very convenient for exemplifying our results with
+items with opposite-sign discrimination parameters.
+
+The two examples are represented in Tables
+\@ref(tab:empirical-table-Reckase2009-out)
+and \@ref(tab:empirical-table-Tezza2018-out), respectively.
+Each of these tables shows the original `r MODEL_ACRONYM` parameters
+[the estimation errors in @tezza_modelo_2018 are omitted for clarity],
+along with the agnostic version of the multidimensional parameters;
+the $`r MDISC_AG_PARAM`$ and the $`r DISTANCE_AG_PARAM`$
+are coincident with their respective original sources
+[see Table 6.1 in @reckase_multidimensional_2009, p. 153, and
+Table 5 in @tezza_modelo_2018, p. 926].
+In the first case, the direction angles are also coincident with the values
+provided in the original [@tezza_modelo_2018 do not provide these results].
+In order to facilitate the comparison,
+the covariance-based version of each parameter is provided in
+the column next to its corresponding agnostic version.
+
+### @reckase_multidimensional_2009
 
@@
 ```
+
+Table \@ref(tab:empirical-table-Reckase2009-out) shows the parameters of the
+30-item test in the @reckase_multidimensional_2009 example:
+There are three ten-item blocks assumed to approximate
+a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
+and items 21-30 dimension 2
+[see Table 6.1, p. 153, in @reckase_multidimensional_2009].
+In his example, Reckase tests the estimation of these items
+on two simulated datasets: In the first one, the latent trait distribution
+is standard with null correlations; however, the second one has
+
+\begin{equation}
+  `r cov_matrix_value_out`,
+\end{equation}
+
+as covariance matrix, which corresponds to a correlation matrix of
+
+\begin{equation}
+  `r cor_matrix_value_out`.
+\end{equation}
+
+As we can see in Table \@ref(tab:empirical-table-Reckase2009-out),
+the agnostic versions of the multidimensional parameters
+(columns labelled _`r AGNOSTIC_ABBR`_)
+coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
+However, when we consider the latent space structure
+(columns labelled $`r COV_MATRIX`$), we notice several differences.
+First, the $`r MDISC_ITEM`$ parameters tend to be larger
+than their agnostic counterparts.
+This happens because all the discrimination parameters and the latent space
+correlations are positive; in this respect, these items are similar to item 1
+in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
+However, for the items mostly aligned with dimension 2 (items 21-30), as the
+$`r MDISC_ITEM`$ parameters are also scaled by a standard deviation smaller
+than 1, their values decrease, being some of them smaller than the agnostic
+versions. The $`r DISTANCE_PARAM`$ parameters decrease or increase, consequently
+(see Equation \@ref(eq:M2PL-formula-polar)), thus being most of them smaller
+than the agnostic ones (again, with the exceptions only happening
+in the items aligned with dimension 2).
+Finally, we see that the direction angles are not as clearly separated as one
+might expect from the agnostic versions. This effect is more pronounced among
+the items aligned with dimensions 1 and 3, which are strongly correlated.
+In this situation, we notice that the angles tend to be much smaller, as the
+high correlation induces an alignment between the two dimensions
+(compare item 1 in the two test spaces in
+Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
 
@@
 #| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: Item parameters for the empirical example
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
 item_params_output
 ```
+
+The latter effect can be better noticed in another case, in which Reckase
+uses the same item parameters to demonstrate the over-specification of
+dimensions (see p. 183).
+In that example, he uses a covariance matrix that practically has rank 1,
+implying that the (alleged) three dimensions collapse into a single one, being
+thus the test practically unidimensional.
+In such a case, the covariance-based direction angles
+are extremely close to 0 for all items, as one would expect
+from a unidimensional test, where all items are strictly aligned with the one
+dimension they measure. The reader interested can check out the parameters
+computed in that case in the table in the Supplementary Material.
+
+### @tezza_modelo_2018
 
@@
 its inner product matrix.
+This implies that the axes defined by the test space basis,
+and not the latent space basis,
+are a geometrical representation of the covariance structure.
+This result is both surprising and counterintuitive
+---Several instances in the MIRT literature suggest that
+"the angle between the axes [of the latent ability space] represents
+the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
+We hope our results help shed a clearer light on
+this often misunderstood topic.
+
 It is worth noting that a similar relationship exists in

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Discusses generalization to unchanging monotony

--- 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 $`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
-Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$
-and, as $`r DISCR_VECTOR`$ is expressed in this basis,
-its algebraic vector values correspond to its coordinates, i.e.,
-$`r DISCR_VECTOR_COORDS_EQ`$.
-
-As we did with $`r LS_ORTH_BASIS`$ before,
+Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
+
+To make @reckase_difficulty_1985's definition a particular case of ours
 we can conveniently assume that the inner product in
-$`r TEST_SPACE_STD_BASIS`$ is the dot product.
-Then, as both $`r TEST_SPACE_STD_BASIS`$ and $`r LS_ORTH_BASIS`$ are
-orthonormal in $`r REAL_N_SPACE`$,
-to make @reckase_difficulty_1985's definition a particular case of ours we can,
-without loss of generality  (i.e., appying an arbitrary rotation),
-consider that $`r BASES_EQ`$.
+$`r TEST_SPACE_STD_BASIS`$ is the dot product,
+as we did with $`r LS_ORTH_BASIS`$.
 As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
@@
 $`r LS_ORTH_BASIS`$.
-Now, applying Proposition 2 (with both $`r TRANSFORM_MATRIX_ALT`$ and
-$`r DIAG_MATRIX_INNER_PROD_ALT`$ equal to the identity matrix) we obtain
+Now, applying Proposition 2 (with $`r TRANSFORM_MATRIX_ALT`$
+an orthonormal matrix and $`r ALT_MATRIX_AS_ID`$) we obtain
 
@@
 \begin{equation}
-  `r SLOPE_MAX_STD_COSINES_EQ`                      (\#eq:slope-max-std-cosines)
-\end{equation}
-
-and, by Equation \@ref(eq:discr-vector-std),
+  `r SLOPE_MAX_STD_COSINES_EQ`.                     (\#eq:slope-max-std-cosines)
+\end{equation}
+
+As $`r DISCR_VECTOR`$ is defined in space $`r TEST_SPACE`$ with basis
+$`r TEST_SPACE_BASIS`$, its algebraic vector values correspond to its
+coordinates, i.e., $`r DISCR_VECTOR_COORDS_EQ`$. Therefore, by Equation
+\@ref(eq:discr-vector-std),
 
@@
 can be superimposed in this case,
-they are still two different spaces.
+they are still two different spaces. <!-- # TODO: To discussion? -->
 
@@
 in the test space:
-Applying Proposition 1, if $`r ANGLE_VEC_TS_ITEM`$ is
+As $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are both
+orthonormal in $`r REAL_N_SPACE`$, without loss of generality
+we can consider that $`r BASES_EQ`$ (save for a rotation).
+We can then map the latent space directly onto the test space and
+apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
 the vector of direction angles of $`r DISCR_VECTOR`$,
@@
   \begin{split}
-    `r COV_MATRIX_TRANSF_ID_EQ`, \\
+    `r COV_MATRIX_TRANSF_ID_EQ` \\
     `r COV_MATRIX_RESULT_EQ`.
@@
 In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
-$`r ORIGIN_ORTH_EQ`$ and $`r END_ORTH_EQ`$, respectively,
+$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
 and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
@@
 i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
-Applying Proposition 2 in the test space, we have that
-
-\begin{equation}
-  `r DIR_COS_ITEM_EQ`.
-\end{equation}
-
-Therefore, the coordinates result in
-
-\begin{equation}
-  `r ORIGIN_ITEM_COORDS_EQ`                          (\#eq:origin-coords-result)
-\end{equation}
-
-and
+Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
+the direction cosines in $`r TEST_SPACE_STD_BASIS`$
+are $`r DIR_COS_ITEM_VEC_STD`$.
+Therefore,
+
+\begin{align}
+  `r ORIGIN_ORTH_EQ` \\
+  `r END_ORTH_EQ`.
+\end{align}
+
+Applying Equation \@ref(eq:dir-cos-vec-std),
+for the origin coordinates we have that
+
+\begin{equation}
+  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
+\end{equation}
+
+which, operating, results in
+
+\begin{align}
+  `r ORIGIN_ITEM_COORDS_EQ`    \\
+  `r ORIGIN_TRANSF_SQUARED_EQ` \\
+  `r ORIGIN_INNERPROD_EQ`      \\
+  `r ORIGIN_INNERPROD_DIAG_EQ`
+\end{align}
+
+and, finally,
+
+\begin{equation}
+  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
+\end{equation}
+
+The end coordinates can be computed by a similar procedure, which results in
 
@@
 Of course,
-from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
+from Equations \@ref(eq:origin-orth-discr) and \@ref(eq:end-coords-result)
 it is easy to express the coordinates in terms of the parameters in the original
@@
 
-The measurement direction is given by
-the signs of $`r DIR_COS_ITEM_VEC_BASIS_TS`$ which,
+The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
 also due to the strict positiveness of $`r MDISC_SYM`$,
@@
 the multivariate latent variable, and (2) to any case of unchanging monotony.
-It is worth noting that @zhang_theoretical_1999 already proposed a formulation
+The latter may seem superfluous, given that several examples in the literature
+already make use of this generalization
+[e.g., the one we have used to illustrate it: @tezza_modelo_2018].
+However, this generalization had never been formalized, up to now.
+Therefore, we highlight here the interest of showing how
+the item psychometric properties are paralleled by their geometrical properties,
+and how this holds for any case of unchanging monotony
+of the item response function.
+Regarding the first generalization,
+it is worth noting that @zhang_theoretical_1999 already proposed a formulation
 of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
@@
 (although constrained to non-decreasing monotony).
-However, they provided no formal proof for this result.
+However, they provided no formal proof for this result either.
 Our results provide this formal derivation, although only for the specific

### 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
Discusses other "allowed" use of agnostic version

@@
 
-With that purpose in mind, we have defined two versions of these parameters:
+With that purpose in mind, we have defined two new versions of these parameters:
 One that takes into account the whole covariance structure,
@@
 As for the agnostic version, we recommend it only in
-the (unlikely) case of having no information about the latent space structure.
+the (unlikely) case of having no information about the latent space structure,
+or when such structure is considered irrelevant
+(e.g., plotting the discrimination parameters after an oblique rotation).
 

### 2024-09-02 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 which is convenient for fidelity when representing the items.
-However, it is worth noting that it has been defined ad-hoc
-to fulfill this specific property.
-Therefore, it lacks other desirable properties of the covariance-based version,
-like being invariant to a latent space scaling transformation,
-or yielding a diagonal latent covariance matrix in the corresponding
-orthogonal latent space basis.
-Therefore, we recommend its use for graphical representation purposes only.
+However, as it has been defined ad-hoc to fulfill this specific property,
+it lacks the other desirable ones the covariance-based version has
+(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
+the corresponding orthogonal basis).
+Therefore, we recommend its use for graphical representation purposes only,
+although in many applications both versions will be coincident, fortunately.
 As for the agnostic version, we recommend it only in
 the (unlikely) case of having no information about the latent space structure,
-or when such structure is considered irrelevant
-(e.g., plotting the discrimination parameters after an oblique rotation).
+or when such structure is irrelevant
+(e.g., plotting the discrimination parameters for interpretation purposes).
 

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd

--- 2024-07-24 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 in polar coordinates, as in @reckase_difficulty_1985.
+
 However, Reckase's Equation 3 implicitly assumes orthogonality.
-To contemplate the general case,
+In order to contemplate the general case,
 we need to introduce a few linear algebra results before proceeding further.
@@
 Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
-the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$.
-
-### Test space definition
-
-Let $`r TEST_SPACE`$ be the vector space made up by
-the set of $`r DISCR_VECTOR`$ vectors,
-with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
-We want @reckase_difficulty_1985's definition of
-the multidimensional parameters in rectangular coordinates to be
-a particular case of our more general definition;
-we can thus consider an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
-in $`r TEST_SPACE`$ such that $`r DISCR_STD_COORDS_ALT_EQ`$
-corresponds in the model to $`r TRAIT_ORTH_COORDS`$.
-Since the response probability to the `r MODEL_ACRONYM`,
-given by Equation \@ref(eq:M2PL-formulation), must be invariant with respect to
-the change of basis expressed in Equation \@ref(eq:change-basis),
-$`r INVARIANCE_EQ`$.
-Therefore, we find a basis $`r TEST_SPACE_BASIS_EQ`$ in $`r TEST_SPACE`$
-defined by
-
-\begin{equation}
-  `r DISCR_STD_COORDS_EQ`,                               (\#eq:discr-vector-std)
-\end{equation}
-
-with $`r DISCR_COORDS_ALT_EQ`$, and
-$`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
-Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
-
-To make @reckase_difficulty_1985's definition a particular case of ours
-we can conveniently assume that the inner product in
-$`r TEST_SPACE_STD_BASIS`$ is the dot product,
-as we did with $`r LS_ORTH_BASIS`$.
-As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
+the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
+$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
 invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
-The norm of $`r DISCR_VECTOR`$ thus fulfills
-
-\begin{equation}
-  `r DISCR_VECTOR_MODULE_EQ`.
-\end{equation}
-
-Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
-$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
-keeps the norm of $`r DISCR_VECTOR`$ invariant.
-Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
-the *test space*, as it is the space where
-the items (i.e., the test elements) are represented.
-
-### Direction cosines in the latent space
 
@@
 
-i. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
-
-i. $`r TRAIT_VECTOR_POLAR_EQ`$
-
-i. $`r TRAIT_VEC_ALT_POLAR_EQ`$
-
-Given i,
+1. $`r TRAIT_VEC_ORTH_POLAR_EQ`$
+
+1. $`r TRAIT_VECTOR_POLAR_EQ`$
+
+1. $`r TRAIT_VEC_ALT_POLAR_EQ`$
+
+Given (1),
 
@@
 
-Using ii and iii we get that
+Using (2) and (3) we get that
 
@@
 
-# Results
-
-## Relationship between test space and latent space
-
-Before presenting the results about the parameters,
-we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
-As shown before, this space is an entity by itself,
-distinct from the latent space $`r LATENT_SPACE`$.
-Despite this, the dot product in the exponent of Equation
-\@ref(eq:M2PL-formulation) induces a bijective relationship between
-their two sets of respective bases.
-Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
-with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
-$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
-with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
-for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
-Only when the two bases are orthonormal,
-the two spaces share a common inner product (i.e., the dot product),
-and thus can be mistaken.
-However, it is worth noting that, although their representation
-can be superimposed in this case,
-they are still two different spaces. <!-- # TODO: To discussion? -->
-
-We must make a few observations here.
-First, the transformation to the orthonormal bases of the two spaces allows
-mapping the $`r MODEL_ACRONYM`$ parameters indistinctly in the two spaces,
-by projecting one space onto the other.
-This allows representing the coordinates of the discrimination vectors
-in the latent space and, conversely,
-the coordinates of the latent trait vectors in the test space.
-As we can see, the invariance of the model holds,
-no matter in which space we represent the coordinates.
-<!-- # TODO: Incluir la condición de invarianza de las coordenadas en los dos espacios (Foto pizarra)? -->
-Second, and most importantly, if the test space is
-where the item parameters naturally lie, they must be represented in this space,
-and not in the latent space.
-Therefore, we may find the direction angles of $`r DISCR_VECTOR`$
-in the test space:
-As $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are both
-orthonormal in $`r REAL_N_SPACE`$, without loss of generality
-we can consider that $`r BASES_EQ`$ (save for a rotation).
-We can then map the latent space directly onto the test space and
-apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
-the vector of direction angles of $`r DISCR_VECTOR`$,
-in the test space $`r TEST_SPACE`$ we get
+Applying Proposition 1,
 
@@
 $`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
-As we will see later, this representation is very convenient, as
-the direction of the item vectors with respect to the coordinate axes
-has a direct, meaningful interpretation.
 
@@
 
-### Correlation-based version
-
-Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
-we have argued that it is still a valid generalization of the IRT parameters.
-However, it may be convenient to find yet another generalization,
-one that accounts for the latent space structure
-while still fulfilling this property.
-To do this, let's assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
-$`r TRANSFORM_MATRIX_AUX_RESULT`$.
-Then, we have that
-
-$$
-  `r CORR_BASED_DERIVATION`.
-$$
-
-That is, defining $`r TRANSFORM_MATRIX`$ as
-$`r TRANSFORM_MATRIX_CORR_BASED_DEF`$ can be interpreted as re-scaling
-the transformed parameter back to its original metric.
-In this case, we get that $`r INNER_PROD_CORR_COND`$.
-Therefore, we can also define a *correlation-based* version of the
-multidimensional parameters as
-
-\begin{equation}
-  \begin{split}
-    `r MDISC_CORR_PARAM_EQ` \\
-    `r MIL_CORR_PARAM_EQ`.
-  \end{split}
-\end{equation}
-
-As we can see, this version also fulfills the scale invariance property,
-as the $`r INNER_PROD_MAT_INV_DIAG_ELEMENT`$ elements are
-the diagonal elements of $`r CORR_MATRIX`$, which are always equal to 1.
-It is important to note though that this transformation does not have
-an also desirable property,
-which we have used to justify the covariance-based version:
-The $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
-$`r TRAIT_COORDS`$ to the orthonormal basis $`r LS_ORTH_BASIS`$
-are not generally uncorrelated, as one would expect
-(save some exceptions, e.g., when all the variances are equal).
-Nevertheless, it is important to note how the $`r MDISC_CORR_PARAM`$ accounts
-for the correlation structure of the latent space
-while still being scale invariant.
-This will be very useful for the purpose of representing the items graphically,
-as we will see later.
-
-## Vector representation
-
-According to @ackerman_creating_1994, the most appropriate way of representing
-multidimensional items is in vector form.
-This allows representing several items altogether and analyzing them
-in terms of their multidimensional parameters [@ackerman_graphical_1996].
-An item vector will be applied at a direction and (signed) distance
-from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter,
-and will have a length equal to its $`r MDISC_SYM`$ parameter.
-Its direction will also be given by the direction component of $`r MIL_PARAM`$,
-so its orientation will always pass through the origin.
-Based on these conditions, we need to compute
-the origin and end coordinates of the vector,
-denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
-
-In an orthogonal basis it is easy to compute the coordinates,
-as they are simply the orthogonal projections onto the corresponding axis.
-In the oblique case it is not so simple though, but we can take advantage of the
-orthornormalization introduced above.
-In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
-$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
-and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
-according to Equation \@ref(eq:discr-vector-std),
-i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
-Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
-the direction cosines in $`r TEST_SPACE_STD_BASIS`$
-are $`r DIR_COS_ITEM_VEC_STD`$.
-Therefore,
-
-\begin{align}
-  `r ORIGIN_ORTH_EQ` \\
-  `r END_ORTH_EQ`.
-\end{align}
-
-Applying Equation \@ref(eq:dir-cos-vec-std),
-for the origin coordinates we have that
-
-\begin{equation}
-  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
-\end{equation}
-
-which, operating, results in
-
-\begin{align}
-  `r ORIGIN_ITEM_COORDS_EQ`    \\
-  `r ORIGIN_TRANSF_SQUARED_EQ` \\
-  `r ORIGIN_INNERPROD_EQ`      \\
-  `r ORIGIN_INNERPROD_DIAG_EQ`
-\end{align}
-
-and, finally,
-
-\begin{equation}
-  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
-\end{equation}
-
-The end coordinates can be computed by a similar procedure, which results in
-
-\begin{equation}
-  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
-\end{equation}
-
-Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
-express the items coordinates in terms of their multidimensional parameters.
-However, it is worth noting that the latent space Gram matrix
-also plays a role in computing these coordinates,
-and will therefore have an effect on the representation.
-Of course,
-from Equations \@ref(eq:origin-orth-discr) and \@ref(eq:end-coords-result)
-it is easy to express the coordinates in terms of the parameters in the original
-model formulation, namely,
-
-\begin{align}
-  `r ORIGIN_ITEM_PARAMS_EQ`                  &(\#eq:origin-from-model-params) \\
-  `r END_ITEM_PARAMS_EQ`.                    &(\#eq:end-from-model-params)
-\end{align}
-
-However, formulating them in terms of the multidimensional parameters allows us
-studying the items in terms of their multidimensional properties.
-In the following, we see how to geometrically interpret
-these multidimensional parameters.
-
-### Geometric properties of the items
-
-We have seen how the two new versions of the parameters fulfill
-the three conditions proposed by @reckase_discriminating_1991
-for valid generalizations of the IRT parameters to the multidimensional case,
-being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
-From a geometrical perspective, the item $`r MDISC_SYM`$ is simply interpreted
-as the length of its corresponding vector.
-Regarding the $`r MIL_PARAM`$ parameter,
-we will consider now two more properties, related to each of its two components:
-the distance to the origin, and the measurement direction.
-
 The (signed) distance to the origin,
-given by the first component of $`r MIL_PARAM`$, has an interesting property:
+given by the second element of $`r MIL_PARAM`$, has an interesting property:
 By the definition of the inner product,
@@
 
-The geometric properties are easier to apprehend with
+The aforementioned properties are easier to apprehend with
 a visual representation of the items.
-To do that, we need to choose a version of the parameters
-that yields their most faithful representation.
-We have already seen that
-the agnostic version disregards the latent space structure.
-Therefore, it will only be useful in the unlikely case that
-we have no information of such structure.
-When we have an estimate of the latent space covariances,
-we may use either the covariance-based or the correlation-based version.
-However, we have already seen how
-the distance to the origin depends on the $`r MDISC_SYM`$.
-Also, in the covariance-based,
-the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
-which, as we have argued before, implies a change of scale.
-In the case of the correlation-based version however,
-$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
-being the coordinates scale-invariant.
-Therefore, for graphical representation purposes,
-the correlation-based version will be preferred.
-
-Plotting on a display usually requires providing
-coordinates in a rectangular Cartesian system.
-Therefore, we will need to compute the equivalent rectangular coordinates
-of our general, possibly non-rectangular coordinates.
-This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
-by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
-the equivalent, orthonormalized coordinates.
-Therefore, we will need to define
-a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
-that allows us to represent the test space structure and the items
-in a rectangular Cartesian system.
-
-The general procedure for plotting an item $`r ITEM_INDEX`$ consists of
-the following steps:
-
-1. Compute $`r MDISC_CORR_PARAM_ITEM`$ and $`r MIL_CORR_PARAM_ITEM`$,
-   the item $`r MDISC_CORR_PARAM`$ and $`r MIL_CORR_PARAM`$ parameters,
-   respectively.
-
-1. Compute the origin coordinates $`r ORIGIN_ITEM_EQ`$, with
-   $`r DISTANCE_CORR_PARAM`$ and $`r DIR_CORR_PARAM`$ the distance and direction
-   component, respectively, of $`r MIL_CORR_PARAM_ITEM`$.
-
-1. Compute the end coordinates $`r END_ITEM_EQ`$.
-
-1. Define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
-   $`r TRANSF_MATRIX_INV_SQ_CORR_EQ`$.
-
-1. Compute the rectangular coordinates
-   $`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$ by pre-multiplying
-   $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively, by
-   the transformation matrix $`r TRANSFORM_MATRIX`$
-   ($`r ORIGIN_TRANSF_EQ`$; $`r END_TRANSF_EQ`$).
-
-Instead of steps 1 to 3,
-one can alternatively compute $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$
-directly from Equations \@ref(eq:origin-from-model-params)
-and \@ref(eq:end-from-model-params).
-Note though that these equations are not independent of $`r CORR_MATRIX`$,
-as it is used in the computation of $`r DISCR_VECTOR_MODULE`$,
-and thus this matrix is necessary in any case to compute the coordinates.
-
-### Graphical representation example
-
-```{r read-chunks-graphical-example}
-read_chunk("src/Graphical_example_paper.R")
-```
-
-```{r libraries}
-#| warning: false
-```
-
-```{r sources}
-```
-
-```{r constants}
-```
-
-```{r graphical-output-conf}
-```
-
-```{r compute-example-items}
-```
+A vector plot [@ackerman_creating_1994] is the most appropriate way of
+representing the items graphically;
+it allows one to plot several items altogether and analyze them
+in terms of their multidimensional parameters [@ackerman_graphical_1996]:
+An item vector will be applied at a direction and (signed) distance
+from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
+and will have a length equal to its $`r MDISC_SYM`$ parameter.
 
@@
 In the first case, the two latent dimensions are independent,
-with $`r TRANSFORM_MATRIX_TRANSP_INV`$ being an identity matrix.
+with $`r TRANSFORM_MATRIX`$ being an identity matrix.
 In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
-To keep the horizontal axis invariant,
-we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
-it only rotates the vertical axis [@harman_modern_1970]:
+In order to keep the horizontal axis invariant,
+we define $`r TRANSFORM_MATRIX`$ such that it only rotates the vertical axis
+[@harman_modern_1970]:
 
@@
 the multivariate latent variable, and (2) to any case of unchanging monotony.
-The latter may seem superfluous, given that several examples in the literature
-already make use of this generalization
-[e.g., the one we have used to illustrate it: @tezza_modelo_2018].
-However, this generalization had never been formalized, up to now.
-Therefore, we highlight here the interest of showing how
-the item psychometric properties are paralleled by their geometrical properties,
-and how this holds for any case of unchanging monotony
-of the item response function.
-Regarding the first generalization,
-it is worth noting that @zhang_theoretical_1999 already proposed a formulation
+It is worth noting that @zhang_theoretical_1999 already proposed a formulation
 of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
@@
 (although constrained to non-decreasing monotony).
-However, they provided no formal proof for this result either.
+However, they provided no formal proof for this result.
 Our results provide this formal derivation, although only for the specific
@@
 
-With that purpose in mind, we have defined two new versions of these parameters:
+With that purpose in mind, we have defined two versions of these parameters:
 One that takes into account the whole covariance structure,
 and another one that only considers the correlations among latent dimensions.
-The covariance-based parameter has the drawback of not fulfilling
+The covariance-based index has the drawback of not fulfilling
 the first of @reckase_discriminating_1991's properties, namely,
-the scale invariance property.
+that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
 However, we have provided a rationale for the violation of this property,
@@
 comparable among different latent spaces.
-More importantly, the covariance-based version
-of the multidimensional parameters is invariant to
-changes of latent-space basis.
 Once again, this is important in applications where
-potentially different bases are involved.
-Given these considerations,
-we propose to adopt the covariance-based version of
-the multidimensional parameters as the most general definition,
-and refer to them simply as
+potentially different latent spaces are involved.
+Given these considerations, we propose to adopt the covariance-based version of
+the multidimensional parameters and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of being scale-invariant,
+has the advantage of preserving the metric of the parameters,
 which is convenient for fidelity when representing the items.
-However, it is worth noting that it has been defined ad-hoc
-to fulfill this specific property.
-Therefore, it lacks other desirable properties of the covariance-based version,
-like being invariant to a latent space scaling transformation,
-or yielding a diagonal latent covariance matrix in the corresponding
-orthogonal latent space basis.
-Therefore, we recommend its use for graphical representation purposes only.
-As for the agnostic version, we recommend it only in
-the (unlikely) case of having no information about the latent space structure,
-or when such structure is considered irrelevant
-(e.g., plotting the discrimination parameters after an oblique rotation).
-
-One may consider that the formal derivations presented here are
-overly complicated,
+Therefore, we recommend using the correlation-based version of the parameters
+for graphical representation purposes only.
+
+One may consider that the formal derivations presented here are overly
+complicated,
 when one might simply transform the space to rectangular Cartesian coordinates
@@
 which may give relevant information if these are actual, substantive traits.
-Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require computing the cosines in
-the original basis afterwards.
-
-On the other hand, establishing the relationship between the original and
-the orthonormalized basis has provided us with an important insight:
+Moreover, strictly following @reckase_difficulty_1985's method in the original,
+oblique space has also provided us with an important insight:
 The distinction between the latent space and the test space.
-Our derivations show that, to properly represent the items,
+Our derivations show that, in order to properly represent the items,
 we need to define a vector space that has
@@
 its inner product matrix.
-This implies that the axes defined by the test space basis,
-and not the latent space basis,
-are a geometrical representation of the covariance structure.
-This result is both surprising and counterintuitive
----Several instances in the MIRT literature suggest that
-"the angle between the axes [of the latent ability space] represents
-the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
-We hope our results help shed a clearer light on
-this often misunderstood topic.
-
 It is worth noting that a similar relationship exists in

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd

--- 2024-09-02 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 currently available technology),
-we limit ourselves to the bidimensional case here.
+we will limit ourselves to the bidimensional case here.
+
 The example in Figure \@ref(fig:item-plot-out) showcases a representation of

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 $`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
-Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.
-
-To make @reckase_difficulty_1985's definition a particular case of ours
+Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$
+and, as $`r DISCR_VECTOR`$ is expressed in this basis,
+its algebraic vector values correspond to its coordinates, i.e.,
+$`r DISCR_VECTOR_COORDS_EQ`$.
+
+As we did with $`r LS_ORTH_BASIS`$ before,
 we can conveniently assume that the inner product in
-$`r TEST_SPACE_STD_BASIS`$ is the dot product,
-as we did with $`r LS_ORTH_BASIS`$.
+$`r TEST_SPACE_STD_BASIS`$ is the dot product.
+Then, as both $`r TEST_SPACE_STD_BASIS`$ and $`r LS_ORTH_BASIS`$ are
+orthonormal in $`r REAL_N_SPACE`$,
+to make @reckase_difficulty_1985's definition a particular case of ours we can,
+without loss of generality  (i.e., appying an arbitrary rotation),
+consider that $`r BASES_EQ`$.
 As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
@@
 in the test space:
-As $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are both
-orthonormal in $`r REAL_N_SPACE`$, without loss of generality
-we can consider that $`r BASES_EQ`$ (save for a rotation).
-We can then map the latent space directly onto the test space and
-apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
+Applying Proposition 1, if $`r ANGLE_VEC_TS_ITEM`$ is
 the vector of direction angles of $`r DISCR_VECTOR`$,
@@
 i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
-Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
-the direction cosines in $`r TEST_SPACE_STD_BASIS`$
-are $`r DIR_COS_ITEM_VEC_STD`$.
-Therefore,
-
-\begin{align}
-  `r ORIGIN_ORTH_EQ` \\
-  `r END_ORTH_EQ`.
-\end{align}
-
-Applying Equation \@ref(eq:dir-cos-vec-std),
-for the origin coordinates we have that
-
-\begin{equation}
-  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
-\end{equation}
-
-which, operating, results in
-
-\begin{align}
-  `r ORIGIN_ITEM_COORDS_EQ`    \\
-  `r ORIGIN_TRANSF_SQUARED_EQ` \\
-  `r ORIGIN_INNERPROD_EQ`      \\
-  `r ORIGIN_INNERPROD_DIAG_EQ`
-\end{align}
-
-and, finally,
-
-\begin{equation}
-  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
-\end{equation}
-
-The end coordinates can be computed by a similar procedure, which results in
+Applying Proposition 2 in the test space, we have that
+
+\begin{equation}
+  `r DIR_COS_ITEM_EQ`.
+\end{equation}
+
+Therefore, the coordinates result in
+
+\begin{equation}
+  `r ORIGIN_ITEM_COORDS_EQ`                          (\#eq:origin-coords-result)
+\end{equation}
+
+and
 

### 2024-09-03 output/paper_draft/Supplementary_material.Rmd

---
title: "Supplementary material"
author: "Daniel Morillo"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  officedown::rdocx_document:
    tables:
      caption:
        pre: "Table S."
      style: APA_Like_style
    mapstyles:
      Normal: First Paragraph
    base_format: bookdown::word_document2
    reference_docx: ../../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
csl:          ../../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| include: false

# Includes:

library(knitr)
library(patchwork)
library(rstudioapi)

ROOT_DIR <- getActiveProject()
DOC_DIR  <- getwd()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  warning  = FALSE,
  cache    = FALSE
)

# Output configuration options:
options(digits = 3)
```

```{r reset-root-wd}
#| cache: false

# Necessary for knitr to find the source file:
opts_knit$set(root.dir = ROOT_DIR)
```

```{r read-chunks-empirical-example}
read_chunk("src/Empirical_example.R")
```

```{r libraries}
```

```{r sources}
```

```{r general-constants}
```

```{r set-cov-matrix}
```

```{r read-items}
```

```{r compute-orthogonal-params}
```

```{r compute-oblique-params}
```

```{r format-covariance-matrix}
```

```{r create-params-table}
```

```{r compose-output-table}
```

Table S.\@ref(tab:empirical-table-Reckase2009-out) is similar to Table 2,
comparing the agnostic version of the multidimensional parameters drawn from
@reckase_multidimensional_2009 with the covariance-based version.
In this case though, we use the almost rank-1 covariance matrix Reckase defines
for representing a *close-to-unidimensional* case, that is,

\begin{equation}
  `r cov_matrix_value_out`,
\end{equation}

which yields the correlation matrix

\begin{equation}
  `r cor_matrix_value_out`.
\end{equation}

As in Table 2, the agnostic versions (columns labelled _`r AGNOSTIC_ABBR`_)
coincide with Reckase's columns A, B, and $\alpha_1$ to $\alpha_3$.
However, the covariance-based versions (columns labelled $`r COV_MATRIX`$)
have much smaller values for the $`r MDISC_ITEM`$ parameters.
This is primarily due to the scaling effect induced by the variances,
which are all smaller than 1.
Even though the discrimination parameters almost sum up to yield
the $`r MDISC_ITEM`$ value, due to the very high alignment of the dimensions,
the scaling effect of the variances has a stronger weight.
The $`r DISTANCE_PARAM`$ parameter values decrease consequently,
as we have seen before.

More interesting is the effect on the direction angles:
We can see how their values are all very close to 0.
This result is as expected,
given that all the three dimensions are highly aligned,
and therefore they almost represent a single dimension.
The items thus all measure in a direction very closely aligned to the dominant
dimension made up by those three, being almost parallel to all the three
dimensions.
Therefore, we see how computing the more appropriate covariance-based version of
the multidimensional parameters can help diagnose correctly
the dimensionality of a test.

```{r set-flextable-wd}
#| cache: false
```

# References

::: {#refs}
:::

\newpage

```{r empirical-table-Reckase2009-out}
#| cache:   false
#| tab.id:  empirical-table-Reckase2009-out
#| tab.cap: |
#|   Agnostic and covariance-based multidimensional item parameters in
#|   Reckase (2009), with an almost-rank-1 covariance matrix.
item_params_output
```

### 2024-09-03 notebooks/Mirt_solution_exploration.Rmd

@@
 output:
+  word_document:
+    df_print:        kable
+    toc:             yes
+    reference_docx:  ../www/Notebook_template.docx
   html_document:
@@
 ```{r setup}
+#| echo: !expr 'knitr::is_html_output()'
 #| message: false
@@
 opts_chunk$set(
-  echo     = TRUE,
+  echo     = is_html_output(),
   message  = FALSE,

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds abstract and keywords (WiP)

@@
     fig_height:      4
+abstract: |
+  Modern measurement theory demands that Multidimensional Item Response Theory
+  provides an accurate representation of covariate spaces with correlated
+  dimensions and/or non-standard variances. As it is also used in applications
+  in the non-cognitive domain, they also need to model negatively-keyeed items
+  in an accurate and interpretable manner. Henceforth we generalize the
+  multidimensional discrimination [@reckase_discriminating_1991] and
+  difficulty [@reckase_difficulty_1985] parameters to account for those two
+  phenomena. We do so by applying Reckase's [-@reckase_difficulty_1985] method
+  in a strictly algebraic sense to the Multidimensional 2-Parameter Logistic
+  Model, represented in a latent space with an arbitrary basis, and
+  then defining such basis to be a concrete geometrical representation of the
+  phenomena we intend to measure. Following @ackerman_creating_1994, we also
+  provide a geometrical representation based on these parameters, and provide
+  applied examples to illustrate the pertinence of these generalized parameters
+  to accurately represent the measurement properties of the items.
+keywords: |
+  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
+  discrimination, difficulty, location, item vector representation
 bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Applies grammarly corrections in abstract

@@
   Modern measurement theory demands that Multidimensional Item Response Theory
-  provides an accurate representation of covariate spaces with correlated
-  dimensions and/or non-standard variances. As it is also used in applications
-  in the non-cognitive domain, they also need to model negatively-keyeed items
-  in an accurate and interpretable manner. Henceforth we generalize the
+  accurately represents covariate spaces with correlated dimensions and/or
+  non-standard variances. As it is also used in applications in the
+  non-cognitive domain, they also need to model negatively-keyeed items
+  accurately and interpretably. Henceforth we generalize the
   multidimensional discrimination [@reckase_discriminating_1991] and
@@
   then defining such basis to be a concrete geometrical representation of the
-  phenomena we intend to measure. Following @ackerman_creating_1994, we also
-  provide a geometrical representation based on these parameters, and provide
-  applied examples to illustrate the pertinence of these generalized parameters
-  to accurately represent the measurement properties of the items.
-keywords: |
-  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
-  discrimination, difficulty, location, item vector representation
+  phenomena we intend to measure. This results in three different versions of
+  the multidimensional parameters: The original one, based on the item
+  parameters solely, one that incorporates the covariance structure of the
+  latent space, and a third one that uses the correlation structure instead.
+  Following @ackerman_creating_1994, we also provide a geometrical
+  representation based on these parameters, and provide applied examples to
+  illustrate the pertinence of these generalized parameters to represent the
+  measurement properties of the items accurately. We recommend using the
+  version that uses the latent covariance structure for describing the
+  properties of the parameters, and the one based on the correlation structure
+  for graphical representation. Finally, we discuss the implications of this
+  generalization for other multidimensional IRT models, and the parallels of our
+  results in common factor model theory.
+keywords: "Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters, item discrimination, item difficulty, item vector representation"
 bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Remove author and date from main text

@@
 title: "A generalized definition of multidimensional item response theory parameters"
-author: "Daniel Morillo"
-date: "`r format(Sys.Date(), '%d/%m/%Y')`"
 output:

### 2024-09-03 output/paper_draft/Supplementary_material.Rmd
Remove author and date from supplementary material

@@
 title: "Supplementary material"
-author: "Daniel Morillo"
-date: "`r format(Sys.Date(), '%d/%m/%Y')`"
 output:

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds keywords inline
Reflows "keywords" yaml field

@@
   results in common factor model theory.
-keywords: "Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters, item discrimination, item difficulty, item vector representation"
+keywords: |
+  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
+  item discrimination, item difficulty, item vector representation
 bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
@@
 ```
+
+**Keywords:** Multidimensional IRT, 2-Parameter Logistic model,
+multidimensional parameters, item discrimination, item difficulty,
+item vector representation
 

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Changes image resolution (min. required = 800)

@@
   dev      = "png",
-  dpi      =   300,
+  dpi      =   900,
   dev.args = list(type = "cairo-png")

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Upates style mapping for abstract output

@@
     mapstyles:
-      Normal: First Paragraph
+      Body Text: ['First Paragraph', 'Abstract']
     base_format: bookdown::word_document2

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects repeated uses of "Therefore"

--- 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 title: "A generalized definition of multidimensional item response theory parameters"
+author: "Daniel Morillo"
+date: "`r format(Sys.Date(), '%d/%m/%Y')`"
 output:
@@
     fig_height:      4
-abstract: |
-  Modern measurement theory demands that Multidimensional Item Response Theory
-  accurately represents covariate spaces with correlated dimensions and/or
-  non-standard variances. As it is also used in applications in the
-  non-cognitive domain, they also need to model negatively-keyeed items
-  accurately and interpretably. Henceforth we generalize the
-  multidimensional discrimination [@reckase_discriminating_1991] and
-  difficulty [@reckase_difficulty_1985] parameters to account for those two
-  phenomena. We do so by applying Reckase's [-@reckase_difficulty_1985] method
-  in a strictly algebraic sense to the Multidimensional 2-Parameter Logistic
-  Model, represented in a latent space with an arbitrary basis, and
-  then defining such basis to be a concrete geometrical representation of the
-  phenomena we intend to measure. This results in three different versions of
-  the multidimensional parameters: The original one, based on the item
-  parameters solely, one that incorporates the covariance structure of the
-  latent space, and a third one that uses the correlation structure instead.
-  Following @ackerman_creating_1994, we also provide a geometrical
-  representation based on these parameters, and provide applied examples to
-  illustrate the pertinence of these generalized parameters to represent the
-  measurement properties of the items accurately. We recommend using the
-  version that uses the latent covariance structure for describing the
-  properties of the parameters, and the one based on the correlation structure
-  for graphical representation. Finally, we discuss the implications of this
-  generalization for other multidimensional IRT models, and the parallels of our
-  results in common factor model theory.
-keywords: |
-  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
-  item discrimination, item difficulty, item vector representation
 bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
@@
 ```
-
-**Keywords:** Multidimensional IRT, 2-Parameter Logistic model,
-multidimensional parameters, item discrimination, item difficulty,
-item vector representation
 
@@
 $4 `r SLOPE_MAX_PARAM`$ [@reckase_discriminating_1991].
-Therefore, we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
+Hence we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
 Equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
@@
 being the coordinates scale-invariant.
-Therefore, for graphical representation purposes,
-the correlation-based version will be preferred.
+For graphical representation purposes,
+the correlation-based version will thus be preferred.
 
@@
 coordinates in a rectangular Cartesian system.
-Therefore, we will need to compute the equivalent rectangular coordinates
+Therefore, we must compute the equivalent rectangular coordinates
 of our general, possibly non-rectangular coordinates.
@@
 the equivalent, orthonormalized coordinates.
-Therefore, we will need to define
-a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
+Hence we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
 that allows us to represent the test space structure and the items
@@
 
-With that purpose in mind, we have defined two new versions of these parameters:
+With that purpose in mind, we have defined two versions of these parameters:
 One that takes into account the whole covariance structure,
@@
 comparable among different latent spaces.
-More importantly, the covariance-based version
-of the multidimensional parameters is invariant to
-changes of latent-space basis.
 Once again, this is important in applications where
-potentially different bases are involved.
-Given these considerations,
-we propose to adopt the covariance-based version of
-the multidimensional parameters as the most general definition,
-and refer to them simply as
+potentially different latent spaces are involved.
+Given these considerations, we propose to adopt the covariance-based version of
+the multidimensional parameters and refer to them simply as
 the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
 The correlation-based version, on the other hand,
-has the advantage of being scale-invariant,
+has the advantage of preserving the metric of the parameters,
 which is convenient for fidelity when representing the items.
-However, as it has been defined ad-hoc to fulfill this specific property,
-it lacks the other desirable ones the covariance-based version has
-(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
-the corresponding orthogonal basis).
-Therefore, we recommend its use for graphical representation purposes only,
-although in many applications both versions will be coincident, fortunately.
-As for the agnostic version, we recommend it only in
-the (unlikely) case of having no information about the latent space structure,
-or when such structure is irrelevant
-(e.g., plotting the discrimination parameters for interpretation purposes).
-
-One may consider that the formal derivations presented here are
-overly complicated,
+Therefore, we recommend using the correlation-based version of the parameters
+for graphical representation purposes only.
+
+One may consider that the formal derivations presented here are overly
+complicated,
 when one might simply transform the space to rectangular Cartesian coordinates

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd

--- 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
   dev      = "png",
-  dpi      =   900,
+  dpi      =   300,
   dev.args = list(type = "cairo-png")

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Drops "methods" and "results" titles

@@
 
-# Method
-
-## Model formulation
+# Model formulation
 
@@
 
-## Linear algebra of non-orthogonal coordinates
+# Linear algebra of non-orthogonal coordinates
 
@@
 
-### Test space definition
+## Test space definition
 
@@
 
-### Direction cosines in the latent space
+## Direction cosines in the latent space
 
@@
 
-## Model recasting into polar coordinates
+# Model recasting into polar coordinates
 
@@
 
-## Point of maximum slope
+# Point of maximum slope
 
@@
 
-# Results
-
-## Relationship between test space and latent space
+# Relationship between test space and latent space
 
@@
 
-## Generalized multidimensional parameters
+# Generalized multidimensional parameters
 
@@
 
-### Agnostic version
+## Agnostic version
 
@@
 
-### Covariance-based version
+## Covariance-based version
 
@@
 
-### Correlation-based version
+## Correlation-based version
 
@@
 
-## Vector representation
+# Vector representation
 
@@
 
-### Geometric properties of the items
+## Geometric properties of the items
 
@@
 
-## Graphical representation
+# Graphical representation
 
@@
 
-### Graphical representation example
+## Graphical representation example
 
@@
 
-## Application to examples from the literature
+# Application to examples from the literature
 
@@
 
-### @reckase_multidimensional_2009
+## @reckase_multidimensional_2009
 
@@
 
-### @tezza_modelo_2018
+## @tezza_modelo_2018
 

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Positions tables better in the main text
The new positioning avoids blanks

@@
   non-standard variances. As it is also used in applications in the
-  non-cognitive domain, they also need to model negatively-keyeed items
+  non-cognitive domain, they also need to model negatively-keyed items
   accurately and interpretably. Henceforth we generalize the
@@
 
-## @reckase_multidimensional_2009
-
 ```{r reset-root-wd}
@@
 ```
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Reckase2009-out}
+#| cache:   false
+#| tab.id:  empirical-table-Reckase2009-out
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
+item_params_output
+```
+
+## @reckase_multidimensional_2009
 
@@
 
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Reckase2009-out}
-#| cache:   false
-#| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
-item_params_output
-```
-
 The latter effect can be better noticed in another case, in which Reckase
@@
 parameters from the agnostic to the covariance-based version
-(see Table \@ref(tab:empirical-table-Tezza2018-out)) ---this happens,
+(see Table \@ref(tab:empirical-table-Tezza2018-out))
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Tezza2018-out}
+#| cache:   false
+#| tab.id:  empirical-table-Tezza2018-out
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Tezza et al. (2018).
+item_params_output
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
+
+---this happens,
 for example, with items 57 and 60. These items also have different-sign
@@
 and Figure \@ref(fig:item-plot-out).
-
-<!---BLOCK_LANDSCAPE_START--->
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Tezza2018-out}
-#| cache:   false
-#| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Tezza et al. (2018).
-item_params_output
-```
-
-<!---BLOCK_LANDSCAPE_STOP--->
 

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
Adds statements and declarations
Change figure output to SVG

@@
   cache    = FALSE,
-  dev      = "png",
-  dpi      =   900,
-  dev.args = list(type = "cairo-png")
+  dev      = "svg"
 )
@@
 
+# Statements and Declarations
+
+## Ethical considerations
+
+Not applicable
+
+## Consent to participate
+
+Not applicable
+
+## Consent for publication
+
+Not applicable
+
+## Declaration of conflicting interest
+
+The author(s) declared no potential conflicts of interest with respect to
+the research, authorship, and/or publication of this article.
+
+## Funding statement
+
+Not applicable
+
+## Data availability statement
+
+Not applicable
+
 \newpage

### 2024-09-03 output/Title_page.Rmd

---
title: "Title page"
---

- Article title
- The full list of authors including names and affiliations of each
    - The listed affiliation should be the institution where the research was
    conducted. If an author has moved to a new institution since completing the
    research, the new affiliation can be included in a note at the end of the
    manuscript – please indicate this on the title page.
    - All persons eligible for authorship must be included at the time of
    submission (please see the authorship section for more information).
    - Contact information for the corresponding author: name, institutional
    address, phone, email
- Statements and Declarations section

### 2024-09-04 output/Title_page.Rmd
Versions title page with title and author data

@@
 title: "Title page"
+output:
+  officedown::rdocx_document:
+    mapstyles:
+      Body Text: ['First Paragraph', 'Abstract']
+    base_format: bookdown::word_document2
+    reference_docx: ../../www/Template.docx
+    number_sections: no
+    keep_md:         no
+    fig_width:       4
+    fig_height:      4
+editor_options: 
+  chunk_output_type: console
 ---
 
-- Article title
-- The full list of authors including names and affiliations of each
-    - The listed affiliation should be the institution where the research was
-    conducted. If an author has moved to a new institution since completing the
-    research, the new affiliation can be included in a note at the end of the
-    manuscript – please indicate this on the title page.
-    - All persons eligible for authorship must be included at the time of
-    submission (please see the authorship section for more information).
-    - Contact information for the corresponding author: name, institutional
-    address, phone, email
-- Statements and Declarations section
+**Article title:**
+A generalized definition of multidimensional item response theory parameters
+
+**Short title:** Generalized MIRT parameters
+
+**List of authors:**
+
+- Daniel Morillo-Cuadrado. Department of Behavioral Science Methodology,
+  Faculty of Psychology, National Remote Education University (UNED), Spain.
+
+- Mario Luzardo-Verde. Department of Mathematics and Applications, La República
+  University, Uruguay.
+
+**Corresponding author:**
+
+Daniel Morillo-Cuadrado \\
+Faculty of Psychology, Office 2.74.
+C. de Juan del Rosal 10, 28040 Madrid
+
+Phone nº: +34 91398 6584
+Email address: dvmorillo@psi.uned.es

### 2024-09-04 output/paper_draft/paper_M2PLM_parameters.Rmd
Moves Statements and Declarations to title page

--- 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
 library(patchwork)
+library(ggplot2)
 library(rstudioapi)
@@
 plot_orth + plot_oblique
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"))
 ```

### 2024-09-04 output/Title_page.Rmd
Moves Statements and Declarations to title page
Corrects formatting issues

@@
 ---
-title: "Title page"
 output:
@@
 
-Daniel Morillo-Cuadrado \\
+Daniel Morillo-Cuadrado \
 Faculty of Psychology, Office 2.74.
@@
 
-Phone nº: +34 91398 6584
+Phone nº: +34 91398 6584 \
 Email address: dvmorillo@psi.uned.es
+
+# Statements and Declarations
+
+## Ethical considerations
+
+Not applicable
+
+## Consent to participate
+
+Not applicable
+
+## Consent for publication
+
+Not applicable
+
+## Declaration of conflicting interest
+
+The author(s) declared no potential conflicts of interest with respect to
+the research, authorship, and/or publication of this article.
+
+## Funding statement
+
+Not applicable
+
+## Data availability statement
+
+Not applicable

### 2024-09-04 output/Title_page.Rmd
Creates title page docx template

@@
     base_format: bookdown::word_document2
-    reference_docx: ../../www/Template.docx
+    reference_docx: ../www/Title_page_template.docx
     number_sections: no

### 2024-09-04 output/Title_page.Rmd
Unifies template use

--- 2024-09-04 output/Title_page.Rmd
@@
     base_format: bookdown::word_document2
-    reference_docx: ../../www/Template.docx
+    reference_docx: ../www/Template.docx
     number_sections: no

### 2024-09-04 output/paper_draft/Supplementary_material.Rmd
Unifies template use

@@
     mapstyles:
-      Normal: First Paragraph
+      Body Text: First Paragraph
     base_format: bookdown::word_document2

### 2024-09-04 output/paper_draft/paper_M2PLM_parameters.Rmd
Corrects minor errata and style issues

@@
 
+\newpage
+
 # Introduction
@@
 interpreting the latent dimensions as substantive constructs is not a goal
-[for for multidimensional linking and equating procedures,
+[for multidimensional linking and equating procedures,
 see @reckase_linking_2009].
@@
 of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
-the item is *inverse* for (i.e., taps the negative pole of)
-latent dimension $`r DIM_INDEX`$ (e.g., a negatively-keyed item).
+the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
 Finally, as responses are not necessarily interpreted as correct or incorrect,

### 2024-09-04 output/paper_draft/Supplementary_material.Rmd
Adds page number to reference

@@
 comparing the agnostic version of the multidimensional parameters drawn from
-@reckase_multidimensional_2009 with the covariance-based version.
+@reckase_multidimensional_2009 [p. 183] with the covariance-based version.
 In this case though, we use the almost rank-1 covariance matrix Reckase defines

### 2024-09-04 output/paper_draft/paper_M2PLM_parameters.Rmd

@@
 [for multidimensional linking and equating procedures,
-see @reckase_linking_2009].
+see Chapter 9 in @reckase_multidimensional_2009].
 As we will argue, the usual definitions of the multidimensional difficulty and

### 2024-09-04 output/paper_draft/Main_text.Rmd

--- 2024-09-04 output/paper_draft/paper_M2PLM_parameters.Rmd
@@
   phenomena we intend to measure. This results in three different versions of
-  the multidimensional parameters: The original one, based on the item
   parameters solely, one that incorporates the covariance structure of the
@@
   results in common factor model theory.
+  the multidimensional parameters: The original one, based on the item
 keywords: |

### 2024-09-04 output/Cover_letter.Rmd
Versions cover letter

---
output:
  officedown::rdocx_document:
    mapstyles:
      Body Text: ['First Paragraph', 'Abstract']
    base_format: bookdown::word_document2
    reference_docx: ../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
editor_options: 
  chunk_output_type: console
---

September 5, 2024.

George A. Marcoulides \
Editor, *Educational and Psychological Measurement* \
University of California, Santa Barbara, USA

Dear Dr. Marcoulides:

Please find enclosed the manuscript entitled
*"A generalized definition of multidimensional item response theory
parameters"*,
which I am submitting for exclusive consideration of publication as an article
in Educational and Psychological Measurement.
Excluding the title page
(submitted separately according to Sage's guidelines),
the manuscript is 24 pages long and includes one figure and three tables.

The authors of this manuscript are myself, Daniel Morillo-Cuadrado,
and my colleague Mario Luzardo-Verde.
You can find all the information about our affiliations in
the title page submitted separately.
We have both agreed to authorship in the indicated order, and
none of us have any interest that might be interpreted
as influencing the research.

Please note that, while all the APA ethical standards were followed in
the conduction of this study, as
the manuscript presents the results of purely formal research,
no ethical considerations were involved,
therefore the lack of ethical approval or consent from actual participants.

Our manuscript presents a generalization of the multidimensional IRT parameters
introduced by Reckase and Reckase & McKinley in their articles
["The difficulty of test items that measure more than one ability" (1985)][1]
and
["The discriminating power of items that measure more than one dimension"
(1991)][2],
respectively.
We believe that our results are still
an important contribution to modern measurement theory,
despite the long time since the publication of these articles.
Our generalization helps clarify the interpretation of
the multidimensional parameters while circumventing several issues
we have found in the literature regarding those multidimensional parameters,
due to implicitly applying them to monotonously non-increasing items and,
more importantly,
to a misconception of how these parameters should be computed and interpreted.
In the course of our research we have also found a crucial result regarding
the interpretation of the latent space and
its distinction from the so-called *test space*,
a duality of spaces almost completely overlooked in the literature,
but critical for understanding Multidimensional IRT and its connection with
the common factor model theory.
We also provide a method for accurately representing the items in vector form
in any test space (regardless of the covariance structure represented),
to help interpret and understand the multidimensional parameters.
We sincerely believe and hope you will find
the results we present in this manuscript interesting and fit for this journal.

[1]: https://doi.org/10.1177/014662168500900409

[2]: https://doi.org/10.1177/014662169101500407

We thank you in advance for your time and effort considering our work.
Please feel free to correspond with me by email (dvmorillo@psi.uned.es).
I will assume all responsibility for keeping
my coauthor and colleague informed of further updates.

Sincerely,

Daniel Morillo, Ph.D.\
Department of Behavioral Science Methodology \
Universidad Nacional de Educación a Distancia \
C. Juan del Rosal 10, Office 2.74 \
28040 Madrid, Spain

### 2024-09-05 output/Title_page.Rmd

@@
 - Daniel Morillo-Cuadrado. Department of Behavioral Science Methodology,
-  Faculty of Psychology, National Remote Education University (UNED), Spain.
+  Faculty of Psychology, Universidad Nacional de Educación a Distancia (UNED),
+  Spain.
 
-- Mario Luzardo-Verde. Department of Mathematics and Applications, La República
-  University, Uruguay.
+- Mario Luzardo-Verde. Department of Mathematics and Applications,
+  Universidad de La República, Uruguay.
 
@@
 Daniel Morillo-Cuadrado \
-Faculty of Psychology, Office 2.74.
+Faculty of Psychology, Office 2.74. \
 C. de Juan del Rosal 10, 28040 Madrid

### 2024-09-05 output/paper_draft/Main_text.Rmd

@@
 
+## @reckase_multidimensional_2009
+
 ```{r set-cov-matrix-Reckase2009}
@@
 ```
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Reckase2009-out}
-#| cache:   false
-#| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
-item_params_output
-```
-
-## @reckase_multidimensional_2009
 
@@
 
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Reckase2009-out}
+#| cache:   false
+#| tab.id:  empirical-table-Reckase2009-out
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
+item_params_output
+```
+
 The latter effect can be better noticed in another case, in which Reckase
@@
 (see Table \@ref(tab:empirical-table-Tezza2018-out))
-<!---BLOCK_LANDSCAPE_START--->
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Tezza2018-out}
-#| cache:   false
-#| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Tezza et al. (2018).
-item_params_output
-```
-
-<!---BLOCK_LANDSCAPE_STOP--->
-
 ---this happens,
@@
 and Figure \@ref(fig:item-plot-out).
+
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Tezza2018-out}
+#| cache:   false
+#| tab.id:  empirical-table-Tezza2018-out
+#| tab.cap: |
+#|   Agnostic and covariance-based multidimensional item parameters in
+#|   Tezza et al. (2018).
+item_params_output
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
 

### 2024-09-05 output/paper_draft/Main_text.Rmd
Updates keywords after agreement with Mario

@@
   phenomena we intend to measure. This results in three different versions of
+  the multidimensional parameters: The original one, based on the item
   parameters solely, one that incorporates the covariance structure of the
@@
   results in common factor model theory.
-  the multidimensional parameters: The original one, based on the item
 keywords: |
-  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
-  item discrimination, item difficulty, item vector representation
+  Multidimensional IRT, 2-Parameter Logistic model,
+  multidimensional item discrimination, multidimensional item difficulty,
+  item vector representation, test space
 bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
@@
 **Keywords:** Multidimensional IRT, 2-Parameter Logistic model,
-multidimensional parameters, item discrimination, item difficulty,
-item vector representation
+multidimensional item discrimination, multidimensional item difficulty,
+item vector representation, test space
 

### 2024-09-05 output/Main_text.Rmd
Reorders main output files
Main output files are moved to the main output folder, instead of a subfolder

--- 2024-09-05 output/paper_draft/Main_text.Rmd
@@
 keywords: |
-  Multidimensional IRT, 2-Parameter Logistic model, multidimensional parameters,
-  item discrimination, item difficulty, item vector representation
-bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
-csl:          ../../www/apa-old-doi-prefix.csl
+  Multidimensional IRT, 2-Parameter Logistic model,
+  multidimensional item discrimination, multidimensional item difficulty,
+  item vector representation, test space
+bibliography: [../www/Multidimensional-parameters-MCLM.bib, ../www/packages.bib]
+csl:          ../www/apa-old-doi-prefix.csl
 editor_options: 
@@
 ```{r write-bib}
-#| cache: true
-
 # automatically create a bib database for R packages

### 2024-09-05 output/Supplementary_material.Rmd
Reorders main output files
Main output files are moved to the main output folder, instead of a subfolder

--- 2024-09-04 output/paper_draft/Supplementary_material.Rmd
@@
     fig_height:      4
-bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
-csl:          ../../www/apa-old-doi-prefix.csl
+bibliography: [../www/Multidimensional-parameters-MCLM.bib, ../www/packages.bib]
+csl:          ../www/apa-old-doi-prefix.csl
 editor_options: 

### 2024-09-05 R/Constants.R
Updates build script
The build script now generates all output files. Also, the files can be rendered with the automating script, not depending on the Rstudio active session which would cause an exception in a non-interactive session

@@ -22,18 +22,23 @@ source("R/Formulae.R", encoding = 'UTF-8')
 ## File system:
 
 ### File names:
-MANUSCRIPT_FILENAME               <- "paper_M2PLM_parameters.Rmd"
+MANUSCRIPT_FILENAME               <- "Main_text.Rmd"
+SUPPLEMENT_FILENAME               <- "Supplementary_material.Rmd"
+TITLEPAGE_FILENAME                <- "Title_page.Rmd"
+COVERLETTER_FILENAME              <- "Cover_letter.Rmd"
 ITEM_PARAMS_RECKASE_2009_FILENAME <- "Table_6.1_Reckase_2009.csv"
 ITEM_PARAMS_TEZZA_2018_FILENAME   <- "Table_5_Tezza_et_al_2018.csv"
 
 ### Project directories:
 DATA_DIR       <- "dat"
 OUTPUT_DIR     <- "output"
-MANUSCRIPT_DIR <- file.path(OUTPUT_DIR, "paper_draft")
 
 ### File paths:
 
-MANUSCRIPT_PATH <- file.path(MANUSCRIPT_DIR, MANUSCRIPT_FILENAME)
+MANUSCRIPT_PATH  <- file.path(OUTPUT_DIR, MANUSCRIPT_FILENAME)
+SUPPLEMENT_PATH  <- file.path(OUTPUT_DIR, SUPPLEMENT_FILENAME)
+TITLEPAGE_PATH   <- file.path(OUTPUT_DIR, TITLEPAGE_FILENAME)
+COVERLETTER_PATH <- file.path(OUTPUT_DIR, COVERLETTER_FILENAME)
 
 ITEM_PARAMS_RECKASE_2009_FILEPATH <- file.path(
   DATA_DIR,

### 2024-09-05 output/Main_text.Rmd
Corrects and updates abstract

--- 2024-09-05 output/paper_draft/Main_text.Rmd
@@
 abstract: |
-  Modern measurement theory demands that Multidimensional Item Response Theory
-  accurately represents covariate spaces with correlated dimensions and/or
-  non-standard variances. As it is also used in applications in the
-  non-cognitive domain, they also need to model negatively-keyed items
-  accurately and interpretably. Henceforth we generalize the
-  multidimensional discrimination [@reckase_discriminating_1991] and
-  difficulty [@reckase_difficulty_1985] parameters to account for those two
-  phenomena. We do so by applying Reckase's [-@reckase_difficulty_1985] method
+  Modern measurement theory demands that
+  Multidimensional Item Response Theory models accurately represent
+  covariance spaces with correlated dimensions and/or non-standard variances.
+  As they are applied in the non-cognitive domain as well,
+  they also need to model negatively-keyed items accurately and interpretably.
+  Henceforth we generalize the multidimensional discrimination
+  [@reckase_discriminating_1991] and difficulty [@reckase_difficulty_1985]
+  parameters to account for those two phenomena.
+  We do so by applying Reckase's [-@reckase_difficulty_1985] method
   in a strictly algebraic sense to the Multidimensional 2-Parameter Logistic
@@
   latent space, and a third one that uses the correlation structure instead.
+  Importantly, we find there is a test space where
+  the items should be properly represented, which must be differentiated from
+  the latent space itself.
   Following @ackerman_creating_1994, we also provide a geometrical
-  representation based on these parameters, and provide applied examples to
-  illustrate the pertinence of these generalized parameters to represent the
-  measurement properties of the items accurately. We recommend using the
-  version that uses the latent covariance structure for describing the
-  properties of the parameters, and the one based on the correlation structure
-  for graphical representation. Finally, we discuss the implications of this
-  generalization for other multidimensional IRT models, and the parallels of our
-  results in common factor model theory.
+  representation based on these parameters in the test space.
+  We apply our results to examples from the literature
+  to illustrate the pertinence of these generalized parameters to get an
+  accurate representation of the measurement properties of the items.
+  We recommend using the version that uses the latent covariance structure
+  for describing the properties of the parameters,
+  and the one based on the correlation structure for graphical representation.
+  Finally, we discuss the implications of this generalization for
+  other multidimensional IRT models
+  and the parallels of our results in common factor model theory.
 keywords: |
@@
   item vector representation, test space
-bibliography: [../../www/Multidimensional-parameters-MCLM.bib, ../../www/packages.bib]
-csl:          ../../www/apa-old-doi-prefix.csl
+bibliography: [../www/Multidimensional-parameters-MCLM.bib, ../www/packages.bib]
+csl:          ../www/apa-old-doi-prefix.csl
 editor_options: 
@@
 
+library(here)
 library(knitr)
@@
 
-ROOT_DIR <- getActiveProject()
-DOC_DIR  <- getwd()
+ROOT_DIR <- here()
+DOC_DIR  <- file.path(ROOT_DIR, "output")
 

### 2024-09-06 output/Main_text.Rmd
Avoids repetition of "aims at"

@@
 a structure matrix as simple as possible. MIRT literature, on the other hand,
-aims at modeling responses to certain stimuli and estimating person scores,
+intends to model responses to certain stimuli and estimate person scores,
 being the model interpretation secondary in relevance.<!--# Reference needed? -->
@@
 to make @reckase_difficulty_1985's definition a particular case of ours we can,
-without loss of generality  (i.e., appying an arbitrary rotation),
+without loss of generality  (i.e., applying an arbitrary rotation),
 consider that $`r BASES_EQ`$.

### 2024-09-06 output/Main_text.Rmd
Corrects use of parentheses following reference

--- 2024-09-05 output/Main_text.Rmd
@@
 Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
-non-negative real numbers in Reckase
-[-@reckase_difficulty_1985, a decrease in the probability of a correct response
+non-negative real numbers in @reckase_difficulty_1985
+---a decrease in the probability of a correct response
 with an increase in ability makes little sense].

### 2024-09-06 output/Main_text.Rmd
Adds libraries to prevent {renv} inconsistencies

--- 2024-09-05 output/Main_text.Rmd
@@
 library(rstudioapi)
+library(rsvg)    # Used by {knitr} and {ggplot2} to render the figure to SVG;
+library(svglite) #   added to prevent {renv} from reporting inconsistent state.
 

### 2024-09-06 output/Main_text.Rmd
Deletes pending in-text comments

@@
 intends to model responses to certain stimuli and estimate person scores,
-being the model interpretation secondary in relevance.<!--# Reference needed? -->
+being the model interpretation secondary in relevance.
 Nevertheless, the confluence of the two theoretical approaches into a common
@@
 no matter in which space we represent the coordinates.
-<!-- # TODO: Incluir la condición de invarianza de las coordenadas en los dos espacios (Foto pizarra)? -->
 Second, and most importantly, if the test space is

### 2024-09-06 output/Main_text.Rmd
Reverts rendering Figure 1 from SVG to PNG

--- 2024-09-06 output/Main_text.Rmd
@@
 library(rstudioapi)
-library(rsvg)    # Used by {knitr} and {ggplot2} to render the figure to SVG;
-library(svglite) #   added to prevent {renv} from reporting inconsistent state.
+library(svglite) # Used by {knitr} and {ggplot2} to render the figure to SVG;
+                 #   added to prevent {renv} from reporting inconsistent state.
 
@@
   cache    = FALSE,
-  dev      = "svg"
+  dev      = "png",
+  dpi      = 1200L,
+  dev.args = list(type = "cairo-png")
 )

### 2024-09-06 output/Main_text.Rmd
Corrects use of saxon genitive in references

@@
 
-Following @reckase_difficulty_1985’s procedure,
+Following Reckase's [-@reckase_difficulty_1985] procedure,
 we must find in the first place
@@
 with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
-We want @reckase_difficulty_1985's definition of
+We want Reckase's [-@reckase_difficulty_1985] definition of
 the multidimensional parameters in rectangular coordinates to be
@@
 orthonormal in $`r REAL_N_SPACE`$,
-to make @reckase_difficulty_1985's definition a particular case of ours we can,
+to make Reckase's [-@reckase_difficulty_1985] definition
+a particular case of ours we can,
 without loss of generality  (i.e., applying an arbitrary rotation),
@@
 involved in the expression of the model in polar coordinates.
-Importantly, this shows that @reckase_difficulty_1985's Equation 3 (p. 403)
+Importantly, this shows that
+Reckase's [-@reckase_difficulty_1985] Equation 3 (p. 403)
 already assumes the orthonormality of the latent space, as we stated before.
@@
 We are thus considering the direction cosines with respect to the orthonormal
-axes, so we can just apply @reckase_difficulty_1985's results, obtaining
+axes, so we can just apply Reckase's [-@reckase_difficulty_1985] results,
+obtaining
 
@@
 then compute the parameters in the transformed space using
-@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
+Reckase's [-@reckase_difficulty_1985] and
+Reckase and McKinley's [-@reckase_discriminating_1991] procedure.
 After all, interpreting the axes of this space as substantive latent traits

### 2024-09-06 output/Main_text.Rmd
Corrects minor errors in rendering of references

@@
 are coincident with their respective original sources
-[see Table 6.1 in @reckase_multidimensional_2009, p. 153, and
+[see Table 6.1 in @reckase_multidimensional_2009, p. 153; and
 Table 5 in @tezza_modelo_2018, p. 926].
 In the first case, the direction angles are also coincident with the values
-provided in the original [@tezza_modelo_2018 do not provide these results].
+provided in the original [@tezza_modelo_2018, do not provide these results].
 In order to facilitate the comparison,
@@
 and another one that only considers the correlations among latent dimensions.
-The covariance-based parameter has the drawback of not fulfilling
-the first of @reckase_discriminating_1991's properties, namely,
-the scale invariance property.
+The covariance-based parameters have the drawback of not fulfilling
+the first of Reckase & McKinley's [-@reckase_discriminating_1991] properties,
+namely, the scale invariance property.
 However, we have provided a rationale for the violation of this property,
@@
 Reckase's [-@reckase_difficulty_1985] and
-Reckase and McKinley's [-@reckase_discriminating_1991] procedure.
+Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
 After all, interpreting the axes of this space as substantive latent traits

### 2024-09-06 output/Main_text.Rmd
Corrects transformation matrix
In the graphical representation algorithm, the latent space transformation matrix is incorrectly referenced, instead of its transpose inverse (i.e. the transformation matrix in the test space)

@@
    $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively, by
-   the transformation matrix $`r TRANSFORM_MATRIX`$
+   the transformation matrix $`r TRANSFORM_MATRIX_TRANSP_INV`$
    ($`r ORIGIN_TRANSF_EQ`$; $`r END_TRANSF_EQ`$).
@@
 Although proposals such as the graded-scale [@samejima_estimation_1968] and
-the nominal response [@bock_estimating_1972] model are closely related to
+the nominal response [@bock_estimating_1972] models are closely related to
 the 2-parameter logistic model, it would be hasty to assume that these

### 2024-09-13 output/Main_text.Rmd
Reduces abstract to 199 words
Maximum permitted by Psychometrika is 200 words (https://www.cambridge.org/core/journals/psychometrika/information/author-instructions/preparing-your-materials)

@@
 abstract: |
-  Modern measurement theory demands that
-  Multidimensional Item Response Theory models accurately represent
-  covariance spaces with correlated dimensions and/or non-standard variances.
-  As they are applied in the non-cognitive domain as well,
-  they also need to model negatively-keyed items accurately and interpretably.
-  Henceforth we generalize the multidimensional discrimination
-  [@reckase_discriminating_1991] and difficulty [@reckase_difficulty_1985]
-  parameters to account for those two phenomena.
-  We do so by applying Reckase's [-@reckase_difficulty_1985] method
-  in a strictly algebraic sense to the Multidimensional 2-Parameter Logistic
-  Model, represented in a latent space with an arbitrary basis, and
-  then defining such basis to be a concrete geometrical representation of the
-  phenomena we intend to measure. This results in three different versions of
-  the multidimensional parameters: The original one, based on the item
-  parameters solely, one that incorporates the covariance structure of the
-  latent space, and a third one that uses the correlation structure instead.
-  Importantly, we find there is a test space where
-  the items should be properly represented, which must be differentiated from
-  the latent space itself.
-  Following @ackerman_creating_1994, we also provide a geometrical
-  representation based on these parameters in the test space.
-  We apply our results to examples from the literature
-  to illustrate the pertinence of these generalized parameters to get an
-  accurate representation of the measurement properties of the items.
-  We recommend using the version that uses the latent covariance structure
-  for describing the properties of the parameters,
-  and the one based on the correlation structure for graphical representation.
-  Finally, we discuss the implications of this generalization for
-  other multidimensional IRT models
-  and the parallels of our results in common factor model theory.
+  In this paper, we generalize the multidimensional discrimination (MDISC) and
+  difficulty (MID) parameters in the Multidimensional 2-Parameter Logistic model
+  to account for non-identity latent covariances and negatively-keyed items.
+  We apply Reckase's maximum discrimination point method to define them in an
+  algebraic sense in an arbitrary basis, and then defining such basis to be a
+  geometrical representation of the measured construct.
+  This results in three different versions of the parameters: The original one,
+  based on the item parameters solely, one that incorporates the covariance
+  structure of the latent space, and one that uses the correlation structure
+  instead. Importantly, we find that the items should be properly represented in
+  a test space, distinct from the latent space.
+  We also provide a procedure for the geometrical representation of the items in
+  the test space, and apply our results to examples from the literature to get a
+  more accurate representation of the measurement properties of the items.
+  We recommend using the covariance structure version for describing the
+  properties of the parameters, and the correlation structure version for
+  graphical representation.
+  Finally, we discuss the implications of this generalization for other
+  multidimensional IRT models and the parallels of our results in common factor
+  model theory.
 keywords: |

### 2024-09-13 output/Main_text.Rmd
Reduces keywords to 5
Maximum permitted by Psychometrika is 5 keywords (https://www.cambridge.org/core/journals/psychometrika/information/author-instructions/preparing-your-materials)

@@
 keywords: |
-  Multidimensional IRT, 2-Parameter Logistic model,
+  Multidimensional 2-Parameter Logistic model,
   multidimensional item discrimination, multidimensional item difficulty,

### 2024-09-16 output/Main_text.Rmd
Update styles to APA 7 format

@@
 ---
-title: "A generalized definition of multidimensional item response theory parameters"
+title: |
+  A Generalized Definition of Multidimensional Item Response Theory Parameters
 output:
@@
       style: APA_Like_style
+      caption:
+        style: Image Caption
+        sep: ' '
+    plots:
+      caption:
+        sep: ' '
     mapstyles:
@@
                  #   added to prevent {renv} from reporting inconsistent state.
+library(officer)
 
@@
   dpi      = 1200L,
-  dev.args = list(type = "cairo-png")
+  dev.args = list(type = "cairo-png"),
+  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE),
+  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
 )
@@
 \newpage
-
-# Introduction
 
@@
 
-# Model formulation
+# Model Formulation
 
@@
 
-# Linear algebra of non-orthogonal coordinates
+# Linear Algebra of Non-Orthogonal Coordinates
 
@@
 
-## Test space definition
+## Test Space Definition
 
@@
 
-## Direction cosines in the latent space
+## Direction Cosines in the Latent Space
 
@@
 
-# Model recasting into polar coordinates
+# Model Recasting Into Polar Coordinates
 
@@
 
-# Point of maximum slope
+# Point of Maximum Slope
 
@@
 
-# Relationship between test space and latent space
+# Relationship Between Test Space and Latent Space
 
@@
 
-# Generalized multidimensional parameters
+# Generalized Multidimensional Parameters
 
@@
 
-## Agnostic version
+## Agnostic Version
 
@@
 
-## Covariance-based version
+## Covariance-Based Version
 
@@
 
-## Correlation-based version
+## Correlation-Based Version
 
@@
 
-# Vector representation
+# Vector Representation
 
@@
 
-## Geometric properties of the items
+## Geometric Properties of the Items
 
@@
 
-# Graphical representation
+# Graphical Representation
 
@@
 
-## Graphical representation example
+## Graphical Representation Example
 
@@
 #| tab_id:  example-items-table-out
-#| tab.cap: Item parameters for the graphical example
+#| tab.cap: Item Parameters for the Graphical Representation Example
 item_params_output
@@
 #| cache:          false
-#| fig.cap:   Item plots in an orthogonal (left) and oblique (right) test space.
+#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
+#| fig.topcaption: true
 #| fig.height:     3.1
@@
 
-# Application to examples from the literature
+# Application to Examples From the Literature
 
@@
 #| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Reckase (2009), with a (rank-complete) three-dimensional covariance matrix.
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Reckase (2009), With a (Rank-Complete) Three-Dimensional Covariance Matrix
 item_params_output
@@
 #| tab.cap: |
-#|   Agnostic and covariance-based multidimensional item parameters in
-#|   Tezza et al. (2018).
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Tezza et al. (2018)
 item_params_output

### 2024-09-16 output/Main_text.Rmd
Put abstract and title in APA 7 style

@@
 ---
-title: |
-  A Generalized Definition of Multidimensional Item Response Theory Parameters
+title: Abstract
 output:
@@
     fig_height:      4
-abstract: |
+keywords: |
+  Multidimensional 2-Parameter Logistic model,
+  multidimensional item discrimination, multidimensional item difficulty,
+  item vector representation, test space
+bibliography: [../www/Multidimensional-parameters-MCLM.bib, ../www/packages.bib]
+csl:          ../www/apa-old-doi-prefix.csl
+editor_options: 
+  chunk_output_type: console
+---
+
+```{r setup}
+#| include: false
+
+# Includes:
+
+library(here)
+library(knitr)
+library(patchwork)
+library(ggplot2)
+library(rstudioapi)
+library(svglite) # Used by {knitr} and {ggplot2} to render the figure to SVG;
+                 #   added to prevent {renv} from reporting inconsistent state.
+library(officer)
+
+ROOT_DIR <- here()
+DOC_DIR  <- file.path(ROOT_DIR, "output")
+
+opts_knit$set(root.dir = ROOT_DIR)
+opts_chunk$set(
+  echo     = FALSE,
+  message  = FALSE,
+  cache    = FALSE,
+  dev      = "png",
+  dpi      = 1200L,
+  dev.args = list(type = "cairo-png"),
+  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE),
+  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
+)
+
+# Output configuration options:
+options(digits = 3)
+```
+
+```{r write-bib}
+# automatically create a bib database for R packages
+write_bib(
+  c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
+  'www/packages.bib'
+)
+```
+
+```{r sources}
+#| cache:   false
+#| include: false
+source("R/Formulae.R",  encoding = 'UTF-8')
+source("R/Constants.R", encoding = 'UTF-8')
+```
+
 In this paper, we generalize the multidimensional discrimination (MDISC) and
@@
 model theory.
-keywords: |
-  Multidimensional 2-Parameter Logistic model,
-  multidimensional item discrimination, multidimensional item difficulty,
-  item vector representation, test space
-bibliography: [../www/Multidimensional-parameters-MCLM.bib, ../www/packages.bib]
-csl:          ../www/apa-old-doi-prefix.csl
-editor_options: 
-  chunk_output_type: console
----
-
-```{r setup}
-#| include: false
-
-# Includes:
-
-library(here)
-library(knitr)
-library(patchwork)
-library(ggplot2)
-library(rstudioapi)
-library(svglite) # Used by {knitr} and {ggplot2} to render the figure to SVG;
-                 #   added to prevent {renv} from reporting inconsistent state.
-library(officer)
-
-ROOT_DIR <- here()
-DOC_DIR  <- file.path(ROOT_DIR, "output")
-
-opts_knit$set(root.dir = ROOT_DIR)
-opts_chunk$set(
-  echo     = FALSE,
-  message  = FALSE,
-  cache    = FALSE,
-  dev      = "png",
-  dpi      = 1200L,
-  dev.args = list(type = "cairo-png"),
-  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE),
-  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
-)
-
-# Output configuration options:
-options(digits = 3)
-```
-
-```{r write-bib}
-# automatically create a bib database for R packages
-write_bib(
-  c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
-  'www/packages.bib'
-)
-```
-
-```{r sources}
-#| cache:   false
-#| include: false
-source("R/Formulae.R",  encoding = 'UTF-8')
-source("R/Constants.R", encoding = 'UTF-8')
-```
 
@@
 \newpage
+
+# A Generalized Definition of Multidimensional Item Response Theory Parameters
 

### 2024-09-16 src/Empirical_example.R
Update table footers to approximate APA 7 style

@@ -254,7 +254,7 @@ item_params_output <- item_params                          |>
   style(
     pr_p = fp_par(
       padding.bottom = 1,
-      padding.top    = 2,
+      padding.top    = 1,
       padding.left   = 3,
       padding.right  = 3
     ),
@@ -264,15 +264,23 @@ item_params_output <- item_params                          |>
     j    = separator_index,
     pr_p = fp_par(
       padding.bottom = 1,
-      padding.top    = 2,
+      padding.top    = 1,
       padding.left   = 1,
       padding.right  = 1
     ),
     part = "all"
   )                                                        |>
+  style(
+    pr_p = fp_par(
+      line_spacing   = 1.25,
+      padding.top    = 2,
+      padding.bottom = 0
+    ),
+    part = "footer",
+  )                                                        |>
   valign(valign = "bottom", part = "header")               |>
   align(align = "center",   part = "header")               |>
   align(align = "right",    part = "body")                 |>
   fontsize(size = 12,       part = "all")                  |>
-  fontsize(size = 10,       part = "body")                 |>
+  fontsize(size =  9,       part = "body")                 |>
   set_table_properties(layout = "autofit")

### 2024-09-16 output/Supplementary_material.Rmd
Updates build script
The build script now generates all output files. Also, the files can be rendered with the automating script, not depending on the Rstudio active session which would cause an exception in a non-interactive session
Update style in Supplementary Material table

@@
 
+library(here)
 library(knitr)
 library(patchwork)
-library(rstudioapi)
+library(officer)
 
-ROOT_DIR <- getActiveProject()
-DOC_DIR  <- getwd()
+ROOT_DIR <- here()
 
@@
   warning  = FALSE,
-  cache    = FALSE
+  cache    = FALSE,
+  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
 )
@@
 
-```{r set-flextable-wd}
-#| cache: false
-```
-
 # References

### 2024-09-16 output/Main_text.Rmd
Update table footers to approximate APA 7 style
Update position of tables in text
Tables need to be re-positioned to be in a single page, avoiding long blanks before
Put "keywords" title in italics as per APA style

@@
 
-**Keywords:** Multidimensional IRT, 2-Parameter Logistic model,
+*Keywords:* Multidimensional IRT, 2-Parameter Logistic model,
 multidimensional item discrimination, multidimensional item difficulty,
@@
 The sample items have the same `r MODEL_ACRONYM` parameters
-in both latent spaces.
+in both test spaces.
 However, their multidimensional parameters differ due to the correlation among
@@
 <br>
-
-```{r compose-oblique-plot}
-```
-
-```{r compose-orthogonal-plot}
-```
-
-```{r item-plot-out}
-#| cache:          false
-#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
-#| fig.topcaption: true
-#| fig.height:     3.1
-#| fig.width:      6.65
-plot_orth + plot_oblique
-
-ggsave(file.path(DOC_DIR, "Figure_1.svg"))
-```
 
@@
 so its $`r MDISC_CORR_PARAM_ITEM`$ is unaffected by the correlation.
+
+```{r compose-oblique-plot}
+```
+
+```{r compose-orthogonal-plot}
+```
+
+```{r item-plot-out}
+#| cache:          false
+#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
+#| fig.topcaption: true
+#| fig.height:     3.1
+#| fig.width:      6.65
+plot_orth + plot_oblique
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"))
+```
 
@@
 
+\newpage
+
 ```{r empirical-table-Reckase2009-out}
@@
 
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Tezza2018-out}
+#| cache:   false
+#| tab.id:  empirical-table-Tezza2018-out
+#| tab.cap: |
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Tezza et al. (2018)
+item_params_output
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
+
 On the other hand, in cases such as items 12 or 59, which have
@@
 and Figure \@ref(fig:item-plot-out).
-
-<!---BLOCK_LANDSCAPE_START--->
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Tezza2018-out}
-#| cache:   false
-#| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: |
-#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
-#|   Tezza et al. (2018)
-item_params_output
-```
-
-<!---BLOCK_LANDSCAPE_STOP--->
 

### 2024-09-16 output/ESM_1.Rmd
Rename Supplementary Material
Rename file and reference to it according to Psychometrika's submission instructions

--- 2024-09-16 output/Supplementary_material.Rmd
@@
 ---
-title: "Supplementary material"
+title: "Online Resource 1"
 output:

### 2024-09-16 output/Title_page.Rmd
Update title page
Update styles in title page to avoid indentation

@@
     mapstyles:
-      Body Text: ['First Paragraph', 'Abstract']
+      First Paragraph: Body Text
     base_format: bookdown::word_document2
@@
 
-# Statements and Declarations
+# Financial Support
 
-## Ethical considerations
+This research received no specific grant funding form any funding agency,
+commercial or not-for-profit sectors.
 
-Not applicable
+# Competing Interests Statement
 
-## Consent to participate
-
-Not applicable
-
-## Consent for publication
-
-Not applicable
-
-## Declaration of conflicting interest
-
-The author(s) declared no potential conflicts of interest with respect to
-the research, authorship, and/or publication of this article.
-
-## Funding statement
-
-Not applicable
-
-## Data availability statement
-
-Not applicable
+Competing interests: The author(s) declare none.

### 2024-09-16 output/Cover_letter.Rmd
Clean up unused knitting parameters

@@
     keep_md:         no
-    fig_width:       4
-    fig_height:      4
 editor_options: 

### 2024-09-16 output/Main_text.Rmd
Rename Supplementary Material
Rename file and reference to it according to Psychometrika's submission instructions
Clean up unused knitting parameters

@@
     keep_md:         no
-    fig_width:       4
-    fig_height:      4
 keywords: |
@@
 dimension they measure. The reader interested can check out the parameters
-computed in that case in the table in the Supplementary Material.
+computed in that case in the table in Online Resource 1.
 

### 2024-09-16 output/ESM_1.Rmd
Update styles in supplementary material

@@
     tables:
+      style: APA_Like_style
       caption:
+        style: Image Caption
         pre: "Table S."
-      style: APA_Like_style
+        sep: ' '
     mapstyles:

### 2024-09-16 output/Main_text.Rmd
Add extra padding around table 1 and figure 1
Update printed keywords to attribute values

@@
 
-*Keywords:* Multidimensional IRT, 2-Parameter Logistic model,
+*Keywords:* Multidimensional 2-Parameter Logistic model,
 multidimensional item discrimination, multidimensional item difficulty,
@@
 
+<br>
+
 ```{r example-items-table-out}
@@
 ```
+
+<br>
 

### 2024-10-21 output/seminar-slideshow.qmd
Version slideshow template

---
title:     |
  Parámetros multidimensionales en Teoría de Respuesta al Ítem:
  ¡Algo estamos haciendo mal!
author:    |
  <table>
    <tr>
      <td>Daniel Morillo, Ph.D.</td>
    </tr>
    <tr>
      <td>
        <center>
          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/DaniMori/)&emsp;
          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0003-3021-3878)
        </center>
      </td>
    </tr>
  </table>
institute: |
  <table>
    <tr>
      <td>
        <center>
          [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="120"}](https://www.uned.es/universidad/facultades/psicologia.html)
        </center>
      </td>
    </tr>
  </table>
date:        "2024-10-21"
date-meta:   "2024-11-27"
date-format: long
editor:      source
lang:        es
knitr:
  opts_knit: 
    root_dir: here::here()
format:
  revealjs:
    auto-stretch:            true
    code-annotations:        hover
    fig-cap-location:        bottom
    incremental:             false
    link-external-newwindow: true
    self-contained:          true
    slide-number:            false
    theme:                   ../www/slideshow-assets/extra-styles.scss
    transition:              none
    view-distance:           3
    template-partials:
      - ../www/slideshow-assets/title-slide.html
---

### 2024-10-22 output/seminar-slideshow.qmd
Add comparison table between FA and MIRT

@@
     root_dir: here::here()
+  opts_chunk: 
+    results:  asis
 format:
@@
     code-annotations:        hover
+    df-print:                paged
     fig-cap-location:        bottom
     incremental:             false
+    keep-md:                 true
     link-external-newwindow: true
@@
 ---
+
+```{r setup}
+library(tibble)
+library(dplyr, warn.conflicts = FALSE)
+library(knitr)
+```
+
+<!-- Slides "Diferencias AF - TRI" -->
+
+```{r irt-fa-table}
+af_tri_differences <- tribble(
+  ~AF,                            ~TRI,
+  "Relaciones entre variables",   "Comparación entre participantes",
+  "Cuestionarios",                "Pruebas (rendimiento)",
+  "Explicación",                  "Medición",
+  "Puntuación compuesta",         "Puntuación latente",
+  "Foco en estructura",           "Foco en diferencias",
+  "Estructura simple",            "Ejes ortogonales"
+)
+
+section_title <- "## Diferencias AF - TRI"
+
+for (slide in af_tri_differences |> nrow() |> seq_len()) {
+
+  cat( "\n", section_title, sep = '')
+
+  af_tri_differences |> slice_head(n = slide) |> kable() |> print()
+}
+```
+
+::: notes
+
+:::

### 2024-10-22 output/seminar-slideshow.qmd
Version introduction template slides

@@
   "Puntuación compuesta",         "Puntuación latente",
-  "Foco en estructura",           "Foco en diferencias",
-  "Estructura simple",            "Ejes ortogonales"
+  "Estructura simple",            "Diferencias individuales"
 )
@@
 ::: notes
+:::
 
+## AF: Estructura simple
+
+<!--# TODO: Ejemplos estructura simple vs. compleja: Path diagram vs. coordenadas -->
+
+::: notes
+Para darle más sentido al modelo factorial es común buscar la "estructura simple".
+Es decir, asumimos que la varianza de cada ítem está explicada solamente por un factor latente.
+Los pesos factoriales en los demás factores son nulos.
+Si representamos los pesos factoriales como coordenadas, el ítem estará alineado con el eje correspondiente al factor en el que pesa.
 :::
+
+## AF: Estructura simple
+
+<!--# TODO: Ejemplo mostrando solución factorial sin rotar y ejes tras rotación oblicua? -->
+
+::: notes
+Para obtener esa estructura simple, muchas veces necesitamos usar una rotación oblicua.
+:::
+
+## AF: Estructura compleja
+
+<!--# TODO: Ejemplo con pesos cruzados: Path diagram vs. coordenadas -->
+
+::: notes
+Sin embargo, en ocasiones no es posible encontrar la estructura simple.
+Es cuando aparecen lo que llamamos "pesos cruzados".
+Esto da lugar a una "estructura compleja", que en realidad es más común que la estructura simple, en general.
+:::
+
+## AF: Estructura compleja
+
+<!--# TODO: Ejemplo con factores de método: Path diagram vs. coordenadas -->
+
+::: notes
+Por ejemplo, si tenemos ítems directos e inversos, podemos modelar lo que llamamos "factores de método": Una variable latente para cada factor sustantivo, y una para cada uno de los tipos de ítem.
+:::
+
+## Modelo factorial clásico
+
+<!--# TODO: Fórmula del modelo factorial -->
+
+<!--# TODO: Fórmula matricial para una variable observable -->
+
+::: notes
+Un modelo factorial no es más que una serie de regresiones lineales (una por cada variable observable) con variables latentes.
+Las variables "respuesta" o "criterio" es la variable observable, y su varianza está explicada o "predicha" por las variables latentes, que son los factores y los términos de error de cada ítem.
+
+Los "coeficientes de regresión" los llamamos "pesos factoriales" y el término error, "unicidad" o "varianza única". Esto se puede expresar en "forma matricial" de esta manera.
+
+Tenemos costumbre de ver el análisis factorial en el contexto de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
+Sin embargo, se aplica también (e idealmente se debería aplicar) a variables discretas ordinales.
+En este caso vamos a concretarlo en variables dicotómicas (p.ej.
+ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
+Esto es una simplificación que rara vez se utiliza en cuestionarios no cognitivos, pero es interesante estudiarlo porque otros modelos pueden considerarse extensiones de este modelo.
+:::
+
+## Análisis factorial de ítems
+
+<!--# TODO: Fórmula del modelo factorial con función de enlace probit -->
+
+<!--# TODO: Función de enlace "normal acumulada" -->
+
+<!--# TODO: Curva Gaussiana y curva normal acumulada -->
+
+::: notes
+El análisis factorial de ítems (dicotómicos) lo que hace es aplicar ese "modelo lineal general" a variables dicotómicas: Asume una "variable latente" lineal, con la misma fórmula anterior, y aplica una "función de umbral" en 0 (la respuesta es correcta o afirmativa, codificada como 1, si la variable latente es mayor que 0, o incorrecta o negativa, codificada como 0, si es menor).
+Eso da una "probabilidad" de responder correcta o "afirmativamente" a un ítem.
+:::
+
+## Teoría de respuesta al ítem
+
+<!--# TODO: Fórmula de modelo 2PL -->
+
+<!--# TODO: Gráfica de discriminación alta/baja, dificultad alta/baja -->
+
+::: notes
+La TRI, por su lado, asume que hay cada persona tiene una puntuación latente, y que su respuesta depende de esa puntuación, con un parámetro de discriminación, que determina cómo de bien o mal mide el ítem, y uno de dificultad (aunque aquí prefiero llamarlo "de posición"), que determina lo fácil o difícil que será dar una respuesta positiva (afirmativa, correcta, codificada como 1).
+
+Como se puede ver, esto también se trata de un "modelo lineal general".
+En este caso, en lugar de una función probit (normal acumulada), la función de enlace es la función "logit".
+Es decir, la respuesta a cada ítem se modela como una regresión logistica donde los predictores son las variables latentes.
+:::
+
+## Teoría de respuesta al ítem multidimensional
+
+<!--# TODO: Fórmula de modelo M2PL -->
+
+<!--# TODO: Gráfica tridimensional de modelo compensatorio -->
+
+::: notes
+Una de las extensiones multidimensionales (más antiguas, y probablemente más obvias) de este modelo consiste simplemente en asumir que la respuesta se modela como una "regresión logística multivariada".
+:::

### 2024-10-22 output/seminar-slideshow.qmd
Add factor analysis score estimation methods

@@
 
+## AF: Puntuaciones factoriales latentes
+
+Métodos:
+
+-   Regresión
+
+-   Bartlett
+
+-   **Máxima Verosimilitud**
+
+-   **Estimación Modal Bayesiana Empírica**
+
+::: notes
+Si alguna vez habéis ajustado un modelo de AF y habéis calculado las puntuaciones factoriales mediante Máxima Verosimilitud, o mediante esimación Modal Bayesiana (SPSS no lo permite), entonces habéis aplicado la Teoría de Respuesta al Ítem, y a continuación veremos por qué.
+
+Con variables categóricas, si habéis usado correlaciones policóricas en lugar de correlaciones de Pearson (como se recomienda por ejemplo en escalas de respuesta tipo Likert, si tenéis pocas categorías de respuesta), esto es lo más habitual.
+:::
+
 ## Teoría de respuesta al ítem

### 2024-10-22 output/seminar-slideshow.qmd
Add further IRT context

@@
 
+<!--# TODO: Fórmula del parámetro de intersección y del modelo en forma pendiente-intersección -->
+
 ::: notes
-La TRI, por su lado, asume que hay cada persona tiene una puntuación latente, y que su respuesta depende de esa puntuación, con un parámetro de discriminación, que determina cómo de bien o mal mide el ítem, y uno de dificultad (aunque aquí prefiero llamarlo "de posición"), que determina lo fácil o difícil que será dar una respuesta positiva (afirmativa, correcta, codificada como 1).
+La TRI, por su lado, asume que hay cada persona tiene una puntuación latente, y que su respuesta depende de esa puntuación, con un parámetro de discriminación que determina cómo de bien o mal mide el ítem, y uno de dificultad (aunque aquí prefiero llamarlo "de posición") que determina lo fácil o difícil que será dar una respuesta positiva (afirmativa, correcta, codificada como 1).
+Esa posición representa el nivel de rasgo latente que da lugar a una probabilidad de .5 de dar la respuesta positiva.
+
+Ahora bien, la posición se puede reformular como un parámetro de intersección.
 
@@
 Una de las extensiones multidimensionales (más antiguas, y probablemente más obvias) de este modelo consiste simplemente en asumir que la respuesta se modela como una "regresión logística multivariada".
+
+Esto da lugar a una familia de modelos llamados "compensatorios" en TRI, ya que un nivel bajo en una puntuación latente se compensa con un nivel alto en otra, dando lugar a la misma probabilidad de respuesta.
 :::
+
+@reckase_difficulty_1985
+
+@reckase_discriminating_1991

### 2024-10-22 output/seminar-slideshow.qmd
Add bibliography file

@@
   </table>
+bibliography: ../www/Multidimensional-parameters-MCLM.bib
+csl:          ../www/apa-old-doi-prefix.csl
 date:        "2024-10-21"
@@
 ```
-
-::: notes
-:::
 

### 2024-10-22 output/seminar-slideshow.qmd
Add editor configuration options

--- 2024-10-22 output/seminar-slideshow.qmd
@@
   </table>
+bibliography:  ../www/Multidimensional-parameters-MCLM.bib
+csl:           ../www/apa-old-doi-prefix.csl
 date:          "2024-10-21"
@@
 date-format:   long
-editor:      source
+editor:
+  mode:        source
+  markdown: 
+    wrap:      none
+    canonical: true
 lang:          es

### 2024-10-22 output/seminar-slideshow.qmd
Configure editor to apply sentence-wrapping

--- 2024-10-22 output/seminar-slideshow.qmd
@@
 date-format:   long
-editor:      source
+editor:
+  mode:        source
+  markdown: 
+    wrap:      sentence
+    canonical: true
 lang:          es
@@
 
-::: notes
-:::
-
 ## AF: Estructura simple
@@
 
-Los "coeficientes de regresión" los llamamos "pesos factoriales" y el término error, "unicidad" o "varianza única". Esto se puede expresar en "forma matricial" de esta manera.
+Los "coeficientes de regresión" los llamamos "pesos factoriales" y el término error, "unicidad" o "varianza única".
+Esto se puede expresar en "forma matricial" de esta manera.
 
@@
 Sin embargo, se aplica también (e idealmente se debería aplicar) a variables discretas ordinales.
-En este caso vamos a concretarlo en variables dicotómicas (p.ej.
-ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
+En este caso vamos a concretarlo en variables dicotómicas (p.ej. ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
 Esto es una simplificación que rara vez se utiliza en cuestionarios no cognitivos, pero es interesante estudiarlo porque otros modelos pueden considerarse extensiones de este modelo.

### 2024-10-22 output/seminar-slideshow.qmd
Version initial notes

@@
 
-```{r setup}
+# Introducción
+
+```{r libraries}
 library(tibble)
@@
 ```
+
+::: notes
+-   Nombre de la presentación: Cero atractivo
+
+-   Nuria me pide un título más llamativo
+
+-   Lo mejor que puedo conseguir
+
+-   ¿Por qué es relevante?
+
+-   TRI muy usada en evaluación e investigación educativa
+
+-   Poco en investigación psicológica (muestras "pequeñas" en comparación)
+
+-   Pero hay un "paralelismo" entre ambas teorías (isomórficas, ante determinados supuestos)
+:::
 

### 2024-10-22 output/seminar-slideshow.qmd
Add notes to "latent score methods"

@@
 
-## AF: Puntuaciones factoriales latentes
+## AF: Puntuaciones factoriales
 
@@
 
+<!--# TODO: Diagrama de CFA y de "CFA inverso" -->
+
 ::: notes
-Si alguna vez habéis ajustado un modelo de AF y habéis calculado las puntuaciones factoriales mediante Máxima Verosimilitud, o mediante esimación Modal Bayesiana (SPSS no lo permite), entonces habéis aplicado la Teoría de Respuesta al Ítem, y a continuación veremos por qué.
+Los dos primeros métodos se aplican en AF lineal y están implementados en SPSS.
+Consisten a grandes rasgos en calcular la puntuación factorial como una suma ponderada de las variables observables.
+Es decir, algo así como "darle la vuelta a las flechas" de causalidad.
+Lo que tiene más sentido, desde el punto de vista estadístico, es "estimar" qué puntuaciones factoriales hacen más probable para una participante dar las respuestas que ha dado.
+
+Eso es lo que hacen los dos métodos de abajo (SPSS no los implementa creo), Máxima Verosimilitud, o esimación Modal Bayesiana.
+Si habéis hecho esto con variables categóricas, entonces habéis aplicado la Teoría de Respuesta al Ítem, y a continuación veremos por qué.
+:::
 

### 2024-10-22 output/seminar-slideshow.qmd
Introduce FA-IRT parallelism, by McDonald (1999)

@@
 
-Con variables categóricas, si habéis usado correlaciones policóricas en lugar de correlaciones de Pearson (como se recomienda por ejemplo en escalas de respuesta tipo Likert, si tenéis pocas categorías de respuesta), esto es lo más habitual.
+## Paralelismo AF - TRI
+
+::::: columns
+::: column
+![](../www/slideshow-assets/cover_McDonald-1999.jpg){height="300px"}
+
+[@mcdonald_test_1999]
 :::
+
+::: column
+<!--# TODO: Citation by McDonald -->
+:::
+:::::
 
@@
 @reckase_discriminating_1991
+
+## References {.smaller}

### 2024-10-22 output/seminar-slideshow.qmd
Add MIRT computation example

@@
 
+<!--# TODO: Explicar cómo se calcula en forma matricial? -->
+
 Tenemos costumbre de ver el análisis factorial en el contexto de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
@@
 
-<!--# TODO: Gráfica tridimensional de modelo compensatorio -->
+<!--# TODO: Gráfica tridimensional de modelo compensatorio (y "curvas de nivel"?) -->
 
@@
 
-@reckase_difficulty_1985
+## TRI multidimensional: Ejemplo
+
+<!--# TODO: Valores, cálculo y gráfica -->
+
+::: notes
+Veamos un ejemplo de cuál sería la probabilidad de respuesta de una persona con puntuaciones de rasgo $\theta_1 = TODO$ y $\theta_2 = TODO$.
+
+(Hacer cálculos matriciales)
+:::
 

### 2024-10-22 output/seminar-slideshow.qmd
Add FA in MIRT metric

@@
 
+## AF en métrica TRI
+
+<!--# TODO: Fórmulas de los modelos y de transformación de parámetros. -->
+
+::: notes
+A partir de las ecuaciones del AF y de TRI, es fácil ver que existe un paralelismo entre ambos modelos.
+Es posible transformar los parámetros de la métrica del AF a métrica de TRI y, de hecho, esto es necesario, y se hace así para hacer la estimación máximo-verosimil de la que hablábamos antes.
+:::
+
 @reckase_discriminating_1991

### 2024-10-22 output/seminar-slideshow.qmd
Add multidimensional IRT parameters

@@
 
+## TRI Multidimensional: Parámetros multidimensionales
+
+<!--# TODO: Fórmula de modelo M2PL -->
+
+<!--# TODO: Fórmula del parámetro D_i -->
+
+> The distance can be interpreted much like a $b$ parameter from unidimensional IRT
+
+---@reckase_difficulty_1985
+
+::: notes
+A pesar de este paralelismo, y de estar clara la utilidad de los modelos de TRI, hay un problema con la formulación del modelo que os he planteado, y es que como habéis podido ver los parámetros no son fácilmente interpretables.
+:::
+
 @reckase_discriminating_1991

### 2024-10-23 output/seminar-slideshow.qmd
Correct reference to Reckase & McKinley (1991)
Complete outline of multidimensional parameters

@@
 
+@reckase_difficulty_1985:
+
+> \[...\] means of describing the characteristics of an item that takes into account the dimensionality of the skills \[...\] can then be used to determine how or if it is possible to compare items that measure different combinations of abilities.
+
+## Dificultad multidimensional
+
+> \[...\] the most reasonable point to use in defining the MID for an item in the multidimensional space is the point where the item is most discriminating.
+
+---(p. 402)
+
 <!--# TODO: Fórmula del parámetro D_i -->
@@
 
----@reckase_difficulty_1985
+---(p. 405)
+
+<!--# TODO: Representación de la curva de inflexión y la posición del ítem -->
+
+::: notes
+A pesar de este paralelismo, y de la utilidad de los modelos de TRI, hay un problema con la formulación del modelo que os he planteado, y es que como habéis podido ver los parámetros no son fácilmente interpretables.
+La intersección del modelo no tiene una interpretación clara, al contrario que el parámetro de posición en la TRI unidimensional, que permite comparar ítems entre sí en cuanto a su dificultad.
+
+Esto en AF no es muy habitual (prestamos poca atención a la "intersección" o "umbral"), porque no nos interesa tanto cómo de probable sea dar una respuesta positiva o negativa a un ítem, pero podría ayudarnos por ejemplo a diagnosticar ítems que estén mal diseñados, ya que nos da una idea de si va a haber muchas o pocas respuestas positivas a un ítem.
+
+Por ese motivo Reckase propone en 1985 obtener un parámetro de "dificultad" (lo llama, para nosotro sería de posición) que permita comparar ítems, y que sea de algún modo equivalente a la dificultad en el modelo unidimensional.
+Ese punto es donde la curva (superficie) de probabilidad de respuesta tiene una inflexión (a lo largo de toda ella, el punto más cercano al origen, donde la recta que lo une al origen es perpendicular a la recta de inflexión de la curva).
+
+Este parámetro se puede entender como una "distancia con signo" desde el origen hasta el punto de posición del ítem.
+:::
+
+## Dificultad multidimensional
+
+<!--# TODO: Fórmula de los cosenos directores -->
+
+<!--# TODO: Representación gráfica de los cosenos directores? -->
+
+<!--# TODO: Conjunto del MID -->
+
+::: notes
+Para representar ese punto no basta con la "distancia (con signo)"; hace falta saber en qué dirección mide ese ítem también.
+Eso viene dado por los llamados "cosenos directores".
+
+Por lo tanto, la "dificultad (posición) multidimensional" viene dada por dos elementos, un vector de cosenos que determina la dirección en la que mide un ítem, y un número entero que representa cuánto se desplaza respecto del origen, y en qué sentido, la recta de inflexión del ítem.
+Así se puede interpretar un ítem, o varios, de manera relativamente sencilla: Todos los participantes cuyas coordenadas estén en la "recta de inflexión" (la recta que pasa por la posición del ítem, perpendicular a su dirección) tendrán una probabilidad de dar una respuesta positiva y negativa de .5 (dificultad media; igual probabilidad de acertar o fallar).
+Aquellos que estén pasada la recta, encontrarán el ítem "relativamente fácil" (tendrán una alta probabilidad de dar una respuesta positiva mayor a .5; lo encontrarán "más fácil que difícil").
+Lo contrario pasa para los que estén al otro lado de esa recta, tendrán más probabilidad de responder negativamente (mal), por lo que lo encontrarán difícil.
+
+De lo que este parámetro no informa es de "cómo de fácil o difícil" va a ser el ítem para aquellos participantes que están muy cerca o muy lejos de esta recta.
+Es decir, una vez estoy pasada la recta, aunque esté cerca, ¿es el ítem muy fácil, o más bien tengo que estar "muy lejos" (tener alta puntuación en uno u otro rasgo) para que sea más probable que de una respuesta positiva?
+:::
+
+## Discriminación multidimensional
+
+@reckase_discriminating_1991:
+
+> the discriminating power of an item indicates how quickly the transition takes place from low probability to high probability of a correct response.
+> \[...\] A highly discriminating item divides the regions clearly-having a narrow region of ambiguity, that is, a region where the probabilities are intermediate in magnitude.
+
+<!-- # TODO: Superficies de ítems con alta y baja discriminación -->
+
+::: notes
+Esta característica viene representada por el parámetro de "discriminación", que es una medida de la "calidad general" de un ítem.
+Los ítems "buenos" serán muy discriminativos, lo que indica que distinguen bien entre la gente que tiene nivel de rasgo alto o bajo, mientras que los "malos" tendrán una superficie de respuesta más "plana".
+
+La discriminación se puede entender con una especie de "dardo" que atraviesa de arriba abajo esa superficie perpendicularmente por esa "recta de inflexión" de la que hablábamos antes, y luego cruza el plano origen.
+Si la superficie es "totalmente plana" el dardo entra por un punto y sale por otro que está justo debajo, en las mismas coordenadas de rasgo latente.
+La "proyección" (si ponemos un foco justo encima, la sombra que forma en el plano) es un punto.
+Es decir, la "discriminación" es cero (el vector del ítem tiene longitud "cero").
+Si la pendiente es muy pronunciada (distingue bien entre niveles de rasgo) el dardo atravesará el plano origen muy lejos.
+La sombra será muy larga (es decir, la discriminación muy alta).
+El caso "extremo" es una superficie "vertical" en la recta de inflexión (un escalón; cosa imposible en ítems reales).
+En este caso, el dardo nunca cortaría al plano origen, y la discriminación sería infinita.
+:::
+
+## Parámetros multidimensionales
+
+@reckase_discriminating_1991:
+
+<!-- # TODO: Fórmula de MDISC -->
+
+<!-- # TODO: Gráfico del cálculo de MDISC como teorema de Pitágoras -->
+
+<!-- # TODO: Fórmula de dificultad con MDISC -->
+
+::: notes
+@reckase_discriminating_1991 hacen la deducción de la discriminación multidimensional, y resulta dar lugar a esta fórmula.
+Como veis es simplemente la "longitud" del vector de parámetros de discriminación, resultante simplemente de aplicar el teorema de Pitágoras.
+El vector es un triángulo rectángulo, $a_1$ sería un cateto, el otro cateto sería igual a $a_2$, y la hipotenusa es la longitud (las dos discriminaciones al cuadrado, sumadas, y luego la raíz cuadrada de eso).
+Esto por cierto también vale para cualquier número de dimensiones.
+Aquí estamos representando sólo 2 porque es más fácil, pero podemos calcular la suma de cuadrados estos parámetros de discriminación (sea cual sea su número), y calcular su raíz cuadrada, y eso nos daría igualmente la discriminación multidimensional (la longitud de ese vector de $a$ del ítem).
+:::
+
+## Representación gráfica
+
+@ackerman_graphical_1996:
+
+::: notes
+La propuesta de representación gráfica completa (y consistente) no llega hasta 1996.
+Ackerman propone que el ítem se represente como un vector de longitud dada por la discriminación multidimensional, aplicado en la posición del ítem (y en la dirección dada por esa posición; es decir, todos los ítems "apuntan" siempre en la dirección que pasa por el origen).
+:::
+
+---@reckase_discriminating_1991
 
@@
 
-@reckase_discriminating_1991
-
 ## References {.smaller}

### 2024-10-23 output/seminar-slideshow.qmd
Add outline of generalization to inverse items

@@
 
----@reckase_discriminating_1991
+## Parámetros multidimensionales: Supuestos
+
+> the probability of answering an item correctly increases monotonically with an increase in each dimension being measured
+
+---@reckase_difficulty_1985 [p. 402]
+
+<!-- # TODO: Ecuación modelo y ejemplo probabilidad constante -->
+
+::: notes
+Ahora bien, como hemos visto, la TRI originalmente sólo trataba con "ítems cognitivos" (lo hemos podido ver en el propio lenguaje que utiliza Reckase: "dificultad", "habilidad", "respuesta correcta").
+Uno de los supuestos que hace es este.
+Eso tiene todo el sentido (casi), ya que si la probabilidad de un ítem "decrece" al aumentar el nivel de habilidad, es que algo está mal en el ítem: No puede ser que alguien que sepa más tenga menos probabilidad de acertar un ítem que alguien que sepa menos.
+Pero cuidado, si el ítem es "puro" de una dimensión, la probabilidad en la otra u otras dimensiones sería constante (discriminación = 0); es decir, la probabilidad puede ser "plana" (es una dimensión que no "está siendo medida por el ítem").
+:::
+
+## Generalización: Ítems inversos
+
+<!-- # TODO: Curva de ítem inverso y ejemplo -->
+
+<!-- # TODO: Fórmulas de los parámetros -->
+
+<!-- # TODO: Ejemplo con a negativa -->
+
+::: notes
+Pero además, si hablamos de ítems "no cognitivos" (p.ej. personalidad) en los que la persona evaluada puede estar "de acuerdo" o "en desacuerdo", sí podemos tener ítems con probabilidad "decreciente".
+
+La generalización a ítems inversos es directa: La discriminación multidimensional "eleva al cuadrado" los "parámetros de discriminación marginales", luego va a ser siempre positiva.
+El parámetro de posición, por tanto, siempre va a tener el signo inverso al que tenga la intersección.
+Los cosenos tendrán, cada uno, el mismo que el parámetro de discriminación de esa dimensión.
+:::
+
+## Generalización: Ítems inversos
+
+<!-- # TODO: Representación de ítems con a (una o varias) negativa (tabla y gráfico) -->
+
+::: notes
+La representación de estos ítems sería como se muestra en este gráfico.
+<!-- # TODO: Descripción de los ítems representados -->
+:::
 

### 2024-10-23 output/seminar-slideshow.qmd
Add "generalization to constant monotony" outline

@@
 Aquí estamos representando sólo 2 porque es más fácil, pero podemos calcular la suma de cuadrados estos parámetros de discriminación (sea cual sea su número), y calcular su raíz cuadrada, y eso nos daría igualmente la discriminación multidimensional (la longitud de ese vector de $a$ del ítem).
+
+Esto es en realidad el producto escalar (o producto punto) de un vector consigo mismo, que consiste en multiplicar una a una las componentes y sumarlos todos.
+En forma matricial se representaría como el vector pre-multiplicado por sí mismo.
+<!-- # TODO: Explicar la multiplicación vectorial (y poner ejemplo?) -->
+
+Esta expresión de la discriminación multidimensional se puede usar para expresar de manera más sencilla la dificultad.
 :::
@@
 
-## Parámetros multidimensionales: Supuestos
+## Supuestos: Monotonía creciente
 
@@
 
+## Ítems "inversos" en la literatura
+
+> the negative relation of Item 15 with geometry achievement is particularly puzzling
+
+---@mcdonald_basis_2000 [p. 109]
+
+> This result occurs because $a_1$ for the item is negative, and thus as an examinee's $\theta_1$ ability increases, the chance of answering the item correctly actually decreases.
+
+---@ackerman_multidimensional_2005-1 [p. 17]
+
+### Generalización: "Monotonía constante"
+
+::: notes
+Esta generalización es bastante evidente, y se puede aplicar sin problema a ítems no cognitivos.
+Sin embargo, hay que tener en cuenta que Reckase explicitaba como condición que los ítems tenían que ser "monotonamente crecientes".
+Pero en la literatura nos encontramos casos como estos, que no lo son, y a los autores no se les ocurre decir que "hay que generalizar esas definiciones".
+Más bien, hacen comentarios como estos, restándole importancia al hecho de que los parámetros de discriminación sean negativos ("particularmente misterioso"), o afirmando lo obvio (el parámetro de discriminación del ítem es negativo para una dimensión, por lo que la probabilidad de respuesta correcta disminuye al incrementarse su habilidad)
+
+En lugar de eso, proponemos que los supuestos de Reckase se "relajen" para considerar ítems con probabilidad "con monotonía constante" (es decir, que no tengan un "pico" u óptimo de rasgo latente para dar una respuesta positiva y luego decrezca), sino que cuanto más extremo sea el nivel de rasgo, más extrema es también la probabilidad (positiva o negativa).
+Este caso general incluye también la probabilidad constante (ya que es monotonía "nula constante", es decir, "el cambio es nulo de manera constante").
+Este tipo de ítems, positivos, negativos, o nulos, se pueden acomodar y modelar fácilmente por un modelo como el logístico de 2 parámetros multidimensional, como hemos visto.
+Ítems que estarían excluidos, por no tener monotonía constante, podrían ser formulaciones como, por ejemplo, "Soy tan extrovertido como la mayoría de mis amigos" (alguien puede decir, "No, soy más introvertido que todos ellos", respuesta negativa con rasgo bajo, o "Qué va, soy el más extrovertido con diferencia", respuesta negativa con rasgo alto).
+
 ::: notes

### 2024-10-23 output/seminar-slideshow.qmd
Add "orthogonality assumption" outline

@@
 Ítems que estarían excluidos, por no tener monotonía constante, podrían ser formulaciones como, por ejemplo, "Soy tan extrovertido como la mayoría de mis amigos" (alguien puede decir, "No, soy más introvertido que todos ellos", respuesta negativa con rasgo bajo, o "Qué va, soy el más extrovertido con diferencia", respuesta negativa con rasgo alto).
+:::
+
+## Supuestos: Ortogonalidad
+
+> before performing the differentiation, the constraint that <!-- # TODO: Sum of cosines^2 = 1 --> is added to the expression for the slope \[...\]
+
+---@reckase_difficulty_1985 [p. 404]
+
+<!-- # TODO: Ejemplo de rotación oblicua con los ejes resultantes oblicuos -->
+
+<!-- # TODO: Coordenadas rectangulares resultantes de una rotación oblicua -->
+
+::: notes
+Otro de los supuestos que hace Reckase originalmente para calcular los parámetros es este: Los cosenos al cuadrado suman 1.
+Es decir, está aplicando el teorema de Pitágoras, pero ello implica de manera implícita que los ejes del espacio latente son "ortogonales" (forman ángulos de 90 grados).
+
+Esto en TRI multidimensional originalmente tenía sentido, ya que no se tenía muy en cuenta la interpretación de los parámetros de rasgo latente.
+O bien no importaba mucho que los ejes se pudieran interpretar como rasgos latentes, o se interpretaban de manera no necesariamente muy precisa.
+
+No obstante, en 1999 llega McDonald, y dice que la TRI y el AF son fundamentalmente la misma cosa.
+Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas que dan lugar a ejes no ortogonales.
+
+Los pesos factoriales resultantes (o discriminaciones) los podemos representar sin ningún problema en coordenadas rectangulares (las llamadas "coordenadas Cartesianas").
+Pero algo parece fallar cuando generalizamos los parámetros multidimensionales de TRI que hemos mostrado antes a esos espacios de coordenadas no ortogonales.
+Si rotamos los ejes de coordenadas, pero los ítems siguen siendo los mismos, es de esperar que la "longitud" del vector de ese ítem (i.e.
+su discriminación), no cambie (esta propiedad se llama invarianza).
+Veámoslo con un ejemplo.
+:::
+
+## Supuestos: Ortogonalidad
+
+<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
+
+<!-- # TODO: Ejemplo de rotación oblicua (gráfico de ejes y matriz de rotación) -->
+
+::: notes
+Supongamos este ítem, en coordenadas ortogonales.
+El vector muestra sus parámetros de discriminación en las dos dimensiones de los ejes representados.
+Esto sería la "solución sin rotar" que resulta típicamente de ajustar un modelo de TRI multdimensional, o un AFE exploratorio.
+La discriminación multidimensional de este vector es (explicar paso a paso).
+
+Supongamos que aplicamos una rotación oblicua.
+Hemos identificado dos rasgos latentes sustantivos, correlacionados entre sí.
+Los ejes resultantes son los que se muestran en este gráfico.
+Los nuevos parámetros de discriminación del ítem están representadas cada una por la "proyección" de ese vector sobre cada uno de los ejes (paralela al otro, u otros, si hay más de dos).
+:::
+
+## Supuestos: Ortogonalidad
+
+<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
+
+<!-- # TODO: Ejemplo de item proyectado en los ejes rotados (gráfico y vectorial) -->
+
+<!-- # TODO: Cálculo de la discriminación en los ejes rotados -->
+
+::: notes
+Hacer una rotación supone "multiplicar" las coordenadas de este vector por la matriz de rotación que representa ese nuevo sistema de coordenadas (la transformación del ortogonal a este nuevo, oblicuo).
+En este caso, esta es la matriza de rotación.
+Como los vectores se representan en columnas, se "pre-multiplica" por la matriz de rotación.
+Esto es parecido a multiplicar por un vector; es como si cada fila de la matriz fuese un vector, y los resultados se ponen cada uno en una fila.
+<!-- # TODO: Explicar paso a paso el producto matricial -->
+
+Si ahora calculamos la discriminación del ítem en estos ejes rotados, nos encontramos con este resultado.
+Es decir, ¿que el ítem discrimina mejor o peor en función de qué ejes quiera yo considerar?
+Eso no parece tener sentido, y sin embargo hay muchos ejemplos en la literatura que calculan esta discriminación multidimensional con la fórmula de Reckase (con "rasgos correlacionados") sin preocuparse por "los ejes" que representan esos rasgos de manera fidedigna.
+<!-- # TODO: Ejemplos de la literatura -->
+:::
+
+## Supuestos: Ortogonalidad
+
+> It has been suggested by some researchers that the angle between the axes represents the degree of correlation.
+> However, for sake of clarity, an orthogonal axes system is used in which distance measures and vectors can be easily calculated, understood, and interpreted.
+
+---@ackerman_multidimensional_2005-1 [p. 16]
+
+### Generalización: "Coordenadas oblicuas"
+
+::: notes
+Para la representación gráfica esto puede valer, como dice Ackerman aquí.
+Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente.
 

### 2024-10-23 output/seminar-slideshow.qmd
Add "olblique coordinates generalization" outline

@@
 Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente.
+:::
+
+## Generalización: Coordenadas oblicuas
+
+> before performing the differentiation, the constraint that <!-- # TODO: Sum of cosines^2 = 1 --> is added to the expression for the slope \[...\]
+
+---@reckase_difficulty_1985 [p. 404]
+
+<!-- # TODO: Expresión de las \thetas transformadas -->
+
+<!-- # TODO: Distribución de \thetas en ejes ortogonales -->
 
 ::: notes
-A pesar de este paralelismo, y de estar clara la utilidad de los modelos de TRI, hay un problema con la formulación del modelo que os he planteado, y es que como habéis podido ver los parámetros no son fácilmente interpretables.
+Hacer esta generalización no es algo tan directo como ocurría con la probabilidad "monótonamente decreciente".
+Si os acordáis, Reckase aplicaba esta restricción de que la suma de cosenos directores era 1 (que se deriva del teorema de Pitágoras).
+Pero para aplicarla hace falta asumir que esos cosenos directores son todos ortogonales entre sí.
+Si no tenemos ángulos rectos, esa condición no se cumple.
+
+Y lo que nosotros estamos asumiendo que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones ($R$, la que sea).
+Normalmente la representamos en coordenadas rectangulares (ejes ortogonales).
+
+, que están representados en un espacio no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el o y el 1 pueda ser distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
+
+¿Cómo podríamos solucionar esto?
+Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de este espacio oblicuo (el que sea, aún no sabemos cuál es) a un espacio ortogonal.
+:::
+
+## Generalización: Coordenadas oblicuas
+
+<!-- # TODO: Expresión de las \thetas transformadas -->
+
+<!-- # TODO: Expresión de la R transformada -->
+
+<!-- # TODO: Distribución (esférica) de las \thetas transformadas -->
+
+::: notes
+Esto es una transformación de las coordenadas, como la que aplicamos antes para calcular las coordenadas del ítem tras aplicar la rotación oblicua, pero aplicada en este caso al vector de rasgo latente.
+
+Estas nuevas coordenadas (transformadas), que ahora están en un espacio ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^B$) sea una matriz identidad (unos en la diagonal, ceros en el resto).
 :::

### 2024-10-23 output/seminar-slideshow.qmd
Improve notes in "oblique coordinates"

@@
 Y lo que nosotros estamos asumiendo que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones ($R$, la que sea).
-Normalmente la representamos en coordenadas rectangulares (ejes ortogonales).
-
-, que están representados en un espacio no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el o y el 1 pueda ser distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
+Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a coordenadas oblicuas tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el o y el 1 pueda ser distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
 

### 2024-10-23 output/seminar-slideshow.qmd
Completes outline of "oblique coordinates"

@@
 Estas nuevas coordenadas (transformadas), que ahora están en un espacio ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^B$) sea una matriz identidad (unos en la diagonal, ceros en el resto).
+Pero hay que tener en cuenta que esto es un "artefacto matemático".
+Estamos haciendo esto por conveniencia, pero las coordenadas que realmente nos interesan como investigadores, porque son las que verdaderamente representan algo que nos interesa investigar, son esas coordenadas originales, que pueden representar cualquier fenómeno que sea de nuestro interés (rasgos de personalidad, síntomas, actitudes, inteligencia, conocimientos...
+lo que estemos investigando).
+:::
+
+## Generalización: Coordenadas oblicuas
+
+<!-- # TODO: Ejemplo de \theta (expresión) y coordenadas en espacio oblicuo -->
+
+::: notes
+Mientras tanto, las coordenadas de los "rasgos latentes" originales se van a representar en otros ejes, en general distintos de los "ejes ortogonales" transformados.
+Esto lo que significa es que las puntuaciones de rasgo latente originales van a representar la posición de un participante respecto de esos ejes originales.
+Por ejemplo, en este caso, si el participante tiene un "vector de puntuaciones" \[1, 2\], para representar sus coordenadas tenemos que desplazarnos, primero una unidad en el primer eje, y luego dos unidades en el segundo eje.
+:::
+
+## Generalización: Coordenadas oblicuas
+
+<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
+
+<!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
+
+::: notes
+En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
+Supongamos una distribución bivariada de rasgos latentes, con una correlación de .8, bastante alta.
+¿Cuál de los siguientes sistemas de coordenadas consideráis que representaría adecuadamente los ejes de una distribución bivariada, de forma que al transformarlos a coordenadas rectangulares obtuviésemos una distribución con correlación de 0?
+:::
+
+## Generalización: Coordenadas oblicuas
+
+<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
+
+<!-- # TODO: Distribución en ejes rectangulares -->
+
+::: notes
+El primero ya sabemos que no, de acuerdo, puesto que es la representación habitual, y sabemos que la nube que forma está "achatada".
+No se aplica ninguna transformación (la transformación es una matriz identidad), ya que los que serían los ejes originales y los transformados son los mismos.
+:::
+
+## Generalización: Coordenadas oblicuas
+
+<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
+
+<!-- # TODO: Animación de distribución en ejes rectangulares a ejes alineados -->
+
+::: notes
+Si pensáis igual que yo, habréis elegido el segundo, porque el coseno del ángulo entre los ejes representa la correlación entre las dimensiones.
+Por lo tanto, ejes muy alineados representan una correlación muy alta.
+
+Sin embargo, fijáos lo que ocurre cuando "transformamos" las coordenadas de los puntos de nuestra distribución a ese sistema de coordenadas.
+Como podéis ver, la distribución tiende a "achatarse" aún más, en lugar de hacerse más "esférica".
+:::
+
+## Generalización: Coordenadas oblicuas
+
+<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
+
+<!-- # TODO: Animación de distribución en ejes rectangulares a ejes correctos -->
+
+<!-- # TODO: Matriz de correlaciones transformada "identidad" -->
+
+::: notes
+El sistema de coordenadas correcto es el tercero, que "compensa" de alguna manera el achatamiento de la nube de puntos, de forma que las coordenadas transformadas que nos quedan (en los ejes ortogonales), los "normales" del gráfico) tienen una distribución "esférica" (con matriz de correlaciones "identidad")
 :::

### 2024-10-23 output/seminar-slideshow.qmd
Items in transformed coordinates slide

@@
 
+## Generalización: Coordenadas oblicuas
+
+<!-- # TODO: Fórmula del modelo en coordenadas "transformadas" y equivalencia -->
+
+<!-- # TODO: Expresión de la transformación de la discriminación a coordenadas rectangulares -->
+
+::: notes
+Lo que ocurre es que el espacio en el que se encuentran estas coordenadas de rasgo latente es diferente del espacio en el que se encuentran los ítems.
+Para que el modelo tenga una probabilidad "invariante", los ítems tamibén tienen que transformarse a un sistema de coordenadas ortogonal.
+Pero podemos ver que su matriz de transformación no es la misma que la de las puntuaciones de rasgo latente.
+Es su inversa (inversa traspuesta, más concretamente)
+:::
+
 ## References {.smaller}

### 2024-10-24 output/seminar-slideshow.qmd
Completes outline

--- 2024-10-23 output/seminar-slideshow.qmd
@@
 
-## Supuestos: Ortogonalidad
-
-> It has been suggested by some researchers that the angle between the axes represents the degree of correlation.
-> However, for sake of clarity, an orthogonal axes system is used in which distance measures and vectors can be easily calculated, understood, and interpreted.
-
----@ackerman_multidimensional_2005-1 [p. 16]
-
 ### Generalización: "Coordenadas oblicuas"
@@
 ::: notes
-El sistema de coordenadas correcto es el tercero, que "compensa" de alguna manera el achatamiento de la nube de puntos, de forma que las coordenadas transformadas que nos quedan (en los ejes ortogonales), los "normales" del gráfico) tienen una distribución "esférica" (con matriz de correlaciones "identidad")
+El sistema de coordenadas correcto es el tercero, que "compensa" de alguna manera el achatamiento de la nube de puntos, de forma que las coordenadas transformadas que nos quedan (en los ejes ortogonales), los "normales" del gráfico) tienen una distribución "esférica" (con matriz de correlaciones "identidad").
+
+Esto, que visto así parece ya evidente, es algo que al menos a mí me ha llevado varias decenas (puede que "cientos") de horas entender, durante al menos dos años trabajando en este problema.
+:::
+
+## Espacio latente vs. espacio test
+
+> It has been suggested by some researchers that the angle between the axes represents the degree of correlation.
+> However, for sake of clarity, an orthogonal axes system is used in which distance measures and vectors can be easily calculated, understood, and interpreted.
+
+---@ackerman_multidimensional_2005-1 [p. 16]
+
+<!-- # TODO: Fórmula del modelo en coordenadas "transformadas" y equivalencia -->
+
+<!-- # TODO: Expresión de la transformación de la discriminación a coordenadas rectangulares -->
+
+::: notes
+Viendo que en la literatura hay afirmaciones como esta, no me resulta totalmente descorazonador.
+Parece que no soy el único al que le ha supuesto un esfuerzo entenderlo (resultados de la "encuesta"?)
+
+Lo que ocurre en realidad, y esto es en realidad también un paralelismo con lo que ocurre en análisis factorial, es que para que el modelo sea invariante, los ítems también tienen que transformarse a un sistema de coordenadas ortogonal.
+Pero podemos ver que su matriz de transformación no es la misma que la de las puntuaciones de rasgo latente.
+Es su inversa (inversa traspuesta, más concretamente).
+
+Y esto nos ha permitido "descubrir" (o darnos cuenta, más bien) de que hay dos "espacios vectoriales" distintos, uno en el que se encuentran las coordenadas de las puntuaciones factoriales, y otro distinto, que es el espacio en el que se encuentran las coordenadas de los ítems, y que es donde en realidad hay que representarlos.
+:::
+
+## Espacio latente vs. espacio test
+
+> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by <!-- # TODO: Inner product -->.
+
+---@zhang_theoretical_1999 [p. 221]
+
+<!-- #TODO: Ref. to Mahalanobis -->
+
+> \[...\] the *inner product* in \[the test space\] is consistently defined by <!-- # TODO: Inner product -->.
+
+---@zhang_theoretical_1999 [p. 221]
+
+::: notes
+Esto, curiosamente, aparece en la literatura como podéis ver; en el año 1999, en un artículo que no tiene nada que ver con esto.
+Lo menciona de soslayo y no profundiza en la deducción de los parámetros, como hemos hecho aquí.S olamente lo utiliza para resolver un problema "parcial" que se encuentra en la deducción de un índice de dimensionalidad, distinto de los parámetros de los que estamos hablando aquí.
+Pero podía haber obviado este problema y haber utilizado coordenadas ortogonales, sin importarle si había o no correlaciones entre los rasgos latentes, como ha hecho todo el mundo.
+
+Así que, todos mis respetos por estos autores.
+Sin embargo claro, ellos estaban hablando de otra cosa, y sólo tocaban este tema de paso.
+Así que al final eso se quedó ahí, y nadie ha hecho caso en 25 años a lo que dijeron Zhang y Stout.
+La gente ha usado los modelos, hecho simulaciones, etc., usando distribuciones correlacionadas, y ha usado las fórmulas de Reckase para transformar entre parámetros multidimensionales y los parámetros "lineales" (podríamos llamarlos, los de la formulación original) del modelo.
+:::
+
+## Parámetros multidimensionales generalizados
+
+<!-- # TODO: Fórmulas de los parámetros multidimensionales -->
+
+::: notes
+Si tenemos en cuenta las correlaciones (más generalmente, covarianzas) de la distribución de puntuaciones factoriales, lo que obtenemos son estas expresiones para los parámetros multidimensionales.
+
+Muy importante, la distancia con signo y la discriminación multidimensional son iguales, no importa en qué espacio queramos obtenerlas y representarlas.
+Pero los cosenos directores es necesario calcularlos en el espacio test, no en el espacio de puntuaciones latentes, como decía Reckase, porque este es el espacio que va a representar en realidad la relación entre las variables latentes.
+:::
+
+## Conclusiones
+
+Parámetros multidimensionales generalizados
+
+-   A ítems con monotonía constante
+
+-   A coordenadas oblicuas
+
+Distinción espacio latente - espacio test
+
+::: notes
+En conclusión, lo que hemos propuesto en este artículo consiste en una generalización:
+
+-   A ítems con monotonía constante, creciente, decreciente o nula.
+    Esto se venía haciendo ya como hemos visto, pero creemos que hacía falta formalizarlo.
+    No basta con hacerlo y punto, si Reckase dijo que las discriminaciones tiene que ser positivas, tenemos que demostrar que también vale para discriminaciones negativas o nulas (monotonía decreciente, o probabilidad constante).
+
+-   A coordenadas oblicuas, las cuales consideramos que son necesarias para representar distribuciones de puntuaciones de rasgo correlacionadas, en general.
+    Hemos visto que esto ya lo mencionaron Zhang y Stout, hace 25 años, pero no proporcionaron una "deducción" de estos parámetros multidimensionales, cosa que nosotros hacemos.
+
+Por último, es imortante recalcar este "descubrimiento" (o "redescubrimiento", más bien) del espacio test, ya que tenemos que ser muy conscientes de que los ítems y las personas están en espacios vectoriales distintos.
+Aunque esto parezca solamente un "artefacto matemático" (y en el fondo lo es), hemos visto que es verdaderamente crucial tenerlo en cuenta para entender bien los modelos que estamos utilizando.
+Además, esta "distinción" se halla también en los modelos factoriales: Existe lo que se denominan "ejes primarios" y "ejes de referencia", propuestos por Thurstone hace años (pero que prácticamente no se utilizan ni tienen en consideración).
+<!-- # TODO: Hace cuántos años?
+(Thurstone) --> Sin embargo, cuidado con esto, porque no está claro que esos espacios en TRI multidimensional se puedan asimilar directamente a los ejes en un modelo de AF.
+:::
+
+## To be continued...
+
+-   Paralelismo TRI multidimensional - AF
+
+-   Extensiones del modelo M2PL: Escala graduada, respuesta nominal, 3PL...
+
+-   Investigación en otros modelos
+
+::: notes
+Esta de hecho puede ser una futura línea de investigación.
+Puede parecer evidente el paralelismo (o al menos a mí me lo parece), pero si una cosa me ha enseñado el estar trabajando en este problema de los parámetros (y toparme con el problema de los dos espacios), es que no hay que fiarse de las apariencias.
+Por eso aplicamos el método científico (el método puramente deductivo, en este caso)
+
+El modelo multidimensional logístico de 2 parámetros se puede considerar como "anidado", o un caso más simple de varios modelos.
+Así que partiendo de este como base, lo interesante es aplicarlo a otros más complejos, pero más útiles en la práctica.
+Por cierto, que esto también lo hace mucha gente, aplica las fórmulas de Reckase a otros modelos y a correr, se quedan "tan anchos".
+
+El modelo de escala graduada, por ejemplo, sería el equivalente al modelo de AF para respuestas ordinales, con varios umbrales o "posiciones" en TRI.
+
+El de crédito parcial, por ejemplo, puede servir también para formatos con varias opciones de respuesta que miden o pesan en factores diferentes cada una de ella.
+
+El logístico de 3 parámetros es como el de 2, pero añadiendo un parámetro de pseudoazar, para modelar el que un participante acierte un ítem por pura suerte, respondiendo "a ciegas" (esto sólo valdría para ítems cognitivos).
+
+Estos modelos, que son "extensiones" de este más básico, deberían partir de estos parámetros como fundamento.
+Pero un investigador no debería dar los parámetros sin más de ese modelo, sin proporcionar una prueba formal de que se aplica esa fórmula y se cumplen los supuestos.
+:::
+
+## Últimas palabras
+
+<!-- # TODO: Imagen ilustrativa -->
+
+::: notes
+A este respecto, me gustaría terminar recalcando esto último que he mencionado, aplicado no sólo a cuestiones psicométricas, o de TRI.
+Cuando estaba empezando a preparar este seminario, una de mis inquietudes era la complejidad, tanto conceptual como formal de esta investigación.
+Compartiendo estas inquietudes con algunos compañeros, me aconsejaron utilizar digamos este discurso como excusa para lanzar este "mensaje" en general: La idea de que tenemos que ser muy precavidos a la hora de aplicar estos métodos, sean cuales sean (cuantitativos, cualitativos, o del tipo que sean) sin reflexionar sobre si son los más adecuados o no.
+Y bueno, a este respecto comentar que, no sólo yo, sino el departamento de Metodología al completo, hacemos una tipo de investigación que precisamente se encarga de eso, de intentar entender estos métodos, saber si son los más adecuados, y saber cuándo, cómo, y sobre todo, por qué, aplicarlos en la investigación aplicada (aplicada para nosotros, al menos).
+
+Así que bueno, ya lo sabéis, estamos disponibles en el departamento para colaborar en esas investigaciones, si lo consideráis oportuno, para aplicar los métodos más apropiados a los problemas de investigación que tengáis.
+En mi caso en concreto, no sólo puedo proporcionar cierto "expertise" psicométrico.
+También me dedico mucho a cuestiones de ciencia abierta y reproduciblidad, así que si queréis darle a vuestra investigación un giro en ese sentido, para hacerla más sostenible y alineada con los objetivos de los organismos financiadores (y si no queréis, deberíais ir queriendo), podéis contactar conmigo, y podemos plantear líneas de colaboración.
+Eso sí, me gustaría pediros, si necesitáis colaboración, no sólo metodológica, sino de cualquier tipo, por favor no nos llaméis cuando tengáis un lío de datos, y no sepáis bien cómo resolver la papeleta (no digo que sea vuestro caso, pero me he encontrado con muchos casos, que son verdaderas historias de terror).
+Contad por favor con nosotros desde el momento de escribir vuestro proyecto, para solicitar la financiación, porque es en ese punto en el que se pueden prevér problemas y anticiparse a ellos.
+Porque si una investigación no está bien diseñada desde el principio, teniendo en cuenta estos aspectos metodológicos, de reproducibilidad, etc., si tenemos mala suerte pueden surgir problemas que se puedan resolver sólo con muchísimo trabajo, o incluso que ya nunca se puedan resolver, y tengamos que tirar a la basura todo ese esfuerzo de investigación.
+
+Eso es todo, muchas gracias.
+Y espero vuestras preguntas y comentarios.
 :::

### 2024-10-24 output/seminar-slideshow.qmd
Add color palette

@@
 library(knitr)
+```{r color-palettes}
+PALETTE_UNED <- c(
+  normal = "#00533e",
+  medium = "#427562",
+  light  = "#86a699",
+  white  = "#cddad5"
+)
+PALETTE_APPLE <- c(
+  normal = "#749f4c",
+  medium = "#8aac5d",
+  light  = "#a2bc7e",
+  white  = "#cddbb8"
+)
+PALETTE_BLUE <- c(
+  normal = "#5c6eb1",
+  medium = "#7683bd",
+  light  = "#929acb",
+  white  = "#d4d6eb"
+)
+PALETTE_TANGERINE <- c(
+  normal = "#d76f47",
+  medium = "#dd8964",
+  light  = "#e4a482",
+  white  = "#f0cfb9"
+)
+PALETTE_STRAWBERRY <- c(
+  normal = "#da5268",
+  medium = "#c87384",
+  light  = "#daa6b1",
+  white  = "#efdbdf"
+)
+PALETTE_RASPBERRY <- c(
+  normal = "#90214a",
+  medium = "#8e4d60",
+  light  = "#af848b",
+  white  = "#dac7c7"
+)
+```
+
 ```

### 2024-10-24 output/seminar-slideshow.qmd
Add graphical output configuration

--- 2024-10-24 output/seminar-slideshow.qmd
@@
       - ../www/slideshow-assets/title-slide.html
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(knitr)
+
+```{r souces}
+source("../R/Output.R", encoding = 'UTF-8')
+```
+
+```{r color-palettes}
+PALETTE_UNED <- c(
+  normal = "#00533e",
+  medium = "#427562",
+  light  = "#86a699",
+  white  = "#cddad5"
+)
+PALETTE_APPLE <- c(
+  normal = "#749f4c",
+  medium = "#8aac5d",
+  light  = "#a2bc7e",
+  white  = "#cddbb8"
+)
+PALETTE_BLUE <- c(
+  normal = "#5c6eb1",
+  medium = "#7683bd",
+  light  = "#929acb",
+  white  = "#d4d6eb"
+)
+PALETTE_TANGERINE <- c(
+  normal = "#d76f47",
+  medium = "#dd8964",
+  light  = "#e4a482",
+  white  = "#f0cfb9"
+)
+PALETTE_STRAWBERRY <- c(
+  normal = "#da5268",
+  medium = "#c87384",
+  light  = "#daa6b1",
+  white  = "#efdbdf"
+)
+PALETTE_RASPBERRY <- c(
+  normal = "#90214a",
+  medium = "#8e4d60",
+  light  = "#af848b",
+  white  = "#dac7c7"
+)
+```
+
+```{r read-chunks}
+read_chunk("../src/Graphical_example_paper.R")
+```
+
+```{r constants}
+```
+
+```{r graphical-output-conf}
 ```

### 2024-10-24 output/seminar-slideshow.qmd
Add FA path diagrams

@@
 ```{r libraries}
+#| include: false
+
 library(tibble)
-library(dplyr, warn.conflicts = FALSE)
+library(dplyr)
 library(knitr)
+library(tidyr)
+library(ggplot2)
+library(lavaan)
+library(semPlot)
+```
 
@@
 <!--# TODO: Ejemplos estructura simple vs. compleja: Path diagram vs. coordenadas -->
+
+```{r simple-complex-structure}
+ITEM_BASE_LOADING <- 1
+ITEM_DIST         <- 0.2
+
+simple_item_loadings <- tribble(
+  ~item, ~factor, ~loading,
+  'I1',  'F1',    ITEM_BASE_LOADING,
+  'I2',  'F1',    ITEM_BASE_LOADING + ITEM_DIST,
+  'I3',  'F1',    ITEM_BASE_LOADING + 2 * ITEM_DIST,
+  'I4',  'F2',    ITEM_BASE_LOADING,
+  'I5',  'F2',    ITEM_BASE_LOADING + ITEM_DIST,
+  'I6',  'F2',    ITEM_BASE_LOADING + 2 * ITEM_DIST
+)
+complex_item_loadings <- simple_item_loadings                                |>
+  bind_rows(tibble(item = 'I3', factor = 'F2', loading = ITEM_BASE_LOADING)) |>
+  complete(item, factor, fill = list(loading = 0))
+
+simple_item_loadings <- simple_item_loadings |>
+  complete(item, factor, fill = list(loading = 0))
+
+# TODO: Syntax from tibble
+simple_spec <- "
+  F1 =~ I1 + I2 + I3
+  F2 =~ I4 + I5 + I6
+
+  F1 ~~ F2
+"
+complex_spec <- "
+  F1 =~ I1 + I2 + I3
+  F2 =~ I3 + I4 + I5 + I6
+
+  F1 ~~ F2
+"
+
+simple_spec      |>
+  semPlotModel() |>
+  semPaths(
+    edge.color     = "black",
+    residuals      = FALSE,
+    thresholds     = FALSE,
+    intercepts     = FALSE,
+    fixedStyle     = 1,
+    whatLabels     = "std"
+  )
+
+simple_item_loadings |>
+  pivot_wider(names_from = factor, values_from = loading) |>
+  ggplot(mapping = aes(F1, F2)) +
+  geom_point(size = I(4))
+
+complex_spec     |>
+  semPlotModel() |>
+  semPaths(
+    edge.color     = c(
+      "black" |> rep(3),
+      PALETTE_RASPBERRY["normal"], # cross-loading
+      "black" |> rep(3)
+    ),
+    residuals      = FALSE,
+    thresholds     = FALSE,
+    intercepts     = FALSE,
+    fixedStyle     = 1,
+    whatLabels     = "std",
+  )
+```
 

### 2024-10-24 output/seminar-slideshow.qmd
Update graphical configuration

@@
 knitr:
-  opts_knit: 
-    root_dir:  here::here()
+  # opts_knit: # Does not work 
+  #   root_dir:  here::here()
   opts_chunk: 
@@
 
-```{r souces}
-source("../R/Output.R", encoding = 'UTF-8')
+```{r setup}
+#| include: false
+opts_knit$set(root.dir = here::here()) # See lines 41-42
 ```
@@
 ```{r read-chunks}
-read_chunk("../src/Graphical_example_paper.R")
+read_chunk("src/Graphical_example_paper.R")
+```
+
+```{r sources}
 ```
@@
 ```{r constants}
+```
+
+```{r set-font}
+GRAPH_FONT <- "arial" # Substitute default font to comply with UNED identity
 ```

### 2024-10-24 R/Output.R

@@ -55,7 +55,7 @@ format_prop_like <- function(values, sig = 3, drop_0 = TRUE) {
     round(digits = sig) |>
     format(digits = sig, nsmall = sig)
   
-  if (drop_0) result <- result |> str_remove(LEADING_ZERO_PATTERN)
+  if (drop_0) result <- result |> stringr::str_remove(LEADING_ZERO_PATTERN)
   
   result
 }

### 2024-10-24 output/seminar-slideshow.qmd
Configure graphical output

--- 2024-10-24 output/seminar-slideshow.qmd
@@
 knitr:
-  opts_knit: 
-    root_dir:  here::here()
+  # opts_knit: # Does not work, see line 88
+  #   root_dir:  here::here()
   opts_chunk: 
     results:   asis
+    cache:     true
+    autodep:   true
 format:
@@
     fig-cap-location:        bottom
+    fig-format:              svg
+    fig-height:              6
+    fig-width:               6
     incremental:             false
@@
     link-external-newwindow: true
+    margin:                  0
     self-contained:          true
@@
 ```{r libraries}
+#| cache:   false
 #| include: false
@@
 library(knitr)
+library(lavaan)
+library(semPlot)
+library(extrafont)
 library(tidyr)
 library(ggplot2)
-library(lavaan)
-library(semPlot)
 ```
 
-```{r souces}
-source("../R/Output.R", encoding = 'UTF-8')
+```{r setup}
+#| cache:   false
+#| include: false
+opts_knit$set(root.dir = here::here()) # See lines 41-42
+
+# Graphical configuration options:
+loadfonts(device = "win") # Load fonts (if necessary)
+par(family = "Arial", adj = 0.5) # Set graphical device params
+
+margins <- c(12, 1.3, 8, 1.3) # margins for the sem path plots
 ```
@@
 ```{r read-chunks}
-read_chunk("../src/Graphical_example_paper.R")
+#| cache: false
+read_chunk("src/Graphical_example_paper.R")
+```
+
+```{r sources}
+#| cache: false
 ```
@@
 ```{r constants}
+#| cache: false
+```
+
+```{r replace-plot-params}
+GRAPH_FONT   <- "Arial" # Replace default font to comply with UNED identity
+FONT_SIZE    <- 24      # Replace default font size for slideshow
+AXIS_LAB_POS <-   .5    # Replace default axis label position for slideshow
+LINE_WIDTH   <-   .75   # Replace default line width for slideshow
 ```
@@
 ```{r graphical-output-conf}
+#| cache: false
 ```

### 2024-10-24 output/seminar-slideshow.qmd
Create simple structure first-slide graphics

@@
 
+  cat( "\n", "<br>", sep = '')
+
   af_tri_differences |> slice_head(n = slide) |> kable() |> print()
@@
 
+# Generate data for simple and complex factorial structures
 simple_item_loadings <- tribble(
@@
 
+```
+
+::::: columns
+::: column
+```{r path-simple-structure}
 # TODO: Syntax from tibble
@@
 "
-complex_spec <- "
-  F1 =~ I1 + I2 + I3
-  F2 =~ I3 + I4 + I5 + I6
 
-  F1 ~~ F2
-"
+F1_COLOR <- PALETTE_BLUE["normal"]
+F2_COLOR <- PALETTE_RASPBERRY["normal"]
+
+item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
+factor_colors <- c('F1' = unname(F1_COLOR), 'F2' = unname(F2_COLOR))
 
@@
   semPaths(
-    edge.color     = "black",
-    residuals      = FALSE,
-    thresholds     = FALSE,
-    intercepts     = FALSE,
-    fixedStyle     = 1,
-    whatLabels     = "std"
-  )
-
-simple_item_loadings |>
-  pivot_wider(names_from = factor, values_from = loading) |>
-  ggplot(mapping = aes(F1, F2)) +
-  geom_point(size = I(4))
-
-complex_spec     |>
-  semPlotModel() |>
-  semPaths(
+    border.color = c(
+      item_colors,       # Item colors
+      F1_COLOR, F2_COLOR # Factor colors
+    ),
     edge.color   = c(
-      "black" |> rep(3),
-      PALETTE_RASPBERRY["normal"], # cross-loading
-      "black" |> rep(3)
+      item_colors, # Loading colors
+      "black"      # Correlation color
     ),
@@
     fixedStyle   = 1,
-    whatLabels     = "std",
+    curve        = 3,
+    label.cex    = 1.6,
+    sizeMan      = 10,
+    sizeLat      = 14,
+    border.width = 3,
+    edge.width   = 3,
+    mar          = margins
   )
 ```
+:::
+
+::: {.column .fragment}
+```{r scatter-simple-structure}
+LIM_INF <- -1.5
+LIM_SUP <-  2
+
+axis_lims <- c(LIM_INF, LIM_SUP)
+lim_range <- LIM_SUP - LIM_INF
+axis_pos  <- (-LIM_INF) / lim_range
+
+item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
+
+simple_item_loadings                                      |>
+  group_by(item)                                          |>
+  arrange(desc(loading))                                  |>
+  mutate(main_factor = first(factor))                     |>
+  pivot_wider(names_from = factor, values_from = loading) |>
+  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
+  geom_point(size = I(4))                                          +
+  geom_text(
+    nudge_x =  .1,
+    nudge_y = -.08,
+    color   = I("black"),
+    size    = I(5)
+  )                                                                +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F2",
+  )                                                                +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F1"
+  )                                                                +
+  scale_color_manual(values = factor_colors, guide = NULL)         +
+  coord_fixed(expand = FALSE, clip = "on")                         +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_text(hjust = axis_pos),
+    axis.title.y       = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = LINE_WIDTH
+    ),
+    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
+    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+  )
+```
+:::
+:::::
 
@@
 Los pesos factoriales en los demás factores son nulos.
+
 Si representamos los pesos factoriales como coordenadas, el ítem estará alineado con el eje correspondiente al factor en el que pesa.

### 2024-10-24 output/seminar-slideshow.qmd
Create complex items data frame

@@
 )
-complex_item_loadings <- simple_item_loadings                                |>
-  bind_rows(tibble(item = 'I3', factor = 'F2', loading = ITEM_BASE_LOADING)) |>
-  complete(item, factor, fill = list(loading = 0))
+complex_item_loadings <- simple_item_loadings |> bind_rows(
+  tibble(
+    item    = 'I3',
+    factor  = 'F2',
+    loading = ITEM_BASE_LOADING - 2 * ITEM_DIST # Secondary loading smaller
+  )
+)
 
@@
 
+complex_item_loadings <- complex_item_loadings |>
+  complete(item, factor, fill = list(loading = 0))
 ```

### 2024-10-24 output/seminar-slideshow.qmd
Add complex structure plots

@@
 
-<!--# TODO: Ejemplo con pesos cruzados: Path diagram vs. coordenadas -->
+::::: columns
+::: column
+```{r path-complex-structure}
+# TODO: Syntax from tibble
+complex_spec <- "
+  F1 =~ I1 + I2 + I3
+  F2 =~ I3 + I4 + I5 + I6
+
+  F1 ~~ F2
+"
+
+edge_colors <- c(
+  F1_COLOR |> rep(3),
+  PALETTE_RASPBERRY["light"], # Cross-loading
+  F2_COLOR |> rep(3)
+)
+
+complex_spec     |>
+  semPlotModel() |>
+  semPaths(
+    border.color = c(
+      item_colors,       # Item colors
+      F1_COLOR, F2_COLOR # Factor colors
+    ),
+    edge.color   = c(
+      edge_colors, # Loading colors
+      "black"      # Correlation color
+    ),
+    residuals    = FALSE,
+    thresholds   = FALSE,
+    intercepts   = FALSE,
+    fixedStyle   = 1,
+    curve        = 3,
+    label.cex    = 1.6,
+    sizeMan      = 10,
+    sizeLat      = 14,
+    border.width = 3,
+    edge.width   = 3,
+    mar          = margins
+  )
+```
+:::
+
+::: {.column .fragment}
+```{r scatter-complex-structure}
+LIM_INF <- -1.5
+LIM_SUP <-  2
+
+axis_lims <- c(LIM_INF, LIM_SUP)
+lim_range <- LIM_SUP - LIM_INF
+axis_pos  <- (-LIM_INF) / lim_range
+
+item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
+
+complex_item_loadings_wide <- complex_item_loadings |>
+  group_by(item)                                    |>
+  arrange(desc(loading))                            |>
+  mutate(main_factor = first(factor))               |>
+  ungroup()                                         |>
+  pivot_wider(names_from = factor, values_from = loading)
+
+item3_spikes <- complex_item_loadings_wide      |>
+  filter(item == 'I3')                          |>
+  select(starts_with('F'))                      |>
+  add_column(type = "end")                      |> 
+  bind_rows(tibble(F1 = 0, F2 = 0))             |>
+  complete(F1, F2, fill = list(type = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+item3_spikes <- item3_spikes                         |> 
+  bind_rows(item3_spikes |> slice(3))                |>
+  mutate(n = (row_number() + 1) %% 2 |> as.factor()) |>
+  pivot_wider(names_from = type, values_from = F1:F2)
+
+spike_colors = c(
+  '1' = unname(PALETTE_BLUE['light']),
+  '0' = unname(PALETTE_RASPBERRY['light'])
+)
+
+complex_item_loadings_wide |>
+  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
+  geom_segment(
+    data    = item3_spikes,
+    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = spike_colors
+  )                                                                +
+  geom_point(size = I(4))                                          +
+  geom_text(
+    nudge_x =  .1,
+    nudge_y = -.08,
+    color   = I("black"),
+    size    = I(5)
+  )                                                                +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F2",
+  )                                                                +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F1"
+  )                                                                +
+  scale_color_manual(values = factor_colors, guide = NULL)         +
+  coord_fixed(expand = FALSE, clip = "on")                         +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_text(hjust = axis_pos),
+    axis.title.y       = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = LINE_WIDTH
+    ),
+    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
+    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+  )
+```
+:::
+:::::
 
@@
 Es cuando aparecen lo que llamamos "pesos cruzados".
+
 Esto da lugar a una "estructura compleja", que en realidad es más común que la estructura simple, en general.

### 2024-10-24 output/seminar-slideshow.qmd
Add oblique rotation example

@@
     fig-format:              svg
+    fig-asp:                 1
     fig-height:              6
@@
 library(ggplot2)
+library(units)
 ```
@@
 
-```{r sources}
-#| cache: false
-```
-
-```{r constants}
-#| cache: false
+```{r sources, cache=FALSE}
+```
+
+```{r constants, cache=FALSE}
 ```
@@
 LINE_WIDTH   <-   .75      # Replace default line width for slideshow
-```
-
-```{r graphical-output-conf}
-#| cache: false
+AXIS_COLOR   <- "darkgray" # Replace default grid line color for slideshow
+```
+
+```{r graphical-output-conf, cache=FALSE}
 ```
@@
 
-<!--# TODO: Ejemplos estructura simple vs. compleja: Path diagram vs. coordenadas -->
+<br>
 
@@
 
-simple_item_loadings                                      |>
+simple_item_loadings_wide <- simple_item_loadings |>
   group_by(item)                                  |>
@@
   mutate(main_factor = first(factor))             |>
-  pivot_wider(names_from = factor, values_from = loading) |>
+  pivot_wider(names_from = factor, values_from = loading)
+
+simple_item_loadings_wide |>
   ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
@@
 Es decir, asumimos que la varianza de cada ítem está explicada solamente por un factor latente.
-Los pesos factoriales en los demás factores son nulos.
+Los pesos factoriales de un ítem en los demás factores son nulos.
 
@@
 
-<!--# TODO: Ejemplo mostrando solución factorial sin rotar y ejes tras rotación oblicua? -->
-
-::: notes
-Para obtener esa estructura simple, muchas veces necesitamos usar una rotación oblicua.
+::: r-center
+```{r simple-structure-rotated}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+AXIS1_DEGREES <- -25
+AXIS2_DEGREES <-  30
+
+axis1_radians <- AXIS1_DEGREES |> as_units("degrees") |> set_units("radians")
+axis2_radians <- AXIS2_DEGREES |> as_units("degrees") |> set_units("radians")
+
+rot_matrix <- c(
+  cos(axis1_radians), cos(axis2_radians),
+  sin(axis1_radians), sin(axis2_radians)
+) |>
+  matrix(nrow = 2, byrow = TRUE)
+
+simple_item_loadings_rot <- simple_item_loadings |>
+  group_by(item) |>
+  mutate(
+    loading = rot_matrix %*% loading |> drop()
+  ) |>
+  left_join(simple_item_loadings_wide |> select(item, main_factor), by = "item")
+
+simple_item_loadings_rot_plot <- simple_item_loadings_rot |>
+  group_by(item)                                          |>
+  pivot_wider(names_from = factor, values_from = loading) |>
+  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
+  geom_point(size = I(4))                                          +
+  geom_text(
+    nudge_x = -.12,
+    # nudge_y = -.08,
+    color   = I("black"),
+    size    = I(5)
+  )                                                                +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "",
+  )                                                                +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = ""
+  )                                                                +
+  scale_color_manual(values = factor_colors, guide = NULL)         +
+  coord_fixed(expand = FALSE, clip = "on")                         +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_text(hjust = axis_pos),
+    axis.title.y       = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = LINE_WIDTH
+    )
+  )
+
+simple_item_loadings_rot_plot
+```
+:::
+
+::: notes
+Ahora bien, la mayoría de procedimientos factoriales, exploratorios al menos, arrojan soluciones de este tipo: un primer factor que maximiza toda la varianza posible, el segundo (ortogonal al primeroo) maximiza la varianza restante, y si hay más, hacen lo mismo, dando ejes siempre ortogonales a los anteriores.
+:::
+
+## AF: Estructura simple
+
+::: r-center
+```{r simple-structure-rotated-axes}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+rotated_axes <- tibble(
+  slope  = c(axis1_radians, axis2_radians) |> tan() |> drop_units(),
+  factor = c('F1', 'F2')
+)
+
+simple_item_loadings_rot_plot +
+  geom_abline(
+    data    = rotated_axes,
+    mapping = aes(slope = slope, intercept = 0, color = factor)
+  )
+```
+:::
+
+::: notes
+Para obtener esa estructura simple, muchas veces necesitamos usar una rotación oblicua de los ejes.
+Cualquier procedimiento de rotación busca alinearse lo más posible con al menos algunos de los ítems, como vemos aquí.
+Pero esto es una situación ideal.
+Nunca van a quedar, en un modelo exploratorio, los ítems perfectamente alineados con un eje, dando cero exacto en los pesos de los factores que no le corresponden.
+En un modelo confirmatorio sí, forzamos que sea así, pero a costa de perder ajuste claro (y normalmente permitiendo también que los factores correlacionen).
 :::
@@
 ## AF: Estructura compleja
+
+<br>
 

### 2024-10-24 output/seminar-slideshow.qmd
Add method factor example

@@
 
-<!--# TODO: Ejemplo con factores de método: Path diagram vs. coordenadas -->
-
-::: notes
-Por ejemplo, si tenemos ítems directos e inversos, podemos modelar lo que llamamos "factores de método": Una variable latente para cada factor sustantivo, y una para cada uno de los tipos de ítem.
+<br>
+
+::::: columns
+::: column
+```{r path-method-factor}
+# TODO: Syntax from tibble
+method_spec <- "
+  F1 =~ I1 + I2 + I3 + I4 + I5 + I6
+  M1 =~ I4 + I5 + I6
+"
+
+edge_colors <- c(
+  F1_COLOR |> rep(6),
+  F2_COLOR |> rep(3)
+)
+
+method_spec      |>
+  semPlotModel() |>
+  semPaths(
+    border.color = c(
+      F1_COLOR |> rep(6), # Item colors
+      F1_COLOR, F2_COLOR  # Factor colors
+    ),
+    edge.color   = edge_colors, # Loading colors
+    residuals    = FALSE,
+    thresholds   = FALSE,
+    intercepts   = FALSE,
+    fixedStyle   = 1,
+    curve        = 3,
+    label.cex    = 1.6,
+    sizeMan      = 10,
+    sizeLat      = 14,
+    border.width = 3,
+    edge.width   = 3,
+    mar          = margins
+  )
+```
+:::
+
+::: column
+```{r scatter-method-factor}
+LIM_INF <- -1.8
+LIM_SUP <-  1.8
+
+axis_lims <- c(LIM_INF, LIM_SUP)
+lim_range <- LIM_SUP - LIM_INF
+axis_pos  <- (-LIM_INF) / lim_range
+
+method_item_loadings <- tribble(
+  ~item, ~factor, ~loading,
+  'I1',  'F1',     ITEM_BASE_LOADING,
+  'I1',  'M1',     0,
+  'I2',  'F1',     ITEM_BASE_LOADING + ITEM_DIST,
+  'I2',  'M1',     0,
+  'I3',  'F1',     ITEM_BASE_LOADING + 2 * ITEM_DIST,
+  'I3',  'M1',     0,
+  'I4',  'F1',    -ITEM_BASE_LOADING,
+  'I4',  'M1',     ITEM_BASE_LOADING - 2 * ITEM_DIST,
+  'I5',  'F1',     ITEM_BASE_LOADING + ITEM_DIST,
+  'I5',  'M1',    -ITEM_BASE_LOADING + 2 * ITEM_DIST,
+  'I6',  'F1',    -ITEM_BASE_LOADING - ITEM_DIST,
+  'I6',  'M1',    -ITEM_BASE_LOADING + 2 * ITEM_DIST
+)
+
+method_item_loadings                                      |>
+  pivot_wider(names_from = factor, values_from = loading) |>
+  ggplot(mapping = aes(F1, M1, label = item))              +
+  geom_point(size = I(4), color = F1_COLOR)                +
+  geom_text(
+    # nudge_x =  .1,
+    nudge_y = -.12,
+    color   = I("black"),
+    size    = I(5)
+  )                                                        +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "M1",
+  )                                                        +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F1"
+  )                                                        +
+  scale_color_manual(values = factor_colors, guide = NULL) +
+  coord_fixed(expand = FALSE, clip = "on")                 +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_text(hjust = axis_pos),
+    axis.title.y       = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = LINE_WIDTH
+    ),
+    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
+    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+  )
+```
+:::
+:::::
+
+::: notes
+Por ejemplo, si tenemos ítems directos e inversos, podemos modelar lo que llamamos "factores de método": Una variable latente para cada factor sustantivo (sólo uno en este caso), y por ejemplo uno para los ítems formulados en negativo.
 :::

### 2024-10-25 output/seminar-slideshow.qmd
Adjust item distances

@@
   'I2',  'F1',    ITEM_BASE_LOADING + ITEM_DIST,
-  'I3',  'F1',    ITEM_BASE_LOADING + 2 * ITEM_DIST,
-  'I4',  'F2',    ITEM_BASE_LOADING,
+  'I3',  'F1',    ITEM_BASE_LOADING + 3 * ITEM_DIST,
+  'I4',  'F2',    ITEM_BASE_LOADING - ITEM_DIST,
   'I5',  'F2',    ITEM_BASE_LOADING + ITEM_DIST,
@@
 ```{r scatter-method-factor}
-LIM_INF <- -1.8
-LIM_SUP <-  1.8
-
-axis_lims <- c(LIM_INF, LIM_SUP)
-lim_range <- LIM_SUP - LIM_INF
-axis_pos  <- (-LIM_INF) / lim_range
-
 method_item_loadings <- tribble(
@@
   'I1',  'M1',     0,
-  'I2',  'F1',     ITEM_BASE_LOADING + ITEM_DIST,
+  'I2',  'F1',     ITEM_BASE_LOADING + 2 * ITEM_DIST,
   'I2',  'M1',     0,
-  'I3',  'F1',     ITEM_BASE_LOADING + 2 * ITEM_DIST,
+  'I3',  'F1',     ITEM_BASE_LOADING + 3 * ITEM_DIST,
   'I3',  'M1',     0,
@@
   'I6',  'F1',    -ITEM_BASE_LOADING - ITEM_DIST,
-  'I6',  'M1',    -ITEM_BASE_LOADING + 2 * ITEM_DIST
+  'I6',  'M1',    -ITEM_BASE_LOADING + 3 * ITEM_DIST
 )

### 2024-10-25 output/seminar-slideshow.qmd
Embed graphical configuration in slideshow file
Avoid dependency (in principle) from the graphical representation example for the paper

@@
 par(family = "Arial", adj = 0.5) # Set graphical device params
-
-margins <- c(12, 1.3, 8, 1.3) # margins for the sem path plots
+```
+
+```{r sources}
+source("R/LaTeX_math.R", encoding = 'UTF-8')
+source("R/Output.R",     encoding = 'UTF-8')
 ```
@@
 ```{r color-palettes}
+MARGINS <- c(12, 1.3, 8, 1.3) # margins for the sem path plots
+
 PALETTE_UNED <- c(
@@
 )
-```
-
-```{r read-chunks}
-#| cache: false
-read_chunk("src/Graphical_example_paper.R")
-```
-
-```{r sources, cache=FALSE}
-```
-
-```{r constants, cache=FALSE}
-```
-
-```{r replace-plot-params}
+
 GRAPH_FONT   <- "Arial"    # Replace default font to comply with UNED identity
@@
 
-```{r graphical-output-conf, cache=FALSE}
+```{r graphical-output-conf}
+theme_set( # `ggplot` output configuration
+  theme_classic(
+    base_size      = FONT_SIZE,
+    base_family    = GRAPH_FONT,
+    base_line_size = LINE_WIDTH
+  ) %+replace%
+    theme(
+      axis.title.x = element_text(hjust = AXIS_LAB_POS),
+      axis.title.y = element_text(vjust = AXIS_LAB_POS, angle = 0),
+      panel.grid   = element_line(color = AXIS_COLOR)
+    )
+)
 ```
@@
     edge.width   = 3,
-    mar          = margins
+    mar          = MARGINS
   )

### 2024-10-25 output/seminar-slideshow.qmd
Add "classic factor analysis" slides

@@
 library(units)
+library(ggtext)
 ```
@@
 ```{r graphical-output-conf}
+#| cache: false
 theme_set( # `ggplot` output configuration
@@
       axis.title.y     = element_text(vjust = AXIS_LAB_POS, angle = 0),
-      panel.grid   = element_line(color = AXIS_COLOR)
+      panel.grid       = element_line(color = AXIS_COLOR),
+      panel.grid.minor = element_line(linetype = "17", color = "grey70")
     )
@@
 
+<!-- # TODO: Añadir ángulos? -->
+
 ::: r-center
@@
   bind_rows(item3_spikes |> slice(3))                  |>
-  mutate(n = (row_number() + 1) %% 2 |> as.factor()) |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
   pivot_wider(names_from = type, values_from = F1:F2)
@@
 spike_colors = c(
-  '1' = unname(PALETTE_BLUE['light']),
-  '0' = unname(PALETTE_RASPBERRY['light'])
+  'F1' = unname(PALETTE_BLUE['light']),
+  'F2' = unname(PALETTE_RASPBERRY['light'])
 )
@@
 
-<!--# TODO: Fórmula del modelo factorial -->
-
-<!--# TODO: Fórmula matricial para una variable observable -->
+<br>
+
+$$
+I_1 = \tau_1 + \lambda_{11} F_1^p + \lambda_{21} F_2^p + ... + \lambda_{n1} F_n^p + U_1^p \\
+\vdots                                                                            \\
+I_k = \tau_k + \lambda_{1k} F_1^p + \lambda_{2k} F_2^p + ... + \lambda_{nk} F_n^p + U_k^p
+$$
+
+<br>
+
+::: fragment
+$$
+I_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
+$$
+:::
 
@@
 Las variables "respuesta" o "criterio" es la variable observable, y su varianza está explicada o "predicha" por las variables latentes, que son los factores y los términos de error de cada ítem.
-
-Los "coeficientes de regresión" los llamamos "pesos factoriales" y el término error, "unicidad" o "varianza única".
-Esto se puede expresar en "forma matricial" de esta manera.
-
-<!--# TODO: Explicar cómo se calcula en forma matricial? -->
-
-Tenemos costumbre de ver el análisis factorial en el contexto de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
-Sin embargo, se aplica también (e idealmente se debería aplicar) a variables discretas ordinales.
+Lo que ocurre es que aquí a "coeficientes de regresión" los llamamos "pesos factoriales" y al término error, "unicidad" o "varianza única".
+
+El producto de los pesos por las puntuaciones factoriales se puede expresar en "forma vectorial", de esta manera.
+La "T" del vector de pesos factoriales significa que está "traspuesto", es decir, que se pone horizontalmente (esto es una convención en álgebra matricial).
+El producto del vector de pesos por el de puntuaciones factoriales es lo que se llama un "producto vectorial", que consiste simplemente en multiplicar dos a dos los términos de cada vector y sumar todos.
+:::
+
+## Modelo factorial clásico {.smaller}
+
+::::::: columns
+::::: column
+$$
+  I_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
+$$
+
+$$
+  \mathbf{F}^1 =
+    \begin{bmatrix}
+      0.5 \\
+      1
+    \end{bmatrix}
+$$
+
+::: {.fragment fragment-index="1"}
+$$
+  \mathbf{\lambda}_{1} =
+    \begin{bmatrix}
+      -1 \\
+       0.5
+    \end{bmatrix}
+$$
+:::
+
+::: fragment
+$$
+  \begin{align}
+    \tau_1 &=  1 \\
+    U_1^1  &= -0.5
+  \end{align}
+$$
+:::
+:::::
+
+::: {.column .fragment fragment-index="2"}
+```{r fa_example_plot}
+#| cache: false
+AXIS_MINOR <- c(-1, 1, 2)
+
+# Generate data for simple and complex factorial structures
+fa_example <- tribble(
+  ~object, ~factor, ~loading,
+  'I~1~',  'F1',    -1,
+  'I~1~',  'F2',     0.5,
+  'F^1^',  'F1',     0.5,
+  'F^1^',  'F2',     1
+)
+
+fa_example_wide <- fa_example |>
+  pivot_wider(names_from = factor, values_from = loading)
+
+fa_example_spikes <- fa_example_wide |>
+  add_column(type = "end")           |> 
+  bind_rows(
+    tribble(
+      ~object, ~F1,  ~F2, ~type,
+      "I~1~",   0,   0.5, "start",
+      "F^1^",   0,   1,   "start",
+      "I~1~",  -1,   0,   "start",
+      "F^1^",   0.5, 0,   "start",
+    )
+  )
+
+fa_example_spikes <- fa_example_spikes |> slice(1:2)        |>
+  bind_rows(fa_example_spikes)                              |>
+  add_column(factor = paste0('F', 1:2 |> rep(2, each = 2))) |>
+  pivot_wider(names_from = type, values_from = F1:F2)
+
+fa_example_wide |>
+  ggplot(mapping = aes(F1, F2, label = object))            +
+  geom_segment(
+    data        = fa_example_spikes,
+    mapping     = aes(
+      F1_start, F2_start,
+      xend  = F1_end, yend = F2_end
+    ),
+    inherit.aes = FALSE,
+    linewidth   = LINE_WIDTH,
+    linetype    = '43',
+    color       = spike_colors |> rep(each = 2)
+  )                                                        +
+  geom_point(size = I(4))                                  +
+  geom_richtext(
+    nudge_x = .15,
+    nudge_y = .10,
+    color   = I("black"),
+    size    = I(5),
+    lineheight = 1.2,
+    fill = NA,
+    label.color = NA
+  )                                                        +
+  scale_x_continuous(
+    limits       = axis_lims,
+    breaks       = 0,
+    labels       = NULL,
+    name         = "F~2~",
+    minor_breaks = AXIS_MINOR
+  )                                                        +
+  scale_y_continuous(
+    limits       = axis_lims,
+    breaks       = 0,
+    labels       = NULL,
+    name         = "F~1~",
+    minor_breaks = AXIS_MINOR
+  )                                                        +
+  scale_color_manual(values = factor_colors, guide = NULL) +
+  coord_fixed(expand = FALSE, clip = "on")                 +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_markdown(hjust = axis_pos),
+    axis.title.y       = element_markdown(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = LINE_WIDTH
+    ),
+    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
+    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+  )
+```
+:::
+:::::::
+
+::: fragment
+$$
+  I_1 = 1 + \left[ -1, 0.5 \right]
+    \begin{bmatrix}
+      0.5 \\
+      1
+    \end{bmatrix}
+    - 0.5 = 1 + (-0.5 + 0.5) - 0.5 = 0.5
+$$
+:::
+
+::: notes
+Veámoslo con un ejemplo.
+Tenemos el participante 1, en nuestro estudio, que tiene esas puntuaciones factoriales.
+Por ejemplo, en el primer rasgo, tiene un 0.5 (nos da igual la distribución pero supongamos que está estandarizada, luego está media desviación típica por encima de la media).
+En el segundo rasgo, tiene una puntuación que está una desviación típica por encima de la media.
+
+Los pesos factoriales del ítem 1 son estos: -1 para el primer rasgo (por lo cual, cuanto mayor sea la puntuación factorial en el primer factor, menor va a ser la respuesta observada en el ítem 1), y 0.5 para el segundo.
+
+Aquí vemos representados el ítem y la persona, en las mismas coordenadas cartesianas.
+Y luego el ítem tiene una intersección de 1, y la persona 1 tiene un -0.5 en el "error" de esa variable $I_1$.
+
+Su respuesta se calcularía así: (hacer los cálculos).
+Luego esta persona (el particpante 1) tiene la respuesta o valor observado 0.5 en esta variable observada.
+Lo mismo podríamos hacer con el resto de variables y de participantes, pero obviamente nosotros lo que tenemos son estas respuestas observadas, y lo que hacemos con el AF es estimar todos estos parámetros del modelo, en lugar de como lo hemos hecho ahora.
+
+Ahora bien, esta respuesta observada que hemos calculado aquí es cuantitativa.
+Tenemos costumbre de ver el análisis factorial en ese contexto, de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
+Sin embargo, se aplica también a variables discretas ordinales (e idealmente se debería aplicar así a ítems tipo Likert, aunque a veces no se haga, por simplificar).
+
 En este caso vamos a concretarlo en variables dicotómicas (p.ej. ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).

### 2024-10-25 output/seminar-slideshow.qmd
Add item factor analysis slide

@@
 library(ggtext)
+library(ggridges)
+library(patchwork)
 ```
@@
 Veámoslo con un ejemplo.
-Tenemos el participante 1, en nuestro estudio, que tiene esas puntuaciones factoriales.
+Tenemos el participante 1, en nuestro estudio, que tiene esas puntuaciones factoriales; usamos el superíndice para denotar al participante.
 Por ejemplo, en el primer rasgo, tiene un 0.5 (nos da igual la distribución pero supongamos que está estandarizada, luego está media desviación típica por encima de la media).
@@
 Tenemos costumbre de ver el análisis factorial en ese contexto, de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
-Sin embargo, se aplica también a variables discretas ordinales (e idealmente se debería aplicar así a ítems tipo Likert, aunque a veces no se haga, por simplificar).
-
-En este caso vamos a concretarlo en variables dicotómicas (p.ej. ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
+Sin embargo, el llamado "AF de ítems" se aplica también a variables discretas ordinales (e idealmente se debería aplicar así a ítems tipo Likert, aunque a veces no se haga, por simplificar).
+:::
+
+## Análisis factorial de ítems
+
+:::::: columns
+:::: column
+$$
+  i_k^* = \tau_k + \mathbf{\lambda}_k^T \mathbf{F} + U_k
+$$
+
+<br>
+
+::: fragment
+$$
+  I_k = \left\{ \begin{matrix}
+      1 &si &i_k^* \geq 0\\
+      0 &si &i_k^* <    0
+    \end{matrix} \right.
+$$
+:::
+::::
+
+::: column
+```{r ifa-curve}
+#| fig-height: 9
+
+PROB_AXIS      <- seq(-3, 3, by = .01)
+ERROR_POINTS   <- seq(-3, 3, length.out = 5)
+ITEM_LOADING   <-  0.35
+ITEM_INTERCEPT <- -0.5
+
+item_unique_var <-  1 - sqrt(ITEM_LOADING)
+
+probit_curve_data <- tibble(
+  score    = PROB_AXIS,
+  response = ITEM_INTERCEPT + ITEM_LOADING * score,
+  prob     = pnorm(response, sd = sqrt(item_unique_var))
+)
+
+latent_response <- tibble(
+  point    = ERROR_POINTS |> rep(100000),
+  response = rnorm(
+    length(point),
+    mean = ITEM_INTERCEPT + ITEM_LOADING * point,
+    sd   = sqrt(item_unique_var)
+  )
+)
+
+# Plot: Upper Panel - Linear regression with Gaussian error distributions
+latent_response_plot <- latent_response |>
+  ggplot(aes(x = response, y = factor(point), fill = ..x.. > 0))           +
+  geom_density_ridges_gradient(
+    scale          = .2,
+    rel_min_height = 0.001,
+    color          = PALETTE_BLUE["normal"]
+  )                                                                        +
+  labs(x = "i~k~^*^", y = NULL)                                            +
+  scale_y_discrete(labels = ERROR_POINTS)                                  +
+  scale_x_continuous(breaks = 0)                                           +
+  scale_fill_manual(
+    values = c(`FALSE` = "white", `TRUE` = unname(PALETTE_BLUE["white"]))) +
+  theme(
+    legend.position = "none",
+    axis.title.y = element_markdown(angle = 90),
+    panel.grid.major.y = element_line(linetype = "17")
+  )                                                                        +
+  coord_flip()
+
+# Plot: Lower Panel - Probit curve
+probit_plot <- probit_curve_data |>
+  ggplot(mapping = aes(x = score, y = prob))                          +
+  geom_line(color = PALETTE_BLUE["normal"], linewidth = LINE_WIDTH)   +
+  labs(x = "i~k~^*^", y = "P(I~k = 1)")                               +
+  scale_y_continuous(breaks = 0:1, minor_breaks = NULL, limits = 0:1) +
+  scale_x_continuous(minor_breaks = ERROR_POINTS, breaks = -1:1 * 3)  +
+  theme(
+    axis.title   = element_markdown(),
+    axis.title.y = element_markdown(angle = 90),
+  )
+
+## TODO: Correct error break points
+latent_response_plot / probit_plot
+```
+:::
+::::::
+
+::: notes
+Lo que hace el análisis factorial de ítems es aplicar ese "modelo lineal" a variables ordinales, haciendo una generalización.
+(En este caso vamos a simplificarlo a variables dicotómicas; p.ej. ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
 Esto es una simplificación que rara vez se utiliza en cuestionarios no cognitivos, pero es interesante estudiarlo porque otros modelos pueden considerarse extensiones de este modelo.
-:::
-
-## Análisis factorial de ítems
-
-<!--# TODO: Fórmula del modelo factorial con función de enlace probit -->
-
-<!--# TODO: Función de enlace "normal acumulada" -->
-
-<!--# TODO: Curva Gaussiana y curva normal acumulada -->
-
-::: notes
-El análisis factorial de ítems (dicotómicos) lo que hace es aplicar ese "modelo lineal general" a variables dicotómicas: Asume una "variable latente" lineal, con la misma fórmula anterior, y aplica una "función de umbral" en 0 (la respuesta es correcta o afirmativa, codificada como 1, si la variable latente es mayor que 0, o incorrecta o negativa, codificada como 0, si es menor).
-Eso da una "probabilidad" de responder correcta o "afirmativamente" a un ítem.
+El AF de ítems asume una nueva "variable latente" lineal ($I_1^*$), con la misma fórmula anterior
+
+Después se aplica una "función de umbral": La respuesta es correcta o afirmativa, codificada como 1, si la variable latente es mayor (o igual) que 0, o incorrecta o negativa, codificada como 0, si es menor.
+Eso lo que da, en lugar de un respuesta observada (cuantitativa), es una "probabilidad" de responder correcta o "afirmativamente" a un ítem.
 :::

### 2024-10-27 output/seminar-slideshow.qmd
Complete "Item Factor Analysis" slide

@@
     fig-asp:                 1
-    fig-height:              6
     fig-width:               6
@@
 :::: column
+<br>
+
 $$
@@
 
-::: fragment
+::: {.fragment fragment-index="1"}
 $$
@@
 
-::: column
+::: {.column .fragment fragment-index="1"}
 ```{r ifa-curve}
-#| fig-height: 9
+#| fig-asp: 1.4
+#| cache: false
 
@@
 latent_response <- tibble(
-  point    = ERROR_POINTS |> rep(100000),
+  point    = ERROR_POINTS |> rep(500000),
   response = rnorm(
@@
 latent_response_plot <- latent_response |>
-  ggplot(aes(x = response, y = factor(point), fill = ..x.. > 0))           +
+  ggplot(aes(x = response, y = point, fill = ..x.. > 0, group = point)) +
   geom_density_ridges_gradient(
@@
   labs(x = "i~k~^*^", y = NULL)                                         +
-  scale_y_discrete(labels = ERROR_POINTS)                                  +
+  scale_y_continuous(
+    breaks       = ERROR_POINTS,
+    minor_breaks = NULL,
+    labels       = NULL,
+    limits       = range(ERROR_POINTS) * 1.12,
+    expand       = expansion()
+    )           +
   scale_x_continuous(breaks = 0)                                        +
   scale_fill_manual(
-    values = c(`FALSE` = "white", `TRUE` = unname(PALETTE_BLUE["white"]))) +
+    values = c(
+      `FALSE` = unname(PALETTE_BLUE["white"]),
+      `TRUE`  = unname(PALETTE_BLUE["normal"])
+    )
+  ) +
   theme(
+    axis.ticks.x = element_blank(),
+    axis.title.y = element_markdown(angle = 90),
     legend.position = "none",
-    axis.title.y = element_markdown(angle = 90),
-    panel.grid.major.y = element_line(linetype = "17")
+    panel.grid.major = element_line(linetype = "17")
   )                                                                     +
@@
   geom_line(color = PALETTE_BLUE["normal"], linewidth = LINE_WIDTH) +
-  labs(x = "i~k~^*^", y = "P(I~k = 1)")                               +
-  scale_y_continuous(breaks = 0:1, minor_breaks = NULL, limits = 0:1) +
-  scale_x_continuous(minor_breaks = ERROR_POINTS, breaks = -1:1 * 3)  +
+  labs(x = "E(*i~k~^*^*)", y = "P(I~k~ = 1)")                       +
+  scale_y_continuous(
+    breaks       = 0:1,
+    minor_breaks = NULL,
+    limits       = 0:1,
+    expand       = expansion(mult = c(0, .05))
+  )  +
+  scale_x_continuous(
+    breaks       = ERROR_POINTS,
+    minor_breaks = NULL,
+    limits = range(ERROR_POINTS) * 1.12,
+    expand = expansion()
+  ) +
   theme(
@@
     axis.title.y     = element_markdown(angle = 90),
-  )
-
-## TODO: Correct error break points
+    panel.grid.major = element_line(linetype = "17"),
+  )
+
+## TODO: Correct error break points and fig height
 latent_response_plot / probit_plot

### 2024-10-27 output/seminar-slideshow.qmd
Add "factor scores" slides content (WiP)

@@
 
+<br>
+
+:::::: columns
+:::: column
 Métodos:
@@
 
+::: {.fragment fragment-index="2"}
 -   **Máxima Verosimilitud**
@@
 -   **Estimación Modal Bayesiana Empírica**
+:::
+::::
+
+::: {.column .fragment fragment-index="1"}
+```{r path-regression-scores}
+# TODO: Syntax from tibble
+simple_spec <- "
+  F1 <~ I1 + I2 + I3
+  F2 <~ I4 + I5 + I6
+
+  F1 ~~ F2
+"
+
+F1_COLOR <- PALETTE_BLUE["normal"]
+F2_COLOR <- PALETTE_RASPBERRY["normal"]
+
+item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
+factor_colors <- c('F1' = unname(F1_COLOR), 'F2' = unname(F2_COLOR))
+
+# TODO: Manual call to qgraph:
+Edgelist <- tribble(
+  ~lhs, ~rhs,
+  7,   1,
+  7,   2,
+  7,   3,
+  8,   4,
+  8,   5,
+  8,   6,
+  7,   8
+)
+nLab <- c("I1", "I2", "I3", "I4", "I5", "I6", "F1", "F2") |> setNames(nm = _)
+Bidir <- FALSE |> rep(6)
+Directed <- TRUE |> rep(6)
+Shape <- c("square", "square", "square", "square", "square", "square", 
+"circle", "circle")
+
+qgraph::qgraph(Edgelist, 
+      labels = nLab, bidirectional = Bidir, directed = Directed, 
+      shape = Shape, layout = Layout, lty = lty, loopRotation = loopRotation, 
+      curve = Curve, edge.labels = eLab, mar = Mar, vsize = vSize, 
+      vsize2 = vSize2, edge.color = eColor, groups = NodeGroups2, 
+      color = Vcolors, residuals = LoopAsResid, residScale = residScale, 
+      residEdge = isResid, edgelist = TRUE, curveDefault = curveDefault, 
+      knots = GroupPars$knot, aspect = layoutSplit, CircleEdgeEnd = CircleEdgeEnd, 
+      curveScale = curveScale, bars = bars, barSide = barSide, 
+      barColor = tColor, barLength = thresholdSize, barsAtSide = ThreshAtSide, 
+      edge.label.cex = edge.label.cex, edgeConnectPoints = ECP, 
+      ...)
+
+# simple_spec      |>
+#   semPlotModel() |>
+#   semPaths(
+#     border.color = c(
+#       item_colors,       # Item colors
+#       F1_COLOR, F2_COLOR # Factor colors
+#     ),
+#     edge.color   = c(
+#       item_colors, # Loading colors
+#       "black"      # Correlation color
+#     ),
+#     residuals    = FALSE,
+#     thresholds   = FALSE,
+#     intercepts   = FALSE,
+#     arrows       = TRUE,
+#     fixedStyle   = 1,
+#     curve        = 3,
+#     label.cex    = 1.6,
+#     sizeMan      = 10,
+#     sizeLat      = 14,
+#     border.width = 3,
+#     edge.width   = 3,
+#     mar          = MARGINS
+#   )
+```
+:::
+::::::
 
@@
 ::: notes
-Los dos primeros métodos se aplican en AF lineal y están implementados en SPSS.
+Una vez tenemos los parámetros del modelo factorial estimado, entre otras cosas, podemos aplicarlo para estimar las puntuaciones factoriales de los participantes.
+Para hacer esto hay varios métodos.
+
+El método de regresión y el de Bartlett se aplican en AF lineal y están implementados en SPSS.
 Consisten a grandes rasgos en calcular la puntuación factorial como una suma ponderada de las variables observables.

### 2024-10-28 output/seminar-slideshow.qmd
Add manual plot of "Factor regression scores"

--- 2024-10-27 output/seminar-slideshow.qmd
@@
 
-F1_COLOR <- PALETTE_BLUE["normal"]
-F2_COLOR <- PALETTE_RASPBERRY["normal"]
+F1_COLOR <- unname(PALETTE_BLUE["normal"])
+F2_COLOR <- unname(PALETTE_RASPBERRY["normal"])
 
 item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
-factor_colors <- c('F1' = unname(F1_COLOR), 'F2' = unname(F2_COLOR))
+factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
 
@@
 
+::: {.fragment fragment-index="2"}
 -   **Máxima Verosimilitud**
 
--   **Estimación Modal Bayesiana Empírica**
+-   **Empirical Modal Bayes**
+:::
+::::
+
+::: {.column .fragment fragment-index="1"}
+```{r path-regression-scores}
+# item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
+# factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
+
+# NOTE: Manual call to qgraph:
+edge_list <- tribble(
+  ~rhs, ~lhs,
+  7,    1,
+  7,    2,
+  7,    3,
+  8,    4,
+  8,    5,
+  8,    6,
+  7,    8,
+  8,    7
+)
+vSize <- c(10 |> rep(6), 14 |> rep(2))
+
+qgraph::qgraph(
+  edge_list,
+  labels = c("I1", "I2", "I3", "I4", "I5", "I6", "F1", "F2") |>
+    setNames(nm = _),
+  label.cex = 1.6,
+  shape = "square" |> rep(8),
+  vsize = vSize, 
+  vsize2 = vSize,
+  color = "background" |> rep(8),
+  border.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), F1_COLOR, F2_COLOR),
+  border.width = 3,
+  bidirectional = c(FALSE |> rep(6), TRUE, TRUE),
+  directed = TRUE |> rep(8),
+  curve = c(NA |> rep(6), 3, 3),
+  edge.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), "black"),
+  edge.width = 3,
+  lty = 1,
+  layout = matrix(
+    c(
+      -1, -0.6, -0.2, 0.2, 0.6, 1, -0.6, 0.6,
+      -0.5 |> rep(6), 0.5 |> rep(2)
+    ),
+    ncol = 2
+  ),
+  mar = c(12, 1.3, 8, 1.3)
+)
+```
+:::
+::::::
 

### 2024-10-28 output/seminar-slideshow.qmd
Include normal probability function in IFA slide

@@
 
-## Análisis factorial de ítems
+## Análisis factorial de ítems {.smaller}
 
 :::::: columns
-:::: column
-<br>
-
+:::: {.column width="40%"}
 $$
@@
 :::
+
+<br>
+
+::: {.fragment fragment-index="2"}
+$$
+  \begin{align}
+    P(I_k = 1) &= \int_{-\infty}^0 \mathcal{N}(i_k^*) \delta i_k^* \\
+               &= \Phi_N (i_k^*)
+  \end{align}
+$$
+:::
 ::::
 
-::: {.column .fragment fragment-index="1"}
-```{r ifa-curve}
-#| fig-asp: 1.4
-#| cache: false
+:::: {.column width="60%"}
+::: {.fragment fragment-index="1"}
+```{r ifa-density}
+#| fig-asp:   0.5
+#| fig-width: 8
 
@@
 ERROR_POINTS   <- seq(-3, 3, length.out = 5)
-ITEM_LOADING   <-  0.35
-ITEM_INTERCEPT <- -0.5
+ITEM_LOADING   <-  0.45
+ITEM_INTERCEPT <- -0.3
 
@@
 # Plot: Upper Panel - Linear regression with Gaussian error distributions
-latent_response_plot <- latent_response |>
-  ggplot(aes(x = response, y = point, fill = ..x.. > 0, group = point)) +
+latent_response |>
+  ggplot(
+    aes(x = response, y = point, fill = after_stat(x) > 0, group = point)
+  )                            +
   geom_density_ridges_gradient(
@@
   coord_flip()
+```
+:::
+
+::: {.fragment fragment-index="2"}
+```{r ifa-curve}
+#| fig-asp:   0.5
+#| fig-width: 8
 
 # Plot: Lower Panel - Probit curve
-probit_plot <- probit_curve_data |>
+probit_curve_data |>
   ggplot(mapping = aes(x = score, y = prob))                        +
@@
   )
-
-## TODO: Correct error break points and fig height
-latent_response_plot / probit_plot
-```
-:::
+```
+:::
+::::
 ::::::

### 2024-10-28 output/seminar-slideshow.qmd
Add 2-parameter logistic model slide

@@
 library(ggridges)
-library(patchwork)
+library(purrr)
 ```
@@
 
-:::::: columns
-:::: {.column width="40%"}
+::::::::: columns
+::::: {.column width="40%"}
 $$
@@
 :::
-::::
-
-:::: {.column width="60%"}
+:::::
+
+::::: {.column width="60%"}
 ::: {.fragment fragment-index="1"}
@@
 item_unique_var <-  1 - sqrt(ITEM_LOADING)
-
-probit_curve_data <- tibble(
-  score    = PROB_AXIS,
-  response = ITEM_INTERCEPT + ITEM_LOADING * score,
-  prob     = pnorm(response, sd = sqrt(item_unique_var))
-)
 
@@
 #| fig-width: 8
+
+probit_curve_data <- tibble(
+  score    = PROB_AXIS,
+  response = ITEM_INTERCEPT + ITEM_LOADING * score,
+  prob     = pnorm(response, sd = sqrt(item_unique_var))
+)
 
@@
 :::
-::::
-::::::
+:::::
+:::::::::
 
@@
 
+<br>
+
 ::::: columns
@@
 
-## Teoría de respuesta al ítem
-
-<!--# TODO: Fórmula de modelo 2PL -->
-
-<!--# TODO: Gráfica de discriminación alta/baja, dificultad alta/baja -->
-
-<!--# TODO: Fórmula del parámetro de intersección y del modelo en forma pendiente-intersección -->
+::: notes
+En este libro, que se publica en 1999 (fijaos en el subtítulo de "unified treatment"), McDonald indica que estos modelos, factoriales, TRI, son a fin de cuentas "isomóficos": Sin ser exactamente iguales ni partir de los mismos supuestos, las predicciones que hacen uno y otro son indistinguibles en la práctica.
+:::
+
+## Teoría de respuesta al ítem {.smaller}
+
+Modelo logístico de 2 parámetros [@birnbaum_latent_1968]:
+
+:::::::::: columns
+::::: {.column width="40%"}
+$$
+  P(I_k = 1) = \Phi_L (a_k (\theta - b_k))
+$$
+
+::: {.fragment fragment-index="1"}
+$$
+  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
+$$
+:::
+
+<br>
+
+::: {.fragment fragment-index="3"}
+
+$$
+  \begin{align}
+    a_k &= 1.7 \frac{\lambda_k}{\sqrt{var(U_k)}} \\
+    a_k b_k &= -\frac{\tau_k}{\lambda}
+  \end{align}
+$$
+:::
+:::::
+
+:::::: {.column .fragment width="60%" fragment-index="2"}
+```{r irt-curves}
+#| fig-asp:   0.6
+#| fig-width: 8
+
+ITEMS_PALETTE <- c(
+  PALETTE_BLUE["normal"],
+  PALETTE_RASPBERRY["normal"],
+  PALETTE_APPLE["normal"]
+) |>
+  unname() |>
+  setNames("prob" |> paste(1:3, sep = '_'))
+
+ITEMS <- tribble(
+  ~item, ~a,   ~b,
+   1,     1,    0,
+   3,     1,    1.5,
+   2,     2.5,  0
+)
+
+logit <- function(x) 1 / (1 + exp(-x))
+irf   <- function(a, b, theta) logit(a * (theta - b))
+
+logit_curve_data <- tibble(
+  latent_trait = PROB_AXIS,
+  prob         = ITEMS |>
+    select(-item) |>
+    pmap(irf, theta = latent_trait) |>
+    bind_cols()
+) |>
+  unnest(
+    prob,
+    names_sep = '_'
+  ) |>
+  rename_with(stringr::str_remove, pattern = '\\.{3}')
+
+logit_curve_data_long <- logit_curve_data |>
+  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
+
+logit_curve_data_long |>
+  ggplot(
+    mapping = aes(
+      latent_trait, prob,
+      color = item,
+      group = after_scale(color)
+    )
+  )                                        +
+  geom_line(linewidth = LINE_WIDTH)        +
+  labs(x = "*&theta;*", y = "P(I~k~ = 1)") +
+  scale_y_continuous(
+    breaks       = 0:1,
+    minor_breaks = NULL,
+    limits       = 0:1,
+    expand       = expansion(mult = c(0, .05))
+  )                                        +
+  scale_x_continuous(
+    breaks       = ERROR_POINTS,
+    minor_breaks = NULL,
+    limits = range(ERROR_POINTS) * 1.12,
+    expand = expansion()
+  )                                        +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  theme(
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(angle = 90),
+    panel.grid.major = element_line(linetype = "17"),
+  )
+```
+
+::: r-stack
+[$a_1$ = 1.0; $b_1$ = 1.0]{style="color:#5c6eb1;"}
+:::
+
+::: r-stack
+[$a_2$ = 1.0; $b_2$ = 1.5]{style="color:#90214a;"}
+:::
+
+::: r-stack
+[$a_3$ = 2.5; $b_3$ = 0.0]{style="color:#749f4c;"}
+:::
+::::::
+::::::::::
 

### 2024-10-28 output/seminar-slideshow.qmd
Correct 2-PL model slide style and formulas
Add interactive 3D IRS in plotly

@@
 library(ggridges)
+library(qgraph)
 library(purrr)
+library(plotly)
 ```
@@
 
-qgraph::qgraph(
+qgraph(
   edge_list,
@@
 :::::::::: columns
-::::: {.column width="40%"}
-$$
-  P(I_k = 1) = \Phi_L (a_k (\theta - b_k))
-$$
-
-::: {.fragment fragment-index="1"}
-$$
-  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
-$$
-:::
-
-<br>
-
-::: {.fragment fragment-index="3"}
-
-$$
-  \begin{align}
-    a_k &= 1.7 \frac{\lambda_k}{\sqrt{var(U_k)}} \\
-    a_k b_k &= -\frac{\tau_k}{\lambda}
-  \end{align}
-$$
-:::
-:::::
-
-:::::: {.column .fragment width="60%" fragment-index="2"}
+:::::: {.column width="60%"}
 ```{r irt-curves}
@@
 ::: r-stack
-[$a_1$ = 1.0; $b_1$ = 1.0]{style="color:#5c6eb1;"}
+[$a_1$ = 1.0; $b_1$ = 0.0]{style="color:#5c6eb1;"}
 :::
@@
 ::::::
+
+::::: {.column width="40%"}
+::: {.fragment fragment-index="1"}
+$$
+  P(I_k = 1) = \Phi_L (a_k (\theta - b_k))
+$$
+:::
+
+<br>
+
+::: {.fragment fragment-index="2"}
+$$
+  \begin{align}
+    d_k &= -a_k b_k \\
+    P(I_k = 1) &= \Phi_L (a_k \theta + d_k))
+  \end{align}
+$$
+:::
+
+<br>
+
+::: {.fragment fragment-index="3"}
+$$
+  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
+$$
+:::
+
+<br>
+
+:::::
 ::::::::::
@@
 
-## Teoría de respuesta al ítem multidimensional
-
-<!--# TODO: Fórmula de modelo M2PL -->
-
-<!--# TODO: Gráfica tridimensional de modelo compensatorio (y "curvas de nivel"?) -->
+## TRI multidimensional
+
+$$
+  P(I_k = 1) = \Phi_L (\mathbf{a}_k^T \mathbf{\theta} + d_k)
+$$
+
+::: columns
+::: column
+```{r mirt-model}
+#| fig-width:  5
+#| fig-height: 5
+PROB_AXIS_2D <- seq(-3, 3, by = .1)
+
+mirt_item <- tribble(
+  ~a_1, ~a_2, ~d,
+  2,    1,    -1.5
+)
+
+irs2d <- function(a_1, a_2, d, theta_1, theta_2) {
+
+  logit(a_1 * theta_1 + a_2 * theta_2 + d)
+}
+
+mirt_item_probs <- expand_grid(
+  trait_2 = PROB_AXIS_2D,
+  trait_1 = PROB_AXIS_2D
+) |>
+  mutate(
+    mirt_item |>
+      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
+      bind_cols()
+  ) |>
+    rename(prob = `...1`)
+
+mirt_prob_surface <- mirt_item_probs |>
+  pivot_wider(names_from = trait_1, values_from = prob) |>
+    select(-trait_2) |>
+    as.matrix()
+
+mirt_item_probs |>
+  plot_ly(
+    x = ~PROB_AXIS_2D,
+    y = ~PROB_AXIS_2D,
+    z = ~mirt_prob_surface,
+    colorscale = list(c(0, 1), PALETTE_BLUE["normal"] |> rep(2)),
+    type = 'surface',
+    showscale = FALSE,
+    hoverinfo = 'none'
+  ) |>
+  layout(
+    scene = list(
+      xaxis = list(title = "<i>θ</i><sub>1</sub>"),
+      yaxis = list(title = "<i>θ</i><sub>2</sub>"),
+      zaxis = list(title = "P(I<sub>k</sub> = 1)"),
+      camera = list(
+        center = list(z = -.2),
+        eye = list(x = -.12, y = -1.8, z = 1)
+      )
+    )
+  ) |>
+  config(
+    displayModeBar = FALSE,
+    scrollZoom = FALSE,
+    mathjax = 'cdn'
+  )
+```
+
+:::
+
+::: column
+:::
+:::
 

### 2024-10-28 output/seminar-slideshow.qmd
Add IRS contour plot

@@
 ::: column
-```{r mirt-model}
+```{r mirt-model-irs-3d}
 #| fig-width:  5
@@
   ~a_1, ~a_2, ~d,
-  2,    1,    -1.5
+  1,    0.5,  -0.5
 )
@@
 
-mirt_item_probs |>
 plot_ly(
@@
 ```
-
 :::
@@
 ::: column
+```{r mirt-model-irs-3d}
+#| fig-width:  5
+#| fig-height: 5
+
+mirt_item_probs |>
+  ggplot(mapping = aes(trait_1, trait_2, z = prob)) +
+  geom_contour() +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    limits = range(PROB_AXIS_2D),
+    expand = expansion()
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    limits = range(PROB_AXIS_2D),
+    expand = expansion()
+  ) +
+  coord_fixed() +
+  labs(x = "*&theta;*~1~", y = "*&theta;*~2~") +
+  theme(
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(angle = 90),
+    panel.grid.major = element_line(linetype = "17"),
+  )
+
+```
 :::

### 2024-10-28 output/seminar-slideshow.qmd
Improve 3d plot with camera view and tooltip

@@
 ```{r mirt-model-irs-3d}
-#| fig-width:  5
-#| fig-height: 5
+#| fig-width:  4.6
+#| fig-height: 4.6
 PROB_AXIS_2D <- seq(-3, 3, by = .1)
@@
   type = 'surface',
+  hovertemplate = 'P(I<sub>k</sub> = 1) = %{z:.2f}<extra></extra>',
   showscale = FALSE,
-  hoverinfo = 'none'
+  hoverinfo = 'Z'
 ) |>
@@
       camera = list(
-        center = list(z = -.2),
+        center = list(x = -.1, z = -.2),
         eye = list(x = -.12, y = -1.8, z = 1)
@@
   ) |>
-  config(
-    displayModeBar = FALSE,
-    scrollZoom = FALSE,
-    mathjax = 'cdn'
-  )
+  config(displayModeBar = FALSE, scrollZoom = FALSE)
 ```

### 2024-10-28 output/seminar-slideshow.qmd
Add labels to IRS contour plot

@@
 library(plotly)
+library(metR)
 ```
@@
 ::: column
-```{r mirt-model-irs-3d}
-#| fig-width:  5
-#| fig-height: 5
-
+```{r mirt-model-irs-contour}
 mirt_item_probs |>
-  ggplot(mapping = aes(trait_1, trait_2, z = prob)) +
-  geom_contour() +
+  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
+  geom_contour2(
+    color = F1_COLOR,
+    linewidth = LINE_WIDTH,
+    breaks = c(.05, .1, .2, .3, .5, .7, .8, .9, .95),
+    label.placer = label_placer_fraction(frac = .18, rot_adjuster = 0)
+  ) +
   scale_x_continuous(
@@
   )
-
 ```

### 2024-10-28 output/seminar-slideshow.qmd
Corrects expression of the FA model
Improve notes explaining MIRT model

@@
 $$
-  \begin{align}
-    P(I_k = 1) &= \int_{-\infty}^0 \mathcal{N}(i_k^*) \delta i_k^* \\
-               &= \Phi_N (i_k^*)
-  \end{align}
+  P(I_k = 1) = \int_{-\infty}^0 \mathcal{N}(i_k^*) \delta i_k^* \\
+             = \Phi_N \left(\frac{\tau_k + \mathbf{\lambda}_k^T \mathbf{F}}{\sqrt{var(U_k)}}\right)
 $$
@@
 ::: notes
-Una de las extensiones multidimensionales (más antiguas, y probablemente más obvias) de este modelo consiste simplemente en asumir que la respuesta se modela como una "regresión logística multivariada".
+Esto se puede extender, de manera bastante directa, asumiendo que la respuesta se modela como una "regresión logística multivariada".
+Aquí aparece otra vez la discriminación pero ahora en forma vectorial, que se multiplica por el vector de rasgo latente de la persona, y un parámetro de intersección, como antes (pero no puede haber un parámetro de dificultad simplemente, o uno por cada dimensión, porque serían indistinguibles entre sí).
+Esta es una de las extensiones multidimensionales (más antiguas, y probablemente más obvias) del modelo logístico de 2 parámetros: El modelo logístico de 2 parámetros multidimensional compensatorio.
+Pertenece a una familia de modelos llamados "compensatorios" en TRI, ya que un nivel bajo en una puntuación latente se compensa con un nivel alto en otra, dando lugar a la misma probabilidad de respuesta.
+:::
 

### 2024-10-28 output/seminar-slideshow.qmd
Add MIRT computation example

@@
 ```{r mirt-model-irs-contour}
-mirt_item_probs |>
+contour_plot <- mirt_item_probs |>
   ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
@@
   )
+
+contour_plot
 ```
@@
 
-Esto da lugar a una familia de modelos llamados "compensatorios" en TRI, ya que un nivel bajo en una puntuación latente se compensa con un nivel alto en otra, dando lugar a la misma probabilidad de respuesta.
-:::
-
-## TRI multidimensional: Ejemplo
-
-<!--# TODO: Valores, cálculo y gráfica -->
-
-::: notes
-Veamos un ejemplo de cuál sería la probabilidad de respuesta de una persona con puntuaciones de rasgo $\theta_1 = TODO$ y $\theta_2 = TODO$.
+## TRI multidimensional: Ejemplo {.smaller}
+
+::: columns
+::: column
+$$
+  P(I_k = 1) = \Phi_L (\mathbf{a}_k^T \mathbf{\theta} + d_k)
+$$
+
+```{r mirt-item-response-example}
+mirt_example <- tribble(
+  ~theta_1, ~theta_2,
+  1,        -1
+)
+
+mirt_example_spikes <- mirt_example |>
+  add_column(type = "end")          |> 
+  bind_rows(
+    tribble(
+      ~theta_1, ~theta_2, ~type,
+      0,        -1,        "start",
+      1,         0,        "start"
+    )
+  )
+
+mirt_example_spikes <- mirt_example_spikes |>
+  slice(1) |>
+  bind_rows(mirt_example_spikes) |>
+  add_column(factor = paste0('F', 1:2 |> rep(2))) |>
+  pivot_wider(names_from = type, values_from = starts_with("theta"))
+
+contour_plot +
+    geom_segment(
+    data    = mirt_example_spikes,
+    mapping = aes(
+      theta_1_start, theta_2_start,
+      xend = theta_1_end,
+      yend = theta_2_end
+    ),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = PALETTE_BLUE["light"]
+  ) +
+  geom_point(
+    data = mirt_example,
+    mapping = aes(theta_1, theta_2),
+    inherit.aes = FALSE,
+    size = I(4)
+  )
+```
+:::
+
+::: column
+\begin{matrix}
+  \mathbf{\theta} =
+    \begin{bmatrix}
+       1 \\
+      -1
+    \end{bmatrix}
+
+  &\begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        0.5
+      \end{bmatrix} \\
+    d_1 &= -0.5
+  \end{align}
+\end{matrix}
+
+::: fragment
+$$
+  \begin{align}
+    P(I_k = 1) &= \Phi_L \left([1, 0.5]
+      \begin{bmatrix}
+         1 \\
+        -1
+      \end{bmatrix}
+      -0.5 \right) \\
+      &= \Phi_L \left(1 · 1 + 0.5 · (-1) -0.5 \right) \\
+      &= \Phi_L \left(1 -0.5 -0.5 \right) = \Phi_L \left(0 \right) \\
+      &= \frac{1}{1 + e^{-0}} = \frac{1}{1 + 1} = \frac{1}{2} = 0.5
+  \end{align}
+$$
+:::
+:::
+:::
+
+::: notes
+Veamos un ejemplo de cuál sería la probabilidad de respuesta de una persona con puntuaciones de rasgo de 1 y -1.
 

### 2024-10-28 output/seminar-slideshow.qmd
Add metric transformation slide content

@@
 
-## AF en métrica TRI
-
-<!--# TODO: Fórmulas de los modelos y de transformación de parámetros. -->
-
-::: notes
-A partir de las ecuaciones del AF y de TRI, es fácil ver que existe un paralelismo entre ambos modelos.
-Es posible transformar los parámetros de la métrica del AF a métrica de TRI y, de hecho, esto es necesario, y se hace así para hacer la estimación máximo-verosimil de la que hablábamos antes.
+## AF en métrica TRI {.smaller}
+
+<br>
+
+$$
+  \begin{align}
+    P(I_k = 1) &= \Phi_N \left(
+      \frac{\mathbf{\lambda}_k^T \mathbf{F} + \tau_k}{\sqrt{var(U_k)}}
+    \right) \\
+    P(I_k = 1) &= \Phi_L (\mathbf{a}_k^T \mathbf{\theta} + d_k)
+  \end{align}
+$$
+
+::: fragment
+@takane_relationship_1987:
+
+$$
+  \begin{align}
+    \mathbf{a}_k &= \frac{1.7}{\sqrt{var(U_k)}} \mathbf{\lambda}_k \\
+    d_k &= \frac{1.7}{\sqrt{var(U_k)}} \tau_k
+  \end{align}
+$$
+:::
+
+::: notes
+Claro, si ponemos las ecuaciones del AF y de TRI lado a lado, es fácil ver que existe un paralelismo entre ambos modelos.
+
+Pues resulta que estas dos curvas, la normal y la logística, son muy parecidas; tanto, que es posible transformar los parámetros de la métrica del AF a métrica de TRI, obteniendo modelos indistinguibles en la práctica.
+Y de hecho, esto es necesario y se hace así para hacer la estimación máximo-verosimil de la que hablábamos antes.
 :::

### 2024-10-28 output/seminar-slideshow.qmd
Add MIRT difficulty parameter plot

@@
 source("R/Output.R",       encoding = 'UTF-8')
+source("R/Mirt_toolbox.R", encoding = 'UTF-8')
 ```
@@
 La TRI, por su lado, asume que hay cada persona tiene una puntuación latente, y que su respuesta depende de esa puntuación, con un parámetro de discriminación que determina cómo de bien o mal mide el ítem, y uno de dificultad (aunque aquí prefiero llamarlo "de posición") que determina lo fácil o difícil que será dar una respuesta positiva (afirmativa, correcta, codificada como 1).
-Esa posición representa el nivel de rasgo latente que da lugar a una probabilidad de .5 de dar la respuesta positiva.
+Esa posición representa el nivel de rasgo latente que da lugar a una probabilidad de .5 de dar la respuesta positiva (en este modelo; en general representa el punto de inflexión de la curva, es decir, donde pasa de hacerse cada vez más pronunciada a empezar a aplanarse).
 
@@
 mirt_item <- tribble(
-  ~a_1, ~a_2, ~d,
-  1,    0.5,  -0.5
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    0.5,  -0.5
 )
@@
     mirt_item |>
+      select(-item) |>
       pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
@@
 
+<br>
+
 ::: fragment
@@
 Pues resulta que estas dos curvas, la normal y la logística, son muy parecidas; tanto, que es posible transformar los parámetros de la métrica del AF a métrica de TRI, obteniendo modelos indistinguibles en la práctica.
+Las dos curvas se diferencian por este factor de 1.7, pero aplicándo ese factor y escalando por la desviación estándar de la unicidad del ítem, obtenemos los mismos parámetros en métrica de TRI mulidimensional con función de respuesta logística.
 Y de hecho, esto es necesario y se hace así para hacer la estimación máximo-verosimil de la que hablábamos antes.
@@
 
-## TRI Multidimensional: Parámetros multidimensionales
-
-<!--# TODO: Fórmula de modelo M2PL -->
-
+## Parámetros multidimensionales de TRI {.smaller}
+
+<br>
+
+$$
+  P(I_k = 1) = \Phi_L (\mathbf{a}_k^T \mathbf{\theta} + d_k)
+$$
+
+::: fragment
+$$
+  P(I_k = 1) = \Phi_L (a_k (\theta - b_k))
+$$
+:::
+
+::: fragment
 @reckase_difficulty_1985:
@@
 > \[...\] means of describing the characteristics of an item that takes into account the dimensionality of the skills \[...\] can then be used to determine how or if it is possible to compare items that measure different combinations of abilities.
+:::
+
+::: notes
+A pesar de este paralelismo que os he mostrado, y de la utilidad de los modelos de TRI, hay un problema con la formulación del modelo que os he planteado, y es que como habéis podido ver los parámetros no son fácilmente interpretables.
+
+La intersección del modelo por ejemplo no tiene una interpretación clara.
+Si recordáis en la TRI unidimensional el parámetro de posición permitía comparar ítems entre sí en cuanto a su dificultad (en el caso de items cognitivos).
+Esto en AF no es muy habitual (prestamos poca atención a la "intersección" o "umbral"), porque no nos interesa tanto cómo de probable sea dar una respuesta positiva o negativa a un ítem, pero podría ayudarnos por ejemplo a diagnosticar ítems que estén mal diseñados, ya que nos da una idea de si va a haber muchas o pocas respuestas positivas a un ítem.
+
+Por eso Reckase dice, ya en 1985, que se necesita una forma de describir las características de los ítems que tenga en cuenta la dimensionalidad de las habilidades (los rasgos latentes en general).
+Daos cuenta de que entonces aún no se sabe nada de ese paralelismo con el AF.
+Además menciona que de esa manera se puede intentar comparar ítems que miden diferentes combinaciones de habilidades (fijaos en la importancia que tiene el dónde están los ítems en el continuo, en TRI, en contraste con AF, como comentábamos al principio).
+:::
+
+## Dificultad multidimensional {.smaller}
+
+@reckase_difficulty_1985:
+
+> \[...\] the most reasonable point to use in defining the [multidimensional difficulty] for an item in the multidimensional space is the point where the item is most discriminating.
+
+---(p. 402)
+
+$$
+MID = \frac{-d_k}{\sqrt{\mathbf{a}_k^T\mathbf{a}_k}}
+$$
+
+::: fragment
+> The distance can be interpreted much like a $b$ parameter from unidimensional IRT
+
+---(p. 405)
+:::
+
+::: notes
+Por ese motivo propone un parámetro de "dificultad" (lo llama, para nosotros sería de posición) que sea de algún modo equivalente a la dificultad en el modelo unidimensional.
+Ese punto es donde la curva (superficie) de probabilidad de respuesta tiene una inflexión (a lo largo de toda ella, el punto más cercano al origen, donde la recta que lo une al origen es perpendicular a la recta de inflexión de la curva).
+
+Este parámetro se puede entender como una "distancia **con signo**" desde el origen hasta el punto de posición del ítem.
+:::
 
@@
 
-> \[...\] the most reasonable point to use in defining the MID for an item in the multidimensional space is the point where the item is most discriminating.
-
----(p. 402)
-
-<!--# TODO: Fórmula del parámetro D_i -->
-
-> The distance can be interpreted much like a $b$ parameter from unidimensional IRT
-
----(p. 405)
-
-<!--# TODO: Representación de la curva de inflexión y la posición del ítem -->
-
-::: notes
-A pesar de este paralelismo, y de la utilidad de los modelos de TRI, hay un problema con la formulación del modelo que os he planteado, y es que como habéis podido ver los parámetros no son fácilmente interpretables.
-La intersección del modelo no tiene una interpretación clara, al contrario que el parámetro de posición en la TRI unidimensional, que permite comparar ítems entre sí en cuanto a su dificultad.
-
-Esto en AF no es muy habitual (prestamos poca atención a la "intersección" o "umbral"), porque no nos interesa tanto cómo de probable sea dar una respuesta positiva o negativa a un ítem, pero podría ayudarnos por ejemplo a diagnosticar ítems que estén mal diseñados, ya que nos da una idea de si va a haber muchas o pocas respuestas positivas a un ítem.
-
-Por ese motivo Reckase propone en 1985 obtener un parámetro de "dificultad" (lo llama, para nosotro sería de posición) que permita comparar ítems, y que sea de algún modo equivalente a la dificultad en el modelo unidimensional.
-Ese punto es donde la curva (superficie) de probabilidad de respuesta tiene una inflexión (a lo largo de toda ella, el punto más cercano al origen, donde la recta que lo une al origen es perpendicular a la recta de inflexión de la curva).
-
-Este parámetro se puede entender como una "distancia con signo" desde el origen hasta el punto de posición del ítem.
-:::
-
-## Dificultad multidimensional
+```{r mid-representation}
+#| fig-height: 8
+#| fig-width:  8
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE)
+
+mirt_item_spikes <- mirt_item_coords |>
+  select(item, trait_1 = origin_transf_1, trait_2 = origin_transf_2) |>
+  bind_rows(tibble(item = 1, trait_1 = 0, trait_2 = 0)) |>
+  add_column(type = c("end", "start")) |>
+  pivot_wider(names_from = type, values_from = starts_with("trait"))
+
+contour_plot +
+    geom_segment(
+    data    = mirt_item_spikes,
+    mapping = aes(
+      trait_1_start, trait_2_start,
+      xend = trait_1_end,
+      yend = trait_2_end
+    ),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = PALETTE_BLUE["light"]
+  ) +
+  geom_point(
+    data = mirt_item_coords,
+    mapping = aes(origin_transf_1, origin_transf_2),
+    inherit.aes = FALSE,
+    size = I(4)
+  ) +
+  scale_x_continuous(
+    breaks = -1:1,
+    minor_breaks = NULL,
+    limits = c(-1.5, 1.5)
+  ) +
+  scale_y_continuous(
+    breaks = -1:1,
+    minor_breaks = NULL,
+    limits = c(-1.5, 1.5)
+  ) +
+  annotate(
+    geom = "text",
+    x = 0.1, y = 0.2,
+    label = "MID",
+    size = 6,
+    angle = mirt_item_params |> pull(deg_1)
+  )
+```
 

### 2024-10-28 output/seminar-slideshow.qmd
Make notation more consistent with article

@@
 $$
-I_1 = \tau_1 + \lambda_{11} F_1^p + \lambda_{21} F_2^p + ... + \lambda_{n1} F_n^p + U_1^p \\
+Y_1 = \tau_1 + \lambda_{11} F_1^p + \lambda_{21} F_2^p + ... + \lambda_{n1} F_n^p + U_1^p \\
 \vdots                                                                            \\
-I_k = \tau_k + \lambda_{1k} F_1^p + \lambda_{2k} F_2^p + ... + \lambda_{nk} F_n^p + U_k^p
+Y_i = \tau_i + \lambda_{1i} F_1^p + \lambda_{2i} F_2^p + ... + \lambda_{ni} F_n^p + U_i^p
 $$
@@
 $$
-I_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
+Y_i = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
 $$
@@
 $$
-  i_k^* = \tau_k + \mathbf{\lambda}_k^T \mathbf{F} + U_k
+  y_i^* = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
 $$
@@
 $$
-  I_k = \left\{ \begin{matrix}
-      1 &si &i_k^* \geq 0\\
-      0 &si &i_k^* <    0
+  Y_i = \left\{ \begin{matrix}
+      1 &si &y_i^* \geq 0\\
+      0 &si &y_i^* <    0
     \end{matrix} \right.
@@
 $$
-  P(I_k = 1) = \int_{-\infty}^0 \mathcal{N}(i_k^*) \delta i_k^* \\
-             = \Phi_N \left(\frac{\tau_k + \mathbf{\lambda}_k^T \mathbf{F}}{\sqrt{var(U_k)}}\right)
+  P(Y_i = 1) = \int_{-\infty}^0 \mathcal{N}(y_i^*) \delta y_i^* \\
+             = \Phi_N \left(\frac{\tau_i + \mathbf{\lambda}_i^T \mathbf{F}}{\sqrt{var(U_i)}}\right)
 $$
@@
   )                              +
-  labs(x = "i~k~^*^", y = NULL)  +
+  labs(x = "y~i~^*^", y = NULL)  +
   scale_y_continuous(
@@
   geom_line(color = PALETTE_BLUE["normal"], linewidth = LINE_WIDTH) +
-  labs(x = "E(*i~k~^*^*)", y = "P(I~k~ = 1)")                       +
+  labs(x = "E(*y~i~^*^*)", y = "P(Y~i~ = 1)")                       +
   scale_y_continuous(
@@
   geom_line(linewidth = LINE_WIDTH)        +
-  labs(x = "*&theta;*", y = "P(I~k~ = 1)") +
+  labs(x = "*&theta;*", y = "P(Y~i~ = 1)") +
   scale_y_continuous(
@@
 $$
-  P(I_k = 1) = \Phi_L (a_k (\theta - b_k))
+  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
 $$
@@
   \begin{align}
-    d_k &= -a_k b_k \\
-    P(I_k = 1) &= \Phi_L (a_k \theta + d_k))
+    d_i &= -a_i b_i \\
+    P(Y_i = 1) &= \Phi_L (a_i \theta + d_i))
   \end{align}
@@
 $$
-  P(I_k = 1) = \Phi_L (\mathbf{a}_k^T \mathbf{\theta} + d_k)
+  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
 $$
@@
   type = 'surface',
-  hovertemplate = 'P(I<sub>k</sub> = 1) = %{z:.2f}<extra></extra>',
+  hovertemplate = 'P(I<sub>i</sub> = 1) = %{z:.2f}<extra></extra>',
   showscale = FALSE,
@@
       yaxis = list(title = "<i>θ</i><sub>2</sub>"),
-      zaxis = list(title = "P(I<sub>k</sub> = 1)"),
+      zaxis = list(title = "P(I<sub>i</sub> = 1)"),
       camera = list(
@@
 $$
-  P(I_k = 1) = \Phi_L (\mathbf{a}_k^T \mathbf{\theta} + d_k)
+  P(Y_1 = 1) = \Phi_L (\mathbf{a}_1^T \mathbf{\theta} + d_1)
 $$
@@
   \begin{align}
-    P(I_k = 1) &= \Phi_L \left([1, 0.5]
+    P(Y_1 = 1) &= \Phi_L \left([1, 0.5]
       \begin{bmatrix}
@@
   \begin{align}
-    P(I_k = 1) &= \Phi_N \left(
-      \frac{\mathbf{\lambda}_k^T \mathbf{F} + \tau_k}{\sqrt{var(U_k)}}
+    P(Y_i = 1) &= \Phi_N \left(
+      \frac{\mathbf{\lambda}_i^T \mathbf{F} + \tau_i}{\sqrt{var(U_i)}}
     \right) \\
-    P(I_k = 1) &= \Phi_L (\mathbf{a}_k^T \mathbf{\theta} + d_k)
+    P(Y_i = 1) &= \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
   \end{align}
@@
   \begin{align}
-    \mathbf{a}_k &= \frac{1.7}{\sqrt{var(U_k)}} \mathbf{\lambda}_k \\
-    d_k &= \frac{1.7}{\sqrt{var(U_k)}} \tau_k
+    \mathbf{a}_i &= \frac{1.7}{\sqrt{var(U_i)}} \mathbf{\lambda}_i \\
+    d_i &= \frac{1.7}{\sqrt{var(U_i)}} \tau_i
   \end{align}
@@
 $$
-MID = \frac{-d_k}{\sqrt{\mathbf{a}_k^T\mathbf{a}_k}}
+  MID = \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
 $$
@@
 
+:::: columns
+::: column
 ```{r mid-representation}
@@
 ```
-
-<!--# TODO: Fórmula de los cosenos directores -->
+:::
+
+::: column
+::: fragment
+$$
+  cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
+$$
+:::
+
+::: fragment
+
+:::
+:::
+::::
 

### 2024-10-28 output/seminar-slideshow.qmd
Add discrimination parameter content (WiP)

@@
 library(metR)
+library(scales)
 ```
@@
 AXIS_COLOR   <- "darkgray" # Replace default grid line color for slideshow
+VECTOR_WIDTH <- 2.5        # Replace default vector width for slideshow
 ```
@@
 
+<br>
+
 ::: fragment
-
+$$
+  MID_i = \begin{Bmatrix}
+    \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}} \\
+    \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
+  \end{Bmatrix}
+$$
 :::
@@
 ::::
-
-<!--# TODO: Representación gráfica de los cosenos directores? -->
-
-<!--# TODO: Conjunto del MID -->
 
@@
 
-Por lo tanto, la "dificultad (posición) multidimensional" viene dada por dos elementos, un vector de cosenos que determina la dirección en la que mide un ítem, y un número entero que representa cuánto se desplaza respecto del origen, y en qué sentido, la recta de inflexión del ítem.
+Por lo tanto, la "dificultad (posición) multidimensional" viene dada por dos elementos, un número real que representa cuánto se desplaza la recta de inflexión del ítem respecto del origen y en qué sentido, y un vector de cosenos que determina la dirección en la que se desplaza.
 Así se puede interpretar un ítem, o varios, de manera relativamente sencilla: Todos los participantes cuyas coordenadas estén en la "recta de inflexión" (la recta que pasa por la posición del ítem, perpendicular a su dirección) tendrán una probabilidad de dar una respuesta positiva y negativa de .5 (dificultad media; igual probabilidad de acertar o fallar).
@@
 
-## Discriminación multidimensional
+## Discriminación multidimensional {.smaller}
 
@@
 
-<!-- # TODO: Superficies de ítems con alta y baja discriminación -->
+:::: {.columns .fragment}
+
+::: column
+<br>
+
+$$
+  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
+$$
+:::
+
+::: column
+```{r mdisc-item}
+#| fig-width:  4
+#| fig-height: 4
+mirt_item_coords |>
+  arrange(desc(item))   |> # To plot them in reverse order
+  ggplot(
+    aes(
+      0, 0,
+      xend = end_transf_1 - origin_transf_1,
+      yend = end_transf_2 - origin_transf_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = LINE_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 2),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    oob          = oob_keep
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 2),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    oob          = oob_keep
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  coord_fixed(expand = FALSE, clip = "on")           +
+  theme(
+    axis.line        = element_blank(),
+    panel.grid.major = element_line(
+      linewidth = LINE_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+::::
 
@@
 
+## Discriminación multidimensional {.smaller}
+
+<!-- # TODO: Superficies de ítems con alta y baja discriminación -->
+
 ## Parámetros multidimensionales

### 2024-10-28 output/seminar-slideshow.qmd
Apply canonical markdown formatting to progress

@@
 
-:::::::::: columns
+::::::::::: columns
 :::::: {.column width="60%"}
@@
 
-::::: {.column width="40%"}
+:::::: {.column width="40%"}
 ::: {.fragment fragment-index="1"}
@@
 <br>
-
-:::::
-::::::::::
+::::::
+:::::::::::
 
@@
 
-::: columns
+::::: columns
 ::: column
@@
 :::
-:::
+:::::
 
@@
 
-::: columns
+:::::: columns
 ::: column
@@
 
-::: column
+:::: column
 \begin{matrix}
@@
 :::
-:::
-:::
+::::
+::::::
 
@@
 
-> \[...\] the most reasonable point to use in defining the [multidimensional difficulty] for an item in the multidimensional space is the point where the item is most discriminating.
+> \[...\] the most reasonable point to use in defining the \[multidimensional difficulty\] for an item in the multidimensional space is the point where the item is most discriminating.
 
@@
 
-:::: columns
+::::::: columns
 ::: column
@@
 
-::: column
+::::: column
 ::: fragment
@@
 :::
-:::
-::::
+:::::
+:::::::
 
@@
 
-:::: {.columns .fragment}
-
+::::: {.columns .fragment}
 ::: column
@@
 :::
-::::
+:::::
 

### 2024-10-28 output/seminar-slideshow.qmd
Version progress point

@@
   scale_x_continuous(
-    limits       = c(-1, 2),
+    limits       = c(-1, 2), ## TODO: Change limits
     breaks       = -1:2,

### 2024-10-29 output/seminar-slideshow.qmd
Correct IFA probability function formula
Mkes several style updates in plots

--- 2024-10-28 output/seminar-slideshow.qmd
@@
 library(qgraph)
+library(stringr)
 library(purrr)
@@
 FONT_SIZE    <- 24       # Replace default font size for slideshow
-AXIS_LAB_POS <-   .5       # Replace default axis label position for slideshow
-LINE_WIDTH   <-   .75      # Replace default line width for slideshow
-AXIS_COLOR   <- "darkgray" # Replace default grid line color for slideshow
-VECTOR_WIDTH <- 2.5        # Replace default vector width for slideshow
+AXIS_LAB_POS <-  0.5     # Replace default axis label position for slideshow
+LINE_WIDTH   <-  0.75    # Replace default line width for slideshow
+GRID_WIDTH   <-  0.5     # Width for grid and axis lines
+AXIS_COLOR   <- "gray70" # Replace default grid line color for slideshow
+VECTOR_WIDTH <- 1.5      # Replace default vector width for slideshow
 ```
@@
       axis.title.y       = element_text(vjust = AXIS_LAB_POS, angle = 0),
+      axis.title.y.right = element_text(vjust = AXIS_LAB_POS, angle = 0),
       panel.grid         = element_line(color = AXIS_COLOR),
-      panel.grid.minor = element_line(linetype = "17", color = "grey70")
+      panel.grid.minor   = element_line(linetype = "17", color = AXIS_COLOR)
     )
@@
     name     = "F2",
+    position = "top"
   )                                                                +
@@
     labels   = NULL,
-    name   = "F1"
+    name     = "F1",
+    position = "right"
   )                                                                +
@@
     axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y       = element_text(vjust = axis_pos),
+    axis.title.y.right = element_text(vjust = axis_pos),
     panel.grid.major   = element_line(
       color     = "black",
-      linewidth = LINE_WIDTH
+      linewidth = GRID_WIDTH
     ),
@@
   ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
-  geom_point(size = I(4))                                          +
   geom_text(
     nudge_x = -.12,
-    # nudge_y = -.08,
     color   = I("black"),
@@
     axis.ticks         = element_blank(),
-    axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y       = element_text(vjust = axis_pos),
     panel.grid.major   = element_line(
       color     = "black",
-      linewidth = LINE_WIDTH
-    )
-  )
-
-simple_item_loadings_rot_plot
+      linewidth = GRID_WIDTH
+    )
+  )
+
+simple_item_loadings_rot_plot + geom_point(size = I(4))
 ```
@@
     data    = rotated_axes,
-    mapping = aes(slope = slope, intercept = 0, color = factor)
+    mapping = aes(slope = slope, intercept = 0),
+    color   = c(PALETTE_BLUE["light"], PALETTE_RASPBERRY["light"]),
+    linewidth = LINE_WIDTH
+  ) +
+  geom_point(size = I(4)) +
+  annotate(
+    "text",
+    x = -1.35, y = c(-.62, .78),
+    label = rotated_axes |> pull(factor),
+    size  = 9
   )
@@
     name   = "F2",
+    position = "top"
   )                                                                +
@@
     labels = NULL,
-    name   = "F1"
+    name   = "F1",
+    position = "right"
   )                                                                +
@@
   geom_text(
-    # nudge_x =  .1,
     nudge_y = -.12,
@@
     name   = "M1",
+    position = "top"
   )                                                        +
@@
     labels = NULL,
-    name   = "F1"
+    name   = "F1",
+    position = "right"
   )                                                        +
@@
 #| cache: false
-AXIS_MINOR <- c(-1, 1, 2)
+AXIS_GRID <- -1:1
 
@@
   ggplot(mapping = aes(F1, F2, label = object))             +
+  geom_hline(
+    yintercept = 0,
+    color      = PALETTE_BLUE["light"],
+    linewidth  = GRID_WIDTH
+  )                                                         +
+  geom_vline(
+    xintercept = 0,
+    color      = PALETTE_RASPBERRY["light"],
+    linewidth  = GRID_WIDTH
+  )                                                         +
   geom_segment(
@@
     limits       = axis_lims,
-    breaks       = 0,
-    labels       = NULL,
-    name         = "F~2~",
-    minor_breaks = AXIS_MINOR
+    breaks       = AXIS_GRID,
+    name         = NULL,
+    minor_breaks = NULL,
+    sec.axis     = dup_axis(name = "F~2~", labels = NULL)
   )                                                        +
@@
     limits       = axis_lims,
-    breaks       = 0,
-    labels       = NULL,
-    name         = "F~1~",
-    minor_breaks = AXIS_MINOR
+    breaks       = AXIS_GRID,
+    name         = NULL,
+    minor_breaks = NULL,
+    sec.axis     = dup_axis(name = "F~1~", labels = NULL)
   )                                                        +
@@
     axis.title.y       = element_markdown(vjust = axis_pos),
+    axis.title.y.right = element_text(vjust = axis_pos),
     panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = LINE_WIDTH
-    ),
-    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
-    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
   )
@@
 ::::::::: columns
-::::: {.column width="40%"}
+::::: {.column width="45%"}
 $$
@@
 $$
-  P(Y_i = 1) = \int_{-\infty}^0 \mathcal{N}(y_i^*) \delta y_i^* \\
-             = \Phi_N \left(\frac{\tau_i + \mathbf{\lambda}_i^T \mathbf{F}}{\sqrt{var(U_i)}}\right)
+  \begin{align}
+    P(Y_i = 1) &= \int_0^{-\infty} \mathcal{N}(y_i^*) \delta y_i^* \\
+               &= \Phi_N \left(
+                 \frac{ \tau_i + \mathbf{\lambda}_i^T \mathbf{F} }
+                   { \sqrt{var(U_i)} }
+               \right)
+  \end{align}
 $$
@@
 
-::::: {.column width="60%"}
+::::: {.column width="55%"}
 ::: {.fragment fragment-index="1"}
@@
     legend.position = "none",
-    panel.grid.major = element_line(linetype = "17")
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
   )                              +
@@
     limits       = 0:1,
-    expand       = expansion(mult = c(0, .05))
+    expand       = expansion(mult = c(0, .05)),
+    oob = oob_keep
   )  +
@@
     axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17"),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
   )
@@
   ) |>
-  rename_with(stringr::str_remove, pattern = '\\.{3}')
+  rename_with(str_remove, pattern = '\\.{3}')
 
@@
   ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
   geom_contour2(
@@
     breaks = c(.05, .1, .2, .3, .5, .7, .8, .9, .95),
-    label.placer = label_placer_fraction(frac = .18, rot_adjuster = 0)
+    label.placer = label_placer_fraction(frac = .90, rot_adjuster = 0)
   ) +
@@
     limits = range(PROB_AXIS_2D),
-    expand = expansion()
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
   ) +
@@
     limits = range(PROB_AXIS_2D),
-    expand = expansion()
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
   ) +
   coord_fixed() +
-  labs(x = "*&theta;*~1~", y = "*&theta;*~2~") +
   theme(
+    axis.ticks       = element_blank(),
+    axis.line        = element_blank(),
     axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17"),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
   )
@@
 :::: column
+
 \begin{matrix}
@@
 mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE)
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
 
 mirt_item_spikes <- mirt_item_coords |>
-  select(item, trait_1 = origin_transf_1, trait_2 = origin_transf_2) |>
+  select(item, trait_1 = origin_1, trait_2 = origin_2) |>
   bind_rows(tibble(item = 1, trait_1 = 0, trait_2 = 0)) |>
@@
     data = mirt_item_coords,
-    mapping = aes(origin_transf_1, origin_transf_2),
+    mapping = aes(origin_1, origin_2),
     inherit.aes = FALSE,
@@
     minor_breaks = NULL,
-    limits = c(-1.5, 1.5)
+    limits       = c(-1.5, 1.5),
+    expand       = expansion(),
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
   ) +
@@
     minor_breaks = NULL,
-    limits = c(-1.5, 1.5)
+    limits       = c(-1.5, 1.5),
+    expand       = expansion(),
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
   ) +
@@
     geom = "text",
-    x = 0.1, y = 0.2,
+    x = 0.18, y = 0.17,
     label = "MID",
@@
 
-@reckase_discriminating_1991:
-
-> the discriminating power of an item indicates how quickly the transition takes place from low probability to high probability of a correct response.
-> \[...\] A highly discriminating item divides the regions clearly-having a narrow region of ambiguity, that is, a region where the probabilities are intermediate in magnitude.
+@reckase_discriminating_1991 [p. 362]:
+
+> the discriminating power of an item indicates how quickly the transition takes place from low probability to high probability \[...\].
+> A highly discriminating item divides the regions clearly---having a narrow region  \[...\] where the probabilities are intermediate in magnitude.
 
@@
 <br>
+<br>
+<br>
 
@@
 ```{r mdisc-item}
-#| fig-width:  4
-#| fig-height: 4
-mirt_item_coords |>
-  arrange(desc(item))   |> # To plot them in reverse order
+#| fig-width:  4.7
+#| fig-height: 4.7
+#| cache: false
+
+item_vector <- mirt_item_coords |> mutate(
+  across(ends_with("1"), ~ . - origin_1),
+  across(ends_with("2"), ~ . - origin_2)
+)
+
+item_spikes <- item_vector             |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+item_vector |>
   ggplot(
@@
       0, 0,
-      xend = end_transf_1 - origin_transf_1,
-      yend = end_transf_2 - origin_transf_2,
+      xend = end_1,
+      yend = end_2,
       color = factor(item), fill = factor(item)
@@
   )                                                  +
-  geom_vline(xintercept = 0, linewidth = LINE_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = LINE_WIDTH) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
   geom_segment(
@@
   scale_x_continuous(
-    limits       = c(-1, 2),
+    limits       = c(-1, 1.5),
     breaks       = -1:2,
@@
     name         = NULL,
-    oob          = oob_keep
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
   ) +
   scale_y_continuous(
-    limits       = c(-1, 2),
+    limits       = c(-1, 1.5),
     breaks       = -1:2,
@@
     name         = NULL,
-    oob          = oob_keep
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
   ) +
   scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(expand = FALSE, clip = "on")           +
+  annotate(
+    "text",
+    x = 0.32, y = 0.3,
+    label = "MDISC",
+    size  = 5.5,
+    angle = mirt_item_params |> pull(deg_1)
+  ) +
+  annotate(
+    "richtext",
+    x = c(0.5, -0.13), y = c(-0.12, 0.25),
+    label         = c("*a*~1*i*~", "*a*~2*i*~"),
+    size          = 4.5,
+    label.colour  = NA,
+    label.padding = unit(0, "lines")
+  ) +
+  coord_fixed(clip = "on") +
   theme(
     axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
     panel.grid.major   = element_line(
-      linewidth = LINE_WIDTH,
+      linewidth = GRID_WIDTH,
       linetype  = "17"
-    )
+    ),
   )

### 2024-10-29 output/seminar-slideshow.qmd
Corrects arrow direction in regression scores plot

@@
   ~rhs, ~lhs,
-  7,    1,
-  7,    2,
-  7,    3,
-  8,    4,
-  8,    5,
-  8,    6,
+  1,    7,
+  2,    7,
+  3,    7,
+  4,    8,
+  5,    8,
+  6,    8,
   7,    8,

### 2024-10-29 output/seminar-slideshow.qmd
Adjust labels of the MID and MDISC plots

@@
   annotate(
-    geom = "text",
-    x = 0.18, y = 0.17,
-    label = "MID",
+    geom = "richtext",
+    x = 0.16, y = 0.15,
+    label = "*MID~i~*",
     size = 6,
-    angle = mirt_item_params |> pull(deg_1)
+    angle = mirt_item_params |> pull(deg_1),
+    label.colour  = NA,
+    fill = NA
   )
@@
   annotate(
-    "text",
-    x = 0.32, y = 0.3,
-    label = "MDISC",
+    "richtext",
+    x = 0.3, y = 0.27,
+    label = "*MDISC~i~*",
     size  = 5.5,
-    angle = mirt_item_params |> pull(deg_1)
+    angle = mirt_item_params |> pull(deg_1),
+    label.colour  = NA,
+    fill = NA
   ) +
@@
     label.colour  = NA,
-    label.padding = unit(0, "lines")
+    label.padding = unit(0, "lines"),
+    fill = NA
   ) +

### 2024-11-08 output/seminar-slideshow.qmd
Apply canonical markdown to new content
Improve notes in "multidimensional discrimination"

@@
 :::: column
-
 \begin{matrix}
@@
 ::: column
-<br>
-<br>
-<br>
+<br> <br> <br>
 
@@
 Esta característica viene representada por el parámetro de "discriminación", que es una medida de la "calidad general" de un ítem.
-Los ítems "buenos" serán muy discriminativos, lo que indica que distinguen bien entre la gente que tiene nivel de rasgo alto o bajo, mientras que los "malos" tendrán una superficie de respuesta más "plana".
-
+Los ítems "buenos" serán muy discriminativos, lo que indica que distinguen bien entre la gente que está en la región de alta o baja probabilidad de respuesta, mientras que los "malos" tendrán una superficie de respuesta más "plana".
+:::
+
+## Discriminación multidimensional {.smaller}
+
+<!-- # TODO: Superficies de ítems con alta y baja discriminación -->
+
+::: notes
 La discriminación se puede entender con una especie de "dardo" que atraviesa de arriba abajo esa superficie perpendicularmente por esa "recta de inflexión" de la que hablábamos antes, y luego cruza el plano origen.
@@
 En este caso, el dardo nunca cortaría al plano origen, y la discriminación sería infinita.
-:::
-
-## Discriminación multidimensional {.smaller}
-
-<!-- # TODO: Superficies de ítems con alta y baja discriminación -->
+
+Pero cuidado, que el ítem "discimine mucho" no significa que sea "bueno" midiendo todas las dimensiones, ya que eso va a depender también de "en qué dirección" mida el ítem, y eso viene dado por el componente de dirección que hemos visto antes.
+:::
 

### 2024-11-08 output/seminar-slideshow.qmd
Add slide about comparison of high and low MDISCs

@@
 ```{r mirt-model-irs-3d}
-#| fig-width:  4.6
-#| fig-height: 4.6
+#| fig-width:  5
+#| fig-height: 5
 PROB_AXIS_2D <- seq(-3, 3, by = .1)
@@
 
-<!-- # TODO: Superficies de ítems con alta y baja discriminación -->
+```{r mdisc-items}
+mdisc_items <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     4,    0.1,     0,
+  2,     0.01, 0.02,  0,
+)
+
+mirt_item_probs <- expand_grid(
+  trait_1 = PROB_AXIS_2D,
+  trait_2 = PROB_AXIS_2D
+) |>
+  mutate(
+    mdisc_items |>
+      select(-item) |>
+      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
+      bind_cols()
+  ) |>
+    rename_with(str_replace, pattern = '\\.{3}', replacement = "prob_")
+```
+
+::::: columns
+::: column
+$$
+  \mathbf{a}_1 = \begin{bmatrix}
+    0.1 \\
+    0.02
+  \end{bmatrix};\quad
+  d_1 = 0
+$$
+
+```{r low-mdisc-irs-contour}
+#| fig-height: 5
+#| fig-width:  5
+mirt_prob_surface <- mirt_item_probs |>
+  select(starts_with("trait"), prob_2) |>
+  pivot_wider(names_from = trait_1, values_from = prob_2) |>
+    select(-trait_2) |>
+    as.matrix()
+
+plot_ly(
+  x = ~PROB_AXIS_2D,
+  y = ~PROB_AXIS_2D,
+  z = ~mirt_prob_surface,
+  colorscale = list(c(0, 1), PALETTE_BLUE["normal"] |> rep(2)),
+  type = 'surface',
+  hovertemplate = 'P(I<sub>1</sub> = 1) = %{z:.2f}<extra></extra>',
+  showscale = FALSE,
+  hoverinfo = 'Z'
+) |>
+  layout(
+    scene = list(
+      xaxis = list(title = "<i>θ</i><sub>1</sub>"),
+      yaxis = list(title = "<i>θ</i><sub>2</sub>"),
+      zaxis = list(
+        range = 0:1,
+        title = "P(I<sub>1</sub> = 1)"
+      ),
+      camera = list(
+        center = list(x = -.1, z = -.2),
+        eye = list(x = -.12, y = -1.8, z = 1)
+      )
+    )
+  ) |>
+  config(displayModeBar = FALSE, scrollZoom = FALSE)
+```
+:::
+
+::: column
+$$
+  \mathbf{a}_2 = \begin{bmatrix}
+    4 \\
+    0.1
+  \end{bmatrix};\quad
+  d_2 = 0
+$$
+
+```{r high-mdisc-irs-contour}
+#| fig-height: 5
+#| fig-width:  5
+mirt_prob_surface <- mirt_item_probs |>
+  select(trait_1:prob_1) |>
+  pivot_wider(names_from = trait_1, values_from = prob_1) |>
+    select(-trait_2) |>
+    as.matrix()
+
+plot_ly(
+  x = ~PROB_AXIS_2D,
+  y = ~PROB_AXIS_2D,
+  z = ~mirt_prob_surface,
+  colorscale = list(c(0, 1), PALETTE_BLUE["normal"] |> rep(2)),
+  type = 'surface',
+  hovertemplate = 'P(I<sub>2</sub> = 1) = %{z:.2f}<extra></extra>',
+  showscale = FALSE,
+  hoverinfo = 'Z'
+) |>
+  layout(
+    scene = list(
+      xaxis = list(title = "<i>θ</i><sub>1</sub>"),
+      yaxis = list(title = "<i>θ</i><sub>2</sub>"),
+      zaxis = list(
+        range = 0:1,
+        title = "P(I<sub>2</sub> = 1)"
+      ),
+      camera = list(
+        center = list(x = -.1, z = -.2),
+        eye = list(x = -.12, y = -1.8, z = 1)
+      )
+    )
+  ) |>
+  config(displayModeBar = FALSE, scrollZoom = FALSE)
+```
+:::
+:::::
 

### 2024-11-08 output/sandbox_slideshow.qmd
Version sandbox document to work on single slides
Intermediate markdown in the main slideshow is not necessary while using this sandbox slidehow for development

--- 2024-10-29 output/seminar-slideshow.qmd
@@
     results:   asis
-    cache:     true
+    cache:     false
     autodep:   true
@@
 ```
-
-::: notes
--   Nombre de la presentación: Cero atractivo
-
--   Nuria me pide un título más llamativo
-
--   Lo mejor que puedo conseguir
-
--   ¿Por qué es relevante?
-
--   TRI muy usada en evaluación e investigación educativa
-
--   Poco en investigación psicológica (muestras "pequeñas" en comparación)
-
--   Pero hay un "paralelismo" entre ambas teorías (isomórficas, ante determinados supuestos)
-:::
-
-<!-- Slides "Diferencias AF - TRI" -->
-
-```{r irt-fa-table}
-af_tri_differences <- tribble(
-  ~AF,                            ~TRI,
-  "Relaciones entre variables",   "Comparación entre participantes",
-  "Cuestionarios",                "Pruebas (rendimiento)",
-  "Explicación",                  "Medición",
-  "Puntuación compuesta",         "Puntuación latente",
-  "Estructura simple",            "Diferencias individuales"
-)
-
-section_title <- "## Diferencias AF - TRI"
-
-for (slide in af_tri_differences |> nrow() |> seq_len()) {
-
-  cat( "\n", section_title, sep = '')
-
-  cat( "\n", "<br>", sep = '')
-
-  af_tri_differences |> slice_head(n = slide) |> kable() |> print()
-}
-```
-
-## AF: Estructura simple
-
-<br>
-
-```{r simple-complex-structure}
-ITEM_BASE_LOADING <- 1
-ITEM_DIST         <- 0.2
-
-# Generate data for simple and complex factorial structures
-simple_item_loadings <- tribble(
-  ~item, ~factor, ~loading,
-  'I1',  'F1',    ITEM_BASE_LOADING,
-  'I2',  'F1',    ITEM_BASE_LOADING + ITEM_DIST,
-  'I3',  'F1',    ITEM_BASE_LOADING + 3 * ITEM_DIST,
-  'I4',  'F2',    ITEM_BASE_LOADING - ITEM_DIST,
-  'I5',  'F2',    ITEM_BASE_LOADING + ITEM_DIST,
-  'I6',  'F2',    ITEM_BASE_LOADING + 2 * ITEM_DIST
-)
-complex_item_loadings <- simple_item_loadings |> bind_rows(
-  tibble(
-    item    = 'I3',
-    factor  = 'F2',
-    loading = ITEM_BASE_LOADING - 2 * ITEM_DIST # Secondary loading smaller
-  )
-)
-
-simple_item_loadings <- simple_item_loadings |>
-  complete(item, factor, fill = list(loading = 0))
-
-complex_item_loadings <- complex_item_loadings |>
-  complete(item, factor, fill = list(loading = 0))
-```
-
-::::: columns
-::: column
-```{r path-simple-structure}
-# TODO: Syntax from tibble
-simple_spec <- "
-  F1 =~ I1 + I2 + I3
-  F2 =~ I4 + I5 + I6
-
-  F1 ~~ F2
-"
-
-F1_COLOR <- unname(PALETTE_BLUE["normal"])
-F2_COLOR <- unname(PALETTE_RASPBERRY["normal"])
-
-item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
-factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
-
-simple_spec      |>
-  semPlotModel() |>
-  semPaths(
-    border.color = c(
-      item_colors,       # Item colors
-      F1_COLOR, F2_COLOR # Factor colors
-    ),
-    edge.color   = c(
-      item_colors, # Loading colors
-      "black"      # Correlation color
-    ),
-    residuals    = FALSE,
-    thresholds   = FALSE,
-    intercepts   = FALSE,
-    fixedStyle   = 1,
-    curve        = 3,
-    label.cex    = 1.6,
-    sizeMan      = 10,
-    sizeLat      = 14,
-    border.width = 3,
-    edge.width   = 3,
-    mar          = MARGINS
-  )
-```
-:::
-
-::: {.column .fragment}
-```{r scatter-simple-structure}
-LIM_INF <- -1.5
-LIM_SUP <-  2
-
-axis_lims <- c(LIM_INF, LIM_SUP)
-lim_range <- LIM_SUP - LIM_INF
-axis_pos  <- (-LIM_INF) / lim_range
-
-item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
-
-simple_item_loadings_wide <- simple_item_loadings |>
-  group_by(item)                                  |>
-  arrange(desc(loading))                          |>
-  mutate(main_factor = first(factor))             |>
-  pivot_wider(names_from = factor, values_from = loading)
-
-simple_item_loadings_wide |>
-  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
-  geom_point(size = I(4))                                          +
-  geom_text(
-    nudge_x =  .1,
-    nudge_y = -.08,
-    color   = I("black"),
-    size    = I(5)
-  )                                                                +
-  scale_x_continuous(
-    limits   = axis_lims,
-    breaks   = 0,
-    labels   = NULL,
-    name     = "F2",
-    position = "top"
-  )                                                                +
-  scale_y_continuous(
-    limits   = axis_lims,
-    breaks   = 0,
-    labels   = NULL,
-    name     = "F1",
-    position = "right"
-  )                                                                +
-  scale_color_manual(values = factor_colors, guide = NULL)         +
-  coord_fixed(expand = FALSE, clip = "on")                         +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    ),
-    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
-    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
-  )
-```
-:::
-:::::
-
-::: notes
-Para darle más sentido al modelo factorial es común buscar la "estructura simple".
-Es decir, asumimos que la varianza de cada ítem está explicada solamente por un factor latente.
-Los pesos factoriales de un ítem en los demás factores son nulos.
-
-Si representamos los pesos factoriales como coordenadas, el ítem estará alineado con el eje correspondiente al factor en el que pesa.
-:::
-
-## AF: Estructura simple
-
-::: r-center
-```{r simple-structure-rotated}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-AXIS1_DEGREES <- -25
-AXIS2_DEGREES <-  30
-
-axis1_radians <- AXIS1_DEGREES |> as_units("degrees") |> set_units("radians")
-axis2_radians <- AXIS2_DEGREES |> as_units("degrees") |> set_units("radians")
-
-rot_matrix <- c(
-  cos(axis1_radians), cos(axis2_radians),
-  sin(axis1_radians), sin(axis2_radians)
-) |>
-  matrix(nrow = 2, byrow = TRUE)
-
-simple_item_loadings_rot <- simple_item_loadings |>
-  group_by(item) |>
-  mutate(
-    loading = rot_matrix %*% loading |> drop()
-  ) |>
-  left_join(simple_item_loadings_wide |> select(item, main_factor), by = "item")
-
-simple_item_loadings_rot_plot <- simple_item_loadings_rot |>
-  group_by(item)                                          |>
-  pivot_wider(names_from = factor, values_from = loading) |>
-  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
-  geom_text(
-    nudge_x = -.12,
-    color   = I("black"),
-    size    = I(5)
-  )                                                                +
-  scale_x_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "",
-  )                                                                +
-  scale_y_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = ""
-  )                                                                +
-  scale_color_manual(values = factor_colors, guide = NULL)         +
-  coord_fixed(expand = FALSE, clip = "on")                         +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    )
-  )
-
-simple_item_loadings_rot_plot + geom_point(size = I(4))
-```
-:::
-
-::: notes
-Ahora bien, la mayoría de procedimientos factoriales, exploratorios al menos, arrojan soluciones de este tipo: un primer factor que maximiza toda la varianza posible, el segundo (ortogonal al primeroo) maximiza la varianza restante, y si hay más, hacen lo mismo, dando ejes siempre ortogonales a los anteriores.
-:::
-
-## AF: Estructura simple
-
-<!-- # TODO: Añadir ángulos? -->
-
-::: r-center
-```{r simple-structure-rotated-axes}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-rotated_axes <- tibble(
-  slope  = c(axis1_radians, axis2_radians) |> tan() |> drop_units(),
-  factor = c('F1', 'F2')
-)
-
-simple_item_loadings_rot_plot +
-  geom_abline(
-    data    = rotated_axes,
-    mapping = aes(slope = slope, intercept = 0),
-    color   = c(PALETTE_BLUE["light"], PALETTE_RASPBERRY["light"]),
-    linewidth = LINE_WIDTH
-  ) +
-  geom_point(size = I(4)) +
-  annotate(
-    "text",
-    x = -1.35, y = c(-.62, .78),
-    label = rotated_axes |> pull(factor),
-    size  = 9
-  )
-```
-:::
-
-::: notes
-Para obtener esa estructura simple, muchas veces necesitamos usar una rotación oblicua de los ejes.
-Cualquier procedimiento de rotación busca alinearse lo más posible con al menos algunos de los ítems, como vemos aquí.
-Pero esto es una situación ideal.
-Nunca van a quedar, en un modelo exploratorio, los ítems perfectamente alineados con un eje, dando cero exacto en los pesos de los factores que no le corresponden.
-En un modelo confirmatorio sí, forzamos que sea así, pero a costa de perder ajuste claro (y normalmente permitiendo también que los factores correlacionen).
-:::
-
-## AF: Estructura compleja
-
-<br>
-
-::::: columns
-::: column
-```{r path-complex-structure}
-# TODO: Syntax from tibble
-complex_spec <- "
-  F1 =~ I1 + I2 + I3
-  F2 =~ I3 + I4 + I5 + I6
-
-  F1 ~~ F2
-"
-
-edge_colors <- c(
-  F1_COLOR |> rep(3),
-  PALETTE_RASPBERRY["light"], # Cross-loading
-  F2_COLOR |> rep(3)
-)
-
-complex_spec     |>
-  semPlotModel() |>
-  semPaths(
-    border.color = c(
-      item_colors,       # Item colors
-      F1_COLOR, F2_COLOR # Factor colors
-    ),
-    edge.color   = c(
-      edge_colors, # Loading colors
-      "black"      # Correlation color
-    ),
-    residuals    = FALSE,
-    thresholds   = FALSE,
-    intercepts   = FALSE,
-    fixedStyle   = 1,
-    curve        = 3,
-    label.cex    = 1.6,
-    sizeMan      = 10,
-    sizeLat      = 14,
-    border.width = 3,
-    edge.width   = 3,
-    mar          = MARGINS
-  )
-```
-:::
-
-::: {.column .fragment}
-```{r scatter-complex-structure}
-LIM_INF <- -1.5
-LIM_SUP <-  2
-
-axis_lims <- c(LIM_INF, LIM_SUP)
-lim_range <- LIM_SUP - LIM_INF
-axis_pos  <- (-LIM_INF) / lim_range
-
-item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
-
-complex_item_loadings_wide <- complex_item_loadings |>
-  group_by(item)                                    |>
-  arrange(desc(loading))                            |>
-  mutate(main_factor = first(factor))               |>
-  ungroup()                                         |>
-  pivot_wider(names_from = factor, values_from = loading)
-
-item3_spikes <- complex_item_loadings_wide      |>
-  filter(item == 'I3')                          |>
-  select(starts_with('F'))                      |>
-  add_column(type = "end")                      |> 
-  bind_rows(tibble(F1 = 0, F2 = 0))             |>
-  complete(F1, F2, fill = list(type = "start")) |>
-  filter(F1 != 0 | F2 != 0)
-
-item3_spikes <- item3_spikes                           |> 
-  bind_rows(item3_spikes |> slice(3))                  |>
-  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
-  pivot_wider(names_from = type, values_from = F1:F2)
-
-spike_colors = c(
-  'F1' = unname(PALETTE_BLUE['light']),
-  'F2' = unname(PALETTE_RASPBERRY['light'])
-)
-
-complex_item_loadings_wide |>
-  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
-  geom_segment(
-    data    = item3_spikes,
-    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = spike_colors
-  )                                                                +
-  geom_point(size = I(4))                                          +
-  geom_text(
-    nudge_x =  .1,
-    nudge_y = -.08,
-    color   = I("black"),
-    size    = I(5)
-  )                                                                +
-  scale_x_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "F2",
-    position = "top"
-  )                                                                +
-  scale_y_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "F1",
-    position = "right"
-  )                                                                +
-  scale_color_manual(values = factor_colors, guide = NULL)         +
-  coord_fixed(expand = FALSE, clip = "on")                         +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    ),
-    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
-    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
-  )
-```
-:::
-:::::
-
-::: notes
-Sin embargo, en ocasiones no es posible encontrar la estructura simple.
-Es cuando aparecen lo que llamamos "pesos cruzados".
-
-Esto da lugar a una "estructura compleja", que en realidad es más común que la estructura simple, en general.
-:::
-
-## AF: Estructura compleja
-
-<br>
-
-::::: columns
-::: column
-```{r path-method-factor}
-# TODO: Syntax from tibble
-method_spec <- "
-  F1 =~ I1 + I2 + I3 + I4 + I5 + I6
-  M1 =~ I4 + I5 + I6
-"
-
-edge_colors <- c(
-  F1_COLOR |> rep(6),
-  F2_COLOR |> rep(3)
-)
-
-method_spec      |>
-  semPlotModel() |>
-  semPaths(
-    border.color = c(
-      F1_COLOR |> rep(6), # Item colors
-      F1_COLOR, F2_COLOR  # Factor colors
-    ),
-    edge.color   = edge_colors, # Loading colors
-    residuals    = FALSE,
-    thresholds   = FALSE,
-    intercepts   = FALSE,
-    fixedStyle   = 1,
-    curve        = 3,
-    label.cex    = 1.6,
-    sizeMan      = 10,
-    sizeLat      = 14,
-    border.width = 3,
-    edge.width   = 3,
-    mar          = MARGINS
-  )
-```
-:::
-
-::: column
-```{r scatter-method-factor}
-method_item_loadings <- tribble(
-  ~item, ~factor, ~loading,
-  'I1',  'F1',     ITEM_BASE_LOADING,
-  'I1',  'M1',     0,
-  'I2',  'F1',     ITEM_BASE_LOADING + 2 * ITEM_DIST,
-  'I2',  'M1',     0,
-  'I3',  'F1',     ITEM_BASE_LOADING + 3 * ITEM_DIST,
-  'I3',  'M1',     0,
-  'I4',  'F1',    -ITEM_BASE_LOADING,
-  'I4',  'M1',     ITEM_BASE_LOADING - 2 * ITEM_DIST,
-  'I5',  'F1',     ITEM_BASE_LOADING + ITEM_DIST,
-  'I5',  'M1',    -ITEM_BASE_LOADING + 2 * ITEM_DIST,
-  'I6',  'F1',    -ITEM_BASE_LOADING - ITEM_DIST,
-  'I6',  'M1',    -ITEM_BASE_LOADING + 3 * ITEM_DIST
-)
-
-method_item_loadings                                      |>
-  pivot_wider(names_from = factor, values_from = loading) |>
-  ggplot(mapping = aes(F1, M1, label = item))              +
-  geom_point(size = I(4), color = F1_COLOR)                +
-  geom_text(
-    nudge_y = -.12,
-    color   = I("black"),
-    size    = I(5)
-  )                                                        +
-  scale_x_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "M1",
-    position = "top"
-  )                                                        +
-  scale_y_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "F1",
-    position = "right"
-  )                                                        +
-  scale_color_manual(values = factor_colors, guide = NULL) +
-  coord_fixed(expand = FALSE, clip = "on")                 +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    ),
-    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
-    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
-  )
-```
-:::
-:::::
-
-::: notes
-Por ejemplo, si tenemos ítems directos e inversos, podemos modelar lo que llamamos "factores de método": Una variable latente para cada factor sustantivo (sólo uno en este caso), y por ejemplo uno para los ítems formulados en negativo.
-:::
-
-## Modelo factorial clásico
-
-<br>
-
-$$
-Y_1 = \tau_1 + \lambda_{11} F_1^p + \lambda_{21} F_2^p + ... + \lambda_{n1} F_n^p + U_1^p \\
-\vdots                                                                            \\
-Y_i = \tau_i + \lambda_{1i} F_1^p + \lambda_{2i} F_2^p + ... + \lambda_{ni} F_n^p + U_i^p
-$$
-
-<br>
-
-::: fragment
-$$
-Y_i = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
-$$
-:::
-
-::: notes
-Un modelo factorial no es más que una serie de regresiones lineales (una por cada variable observable) con variables latentes.
-Las variables "respuesta" o "criterio" es la variable observable, y su varianza está explicada o "predicha" por las variables latentes, que son los factores y los términos de error de cada ítem.
-Lo que ocurre es que aquí a "coeficientes de regresión" los llamamos "pesos factoriales" y al término error, "unicidad" o "varianza única".
-
-El producto de los pesos por las puntuaciones factoriales se puede expresar en "forma vectorial", de esta manera.
-La "T" del vector de pesos factoriales significa que está "traspuesto", es decir, que se pone horizontalmente (esto es una convención en álgebra matricial).
-El producto del vector de pesos por el de puntuaciones factoriales es lo que se llama un "producto vectorial", que consiste simplemente en multiplicar dos a dos los términos de cada vector y sumar todos.
-:::
-
-## Modelo factorial clásico {.smaller}
-
-::::::: columns
-::::: column
-$$
-  I_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
-$$
-
-$$
-  \mathbf{F}^1 =
-    \begin{bmatrix}
-      0.5 \\
-      1
-    \end{bmatrix}
-$$
-
-::: {.fragment fragment-index="1"}
-$$
-  \mathbf{\lambda}_{1} =
-    \begin{bmatrix}
-      -1 \\
-       0.5
-    \end{bmatrix}
-$$
-:::
-
-::: fragment
-$$
-  \begin{align}
-    \tau_1 &=  1 \\
-    U_1^1  &= -0.5
-  \end{align}
-$$
-:::
-:::::
-
-::: {.column .fragment fragment-index="2"}
-```{r fa_example_plot}
-#| cache: false
-AXIS_GRID <- -1:1
-
-# Generate data for simple and complex factorial structures
-fa_example <- tribble(
-  ~object, ~factor, ~loading,
-  'I~1~',  'F1',    -1,
-  'I~1~',  'F2',     0.5,
-  'F^1^',  'F1',     0.5,
-  'F^1^',  'F2',     1
-)
-
-fa_example_wide <- fa_example |>
-  pivot_wider(names_from = factor, values_from = loading)
-
-fa_example_spikes <- fa_example_wide |>
-  add_column(type = "end")           |> 
-  bind_rows(
-    tribble(
-      ~object, ~F1,  ~F2, ~type,
-      "I~1~",   0,   0.5, "start",
-      "F^1^",   0,   1,   "start",
-      "I~1~",  -1,   0,   "start",
-      "F^1^",   0.5, 0,   "start",
-    )
-  )
-
-fa_example_spikes <- fa_example_spikes |> slice(1:2)        |>
-  bind_rows(fa_example_spikes)                              |>
-  add_column(factor = paste0('F', 1:2 |> rep(2, each = 2))) |>
-  pivot_wider(names_from = type, values_from = F1:F2)
-
-fa_example_wide |>
-  ggplot(mapping = aes(F1, F2, label = object))             +
-  geom_hline(
-    yintercept = 0,
-    color      = PALETTE_BLUE["light"],
-    linewidth  = GRID_WIDTH
-  )                                                         +
-  geom_vline(
-    xintercept = 0,
-    color      = PALETTE_RASPBERRY["light"],
-    linewidth  = GRID_WIDTH
-  )                                                         +
-  geom_segment(
-    data        = fa_example_spikes,
-    mapping     = aes(
-      F1_start, F2_start,
-      xend  = F1_end, yend = F2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth   = LINE_WIDTH,
-    linetype    = '43',
-    color       = spike_colors |> rep(each = 2)
-  )                                                        +
-  geom_point(size = I(4))                                  +
-  geom_richtext(
-    nudge_x = .15,
-    nudge_y = .10,
-    color   = I("black"),
-    size    = I(5),
-    lineheight = 1.2,
-    fill = NA,
-    label.color = NA
-  )                                                        +
-  scale_x_continuous(
-    limits       = axis_lims,
-    breaks       = AXIS_GRID,
-    name         = NULL,
-    minor_breaks = NULL,
-    sec.axis     = dup_axis(name = "F~2~", labels = NULL)
-  )                                                        +
-  scale_y_continuous(
-    limits       = axis_lims,
-    breaks       = AXIS_GRID,
-    name         = NULL,
-    minor_breaks = NULL,
-    sec.axis     = dup_axis(name = "F~1~", labels = NULL)
-  )                                                        +
-  scale_color_manual(values = factor_colors, guide = NULL) +
-  coord_fixed(expand = FALSE, clip = "on")                 +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_markdown(hjust = axis_pos),
-    axis.title.y       = element_markdown(vjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    )
-  )
-```
-:::
-:::::::
-
-::: fragment
-$$
-  I_1 = 1 + \left[ -1, 0.5 \right]
-    \begin{bmatrix}
-      0.5 \\
-      1
-    \end{bmatrix}
-    - 0.5 = 1 + (-0.5 + 0.5) - 0.5 = 0.5
-$$
-:::
-
-::: notes
-Veámoslo con un ejemplo.
-Tenemos el participante 1, en nuestro estudio, que tiene esas puntuaciones factoriales; usamos el superíndice para denotar al participante.
-Por ejemplo, en el primer rasgo, tiene un 0.5 (nos da igual la distribución pero supongamos que está estandarizada, luego está media desviación típica por encima de la media).
-En el segundo rasgo, tiene una puntuación que está una desviación típica por encima de la media.
-
-Los pesos factoriales del ítem 1 son estos: -1 para el primer rasgo (por lo cual, cuanto mayor sea la puntuación factorial en el primer factor, menor va a ser la respuesta observada en el ítem 1), y 0.5 para el segundo.
-
-Aquí vemos representados el ítem y la persona, en las mismas coordenadas cartesianas.
-Y luego el ítem tiene una intersección de 1, y la persona 1 tiene un -0.5 en el "error" de esa variable $I_1$.
-
-Su respuesta se calcularía así: (hacer los cálculos).
-Luego esta persona (el particpante 1) tiene la respuesta o valor observado 0.5 en esta variable observada.
-Lo mismo podríamos hacer con el resto de variables y de participantes, pero obviamente nosotros lo que tenemos son estas respuestas observadas, y lo que hacemos con el AF es estimar todos estos parámetros del modelo, en lugar de como lo hemos hecho ahora.
-
-Ahora bien, esta respuesta observada que hemos calculado aquí es cuantitativa.
-Tenemos costumbre de ver el análisis factorial en ese contexto, de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
-Sin embargo, el llamado "AF de ítems" se aplica también a variables discretas ordinales (e idealmente se debería aplicar así a ítems tipo Likert, aunque a veces no se haga, por simplificar).
-:::
-
-## Análisis factorial de ítems {.smaller}
-
-::::::::: columns
-::::: {.column width="45%"}
-$$
-  y_i^* = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
-$$
-
-<br>
-
-::: {.fragment fragment-index="1"}
-$$
-  Y_i = \left\{ \begin{matrix}
-      1 &si &y_i^* \geq 0\\
-      0 &si &y_i^* <    0
-    \end{matrix} \right.
-$$
-:::
-
-<br>
-
-::: {.fragment fragment-index="2"}
-$$
-  \begin{align}
-    P(Y_i = 1) &= \int_0^{-\infty} \mathcal{N}(y_i^*) \delta y_i^* \\
-               &= \Phi_N \left(
-                 \frac{ \tau_i + \mathbf{\lambda}_i^T \mathbf{F} }
-                   { \sqrt{var(U_i)} }
-               \right)
-  \end{align}
-$$
-:::
-:::::
-
-::::: {.column width="55%"}
-::: {.fragment fragment-index="1"}
-```{r ifa-density}
-#| fig-asp:   0.5
-#| fig-width: 8
-
-PROB_AXIS      <- seq(-3, 3, by = .01)
-ERROR_POINTS   <- seq(-3, 3, length.out = 5)
-ITEM_LOADING   <-  0.45
-ITEM_INTERCEPT <- -0.3
-
-item_unique_var <-  1 - sqrt(ITEM_LOADING)
-
-latent_response <- tibble(
-  point    = ERROR_POINTS |> rep(500000),
-  response = rnorm(
-    length(point),
-    mean = ITEM_INTERCEPT + ITEM_LOADING * point,
-    sd   = sqrt(item_unique_var)
-  )
-)
-
-# Plot: Upper Panel - Linear regression with Gaussian error distributions
-latent_response |>
-  ggplot(
-    aes(x = response, y = point, fill = after_stat(x) > 0, group = point)
-  )                            +
-  geom_density_ridges_gradient(
-    scale          =  .2,
-    rel_min_height = 0.001,
-    color          = PALETTE_BLUE["normal"]
-  )                              +
-  labs(x = "y~i~^*^", y = NULL)  +
-  scale_y_continuous(
-    breaks       = ERROR_POINTS,
-    minor_breaks = NULL,
-    labels       = NULL,
-    limits       = range(ERROR_POINTS) * 1.12,
-    expand       = expansion()
-  )           +
-  scale_x_continuous(breaks = 0) +
-  scale_fill_manual(
-    values = c(
-      `FALSE` = unname(PALETTE_BLUE["white"]),
-      `TRUE`  = unname(PALETTE_BLUE["normal"])
-    )
-  )                              +
-  theme(
-    axis.ticks.x = element_blank(),
-    axis.title.y = element_markdown(angle = 90),
-    legend.position = "none",
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )                              +
-  coord_flip()
-```
-:::
-
-::: {.fragment fragment-index="2"}
-```{r ifa-curve}
-#| fig-asp:   0.5
-#| fig-width: 8
-
-probit_curve_data <- tibble(
-  score    = PROB_AXIS,
-  response = ITEM_INTERCEPT + ITEM_LOADING * score,
-  prob     = pnorm(response, sd = sqrt(item_unique_var))
-)
-
-# Plot: Lower Panel - Probit curve
-probit_curve_data |>
-  ggplot(mapping = aes(x = score, y = prob))                        +
-  geom_line(color = PALETTE_BLUE["normal"], linewidth = LINE_WIDTH) +
-  labs(x = "E(*y~i~^*^*)", y = "P(Y~i~ = 1)")                       +
-  scale_y_continuous(
-    breaks       = 0:1,
-    minor_breaks = NULL,
-    limits       = 0:1,
-    expand       = expansion(mult = c(0, .05)),
-    oob = oob_keep
-  )  +
-  scale_x_continuous(
-    breaks       = ERROR_POINTS,
-    minor_breaks = NULL,
-    limits = range(ERROR_POINTS) * 1.12,
-    expand = expansion()
-  ) +
-  theme(
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
-  )
-```
-:::
-:::::
-:::::::::
-
-::: notes
-Lo que hace el análisis factorial de ítems es aplicar ese "modelo lineal" a variables ordinales, haciendo una generalización.
-(En este caso vamos a simplificarlo a variables dicotómicas; p.ej. ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
-Esto es una simplificación que rara vez se utiliza en cuestionarios no cognitivos, pero es interesante estudiarlo porque otros modelos pueden considerarse extensiones de este modelo.
-El AF de ítems asume una nueva "variable latente" lineal ($I_1^*$), con la misma fórmula anterior
-
-Después se aplica una "función de umbral": La respuesta es correcta o afirmativa, codificada como 1, si la variable latente es mayor (o igual) que 0, o incorrecta o negativa, codificada como 0, si es menor.
-Eso lo que da, en lugar de un respuesta observada (cuantitativa), es una "probabilidad" de responder correcta o "afirmativamente" a un ítem.
-:::
-
-## AF: Puntuaciones factoriales
-
-<br>
-
-:::::: columns
-:::: column
-Métodos:
-
--   Regresión
-
--   Bartlett
-
-::: {.fragment fragment-index="2"}
--   **Máxima Verosimilitud**
-
--   **Empirical Modal Bayes**
-:::
-::::
-
-::: {.column .fragment fragment-index="1"}
-```{r path-regression-scores}
-# item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
-# factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
-
-# NOTE: Manual call to qgraph:
-edge_list <- tribble(
-  ~rhs, ~lhs,
-  1,    7,
-  2,    7,
-  3,    7,
-  4,    8,
-  5,    8,
-  6,    8,
-  7,    8,
-  8,    7
-)
-vSize <- c(10 |> rep(6), 14 |> rep(2))
-
-qgraph(
-  edge_list,
-  labels = c("I1", "I2", "I3", "I4", "I5", "I6", "F1", "F2") |>
-    setNames(nm = _),
-  label.cex = 1.6,
-  shape = "square" |> rep(8),
-  vsize = vSize, 
-  vsize2 = vSize,
-  color = "background" |> rep(8),
-  border.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), F1_COLOR, F2_COLOR),
-  border.width = 3,
-  bidirectional = c(FALSE |> rep(6), TRUE, TRUE),
-  directed = TRUE |> rep(8),
-  curve = c(NA |> rep(6), 3, 3),
-  edge.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), "black"),
-  edge.width = 3,
-  lty = 1,
-  layout = matrix(
-    c(
-      -1, -0.6, -0.2, 0.2, 0.6, 1, -0.6, 0.6,
-      -0.5 |> rep(6), 0.5 |> rep(2)
-    ),
-    ncol = 2
-  ),
-  mar = c(12, 1.3, 8, 1.3)
-)
-```
-:::
-::::::
-
-<!--# TODO: Diagrama de CFA y de "CFA inverso" -->
-
-::: notes
-Una vez tenemos los parámetros del modelo factorial estimado, entre otras cosas, podemos aplicarlo para estimar las puntuaciones factoriales de los participantes.
-Para hacer esto hay varios métodos.
-
-El método de regresión y el de Bartlett se aplican en AF lineal y están implementados en SPSS.
-Consisten a grandes rasgos en calcular la puntuación factorial como una suma ponderada de las variables observables.
-Es decir, algo así como "darle la vuelta a las flechas" de causalidad.
-Lo que tiene más sentido, desde el punto de vista estadístico, es "estimar" qué puntuaciones factoriales hacen más probable para una participante dar las respuestas que ha dado.
-
-Eso es lo que hacen los dos métodos de abajo (SPSS no los implementa creo), Máxima Verosimilitud, o esimación Modal Bayesiana.
-Si habéis hecho esto con variables categóricas, entonces habéis aplicado la Teoría de Respuesta al Ítem, y a continuación veremos por qué.
-:::
-
-## Paralelismo AF - TRI
-
-<br>
-
-::::: columns
-::: column
-![](../www/slideshow-assets/cover_McDonald-1999.jpg){height="300px"}
-
-[@mcdonald_test_1999]
-:::
-
-::: column
-<!--# TODO: Citation by McDonald -->
-:::
-:::::
-
-::: notes
-En este libro, que se publica en 1999 (fijaos en el subtítulo de "unified treatment"), McDonald indica que estos modelos, factoriales, TRI, son a fin de cuentas "isomóficos": Sin ser exactamente iguales ni partir de los mismos supuestos, las predicciones que hacen uno y otro son indistinguibles en la práctica.
-:::
-
-## Teoría de respuesta al ítem {.smaller}
-
-Modelo logístico de 2 parámetros [@birnbaum_latent_1968]:
-
-::::::::::: columns
-:::::: {.column width="60%"}
-```{r irt-curves}
-#| fig-asp:   0.6
-#| fig-width: 8
-
-ITEMS_PALETTE <- c(
-  PALETTE_BLUE["normal"],
-  PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"]
-) |>
-  unname() |>
-  setNames("prob" |> paste(1:3, sep = '_'))
-
-ITEMS <- tribble(
-  ~item, ~a,   ~b,
-   1,     1,    0,
-   3,     1,    1.5,
-   2,     2.5,  0
-)
-
-logit <- function(x) 1 / (1 + exp(-x))
-irf   <- function(a, b, theta) logit(a * (theta - b))
-
-logit_curve_data <- tibble(
-  latent_trait = PROB_AXIS,
-  prob         = ITEMS |>
-    select(-item) |>
-    pmap(irf, theta = latent_trait) |>
-    bind_cols()
-) |>
-  unnest(
-    prob,
-    names_sep = '_'
-  ) |>
-  rename_with(str_remove, pattern = '\\.{3}')
-
-logit_curve_data_long <- logit_curve_data |>
-  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
-
-logit_curve_data_long |>
-  ggplot(
-    mapping = aes(
-      latent_trait, prob,
-      color = item,
-      group = after_scale(color)
-    )
-  )                                        +
-  geom_line(linewidth = LINE_WIDTH)        +
-  labs(x = "*&theta;*", y = "P(Y~i~ = 1)") +
-  scale_y_continuous(
-    breaks       = 0:1,
-    minor_breaks = NULL,
-    limits       = 0:1,
-    expand       = expansion(mult = c(0, .05))
-  )                                        +
-  scale_x_continuous(
-    breaks       = ERROR_POINTS,
-    minor_breaks = NULL,
-    limits = range(ERROR_POINTS) * 1.12,
-    expand = expansion()
-  )                                        +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  theme(
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
-  )
-```
-
-::: r-stack
-[$a_1$ = 1.0; $b_1$ = 0.0]{style="color:#5c6eb1;"}
-:::
-
-::: r-stack
-[$a_2$ = 1.0; $b_2$ = 1.5]{style="color:#90214a;"}
-:::
-
-::: r-stack
-[$a_3$ = 2.5; $b_3$ = 0.0]{style="color:#749f4c;"}
-:::
-::::::
-
-:::::: {.column width="40%"}
-::: {.fragment fragment-index="1"}
-$$
-  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
-$$
-:::
-
-<br>
-
-::: {.fragment fragment-index="2"}
-$$
-  \begin{align}
-    d_i &= -a_i b_i \\
-    P(Y_i = 1) &= \Phi_L (a_i \theta + d_i))
-  \end{align}
-$$
-:::
-
-<br>
-
-::: {.fragment fragment-index="3"}
-$$
-  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
-$$
-:::
-
-<br>
-::::::
-:::::::::::
-
-::: notes
-La TRI, por su lado, asume que hay cada persona tiene una puntuación latente, y que su respuesta depende de esa puntuación, con un parámetro de discriminación que determina cómo de bien o mal mide el ítem, y uno de dificultad (aunque aquí prefiero llamarlo "de posición") que determina lo fácil o difícil que será dar una respuesta positiva (afirmativa, correcta, codificada como 1).
-Esa posición representa el nivel de rasgo latente que da lugar a una probabilidad de .5 de dar la respuesta positiva (en este modelo; en general representa el punto de inflexión de la curva, es decir, donde pasa de hacerse cada vez más pronunciada a empezar a aplanarse).
-
-Ahora bien, la posición se puede reformular como un parámetro de intersección.
-
-Como se puede ver, esto también se trata de un "modelo lineal general".
-En este caso, en lugar de una función probit (normal acumulada), la función de enlace es la función "logit".
-Es decir, la respuesta a cada ítem se modela como una regresión logistica donde los predictores son las variables latentes.
-:::
-
-## TRI multidimensional
-
-$$
-  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
-$$
-
-::::: columns
-::: column
-```{r mirt-model-irs-3d}
-#| fig-width:  4.6
-#| fig-height: 4.6
-PROB_AXIS_2D <- seq(-3, 3, by = .1)
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1,    0.5,  -0.5
-)
-
-irs2d <- function(a_1, a_2, d, theta_1, theta_2) {
-
-  logit(a_1 * theta_1 + a_2 * theta_2 + d)
-}
-
-mirt_item_probs <- expand_grid(
-  trait_2 = PROB_AXIS_2D,
-  trait_1 = PROB_AXIS_2D
-) |>
-  mutate(
-    mirt_item |>
-      select(-item) |>
-      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
-      bind_cols()
-  ) |>
-    rename(prob = `...1`)
-
-mirt_prob_surface <- mirt_item_probs |>
-  pivot_wider(names_from = trait_1, values_from = prob) |>
-    select(-trait_2) |>
-    as.matrix()
-
-plot_ly(
-  x = ~PROB_AXIS_2D,
-  y = ~PROB_AXIS_2D,
-  z = ~mirt_prob_surface,
-  colorscale = list(c(0, 1), PALETTE_BLUE["normal"] |> rep(2)),
-  type = 'surface',
-  hovertemplate = 'P(I<sub>i</sub> = 1) = %{z:.2f}<extra></extra>',
-  showscale = FALSE,
-  hoverinfo = 'Z'
-) |>
-  layout(
-    scene = list(
-      xaxis = list(title = "<i>θ</i><sub>1</sub>"),
-      yaxis = list(title = "<i>θ</i><sub>2</sub>"),
-      zaxis = list(title = "P(I<sub>i</sub> = 1)"),
-      camera = list(
-        center = list(x = -.1, z = -.2),
-        eye = list(x = -.12, y = -1.8, z = 1)
-      )
-    )
-  ) |>
-  config(displayModeBar = FALSE, scrollZoom = FALSE)
-```
-:::
-
-::: column
-```{r mirt-model-irs-contour}
-contour_plot <- mirt_item_probs |>
-  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_contour2(
-    color = F1_COLOR,
-    linewidth = LINE_WIDTH,
-    breaks = c(.05, .1, .2, .3, .5, .7, .8, .9, .95),
-    label.placer = label_placer_fraction(frac = .90, rot_adjuster = 0)
-  ) +
-  scale_x_continuous(
-    minor_breaks = NULL,
-    limits = range(PROB_AXIS_2D),
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    minor_breaks = NULL,
-    limits = range(PROB_AXIS_2D),
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  coord_fixed() +
-  theme(
-    axis.ticks       = element_blank(),
-    axis.line        = element_blank(),
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )
-
-contour_plot
-```
-:::
-:::::
-
-::: notes
-Esto se puede extender, de manera bastante directa, asumiendo que la respuesta se modela como una "regresión logística multivariada".
-Aquí aparece otra vez la discriminación pero ahora en forma vectorial, que se multiplica por el vector de rasgo latente de la persona, y un parámetro de intersección, como antes (pero no puede haber un parámetro de dificultad simplemente, o uno por cada dimensión, porque serían indistinguibles entre sí).
-Esta es una de las extensiones multidimensionales (más antiguas, y probablemente más obvias) del modelo logístico de 2 parámetros: El modelo logístico de 2 parámetros multidimensional compensatorio.
-Pertenece a una familia de modelos llamados "compensatorios" en TRI, ya que un nivel bajo en una puntuación latente se compensa con un nivel alto en otra, dando lugar a la misma probabilidad de respuesta.
-:::
-
-## TRI multidimensional: Ejemplo {.smaller}
-
-:::::: columns
-::: column
-$$
-  P(Y_1 = 1) = \Phi_L (\mathbf{a}_1^T \mathbf{\theta} + d_1)
-$$
-
-```{r mirt-item-response-example}
-mirt_example <- tribble(
-  ~theta_1, ~theta_2,
-  1,        -1
-)
-
-mirt_example_spikes <- mirt_example |>
-  add_column(type = "end")          |> 
-  bind_rows(
-    tribble(
-      ~theta_1, ~theta_2, ~type,
-      0,        -1,        "start",
-      1,         0,        "start"
-    )
-  )
-
-mirt_example_spikes <- mirt_example_spikes |>
-  slice(1) |>
-  bind_rows(mirt_example_spikes) |>
-  add_column(factor = paste0('F', 1:2 |> rep(2))) |>
-  pivot_wider(names_from = type, values_from = starts_with("theta"))
-
-contour_plot +
-  geom_segment(
-    data    = mirt_example_spikes,
-    mapping = aes(
-      theta_1_start, theta_2_start,
-      xend = theta_1_end,
-      yend = theta_2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = PALETTE_BLUE["light"]
-  ) +
-  geom_point(
-    data = mirt_example,
-    mapping = aes(theta_1, theta_2),
-    inherit.aes = FALSE,
-    size = I(4)
-  )
-```
-:::
-
-:::: column
-
-\begin{matrix}
-  \mathbf{\theta} =
-    \begin{bmatrix}
-      1 \\
-      -1
-    \end{bmatrix}
-
-  &\begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1 \\
-        0.5
-      \end{bmatrix} \\
-    d_1 &= -0.5
-  \end{align}
-\end{matrix}
-
-::: fragment
-$$
-  \begin{align}
-    P(Y_1 = 1) &= \Phi_L \left([1, 0.5]
-      \begin{bmatrix}
-         1 \\
-        -1
-      \end{bmatrix}
-      -0.5 \right) \\
-      &= \Phi_L \left(1 · 1 + 0.5 · (-1) -0.5 \right) \\
-      &= \Phi_L \left(1 -0.5 -0.5 \right) = \Phi_L \left(0 \right) \\
-      &= \frac{1}{1 + e^{-0}} = \frac{1}{1 + 1} = \frac{1}{2} = 0.5
-  \end{align}
-$$
-:::
-::::
-::::::
-
-::: notes
-Veamos un ejemplo de cuál sería la probabilidad de respuesta de una persona con puntuaciones de rasgo de 1 y -1.
-
-(Hacer cálculos matriciales)
-:::
-
-## AF en métrica TRI {.smaller}
-
-<br>
-
-$$
-  \begin{align}
-    P(Y_i = 1) &= \Phi_N \left(
-      \frac{\mathbf{\lambda}_i^T \mathbf{F} + \tau_i}{\sqrt{var(U_i)}}
-    \right) \\
-    P(Y_i = 1) &= \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
-  \end{align}
-$$
-
-<br>
-
-::: fragment
-@takane_relationship_1987:
-
-$$
-  \begin{align}
-    \mathbf{a}_i &= \frac{1.7}{\sqrt{var(U_i)}} \mathbf{\lambda}_i \\
-    d_i &= \frac{1.7}{\sqrt{var(U_i)}} \tau_i
-  \end{align}
-$$
-:::
-
-::: notes
-Claro, si ponemos las ecuaciones del AF y de TRI lado a lado, es fácil ver que existe un paralelismo entre ambos modelos.
-
-Pues resulta que estas dos curvas, la normal y la logística, son muy parecidas; tanto, que es posible transformar los parámetros de la métrica del AF a métrica de TRI, obteniendo modelos indistinguibles en la práctica.
-Las dos curvas se diferencian por este factor de 1.7, pero aplicándo ese factor y escalando por la desviación estándar de la unicidad del ítem, obtenemos los mismos parámetros en métrica de TRI mulidimensional con función de respuesta logística.
-Y de hecho, esto es necesario y se hace así para hacer la estimación máximo-verosimil de la que hablábamos antes.
-:::
-
-## Parámetros multidimensionales de TRI {.smaller}
-
-<br>
-
-$$
-  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
-$$
-
-::: fragment
-$$
-  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
-$$
-:::
-
-::: fragment
-@reckase_difficulty_1985:
-
-> \[...\] means of describing the characteristics of an item that takes into account the dimensionality of the skills \[...\] can then be used to determine how or if it is possible to compare items that measure different combinations of abilities.
-:::
-
-::: notes
-A pesar de este paralelismo que os he mostrado, y de la utilidad de los modelos de TRI, hay un problema con la formulación del modelo que os he planteado, y es que como habéis podido ver los parámetros no son fácilmente interpretables.
-
-La intersección del modelo por ejemplo no tiene una interpretación clara.
-Si recordáis en la TRI unidimensional el parámetro de posición permitía comparar ítems entre sí en cuanto a su dificultad (en el caso de items cognitivos).
-Esto en AF no es muy habitual (prestamos poca atención a la "intersección" o "umbral"), porque no nos interesa tanto cómo de probable sea dar una respuesta positiva o negativa a un ítem, pero podría ayudarnos por ejemplo a diagnosticar ítems que estén mal diseñados, ya que nos da una idea de si va a haber muchas o pocas respuestas positivas a un ítem.
-
-Por eso Reckase dice, ya en 1985, que se necesita una forma de describir las características de los ítems que tenga en cuenta la dimensionalidad de las habilidades (los rasgos latentes en general).
-Daos cuenta de que entonces aún no se sabe nada de ese paralelismo con el AF.
-Además menciona que de esa manera se puede intentar comparar ítems que miden diferentes combinaciones de habilidades (fijaos en la importancia que tiene el dónde están los ítems en el continuo, en TRI, en contraste con AF, como comentábamos al principio).
-:::
-
-## Dificultad multidimensional {.smaller}
-
-@reckase_difficulty_1985:
-
-> \[...\] the most reasonable point to use in defining the \[multidimensional difficulty\] for an item in the multidimensional space is the point where the item is most discriminating.
-
----(p. 402)
-
-$$
-  MID = \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
-$$
-
-::: fragment
-> The distance can be interpreted much like a $b$ parameter from unidimensional IRT
-
----(p. 405)
-:::
-
-::: notes
-Por ese motivo propone un parámetro de "dificultad" (lo llama, para nosotros sería de posición) que sea de algún modo equivalente a la dificultad en el modelo unidimensional.
-Ese punto es donde la curva (superficie) de probabilidad de respuesta tiene una inflexión (a lo largo de toda ella, el punto más cercano al origen, donde la recta que lo une al origen es perpendicular a la recta de inflexión de la curva).
-
-Este parámetro se puede entender como una "distancia **con signo**" desde el origen hasta el punto de posición del ítem.
-:::
-
-## Dificultad multidimensional
-
-::::::: columns
-::: column
-```{r mid-representation}
-#| fig-height: 8
-#| fig-width:  8
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mirt_item_spikes <- mirt_item_coords |>
-  select(item, trait_1 = origin_1, trait_2 = origin_2) |>
-  bind_rows(tibble(item = 1, trait_1 = 0, trait_2 = 0)) |>
-  add_column(type = c("end", "start")) |>
-  pivot_wider(names_from = type, values_from = starts_with("trait"))
-
-contour_plot +
-  geom_segment(
-    data    = mirt_item_spikes,
-    mapping = aes(
-      trait_1_start, trait_2_start,
-      xend = trait_1_end,
-      yend = trait_2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = PALETTE_BLUE["light"]
-  ) +
-  geom_point(
-    data = mirt_item_coords,
-    mapping = aes(origin_1, origin_2),
-    inherit.aes = FALSE,
-    size = I(4)
-  ) +
-  scale_x_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  annotate(
-    geom = "richtext",
-    x = 0.16, y = 0.15,
-    label = "*MID~i~*",
-    size = 6,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  )
-```
-:::
-
-::::: column
-::: fragment
-$$
-  cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
-$$
-:::
-
-<br>
-
-::: fragment
-$$
-  MID_i = \begin{Bmatrix}
-    \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}} \\
-    \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
-  \end{Bmatrix}
-$$
-:::
-:::::
-:::::::
-
-::: notes
-Para representar ese punto no basta con la "distancia (con signo)"; hace falta saber en qué dirección mide ese ítem también.
-Eso viene dado por los llamados "cosenos directores".
-
-Por lo tanto, la "dificultad (posición) multidimensional" viene dada por dos elementos, un número real que representa cuánto se desplaza la recta de inflexión del ítem respecto del origen y en qué sentido, y un vector de cosenos que determina la dirección en la que se desplaza.
-Así se puede interpretar un ítem, o varios, de manera relativamente sencilla: Todos los participantes cuyas coordenadas estén en la "recta de inflexión" (la recta que pasa por la posición del ítem, perpendicular a su dirección) tendrán una probabilidad de dar una respuesta positiva y negativa de .5 (dificultad media; igual probabilidad de acertar o fallar).
-Aquellos que estén pasada la recta, encontrarán el ítem "relativamente fácil" (tendrán una alta probabilidad de dar una respuesta positiva mayor a .5; lo encontrarán "más fácil que difícil").
-Lo contrario pasa para los que estén al otro lado de esa recta, tendrán más probabilidad de responder negativamente (mal), por lo que lo encontrarán difícil.
-
-De lo que este parámetro no informa es de "cómo de fácil o difícil" va a ser el ítem para aquellos participantes que están muy cerca o muy lejos de esta recta.
-Es decir, una vez estoy pasada la recta, aunque esté cerca, ¿es el ítem muy fácil, o más bien tengo que estar "muy lejos" (tener alta puntuación en uno u otro rasgo) para que sea más probable que de una respuesta positiva?
-:::
-
-## Discriminación multidimensional {.smaller}
-
-@reckase_discriminating_1991 [p. 362]:
-
-> the discriminating power of an item indicates how quickly the transition takes place from low probability to high probability \[...\].
-> A highly discriminating item divides the regions clearly---having a narrow region  \[...\] where the probabilities are intermediate in magnitude.
-
-::::: {.columns .fragment}
-::: column
-<br>
-<br>
-<br>
-
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-:::
-
-::: column
-```{r mdisc-item}
-#| fig-width:  4.7
-#| fig-height: 4.7
-#| cache: false
-
-item_vector <- mirt_item_coords |> mutate(
-  across(ends_with("1"), ~ . - origin_1),
-  across(ends_with("2"), ~ . - origin_2)
-)
-
-item_spikes <- item_vector             |>
-  slice(1 |> rep(2))                   |>
-  add_column(dim = 'F' |> paste0(1:2)) |>
-  mutate(
-    origin_1 = if_else(dim == 'F1', end_1, origin_1),
-    origin_2 = if_else(dim == 'F2', end_2, origin_2),
-  )
-
-item_vector |>
-  ggplot(
-    aes(
-      0, 0,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data = item_spikes,
-    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color     = PALETTE_BLUE["light"]
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    "richtext",
-    x = 0.3, y = 0.27,
-    label = "*MDISC~i~*",
-    size  = 5.5,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  annotate(
-    "richtext",
-    x = c(0.5, -0.13), y = c(-0.12, 0.25),
-    label         = c("*a*~1*i*~", "*a*~2*i*~"),
-    size          = 4.5,
-    label.colour  = NA,
-    label.padding = unit(0, "lines"),
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::
-
-::: notes
-Esta característica viene representada por el parámetro de "discriminación", que es una medida de la "calidad general" de un ítem.
-Los ítems "buenos" serán muy discriminativos, lo que indica que distinguen bien entre la gente que tiene nivel de rasgo alto o bajo, mientras que los "malos" tendrán una superficie de respuesta más "plana".
-
-La discriminación se puede entender con una especie de "dardo" que atraviesa de arriba abajo esa superficie perpendicularmente por esa "recta de inflexión" de la que hablábamos antes, y luego cruza el plano origen.
-Si la superficie es "totalmente plana" el dardo entra por un punto y sale por otro que está justo debajo, en las mismas coordenadas de rasgo latente.
-La "proyección" (si ponemos un foco justo encima, la sombra que forma en el plano) es un punto.
-Es decir, la "discriminación" es cero (el vector del ítem tiene longitud "cero").
-Si la pendiente es muy pronunciada (distingue bien entre niveles de rasgo) el dardo atravesará el plano origen muy lejos.
-La sombra será muy larga (es decir, la discriminación muy alta).
-El caso "extremo" es una superficie "vertical" en la recta de inflexión (un escalón; cosa imposible en ítems reales).
-En este caso, el dardo nunca cortaría al plano origen, y la discriminación sería infinita.
-:::
-
-## Discriminación multidimensional {.smaller}
-
-<!-- # TODO: Superficies de ítems con alta y baja discriminación -->
-
-## Parámetros multidimensionales
-
-@reckase_discriminating_1991:
-
-<!-- # TODO: Fórmula de MDISC -->
-
-<!-- # TODO: Gráfico del cálculo de MDISC como teorema de Pitágoras -->
-
-<!-- # TODO: Fórmula de dificultad con MDISC -->
-
-::: notes
-@reckase_discriminating_1991 hacen la deducción de la discriminación multidimensional, y resulta dar lugar a esta fórmula.
-Como veis es simplemente la "longitud" del vector de parámetros de discriminación, resultante simplemente de aplicar el teorema de Pitágoras.
-El vector es un triángulo rectángulo, $a_1$ sería un cateto, el otro cateto sería igual a $a_2$, y la hipotenusa es la longitud (las dos discriminaciones al cuadrado, sumadas, y luego la raíz cuadrada de eso).
-Esto por cierto también vale para cualquier número de dimensiones.
-Aquí estamos representando sólo 2 porque es más fácil, pero podemos calcular la suma de cuadrados estos parámetros de discriminación (sea cual sea su número), y calcular su raíz cuadrada, y eso nos daría igualmente la discriminación multidimensional (la longitud de ese vector de $a$ del ítem).
-
-Esto es en realidad el producto escalar (o producto punto) de un vector consigo mismo, que consiste en multiplicar una a una las componentes y sumarlos todos.
-En forma matricial se representaría como el vector pre-multiplicado por sí mismo.
-<!-- # TODO: Explicar la multiplicación vectorial (y poner ejemplo?) -->
-
-Esta expresión de la discriminación multidimensional se puede usar para expresar de manera más sencilla la dificultad.
-:::
-
-## Representación gráfica
-
-@ackerman_graphical_1996:
-
-::: notes
-La propuesta de representación gráfica completa (y consistente) no llega hasta 1996.
-Ackerman propone que el ítem se represente como un vector de longitud dada por la discriminación multidimensional, aplicado en la posición del ítem (y en la dirección dada por esa posición; es decir, todos los ítems "apuntan" siempre en la dirección que pasa por el origen).
-:::
-
-## Supuestos: Monotonía creciente
-
-> the probability of answering an item correctly increases monotonically with an increase in each dimension being measured
-
----@reckase_difficulty_1985 [p. 402]
-
-<!-- # TODO: Ecuación modelo y ejemplo probabilidad constante -->
-
-::: notes
-Ahora bien, como hemos visto, la TRI originalmente sólo trataba con "ítems cognitivos" (lo hemos podido ver en el propio lenguaje que utiliza Reckase: "dificultad", "habilidad", "respuesta correcta").
-Uno de los supuestos que hace es este.
-Eso tiene todo el sentido (casi), ya que si la probabilidad de un ítem "decrece" al aumentar el nivel de habilidad, es que algo está mal en el ítem: No puede ser que alguien que sepa más tenga menos probabilidad de acertar un ítem que alguien que sepa menos.
-Pero cuidado, si el ítem es "puro" de una dimensión, la probabilidad en la otra u otras dimensiones sería constante (discriminación = 0); es decir, la probabilidad puede ser "plana" (es una dimensión que no "está siendo medida por el ítem").
-:::
-
-## Generalización: Ítems inversos
-
-<!-- # TODO: Curva de ítem inverso y ejemplo -->
-
-<!-- # TODO: Fórmulas de los parámetros -->
-
-<!-- # TODO: Ejemplo con a negativa -->
-
-::: notes
-Pero además, si hablamos de ítems "no cognitivos" (p.ej. personalidad) en los que la persona evaluada puede estar "de acuerdo" o "en desacuerdo", sí podemos tener ítems con probabilidad "decreciente".
-
-La generalización a ítems inversos es directa: La discriminación multidimensional "eleva al cuadrado" los "parámetros de discriminación marginales", luego va a ser siempre positiva.
-El parámetro de posición, por tanto, siempre va a tener el signo inverso al que tenga la intersección.
-Los cosenos tendrán, cada uno, el mismo que el parámetro de discriminación de esa dimensión.
-:::
-
-## Generalización: Ítems inversos
-
-<!-- # TODO: Representación de ítems con a (una o varias) negativa (tabla y gráfico) -->
-
-::: notes
-La representación de estos ítems sería como se muestra en este gráfico.
-<!-- # TODO: Descripción de los ítems representados -->
-:::
-
-## Ítems "inversos" en la literatura
-
-> the negative relation of Item 15 with geometry achievement is particularly puzzling
-
----@mcdonald_basis_2000 [p. 109]
-
-> This result occurs because $a_1$ for the item is negative, and thus as an examinee's $\theta_1$ ability increases, the chance of answering the item correctly actually decreases.
-
----@ackerman_multidimensional_2005-1 [p. 17]
-
-### Generalización: "Monotonía constante"
-
-::: notes
-Esta generalización es bastante evidente, y se puede aplicar sin problema a ítems no cognitivos.
-Sin embargo, hay que tener en cuenta que Reckase explicitaba como condición que los ítems tenían que ser "monotonamente crecientes".
-Pero en la literatura nos encontramos casos como estos, que no lo son, y a los autores no se les ocurre decir que "hay que generalizar esas definiciones".
-Más bien, hacen comentarios como estos, restándole importancia al hecho de que los parámetros de discriminación sean negativos ("particularmente misterioso"), o afirmando lo obvio (el parámetro de discriminación del ítem es negativo para una dimensión, por lo que la probabilidad de respuesta correcta disminuye al incrementarse su habilidad)
-
-En lugar de eso, proponemos que los supuestos de Reckase se "relajen" para considerar ítems con probabilidad "con monotonía constante" (es decir, que no tengan un "pico" u óptimo de rasgo latente para dar una respuesta positiva y luego decrezca), sino que cuanto más extremo sea el nivel de rasgo, más extrema es también la probabilidad (positiva o negativa).
-Este caso general incluye también la probabilidad constante (ya que es monotonía "nula constante", es decir, "el cambio es nulo de manera constante").
-Este tipo de ítems, positivos, negativos, o nulos, se pueden acomodar y modelar fácilmente por un modelo como el logístico de 2 parámetros multidimensional, como hemos visto.
-Ítems que estarían excluidos, por no tener monotonía constante, podrían ser formulaciones como, por ejemplo, "Soy tan extrovertido como la mayoría de mis amigos" (alguien puede decir, "No, soy más introvertido que todos ellos", respuesta negativa con rasgo bajo, o "Qué va, soy el más extrovertido con diferencia", respuesta negativa con rasgo alto).
-:::
-
-## Supuestos: Ortogonalidad
-
-> before performing the differentiation, the constraint that <!-- # TODO: Sum of cosines^2 = 1 --> is added to the expression for the slope \[...\]
-
----@reckase_difficulty_1985 [p. 404]
-
-<!-- # TODO: Ejemplo de rotación oblicua con los ejes resultantes oblicuos -->
-
-<!-- # TODO: Coordenadas rectangulares resultantes de una rotación oblicua -->
-
-::: notes
-Otro de los supuestos que hace Reckase originalmente para calcular los parámetros es este: Los cosenos al cuadrado suman 1.
-Es decir, está aplicando el teorema de Pitágoras, pero ello implica de manera implícita que los ejes del espacio latente son "ortogonales" (forman ángulos de 90 grados).
-
-Esto en TRI multidimensional originalmente tenía sentido, ya que no se tenía muy en cuenta la interpretación de los parámetros de rasgo latente.
-O bien no importaba mucho que los ejes se pudieran interpretar como rasgos latentes, o se interpretaban de manera no necesariamente muy precisa.
-
-No obstante, en 1999 llega McDonald, y dice que la TRI y el AF son fundamentalmente la misma cosa.
-Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas que dan lugar a ejes no ortogonales.
-
-Los pesos factoriales resultantes (o discriminaciones) los podemos representar sin ningún problema en coordenadas rectangulares (las llamadas "coordenadas Cartesianas").
-Pero algo parece fallar cuando generalizamos los parámetros multidimensionales de TRI que hemos mostrado antes a esos espacios de coordenadas no ortogonales.
-Si rotamos los ejes de coordenadas, pero los ítems siguen siendo los mismos, es de esperar que la "longitud" del vector de ese ítem (i.e.
-su discriminación), no cambie (esta propiedad se llama invarianza).
-Veámoslo con un ejemplo.
-:::
-
-## Supuestos: Ortogonalidad
-
-<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
-
-<!-- # TODO: Ejemplo de rotación oblicua (gráfico de ejes y matriz de rotación) -->
-
-::: notes
-Supongamos este ítem, en coordenadas ortogonales.
-El vector muestra sus parámetros de discriminación en las dos dimensiones de los ejes representados.
-Esto sería la "solución sin rotar" que resulta típicamente de ajustar un modelo de TRI multdimensional, o un AFE exploratorio.
-La discriminación multidimensional de este vector es (explicar paso a paso).
-
-Supongamos que aplicamos una rotación oblicua.
-Hemos identificado dos rasgos latentes sustantivos, correlacionados entre sí.
-Los ejes resultantes son los que se muestran en este gráfico.
-Los nuevos parámetros de discriminación del ítem están representadas cada una por la "proyección" de ese vector sobre cada uno de los ejes (paralela al otro, u otros, si hay más de dos).
-:::
-
-## Supuestos: Ortogonalidad
-
-<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
-
-<!-- # TODO: Ejemplo de item proyectado en los ejes rotados (gráfico y vectorial) -->
-
-<!-- # TODO: Cálculo de la discriminación en los ejes rotados -->
-
-::: notes
-Hacer una rotación supone "multiplicar" las coordenadas de este vector por la matriz de rotación que representa ese nuevo sistema de coordenadas (la transformación del ortogonal a este nuevo, oblicuo).
-En este caso, esta es la matriza de rotación.
-Como los vectores se representan en columnas, se "pre-multiplica" por la matriz de rotación.
-Esto es parecido a multiplicar por un vector; es como si cada fila de la matriz fuese un vector, y los resultados se ponen cada uno en una fila.
-<!-- # TODO: Explicar paso a paso el producto matricial -->
-
-Si ahora calculamos la discriminación del ítem en estos ejes rotados, nos encontramos con este resultado.
-Es decir, ¿que el ítem discrimina mejor o peor en función de qué ejes quiera yo considerar?
-Eso no parece tener sentido, y sin embargo hay muchos ejemplos en la literatura que calculan esta discriminación multidimensional con la fórmula de Reckase (con "rasgos correlacionados") sin preocuparse por "los ejes" que representan esos rasgos de manera fidedigna.
-<!-- # TODO: Ejemplos de la literatura -->
-:::
-
-### Generalización: "Coordenadas oblicuas"
-
-::: notes
-Para la representación gráfica esto puede valer, como dice Ackerman aquí.
-Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente.
-:::
-
-## Generalización: Coordenadas oblicuas
-
-> before performing the differentiation, the constraint that <!-- # TODO: Sum of cosines^2 = 1 --> is added to the expression for the slope \[...\]
-
----@reckase_difficulty_1985 [p. 404]
-
-<!-- # TODO: Expresión de las \thetas transformadas -->
-
-<!-- # TODO: Distribución de \thetas en ejes ortogonales -->
-
-::: notes
-Hacer esta generalización no es algo tan directo como ocurría con la probabilidad "monótonamente decreciente".
-Si os acordáis, Reckase aplicaba esta restricción de que la suma de cosenos directores era 1 (que se deriva del teorema de Pitágoras).
-Pero para aplicarla hace falta asumir que esos cosenos directores son todos ortogonales entre sí.
-Si no tenemos ángulos rectos, esa condición no se cumple.
-
-Y lo que nosotros estamos asumiendo que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones ($R$, la que sea).
-Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a coordenadas oblicuas tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el o y el 1 pueda ser distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
-
-¿Cómo podríamos solucionar esto?
-Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de este espacio oblicuo (el que sea, aún no sabemos cuál es) a un espacio ortogonal.
-:::
-
-## Generalización: Coordenadas oblicuas
-
-<!-- # TODO: Expresión de las \thetas transformadas -->
-
-<!-- # TODO: Expresión de la R transformada -->
-
-<!-- # TODO: Distribución (esférica) de las \thetas transformadas -->
-
-::: notes
-Esto es una transformación de las coordenadas, como la que aplicamos antes para calcular las coordenadas del ítem tras aplicar la rotación oblicua, pero aplicada en este caso al vector de rasgo latente.
-
-Estas nuevas coordenadas (transformadas), que ahora están en un espacio ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^B$) sea una matriz identidad (unos en la diagonal, ceros en el resto).
-Pero hay que tener en cuenta que esto es un "artefacto matemático".
-Estamos haciendo esto por conveniencia, pero las coordenadas que realmente nos interesan como investigadores, porque son las que verdaderamente representan algo que nos interesa investigar, son esas coordenadas originales, que pueden representar cualquier fenómeno que sea de nuestro interés (rasgos de personalidad, síntomas, actitudes, inteligencia, conocimientos...
-lo que estemos investigando).
-:::
-
-## Generalización: Coordenadas oblicuas
-
-<!-- # TODO: Ejemplo de \theta (expresión) y coordenadas en espacio oblicuo -->
-
-::: notes
-Mientras tanto, las coordenadas de los "rasgos latentes" originales se van a representar en otros ejes, en general distintos de los "ejes ortogonales" transformados.
-Esto lo que significa es que las puntuaciones de rasgo latente originales van a representar la posición de un participante respecto de esos ejes originales.
-Por ejemplo, en este caso, si el participante tiene un "vector de puntuaciones" \[1, 2\], para representar sus coordenadas tenemos que desplazarnos, primero una unidad en el primer eje, y luego dos unidades en el segundo eje.
-:::
-
-## Generalización: Coordenadas oblicuas
-
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
-
-::: notes
-En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
-Supongamos una distribución bivariada de rasgos latentes, con una correlación de .8, bastante alta.
-¿Cuál de los siguientes sistemas de coordenadas consideráis que representaría adecuadamente los ejes de una distribución bivariada, de forma que al transformarlos a coordenadas rectangulares obtuviésemos una distribución con correlación de 0?
-:::
-
-## Generalización: Coordenadas oblicuas
-
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Distribución en ejes rectangulares -->
-
-::: notes
-El primero ya sabemos que no, de acuerdo, puesto que es la representación habitual, y sabemos que la nube que forma está "achatada".
-No se aplica ninguna transformación (la transformación es una matriz identidad), ya que los que serían los ejes originales y los transformados son los mismos.
-:::
-
-## Generalización: Coordenadas oblicuas
-
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Animación de distribución en ejes rectangulares a ejes alineados -->
-
-::: notes
-Si pensáis igual que yo, habréis elegido el segundo, porque el coseno del ángulo entre los ejes representa la correlación entre las dimensiones.
-Por lo tanto, ejes muy alineados representan una correlación muy alta.
-
-Sin embargo, fijáos lo que ocurre cuando "transformamos" las coordenadas de los puntos de nuestra distribución a ese sistema de coordenadas.
-Como podéis ver, la distribución tiende a "achatarse" aún más, en lugar de hacerse más "esférica".
-:::
-
-## Generalización: Coordenadas oblicuas
-
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Animación de distribución en ejes rectangulares a ejes correctos -->
-
-<!-- # TODO: Matriz de correlaciones transformada "identidad" -->
-
-::: notes
-El sistema de coordenadas correcto es el tercero, que "compensa" de alguna manera el achatamiento de la nube de puntos, de forma que las coordenadas transformadas que nos quedan (en los ejes ortogonales), los "normales" del gráfico) tienen una distribución "esférica" (con matriz de correlaciones "identidad").
-
-Esto, que visto así parece ya evidente, es algo que al menos a mí me ha llevado varias decenas (puede que "cientos") de horas entender, durante al menos dos años trabajando en este problema.
-:::
-
-## Espacio latente vs. espacio test
-
-> It has been suggested by some researchers that the angle between the axes represents the degree of correlation.
-> However, for sake of clarity, an orthogonal axes system is used in which distance measures and vectors can be easily calculated, understood, and interpreted.
-
----@ackerman_multidimensional_2005-1 [p. 16]
-
-<!-- # TODO: Fórmula del modelo en coordenadas "transformadas" y equivalencia -->
-
-<!-- # TODO: Expresión de la transformación de la discriminación a coordenadas rectangulares -->
-
-::: notes
-Viendo que en la literatura hay afirmaciones como esta, no me resulta totalmente descorazonador.
-Parece que no soy el único al que le ha supuesto un esfuerzo entenderlo (resultados de la "encuesta"?)
-
-Lo que ocurre en realidad, y esto es en realidad también un paralelismo con lo que ocurre en análisis factorial, es que para que el modelo sea invariante, los ítems también tienen que transformarse a un sistema de coordenadas ortogonal.
-Pero podemos ver que su matriz de transformación no es la misma que la de las puntuaciones de rasgo latente.
-Es su inversa (inversa traspuesta, más concretamente).
-
-Y esto nos ha permitido "descubrir" (o darnos cuenta, más bien) de que hay dos "espacios vectoriales" distintos, uno en el que se encuentran las coordenadas de las puntuaciones factoriales, y otro distinto, que es el espacio en el que se encuentran las coordenadas de los ítems, y que es donde en realidad hay que representarlos.
-:::
-
-## Espacio latente vs. espacio test
-
-> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by <!-- # TODO: Inner product -->.
-
----@zhang_theoretical_1999 [p. 221]
-
-<!-- #TODO: Ref. to Mahalanobis -->
-
-> \[...\] the *inner product* in \[the test space\] is consistently defined by <!-- # TODO: Inner product -->.
-
----@zhang_theoretical_1999 [p. 221]
-
-::: notes
-Esto, curiosamente, aparece en la literatura como podéis ver; en el año 1999, en un artículo que no tiene nada que ver con esto.
-Lo menciona de soslayo y no profundiza en la deducción de los parámetros, como hemos hecho aquí.S olamente lo utiliza para resolver un problema "parcial" que se encuentra en la deducción de un índice de dimensionalidad, distinto de los parámetros de los que estamos hablando aquí.
-Pero podía haber obviado este problema y haber utilizado coordenadas ortogonales, sin importarle si había o no correlaciones entre los rasgos latentes, como ha hecho todo el mundo.
-
-Así que, todos mis respetos por estos autores.
-Sin embargo claro, ellos estaban hablando de otra cosa, y sólo tocaban este tema de paso.
-Así que al final eso se quedó ahí, y nadie ha hecho caso en 25 años a lo que dijeron Zhang y Stout.
-La gente ha usado los modelos, hecho simulaciones, etc., usando distribuciones correlacionadas, y ha usado las fórmulas de Reckase para transformar entre parámetros multidimensionales y los parámetros "lineales" (podríamos llamarlos, los de la formulación original) del modelo.
-:::
-
-## Parámetros multidimensionales generalizados
-
-<!-- # TODO: Fórmulas de los parámetros multidimensionales -->
-
-::: notes
-Si tenemos en cuenta las correlaciones (más generalmente, covarianzas) de la distribución de puntuaciones factoriales, lo que obtenemos son estas expresiones para los parámetros multidimensionales.
-
-Muy importante, la distancia con signo y la discriminación multidimensional son iguales, no importa en qué espacio queramos obtenerlas y representarlas.
-Pero los cosenos directores es necesario calcularlos en el espacio test, no en el espacio de puntuaciones latentes, como decía Reckase, porque este es el espacio que va a representar en realidad la relación entre las variables latentes.
-:::
-
-## Conclusiones
-
-Parámetros multidimensionales generalizados
-
--   A ítems con monotonía constante
-
--   A coordenadas oblicuas
-
-Distinción espacio latente - espacio test
-
-::: notes
-En conclusión, lo que hemos propuesto en este artículo consiste en una generalización:
-
--   A ítems con monotonía constante, creciente, decreciente o nula.
-    Esto se venía haciendo ya como hemos visto, pero creemos que hacía falta formalizarlo.
-    No basta con hacerlo y punto, si Reckase dijo que las discriminaciones tiene que ser positivas, tenemos que demostrar que también vale para discriminaciones negativas o nulas (monotonía decreciente, o probabilidad constante).
-
--   A coordenadas oblicuas, las cuales consideramos que son necesarias para representar distribuciones de puntuaciones de rasgo correlacionadas, en general.
-    Hemos visto que esto ya lo mencionaron Zhang y Stout, hace 25 años, pero no proporcionaron una "deducción" de estos parámetros multidimensionales, cosa que nosotros hacemos.
-
-Por último, es imortante recalcar este "descubrimiento" (o "redescubrimiento", más bien) del espacio test, ya que tenemos que ser muy conscientes de que los ítems y las personas están en espacios vectoriales distintos.
-Aunque esto parezca solamente un "artefacto matemático" (y en el fondo lo es), hemos visto que es verdaderamente crucial tenerlo en cuenta para entender bien los modelos que estamos utilizando.
-Además, esta "distinción" se halla también en los modelos factoriales: Existe lo que se denominan "ejes primarios" y "ejes de referencia", propuestos por Thurstone hace años (pero que prácticamente no se utilizan ni tienen en consideración).
-<!-- # TODO: Hace cuántos años?
-(Thurstone) --> Sin embargo, cuidado con esto, porque no está claro que esos espacios en TRI multidimensional se puedan asimilar directamente a los ejes en un modelo de AF.
-:::
-
-## To be continued...
-
--   Paralelismo TRI multidimensional - AF
-
--   Extensiones del modelo M2PL: Escala graduada, respuesta nominal, 3PL...
-
--   Investigación en otros modelos
-
-::: notes
-Esta de hecho puede ser una futura línea de investigación.
-Puede parecer evidente el paralelismo (o al menos a mí me lo parece), pero si una cosa me ha enseñado el estar trabajando en este problema de los parámetros (y toparme con el problema de los dos espacios), es que no hay que fiarse de las apariencias.
-Por eso aplicamos el método científico (el método puramente deductivo, en este caso)
-
-El modelo multidimensional logístico de 2 parámetros se puede considerar como "anidado", o un caso más simple de varios modelos.
-Así que partiendo de este como base, lo interesante es aplicarlo a otros más complejos, pero más útiles en la práctica.
-Por cierto, que esto también lo hace mucha gente, aplica las fórmulas de Reckase a otros modelos y a correr, se quedan "tan anchos".
-
-El modelo de escala graduada, por ejemplo, sería el equivalente al modelo de AF para respuestas ordinales, con varios umbrales o "posiciones" en TRI.
-
-El de crédito parcial, por ejemplo, puede servir también para formatos con varias opciones de respuesta que miden o pesan en factores diferentes cada una de ella.
-
-El logístico de 3 parámetros es como el de 2, pero añadiendo un parámetro de pseudoazar, para modelar el que un participante acierte un ítem por pura suerte, respondiendo "a ciegas" (esto sólo valdría para ítems cognitivos).
-
-Estos modelos, que son "extensiones" de este más básico, deberían partir de estos parámetros como fundamento.
-Pero un investigador no debería dar los parámetros sin más de ese modelo, sin proporcionar una prueba formal de que se aplica esa fórmula y se cumplen los supuestos.
-:::
-
-## Últimas palabras
-
-<!-- # TODO: Imagen ilustrativa -->
-
-::: notes
-A este respecto, me gustaría terminar recalcando esto último que he mencionado, aplicado no sólo a cuestiones psicométricas, o de TRI.
-Cuando estaba empezando a preparar este seminario, una de mis inquietudes era la complejidad, tanto conceptual como formal de esta investigación.
-Compartiendo estas inquietudes con algunos compañeros, me aconsejaron utilizar digamos este discurso como excusa para lanzar este "mensaje" en general: La idea de que tenemos que ser muy precavidos a la hora de aplicar estos métodos, sean cuales sean (cuantitativos, cualitativos, o del tipo que sean) sin reflexionar sobre si son los más adecuados o no.
-Y bueno, a este respecto comentar que, no sólo yo, sino el departamento de Metodología al completo, hacemos una tipo de investigación que precisamente se encarga de eso, de intentar entender estos métodos, saber si son los más adecuados, y saber cuándo, cómo, y sobre todo, por qué, aplicarlos en la investigación aplicada (aplicada para nosotros, al menos).
-
-Así que bueno, ya lo sabéis, estamos disponibles en el departamento para colaborar en esas investigaciones, si lo consideráis oportuno, para aplicar los métodos más apropiados a los problemas de investigación que tengáis.
-En mi caso en concreto, no sólo puedo proporcionar cierto "expertise" psicométrico.
-También me dedico mucho a cuestiones de ciencia abierta y reproduciblidad, así que si queréis darle a vuestra investigación un giro en ese sentido, para hacerla más sostenible y alineada con los objetivos de los organismos financiadores (y si no queréis, deberíais ir queriendo), podéis contactar conmigo, y podemos plantear líneas de colaboración.
-Eso sí, me gustaría pediros, si necesitáis colaboración, no sólo metodológica, sino de cualquier tipo, por favor no nos llaméis cuando tengáis un lío de datos, y no sepáis bien cómo resolver la papeleta (no digo que sea vuestro caso, pero me he encontrado con muchos casos, que son verdaderas historias de terror).
-Contad por favor con nosotros desde el momento de escribir vuestro proyecto, para solicitar la financiación, porque es en ese punto en el que se pueden prevér problemas y anticiparse a ellos.
-Porque si una investigación no está bien diseñada desde el principio, teniendo en cuenta estos aspectos metodológicos, de reproducibilidad, etc., si tenemos mala suerte pueden surgir problemas que se puedan resolver sólo con muchísimo trabajo, o incluso que ya nunca se puedan resolver, y tengamos que tirar a la basura todo ese esfuerzo de investigación.
-
-Eso es todo, muchas gracias.
-Y espero vuestras preguntas y comentarios.
-:::
-
-## References {.smaller}

### 2024-11-08 output/seminar-slideshow.qmd
Version sandbox document to work on single slides
Intermediate markdown in the main slideshow is not necessary while using this sandbox slidehow for development

@@
     incremental:             false
-    keep-md:                 true
+    keep-md:                 false
     link-external-newwindow: true

### 2024-11-15 output/seminar-slideshow.qmd
Improves explanation of scalar product throughout

@@
 El producto de los pesos por las puntuaciones factoriales se puede expresar en "forma vectorial", de esta manera.
-La "T" del vector de pesos factoriales significa que está "traspuesto", es decir, que se pone horizontalmente (esto es una convención en álgebra matricial).
-El producto del vector de pesos por el de puntuaciones factoriales es lo que se llama un "producto vectorial", que consiste simplemente en multiplicar dos a dos los términos de cada vector y sumar todos.
+Este término significa que el vector de puntuaciones factoriales está pre-multiplicado por el vector de pesos factoriales.
+Esta operación es lo que se llama un "producto escalar", que consiste simplemente en multiplicar dos a dos los términos de cada vector y sumar todos.
 :::
@@
 
-Su respuesta se calcularía así: (hacer los cálculos).
+\[NEXT\]
+
+Su respuesta se calcularía así: Primero tenemos el parámetro de intersección que vale 1.
+La "T" del vector de pesos factoriales significa que está "traspuesto", es decir, que se pone horizontalmente (esto es una convención en álgebra matricial).
+El producto del vector de pesos por el de puntuaciones factoriales es (...).
+Y por último sumamos la puntuación de error de ese participante en la variable.
 Luego esta persona (el particpante 1) tiene la respuesta o valor observado 0.5 en esta variable observada.
+
 Lo mismo podríamos hacer con el resto de variables y de participantes, pero obviamente nosotros lo que tenemos son estas respuestas observadas, y lo que hacemos con el AF es estimar todos estos parámetros del modelo, en lugar de como lo hemos hecho ahora.
@@
 Esto se puede extender, de manera bastante directa, asumiendo que la respuesta se modela como una "regresión logística multivariada".
-Aquí aparece otra vez la discriminación pero ahora en forma vectorial, que se multiplica por el vector de rasgo latente de la persona, y un parámetro de intersección, como antes (pero no puede haber un parámetro de dificultad simplemente, o uno por cada dimensión, porque serían indistinguibles entre sí).
+Aquí aparece otra vez la discriminación pero ahora en forma vectorial, que se pre-multiplica (recordad, producto escalar) por el vector de rasgo latente de la persona, y un parámetro de intersección, como antes (pero no puede haber un parámetro de dificultad simplemente, o uno por cada dimensión, porque serían indistinguibles entre sí).
 Esta es una de las extensiones multidimensionales (más antiguas, y probablemente más obvias) del modelo logístico de 2 parámetros: El modelo logístico de 2 parámetros multidimensional compensatorio.

### 2024-11-15 output/seminar-slideshow.qmd
Improve explanation of multidimensional difficulty

@@
 
+::: fragment
 $$
@@
 $$
+:::
 
@@
 ::: notes
-Por ese motivo propone un parámetro de "dificultad" (lo llama, para nosotros sería de posición) que sea de algún modo equivalente a la dificultad en el modelo unidimensional.
-Ese punto es donde la curva (superficie) de probabilidad de respuesta tiene una inflexión (a lo largo de toda ella, el punto más cercano al origen, donde la recta que lo une al origen es perpendicular a la recta de inflexión de la curva).
-
-Este parámetro se puede entender como una "distancia **con signo**" desde el origen hasta el punto de posición del ítem.
+Lo que hace Reckase entonces es intentar definir un parámetro de "dificultad" (lo llama, para nosotros sería de posición), que sea de algún modo equivalente a la dificultad en el modelo unidimensional, es decir, el lugar donde el ítem es lo más discriminativo posible.
+Donde la curva de probabilidad de respuesta pasa de ser cada vez más abrupta a cada vez más plana.
+
+\[NEXT\]
+
+Y llega a esta definición, que consiste simplemente en el parámetro de intersección (con signo negativo), dividido entre la raíz del producto escalar (vuelve a aparecer el producto escalar), en este caso, del vector de parámetros de discriminación consigo mismo.
+Toda esta expresión se denomina "norma" del vector.
+
+\[NEXT\]
+
+Este parámetro, de "discriminación" o "dificultad", se puede interpretar de igual manera a la dificultad unidimensional.
 :::
@@
 ::: notes
-Para representar ese punto no basta con la "distancia (con signo)"; hace falta saber en qué dirección mide ese ítem también.
+Geométricamente consiste en una "distancia **con signo**" (puede ser hacia cualquiera de ambos sentidos) desde el origen hasta ese punto donde está la posición del ítem, sobre la recta de inflexión (donde la probabilidad es igual a 0.5).
+
+\[NEXT\]
+
+A lo largo de toda esa recta podría ser cualquier punto, pero en realidad es el punto más cercano al origen, donde la recta que lo une al origen es perpendicular a esa misma recta de inflexión).
+Para ello claro, no basta con la "distancia (con signo)"; hace falta saber en qué dirección mide ese ítem también.
 Eso viene dado por los llamados "cosenos directores".
-
-Por lo tanto, la "dificultad (posición) multidimensional" viene dada por dos elementos, un número real que representa cuánto se desplaza la recta de inflexión del ítem respecto del origen y en qué sentido, y un vector de cosenos que determina la dirección en la que se desplaza.
+Fijáos que vuelve a aparecer la norma de este vector en el denominador.
+
+\[NEXT\]
+
+Por lo tanto, la "dificultad (posición) multidimensional" está compuesta por dos elementos, un número real que representa cuánto se desplaza la recta de inflexión del ítem respecto del origen y en qué sentido, y un vector de cosenos que determina la dirección en la que se desplaza.
 Así se puede interpretar un ítem, o varios, de manera relativamente sencilla: Todos los participantes cuyas coordenadas estén en la "recta de inflexión" (la recta que pasa por la posición del ítem, perpendicular a su dirección) tendrán una probabilidad de dar una respuesta positiva y negativa de .5 (dificultad media; igual probabilidad de acertar o fallar).

### 2024-11-15 output/seminar-slideshow.qmd
Add discrimination slide with Pythagoras theorem

@@
 ::: column
-```{r mdisc-item}
+```{r low-disc}
 #| fig-width:  4.7
@@
 
-## Parámetros multidimensionales
-
+## Discriminación multidimensional
+
+::::::: columns
+::: column
+```{r mdisc-item}
+#| fig-width:  8
+#| fig-height: 8
+#| cache: false
+
+item_vector <- mirt_item_coords |> mutate(
+  across(ends_with("1"), ~ . - origin_1),
+  across(ends_with("2"), ~ . - origin_2)
+)
+
+item_spikes <- item_vector             |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+item_vector |>
+  ggplot(
+    aes(
+      0, 0,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = 0.3, y = 0.27,
+    label = "*MDISC~i~*",
+    size  = 9,
+    angle = mirt_item_params |> pull(deg_1),
+    label.colour  = NA,
+    fill = NA
+  ) +
+  annotate(
+    "richtext",
+    x = c(0.5, -0.13), y = c(-0.12, 0.25),
+    label         = c("*a*~1*i*~", "*a*~2*i*~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+
+::::: column
+::: fragment
 @reckase_discriminating_1991:
 
-<!-- # TODO: Fórmula de MDISC -->
-
-<!-- # TODO: Gráfico del cálculo de MDISC como teorema de Pitágoras -->
-
-<!-- # TODO: Fórmula de dificultad con MDISC -->
-
-::: notes
-@reckase_discriminating_1991 hacen la deducción de la discriminación multidimensional, y resulta dar lugar a esta fórmula.
-Como veis es simplemente la "longitud" del vector de parámetros de discriminación, resultante simplemente de aplicar el teorema de Pitágoras.
-El vector es un triángulo rectángulo, $a_1$ sería un cateto, el otro cateto sería igual a $a_2$, y la hipotenusa es la longitud (las dos discriminaciones al cuadrado, sumadas, y luego la raíz cuadrada de eso).
-Esto por cierto también vale para cualquier número de dimensiones.
-Aquí estamos representando sólo 2 porque es más fácil, pero podemos calcular la suma de cuadrados estos parámetros de discriminación (sea cual sea su número), y calcular su raíz cuadrada, y eso nos daría igualmente la discriminación multidimensional (la longitud de ese vector de $a$ del ítem).
-
-Esto es en realidad el producto escalar (o producto punto) de un vector consigo mismo, que consiste en multiplicar una a una las componentes y sumarlos todos.
+<br>
+
+$$
+  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
+$$
+:::
+
+::: fragment
+$$
+  MID_i = \begin{Bmatrix}
+    \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}} \\
+    \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
+  \end{Bmatrix}
+$$
+:::
+:::::
+:::::::
+
+::: notes
+@reckase_discriminating_1991 hacen la deducción de la discriminación multidimensional, y el resultado es simplemente la "longitud" del vector de parámetros de discriminación.
+El cálculo puede hacerse simplemente aplicando el teorema de Pitágoras.
+El vector es un triángulo rectángulo, $a_1$ sería un cateto, el otro cateto (SEÑALAR "SPIKE") sería igual a $a_2$, y la hipotenusa es la longitud (las dos discriminaciones al cuadrado, sumadas, y luego la raíz cuadrada de eso).
+
+En forma matricial, esto no es más que el producto escalar del vector consigo mismo, es decir, que consiste en multiplicar una a una las componentes y sumarlos todos.
 En forma matricial se representaría como el vector pre-multiplicado por sí mismo.
-<!-- # TODO: Explicar la multiplicación vectorial (y poner ejemplo?) -->
-
-Esta expresión de la discriminación multidimensional se puede usar para expresar de manera más sencilla la dificultad.
+Por ejemplo, aquí $a_1$ es 1, al cuadrado 1, más 0.5 ($a_2$) al cuadrado que es 0,25, luego la raíz cuadrada de 1,25 (sale 1,12 aproximadamente).
+Esta fórmula por cierto vale para cualquier número de dimensiones.
+Aquí estamos representando sólo 2 porque es más fácil, pero podemos calcular el producto escalar de este vector de parámetros de discriminación consigo mismo (sea cual sea su dimensión), y calcular su raíz cuadrada, y eso nos daría igualmente la discriminación multidimensional (la longitud de ese vector de $\mathbf{a}$ del ítem).
+
+Si recordáis además, esta raíz del producto escalar del ítem consigo mismo aparecía en el denominador de la dificultad multidimensional (en ambos componentes).
+Por lo tanto, esta expresión de la discriminación multidimensional se puede usar para expresar de manera más sencilla la dificultad.
+:::
+
 :::

### 2024-11-15 output/sandbox_slideshow.qmd
Avoid references from appearing in previous slide
Adding the "reference" slide prevents that added references appear at the end of the slide being edited
lighten up sandbox slideshow

@@
 title:     |
-  Parámetros multidimensionales en Teoría de Respuesta al Ítem:
-  ¡Algo estamos haciendo mal!
-author:    |
-  <table>
-    <tr>
-      <td>Daniel Morillo, Ph.D.</td>
-    </tr>
-    <tr>
-      <td>
-        <center>
-          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/DaniMori/)&emsp;
-          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0003-3021-3878)
-        </center>
-      </td>
-    </tr>
-  </table>
-institute: |
-  <table>
-    <tr>
-      <td>
-        <center>
-          [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="120"}](https://www.uned.es/universidad/facultades/psicologia.html)
-        </center>
-      </td>
-    </tr>
-  </table>
+  Sandbox slide
 bibliography:  ../www/Multidimensional-parameters-MCLM.bib
@@
 ---
-
-# Introducción
 
@@
 ```
+
+## References

### 2024-11-15 output/seminar-slideshow.qmd
Add slide with complete formulation of parameters
Make minor tweaks to discrimination final slide

@@
       &= \Phi_L \left(1 -0.5 -0.5 \right) = \Phi_L \left(0 \right) \\
-      &= \frac{1}{1 + e^{-0}} = \frac{1}{1 + 1} = \frac{1}{2} = 0.5
+      &= \frac{1}{1 + e^{-0}} = \frac{1}{1 + 1} = \frac{1}{2} \\
+      &= 0.5
   \end{align}
@@
 #| fig-height: 8
-#| cache: false
 
@@
     "richtext",
-    x = 0.3, y = 0.27,
+    x = 0.32, y = 0.25,
     label = "*MDISC~i~*",
@@
 @reckase_discriminating_1991 hacen la deducción de la discriminación multidimensional, y el resultado es simplemente la "longitud" del vector de parámetros de discriminación.
+Esto sería, en cierto modo, equivalente a la "representación factorial" del ítem; aquí la punta del vector está donde estarían las "bolitas" que usábamos antes para representar los pesos factoriales en el espacio de factores comunes, excepto que, para un mismo ítem, estaría escalado por la varianza única del ítem y el factor aquel de 1.7 que salía en la fórmula, pero por lo demás es una representación equivalente a la del modelo factorial.
+
 El cálculo puede hacerse simplemente aplicando el teorema de Pitágoras.
 El vector es un triángulo rectángulo, $a_1$ sería un cateto, el otro cateto (SEÑALAR "SPIKE") sería igual a $a_2$, y la hipotenusa es la longitud (las dos discriminaciones al cuadrado, sumadas, y luego la raíz cuadrada de eso).
-
-En forma matricial, esto no es más que el producto escalar del vector consigo mismo, es decir, que consiste en multiplicar una a una las componentes y sumarlos todos.
-En forma matricial se representaría como el vector pre-multiplicado por sí mismo.
 Por ejemplo, aquí $a_1$ es 1, al cuadrado 1, más 0.5 ($a_2$) al cuadrado que es 0,25, luego la raíz cuadrada de 1,25 (sale 1,12 aproximadamente).
-Esta fórmula por cierto vale para cualquier número de dimensiones.
+
+\[NEXT\]
+
+En forma matricial, esto no es más que el producto escalar del vector consigo mismo, es decir, aparece otra vez la norma del vector, nuestra vieja amiga.
+Y esta fórmula por cierto vale para cualquier número de dimensiones.
 Aquí estamos representando sólo 2 porque es más fácil, pero podemos calcular el producto escalar de este vector de parámetros de discriminación consigo mismo (sea cual sea su dimensión), y calcular su raíz cuadrada, y eso nos daría igualmente la discriminación multidimensional (la longitud de ese vector de $\mathbf{a}$ del ítem).
 
-Si recordáis además, esta raíz del producto escalar del ítem consigo mismo aparecía en el denominador de la dificultad multidimensional (en ambos componentes).
-Por lo tanto, esta expresión de la discriminación multidimensional se puede usar para expresar de manera más sencilla la dificultad.
-:::
-
+Si recordáis además, esta norma aparecía en el denominador de la dificultad multidimensional (en ambos componentes).
+:::
+
+## Parámetros multidimensionales
+
+<br>
+
+$$
+  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
+$$
+
+<br>
+
+$$
+  MID_i = \begin{Bmatrix}
+    \frac{-d_i}{MDISC_i} \\
+    \frac{\mathbf{a}_i}{MDISC_i}
+  \end{Bmatrix}
+$$
+
+::: notes
+Luego esta expresión de la discriminación multidimensional se puede usar para expresar de manera más sencilla la dificultad.
+Y así, esta sería la expresión completa de los parámetros multidimensionales de un ítem según el modelo de TRI logístico de 2 parámetros.
 :::

### 2024-11-15 output/seminar-slideshow.qmd
Add labels to all contour plots in difficulty plot

@@
 
-contour_plot +
+mirt_item_probs |>
+  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_contour2(
+    color = F1_COLOR,
+    linewidth = LINE_WIDTH,
+    skip = 0,
+    breaks = 1:9 / 10,
+    label.placer = label_placer_fraction(frac = .90, rot_adjuster = 0)
+  ) +
   geom_segment(
@@
     fill = NA
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks       = element_blank(),
+    axis.line        = element_blank(),
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
   )

### 2024-11-22 output/sandbox_slideshow.qmd
Drop unused config options from front matter

@@
 ---
-title:     |
-  Sandbox slide
 bibliography:  ../www/Multidimensional-parameters-MCLM.bib
 csl:           ../www/apa-old-doi-prefix.csl
-date:          "2024-10-21"
 date-meta:     "2024-11-27"
-date-format:   long
 editor:
@@
 knitr:
-  # opts_knit: # Does not work, see line 88
-  #   root_dir:  here::here()
   opts_chunk: 
@@
     fig-width:               6
-    incremental:             false
     keep-md:                 true
@@
     theme:                   ../www/slideshow-assets/extra-styles.scss
-    transition:              none
-    view-distance:           3
-    template-partials:
-      - ../www/slideshow-assets/title-slide.html
 editor_options: 
@@
 
+```{r pre}
+PROB_AXIS_2D <- seq(-3, 3, by = .1)
+F1_COLOR <- unname(PALETTE_BLUE["normal"])
+
+logit <- function(x) 1 / (1 + exp(-x))
+irf   <- function(a, b, theta) logit(a * (theta - b))
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    0.5,  -0.5
+)
+
+mirt_item_probs <- expand_grid(
+  trait_2 = PROB_AXIS_2D,
+  trait_1 = PROB_AXIS_2D
+) |>
+  mutate(
+    mirt_item |>
+      select(-item) |>
+      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
+      bind_cols()
+  ) |>
+    rename(prob = `...1`)
+
+```
+
 ## References

### 2024-11-22 output/seminar-slideshow.qmd
Correct reference to signed distance in figures

--- 2024-11-15 output/seminar-slideshow.qmd
@@
     canonical: true
+execute: 
+  cache:       true
 lang:          es
@@
     results:   asis
-    cache:     true
     autodep:   true
@@
 $$
-  MID = \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
+  D_i = \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
 $$
@@
 
-mirt_item_spikes <- mirt_item_coords |>
+d_i_spike <- mirt_item_coords |>
   select(item, trait_1 = origin_1, trait_2 = origin_2) |>
@@
 
-contour_plot +
+mirt_item_probs |>
+  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_contour2(
+    color = F1_COLOR,
+    linewidth = LINE_WIDTH,
+    skip = 0,
+    breaks = 1:9 / 10,
+    label.placer = label_placer_fraction(frac = .90, rot_adjuster = 0)
+  ) +
   geom_segment(
-    data    = mirt_item_spikes,
+    data    = d_i_spike,
     mapping = aes(
@@
     geom = "richtext",
-    x = 0.16, y = 0.15,
-    label = "*MID~i~*",
-    size = 6,
+    x = 0.15, y = 0.15,
+    label = "*D~i~*",
+    size = 9,
     angle = mirt_item_params |> pull(deg_1),
@@
 
-::::: {.columns .fragment}
-::: column
-<br> <br> <br>
-
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-:::
-
-::: column
-```{r low-disc}
+::::: columns
+::: {.column .fragment}
+```{r low-disc-contour}
 #| fig-width:  4.7
 #| fig-height: 4.7
-#| cache: false
-
-item_vector <- mirt_item_coords |> mutate(
-  across(ends_with("1"), ~ . - origin_1),
-  across(ends_with("2"), ~ . - origin_2)
-)
-
-item_spikes <- item_vector             |>
-  slice(1 |> rep(2))                   |>
-  add_column(dim = 'F' |> paste0(1:2)) |>
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     0.5,  0.25,  -0.125
+)
+
+mirt_item_probs <- expand_grid(
+  trait_2 = PROB_AXIS_2D,
+  trait_1 = PROB_AXIS_2D
+) |>
   mutate(
-    origin_1 = if_else(dim == 'F1', end_1, origin_1),
-    origin_2 = if_else(dim == 'F2', end_2, origin_2),
-  )
-
-item_vector |>
-  ggplot(
-    aes(
-      0, 0,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
+    mirt_item |>
+      select(-item) |>
+      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
+      bind_cols()
+  ) |>
+    rename(prob = `...1`)
+
+mirt_item_probs |>
+  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
   geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
   geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data = item_spikes,
-    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+  geom_contour2(
+    color = F1_COLOR,
     linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color     = PALETTE_BLUE["light"]
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
+    breaks = 1:9 / 10,
+    skip = 0,
+    label.placer = label_placer_fraction(frac = .80, rot_adjuster = 0)
   ) +
   scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
+    breaks       = -1:1,
     minor_breaks = NULL,
+    limits       = c(-1.5, 1.5),
+    expand       = expansion(),
     name         = NULL,
@@
   scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
+    breaks       = -1:1,
     minor_breaks = NULL,
+    limits       = c(-1.5, 1.5),
+    expand       = expansion(),
     name         = NULL,
@@
   ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    "richtext",
-    x = 0.3, y = 0.27,
-    label = "*MDISC~i~*",
-    size  = 5.5,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  annotate(
-    "richtext",
-    x = c(0.5, -0.13), y = c(-0.12, 0.25),
-    label         = c("*a*~1*i*~", "*a*~2*i*~"),
-    size          = 4.5,
-    label.colour  = NA,
-    label.padding = unit(0, "lines"),
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
+  coord_fixed() +
   theme(
+    axis.ticks       = element_blank(),
     axis.line        = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
+  )
+```
+:::
+
+::: {.column .fragment}
+```{r high-disc-contour}
+#| fig-width:  4.7
+#| fig-height: 4.7
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1.5,  0.75,  -0.375
+)
+
+mirt_item_probs <- expand_grid(
+  trait_2 = PROB_AXIS_2D,
+  trait_1 = PROB_AXIS_2D
+) |>
+  mutate(
+    mirt_item |>
+      select(-item) |>
+      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
+      bind_cols()
+  ) |>
+    rename(prob = `...1`)
+
+mirt_item_probs |>
+  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_contour2(
+    color = F1_COLOR,
+    linewidth = LINE_WIDTH,
+    breaks = 1:9 / 10,
+    # skip = 0,
+    label.placer = label_placer_fraction(frac = .90, rot_adjuster = 0)
+  ) +
+  scale_x_continuous(
+    breaks       = -1:1,
+    minor_breaks = NULL,
+    limits       = c(-1.5, 1.5),
+    expand       = expansion(),
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    breaks       = -1:1,
+    minor_breaks = NULL,
+    limits       = c(-1.5, 1.5),
+    expand       = expansion(),
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  coord_fixed() +
+  theme(
     axis.ticks       = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
+    axis.line        = element_blank(),
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
   )
@@
 Esta característica viene representada por el parámetro de "discriminación", que es una medida de la "calidad general" de un ítem.
-Los ítems "buenos" serán muy discriminativos, lo que indica que distinguen bien entre la gente que está en la región de alta o baja probabilidad de respuesta, mientras que los "malos" tendrán una superficie de respuesta más "plana".
+Los ítems "malos", o poco discriminantivos, tendrán una superficie de respuesta bastante "plana", con una transición suave entre la región de baja y la de alta probabilidad de respuesta, mientras que los muy discriminativos distinguirán bien entre la gente que está en la región de alta o baja probabilidad de respuesta, con zona de transición más estrecha, más abrupta.
 :::

### 2024-11-22 output/seminar-slideshow.qmd
Add M2PL model graphical representation slide

@@
 
+::: r-center
+```{r m2pl-representation}
+#| fig-align:  center
+#| fig-height: 7.2
+#| fig-width:  7.2
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data    = d_i_spike,
+    mapping = aes(
+      trait_1_start, trait_2_start,
+      xend = trait_1_end,
+      yend = trait_2_end
+    ),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = PALETTE_BLUE["light"]
+  ) +
+  geom_point(
+    data = mirt_item_coords,
+    mapping = aes(origin_1, origin_2),
+    inherit.aes = FALSE,
+    size = I(4)
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    geom = "richtext",
+    x = 0.16, y = 0.20,
+    label = "*D~i~*",
+    size = 9,
+    angle = mirt_item_params |> pull(deg_1),
+    label.colour  = NA,
+    fill = NA
+  ) +
+  annotate(
+    "richtext",
+    x = mirt_item_coords |> pull(origin_1) + 0.40,
+    y = mirt_item_coords |> pull(origin_2) + 0.32,
+    label = "*MDISC~i~*",
+    size  = 9,
+    angle = mirt_item_params |> pull(deg_1),
+    label.colour  = NA,
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+
 ::: notes

### 2024-11-29 output/seminar-slideshow.qmd
Add second "graphical representation" slide
(with applied item vector only, to show the final, actual representation)

@@
 ## Supuestos: Monotonía creciente
+## Representación gráfica
+
+@ackerman_graphical_1996:
+
+::: r-center
+```{r m2pl-representation-final}
+#| fig-align:  center
+#| fig-height: 7.2
+#| fig-width:  7.2
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+
+::: notes
+La representación como tal consiste solamente en el vector, aplicado en ese punto, obviamente.
+:::
 

### 2024-11-29 output/seminar-slideshow.qmd
Complete "Increasing monotony" slide

@@
 
-## Supuestos: Monotonía creciente
 ## Representación gráfica
@@
 
+## Supuestos: Monotonía creciente {.smaller}
+
 > the probability of answering an item correctly increases monotonically with an increase in each dimension being measured
@@
 
-<!-- # TODO: Ecuación modelo y ejemplo probabilidad constante -->
+::::: {.columns .fragment}
+::: column
+<br>
+
+$$
+  \begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1.5 \\
+        0
+      \end{bmatrix} \\
+      \\
+      \\
+    d_1 &= 0.75
+  \end{align}
+$$
+:::
+
+::: column
+```{r m2pl-representation-unidimensional}
+#| fig-align:  center
+#| fig-height: 7.2
+#| fig-width:  6
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1.5,  0,    0.75
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1),
+    breaks       = NULL,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::
 
@@
 Uno de los supuestos que hace es este.
-Eso tiene todo el sentido (casi), ya que si la probabilidad de un ítem "decrece" al aumentar el nivel de habilidad, es que algo está mal en el ítem: No puede ser que alguien que sepa más tenga menos probabilidad de acertar un ítem que alguien que sepa menos.
-Pero cuidado, si el ítem es "puro" de una dimensión, la probabilidad en la otra u otras dimensiones sería constante (discriminación = 0); es decir, la probabilidad puede ser "plana" (es una dimensión que no "está siendo medida por el ítem").
+Esto tiene todo el sentido, de acuerdo, ya que si la probabilidad de un ítem "decrece" al aumentar el nivel de habilidad, es que algo está mal en el ítem: No puede ser que alguien más inteligente (pongamos como ejemplo) tenga menos probabilidad de acertar un ítem que alguien menos inteligente.
+
+\[NEXT\]
+
+Pero cuidado, si el ítem es "puro" de una dimensión (o no mide "todas" las dimensiones del test), la probabilidad en la otra (u otras) dimensiones sel test sería constante (porque su discriminación sería 0); es decir, la probabilidad puede ser "plana" (es una dimensión que no "está siendo medida por el ítem").
+En este ejemplo, las rectas de isoprobabilidad serían paralelas al eje $\theta_2$.
 :::

### 2024-12-05 output/seminar-slideshow.qmd
Extend probability curves through whole axis range
Previously, their domain seemed to be delimited to an interval constrained between the range of the axis breaks

@@
 
-PROB_AXIS      <- seq(-3, 3, by = .01)
+PROB_AXIS      <- seq(-3.3, 3.3, by = .01)
 ERROR_POINTS   <- seq(-3, 3, length.out = 5)
@@
     labels       = NULL,
-    limits       = range(ERROR_POINTS) * 1.12,
+    limits       = range(PROB_AXIS),
     expand       = expansion()
@@
     minor_breaks = NULL,
-    limits = range(ERROR_POINTS) * 1.12,
+    limits = range(PROB_AXIS),
     expand = expansion()

### 2024-12-05 output/seminar-slideshow.qmd
Version inverse items generalization slide (WiP)

@@
 
+<!-- TODO: Put in column and make figure in 2nd column taller -->
+
 ---@reckase_difficulty_1985 [p. 402]
@@
 
-## Generalización: Ítems inversos
-
-<!-- # TODO: Curva de ítem inverso y ejemplo -->
+## Generalización: Ítems inversos {.smaller}
+
+:::::: columns
+:::: column
+> "*I neglect my duties*"
+
+-- [International Personality Item Pool](https://ipip.ori.org/newAB5CKey.htm#Conscientiousness)
+
+<br> <br>
+
+::: fragment
+```{r inverse-item-curve}
+#| fig-asp:   0.74
+#| fig-width: 6.5
+
+ITEMS_PALETTE <- PALETTE_BLUE["normal"] |>
+  unname() |>
+  setNames("prob_1")
+
+ITEMS <- tribble(
+  ~item, ~a,   ~b,
+   1,    -2,   0
+)
+
+logit_curve_data <- tibble(
+  latent_trait = PROB_AXIS,
+  prob         = ITEMS |>
+    select(-item) |>
+    pmap(irf, theta = latent_trait) |>
+    bind_cols()
+) |>
+  unnest(
+    prob,
+    names_sep = '_'
+  ) |>
+  rename_with(str_remove, pattern = '\\.{3}')
+
+logit_curve_data_long <- logit_curve_data |>
+  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
+
+logit_curve_data_long |>
+  ggplot(
+    mapping = aes(
+      latent_trait, prob,
+      color = item,
+      group = after_scale(color)
+    )
+  )                                                        +
+  geom_line(linewidth = LINE_WIDTH)                        +
+  labs(x = "Responsabilidad", y = "P(Y~i~ = 1)")           +
+  scale_y_continuous(
+    breaks       = 0:1,
+    minor_breaks = NULL,
+    limits       = 0:1,
+    expand       = expansion(mult = c(0, .05))
+  )                                                        +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    expand = expansion()
+  )                                                        +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  theme(
+    axis.title       = element_markdown(),
+    axis.title.x     = element_text(margin = margin(t = 1, unit = "lines")),
+    axis.title.y     = element_markdown(angle = 90),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
+    # The following is necessary to avoid showing the axis breaks, as dropping
+    #   them prevents the axis line from being drawn (see
+    #   https://github.com/tidyverse/ggplot2/issues/2983)
+    axis.text.x.bottom = element_blank(),
+    axis.ticks.length.x.bottom = unit(0, units = "cm"),
+    panel.grid.major.x = element_blank()
+  )
+```
+:::
+::::
+
+::: {.column .fragment}
+$$
+  \begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        -1 \\
+        1
+      \end{bmatrix} \\
+    d_1 &= 0
+  \end{align}
+$$
+
+<br>
+
+```{r m2pl-representation-final}
+#| fig-align:  center
+#| fig-height: 5
+#| fig-width:  5
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     -1,    1,   0
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1.5, 1),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+::::::
 

### 2024-12-05 output/seminar-slideshow.qmd
Add MDISC properties slide

@@
 La representación como tal consiste solamente en el vector, aplicado en ese punto, obviamente.
-:::
-
-## Supuestos: Monotonía creciente {.smaller}
-
-> the probability of answering an item correctly increases monotonically with an increase in each dimension being measured
-
-<!-- TODO: Put in column and make figure in 2nd column taller -->
-
----@reckase_difficulty_1985 [p. 402]
-
-::::: {.columns .fragment}
-::: column
-<br>
-
+Y lo interesante que tiene esta representación es que nos permite visualizar con claridad las propiedades de estos parámetros multidimensionales.
+:::
+
+## Propiedades {.smaller}
+
+:::::: columns
+:::: column
+$$
+  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
+$$
+
+::: {.fragment .r-stack style="color:#5c6eb1;" fragment-index="1"}
 $$
   \begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1.5 \\
-        0
+    \mathbf{a}_1 &= \begin{bmatrix}
+      1 \\
+      1
     \end{bmatrix} \\
-      \\
-      \\
-    d_1 &= 0.75
+    MDISC_1 &\approx 1,41
   \end{align}
@@
 :::
-
-::: column
-```{r m2pl-representation-unidimensional}
+::::
+
+::: {.column .fragment fragment-index="1"}
+```{r mdisc-comparison-1}
 #| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  6
-
-mirt_item <- tribble(
+#| fig-height: 8
+#| fig-width:  8
+
+mdisc_diff_items <- tribble(
   ~item, ~a_1, ~a_2, ~d,
-  1,     1.5,  0,    0.75
-)
-
-mirt_item_params <- mirt_item |>
+  1,     1,    1,    0,
+  2,     2,    2,    0,
+)
+
+mdisc_diff_params <- mdisc_diff_items |>
   compute_mirt_params(d, starts_with('a'))
 
-mirt_item_coords <- mirt_item_params |>
+mdisc_diff_coords <- mdisc_diff_params |>
   compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
@@
 
-mirt_item_coords |>
+mdisc_diff_coords |>
+  filter(item == 1) |>
   ggplot(
@@
   scale_x_continuous(
-    limits       = c(-1, 1.5),
+    limits       = c(-0.2, 2.2),
     breaks       = -1:2,
@@
   scale_y_continuous(
-    limits       = c(-1, 1),
-    breaks       = NULL,
+    limits       = c(-0.2, 2.2),
+    breaks       = -1:2,
     minor_breaks = NULL,
@@
 :::
+::::::
+
+::: notes
+Por ejemplo, la discriminación multidimensional, vemos que depende solamente del parámetro de discriminación,
+
+\[NEXT\]
+
+que en este ejemplo tiene dos componentes.
+:::
+
+## Propiedades {.smaller}
+
+::::::: columns
+::::: column
+$$
+  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
+$$
+
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{align}
+    \mathbf{a}_1 &= \begin{bmatrix}
+      1 \\
+      1
+    \end{bmatrix} \\
+    MDISC_1 &\approx 1,41
+  \end{align}
+$$
+:::
+
+::: {.r-stack style="color:#90214a;"}
+$$
+  \begin{align}
+    \mathbf{a}_2 &= \begin{bmatrix}
+      2 \\
+      2
+    \end{bmatrix} \\
+    MDISC_2 &\approx 2,83
+  \end{align}
+$$
+:::
+:::::
+
+::: column
+```{r mdisc-comparison-2}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+ITEMS_PALETTE <- c(
+  PALETTE_BLUE["normal"],
+  PALETTE_RASPBERRY["normal"],
+  PALETTE_APPLE["normal"]
+) |>
+  unname() |>
+  setNames(paste(1:3, sep = '_'))
+
+mdisc_diff_coords |>
+  arrange(item |> desc()) |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.2, 2.2),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.2, 2.2),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::::
+
+::: notes
+Cuanto más grandes sean los componentes del parámetro de discriminación, mayor será también la discriminación multidimensional.
+Hasta aquí, todo es sencillo (creo).
+:::
+
+## Supuestos: Monotonía creciente {.smaller}
+
+> the probability of answering an item correctly increases monotonically with an increase in each dimension being measured
+
+<!-- TODO: Put in column and make figure in 2nd column taller -->
+
+---@reckase_difficulty_1985 [p. 402]
+
+::::: {.columns .fragment}
+::: column
+<br>
+
+$$
+  \begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1.5 \\
+        0
+      \end{bmatrix} \\
+      \\
+      \\
+    d_1 &= 0.75
+  \end{align}
+$$
+:::
+
+::: column
+```{r m2pl-representation-unidimensional}
+#| fig-align:  center
+#| fig-height: 7.2
+#| fig-width:  6
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1.5,  0,    0.75
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1),
+    breaks       = NULL,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
 :::::

### 2024-12-13 output/seminar-slideshow.qmd
Correct symbol for continuous FA response variable
Explain multidimensional (partially) inverse items

@@
 $$
-  I_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
+  Y_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
 $$
@@
 $$
-  I_1 = 1 + \left[ -1, 0.5 \right]
+  Y_1 = 1 + \left[ -1, 0.5 \right]
     \begin{bmatrix}
@@
 
-:::::: columns
-:::: column
+<br>
+
 > "*I neglect my duties*"
@@
 
-<br> <br>
-
-::: fragment
+<br>
+
+::: {.fragment .r-center}
 ```{r inverse-item-curve}
+#| fig-align: center
 #| fig-asp:   0.74
@@
 :::
-::::
-
-::: {.column .fragment}
+
+::: notes
+Pero además, si hablamos de ítems "no cognitivos" (p.ej. personalidad) en los que la persona evaluada puede estar "de acuerdo" o "en desacuerdo", sí podemos tener ítems con probabilidad "decreciente".
+
+Llevado al caso multidimensional, lo que nos podemos encontrar es con ítems que por ejemplo tengan todos los valores de discriminación negativos, o algunos negativos y otros positivos.
+Pensad por ejemplo que tenemos el problema de los "sesgos de respuesta".
+Vamos a imaginar que alguien estuviera respondiendo con un sesgo de "deseabilidad social" (intenta dar una buena imagen de sí mismo o de sí misma, de manera consciente o no).
+A un ítem como este, tendería a responder que no, al margen de cómo autoevalúe su verdadero nivel de responsabilidad.
+Por lo tanto, tendríamos un ítems con dos valores negativos de discriminación, uno en la dimensión de responsabilidad, y otro en la dimensión de deseabilidad social.
+Sin embargo, consideremos ahora un sesgo de "aquiescencia", que hace que la gente tienda a estar de acuerdo con todo, o a decir que sí a todo sin un criterio muy reflexivo ("¿Descuidas tus obligaciones?
+Mmmmm...
+sí"; "¿Atiendes tus obligaciones con diligencia?
+Mmmmm...
+sí").
+Cuanto más aquiescente sea una persona (a igual responsabilidad) mayor será su tendencia o probabilidad a responder afirmativamente al ítem.
+Sin embargo, cuanto más responsable sea una persona (a igual aquiescencia) menor será su tendencia (o probabilidad) a responder afirmativamente.
+En ese caso, tendríamos un ítem con un valor negativo de discriminación en la dimensión de responsabilidad, y uno positivo en la de aquiescencia.
+:::
+
 $$

### 2024-12-13 output/seminar-slideshow.qmd
Explain inverse item parameter properties

@@
 
+## Generalización: Ítems inversos
+
+<br>
+
+$$
+  MDISC_i = \sqrt{a_{1i}^2 + ... + a_{ni}^2}
+$$
+
+::: fragment
+$$
+  D_i = \frac{-d_i}{MDISC_i}
+$$
+:::
+
+::: fragment
+$$
+  cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
+$$
+:::
+
+::: notes
+Si vemos la expresión de los parámetros multidimensionales, podemos prevér cómo van a ser en estos casos.
+Aquí he preferido representar la discriminación como suma de cuadrados, en lugar de como el producto matricial, pero es lo mismo, ¿os acordáis?
+El primer elemento de un vector por el primero del otro, y asi sucesivamente...
+todos sumados.
+Sólo que en este caso el primer y el segundo vector son el mismo, así que es la suma de los elementos del vector al cuadrado.
+Por lo tanto, la discriminación multidimensional siempre va a ser positiva.
+¿Lo veis claro?
+
+Por otro lado, la distancia (con signo), al depender sólo de la intersección y la discriminación multidimensional, nos da igual que los valores de la discriminación sean negativos.
+Siempre va a cumplir que su desplazamiento es "inverso" al valor indicado por la intersección, como indicábamos antes.
+
+La dirección sin embargo tiene una peculiaridad, y es que vemos que cada uno de los cosenos puede tener un signo distinto, que va a ser el mismo que el del valor de la discriminación para esa dimensión.
+Eso lo que va a hacer es que el sentido del item en cada una de esas dimensiones venga determinado por el signo de ese valor.
+:::
+
 $$

### 2024-12-13 output/seminar-slideshow.qmd
Draft representation of (partially) inverse items

@@
 
+## Generalización: Ítems inversos
+
+<!-- # TODO: Representación de ítems con a (una o varias) negativa (tabla y gráfico) -->
+
 $$
@@
 
-```{r m2pl-representation-final}
+```{r inverse-item-representation}
 #| fig-align:  center
@@
 ```
-:::
-::::::
-
-<!-- # TODO: Fórmulas de los parámetros -->
-
-<!-- # TODO: Ejemplo con a negativa -->
-
-::: notes
-Pero además, si hablamos de ítems "no cognitivos" (p.ej. personalidad) en los que la persona evaluada puede estar "de acuerdo" o "en desacuerdo", sí podemos tener ítems con probabilidad "decreciente".
+
+::: notes
+La representación de estos ítems sería como se muestra en este gráfico.
+<!-- # TODO: Descripción de los ítems representados -->
 
@@
 Los cosenos tendrán, cada uno, el mismo que el parámetro de discriminación de esa dimensión.
-:::
-
-## Generalización: Ítems inversos
-
-<!-- # TODO: Representación de ítems con a (una o varias) negativa (tabla y gráfico) -->
-
-::: notes
-La representación de estos ítems sería como se muestra en este gráfico.
-<!-- # TODO: Descripción de los ítems representados -->
 :::

### 2024-12-20 output/seminar-slideshow.qmd
Add parameter name to "Properties" slide title
Adjust axis label position in properties plots

@@
 
-## Propiedades {.smaller}
+## Propiedades: *MDISC* {.smaller}
 
@@
     axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
+    axis.title.x       = element_markdown(hjust = 1/12),
+    axis.title.y.right = element_markdown(vjust = 1/12),
     axis.ticks         = element_blank(),
@@
 Cuanto más grandes sean los componentes del parámetro de discriminación, mayor será también la discriminación multidimensional.
+(Aunque en este ejemplo sean iguales entre sí, no tienen por qué serlo).
 Hasta aquí, todo es sencillo (creo).

### 2024-12-20 output/sandbox_slideshow.qmd
Add probability axis elements for unidimensional representation
Add item palette to sandbox slide

@@
 ```{r pre}
+PROB_AXIS    <- seq(-3, 3, by = .01)
+ERROR_POINTS <- seq(-3, 3, length.out = 5)
 PROB_AXIS_2D <- seq(-3, 3, by = .1)
@@
 
+ITEMS_PALETTE <- c(
+  PALETTE_BLUE["normal"],
+  PALETTE_RASPBERRY["normal"],
+  PALETTE_APPLE["normal"],
+  PALETTE_TANGERINE["normal"]
+) |>
+  unname() |>
+  setNames(paste(1:4, sep = '_'))
 ```

### 2024-12-20 output/seminar-slideshow.qmd
Add D (signed distance) properties slides

@@
 
+## Propiedades: *D* {.smaller}
+
+:::::: columns
+:::: column
+$$
+  D_i = \frac{-d_i}{MDISC_i}
+$$
+
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_1     &= 2 \\
+      MDISC_1 &= \sqrt{2}
+    \end{align} &
+    D_1 \approx -1,41
+  \end{matrix}
+$$
+:::
+::::
+
+::: column
+```{r d-comparison-1}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+mdisc_diff_items <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    1,    2,
+  2,     1,    1,    1,
+  3,     1,    1,    -1,
+  4,     2,    2,    -1,
+)
+
+mdisc_diff_params <- mdisc_diff_items |>
+  compute_mirt_params(d, starts_with('a'))
+
+mdisc_diff_coords <- mdisc_diff_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+mdisc_diff_coords |>
+  filter(item == 1) |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1.1/3.5),
+    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+::::::
+
+::: notes
+En cuanto a la distancia con signo, obviamente es proporcional a la intersección del modelo, eso creo que también es muy evidente
+:::
+
+## Propiedades: *D* {.smaller}
+
+::::::: columns
+::::: column
+$$
+  D_i = \frac{-d_i}{MDISC_i}
+$$
+
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_1     &= 2 \\
+      MDISC_1 &= \sqrt{2}
+    \end{align} &
+    D_1 \approx -1,41
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#90214a;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_2     &= 1 \\
+      MDISC_2 &= \sqrt{2}
+    \end{align} &
+    D_2 \approx -0,71
+  \end{matrix}
+$$
+:::
+:::::
+
+::: column
+```{r d-comparison-2}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+mdisc_diff_coords |>
+  filter(item %in% 1:2) |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1.1/3.5),
+    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::::
+
+::: notes
+Aquí vemos un ítem con la mitad del valor de la intersección que el primero.
+Por lo tanto, su distancia al origen es la mitad.
+(Los ítems están un poco solapados, pero empieza aquí; espero que se vea bien).
+:::
+
+## Propiedades: *D* {.smaller}
+
+:::::::: columns
+:::::: column
+$$
+  D_i = \frac{-d_i}{MDISC_i}
+$$
+
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_1     &= 2 \\
+      MDISC_1 &= \sqrt{2}
+    \end{align} &
+    D_1 \approx -1,41
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#90214a;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_2     &= 1 \\
+      MDISC_2 &= \sqrt{2}
+    \end{align} &
+    D_2 \approx -0,71
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#749f4c;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_3     &= -1 \\
+      MDISC_3 &= \sqrt{2}
+    \end{align} &
+    D_3 \approx 0,71
+  \end{matrix}
+$$
+:::
+::::::
+
+::: column
+```{r d-comparison-3}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+mdisc_diff_coords |>
+  filter(item %in% 1:3) |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1.1/3.5),
+    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+::::::::
+
+::: notes
+Y además, el desplazamiento respecto del origen es en sentido opuesto al signo de la intersección.
+(Si consideramos el sentido positivo hacia donde el ítem apunta, que es el ascendente para ambos rasgos latentes, o sea, hacia arriba a la derecha).
+Este tercer ítem se encuentra aquí, justo en la punta del segundo.
+:::
+
+## Propiedades: *D* {.smaller}
+
+::::::: columns
+::::: column
+$$
+  D_i = \frac{-d_i}{MDISC_i}
+$$
+
+::: {.r-stack style="color:#749f4c;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_3     &= -1 \\
+      MDISC_3 &= \sqrt{2}
+    \end{align} &
+    D_3 \approx 0,71
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#d76f47;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_4     &= -1 \\
+      MDISC_4 &= \sqrt{8}
+    \end{align} &
+    D_4 \approx 0,35
+  \end{matrix}
+$$
+:::
+:::::
+
+::: column
+```{r d-comparison-4}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+mdisc_diff_coords |>
+  slice(4:3) |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1.1/3.5),
+    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::::
+
+::: notes
+Y además, el desplazamiento respecto del origen es en sentido opuesto al signo de la intersección.
+(Si consideramos el sentido positivo hacia donde el ítem apunta, que es el ascendente para ambos rasgos latentes, o sea, hacia arriba a la derecha).
+Este tercer ítem se encuentra aquí, justo en la punta del segundo.
+:::
+
 ## Supuestos: Monotonía creciente {.smaller}

### 2024-12-20 README.Rmd

---
output: github_document
editor_options:
  markdown:
    wrap:      80
    canonical: false
---

```{r config}
#| echo: false

REPOSITORY_NAME <- "mirt-parameters"
GITHUB_USERNAME <- "DaniMori"
AUTHOR_NAME     <- "Daniel Morillo"
GITHUB_OWNER    <- GITHUB_USERNAME

# Objects created automatically (no need to edit)

## Repository configuration objects
profile_url     <- paste0("https://github.com/", GITHUB_OWNER)
repo_url        <- paste(profile_url, REPOSITORY_NAME, sep = '/')
https_clone_url <- paste0(repo_url, '.git')
creation_year   <- lubridate::now() |> lubridate::year()

## Software version objects
r_version     <- paste(version$major, version$minor, sep = '.')
rs_version    <- rstudioapi::versionInfo()$long_version
rs_version_dl <- rs_version |> stringr::str_replace('\\+', '-')
q_version     <- quarto::quarto_version()
git_version   <- gitr::git_version()
```

# Repository ``r REPOSITORY_NAME``

Paper "A generalized definition of multidimensional item response
theory parameters"

# License

<!-- TODO: Choose a license and update this section, if necessary -->

[Repository `r REPOSITORY_NAME`](`r repo_url`) © `r creation_year` by
[`r AUTHOR_NAME`](`r profile_url`) is licensed under [CC BY 4.0][license].
Please see the [license file][licfile].

[license]: https://creativecommons.org/licenses/by/4.0

[licfile]: LICENSE.md

## Attributions

### Rstudio project template

This project makes use of the [rproj-template][selflink] Github template
created by [Daniel Morillo][profile] and licensed under the 
[Creative Commons Attribution 4.0 International license][repo_lic].

[selflink]: https://github.com/DaniMori/rproj-template

[profile]: https://github.com/DaniMori

[repo_lic]: https://creativecommons.org/licenses/by/4.0

<!-- TODO: Add attributions to other software components, if necessary -->

### <Software component>

# Project installation

## Software components

Start by installing the following software components:

- [R version `r r_version`][R]:
  In Windows, using the [binary installer][inst] is recommended.

[R]: https://cran.rstudio.com/bin/windows/base/old/`r r_version`/
[inst]: https://cran.rstudio.com/bin/windows/base/old/`r r_version`/R-`r r_version`-win.exe

- [Rstudio Desktop][RS]: Although not strictly necessary, it is recommended
  to install the Rstudio IDE; for strict reproducibility, use build
  [`r rs_version` for Windows 10/11][RSv].

[RS]: https://posit.co/download/rstudio-desktop/

[RSv]: https://download1.rstudio.org/electron/windows/RStudio-`r rs_version_dl`.exe

- [Quarto publishing system][Q]: An additional component used by Rstudio to
generate and publish literate computing outputs. For strict reproducibility
please use build `r q_version`; On Windows, use [the 64-bit installer][QW].

[Q]: https://quarto.org/

[QW]: https://github.com/quarto-dev/quarto-cli/releases/download/v`r q_version`/quarto-`r q_version`-win.msi

- [Git client][G]: Install the Git client in order to be able to clone locally
  the project repository.
  On Windows, use [the 64-bit Windows installer][GW].

[G]: https://git-scm.com/download

[GW]: https://github.com/git-for-windows/git/releases/download/v`r git_version`.windows.1/Git-`r git_version`-64-bit.exe

## Installing the project locally

This project is hosted as a GitHub repository.
It can be cloned as a local Git repository following [these instructions][CR]
(steps 2 through 7).
Note that this will create a local copy of ('clone') the GitHub repository as an
Rstudio project in the folder specified.
The URL that must be entered into the `Repository URL` text box is:

```
`r https_clone_url`
```

[CR]: https://book.cds101.com/using-rstudio-server-to-clone-a-github-repo-as-a-new-project.html#step---2

**IMPORTANT:** It is totally unrecommended to clone a git repository inside a
cloud storage folder (e.g., Dropbox, OneDrive). Please note that GitHub serves
the purpose of backing up the repository, so no cloud storage is necessary.
Similarly, cloning the repository in a network folder may cause problems with
the `renv` environment (see below); do it at your own risk!

After cloning the repository,
the Rstudio project will open automatically in the Rstudio IDE.
If it doesn't, or you want to return later to the project in Rstudio,
you can do so by double clicking on the file `rstudio_project.Rproj`
that has been created in the project folder when cloning the repository.

**NOTE:** It is common practice to avoid using and versioning `.Rprofile` files.
However, this project uses [package `renv`][renv]
to create a reproducible environment,
which needs the `.Rprofile` file that lives in the root directory of the
project. **Please DO NOT delete or edit this file**; it will install and
activate the `renv` package and make it ready for restoring the environment.

[renv]: https://cran.r-project.org/package=renv

## Restoring the environment

The reproducible environment created by `renv` must be restored to install all
the packages this project needs to be built properly. If `renv` does not
initialize automatically (check the console for messages about this), you will
need to manually install the package first:

```r
install.packages("renv")
```

Once it is successfully installed, use the "renv" -\> "Restore library..."
button in Rstudio's "Packages" tab to restore the environment. Alternatively,
you can type in the console:

```r         
renv::restore(prompt = FALSE)
```

# Repository structure

The file structure of this repository is as follows:

```
`r REPOSITORY_NAME`
|
|--- apps         (To store apps, e.g. in Shiny)
|
|--- dat          (To store input datasets; must NEVER be checked-in to Github)
|
|--- doc          (To store important documentation of the project)
|    |
|    |--- minutes (To store meeting minutes)
|
|--- notebooks    (Notebooks to explore data and test processes live here)
|
|--- output       (Processing outputs; files must be individually "checked-in"
|                 when necessary)
|
|--- R            (R functions created for this project live here)
|
|--- renv         (System library necesssary for `renv` to work. DON'T TOUCH)
|
|--- src          (Source scripts that implement the main processes)
|
|--- www          (Project assets, e.g., images, bibliography files, etc.)
```

Use the folders as indicated to store the different files and generate the
outputs of the processes.

# How to use this project

## Main output

The main file is
[`output/paper_draft/paper_M2PLM_parameters.Rmd`][paper],
which renders the article main text output.

[paper]: output/paper_draft/paper_M2PLM_parameters.Rmd

The output of this file is the document that will eventually
be submitted to a journal,
along with the necessary additional files (i.e., cover letter).
If the rendering is successful, the output will be generated in file
"output/Main_text.docx".

**NOTE:** When opening the output document,
Word will prompt to "Update the fields".
This is [set by design in the R package `officer`][update_prompt],
used to render the document, for security reasons.
It is enough to click on "Yes" in the pop-up window.

[update_prompt]: https://ardata-fr.github.io/officeverse/faq.html#update-fields

## Rendering the output

A script for building the project is provided in file
[`src/Build_project.R`](src/Build_project.R).
The easiest way to render the output is by sourcing this script, running the
following command:

```r
source("src/Build_project.R", encoding = 'UTF-8')
```

The output file generated will be in the following path:

```
"output/Main_text.docx"
```

**NOTE:** Rendering the project may take a while, especially the first time.
This is due to the `renv` package restoring the environment, which implies
downloading and installing several R packages. Please, be patient!

### 2024-12-20 output/seminar-slideshow.qmd
Correct cosine expressions in LaTeX
Versions "template" slide of cosines properties

--- 2024-12-13 output/seminar-slideshow.qmd
@@
     canonical: true
-execute: 
-  cache:       true
 lang:          es
@@
     results:   asis
+    cache:     true
     autodep:   true
@@
 $$
-  Y_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
+  I_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
 $$
@@
 $$
-  Y_1 = 1 + \left[ -1, 0.5 \right]
+  I_1 = 1 + \left[ -1, 0.5 \right]
     \begin{bmatrix}
@@
 
-PROB_AXIS      <- seq(-3.3, 3.3, by = .01)
+PROB_AXIS      <- seq(-3, 3, by = .01)
 ERROR_POINTS   <- seq(-3, 3, length.out = 5)
@@
     labels       = NULL,
-    limits       = range(PROB_AXIS),
+    limits       = range(ERROR_POINTS) * 1.12,
     expand       = expansion()
@@
     minor_breaks = NULL,
-    limits = range(PROB_AXIS),
+    limits = range(ERROR_POINTS) * 1.12,
     expand = expansion()
@@
 $$
-  D_i = \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
+  MID = \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
 $$
@@
 
-d_i_spike <- mirt_item_coords |>
+mirt_item_spikes <- mirt_item_coords |>
   select(item, trait_1 = origin_1, trait_2 = origin_2) |>
@@
   geom_segment(
-    data    = d_i_spike,
+    data    = mirt_item_spikes,
     mapping = aes(
@@
     geom = "richtext",
-    x = 0.15, y = 0.15,
-    label = "*D~i~*",
-    size = 9,
+    x = 0.16, y = 0.15,
+    label = "*MID~i~*",
+    size = 6,
     angle = mirt_item_params |> pull(deg_1),
@@
 
-::: r-center
-```{r m2pl-representation}
-#| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  7.2
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data    = d_i_spike,
-    mapping = aes(
-      trait_1_start, trait_2_start,
-      xend = trait_1_end,
-      yend = trait_2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = PALETTE_BLUE["light"]
-  ) +
-  geom_point(
-    data = mirt_item_coords,
-    mapping = aes(origin_1, origin_2),
-    inherit.aes = FALSE,
-    size = I(4)
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    geom = "richtext",
-    x = 0.16, y = 0.20,
-    label = "*D~i~*",
-    size = 9,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  annotate(
-    "richtext",
-    x = mirt_item_coords |> pull(origin_1) + 0.40,
-    y = mirt_item_coords |> pull(origin_2) + 0.32,
-    label = "*MDISC~i~*",
-    size  = 9,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-
 ::: notes
@@
 
-## Representación gráfica
-
-@ackerman_graphical_1996:
-
-::: r-center
-```{r m2pl-representation-final}
-#| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  7.2
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-
-::: notes
-La representación como tal consiste solamente en el vector, aplicado en ese punto, obviamente.
-Y lo interesante que tiene esta representación es que nos permite visualizar con claridad las propiedades de estos parámetros multidimensionales.
-:::
-
-## Propiedades {.smaller}
-
-:::::: columns
-:::: column
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-
-::: {.fragment .r-stack style="color:#5c6eb1;" fragment-index="1"}
-$$
-  \begin{align}
-    \mathbf{a}_1 &= \begin{bmatrix}
-      1 \\
-      1
-    \end{bmatrix} \\
-    MDISC_1 &\approx 1,41
-  \end{align}
-$$
-:::
-::::
-
-::: {.column .fragment fragment-index="1"}
-```{r mdisc-comparison-1}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-mdisc_diff_items <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1,    1,    0,
-  2,     2,    2,    0,
-)
-
-mdisc_diff_params <- mdisc_diff_items |>
-  compute_mirt_params(d, starts_with('a'))
-
-mdisc_diff_coords <- mdisc_diff_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mdisc_diff_coords |>
-  filter(item == 1) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-::::::
-
-::: notes
-Por ejemplo, la discriminación multidimensional, vemos que depende solamente del parámetro de discriminación,
-
-\[NEXT\]
-
-que en este ejemplo tiene dos componentes.
-:::
-
-## Propiedades {.smaller}
-
-::::::: columns
-::::: column
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{align}
-    \mathbf{a}_1 &= \begin{bmatrix}
-      1 \\
-      1
-    \end{bmatrix} \\
-    MDISC_1 &\approx 1,41
-  \end{align}
-$$
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$$
-  \begin{align}
-    \mathbf{a}_2 &= \begin{bmatrix}
-      2 \\
-      2
-    \end{bmatrix} \\
-    MDISC_2 &\approx 2,83
-  \end{align}
-$$
-:::
-:::::
-
-::: column
-```{r mdisc-comparison-2}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-ITEMS_PALETTE <- c(
-  PALETTE_BLUE["normal"],
-  PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"]
-) |>
-  unname() |>
-  setNames(paste(1:3, sep = '_'))
-
-mdisc_diff_coords |>
-  arrange(item |> desc()) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::::
-
-::: notes
-Cuanto más grandes sean los componentes del parámetro de discriminación, mayor será también la discriminación multidimensional.
-Hasta aquí, todo es sencillo (creo).
-:::
-
-## Supuestos: Monotonía creciente {.smaller}
+## Supuestos: Monotonía creciente
 
@@
 
-<!-- TODO: Put in column and make figure in 2nd column taller -->
-
 ---@reckase_difficulty_1985 [p. 402]
 
-::::: {.columns .fragment}
-::: column
-<br>
-
-$$
-  \begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1.5 \\
-        0
-      \end{bmatrix} \\
-      \\
-      \\
-    d_1 &= 0.75
-  \end{align}
-$$
-:::
-
-::: column
-```{r m2pl-representation-unidimensional}
-#| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  6
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1.5,  0,    0.75
-)
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1),
-    breaks       = NULL,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::
+<!-- # TODO: Ecuación modelo y ejemplo probabilidad constante -->
 
@@
 Uno de los supuestos que hace es este.
-Esto tiene todo el sentido, de acuerdo, ya que si la probabilidad de un ítem "decrece" al aumentar el nivel de habilidad, es que algo está mal en el ítem: No puede ser que alguien más inteligente (pongamos como ejemplo) tenga menos probabilidad de acertar un ítem que alguien menos inteligente.
-
-\[NEXT\]
-
-Pero cuidado, si el ítem es "puro" de una dimensión (o no mide "todas" las dimensiones del test), la probabilidad en la otra (u otras) dimensiones sel test sería constante (porque su discriminación sería 0); es decir, la probabilidad puede ser "plana" (es una dimensión que no "está siendo medida por el ítem").
-En este ejemplo, las rectas de isoprobabilidad serían paralelas al eje $\theta_2$.
-:::
-
-## Generalización: Ítems inversos {.smaller}
-
-<br>
-
-> "*I neglect my duties*"
-
--- [International Personality Item Pool](https://ipip.ori.org/newAB5CKey.htm#Conscientiousness)
-
-<br>
-
-::: {.fragment .r-center}
-```{r inverse-item-curve}
-#| fig-align: center
-#| fig-asp:   0.74
-#| fig-width: 6.5
-
-ITEMS_PALETTE <- PALETTE_BLUE["normal"] |>
-  unname() |>
-  setNames("prob_1")
-
-ITEMS <- tribble(
-  ~item, ~a,   ~b,
-   1,    -2,   0
-)
-
-logit_curve_data <- tibble(
-  latent_trait = PROB_AXIS,
-  prob         = ITEMS |>
-    select(-item) |>
-    pmap(irf, theta = latent_trait) |>
-    bind_cols()
-) |>
-  unnest(
-    prob,
-    names_sep = '_'
-  ) |>
-  rename_with(str_remove, pattern = '\\.{3}')
-
-logit_curve_data_long <- logit_curve_data |>
-  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
-
-logit_curve_data_long |>
-  ggplot(
-    mapping = aes(
-      latent_trait, prob,
-      color = item,
-      group = after_scale(color)
-    )
-  )                                                        +
-  geom_line(linewidth = LINE_WIDTH)                        +
-  labs(x = "Responsabilidad", y = "P(Y~i~ = 1)")           +
-  scale_y_continuous(
-    breaks       = 0:1,
-    minor_breaks = NULL,
-    limits       = 0:1,
-    expand       = expansion(mult = c(0, .05))
-  )                                                        +
-  scale_x_continuous(
-    minor_breaks = NULL,
-    expand = expansion()
-  )                                                        +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  theme(
-    axis.title       = element_markdown(),
-    axis.title.x     = element_text(margin = margin(t = 1, unit = "lines")),
-    axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
-    # The following is necessary to avoid showing the axis breaks, as dropping
-    #   them prevents the axis line from being drawn (see
-    #   https://github.com/tidyverse/ggplot2/issues/2983)
-    axis.text.x.bottom = element_blank(),
-    axis.ticks.length.x.bottom = unit(0, units = "cm"),
-    panel.grid.major.x = element_blank()
-  )
-```
-:::
+Eso tiene todo el sentido (casi), ya que si la probabilidad de un ítem "decrece" al aumentar el nivel de habilidad, es que algo está mal en el ítem: No puede ser que alguien que sepa más tenga menos probabilidad de acertar un ítem que alguien que sepa menos.
+Pero cuidado, si el ítem es "puro" de una dimensión, la probabilidad en la otra u otras dimensiones sería constante (discriminación = 0); es decir, la probabilidad puede ser "plana" (es una dimensión que no "está siendo medida por el ítem").
+:::
+
+## Generalización: Ítems inversos
+
+<!-- # TODO: Curva de ítem inverso y ejemplo -->
+
+<!-- # TODO: Fórmulas de los parámetros -->
+
+<!-- # TODO: Ejemplo con a negativa -->
 
@@
 Pero además, si hablamos de ítems "no cognitivos" (p.ej. personalidad) en los que la persona evaluada puede estar "de acuerdo" o "en desacuerdo", sí podemos tener ítems con probabilidad "decreciente".
-
-Llevado al caso multidimensional, lo que nos podemos encontrar es con ítems que por ejemplo tengan todos los valores de discriminación negativos, o algunos negativos y otros positivos.
-Pensad por ejemplo que tenemos el problema de los "sesgos de respuesta".
-Vamos a imaginar que alguien estuviera respondiendo con un sesgo de "deseabilidad social" (intenta dar una buena imagen de sí mismo o de sí misma, de manera consciente o no).
-A un ítem como este, tendería a responder que no, al margen de cómo autoevalúe su verdadero nivel de responsabilidad.
-Por lo tanto, tendríamos un ítems con dos valores negativos de discriminación, uno en la dimensión de responsabilidad, y otro en la dimensión de deseabilidad social.
-Sin embargo, consideremos ahora un sesgo de "aquiescencia", que hace que la gente tienda a estar de acuerdo con todo, o a decir que sí a todo sin un criterio muy reflexivo ("¿Descuidas tus obligaciones?
-Mmmmm...
-sí"; "¿Atiendes tus obligaciones con diligencia?
-Mmmmm...
-sí").
-Cuanto más aquiescente sea una persona (a igual responsabilidad) mayor será su tendencia o probabilidad a responder afirmativamente al ítem.
-Sin embargo, cuanto más responsable sea una persona (a igual aquiescencia) menor será su tendencia (o probabilidad) a responder afirmativamente.
-En ese caso, tendríamos un ítem con un valor negativo de discriminación en la dimensión de responsabilidad, y uno positivo en la de aquiescencia.
-:::
-
-## Generalización: Ítems inversos
-
-<br>
-
-$$
-  MDISC_i = \sqrt{a_{1i}^2 + ... + a_{ni}^2}
-$$
-
-::: fragment
-$$
-  D_i = \frac{-d_i}{MDISC_i}
-$$
-:::
-
-::: fragment
-$$
-  cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
-$$
-:::
-
-::: notes
-Si vemos la expresión de los parámetros multidimensionales, podemos prevér cómo van a ser en estos casos.
-Aquí he preferido representar la discriminación como suma de cuadrados, en lugar de como el producto matricial, pero es lo mismo, ¿os acordáis?
-El primer elemento de un vector por el primero del otro, y asi sucesivamente...
-todos sumados.
-Sólo que en este caso el primer y el segundo vector son el mismo, así que es la suma de los elementos del vector al cuadrado.
-Por lo tanto, la discriminación multidimensional siempre va a ser positiva.
-¿Lo veis claro?
-
-Por otro lado, la distancia (con signo), al depender sólo de la intersección y la discriminación multidimensional, nos da igual que los valores de la discriminación sean negativos.
-Siempre va a cumplir que su desplazamiento es "inverso" al valor indicado por la intersección, como indicábamos antes.
-
-La dirección sin embargo tiene una peculiaridad, y es que vemos que cada uno de los cosenos puede tener un signo distinto, que va a ser el mismo que el del valor de la discriminación para esa dimensión.
-Eso lo que va a hacer es que el sentido del item en cada una de esas dimensiones venga determinado por el signo de ese valor.
-:::
-
-## Generalización: Ítems inversos
-
-<!-- # TODO: Representación de ítems con a (una o varias) negativa (tabla y gráfico) -->
-
-$$
-  \begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        -1 \\
-        1
-      \end{bmatrix} \\
-    d_1 &= 0
-  \end{align}
-$$
-
-<br>
-
-```{r inverse-item-representation}
-#| fig-align:  center
-#| fig-height: 5
-#| fig-width:  5
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     -1,    1,   0
-)
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1.5, 1),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-
-::: notes
-La representación de estos ítems sería como se muestra en este gráfico.
-<!-- # TODO: Descripción de los ítems representados -->
 
@@
 Los cosenos tendrán, cada uno, el mismo que el parámetro de discriminación de esa dimensión.
+:::
+
+## Generalización: Ítems inversos
+
+<!-- # TODO: Representación de ítems con a (una o varias) negativa (tabla y gráfico) -->
+
+::: notes
+La representación de estos ítems sería como se muestra en este gráfico.
+<!-- # TODO: Descripción de los ítems representados -->
 :::

### 2025-01-07 output/Letter_to_editor_reviewers.Rmd
Version response to reviewers initial draft

---
output:
  officedown::rdocx_document:
    mapstyles:
      First Paragraph: Body Text
    base_format: bookdown::word_document2
    reference_docx: ../www/Template.docx
    number_sections: no
    keep_md:         no
editor_options: 
  chunk_output_type: console
---

September 16, 2024. <!-- TODO: Update date -->

<br>

Sandip Sinharay, PhD \
Editor, *Psychometrika* \
Educational Testing Service, USA

<br>

Dear Dr. Sinharay:

Please find attached the updated version of our manuscript
*"A generalized definition of multidimensional item response theory
parameters"*.
We hope that this new version lives up to the expectations of
the action editor and the reviewers;
we have put as much effort as we have been able to address all the comments and
suggestions, which we believe have helped improve the quality of our paper.
<!-- TODO: Complete -->

Our manuscript presents a generalization of the multidimensional IRT parameters
introduced by Reckase and Reckase & McKinley in their articles
["The difficulty of test items that measure more than one ability" (1985)][1]
and
["The discriminating power of items that measure more than one dimension"
(1991)][2],
respectively.

[1]: https://doi.org/10.1177/014662168500900409

[2]: https://doi.org/10.1177/014662169101500407

We thank you in advance for considering our reviewed manuscript.
Please feel free to contact us again at my aforementioned email address
(dvmorillo@psi.uned.es).
We will be happy to address any further comments or suggestions
the action editor or the reviewers may deem necesssary.

<br>

Sincerely,

Daniel Morillo, Ph.D.\
Department of Behavioral Science Methodology \
Universidad Nacional de Educación a Distancia (UNED) \
C. Juan del Rosal 10, Office 2.74 \
28040 Madrid, Spain

Response to the action editor:

> I have a concern with making the latent ability axes nonorthogonal to represent the covariance structure between the latent abilities.
> Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.

> The statement (p.13, line 8) about “the test space where item parameters naturally lie” needs greater explanation with examples.

> Like Reviewer 3, I do not understand how the magnitude of discrimination values (p 29 , 7) can be compared/interpreted values of discrimination are comparable from the two representations.
> It seems that the metric of the represented space has changed, and equivalent numbers will not have equivalent interpretations.

> I also wonder about how one compares the standard errors of the item parameters when the covariance covariance structures are different.  Are they directly comparable?

> I would like to see the motivation and psychometric utility of this approach.
> Does it improve the interpretation of the latent abilities and thus result in more insight into how the results are used?
> The author needs to address this with legitimate examples for testing practitioners.

> What implications does your approach have for standard psychometric procedures like equating.

> Are there standard computer programs that you would recommend to do the estimation/modeling you are proposing?

Response to Reviwer 1:

> 1. Introduction, too lengthy, 1-2 pages will be fine.

> 2. Introduction, do you have any hypothesis for the current study? Please briefly describe it at the end.

> 3. The details about the datasets should be described here.

> 4. Within or between multidimensional latent? This should be detailed.

> 5. Limitation is missing and should be added here.

> 6. Conclusion is missing and should be added here.

Response to Reviwer 2:

> [...] it is difficult to ascertain its practical value, as it is unclear how it can be applied in the real world.

> While the manuscript includes a graphical representation example with two-dimensional latent variables, real-world applications often involve higher dimensions (e.g., the Big Five personality traits or
Cattell’s 16 personality factors). If the current results cannot be applied to the
high-dimensional cases, would that significantly limit their practical value and
applicability?

> 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect.
> It should be “$\mathcal{C}$”.

> 2. The notations in the manuscript lack clarity, particularly in Eq. (19) on page 12.
> Based on Eq. (5) and Eq. (15), $\mathbf{a}_i$ in the numerator of Eq. (19) should be $\mathbf{a}_i^{\mathcal{B}*}$.

> Similar issues occur in Eqs. (26) and (27) on page 17.

> 3. In the second graph on page 18, the authors state, “*That is, the less discriminating the item, the further it will be from the origin.*”
> However, the value of $D_i$ is influenced not only by the $MDISC_i$ parameter but also by $d_i$.
> For instance, in Table 1 on page 20, $d_i$ for item 1 is zero, meaning that regardless of $MDISC_i$, $D_i$ remains zero.

Response to Reviwer 3:

> The manuscript highlights the fact that the coordinate system used to represent results typical of educational achievement and other types of social measures is at the discretion of the person performing the analysis, or the entities developing computer software, if default settings are used in the analysis.

> This is a very important issue and it is not frequently discussed in the research literature. Two researchers can administer the identical instrument to samples of individuals with different characteristics – say typical high school students versus those who have a different native language – and get different representations of the test results from analyses programs, such as the correlations between estimated constructs.  This manuscript stimulates much thought related to this problem.

> I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes.
> The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals.
> How are these features interpreted when the coordinate axes are non-orthogonal?

> Care needs to be taken when using nonorthogonal coordinates that the distances between points in the space are correctly represented.
> What formula should be used for the distance between two points in each case?

> Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005).
> It would help if a page number were provided for the discussion.
> I also wonder about the Harman (1970) reference.
> I know that he shows a plot of length of arm against length of leg using orthogonal axes for the plot even though the two measures are correlated.
> I am not not sure how the axes should be oriented to reflect the correlation between the two measures.

> Page 6  Line 21 The expression on this line needs more explanation.
> I am not sure what the expression is supposed to represent.

> Page 7  Line 24 The meaning of the superscripts is not clear in this expression.
> What is the source of the -1T?  Is that the transpose of the inverse?
> Do the -1 superscripts indicate an inverse?

> Page 13 Line 4 The implications of the word “indistinctly” in this sentence are not clear.

> Page 13 Line 8 The statement “the test space is where the item parameters naturally lie” seems to imply there is a correct representation of the item parameters.
> But, the correlations between constructs changes with the sample of examinees and the coordinates can easily be rotated and translated giving different representations of the item parameters.
> I would like to see support for the statement.

> Page 19 Line 8 There is an interesting distinction that seems to be lost here.
> Attributes can be measured independently, such as height and weight and yield correlated results.
> Under those circumstances, should axes be represented as orthogonal, or should the axes represent the correlation.
> This has an implication for the formula that is used to compute the distance between points in the space.

> Page 22 Figure 1 It seems that both of these plots are shown with orthogonal coordinate axes.
> But the orthogonal axes have changed so that the nonorthogonal axes can be shown in the orthogonal representation.

> Page 26 Line 8 This discussion seems to miss the distinction between correlations between traits and dimensions  of sensitivity of the test items.

> Page 29 Line 7 I am curious if the values of discrimination are comparable from the two representations.
> It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.

> Page 30 Line 8 It would be helpful to show the orthogonal and nonorthogonal representations of items and then discuss the interpretation of the parameter estimates based on each of the representations.
> I have been unable to determine the value of using the nonorthogonal representation.
> The nonorthogonal representation is more complex, and it seems sample dependent because it uses the observed correlations/covariances.
> What is the value added of using the more complex representation?

### 2025-01-10 output/seminar-slideshow.qmd
Update items color palette to plot up to 5 items

--- 2024-12-20 output/seminar-slideshow.qmd
@@
 $$
-  cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
+  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
 $$
@@
   PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"]
+  PALETTE_APPLE["normal"],
+  PALETTE_TANGERINE["normal"],
+  PALETTE_UNED["normal"]
 ) |>
   unname() |>
-  setNames(paste(1:3, sep = '_'))
+  setNames(1:5)
 
@@
 Este tercer ítem se encuentra aquí, justo en la punta del segundo.
+:::
+
+## Propiedades: $\cos \, \mathbf{\alpha}$ {.smaller}
+
+:::::::: columns
+:::::: column
+$$
+  D_i = \frac{-d_i}{MDISC_i}
+$$
+
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_1     &= 2 \\
+      MDISC_1 &= \sqrt{2}
+    \end{align} &
+    D_1 \approx -1,41
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#90214a;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_2     &= 1 \\
+      MDISC_2 &= \sqrt{2}
+    \end{align} &
+    D_2 \approx -0,71
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#749f4c;"}
+$$
+  \begin{matrix}
+    \begin{align}
+      d_3     &= -1 \\
+      MDISC_3 &= \sqrt{2}
+    \end{align} &
+    D_3 \approx 0,71
+  \end{matrix}
+$$
+:::
+::::::
+
+::: column
+```{r d-comparison-3}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+mdisc_diff_items <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    1,    2,
+  2,     1,    1,    1,
+  3,     1,    1,    -1,
+  4,     2,    2,    -1,
+)
+
+mdisc_diff_params <- mdisc_diff_items |>
+  compute_mirt_params(d, starts_with('a'))
+
+mdisc_diff_coords <- mdisc_diff_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+mdisc_diff_coords |>
+  filter(item %in% 1:3) |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1.1, 2.4),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1.1/3.5),
+    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+::::::::
+
+::: notes
 :::
@@
 $$
-  cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
+  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
 $$

### 2025-01-10 output/seminar-slideshow.qmd
Add direction cosine vector properties slides

@@
 Si la superficie es "totalmente plana" el dardo entra por un punto y sale por otro que está justo debajo, en las mismas coordenadas de rasgo latente.
+<!-- TODO: Añadir los vectores perpendiculares en las gráficas? -->.
 La "proyección" (si ponemos un foco justo encima, la sombra que forma en el plano) es un punto.
@@
 
-mdisc_diff_coords |>
+# TODO: Cambiar grosor de los ítems
+mdisc_diff_plot <- mdisc_diff_coords |>
   arrange(item |> desc()) |>
@@
   )
+
+mdisc_diff_plot
 ```
@@
 ::: {.r-stack style="color:#749f4c;"}
+<!-- TODO: Align D_4 value (add space with equal width as minus signs) -->
+
 $$
@@
 ::: {.r-stack style="color:#749f4c;"}
+<!-- TODO: Align D_4 value with prevous slide -->
+
 $$
@@
 
+# TODO: Cambiar grosor de los ítems
 mdisc_diff_coords |>
@@
 
-:::::::: columns
-:::::: column
-$$
-  D_i = \frac{-d_i}{MDISC_i}
+::::::: columns
+::::: column
+$$
+  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
 $$
@@
   \begin{matrix}
-    \begin{align}
-      d_1     &= 2 \\
-      MDISC_1 &= \sqrt{2}
-    \end{align} &
-    D_1 \approx -1,41
+    \mathbf{a}_1 \approx \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix} &
+      \mathbf{\alpha}_1 =  \begin{bmatrix}
+        45º \\
+        45º
+      \end{bmatrix}
   \end{matrix}
@@
   \begin{matrix}
-    \begin{align}
-      d_2     &= 1 \\
-      MDISC_2 &= \sqrt{2}
-    \end{align} &
-    D_2 \approx -0,71
+    \mathbf{a}_2 \approx \begin{bmatrix}
+        2 \\
+        2
+      \end{bmatrix} &
+      \mathbf{\alpha}_2 =  \begin{bmatrix}
+        45º \\
+        45º
+      \end{bmatrix}
   \end{matrix}
@@
 :::
-
-::: {.r-stack style="color:#749f4c;"}
-$$
-  \begin{matrix}
-    \begin{align}
-      d_3     &= -1 \\
-      MDISC_3 &= \sqrt{2}
-    \end{align} &
-    D_3 \approx 0,71
-  \end{matrix}
-$$
-:::
-::::::
+:::::
 
 ::: column
-```{r d-comparison-3}
+```{r dir-comparison-1}
 #| fig-align:  center
@@
 
-mdisc_diff_items <- tribble(
+mdisc_diff_plot
+```
+:::
+:::::::
+
+::: notes
+Por último, en cuanto a la dirección del vector: Vemos que, mientras que los componentes del parámetro de discriminación se mantengan proporcionales, la dirección no va a cambiar, a pesar de que cambie la discriminación multidimensional.
+:::
+
+## Propiedades: $\cos \, \mathbf{\alpha}$ {.smaller}
+
+::::::: columns
+::::: column
+$$
+  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
+$$
+
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 \approx \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix} &
+      \mathbf{\alpha}_1 =  \begin{bmatrix}
+        45º \\
+        45º
+      \end{bmatrix}
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#749f4c;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_3 \approx \begin{bmatrix}
+        1,329 \\
+        0,484
+      \end{bmatrix} &
+      \mathbf{\alpha}_3 =  \begin{bmatrix}
+        20º \\
+        70º
+      \end{bmatrix}
+  \end{matrix}
+$$
+:::
+:::::
+
+::: column
+```{r dir-comparison-2}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+cos_diff_items <- tribble(
   ~item, ~a_1,                       ~a_2,                       ~d,
-  1,     1,    1,    2,
-  2,     1,    1,    1,
-  3,     1,    1,    -1,
-  4,     2,    2,    -1,
-)
-
-mdisc_diff_params <- mdisc_diff_items |>
+  1,     1,                          1,                          0,
+  2,     2,                          2,                          0,
+  3,     cos((20/180*pi)) * sqrt(2), cos((70/180*pi)) * sqrt(2), 0,
+  4,     0,                          2,                          0,
+  5,     2,                          0,                          0
+)
+
+cos_diff_params <- cos_diff_items |>
   compute_mirt_params(d, starts_with('a'))
 
-mdisc_diff_coords <- mdisc_diff_params |>
+cos_diff_coords <- cos_diff_params |>
   compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
@@
 
-mdisc_diff_coords |>
-  filter(item %in% 1:3) |>
+cos_diff_coords |>
+  filter(item %in% c(1, 3)) |>
   ggplot(
@@
   scale_x_continuous(
-    limits       = c(-1.1, 2.4),
+    limits       = c(-0.2, 2.2),
     breaks       = -1:2,
@@
   scale_y_continuous(
-    limits       = c(-1.1, 2.4),
+    limits       = c(-0.2, 2.2),
     breaks       = -1:2,
@@
     axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1.1/3.5),
-    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
+    axis.title.x       = element_markdown(hjust = 1/12),
+    axis.title.y.right = element_markdown(vjust = 1/12),
     axis.ticks         = element_blank(),
@@
 :::
-::::::::
-
-::: notes
+:::::::
+
+::: notes
+Ahora bien, si cambia la proporción entre ellos, la dirección sÍ va a cambiar; y eso es independiente de que la disciminación multidimensional se mantenga constante, como vemos en este ejemplo: misma longitud del vector, distinta dirección.
+:::
+
+## Propiedades: $\cos \, \mathbf{\alpha}$ {.smaller}
+
+::::::: columns
+::::: column
+$$
+  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
+$$
+
+::: {.r-stack style="color:#d76f47;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_4 \approx \begin{bmatrix}
+        0 \\
+        2
+      \end{bmatrix} &
+      \mathbf{\alpha}_4 =  \begin{bmatrix}
+        90º \\
+         0º
+      \end{bmatrix}
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#00533e;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_5 \approx \begin{bmatrix}
+        2 \\
+        0
+      \end{bmatrix} &
+      \mathbf{\alpha}_5 =  \begin{bmatrix}
+         0º \\
+        90º
+      \end{bmatrix}
+  \end{matrix}
+$$
+:::
+:::::
+
+::: column
+```{r dir-comparison-3}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+cos_diff_coords |>
+  filter(item %in% 4:5) |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.2, 2.2),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.2, 2.2),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1/12),
+    axis.title.y.right = element_markdown(vjust = 1/12),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::::
+
+::: notes
+Por último, vemos que, cuando uno de los valores de la discriminación es cero, el vector es perpendicular al eje correspondiente (además de ser paralelo al otro eje, pero esto es una simplificación por tener dos dimensones solamente).
+Esto es muy importante, ya que nos viene a indicar que, cuando un ítem no mide en absoluto alguna de las dimensiones del instrumento que estamos considerando, entonces no proporciona absolutamente ninguna información del nivel de rasgo en esa dimensión de las personas que están respondiendo al instrumento.
 :::

### 2025-01-10 output/seminar-slideshow.qmd
Tweak style in direction cosine properties slide

@@
   PALETTE_TANGERINE["normal"],
-  PALETTE_UNED["normal"]
+  PALETTE_UNED["medium"]
 ) |>
@@
 
+<br>
+
 ::: {.r-stack style="color:#5c6eb1;"}
@@
 $$
+
+<br>
 
@@
 
+<br>
+
 ::: {.r-stack style="color:#d76f47;"}
@@
 
-::: {.r-stack style="color:#00533e;"}
+::: {.r-stack style="color:#427562;"}
 $$

### 2025-01-10 output/seminar-slideshow.qmd
Show better equivalence between MIRT and FA
By decomposing the parameters in the FA formula, it is better shown how they translate to the MIRT formulation
Improve direction cosine properties speaker notes

@@
       \frac{\mathbf{\lambda}_i^T \mathbf{F} + \tau_i}{\sqrt{var(U_i)}}
-    \right) \\
+    \right) =
+      \left(
+        \frac{1}{\sqrt{var(U_i)}} \mathbf{\lambda}_i^T \mathbf{F} +
+          \frac{1}{\sqrt{var(U_i)}} \tau_i
+      \right)\\
     P(Y_i = 1) &= \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
@@
 ::: notes
-Por último, en cuanto a la dirección del vector: Vemos que, mientras que los componentes del parámetro de discriminación se mantengan proporcionales, la dirección no va a cambiar, a pesar de que cambie la discriminación multidimensional.
+Por último, en cuanto a la dirección del vector: Vemos que, mientras que los componentes del parámetro de discriminación se mantengan proporcionales, la dirección no va a cambiar, a pesar de que la discriminación multidimensional sí cambia.
 :::
@@
 ::: notes
-Ahora bien, si cambia la proporción entre ellos, la dirección sÍ va a cambiar; y eso es independiente de que la disciminación multidimensional se mantenga constante, como vemos en este ejemplo: misma longitud del vector, distinta dirección.
+Ahora bien, si cambia la proporción entre ellos, la dirección sí va a cambiar; y eso va a ser independiente de que la disciminación multidimensional se mantenga constante, como vemos en este ejemplo: misma longitud del vector, distinta dirección.
+Es decir, que por ejemplo un ítem puede medir "igual de bien" dos constructos, y otro, que mida más bien uno de ellos, estaría más alineado con esa dirección, a pesar de tener la misma discriminación multidimensional, de ser igual de discriminante "en términos globales", diríamos.
 :::
@@
 ::: notes
-Por último, vemos que, cuando uno de los valores de la discriminación es cero, el vector es perpendicular al eje correspondiente (además de ser paralelo al otro eje, pero esto es una simplificación por tener dos dimensones solamente).
+Y por último, vemos que, cuando uno de los valores de la discriminación es cero, el vector es perpendicular al eje correspondiente (además de ser paralelo al otro eje, pero esto es una simplificación por tener dos dimensones solamente).
 Esto es muy importante, ya que nos viene a indicar que, cuando un ítem no mide en absoluto alguna de las dimensiones del instrumento que estamos considerando, entonces no proporciona absolutamente ninguna información del nivel de rasgo en esa dimensión de las personas que están respondiendo al instrumento.

### 2025-01-17 output/seminar-slideshow.qmd
Add TODO markers for some pending improvements

@@
 
+<!-- TODO: Como ecuaciones (sin punto y coma?) -->
+
 ::: r-stack
@@
 mdisc_diff_coords |>
+  # TODO: Item 3 on top
   filter(item %in% 1:3) |>
@@
 <br>
+
+<!-- TODO: Same color as item? -->
 

### 2025-01-17 output/seminar-slideshow.qmd
Update "inverse items" introductory slide

@@
 Es decir, la respuesta a cada ítem se modela como una regresión logistica donde los predictores son las variables latentes.
+Esto es lo que se llama "Función Característica del Ítem".
 :::
@@
 
-::: {.fragment .r-center}
+::::: columns
+::: {.column .fragment .r-center}
 ```{r inverse-item-curve}
 #| fig-align:  center
-#| fig-asp:   0.74
-#| fig-width: 6.5
-
-ITEMS_PALETTE <- PALETTE_BLUE["normal"] |>
-  unname() |>
-  setNames("prob_1")
+#| fig-asp:    0.5344
+#| fig-width:  9
 
@@
   ~item, ~a,   ~b,
-   1,    -2,   0
+   1,    -1.5, 0
 )
@@
   ) |>
-  rename_with(str_remove, pattern = '\\.{3}')
+  rename_with(str_remove, pattern = 'prob_\\.{3}')
 
 logit_curve_data_long <- logit_curve_data |>
-  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
+  pivot_longer(matches('\\d'), names_to = "item", values_to = "prob")
 
@@
 
+::: {.column .fragment .r-center}
+<br> <br>
+
+$$
+  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
+$$
+:::
+:::::
+
 ::: notes
 Pero además, si hablamos de ítems "no cognitivos" (p.ej. personalidad) en los que la persona evaluada puede estar "de acuerdo" o "en desacuerdo", sí podemos tener ítems con probabilidad "decreciente".
+
+\[NEXT\]
+
+Formalmente, esto vendría no sería más que un parámetro de discriminación negativo.
+
+\[NEXT\]
+
+Si recordáis la "función característica del ítem", esto consiste simplemente en un valor negativo en el parámetro $a$; eso hace que la probabilidad sea decreciente.
 
@@
 Vamos a imaginar que alguien estuviera respondiendo con un sesgo de "deseabilidad social" (intenta dar una buena imagen de sí mismo o de sí misma, de manera consciente o no).
-A un ítem como este, tendería a responder que no, al margen de cómo autoevalúe su verdadero nivel de responsabilidad.
-Por lo tanto, tendríamos un ítems con dos valores negativos de discriminación, uno en la dimensión de responsabilidad, y otro en la dimensión de deseabilidad social.
-Sin embargo, consideremos ahora un sesgo de "aquiescencia", que hace que la gente tienda a estar de acuerdo con todo, o a decir que sí a todo sin un criterio muy reflexivo ("¿Descuidas tus obligaciones?
+A un ítem como este, tendería a responder que no, dado ese sesgo de deseabilidad social, al margen de cómo autoevalúe su verdadero nivel de responsabilidad.
+Por lo tanto, tendríamos un ítem con dos valores negativos de discriminación, uno en la dimensión de responsabilidad, y otro en la dimensión de deseabilidad social.
+Sin embargo, consideremos ahora que estamos modelando un sesgo de "aquiescencia", que hace que la gente tienda a estar de acuerdo con todo, o a decir que sí a todo sin un criterio muy reflexivo ("¿Descuidas tus obligaciones?
 Mmmmm...

### 2025-01-17 output/seminar-slideshow.qmd
Add inverse items examples

@@
 
-## Generalización: Ítems inversos
-
-<!-- # TODO: Representación de ítems con a (una o varias) negativa (tabla y gráfico) -->
-
-$$
-  \begin{align}
+## Generalización: Ítems inversos {.smaller}
+
+::::::::: columns
+::::::: column
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix} &
+    d_1 = 0
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#90214a;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+         1 \\
+        -1
+      \end{bmatrix} &
+    d_1 = 0
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#749f4c;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        -1 \\
+        -1
+      \end{bmatrix} &
+    d_1 = 0
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#d76f47;"}
+$$
+  \begin{matrix}
     \mathbf{a}_1 &=
@@
          1
-      \end{bmatrix} \\
-    d_1 &= 0
-  \end{align}
-$$
-
-<br>
-
+      \end{bmatrix} &
+    d_1 = 0
+  \end{matrix}
+$$
+:::
+:::::::
+
+::: column
 ```{r inverse-item-representation}
 #| fig-align:  center
-#| fig-height: 5
-#| fig-width:  5
+#| fig-height: 8
+#| fig-width:  8
 
@@
   ~item, ~a_1, ~a_2, ~d,
-  1,     -1,    1,   0
+  1,      1,    1,   0,
+  2,      1,   -1,   0,
+  3,     -1,   -1,   0,
+  4,     -1,    1,   0
 )
@@
   scale_x_continuous(
-    limits       = c(-1.5, 1),
+    limits       = c(-1.5, 1.5),
     breaks       = -1:2,
@@
   scale_y_continuous(
-    limits       = c(-1, 1.5),
+    limits       = c(-1.5, 1.5),
     breaks       = -1:2,
@@
   ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
   coord_fixed(clip = "on") +
@@
     axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
+    axis.title.x       = element_markdown(hjust = .5),
+    axis.title.y.right = element_markdown(vjust = .5),
     axis.ticks         = element_blank(),
@@
 ```
-
-::: notes
-La representación de estos ítems sería como se muestra en este gráfico.
-<!-- # TODO: Descripción de los ítems representados -->
-
-La generalización a ítems inversos es directa: La discriminación multidimensional "eleva al cuadrado" los "parámetros de discriminación marginales", luego va a ser siempre positiva.
-El parámetro de posición, por tanto, siempre va a tener el signo inverso al que tenga la intersección.
-Los cosenos tendrán, cada uno, el mismo que el parámetro de discriminación de esa dimensión.
+:::
+:::::::::
+
+::: notes
+Veamos un ejemplo: En el ítem 1 ambos cosenos son positivos, luego ambos ánculos son menores de 90º.
+El ítem está en el primer cuadrante.
+
+El segundo ítem va a tener un coseno positivo (ángulo menor que 90º) y uno negativo, por lo que será mayor que 90º; por tanto, apunta en el sentido negativo de ese eje.
+El ángulo es 180º menos el ángulo que tendría ese mismo coseno si fuera positivo.
+
+En el tercer ítem pasa lo mismo, pero con los dos cosenos negativos (y por tanto, ángulos mayores que 90º), y en el cuarto como el segundo pero al revés.
+
+Es decir, las propiedades algebraicas de la discriminación determinan las propiedades geométricas; y vemos cómo se puede visualizar gráficamente con facilidad.
+:::
+
+## Generalización: Ítems inversos {.smaller}
+
+::::::::: columns
+::::::: column
+::: {.r-stack style="color:#5c6eb1;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix} &
+    d_1 = 0.25
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#90214a;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+         1 \\
+        -1
+      \end{bmatrix} &
+    d_1 = -0.5
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#749f4c;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        -1 \\
+        -1
+      \end{bmatrix} &
+    d_1 = -0.5
+  \end{matrix}
+$$
+:::
+
+::: {.r-stack style="color:#d76f47;"}
+$$
+  \begin{matrix}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        -1 \\
+         1
+      \end{bmatrix} &
+    d_1 = 0
+  \end{matrix}
+$$
+:::
+:::::::
+
+::: column
+```{r inverse-item-representation-distance}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,      1,    1,    0.25,
+  2,      1,   -1,   -0.5,
+  3,     -1,   -1,   -0.5,
+  4,     -1,    1,    0
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+  rename_with(str_remove, pattern = "transf_")
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1.5, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1.5, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .5),
+    axis.title.y.right = element_markdown(vjust = .5),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::::::
+
+::: notes
+Si la intersección es distinta de cero, como decíamos antes: El desplazamiento del ítem es opuesto al sentido en el que apunta el ítem.
+
+Si la intersección es positiva, como en el ítem 1, se desplaza "hacia atrás" respecto de la cabeza de la flecha, o sea, el ítem es "en general más fácil o atractivo": La región donde es fácil acertar o estar de acuerdo con el ítem se amplía.
+
+Si la intersección es negativa, como en los ítems 2 y 3, se desplaza "hacia adelante" en el sentido del ítem.
+El ítem es "más difícil en general, o 'repulsivo'", o la región donde es fácil acertar o estar de acuerdo disminuye.
 :::

### 2025-01-17 output/seminar-slideshow.qmd
Completes conclusive slide about "inverse items"

@@
 El ítem es "más difícil en general, o 'repulsivo'", o la región donde es fácil acertar o estar de acuerdo disminuye.
-:::
-
-## Ítems "inversos" en la literatura
-
-> the negative relation of Item 15 with geometry achievement is particularly puzzling
-
----@mcdonald_basis_2000 [p. 109]
-
-> This result occurs because $a_1$ for the item is negative, and thus as an examinee's $\theta_1$ ability increases, the chance of answering the item correctly actually decreases.
-
----@ackerman_multidimensional_2005-1 [p. 17]
-
-### Generalización: "Monotonía constante"
-
-::: notes
+
 Esta generalización es bastante evidente, y se puede aplicar sin problema a ítems no cognitivos.
 Sin embargo, hay que tener en cuenta que Reckase explicitaba como condición que los ítems tenían que ser "monotonamente crecientes".
+:::
+
+## Ítems "inversos" en la literatura {.smaller}
+
+> the negative relation of Item 15 with geometry achievement is particularly puzzling
+
+---@mcdonald_basis_2000 [p. 109]
+
+<br>
+
+::: fragment
+> This result occurs because $a_1$ for the item is negative, and thus as an examinee's $\theta_1$ ability increases, the chance of answering the item correctly actually decreases.
+
+---@ackerman_multidimensional_2005-1 [p. 17]
+:::
+
+<br>
+
+::: {.fragment .r-center}
+### Generalización: "Monotonía constante"
+:::
+
+::: notes
 Pero en la literatura nos encontramos casos como estos, que no lo son, y a los autores no se les ocurre decir que "hay que generalizar esas definiciones".
-Más bien, hacen comentarios como estos, restándole importancia al hecho de que los parámetros de discriminación sean negativos ("particularmente misterioso"), o afirmando lo obvio (el parámetro de discriminación del ítem es negativo para una dimensión, por lo que la probabilidad de respuesta correcta disminuye al incrementarse su habilidad)
+Más bien, hacen comentarios como estos, restándole importancia al hecho de que los parámetros de discriminación sean negativos ("particularmente misterioso"),...
+
+\[NEXT\]
+
+o afirmando lo obvio (el parámetro de discriminación del ítem es negativo para una dimensión, por lo que la probabilidad de respuesta correcta disminuye al incrementarse su habilidad)
+
+\[NEXT\]
 
@@
 
-### Generalización: "Coordenadas oblicuas"
+## Generalización: "Coordenadas oblicuas"
 

### 2025-01-17 output/seminar-slideshow.qmd
Improve signed distance example with inverse items

@@
       \end{bmatrix} &
-    d_1 = 0.25
+    d_1 = 0
   \end{matrix}
@@
       \end{bmatrix} &
-    d_1 = 0
+    d_1 = 0.25
   \end{matrix}
@@
   ~item, ~a_1, ~a_2, ~d,
-  1,      1,    1,    0.25,
+  1,      1,    1,    0,
   2,      1,   -1,   -0.5,
   3,     -1,   -1,   -0.5,
-  4,     -1,    1,    0
+  4,     -1,    1,    0.25
 )
@@
 
-Si la intersección es positiva, como en el ítem 1, se desplaza "hacia atrás" respecto de la cabeza de la flecha, o sea, el ítem es "en general más fácil o atractivo": La región donde es fácil acertar o estar de acuerdo con el ítem se amplía.
-
-Si la intersección es negativa, como en los ítems 2 y 3, se desplaza "hacia adelante" en el sentido del ítem.
-El ítem es "más difícil en general, o 'repulsivo'", o la región donde es fácil acertar o estar de acuerdo disminuye.
+Si la intersección es negativa, como en los ítems 2 y 3, el vector se desplaza "hacia adelante" en el sentido de la cabeza de la flecha.
+Es decir, el ítem es "más difícil en general, o menos atractivo", y la región donde es relativamente fácil acertar el ítem (o estar de acuerdo con él) disminuye.
+
+Si la intersección es negativa, como en el ítem 1, el vector se desplaza "hacia atrás"; o sea, el ítem es "en general más fácil o atractivo": La región donde es fácil acertar o estar de acuerdo con el ítem se amplía.
+
+Y da igual en qué dirección apunten los vectores; esto se cumple en todos los casos.
 

### 2025-01-20 output/Letter_to_editor_reviewers.Rmd
Set apart comments for responding independently
Version responses to the editor (WiP)

@@
 
-Response to the action editor:
+# Responses to the action editor:
 
@@
 > Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.
+
+We agree with the editor that different groups of examinees can result in different covariance structures, due to the interaction between the examinees and the items.
+However, in our understanding, the same could be predicated from the item parameters themselves, without taking into account the covariance structure.
+For example, in unidimensional IRT, the item parameters may be different for different groups of examinees, even if the variance is constrained to be equal across groups; this is indeed the basis of methods such as multigroup IRT.
+It is our stance that the covariance structure, understood as a property of the data of course, should be taken into account when determining the multidimensional properties of the items.
+That is what we do when we posit that the covariance structure is taken into account when computing the multidimensional parameters.
+The method we propose thus so, by considering an algebraic basis that conveys information of the covariance structure.
+<!-- TODO: Changes to the main text? -->
 
@@
 
+Indeed, we propose that the basis used to represent the test space changes its metric.
+We also understand that the metrics of the two representation are not comparable, and thus one should not compare items represented with the agnostic and the covariance-base version of the multidimensional parameters.
+However, when making multi-group comparisons, we think that taking into account the covariance-based structure of the latent space may provide more accurate information.
+
+There are several references in the literature that have found contradictory results that, we believe, are due to disregarding the latent covariance structure (i.e., using Reckase & McKinley's formulae in a non-orthonormal space).
+<!-- TODO: Examples? -->
+While we consider this is out of the scope of the current paper, we will gladly add furthere detail to explain why these results are contradictory and how our method may solve the problem.
+<!-- TODO: Add examples of contradictory results? -->
+
 > I also wonder about how one compares the standard errors of the item parameters when the covariance covariance structures are different.  Are they directly comparable?
+
+The standard errors of the multidimensional parameters are, without a doubt, more difficult to compute for the covariance-based case than for the agnostic one (if we assume the covariance structure is estimated from the model).
+Computing the covariance errors in this case requires taking into account the standard errors of the item parameters themselves and at least the standard errors of the correlations (in the best-case scenario, assuming standard latent dimensions).
+However, to our knowledge, multidimensional item parameters, as defined by @reckase_difficulty_1985 and @reckase_discriminating_1991 have been previously defined and used only for a theoretical perspective, without ever taking into account the uncertainty resulting from the model estimation.
+
+Regarding the "original" item parameters in the classical M2PL metric, we understand that the standard errors of the item parameters are independent of the definition of the multidimensional parameters used.
+Software such as R package {mirt} <!-- TODO: Add reference --> use a method similar to exploratory factor analysis to compute the item parameter estimtates and the standard errors, based on an oblique rotation to approximate simple structure.
+As far as we know, these standard error estimates are computed using the usual methods, involving Taylor series, bootstraping, and similar.
+<!-- TODO: Check methods; complete? -->
 
@@
 
-Response to Reviwer 1:
+Our manuscript does not propose an estimation or modeling method.
+On the contrary, what we propose is a way to compute and represent the multidimensional item parameters without disregarding the covariance structure of the latent space.
+Implementing the formulae we propose, using the standard output of most MIRT software, should be straightforward for most IRT researchers.
+
+We have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, examples, and even the submitted version of the manuscript.
+Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software (e.g., {mirt} <!-- TODO: Add reference -->, FlexMIRT <!-- TODO: Add reference -->).
+As the journal guidelines require a specific format for the Supplementary Material,
+<!-- TODO: Check this -->
+we have preferred not to include the repository as additional material.
+However, we can do so if the journal accepts linking an external repository as supplenetary material, or otherwise grant access to the editor and reviewers to this repository.
+<!-- TODO: Publish repo and link Software Legacy DOI? -->
+
+# Responses to Reviwer 1:
 
@@
 
-Response to Reviwer 2:
+# Responses to Reviwer 2:
 
@@
 
-Response to Reviwer 3:
+# Responses to Reviwer 3:
 
@@
 > It would help if a page number were provided for the discussion.
+
 > I also wonder about the Harman (1970) reference.

### 2025-01-23 output/Letter_to_editor_reviewers.Rmd
Complete first draft of response to editor

@@
 
+We think the motivation and psychometric utility of our approach are more
+related with the interpretation of the item parameters than with the latent
+abilities.
+To our understanding, the better interpretation of the item parameters that
+results from our approach provides indeed a better insight into how the results
+are used.
+<!-- TODO: Complete -->
+
 > What implications does your approach have for standard psychometric procedures like equating.
+
+<!-- TODO: Reply -->
 

### 2025-01-23 output/Main_text.Rmd
Correct use of curly double quote characters

@@
 "are constrained to be positive" [@ackerman_graphical_1996, p. 315],
-or that “in an [independent-clusters] solution, the loadings can always
-be selected to be positive” [@mcdonald_basis_2000, p. 110].
+or that "in an [independent-clusters] solution, the loadings can always
+be selected to be positive" [@mcdonald_basis_2000, p. 110].
 However, when such instances of negative parameters appear,
-they are often dismissed as “particularly puzzling” (p. 109),
+they are often dismissed as "particularly puzzling" (p. 109),
 or implicitly assuming item malfunctioning without further discussion

### 2025-01-23 output/Letter_to_editor_reviewers.Rmd
Version responses to Reviewer 1

@@
 
-We think the motivation and psychometric utility of our approach are more
-related with the interpretation of the item parameters than with the latent
-abilities.
-To our understanding, the better interpretation of the item parameters that
-results from our approach provides indeed a better insight into how the results
-are used.
+We think the motivation and psychometric utility of our approach are more related with the interpretation of the item parameters than with the latent abilities.
+To our understanding, the better interpretation of the item parameters that results from our approach provides indeed a better insight into how the results are used.
+A multidimensional representation of the items provides benchmarks for comparing different latent trait levels, as discussed by @reckase_difficulty_1985.
+Therefore, the more accurate such representation is, the better the latent trait interpretation will be.
 <!-- TODO: Complete -->
@@
 
+Our introduction is a bit less than four pages long.
+We have provided concise paragraphs about the Multidimensional IRT model and its parameters, their assumptions, and our rationale to relax them.
+Finally, a last paragraph presents the objectives of our research.
+We honestly believe we have made a thorough effort to summarize the motivation of our research, and we find it difficult to shorten it further without losing relevant information to justify our work.
+Therefore, we are sorry to say that we cannot comply with the reviewer's request to shorten the introduction.
+
 > 2. Introduction, do you have any hypothesis for the current study? Please briefly describe it at the end.
+
+Our work is completely theoretical and thus uses a purely deductive method.
+Therefore, we do not have any hypothesis to test, but the results of the formal derivations that we present, adressing the objective crearly described at the end of our introduction.
 
@@
 
+The datasets we use in our paper are examples of item parameter estimates, taken from the literature.
+As they only serve exemplification purposes, and are not the main focus of our article, we deem it more appropriate to describe their details in the corresponding section of the manuscript.
+We have added a short reference to them in the introduction, to make sure their pertinence as examples is clear.
+
 > 4. Within or between multidimensional latent? This should be detailed.
+
+We think the context makes it clear that we are referring to "within-item" multidimensionality.
+In fact, it is clear from the manuscript that the results are trivial in the "between-item" multidimensionality case,
+as the `r MDISC_SYM` parameter reduces to the only non-null component of the $`r DISCR_VECTOR`$ parameter.
+However, we acknowledge this terminology is prevalent in the current MIRT literature;
+therefore, we have now explained this in our "Model formulation" section.
+The end of the second paragraph reads now as follows:
+
+    > Third, as responses are not necessarily interpreted as correct or incorrect,
+    parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
+    instead of the item difficulty.
+    Finally, it is worth noting that
+    this formulation assumes *within-item multidimensionality* in general,
+    which will be present whenever more than one component of $`r DISCR_VECTOR`$
+    is non-null.
+    However, the model can also account for *between-item multidimensionality*,
+    if all items are unidimensional
+    (i.e., their $`r DISCR_VECTOR`$ parameters only have one non-null component).
 
@@
 
+Our last two paragraphs mention the most important limitations of our work.
+We clearly demarcate the limited scope of our conclusions, and warn against overgeneralizing our results to other MIRT models or the common factor model.
+We have put a great effort into writing our limitations in a way that is elegant and, at the same time, avoids cliché sentences that add little value.
+Therefore, we would rather not add a separate section for limitations.
+
 > 6. Conclusion is missing and should be added here.
+
+Similarly to the previous comment, we believe that our conclusions are thoroughly and carefully elaborated in our "Discussion" section, and we would prefer not to add an explicit "Conclusions" header.
 

### 2025-01-23 output/Main_text.Rmd
Version responses to Reviewer 1

@@
 and present two examples of application of our results.
+These examples, from the cognitive and the non-cognitive domain respectively,
+are taken from the MIRT literature:
+One of them is a classic dataset, often used to exemplify
+the use of the `r MODEL_ACRONYM` model [@reckase_multidimensional_2009];
+the other one, an application in marketing research [@tezza_modelo_2018].
 We round up with a discussion about
@@
 the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
-Finally, as responses are not necessarily interpreted as correct or incorrect,
+Third, as responses are not necessarily interpreted as correct or incorrect,
 parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
 instead of the item difficulty.
+Finally, it is worth highlighting that
+this model generally assumes *within-item multidimensionality*,
+which will be present whenever more than one component of $`r DISCR_VECTOR`$
+is non-null.
+However, the model can also account for *between-item multidimensionality*,
+if the $`r DISCR_VECTOR`$ parameter only has one non-null component.
 

### 2025-01-23 output/Letter_to_editor_reviewers.Rmd
Version responses to Reviewer 2 (WiP)

@@
 ---
+
+```{r setup}
+#| include: false
+
+# Includes:
+
+library(here)
+library(knitr)
+library(patchwork)
+library(ggplot2)
+library(rstudioapi)
+library(svglite) # Used by {knitr} and {ggplot2} to render the figure to SVG;
+                 #   added to prevent {renv} from reporting inconsistent state.
+library(officer)
+
+ROOT_DIR <- here()
+DOC_DIR  <- file.path(ROOT_DIR, "output")
+
+opts_knit$set(root.dir = ROOT_DIR)
+opts_chunk$set(
+  echo     = FALSE,
+  message  = FALSE,
+  cache    = FALSE,
+  dev      = "png",
+  dpi      = 1200L,
+  dev.args = list(type = "cairo-png"),
+  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE),
+  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
+)
+
+# Output configuration options:
+options(digits = 3)
+```
+
+```{r write-bib}
+# automatically create a bib database for R packages
+write_bib(
+  c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
+  'www/packages.bib'
+)
+```
+
+```{r sources}
+#| cache:   false
+#| include: false
+source("R/Formulae.R",  encoding = 'UTF-8')
+source("R/Constants.R", encoding = 'UTF-8')
+```
 
@@
 <!-- TODO: Complete -->
-
-Our manuscript presents a generalization of the multidimensional IRT parameters
-introduced by Reckase and Reckase & McKinley in their articles
-["The difficulty of test items that measure more than one ability" (1985)][1]
-and
-["The discriminating power of items that measure more than one dimension"
-(1991)][2],
-respectively.
-
-[1]: https://doi.org/10.1177/014662168500900409
-
-[2]: https://doi.org/10.1177/014662169101500407
 
@@
 
+<!-- TODO: Reply -->
+
 > While the manuscript includes a graphical representation example with two-dimensional latent variables, real-world applications often involve higher dimensions (e.g., the Big Five personality traits or
@@
 
+Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that may account for any number of dimensions.
+we have found our previous version to not be explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:
+
+    > Consider a basis $`r LS_BASIS_EQ`$,
+    with $`r LS_BASIS`$ generally non-orthonormal.
+    Let $`r LATENT_SPACE`$ be a $`r N_DIMS`$-dimensional latent space
+    in basis $`r LS_BASIS`$ (with $`r N_DIMS`$ any positive integer),
+    where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
+    According to the `r MODEL_ACRONYM` model, the probability of
+    a positive response to item $`r ITEM_INDEX`$ is [...]
+
 > 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect.
 > It should be “$\mathcal{C}$”.
+
+Thank you so much to Reviewer 2 for pointing out this typo.
+We have corrected it in the manuscript accordingly.
 
@@
 
+This case was also a typo; we have now corrected it in the manuscript as well.
+
 > Similar issues occur in Eqs. (26) and (27) on page 17.
+
+Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
+To do that, we have made use of the expression $`r DISCR_VECTOR_COORDS_EQ`$, stated in p. 7.
+(That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
+However, we acknowledge that we have skipped a step in our discrourse, thus making harder than necessary to follow along.
+To correct this, we have modified the immediately preceding paragraph to the following:
+
+    > Of course,
+    from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
+    it is easy to express the coordinates in terms of the parameters in the original
+    model formulation.
+    Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as
 

### 2025-01-23 output/Main_text.Rmd
Version responses to Reviewer 2 (WiP)

@@
 
-Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
-in basis $`r LS_BASIS_EQ`$, being $`r LS_BASIS`$ not necessarily orthonormal,
+Consider a basis $`r LS_BASIS_EQ`$,
+with $`r LS_BASIS`$ generally non-orthonormal.
+Let $`r LATENT_SPACE`$ be a $`r N_DIMS`$-dimensional latent space
+in basis $`r LS_BASIS`$ (with $`r N_DIMS`$ any positive integer),
 where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.

### 2025-01-23 R/Formulae.R

@@ -567,7 +567,11 @@ TRAIT_TRANSFORM_ALT    <- latex(TRANSFORM_MATRIX_ALT, TRAIT_VECTOR_ALT)
 TRAIT_TRANSFORM_ALT_EQ <- latex_eq(TRAIT_ORTH_COORDS, TRAIT_TRANSFORM_ALT)
 
 # Inner product in new basis:
-
+INNER_PROD_TRAIT_ALT_BASIS       <- latex_innerprod(
+  TRAIT_VECTOR_J,
+  TRAIT_VECTOR_K,
+  basis = ALT_BASIS
+)
 TRAIT_VECTOR_J_COORDS_ALT    <- latex_raised_to(TRAIT_VECTOR_J, exp = ALT_BASIS)
 TRAIT_VECTOR_J_COORDS_ALT_TRANSP <- latex_transp(TRAIT_VECTOR_J_COORDS_ALT)
 TRANSFORM_MATRIX_ALT_TRANSP      <- latex_transp(TRANSFORM_MATRIX_ALT)
@@ -581,7 +585,10 @@ INNER_PROD_ALT_DEF               <- latex(
   INNER_PROD_MATRIX_ALT_DEF,
   TRAIT_VECTOR_K_COORDS_ALT
 )
-INNER_PROD_TRAIT_ALT_EQ        <- latex_eq(INNER_PROD_TRAIT, INNER_PROD_ALT_DEF)
+INNER_PROD_TRAIT_ALT_EQ        <- latex_eq(
+  INNER_PROD_TRAIT_ALT_BASIS,
+  INNER_PROD_ALT_DEF
+)
 
 # Inner product matrix in new basis:
 INNER_PROD_MATRIX_ALT    <- latex_bf('K')
@@ -753,7 +760,7 @@ DIR_COS_VEC_STD_EQ       <- latex_eq(
 DIR_ANGLE_ITEM_VEC_EQ   <- latex_eq(DIR_ANGLE_VEC_ORG, DIR_ANGLE_ITEM_VEC)
 DISCR_VECTOR_MODULE_DEF <- latex_sqrt(DISCR_VECTOR_INNER_PROD)
 DIR_COS_ITEM_VEC_DEF    <- latex_frac(
-  "$DIAG_MATRIX_INNER_PROD_SR_INV$ $DISCR_VECTOR$",
[…]

### 2025-01-23 output/Main_text.Rmd
Add progress on responses to Reviewer 2 (WiP)

@@
 it is easy to express the coordinates in terms of the parameters in the original
-model formulation, namely,
+model formulation.
+Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as
 

### 2025-01-23 R/Formulae.R

@@ -760,7 +760,7 @@ DIR_COS_VEC_STD_EQ       <- latex_eq(
 DIR_ANGLE_ITEM_VEC_EQ   <- latex_eq(DIR_ANGLE_VEC_ORG, DIR_ANGLE_ITEM_VEC)
 DISCR_VECTOR_MODULE_DEF <- latex_sqrt(DISCR_VECTOR_INNER_PROD)
 DIR_COS_ITEM_VEC_DEF    <- latex_frac(
-  latex("$DIAG_MATRIX_INNER_PROD_SR_INV$ $DISCR_VECTOR$"),
+  latex("$DIAG_MATRIX_INNER_PROD_SR_INV$ $DISCR_COORDS$"),
   DISCR_VECTOR_MODULE_DEF
 )
 DIR_COS_ITEM_VEC_EQ     <- latex_eq(DIR_COS_ITEM_VEC, DIR_COS_ITEM_VEC_DEF)

### 2025-01-24 output/Letter_to_editor_reviewers.Rmd
Correct redundancy in response to editor

--- 2025-01-23 output/Letter_to_editor_reviewers.Rmd
@@
 
-We think the motivation and psychometric utility of our approach are more
-related with the interpretation of the item parameters than with the latent
-abilities.
-To our understanding, the better interpretation of the item parameters that
-results from our approach provides indeed a better insight into how the results
-are used.
-<!-- TODO: Complete -->
-
 > What implications does your approach have for standard psychometric procedures like equating.
-
-<!-- TODO: Reply -->
 
@@
 
-We have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, examples, and even the submitted version of the manuscript.
+We have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, and even the submitted versions of the manuscript.
 Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software (e.g., {mirt} <!-- TODO: Add reference -->, FlexMIRT <!-- TODO: Add reference -->).

### 2025-01-24 output/seminar-slideshow.qmd
Improve example of "inverse items" in literature
Add item plot

@@
 
-> the negative relation of Item 15 with geometry achievement is particularly puzzling
+:::::: columns
+:::: {.column width="62%"}
+> \[...\] the negative relation of Item 15 with geometry achievement is particularly puzzling.
 
 ---@mcdonald_basis_2000 [p. 109]
-
-<br>
 
@@
 :::
-
-<br>
+::::
+
+::: {.column width="38%"}
+![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="left"}
+:::
+::::::
 
@@
 ### Generalización: "Monotonía constante"
+:::
+
+::: aside
+Figura tomada de @ackerman_multidimensional_2005-1
 :::

### 2025-01-24 output/seminar-slideshow.qmd
Expand "orthogonality assumption" intro notes
These notes help ease the transition from the "increasing monotony assumption" section to the "orthogonality assumption" section.

@@
 Ítems que estarían excluidos, por no tener monotonía constante, podrían ser formulaciones como, por ejemplo, "Soy tan extrovertido como la mayoría de mis amigos" (alguien puede decir, "No, soy más introvertido que todos ellos", respuesta negativa con rasgo bajo, o "Qué va, soy el más extrovertido con diferencia", respuesta negativa con rasgo alto).
+
+Bien, esto por lo que respecta a los ítems que denominamos "inversos" (o que tienen parámetros de discrimiinación / pesos factoriales inversos).
+Como veis, esta generalización es necesaria cuando hablamos de medición en el dominio "no cognitivos"; en el cognitivo, ya vimos que no tenía sentido, y de ahí que hiciera falta esta matización.
 :::
@@
 
-> before performing the differentiation, the constraint that <!-- # TODO: Sum of cosines^2 = 1 --> is added to the expression for the slope \[...\]
+> \[...\] before performing the differentiation, the constraint that $\cos^2 \, \mathbf{\alpha} = 1$ is added to the expression for the slope \[...\]
 
@@
 
-::: notes
-Otro de los supuestos que hace Reckase originalmente para calcular los parámetros es este: Los cosenos al cuadrado suman 1.
-Es decir, está aplicando el teorema de Pitágoras, pero ello implica de manera implícita que los ejes del espacio latente son "ortogonales" (forman ángulos de 90 grados).
-
-Esto en TRI multidimensional originalmente tenía sentido, ya que no se tenía muy en cuenta la interpretación de los parámetros de rasgo latente.
-O bien no importaba mucho que los ejes se pudieran interpretar como rasgos latentes, o se interpretaban de manera no necesariamente muy precisa.
+::::: columns
+::: column
+:::
+
+::: column
+:::
+:::::
+
+::: notes
+Ahora, la TRI multidimensional también dijimos que se interesaba sobre todo por medir (y ordenar) a las personas, no tanto por interpretar sustantivamente esas dimensiones de rasgo latente.
+De ahí que estimar un modelo con suficientes dimensiones para explicar la variabilidad entre los sujetos fuese importante, pero no era tan importante cuáles fueran esas dimensiones.
+Por eso, hay un supuesto, implícito muchas veces en TRI, de que las dimensiones no están correlacionadas: Se estima el modelo con N dimensiones, las que haga falta, añadiendo restricciones a los ítems para identificarlas, y eso da lugar a N dimensiones independientes entre sí (con correlaciones de 0), representadas en N ejes ortogonales (ángulos rectos todos ellos).
+Que esos ejes representen competencia o habilidad en esto o lo otro, en principio, nos da igual.
+Con asegurarnos de que hay suficientes dimensiones para "capturar" toda la variabilidad debida a esos rasgos nos basta.
+
+Pues bien, ese es el otro supuesto que hace Reckase para calcular los parámetros multidimensionales: Los cosenos directores del ítem al cuadrado suman 1.
+Es decir, está aplicando el teorema de Pitágoras.
+Esta expresión implica, de manera implícita, que todos los ejes del espacio latente son "ortogonales" entre sí.
+Como digo, en TRI originalmente no importaba mucho que los ejes se pudieran interpretar como rasgos latentes.
+
+\[NEXT\]
 
 No obstante, en 1999 llega McDonald, y dice que la TRI y el AF son fundamentalmente la misma cosa.
-Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas que dan lugar a ejes no ortogonales.
+Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas, que dan lugar a ejes no ortogonales.
+
+\[NEXT\]
 

### 2025-01-24 output/sandbox_slideshow.qmd
Version "oblique rotation" slide (WiP)
The "spikes" of the oblique coordinates must be corrected (I can't seem to find the correct transformation)

@@
   PALETTE_APPLE["normal"],
-  PALETTE_TANGERINE["normal"]
+  PALETTE_TANGERINE["normal"],
+  PALETTE_UNED["normal"]
 ) |>
   unname() |>
-  setNames(paste(1:4, sep = '_'))
+  setNames(1:5)
 ```
 
+## Supuestos: Ortogonalidad {.smaller}
+
+> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
+
+---@reckase_difficulty_1985 [p. 404]
+
+::::: columns
+::: column
+```{r oblique-rotation-axis-1}
+LIM_INF <- -0.5
+LIM_SUP <-  2.5
+
+axis_lims <- c(LIM_INF, LIM_SUP)
+
+point <- tibble(F1 = 2, F2 = .2)
+
+point_spikes <- point      |>
+  add_column(type = "end")                      |> 
+  bind_rows(tibble(F1 = 0, F2 = 0))             |>
+  complete(F1, F2, fill = list(type = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+point_spikes <- point_spikes                           |> 
+  bind_rows(point_spikes |> slice(3))                  |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+  pivot_wider(names_from = type, values_from = F1:F2)
+
+point |>
+  ggplot(mapping = aes(F1, F2))            +
+  geom_segment(
+    data    = point_spikes,
+    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+  )                                        +
+  geom_point(size = I(4))                  +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  )                                        +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  )                                        +
+  coord_fixed(expand = FALSE, clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = GRID_WIDTH
+    )
+  )
+```
+:::
+
+::: column
+:::
+:::::
+
+::: notes
+Ahora, la TRI multidimensional también dijimos que se interesaba sobre todo por medir (y ordenar) a las personas, no tanto por interpretar sustantivamente esas dimensiones de rasgo latente.
+De ahí que estimar un modelo con suficientes dimensiones para explicar la variabilidad entre los sujetos fuese importante, pero no era tan importante cuáles fueran esas dimensiones.
+Por eso, hay un supuesto, implícito muchas veces en TRI, de que las dimensiones no están correlacionadas: Se estima el modelo con N dimensiones, las que haga falta, añadiendo restricciones a los ítems para identificarlas, y eso da lugar a N dimensiones independientes entre sí (con correlaciones de 0), representadas en N ejes ortogonales (ángulos rectos todos ellos).
+Que esos ejes representen competencia o habilidad en esto o lo otro, en principio, nos da igual.
+Con asegurarnos de que hay suficientes dimensiones para "capturar" toda la variabilidad debida a esos rasgos nos basta.
+
+Pues bien, ese es el otro supuesto que hace Reckase para calcular los parámetros multidimensionales: Los cosenos directores del ítem al cuadrado suman 1.
+Es decir, está aplicando el teorema de Pitágoras.
+Esta expresión implica, de manera implícita, que todos los ejes del espacio latente son "ortogonales" entre sí.
+Como digo, en TRI originalmente no importaba mucho que los ejes se pudieran interpretar como rasgos latentes.
+No obstante, en 1999 llega McDonald, y dice que la TRI y el AF son fundamentalmente la misma cosa.
+:::
+
+## Supuestos: Ortogonalidad {.smaller}
+
+<!-- TODO: Equation as LaTeX -->
+
+> ~~\[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]~~
+
+---@reckase_difficulty_1985 [p. 404]
+
+<!-- # TODO: Coordenadas rectangulares resultantes de una rotación oblicua -->
+
+::::: columns
+::: column
+```{r oblique-rotation-axis-2}
+SEC_COLOR <- "grey60"
+
+BASIS_VEC_1 <- c(1, 0)
+BASIS_VEC_2 <- c(4, 1) / sqrt(17)
+
+axis_lims  <- c(LIM_INF, LIM_SUP)
+axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
+transf_matrix <- cbind(BASIS_VEC_1, BASIS_VEC_2) |> unname() |> t() |> solve()
+
+# point_transf <- point |>
+#   pivot_longer(everything()) |>
+#   mutate(value = transf_matrix %*% value |> drop()) |>
+#   pivot_wider()
+# 
+# point_spikes_transf <- point_transf             |>
+#   add_column(type = "end")                      |> 
+#   bind_rows(tibble(F1 = 0, F2 = 0))             |>
+#   complete(F1, F2, fill = list(type = "start")) |>
+#   filter(F1 != 0 | F2 != 0)
+# 
+# point_spikes_transf <- point_spikes_transf             |> 
+#   bind_rows(point_spikes_transf |> slice(3))           |>
+#   mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+#   pivot_wider(names_from = type, values_from = F1:F2)
+
+point_spikes_transf <-
+  point_spikes |>
+    pivot_longer(starts_with("F")) |>
+    separate(name, into = c("dim", "coord")) |>
+    pivot_wider(names_from = coord, values_from = value) |>
+    group_by(n) |>
+    mutate(start = (transf_matrix %*% start |> drop())) |>
+  pivot_wider(names_from = dim, values_from = start:end)
+
+point |>
+  ggplot(mapping = aes(F1, F2)) +
+  geom_abline(slope = axis_slope) +
+  geom_segment(
+    data    = point_spikes_transf,
+    mapping = aes(start_F1, start_F2, xend = end_F1, yend = end_F2),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = SEC_COLOR
+  ) +
+  geom_point(size = I(4)) +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  ) +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  ) +
+  coord_fixed(expand = FALSE, clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(linewidth = GRID_WIDTH),
+    panel.grid.major.x = element_line(color = SEC_COLOR),
+    panel.grid.major.y = element_line(color = "black")
+  )
+```
+:::
+
+::: column
+:::
+:::::
+
+::: notes
+Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas, que dan lugar a ejes no ortogonales.
+Luego esta restricción que propone Reckase parece que en principio no se podría aplicar.
+
+\[NEXT\]
+
+Las coordenadas resultantes de cualquier rotación oblicua las podemos representar sin ningún problema en ejes rectangulares (las llamadas "coordenadas Cartesianas").
+Pero algo parece fallar cuando generalizamos los parámetros multidimensionales de TRI que hemos mostrado antes a esos espacios de coordenadas no ortogonales.
+Si rotamos los ejes de coordenadas, pero los ítems siguen siendo los mismos, es de esperar que la discriminación multidimensional de un ítem (i.e., la "longitud" de su vector) no cambie.
+Esta propiedad se llama invarianza.
+Veámoslo con un ejemplo.
+:::
+
 ## References

### 2025-01-24 output/sandbox_slideshow.qmd
Complete "orthogonality" introductory slide

--- 2024-12-20 output/sandbox_slideshow.qmd
@@
   PALETTE_APPLE["normal"],
-  PALETTE_TANGERINE["normal"]
+  PALETTE_TANGERINE["normal"],
+  PALETTE_UNED["normal"]
 ) |>
   unname() |>
-  setNames(paste(1:4, sep = '_'))
+  setNames(1:5)
 ```

### 2025-01-24 output/seminar-slideshow.qmd
Complete "orthogonality" introductory slide

@@
 
-## Supuestos: Ortogonalidad
-
-> \[...\] before performing the differentiation, the constraint that $\cos^2 \, \mathbf{\alpha} = 1$ is added to the expression for the slope \[...\]
+## Supuestos: Ortogonalidad {.smaller}
+
+> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
 
 ---@reckase_difficulty_1985 [p. 404]
-
-<!-- # TODO: Ejemplo de rotación oblicua con los ejes resultantes oblicuos -->
-
-<!-- # TODO: Coordenadas rectangulares resultantes de una rotación oblicua -->
 
@@
 ::: column
+```{r oblique-rotation-axis-1}
+LIM_INF <- -0.5
+LIM_SUP <-  2.5
+
+axis_lims <- c(LIM_INF, LIM_SUP)
+
+point <- tibble(F1 = 2.4, F2 = .1)
+
+point_spikes <- point      |>
+  add_column(type = "end")                      |> 
+  bind_rows(tibble(F1 = 0, F2 = 0))             |>
+  complete(F1, F2, fill = list(type = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+point_spikes <- point_spikes                           |> 
+  bind_rows(point_spikes |> slice(3))                  |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+  pivot_wider(names_from = type, values_from = F1:F2)
+
+point |>
+  ggplot(mapping = aes(F1, F2))            +
+  geom_segment(
+    data    = point_spikes,
+    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+  )                                        +
+  geom_point(size = I(4))                  +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  )                                        +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  )                                        +
+  coord_fixed(expand = FALSE, clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = GRID_WIDTH
+    )
+  )
+```
 :::
@@
 Como digo, en TRI originalmente no importaba mucho que los ejes se pudieran interpretar como rasgos latentes.
-
-\[NEXT\]
-
 No obstante, en 1999 llega McDonald, y dice que la TRI y el AF son fundamentalmente la misma cosa.
+:::
+
+## Supuestos: Ortogonalidad {.smaller}
+
+<!-- TODO: Equation as LaTeX -->
+
+> ~~\[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]~~
+
+---@reckase_difficulty_1985 [p. 404]
+
+::::: columns
+::: column
+<!-- TODO: Animate? -->
+
+```{r oblique-rotation-axis-2}
+SEC_COLOR <- "grey60"
+
+BASIS_VEC_1 <- c(1, 0)
+BASIS_VEC_2 <- c(12, 1)
+BASIS_VEC_2 <- BASIS_VEC_2 / sqrt(sum(BASIS_VEC_2^2))
+
+axis_lims  <- c(LIM_INF, LIM_SUP)
+axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
+transf_matrix <- cbind(BASIS_VEC_1, BASIS_VEC_2) |> unname()
+
+point_transf <- point |>
+  pivot_longer(everything()) |>
+  mutate(value = solve(transf_matrix) %*% value |> drop()) |>
+  pivot_wider()
+
+point_spikes_transf <- point_transf             |>
+  add_column(type = "end")                      |>
+  bind_rows(tibble(F1 = 0, F2 = 0))             |>
+  complete(F1, F2, fill = list(type = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+point_spikes_transf <- point_spikes_transf             |>
+  bind_rows(point_spikes_transf |> slice(3))           |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+  pivot_wider(names_from = type, values_from = F1:F2)
+
+point_spikes_transf_ob <- point_spikes_transf |>
+  pivot_longer(starts_with("F")) |>
+  separate(name, into = c("dim", "coord")) |>
+  pivot_wider(names_from = coord, values_from = value) |>
+  group_by(n) |>
+  mutate(
+    start = (transf_matrix %*% start |> drop()),
+    end   = (transf_matrix %*% end   |> drop())
+  ) |>
+  pivot_wider(names_from = dim, values_from = start:end)
+
+point |>
+  ggplot(mapping = aes(F1, F2)) +
+  geom_abline(slope = axis_slope) +
+  geom_segment(
+    data    = point_spikes,
+    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = SEC_COLOR
+  ) +
+  geom_segment(
+    data    = point_spikes_transf_ob,
+    mapping = aes(start_F1, start_F2, xend = end_F1, yend = end_F2),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43'
+  ) +
+  geom_point(size = I(4)) +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  ) +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  ) +
+  coord_fixed(expand = FALSE, clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    # panel.grid.major   = element_line(linewidth = GRID_WIDTH),
+    panel.grid.major.x = element_line(
+      linewidth = GRID_WIDTH,
+      color = SEC_COLOR
+    ),
+    panel.grid.major.y = element_line(linewidth = GRID_WIDTH, color = "black")
+  )
+```
+:::
+
+::: column
+:::
+:::::
+
+::: notes
 Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas, que dan lugar a ejes no ortogonales.
-
-\[NEXT\]
-
-Los pesos factoriales resultantes (o discriminaciones) los podemos representar sin ningún problema en coordenadas rectangulares (las llamadas "coordenadas Cartesianas").
+Luego esta restricción que propone Reckase parece que en principio no se podría aplicar.
+:::
+
+## Supuestos: Ortogonalidad {.smaller}
+
+> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
+
+---@reckase_difficulty_1985 [p. 404]
+
+::::: columns
+::: column
+```{r oblique-rotation-axis-2}
+```
+:::
+
+::: column
+<!-- TODO: Animate? -->
+
+```{r oblique-rotation-rect-coords}
+point_transf |>
+  ggplot(mapping = aes(F1, F2))            +
+  geom_segment(
+    data    = point_spikes_transf,
+    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+  )                                        +
+  geom_point(size = I(4))                  +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  )                                        +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = NULL
+  )                                        +
+  coord_fixed(expand = FALSE, clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = GRID_WIDTH
+    )
+  )
+```
+:::
+:::::
+
+::: notes
+Las coordenadas resultantes de cualquier rotación oblicua las podemos representar sin ningún problema en ejes rectangulares (las llamadas "coordenadas Cartesianas rectangulares").
+Son coordenadas, nada más; el espacio en el que yo las represente en realidad da igual.
+Así que si represento un ítem en ejes rectangulares, nada me impide aplicar las fórmulas de Reckase y obtener sus parámetros multidimensionales.
+
 Pero algo parece fallar cuando generalizamos los parámetros multidimensionales de TRI que hemos mostrado antes a esos espacios de coordenadas no ortogonales.
-Si rotamos los ejes de coordenadas, pero los ítems siguen siendo los mismos, es de esperar que la "longitud" del vector de ese ítem (i.e.
-su discriminación), no cambie (esta propiedad se llama invarianza).
+Si rotamos los ejes de coordenadas, pero los ítems siguen siendo los mismos, es de esperar que la discriminación multidimensional de un ítem (i.e., la "longitud" de su vector) no cambie.
+Esta propiedad se llama invarianza.
 Veámoslo con un ejemplo.

### 2025-01-24 output/seminar-slideshow.qmd
Changes example to exaggerate difference in length
The new example of the coordinate transformation takes a point with very small coordinates and makes its transformed coordinates much bigger

@@
 ```{r oblique-rotation-axis-1}
-LIM_INF <- -0.5
-LIM_SUP <-  2.5
-
-axis_lims <- c(LIM_INF, LIM_SUP)
-
-point <- tibble(F1 = 2.4, F2 = .1)
+LIM_INF_X <- -2.7
+LIM_INF_Y <- -0.5
+LIM_SUP   <-  2.7
+
+axis_lims_x <- c(LIM_INF_X, LIM_SUP)
+axis_lims_y <- c(LIM_INF_Y, LIM_SUP)
+
+point <- tibble(F1 = .1, F2 = .3)
 
@@
   scale_x_continuous(
-    limits = axis_lims,
+    limits = axis_lims_x,
     breaks = 0,
@@
   scale_y_continuous(
-    limits = axis_lims,
+    limits = axis_lims_y,
     breaks = 0,
@@
 BASIS_VEC_1 <- c(1, 0)
-BASIS_VEC_2 <- c(12, 1)
+BASIS_VEC_2 <- c(-8, 1)
 BASIS_VEC_2 <- BASIS_VEC_2 / sqrt(sum(BASIS_VEC_2^2))
 
-axis_lims  <- c(LIM_INF, LIM_SUP)
 axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
@@
   ggplot(mapping = aes(F1, F2)) +
-  geom_abline(slope = axis_slope) +
+  geom_abline(slope = axis_slope, linewidth = GRID_WIDTH) +
   geom_segment(

### 2025-01-31 output/seminar-slideshow.qmd
Add "item rotation" example

@@
 ::: {.column width="38%"}
+<!-- TODO: Annotate figure? -->
+
 ![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="left"}
@@
 
-## Supuestos: Ortogonalidad
-
-<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
-
-<!-- # TODO: Ejemplo de rotación oblicua (gráfico de ejes y matriz de rotación) -->
-
-::: notes
-Supongamos este ítem, en coordenadas ortogonales.
-El vector muestra sus parámetros de discriminación en las dos dimensiones de los ejes representados.
-Esto sería la "solución sin rotar" que resulta típicamente de ajustar un modelo de TRI multdimensional, o un AFE exploratorio.
+## Supuestos: Ortogonalidad {.smaller}
+
+::::::: columns
+::: column
+```{r item-oblique-axes-1}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     0.5,  1,    0
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+item_spikes <- mirt_item_coords        |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(0.5, -0.1), y = c(-0.1, 1),
+    label         = c("*a*^*^~11~", "*a*^*^~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1/3),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+::::: column
+::: fragment
+$$
+  \begin{align}
+    \mathbf{a}^*_1 &=
+      \begin{bmatrix}
+        0.5 \\
+        1
+      \end{bmatrix}\\
+      d^*_1 &= 0
+  \end{align}
+$$
+:::
+
+<br>
+
+::: fragment
+$$
+  \begin{align}
+    MDISC^*_1 &= \sqrt{
+        [0.5, 1]
+        \begin{bmatrix}
+          0.5 \\
+          1
+        \end{bmatrix}
+      } \\
+      &= \sqrt{ 0.5 · 0.5 + 1 · 1 } \\
+      &= \sqrt{ 0.25 + 1 } \\
+      &= \sqrt{ 1.25 } \\
+      &\approx 1.12
+  \end{align}
+$$
+:::
+:::::
+:::::::
+
+::: notes
+Supongamos un ítem, en coordenadas ortogonales (de una solución sin rotar).
+(Con vuestro permiso, le voy a llamar "a asterico", para representar las coordenadas "originales" de una solución ortogonal, antes de una rotación).
+El vector muestra los componentes de su parámetrosde discriminación en las dos dimensiones de los ejes representados.
+Esto sería la "solución sin rotar" que resulta típicamente de ajustar un modelo de TRI multdimensional, o un análisis factorial exploratorio (si lo convertimos a métrica de TRI).
 La discriminación multidimensional de este vector es (explicar paso a paso).
@@
 Supongamos que aplicamos una rotación oblicua.
-Hemos identificado dos rasgos latentes sustantivos, correlacionados entre sí.
+Imaginad, por ejemplo, que la solución incluye más ítems aparte de este, y cada uno de los ejes se alinea con un grupo de ítems después de hacer la rotación, de forma que da una solución próxima a la estructura simple.
+Esa rotación como digo es oblicua, da lugar a dos ejes no ortogonales, y cada eje se identifica con un rasgo latente sustantivo.
+Como son oblicuos, la solución representa que esos dos rasgos latentes están correlacionados entre sí, con una correlación "la que sea", no nos importa cuál en este momento.
+:::
+
+## Supuestos: Ortogonalidad {.smaller}
+
+:::::::: columns
+::: column
+```{r item-oblique-axes-2}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+BASIS_VEC_1 <- c(1, 0)
+BASIS_VEC_2 <- c(-.5, 1)
+
+axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
+transf_matrix <- cbind(BASIS_VEC_1, BASIS_VEC_2) |> unname()
+
+item_coords_transf <- mirt_item_coords |>
+  pivot_longer(-item) |>
+  separate(name, into = c("coord", "dim")) |>
+  group_by(coord) |>
+  mutate(
+    dim   = paste0('F', dim),
+    value = solve(transf_matrix) %*% value |> drop()
+  ) |>
+  ungroup() |>
+  pivot_wider(names_from = dim)
+
+item_spikes_transf <- item_coords_transf         |>
+  select(-item)                                  |>
+  filter(coord == "end")                         |>
+  bind_rows(tibble(F1 = 0, F2 = 0))              |>
+  complete(F1, F2, fill = list(coord = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+item_spikes_transf <- item_spikes_transf               |>
+  bind_rows(item_spikes_transf |> slice(3))            |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+  pivot_wider(names_from = coord, values_from = F1:F2)
+
+item_spikes_transf_ob <- item_spikes_transf |>
+  pivot_longer(starts_with("F")) |>
+  separate(name, into = c("dim", "coord")) |>
+  pivot_wider(names_from = coord, values_from = value) |>
+  group_by(n) |>
+  mutate(
+    start = (transf_matrix %*% start |> drop()),
+    end   = (transf_matrix %*% end   |> drop())
+  ) |>
+  pivot_wider(names_from = dim, values_from = start:end)
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  ) +
+  geom_abline(
+    intercept = -2:3,
+    slope     = -2,
+    linewidth = GRID_WIDTH,
+    linetype  = "17",
+    color     = AXIS_COLOR
+  ) +
+  geom_abline(slope = -2, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0,  linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes_transf_ob,
+    mapping = aes(start_F1, start_F2, xend = end_F1, yend = end_F2),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-2:2)/2 + .375,
+    labels       = ~. - .375,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:2)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(1, -0.59), y = c(-0.07, 1),
+    label         = c("*a*~11~", "*a*~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 0),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major.y = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+:::::: column
+::: fragment
+$$
+  \mathbf{Q} = \begin{bmatrix}
+      1 & 0.5 \\
+      0 & 1
+    \end{bmatrix}
+$$
+:::
+
+::: fragment
+$$
+  \begin{align}
+    \mathbf{a}_1 =
+    \mathbf{Q} \mathbf{a}^*_1 &= \begin{bmatrix}
+      1 & 0.5 \\
+      0 & 1
+    \end{bmatrix}
+    \begin{bmatrix}
+      0.5 \\
+      1
+    \end{bmatrix} \\
+    &= \begin{bmatrix}
+      1 · 0.5 + 0.5 · 1 \\
+      0 · 0.5 + 1 · 1
+    \end{bmatrix} \\
+    &= \begin{bmatrix}
+      0.5 + 0.5 \\
+      0   + 1
+    \end{bmatrix}
+    = \begin{bmatrix}
+      1 \\
+      1
+    \end{bmatrix}
+  \end{align}
+$$
+:::
+
+::: fragment
+$$
+  \begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix}\\
+      d_1 &= d^*_1 = 0
+  \end{align}
+$$
+:::
+::::::
+::::::::
+
+::: notes
 Los ejes resultantes son los que se muestran en este gráfico.
-Los nuevos parámetros de discriminación del ítem están representadas cada una por la "proyección" de ese vector sobre cada uno de los ejes (paralela al otro, u otros, si hay más de dos).
-:::
-
-## Supuestos: Ortogonalidad
-
-<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
-
-<!-- # TODO: Ejemplo de item proyectado en los ejes rotados (gráfico y vectorial) -->
-
-<!-- # TODO: Cálculo de la discriminación en los ejes rotados -->
-
-::: notes
+No es exactamente una rotación, ya que el eje oblicuo \[señalar\] es un poco más largo (si lo rotáramos solamente acabaría por aquí \[señalar\]), pero para el caso es una transformación de los ejes, y cualquier transformación lineal nos sirve para ejemplificar, y así simplificamos el ejemplo.
+Los nuevos valores del parámetros de discriminación del ítem, en este caso ya sí lo llamo "a", porque esta es la solución rotada, la que me interesa, la que es interpretable según mi teoría, están representados cada uno por la "proyección" de ese vector sobre cada uno de los ejes (proyección paralela al otro eje, u otros, si hubiera más de dos).
+
 Hacer una rotación supone "multiplicar" las coordenadas de este vector por la matriz de rotación que representa ese nuevo sistema de coordenadas (la transformación del ortogonal a este nuevo, oblicuo).
-En este caso, esta es la matriza de rotación.
+
+\[NEXT\]
+
+En este caso, la matriz de rotación es esta (la llamamos matriz Q).
+Esta es la matriz que convierte las coordenadas del espacio ortogonal original a las coordenadas transformadas (coordenadas "rotadas").
+
+\[NEXT\]
+
 Como los vectores se representan en columnas, se "pre-multiplica" por la matriz de rotación.
 Esto es parecido a multiplicar por un vector; es como si cada fila de la matriz fuese un vector, y los resultados se ponen cada uno en una fila.
-<!-- # TODO: Explicar paso a paso el producto matricial -->
-
+\[Explicar paso a paso el producto matricial\]
+
+\[NEXT\]
+
+El resultado es, como vemos en el gráfico, (1, 1).
+El parámetro de discriminación de este ítem, en su solución rotada, interpretable, es este, (1, 1).
+El valor de la intersección del modelo no cambia.
+:::
+
+## Supuestos: Ortogonalidad
+
+<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
+
+<!-- # TODO: Ejemplo de item proyectado en los ejes rotados (gráfico y vectorial) -->
+
+<!-- # TODO: Cálculo de la discriminación en los ejes rotados -->
+
+::: notes
 Si ahora calculamos la discriminación del ítem en estos ejes rotados, nos encontramos con este resultado.

### 2025-02-07 output/seminar-slideshow.qmd
Add slide with "agnostic MDISC" in rotated axes

@@
 
-## Supuestos: Ortogonalidad
-
-<!-- # TODO: Ejemplo de item en coordenadas ortogonales (gráfico y vectorial) -->
-
-<!-- # TODO: Ejemplo de item proyectado en los ejes rotados (gráfico y vectorial) -->
-
-<!-- # TODO: Cálculo de la discriminación en los ejes rotados -->
-
-::: notes
-Si ahora calculamos la discriminación del ítem en estos ejes rotados, nos encontramos con este resultado.
+## Supuestos: Ortogonalidad {.smaller}
+
+::::::: columns
+::: column
+```{r item-oblique-axes-3}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    1,    0
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+item_spikes <- mirt_item_coords        |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(0.5, -0.1), y = c(-0.1, 1),
+    label         = c("*a*^*^~11~", "*a*^*^~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1/3),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+::::: column
+::: fragment
+$$
+  \begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix}\\
+      d_1 &= 0
+  \end{align}
+$$
+:::
+
+<br>
+
+::: fragment
+$$
+  \begin{align}
+    MDISC^*_1 &= \sqrt{
+        [1, 1]
+        \begin{bmatrix}
+          1 \\
+          1
+        \end{bmatrix}
+      } \\
+      &= \sqrt{ 1 · 1 + 1 · 1 } \\
+      &= \sqrt{ 1 + 1 } \\
+      &= \sqrt{ 2 } \\
+      &\approx 1.41
+  \end{align}
+$$
+:::
+:::::
+:::::::
+
+::: notes
+Ahora bien, al igual que vimos antes, nadie me impide representar este ítem en coordenadas rectangulares.
+
+\[NEXT\]
+
+Y si el ítem está representado en coordenadas rectangulares, puedo aplicar, como vimos, las fórmulas de Reckase.
+Pero si hago eso, me encuentro con que el resultado de la discriminación es este, raíz de 2.
+Antes, recordaréis que era raíz de 1,25, aproximadamente 1,12.
+
 Es decir, ¿que el ítem discrimina mejor o peor en función de qué ejes quiera yo considerar?
-Eso no parece tener sentido, y sin embargo hay muchos ejemplos en la literatura que calculan esta discriminación multidimensional con la fórmula de Reckase (con "rasgos correlacionados") sin preocuparse por "los ejes" que representan esos rasgos de manera fidedigna.
+El ítem es exactamente el mismo, y por lo tanto su discriminación (multidimensional al menos, ya que no depende de la "dirección" a considerar, como los valores del vector de discriminación) debería ser la misma.
+Esta propiedad se llama "invarianza" del ítem.
+
+No tien sentido que cambie su discriminación multidimensional, y sin embargo hay muchos ejemplos en la literatura que la calculan con la fórmula de Reckase (con "rasgos correlacionados") sin preocuparse por "los ejes" que representan esos rasgos de manera fidedigna.
 <!-- # TODO: Ejemplos de la literatura -->

### 2025-02-07 output/seminar-slideshow.qmd
Add comments of pending changes in header

@@
 date:          "2024-10-21"
-date-meta:     "2024-11-27"
+date-meta:     "2024-11-27" # TODO: Update
 date-format:   long
@@
   markdown: 
-    wrap:      sentence
-    canonical: true
+    wrap:      sentence  # TODO: Restore
+    canonical: true # TODO: Restore
 execute: 

### 2025-02-07 output/seminar-slideshow.qmd
Improve comments in "agnostic MDISC" slide

@@
 
-No tien sentido que cambie su discriminación multidimensional, y sin embargo hay muchos ejemplos en la literatura que la calculan con la fórmula de Reckase (con "rasgos correlacionados") sin preocuparse por "los ejes" que representan esos rasgos de manera fidedigna.
-<!-- # TODO: Ejemplos de la literatura -->
+No tien sentido que cambie su discriminación multidimensional, y sin embargo la literatura está llena de ejemplos que calculan este parámetro con la fórmula de Reckase (en un espacio de "rasgos correlacionados", que en seguida veremos por qué eso es importante), y que no se preocupan por "los ejes" que representan esos rasgos de manera fidedigna.
+<!-- # TODO: Ejemplos de la literatura? -->
 :::

### 2025-02-14 output/seminar-slideshow.qmd
Add isometric perspective example slide

@@
 ::::: columns
-::: column
+::: {.column .fragment}
 ```{r oblique-rotation-axis-1}
@@
 Por eso, hay un supuesto, implícito muchas veces en TRI, de que las dimensiones no están correlacionadas: Se estima el modelo con N dimensiones, las que haga falta, añadiendo restricciones a los ítems para identificarlas, y eso da lugar a N dimensiones independientes entre sí (con correlaciones de 0), representadas en N ejes ortogonales (ángulos rectos todos ellos).
+
+\[NEXT\]
+
 Que esos ejes representen competencia o habilidad en esto o lo otro, en principio, nos da igual.
@@
 Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas, que dan lugar a ejes no ortogonales.
+
+Aquí vemos un ejemplo de ejes oblicuos (muy exagerado, eso sí).
+Imaginad que esto que estamos representando aquí es un punto sobre La Tierra, y sus coordenadas rectangulares son la longitud y latitud.
+
+Es posible que por algún motivo (no sé cuál, pero imaginémoslo) nos interesen las coordenadas en estos "ejes oblicuos": Este-Oeste, y uno que podría ser algo así como "Este-Sureste este este - Oeste-Noroeste oeste oeste".
+¿Se entiende claramente?
+Proyectando cada coordenada, de forma paralela a los ejes, obtendríamos las coordenadas en este nuevo sistema de "coordenadas geodésicas oblicuas".
+
 Luego esta restricción que propone Reckase parece que en principio no se podría aplicar.
+:::
+
+## (Pero, un inciso...)
+
+![](../www/slideshow-assets/isometric-cube.jpg){height="600" fig-align="center"}
+
+::: aside
+Figura tomada de [Vecteezy](https://www.vecteezy.com/vector-art/3709640-isometric-cube-on-a-white-background)
+:::
+
+::: notes
+Sin embargo, mirad este ejemplo:
+
+¿Qué veis?
+¿Un cubo, verdad?
+¿Son sus lados todos iguales?
+¿Y sus ángulos?
+Son todos ángulos rectos, se entiende.
+Pero si cojo un transportador de ángulos y mido, voy a ver que este mide 120 grados, y este 60.
+
+Este cubo está representado en un plano en perspectiva isométrica (Creo que todo el mundo sabe lo que es la perspectiva isométrica, si no de dibujo en bachillerato, de jugar a los SIMS, aunque sea).
+
+Bien, si ahora os pregunto por las diagonales de una cara, ¿diríais que son iguales?
+Y sin embargo, si mido (en este plano) las dos diagonales, vemos que son distintas.
+¿Estáis de acuerdo?
+
+Esta representación no es errónea: De hecho, se necesita algún sistema de perspectiva para representar un cuerpo sólido en dos dimensiones; la isométrica es solamente una de ellas.
+
+Es decir, que nada me impide, si lo necesito, representar unos ejes con un ángulo cualquiera (recto en este caso) como otro ángulo (120 grados).
 :::

### 2025-02-14 output/seminar-slideshow.qmd
Add alt-text to cube in isometric perspective
Improve notes of "agnostic MDISC"

@@
 
-![](../www/slideshow-assets/isometric-cube.jpg){height="600" fig-align="center"}
+![Un cubo rojo en perspectiva isométrica](../www/slideshow-assets/isometric-cube.jpg){height="600" fig-align="center"}
 
@@
 ::: notes
-Las coordenadas resultantes de cualquier rotación oblicua las podemos representar sin ningún problema en ejes rectangulares (las llamadas "coordenadas Cartesianas rectangulares").
+Lo mismo pasa al representar los ejes de una rotación oblicua:
+Nada me impide representar las coordenadas resultantes en ejes rectangulares (las llamadas "coordenadas Cartesianas rectangulares").
 Son coordenadas, nada más; el espacio en el que yo las represente en realidad da igual.
+Y de hecho, esto lo habréis visto muchas veces; por ejemplo, SPSS representa los pesos factoriales siempre en coordenadas rectangulares, incluso después de aplicar una rotación oblicua.
+
 Así que si represento un ítem en ejes rectangulares, nada me impide aplicar las fórmulas de Reckase y obtener sus parámetros multidimensionales.

### 2025-02-14 output/seminar-slideshow.qmd
Correct MDISC and discrimination super-indexes

@@
     x = c(0.5, -0.1), y = c(-0.1, 1),
-    label         = c("*a*^*^~11~", "*a*^*^~21~"),
+    label         = c("*a*~11~", "*a*~21~"),
     size          = 8,
@@
   \begin{align}
-    MDISC^*_1 &= \sqrt{
+    MDISC^_1 &= \sqrt{
         [1, 1]

### 2025-02-14 output/Letter_to_editor_reviewers.Rmd
Version last response to reviewer 2

@@
 
+The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may be confusing to the reader to refer to the $MDISC_i$ and obviate the influence of $d_i$.
+We have added a clarification in this respect to the end of the paragraph, which now reads:
+
+    > That is, the less discriminating the item is,
+    the further away it will be from the origin.
+    Of course, it is also directly proportional to
+    the intercept parameter $`r INTERCEPT_PARAM`$.
+    This implies that, if $`r INTERCEPT_NULL_EQ`$,
+    $`r DISTANCE_NULL_EQ`$ as well,
+    regardless of the discrimination.
+
 # Responses to Reviwer 3:
@@
 > The manuscript highlights the fact that the coordinate system used to represent results typical of educational achievement and other types of social measures is at the discretion of the person performing the analysis, or the entities developing computer software, if default settings are used in the analysis.
+
+Although we argue indeed that the coordinate system is arbitrary, our stance is also that there is a more interpretable approach than simply assuming orthogonal axes.
+This approach implies that there is a geometric basis (and thus a coordinate system) that can be considered a faithful representation of the covariance structure.
 
@@
 
-> I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes.
+We are glad that the reviewer has found our work thought-provoking.
+We argue indeed that different samples may yield different latent covariance structures.
+This problem has received much attention in the factor analysis literature (i.e., factorial invariance), it has been largely overlooked in the MIRT literature, where any covariance structure is usually represented in orthogonal axes (implying an orthonormal basis).
+
+> [...] I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes.
 > The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals.
 > How are these features interpreted when the coordinate axes are non-orthogonal?
+
+We believe that non-orthogonal axes provide a faithful representation of the latent covariance structure:
+In geometrical terms, the cosine between axes directly maps to the correlation between the two dimensions represented by those axes.
+In the orthogonal space, the angles between items are "distorted" and do not faithfully represent the correlation among those items.
+Consider the following example: Two simple items, each measuring a single dimension of a two-dimensional construct, will always be represented by a 90-degree angle in the orthogonal coordinate system.
+If those two dimensions are uncorrelated, so the items will also be.
+However, suppose those two dimensions are highly correlated; in that case, the items would also have to be equally highly correlated, but would still be represented by a 90-degree angle in orthogonal coordinates.
+Therefore, our argument is that the correlation among items is better represented by the cosines of their vectors in the non-orthogonal space.
+
+Regarding the person parameters, the interpretation is more complicated:
+If the original, correlated coordinates of the persons are represented in the orthogonal space, the shape of the scatter plot indeed represents the correlation between the latent trait coordinates.
+However, our rationale is that a basis may be used such that, when the coordinates are transformed to an orthonormal one, the resulting scatter plot is spherical (i.e.
+its covariance is an identity matrix).
+This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being the "inverse transpose" of the basis where we argue the items are faithfully represented.
 
@@
 
+The distances between points are given by the norm, which we argue to be invariant to changes of bases.
+Interestingly, when considering the covariance-based version of the multidimensional parameters, the distance between points in the latent space is given by the Mahalanobis distance.
+
 > Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005).
 > It would help if a page number were provided for the discussion.
+
+We apologize for the lack of clarity.
+We intended to refer to footnote 1 in page 16, we have correctly referenced the page and footnote in the new version.
 
@@
 
+We intended to refer to <!-- # TODO: Complete -->
+Please note that, as we do not have access to the 2nd edition anymore, we have used the 3rd one in our new version, and updated all the references accordingly.
+
 > Page 6  Line 21 The expression on this line needs more explanation.
 > I am not sure what the expression is supposed to represent.
+
+This is the expression of a vector represented in a basis, in this case, representing that the vector can be represented in each of the two bases indistinctly.
+Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:
+
+    > [...] such that $`r TRAIT_VECTOR_DEF_EQ`$.
+    That is, vector $`r TRAIT_VECTOR`$ is represented in each of the two bases by
+    the linear combination of its element vectors
+    scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
 
@@
 
+The -1 superscripts indeed indicate the inverse of a matrix.
+This has been an unfortunate typesetting problem of the submission system,
+that renders the -1 and the *T* at the same height.
+We have sorted this out by using parentheses, as you will be able to see in our new submission.
+
 > Page 13 Line 4 The implications of the word “indistinctly” in this sentence are not clear.
+
+<!-- # TODO: Add response -->
 
@@
 
+What we want to imply is that there are two different spaces, the latent space, and the test space, and that, unlike the latent traits, which "naturally lie" in the latent space, the item parameters "naturally lie in the test space", even though they can be projected onto the latent space.
+This happens even when the basis of the latent space (and, consequently, that of the test space) is orthonormal, whih we argue is overlooked (due to a lack of need to make this distinction) in Reckase's original formulations.
+
 > Page 19 Line 8 There is an interesting distinction that seems to be lost here.
@@
 > This has an implication for the formula that is used to compute the distance between points in the space.
+
+As we mentioned above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance.
+One should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
+It is when we represent the items that we should use axes that "faithfully represent the covariance structure", and not when we represent the attributes.
 

### 2025-02-14 R/Formulae.R

@@ -1169,6 +1169,10 @@ END_ITEM_PARAMS_EQ     <- latex_eq(
 
 #### Geometric properties of the parameters: ----
 
+# Signed distance:
+INTERCEPT_NULL_EQ <- latex_eq(INTERCEPT_PARAM, 0)
+DISTANCE_NULL_EQ  <- latex_eq(DISTANCE_PARAM, 0)
+
 # Direction cosines:
 ANGLE_VECTORS_ITEM   <- latex_sub(ANGLE_TS, "$ITEM_INDEX$$DIM_INDEX$")
 COS_VECTORS_ITEM     <- latex_cos(ANGLE_VECTORS_ITEM)

### 2025-02-14 output/Main_text.Rmd
Version last response to reviewer 2
Version responses to reviewer 3 (WiP)

@@
 accurate on a geometrical coordinate system with non-orthogonal axes
-[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
+[@ackerman_multidimensional_2005-1, p. 16, see footnote 1]
+and (possibly) non-standard units,
 often referred to as *general Cartesian coordinates* [@harman_modern_1970].
@@
 respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
+That is, vector $`r TRAIT_VECTOR`$ is represented in each of the two bases by
+the linear combination of its element vectors
+scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
 
@@
 the further away it will be from the origin.
+Of course, it is also directly proportional to
+the intercept parameter $`r INTERCEPT_PARAM`$.
+This implies that, if $`r INTERCEPT_NULL_EQ`$, $`r DISTANCE_NULL_EQ`$ as well,
+regardless of the discrimination.
 

### 2025-02-14 R/Formulae.R

@@ -289,7 +289,10 @@ TEST_SPACE_BASIS_EQ       <- latex_eq(TEST_SPACE_BASIS, TEST_SPACE_BASIS_SET)
 
 # Standardized discrimination vector:
 TRANSFORM_MATRIX_INV        <- latex_inverse(TRANSFORM_MATRIX)
-TRANSFORM_MATRIX_TRANSP_INV <- latex_transp("{$TRANSFORM_MATRIX_INV$}")
+TRANSFORM_MATRIX_TRANSP_INV <- latex_transp(
+  "{$TRANSFORM_MATRIX_INV$}",
+  .par = TRUE
+)
 DISCR_STD_COORDS_DEF        <- latex(TRANSFORM_MATRIX_TRANSP_INV, DISCR_COORDS)
 DISCR_STD_COORDS_EQ         <- latex_eq(
   DISCR_STD_COORDS,

### 2025-02-15 output/Letter_to_editor_reviewers.Rmd
Finish adding (initial) responses to reviewer 3

@@
 We apologize for the lack of clarity.
-We intended to refer to footnote 1 in page 16, we have correctly referenced the page and footnote in the new version.
+We intended to refer to footnote 1 in page 16, we have now completed the reference.
 
@@
 
+That is exactly the case.
+All graphical devices require that coordinates are represented in a rectangular Cartesian system such that the first coordinate maps to the horizontal dimension and the second coordinate to the vertical one.
+Therefore, even when computing the coordinates in a non-orthonormal basis, we need to transform them to the canonical one to plot them.
+This is explained in our manuscript, in p. 19 (lines 9-14) in the original version, and in p. <!-- # TODO: Complete --> (lines <!-- # TODO: Complete -->) in the new submission.
+We are unsure about how to make this clearer, we are afraid, so we would appreciate any suggestions.
+
 > Page 26 Line 8 This discussion seems to miss the distinction between correlations between traits and dimensions  of sensitivity of the test items.
+
+<!-- # NOTE: Help needed -->
 
@@
 > It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.
+
+<!-- # NOTE: Help needed -->
 
@@
 > What is the value added of using the more complex representation?
+
+To our understanding, the representation is always sample dependent, because item parameters are in turn.
+Our argument is that the non-orthogonal coordinate system is a more faithful representation of the latent covariance structure, as the cosines between the axes are the geometric equivalent of the latent trait correlations.
+This is, to our understanding, the value of using this more complex representation, and so we have tried to convey in (please see p. 31, line 19 to p. 32, line 4, in the original version.
+
+We acknowledge though that this may have been explained without enough clarity, and therefore strived to improve this discussion.
+The aforementioned paragraph now reads:
+
+    > [...] This implies that the axes defined by the test space basis,
+    and not the ones defined by the latent space basis,
+    are a geometrical representation of the covariance structure:
+    Indeed, the cosines of the angles among the test space axes are
+    the geometric equivalent of the correlations. [...]

### 2025-02-15 output/Main_text.Rmd
Finish adding (initial) responses to reviewer 3

@@
 This implies that the axes defined by the test space basis,
-and not the latent space basis,
-are a geometrical representation of the covariance structure.
+and not the ones defined by the latent space basis,
+are a geometrical representation of the covariance structure:
+Indeed, the cosines of the angles among the test space axes are
+the geometric equivalent of the correlations.
 This result is both surprising and counterintuitive

### 2025-02-15 output/Letter_to_editor_reviewers.Rmd
Drop package bibliography rendering chunk
This code chunk is (obviously) unnecessary
Update letter to editor styles

@@
 
-```{r write-bib}
-# automatically create a bib database for R packages
-write_bib(
-  c(.packages(), 'bookdown', 'knitr', 'rmarkdown'),
-  'www/packages.bib'
-)
-```
-
 ```{r sources}
@@
 We will be happy to address any further comments or suggestions
-the action editor or the reviewers may deem necesssary.
+the action editor or the reviewers may deem necessary.
 
@@
 
-> I have a concern with making the latent ability axes nonorthogonal to represent the covariance structure between the latent abilities.
-> Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.
+###### I have a concern with making the latent ability axes nonorthogonal to represent the covariance structure between the latent abilities.
+###### Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.
 
@@
 That is what we do when we posit that the covariance structure is taken into account when computing the multidimensional parameters.
-The method we propose thus so, by considering an algebraic basis that conveys information of the covariance structure.
+The method we propose does so, by considering an algebraic basis that conveys information of the covariance structure.
 <!-- TODO: Changes to the main text? -->
 
-> The statement (p.13, line 8) about “the test space where item parameters naturally lie” needs greater explanation with examples.
+###### The statement (p.13, line 8) about “the test space where item parameters naturally lie” needs greater explanation with examples.
 
-> Like Reviewer 3, I do not understand how the magnitude of discrimination values (p 29 , 7) can be compared/interpreted values of discrimination are comparable from the two representations.
-> It seems that the metric of the represented space has changed, and equivalent numbers will not have equivalent interpretations.
+###### Like Reviewer 3, I do not understand how the magnitude of discrimination values (p 29 , 7) can be compared/interpreted values of discrimination are comparable from the two representations.
+###### It seems that the metric of the represented space has changed, and equivalent numbers will not have equivalent interpretations.
 
@@
 
-> I also wonder about how one compares the standard errors of the item parameters when the covariance covariance structures are different.  Are they directly comparable?
+###### I also wonder about how one compares the standard errors of the item parameters when the covariance covariance structures are different.  Are they directly comparable?
 
@@
 
-> I would like to see the motivation and psychometric utility of this approach.
-> Does it improve the interpretation of the latent abilities and thus result in more insight into how the results are used?
-> The author needs to address this with legitimate examples for testing practitioners.
+###### I would like to see the motivation and psychometric utility of this approach.
+###### Does it improve the interpretation of the latent abilities and thus result in more insight into how the results are used?
+###### The author needs to address this with legitimate examples for testing practitioners.
 
@@
 
-> What implications does your approach have for standard psychometric procedures like equating.
+###### What implications does your approach have for standard psychometric procedures like equating.
 
@@
 
-> Are there standard computer programs that you would recommend to do the estimation/modeling you are proposing?
+###### Are there standard computer programs that you would recommend to do the estimation/modeling you are proposing?
 
@@
 
-> 1. Introduction, too lengthy, 1-2 pages will be fine.
+###### 1. Introduction, too lengthy, 1-2 pages will be fine.
 
@@
 
-> 2. Introduction, do you have any hypothesis for the current study? Please briefly describe it at the end.
+###### 2. Introduction, do you have any hypothesis for the current study? Please briefly describe it at the end.
 
@@
 
-> 3. The details about the datasets should be described here.
+###### 3. The details about the datasets should be described here.
 
@@
 
-> 4. Within or between multidimensional latent? This should be detailed.
+###### 4. Within or between multidimensional latent? This should be detailed.
 
@@
 
-> 5. Limitation is missing and should be added here.
+###### 5. Limitation is missing and should be added here.
 
@@
 
-> 6. Conclusion is missing and should be added here.
+###### 6. Conclusion is missing and should be added here.
 
@@
 
-> [...] it is difficult to ascertain its practical value, as it is unclear how it can be applied in the real world.
+###### [...] it is difficult to ascertain its practical value, as it is unclear how it can be applied in the real world.
 
@@
 
-> While the manuscript includes a graphical representation example with two-dimensional latent variables, real-world applications often involve higher dimensions (e.g., the Big Five personality traits or
-Cattell’s 16 personality factors). If the current results cannot be applied to the
-high-dimensional cases, would that significantly limit their practical value and
-applicability?
+###### While the manuscript includes a graphical representation example with two-dimensional latent variables, real-world applications often involve higher dimensions (e.g., the Big Five personality traits or Cattell’s 16 personality factors).
+###### If the current results cannot be applied to the high-dimensional cases, would that significantly limit their practical value and applicability?
 
@@
 
-> 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect.
-> It should be “$\mathcal{C}$”.
+###### 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect.
+###### It should be “$\mathcal{C}$”.
 
@@
 
-> 2. The notations in the manuscript lack clarity, particularly in Eq. (19) on page 12.
-> Based on Eq. (5) and Eq. (15), $\mathbf{a}_i$ in the numerator of Eq. (19) should be $\mathbf{a}_i^{\mathcal{B}*}$.
+###### 2. The notations in the manuscript lack clarity, particularly in Eq. (19) on page 12.
+###### Based on Eq. (5) and Eq. (15), $\mathbf{a}_i$ in the numerator of Eq. (19) should be $\mathbf{a}_i^{\mathcal{B}*}$.
 
@@
 
-> Similar issues occur in Eqs. (26) and (27) on page 17.
+###### Similar issues occur in Eqs. (26) and (27) on page 17.
 
@@
 
+<!-- # TODO: Substitute refs. by manual values -->
 > Of course,
@@
 
-> 3. In the second graph on page 18, the authors state, “*That is, the less discriminating the item, the further it will be from the origin.*”
-> However, the value of $D_i$ is influenced not only by the $MDISC_i$ parameter but also by $d_i$.
-> For instance, in Table 1 on page 20, $d_i$ for item 1 is zero, meaning that regardless of $MDISC_i$, $D_i$ remains zero.
+###### 3. In the second graph on page 18, the authors state, “*That is, the less discriminating the item, the further it will be from the origin.*”
+###### However, the value of $D_i$ is influenced not only by the $MDISC_i$ parameter but also by $d_i$.
+###### For instance, in Table 1 on page 20, $d_i$ for item 1 is zero, meaning that regardless of $MDISC_i$, $D_i$ remains zero.
 
@@
 
-> The manuscript highlights the fact that the coordinate system used to represent results typical of educational achievement and other types of social measures is at the discretion of the person performing the analysis, or the entities developing computer software, if default settings are used in the analysis.
+###### The manuscript highlights the fact that the coordinate system used to represent results typical of educational achievement and other types of social measures is at the discretion of the person performing the analysis, or the entities developing computer software, if default settings are used in the analysis.
 
@@
 
-> This is a very important issue and it is not frequently discussed in the research literature.
+###### This is a very important issue and it is not frequently discussed in the research literature.
 Two researchers can administer the identical instrument to samples of individuals with different characteristics – say typical high school students versus those who have a different native language – and get different representations of the test results from analyses programs, such as the correlations between estimated constructs.
@@
 
-> [...] I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes.
-> The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals.
-> How are these features interpreted when the coordinate axes are non-orthogonal?
+###### [...] I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes.
+###### The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals.
+###### How are these features interpreted when the coordinate axes are non-orthogonal?
 
@@
 
-> Care needs to be taken when using nonorthogonal coordinates that the distances between points in the space are correctly represented.
-> What formula should be used for the distance between two points in each case?
+###### Care needs to be taken when using nonorthogonal coordinates that the distances between points in the space are correctly represented.
+###### What formula should be used for the distance between two points in each case?
 
@@
 
-> Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005).
-> It would help if a page number were provided for the discussion.
+###### Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005).
+###### It would help if a page number were provided for the discussion.
 
@@
 
-> I also wonder about the Harman (1970) reference.
-> I know that he shows a plot of length of arm against length of leg using orthogonal axes for the plot even though the two measures are correlated.
-> I am not not sure how the axes should be oriented to reflect the correlation between the two measures.
+###### I also wonder about the Harman (1970) reference.
+###### I know that he shows a plot of length of arm against length of leg using orthogonal axes for the plot even though the two measures are correlated.
+###### I am not not sure how the axes should be oriented to reflect the correlation between the two measures.
 
@@
 
-> Page 6  Line 21 The expression on this line needs more explanation.
-> I am not sure what the expression is supposed to represent.
+###### Page 6  Line 21 The expression on this line needs more explanation.
+###### I am not sure what the expression is supposed to represent.
 
@@
 
-> Page 7  Line 24 The meaning of the superscripts is not clear in this expression.
-> What is the source of the -1T?  Is that the transpose of the inverse?
-> Do the -1 superscripts indicate an inverse?
+###### Page 7  Line 24 The meaning of the superscripts is not clear in this expression.
+###### What is the source of the -1T?  Is that the transpose of the inverse?
+###### Do the -1 superscripts indicate an inverse?
 
@@
 
-> Page 13 Line 4 The implications of the word “indistinctly” in this sentence are not clear.
+###### Page 13 Line 4 The implications of the word “indistinctly” in this sentence are not clear.
 
@@
 
-> Page 13 Line 8 The statement “the test space is where the item parameters naturally lie” seems to imply there is a correct representation of the item parameters.
-> But, the correlations between constructs changes with the sample of examinees and the coordinates can easily be rotated and translated giving different representations of the item parameters.
-> I would like to see support for the statement.
+###### Page 13 Line 8 The statement “the test space is where the item parameters naturally lie” seems to imply there is a correct representation of the item parameters.
+###### But, the correlations between constructs changes with the sample of examinees and the coordinates can easily be rotated and translated giving different representations of the item parameters.
+###### I would like to see support for the statement.
 
@@
 
-> Page 19 Line 8 There is an interesting distinction that seems to be lost here.
-> Attributes can be measured independently, such as height and weight and yield correlated results.
-> Under those circumstances, should axes be represented as orthogonal, or should the axes represent the correlation.
-> This has an implication for the formula that is used to compute the distance between points in the space.
+###### Page 19 Line 8 There is an interesting distinction that seems to be lost here.
+###### Attributes can be measured independently, such as height and weight and yield correlated results.
+###### Under those circumstances, should axes be represented as orthogonal, or should the axes represent the correlation.
+###### This has an implication for the formula that is used to compute the distance between points in the space.
 
@@
 
-> Page 22 Figure 1 It seems that both of these plots are shown with orthogonal coordinate axes.
-> But the orthogonal axes have changed so that the nonorthogonal axes can be shown in the orthogonal representation.
+###### Page 22 Figure 1 It seems that both of these plots are shown with orthogonal coordinate axes.
+###### But the orthogonal axes have changed so that the nonorthogonal axes can be shown in the orthogonal representation.
 
@@
 
-> Page 26 Line 8 This discussion seems to miss the distinction between correlations between traits and dimensions  of sensitivity of the test items.
+###### Page 26 Line 8 This discussion seems to miss the distinction between correlations between traits and dimensions  of sensitivity of the test items.
 
@@
 
-> Page 29 Line 7 I am curious if the values of discrimination are comparable from the two representations.
-> It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.
+###### Page 29 Line 7 I am curious if the values of discrimination are comparable from the two representations.
+###### It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.
 
@@
 
-> Page 30 Line 8 It would be helpful to show the orthogonal and nonorthogonal representations of items and then discuss the interpretation of the parameter estimates based on each of the representations.
-> I have been unable to determine the value of using the nonorthogonal representation.
-> The nonorthogonal representation is more complex, and it seems sample dependent because it uses the observed correlations/covariances.
-> What is the value added of using the more complex representation?
+###### Page 30 Line 8 It would be helpful to show the orthogonal and nonorthogonal representations of items and then discuss the interpretation of the parameter estimates based on each of the representations.
+###### I have been unable to determine the value of using the nonorthogonal representation.
+###### The nonorthogonal representation is more complex, and it seems sample dependent because it uses the observed correlations/covariances.
+###### What is the value added of using the more complex representation?
 

### 2025-02-16 output/Letter_to_editor_reviewers.Rmd
Improve responses to editor and reviewers

@@
     keep_md:         no
+bibliography: ../www/Multidimensional-parameters-MCLM.bib
+csl:          ../www/apa-old-doi-prefix.csl
 editor_options: 
@@
 
-We agree with the editor that different groups of examinees can result in different covariance structures, due to the interaction between the examinees and the items.
-However, in our understanding, the same could be predicated from the item parameters themselves, without taking into account the covariance structure.
+Indeed, different groups of examinees can result in different covariance structures.
+However, in our understanding, the same fact can be predicated about the item parameters themselves, even without considering the covariance structure at all.
 For example, in unidimensional IRT, the item parameters may be different for different groups of examinees, even if the variance is constrained to be equal across groups; this is indeed the basis of methods such as multigroup IRT.
+
 It is our stance that the covariance structure, understood as a property of the data of course, should be taken into account when determining the multidimensional properties of the items.
-That is what we do when we posit that the covariance structure is taken into account when computing the multidimensional parameters.
-The method we propose does so, by considering an algebraic basis that conveys information of the covariance structure.
-<!-- TODO: Changes to the main text? -->
+That is what we propose with the covariance-based (and correlation-based) versions of the multidimensional parameters.
 
 ###### The statement (p.13, line 8) about “the test space where item parameters naturally lie” needs greater explanation with examples.
+
+<!-- TODO: Reply -->
 
@@
 
-Indeed, we propose that the basis used to represent the test space changes its metric.
-We also understand that the metrics of the two representation are not comparable, and thus one should not compare items represented with the agnostic and the covariance-base version of the multidimensional parameters.
-However, when making multi-group comparisons, we think that taking into account the covariance-based structure of the latent space may provide more accurate information.
+As the editor says, the basis used to represent the test space changes its metric.
+We also understand that the metrics of the two representations are not comparable, and thus one should not compare items represented with the agnostic and the covariance-based version of the multidimensional parameters.
+However, when making multi-group comparisons, we think that taking into account the covariance-based structure of the latent space will provide more realistic information, and thus one should use the covariance-based version.
+
+In addition to other (minor) changes, we have added the following paragraph to our introduction to give an intuitive explanation of this fact:
+
+> Take the following as an example:
+A hypothetical complex item measures two latent traits with equal validity
+---i.e., its discrimination parameters are equal with value $a$.
+Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
+we would find that $`r MDISC_SYM` = \sqrt{2} a_i$.
+Assume we apply this item in two groups where, taken to the extreme,
+the two measured dimensions are orthogonal in the first group
+and perfectly correlated in the second one.
+While the previous formula seems correct in the first group,
+one can intuitively say that,
+because the two dimensions are *perfectly aligned* and thus
+compound up in the second group,
+the item `r MDISC_SYM` should be the algebraic sum of both components,
+that is, $2 a_i$.
+Similarly, the correlation between two simple items items,
+each one measuring a distinct dimension,
+should be uncorrelated in the first group,
+but perfectly correlated in the second one.
+However, using an orthogonal representation
+would yield null cosines in both cases.
 
 There are several references in the literature that have found contradictory results that, we believe, are due to disregarding the latent covariance structure (i.e., using Reckase & McKinley's formulae in a non-orthonormal space).
-<!-- TODO: Examples? -->
-While we consider this is out of the scope of the current paper, we will gladly add furthere detail to explain why these results are contradictory and how our method may solve the problem.
-<!-- TODO: Add examples of contradictory results? -->
+For example, 
+<!-- # TODO: Example(s) from Gecko search -->
+We believe this is also what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution:
+As you have noted yourself [@ackerman_multidimensional_2005-1, p. 7].
+<!-- # TODO: Correctly expressed? -->
+While we consider this is out of the scope of the current paper, we will gladly add further detail to explain why these results are contradictory and how our method may solve the problem.
+<!-- # TODO: Add examples of contradictory results? -->
 
@@
 The standard errors of the multidimensional parameters are, without a doubt, more difficult to compute for the covariance-based case than for the agnostic one (if we assume the covariance structure is estimated from the model).
-Computing the covariance errors in this case requires taking into account the standard errors of the item parameters themselves and at least the standard errors of the correlations (in the best-case scenario, assuming standard latent dimensions).
-However, to our knowledge, multidimensional item parameters, as defined by @reckase_difficulty_1985 and @reckase_discriminating_1991 have been previously defined and used only for a theoretical perspective, without ever taking into account the uncertainty resulting from the model estimation.
+Computing the covariance errors in the former would require taking into account the standard errors of the item parameters themselves and at least the standard errors of the correlations (in the best-case scenario, assuming standard latent dimensions; otherwise, those of the variances and covariances).
+However, to our knowledge, multidimensional item parameters, as defined by @reckase_difficulty_1985 and @reckase_discriminating_1991 have been previously defined and used only from a theoretical perspective, without ever taking into account the uncertainty resulting from the model estimation.
 
 Regarding the "original" item parameters in the classical M2PL metric, we understand that the standard errors of the item parameters are independent of the definition of the multidimensional parameters used.
-Software such as R package {mirt} <!-- TODO: Add reference --> use a method similar to exploratory factor analysis to compute the item parameter estimtates and the standard errors, based on an oblique rotation to approximate simple structure.
-As far as we know, these standard error estimates are computed using the usual methods, involving Taylor series, bootstraping, and similar.
-<!-- TODO: Check methods; complete? -->
+Software such as R package {mirt} [@chalmers_mirt_2012] use a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
+As far as we know, these standard error estimates are computed by the usual methods, involving e.g. Taylor series, bootstrapping, or similar.
+<!-- TODO: Check SE methods; complete? -->
 
@@
 
-We think the motivation and psychometric utility of our approach are more related with the interpretation of the item parameters than with the latent abilities.
+We think the motivation and psychometric utility of our approach are more related to the interpretation of the item parameters than with the latent abilities.
 To our understanding, the better interpretation of the item parameters that results from our approach provides indeed a better insight into how the results are used.
 A multidimensional representation of the items provides benchmarks for comparing different latent trait levels, as discussed by @reckase_difficulty_1985.
-Therefore, the more accurate such representation is, the better the latent trait interpretation will be.
+Therefore, the more accurate such a representation is, the better the latent trait interpretation will be.
+
+Please see our answer in p. 3 of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
 <!-- TODO: Complete -->
@@
 On the contrary, what we propose is a way to compute and represent the multidimensional item parameters without disregarding the covariance structure of the latent space.
-Implementing the formulae we propose, using the standard output of most MIRT software, should be straightforward for most IRT researchers.
+Implementing the formulae we propose should be rather straightforward for most IRT researchers, simply using the standard output of most MIRT software.
+We have now added a discussion about this at the end of our fourth paragraph in the "Discussion" section. This addition reads:
 
-We have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, and even the submitted versions of the manuscript.
-Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software (e.g., {mirt} <!-- TODO: Add reference -->, FlexMIRT <!-- TODO: Add reference -->).
+> [...] Computing the parameters in the orthonormalized (or just orthogonalized) space,
+by contrast, would require transforming the cosines to
+the original, non-orthonormal basis afterwards.
+Nevertheless, most available MIRT software solutions,
+such as FlexMIRT <!-- TODO: Add reference --> or R package {mirt} [@chalmers_mirt_2012],
+initially estimate an uncorrelated solution and apply
+an oblique rotation afterwards, at least in exploratory solutions.
+Taking advantage of the invariance property,
+we may compute the parameters in the orthogonal space
+(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
+Then, we may compute the direction cosines in the rotated solution applying
+(the covariance-based version of) Equation \@ref(eq:dir-cos-vec-test-space).
+<!-- TODO: Number and reference cov-based version of the cosines? -->
+What a researcher should be aware of, though,
+is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
+after any non-orthonormal transformation will yield
+a $`r MDISC_AG_PARAM`$ value specific to that basis,
+instead of the invariant value of $`r MDISC_COV_PARAM`$.
+
+Notwithstanding, we have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, and even the submitted versions of the manuscript.
+Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software (e.g., {mirt} [@chalmers_mirt_2012], FlexMIRT <!-- TODO: Add reference -->).
 As the journal guidelines require a specific format for the Supplementary Material,
@@
 We clearly demarcate the limited scope of our conclusions, and warn against overgeneralizing our results to other MIRT models or the common factor model.
-We have put a great effort into writing our limitations in a way that is elegant and, at the same time, avoids cliché sentences that add little value.
-Therefore, we would rather not add a separate section for limitations.
+We have put a great effort into writing our limitations in a way that is elegant and, at the same time, avoids cliché sentences that add no value to our manuscript.
+Therefore, we would rather not add a separate "Limitations" section.
 
@@
 Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that may account for any number of dimensions.
-we have found our previous version to not be explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:
+we have found that our previous version was not explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:
 
@@
 
-<!-- # TODO: Substitute refs. by manual values -->
-> Of course,
-from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
-it is easy to express the coordinates in terms of the parameters in the original
-model formulation.
+<!-- # TODO: Confirm equation references -->
+> Of course, from Equations (24) and (25) it is easy to express
+the coordinates in terms of the parameters in the original model formulation.
 Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as
@@
 
-The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may be confusing to the reader to refer to the $MDISC_i$ and obviate the influence of $d_i$.
+The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may be confuse the reader to refer to the $MDISC_i$ and obviate the influence of $d_i$.
 We have added a clarification in this respect to the end of the paragraph, which now reads:
@@
 Although we argue indeed that the coordinate system is arbitrary, our stance is also that there is a more interpretable approach than simply assuming orthogonal axes.
-This approach implies that there is a geometric basis (and thus a coordinate system) that can be considered a faithful representation of the covariance structure.
+Such an approach implies that there is a geometric basis (and thus a coordinate system) that can be considered a faithful representation of the covariance structure.
 
@@
 We argue indeed that different samples may yield different latent covariance structures.
-This problem has received much attention in the factor analysis literature (i.e., factorial invariance), it has been largely overlooked in the MIRT literature, where any covariance structure is usually represented in orthogonal axes (implying an orthonormal basis).
+While this problem has received much attention in the factor analysis literature (i.e., factorial invariance), it has been largely overlooked in the MIRT literature, where any covariance structure is usually represented in orthogonal axes (implying an orthonormal basis).
 
@@
 
-We believe that non-orthogonal axes provide a faithful representation of the latent covariance structure:
+We argue that non-orthogonal axes provide a faithful representation of the latent covariance structure.
+We have made the following addition in the third paragraph of our introduction to reflect this fact:
+
+> [...] the assimilation of correlated latent traits to non-orthogonal
+axes is prevalent in the common factor literature
+[see, e.g., @harman_modern_1970]:
+Indeed, the cosine between any two axes (or, more generally, vectors)
+is considered as the geometrical equivalent of the correlation between
+the variables they represent. [...]
+
 In geometrical terms, the cosine between axes directly maps to the correlation between the two dimensions represented by those axes.
 In the orthogonal space, the angles between items are "distorted" and do not faithfully represent the correlation among those items.
-Consider the following example: Two simple items, each measuring a single dimension of a two-dimensional construct, will always be represented by a 90-degree angle in the orthogonal coordinate system.
-If those two dimensions are uncorrelated, so the items will also be.
-However, suppose those two dimensions are highly correlated; in that case, the items would also have to be equally highly correlated, but would still be represented by a 90-degree angle in orthogonal coordinates.
-Therefore, our argument is that the correlation among items is better represented by the cosines of their vectors in the non-orthogonal space.
+We have now added a paragraph to our introduction to try to explain this fact intuitively.
+Please refer to our answer to the editor in p. 3 of this letter for more details on this addition.
 
@@
 its covariance is an identity matrix).
-This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being the "inverse transpose" of the basis where we argue the items are faithfully represented.
+This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
+
+Our first version tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p.
+14, lines 15-22, or p.
+<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
 
@@
 
-The distances between points are given by the norm, which we argue to be invariant to changes of bases.
+The distances between points are given by the norm, which we argue to be invariant to a change of basis.
 Interestingly, when considering the covariance-based version of the multidimensional parameters, the distance between points in the latent space is given by the Mahalanobis distance.
+We have reflected this in our new version, by adding the following to the third paragraph in our Discussion:
+
+> [...] Once again, this is relevant in applications where
+potentially different bases are involved.
+Furthermore, not that, when using the covariance-based version formulation,
+the resulting norm in the latent space is the Mahalanobis distance
+[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
+thus taking the latent covariance structure into account. 
+Given these considerations, [...]
 
@@
 We apologize for the lack of clarity.
-We intended to refer to footnote 1 in page 16, we have now completed the reference.
+We intended to refer to footnote 1 in page 16; we have now completed the reference.
 
@@
 
+We think that our word choice has been unfortunate here.
 What we want to imply is that there are two different spaces, the latent space, and the test space, and that, unlike the latent traits, which "naturally lie" in the latent space, the item parameters "naturally lie in the test space", even though they can be projected onto the latent space.
-This happens even when the basis of the latent space (and, consequently, that of the test space) is orthonormal, whih we argue is overlooked (due to a lack of need to make this distinction) in Reckase's original formulations.
+This is implicit in the definition of the test space itself (see p.
+7 line 8 in our original submission, or p.
+<!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen.
+
+We argue that the test space and the latent space are two distinct spaces, even when the basis of the latent space (and, consequently, that of the test space) is orthonormal.
+It is our understanding that,  due to a lack of need for this distinction, this is overlooked in Reckase's [-@reckase_difficulty_1985] and Reckase & McKinley's [-@reckase_discriminating_1991] original formulations.
+We have added this latter clarification in our new version (p. <!-- TODO: Complete --> lines <!-- TODO: Complete -->):
+
+> [...] the latent space,
+where the person parameters are represented,
+has the inverse of the covariance (correlation) matrix as
+its inner product matrix.
+(Note that in the original derivations by @reckase_difficulty_1985
+and @reckase_discriminating_1991 both matrices simplify to the identity matrix,
+so there was no need to make this distinction.)
+This implies that the axes defined by the test space basis, [...]
 
@@
 
-As we mentioned above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance.
-One should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
-It is when we represent the items that we should use axes that "faithfully represent the covariance structure", and not when we represent the attributes.
+As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response in p.
+<!-- TODO: Complete --> in this letter).
+In this respect, we should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
+Our argument is that it is when we represent the items that we should use axes that "faithfully represent the covariance structure", and not when we represent the attributes.
 
@@
 
-<!-- # NOTE: Help needed -->
+<!-- TODO: Reply -->
 
@@
 
-To our understanding, the representation is always sample dependent, because item parameters are in turn.
+To our understanding, the representation is always sample dependent because item parameters also are.
 Our argument is that the non-orthogonal coordinate system is a more faithful representation of the latent covariance structure, as the cosines between the axes are the geometric equivalent of the latent trait correlations.
-This is, to our understanding, the value of using this more complex representation, and so we have tried to convey in (please see p. 31, line 19 to p. 32, line 4, in the original version.
+We argue that this is the value of using this more complex representation, and so we have tried to convey in <!-- # TODO: Complete --> (please see p.
+31, line 19 to p.
+32, line 4, in the original version).
 
-We acknowledge though that this may have been explained without enough clarity, and therefore strived to improve this discussion.
+We acknowledge though that this may have been explained without enough clarity, and therefore striven to improve this discussion.
 The aforementioned paragraph now reads:
@@
 the geometric equivalent of the correlations. [...]
+
+# References

### 2025-02-16 output/Main_text.Rmd
Improve responses to editor and reviewers

--- 2025-02-14 output/Main_text.Rmd
@@
 axes is prevalent in the common factor literature
-[see, e.g., @harman_modern_1970].
+[see, e.g., @harman_modern_1970]:
+Indeed, the cosine between any two axes (or, more generally, vectors)
+is considered as the geometrical equivalent of the correlation between
+the variables they represent.
 This assumption is not so easily found in the MIRT literature though,
@@
 
-The representation of a latent space where the traits are correlated is more
-accurate on a geometrical coordinate system with non-orthogonal axes
+Take the following as an example:
+A hypothetical complex item measures two latent traits with equal validity
+---i.e., its discrimination parameters are equal with value $a$.
+Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
+we would find that $`r MDISC_SYM` = \sqrt{2} a_i$.
+Assume we apply this item in two groups where, taken to the extreme,
+the two measured dimensions are orthogonal in the first group
+and perfectly correlated in the second one.
+While the previous formula seems correct in the first group,
+one can intuitively say that,
+because the two dimensions are *perfectly aligned* and thus
+compound up in the second group,
+the item `r MDISC_SYM` should be the algebraic sum of both components,
+that is, $2 a_i$.
+Similarly, the correlation between two simple items items,
+each one measuring a distinct dimension,
+should be uncorrelated in the first group,
+but perfectly correlated in the second one.
+However, using an orthogonal representation
+would yield null cosines in both cases.
+
+Therefore, the representation of a latent space where the traits are correlated
+is more accurate on a geometrical coordinate system with non-orthogonal axes
 [@ackerman_multidimensional_2005-1, p. 16, see footnote 1]
@@
 changes of latent-space basis.
-Once again, this is important in applications where
+Once again, this is relevant in applications where
 potentially different bases are involved.
+Furthermore, not that, when using the covariance-based version formulation,
+the resulting norm in the latent space is the Mahalanobis distance
+[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
+thus taking the latent covariance structure into account. 
 Given these considerations,
@@
 Such a procedure will of course yield identical results,
-given the invariance of the parameters to a change of basis.
+given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
 However, the computations in the oblique space allow us to obtain
@@
 Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require computing the cosines in
-the original basis afterwards.
+by contrast, would require transforming the cosines to
+the original, non-orthonormal basis afterwards.
+Nevertheless, most available MIRT software solutions,
+such as FlexMIRT <!-- TODO: Add reference --> or R package {mirt} [@chalmers_mirt_2012],
+initially estimate an uncorrelated solution and apply
+an oblique rotation afterwards, at least in exploratory solutions.
+Taking advantage of the invariance property,
+we may compute the parameters in the orthogonal space
+(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
+Then, we may compute the direction cosines in the rotated solution applying
+(the covariance-based version of) Equation \@ref(eq:dir-cos-vec-test-space).
+<!-- TODO: Number and reference cov-based version of the cosines? -->
+What a researcher should be aware of, though,
+is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
+after any non-orthonormal transformation will yield
+a $`r MDISC_AG_PARAM`$ value specific to that basis,
+instead of the invariant value of $`r MDISC_COV_PARAM`$.
 
@@
 its inner product matrix.
+(Note that in the original derivations by @reckase_difficulty_1985
+and @reckase_discriminating_1991 both matrices simplify to the identity matrix,
+so there was no need to make this distinction.)
 This implies that the axes defined by the test space basis,
-and not the latent space basis,
-are a geometrical representation of the covariance structure.
+and not the ones defined by the latent space basis,
+are a geometrical representation of the covariance structure:
+Indeed, the cosines of the angles among the test space axes are
+the geometric equivalent of the correlations.
 This result is both surprising and counterintuitive

### 2025-02-16 R/Formulae.R

@@ -181,24 +181,18 @@ TRAIT_NORM_EQ           <- latex_eq(
 )
 
 # Latent vector invariance condition:
-TRAIT_NORM              <- latex_norm(TRAIT_VECTOR)
-TRAIT_NORM_SQ           <- latex_squared(TRAIT_NORM)
-TRAIT_VECTOR_ORTH_TRANSP <- latex_raised_to(
-  TRAIT_VECTOR, # TODO: Transposing a "vector in basis" may typeset wrongly
-  exp = "{$LS_ORTH_BASIS$T}"
-)
+TRAIT_NORM               <- latex_norm(TRAIT_VECTOR)
+TRAIT_NORM_SQ            <- latex_squared(TRAIT_NORM)
+TRAIT_VECTOR_ORTH_TRANSP <- latex_transp(TRAIT_ORTH_COORDS, .par = TRUE)
 TRAIT_VECTOR_ORTH_SQ     <- latex(TRAIT_VECTOR_ORTH_TRANSP, TRAIT_ORTH_COORDS)
-TRAIT_COORDS_TRANSP     <- latex_raised_to(
-  TRAIT_VECTOR, # TODO: Transposing a "vector in basis" may typeset wrongly
-  exp = "{$LS_BASIS$T}"
-)
-INNER_PROD_TRANSF_DEF   <- latex(TRANSFORM_MATRIX_TRANSP, TRANSFORM_MATRIX)
-TRAIT_TRANSF_TRAIT      <- latex(
+TRAIT_COORDS_TRANSP      <- latex_transp(TRAIT_COORDS, .par = TRUE)
+INNER_PROD_TRANSF_DEF    <- latex(TRANSFORM_MATRIX_TRANSP, TRANSFORM_MATRIX)
+TRAIT_TRANSF_TRAIT       <- latex(
   TRAIT_COORDS_TRANSP,
   INNER_PROD_TRANSF_DEF,
   TRAIT_COORDS
 )
-TRAIT_NORM_SQ_EQ    <- latex_eq(
+TRAIT_NORM_SQ_EQ         <- latex_eq(
   TRAIT_NORM_SQ,
   TRAIT_VECTOR_ORTH_SQ,
   TRAIT_TRANSF_TRAIT
@@ -1190,7 +1184,7 @@ CORR_MATRIX_INV        <- latex_inverse(CORR_MATRIX)
[…]

### 2025-02-16 output/Main_text.Rmd
Create formulae for introduction example

@@
 A hypothetical complex item measures two latent traits with equal validity
----i.e., its discrimination parameters are equal with value $a$.
+---i.e., its discrimination parameters are equal with value $`r DISCR_SYM`$.
 Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
-we would find that $`r MDISC_SYM` = \sqrt{2} a_i$.
+we would find that $`r AG_MDISC_EXAMPLE`$.
 Assume we apply this item in two groups where, taken to the extreme,
@@
 the item `r MDISC_SYM` should be the algebraic sum of both components,
-that is, $2 a_i$.
+that is, $`r MDISC_VALUE_COLLIN`$.
 Similarly, the correlation between two simple items items,

### 2025-02-16 R/Formulae.R

@@ -26,6 +26,15 @@ source("R/LaTeX_math.R")
 
 FRAC_1_2 <- latex_frac(1, 2, .sep = NO_SEP)
 FRAC_1_4 <- latex_frac(1, 4, .sep = NO_SEP)
+SQRT_2   <- latex_sqrt(2)
+
+### Introduction: ----
+
+# Discrimination example:
+MDISC_SYM          <- latex("MDISC")
+DISCR_SYM          <- latex('a')
+AG_MDISC_EXAMPLE   <- latex_eq(MDISC_SYM, "$SQRT_2$ $DISCR_SYM$")
+MDISC_VALUE_COLLIN <- latex(2, DISCR_SYM)
 
 ### Latent space and variable definition ----
 
@@ -55,9 +64,9 @@ TRAIT_VEC_IN_LS <- latex_in(TRAIT_VECTOR, LATENT_SPACE)
 ITEM_INDEX          <- latex('i')
 DIM_INDEX           <- latex('k')
 INTERCEPT_PARAM     <- latex_sub('d', ITEM_INDEX)
-DISCR_VECTOR_ANY    <- latex_bf('a')
+DISCR_VECTOR_ANY    <- latex_bf(DISCR_SYM)
 DISCR_VECTOR        <- latex_sub(DISCR_VECTOR_ANY, ITEM_INDEX)
-DISCR_PARAM         <- latex_sub('a', "$ITEM_INDEX$$DIM_INDEX$")
+DISCR_PARAM         <- latex_sub(DISCR_SYM, "$ITEM_INDEX$$DIM_INDEX$")
 DISCR_VECTOR_TRANSP <- latex_transp(DISCR_VECTOR)
 
 # IRF denotation:
@@ -886,7 +895,6 @@ MATRIX_INNER_PROD_INV_EL_DIAG_EQ  <- latex_eq(
 ### Generalized multidimensional parameters ----
 
 # Parameters:
-MDISC_SYM  <- latex("MDISC")
 MDISC_ITEM <- latex_sub(MDISC_SYM, ITEM_INDEX)
 MIL_PARAM  <- latex("MIL")
 

### 2025-02-16 output/Main_text.Rmd
Correct grammar and errata from Word and Grammarly

@@
 ---a decrease in the probability of a correct response
-with an increase in ability makes little sense].
+with an increase in ability makes little sense.
 In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
@@
 Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
-express the items coordinates in terms of their multidimensional parameters.
+express the item coordinates in terms of their multidimensional parameters.
 However, it is worth noting that the latent space Gram matrix
@@
 The implication is that the item is displaced along
-its measurement direction forwards or backwards with respect to the origin.
+its measurement direction forward or backward with respect to the origin.
 As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
@@
 which, as we have argued before, implies a change of scale.
-In the case of the correlation-based version however,
+In the case of the correlation-based version, however,
 $`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
@@
 the equivalent, orthonormalized coordinates.
-Hence we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
+Hence, we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
 that allows us to represent the test space structure and the items
@@
 implying that the (alleged) three dimensions collapse into a single one, being
-thus the test practically unidimensional.
+thus the test unidimensional in practice.
 In such a case, the covariance-based direction angles
@@
 Given these considerations,
-we propose to adopt the covariance-based version of
+we propose adopting the covariance-based version of
 the multidimensional parameters as the most general definition,

### 2025-02-16 output/Letter_to_editor_reviewers.Rmd
Correct erratum

@@
 <!-- # TODO: Correctly expressed? -->
-While we consider this is out of the scope of the current paper, we will gladly add further detail to explain why these results are contradictory and how our method may solve the problem.
-<!-- # TODO: Add examples of contradictory results? -->
+While we consider this is out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our method may solve the problem.
 

### 2025-02-17 output/Letter_to_editor_reviewers.Rmd
Add proper citations to flexMIRT

@@
 Nevertheless, most available MIRT software solutions,
-such as FlexMIRT <!-- TODO: Add reference --> or R package {mirt} [@chalmers_mirt_2012],
+such as flexMIRT [@cai_flexmirttm_2024] or R package {mirt} [@chalmers_mirt_2012],
 initially estimate an uncorrelated solution and apply
@@
 Notwithstanding, we have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, and even the submitted versions of the manuscript.
-Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software (e.g., {mirt} [@chalmers_mirt_2012], FlexMIRT <!-- TODO: Add reference -->).
+Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software [e.g., {mirt}, @chalmers_mirt_2012; flexMIRT, @cai_flexmirttm_2024].
 As the journal guidelines require a specific format for the Supplementary Material,

### 2025-02-17 output/Letter_to_editor_reviewers.Rmd
Correct "Trademark" symbol in flexMIRT .bib entry

@@
 Nevertheless, most available MIRT software solutions,
-such as flexMIRT [@cai_flexmirttm_2024] or R package {mirt} [@chalmers_mirt_2012],
+such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
 initially estimate an uncorrelated solution and apply
@@
 Notwithstanding, we have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, and even the submitted versions of the manuscript.
-Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software [e.g., {mirt}, @chalmers_mirt_2012; flexMIRT, @cai_flexmirttm_2024].
+Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software [e.g., {mirt}, @chalmers_mirt_2012; flexMIRT, @cai_flexmirt_2024].
 As the journal guidelines require a specific format for the Supplementary Material,

### 2025-02-17 output/Letter_to_editor_reviewers.Rmd
Correct question styles, using one line per item

@@
 ###### I have a concern with making the latent ability axes nonorthogonal to represent the covariance structure between the latent abilities.
-###### Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.
+Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.
 
@@
 ###### Like Reviewer 3, I do not understand how the magnitude of discrimination values (p 29 , 7) can be compared/interpreted values of discrimination are comparable from the two representations.
-###### It seems that the metric of the represented space has changed, and equivalent numbers will not have equivalent interpretations.
+It seems that the metric of the represented space has changed, and equivalent numbers will not have equivalent interpretations.
 
@@
 ###### I would like to see the motivation and psychometric utility of this approach.
-###### Does it improve the interpretation of the latent abilities and thus result in more insight into how the results are used?
-###### The author needs to address this with legitimate examples for testing practitioners.
+Does it improve the interpretation of the latent abilities and thus result in more insight into how the results are used?
+The author needs to address this with legitimate examples for testing practitioners.
 
@@
 ###### While the manuscript includes a graphical representation example with two-dimensional latent variables, real-world applications often involve higher dimensions (e.g., the Big Five personality traits or Cattell’s 16 personality factors).
-###### If the current results cannot be applied to the high-dimensional cases, would that significantly limit their practical value and applicability?
+If the current results cannot be applied to the high-dimensional cases, would that significantly limit their practical value and applicability?
 
@@
 
-###### 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect.
-###### It should be “$\mathcal{C}$”.
+###### 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect. It should be “$\mathcal{C}$”.
 
@@
 
-###### 2. The notations in the manuscript lack clarity, particularly in Eq. (19) on page 12.
-###### Based on Eq. (5) and Eq. (15), $\mathbf{a}_i$ in the numerator of Eq. (19) should be $\mathbf{a}_i^{\mathcal{B}*}$.
+###### 2.
+The notations in the manuscript lack clarity, particularly in Eq.
+(19) on page 12.
+Based on Eq.
+(5) and Eq.
+(15), $\mathbf{a}_i$ in the numerator of Eq.
+(19) should be $\mathbf{a}_i^{\mathcal{B}*}$.
 
@@
 
-###### 3. In the second graph on page 18, the authors state, “*That is, the less discriminating the item, the further it will be from the origin.*”
-###### However, the value of $D_i$ is influenced not only by the $MDISC_i$ parameter but also by $d_i$.
-###### For instance, in Table 1 on page 20, $d_i$ for item 1 is zero, meaning that regardless of $MDISC_i$, $D_i$ remains zero.
+###### 3.
+In the second graph on page 18, the authors state, “*That is, the less discriminating the item, the further it will be from the origin.*” However, the value of $D_i$ is influenced not only by the $MDISC_i$ parameter but also by $d_i$.
+For instance, in Table 1 on page 20, $d_i$ for item 1 is zero, meaning that regardless of $MDISC_i$, $D_i$ remains zero.
 
@@
 ###### [...] I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes.
-###### The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals.
-###### How are these features interpreted when the coordinate axes are non-orthogonal?
+The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals.
+How are these features interpreted when the coordinate axes are non-orthogonal?
 
@@
 ###### Care needs to be taken when using nonorthogonal coordinates that the distances between points in the space are correctly represented.
-###### What formula should be used for the distance between two points in each case?
+What formula should be used for the distance between two points in each case?
 
@@
 
-###### Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005).
-###### It would help if a page number were provided for the discussion.
+###### Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005). It would help if a page number were provided for the discussion.
 
@@
 ###### I also wonder about the Harman (1970) reference.
-###### I know that he shows a plot of length of arm against length of leg using orthogonal axes for the plot even though the two measures are correlated.
-###### I am not not sure how the axes should be oriented to reflect the correlation between the two measures.
+I know that he shows a plot of length of arm against length of leg using orthogonal axes for the plot even though the two measures are correlated.
+I am not not sure how the axes should be oriented to reflect the correlation between the two measures.
 
@@
 
-###### Page 6  Line 21 The expression on this line needs more explanation.
-###### I am not sure what the expression is supposed to represent.
+###### Page 6  Line 21 The expression on this line needs more explanation. I am not sure what the expression is supposed to represent.
 
@@
 ###### Page 7  Line 24 The meaning of the superscripts is not clear in this expression.
-###### What is the source of the -1T?  Is that the transpose of the inverse?
-###### Do the -1 superscripts indicate an inverse?
+What is the source of the -1T?
+Is that the transpose of the inverse?
+Do the -1 superscripts indicate an inverse?
 
@@
 ###### Page 13 Line 8 The statement “the test space is where the item parameters naturally lie” seems to imply there is a correct representation of the item parameters.
-###### But, the correlations between constructs changes with the sample of examinees and the coordinates can easily be rotated and translated giving different representations of the item parameters.
-###### I would like to see support for the statement.
+But, the correlations between constructs changes with the sample of examinees and the coordinates can easily be rotated and translated giving different representations of the item parameters.
+I would like to see support for the statement.
 
@@
 ###### Page 19 Line 8 There is an interesting distinction that seems to be lost here.
-###### Attributes can be measured independently, such as height and weight and yield correlated results.
-###### Under those circumstances, should axes be represented as orthogonal, or should the axes represent the correlation.
-###### This has an implication for the formula that is used to compute the distance between points in the space.
+Attributes can be measured independently, such as height and weight and yield correlated results.
+Under those circumstances, should axes be represented as orthogonal, or should the axes represent the correlation.
+This has an implication for the formula that is used to compute the distance between points in the space.
 
@@
 ###### Page 22 Figure 1 It seems that both of these plots are shown with orthogonal coordinate axes.
-###### But the orthogonal axes have changed so that the nonorthogonal axes can be shown in the orthogonal representation.
+But the orthogonal axes have changed so that the nonorthogonal axes can be shown in the orthogonal representation.
 
@@
 ###### Page 29 Line 7 I am curious if the values of discrimination are comparable from the two representations.
-###### It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.
+It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.
 
@@
 ###### Page 30 Line 8 It would be helpful to show the orthogonal and nonorthogonal representations of items and then discuss the interpretation of the parameter estimates based on each of the representations.
-###### I have been unable to determine the value of using the nonorthogonal representation.
-###### The nonorthogonal representation is more complex, and it seems sample dependent because it uses the observed correlations/covariances.
-###### What is the value added of using the more complex representation?
+I have been unable to determine the value of using the nonorthogonal representation.
+The nonorthogonal representation is more complex, and it seems sample dependent because it uses the observed correlations/covariances.
+What is the value added of using the more complex representation?
 

### 2025-02-17 output/Main_text.Rmd
Correct errata

@@
 potentially different bases are involved.
-Furthermore, not that, when using the covariance-based version formulation,
+Furthermore, note that, when using the covariance-based version formulation,
 the resulting norm in the latent space is the Mahalanobis distance
@@
 its inner product matrix.
-(Note that in the original derivations by @reckase_difficulty_1985
-and @reckase_discriminating_1991 both matrices simplify to the identity matrix,
-so there was no need to make this distinction.)
+[Note that in the original derivations by @reckase_difficulty_1985;
+and @reckase_discriminating_1991, both matrices simplify to the identity matrix,
+so there was no need to make this distinction.]
 This implies that the axes defined by the test space basis,

### 2025-02-17 output/Main_text.Rmd
Update Harman reference to 3rd edition

@@
 axes is prevalent in the common factor literature
-[see, e.g., @harman_modern_1970]:
+[see, e.g., @harman_modern_1976]:
 Indeed, the cosine between any two axes (or, more generally, vectors)
@@
 and (possibly) non-standard units,
-often referred to as *general Cartesian coordinates* [@harman_modern_1970].
+often referred to as *general Cartesian coordinates*
+[@harman_modern_1976, p. 60].
 Such a space has the advantage that the basis vectors and the axes they yield
@@
 we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
-it only rotates the vertical axis [@harman_modern_1970]:
+it only rotates the vertical axis [@harman_modern_1976]:
 
@@
 between the *primary factors* and Thurstone's *reference axes*
-[@harman_modern_1970, see p. 291].
+[@harman_modern_1976, see p. 276].
 Nevertheless, one must not make the mistake of assimilating the test space and

### 2025-02-21 notebooks/Graphical_representation_alternatives.Rmd

@@
 ```{r setup}
+#| echo: !expr 'knitr::is_html_output()'
 library(knitr)
@@
 # Output configuration options:
-options(digits = 3)
+options(digits = 2)
 ```

### 2025-02-21 notebooks/example_script.Rmd

@@
 library(pracma)
-library(matlib)
 library(rgl)
 set.seed(123)
+
+source("R/Math.R")
 ```
@@
 
-Estos vectores tienen norma 1 y un ángulo entre ellos de `r round(matlib::angle(Base_1[,1],Base_1[,2]),2)` grados
+Estos vectores tienen norma 1 y un ángulo entre ellos de `r round(angle(Base_1[,1],Base_1[,2]),2)` grados
 
@@
 ```{r angulos, include=FALSE}
-alfa_1_1=matlib::angle(Base_1[,1],Base_C[,1])
-alfa_1_2=matlib::angle(Base_1[,1],Base_C[,2])
-alfa_2_1=matlib::angle(Base_1[,2],Base_C[,1])
-alfa_2_2=matlib::angle(Base_1[,2],Base_C[,2])
+alfa_1_1=angle(Base_1[,1],Base_C[,1])
+alfa_1_2=angle(Base_1[,1],Base_C[,2])
+alfa_2_1=angle(Base_1[,2],Base_C[,1])
+alfa_2_2=angle(Base_1[,2],Base_C[,2])
 ```
@@
 ```{r , echo=FALSE, include=TRUE}
-print(matlib::angle(Q[,1],Q[,2]))
+print(angle(Q[,1],Q[,2]))
 ```

### 2025-02-21 output/Cover_letter.Rmd

@@
 
-September 5, 2024.
+September 16, 2024.
 
-George A. Marcoulides \
-Editor, *Educational and Psychological Measurement* \
-University of California, Santa Barbara, USA
+<br>
 
-Dear Dr. Marcoulides:
+Sandip Sinharay, PhD \
+Editor, *Psychometrika* \
+Educational Testing Service, USA
+
+<br>
+
+Dear Dr. Sinharay:
 
@@
 which I am submitting for exclusive consideration of publication as an article
-in Educational and Psychological Measurement.
+in Psychometrika.
 Excluding the title page
 (submitted separately according to Sage's guidelines),
-the manuscript is 24 pages long and includes one figure and three tables.
+the manuscript is 35 pages long and includes one figure and three tables.
 
@@
 more importantly,
-to a misconception of how these parameters should be computed and interpreted.
+to a misconception about their computation and interpretation in applications
+with non-identity covariance latent matrices.
 In the course of our research we have also found a crucial result regarding
@@
 
+<br>
+
 Sincerely,
@@
 Department of Behavioral Science Methodology \
-Universidad Nacional de Educación a Distancia \
+Universidad Nacional de Educación a Distancia (UNED) \
 C. Juan del Rosal 10, Office 2.74 \

### 2025-02-21 output/ESM_1.Rmd

@@
 
-# References
-
-::: {#refs}
-:::
-
 \newpage
@@
 #|   Agnostic and covariance-based multidimensional item parameters in
-#|   Reckase (2009), with an almost-rank-1 covariance matrix.
+#|   Reckase (2009), with an almost-rank-1 covariance matrix
 item_params_output
 ```
+
+# References

### 2025-02-21 output/Main_text.Rmd

--- 2025-02-17 output/Main_text.Rmd
@@
 axes is prevalent in the common factor literature
-[see, e.g., @harman_modern_1970]:
-Indeed, the cosine between any two axes (or, more generally, vectors)
-is considered as the geometrical equivalent of the correlation between
-the variables they represent.
+[see, e.g., @harman_modern_1970].
 This assumption is not so easily found in the MIRT literature though,
@@
 
-Take the following as an example:
-A hypothetical complex item measures two latent traits with equal validity
----i.e., its discrimination parameters are equal with value $`r DISCR_SYM`$.
-Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
-we would find that $`r AG_MDISC_EXAMPLE`$.
-Assume we apply this item in two groups where, taken to the extreme,
-the two measured dimensions are orthogonal in the first group
-and perfectly correlated in the second one.
-While the previous formula seems correct in the first group,
-one can intuitively say that,
-because the two dimensions are *perfectly aligned* and thus
-compound up in the second group,
-the item `r MDISC_SYM` should be the algebraic sum of both components,
-that is, $`r MDISC_VALUE_COLLIN`$.
-Similarly, the correlation between two simple items items,
-each one measuring a distinct dimension,
-should be uncorrelated in the first group,
-but perfectly correlated in the second one.
-However, using an orthogonal representation
-would yield null cosines in both cases.
-
-Therefore, the representation of a latent space where the traits are correlated
-is more accurate on a geometrical coordinate system with non-orthogonal axes
-[@ackerman_multidimensional_2005-1, p. 16, see footnote 1]
-and (possibly) non-standard units,
+The representation of a latent space where the traits are correlated is more
+accurate on a geometrical coordinate system with non-orthogonal axes
+[@ackerman_multidimensional_2005-1] and (possibly) non-standard units,
 often referred to as *general Cartesian coordinates* [@harman_modern_1970].
@@
 "are constrained to be positive" [@ackerman_graphical_1996, p. 315],
-or that "in an [independent-clusters] solution, the loadings can always
-be selected to be positive" [@mcdonald_basis_2000, p. 110].
+or that “in an [independent-clusters] solution, the loadings can always
+be selected to be positive” [@mcdonald_basis_2000, p. 110].
 However, when such instances of negative parameters appear,
-they are often dismissed as "particularly puzzling" (p. 109),
+they are often dismissed as “particularly puzzling” (p. 109),
 or implicitly assuming item malfunctioning without further discussion
@@
 and present two examples of application of our results.
-These examples, from the cognitive and the non-cognitive domain respectively,
-are taken from the MIRT literature:
-One of them is a classic dataset, often used to exemplify
-the use of the `r MODEL_ACRONYM` model [@reckase_multidimensional_2009];
-the other one, an application in marketing research [@tezza_modelo_2018].
 We round up with a discussion about
@@
 
-Consider a basis $`r LS_BASIS_EQ`$,
-with $`r LS_BASIS`$ generally non-orthonormal.
-Let $`r LATENT_SPACE`$ be a $`r N_DIMS`$-dimensional latent space
-in basis $`r LS_BASIS`$ (with $`r N_DIMS`$ any positive integer),
+Consider a $`r N_DIMS`$-dimensional latent space $`r LATENT_SPACE`$
+in basis $`r LS_BASIS_EQ`$, being $`r LS_BASIS`$ not necessarily orthonormal,
 where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
@@
 ---a decrease in the probability of a correct response
-with an increase in ability makes little sense.
+with an increase in ability makes little sense].
 In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
@@
 the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
-Third, as responses are not necessarily interpreted as correct or incorrect,
+Finally, as responses are not necessarily interpreted as correct or incorrect,
 parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
 instead of the item difficulty.
-Finally, it is worth highlighting that
-this model generally assumes *within-item multidimensionality*,
-which will be present whenever more than one component of $`r DISCR_VECTOR`$
-is non-null.
-However, the model can also account for *between-item multidimensionality*,
-if the $`r DISCR_VECTOR`$ parameter only has one non-null component.
 
@@
 respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
-That is, vector $`r TRAIT_VECTOR`$ is represented in each of the two bases by
-the linear combination of its element vectors
-scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
 
@@
 Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
-express the item coordinates in terms of their multidimensional parameters.
+express the items coordinates in terms of their multidimensional parameters.
 However, it is worth noting that the latent space Gram matrix
@@
 it is easy to express the coordinates in terms of the parameters in the original
-model formulation.
-Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as
+model formulation, namely,
 
@@
 The implication is that the item is displaced along
-its measurement direction forward or backward with respect to the origin.
+its measurement direction forwards or backwards with respect to the origin.
 As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
@@
 the further away it will be from the origin.
-Of course, it is also directly proportional to
-the intercept parameter $`r INTERCEPT_PARAM`$.
-This implies that, if $`r INTERCEPT_NULL_EQ`$, $`r DISTANCE_NULL_EQ`$ as well,
-regardless of the discrimination.
 
@@
 which, as we have argued before, implies a change of scale.
-In the case of the correlation-based version, however,
+In the case of the correlation-based version however,
 $`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
@@
 the equivalent, orthonormalized coordinates.
-Hence, we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
+Hence we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
 that allows us to represent the test space structure and the items
@@
 implying that the (alleged) three dimensions collapse into a single one, being
-thus the test unidimensional in practice.
+thus the test practically unidimensional.
 In such a case, the covariance-based direction angles
@@
 changes of latent-space basis.
-Once again, this is relevant in applications where
+Once again, this is important in applications where
 potentially different bases are involved.
-Furthermore, not that, when using the covariance-based version formulation,
-the resulting norm in the latent space is the Mahalanobis distance
-[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
-thus taking the latent covariance structure into account. 
 Given these considerations,
-we propose adopting the covariance-based version of
+we propose to adopt the covariance-based version of
 the multidimensional parameters as the most general definition,
@@
 Such a procedure will of course yield identical results,
-given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
+given the invariance of the parameters to a change of basis.
 However, the computations in the oblique space allow us to obtain
@@
 Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require transforming the cosines to
-the original, non-orthonormal basis afterwards.
-Nevertheless, most available MIRT software solutions,
-such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
-initially estimate an uncorrelated solution and apply
-an oblique rotation afterwards, at least in exploratory solutions.
-Taking advantage of the invariance property,
-we may compute the parameters in the orthogonal space
-(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
-Then, we may compute the direction cosines in the rotated solution applying
-(the covariance-based version of) Equation \@ref(eq:dir-cos-vec-test-space).
-<!-- TODO: Number and reference cov-based version of the cosines? -->
-What a researcher should be aware of, though,
-is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
-after any non-orthonormal transformation will yield
-a $`r MDISC_AG_PARAM`$ value specific to that basis,
-instead of the invariant value of $`r MDISC_COV_PARAM`$.
+by contrast, would require computing the cosines in
+the original basis afterwards.
 
@@
 its inner product matrix.
-(Note that in the original derivations by @reckase_difficulty_1985
-and @reckase_discriminating_1991 both matrices simplify to the identity matrix,
-so there was no need to make this distinction.)
 This implies that the axes defined by the test space basis,
-and not the ones defined by the latent space basis,
-are a geometrical representation of the covariance structure:
-Indeed, the cosines of the angles among the test space axes are
-the geometric equivalent of the correlations.
+and not the latent space basis,
+are a geometrical representation of the covariance structure.
 This result is both surprising and counterintuitive

### 2025-02-21 output/sandbox_slideshow.qmd

--- 2024-12-20 output/sandbox_slideshow.qmd
@@
 ---
+title:     |
+  Sandbox slide
 bibliography:  ../www/Multidimensional-parameters-MCLM.bib
 csl:           ../www/apa-old-doi-prefix.csl
+date:          "2024-10-21"
 date-meta:     "2024-11-27"
+date-format:   long
 editor:
@@
 knitr:
+  # opts_knit: # Does not work, see line 88
+  #   root_dir:  here::here()
   opts_chunk: 
@@
     fig-width:               6
+    incremental:             false
     keep-md:                 true
@@
     theme:                   ../www/slideshow-assets/extra-styles.scss
+    transition:              none
+    view-distance:           3
+    template-partials:
+      - ../www/slideshow-assets/title-slide.html
 editor_options: 
@@
 ```{r pre}
-PROB_AXIS    <- seq(-3, 3, by = .01)
-ERROR_POINTS <- seq(-3, 3, length.out = 5)
 PROB_AXIS_2D <- seq(-3, 3, by = .1)
@@
 
-ITEMS_PALETTE <- c(
-  PALETTE_BLUE["normal"],
-  PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"],
-  PALETTE_TANGERINE["normal"]
-) |>
-  unname() |>
-  setNames(paste(1:4, sep = '_'))
 ```

### 2025-02-21 output/IMPS-2023_proposal/IMPS-2025_abstract.Rmd
Version IMPS 2025 form responses

---
title: IMPS 2025 proposal
author: "Daniel Morillo, Mario Luzardo"
date: "2023-02-27"
output: word_document
bibliography: ../../www/Multidimensional-parameters-MCLM.bib
csl:          ../../www/apa-old-doi-prefix.csl
---

```{r setup}
#| include: false

# Includes:

library(knitr)
library(here)

ROOT_DIR <- here()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE
)
```

```{r sources}
#| cache:   false
#| include: false
source("R/Constants.R", encoding = 'UTF-8')
```

# Format

## Which presentation format would you prefer?

- [X] Oral

- [ ] Poster

- [ ] Symposia

- [ ] Symposium Overview

# Title & Abstract

## Title

A generalization of multidimensional item response theory parameters

<!-- # 8/10 words -->

## Abstract

The Multidimensional 2-Parameter Logistic (`r MODEL_ACRONYM`, McKinley &
Reckase, 1983) model is one of the most important and widely used models in
Multidimensional Item Response Theory (MIRT).
Reckase (1985) and Reckase & McKinley (1991) gave a precise definition of a
Multidimensional Item Difficulty (MID) and Multidimensional Discrimination
(MDISC) parameter, respectively, and derived their expressions for the M2PL
model.
However, we argue that their derivations (1) only consider the case of items
with conditional monotonously increasing probability of a positive response, and
(2) implicitly assume that the items are represented in an orthogonal basis.

We argue that these assumptions, which restrict the application of the model to
the assessment of non-cognitive traits, can be relaxed.
We thus extend the expressions to the general case of unchanging monotony items
in any non-orthonormal space, which allows to properly represent disparate
covariance structures.
We then use these definitions in two applications: to represent M2PL items in
non-orthonormal bases, and to compute the multidimensional parameters in various
applications, showing how they provide better properties compared with their
"covariance-agnostic" counterparts.

These results become increasingly relevant as the modern measurement theory in
social and behavioral sciences experiences as a convergence between MIRT and the
common factor model.
Importantly, our derivations show that the space where the items are represented
is necessarily distinct from the latent space.
Nevertheless, we warn against generalizing our results to
other MIRT models (i.e. beyond the M2PL), without carefully considering each
particular case.

<!-- # 239/250 words -->

# Authors

## Is this a student submission?

- [ ] Yes

- [X] No

## Please nominate all the authors that contributed to this submission below.

Daniel Morillo

danielmorillo.ac@gmail.com

### Title

Dr

### Affiliation

Universidad Nacional de Educación a Distancia (UNED)

### Country

Spain

## Add another author

### Please type the new author's email

mluzardov@gmail.com

### Title

Dr

### First Name

Mario

### Last Name

Luzardo-Verde

### Affiliation

Universidad de la República (UdelaR)

### Country

Uruguay

# Topics

Please choose a maximum of three topics from the list below.

- [ ] Applications
- [ ] Artificial Intelligence Applications
- [ ] Bayesian Statistics
- [ ] Categorical Data Analysis
- [ ] Causal Inference and Mediation
- [ ] Classical Test Theory
- [ ] Classification
- [ ] Clustering
- [ ] Cognitive Diagnostic Models
- [ ] Conceptual and Philosophical Issues
- [ ] Computational modeling
- [ ] Differential Item Functioning
- [ ] Dynamical Systems Modeling
- [ ] Computer-based Testing
- [ ] Estimation and Computational Methods
- [ ] Factor Analysis
- [ ] Generalized Linear Models
- [ ] Graphical Models
- [X] Item Response Theory
- [ ] Latent Class Models
- [ ] Longitudinal Data Analysis
- [ ] Measurement Invariance and DIF
- [ ] Missing Data
- [X] Multidimensional Analysis
- [ ] Multilevel/Hierarchical/Mixed Models
- [ ] Multivariate Analysis
- [ ] Social Network Analysis
- [ ] Prediction
- [ ] Network Psychometrics
- [ ] Optimal/Dual Scaling
- [ ] Principal Components/Correspondence Analysis
- [ ] Patient-Reported Outcomes
- [ ] Statistical and Machine Learning
- [ ] Test Equating
- [ ] Time Series Analysis
- [ ] Validity and Reliability

# Additional Information

Links to author's websites.

Please provide a link to any professional websites you would like to share in the app.

# Application for Student Travel Award

(NA)

Proceedings

- [ ] I plan to submit a chapter based on my presentation for the 2025 IMPS Proceedings volume.

<!-- # TODO: Decide -->

# References

### 2025-02-21 output/seminar-slideshow.qmd
Add pending refs. in oblique coordinates slide
Improve representation of item in transformed axes

--- 2025-02-14 output/seminar-slideshow.qmd
@@
     limits       = c(-0.75, 1.5),
-    breaks       = (-1:3)/2,
+    breaks       = (-1:2)/2,
     minor_breaks = NULL,
@@
       &= \sqrt{ 0.25 + 1 } \\
-      &= \sqrt{ 1.25 } \\
-      &\approx 1.12
+      &= \sqrt{ 1.25 } \approx 1.12
   \end{align}
@@
     breaks       = (-2:2)/2 + .375,
-    labels       = ~. - .375,
+    labels       = ~number(. - .375, .1),
     minor_breaks = NULL,
@@
     "richtext",
-    x = c(0.5, -0.1), y = c(-0.1, 1),
-    label         = c("*a*^*^~11~", "*a*^*^~21~"),
+    x = c(1, -0.1), y = c(-0.1, 1),
+    label         = c("*a*~11~", "*a*~21~"),
     size          = 8,
@@
 No tien sentido que cambie su discriminación multidimensional, y sin embargo la literatura está llena de ejemplos que calculan este parámetro con la fórmula de Reckase (en un espacio de "rasgos correlacionados", que en seguida veremos por qué eso es importante), y que no se preocupan por "los ejes" que representan esos rasgos de manera fidedigna.
-<!-- # TODO: Ejemplos de la literatura? -->
-:::
-
-## Generalización: "Coordenadas oblicuas"
-
-::: notes
+:::
+
+## Generalización: Coordenadas oblicuas
+
+<!-- TODO: ACT en McDonald, 1999? -->
+
+::: notes
+<!-- # TODO: Corregir referencia de Ackerman? -->
 Para la representación gráfica esto puede valer, como dice Ackerman aquí.

### 2025-02-21 output/seminar-slideshow.qmd
Add explanation of "arbitrary" angles in MIRT
Correct MDISC expression in transformed basis
Put aside orthogonal coordinates example slide

@@
 
-## (Pero, un inciso...)
-
-![Un cubo rojo en perspectiva isométrica](../www/slideshow-assets/isometric-cube.jpg){height="600" fig-align="center"}
+## (Pero, un inciso...) {.smaller}
+
+::: columns
+::: column
+![Un cubo rojo en perspectiva isométrica](../www/slideshow-assets/isometric-cube.jpg){fig-align="center"}
+:::
+
+::: {.column .fragment}
+<br>
+
+> It has been suggested by some researchers that the angle between the axes represents the degree of correlation.
+> However, for sake of clarity, an orthogonal axes system is used in which distance measures and vectors can be easily calculated, understood, and interpreted.
+
+---@ackerman_multidimensional_2005-1 [p. 16]
+:::
+:::
 
@@
   \mathbf{Q} = \begin{bmatrix}
+      1 & -0.5 \\
+      0 & 1
+    \end{bmatrix}; \quad
+  \mathbf{Q}^{-1} = \begin{bmatrix}
       1 & 0.5 \\
@@
     \mathbf{a}_1 =
-    \mathbf{Q} \mathbf{a}^*_1 &= \begin{bmatrix}
+    \mathbf{Q}^{-1} \mathbf{a}^*_1 &= \begin{bmatrix}
       1 & 0.5 \\
@@
 $$
-  \begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
+  \mathbf{a}_1 = \begin{bmatrix}
     1 \\
     1
-      \end{bmatrix}\\
-      d_1 &= 0
-  \end{align}
+  \end{bmatrix};\quad
+  d_1 = 0
 $$
@@
   \begin{align}
-    MDISC^_1 &= \sqrt{
+    MDISC_1 &= \sqrt{
         [1, 1]
@@
       &= \sqrt{ 1 + 1 } \\
-      &= \sqrt{ 2 } \\
-      &\approx 1.41
+      &= \sqrt{ 2 } \approx 1.41
   \end{align}
@@
 Esta propiedad se llama "invarianza" del ítem.
-
-No tien sentido que cambie su discriminación multidimensional, y sin embargo la literatura está llena de ejemplos que calculan este parámetro con la fórmula de Reckase (en un espacio de "rasgos correlacionados", que en seguida veremos por qué eso es importante), y que no se preocupan por "los ejes" que representan esos rasgos de manera fidedigna.
-:::
-
-## Generalización: Coordenadas oblicuas
-
-<!-- TODO: ACT en McDonald, 1999? -->
-
-::: notes
-<!-- # TODO: Corregir referencia de Ackerman? -->
-Para la representación gráfica esto puede valer, como dice Ackerman aquí.
-Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente.
 :::
@@
 Y esto nos ha permitido "descubrir" (o darnos cuenta, más bien) de que hay dos "espacios vectoriales" distintos, uno en el que se encuentran las coordenadas de las puntuaciones factoriales, y otro distinto, que es el espacio en el que se encuentran las coordenadas de los ítems, y que es donde en realidad hay que representarlos.
+:::
+
+## Supuestos: Ortogonalidad
+
+<!-- # TODO: Review/delete slide? -->
+
+<!-- TODO: ACT en McDonald, 1999? -->
+
+::: r-center
+![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height=400px}
+:::
+
+::: aside
+Figura tomada de @ackerman_multidimensional_2005-1
+:::
+
+::: notes
+No tiene sentido que cambie su discriminación multidimensional, y sin embargo la literatura está llena de ejemplos que calculan este parámetro con la fórmula de Reckase (en un espacio de "rasgos correlacionados", que en seguida veremos por qué eso es importante), y que no se preocupan por "los ejes" que representan esos rasgos de manera fidedigna.
+
+Para la representación gráfica esto puede valer, como dice Ackerman aquí.
+Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente; obtenemos parámetros de discriminación multidimensional "distorsionados".
 :::

### 2025-02-21 output/seminar-slideshow.qmd
Cite Ackerman for orthogonal MDISC computation
Fit "oblique axes" title in one line

@@
 
-<br>
+::: fragment
+> Using orthogonal axes, discrimination corresponds
+to the length of the item response vector
+
+---@ackerman_multidimensional_2005 [p. 6]
+:::
 
@@
 
-## Generalización: Coordenadas oblicuas
-
-> before performing the differentiation, the constraint that <!-- # TODO: Sum of cosines^2 = 1 --> is added to the expression for the slope \[...\]
+## Generalización: Ejes oblicuos {.smaller}
+
+> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
 
@@
 Si os acordáis, Reckase aplicaba esta restricción de que la suma de cosenos directores era 1 (que se deriva del teorema de Pitágoras).
-Pero para aplicarla hace falta asumir que esos cosenos directores son todos ortogonales entre sí.
+Pero para aplicarla hace falta asumir que esos cosenos directores son todos ortogonales entre sí, como ya hemos visto.
 Si no tenemos ángulos rectos, esa condición no se cumple.
 
-Y lo que nosotros estamos asumiendo que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones ($R$, la que sea).
-Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a coordenadas oblicuas tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el o y el 1 pueda ser distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
+[NEXT]
+
+Y lo que nosotros estamos asumiendo es que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones $R$ (la que sea).
+Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a Ejes oblicuos tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el o y el 1 pueda ser distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
 
@@
 
-## Generalización: Coordenadas oblicuas
+## Generalización: Ejes oblicuos
 
@@
 
--   A coordenadas oblicuas
+-   A Ejes oblicuos
 
@@
 
--   A coordenadas oblicuas, las cuales consideramos que son necesarias para representar distribuciones de puntuaciones de rasgo correlacionadas, en general.
+-   A Ejes oblicuos, las cuales consideramos que son necesarias para representar distribuciones de puntuaciones de rasgo correlacionadas, en general.
     Hemos visto que esto ya lo mencionaron Zhang y Stout, hace 25 años, pero no proporcionaron una "deducción" de estos parámetros multidimensionales, cosa que nosotros hacemos.

### 2025-02-21 output/seminar-slideshow.qmd
Add "orthogonalization" slide

@@
 library(scales)
+library(mvtnorm)
 ```
@@
 
-<!-- # TODO: Expresión de las \thetas transformadas -->
-
-<!-- # TODO: Distribución de \thetas en ejes ortogonales -->
+:::::::::::: columns
+::::::: column
+::: {.fragment fragment-index="1"}
+$$
+  cor (\mathbf{\theta}) = \mathbf{R}
+$$
+:::
+
+::: {.fragment fragment-index="2"}
+$$
+  \mathbf{R} = \begin{bmatrix}
+      1 & 0.5 \\
+      0.5 & 1
+    \end{bmatrix}
+$$
+:::
+
+::: {.fragment fragment-index="3"}
+$$
+  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
+$$
+:::
+
+::: {.fragment fragment-index="4"}
+$$
+  cor (\mathbf{\theta}^*) =
+    \mathbf{R}^* =
+    \begin{bmatrix}
+      1 & 0 \\
+      0 & 1
+    \end{bmatrix}
+$$
+:::
+:::::::
+
+:::::: column
+::::: r-stack
+::: {.fragment fragment-index="2"}
+```{r corr-traits-example}
+#| fig-width:  5.5
+#| fig-height: 5.5
+
+set.seed(7117)
+
+corr_matrix <- matrix(c(1, .5, .5, 1), nrow = 2)
+
+thetas <- rmvnorm(10^4, sigma = corr_matrix) |>
+  as_tibble(.name_repair = ~paste0("trait_", 1:2))
+
+thetas |>
+  ggplot(mapping = aes(trait_1, trait_2)) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks       = element_blank(),
+    axis.line        = element_blank(),
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
+  )
+```
+:::
+
+::: {.fragment .fade-in fragment-index="4"}
+```{r corr-traits-transf-example}
+#| fig-width:  5.5
+#| fig-height: 5.5
+
+set.seed(7117)
+
+# This is a square root of the correlation matrix, then inverted and transposed:
+orth_matrix <- c(1, 0, .5, sqrt(.75)) |> matrix(nrow = 2) |> solve() |> t()
+
+thetas_transf <- thetas |>
+  rownames_to_column("i") |>
+  pivot_longer(starts_with("trait")) |>
+  group_by(i) |>
+  mutate(value = orth_matrix %*% value |> drop()) |>
+  pivot_wider()
+
+thetas_transf |>
+  ggplot(mapping = aes(trait_1, trait_2)) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*^*^~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*^*^~1~", labels = NULL)
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks       = element_blank(),
+    axis.line        = element_blank(),
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
+  )
+```
+:::
+:::::
+::::::
+::::::::::::
 
@@
 
-[NEXT]
+\[NEXT\]
 
 Y lo que nosotros estamos asumiendo es que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones $R$ (la que sea).
-Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a Ejes oblicuos tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el o y el 1 pueda ser distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
+Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a ejes oblicuos tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no" estandarizadas", es decir, que la distancia entre el 0 y el 1 pueda ser "distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
 
 ¿Cómo podríamos solucionar esto?
-Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de este espacio oblicuo (el que sea, aún no sabemos cuál es) a un espacio ortogonal.
+Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de ese espacio oblicuo (el que sea, aún no sabemos cuál es) a un espacio ortogonal.
 :::

### 2025-02-21 output/seminar-slideshow.qmd
Apply canonical markdown

@@
 ::: notes
-Lo mismo pasa al representar los ejes de una rotación oblicua:
-Nada me impide representar las coordenadas resultantes en ejes rectangulares (las llamadas "coordenadas Cartesianas rectangulares").
+Lo mismo pasa al representar los ejes de una rotación oblicua: Nada me impide representar las coordenadas resultantes en ejes rectangulares (las llamadas "coordenadas Cartesianas rectangulares").
 Son coordenadas, nada más; el espacio en el que yo las represente en realidad da igual.
@@
 
-::::::: columns
+:::::::: columns
 ::: column
@@
 
-::::: column
+:::::: column
 ::: fragment
@@
 ::: fragment
-> Using orthogonal axes, discrimination corresponds
-to the length of the item response vector
+> Using orthogonal axes, discrimination corresponds to the length of the item response vector
 
@@
 :::
-:::::
-:::::::
+::::::
+::::::::
 
@@
 ::: r-center
-![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height=400px}
+![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height="400px"}
 :::

### 2025-02-28 output/seminar-slideshow.qmd
Reorder slide about "correlated traits"

@@
 :::::::::::: columns
-::::::: column
-::: {.fragment fragment-index="1"}
-$$
-  cor (\mathbf{\theta}) = \mathbf{R}
-$$
-:::
-
-::: {.fragment fragment-index="2"}
-$$
-  \mathbf{R} = \begin{bmatrix}
-      1 & 0.5 \\
-      0.5 & 1
-    \end{bmatrix}
-$$
-:::
-
-::: {.fragment fragment-index="3"}
-$$
-  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
-$$
-:::
-
-::: {.fragment fragment-index="4"}
-$$
-  cor (\mathbf{\theta}^*) =
-    \mathbf{R}^* =
-    \begin{bmatrix}
-      1 & 0 \\
-      0 & 1
-    \end{bmatrix}
-$$
-:::
-:::::::
-
 :::::: column
 ::::: r-stack
-::: {.fragment fragment-index="2"}
+::: {.fragment fragment-index="1"}
 ```{r corr-traits-example}
@@
 ::::::
+
+::::::: column
+::: {.fragment fragment-index="1"}
+$$
+  cor (\mathbf{\theta}) = \mathbf{R}
+$$
+:::
+
+::: {.fragment fragment-index="2"}
+$$
+  \mathbf{R} = \begin{bmatrix}
+      1 & 0.5 \\
+      0.5 & 1
+    \end{bmatrix}
+$$
+:::
+
+::: {.fragment fragment-index="3"}
+$$
+  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
+$$
+:::
+
+::: {.fragment fragment-index="4"}
+$$
+  cor (\mathbf{\theta}^*) =
+    \mathbf{R}^* =
+    \begin{bmatrix}
+      1 & 0 \\
+      0 & 1
+    \end{bmatrix}
+$$
+:::
+:::::::
 ::::::::::::
@@
 Y lo que nosotros estamos asumiendo es que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones $R$ (la que sea).
+
+\[NEXT\]
+
+(En este ejemplo, vamos a usar una distribución normal estandarizada con correlación de 0,5).
+
 Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a ejes oblicuos tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no" estandarizadas", es decir, que la distancia entre el 0 y el 1 pueda ser "distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
@@
 ¿Cómo podríamos solucionar esto?
+
+\[NEXT\]
+
 Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de ese espacio oblicuo (el que sea, aún no sabemos cuál es) a un espacio ortogonal.
-:::
-
-## Generalización: Ejes oblicuos
-
-<!-- # TODO: Expresión de las \thetas transformadas -->
-
-<!-- # TODO: Expresión de la R transformada -->
-
-<!-- # TODO: Distribución (esférica) de las \thetas transformadas -->
-
-::: notes
 Esto es una transformación de las coordenadas, como la que aplicamos antes para calcular las coordenadas del ítem tras aplicar la rotación oblicua, pero aplicada en este caso al vector de rasgo latente.
 
-Estas nuevas coordenadas (transformadas), que ahora están en un espacio ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^B$) sea una matriz identidad (unos en la diagonal, ceros en el resto).
+Ahora bien, como recordaréis, queremos representar los ítems en unos ejes que sean, por los ángulos relativos entre ellos, una representación fidedigna, de alguna manera, de las correlaciones entre las dimensiones de rasgo latente.
+
+Estas nuevas coordenadas (transformadas), que ahora están en un espacio ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^*$ la vamos a llamar) sea una matriz identidad (unos en la diagonal, ceros en el resto).
 Pero hay que tener en cuenta que esto es un "artefacto matemático".
@@
 lo que estemos investigando).
-:::
-
-## Generalización: Ejes oblicuos
-
-<!-- # TODO: Ejemplo de \theta (expresión) y coordenadas en espacio oblicuo -->
-
-::: notes
-Mientras tanto, las coordenadas de los "rasgos latentes" originales se van a representar en otros ejes, en general distintos de los "ejes ortogonales" transformados.
-Esto lo que significa es que las puntuaciones de rasgo latente originales van a representar la posición de un participante respecto de esos ejes originales.
-Por ejemplo, en este caso, si el participante tiene un "vector de puntuaciones" \[1, 2\], para representar sus coordenadas tenemos que desplazarnos, primero una unidad en el primer eje, y luego dos unidades en el segundo eje.
 :::

### 2025-02-28 output/seminar-slideshow.qmd
Version coordinate axes candidates (WiP)

@@
 
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
+¿Cuál es el sistema de coordenadas de θ para que θ* tenga correlaciones nulas?
+
+:::: {.columns .fragment}
+::: column
+```{r}
+x_limits <- c(-4, 4)
+y_limits <- c(-4, 4)
+
+vec_1 <- c(1, 0)
+vec_2 <- c(.5, sqrt(.75))
+basis <- cbind(vec_1, vec_2) |> unname()
+
+vec_1_tan <- vec_1[2]/vec_1[1]
+vec_2_tan <- vec_2[2]/vec_2[1]
+
+h_grid_limits <- trunc(x_limits / vec_2[2]) + c(-1, 1)
+h_grid_values <- h_grid_limits[1]:h_grid_limits[2]
+h_grid_interc <- h_grid_values * vec_2[2]
+h_grid_labpos <- h_grid_values + vec_1_tan * x_limits[1]
+
+v_grid_limits <- trunc(y_limits / vec_1[1]) + c(-1, 1)
+v_grid_values <- v_grid_limits[1]:v_grid_limits[2]
+v_grid_xcoord <- v_grid_values * vec_1[1]
+v_grid_interc <- - v_grid_xcoord * vec_2_tan
+v_grid_labpos <- v_grid_values # TODO
+
+ggplot() +
+  geom_abline(intercept = h_grid_interc, slope = vec_1_tan) +
+  geom_abline(intercept = v_grid_interc, slope = vec_2_tan) +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    breaks = v_grid_labpos,
+    labels = v_grid_values,
+    limits = x_limits,
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    breaks = h_grid_labpos,
+    labels = h_grid_values,
+    limits = y_limits,
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks       = element_blank(),
+    axis.line        = element_blank(),
+    axis.title       = element_markdown(valign = vec_2_tan * y_limits[2]), # TODO
+    axis.title.y     = element_markdown(halign = vec_1_tan * x_limits[2])
+  )
+```
+:::
+::: column
+:::
+::: column
+:::
+::::
 <!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
@@
 ::: notes
+Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
+Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
+(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
+
+[NEXT]
+
 En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Add remark about implicit canonical basis in model

@@
 
-> Third, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
-instead of the item difficulty.
-Finally, it is worth noting that
+> [...] Finally, it is worth noting that
 this formulation assumes *within-item multidimensionality* in general,

### 2025-03-02 output/Main_text.Rmd
Add remark about implicit canonical basis in model

--- 2025-02-17 output/Main_text.Rmd
@@
 in a non-cognitive domain.
-Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
+Second, as responses are not necessarily interpreted as correct or incorrect,
+parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
+instead of the item difficulty.
+Third, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
 non-negative real numbers in @reckase_difficulty_1985
@@
 the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
-Third, as responses are not necessarily interpreted as correct or incorrect,
-parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
-instead of the item difficulty.
+Fourth, $`r DISCR_VECTOR`$ is expressed on an algebraic basis,
+which the MIRT literature implicitly assumes to be the canonical basis
+(we will show that this assumption is not necessary though).
 Finally, it is worth highlighting that

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Improve clarification of geometric coordinates

@@
 > [...] such that $`r TRAIT_VECTOR_DEF_EQ`$.
-That is, vector $`r TRAIT_VECTOR`$ is represented in each of the two bases by
-the linear combination of its element vectors
+That is, the geometric representation of the algebraic vector $`r TRAIT_VECTOR`$
+in each of the two bases is given by the linear combination of
+the element vectors of the basis,
 scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
+Note thouth that, as $`r LATENT_SPACE`$ is originally represented in
+$`r LS_BASIS`$, the geometric coordinates $`r TRAIT_COORDS`$ of
+$`r TRAIT_VECTOR`$ in $`r LS_BASIS`$ are equal to the algebraic ones,
+i.e., $`r TRAIT_GEOM_ALGEBR_EQ`$.
 

### 2025-03-02 output/Main_text.Rmd
Improve clarification of geometric coordinates

@@
 respectively, such that $`r TRAIT_VECTOR_DEF_EQ`$.
-That is, vector $`r TRAIT_VECTOR`$ is represented in each of the two bases by
-the linear combination of its element vectors
+That is, the geometric representation of the algebraic vector $`r TRAIT_VECTOR`$
+in each of the two bases is given by the linear combination of
+the element vectors of the basis,
 scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
+Note thouth that, as $`r LATENT_SPACE`$ is originally represented in
+$`r LS_BASIS`$, the geometric coordinates $`r TRAIT_COORDS`$ of
+$`r TRAIT_VECTOR`$ in $`r LS_BASIS`$ are equal to the algebraic ones,
+i.e., $`r TRAIT_GEOM_ALGEBR_EQ`$.
 

### 2025-03-02 R/Formulae.R

@@ -141,6 +141,7 @@ TRAIT_COORDS_EQ       <- latex_eq(TRAIT_COORDS, TRAIT_COORDS_DEF)
 TRAIT_ORTH_COORDS     <- latex_raised_to(TRAIT_VECTOR, exp = LS_ORTH_BASIS)
 TRAIT_COORDS_ORTH_DEF <- latex_coords(TRAIT_VECTOR, LS_ORTH_BASIS)
 TRAIT_COORDS_ORTH_EQ  <- latex_eq(TRAIT_ORTH_COORDS, TRAIT_COORDS_ORTH_DEF)
+TRAIT_GEOM_ALGEBR_EQ  <- latex_eq(TRAIT_VECTOR, TRAIT_COORDS)
 
 # Change of basis matrix:
 TRANSFORM_MATRIX <- latex_bf('P')

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Explain resulting Mahalanobis distance better
The Mahalanobis distance, which results from the covariance-based version of the parameters, is explained in the "Covariance-based version" instead of in the discussion.

@@
 
-The distances between points are given by the norm, which we argue to be invariant to a change of basis.
-Interestingly, when considering the covariance-based version of the multidimensional parameters, the distance between points in the latent space is given by the Mahalanobis distance.
-We have reflected this in our new version, by adding the following to the third paragraph in our Discussion:
+The distances between points are given by the norm,
+be it in the latent space or in the test space.
+We propose that the norm of both items and trait vectors should be invariant to
+changes of basis and, interestingly, when considering the covariance-based version of the multidimensional parameters, the distance between points in the latent space is given by the Mahalanobis distance.
+We have reflected this in our new version, by adding the following to the end of the "Covariance-based Version" section:
 
-> [...] Once again, this is relevant in applications where
-potentially different bases are involved.
-Furthermore, not that, when using the covariance-based version formulation,
-the resulting norm in the latent space is the Mahalanobis distance
-[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
-thus taking the latent covariance structure into account. 
-Given these considerations, [...]
+> [...] Moreover, it is important to notice that,
+when the latent trait inner product Gram matrix
+is defined as $`r INNER_PROD_COV_COND`$,
+the norm of a trait vector in the latent space is given by
+the Mahalanobis distance, i.e., $`r MAH_DIST_EQ`$ [@mahalanobis_reprint_2018].
+
+Also, we have made changes and additions to
+the third paragraph of our Discussion, which now reads:
+
+> [...] Once again, this is relevantimportant in applications where potentially different bases are involved.
+Furthermore, we have found that using this version yields a norm in
+the latent space given by the Mahalanobis distance.
+This statistical distance not only takes
+the latent covariance structure into account,
+but also makes the latent space norm invariant to changes of basis. [...]
 

### 2025-03-02 output/Main_text.Rmd
Explain resulting Mahalanobis distance better
The Mahalanobis distance, which results from the covariance-based version of the parameters, is explained in the "Covariance-based version" instead of in the discussion.

@@
 (i.e., referred to a unitary variance space).
+Moreover, it is important to notice that,
+when the latent trait inner product Gram matrix
+is defined as $`r INNER_PROD_COV_COND`$,
+the norm of a trait vector in the latent space is given by
+the Mahalanobis distance, i.e., $`r MAH_DIST_EQ`$ [@mahalanobis_reprint_2018].
 
@@
 potentially different bases are involved.
-Furthermore, note that, when using the covariance-based version formulation,
-the resulting norm in the latent space is the Mahalanobis distance
-[i.e., $`r MAH_DIST_EQ`$, @mahalanobis_reprint_2018],
-thus taking the latent covariance structure into account. 
+Furthermore, we have found that using this version yields a norm in
+the latent space given by the Mahalanobis distance.
+This statistical distance not only takes
+the latent covariance structure into account,
+but also makes the latent space norm invariant to changes of basis.
 Given these considerations,

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Add covariance-based direction cosines expression
Better to add this numbered equation and refer to it in the discussion, instead of "(the covariance-based version of) Equation (23)" as it was in the previous version

@@
 
-> [...] Computing the parameters in the orthonormalized (or just orthogonalized) space,
+> [...] Computing the parameters in the orthonormalized
+(or just orthogonalized) space,
 by contrast, would require transforming the cosines to
@@
 Then, we may compute the direction cosines in the rotated solution applying
-(the covariance-based version of) Equation \@ref(eq:dir-cos-vec-test-space).
-<!-- TODO: Number and reference cov-based version of the cosines? -->
-What a researcher should be aware of, though,
+Equation (24).
+<!-- TODO: Check Eq. nº afterwards -->
+What a practitioner should be aware of, though,
 is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
@@
 a $`r MDISC_AG_PARAM`$ value specific to that basis,
-instead of the invariant value of $`r MDISC_COV_PARAM`$.
+unlike the invariant value of $`r MDISC_COV_PARAM`$.
 

### 2025-03-02 output/Main_text.Rmd
Add covariance-based direction cosines expression
Better to add this numbered equation and refer to it in the discussion, instead of "(the covariance-based version of) Equation (23)" as it was in the previous version

@@
 $`r INNER_PROD_COV_COND`$;
-in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$, resulting in a
-*covariance-based* version of the parameters:
+in this case, $`r DIAG_MATRIX_INNER_PROD_INV_VAR_EQ`$,
+resulting in the direction cosines defined by
+
+\begin{equation}
+  `r DIR_COS_VEC_COV_EQ`.                                 (\#eq:dir-cos-vec-cov)
+\end{equation}
+
+Along with the expressions for the $`r MDISC_SYM`$ and the signed distance,
+this results in a *covariance-based* version of the parameters:
 
@@
 Then, we may compute the direction cosines in the rotated solution applying
-(the covariance-based version of) Equation \@ref(eq:dir-cos-vec-test-space).
-<!-- TODO: Number and reference cov-based version of the cosines? -->
+Equation \@ref(eq:dir-cos-vec-cov).
 What a researcher should be aware of, though,

### 2025-03-02 R/Formulae.R

@@ -974,11 +974,8 @@ MDISC_COV_PARAM_EQ          <- latex_def(
 )
 
 # MIL:
-MIL_COV_PARAM                 <- latex_sub(MIL_PARAM, COV_MATRIX)
-DISTANCE_COV_DEF              <- latex_frac(
-  "- $INTERCEPT_PARAM$",
-  MDISC_COV_PARAM
-)
+MIL_COV_PARAM                     <- latex_sub(MIL_PARAM, COV_MATRIX)
+DISTANCE_COV_DEF           <- latex_frac("- $INTERCEPT_PARAM$", MDISC_COV_PARAM)
 SD_MATRIX_SQUARED                 <- latex_squared(SD_MATRIX)
 DIAG_MATRIX_INNER_PROD_INV_VAR_EQ <- latex_eq(
   DIAG_MATRIX_INNER_PROD_INV,
@@ -989,6 +986,7 @@ DIR_COS_ITEM_VEC_COV_DEF          <- latex_frac(
   "$SD_MATRIX_INV$ $COV_MATRIX$ $DISCR_VECTOR$",
   MDISC_COV_PARAM
 )
+DIR_COS_VEC_COV_EQ         <- latex_eq(DIR_COS_VEC_TS, DIR_COS_ITEM_VEC_COV_DEF)
 MIL_COV_PARAM_EQ                  <- latex_def(
   MIL_COV_PARAM,
   latex_curlybraces("$DISTANCE_COV_DEF$, $DIR_COS_ITEM_VEC_COV_DEF$")

### 2025-03-02 output/Main_text.Rmd
Correct reference to scale invariance
"Norm invariance" was incorrectly referred to as "scale invariance", making the discourse contradictory.
(Also, a minor reference to scale invariance in the "Covariance-based version" section is improved)

@@
 $`r MDISC_COV_PARAM`$ does not fulfill the first property
-(we shall refer to this property as *scale invariance*, from now on).
-Nevertheless, it can be interpreted as a *scaled version* of the parameter,
+(we shall refer to this property as *scale invariance*, henceforth).
+Nevertheless, $`r MDISC_COV_PARAM`$ can be interpreted as
+a *scaled version* of the parameter,
 with the standard deviation of the corresponding dimension as scaling factor.
@@
 it lacks the other desirable ones the covariance-based version has
-(i.e., scale invariance, and yielding a diagonal latent covariance matrix in
+(i.e., norm invariance, and yielding a diagonal latent covariance matrix in
 the corresponding orthogonal basis).

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Clarify dimensionality in graphical example
In response to Reviewer 2, we clarify that all the derivations are in n-dimensions, except for the graphical representation example

@@
 <!-- TODO: Check Eq. nº afterwards -->
-What a practitioner should be aware of, though,
+What a test practitioner should be aware of, though,
 is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
@@
 
-Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that may account for any number of dimensions.
-we have found that our previous version was not explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:
+Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that accounts for any number of dimensions.
+We have found that our previous version was not explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:
 
@@
 a positive response to item $`r ITEM_INDEX`$ is [...]
+
+Also, the "Graphical Representation Example" section now begins:
+
+> Up to now, all our derivations
+have considered $`r N_DIMS`$-dimensional spaces;
+that is, multidimensional parameters and item vector coordinates
+can be computed in an arbitrary large number of dimensions.
+However, as plotting more than two dimensions [...]
 

### 2025-03-02 output/Main_text.Rmd
Clarify dimensionality in graphical example
In response to Reviewer 2, we clarify that all the derivations are in n-dimensions, except for the graphical representation example

@@
 
-As plotting more than two dimensions is difficult on a bidimensional display
+Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
+that is, multidimensional parameters and item vector coordinates
+can be computed in an arbitrary large number of dimensions.
+However, as plotting more than two dimensions
+is difficult on a bidimensional display
 (and hardly possible at all for more than three dimensions with the
@@
 Equation \@ref(eq:dir-cos-vec-cov).
-What a researcher should be aware of, though,
+What a test practitioner should be aware of, though,
 is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Clarify better the item coordinates expressions
State explicitly the equivalence between discrimination geometric and algebraic coordinates, to make use of the latter in the definition of the item geometric coordinates.

@@
 Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
-To do that, we have made use of the expression $`r DISCR_VECTOR_COORDS_EQ`$, stated in p. 7.
+To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated in p. 7.
 (That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
@@
 the coordinates in terms of the parameters in the original model formulation.
-Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as
+Given $`r DISCR_VECTOR_COORDS_EQ`$,
+we can express the coordinates in terms of the model parameters as
 

### 2025-03-02 output/Letter_to_editor_reviewers.Rmd
Add minor changes about signed distance
In response to Reviewer 2's 3rd point

@@
 
-The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may be confuse the reader to refer to the $MDISC_i$ and obviate the influence of $d_i$.
+The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may confuse the reader to refer to the $MDISC_i$ and overlook the influence of $d_i$.
 We have added a clarification in this respect to the end of the paragraph, which now reads:
@@
 $`r DISTANCE_NULL_EQ`$ as well,
-regardless of the discrimination.
+regardless of the multidimensional discrimination.
 

### 2025-03-02 output/Main_text.Rmd
Clarify better the item coordinates expressions
State explicitly the equivalence between discrimination geometric and algebraic coordinates, to make use of the latter in the definition of the item geometric coordinates.
Add minor changes about signed distance
In response to Reviewer 2's 3rd point

@@
 model formulation.
-Given $`r DISCR_VECTOR_COORDS_EQ`$, we can express the coordinates as
+Given $`r DISCR_VECTOR_COORDS_EQ`$,
+we can express the coordinates in terms of the model parameters as
 
@@
 This implies that, if $`r INTERCEPT_NULL_EQ`$, $`r DISTANCE_NULL_EQ`$ as well,
-regardless of the discrimination.
+regardless of the multidimensional discrimination.
 

### 2025-03-07 output/sandbox_slideshow.qmd
Experiments with grid plotting

--- 2025-01-24 output/sandbox_slideshow.qmd
@@
 
+## Generalización: Ejes oblicuos
+
+¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
+
+:::::: {.columns .fragment}
+::: column
+```{r}
+x_limits   <- c(-4, 4)
+y_limits   <- c(-4, 4)
+break_step <- 2
+
+# Define grid limits as a function of the axes limits:
+grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
+  x_lim = if_else(x == x_limits[1], "inf", "sup"),
+  y_lim = if_else(y == y_limits[1], "inf", "sup")
+)
+
+vec_1 <- c(sqrt(1), sqrt(0))
+vec_2 <- c(sqrt(.25), sqrt(.75))
+basis <- cbind(vec_1, vec_2) |> unname()
+# basis <- basis |> solve() |> t()
+# vec_1 <- basis[, 1]
+# vec_2 <- basis[, 2]
+
+vec_1_tan <- vec_1[2] / vec_1[1]
+vec_2_tan <- vec_2[2] / vec_2[1]
+vec_1_mod <- vec_1^2 |> sum() |> sqrt()
+vec_2_mod <- vec_2^2 |> sum() |> sqrt()
+
+linear_transf <- function(points, basis, x = x, y = y) {
+
+  any_group_cols <- points |> select(-({{x}}:{{y}})) |> ncol()
+
+  if (!any_group_cols) points <- points |> rownames_to_column()
+
+  points |>
+    pivot_longer({{x}}:{{y}}) |>
+    group_by(across(-(name:value))) |>
+    mutate(value = basis %*% value |> drop()) |>
+    pivot_wider()    
+}
+
+# Select delimiting grid vertices, based on the basis vector directions
+grid_bounds <- bind_rows(
+  h = grid_box |>
+    filter(xor(vec_1_tan > 0, x_lim == y_lim)) |>
+    mutate(limit = y_lim, slope = vec_1_tan),
+  v = grid_box |>
+    filter(xor(vec_2_tan > 0, x_lim == y_lim)) |>
+    mutate(limit = x_lim, slope = vec_2_tan),
+  .id = "grid"
+) |>
+  # linear_transf(basis) |>
+  mutate(intercept = y - slope * x) # Compute their intercept
+
+# x_axis_titpos <- -y_limits[1] +
+#   if(is.infinite(vec_1_tan)) 0 else x_limits[2] * vec_1_tan
+# x_axis_titjust <- x_axis_titpos / x_axis_span
+# y_axis_titpos  <- -x_limits[1] +
+#   if(is.infinite(vec_2_tan)) 0 else y_limits[2] / vec_2_tan
+# y_axis_titjust <- y_axis_titpos / y_axis_span
+
+base_intercepts <- tibble(
+  h = vec_2[2] - vec_1_tan * vec_2[1],
+  v = vec_1[2] - vec_2_tan * vec_1[1]
+) |>
+  pivot_longer(everything(), names_to = "grid", values_to = "intercept")
+
+grid_specs <- grid_bounds |>
+  select(grid, limit:intercept) |>
+  left_join(base_intercepts, by = "grid", suffix = c("_limit", "_base")) |>
+  mutate(
+    n_limit = trunc(intercept_limit / intercept_base) %/%
+      break_step *
+      break_step
+  )
+
+grid <- grid_specs |>
+  select(-intercept_limit) |>
+  pivot_wider(names_from = limit, values_from = n_limit) |>
+  group_by(grid) |>
+  reframe(
+    slope,
+    coord = seq(from = inf, to = sup, by = break_step),
+    intercept = coord * intercept_base
+  ) |>
+  mutate(
+    label_pos = if_else(
+      grid == "h",
+      intercept + x_limits[1] * vec_1_tan,
+      intercept + y_limits[1] / vec_2_tan
+    ),
+    ref = as.numeric(!intercept)
+  )
+
+# h_grid <- tibble(
+#   pos = seq(from = h_grid_limits[1], by = break_step, to = h_grid_limits[2]),
+#   x   = 0L,
+#   y   = pos,
+#   ref = as.numeric(!y)
+# ) |>
+#   pivot_longer(
+#     x:y
+#   ) |>
+#     group_by(pos) |>
+#     mutate(value = basis %*% value |> drop()) |>
+#     pivot_wider() |>
+#     mutate(
+#       slope     = vec_1_tan,
+#       intercept = y - vec_1_tan * x 
+#     )
+# 
+# 
+# h_grid_interc <- h_grid_values * vec_2[2]
+# h_grid_labpos <- h_grid_interc + vec_1_tan * x_limits[1]
+# h_linetype    <- if_else(h_grid_interc == 0, 'solid', '13')
+# 
+# v_grid_limits <- (trunc(y_limits / abs(vec_1[1]) + x_limits / abs(vec_2_tan)) + c(-1, 1)) %/% break_step * break_step
+# v_grid_values <- seq(from = v_grid_limits[1], by = break_step, to = v_grid_limits[2])
+# v_grid_xcoord <- v_grid_values * vec_1[1]
+# v_grid_labpos <- v_grid_xcoord # Initial position of the x labels
+scale_linetype <- c(`1` = 'solid', `0` = 'dashed')
+
+if (is.infinite(vec_2_tan)) {
+
+  v_grid <- grid_bounds |>
+    filter(grid == "v") |>
+    mutate(
+      unit    = vec_1[1],
+      x_coord = trunc(x / unit) %/% break_step * break_step
+    ) |>
+    select(grid, limit, x_coord, unit) |>
+    pivot_wider(names_from = "limit", values_from = x_coord) |>
+    reframe(
+      x_coord = seq(from = inf, to = sup, by = break_step) * unit
+    ) |>
+    mutate(ref = as.numeric(!x_coord))
+
+  geom_vgrid <- geom_vline(
+    mapping = aes(xintercept = x_coord, linetype = ref),
+    data = v_grid
+  )
+
+}
+
+# else {
+#   
+#   v_grid_labpos <- v_grid_labpos + y_limits[1] / vec_2_tan
+#   v_grid_interc <- -v_grid_xcoord * vec_2_tan
+#   v_linetype    <- if_else(v_grid_interc == 0, 'solid', '13')
+#   geom_vgrid    <- geom_abline(
+#     aes(
+#       intercept = v_grid_interc,
+#       slope     = vec_2_tan,
+#       linetype  = v_grid_interc == 0
+#     )
+#   )
+# }
+
+grid_plot <- grid |>
+  ggplot(
+    mapping = aes(
+      intercept = intercept,
+      slope     = slope
+      # linetype  = ref
+    )
+  ) +
+  geom_abline()
+
+if (is.infinite(vec_2_tan)) grid_plot <- grid_plot + geom_vgrid
+
+grid_plot +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    breaks = grid |> filter(grid == "v") |> pull(label_pos),
+    labels = grid |> filter(grid == "v") |> pull(coord),
+    limits = x_limits,
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    breaks = grid |> filter(grid == "h") |> pull(label_pos),
+    labels = grid |> filter(grid == "h") |> pull(coord),
+    limits = y_limits,
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  # scale_linetype_manual(values = scale_linetype, guide = NULL) +
+  coord_fixed() +
+  theme(
+    axis.ticks   = element_blank(),
+    axis.line    = element_blank(),
+    # axis.title   = element_markdown(vjust = y_axis_titjust),
+    # axis.title.y = element_markdown(hjust = x_axis_titjust)
+    axis.title   = element_markdown(),
+    axis.title.y = element_markdown()
+  )
+```
+:::
+
+::: column
+:::
+
+::: column
+:::
+::::::
+
+<!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
+
+::: notes
+Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
+Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
+(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
+
+\[NEXT\]
+
+En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
+Supongamos una distribución bivariada de rasgos latentes, con una correlación de .8, bastante alta.
+¿Cuál de los siguientes sistemas de coordenadas consideráis que representaría adecuadamente los ejes de una distribución bivariada, de forma que al transformarlos a coordenadas rectangulares obtuviésemos una distribución con correlación de 0?
+:::
+
 ## References

### 2025-03-08 output/sandbox_slideshow.qmd
Create grid correctly

--- 2025-01-24 output/sandbox_slideshow.qmd
@@
 
+## Generalización: Ejes oblicuos
+
+¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
+
+:::::: {.columns .fragment}
+::: column
+```{r}
+x_limits   <- c(-4, 4) # Both limits expressed in canonical metric
+y_limits   <- c(-4, 4) # Both limits expressed in canonical metric
+break_step <- 2 # "Break units" for both grids, expressed in transformed metric
+
+vec_1 <- c(sqrt(1), sqrt(0))
+vec_2 <- c(sqrt(.25), sqrt(.75))
+basis <- cbind(vec_1, vec_2) |> unname()
+
+# Basis vectors tangents and norms
+vec_tan <- basis[2, ] / basis[1, ]
+vec_mod <- basis^2 |> colSums() |> sqrt()
+
+# Define grid limits as a function of the axes limits (in canonical metric):
+grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
+  x_lim = if_else(x == x_limits[1], "inf", "sup"),
+  y_lim = if_else(y == y_limits[1], "inf", "sup")
+)
+
+# Select delimiting grid vertices, based on the basis vector directions
+grid_bounds <- bind_rows(
+  h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
+  v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
+  .id = "grid"
+) |>
+  filter(xor(slope > 0, x_lim == y_lim)) |>
+  mutate(limit_intercept = y - slope * x) # Compute their intercept
+
+# x_axis_titpos <- -y_limits[1] +
+#   if(is.infinite(vec_1_tan)) 0 else x_limits[2] * vec_1_tan
+# x_axis_titjust <- x_axis_titpos / x_axis_span
+# y_axis_titpos  <- -x_limits[1] +
+#   if(is.infinite(vec_2_tan)) 0 else y_limits[2] / vec_2_tan
+# y_axis_titjust <- y_axis_titpos / y_axis_span
+
+# Transformed (1, 1), expressed in the canonical basis
+unit_point <- basis %*% c(1, 1) |>
+  as_tibble() |>
+  add_column(name = c('x', 'y') |> paste0('_unit')) |>
+  pivot_wider(values_from = V1)
+
+# Grid specifications, in canonical basis
+grid_specs <- grid_bounds |>
+  bind_cols(unit_point) |>
+  mutate(unit_intercept = y_unit - slope * x_unit) |>
+  select(-ends_with("unit")) |>
+  mutate(
+    n_units  = trunc(limit_intercept / unit_intercept),
+    n_breaks = n_units %/% break_step * break_step
+  )
+
+if (is.infinite(vec_tan[2])) {
+
+  v_grid <- grid_bounds |>
+    filter(grid == "v") |>
+    mutate(
+      unit     = vec_1[1],
+      n_units  = trunc(x / unit),
+      n_breaks = n_units %/% break_step * break_step
+    ) |>
+    select(grid, limit, n_breaks, unit) |>
+    pivot_wider(names_from = "limit", values_from = n_breaks) |>
+    reframe(
+      coord   = seq(from = inf, to = sup, by = break_step),
+      x_coord = coord * unit
+    ) |>
+  mutate(label_pos = x_coord, ref = as.character(!coord))
+
+  grid_specs <- grid_specs |> filter(!is.infinite(slope))
+
+  geom_vgrid <- geom_vline(
+    mapping = aes(xintercept = x_coord, linetype = ref),
+    data = v_grid
+  )
+}
+
+grid <- grid_specs |>
+  select(grid, limit, slope, unit_intercept, n_breaks) |>
+  pivot_wider(names_from = limit, values_from = n_breaks) |>
+  group_by(grid) |>
+  reframe(
+    slope,
+    coord = seq(from = inf, to = sup, by = break_step),
+    intercept = coord * unit_intercept
+  ) |>
+  mutate(
+    label_pos = if_else(
+      grid == "h",
+       intercept   + x_limits[1] * vec_tan[1],
+      (y_limits[1] - intercept)  / vec_tan[2]
+    ),
+    ref = as.character(!intercept)
+  )
+
+scale_linetype <- c(`TRUE` = 'solid', `FALSE` = 'dashed')
+
+grid_plot <- grid |>
+  ggplot() +
+  geom_abline(
+    mapping = aes(
+      intercept = intercept,
+      slope     = slope,
+      linetype  = ref
+    )
+  )
+
+if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid
+
+grid_plot +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    breaks = grid |> filter(grid == "v") |> pull(label_pos),
+    labels = grid |> filter(grid == "v") |> pull(coord),
+    limits = x_limits,
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    breaks = grid |> filter(grid == "h") |> pull(label_pos),
+    labels = grid |> filter(grid == "h") |> pull(coord),
+    limits = y_limits,
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_linetype_manual(values = scale_linetype, guide = NULL) +
+  coord_fixed() +
+  theme(
+    axis.ticks   = element_blank(),
+    axis.line    = element_blank(),
+    # axis.title   = element_markdown(vjust = y_axis_titjust),
+    # axis.title.y = element_markdown(hjust = x_axis_titjust)
+    axis.title   = element_markdown(),
+    axis.title.y = element_markdown()
+  )
+```
+:::
+
+::: column
+:::
+
+::: column
+:::
+::::::
+
+<!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
+
+::: notes
+Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
+Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
+(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
+
+\[NEXT\]
+
+En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
+Supongamos una distribución bivariada de rasgos latentes, con una correlación de .8, bastante alta.
+¿Cuál de los siguientes sistemas de coordenadas consideráis que representaría adecuadamente los ejes de una distribución bivariada, de forma que al transformarlos a coordenadas rectangulares obtuviésemos una distribución con correlación de 0?
+:::
+
 ## References

### 2025-03-08 output/sandbox_slideshow.qmd
Create final version of space transform choices

@@
 
-## Generalización: Ejes oblicuos
+## Generalización: Ejes oblicuos {.smaller}
 
-¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
+```{r create-grid-function}
+transform_grid <- function(basis,
+                           x_limits = c(-4, 4),
+                           y_limits = c(-4, 4),
+                           break_step = 2,
+                           linetype_axis = 'solid',
+                           linetype_grid = 'dashed') {
+  # Argument checking and formatting: ----
 
-:::::: {.columns .fragment}
-::: column
-```{r}
-x_limits   <- c(-4, 4) # Both limits expressed in canonical metric
-y_limits   <- c(-4, 4) # Both limits expressed in canonical metric
-break_step <- 2 # "Break units" for both grids, expressed in transformed metric
+  ## Grid plot linetypes
+  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
 
-vec_1 <- c(sqrt(1), sqrt(0))
-vec_2 <- c(sqrt(.25), sqrt(.75))
-basis <- cbind(vec_1, vec_2) |> unname()
+  ## Basis vectors tangents and norms
+  vec_tan <- basis[2, ] / basis[1, ]
 
-# Basis vectors tangents and norms
-vec_tan <- basis[2, ] / basis[1, ]
-vec_mod <- basis^2 |> colSums() |> sqrt()
+  ## Axis title positions
+  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
+    (x_limits[2] - x_limits[1])
+  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
+    (y_limits[2] - y_limits[1])
+
+  # Main: ----
 
@@
     mutate(limit_intercept = y - slope * x) # Compute their intercept
-
-# x_axis_titpos <- -y_limits[1] +
-#   if(is.infinite(vec_1_tan)) 0 else x_limits[2] * vec_1_tan
-# x_axis_titjust <- x_axis_titpos / x_axis_span
-# y_axis_titpos  <- -x_limits[1] +
-#   if(is.infinite(vec_2_tan)) 0 else y_limits[2] / vec_2_tan
-# y_axis_titjust <- y_axis_titpos / y_axis_span
 
@@
 
+  # Create vertical grid, if the vertical basis vector is completely vertical
   if (is.infinite(vec_tan[2])) {
 
+    # Create vertical grid data
     v_grid <- grid_bounds |>
@@
       mutate(
-      unit     = vec_1[1],
+        unit     = basis[1, 1],
         n_units  = trunc(x / unit),
@@
 
+    # Filter out non-valid grid values
     grid_specs <- grid_specs |> filter(!is.infinite(slope))
 
+    # Assign x axis label values and coordinates (for the vertical case)
+    x_labels <- v_grid |> select(coord, label_pos)
+
+    # Create vertical grid geometry
     geom_vgrid <- geom_vline(
@@
 
+  # Create grid data
   grid <- grid_specs |>
@@
 
-scale_linetype <- c(`TRUE` = 'solid', `FALSE` = 'dashed')
+  # Assign x axis label values and coordinates (accounting for the vertical
+  #   special case)
+  x_labels <- if (is.infinite(vec_tan[2]))    v_grid |> select(coord, label_pos)
+              else       grid |> filter(grid == "v") |> select(coord, label_pos)
 
@@
       minor_breaks = NULL,
-    breaks = grid |> filter(grid == "v") |> pull(label_pos),
-    labels = grid |> filter(grid == "v") |> pull(coord),
+      breaks = x_labels |> pull(label_pos),
+      labels = x_labels |> pull(coord),
       limits = x_limits,
@@
     ) +
-  scale_linetype_manual(values = scale_linetype, guide = NULL) +
+    scale_linetype_manual(values = linetypes, guide = NULL) +
     coord_fixed() +
@@
       axis.line    = element_blank(),
-    # axis.title   = element_markdown(vjust = y_axis_titjust),
-    # axis.title.y = element_markdown(hjust = x_axis_titjust)
-    axis.title   = element_markdown(),
-    axis.title.y = element_markdown()
+      axis.title.x       = element_markdown(hjust = y_axis_titpos),
+      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
     )
+}
+```
+
+¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
+
+<br>
+
+:::::: columns
+::: {.column .fragment width="33%"}
+<center>A</center>
+
+```{r grid-orthogonal-rotation}
+#| fig-width: 5
+basis <- cbind(
+  c(sqrt(.85), -sqrt(.15)),
+  c(sqrt(.15), sqrt(.85))
+)
+
+transform_grid(basis)
 ```
@@
 
-::: column
+::: {.column .fragment width="33%"}
+<center>B</center>
+
+```{r grid-ts-basis}
+#| fig-width: 5
+basis <- cbind(
+  c(1, 0),
+  c(sqrt(.25), sqrt(.75))
+)
+
+transform_grid(basis)
+```
 :::
 
-::: column
+::: {.column .fragment width="33%"}
+<center>C</center>
+
+```{r grid-ls-basis}
+#| fig-width: 5
+basis <- t(solve(basis))
+
+transform_grid(basis)
+```
 :::
 ::::::
-
-<!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
 

### 2025-03-08 output/sandbox_slideshow.qmd
Move slide to main presentation

--- 2025-03-08 output/sandbox_slideshow.qmd
@@
 
-## Generalización: Ejes oblicuos
-
-¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
-
-:::::: {.columns .fragment}
-::: column
-```{r}
-x_limits   <- c(-4, 4) # Both limits expressed in canonical metric
-y_limits   <- c(-4, 4) # Both limits expressed in canonical metric
-break_step <- 2 # "Break units" for both grids, expressed in transformed metric
-
-vec_1 <- c(sqrt(1), sqrt(0))
-vec_2 <- c(sqrt(.25), sqrt(.75))
-basis <- cbind(vec_1, vec_2) |> unname()
-
-# Basis vectors tangents and norms
-vec_tan <- basis[2, ] / basis[1, ]
-vec_mod <- basis^2 |> colSums() |> sqrt()
-
-# Define grid limits as a function of the axes limits (in canonical metric):
-grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
-  x_lim = if_else(x == x_limits[1], "inf", "sup"),
-  y_lim = if_else(y == y_limits[1], "inf", "sup")
-)
-
-# Select delimiting grid vertices, based on the basis vector directions
-grid_bounds <- bind_rows(
-  h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
-  v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
-  .id = "grid"
-) |>
-  filter(xor(slope > 0, x_lim == y_lim)) |>
-  mutate(limit_intercept = y - slope * x) # Compute their intercept
-
-# x_axis_titpos <- -y_limits[1] +
-#   if(is.infinite(vec_1_tan)) 0 else x_limits[2] * vec_1_tan
-# x_axis_titjust <- x_axis_titpos / x_axis_span
-# y_axis_titpos  <- -x_limits[1] +
-#   if(is.infinite(vec_2_tan)) 0 else y_limits[2] / vec_2_tan
-# y_axis_titjust <- y_axis_titpos / y_axis_span
-
-# Transformed (1, 1), expressed in the canonical basis
-unit_point <- basis %*% c(1, 1) |>
-  as_tibble() |>
-  add_column(name = c('x', 'y') |> paste0('_unit')) |>
-  pivot_wider(values_from = V1)
-
-# Grid specifications, in canonical basis
-grid_specs <- grid_bounds |>
-  bind_cols(unit_point) |>
-  mutate(unit_intercept = y_unit - slope * x_unit) |>
-  select(-ends_with("unit")) |>
-  mutate(
-    n_units  = trunc(limit_intercept / unit_intercept),
-    n_breaks = n_units %/% break_step * break_step
-  )
-
-if (is.infinite(vec_tan[2])) {
-
-  v_grid <- grid_bounds |>
-    filter(grid == "v") |>
-    mutate(
-      unit     = vec_1[1],
-      n_units  = trunc(x / unit),
-      n_breaks = n_units %/% break_step * break_step
-    ) |>
-    select(grid, limit, n_breaks, unit) |>
-    pivot_wider(names_from = "limit", values_from = n_breaks) |>
-    reframe(
-      coord   = seq(from = inf, to = sup, by = break_step),
-      x_coord = coord * unit
-    ) |>
-  mutate(label_pos = x_coord, ref = as.character(!coord))
-
-  grid_specs <- grid_specs |> filter(!is.infinite(slope))
-
-  geom_vgrid <- geom_vline(
-    mapping = aes(xintercept = x_coord, linetype = ref),
-    data = v_grid
-  )
-}
-
-grid <- grid_specs |>
-  select(grid, limit, slope, unit_intercept, n_breaks) |>
-  pivot_wider(names_from = limit, values_from = n_breaks) |>
-  group_by(grid) |>
-  reframe(
-    slope,
-    coord = seq(from = inf, to = sup, by = break_step),
-    intercept = coord * unit_intercept
-  ) |>
-  mutate(
-    label_pos = if_else(
-      grid == "h",
-       intercept   + x_limits[1] * vec_tan[1],
-      (y_limits[1] - intercept)  / vec_tan[2]
-    ),
-    ref = as.character(!intercept)
-  )
-
-scale_linetype <- c(`TRUE` = 'solid', `FALSE` = 'dashed')
-
-grid_plot <- grid |>
-  ggplot() +
-  geom_abline(
-    mapping = aes(
-      intercept = intercept,
-      slope     = slope,
-      linetype  = ref
-    )
-  )
-
-if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid
-
-grid_plot +
-  scale_x_continuous(
-    minor_breaks = NULL,
-    breaks = grid |> filter(grid == "v") |> pull(label_pos),
-    labels = grid |> filter(grid == "v") |> pull(coord),
-    limits = x_limits,
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    minor_breaks = NULL,
-    breaks = grid |> filter(grid == "h") |> pull(label_pos),
-    labels = grid |> filter(grid == "h") |> pull(coord),
-    limits = y_limits,
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_linetype_manual(values = scale_linetype, guide = NULL) +
-  coord_fixed() +
-  theme(
-    axis.ticks   = element_blank(),
-    axis.line    = element_blank(),
-    # axis.title   = element_markdown(vjust = y_axis_titjust),
-    # axis.title.y = element_markdown(hjust = x_axis_titjust)
-    axis.title   = element_markdown(),
-    axis.title.y = element_markdown()
-  )
-```
-:::
-
-::: column
-:::
-
-::: column
-:::
-::::::
-
-<!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
-
-::: notes
-Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
-Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
-(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
-
-\[NEXT\]
-
-En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
-Supongamos una distribución bivariada de rasgos latentes, con una correlación de .8, bastante alta.
-¿Cuál de los siguientes sistemas de coordenadas consideráis que representaría adecuadamente los ejes de una distribución bivariada, de forma que al transformarlos a coordenadas rectangulares obtuviésemos una distribución con correlación de 0?
-:::
-
 ## References

### 2025-03-08 output/seminar-slideshow.qmd

--- 2025-02-28 output/seminar-slideshow.qmd
@@
 
-## Generalización: Ejes oblicuos
-
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Opciones de elección de bases oblicuas: rectangular, ejes muy alineados, ejes correctos -->
-
-::: notes
+## Generalización: Ejes oblicuos {.smaller}
+
+```{r create-grid-function}
+transform_grid <- function(basis,
+                           x_limits = c(-4, 4),
+                           y_limits = c(-4, 4),
+                           break_step = 2,
+                           linetype_axis = 'solid',
+                           linetype_grid = 'dashed') {
+  # Argument checking and formatting: ----
+
+  ## Grid plot linetypes
+  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
+
+  ## Basis vectors tangents and norms
+  vec_tan <- basis[2, ] / basis[1, ]
+
+  ## Axis title positions
+  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
+    (x_limits[2] - x_limits[1])
+  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
+    (y_limits[2] - y_limits[1])
+
+  # Main: ----
+
+  # Define grid limits as a function of the axes limits (in canonical metric):
+  grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
+    x_lim = if_else(x == x_limits[1], "inf", "sup"),
+    y_lim = if_else(y == y_limits[1], "inf", "sup")
+  )
+
+  # Select delimiting grid vertices, based on the basis vector directions
+  grid_bounds <- bind_rows(
+    h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
+    v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
+    .id = "grid"
+  ) |>
+    filter(xor(slope > 0, x_lim == y_lim)) |>
+    mutate(limit_intercept = y - slope * x) # Compute their intercept
+
+  # Transformed (1, 1), expressed in the canonical basis
+  unit_point <- basis %*% c(1, 1) |>
+    as_tibble() |>
+    add_column(name = c('x', 'y') |> paste0('_unit')) |>
+    pivot_wider(values_from = V1)
+
+  # Grid specifications, in canonical basis
+  grid_specs <- grid_bounds |>
+    bind_cols(unit_point) |>
+    mutate(unit_intercept = y_unit - slope * x_unit) |>
+    select(-ends_with("unit")) |>
+    mutate(
+      n_units  = trunc(limit_intercept / unit_intercept),
+      n_breaks = n_units %/% break_step * break_step
+    )
+
+  # Create vertical grid, if the vertical basis vector is completely vertical
+  if (is.infinite(vec_tan[2])) {
+
+    # Create vertical grid data
+    v_grid <- grid_bounds |>
+      filter(grid == "v") |>
+      mutate(
+        unit     = basis[1, 1],
+        n_units  = trunc(x / unit),
+        n_breaks = n_units %/% break_step * break_step
+      ) |>
+      select(grid, limit, n_breaks, unit) |>
+      pivot_wider(names_from = "limit", values_from = n_breaks) |>
+      reframe(
+        coord   = seq(from = inf, to = sup, by = break_step),
+        x_coord = coord * unit
+      ) |>
+      mutate(label_pos = x_coord, ref = as.character(!coord))
+
+    # Filter out non-valid grid values
+    grid_specs <- grid_specs |> filter(!is.infinite(slope))
+
+    # Assign x axis label values and coordinates (for the vertical case)
+    x_labels <- v_grid |> select(coord, label_pos)
+
+    # Create vertical grid geometry
+    geom_vgrid <- geom_vline(
+      mapping = aes(xintercept = x_coord, linetype = ref),
+      data = v_grid
+    )
+  }
+
+  # Create grid data
+  grid <- grid_specs |>
+    select(grid, limit, slope, unit_intercept, n_breaks) |>
+    pivot_wider(names_from = limit, values_from = n_breaks) |>
+    group_by(grid) |>
+    reframe(
+      slope,
+      coord = seq(from = inf, to = sup, by = break_step),
+      intercept = coord * unit_intercept
+    ) |>
+    mutate(
+      label_pos = if_else(
+        grid == "h",
+         intercept   + x_limits[1] * vec_tan[1],
+        (y_limits[1] - intercept)  / vec_tan[2]
+      ),
+      ref = as.character(!intercept)
+    )
+
+  # Assign x axis label values and coordinates (accounting for the vertical
+  #   special case)
+  x_labels <- if (is.infinite(vec_tan[2]))    v_grid |> select(coord, label_pos)
+              else       grid |> filter(grid == "v") |> select(coord, label_pos)
+
+  grid_plot <- grid |>
+    ggplot() +
+    geom_abline(
+      mapping = aes(
+        intercept = intercept,
+        slope     = slope,
+        linetype  = ref
+      )
+    )
+
+  if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid
+
+  grid_plot +
+    scale_x_continuous(
+      minor_breaks = NULL,
+      breaks = x_labels |> pull(label_pos),
+      labels = x_labels |> pull(coord),
+      limits = x_limits,
+      expand = expansion(),
+      name   = NULL,
+      sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
+    ) +
+    scale_y_continuous(
+      minor_breaks = NULL,
+      breaks = grid |> filter(grid == "h") |> pull(label_pos),
+      labels = grid |> filter(grid == "h") |> pull(coord),
+      limits = y_limits,
+      expand = expansion(),
+      name   = NULL,
+      sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
+    ) +
+    scale_linetype_manual(values = linetypes, guide = NULL) +
+    coord_fixed() +
+    theme(
+      axis.ticks   = element_blank(),
+      axis.line    = element_blank(),
+      axis.title.x       = element_markdown(hjust = y_axis_titpos),
+      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
+    )
+}
+```
+
+¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
+
+<br>
+
+:::::: columns
+::: {.column .fragment width="33%"}
+<center>A</center>
+
+```{r grid-orthogonal-rotation}
+#| fig-width: 5
+basis <- cbind(
+  c(sqrt(.85), -sqrt(.15)),
+  c(sqrt(.15), sqrt(.85))
+)
+
+transform_grid(basis)
+```
+:::
+
+::: {.column .fragment width="33%"}
+<center>B</center>
+
+```{r grid-ts-basis}
+#| fig-width: 5
+basis <- cbind(
+  c(1, 0),
+  c(sqrt(.25), sqrt(.75))
+)
+
+transform_grid(basis)
+```
+:::
+
+::: {.column .fragment width="33%"}
+<center>C</center>
+
+```{r grid-ls-basis}
+#| fig-width: 5
+basis <- t(solve(basis))
+
+transform_grid(basis)
+```
+:::
+::::::
+
+::: notes
+Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
+Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
+(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
+
+\[NEXT\]
+
 En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.

### 2025-03-21 output/sandbox_slideshow.qmd
Add {gganimate} example to sandbox slide
There's a problem with the paths when the document is rendered in the project path, so instead of showing the chunk output, we need to insert the image aside, immediately afterwards.

--- 2025-03-08 output/sandbox_slideshow.qmd
@@
 library(scales)
-library(mvtnorm)
 ```
@@
 ```
-
-## Generalización: Ejes oblicuos {.smaller}
 
@@
 
-¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
+## Generalización: Ejes oblicuos
 
@@
 :::::: columns
-::: {.column .fragment width="33%"}
+::: {.column width="33%"}
 <center>A</center>
@@
 basis <- cbind(
-  c(sqrt(.85), -sqrt(.15)),
-  c(sqrt(.15), sqrt(.85))
+  c(sqrt(.75), -sqrt(.25)),
+  c(sqrt(.25), sqrt(.75))
 )
@@
 ::: notes
-Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
-Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
-(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
-
-\[NEXT\]
-
-En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
-Supongamos una distribución bivariada de rasgos latentes, con una correlación de .8, bastante alta.
-¿Cuál de los siguientes sistemas de coordenadas consideráis que representaría adecuadamente los ejes de una distribución bivariada, de forma que al transformarlos a coordenadas rectangulares obtuviésemos una distribución con correlación de 0?
+El primero yo creo que estamos de acuerdo en que no, ya que solamente es una "rotación ortogonal" (como la llamaríamos en el ámbito psicométrico).
+La transformación que se aplica podría dar lugar a correlaciones nulas (es lo que haría la transformación de los ejes mediante Análisis de Componentes Principales), pero las varianzas resultantes no serían estandarizadas.
 :::
 
+## Example
+
+```{r}
+#| fig-format: png
+#| results:    hide
+library(gganimate, quietly = TRUE)
+
+p <- ggplot(mtcars, aes(factor(cyl), mpg)) +
+  geom_boxplot() +
+  # Here comes the gganimate code
+  transition_states(
+    gear,
+    transition_length = 2,
+    state_length = 1
+  ) +
+  enter_fade() +
+  exit_shrink() +
+  ease_aes('sine-in-out')
+
+animate(p, renderer = gifski_renderer(file = here::here("www/slideshow-assets/anim_plot.gif"), loop = FALSE))
+```
+
+![](../www/slideshow-assets/anim_plot.gif)
+
 ## References

### 2025-03-23 output/seminar-slideshow.qmd
Add vertical alignment in slide

@@
 
-set.seed(7117)
-
 # This is a square root of the correlation matrix, then inverted and transposed:
@@
 lo que estemos investigando).
+
+Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
+Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
+(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
+:::
+
+## Generalización: Ejes oblicuos {.smaller}
+
+<center>
+
+<br>
+
+[bit.ly/vote-coords](https://bit.ly/vote-coords) ([slido.com](https://slido.com) : #1904926)
+
+<br>
+
+![](../www/slideshow-assets/bit.ly_vote-coords.svg){width="40%" height="40%"}
+
+</center>
+
+::: notes
+En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
+Aquí tenéis un enlace y QR para participar en una breve "encuesta".
+:::
+
+## Generalización: Ejes oblicuos {.smaller}
+
+<br>
+
+:::::: columns
+::: {.column .fragment width="33%"}
+:::
+
+::: {.column .fragment width="33%"}
+:::
+
+::: {.column .fragment width="33%"}
+```{r corr-traits-transf-example}
+#| fig-width:  5.5
+#| fig-height: 5.5
+```
+:::
+::::::
+
+¿Cuál es el sistema de coordenadas de *θ* para que *θ\** sea "esférica"?
+
+::: notes
+Supongamos una distribución bivariada de rasgos latentes, con una correlación de .5, como la mostrada anteriormente.
+Queremos saber cuál es la matriz $\mathbf{P}$ que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$.
+
+Esto es equivalente a preguntar:
+
+\[NEXT\]
+
+¿Cuál es el sistema de coordenadas que representaría adecuadamente los ejes de una distribución bivariada,...
+
+\[NEXT\]
+
+de forma que al transformarlos a coordenadas rectangulares (mediante la matriz $\mathbf{P}$) obtuviésemos una distribución con correlaciones de 0 y varianzas de 1, es decir, una distribución "esférica"?
 :::
@@
 
-¿Cuál es el sistema de coordenadas de θ para que θ\* tenga correlaciones nulas?
-
 <br>
@@
 
-::: notes
-Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
-Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
-(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
+<br>
+
+::: fragment
+[Respuestas en 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
+:::
+
+::: notes
+-   ¿Será el A?
 
@@
 
-En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
-Supongamos una distribución bivariada de rasgos latentes, con una correlación de .8, bastante alta.
-¿Cuál de los siguientes sistemas de coordenadas consideráis que representaría adecuadamente los ejes de una distribución bivariada, de forma que al transformarlos a coordenadas rectangulares obtuviésemos una distribución con correlación de 0?
+-   ¿O puede ser el B?
+
+\[NEXT\]
+
+-   ¿O es el C?
+
+\[NEXT\]
+
+Tomáos vuestro tiempo y en seguida vemos las respuestas.
 :::

### 2025-03-23 output/seminar-slideshow.qmd
Separate grid transformation in functions

@@
   mutate(value = orth_matrix %*% value |> drop()) |>
-  pivot_wider()
+  pivot_wider() |>
+  ungroup()
 
@@
 ```{r create-grid-function}
-transform_grid <- function(basis,
+create_grid <- function(basis,
                         x_limits = c(-4, 4),
                         y_limits = c(-4, 4),
-                           break_step = 2,
-                           linetype_axis = 'solid',
-                           linetype_grid = 'dashed') {
+                        break_step = 2) {
   # Argument checking and formatting: ----
-
-  ## Grid plot linetypes
-  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
 
@@
   vec_tan <- basis[2, ] / basis[1, ]
-
-  ## Axis title positions
-  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
-    (x_limits[2] - x_limits[1])
-  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
-    (y_limits[2] - y_limits[1])
 
@@
   # Transformed (1, 1), expressed in the canonical basis
-  unit_point <- basis %*% c(1, 1) |>
-    as_tibble() |>
+  unit_point <- basis %*% c(1, 1) |> drop() |>
+    as_tibble(.name_repair = "minimal") |>
     add_column(name = c('x', 'y') |> paste0('_unit')) |>
-    pivot_wider(values_from = V1)
+    pivot_wider(values_from = value)
 
@@
 
-  # Create vertical grid, if the vertical basis vector is completely vertical
+  # Create vertical grid data, if the second basis vector is completely vertical
   if (is.infinite(vec_tan[2])) {
@@
       pivot_wider(names_from = "limit", values_from = n_breaks) |>
-      reframe(
-        coord   = seq(from = inf, to = sup, by = break_step),
-        x_coord = coord * unit
-      ) |>
-      mutate(label_pos = x_coord, ref = as.character(!coord))
+      reframe(coord = seq(from = inf, to = sup, by = break_step) * unit) |>
+      mutate(
+        grid      = 'v',
+        slope     = Inf,
+        intercept = NA_real_,
+        label_pos = coord,
+        ref       = as.character(!coord)
+      )
 
@@
     grid_specs <- grid_specs |> filter(!is.infinite(slope))
-
-    # Assign x axis label values and coordinates (for the vertical case)
-    x_labels <- v_grid |> select(coord, label_pos)
-
-    # Create vertical grid geometry
-    geom_vgrid <- geom_vline(
-      mapping = aes(xintercept = x_coord, linetype = ref),
-      data = v_grid
-    )
   }
@@
 
-  # Assign x axis label values and coordinates (accounting for the vertical
+    if (is.infinite(vec_tan[2])) return(grid |> bind_rows(v_grid))
+
+    grid
+}
+
+plot_grid <- function(grid,
+                      x_limits = c(-4, 4),
+                      y_limits = c(-4, 4),
+                      break_step = 2,
+                      linetype_axis = 'solid',
+                      linetype_grid = 'dashed') {
+
+  # Argument checking and formatting: ----
+
+  ## Grid plot linetypes
+  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
+
+  ## Basis vectors tangents and norms
+  vec_tan <- grid |> distinct(grid, slope) |> deframe()
+
+  ## Axis title positions
+  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
+    (x_limits[2] - x_limits[1])
+  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
+    (y_limits[2] - y_limits[1])
+
+  # Main: ----
+
+  # Assign x axis label values and coordinates (to account for the vertical
   #   special case)
-  x_labels <- if (is.infinite(vec_tan[2]))    v_grid |> select(coord, label_pos)
-              else       grid |> filter(grid == "v") |> select(coord, label_pos)
+  x_labels <- grid |> filter(grid == "v") |> select(coord, label_pos)
+
+  # Create vertical grid geometry if the second vector is completely vertical
+  if (is.infinite(vec_tan[2])) {
+
+    geom_vgrid <- geom_vline(
+      mapping = aes(xintercept = coord, linetype = ref),
+      data = grid |> filter(grid == "v")
+    )
+
+    grid <- grid |> filter(grid == "h")
+  }
 
@@
 }
+
+transform_grid <- function(grid,
+                           x_limits = c(-4, 4),
+                           y_limits = c(-4, 4),
+                           break_step = 2,
+                           linetype_axis = 'solid',
+                           linetype_grid = 'dashed') {
+
+  # Argument checking and formatting: ----
+
+  # Main: ----
+  basis |>
+    create_grid(
+      x_limits   = x_limits,
+      y_limits   = y_limits,
+      break_step = break_step
+    ) |>
+    plot_grid(
+      x_limits   = x_limits,
+      y_limits   = y_limits,
+      break_step = break_step,
+      linetype_axis = 'solid',
+      linetype_grid = 'dashed'
+    )
+}
 ```

### 2025-03-23 output/seminar-slideshow.qmd
Add scatter plots with tranformed traits
Demonstrates that the transform matrix is not the square root matrix of the covariance, but its inverse

@@
 thetas <- rmvnorm(10^4, sigma = corr_matrix) |>
-  as_tibble(.name_repair = ~paste0("trait_", 1:2))
+  as_tibble(.name_repair = ~paste0("trait_", 1:2)) |>
+    rownames_to_column("i")
 
@@
 thetas_transf <- thetas |>
-  rownames_to_column("i") |>
   pivot_longer(starts_with("trait")) |>
@@
 <br>
+
+<!-- # TODO: Faltan dos primeros paneles -->
 
@@
 
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Distribución en ejes rectangulares -->
-
-::: notes
-El primero ya sabemos que no, de acuerdo, puesto que es la representación habitual, y sabemos que la nube que forma está "achatada".
-No se aplica ninguna transformación (la transformación es una matriz identidad), ya que los que serían los ejes originales y los transformados son los mismos.
+<br>
+
+::::::::: columns
+::: {.column width="33%"}
+<center>A</center>
+
+```{r scatter-orthogonal-rotation}
+#| fig-width: 5
+basis <- cbind(
+  c(sqrt(.75), -sqrt(.25)),
+  c(sqrt(.25),  sqrt(.75))
+)
+
+thetas_transf <- thetas |>
+  group_by(i) |>
+  pivot_longer(starts_with("trait")) |>
+  mutate(value = basis %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
+transform_grid(basis) +
+  geom_point(
+    data    = thetas_transf,
+    mapping = aes(trait_1, trait_2),
+    alpha   = .33,
+    shape   = 16,
+    colour = I(PALETTE_BLUE["normal"])
+  )
+```
+:::
+
+:::::: {.column width="33%"}
+<center>B</center>
+
+::::: r-stack
+::: fragment
+```{r grid-ts-basis}
+#| fig-width: 5
+```
+:::
+
+::: fragment
+```{r scatter-ts-basis}
+#| fig-width: 5
+basis <- cbind(
+  c(1, 0),
+  c(sqrt(.25), sqrt(.75))
+)
+
+thetas_transf <- thetas |>
+  group_by(i) |>
+  pivot_longer(starts_with("trait")) |>
+  mutate(value = basis %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
+transform_grid(basis) +
+  geom_point(
+    data    = thetas_transf,
+    mapping = aes(trait_1, trait_2),
+    alpha   = .33,
+    shape   = 16,
+    colour = I(PALETTE_BLUE["normal"])
+  )
+```
+:::
+:::::
+::::::
+
+::: {.column .fragment width="33%"}
+<center>C</center>
+
+```{r scatter-ls-basis}
+#| fig-width: 5
+basis <- t(solve(basis))
+
+thetas_transf <- thetas |>
+  group_by(i) |>
+  pivot_longer(starts_with("trait")) |>
+  mutate(value = basis %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
+transform_grid(basis) +
+  geom_point(
+    data    = thetas_transf,
+    mapping = aes(trait_1, trait_2),
+    alpha   = .33,
+    shape   = 16,
+    colour = I(PALETTE_BLUE["normal"])
+  )
+```
+:::
+:::::::::
+
+::: notes
+El primero yo creo que estamos de acuerdo en que no, ya que solamente es una "rotación ortogonal" (como la llamaríamos en el ámbito psicométrico).
+La transformación que se aplica podría dar lugar a correlaciones nulas (es lo que haría la transformación de los ejes mediante Análisis de Componentes Principales), pero las varianzas resultantes no serían estandarizadas.
+
+\[NEXT\]
+
+El segundo caso rotaría el eje vertical de manera que ambos ejes formarían un ángulo de 60º, que tiene un coseno igual a 0.5.
+La gente que ha pensado que estos son los ejes correctos, pensáis como yo.
+
+\[NEXT\]
+
+Sin embargo, mirad lo que ocurre cuando "proyectamos" esa nube de puntos sobre esos ejes.
+La "nube de puntos" se achata y se alarga aún más.
+
+\[NEXT\]
+
+La respuesta correcta era la C, que "deshace" el achatamiento de la nube de puntos, convirtiéndola en esférica.
 :::

### 2025-03-28 output/seminar-slideshow.qmd
Drop duplicated slides about coordinate transform

@@
 
-## Generalización: Ejes oblicuos {.smaller}
+## Generalización: Ejes oblicuos
 
@@
 
-<br>
-
 ::: fragment
@@
 
-:::::: {.column width="33%"}
+:::::: {.column .fragment width="33%"}
 <center>B</center>
@@
 ::::: r-stack
-::: fragment
 ```{r grid-ts-basis}
@@
 ```
-:::
 
@@
 El segundo caso rotaría el eje vertical de manera que ambos ejes formarían un ángulo de 60º, que tiene un coseno igual a 0.5.
-La gente que ha pensado que estos son los ejes correctos, pensáis como yo.
+La gente que ha pensado que estos son los ejes correctos, pensáis como yo: que el coseno del ángulo entre los ejes representa la correlación entre las dimensiones.
+Por lo tanto, cuanto más alta la correlación, más alineados deben estar los ejes.
 
@@
 
-La respuesta correcta era la C, que "deshace" el achatamiento de la nube de puntos, convirtiéndola en esférica.
-:::
-
-## Generalización: Ejes oblicuos
-
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Animación de distribución en ejes rectangulares a ejes alineados -->
-
-::: notes
-Si pensáis igual que yo, habréis elegido el segundo, porque el coseno del ángulo entre los ejes representa la correlación entre las dimensiones.
-Por lo tanto, ejes muy alineados representan una correlación muy alta.
-
-Sin embargo, fijáos lo que ocurre cuando "transformamos" las coordenadas de los puntos de nuestra distribución a ese sistema de coordenadas.
-Como podéis ver, la distribución tiende a "achatarse" aún más, en lugar de hacerse más "esférica".
-:::
-
-## Generalización: Ejes oblicuos
-
-<!-- # TODO: Matriz de correlaciones con correlación muy alta -->
-
-<!-- # TODO: Animación de distribución en ejes rectangulares a ejes correctos -->
-
-<!-- # TODO: Matriz de correlaciones transformada "identidad" -->
-
-::: notes
-El sistema de coordenadas correcto es el tercero, que "compensa" de alguna manera el achatamiento de la nube de puntos, de forma que las coordenadas transformadas que nos quedan (en los ejes ortogonales), los "normales" del gráfico) tienen una distribución "esférica" (con matriz de correlaciones "identidad").
+La respuesta correcta era la C, que "compensa" de alguna manera el achatamiento de la nube de puntos, de forma que las coordenadas transformadas que nos quedan (en los ejes ortogonales, los "normales" del gráfico) tienen una distribución "esférica" (con matriz de correlaciones "identidad").
 

### 2025-03-28 output/seminar-slideshow.qmd
Add "latent space vs. test space" (first) slide

@@
 
-## Espacio latente vs. espacio test
+## Espacio latente vs. espacio test {.smaller}
 
@@
 
-<!-- # TODO: Fórmula del modelo en coordenadas "transformadas" y equivalencia -->
-
-<!-- # TODO: Expresión de la transformación de la discriminación a coordenadas rectangulares -->
+::::::::: columns
+::::: column
+::: {.fragment fragment-index="1"}
+$$
+  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
+$$
+:::
+
+::: {.fragment fragment-index="2"}
+$$
+  P(Y_i = 1) = \Phi_L ({\mathbf{a}^*_i}^T \mathbf{\theta}^* + d_i)
+$$
+:::
+:::::
+
+::::: column
+::: {.fragment fragment-index="2"}
+$$
+  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
+$$
+:::
+
+::: {.fragment fragment-index="3"}
+$$
+  \mathbf{a}^*_i = \mathbf{Q} \mathbf{a}_i \\
+$$
+:::
+:::::
+:::::::::
+
+::: fragment
+$$
+  \mathbf{a}_i^T \mathbf{\theta} =
+    {\mathbf{a}^*_i}^T \mathbf{\theta}^* =
+    \mathbf{a}_i^T \mathbf{Q}^T \mathbf{P} \mathbf{\theta}
+$$
+:::
+
+::: fragment
+$$
+  \mathbf{Q} = (\mathbf{P}^{-1})^T
+$$
+:::
 
@@
 Viendo que en la literatura hay afirmaciones como esta, no me resulta totalmente descorazonador.
-Parece que no soy el único al que le ha supuesto un esfuerzo entenderlo (resultados de la "encuesta"?)
-
-Lo que ocurre en realidad, y esto es en realidad también un paralelismo con lo que ocurre en análisis factorial, es que para que el modelo sea invariante, los ítems también tienen que transformarse a un sistema de coordenadas ortogonal.
-Pero podemos ver que su matriz de transformación no es la misma que la de las puntuaciones de rasgo latente.
-Es su inversa (inversa traspuesta, más concretamente).
+Parece que no soy el único al que le ha supuesto un esfuerzo entenderlo (resultados de la "encuesta"?).
+
+Pues bien, lo que ocurre en realidad es lo siguiente:
+
+\[NEXT\]
+
+Este es el modelo de TRI multidimensional, recordaréis (ya lo presentamos, es como un modelo de regresión multivariada, donde los predcitores, $\theta$ son variables latentes).
+El modelo tiene que ser invariante; es decir, tiene que dar igual qué sistema de coordenadas usemos para representarlo (y esto por cierto también tiene su paralelismo en análisis factorial).
+
+\[NEXT\]
+
+Así que si transformamos los rasgos latentes de las personas a un sistema de coordenadas ortogonal, los ítems tienen que transformarse también.
+
+\[NEXT\]
+
+Pero recordaréis que llamé Q a su matriz de transformación, mientras que a la de los rasgos latentes la llamé P.
+
+\[NEXT\]
+
+Claro, para que se cumpla la invarianza, se tiene que cumplir esta igualdad: Este término debe ser el mismo, da igual el sistema de coordenadas que utilice, y por lo tanto P y Q no pueden ser la misma matriz.
+
+\[NEXT\]
+
+En realidad, para transformar los ítems hay que usar la matriz inversa (inversa traspuesta, más concretamente) de la matriz de transformación de los rasgos latentes.
+
+<!-- TODO: Next slide -->
 

### 2025-04-04 output/seminar-slideshow.qmd
Move figure illustrating correlation "disregard"
Add slide comparing latent and test spaces

@@
 
+## Supuestos: Ortogonalidad
+
+<!-- TODO: ACT en McDonald, 1999? -->
+
+::: r-center
+![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height="400px"}
+:::
+
+::: aside
+Figura tomada de @ackerman_multidimensional_2005-1
+:::
+
+::: notes
+No tiene sentido que cambie su discriminación multidimensional, y sin embargo la literatura está llena de ejemplos que calculan este parámetro con la fórmula de Reckase (en un espacio de "rasgos correlacionados", que en seguida veremos por qué eso es importante), y que no se preocupan por "los ejes" que representan esos rasgos de manera fidedigna.
+
+Para la representación gráfica esto puede valer, como dice Ackerman aquí.
+Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente; obtenemos parámetros de discriminación multidimensional "distorsionados".
+:::
+
 ## Generalización: Ejes oblicuos {.smaller}
@@
 
-transform_grid <- function(grid,
+transform_grid <- function(basis,
                            x_limits = c(-4, 4),
@@
 #| fig-width: 5
-basis <- cbind(
+rot_basis <- cbind(
   c(sqrt(.75), -sqrt(.25)),
@@
 
-transform_grid(basis)
+rot_grid <- transform_grid(rot_basis)
+
+rot_grid
 ```
@@
 #| fig-width: 5
-basis <- cbind(
+ts_basis <- cbind(
   c(1, 0),
@@
 
-transform_grid(basis)
+ts_grid <- transform_grid(ts_basis)
+
+ts_grid
 ```
@@
 #| fig-width: 5
-basis <- t(solve(basis))
-
-transform_grid(basis)
+ls_basis <- t(solve(ts_basis))
+
+ls_grid <- transform_grid(ls_basis)
+
+ls_grid
 ```
@@
 #| fig-width: 5
-basis <- cbind(
-  c(sqrt(.75), -sqrt(.25)),
-  c(sqrt(.25),  sqrt(.75))
-)
-
 thetas_transf <- thetas |>
@@
   pivot_longer(starts_with("trait")) |>
-  mutate(value = basis %*% value |> drop()) |>
+  mutate(value = rot_basis %*% value |> drop()) |>
   pivot_wider() |>
@@
 
-transform_grid(basis) +
+rot_grid +
   geom_point(
@@
 #| fig-width: 5
-basis <- cbind(
-  c(1, 0),
-  c(sqrt(.25), sqrt(.75))
-)
-
 thetas_transf <- thetas |>
@@
   pivot_longer(starts_with("trait")) |>
-  mutate(value = basis %*% value |> drop()) |>
+  mutate(value = ts_basis %*% value |> drop()) |>
   pivot_wider() |>
@@
 
-transform_grid(basis) +
+ts_grid +
   geom_point(
@@
 #| fig-width: 5
-basis <- t(solve(basis))
-
 thetas_transf <- thetas |>
@@
   pivot_longer(starts_with("trait")) |>
-  mutate(value = basis %*% value |> drop()) |>
+  mutate(value = ls_basis %*% value |> drop()) |>
   pivot_wider() |>
@@
 
-transform_grid(basis) +
+ls_scatter <- ls_grid +
   geom_point(
@@
   )
+
+ls_scatter
 ```
@@
 
-::: {.fragment fragment-index="2"}
+::: {.fragment fragment-index="3"}
 $$
@@
 
-::: {.fragment fragment-index="3"}
+::: {.fragment fragment-index="4"}
 $$
@@
 En realidad, para transformar los ítems hay que usar la matriz inversa (inversa traspuesta, más concretamente) de la matriz de transformación de los rasgos latentes.
-
-<!-- TODO: Next slide -->
-
-Y esto nos ha permitido "descubrir" (o darnos cuenta, más bien) de que hay dos "espacios vectoriales" distintos, uno en el que se encuentran las coordenadas de las puntuaciones factoriales, y otro distinto, que es el espacio en el que se encuentran las coordenadas de los ítems, y que es donde en realidad hay que representarlos.
-:::
-
-## Supuestos: Ortogonalidad
-
-<!-- # TODO: Review/delete slide? -->
-
-<!-- TODO: ACT en McDonald, 1999? -->
-
-::: r-center
-![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height="400px"}
-:::
-
-::: aside
-Figura tomada de @ackerman_multidimensional_2005-1
-:::
-
-::: notes
-No tiene sentido que cambie su discriminación multidimensional, y sin embargo la literatura está llena de ejemplos que calculan este parámetro con la fórmula de Reckase (en un espacio de "rasgos correlacionados", que en seguida veremos por qué eso es importante), y que no se preocupan por "los ejes" que representan esos rasgos de manera fidedigna.
-
-Para la representación gráfica esto puede valer, como dice Ackerman aquí.
-Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente; obtenemos parámetros de discriminación multidimensional "distorsionados".
-:::
-
-## Espacio latente vs. espacio test
+:::
+
+## Espacio latente vs. espacio test {.smaller}
+
+<br>
+
+::::::: columns
+::::: column
+<center>Espacio latente</center>
+
+:::: r-stack
+```{r grid-latent-space}
+#| fig-width: 6.5
+ls_grid
+```
+
+::: fragment
+```{r scatter-latent-space}
+#| fig-width: 6.5
+ls_scatter
+```
+:::
+::::
+:::::
+
+::: {.column .fragment}
+<center>Espacio test</center>
+
+```{r grid-test-space}
+#| fig-width: 6.5
+ts_grid
+```
+:::
+:::::::
+
+::: notes
+Y esto nos ha permitido "descubrir" (o darnos cuenta, más bien) de que hay dos "espacios vectoriales" distintos, uno en el que se encuentran las coordenadas de rasgo latente (las puntuaciones factoriales),...
+
+\[NEXT\]
+
+...
+y otro distinto, que es el espacio en el que se encuentran las coordenadas de los ítems, y que es donde en realidad habría que representarlos.
+
+Porque en este espacio es donde los ángulos entre los ejes son los que verdaderamente son una representación geométrica de las correlaciones entre las variables latentes.
+:::
+
+## Espacio latente vs. espacio test {.smaller}
 

### 2025-04-04 output/Letter_to_editor_reviewers.Rmd

@@
 
-To our understanding, the representation is always sample dependent because item parameters also are.
+We have found this comment puzzling as, to our understanding, what Reviewer 3 is asking is what we do in Figure 1 (p. 22 in our original submission, p. <!-- # TODO: Complete --> in the new one).
+We have shown the same items, with as many different cases as possible, represented in both orthogonal and nonorthogonal coordinates, and explained the interpretation of the multidimensional parameters in both spaces.
+We would appreciate if Reviwer 3 would kindly clarify this comment, so we can address it properly.
+
+On the other hand, to our understanding, the representation is always sample dependent because the item parameters also are.
 Our argument is that the non-orthogonal coordinate system is a more faithful representation of the latent covariance structure, as the cosines between the axes are the geometric equivalent of the latent trait correlations.
-We argue that this is the value of using this more complex representation, and so we have tried to convey in <!-- # TODO: Complete --> (please see p.
+We argue that this is the value of using this more complex representation, and so we have tried to convey in our Discussion section (please see p.
 31, line 19 to p.

### 2025-04-04 output/sandbox_slideshow.qmd
Versions latent vs. test space expanation slide

--- 2025-03-08 output/sandbox_slideshow.qmd
@@
 ```{r sources}
-source("R/LaTeX_math.R",   encoding = 'UTF-8')
+source("R/Formulae.R",     encoding = 'UTF-8')
 source("R/Output.R",       encoding = 'UTF-8')

### 2025-04-04 output/seminar-slideshow.qmd
Versions latent vs. test space expanation slide

@@
 
-::::::::: columns
+:::::::: columns
 ::: {.column width="33%"}
@@
 
-:::::: {.column .fragment width="33%"}
+::::: {.column .fragment width="33%"}
 <center>B</center>
 
-::::: r-stack
+:::: r-stack
 ```{r grid-ts-basis}
@@
 :::
+::::
 :::::
-::::::
 
@@
 :::
-:::::::::
+::::::::
 
@@
 
-> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by <!-- # TODO: Inner product -->.
+```{r space-inner-products}
+# Mahalanobis distance:
+TRAIT_VECTOR        <- latex_bf("\\theta")
+TRAIT_NORM          <- latex_norm(TRAIT_VECTOR)
+TRAIT_VECTOR_T      <- latex_transp(TRAIT_VECTOR)
+TRAIT_COV_INV_TRAIT <- latex(TRAIT_VECTOR_T, COV_MATRIX_INV, TRAIT_VECTOR)
+MAH_DIST_DEF        <- latex_sqrt(TRAIT_COV_INV_TRAIT)
+MAH_DIST_EQ         <- latex_eq(TRAIT_NORM, MAH_DIST_DEF)
+
+INNER_PROD_INV_COV_EQ <- latex_eq(COV_MATRIX_INV, INNER_PROD_TRANSF_DEF)
+
+# Latent space:
+TRAIT_VECTOR_1    <- latex_sub(TRAIT_VECTOR, 1)
+TRAIT_VECTOR_2    <- latex_sub(TRAIT_VECTOR, 2)
+TRAIT_VECTOR_1_T  <- latex_transp(TRAIT_VECTOR_1)
+
+INNER_PROD     <- latex_innerprod(TRAIT_VECTOR_1, TRAIT_VECTOR_2)
+INNER_PROD_DEF <- latex(TRAIT_VECTOR_1_T, COV_MATRIX_INV, TRAIT_VECTOR_2)
+INNER_PROD_EQ  <- latex_eq(INNER_PROD, INNER_PROD_DEF)
+
+# Test space:
+DISCR_VECTOR_J <- latex_sub(DISCR_VECTOR_ANY, AUX_INDEX)
+DISCR_VECTOR_T <- latex_transp(DISCR_VECTOR)
+
+INNER_PROD_DISCR     <- latex_innerprod(DISCR_VECTOR, DISCR_VECTOR_J)
+INNER_PROD_DISCR_DEF <- latex(DISCR_VECTOR_T, COV_MATRIX, DISCR_VECTOR_J)
+INNER_PROD_DISCR_EQ  <- latex_eq(INNER_PROD_DISCR, INNER_PROD_DISCR_DEF)
+```
+
+<br>
+
+@mahalanobis_reprint_2018:
+
+::::: columns
+::: {.column width="50%"}
+<center>$`r MAH_DIST_EQ`$</center>
+:::
+
+::: {.column .fragment width="50%"}
+$`r INNER_PROD_INV_COV_EQ`$
+:::
+:::::
+
+<br>
+
+::: fragment
+> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by $`r INNER_PROD_EQ`$.
 
 ---@zhang_theoretical_1999 [p. 221]
-
-<!-- #TODO: Ref. to Mahalanobis -->
-
-> \[...\] the *inner product* in \[the test space\] is consistently defined by <!-- # TODO: Inner product -->.
+:::
+
+::: fragment
+> \[...\] the *inner product* in \[the test space\] is consistently defined by
+>
+> $`r INNER_PROD_DISCR_EQ`$
 
 ---@zhang_theoretical_1999 [p. 221]
-
-::: notes
-Esto, curiosamente, aparece en la literatura como podéis ver; en el año 1999, en un artículo que no tiene nada que ver con esto.
-Lo menciona de soslayo y no profundiza en la deducción de los parámetros, como hemos hecho aquí.S olamente lo utiliza para resolver un problema "parcial" que se encuentra en la deducción de un índice de dimensionalidad, distinto de los parámetros de los que estamos hablando aquí.
-Pero podía haber obviado este problema y haber utilizado coordenadas ortogonales, sin importarle si había o no correlaciones entre los rasgos latentes, como ha hecho todo el mundo.
+:::
+
+::: notes
+Esto da lugar a algo muy curioso, y es que la "norma" de un punto en el espacio latente (la longitud de un vector aplicado en el origen que termina en ese punto) viene determinada por la distancia de Mahalanobis (esto es una reimpresión de 1936, no penséis que es una cosa de anteayer).
+
+\[NEXT\]
+
+Y eso se debe a que, para que la transformación en el espacio latente se cumpla, esa matriz de transformación (la que llamababmos P) tiene que ser "raíz cuadrada" de la inversa de la matriz de covarianzas.
+
+\[NEXT\]
+
+Y, curiosamente también, aparece en la literatura, en el año 1999, en un artículo que no tiene nada que ver con esto.
+Lo menciona de soslayo y no profundiza en su deducción como hemos hecho aquí.
+Solamente lo utiliza para resolver un problema "parcial" que se encuentra en la deducción de un índice de dimensionalidad, distinto de los parámetros de los que estamos hablando.
+Pero podía haber obviado este problema y haber utilizado coordenadas ortogonales, sin importarle si había o no correlaciones entre los rasgos latentes, como ha hecho todo el mundo (antes y después que ellos).
 
 Así que, todos mis respetos por estos autores.
+
 Sin embargo claro, ellos estaban hablando de otra cosa, y sólo tocaban este tema de paso.
@@
 La gente ha usado los modelos, hecho simulaciones, etc., usando distribuciones correlacionadas, y ha usado las fórmulas de Reckase para transformar entre parámetros multidimensionales y los parámetros "lineales" (podríamos llamarlos, los de la formulación original) del modelo.
+
+\[NEXT\]
+
+Y lo más importante: Esto nos demuestra efectivamente que tenemos dos espacios distintos: Uno para los parámetros de las personas (espacio latente) y otro para los ítems (espacio test).
+Y esto es así porque la operación del producto interno está definida de manera distinta en los dos espacios; si la matriz del producto interno es la inversa de las covarianzas en el espacio latente, entonces en el espacio test, donde la transformación del parámetro de discriminación es inversa (como recordaréis), pues la matriz del producto interno es la matriz de covarianzas.
 :::

### 2025-04-10 output/Letter_to_editor_reviewers.Rmd
Add references taken from Harman 1976

@@
 
-We intended to refer to <!-- # TODO: Complete -->
+We intended to refer to section 4.8 (*Distance and Angle in General Cartesian Coordinates*), where it is explained how to compute coordinates that "involve the inclination of the reference axes" [@harman_modern_1976, p.
+60].
 Please note that, as we do not have access to the 2nd edition anymore, we have used the 3rd one in our new version, and updated all the references accordingly.
+
+Regarding the scatter plot you mention, we have not been able to find in (at least not in the 3^rd^ one we are using now, even though the example you refer to appears at the end of section 2.5, pp.
+22-23).
+However, we acknowledge such a plot is the most common representation of the values realized by different cases (persons, in our case) in two variables, correlated or not.
+In this regard, we would kindly refer you to our third response to your comments, in p.
+<!-- TODO: Complete --> of this letter, where we discuss the representation of person parameters in the last two paragraphs.
+<!-- TODO: Possibly complete with references to other responses? -->
 

### 2025-04-24 output/seminar-slideshow.qmd
Add parameter expressions and item representation

@@
 library(mvtnorm)
+library(patchwork)
 ```
@@
 source("R/Mirt_toolbox.R",              encoding = 'UTF-8')
+source("src/Graphical_example_paper.R", encoding = 'UTF-8')
 ```
@@
 
-<!-- # TODO: Fórmulas de los parámetros multidimensionales -->
+<br>
+
+$$
+`r latex_eq(MDISC_COV_PARAM, DISCR_VECTOR_COV_MODULE)`
+$$
+
+<br>
+
+::: fragment
+$$
+`r latex_eq(
+     MIL_COV_PARAM,
+     latex_curlybraces("$DISTANCE_COV_DEF$, $DIR_COS_ITEM_VEC_COV_DEF$")
+   )`
+$$
+:::
 
@@
 
-Muy importante, la distancia con signo y la discriminación multidimensional son iguales, no importa en qué espacio queramos obtenerlas y representarlas.
+Muy importante, la distancia con signo y la discriminación multidimensional son iguales, no importa en qué espacio (latente, o test) queramos obtenerlas y representarlas.
 Pero los cosenos directores es necesario calcularlos en el espacio test, no en el espacio de puntuaciones latentes, como decía Reckase, porque este es el espacio que va a representar en realidad la relación entre las variables latentes.
 :::
+
+## Representación gráfica
+
+```{r item-plot-out}
+#| fig-cap:   Espacio test ortogonal (izquierda) y oblicuo (derecha)
+#| fig-align: center
+#| fig-asp:     .5
+#| fig-width: 13
+
+plot_orth + plot_oblique ## TODO: Tema ggplot
+```
 

### 2025-04-24 output/seminar-slideshow.qmd
Correct notes in latent space basis deduction

@@
 
-::: column
+::: {.column .fragment}
 <!-- TODO: Animate? -->
@@
 
-Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a ejes oblicuos tenemos que representados en un espacio no ortogonal (aquí también hablaría de "unidades no" estandarizadas", es decir, que la distancia entre el 0 y el 1 pueda ser "distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
+Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a ejes oblicuos tenemos que representados en una base no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el 0 y el 1 pueda ser "distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
 
@@
 
-Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de ese espacio oblicuo (el que sea, aún no sabemos cuál es) a un espacio ortogonal.
+Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de esa base oblicua (la que sea, aún no sabemos cuál es) a una base ortogonal.
 Esto es una transformación de las coordenadas, como la que aplicamos antes para calcular las coordenadas del ítem tras aplicar la rotación oblicua, pero aplicada en este caso al vector de rasgo latente.
@@
 
-Estas nuevas coordenadas (transformadas), que ahora están en un espacio ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^*$ la vamos a llamar) sea una matriz identidad (unos en la diagonal, ceros en el resto).
+Estas nuevas coordenadas (transformadas), que ahora están en una base ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^*$ la vamos a llamar) sea una matriz identidad (unos en la diagonal, ceros en el resto).
 Pero hay que tener en cuenta que esto es un "artefacto matemático".

### 2025-04-25 output/seminar-slideshow.qmd
Correct item indices in negative item parameters
Update "quizz" prompt slide

@@
   \begin{matrix}
-    \mathbf{a}_1 &=
+    \mathbf{a}_2 &=
       \begin{bmatrix}
@@
       \end{bmatrix} &
-    d_1 = 0
+    d_2 = 0
   \end{matrix}
@@
   \begin{matrix}
-    \mathbf{a}_1 &=
+    \mathbf{a}_3 &=
       \begin{bmatrix}
@@
       \end{bmatrix} &
-    d_1 = 0
+    d_3 = 0
   \end{matrix}
@@
   \begin{matrix}
-    \mathbf{a}_1 &=
+    \mathbf{a}_4 &=
       \begin{bmatrix}
@@
       \end{bmatrix} &
-    d_1 = 0
+    d_4 = 0
   \end{matrix}
@@
       \end{bmatrix} &
-    d_1 = -0.5
+    d_2 = -0.5
   \end{matrix}
@@
       \end{bmatrix} &
-    d_1 = -0.5
+    d_3 = -0.5
   \end{matrix}
@@
       \end{bmatrix} &
-    d_1 = 0.25
+    d_4 = 0.25
   \end{matrix}
@@
 :::::: columns
+::: {.column width="33%"}
+```{r}
+#| fig-width:  5.5
+#| fig-height: 5.5
+thetas_transf |>
+  ggplot(mapping = aes(trait_1, trait_2)) +
+  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    labels = \(breaks) '' |> rep(length(breaks)),
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~}", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    labels = \(breaks) '' |> rep(length(breaks)),
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks   = element_blank(),
+    axis.line    = element_blank(),
+    axis.title   = element_markdown(),
+    axis.title.y = element_markdown(),
+    panel.grid   = element_blank()
+  )
+```
+:::
+
 ::: {.column .fragment width="33%"}
-:::
-
-::: {.column .fragment width="33%"}
+```{r question-mark}
+fa(
+  "question",
+  fill           = "#929acb",
+  height         = "200px",
+  vertical_align = "-10em",
+  margin_left    = "4em"
+)
+```
 :::

### 2025-04-25 output/seminar-slideshow.qmd
Correct minor errata

@@
 knitr:
-  # opts_knit: # Does not work, see line 88
+  # opts_knit: # Does not work, see line 102
   #   root_dir:  here::here()
@@
 library(scales)
+library(fontawesome)
 library(mvtnorm)
@@
 #| include: false
-opts_knit$set(root.dir = here::here()) # See lines 41-42
+opts_knit$set(root.dir = here::here()) # See lines 43-44
 
@@
     \right) =
-      \left(
+      \Phi_N \left(
         \frac{1}{\sqrt{var(U_i)}} \mathbf{\lambda}_i^T \mathbf{F} +

### 2025-04-25 output/seminar-slideshow.qmd
Add oblique basis representation

@@
 
-## Representación gráfica
+## Representación gráfica completa
 
@@
 #| fig-width: 13
+
+plot_orth <- plot_orth +
+  scale_color_manual(values = ITEMS_PALETTE) +
+  theme(
+    axis.ticks       = element_blank(),
+    legend.title     = element_blank(),
+    panel.grid.minor = element_blank()
+  )
+
+plot_oblique <- plot_oblique +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  theme(axis.ticks = element_blank())
 

### 2025-04-25 output/seminar-slideshow.qmd
Improve flow in concluding slides

@@
 
+<br>
+
 Parámetros multidimensionales generalizados
@@
 
+<br>
+
+::: fragment
 Distinción espacio latente - espacio test
+:::
 
@@
     Hemos visto que esto ya lo mencionaron Zhang y Stout, hace 25 años, pero no proporcionaron una "deducción" de estos parámetros multidimensionales, cosa que nosotros hacemos.
+
+\[NEXT\]
 
@@
 
+<br>
+
 -   Paralelismo TRI multidimensional - AF
 
+<br>
+
+::: fragment
 -   Extensiones del modelo M2PL: Escala graduada, respuesta nominal, 3PL...
-
+:::
+
+<br>
+
+::: fragment
 -   Investigación en otros modelos
+:::
 
@@
 
+\[NEXT\]
+
 El modelo multidimensional logístico de 2 parámetros se puede considerar como "anidado", o un caso más simple de varios modelos.
@@
 Pero un investigador no debería dar los parámetros sin más de ese modelo, sin proporcionar una prueba formal de que se aplica esa fórmula y se cumplen los supuestos.
+
+\[NEXT\]
+
+Y por supuesto, existen muchos otros modelos psicométricos, muy útiles y aplicados en la investigación en Psicología y Educación, en los cuales no está claro aún cómo generalizar estos resultados al ámbito multidimensional.
 :::

### 2025-05-06 output/Main_text.Rmd
Changed agreed with @mluzardov in meeting

@@
 the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
-Fourth, $`r DISCR_VECTOR`$ is expressed on an algebraic basis,
+Fourth, $`r DISCR_VECTOR`$ is expressed in some algebraic basis,
 which the MIRT literature implicitly assumes to be the canonical basis
-(we will show that this assumption is not necessary though).
+(and also the same as $`r LS_BASIS`$;
+we will show that neither of these assumptions is necessary though).
 Finally, it is worth highlighting that

### 2025-05-14 output/Main_text.Rmd
Add changes agreed with @mluzardov in meeting

@@
 as a function of how we define and interpret the latent space.
+(Interestingly, we will show that, when the latent space basis is defined
+as a geometric mapping of the covariance structure,
+we arrive at a definition of the norm that is given by
+the Mahalanobis [-@mahalanobis_reprint_2018] distance.)
 Next, we discuss their properties,
@@
 Consider a basis $`r LS_BASIS_EQ`$,
-with $`r LS_BASIS`$ generally non-orthonormal.
+with $`r LS_BASIS`$ not necessarily orthonormal.
 Let $`r LATENT_SPACE`$ be a $`r N_DIMS`$-dimensional latent space
@@
 $`r LS_BASIS`$, the geometric coordinates $`r TRAIT_COORDS`$ of
-$`r TRAIT_VECTOR`$ in $`r LS_BASIS`$ are equal to the algebraic ones,
-i.e., $`r TRAIT_GEOM_ALGEBR_EQ`$.
+$`r TRAIT_VECTOR`$ in $`r LS_BASIS`$ are equal to the algebraic ones, i.e.,
+in Equation \@ref(eq:M2PL-formulation) we identify $`r TRAIT_GEOM_ALGEBR_EQ`$.
 
@@
 First, the transformation to the orthonormal bases of the two spaces allows
-mapping the $`r MODEL_ACRONYM`$ parameters indistinctly in the two spaces,
+mapping the $`r MODEL_ACRONYM`$ parameters interchangeably in the two spaces,
 by projecting one space onto the other.
@@
 no matter in which space we represent the coordinates.
-Second, and most importantly, if the test space is
-where the item parameters naturally lie, they must be represented in this space,
-and not in the latent space.
+Second, and most importantly,
+given that the test space is defined as the set of  item parameters,
+as we did before (see section Test Space Definition),
+then they should be represented in this space, and not in the latent space.
 Therefore, we may find the direction angles of $`r DISCR_VECTOR`$

### 2025-05-14 output/Letter_to_editor_reviewers.Rmd
Make minor rephrasings in letter

@@
 
-<!-- TODO: Reply -->
+We acknowledge and apologize for the lack of clarity in this sentence.
+Please refer to our 10th response to Reviewer 3 in p. <!-- TODO: Complete --> of this letter.
 
@@
 
-We think the motivation and psychometric utility of our approach are more related to the interpretation of the item parameters than with the latent abilities.
-To our understanding, the better interpretation of the item parameters that results from our approach provides indeed a better insight into how the results are used.
+We think the motivation and psychometric utility of our approach are more related to the interpretation of the item parameters than with the latent abilities, which we believe exceeds the scope of this manuscript and could be the focus of another paper.
+Please note that we only use the transformation in the latent space to an orthonormal basis as a reasoning artifact to obtain the proper expression of the item parameters in the general case, and that we do it in an abstract sense, before even considering whether the original basis is a representation of the covariance matrix.
+
+However, to our understanding, the more precise interpretation of the item parameters that results from our approach provides indeed a better insight into how the results are used.
 A multidimensional representation of the items provides benchmarks for comparing different latent trait levels, as discussed by @reckase_difficulty_1985.
@@
 
-Please see our answer in p. 3 of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
+Please see our answer in p. <!-- TODO: Complete (3?) --> of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
 <!-- TODO: Complete -->
@@
 
-We intended to refer to section 4.8 (*Distance and Angle in General Cartesian Coordinates*), where it is explained how to compute coordinates that "involve the inclination of the reference axes" [@harman_modern_1976, p.
+We intended to refer to section 4.8 (*Distance and Angle in General Cartesian Coordinates*), where he explains how to compute coordinates that "involve the inclination of the reference axes" [@harman_modern_1976, p.
 60].
@@
 7 line 8 in our original submission, or p.
-<!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen.
+<!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen, which does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction.
 
-We argue that the test space and the latent space are two distinct spaces, even when the basis of the latent space (and, consequently, that of the test space) is orthonormal.
+We also argue that the test space and the latent space are two distinct spaces, even when the basis of the latent space (and, consequently, that of the test space) is orthonormal.
 It is our understanding that,  due to a lack of need for this distinction, this is overlooked in Reckase's [-@reckase_difficulty_1985] and Reckase & McKinley's [-@reckase_discriminating_1991] original formulations.

### 2025-05-22 output/Letter_to_editor_reviewers.Rmd
Implement Grammarly corrections

@@
 We acknowledge and apologize for the lack of clarity in this sentence.
-Please refer to our 10th response to Reviewer 3 in p. <!-- TODO: Complete --> of this letter.
+Please refer to our 10th response to Reviewer 3 on p. <!-- TODO: Complete --> of this letter.
 
@@
 that is, $2 a_i$.
-Similarly, the correlation between two simple items items,
+Similarly, the correlation between two simple items,
 each one measuring a distinct dimension,
@@
 For example, 
-<!-- # TODO: Example(s) from Gecko search -->
-We believe this is also what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution:
-As you have noted yourself [@ackerman_multidimensional_2005-1, p. 7].
+<!-- # TODO: Example(s) from Gecko search -->.
+We believe this is also what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution, as you have noted yourself [@ackerman_multidimensional_2005-1, p.
+7].
 <!-- # TODO: Correctly expressed? -->
-While we consider this is out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our method may solve the problem.
+While we consider this to be out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our method may solve the problem.
 
@@
 Computing the covariance errors in the former would require taking into account the standard errors of the item parameters themselves and at least the standard errors of the correlations (in the best-case scenario, assuming standard latent dimensions; otherwise, those of the variances and covariances).
-However, to our knowledge, multidimensional item parameters, as defined by @reckase_difficulty_1985 and @reckase_discriminating_1991 have been previously defined and used only from a theoretical perspective, without ever taking into account the uncertainty resulting from the model estimation.
+However, to our knowledge, multidimensional item parameters, as defined by @reckase_difficulty_1985 and @reckase_discriminating_1991, have been previously defined and used only from a theoretical perspective, without ever taking into account the uncertainty resulting from the model estimation.
 
 Regarding the "original" item parameters in the classical M2PL metric, we understand that the standard errors of the item parameters are independent of the definition of the multidimensional parameters used.
-Software such as R package {mirt} [@chalmers_mirt_2012] use a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
-As far as we know, these standard error estimates are computed by the usual methods, involving e.g. Taylor series, bootstrapping, or similar.
+Software such as the R package {mirt} [@chalmers_mirt_2012] uses a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
+As far as we know, these standard error estimates are computed by the usual methods, involving, e.g., Taylor series, bootstrapping, or similar.
 <!-- TODO: Check SE methods; complete? -->
@@
 
-However, to our understanding, the more precise interpretation of the item parameters that results from our approach provides indeed a better insight into how the results are used.
+However, to our understanding, the more precise interpretation of the item parameters that results from our approach indeed provides a better insight into how the results are used.
 A multidimensional representation of the items provides benchmarks for comparing different latent trait levels, as discussed by @reckase_difficulty_1985.
@@
 
-Please see our answer in p. <!-- TODO: Complete (3?) --> of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
+Please see our answer on p. <!-- TODO: Complete (3?) --> of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
 <!-- TODO: Complete -->
@@
 (in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
-Then, we may compute the direction cosines in the rotated solution applying
+Then, we may compute the direction cosines in the rotated solution by applying
 Equation (24).
@@
 we have preferred not to include the repository as additional material.
-However, we can do so if the journal accepts linking an external repository as supplenetary material, or otherwise grant access to the editor and reviewers to this repository.
+However, we can do so if the journal accepts linking an external repository as supplementary material, or otherwise grant access to the editor and reviewers to this repository.
 <!-- TODO: Publish repo and link Software Legacy DOI? -->
 
-# Responses to Reviwer 1:
+# Responses to Reviewer 1:
 
@@
 Our introduction is a bit less than four pages long.
-We have provided concise paragraphs about the Multidimensional IRT model and its parameters, their assumptions, and our rationale to relax them.
+We have provided concise paragraphs about the Multidimensional IRT model and its parameters, their assumptions, and our rationale for relaxing them.
 Finally, a last paragraph presents the objectives of our research.
@@
 Our work is completely theoretical and thus uses a purely deductive method.
-Therefore, we do not have any hypothesis to test, but the results of the formal derivations that we present, adressing the objective crearly described at the end of our introduction.
+Therefore, we do not have any hypothesis to test, but the results of the formal derivations that we present, which address the objective described at the end of our introduction.
 
@@
 The datasets we use in our paper are examples of item parameter estimates, taken from the literature.
-As they only serve exemplification purposes, and are not the main focus of our article, we deem it more appropriate to describe their details in the corresponding section of the manuscript.
+As they only serve exemplification purposes and are not the main focus of our article, we deem it more appropriate to describe their details in the corresponding section of the manuscript.
 We have added a short reference to them in the introduction, to make sure their pertinence as examples is clear.
+
+<!-- TODO: Include text added -->
 
@@
 We think the context makes it clear that we are referring to "within-item" multidimensionality.
-In fact, it is clear from the manuscript that the results are trivial in the "between-item" multidimensionality case,
+In fact, the results are trivial in the "between-item" multidimensionality case,
 as the `r MDISC_SYM` parameter reduces to the only non-null component of the $`r DISCR_VECTOR`$ parameter.
@@
 is non-null.
-However, the model can also account for *between-item multidimensionality*,
+However, the model can also account for *between-item multidimensionality*
 if all items are unidimensional
@@
 
-# Responses to Reviwer 2:
+# Responses to Reviewer 2:
 
@@
 Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
-To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated in p. 7.
+To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7.
 (That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
-However, we acknowledge that we have skipped a step in our discrourse, thus making harder than necessary to follow along.
+However, we acknowledge that we have skipped a step in our discourse, thus making it harder than necessary to follow along.
 To correct this, we have modified the immediately preceding paragraph to the following:
@@
 <!-- # TODO: Confirm equation references -->
-> Of course, from Equations (24) and (25) it is easy to express
+> Of course, from Equations (25) and (26), it is easy to express
 the coordinates in terms of the parameters in the original model formulation.
@@
 
-# Responses to Reviwer 3:
+# Responses to Reviewer 3:
 
@@
 We have now added a paragraph to our introduction to try to explain this fact intuitively.
-Please refer to our answer to the editor in p. 3 of this letter for more details on this addition.
+Please refer to our answer to the editor on p. 3 of this letter for more details on this addition.
 
@@
 If the original, correlated coordinates of the persons are represented in the orthogonal space, the shape of the scatter plot indeed represents the correlation between the latent trait coordinates.
-However, our rationale is that a basis may be used such that, when the coordinates are transformed to an orthonormal one, the resulting scatter plot is spherical (i.e.
-its covariance is an identity matrix).
+However, our rationale is that a basis may be used such that, when the coordinates are transformed to an orthonormal one, the resulting scatter plot is spherical (i.e., its covariance is an identity matrix).
 This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
@@
 The distances between points are given by the norm,
-be it in the latent space or in the test space.
+be it in the latent space or the test space.
 We propose that the norm of both items and trait vectors should be invariant to
 changes of basis and, interestingly, when considering the covariance-based version of the multidimensional parameters, the distance between points in the latent space is given by the Mahalanobis distance.
-We have reflected this in our new version, by adding the following to the end of the "Covariance-based Version" section:
+We have reflected this in our new version by adding the following to the end of the "Covariance-based Version" section:
 
@@
 
-> [...] Once again, this is relevantimportant in applications where potentially different bases are involved.
+> [...] Once again, this is relevant in applications where potentially different bases are involved.
 Furthermore, we have found that using this version yields a norm in
@@
 We apologize for the lack of clarity.
-We intended to refer to footnote 1 in page 16; we have now completed the reference.
+We intended to refer to footnote 1 on page 16; we have now completed the reference.
 
@@
 
-Regarding the scatter plot you mention, we have not been able to find in (at least not in the 3^rd^ one we are using now, even though the example you refer to appears at the end of section 2.5, pp.
+Regarding the scatter plot you mention, we have not been able to find it (at least not in the 3^rd^ one we are using now, even though the example you refer to appears at the end of section 2.5, pp.
 22-23).
-However, we acknowledge such a plot is the most common representation of the values realized by different cases (persons, in our case) in two variables, correlated or not.
-In this regard, we would kindly refer you to our third response to your comments, in p.
+However, we acknowledge that such a plot is the most common representation of the values realized by different cases (persons, in our case) in two variables, correlated or not.
+In this regard, we would kindly refer you to our third response to your comments, on p.
 <!-- TODO: Complete --> of this letter, where we discuss the representation of person parameters in the last two paragraphs.
@@
 scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
-Note thouth that, as $`r LATENT_SPACE`$ is originally represented in
+However, note that, as $`r LATENT_SPACE`$ is originally represented in
 $`r LS_BASIS`$, the geometric coordinates $`r TRAIT_COORDS`$ of
@@
 its inner product matrix.
-(Note that in the original derivations by @reckase_difficulty_1985
-and @reckase_discriminating_1991 both matrices simplify to the identity matrix,
-so there was no need to make this distinction.)
+[Note that in the original derivations by @reckase_difficulty_1985;
+and @reckase_discriminating_1991, both matrices simplify to the identity matrix,
+so there was no need to make this distinction.]
 This implies that the axes defined by the test space basis, [...]
@@
 
-As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response in p.
+As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response on p.
 <!-- TODO: Complete --> in this letter).
@@
 Therefore, even when computing the coordinates in a non-orthonormal basis, we need to transform them to the canonical one to plot them.
-This is explained in our manuscript, in p. 19 (lines 9-14) in the original version, and in p. <!-- # TODO: Complete --> (lines <!-- # TODO: Complete -->) in the new submission.
+This is explained in our manuscript, on p. 19 (lines 9-14) in the original version, and on p. <!-- # TODO: Complete --> (lines <!-- # TODO: Complete -->) in the new submission.
 We are unsure about how to make this clearer, we are afraid, so we would appreciate any suggestions.
@@
 We have shown the same items, with as many different cases as possible, represented in both orthogonal and nonorthogonal coordinates, and explained the interpretation of the multidimensional parameters in both spaces.
-We would appreciate if Reviwer 3 would kindly clarify this comment, so we can address it properly.
+We would appreciate it if Reviewer 3 would kindly clarify this comment, so we can address it properly.
 
@@
 
-We acknowledge though that this may have been explained without enough clarity, and therefore striven to improve this discussion.
+We acknowledge, though, that this may have been explained without enough clarity, and therefore striven to improve this discussion.
 The aforementioned paragraph now reads:
@@
 and not the ones defined by the latent space basis,
-are a geometrical representation of the covariance structure:
+are a geometrical representation of the covariance structure.
 Indeed, the cosines of the angles among the test space axes are

### 2025-05-22 output/Main_text.Rmd
Implement Grammarly corrections

@@
 that is, $`r MDISC_VALUE_COLLIN`$.
-Similarly, the correlation between two simple items items,
+Similarly, the correlation between two simple items,
 each one measuring a distinct dimension,
@@
 is non-null.
-However, the model can also account for *between-item multidimensionality*,
+However, the model can also account for *between-item multidimensionality*
 if the $`r DISCR_VECTOR`$ parameter only has one non-null component.
@@
 and, as $`r DISCR_VECTOR`$ is expressed in this basis,
-its algebraic vector values correspond to its coordinates, i.e.,
+its geometric coordinates in $`r TEST_SPACE_BASIS`$
+are equal to its algebraic ones,
+i.e., in Equation \@ref(eq:M2PL-formulation) we identify
 $`r DISCR_VECTOR_COORDS_EQ`$.
@@
 Second, and most importantly,
-given that the test space is defined as the set of  item parameters,
-as we did before (see section Test Space Definition),
+given that the test space is defined as the set of item parameters
+(see section *Test Space Definition*),
 then they should be represented in this space, and not in the latent space.
@@
 Of course,
-from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
+from Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result),
 it is easy to express the coordinates in terms of the parameters in the original
@@
 (in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
-Then, we may compute the direction cosines in the rotated solution applying
+Then, we may compute the direction cosines in the rotated solution by applying
 Equation \@ref(eq:dir-cos-vec-cov).

### 2025-05-22 output/Letter_to_editor_reviewers.Rmd
Add example of orthogonal (miss)-interpretation

@@
 There are several references in the literature that have found contradictory results that, we believe, are due to disregarding the latent covariance structure (i.e., using Reckase & McKinley's formulae in a non-orthonormal space).
-For example, 
-<!-- # TODO: Example(s) from Gecko search -->.
-We believe this is also what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution, as you have noted yourself [@ackerman_multidimensional_2005-1, p.
-7].
-<!-- # TODO: Correctly expressed? -->
+For example, @min_evaluation_2007 [p.
+49] computes the item parameters in the M2PL from a priori fixed values of $MDISC_i$ and $MDIFF_i$ (the name they use for $MID_i$), using @reckase_difficulty_1985 and @reckase_discriminating_1991's formulae.
+They find that "[...] bias [...] was more apparent when ability distribution was separated from the
+default distribution, such as was the case in groups 2 to 5".
+Although they attribute this to a possible "degree of confounding among errors of parameter estimation and scale
+transformation", their groups 2 to 5 have all correlations equal (or almost equal) to .5, as it happens.
+
+We believe this is also what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution, as this editor himself has noted before [@ackerman_multidimensional_2005-1, p.
+11].
 While we consider this to be out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our method may solve the problem.

### 2025-05-22 output/Letter_to_editor_reviewers.Rmd
Reference GitHub repo as supplementary material

@@
 Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software [e.g., {mirt}, @chalmers_mirt_2012; flexMIRT, @cai_flexmirt_2024].
-As the journal guidelines require a specific format for the Supplementary Material,
-<!-- TODO: Check this -->
-we have preferred not to include the repository as additional material.
-However, we can do so if the journal accepts linking an external repository as supplementary material, or otherwise grant access to the editor and reviewers to this repository.
-<!-- TODO: Publish repo and link Software Legacy DOI? -->
+As the journal guidelines [require a specific format for the Supplementary Material](https://www.cambridge.org/core/services/authors/publishing-supplementary-material),
+we chose not to include the repository as additional material.
+However, we have checked and the journal accepts and also encourages publishing additional materiales in [a proper, trusted repository](https://www.cambridge.org/core/services/where-to-share-your-data).
+We have thus made public [our GitHub repository for the manuscript](https://github.com/DaniMori/mirt-parameters) and [indexed it in Software Heritage](https://archive.softwareheritage.org/swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af) for being identified with a SWHID.
+can do so if the journal accepts linking an external repository as supplementary material, or otherwise grant access to the editor and reviewers to this repository.
 

### 2025-05-22 output/Letter_to_editor_reviewers.Rmd
Add verbatim of reference to datasets in paper

@@
 As they only serve exemplification purposes and are not the main focus of our article, we deem it more appropriate to describe their details in the corresponding section of the manuscript.
-We have added a short reference to them in the introduction, to make sure their pertinence as examples is clear.
+We have added a short reference to them in the last paragraph of our introduction (p. <!-- TODO: Complete -->, lines <!-- TODO: Complete -->), to make sure their pertinence as examples is clear.
 
-<!-- TODO: Include text added -->
+> These examples, from the cognitive and the non-cognitive domain, respectively,
+are taken from the MIRT literature:
+One of them is a classic dataset, often used to exemplify
+the use of the `r MODEL_ACRONYM` model [@reckase_multidimensional_2009];
+the other one, an application in marketing research [@tezza_modelo_2018].
 

### 2025-05-22 output/Letter_to_editor_reviewers.Rmd
Explain misuse of "indistinctly"

@@
 
-This is the expression of a vector represented in a basis, in this case, representing that the vector can be represented in each of the two bases indistinctly.
+This is the expression of a vector represented in a basis, in this case, representing that the vector can be represented in each of the two bases interchangeably.
 Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:
@@
 
-<!-- # TODO: Add response -->
+This is a mistake on our side, we are afraid to admit, product of a careless translation of a false friend from our mother tongue Spanish ("indistintamente"), which means "indifferently".
+We have changed the term by the mathematically appropriate "interchangeably" in our new version.
 

### 2025-05-22 output/Main_text.Rmd
Reply to "comparability of metrics" in spaces

@@
 and present two examples of application of our results.
-These examples, from the cognitive and the non-cognitive domain respectively,
+These examples, from the cognitive and the non-cognitive domain, respectively,
 are taken from the MIRT literature:

### 2025-05-23 output/Letter_to_editor_reviewers.Rmd
Address "dimensional sensitivity" comment

@@
 
-<!-- TODO: Reply -->
+We acknowledge we may have overlooked this distinction.
+To our understanding, a highly correlated latent space may imply a restricted dimensionality (as indicated by the rank of the covariance matrix) of the latent space.
+We consider that this example, being an edge case (of unidimensionality), is very illustrative for our purposes, because it allows the reader to understand easily that all items should be aligned with the only dimension being measured and thus that grasp the intuitive notion that all direction cosines should be expected to be equal to 0.
+However, in doing so, we have skipped a step in our reasoning, assimilating the dimensionality of the latent space with the dimensional sensitivity of the items.
+
+To clarify better this point, we have tried to express better this distintion, emphasizing that the dimensionality of the latent space is an upper limit of the dimensional sensitivity of the items.
+The aforementioned paragraph now reads (p. <!-- # TODO: Complete -->, lines <!-- # TODO: Complete -->):
+
+> [...] In that example,
+he uses a covariance matrix that practically has rank 1,
+implying that the (alleged) three dimensions collapse into a single one.
+This in turn makes the latent space unidimensional in practice.
+As the dimensional sensitivity of the latent space sets the upper limit of
+the dimensional sensitivity of a test measuring those traits,
+it follows that the test must be unidimensional.
+In such a case, when all the items are strictly aligned with
+the one dimension they measure,
+we should expect the direction angles to be extremely close to 0 for all items.
+The reader interested can check that this is the case for
+the covariance-based version of the direction angles
+in the table in Online Resource 1.
 
@@
 
-<!-- TODO: Reply -->
+We are not completely sure to have understood this comment.
+The discrimination values in the original `r MODEL_ACRONYM` metric are comparable, in the sense that they yield the same model probabilities for the same latent trait coordinates, independently of the covariance structure of the latent space.
+However, the covariance-version of the multidimensional discrimination is different, and will indeed be dependent on the metric of the test space, which is defined by its geometrical basis, which determines the directions and scales of the axes.
+In this case, the basis has been chosen to be a geometric representation of the correlation matrix (as given by the expression $`r INNER_PROD_CORR_COND`$, please see p.
+<!-- # TODO: Complete -->, line <!-- # TODO: Complete -->, in the new version), and therefore, the scales are not affected (only the relative directions among the axes).
+
+We would kindly ask to elaborate further on this question if the Reviewer feels it has not been completely addressed.
 

### 2025-05-23 output/Main_text.Rmd
Address "dimensional sensitivity" comment

@@
 In that example, he uses a covariance matrix that practically has rank 1,
-implying that the (alleged) three dimensions collapse into a single one, being
-thus the test unidimensional in practice.
-In such a case, the covariance-based direction angles
-are extremely close to 0 for all items, as one would expect
-from a unidimensional test, where all items are strictly aligned with the one
-dimension they measure. The reader interested can check out the parameters
-computed in that case in the table in Online Resource 1.
+implying that the (alleged) three dimensions collapse into a single one.
+This in turn makes the latent space unidimensional in practice.
+As the dimensional sensitivity of the latent space sets the upper limit of
+the dimensional sensitivity of a test measuring those traits,
+it follows that the test must be unidimensional.
+In such a case, when all the items are strictly aligned with
+the one dimension they measure,
+we should expect the direction angles to be extremely close to 0 for all items.
+The reader interested can check that this is the case for
+the covariance-based version of the direction angles
+in the table in Online Resource 1.
 

### 2025-05-26 output/Main_text.Rmd
Correct citation inside parenthesis
(Used APA 6th style reference, unable to find APA 7th style but assumed it must not be very different.)

@@
 as a function of how we define and interpret the latent space.
-(Interestingly, we will show that, when the latent space basis is defined
+[Interestingly, we will show that, when the latent space basis is defined
 as a geometric mapping of the covariance structure,
 we arrive at a definition of the norm that is given by
-the Mahalanobis [-@mahalanobis_reprint_2018] distance.)
+the @mahalanobis_reprint_2018, distance.]
 Next, we discuss their properties,

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Address "dimensional sensitivity" comment properly
After reading carefully Reckase's (2009) chapter 7, the dimensional sensitiviy comment is properly addressed.

@@
 
-<!-- TODO: Reply -->
+To our understanding, one may assume a theoretical dimensional sensitivity of a test space that is higher than the dimensionality of the latent space.
+We acknowledge we may have overlooked this distinction.
+However, in such a case, the (minimum) dimensionality of the data would be "the smaller of the dimensions or variation for the people and the dimensions of sensitivity for the test items" (@reckase_multidimensional_2009, p.
+182).
+That is, the dimensional sensitivity of the data would nevertheless be restricted by the dimensionality of the latent space.
+This implies that the parameter values one can estimate for the items will represent a lower-dimensional solution, which would be the projection of the theoretically higher-dimensionally sensitive parameters on the lower-dimensional latent trait space, as discussed in p.
+182 of @reckase_multidimensional_2009.
+In practice, this implies that this solution is indistinguishable from one where the items are simply sensitive to the lower number of dimensions, with their parameters given by such lower-dimensional solution which would correspond to their projections from the alleged higher-dimensional space onto the lower-dimensional one.
+
+As Reviewer 3 points out, the high correlations between traits, and the restricted dimensionality they imply, is different from the dimensional sensitivity of the test items.
+However, a highly correlated latent space, as in dataset 3d1 in @reckase_multidimensional_2009, may induce a dimensional restriction on the latent space (as indicated by the rank-1 covariance matrix), which is indistinguishable from a lower-dimensional (unidimensional, in this case) solution.
+These two facts, the indistinguishability of the "projected high-dimensional" solution and the lower-dimensional one, on the one hand, and the indistinguishability of the "high-dimensional solution" with a low-rank covariance matrix and the "low-dimensional solution", on the other, render the theoretical implications of a higher-dimensional item sensitivity impossible to test in a case like this example.
+This is acknowledged by @reckase_multidimensional_2009, who states that "if the process used to generate the data were not known, the interpretation of the analysis results would likely be that the test items all measured the same construct and that the data could be well fit by a unidimensional IRT model" (p.
+189).
+
+However, in our manuscript we intend to go one step further, and show how the latent space covariance affects how the item parameters are interpreted.
+We have shown that non-orthonormal axes are a more faithful geometrical representation of the test space when the latent covariance is not an identity matrix.
+From the model formulation, one can note that the components of the discrimination parameter serve as a scaling factor on the effect each latent trait has on the response probability function.
+As such, if they were interpreted as abstract parameters in a high-dimensional space that can be projected onto the latent trait axis, then the model formulation would consider those projections instead of the original item parameters.
+Therefore, the discrimination parameter values should be considered as being "parallel" to the latent space axis.
+In other words, each discrimination parameter value is itself the projection of the item vector length onto its corresponding axis.
+
+In this sense, having such high correlations among latent traits not only makes them practically unidimensional; it also implies that the item parameters must be practically aligned with the "unidimensional" latent space axis.
+In other words, even if a test is designed to be sensitive to three dimensions, if those three dimensions are highly correlated, the items will be practically sensitive to a single dimension.
+This will happen regardless of whether one considers that there is a higher-dimensional test space from which the alleged single dimension is a subspace (where the item parameters will be a projection of the higher-dimensionally sensitive ones), or whether one considers that the items are simply sensitive to a single dimension, with the item parameters given simply by the 2PL model.
+We believe this is the reason why estimating such a model would yield "a first dimension with high positive aparameters and other dimensions with much lower a-parameter estimates with many around zero" [@reckase_multidimensional_2009, p.
+189].
+In fact, we predict that the non-zero discrimination value estimate will not match any of the values of the generating discrimination parameter, but $`r MDISC_COV_PARAM`$ which, in this case, actually matches the discrimination parameter of the unidimensional 2PL model.
+Although we acknowledge the lack of time to properly check that this prediction is fulfilled, we will gladly do so in another review round if the editor and Reviewer 3 consider it necessary.
+
+The aforementioned considerations have led us to use this edge case as a very illustrative example for our purposes, because it allows the reader to understand easily that all items should be aligned with the only dimension being measured.
+It is therefore easy to grasp the intuitive notion that all direction cosines should be expected to be equal to 0, and in the more general case, to represent the relative alignment of the item vector with all the axes.
+However, in doing so, we may have skipped a step in our reasoning, assimilating the dimensionality of the latent space with the dimensional sensitivity of the items, as Reviwer 3 points out.
+
+To clarify better this point, we have tried to express better this distinction, emphasizing that, in this case, the unidimensionality induced by the rank-limited covariance matrix restricts the dimensional sensitivity of the test space.
+The aforementioned paragraph now reads (p. <!-- # TODO: Complete -->, lines <!-- # TODO: Complete -->):
+
+> [...] In that example,
+he uses a covariance matrix that practically has rank 1,
+implying that the (alleged) three latent dimensions collapse into a single one.
+This makes the latent space unidimensional in practice,
+which in turn resticts the dimensionality of any response dataset
+generated from a test measuring those traits.
+Therefore, "the data could be well fit by a unidimensional IRT model"
+[@reckase_multidimensional_2009, p. 189].
+In such a case, when all the items are strictly aligned with
+the one dimension being measured,
+we should expect the direction angles to be extremely close to 0 for all items.
+The reader interested can check that this is the case for
+the covariance-based version of the direction angles
+in the table in Online Resource 1.
 

### 2025-05-27 output/Main_text.Rmd
Address "dimensional sensitivity" comment properly
After reading carefully Reckase's (2009) chapter 7, the dimensional sensitiviy comment is properly addressed.

--- 2025-05-26 output/Main_text.Rmd
@@
 In that example, he uses a covariance matrix that practically has rank 1,
-implying that the (alleged) three dimensions collapse into a single one.
-This in turn makes the latent space unidimensional in practice.
-As the dimensional sensitivity of the latent space sets the upper limit of
-the dimensional sensitivity of a test measuring those traits,
-it follows that the test must be unidimensional.
+implying that the (alleged) three latent dimensions collapse into a single one.
+This makes the latent space unidimensional in practice,
+which in turn resticts the dimensionality of any response dataset
+generated from a test measuring those traits.
+Therefore, "the data could be well fit by a unidimensional IRT model"
+[@reckase_multidimensional_2009, p. 189].
 In such a case, when all the items are strictly aligned with
-the one dimension they measure,
+the one dimension being measured,
 we should expect the direction angles to be extremely close to 0 for all items.

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Correct errata and style
In response to "dimensional sensitivity" comment

@@
 
-To our understanding, one may assume a theoretical dimensional sensitivity of a test space that is higher than the dimensionality of the latent space.
-We acknowledge we may have overlooked this distinction.
+To our understanding, one may assume a theoretical dimensional sensitivity of a test space that is higher than the dimensionality of the latent space, which we acknowledge we may have overlooked.
 However, in such a case, the (minimum) dimensionality of the data would be "the smaller of the dimensions or variation for the people and the dimensions of sensitivity for the test items" (@reckase_multidimensional_2009, p.
@@
 This makes the latent space unidimensional in practice,
-which in turn resticts the dimensionality of any response dataset
+which in turn restricts the dimensionality of any response dataset
 generated from a test measuring those traits.

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Apply Grammarly corrections

@@
 To our understanding, one may assume a theoretical dimensional sensitivity of a test space that is higher than the dimensionality of the latent space, which we acknowledge we may have overlooked.
-However, in such a case, the (minimum) dimensionality of the data would be "the smaller of the dimensions or variation for the people and the dimensions of sensitivity for the test items" (@reckase_multidimensional_2009, p.
-182).
+However, in such a case, the (minimum) dimensionality of the data would be "the smaller of the dimensions or variation for the people and the dimensions of sensitivity for the test items" [@reckase_multidimensional_2009, p.
+182].
 That is, the dimensional sensitivity of the data would nevertheless be restricted by the dimensionality of the latent space.
-This implies that the parameter values one can estimate for the items will represent a lower-dimensional solution, which would be the projection of the theoretically higher-dimensionally sensitive parameters on the lower-dimensional latent trait space, as discussed in p.
+This implies that the parameter values one can estimate for the items will represent a lower-dimensional solution, which would be the projection of the theoretically higher-dimensionally sensitive parameters on the lower-dimensional latent trait space, as discussed on p.
 182 of @reckase_multidimensional_2009.
-In practice, this implies that this solution is indistinguishable from one where the items are simply sensitive to the lower number of dimensions, with their parameters given by such lower-dimensional solution which would correspond to their projections from the alleged higher-dimensional space onto the lower-dimensional one.
+In practice, this implies that this solution is indistinguishable from one where the items are simply sensitive to the lower number of dimensions, with their parameters given by such a lower-dimensional solution, which would correspond to their projections from the alleged higher-dimensional space onto the lower-dimensional one.
 
-As Reviewer 3 points out, the high correlations between traits, and the restricted dimensionality they imply, is different from the dimensional sensitivity of the test items.
+As Reviewer 3 points out, the high correlations between traits and the restricted dimensionality they imply are different from the dimensional sensitivity of the test items.
 However, a highly correlated latent space, as in dataset 3d1 in @reckase_multidimensional_2009, may induce a dimensional restriction on the latent space (as indicated by the rank-1 covariance matrix), which is indistinguishable from a lower-dimensional (unidimensional, in this case) solution.
@@
 
-However, in our manuscript we intend to go one step further, and show how the latent space covariance affects how the item parameters are interpreted.
+However, in our manuscript, we intend to go one step further and show how the latent space covariance affects how the item parameters are interpreted.
 We have shown that non-orthonormal axes are a more faithful geometrical representation of the test space when the latent covariance is not an identity matrix.
@@
 This will happen regardless of whether one considers that there is a higher-dimensional test space from which the alleged single dimension is a subspace (where the item parameters will be a projection of the higher-dimensionally sensitive ones), or whether one considers that the items are simply sensitive to a single dimension, with the item parameters given simply by the 2PL model.
-We believe this is the reason why estimating such a model would yield "a first dimension with high positive aparameters and other dimensions with much lower a-parameter estimates with many around zero" [@reckase_multidimensional_2009, p.
+We believe this is the reason why estimating such a model would yield "a first dimension with high positive $`r DISCR_SYM`$-parameters and other dimensions with much lower $`r DISCR_SYM`$-parameter estimates with many around zero" [@reckase_multidimensional_2009, p.
 189].
@@
 It is therefore easy to grasp the intuitive notion that all direction cosines should be expected to be equal to 0, and in the more general case, to represent the relative alignment of the item vector with all the axes.
-However, in doing so, we may have skipped a step in our reasoning, assimilating the dimensionality of the latent space with the dimensional sensitivity of the items, as Reviwer 3 points out.
+However, in doing so, we may have skipped a step in our reasoning, assimilating the dimensionality of the latent space with the dimensional sensitivity of the items, as Reviewer 3 points out.
 
-To clarify better this point, we have tried to express better this distinction, emphasizing that, in this case, the unidimensionality induced by the rank-limited covariance matrix restricts the dimensional sensitivity of the test space.
+To clarify this point better, we have tried to better express this distinction, emphasizing that, in this case, the unidimensionality induced by the rank-limited covariance matrix restricts the dimensional sensitivity of the test space.
 The aforementioned paragraph now reads (p. <!-- # TODO: Complete -->, lines <!-- # TODO: Complete -->):

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Clarify editor's "comparability" concerns

@@
 but perfectly correlated in the second one.
+As the cosine between variables is usually understood as
+the geometrical representation of their correlation,
+the cosine between the two items should be 1 in the second group.
 However, using an orthogonal representation
@@
 
-There are several references in the literature that have found contradictory results that, we believe, are due to disregarding the latent covariance structure (i.e., using Reckase & McKinley's formulae in a non-orthonormal space).
+There are several references in the literature that have found contradictory results which, we believe, are due to disregarding the latent covariance structure (i.e., using Reckase & McKinley's formulae in a non-orthonormal space).
 For example, @min_evaluation_2007 [p.
@@
 default distribution, such as was the case in groups 2 to 5".
-Although they attribute this to a possible "degree of confounding among errors of parameter estimation and scale
-transformation", their groups 2 to 5 have all correlations equal (or almost equal) to .5, as it happens.
+They attribute this to a possible "degree of confounding among errors of parameter estimation and scale transformation", but their groups 2 to 5 happen to all have correlations equal (or almost equal) to .5.
+(We consider that discussing these examples is out of the scope of this manuscript, but we will be glad to consider some of them if this editor deems it necessary.)
 
-We believe this is also what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution, as this editor himself has noted before [@ackerman_multidimensional_2005-1, p.
+Also, we believe this is what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution, as this editor himself has noted elsewhere [@ackerman_multidimensional_2005-1, p.
 11].
-While we consider this to be out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our method may solve the problem.
+While we consider this to be also out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our proposal may solve the problem.
 

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Complete letter to editor
Make minor amendments to responses to editor

@@
 suggestions, which we believe have helped improve the quality of our paper.
-<!-- TODO: Complete -->
+However, we acknowledge we may not have resolved satisfactorily all the action
+editor's and reviewers' concerns.
+Therefore, we would kindly ask you to forward to them our good will
+and predisposition to address any further comments or suggestions,
+or clarify futher on the previous ones.
 
@@
 
-It is our stance that the covariance structure, understood as a property of the data of course, should be taken into account when determining the multidimensional properties of the items.
+It is our stance that the covariance structure, understood as a property of the data, of course, should be taken into account when determining the multidimensional properties of the items.
 That is what we propose with the covariance-based (and correlation-based) versions of the multidimensional parameters.
@@
 We acknowledge and apologize for the lack of clarity in this sentence.
-Please refer to our 10th response to Reviewer 3 on p. <!-- TODO: Complete --> of this letter.
+Please refer to our response to Reviewer 3 on p. <!-- TODO: Complete --> of this letter.
 
@@
 
-The standard errors of the multidimensional parameters are, without a doubt, more difficult to compute for the covariance-based case than for the agnostic one (if we assume the covariance structure is estimated from the model).
+The standard errors of the multidimensional parameters are, without a doubt, more difficult to compute for the covariance-based case than for the agnostic one.
 Computing the covariance errors in the former would require taking into account the standard errors of the item parameters themselves and at least the standard errors of the correlations (in the best-case scenario, assuming standard latent dimensions; otherwise, those of the variances and covariances).
@@
 Software such as the R package {mirt} [@chalmers_mirt_2012] uses a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
-As far as we know, these standard error estimates are computed by the usual methods, involving, e.g., Taylor series, bootstrapping, or similar.
-<!-- TODO: Check SE methods; complete? -->
+One can see the help of the [`mirt()` function](https://www.rdocumentation.org/packages/mirt/versions/1.43/topics/mirt) therein (se description of argument `SE.type`) for a detailed description of how this package computes the standard errors.
 
@@
 However, we have checked and the journal accepts and also encourages publishing additional materiales in [a proper, trusted repository](https://www.cambridge.org/core/services/where-to-share-your-data).
-We have thus made public [our GitHub repository for the manuscript](https://github.com/DaniMori/mirt-parameters) and [indexed it in Software Heritage](https://archive.softwareheritage.org/swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af) for being identified with a SWHID.
-can do so if the journal accepts linking an external repository as supplementary material, or otherwise grant access to the editor and reviewers to this repository.
+We have thus made public [our GitHub repository for the manuscript](https://github.com/DaniMori/mirt-parameters) and [indexed it in Software Heritage](https://archive.softwareheritage.org/swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af) for identification with a SWHID.
 

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Adds reference to SW repository in manuscript

@@
 We have thus made public [our GitHub repository for the manuscript](https://github.com/DaniMori/mirt-parameters) and [indexed it in Software Heritage](https://archive.softwareheritage.org/swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af) for identification with a SWHID.
+For the sake of transparency, we have also made this addition to the manuscript, in section "Applications to Examples from the Literature", before the @reckase_multidimensional_2009 example:
+
+> The interested reader can explore the complete code for these examples in the
+indexed Software Heritage repository for this paper, specifically:
+<https://archive.softwareheritage.org/swh:1:cnt:b104aa768ded8fc19e2cafe8cdc9e64724c4fa47;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af;path=/src/Empirical_example.R>
 

### 2025-05-27 output/Main_text.Rmd
Clarify editor's "comparability" concerns
Adds reference to SW repository in manuscript

@@
 the column next to its corresponding agnostic version.
+The interested reader can explore the complete code for these examples in the
+indexed Software Heritage repository for this paper, specifically:
+<https://archive.softwareheritage.org/swh:1:cnt:b104aa768ded8fc19e2cafe8cdc9e64724c4fa47;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af;path=/src/Empirical_example.R>
 

### 2025-05-27 output/Letter_to_editor_reviewers.Rmd
Make minor changes to response letter style

@@
 
-September 16, 2024. <!-- TODO: Update date -->
+June 2, 2025. <!-- TODO: Update date -->
 
@@
 Software such as the R package {mirt} [@chalmers_mirt_2012] uses a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
-One can see the help of the [`mirt()` function](https://www.rdocumentation.org/packages/mirt/versions/1.43/topics/mirt) therein (se description of argument `SE.type`) for a detailed description of how this package computes the standard errors.
+One can see the help of the [`mirt()` function](https://www.rdocumentation.org/packages/mirt/versions/1.43/topics/mirt) therein (please see description of argument `SE.type`) for a detailed description of how this package computes the standard errors.
 
@@
 
-Our work is completely theoretical and thus uses a purely deductive method.
-Therefore, we do not have any hypothesis to test, but the results of the formal derivations that we present, which address the objective described at the end of our introduction.
+Our work is entirely theoretical and thus employs a purely deductive method.
+Consequently, we do not have any hypothesis to test; rather, we present the results of our formal derivations, which address the objective described at the end of our introduction.
 
@@
 
+<!-- # TODO: Equation references change? -->
 Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
@@
 The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may confuse the reader to refer to the $MDISC_i$ and overlook the influence of $d_i$.
-We have added a clarification in this respect to the end of the paragraph, which now reads:
+We have added a clarification to this at the end of the paragraph, which now reads:
 
@@
 We have now added a paragraph to our introduction to try to explain this fact intuitively.
-Please refer to our answer to the editor on p. 3 of this letter for more details on this addition.
+Please refer to our answer to the editor on pp. 3-4 of this letter for more details on this addition.
 
@@
 This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
-
-Our first version tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p.
+We have tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p.
 14, lines 15-22, or p.
@@
 
-This is the expression of a vector represented in a basis, in this case, representing that the vector can be represented in each of the two bases interchangeably.
+This is the expression of a vector represented in a basis, in this case, stating that the vector can be represented in each of the two bases interchangeably.
 Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:
@@
 
-This is a mistake on our side, we are afraid to admit, product of a careless translation of a false friend from our mother tongue Spanish ("indistintamente"), which means "indifferently".
-We have changed the term by the mathematically appropriate "interchangeably" in our new version.
+This is a mistake on our side, we are afraid to admit, the product of a careless translation of a false friend from our mother tongue Spanish ("indistintamente"), which means "indifferently".
+We have changed the term to the mathematically appropriate "interchangeably" in our new version.
 
@@
 7 line 8 in our original submission, or p.
-<!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen, which does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction.
+<!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen to represent the items.
+This basis does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction, so it can be the canonical basis, or any other, orthonormal or not, and not necessarily representing the covariance structure.
 
@@
 In this respect, we should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
-Our argument is that it is when we represent the items that we should use axes that "faithfully represent the covariance structure", and not when we represent the attributes.
+Our argument is that it is when we represent the items that we should use axes that "faithfully represent the covariance structure", regardless of how we represent the attributes.
 
@@
 We have shown that non-orthonormal axes are a more faithful geometrical representation of the test space when the latent covariance is not an identity matrix.
-From the model formulation, one can note that the components of the discrimination parameter serve as a scaling factor on the effect each latent trait has on the response probability function.
+From the model formulation, one can note that the components of the discrimination parameter serve as a scaling factor of the effect each latent trait has on the response probability function.
 As such, if they were interpreted as abstract parameters in a high-dimensional space that can be projected onto the latent trait axis, then the model formulation would consider those projections instead of the original item parameters.

### 2025-05-28 output/Letter_to_editor_reviewers.Rmd
Reply to editor's multi-group comment

@@
 
-<!-- TODO: Reply -->
+Fitting a single-group model where the latent dimensions are correlated affects the computation of the multidimensional parameters we propose here.
+Given the scale underidentification in IRT, it is relatively common to fix the latent variances to 1, so it is unlikely that non-standard covariances affect these applications.
+In this sense, we have mentioned in our discussion that "in many applications both [the covariance- and correlation-based] versions will be coincident, fortunately" (p.
+31, lines 2-3, or p.
+<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
+However, in multigroup procedures, such as equating and linking, non-standard (in some of the groups at least) may indeed affect the resulting multidimensional parameters.
+Additionally, it is worth noting that the correlation matrix may differ among groups, which can also have a substantial effect on the multidimensional parameter values.
+This is relevant because, even if the item parameters are equal across groups, the covariance structure may differ; thus, the multidimensional parameters will not be the same.
+Conversely, the same multidimensional parameters may correspond to different parameter models if the latent covariance structure is different across groups.
+Although it would be interesting to explore whether it is more realistic to assume equal M2PL parameters across groups (and thus possibly different multidimensional parameters) or vice-versa, we believe that this is an empirical question that lies outside the scope of our current manuscript.
+
+We have probably made the mistake of assuming that this was implicitly understood, for which we apologize to the editor.
+To improve and address this point better, we have made some additions to the second paragraph <!-- TODO: Third now?
+--> of our "Discussion" section (p.
+30, lines 5-9 in the original submission, or p.
+<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new one), which now reads:
+
+> [...] Moreover, in applications where the latent space structure is critical
+(e.g., multi-group IRT, equating, and linking),
+the agnostic version may overlook relevant differences in
+the values of the multidimensional parameters.
+Different covariance structures across groups may significantly affect
+the value of these parameters,
+due to differences in either the variances or the correlation structure.
+Therefore, even when the model parameters are equal across groups,
+the multidimensional parameters may not be the same
+if the covariance structure differs across groups.
+Conversely, the same multidimensional parameters may correspond
+to different parameter models.
+The subsequent invariance of either the M2PL parameters
+or the multidimensional ones may lead to different interpretations of
+the same model when applied to different groups.
+Although this is an empirical question that requires further research,
+it could have significant implications for future theoretical developments
+in the understanding of MIRT theory.
 

### 2025-05-28 output/Main_text.Rmd
Reply to editor's multi-group comment

@@
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g., multigroup equating) even when
+non-identity latent covariance matrices (e.g., multi-group equating) even when
 interpreting the latent dimensions as substantive constructs is not a goal
@@
 the values of the multidimensional parameters.
-Thus, the covariance structure of the latent space should be considered
-when computing the multidimensional parameters of a MIRT model.
+Different covariance structures across groups may significantly affect
+the value of these parameters,
+due to differences in either the variances or the correlation structure.
+Therefore, even when the model parameters are equal across groups,
+the multidimensional parameters may not be the same
+if the covariance structure differs across groups.
+Conversely, the same multidimensional parameters may correspond
+to different parameter models.
+The subsequent invariance of either the M2PL parameters
+or the multidimensional ones may lead to different interpretations of
+the same model when applied to different groups.
+Although this is an empirical question that requires further research,
+it could have significant implications for future theoretical developments
+in the understanding of MIRT theory.
 

### 2025-05-28 output/Letter_to_editor_reviewers.Rmd
Add pending tasks comments in letter

@@
 We have now added a paragraph to our introduction to try to explain this fact intuitively.
-Please refer to our answer to the editor on pp. 3-4 of this letter for more details on this addition.
+Please refer to our answer to the editor on pp. 3-4 <!-- TODO: Review --> of this letter for more details on this addition.
 
@@
 Also, we have made changes and additions to
-the third paragraph of our Discussion, which now reads:
+the third  <!-- TODO: Review --> paragraph of our Discussion, which now reads:
 
@@
 However, we acknowledge that such a plot is the most common representation of the values realized by different cases (persons, in our case) in two variables, correlated or not.
-In this regard, we would kindly refer you to our third response to your comments, on p.
+In this regard, we would kindly refer you to our third <!-- TODO: Expressed ok?
+--> response to your comments, on p.
 <!-- TODO: Complete --> of this letter, where we discuss the representation of person parameters in the last two paragraphs.

### 2025-05-28 output/Letter_to_editor_reviewers.Rmd
Reply to "partical value" comment from Reviewer 2

@@
 
-<!-- TODO: Reply -->
+We understand that it may be difficult to see the practical value of our proposal.
+However, we argue that our main contribution is solidly founded on sound formal derivations, and that its value lies in the theoretical contribution it represents.
+In practical terms, we believe that our results may help better diagnose the validity of the items in multidimensional tests, as both the multidimensional parameters and the vector representation we put forward are a more faithful representation of the properties of those items.
+This happens because item parameters should be interpreted in their context, that is, as properties of the response data that result from the interaction of an instrument and the sample where it is applied, as argued by @reckase_multidimensional_2009.
+
+We expect that the changes we have made in the new version of our manuscript help clarify the practical value of our proposal.
+In this respect, please see especially the responses to the editor in pp. <!-- TODO: Complete (3rd response, last, and 2nd to last) --> in this letter.
 

### 2025-05-28 output/Letter_to_editor_reviewers.Rmd
Clarify "reasoning artifact" nature
Referring to the "change of basis" in the latent space.

@@
 This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
+However, please bear in mind that this transformation in the latent space is a reasoning artifact to obtain the proper expression of the item parameters in the general case; for further explanation about this, please see our response to the editor in pp.
+<!-- TODO: Complete (5th response) --> lines  <!-- TODO: Complete --> of this letter.
 We have tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p.

### 2025-05-28 output/Letter_to_editor_reviewers.Rmd
Addd IRTRPO to software packages discussed

@@
 Nevertheless, most available MIRT software solutions,
-such as flexMIRT [@cai_flexmirt_2024] or R package {mirt} [@chalmers_mirt_2012],
-initially estimate an uncorrelated solution and apply
-an oblique rotation afterwards, at least in exploratory solutions.
+such as flexMIRT [@cai_flexmirt_2024], IRTPRO [@cai_irtpro_2017],
+or R package {mirt} [@chalmers_mirt_2012],
+initially estimate an uncorrelated solution (at least in the exploratory case)
+and apply an oblique rotation afterwards.
 Taking advantage of the invariance property,

### 2025-05-28 output/Title_page.Rmd
Update "running title" title in title page
Addd IRTRPO to software packages discussed

@@
 **Article title:**
-A generalized definition of multidimensional item response theory parameters
+A Generalized Definition of Multidimensional Item Response Theory Parameters
 
-**Short title:** Generalized MIRT parameters
+**Running title:** Generalized MIRT parameters
 
-**List of authors:**
+# List of authors
 
@@
 
-**Corresponding author:**
+## Corresponding author:
 
@@
 
+# Acknowledgements
+
+The authors would like to thank Patricia Recio-Saboya, Ph.D.
+(ORCID: [0000-0002-0530-4404](https://orcid.org/0000-0002-0530-4404))
+for her valuable guidance regarding multidimensional estimation
+in IRT software packages.
+
 # Financial Support
@@
 
-Competing interests: The author(s) declare none.
+Competing interests: The authors declare none.

### 2025-05-28 output/Letter_to_editor_reviewers.Rmd
Add data availability statement
Obfuscate at the same time all repository URLs

@@
 However, we have checked and the journal accepts and also encourages publishing additional materiales in [a proper, trusted repository](https://www.cambridge.org/core/services/where-to-share-your-data).
-We have thus made public [our GitHub repository for the manuscript](https://github.com/DaniMori/mirt-parameters) and [indexed it in Software Heritage](https://archive.softwareheritage.org/swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af) for identification with a SWHID.
+We have thus made public our GitHub repository for the manuscript [Link URL obfuscated for blind peer-review] and indexed it in Software Heritage [Link URL obfuscated for blind peer-review] for identification with a SWHID.
 For the sake of transparency, we have also made this addition to the manuscript, in section "Applications to Examples from the Literature", before the @reckase_multidimensional_2009 example:
@@
 indexed Software Heritage repository for this paper, specifically:
-<https://archive.softwareheritage.org/swh:1:cnt:b104aa768ded8fc19e2cafe8cdc9e64724c4fa47;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af;path=/src/Empirical_example.R>
+[URL obfuscated for blind peer-review]
 

### 2025-05-28 output/Response_to_action_editor_reviewers.Rmd
Set apart comments from response letter to editor

---
output:
  officedown::rdocx_document:
    mapstyles:
      First Paragraph: Body Text
    base_format: bookdown::word_document2
    reference_docx: ../www/Template.docx
    number_sections: no
    keep_md:         no
bibliography: ../www/Multidimensional-parameters-MCLM.bib
csl:          ../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| include: false

# Includes:

library(here)
library(knitr)
library(patchwork)
library(ggplot2)
library(rstudioapi)
library(svglite) # Used by {knitr} and {ggplot2} to render the figure to SVG;
                 #   added to prevent {renv} from reporting inconsistent state.
library(officer)

ROOT_DIR <- here()
DOC_DIR  <- file.path(ROOT_DIR, "output")

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      = 1200L,
  dev.args = list(type = "cairo-png"),
  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE),
  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
)

# Output configuration options:
options(digits = 3)
```

```{r sources}
#| cache:   false
#| include: false
source("R/Formulae.R",  encoding = 'UTF-8')
source("R/Constants.R", encoding = 'UTF-8')
```

# Responses to the action editor:

###### I have a concern with making the latent ability axes nonorthogonal to represent the covariance structure between the latent abilities. Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.

Indeed, different groups of examinees can result in different covariance structures.
However, in our understanding, the same fact can be predicated about the item parameters themselves, even without considering the covariance structure at all.
For example, in unidimensional IRT, the item parameters may be different for different groups of examinees, even if the variance is constrained to be equal across groups; this is indeed the basis of methods such as multigroup IRT.

It is our stance that the covariance structure, understood as a property of the data, of course, should be taken into account when determining the multidimensional properties of the items.
That is what we propose with the covariance-based (and correlation-based) versions of the multidimensional parameters.

###### The statement (p.13, line 8) about “the test space where item parameters naturally lie” needs greater explanation with examples.

We acknowledge and apologize for the lack of clarity in this sentence.
Please refer to our response to Reviewer 3 on p. <!-- TODO: Complete --> of this letter.

###### Like Reviewer 3, I do not understand how the magnitude of discrimination values (p 29 , 7) can be compared/interpreted values of discrimination are comparable from the two representations. It seems that the metric of the represented space has changed, and equivalent numbers will not have equivalent interpretations.

As the editor says, the basis used to represent the test space changes its metric.
We also understand that the metrics of the two representations are not comparable, and thus one should not compare items represented with the agnostic and the covariance-based version of the multidimensional parameters.
However, when making multi-group comparisons, we think that taking into account the covariance-based structure of the latent space will provide more realistic information, and thus one should use the covariance-based version.

In addition to other (minor) changes, we have added the following paragraph to our introduction to give an intuitive explanation of this fact:

> Take the following as an example:
A hypothetical complex item measures two latent traits with equal validity
---i.e., its discrimination parameters are equal with value $a$.
Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
we would find that $`r MDISC_SYM` = \sqrt{2} a_i$.
Assume we apply this item in two groups where, taken to the extreme,
the two measured dimensions are orthogonal in the first group
and perfectly correlated in the second one.
While the previous formula seems correct in the first group,
one can intuitively say that,
because the two dimensions are *perfectly aligned* and thus
compound up in the second group,
the item `r MDISC_SYM` should be the algebraic sum of both components,
that is, $2 a_i$.
Similarly, the correlation between two simple items,
each one measuring a distinct dimension,
should be uncorrelated in the first group,
but perfectly correlated in the second one.
As the cosine between variables is usually understood as
the geometrical representation of their correlation,
the cosine between the two items should be 1 in the second group.
However, using an orthogonal representation
would yield null cosines in both cases.

There are several references in the literature that have found contradictory results which, we believe, are due to disregarding the latent covariance structure (i.e., using Reckase & McKinley's formulae in a non-orthonormal space).
For example, @min_evaluation_2007 [p. 49] computes the item parameters in the M2PL from a priori fixed values of $MDISC_i$ and $MDIFF_i$ (the name they use for $MID_i$), using @reckase_difficulty_1985 and @reckase_discriminating_1991's formulae.
They find that "[...] bias [...] was more apparent when ability distribution was separated from the
default distribution, such as was the case in groups 2 to 5".
They attribute this to a possible "degree of confounding among errors of parameter estimation and scale transformation", but their groups 2 to 5 happen to all have correlations equal (or almost equal) to .5.
(We consider that discussing these examples is out of the scope of this manuscript, but we will be glad to consider some of them if this editor deems it necessary.)

Also, we believe this is what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution, as this editor himself has noted elsewhere [@ackerman_multidimensional_2005-1, p. 11].
While we consider this to be also out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our proposal may solve the problem.

###### I also wonder about how one compares the standard errors of the item parameters when the covariance covariance structures are different.  Are they directly comparable?

The standard errors of the multidimensional parameters are, without a doubt, more difficult to compute for the covariance-based case than for the agnostic one.
Computing the covariance errors in the former would require taking into account the standard errors of the item parameters themselves and at least the standard errors of the correlations (in the best-case scenario, assuming standard latent dimensions; otherwise, those of the variances and covariances).
However, to our knowledge, multidimensional item parameters, as defined by @reckase_difficulty_1985 and @reckase_discriminating_1991, have been previously defined and used only from a theoretical perspective, without ever taking into account the uncertainty resulting from the model estimation.

Regarding the "original" item parameters in the classical M2PL metric, we understand that the standard errors of the item parameters are independent of the definition of the multidimensional parameters used.
Software such as the R package {mirt} [@chalmers_mirt_2012] uses a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
One can see the help of the [`mirt()` function](https://www.rdocumentation.org/packages/mirt/versions/1.43/topics/mirt) therein (please see description of argument `SE.type`) for a detailed description of how this package computes the standard errors.

###### I would like to see the motivation and psychometric utility of this approach. Does it improve the interpretation of the latent abilities and thus result in more insight into how the results are used? The author needs to address this with legitimate examples for testing practitioners.

We think the motivation and psychometric utility of our approach are more related to the interpretation of the item parameters than with the latent abilities, which we believe exceeds the scope of this manuscript and could be the focus of another paper.
Please note that we only use the transformation in the latent space to an orthonormal basis as a reasoning artifact to obtain the proper expression of the item parameters in the general case, and that we do it in an abstract sense, before even considering whether the original basis is a representation of the covariance matrix.

However, to our understanding, the more precise interpretation of the item parameters that results from our approach indeed provides a better insight into how the results are used.
A multidimensional representation of the items provides benchmarks for comparing different latent trait levels, as discussed by @reckase_difficulty_1985.
Therefore, the more accurate such a representation is, the better the latent trait interpretation will be.

Please see our answer on p. <!-- TODO: Complete (3?) --> of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
<!-- TODO: Complete -->

###### What implications does your approach have for standard psychometric procedures like equating.

Fitting a single-group model where the latent dimensions are correlated affects the computation of the multidimensional parameters we propose here.
Given the scale underidentification in IRT, it is relatively common to fix the latent variances to 1, so it is unlikely that non-standard covariances affect these applications.
In this sense, we have mentioned in our discussion that "in many applications both [the covariance- and correlation-based] versions will be coincident, fortunately" (p. 31, lines 2-3, or p. <!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
However, in multigroup procedures, such as equating and linking, non-standard (in some of the groups at least) may indeed affect the resulting multidimensional parameters.
Additionally, it is worth noting that the correlation matrix may differ among groups, which can also have a substantial effect on the multidimensional parameter values.
This is relevant because, even if the item parameters are equal across groups, the covariance structure may differ; thus, the multidimensional parameters will not be the same.
Conversely, the same multidimensional parameters may correspond to different parameter models if the latent covariance structure is different across groups.
Although it would be interesting to explore whether it is more realistic to assume equal M2PL parameters across groups (and thus possibly different multidimensional parameters) or vice-versa, we believe that this is an empirical question that lies outside the scope of our current manuscript.

We have probably made the mistake of assuming that this was implicitly understood, for which we apologize to the editor.
To improve and address this point better, we have made some additions to the second paragraph <!-- TODO: Third now? --> of our "Discussion" section (p. 30, lines 5-9 in the original submission, or p. <!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new one), which now reads:

> [...] Moreover, in applications where the latent space structure is critical
(e.g., multi-group IRT, equating, and linking),
the agnostic version may overlook relevant differences in
the values of the multidimensional parameters.
Different covariance structures across groups may significantly affect
the value of these parameters,
due to differences in either the variances or the correlation structure.
Therefore, even when the model parameters are equal across groups,
the multidimensional parameters may not be the same
if the covariance structure differs across groups.
Conversely, the same multidimensional parameters may correspond
to different parameter models.
The subsequent invariance of either the M2PL parameters
or the multidimensional ones may lead to different interpretations of
the same model when applied to different groups.
Although this is an empirical question that requires further research,
it could have significant implications for future theoretical developments
in the understanding of MIRT theory.

###### Are there standard computer programs that you would recommend to do the estimation/modeling you are proposing?

Our manuscript does not propose an estimation or modeling method.
On the contrary, what we propose is a way to compute and represent the multidimensional item parameters without disregarding the covariance structure of the latent space.
Implementing the formulae we propose should be rather straightforward for most IRT researchers, simply using the standard output of most MIRT software.
We have now added a discussion about this at the end of our fourth paragraph in the "Discussion" section. This addition reads:

> [...] Computing the parameters in the orthonormalized
(or just orthogonalized) space,
by contrast, would require transforming the cosines to
the original, non-orthonormal basis afterwards.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024], IRTPRO [@cai_irtpro_2017],
or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution (at least in the exploratory case)
and apply an oblique rotation afterwards.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution by applying
Equation (24).
<!-- TODO: Check Eq. nº afterwards -->
What a test practitioner should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
unlike the invariant value of $`r MDISC_COV_PARAM`$.

Notwithstanding, we have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, and even the submitted versions of the manuscript.
Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software [e.g., {mirt}, @chalmers_mirt_2012; flexMIRT, @cai_flexmirt_2024].
As the journal guidelines [require a specific format for the Supplementary Material](https://www.cambridge.org/core/services/authors/publishing-supplementary-material),
we chose not to include the repository as additional material.
However, we have checked and the journal accepts and also encourages publishing additional materiales in [a proper, trusted repository](https://www.cambridge.org/core/services/where-to-share-your-data). 
We have thus made public our GitHub repository for the manuscript [Link URL obfuscated for blind peer-review] and indexed it in Software Heritage [Link URL obfuscated for blind peer-review] for identification with a SWHID.
For the sake of transparency, we have also made this addition to the manuscript, in section "Applications to Examples from the Literature", before the @reckase_multidimensional_2009 example:

> The interested reader can explore the complete code for these examples in the
indexed Software Heritage repository for this paper, specifically:
[URL obfuscated for blind peer-review]

# Responses to Reviewer 1:

###### 1. Introduction, too lengthy, 1-2 pages will be fine.

Our introduction is a bit less than four pages long.
We have provided concise paragraphs about the Multidimensional IRT model and its parameters, their assumptions, and our rationale for relaxing them.
Finally, a last paragraph presents the objectives of our research.
We honestly believe we have made a thorough effort to summarize the motivation of our research, and we find it difficult to shorten it further without losing relevant information to justify our work.
Therefore, we are sorry to say that we cannot comply with the reviewer's request to shorten the introduction.

###### 2. Introduction, do you have any hypothesis for the current study? Please briefly describe it at the end.

Our work is entirely theoretical and thus employs a purely deductive method.
Consequently, we do not have any hypothesis to test; rather, we present the results of our formal derivations, which address the objective described at the end of our introduction.

###### 3. The details about the datasets should be described here.

The datasets we use in our paper are examples of item parameter estimates, taken from the literature.
As they only serve exemplification purposes and are not the main focus of our article, we deem it more appropriate to describe their details in the corresponding section of the manuscript.
We have added a short reference to them in the last paragraph of our introduction (p. <!-- TODO: Complete -->, lines <!-- TODO: Complete -->), to make sure their pertinence as examples is clear.

> These examples, from the cognitive and the non-cognitive domain, respectively,
are taken from the MIRT literature:
One of them is a classic dataset, often used to exemplify
the use of the `r MODEL_ACRONYM` model [@reckase_multidimensional_2009];
the other one, an application in marketing research [@tezza_modelo_2018].

###### 4. Within or between multidimensional latent? This should be detailed.

We think the context makes it clear that we are referring to "within-item" multidimensionality.
In fact, the results are trivial in the "between-item" multidimensionality case,
as the `r MDISC_SYM` parameter reduces to the only non-null component of the $`r DISCR_VECTOR`$ parameter.
However, we acknowledge this terminology is prevalent in the current MIRT literature;
therefore, we have now explained this in our "Model formulation" section.
The end of the second paragraph reads now as follows:

> [...] Finally, it is worth noting that
this formulation assumes *within-item multidimensionality* in general,
which will be present whenever more than one component of $`r DISCR_VECTOR`$
is non-null.
However, the model can also account for *between-item multidimensionality*
if all items are unidimensional
(i.e., their $`r DISCR_VECTOR`$ parameters only have one non-null component).

###### 5. Limitation is missing and should be added here.

Our last two paragraphs mention the most important limitations of our work.
We clearly demarcate the limited scope of our conclusions, and warn against overgeneralizing our results to other MIRT models or the common factor model.
We have put a great effort into writing our limitations in a way that is elegant and, at the same time, avoids cliché sentences that add no value to our manuscript.
Therefore, we would rather not add a separate "Limitations" section.

###### 6. Conclusion is missing and should be added here.

Similarly to the previous comment, we believe that our conclusions are thoroughly and carefully elaborated in our "Discussion" section, and we would prefer not to add an explicit "Conclusions" header.

# Responses to Reviewer 2:

###### [...] it is difficult to ascertain its practical value, as it is unclear how it can be applied in the real world.

We understand that it may be difficult to see the practical value of our proposal.
However, we argue that our main contribution is solidly founded on sound formal derivations, and that its value lies in the theoretical contribution it represents.
In practical terms, we believe that our results may help better diagnose the validity of the items in multidimensional tests, as both the multidimensional parameters and the vector representation we put forward are a more faithful representation of the properties of those items.
This happens because item parameters should be interpreted in their context, that is, as properties of the response data that result from the interaction of an instrument and the sample where it is applied, as argued by @reckase_multidimensional_2009.

We expect that the changes we have made in the new version of our manuscript help clarify the practical value of our proposal.
In this respect, please see especially the responses to the editor in pp. <!-- TODO: Complete (3rd response, last, and 2nd to last) --> in this letter.

###### While the manuscript includes a graphical representation example with two-dimensional latent variables, real-world applications often involve higher dimensions (e.g., the Big Five personality traits or Cattell’s 16 personality factors). If the current results cannot be applied to the high-dimensional cases, would that significantly limit their practical value and applicability?

Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that accounts for any number of dimensions.
We have found that our previous version was not explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:

> Consider a basis $`r LS_BASIS_EQ`$,
with $`r LS_BASIS`$ generally non-orthonormal.
Let $`r LATENT_SPACE`$ be a $`r N_DIMS`$-dimensional latent space
in basis $`r LS_BASIS`$ (with $`r N_DIMS`$ any positive integer),
where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
According to the `r MODEL_ACRONYM` model, the probability of
a positive response to item $`r ITEM_INDEX`$ is [...]

Also, the "Graphical Representation Example" section now begins:

> Up to now, all our derivations
have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions [...]

###### 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect. It should be “$\mathcal{C}$”.

Thank you so much to Reviewer 2 for pointing out this typo.
We have corrected it in the manuscript accordingly.

###### 2. The notations in the manuscript lack clarity, particularly in Eq. (19) on page 12. Based on Eq. (5) and Eq. (15), $\mathbf{a}_i$ in the numerator of Eq. (19) should be $\mathbf{a}_i^{\mathcal{B}*}$.

This case was also a typo; we have now corrected it in the manuscript as well.

###### Similar issues occur in Eqs. (26) and (27) on page 17.

<!-- # TODO: Equation references change? -->
Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7.
(That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
However, we acknowledge that we have skipped a step in our discourse, thus making it harder than necessary to follow along.
To correct this, we have modified the immediately preceding paragraph to the following:

<!-- # TODO: Confirm equation references -->
> Of course, from Equations (25) and (26), it is easy to express
the coordinates in terms of the parameters in the original model formulation.
Given $`r DISCR_VECTOR_COORDS_EQ`$,
we can express the coordinates in terms of the model parameters as

###### 3. In the second graph on page 18, the authors state, “*That is, the less discriminating the item, the further it will be from the origin.*” However, the value of $D_i$ is influenced not only by the $MDISC_i$ parameter but also by $d_i$. For instance, in Table 1 on page 20, $d_i$ for item 1 is zero, meaning that regardless of $MDISC_i$, $D_i$ remains zero.

The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may confuse the reader to refer to the $MDISC_i$ and overlook the influence of $d_i$.
We have added a clarification to this at the end of the paragraph, which now reads:

> That is, the less discriminating the item is,
the further away it will be from the origin.
Of course, it is also directly proportional to
the intercept parameter $`r INTERCEPT_PARAM`$.
This implies that, if $`r INTERCEPT_NULL_EQ`$,
$`r DISTANCE_NULL_EQ`$ as well,
regardless of the multidimensional discrimination.

# Responses to Reviewer 3:

###### The manuscript highlights the fact that the coordinate system used to represent results typical of educational achievement and other types of social measures is at the discretion of the person performing the analysis, or the entities developing computer software, if default settings are used in the analysis.

Although we argue indeed that the coordinate system is arbitrary, our stance is also that there is a more interpretable approach than simply assuming orthogonal axes.
Such an approach implies that there is a geometric basis (and thus a coordinate system) that can be considered a faithful representation of the covariance structure.

###### This is a very important issue and it is not frequently discussed in the research literature. Two researchers can administer the identical instrument to samples of individuals with different characteristics – say typical high school students versus those who have a different native language – and get different representations of the test results from analyses programs, such as the correlations between estimated constructs.  This manuscript stimulates much thought related to this problem.

We are glad that the reviewer has found our work thought-provoking.
We argue indeed that different samples may yield different latent covariance structures.
While this problem has received much attention in the factor analysis literature (i.e., factorial invariance), it has been largely overlooked in the MIRT literature, where any covariance structure is usually represented in orthogonal axes (implying an orthonormal basis).

###### [...] I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes. The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals. How are these features interpreted when the coordinate axes are non-orthogonal?

We argue that non-orthogonal axes provide a faithful representation of the latent covariance structure.
We have made the following addition in the third paragraph of our introduction to reflect this fact:

> [...] the assimilation of correlated latent traits to non-orthogonal
axes is prevalent in the common factor literature
[see, e.g., @harman_modern_1976]:
Indeed, the cosine between any two axes (or, more generally, vectors)
is considered as the geometrical equivalent of the correlation between
the variables they represent. [...]

In geometrical terms, the cosine between axes directly maps to the correlation between the two dimensions represented by those axes.
In the orthogonal space, the angles between items are "distorted" and do not faithfully represent the correlation among those items.
We have now added a paragraph to our introduction to try to explain this fact intuitively.
Please refer to our answer to the editor on pp. 3-4 <!-- TODO: Review --> of this letter for more details on this addition.

Regarding the person parameters, the interpretation is more complicated:
If the original, correlated coordinates of the persons are represented in the orthogonal space, the shape of the scatter plot indeed represents the correlation between the latent trait coordinates.
However, our rationale is that a basis may be used such that, when the coordinates are transformed to an orthonormal one, the resulting scatter plot is spherical (i.e., its covariance is an identity matrix).
This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
However, please bear in mind that this transformation in the latent space is a reasoning artifact to obtain the proper expression of the item parameters in the general case; for further explanation about this, please see our response to the editor in pp.  <!-- TODO: Complete (5th response) --> lines  <!-- TODO: Complete --> of this letter.
We have tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p. 14, lines 15-22, or p. <!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).

###### Care needs to be taken when using nonorthogonal coordinates that the distances between points in the space are correctly represented. What formula should be used for the distance between two points in each case?

The distances between points are given by the norm,
be it in the latent space or the test space.
We propose that the norm of both items and trait vectors should be invariant to
changes of basis and, interestingly, when considering the covariance-based version of the multidimensional parameters, the distance between points in the latent space is given by the Mahalanobis distance.
We have reflected this in our new version by adding the following to the end of the "Covariance-based Version" section:

> [...] Moreover, it is important to notice that,
when the latent trait inner product Gram matrix
is defined as $`r INNER_PROD_COV_COND`$,
the norm of a trait vector in the latent space is given by
the Mahalanobis distance, i.e., $`r MAH_DIST_EQ`$ [@mahalanobis_reprint_2018].

Also, we have made changes and additions to
the third  <!-- TODO: Review --> paragraph of our Discussion, which now reads:

> [...] Once again, this is relevant in applications where potentially different bases are involved.
Furthermore, we have found that using this version yields a norm in
the latent space given by the Mahalanobis distance.
This statistical distance not only takes
the latent covariance structure into account,
but also makes the latent space norm invariant to changes of basis. [...]

###### Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005). It would help if a page number were provided for the discussion.

We apologize for the lack of clarity.
We intended to refer to footnote 1 on page 16; we have now completed the reference.

###### I also wonder about the Harman (1970) reference. I know that he shows a plot of length of arm against length of leg using orthogonal axes for the plot even though the two measures are correlated. I am not not sure how the axes should be oriented to reflect the correlation between the two measures.

We intended to refer to section 4.8 (*Distance and Angle in General Cartesian Coordinates*), where he explains how to compute coordinates that "involve the inclination of the reference axes" [@harman_modern_1976, p. 60].
Please note that, as we do not have access to the 2nd edition anymore, we have used the 3rd one in our new version, and updated all the references accordingly.

Regarding the scatter plot you mention, we have not been able to find it (at least not in the 3^rd^ one we are using now, even though the example you refer to appears at the end of section 2.5, pp. 22-23).
However, we acknowledge that such a plot is the most common representation of the values realized by different cases (persons, in our case) in two variables, correlated or not.
In this regard, we would kindly refer you to our third <!-- TODO: Expressed ok? --> response to your comments, on p. <!-- TODO: Complete --> of this letter, where we discuss the representation of person parameters in the last two paragraphs.
<!-- TODO: Possibly complete with references to other responses? -->

###### Page 6  Line 21 The expression on this line needs more explanation. I am not sure what the expression is supposed to represent.

This is the expression of a vector represented in a basis, in this case, stating that the vector can be represented in each of the two bases interchangeably.
Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:

> [...] such that $`r TRAIT_VECTOR_DEF_EQ`$.
That is, the geometric representation of the algebraic vector $`r TRAIT_VECTOR`$
in each of the two bases is given by the linear combination of
the element vectors of the basis,
scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
However, note that, as $`r LATENT_SPACE`$ is originally represented in
$`r LS_BASIS`$, the geometric coordinates $`r TRAIT_COORDS`$ of
$`r TRAIT_VECTOR`$ in $`r LS_BASIS`$ are equal to the algebraic ones,
i.e., $`r TRAIT_GEOM_ALGEBR_EQ`$.

###### Page 7  Line 24 The meaning of the superscripts is not clear in this expression. What is the source of the -1T?  Is that the transpose of the inverse? Do the -1 superscripts indicate an inverse?

The -1 superscripts indeed indicate the inverse of a matrix.
This has been an unfortunate typesetting problem of the submission system,
that renders the -1 and the *T* at the same height.
We have sorted this out by using parentheses, as you will be able to see in our new submission.

###### Page 13 Line 4 The implications of the word “indistinctly” in this sentence are not clear.

This is a mistake on our side, we are afraid to admit, the product of a careless translation of a false friend from our mother tongue Spanish ("indistintamente"), which means "indifferently".
We have changed the term to the mathematically appropriate "interchangeably" in our new version.

###### Page 13 Line 8 The statement “the test space is where the item parameters naturally lie” seems to imply there is a correct representation of the item parameters. But, the correlations between constructs changes with the sample of examinees and the coordinates can easily be rotated and translated giving different representations of the item parameters. I would like to see support for the statement.

We think that our word choice has been unfortunate here.
What we want to imply is that there are two different spaces, the latent space, and the test space, and that, unlike the latent traits, which "naturally lie" in the latent space, the item parameters "naturally lie in the test space", even though they can be projected onto the latent space.
This is implicit in the definition of the test space itself (see p. 7 line 8 in our original submission, or p. <!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen to represent the items.
This basis does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction, so it can be the canonical basis, or any other, orthonormal or not, and not necessarily representing the covariance structure.

We also argue that the test space and the latent space are two distinct spaces, even when the basis of the latent space (and, consequently, that of the test space) is orthonormal.
It is our understanding that,  due to a lack of need for this distinction, this is overlooked in Reckase's [-@reckase_difficulty_1985] and Reckase & McKinley's [-@reckase_discriminating_1991] original formulations.
We have added this latter clarification in our new version (p. <!-- TODO: Complete --> lines <!-- TODO: Complete -->):

> [...] the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
[Note that in the original derivations by @reckase_difficulty_1985;
and @reckase_discriminating_1991, both matrices simplify to the identity matrix,
so there was no need to make this distinction.]
This implies that the axes defined by the test space basis, [...]

###### Page 19 Line 8 There is an interesting distinction that seems to be lost here. Attributes can be measured independently, such as height and weight and yield correlated results. Under those circumstances, should axes be represented as orthogonal, or should the axes represent the correlation. This has an implication for the formula that is used to compute the distance between points in the space.

As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response on p. <!-- TODO: Complete --> in this letter).
In this respect, we should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
Our argument is that it is when we represent the items that we should use axes that "faithfully represent the covariance structure", regardless of how we represent the attributes.

###### Page 22 Figure 1 It seems that both of these plots are shown with orthogonal coordinate axes. But the orthogonal axes have changed so that the nonorthogonal axes can be shown in the orthogonal representation.

That is exactly the case.
All graphical devices require that coordinates are represented in a rectangular Cartesian system such that the first coordinate maps to the horizontal dimension and the second coordinate to the vertical one.
Therefore, even when computing the coordinates in a non-orthonormal basis, we need to transform them to the canonical one to plot them.
This is explained in our manuscript, on p. 19 (lines 9-14) in the original version, and on p. <!-- # TODO: Complete --> (lines <!-- # TODO: Complete -->) in the new submission.
We are unsure about how to make this clearer, we are afraid, so we would appreciate any suggestions.

###### Page 26 Line 8 This discussion seems to miss the distinction between correlations between traits and dimensions of sensitivity of the test items.

To our understanding, one may assume a theoretical dimensional sensitivity of a test space that is higher than the dimensionality of the latent space, which we acknowledge we may have overlooked.
However, in such a case, the (minimum) dimensionality of the data would be "the smaller of the dimensions or variation for the people and the dimensions of sensitivity for the test items" [@reckase_multidimensional_2009, p. 182].
That is, the dimensional sensitivity of the data would nevertheless be restricted by the dimensionality of the latent space.
This implies that the parameter values one can estimate for the items will represent a lower-dimensional solution, which would be the projection of the theoretically higher-dimensionally sensitive parameters on the lower-dimensional latent trait space, as discussed on p. 182 of @reckase_multidimensional_2009.
In practice, this implies that this solution is indistinguishable from one where the items are simply sensitive to the lower number of dimensions, with their parameters given by such a lower-dimensional solution, which would correspond to their projections from the alleged higher-dimensional space onto the lower-dimensional one.

As Reviewer 3 points out, the high correlations between traits and the restricted dimensionality they imply are different from the dimensional sensitivity of the test items.
However, a highly correlated latent space, as in dataset 3d1 in @reckase_multidimensional_2009, may induce a dimensional restriction on the latent space (as indicated by the rank-1 covariance matrix), which is indistinguishable from a lower-dimensional (unidimensional, in this case) solution.
These two facts, the indistinguishability of the "projected high-dimensional" solution and the lower-dimensional one, on the one hand, and the indistinguishability of the "high-dimensional solution" with a low-rank covariance matrix and the "low-dimensional solution", on the other, render the theoretical implications of a higher-dimensional item sensitivity impossible to test in a case like this example.
This is acknowledged by @reckase_multidimensional_2009, who states that "if the process used to generate the data were not known, the interpretation of the analysis results would likely be that the test items all measured the same construct and that the data could be well fit by a unidimensional IRT model" (p. 189).

However, in our manuscript, we intend to go one step further and show how the latent space covariance affects how the item parameters are interpreted.
We have shown that non-orthonormal axes are a more faithful geometrical representation of the test space when the latent covariance is not an identity matrix.
From the model formulation, one can note that the components of the discrimination parameter serve as a scaling factor of the effect each latent trait has on the response probability function.
As such, if they were interpreted as abstract parameters in a high-dimensional space that can be projected onto the latent trait axis, then the model formulation would consider those projections instead of the original item parameters.
Therefore, the discrimination parameter values should be considered as being "parallel" to the latent space axis.
In other words, each discrimination parameter value is itself the projection of the item vector length onto its corresponding axis.

In this sense, having such high correlations among latent traits not only makes them practically unidimensional; it also implies that the item parameters must be practically aligned with the "unidimensional" latent space axis.
In other words, even if a test is designed to be sensitive to three dimensions, if those three dimensions are highly correlated, the items will be practically sensitive to a single dimension.
This will happen regardless of whether one considers that there is a higher-dimensional test space from which the alleged single dimension is a subspace (where the item parameters will be a projection of the higher-dimensionally sensitive ones), or whether one considers that the items are simply sensitive to a single dimension, with the item parameters given simply by the 2PL model.
We believe this is the reason why estimating such a model would yield "a first dimension with high positive $`r DISCR_SYM`$-parameters and other dimensions with much lower $`r DISCR_SYM`$-parameter estimates with many around zero" [@reckase_multidimensional_2009, p. 189].
In fact, we predict that the non-zero discrimination value estimate will not match any of the values of the generating discrimination parameter, but $`r MDISC_COV_PARAM`$ which, in this case, actually matches the discrimination parameter of the unidimensional 2PL model.
Although we acknowledge the lack of time to properly check that this prediction is fulfilled, we will gladly do so in another review round if the editor and Reviewer 3 consider it necessary.

The aforementioned considerations have led us to use this edge case as a very illustrative example for our purposes, because it allows the reader to understand easily that all items should be aligned with the only dimension being measured.
It is therefore easy to grasp the intuitive notion that all direction cosines should be expected to be equal to 0, and in the more general case, to represent the relative alignment of the item vector with all the axes.
However, in doing so, we may have skipped a step in our reasoning, assimilating the dimensionality of the latent space with the dimensional sensitivity of the items, as Reviewer 3 points out.

To clarify this point better, we have tried to better express this distinction, emphasizing that, in this case, the unidimensionality induced by the rank-limited covariance matrix restricts the dimensional sensitivity of the test space.
The aforementioned paragraph now reads (p. <!-- # TODO: Complete -->, lines <!-- # TODO: Complete -->):

> [...] In that example,
he uses a covariance matrix that practically has rank 1,
implying that the (alleged) three latent dimensions collapse into a single one.
This makes the latent space unidimensional in practice,
which in turn restricts the dimensionality of any response dataset
generated from a test measuring those traits.
Therefore, "the data could be well fit by a unidimensional IRT model"
[@reckase_multidimensional_2009, p. 189].
In such a case, when all the items are strictly aligned with
the one dimension being measured,
we should expect the direction angles to be extremely close to 0 for all items.
The reader interested can check that this is the case for
the covariance-based version of the direction angles
in the table in Online Resource 1.

###### Page 29 Line 7 I am curious if the values of discrimination are comparable from the two representations. It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.

We are not completely sure to have understood this comment.
The discrimination values in the original `r MODEL_ACRONYM` metric are comparable, in the sense that they yield the same model probabilities for the same latent trait coordinates, independently of the covariance structure of the latent space.
However, the covariance-version of the multidimensional discrimination is different, and will indeed be dependent on the metric of the test space, which is defined by its geometrical basis, which determines the directions and scales of the axes.
In this case, the basis has been chosen to be a geometric representation of the correlation matrix (as given by the expression $`r INNER_PROD_CORR_COND`$, please see p. <!-- # TODO: Complete -->, line <!-- # TODO: Complete -->, in the new version), and therefore, the scales are not affected (only the relative directions among the axes).

We would kindly ask to elaborate further on this question if the Reviewer feels it has not been completely addressed.

###### Page 30 Line 8 It would be helpful to show the orthogonal and nonorthogonal representations of items and then discuss the interpretation of the parameter estimates based on each of the representations. I have been unable to determine the value of using the nonorthogonal representation. The nonorthogonal representation is more complex, and it seems sample dependent because it uses the observed correlations/covariances. What is the value added of using the more complex representation?

We have found this comment puzzling as, to our understanding, what Reviewer 3 is asking is what we do in Figure 1 (p. 22 in our original submission, p. <!-- # TODO: Complete --> in the new one).
We have shown the same items, with as many different cases as possible, represented in both orthogonal and nonorthogonal coordinates, and explained the interpretation of the multidimensional parameters in both spaces.
We would appreciate it if Reviewer 3 would kindly clarify this comment, so we can address it properly.

On the other hand, to our understanding, the representation is always sample dependent because the item parameters also are.
Our argument is that the non-orthogonal coordinate system is a more faithful representation of the latent covariance structure, as the cosines between the axes are the geometric equivalent of the latent trait correlations.
We argue that this is the value of using this more complex representation, and so we have tried to convey in our Discussion section (please see p. 31, line 19 to p. 32, line 4, in the original version).

We acknowledge, though, that this may have been explained without enough clarity, and therefore striven to improve this discussion.
The aforementioned paragraph now reads:

> [...] This implies that the axes defined by the test space basis,
and not the ones defined by the latent space basis,
are a geometrical representation of the covariance structure.
Indeed, the cosines of the angles among the test space axes are
the geometric equivalent of the correlations. [...]

# References

### 2025-05-28 output/Letter_to_editor.Rmd
Set apart comments from response letter to editor

--- 2025-05-28 output/Letter_to_editor_reviewers.Rmd
@@
 ---
-
-```{r setup}
-#| include: false
-
-# Includes:
-
-library(here)
-library(knitr)
-library(patchwork)
-library(ggplot2)
-library(rstudioapi)
-library(svglite) # Used by {knitr} and {ggplot2} to render the figure to SVG;
-                 #   added to prevent {renv} from reporting inconsistent state.
-library(officer)
-
-ROOT_DIR <- here()
-DOC_DIR  <- file.path(ROOT_DIR, "output")
-
-opts_knit$set(root.dir = ROOT_DIR)
-opts_chunk$set(
-  echo     = FALSE,
-  message  = FALSE,
-  cache    = FALSE,
-  dev      = "png",
-  dpi      = 1200L,
-  dev.args = list(type = "cairo-png"),
-  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE),
-  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
-)
-
-# Output configuration options:
-options(digits = 3)
-```
-
-```{r sources}
-#| cache:   false
-#| include: false
-source("R/Formulae.R",  encoding = 'UTF-8')
-source("R/Constants.R", encoding = 'UTF-8')
-```
 
@@
 28040 Madrid, Spain
-
-# Responses to the action editor:
-
-###### I have a concern with making the latent ability axes nonorthogonal to represent the covariance structure between the latent abilities.
-Recognizing that the data used to estimate MIRT parameters represents an interaction between the examinees and the items and that different groups of examinees can result in different covariance structures and thus quite different item parameter estimates.
-
-Indeed, different groups of examinees can result in different covariance structures.
-However, in our understanding, the same fact can be predicated about the item parameters themselves, even without considering the covariance structure at all.
-For example, in unidimensional IRT, the item parameters may be different for different groups of examinees, even if the variance is constrained to be equal across groups; this is indeed the basis of methods such as multigroup IRT.
-
-It is our stance that the covariance structure, understood as a property of the data, of course, should be taken into account when determining the multidimensional properties of the items.
-That is what we propose with the covariance-based (and correlation-based) versions of the multidimensional parameters.
-
-###### The statement (p.13, line 8) about “the test space where item parameters naturally lie” needs greater explanation with examples.
-
-We acknowledge and apologize for the lack of clarity in this sentence.
-Please refer to our response to Reviewer 3 on p. <!-- TODO: Complete --> of this letter.
-
-###### Like Reviewer 3, I do not understand how the magnitude of discrimination values (p 29 , 7) can be compared/interpreted values of discrimination are comparable from the two representations.
-It seems that the metric of the represented space has changed, and equivalent numbers will not have equivalent interpretations.
-
-As the editor says, the basis used to represent the test space changes its metric.
-We also understand that the metrics of the two representations are not comparable, and thus one should not compare items represented with the agnostic and the covariance-based version of the multidimensional parameters.
-However, when making multi-group comparisons, we think that taking into account the covariance-based structure of the latent space will provide more realistic information, and thus one should use the covariance-based version.
-
-In addition to other (minor) changes, we have added the following paragraph to our introduction to give an intuitive explanation of this fact:
-
-> Take the following as an example:
-A hypothetical complex item measures two latent traits with equal validity
----i.e., its discrimination parameters are equal with value $a$.
-Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
-we would find that $`r MDISC_SYM` = \sqrt{2} a_i$.
-Assume we apply this item in two groups where, taken to the extreme,
-the two measured dimensions are orthogonal in the first group
-and perfectly correlated in the second one.
-While the previous formula seems correct in the first group,
-one can intuitively say that,
-because the two dimensions are *perfectly aligned* and thus
-compound up in the second group,
-the item `r MDISC_SYM` should be the algebraic sum of both components,
-that is, $2 a_i$.
-Similarly, the correlation between two simple items,
-each one measuring a distinct dimension,
-should be uncorrelated in the first group,
-but perfectly correlated in the second one.
-As the cosine between variables is usually understood as
-the geometrical representation of their correlation,
-the cosine between the two items should be 1 in the second group.
-However, using an orthogonal representation
-would yield null cosines in both cases.
-
-There are several references in the literature that have found contradictory results which, we believe, are due to disregarding the latent covariance structure (i.e., using Reckase & McKinley's formulae in a non-orthonormal space).
-For example, @min_evaluation_2007 [p.
-49] computes the item parameters in the M2PL from a priori fixed values of $MDISC_i$ and $MDIFF_i$ (the name they use for $MID_i$), using @reckase_difficulty_1985 and @reckase_discriminating_1991's formulae.
-They find that "[...] bias [...] was more apparent when ability distribution was separated from the
-default distribution, such as was the case in groups 2 to 5".
-They attribute this to a possible "degree of confounding among errors of parameter estimation and scale transformation", but their groups 2 to 5 happen to all have correlations equal (or almost equal) to .5.
-(We consider that discussing these examples is out of the scope of this manuscript, but we will be glad to consider some of them if this editor deems it necessary.)
-
-Also, we believe this is what happens when MIRT software such as NOHARM tries to force a correlated latent trait structure into its default orthogonal solution, as this editor himself has noted elsewhere [@ackerman_multidimensional_2005-1, p.
-11].
-While we consider this to be also out of the scope of the current paper, we will gladly add further details to explain why these results are contradictory and how our proposal may solve the problem.
-
-###### I also wonder about how one compares the standard errors of the item parameters when the covariance covariance structures are different.  Are they directly comparable?
-
-The standard errors of the multidimensional parameters are, without a doubt, more difficult to compute for the covariance-based case than for the agnostic one.
-Computing the covariance errors in the former would require taking into account the standard errors of the item parameters themselves and at least the standard errors of the correlations (in the best-case scenario, assuming standard latent dimensions; otherwise, those of the variances and covariances).
-However, to our knowledge, multidimensional item parameters, as defined by @reckase_difficulty_1985 and @reckase_discriminating_1991, have been previously defined and used only from a theoretical perspective, without ever taking into account the uncertainty resulting from the model estimation.
-
-Regarding the "original" item parameters in the classical M2PL metric, we understand that the standard errors of the item parameters are independent of the definition of the multidimensional parameters used.
-Software such as the R package {mirt} [@chalmers_mirt_2012] uses a method similar to exploratory factor analysis to compute the item parameter estimates and the standard errors, based on an oblique rotation to approximate simple structure.
-One can see the help of the [`mirt()` function](https://www.rdocumentation.org/packages/mirt/versions/1.43/topics/mirt) therein (please see description of argument `SE.type`) for a detailed description of how this package computes the standard errors.
-
-###### I would like to see the motivation and psychometric utility of this approach.
-Does it improve the interpretation of the latent abilities and thus result in more insight into how the results are used?
-The author needs to address this with legitimate examples for testing practitioners.
-
-We think the motivation and psychometric utility of our approach are more related to the interpretation of the item parameters than with the latent abilities, which we believe exceeds the scope of this manuscript and could be the focus of another paper.
-Please note that we only use the transformation in the latent space to an orthonormal basis as a reasoning artifact to obtain the proper expression of the item parameters in the general case, and that we do it in an abstract sense, before even considering whether the original basis is a representation of the covariance matrix.
-
-However, to our understanding, the more precise interpretation of the item parameters that results from our approach indeed provides a better insight into how the results are used.
-A multidimensional representation of the items provides benchmarks for comparing different latent trait levels, as discussed by @reckase_difficulty_1985.
-Therefore, the more accurate such a representation is, the better the latent trait interpretation will be.
-
-Please see our answer on p. <!-- TODO: Complete (3?) --> of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
-<!-- TODO: Complete -->
-
-###### What implications does your approach have for standard psychometric procedures like equating.
-
-Fitting a single-group model where the latent dimensions are correlated affects the computation of the multidimensional parameters we propose here.
-Given the scale underidentification in IRT, it is relatively common to fix the latent variances to 1, so it is unlikely that non-standard covariances affect these applications.
-In this sense, we have mentioned in our discussion that "in many applications both [the covariance- and correlation-based] versions will be coincident, fortunately" (p.
-31, lines 2-3, or p.
-<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
-However, in multigroup procedures, such as equating and linking, non-standard (in some of the groups at least) may indeed affect the resulting multidimensional parameters.
-Additionally, it is worth noting that the correlation matrix may differ among groups, which can also have a substantial effect on the multidimensional parameter values.
-This is relevant because, even if the item parameters are equal across groups, the covariance structure may differ; thus, the multidimensional parameters will not be the same.
-Conversely, the same multidimensional parameters may correspond to different parameter models if the latent covariance structure is different across groups.
-Although it would be interesting to explore whether it is more realistic to assume equal M2PL parameters across groups (and thus possibly different multidimensional parameters) or vice-versa, we believe that this is an empirical question that lies outside the scope of our current manuscript.
-
-We have probably made the mistake of assuming that this was implicitly understood, for which we apologize to the editor.
-To improve and address this point better, we have made some additions to the second paragraph <!-- TODO: Third now?
---> of our "Discussion" section (p.
-30, lines 5-9 in the original submission, or p.
-<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new one), which now reads:
-
-> [...] Moreover, in applications where the latent space structure is critical
-(e.g., multi-group IRT, equating, and linking),
-the agnostic version may overlook relevant differences in
-the values of the multidimensional parameters.
-Different covariance structures across groups may significantly affect
-the value of these parameters,
-due to differences in either the variances or the correlation structure.
-Therefore, even when the model parameters are equal across groups,
-the multidimensional parameters may not be the same
-if the covariance structure differs across groups.
-Conversely, the same multidimensional parameters may correspond
-to different parameter models.
-The subsequent invariance of either the M2PL parameters
-or the multidimensional ones may lead to different interpretations of
-the same model when applied to different groups.
-Although this is an empirical question that requires further research,
-it could have significant implications for future theoretical developments
-in the understanding of MIRT theory.
-
-###### Are there standard computer programs that you would recommend to do the estimation/modeling you are proposing?
-
-Our manuscript does not propose an estimation or modeling method.
-On the contrary, what we propose is a way to compute and represent the multidimensional item parameters without disregarding the covariance structure of the latent space.
-Implementing the formulae we propose should be rather straightforward for most IRT researchers, simply using the standard output of most MIRT software.
-We have now added a discussion about this at the end of our fourth paragraph in the "Discussion" section. This addition reads:
-
-> [...] Computing the parameters in the orthonormalized
-(or just orthogonalized) space,
-by contrast, would require transforming the cosines to
-the original, non-orthonormal basis afterwards.
-Nevertheless, most available MIRT software solutions,
-such as flexMIRT [@cai_flexmirt_2024], IRTPRO [@cai_irtpro_2017],
-or R package {mirt} [@chalmers_mirt_2012],
-initially estimate an uncorrelated solution (at least in the exploratory case)
-and apply an oblique rotation afterwards.
-Taking advantage of the invariance property,
-we may compute the parameters in the orthogonal space
-(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
-Then, we may compute the direction cosines in the rotated solution by applying
-Equation (24).
-<!-- TODO: Check Eq. nº afterwards -->
-What a test practitioner should be aware of, though,
-is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
-after any non-orthonormal transformation will yield
-a $`r MDISC_AG_PARAM`$ value specific to that basis,
-unlike the invariant value of $`r MDISC_COV_PARAM`$.
-
-Notwithstanding, we have written extensive code in R for computing these parameters, which we intend to open up for public use and scrutiny in a GitHub repository that includes all the code, data, and examples (including all the R packages to re-create the necessary environment to grant computational reproducibility), allowing to generate all the tables, plots, and even the submitted versions of the manuscript.
-Using dedicated functions in our code, any interested researcher will be able to compute the multidimensional parameters from the output of any MIRT software [e.g., {mirt}, @chalmers_mirt_2012; flexMIRT, @cai_flexmirt_2024].
-As the journal guidelines [require a specific format for the Supplementary Material](https://www.cambridge.org/core/services/authors/publishing-supplementary-material),
-we chose not to include the repository as additional material.
-However, we have checked and the journal accepts and also encourages publishing additional materiales in [a proper, trusted repository](https://www.cambridge.org/core/services/where-to-share-your-data).
-We have thus made public our GitHub repository for the manuscript [Link URL obfuscated for blind peer-review] and indexed it in Software Heritage [Link URL obfuscated for blind peer-review] for identification with a SWHID.
-For the sake of transparency, we have also made this addition to the manuscript, in section "Applications to Examples from the Literature", before the @reckase_multidimensional_2009 example:
-
-> The interested reader can explore the complete code for these examples in the
-indexed Software Heritage repository for this paper, specifically:
-[URL obfuscated for blind peer-review]
-
-# Responses to Reviewer 1:
-
-###### 1. Introduction, too lengthy, 1-2 pages will be fine.
-
-Our introduction is a bit less than four pages long.
-We have provided concise paragraphs about the Multidimensional IRT model and its parameters, their assumptions, and our rationale for relaxing them.
-Finally, a last paragraph presents the objectives of our research.
-We honestly believe we have made a thorough effort to summarize the motivation of our research, and we find it difficult to shorten it further without losing relevant information to justify our work.
-Therefore, we are sorry to say that we cannot comply with the reviewer's request to shorten the introduction.
-
-###### 2. Introduction, do you have any hypothesis for the current study? Please briefly describe it at the end.
-
-Our work is entirely theoretical and thus employs a purely deductive method.
-Consequently, we do not have any hypothesis to test; rather, we present the results of our formal derivations, which address the objective described at the end of our introduction.
-
-###### 3. The details about the datasets should be described here.
-
-The datasets we use in our paper are examples of item parameter estimates, taken from the literature.
-As they only serve exemplification purposes and are not the main focus of our article, we deem it more appropriate to describe their details in the corresponding section of the manuscript.
-We have added a short reference to them in the last paragraph of our introduction (p. <!-- TODO: Complete -->, lines <!-- TODO: Complete -->), to make sure their pertinence as examples is clear.
-
-> These examples, from the cognitive and the non-cognitive domain, respectively,
-are taken from the MIRT literature:
-One of them is a classic dataset, often used to exemplify
-the use of the `r MODEL_ACRONYM` model [@reckase_multidimensional_2009];
-the other one, an application in marketing research [@tezza_modelo_2018].
-
-###### 4. Within or between multidimensional latent? This should be detailed.
-
-We think the context makes it clear that we are referring to "within-item" multidimensionality.
-In fact, the results are trivial in the "between-item" multidimensionality case,
-as the `r MDISC_SYM` parameter reduces to the only non-null component of the $`r DISCR_VECTOR`$ parameter.
-However, we acknowledge this terminology is prevalent in the current MIRT literature;
-therefore, we have now explained this in our "Model formulation" section.
-The end of the second paragraph reads now as follows:
-
-> [...] Finally, it is worth noting that
-this formulation assumes *within-item multidimensionality* in general,
-which will be present whenever more than one component of $`r DISCR_VECTOR`$
-is non-null.
-However, the model can also account for *between-item multidimensionality*
-if all items are unidimensional
-(i.e., their $`r DISCR_VECTOR`$ parameters only have one non-null component).
-
-###### 5. Limitation is missing and should be added here.
-
-Our last two paragraphs mention the most important limitations of our work.
-We clearly demarcate the limited scope of our conclusions, and warn against overgeneralizing our results to other MIRT models or the common factor model.
-We have put a great effort into writing our limitations in a way that is elegant and, at the same time, avoids cliché sentences that add no value to our manuscript.
-Therefore, we would rather not add a separate "Limitations" section.
-
-###### 6. Conclusion is missing and should be added here.
-
-Similarly to the previous comment, we believe that our conclusions are thoroughly and carefully elaborated in our "Discussion" section, and we would prefer not to add an explicit "Conclusions" header.
-
-# Responses to Reviewer 2:
-
-###### [...] it is difficult to ascertain its practical value, as it is unclear how it can be applied in the real world.
-
-We understand that it may be difficult to see the practical value of our proposal.
-However, we argue that our main contribution is solidly founded on sound formal derivations, and that its value lies in the theoretical contribution it represents.
-In practical terms, we believe that our results may help better diagnose the validity of the items in multidimensional tests, as both the multidimensional parameters and the vector representation we put forward are a more faithful representation of the properties of those items.
-This happens because item parameters should be interpreted in their context, that is, as properties of the response data that result from the interaction of an instrument and the sample where it is applied, as argued by @reckase_multidimensional_2009.
-
-We expect that the changes we have made in the new version of our manuscript help clarify the practical value of our proposal.
-In this respect, please see especially the responses to the editor in pp. <!-- TODO: Complete (3rd response, last, and 2nd to last) --> in this letter.
-
-###### While the manuscript includes a graphical representation example with two-dimensional latent variables, real-world applications often involve higher dimensions (e.g., the Big Five personality traits or Cattell’s 16 personality factors).
-If the current results cannot be applied to the high-dimensional cases, would that significantly limit their practical value and applicability?
-
-Although the graphical representation is limited to two dimensions, we propose a general algebraic formulation that accounts for any number of dimensions.
-We have found that our previous version was not explicit enough about this, so we have updated the first paragraph of the "Model Formulation" section to read as follows:
-
-> Consider a basis $`r LS_BASIS_EQ`$,
-with $`r LS_BASIS`$ generally non-orthonormal.
-Let $`r LATENT_SPACE`$ be a $`r N_DIMS`$-dimensional latent space
-in basis $`r LS_BASIS`$ (with $`r N_DIMS`$ any positive integer),
-where a respondent is represented by vector $`r TRAIT_VEC_IN_LS`$.
-According to the `r MODEL_ACRONYM` model, the probability of
-a positive response to item $`r ITEM_INDEX`$ is [...]
-
-Also, the "Graphical Representation Example" section now begins:
-
-> Up to now, all our derivations
-have considered $`r N_DIMS`$-dimensional spaces;
-that is, multidimensional parameters and item vector coordinates
-can be computed in an arbitrary large number of dimensions.
-However, as plotting more than two dimensions [...]
-
-###### 1. In the 8th line from the bottom on page 9, the subscript in the formula is incorrect. It should be “$\mathcal{C}$”.
-
-Thank you so much to Reviewer 2 for pointing out this typo.
-We have corrected it in the manuscript accordingly.
-
-###### 2.
-The notations in the manuscript lack clarity, particularly in Eq.
-(19) on page 12.
-Based on Eq.
-(5) and Eq.
-(15), $\mathbf{a}_i$ in the numerator of Eq.
-(19) should be $\mathbf{a}_i^{\mathcal{B}*}$.
-
-This case was also a typo; we have now corrected it in the manuscript as well.
-
-###### Similar issues occur in Eqs. (26) and (27) on page 17.
-
-<!-- # TODO: Equation references change? -->
-Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
-To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7.
-(That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
-However, we acknowledge that we have skipped a step in our discourse, thus making it harder than necessary to follow along.
-To correct this, we have modified the immediately preceding paragraph to the following:
-
-<!-- # TODO: Confirm equation references -->
-> Of course, from Equations (25) and (26), it is easy to express
-the coordinates in terms of the parameters in the original model formulation.
-Given $`r DISCR_VECTOR_COORDS_EQ`$,
-we can express the coordinates in terms of the model parameters as
-
-###### 3.
-In the second graph on page 18, the authors state, “*That is, the less discriminating the item, the further it will be from the origin.*” However, the value of $D_i$ is influenced not only by the $MDISC_i$ parameter but also by $d_i$.
-For instance, in Table 1 on page 20, $d_i$ for item 1 is zero, meaning that regardless of $MDISC_i$, $D_i$ remains zero.
-
-The reviewer is right in this respect; we made the mistake of assuming this relationship was obvious enough, but we acknowledge it may confuse the reader to refer to the $MDISC_i$ and overlook the influence of $d_i$.
-We have added a clarification to this at the end of the paragraph, which now reads:
-
-> That is, the less discriminating the item is,
-the further away it will be from the origin.
-Of course, it is also directly proportional to
-the intercept parameter $`r INTERCEPT_PARAM`$.
-This implies that, if $`r INTERCEPT_NULL_EQ`$,
-$`r DISTANCE_NULL_EQ`$ as well,
-regardless of the multidimensional discrimination.
-
-# Responses to Reviewer 3:
-
-###### The manuscript highlights the fact that the coordinate system used to represent results typical of educational achievement and other types of social measures is at the discretion of the person performing the analysis, or the entities developing computer software, if default settings are used in the analysis.
-
-Although we argue indeed that the coordinate system is arbitrary, our stance is also that there is a more interpretable approach than simply assuming orthogonal axes.
-Such an approach implies that there is a geometric basis (and thus a coordinate system) that can be considered a faithful representation of the covariance structure.
-
-###### This is a very important issue and it is not frequently discussed in the research literature.
-Two researchers can administer the identical instrument to samples of individuals with different characteristics – say typical high school students versus those who have a different native language – and get different representations of the test results from analyses programs, such as the correlations between estimated constructs.
-This manuscript stimulates much thought related to this problem.
-
-We are glad that the reviewer has found our work thought-provoking.
-We argue indeed that different samples may yield different latent covariance structures.
-While this problem has received much attention in the factor analysis literature (i.e., factorial invariance), it has been largely overlooked in the MIRT literature, where any covariance structure is usually represented in orthogonal axes (implying an orthonormal basis).
-
-###### [...] I was at a loss at the end of the discussion to identify the interpretive or methodological value that was gained by working with non-orthogonal axes.
-The correlation between items can be represented by the angle between item vectors in the space with orthogonal axes, and the correlation between coordinates for persons can be represented by the shape of the scatter plot of points representing the individuals.
-How are these features interpreted when the coordinate axes are non-orthogonal?
-
-We argue that non-orthogonal axes provide a faithful representation of the latent covariance structure.
-We have made the following addition in the third paragraph of our introduction to reflect this fact:
-
-> [...] the assimilation of correlated latent traits to non-orthogonal
-axes is prevalent in the common factor literature
-[see, e.g., @harman_modern_1976]:
-Indeed, the cosine between any two axes (or, more generally, vectors)
-is considered as the geometrical equivalent of the correlation between
-the variables they represent. [...]
-
-In geometrical terms, the cosine between axes directly maps to the correlation between the two dimensions represented by those axes.
-In the orthogonal space, the angles between items are "distorted" and do not faithfully represent the correlation among those items.
-We have now added a paragraph to our introduction to try to explain this fact intuitively.
-Please refer to our answer to the editor on pp. 3-4 <!-- TODO: Review --> of this letter for more details on this addition.
-
-Regarding the person parameters, the interpretation is more complicated:
-If the original, correlated coordinates of the persons are represented in the orthogonal space, the shape of the scatter plot indeed represents the correlation between the latent trait coordinates.
-However, our rationale is that a basis may be used such that, when the coordinates are transformed to an orthonormal one, the resulting scatter plot is spherical (i.e., its covariance is an identity matrix).
-This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
-However, please bear in mind that this transformation in the latent space is a reasoning artifact to obtain the proper expression of the item parameters in the general case; for further explanation about this, please see our response to the editor in pp.
-<!-- TODO: Complete (5th response) --> lines  <!-- TODO: Complete --> of this letter.
-We have tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p.
-14, lines 15-22, or p.
-<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
-
-###### Care needs to be taken when using nonorthogonal coordinates that the distances between points in the space are correctly represented.
-What formula should be used for the distance between two points in each case?
-
-The distances between points are given by the norm,
-be it in the latent space or the test space.
-We propose that the norm of both items and trait vectors should be invariant to
-changes of basis and, interestingly, when considering the covariance-based version of the multidimensional parameters, the distance between points in the latent space is given by the Mahalanobis distance.
-We have reflected this in our new version by adding the following to the end of the "Covariance-based Version" section:
-
-> [...] Moreover, it is important to notice that,
-when the latent trait inner product Gram matrix
-is defined as $`r INNER_PROD_COV_COND`$,
-the norm of a trait vector in the latent space is given by
-the Mahalanobis distance, i.e., $`r MAH_DIST_EQ`$ [@mahalanobis_reprint_2018].
-
-Also, we have made changes and additions to
-the third  <!-- TODO: Review --> paragraph of our Discussion, which now reads:
-
-> [...] Once again, this is relevant in applications where potentially different bases are involved.
-Furthermore, we have found that using this version yields a norm in
-the latent space given by the Mahalanobis distance.
-This statistical distance not only takes
-the latent covariance structure into account,
-but also makes the latent space norm invariant to changes of basis. [...]
-
-###### Page 3  Line 20 I could not find a discussion of nonorthogonal dimensions in Ackerman (2005). It would help if a page number were provided for the discussion.
-
-We apologize for the lack of clarity.
-We intended to refer to footnote 1 on page 16; we have now completed the reference.
-
-###### I also wonder about the Harman (1970) reference.
-I know that he shows a plot of length of arm against length of leg using orthogonal axes for the plot even though the two measures are correlated.
-I am not not sure how the axes should be oriented to reflect the correlation between the two measures.
-
-We intended to refer to section 4.8 (*Distance and Angle in General Cartesian Coordinates*), where he explains how to compute coordinates that "involve the inclination of the reference axes" [@harman_modern_1976, p.
-60].
-Please note that, as we do not have access to the 2nd edition anymore, we have used the 3rd one in our new version, and updated all the references accordingly.
-
-Regarding the scatter plot you mention, we have not been able to find it (at least not in the 3^rd^ one we are using now, even though the example you refer to appears at the end of section 2.5, pp.
-22-23).
-However, we acknowledge that such a plot is the most common representation of the values realized by different cases (persons, in our case) in two variables, correlated or not.
-In this regard, we would kindly refer you to our third <!-- TODO: Expressed ok?
---> response to your comments, on p.
-<!-- TODO: Complete --> of this letter, where we discuss the representation of person parameters in the last two paragraphs.
-<!-- TODO: Possibly complete with references to other responses? -->
-
-###### Page 6  Line 21 The expression on this line needs more explanation. I am not sure what the expression is supposed to represent.
-
-This is the expression of a vector represented in a basis, in this case, stating that the vector can be represented in each of the two bases interchangeably.
-Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:
-
-> [...] such that $`r TRAIT_VECTOR_DEF_EQ`$.
-That is, the geometric representation of the algebraic vector $`r TRAIT_VECTOR`$
-in each of the two bases is given by the linear combination of
-the element vectors of the basis,
-scaled by the coordinates of $`r TRAIT_VECTOR`$ in that basis.
-However, note that, as $`r LATENT_SPACE`$ is originally represented in
-$`r LS_BASIS`$, the geometric coordinates $`r TRAIT_COORDS`$ of
-$`r TRAIT_VECTOR`$ in $`r LS_BASIS`$ are equal to the algebraic ones,
-i.e., $`r TRAIT_GEOM_ALGEBR_EQ`$.
-
-###### Page 7  Line 24 The meaning of the superscripts is not clear in this expression.
-What is the source of the -1T?
-Is that the transpose of the inverse?
-Do the -1 superscripts indicate an inverse?
-
-The -1 superscripts indeed indicate the inverse of a matrix.
-This has been an unfortunate typesetting problem of the submission system,
-that renders the -1 and the *T* at the same height.
-We have sorted this out by using parentheses, as you will be able to see in our new submission.
-
-###### Page 13 Line 4 The implications of the word “indistinctly” in this sentence are not clear.
-
-This is a mistake on our side, we are afraid to admit, the product of a careless translation of a false friend from our mother tongue Spanish ("indistintamente"), which means "indifferently".
-We have changed the term to the mathematically appropriate "interchangeably" in our new version.
-
-###### Page 13 Line 8 The statement “the test space is where the item parameters naturally lie” seems to imply there is a correct representation of the item parameters.
-But, the correlations between constructs changes with the sample of examinees and the coordinates can easily be rotated and translated giving different representations of the item parameters.
-I would like to see support for the statement.
-
-We think that our word choice has been unfortunate here.
-What we want to imply is that there are two different spaces, the latent space, and the test space, and that, unlike the latent traits, which "naturally lie" in the latent space, the item parameters "naturally lie in the test space", even though they can be projected onto the latent space.
-This is implicit in the definition of the test space itself (see p.
-7 line 8 in our original submission, or p.
-<!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen to represent the items.
-This basis does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction, so it can be the canonical basis, or any other, orthonormal or not, and not necessarily representing the covariance structure.
-
-We also argue that the test space and the latent space are two distinct spaces, even when the basis of the latent space (and, consequently, that of the test space) is orthonormal.
-It is our understanding that,  due to a lack of need for this distinction, this is overlooked in Reckase's [-@reckase_difficulty_1985] and Reckase & McKinley's [-@reckase_discriminating_1991] original formulations.
-We have added this latter clarification in our new version (p. <!-- TODO: Complete --> lines <!-- TODO: Complete -->):
-
-> [...] the latent space,
-where the person parameters are represented,
-has the inverse of the covariance (correlation) matrix as
-its inner product matrix.
-[Note that in the original derivations by @reckase_difficulty_1985;
-and @reckase_discriminating_1991, both matrices simplify to the identity matrix,
-so there was no need to make this distinction.]
-This implies that the axes defined by the test space basis, [...]
-
-###### Page 19 Line 8 There is an interesting distinction that seems to be lost here.
-Attributes can be measured independently, such as height and weight and yield correlated results.
-Under those circumstances, should axes be represented as orthogonal, or should the axes represent the correlation.
-This has an implication for the formula that is used to compute the distance between points in the space.
-
-As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response on p.
-<!-- TODO: Complete --> in this letter).
-In this respect, we should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
-Our argument is that it is when we represent the items that we should use axes that "faithfully represent the covariance structure", regardless of how we represent the attributes.
-
-###### Page 22 Figure 1 It seems that both of these plots are shown with orthogonal coordinate axes.
-But the orthogonal axes have changed so that the nonorthogonal axes can be shown in the orthogonal representation.
-
-That is exactly the case.
-All graphical devices require that coordinates are represented in a rectangular Cartesian system such that the first coordinate maps to the horizontal dimension and the second coordinate to the vertical one.
-Therefore, even when computing the coordinates in a non-orthonormal basis, we need to transform them to the canonical one to plot them.
-This is explained in our manuscript, on p. 19 (lines 9-14) in the original version, and on p. <!-- # TODO: Complete --> (lines <!-- # TODO: Complete -->) in the new submission.
-We are unsure about how to make this clearer, we are afraid, so we would appreciate any suggestions.
-
-###### Page 26 Line 8 This discussion seems to miss the distinction between correlations between traits and dimensions of sensitivity of the test items.
-
-To our understanding, one may assume a theoretical dimensional sensitivity of a test space that is higher than the dimensionality of the latent space, which we acknowledge we may have overlooked.
-However, in such a case, the (minimum) dimensionality of the data would be "the smaller of the dimensions or variation for the people and the dimensions of sensitivity for the test items" [@reckase_multidimensional_2009, p.
-182].
-That is, the dimensional sensitivity of the data would nevertheless be restricted by the dimensionality of the latent space.
-This implies that the parameter values one can estimate for the items will represent a lower-dimensional solution, which would be the projection of the theoretically higher-dimensionally sensitive parameters on the lower-dimensional latent trait space, as discussed on p.
-182 of @reckase_multidimensional_2009.
-In practice, this implies that this solution is indistinguishable from one where the items are simply sensitive to the lower number of dimensions, with their parameters given by such a lower-dimensional solution, which would correspond to their projections from the alleged higher-dimensional space onto the lower-dimensional one.
-
-As Reviewer 3 points out, the high correlations between traits and the restricted dimensionality they imply are different from the dimensional sensitivity of the test items.
-However, a highly correlated latent space, as in dataset 3d1 in @reckase_multidimensional_2009, may induce a dimensional restriction on the latent space (as indicated by the rank-1 covariance matrix), which is indistinguishable from a lower-dimensional (unidimensional, in this case) solution.
-These two facts, the indistinguishability of the "projected high-dimensional" solution and the lower-dimensional one, on the one hand, and the indistinguishability of the "high-dimensional solution" with a low-rank covariance matrix and the "low-dimensional solution", on the other, render the theoretical implications of a higher-dimensional item sensitivity impossible to test in a case like this example.
-This is acknowledged by @reckase_multidimensional_2009, who states that "if the process used to generate the data were not known, the interpretation of the analysis results would likely be that the test items all measured the same construct and that the data could be well fit by a unidimensional IRT model" (p.
-189).
-
-However, in our manuscript, we intend to go one step further and show how the latent space covariance affects how the item parameters are interpreted.
-We have shown that non-orthonormal axes are a more faithful geometrical representation of the test space when the latent covariance is not an identity matrix.
-From the model formulation, one can note that the components of the discrimination parameter serve as a scaling factor of the effect each latent trait has on the response probability function.
-As such, if they were interpreted as abstract parameters in a high-dimensional space that can be projected onto the latent trait axis, then the model formulation would consider those projections instead of the original item parameters.
-Therefore, the discrimination parameter values should be considered as being "parallel" to the latent space axis.
-In other words, each discrimination parameter value is itself the projection of the item vector length onto its corresponding axis.
-
-In this sense, having such high correlations among latent traits not only makes them practically unidimensional; it also implies that the item parameters must be practically aligned with the "unidimensional" latent space axis.
-In other words, even if a test is designed to be sensitive to three dimensions, if those three dimensions are highly correlated, the items will be practically sensitive to a single dimension.
-This will happen regardless of whether one considers that there is a higher-dimensional test space from which the alleged single dimension is a subspace (where the item parameters will be a projection of the higher-dimensionally sensitive ones), or whether one considers that the items are simply sensitive to a single dimension, with the item parameters given simply by the 2PL model.
-We believe this is the reason why estimating such a model would yield "a first dimension with high positive $`r DISCR_SYM`$-parameters and other dimensions with much lower $`r DISCR_SYM`$-parameter estimates with many around zero" [@reckase_multidimensional_2009, p.
-189].
-In fact, we predict that the non-zero discrimination value estimate will not match any of the values of the generating discrimination parameter, but $`r MDISC_COV_PARAM`$ which, in this case, actually matches the discrimination parameter of the unidimensional 2PL model.
-Although we acknowledge the lack of time to properly check that this prediction is fulfilled, we will gladly do so in another review round if the editor and Reviewer 3 consider it necessary.
-
-The aforementioned considerations have led us to use this edge case as a very illustrative example for our purposes, because it allows the reader to understand easily that all items should be aligned with the only dimension being measured.
-It is therefore easy to grasp the intuitive notion that all direction cosines should be expected to be equal to 0, and in the more general case, to represent the relative alignment of the item vector with all the axes.
-However, in doing so, we may have skipped a step in our reasoning, assimilating the dimensionality of the latent space with the dimensional sensitivity of the items, as Reviewer 3 points out.
-
-To clarify this point better, we have tried to better express this distinction, emphasizing that, in this case, the unidimensionality induced by the rank-limited covariance matrix restricts the dimensional sensitivity of the test space.
-The aforementioned paragraph now reads (p. <!-- # TODO: Complete -->, lines <!-- # TODO: Complete -->):
-
-> [...] In that example,
-he uses a covariance matrix that practically has rank 1,
-implying that the (alleged) three latent dimensions collapse into a single one.
-This makes the latent space unidimensional in practice,
-which in turn restricts the dimensionality of any response dataset
-generated from a test measuring those traits.
-Therefore, "the data could be well fit by a unidimensional IRT model"
-[@reckase_multidimensional_2009, p. 189].
-In such a case, when all the items are strictly aligned with
-the one dimension being measured,
-we should expect the direction angles to be extremely close to 0 for all items.
-The reader interested can check that this is the case for
-the covariance-based version of the direction angles
-in the table in Online Resource 1.
-
-###### Page 29 Line 7 I am curious if the values of discrimination are comparable from the two representations.
-It seems that the metric of the representational space has changed, and equivalent numbers will not have equivalent meaning.
-
-We are not completely sure to have understood this comment.
-The discrimination values in the original `r MODEL_ACRONYM` metric are comparable, in the sense that they yield the same model probabilities for the same latent trait coordinates, independently of the covariance structure of the latent space.
-However, the covariance-version of the multidimensional discrimination is different, and will indeed be dependent on the metric of the test space, which is defined by its geometrical basis, which determines the directions and scales of the axes.
-In this case, the basis has been chosen to be a geometric representation of the correlation matrix (as given by the expression $`r INNER_PROD_CORR_COND`$, please see p.
-<!-- # TODO: Complete -->, line <!-- # TODO: Complete -->, in the new version), and therefore, the scales are not affected (only the relative directions among the axes).
-
-We would kindly ask to elaborate further on this question if the Reviewer feels it has not been completely addressed.
-
-###### Page 30 Line 8 It would be helpful to show the orthogonal and nonorthogonal representations of items and then discuss the interpretation of the parameter estimates based on each of the representations.
-I have been unable to determine the value of using the nonorthogonal representation.
-The nonorthogonal representation is more complex, and it seems sample dependent because it uses the observed correlations/covariances.
-What is the value added of using the more complex representation?
-
-We have found this comment puzzling as, to our understanding, what Reviewer 3 is asking is what we do in Figure 1 (p. 22 in our original submission, p. <!-- # TODO: Complete --> in the new one).
-We have shown the same items, with as many different cases as possible, represented in both orthogonal and nonorthogonal coordinates, and explained the interpretation of the multidimensional parameters in both spaces.
-We would appreciate it if Reviewer 3 would kindly clarify this comment, so we can address it properly.
-
-On the other hand, to our understanding, the representation is always sample dependent because the item parameters also are.
-Our argument is that the non-orthogonal coordinate system is a more faithful representation of the latent covariance structure, as the cosines between the axes are the geometric equivalent of the latent trait correlations.
-We argue that this is the value of using this more complex representation, and so we have tried to convey in our Discussion section (please see p.
-31, line 19 to p.
-32, line 4, in the original version).
-
-We acknowledge, though, that this may have been explained without enough clarity, and therefore striven to improve this discussion.
-The aforementioned paragraph now reads:
-
-> [...] This implies that the axes defined by the test space basis,
-and not the ones defined by the latent space basis,
-are a geometrical representation of the covariance structure.
-Indeed, the cosines of the angles among the test space axes are
-the geometric equivalent of the correlations. [...]
-
-# References

### 2025-05-29 output/Response_to_action_editor_reviewers.Rmd
Review after @mluzardov's changes

@@
 Fitting a single-group model where the latent dimensions are correlated affects the computation of the multidimensional parameters we propose here.
-Given the scale underidentification in IRT, it is relatively common to fix the latent variances to 1, so it is unlikely that non-standard covariances affect these applications.
+Given the scale underidentification in IRT, it is relatively common to fix the latent variances to 1, so it is unlikely that non-standard distributions affect these applications.
 In this sense, we have mentioned in our discussion that "in many applications both [the covariance- and correlation-based] versions will be coincident, fortunately" (p.
@@
 <!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
-However, in multigroup procedures, such as equating and linking, non-standard (in some of the groups at least) may indeed affect the resulting multidimensional parameters.
+However, in multi-group procedures, such as equating and linking, non-unitary variances (in some of the groups at least) may indeed affect the resulting multidimensional parameters.
 Additionally, it is worth noting that the correlation matrix may differ among groups, which can also have a substantial effect on the multidimensional parameter values.
@@
 <!-- # TODO: Equation references change? -->
-Our intention in Equations (26) and (27) was to express the item vector coordinates in terms of the original model parameters directly.
+Our intention in Equations (26) and (27), numbered (27) and (28) in the new version, was to express the item vector coordinates in terms of the original model parameters directly.
 To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7.
@@
 However, we acknowledge that we have skipped a step in our discourse, thus making it harder than necessary to follow along.
-To correct this, we have modified the immediately preceding paragraph to the following:
+To correct this, we have modified the immediately preceding paragraph to the following (please note that the equation numbering has also increased by 1):
 
@@
 
-This is the expression of a vector represented in a basis, in this case, stating that the vector can be represented in each of the two bases interchangeably.
+This is the expression of a vector represented as a linear combination of
+the vectors that make up the basis in which it is represented
+(with the coefficients $`r TRAIT_VEC_COMP`$ and $`r TRAIT_VEC_ORTH_COMP`$ being
+the scalars corresponding to each basis).
+In this case, we intend to state that the vector can be represented in each of the two bases interchangeably.
 Following the reviewer's suggestion, we have added this clarification to the text, so it now reads:
@@
 <!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen to represent the items.
-This basis does not necessarily need to be a geometric representation of the covariance structure, as we argue in our introduction, so it can be the canonical basis, or any other, orthonormal or not, and not necessarily representing the covariance structure.
+This basis does not have to be necessarily a geometric representation of the covariance structure, as we argue in our introduction, so it can be the canonical basis, or any other, orthonormal or not, and not necessarily representing the covariance structure.
 

### 2025-05-30 output/Main_text.Rmd
Add discussion about multi-group procedures

@@
 
+Making a distinction between the two spaces can also have a significant impact
+on our understanding of MIRT applications that involve several latent spaces,
+such as multi-group equating and linking.
+Although the usual transformations among coordinate systems still apply
+under this consideration
+[see Chapter 8 of @reckase_multidimensional_2009,
+for an exhaustive discussion on the topic],
+the different geometric bases on which each of the two spaces is represented
+must be considered.
+However, as they are mutually dependent,
+any transformation in any of the two spaces will imply
+the inverse transformation in the other one,
+thus preserving the relationship between the two spaces.
+Nevertheless, using non-orthogonal rotation matrices for aligning the items
+(in a common-item design procedure) or persons
+[in a common-person one, see Chapter 9 in @reckase_multidimensional_2009]
+may have additional implications;
+for example, besides a rescaling and an origin translation, a transformation
+as defined, e.g., in Equation (8.31) of @reckase_multidimensional_2009,
+may imply a rotation that can be non-orthogonal,
+thus affecting the relative alignment among the test space axes,
+and subsequently among the latent space ones.
+Moreover, even if the transformation in one space is scale-invariant
+(i.e., the transform matrix has unitary diagonal elements),
+it may induce a rescaling in the other one,
+given by the diagonal elements of the inverse transform matrix.
+These implications will require future exploration of
+the relationships between the latent and test spaces
+as mathematically separate entities.
+
 Finally, we must also warn against generalizing these results

### 2025-05-30 output/Response_to_action_editor_reviewers.Rmd
Add discussion on multi-group procedures to letter
This includes suggestions and edits by @mluzardov.

@@
 This is relevant because, even if the item parameters are equal across groups, the covariance structure may differ; thus, the multidimensional parameters will not be the same.
+It is our understanding that both common item design and common person design procedures can be applied soundly under this approach.
+Despite the latent and the test space being (generally) represented in a different geometric basis, note that if one of them is known, the other one is determined.
+This linkage between the two spaces, along with the model invariance, implies that any transformation in one space must be accompanied by the corresponding transformation in the other.
+These relationships allow for both common-item and common-person design equating procedures, derived according to Reckase [-@reckase_multidimensional_2009, see Chapter 8], and do not change in our approach.
+For instance, if the common item-design procedure is used (where the item parameters in Form 2 of a test are transformed to resemble as closely as possible those in Form 1), both the item parameters and the covariance matrix will be affected.
+Although it is out of the scope of our paper to explore in depth the implications for these psychometric procedures, we have added a new paragraph to our discussion briefly explaining this fact and acknowledging our limitation.
+This new paragraph reads:
+
+> Making a distinction between the two spaces can also have a significant impact
+on our understanding of MIRT applications that involve several latent spaces,
+such as multi-group equating and linking.
+Although the usual transformations among coordinate systems still apply
+under this consideration
+[see Chapter 8 of @reckase_multidimensional_2009,
+for an exhaustive discussion on the topic],
+the different geometric bases on which each of the two spaces is represented
+must be considered.
+However, as they are mutually dependent,
+any transformation in any of the two spaces will imply
+the inverse transformation in the other one,
+thus preserving the relationship between the two spaces.
+Nevertheless, using non-orthogonal rotation matrices for aligning the items
+(in a common-item design procedure) or persons
+[in a common-person one, see Chapter 9 in @reckase_multidimensional_2009]
+may have additional implications;
+for example, besides a rescaling and an origin translation, a transformation
+as defined, e.g., in Equation (8.31) of @reckase_multidimensional_2009,
+may imply a rotation that can be non-orthogonal,
+thus affecting the relative alignment among the test space axes,
+and subsequently among the latent space ones.
+Moreover, even if the transformation in one space is scale-invariant
+(i.e., the transform matrix has unitary diagonal elements),
+it may induce a rescaling in the other one,
+given by the diagonal elements of the inverse transform matrix.
+These implications will require future exploration of
+the relationships between the latent and test spaces
+as mathematically separate entities.
+
 Conversely, the same multidimensional parameters may correspond to different parameter models if the latent covariance structure is different across groups.

### 2025-05-30 output/Response_to_action_editor_reviewers.Rmd
Adapt response on parameters in multi-group MIRT
This part has been re-written after including the response (and edits) added by @mluzardov.

@@
 
-Fitting a single-group model where the latent dimensions are correlated affects the computation of the multidimensional parameters we propose here.
-Given the scale underidentification in IRT, it is relatively common to fix the latent variances to 1, so it is unlikely that non-standard distributions affect these applications.
-In this sense, we have mentioned in our discussion that "in many applications both [the covariance- and correlation-based] versions will be coincident, fortunately" (p.
-31, lines 2-3, or p.
-<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
-However, in multi-group procedures, such as equating and linking, non-unitary variances (in some of the groups at least) may indeed affect the resulting multidimensional parameters.
-Additionally, it is worth noting that the correlation matrix may differ among groups, which can also have a substantial effect on the multidimensional parameter values.
-This is relevant because, even if the item parameters are equal across groups, the covariance structure may differ; thus, the multidimensional parameters will not be the same.
 It is our understanding that both common item design and common person design procedures can be applied soundly under this approach.
@@
 
+Regarding the multidimensional parameters, it is clear that different covariance matrices may have a substantial effect on the multidimensional parameter values.
+This is relevant for multi-group procedures because, even if the item parameters are equal, if the covariance structure differs across groups, the multidimensional parameters will not be the same.
 Conversely, the same multidimensional parameters may correspond to different parameter models if the latent covariance structure is different across groups.
-Although it would be interesting to explore whether it is more realistic to assume equal M2PL parameters across groups (and thus possibly different multidimensional parameters) or vice-versa, we believe that this is an empirical question that lies outside the scope of our current manuscript.
+Although it could be interesting to explore whether it is more realistic to assume equal M2PL parameters across groups (and thus possibly different multidimensional parameters) or vice-versa, we believe that this is an empirical question that also lies outside the scope of our current manuscript.
 

### 2025-06-02 output/Main_text.Rmd
Correct text size in body in example tables
Makes the table fill in better the size of the page
Reorder tables and figures
Makes a better use of the space in the manuscript
Apply "Header 1" style to document title
The document title is styled "normal", so it is changed to an inline "Header 1" instead

@@
 ---
-title: Abstract
 output:
@@
 ```
+
+# Abstract
 
@@
 
-```{r compose-example-items-table}
-```
-
-```{r set-flextable-wd}
-#| cache: false
-
-# Necessary for officedown to find the template file:
-opts_knit$set(root.dir = DOC_DIR)
-```
-
-<br>
-
-```{r example-items-table-out}
-#| cache:   false
-#| tab_id:  example-items-table-out
-#| tab.cap: Item Parameters for the Graphical Representation Example
-item_params_output
-```
-
-<br>
-
 Figure \@ref(fig:item-plot-out) illustrates the effect of
@@
 
-```{r compose-oblique-plot}
-```
-
-```{r compose-orthogonal-plot}
+```{r compose-example-items-table}
+```
+
+```{r set-flextable-wd}
+#| cache: false
+
+# Necessary for officedown to find the template file:
+opts_knit$set(root.dir = DOC_DIR)
 ```
@@
 
-```{r item-plot-out}
+```{r example-items-table-out}
 #| cache:   false
-#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
-#| fig.topcaption: true
-#| fig.height:     3.1
-#| fig.width:      6.65
-plot_orth + plot_oblique
-
-ggsave(file.path(DOC_DIR, "Figure_1.svg"))
-```
+#| tab_id:  example-items-table-out
+#| tab.cap: Item Parameters for the Graphical Representation Example
+item_params_output
+```
+
+<br>
 
@@
 even when they are not parallel to the other axis.
+
+```{r compose-oblique-plot}
+```
+
+```{r compose-orthogonal-plot}
+```
+
+<br>
+
+```{r item-plot-out}
+#| cache:          false
+#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
+#| fig.topcaption: true
+#| fig.height:     3.1
+#| fig.width:      6.65
+plot_orth + plot_oblique
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"))
+```
 
@@
 #|   Tezza et al. (2018)
-item_params_output
+item_params_output |>
+  fontsize(size = 11, part = "body")
 ```

### 2025-06-02 output/Letter_to_editor.Rmd
Make minor changes to letter to editor
Avoids redundancies

@@
 suggestions, which we believe have helped improve the quality of our paper.
+You will find our thorough responses in the document attached to this letter.
 However, we acknowledge we may not have resolved satisfactorily all the action
@@
 Therefore, we would kindly ask you to forward to them our good will
-and predisposition to address any further comments or suggestions,
-or clarify futher on the previous ones.
+and predisposition to address any other comments or suggestions,
+or clarify further on the previous ones.
 
@@
 (dvmorillo@psi.uned.es).
-We will be happy to address any further comments or suggestions
+We will be happy to address any additional comments or suggestions
 the action editor or the reviewers may deem necessary.

### 2025-06-02 output/Response_to_action_editor_reviewers.Rmd
Add corrections after comparing with Gdocs version
Update all cross references in response letter
To letter and manuscript pages, lines, and paragraphs

@@
 We acknowledge and apologize for the lack of clarity in this sentence.
-Please refer to our response to Reviewer 3 on p. <!-- TODO: Complete --> of this letter.
+Please refer to our response to Reviewer 3 on p. 18 of this letter.
 
@@
 
-As the editor says, the basis used to represent the test space changes its metric.
+As the editor says, the basis used to represent the test space changes its metric, which is determined by the Gram matrix ($`r INNER_PROD_MATRIX_INV`$, in contrast to $`r INNER_PROD_MATRIX`$ in the latent space).
 We also understand that the metrics of the two representations are not comparable, and thus one should not compare items represented with the agnostic and the covariance-based version of the multidimensional parameters.
@@
 
-Please see our answer on p. <!-- TODO: Complete (3?) --> of this letter for an example of how the interpretation of the item parameters may change when considering the covariance structure.
-<!-- TODO: Complete -->
+Please see our next answer for further clarification on the interpretation of the item parameters, as well as our response above, on pp.
+2-3 of this letter, for an example of how the interpretation of the item parameters may change when considering the covariance structure.
 
@@
 We have probably made the mistake of assuming that this was implicitly understood, for which we apologize to the editor.
-To improve and address this point better, we have made some additions to the second paragraph <!-- TODO: Third now?
---> of our "Discussion" section (p.
+To improve and address this point better, we have made some additions to the second paragraph of our "Discussion" section (p.
 30, lines 5-9 in the original submission, or p.
-<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new one), which now reads:
+31, lines 1-12 in the new one), which now reads:
 
@@
 Equation (24).
-<!-- TODO: Check Eq. nº afterwards -->
 What a test practitioner should be aware of, though,
@@
 As they only serve exemplification purposes and are not the main focus of our article, we deem it more appropriate to describe their details in the corresponding section of the manuscript.
-We have added a short reference to them in the last paragraph of our introduction (p. <!-- TODO: Complete -->, lines <!-- TODO: Complete -->), to make sure their pertinence as examples is clear.
+We have added a short reference to them in the last paragraph of our introduction (p. 6, lines 10-13), to make sure their pertinence as examples is clear.
 
@@
 We expect that the changes we have made in the new version of our manuscript help clarify the practical value of our proposal.
-In this respect, please see especially the responses to the editor in pp. <!-- TODO: Complete (3rd response, last, and 2nd to last) --> in this letter.
+In this respect, please see especially the responses to the editor in pp. 4-7 in this letter.
 
@@
 
-<!-- # TODO: Equation references change? -->
 Our intention in Equations (26) and (27), numbered (27) and (28) in the new version, was to express the item vector coordinates in terms of the original model parameters directly.
-To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7.
+To do that, we have made use of the equality $`r DISCR_VECTOR_COORDS_EQ`$, stated on p. 7, or p. 8, line 23 (third by the end) in the new submission.
 (That is, the geometric coordinates in the original test-space basis are equal to the algebraic values of the discrimination parameter.)
@@
 
-<!-- # TODO: Confirm equation references -->
 > Of course, from Equations (25) and (26), it is easy to express
@@
 We have now added a paragraph to our introduction to try to explain this fact intuitively.
-Please refer to our answer to the editor on pp. 3-4 <!-- TODO: Review --> of this letter for more details on this addition.
+Please refer to our answer to the editor on pp. 2-3 of this letter for more details on this addition.
 
@@
 This results in a basis (and an associated coordinate system) that does not have a direct interpretation, save for being "the inverse transpose" of the basis where we argue the items are faithfully represented.
+
 However, please bear in mind that this transformation in the latent space is a reasoning artifact to obtain the proper expression of the item parameters in the general case; for further explanation about this, please see our response to the editor in pp.
-<!-- TODO: Complete (5th response) --> lines  <!-- TODO: Complete --> of this letter.
-We have tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p.
-14, lines 15-22, or p.
-<!-- TODO: Complete -->, lines <!-- TODO: Complete --> in the new version).
+4-5 of this letter.
+We have tried to convey this when explaining the rationale behind the covariance-based version of the parameters (see p. 14, lines 15-22, or p. 15, line 22, to p. 16, line 2, in the new version).
 
@@
 Also, we have made changes and additions to
-the third  <!-- TODO: Review --> paragraph of our Discussion, which now reads:
+the third paragraph of our Discussion, which now reads (p. 31, line 22, to p. 32, line 1):
 
@@
 However, we acknowledge that such a plot is the most common representation of the values realized by different cases (persons, in our case) in two variables, correlated or not.
-In this regard, we would kindly refer you to our third <!-- TODO: Expressed ok?
---> response to your comments, on p.
-<!-- TODO: Complete --> of this letter, where we discuss the representation of person parameters in the last two paragraphs.
-<!-- TODO: Possibly complete with references to other responses? -->
+In this regard, we would kindly refer Reviewer 3 to our response on pp. 15-16 of this letter, where we discuss the representation of person parameters in the last two paragraphs.
 
@@
 7 line 8 in our original submission, or p.
-<!-- TODO: Complete --> line <!-- TODO: Complete --> in the new version), and is independent of the basis chosen to represent the items.
+8, line 12 in the new version), and is independent of the basis chosen to represent the items.
 This basis does not have to be necessarily a geometric representation of the covariance structure, as we argue in our introduction, so it can be the canonical basis, or any other, orthonormal or not, and not necessarily representing the covariance structure.
@@
 It is our understanding that,  due to a lack of need for this distinction, this is overlooked in Reckase's [-@reckase_difficulty_1985] and Reckase & McKinley's [-@reckase_discriminating_1991] original formulations.
-We have added this latter clarification in our new version (p. <!-- TODO: Complete --> lines <!-- TODO: Complete -->):
+We have added this latter clarification in our new version (p. 34, lines 13-16):
 
@@
 
-As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response on p.
-<!-- TODO: Complete --> in this letter).
+As we mention above, when the attributes are correlated, a distance measure that keeps the norm invariant in the latent space is the Mahalanobis distance (please see our response on pp.
+16-17 in this letter).
 In this respect, we should make the distinction between representing the attributes being measured (i.e., the latent traits) and the variables measuring those attributes (in our case, the items).
@@
 Therefore, even when computing the coordinates in a non-orthonormal basis, we need to transform them to the canonical one to plot them.
-This is explained in our manuscript, on p. 19 (lines 9-14) in the original version, and on p. <!-- # TODO: Complete --> (lines <!-- # TODO: Complete -->) in the new submission.
+This is explained in our manuscript, on p. 19 (lines 9-14) in the original version, and on p. 20 (last paragraph) in the new submission.
 We are unsure about how to make this clearer, we are afraid, so we would appreciate any suggestions.
@@
 To clarify this point better, we have tried to better express this distinction, emphasizing that, in this case, the unidimensionality induced by the rank-limited covariance matrix restricts the dimensional sensitivity of the test space.
-The aforementioned paragraph now reads (p. <!-- # TODO: Complete -->, lines <!-- # TODO: Complete -->):
+The aforementioned paragraph now reads (p. 29, lines 2-10):
 
@@
 In this case, the basis has been chosen to be a geometric representation of the correlation matrix (as given by the expression $`r INNER_PROD_CORR_COND`$, please see p.
-<!-- # TODO: Complete -->, line <!-- # TODO: Complete -->, in the new version), and therefore, the scales are not affected (only the relative directions among the axes).
+20, last two paragraphs, in the new version), and therefore, the scales are not affected (only the relative directions among the axes).
 
@@
 
-We have found this comment puzzling as, to our understanding, what Reviewer 3 is asking is what we do in Figure 1 (p. 22 in our original submission, p. <!-- # TODO: Complete --> in the new one).
+We have found this comment puzzling as, to our understanding, what Reviewer 3 is asking is what we do in Figure 1 (p. 22 in our original submission, p. 23 in the new one).
 We have shown the same items, with as many different cases as possible, represented in both orthogonal and nonorthogonal coordinates, and explained the interpretation of the multidimensional parameters in both spaces.
@@
 
-We acknowledge, though, that this may have been explained without enough clarity, and therefore striven to improve this discussion.
-The aforementioned paragraph now reads:
+We acknowledge, though, that this may have been explained without enough clarity, and therefore we have striven to improve this discussion.
+The aforementioned paragraph now reads (p. 33, lines 16-19):
 

### 2025-06-03 output/ESM_1.Rmd

@@
 ---
-title: "Online Resource 1"
 output:
@@
 
+# Online Resource 1
+
 Table S.\@ref(tab:empirical-table-Reckase2009-out) is similar to Table 2,

### 2025-07-03 output/seminar-slideshow.qmd
Update title slide and translate ton EN
Vertical alignment of logos pending

@@
 title:     |
-  Parámetros multidimensionales en Teoría de Respuesta al Ítem:
-  ¡Algo estamos haciendo mal!
+  A generalization of multidimensional item response theory parameters
 author:    |
@@
       <td>Daniel Morillo, Ph.D.</td>
+      <td>Mario Luzardo, Ph.D.</td>
     </tr>
@@
       </td>
+      <td>
+        <center>
+          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/mluzardov/)&emsp;
+          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0002-9360-2806)
+        </center>
+      </td>
     </tr>
@@
   <table>
+    <!-- # TODO: Fix vertical alignment in logos-->
     <tr>
-      <td>
-        <center>
-          [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="120"}](https://www.uned.es/universidad/facultades/psicologia.html)
-        </center>
+      <td style="width: 400px">
+        [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="60"}](https://www.uned.es/universidad/facultades/psicologia.html)
+      </td>
+      <td style="width: 100px">
+        [![](../www/slideshow-assets/logo_udelar.svg){height="120"}](https://udelar.edu.uy)
       </td>
@@
 csl:           ../www/apa-old-doi-prefix.csl
-date:          "2024-10-21"
-date-meta:     "2024-11-27" # TODO: Update
+date:          "2024-07-17"
+date-meta:     "2024-11-27"
 date-format:   long
@@
   cache:       true
-lang:          es
+lang:          en
 knitr:
@@
 
+# Factor axis colors:
+F1_COLOR <- unname(PALETTE_BLUE["normal"])
+F2_COLOR <- unname(PALETTE_RASPBERRY["normal"])
+
 GRAPH_FONT   <- "Arial"  # Replace default font to comply with UNED identity
@@
 ```
-
-::: notes
--   Nombre de la presentación: Cero atractivo
-
--   Nuria me pide un título más llamativo
-
--   Lo mejor que puedo conseguir
-
--   ¿Por qué es relevante?
-
--   TRI muy usada en evaluación e investigación educativa
-
--   Poco en investigación psicológica (muestras "pequeñas" en comparación)
-
--   Pero hay un "paralelismo" entre ambas teorías (isomórficas, ante determinados supuestos)
-:::
-
-<!-- Slides "Diferencias AF - TRI" -->
-
-```{r irt-fa-table}
-af_tri_differences <- tribble(
-  ~AF,                            ~TRI,
-  "Relaciones entre variables",   "Comparación entre participantes",
-  "Cuestionarios",                "Pruebas (rendimiento)",
-  "Explicación",                  "Medición",
-  "Puntuación compuesta",         "Puntuación latente",
-  "Estructura simple",            "Diferencias individuales"
-)
-
-section_title <- "## Diferencias AF - TRI"
-
-for (slide in af_tri_differences |> nrow() |> seq_len()) {
-
-  cat( "\n", section_title, sep = '')
-
-  cat( "\n", "<br>", sep = '')
-
-  af_tri_differences |> slice_head(n = slide) |> kable() |> print()
-}
-```
-
-## AF: Estructura simple
-
-<br>
-
-```{r simple-complex-structure}
-ITEM_BASE_LOADING <- 1
-ITEM_DIST         <- 0.2
-
-# Generate data for simple and complex factorial structures
-simple_item_loadings <- tribble(
-  ~item, ~factor, ~loading,
-  'I1',  'F1',    ITEM_BASE_LOADING,
-  'I2',  'F1',    ITEM_BASE_LOADING + ITEM_DIST,
-  'I3',  'F1',    ITEM_BASE_LOADING + 3 * ITEM_DIST,
-  'I4',  'F2',    ITEM_BASE_LOADING - ITEM_DIST,
-  'I5',  'F2',    ITEM_BASE_LOADING + ITEM_DIST,
-  'I6',  'F2',    ITEM_BASE_LOADING + 2 * ITEM_DIST
-)
-complex_item_loadings <- simple_item_loadings |> bind_rows(
-  tibble(
-    item    = 'I3',
-    factor  = 'F2',
-    loading = ITEM_BASE_LOADING - 2 * ITEM_DIST # Secondary loading smaller
-  )
-)
-
-simple_item_loadings <- simple_item_loadings |>
-  complete(item, factor, fill = list(loading = 0))
-
-complex_item_loadings <- complex_item_loadings |>
-  complete(item, factor, fill = list(loading = 0))
-```
-
-::::: columns
-::: column
-```{r path-simple-structure}
-# TODO: Syntax from tibble
-simple_spec <- "
-  F1 =~ I1 + I2 + I3
-  F2 =~ I4 + I5 + I6
-
-  F1 ~~ F2
-"
-
-F1_COLOR <- unname(PALETTE_BLUE["normal"])
-F2_COLOR <- unname(PALETTE_RASPBERRY["normal"])
-
-item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
-factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
-
-simple_spec      |>
-  semPlotModel() |>
-  semPaths(
-    border.color = c(
-      item_colors,       # Item colors
-      F1_COLOR, F2_COLOR # Factor colors
-    ),
-    edge.color   = c(
-      item_colors, # Loading colors
-      "black"      # Correlation color
-    ),
-    residuals    = FALSE,
-    thresholds   = FALSE,
-    intercepts   = FALSE,
-    fixedStyle   = 1,
-    curve        = 3,
-    label.cex    = 1.6,
-    sizeMan      = 10,
-    sizeLat      = 14,
-    border.width = 3,
-    edge.width   = 3,
-    mar          = MARGINS
-  )
-```
-:::
-
-::: {.column .fragment}
-```{r scatter-simple-structure}
-LIM_INF <- -1.5
-LIM_SUP <-  2
-
-axis_lims <- c(LIM_INF, LIM_SUP)
-lim_range <- LIM_SUP - LIM_INF
-axis_pos  <- (-LIM_INF) / lim_range
-
-item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
-
-simple_item_loadings_wide <- simple_item_loadings |>
-  group_by(item)                                  |>
-  arrange(desc(loading))                          |>
-  mutate(main_factor = first(factor))             |>
-  pivot_wider(names_from = factor, values_from = loading)
-
-simple_item_loadings_wide |>
-  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
-  geom_point(size = I(4))                                          +
-  geom_text(
-    nudge_x =  .1,
-    nudge_y = -.08,
-    color   = I("black"),
-    size    = I(5)
-  )                                                                +
-  scale_x_continuous(
-    limits   = axis_lims,
-    breaks   = 0,
-    labels   = NULL,
-    name     = "F2",
-    position = "top"
-  )                                                                +
-  scale_y_continuous(
-    limits   = axis_lims,
-    breaks   = 0,
-    labels   = NULL,
-    name     = "F1",
-    position = "right"
-  )                                                                +
-  scale_color_manual(values = factor_colors, guide = NULL)         +
-  coord_fixed(expand = FALSE, clip = "on")                         +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    ),
-    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
-    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
-  )
-```
-:::
-:::::
-
-::: notes
-Para darle más sentido al modelo factorial es común buscar la "estructura simple".
-Es decir, asumimos que la varianza de cada ítem está explicada solamente por un factor latente.
-Los pesos factoriales de un ítem en los demás factores son nulos.
-
-Si representamos los pesos factoriales como coordenadas, el ítem estará alineado con el eje correspondiente al factor en el que pesa.
-:::
-
-## AF: Estructura simple
-
-::: r-center
-```{r simple-structure-rotated}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-AXIS1_DEGREES <- -25
-AXIS2_DEGREES <-  30
-
-axis1_radians <- AXIS1_DEGREES |> as_units("degrees") |> set_units("radians")
-axis2_radians <- AXIS2_DEGREES |> as_units("degrees") |> set_units("radians")
-
-rot_matrix <- c(
-  cos(axis1_radians), cos(axis2_radians),
-  sin(axis1_radians), sin(axis2_radians)
-) |>
-  matrix(nrow = 2, byrow = TRUE)
-
-simple_item_loadings_rot <- simple_item_loadings |>
-  group_by(item) |>
-  mutate(
-    loading = rot_matrix %*% loading |> drop()
-  ) |>
-  left_join(simple_item_loadings_wide |> select(item, main_factor), by = "item")
-
-simple_item_loadings_rot_plot <- simple_item_loadings_rot |>
-  group_by(item)                                          |>
-  pivot_wider(names_from = factor, values_from = loading) |>
-  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
-  geom_text(
-    nudge_x = -.12,
-    color   = I("black"),
-    size    = I(5)
-  )                                                                +
-  scale_x_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "",
-  )                                                                +
-  scale_y_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = ""
-  )                                                                +
-  scale_color_manual(values = factor_colors, guide = NULL)         +
-  coord_fixed(expand = FALSE, clip = "on")                         +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    )
-  )
-
-simple_item_loadings_rot_plot + geom_point(size = I(4))
-```
-:::
-
-::: notes
-Ahora bien, la mayoría de procedimientos factoriales, exploratorios al menos, arrojan soluciones de este tipo: un primer factor que maximiza toda la varianza posible, el segundo (ortogonal al primeroo) maximiza la varianza restante, y si hay más, hacen lo mismo, dando ejes siempre ortogonales a los anteriores.
-:::
-
-## AF: Estructura simple
-
-<!-- # TODO: Añadir ángulos? -->
-
-::: r-center
-```{r simple-structure-rotated-axes}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-rotated_axes <- tibble(
-  slope  = c(axis1_radians, axis2_radians) |> tan() |> drop_units(),
-  factor = c('F1', 'F2')
-)
-
-simple_item_loadings_rot_plot +
-  geom_abline(
-    data    = rotated_axes,
-    mapping = aes(slope = slope, intercept = 0),
-    color   = c(PALETTE_BLUE["light"], PALETTE_RASPBERRY["light"]),
-    linewidth = LINE_WIDTH
-  ) +
-  geom_point(size = I(4)) +
-  annotate(
-    "text",
-    x = -1.35, y = c(-.62, .78),
-    label = rotated_axes |> pull(factor),
-    size  = 9
-  )
-```
-:::
-
-::: notes
-Para obtener esa estructura simple, muchas veces necesitamos usar una rotación oblicua de los ejes.
-Cualquier procedimiento de rotación busca alinearse lo más posible con al menos algunos de los ítems, como vemos aquí.
-Pero esto es una situación ideal.
-Nunca van a quedar, en un modelo exploratorio, los ítems perfectamente alineados con un eje, dando cero exacto en los pesos de los factores que no le corresponden.
-En un modelo confirmatorio sí, forzamos que sea así, pero a costa de perder ajuste claro (y normalmente permitiendo también que los factores correlacionen).
-:::
-
-## AF: Estructura compleja
-
-<br>
-
-::::: columns
-::: column
-```{r path-complex-structure}
-# TODO: Syntax from tibble
-complex_spec <- "
-  F1 =~ I1 + I2 + I3
-  F2 =~ I3 + I4 + I5 + I6
-
-  F1 ~~ F2
-"
-
-edge_colors <- c(
-  F1_COLOR |> rep(3),
-  PALETTE_RASPBERRY["light"], # Cross-loading
-  F2_COLOR |> rep(3)
-)
-
-complex_spec     |>
-  semPlotModel() |>
-  semPaths(
-    border.color = c(
-      item_colors,       # Item colors
-      F1_COLOR, F2_COLOR # Factor colors
-    ),
-    edge.color   = c(
-      edge_colors, # Loading colors
-      "black"      # Correlation color
-    ),
-    residuals    = FALSE,
-    thresholds   = FALSE,
-    intercepts   = FALSE,
-    fixedStyle   = 1,
-    curve        = 3,
-    label.cex    = 1.6,
-    sizeMan      = 10,
-    sizeLat      = 14,
-    border.width = 3,
-    edge.width   = 3,
-    mar          = MARGINS
-  )
-```
-:::
-
-::: {.column .fragment}
-```{r scatter-complex-structure}
-LIM_INF <- -1.5
-LIM_SUP <-  2
-
-axis_lims <- c(LIM_INF, LIM_SUP)
-lim_range <- LIM_SUP - LIM_INF
-axis_pos  <- (-LIM_INF) / lim_range
-
-item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
-
-complex_item_loadings_wide <- complex_item_loadings |>
-  group_by(item)                                    |>
-  arrange(desc(loading))                            |>
-  mutate(main_factor = first(factor))               |>
-  ungroup()                                         |>
-  pivot_wider(names_from = factor, values_from = loading)
-
-item3_spikes <- complex_item_loadings_wide      |>
-  filter(item == 'I3')                          |>
-  select(starts_with('F'))                      |>
-  add_column(type = "end")                      |> 
-  bind_rows(tibble(F1 = 0, F2 = 0))             |>
-  complete(F1, F2, fill = list(type = "start")) |>
-  filter(F1 != 0 | F2 != 0)
-
-item3_spikes <- item3_spikes                           |> 
-  bind_rows(item3_spikes |> slice(3))                  |>
-  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
-  pivot_wider(names_from = type, values_from = F1:F2)
-
-spike_colors = c(
-  'F1' = unname(PALETTE_BLUE['light']),
-  'F2' = unname(PALETTE_RASPBERRY['light'])
-)
-
-complex_item_loadings_wide |>
-  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
-  geom_segment(
-    data    = item3_spikes,
-    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = spike_colors
-  )                                                                +
-  geom_point(size = I(4))                                          +
-  geom_text(
-    nudge_x =  .1,
-    nudge_y = -.08,
-    color   = I("black"),
-    size    = I(5)
-  )                                                                +
-  scale_x_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "F2",
-    position = "top"
-  )                                                                +
-  scale_y_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "F1",
-    position = "right"
-  )                                                                +
-  scale_color_manual(values = factor_colors, guide = NULL)         +
-  coord_fixed(expand = FALSE, clip = "on")                         +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    ),
-    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
-    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
-  )
-```
-:::
-:::::
-
-::: notes
-Sin embargo, en ocasiones no es posible encontrar la estructura simple.
-Es cuando aparecen lo que llamamos "pesos cruzados".
-
-Esto da lugar a una "estructura compleja", que en realidad es más común que la estructura simple, en general.
-:::
-
-## AF: Estructura compleja
-
-<br>
-
-::::: columns
-::: column
-```{r path-method-factor}
-# TODO: Syntax from tibble
-method_spec <- "
-  F1 =~ I1 + I2 + I3 + I4 + I5 + I6
-  M1 =~ I4 + I5 + I6
-"
-
-edge_colors <- c(
-  F1_COLOR |> rep(6),
-  F2_COLOR |> rep(3)
-)
-
-method_spec      |>
-  semPlotModel() |>
-  semPaths(
-    border.color = c(
-      F1_COLOR |> rep(6), # Item colors
-      F1_COLOR, F2_COLOR  # Factor colors
-    ),
-    edge.color   = edge_colors, # Loading colors
-    residuals    = FALSE,
-    thresholds   = FALSE,
-    intercepts   = FALSE,
-    fixedStyle   = 1,
-    curve        = 3,
-    label.cex    = 1.6,
-    sizeMan      = 10,
-    sizeLat      = 14,
-    border.width = 3,
-    edge.width   = 3,
-    mar          = MARGINS
-  )
-```
-:::
-
-::: column
-```{r scatter-method-factor}
-method_item_loadings <- tribble(
-  ~item, ~factor, ~loading,
-  'I1',  'F1',     ITEM_BASE_LOADING,
-  'I1',  'M1',     0,
-  'I2',  'F1',     ITEM_BASE_LOADING + 2 * ITEM_DIST,
-  'I2',  'M1',     0,
-  'I3',  'F1',     ITEM_BASE_LOADING + 3 * ITEM_DIST,
-  'I3',  'M1',     0,
-  'I4',  'F1',    -ITEM_BASE_LOADING,
-  'I4',  'M1',     ITEM_BASE_LOADING - 2 * ITEM_DIST,
-  'I5',  'F1',     ITEM_BASE_LOADING + ITEM_DIST,
-  'I5',  'M1',    -ITEM_BASE_LOADING + 2 * ITEM_DIST,
-  'I6',  'F1',    -ITEM_BASE_LOADING - ITEM_DIST,
-  'I6',  'M1',    -ITEM_BASE_LOADING + 3 * ITEM_DIST
-)
-
-method_item_loadings                                      |>
-  pivot_wider(names_from = factor, values_from = loading) |>
-  ggplot(mapping = aes(F1, M1, label = item))              +
-  geom_point(size = I(4), color = F1_COLOR)                +
-  geom_text(
-    nudge_y = -.12,
-    color   = I("black"),
-    size    = I(5)
-  )                                                        +
-  scale_x_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "M1",
-    position = "top"
-  )                                                        +
-  scale_y_continuous(
-    limits = axis_lims,
-    breaks = 0,
-    labels = NULL,
-    name   = "F1",
-    position = "right"
-  )                                                        +
-  scale_color_manual(values = factor_colors, guide = NULL) +
-  coord_fixed(expand = FALSE, clip = "on")                 +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_text(hjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    ),
-    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
-    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
-  )
-```
-:::
-:::::
-
-::: notes
-Por ejemplo, si tenemos ítems directos e inversos, podemos modelar lo que llamamos "factores de método": Una variable latente para cada factor sustantivo (sólo uno en este caso), y por ejemplo uno para los ítems formulados en negativo.
-:::
-
-## Modelo factorial clásico
-
-<br>
-
-$$
-Y_1 = \tau_1 + \lambda_{11} F_1^p + \lambda_{21} F_2^p + ... + \lambda_{n1} F_n^p + U_1^p \\
-\vdots                                                                            \\
-Y_i = \tau_i + \lambda_{1i} F_1^p + \lambda_{2i} F_2^p + ... + \lambda_{ni} F_n^p + U_i^p
-$$
-
-<br>
-
-::: fragment
-$$
-Y_i = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
-$$
-:::
-
-::: notes
-Un modelo factorial no es más que una serie de regresiones lineales (una por cada variable observable) con variables latentes.
-Las variables "respuesta" o "criterio" es la variable observable, y su varianza está explicada o "predicha" por las variables latentes, que son los factores y los términos de error de cada ítem.
-Lo que ocurre es que aquí a "coeficientes de regresión" los llamamos "pesos factoriales" y al término error, "unicidad" o "varianza única".
-
-El producto de los pesos por las puntuaciones factoriales se puede expresar en "forma vectorial", de esta manera.
-Este término significa que el vector de puntuaciones factoriales está pre-multiplicado por el vector de pesos factoriales.
-Esta operación es lo que se llama un "producto escalar", que consiste simplemente en multiplicar dos a dos los términos de cada vector y sumar todos.
-:::
-
-## Modelo factorial clásico {.smaller}
-
-::::::: columns
-::::: column
-$$
-  Y_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
-$$
-
-$$
-  \mathbf{F}^1 =
-    \begin{bmatrix}
-      0.5 \\
-      1
-    \end{bmatrix}
-$$
-
-::: {.fragment fragment-index="1"}
-$$
-  \mathbf{\lambda}_{1} =
-    \begin{bmatrix}
-      -1 \\
-       0.5
-    \end{bmatrix}
-$$
-:::
-
-::: fragment
-$$
-  \begin{align}
-    \tau_1 &=  1 \\
-    U_1^1  &= -0.5
-  \end{align}
-$$
-:::
-:::::
-
-::: {.column .fragment fragment-index="2"}
-```{r fa_example_plot}
-#| cache: false
-AXIS_GRID <- -1:1
-
-# Generate data for simple and complex factorial structures
-fa_example <- tribble(
-  ~object, ~factor, ~loading,
-  'I~1~',  'F1',    -1,
-  'I~1~',  'F2',     0.5,
-  'F^1^',  'F1',     0.5,
-  'F^1^',  'F2',     1
-)
-
-fa_example_wide <- fa_example |>
-  pivot_wider(names_from = factor, values_from = loading)
-
-fa_example_spikes <- fa_example_wide |>
-  add_column(type = "end")           |> 
-  bind_rows(
-    tribble(
-      ~object, ~F1,  ~F2, ~type,
-      "I~1~",   0,   0.5, "start",
-      "F^1^",   0,   1,   "start",
-      "I~1~",  -1,   0,   "start",
-      "F^1^",   0.5, 0,   "start",
-    )
-  )
-
-fa_example_spikes <- fa_example_spikes |> slice(1:2)        |>
-  bind_rows(fa_example_spikes)                              |>
-  add_column(factor = paste0('F', 1:2 |> rep(2, each = 2))) |>
-  pivot_wider(names_from = type, values_from = F1:F2)
-
-fa_example_wide |>
-  ggplot(mapping = aes(F1, F2, label = object))             +
-  geom_hline(
-    yintercept = 0,
-    color      = PALETTE_BLUE["light"],
-    linewidth  = GRID_WIDTH
-  )                                                         +
-  geom_vline(
-    xintercept = 0,
-    color      = PALETTE_RASPBERRY["light"],
-    linewidth  = GRID_WIDTH
-  )                                                         +
-  geom_segment(
-    data        = fa_example_spikes,
-    mapping     = aes(
-      F1_start, F2_start,
-      xend  = F1_end, yend = F2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth   = LINE_WIDTH,
-    linetype    = '43',
-    color       = spike_colors |> rep(each = 2)
-  )                                                        +
-  geom_point(size = I(4))                                  +
-  geom_richtext(
-    nudge_x = .15,
-    nudge_y = .10,
-    color   = I("black"),
-    size    = I(5),
-    lineheight = 1.2,
-    fill = NA,
-    label.color = NA
-  )                                                        +
-  scale_x_continuous(
-    limits       = axis_lims,
-    breaks       = AXIS_GRID,
-    name         = NULL,
-    minor_breaks = NULL,
-    sec.axis     = dup_axis(name = "F~2~", labels = NULL)
-  )                                                        +
-  scale_y_continuous(
-    limits       = axis_lims,
-    breaks       = AXIS_GRID,
-    name         = NULL,
-    minor_breaks = NULL,
-    sec.axis     = dup_axis(name = "F~1~", labels = NULL)
-  )                                                        +
-  scale_color_manual(values = factor_colors, guide = NULL) +
-  coord_fixed(expand = FALSE, clip = "on")                 +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    axis.title.x       = element_markdown(hjust = axis_pos),
-    axis.title.y       = element_markdown(vjust = axis_pos),
-    axis.title.y.right = element_text(vjust = axis_pos),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    )
-  )
-```
-:::
-:::::::
-
-::: fragment
-$$
-  Y_1 = 1 + \left[ -1, 0.5 \right]
-    \begin{bmatrix}
-      0.5 \\
-      1
-    \end{bmatrix}
-    - 0.5 = 1 + (-0.5 + 0.5) - 0.5 = 0.5
-$$
-:::
-
-::: notes
-Veámoslo con un ejemplo.
-Tenemos el participante 1, en nuestro estudio, que tiene esas puntuaciones factoriales; usamos el superíndice para denotar al participante.
-Por ejemplo, en el primer rasgo, tiene un 0.5 (nos da igual la distribución pero supongamos que está estandarizada, luego está media desviación típica por encima de la media).
-En el segundo rasgo, tiene una puntuación que está una desviación típica por encima de la media.
-
-Los pesos factoriales del ítem 1 son estos: -1 para el primer rasgo (por lo cual, cuanto mayor sea la puntuación factorial en el primer factor, menor va a ser la respuesta observada en el ítem 1), y 0.5 para el segundo.
-
-Aquí vemos representados el ítem y la persona, en las mismas coordenadas cartesianas.
-Y luego el ítem tiene una intersección de 1, y la persona 1 tiene un -0.5 en el "error" de esa variable $I_1$.
-
-\[NEXT\]
-
-Su respuesta se calcularía así: Primero tenemos el parámetro de intersección que vale 1.
-La "T" del vector de pesos factoriales significa que está "traspuesto", es decir, que se pone horizontalmente (esto es una convención en álgebra matricial).
-El producto del vector de pesos por el de puntuaciones factoriales es (...).
-Y por último sumamos la puntuación de error de ese participante en la variable.
-Luego esta persona (el particpante 1) tiene la respuesta o valor observado 0.5 en esta variable observada.
-
-Lo mismo podríamos hacer con el resto de variables y de participantes, pero obviamente nosotros lo que tenemos son estas respuestas observadas, y lo que hacemos con el AF es estimar todos estos parámetros del modelo, en lugar de como lo hemos hecho ahora.
-
-Ahora bien, esta respuesta observada que hemos calculado aquí es cuantitativa.
-Tenemos costumbre de ver el análisis factorial en ese contexto, de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
-Sin embargo, el llamado "AF de ítems" se aplica también a variables discretas ordinales (e idealmente se debería aplicar así a ítems tipo Likert, aunque a veces no se haga, por simplificar).
-:::
-
-## Análisis factorial de ítems {.smaller}
-
-::::::::: columns
-::::: {.column width="45%"}
-$$
-  y_i^* = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
-$$
-
-<br>
-
-::: {.fragment fragment-index="1"}
-$$
-  Y_i = \left\{ \begin{matrix}
-      1 &si &y_i^* \geq 0\\
-      0 &si &y_i^* <    0
-    \end{matrix} \right.
-$$
-:::
-
-<br>
-
-::: {.fragment fragment-index="2"}
-$$
-  \begin{align}
-    P(Y_i = 1) &= \int_0^{-\infty} \mathcal{N}(y_i^*) \delta y_i^* \\
-               &= \Phi_N \left(
-                 \frac{ \tau_i + \mathbf{\lambda}_i^T \mathbf{F} }
-                   { \sqrt{var(U_i)} }
-               \right)
-  \end{align}
-$$
-:::
-:::::
-
-::::: {.column width="55%"}
-::: {.fragment fragment-index="1"}
-```{r ifa-density}
-#| fig-asp:   0.5
-#| fig-width: 8
-
-PROB_AXIS      <- seq(-3.3, 3.3, by = .01)
-ERROR_POINTS   <- seq(-3, 3, length.out = 5)
-ITEM_LOADING   <-  0.45
-ITEM_INTERCEPT <- -0.3
-
-item_unique_var <-  1 - sqrt(ITEM_LOADING)
-
-latent_response <- tibble(
-  point    = ERROR_POINTS |> rep(500000),
-  response = rnorm(
-    length(point),
-    mean = ITEM_INTERCEPT + ITEM_LOADING * point,
-    sd   = sqrt(item_unique_var)
-  )
-)
-
-# Plot: Upper Panel - Linear regression with Gaussian error distributions
-latent_response |>
-  ggplot(
-    aes(x = response, y = point, fill = after_stat(x) > 0, group = point)
-  )                            +
-  geom_density_ridges_gradient(
-    scale          =  .2,
-    rel_min_height = 0.001,
-    color          = PALETTE_BLUE["normal"]
-  )                              +
-  labs(x = "y~i~^*^", y = NULL)  +
-  scale_y_continuous(
-    breaks       = ERROR_POINTS,
-    minor_breaks = NULL,
-    labels       = NULL,
-    limits       = range(PROB_AXIS),
-    expand       = expansion()
-  )           +
-  scale_x_continuous(breaks = 0) +
-  scale_fill_manual(
-    values = c(
-      `FALSE` = unname(PALETTE_BLUE["white"]),
-      `TRUE`  = unname(PALETTE_BLUE["normal"])
-    )
-  )                              +
-  theme(
-    axis.ticks.x = element_blank(),
-    axis.title.y = element_markdown(angle = 90),
-    legend.position = "none",
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )                              +
-  coord_flip()
-```
-:::
-
-::: {.fragment fragment-index="2"}
-```{r ifa-curve}
-#| fig-asp:   0.5
-#| fig-width: 8
-
-probit_curve_data <- tibble(
-  score    = PROB_AXIS,
-  response = ITEM_INTERCEPT + ITEM_LOADING * score,
-  prob     = pnorm(response, sd = sqrt(item_unique_var))
-)
-
-# Plot: Lower Panel - Probit curve
-probit_curve_data |>
-  ggplot(mapping = aes(x = score, y = prob))                        +
-  geom_line(color = PALETTE_BLUE["normal"], linewidth = LINE_WIDTH) +
-  labs(x = "E(*y~i~^*^*)", y = "P(Y~i~ = 1)")                       +
-  scale_y_continuous(
-    breaks       = 0:1,
-    minor_breaks = NULL,
-    limits       = 0:1,
-    expand       = expansion(mult = c(0, .05)),
-    oob = oob_keep
-  )  +
-  scale_x_continuous(
-    breaks       = ERROR_POINTS,
-    minor_breaks = NULL,
-    limits = range(PROB_AXIS),
-    expand = expansion()
-  ) +
-  theme(
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
-  )
-```
-:::
-:::::
-:::::::::
-
-::: notes
-Lo que hace el análisis factorial de ítems es aplicar ese "modelo lineal" a variables ordinales, haciendo una generalización.
-(En este caso vamos a simplificarlo a variables dicotómicas; p.ej. ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
-Esto es una simplificación que rara vez se utiliza en cuestionarios no cognitivos, pero es interesante estudiarlo porque otros modelos pueden considerarse extensiones de este modelo.
-El AF de ítems asume una nueva "variable latente" lineal ($I_1^*$), con la misma fórmula anterior
-
-Después se aplica una "función de umbral": La respuesta es correcta o afirmativa, codificada como 1, si la variable latente es mayor (o igual) que 0, o incorrecta o negativa, codificada como 0, si es menor.
-Eso lo que da, en lugar de un respuesta observada (cuantitativa), es una "probabilidad" de responder correcta o "afirmativamente" a un ítem.
-:::
-
-## AF: Puntuaciones factoriales
-
-<br>
-
-:::::: columns
-:::: column
-Métodos:
-
--   Regresión
-
--   Bartlett
-
-::: {.fragment fragment-index="2"}
--   **Máxima Verosimilitud**
-
--   **Empirical Modal Bayes**
-:::
-::::
-
-::: {.column .fragment fragment-index="1"}
-```{r path-regression-scores}
-# item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
-# factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
-
-# NOTE: Manual call to qgraph:
-edge_list <- tribble(
-  ~rhs, ~lhs,
-  1,    7,
-  2,    7,
-  3,    7,
-  4,    8,
-  5,    8,
-  6,    8,
-  7,    8,
-  8,    7
-)
-vSize <- c(10 |> rep(6), 14 |> rep(2))
-
-qgraph(
-  edge_list,
-  labels = c("I1", "I2", "I3", "I4", "I5", "I6", "F1", "F2") |>
-    setNames(nm = _),
-  label.cex = 1.6,
-  shape = "square" |> rep(8),
-  vsize = vSize, 
-  vsize2 = vSize,
-  color = "background" |> rep(8),
-  border.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), F1_COLOR, F2_COLOR),
-  border.width = 3,
-  bidirectional = c(FALSE |> rep(6), TRUE, TRUE),
-  directed = TRUE |> rep(8),
-  curve = c(NA |> rep(6), 3, 3),
-  edge.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), "black"),
-  edge.width = 3,
-  lty = 1,
-  layout = matrix(
-    c(
-      -1, -0.6, -0.2, 0.2, 0.6, 1, -0.6, 0.6,
-      -0.5 |> rep(6), 0.5 |> rep(2)
-    ),
-    ncol = 2
-  ),
-  mar = c(12, 1.3, 8, 1.3)
-)
-```
-:::
-::::::
-
-<!--# TODO: Diagrama de CFA y de "CFA inverso" -->
-
-::: notes
-Una vez tenemos los parámetros del modelo factorial estimado, entre otras cosas, podemos aplicarlo para estimar las puntuaciones factoriales de los participantes.
-Para hacer esto hay varios métodos.
-
-El método de regresión y el de Bartlett se aplican en AF lineal y están implementados en SPSS.
-Consisten a grandes rasgos en calcular la puntuación factorial como una suma ponderada de las variables observables.
-Es decir, algo así como "darle la vuelta a las flechas" de causalidad.
-Lo que tiene más sentido, desde el punto de vista estadístico, es "estimar" qué puntuaciones factoriales hacen más probable para una participante dar las respuestas que ha dado.
-
-Eso es lo que hacen los dos métodos de abajo (SPSS no los implementa creo), Máxima Verosimilitud, o esimación Modal Bayesiana.
-Si habéis hecho esto con variables categóricas, entonces habéis aplicado la Teoría de Respuesta al Ítem, y a continuación veremos por qué.
-:::
-
-## Paralelismo AF - TRI
-
-<br>
-
-::::: columns
-::: column
-![](../www/slideshow-assets/cover_McDonald-1999.jpg){height="300px"}
-
-[@mcdonald_test_1999]
-:::
-
-::: column
-<!--# TODO: Citation by McDonald -->
-:::
-:::::
-
-::: notes
-En este libro, que se publica en 1999 (fijaos en el subtítulo de "unified treatment"), McDonald indica que estos modelos, factoriales, TRI, son a fin de cuentas "isomóficos": Sin ser exactamente iguales ni partir de los mismos supuestos, las predicciones que hacen uno y otro son indistinguibles en la práctica.
-:::
-
-## Teoría de respuesta al ítem {.smaller}
-
-Modelo logístico de 2 parámetros [@birnbaum_latent_1968]:
-
-::::::::::: columns
-:::::: {.column width="60%"}
-```{r irt-curves}
-#| fig-asp:   0.6
-#| fig-width: 8
-
-ITEMS_PALETTE <- c(
-  PALETTE_BLUE["normal"],
-  PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"]
-) |>
-  unname() |>
-  setNames("prob" |> paste(1:3, sep = '_'))
-
-ITEMS <- tribble(
-  ~item, ~a,   ~b,
-   1,     1,    0,
-   3,     1,    1.5,
-   2,     2.5,  0
-)
-
-logit <- function(x) 1 / (1 + exp(-x))
-irf   <- function(a, b, theta) logit(a * (theta - b))
-
-logit_curve_data <- tibble(
-  latent_trait = PROB_AXIS,
-  prob         = ITEMS |>
-    select(-item) |>
-    pmap(irf, theta = latent_trait) |>
-    bind_cols()
-) |>
-  unnest(
-    prob,
-    names_sep = '_'
-  ) |>
-  rename_with(str_remove, pattern = '\\.{3}')
-
-logit_curve_data_long <- logit_curve_data |>
-  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
-
-logit_curve_data_long |>
-  ggplot(
-    mapping = aes(
-      latent_trait, prob,
-      color = item,
-      group = after_scale(color)
-    )
-  )                                        +
-  geom_line(linewidth = LINE_WIDTH)        +
-  labs(x = "*&theta;*", y = "P(Y~i~ = 1)") +
-  scale_y_continuous(
-    breaks       = 0:1,
-    minor_breaks = NULL,
-    limits       = 0:1,
-    expand       = expansion(mult = c(0, .05))
-  )                                        +
-  scale_x_continuous(
-    breaks       = ERROR_POINTS,
-    minor_breaks = NULL,
-    limits = range(PROB_AXIS),
-    expand = expansion()
-  )                                        +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  theme(
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
-  )
-```
-
-<!-- TODO: Como ecuaciones (sin punto y coma?) -->
-
-::: r-stack
-[$a_1$ = 1.0; $b_1$ = 0.0]{style="color:#5c6eb1;"}
-:::
-
-::: r-stack
-[$a_2$ = 1.0; $b_2$ = 1.5]{style="color:#90214a;"}
-:::
-
-::: r-stack
-[$a_3$ = 2.5; $b_3$ = 0.0]{style="color:#749f4c;"}
-:::
-::::::
-
-:::::: {.column width="40%"}
-::: {.fragment fragment-index="1"}
-$$
-  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
-$$
-:::
-
-<br>
-
-::: {.fragment fragment-index="2"}
-$$
-  \begin{align}
-    d_i &= -a_i b_i \\
-    P(Y_i = 1) &= \Phi_L (a_i \theta + d_i))
-  \end{align}
-$$
-:::
-
-<br>
-
-::: {.fragment fragment-index="3"}
-$$
-  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
-$$
-:::
-
-<br>
-::::::
-:::::::::::
-
-::: notes
-La TRI, por su lado, asume que hay cada persona tiene una puntuación latente, y que su respuesta depende de esa puntuación, con un parámetro de discriminación que determina cómo de bien o mal mide el ítem, y uno de dificultad (aunque aquí prefiero llamarlo "de posición") que determina lo fácil o difícil que será dar una respuesta positiva (afirmativa, correcta, codificada como 1).
-Esa posición representa el nivel de rasgo latente que da lugar a una probabilidad de .5 de dar la respuesta positiva (en este modelo; en general representa el punto de inflexión de la curva, es decir, donde pasa de hacerse cada vez más pronunciada a empezar a aplanarse).
-
-Ahora bien, la posición se puede reformular como un parámetro de intersección.
-
-Como se puede ver, esto también se trata de un "modelo lineal general".
-En este caso, en lugar de una función probit (normal acumulada), la función de enlace es la función "logit".
-Es decir, la respuesta a cada ítem se modela como una regresión logistica donde los predictores son las variables latentes.
-Esto es lo que se llama "Función Característica del Ítem".
-:::
 
@@
 Pertenece a una familia de modelos llamados "compensatorios" en TRI, ya que un nivel bajo en una puntuación latente se compensa con un nivel alto en otra, dando lugar a la misma probabilidad de respuesta.
-:::
-
-## TRI multidimensional: Ejemplo {.smaller}
-
-:::::: columns
-::: column
-$$
-  P(Y_1 = 1) = \Phi_L (\mathbf{a}_1^T \mathbf{\theta} + d_1)
-$$
-
-```{r mirt-item-response-example}
-mirt_example <- tribble(
-  ~theta_1, ~theta_2,
-  1,        -1
-)
-
-mirt_example_spikes <- mirt_example |>
-  add_column(type = "end")          |> 
-  bind_rows(
-    tribble(
-      ~theta_1, ~theta_2, ~type,
-      0,        -1,        "start",
-      1,         0,        "start"
-    )
-  )
-
-mirt_example_spikes <- mirt_example_spikes |>
-  slice(1) |>
-  bind_rows(mirt_example_spikes) |>
-  add_column(factor = paste0('F', 1:2 |> rep(2))) |>
-  pivot_wider(names_from = type, values_from = starts_with("theta"))
-
-contour_plot +
-  geom_segment(
-    data    = mirt_example_spikes,
-    mapping = aes(
-      theta_1_start, theta_2_start,
-      xend = theta_1_end,
-      yend = theta_2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = PALETTE_BLUE["light"]
-  ) +
-  geom_point(
-    data = mirt_example,
-    mapping = aes(theta_1, theta_2),
-    inherit.aes = FALSE,
-    size = I(4)
-  )
-```
-:::
-
-:::: column
-\begin{matrix}
-  \mathbf{\theta} =
-    \begin{bmatrix}
-      1 \\
-      -1
-    \end{bmatrix}
-
-  &\begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1 \\
-        0.5
-      \end{bmatrix} \\
-    d_1 &= -0.5
-  \end{align}
-\end{matrix}
-
-<br>
-
-::: fragment
-$$
-  \begin{align}
-    P(Y_1 = 1) &= \Phi_L \left([1, 0.5]
-      \begin{bmatrix}
-         1 \\
-        -1
-      \end{bmatrix}
-      -0.5 \right) \\
-      &= \Phi_L \left(1 · 1 + 0.5 · (-1) -0.5 \right) \\
-      &= \Phi_L \left(1 -0.5 -0.5 \right) = \Phi_L \left(0 \right) \\
-      &= \frac{1}{1 + e^{-0}} = \frac{1}{1 + 1} = \frac{1}{2} \\
-      &= 0.5
-  \end{align}
-$$
-:::
-::::
-::::::
-
-::: notes
-Veamos un ejemplo de cuál sería la probabilidad de respuesta de una persona con puntuaciones de rasgo de 1 y -1.
-
-(Hacer cálculos matriciales)
-:::
-
-## AF en métrica TRI {.smaller}
-
-<br>
-
-$$
-  \begin{align}
-    P(Y_i = 1) &= \Phi_N \left(
-      \frac{\mathbf{\lambda}_i^T \mathbf{F} + \tau_i}{\sqrt{var(U_i)}}
-    \right) =
-      \Phi_N \left(
-        \frac{1}{\sqrt{var(U_i)}} \mathbf{\lambda}_i^T \mathbf{F} +
-          \frac{1}{\sqrt{var(U_i)}} \tau_i
-      \right)\\
-    P(Y_i = 1) &= \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
-  \end{align}
-$$
-
-<br>
-
-::: fragment
-@takane_relationship_1987:
-
-$$
-  \begin{align}
-    \mathbf{a}_i &= \frac{1.7}{\sqrt{var(U_i)}} \mathbf{\lambda}_i \\
-    d_i &= \frac{1.7}{\sqrt{var(U_i)}} \tau_i
-  \end{align}
-$$
-:::
-
-::: notes
-Claro, si ponemos las ecuaciones del AF y de TRI lado a lado, es fácil ver que existe un paralelismo entre ambos modelos.
-
-Pues resulta que estas dos curvas, la normal y la logística, son muy parecidas; tanto, que es posible transformar los parámetros de la métrica del AF a métrica de TRI, obteniendo modelos indistinguibles en la práctica.
-Las dos curvas se diferencian por este factor de 1.7, pero aplicándo ese factor y escalando por la desviación estándar de la unicidad del ítem, obtenemos los mismos parámetros en métrica de TRI mulidimensional con función de respuesta logística.
-Y de hecho, esto es necesario y se hace así para hacer la estimación máximo-verosimil de la que hablábamos antes.
 :::

### 2025-07-03 output/seminar-slideshow.qmd
Drop "concluding remarks" slide

@@
 
-## Últimas palabras
-
-<!-- # TODO: Imagen ilustrativa -->
-
-::: notes
-A este respecto, me gustaría terminar recalcando esto último que he mencionado, aplicado no sólo a cuestiones psicométricas, o de TRI.
-Cuando estaba empezando a preparar este seminario, una de mis inquietudes era la complejidad, tanto conceptual como formal de esta investigación.
-Compartiendo estas inquietudes con algunos compañeros, me aconsejaron utilizar digamos este discurso como excusa para lanzar este "mensaje" en general: La idea de que tenemos que ser muy precavidos a la hora de aplicar estos métodos, sean cuales sean (cuantitativos, cualitativos, o del tipo que sean) sin reflexionar sobre si son los más adecuados o no.
-Y bueno, a este respecto comentar que, no sólo yo, sino el departamento de Metodología al completo, hacemos una tipo de investigación que precisamente se encarga de eso, de intentar entender estos métodos, saber si son los más adecuados, y saber cuándo, cómo, y sobre todo, por qué, aplicarlos en la investigación aplicada (aplicada para nosotros, al menos).
-
-Así que bueno, ya lo sabéis, estamos disponibles en el departamento para colaborar en esas investigaciones, si lo consideráis oportuno, para aplicar los métodos más apropiados a los problemas de investigación que tengáis.
-En mi caso en concreto, no sólo puedo proporcionar cierto "expertise" psicométrico.
-También me dedico mucho a cuestiones de ciencia abierta y reproduciblidad, así que si queréis darle a vuestra investigación un giro en ese sentido, para hacerla más sostenible y alineada con los objetivos de los organismos financiadores (y si no queréis, deberíais ir queriendo), podéis contactar conmigo, y podemos plantear líneas de colaboración.
-Eso sí, me gustaría pediros, si necesitáis colaboración, no sólo metodológica, sino de cualquier tipo, por favor no nos llaméis cuando tengáis un lío de datos, y no sepáis bien cómo resolver la papeleta (no digo que sea vuestro caso, pero me he encontrado con muchos casos, que son verdaderas historias de terror).
-Contad por favor con nosotros desde el momento de escribir vuestro proyecto, para solicitar la financiación, porque es en ese punto en el que se pueden prevér problemas y anticiparse a ellos.
-Porque si una investigación no está bien diseñada desde el principio, teniendo en cuenta estos aspectos metodológicos, de reproducibilidad, etc., si tenemos mala suerte pueden surgir problemas que se puedan resolver sólo con muchísimo trabajo, o incluso que ya nunca se puedan resolver, y tengamos que tirar a la basura todo ese esfuerzo de investigación.
-
-Eso es todo, muchas gracias.
-Y espero vuestras preguntas y comentarios.
-:::
-
 ## References {.smaller}

### 2025-07-07 output/IMPS-2025-slideshow.qmd
Creates slideshow version in EN for the IMPS-2025

--- 2025-07-03 output/seminar-slideshow.qmd
@@
 
-## TRI multidimensional
+## Multidimensional IRT
 
@@
 
-## Parámetros multidimensionales de TRI {.smaller}
+## Multidimensional IRT parameters {.smaller}
 
@@
 
-## Dificultad multidimensional {.smaller}
+## Multidimensional difficulty {.smaller}
 
@@
 
-## Dificultad multidimensional
+## Multidimensional difficulty
 
@@
 
-## Discriminación multidimensional {.smaller}
+## Multidimensional discrimination {.smaller}
 
@@
 
-## Discriminación multidimensional
+## Multidimensional discrimination
 
@@
 
-## Parámetros multidimensionales
+## Multidimensional parameters
 
@@
 
-## Representación gráfica
+## Graphical representation
 
@@
 
-## Propiedades: *MDISC* {.smaller}
+## Properties: *MDISC* {.smaller}
 
@@
 
-## Propiedades: *D* {.smaller}
+## Properties: *D* {.smaller}
 
@@
 
-## Propiedades: $\cos \, \mathbf{\alpha}$ {.smaller}
+## Properties: $\cos \, \mathbf{\alpha}$ {.smaller}
 
@@
 
-## Supuestos: Monotonía creciente {.smaller}
+## Assumptions: Increasing monotony {.smaller}
 
@@
 
-## Generalización: Ítems inversos {.smaller}
+## Generalization: Inverse items {.smaller}
 
@@
 #| fig-width:  9
+
+PROB_AXIS <- seq(-3.3, 3.3, by = .01)
 
@@
 
-## Generalización: Ítems inversos
+## Generalization: Inverse items
 
@@
 
-## Ítems "inversos" en la literatura {.smaller}
+## "Inverse" items in the literature {.smaller}
 
@@
 ::: {.fragment .r-center}
-### Generalización: "Monotonía constante"
+### Generalization: "Constant monotony"
 :::
@@
 ::: aside
-Figura tomada de @ackerman_multidimensional_2005-1
+Figure taken from @ackerman_multidimensional_2005-1
 :::
@@
 
-## Supuestos: Ortogonalidad {.smaller}
+## Assumptions: Orthogonality {.smaller}
 
@@
 
-## (Pero, un inciso...) {.smaller}
+## (But, an aside...) {.smaller}
 
@@
 ::: aside
-Figura tomada de [Vecteezy](https://www.vecteezy.com/vector-art/3709640-isometric-cube-on-a-white-background)
+Figure taken from [Vecteezy](https://www.vecteezy.com/vector-art/3709640-isometric-cube-on-a-white-background)
 :::
@@
 
-## Supuestos: Ortogonalidad
+## Assumptions: Orthogonality
 
@@
 
-## Generalización: Ejes oblicuos {.smaller}
+## Generalization: Oblique axes {.smaller}
 
@@
 
-## Generalización: Ejes oblicuos
+## Generalization: Oblique axes
 
@@
 
-## Espacio latente vs. espacio test {.smaller}
+## Latent space vs. test space {.smaller}
 
@@
 
-## Parámetros multidimensionales generalizados
+## Generalized multidimensional parameters
 
@@
 
-## Representación gráfica completa
+## Complete graphical representation
 
@@
 
-## Conclusiones
+## Conclusions
 
@@
 
--   Paralelismo TRI multidimensional - AF
+-   Multidimensional IRT - FA parallelism
 
@@
 ::: fragment
--   Extensiones del modelo M2PL: Escala graduada, respuesta nominal, 3PL...
+-   M2PL model extensions: Graded scale, nominal response, 3PL...
 :::
@@
 ::: fragment
--   Investigación en otros modelos
+-   Other models
 :::

### 2025-07-07 output/IMPS-2025-slideshow.qmd
Update graphical output configuration

@@
 
-# Introducción
+## Multidimensional IRT
 
@@
 
-```{r color-palettes}
-MARGINS <- c(12, 1.3, 8, 1.3) # margins for the sem path plots
+```{r graphical-constants}
+# Color palettes:
 
@@
 
+# Probability axes:
+PROB_AXIS <- seq(-3.3, 3.3, by = .01)
+PROB_AXIS_2D <- seq(-1.5, 1.5, by = 0.05) # Axis for the probability contour plots
+```
+
+```{r graphical-output-conf}
+#| cache: false
+
+# Graphical output constants:
 GRAPH_FONT   <- "Arial"  # Replace default font to comply with UNED identity
@@
 VECTOR_WIDTH <- 1.5      # Replace default vector width for slideshow
-```
-
-```{r graphical-output-conf}
-#| cache: false
+
 theme_set( # `ggplot` output configuration
@@
 )
-```
-
-## Multidimensional IRT
+
+MARGINS <- c(12, 1.3, 8, 1.3) # margins for the sem path plots
+```
+
+M2PL model (mckinley_extension_1983):
 
@@
 ```
-:::
-:::::
 
@@
 #| fig-width:  9
-
-PROB_AXIS <- seq(-3.3, 3.3, by = .01)
 

### 2025-07-07 output/IMPS-2025-slideshow.qmd
Translate pending verbatim in slides

@@
 PROB_AXIS <- seq(-3.3, 3.3, by = .01)
-PROB_AXIS_2D <- seq(-1.5, 1.5, by = 0.05) # Axis for the probability contour plots
+PROB_AXIS_2D <- seq(-3, 3, by = 0.05) # Axis for the probability contour plots
 ```
@@
 
-M2PL model (mckinley_extension_1983):
-
-$$
-  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
-$$
+M2PL model [@mckinley_extension_1983]: $P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)$
 
@@
 #| fig-height: 5
-PROB_AXIS_2D <- seq(-3, 3, by = .1)
-
 mirt_item <- tribble(
@@
 ```
+:::
+:::::
 
@@
   geom_line(linewidth = LINE_WIDTH)                        +
-  labs(x = "Responsabilidad", y = "P(Y~i~ = 1)")           +
+  labs(x = "Conscientiousness", y = "P(Y~i~ = 1)")           +
   scale_y_continuous(
@@
 
-![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="left"}
+!["ACT Mathematics Usage Test" items](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="left"}
 :::
@@
 ::: column
-![Un cubo rojo en perspectiva isométrica](../www/slideshow-assets/isometric-cube.jpg){fig-align="center"}
+![A red cube in isometric perspective](../www/slideshow-assets/isometric-cube.jpg){fig-align="center"}
 :::
@@
 ::: r-center
-![Ítems de "ACT Mathematics Usage Test"](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height="400px"}
+!["ACT Mathematics Usage Test" items](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height="400px"}
 :::
@@
 
-¿Cuál es el sistema de coordenadas de *θ* para que *θ\** sea "esférica"?
+¿What is the coordinate system of *θ* to make *θ\** "spherical"?
 
@@
 ::: fragment
-[Respuestas en 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
+[Answers in 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
 :::
@@
 ```{r item-plot-out}
-#| fig-cap:   Espacio test ortogonal (izquierda) y oblicuo (derecha)
+#| fig-cap:   Orthogonal (left) and oblique (right) test space
 #| fig-align: center
@@
 
-Parámetros multidimensionales generalizados
-
--   A ítems con monotonía constante
-
--   A Ejes oblicuos
+Generalization of multidimensional parameters
+
+-   To constant-monotony items
+
+-   To oblique axes
 
@@
 ::: fragment
-Distinción espacio latente - espacio test
+Distinction latent space - test space
 :::

### 2025-07-09 output/seminar-slideshow.qmd
Revert latest commits on "reseach seminar" slides

--- 2025-07-03 output/seminar-slideshow.qmd
@@
 title:     |
-  A generalization of multidimensional item response theory parameters
+  Parámetros multidimensionales en Teoría de Respuesta al Ítem:
+  ¡Algo estamos haciendo mal!
 author:    |
@@
       <td>Daniel Morillo, Ph.D.</td>
-      <td>Mario Luzardo, Ph.D.</td>
     </tr>
@@
       </td>
-      <td>
-        <center>
-          [![](../www/slideshow-assets/github-logo.png){height="50"}](https://github.com/mluzardov/)&emsp;
-          [![](../www/slideshow-assets/orcid.png){height="50"}](https://orcid.org/0000-0002-9360-2806)
-        </center>
-      </td>
     </tr>
@@
   <table>
-    <!-- # TODO: Fix vertical alignment in logos-->
     <tr>
-      <td style="width: 400px">
-        [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="60"}](https://www.uned.es/universidad/facultades/psicologia.html)
-      </td>
-      <td style="width: 100px">
-        [![](../www/slideshow-assets/logo_udelar.svg){height="120"}](https://udelar.edu.uy)
+      <td>
+        <center>
+          [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="120"}](https://www.uned.es/universidad/facultades/psicologia.html)
+        </center>
       </td>
@@
 csl:           ../www/apa-old-doi-prefix.csl
-date:          "2024-07-17"
-date-meta:     "2024-11-27"
+date:          "2024-10-21"
+date-meta:     "2024-11-27" # TODO: Update
 date-format:   long
@@
   cache:       true
-lang:          en
+lang:          es
 knitr:

### 2025-07-09 output/IMPS-2025-slideshow.qmd
Version new "Introductory" IRT slides

@@
 
-## Multidimensional IRT
-
 ```{r libraries}
@@
 
+# Grid breaks:
+X_BREAKS <- seq(-3, 3, length.out = 5)
+
 # Probability axes:
@@
 
-M2PL model [@mckinley_extension_1983]: $P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)$
-
-::::: columns
-::: column
+## Unidimensional IRT {.smaller}
+
+-   2-parameter logistic (2PL) model [@birnbaum_latent_1968]:
+
+::::::::: columns
+::::::: {.column width="40%"}
+$$
+  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
+$$
+
+$$
+  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
+$$
+
+<br>
+
+:::::: {.fragment fragment-index="1"}
+::: r-stack
+[$a_1$ = 1.0; $b_1$ = 0.0]{style="color:#5c6eb1;"}
+:::
+
+::: r-stack
+[$a_2$ = 1.0; $b_2$ = 1.5]{style="color:#90214a;"}
+:::
+
+::: r-stack
+[$a_3$ = 2.5; $b_3$ = 0.0]{style="color:#749f4c;"}
+:::
+::::::
+:::::::
+
+::: {.column .fragment width="60%" fragment-index="1"}
+```{r irt-curves}
+#| fig-asp:   0.6
+#| fig-width: 8
+
+ITEMS_PALETTE <- c(
+  PALETTE_BLUE["normal"],
+  PALETTE_RASPBERRY["normal"],
+  PALETTE_APPLE["normal"]
+) |>
+  unname() |>
+  setNames("prob" |> paste(1:3, sep = '_'))
+
+ITEMS <- tribble(
+  ~item, ~a,   ~b,
+   1,     1,    0,
+   3,     1,    1.5,
+   2,     2.5,  0
+)
+
+logit <- function(x) 1 / (1 + exp(-x))
+irf   <- function(a, b, theta) logit(a * (theta - b))
+
+logit_curve_data <- tibble(
+  latent_trait = PROB_AXIS,
+  prob         = ITEMS |>
+    select(-item) |>
+    pmap(irf, theta = latent_trait) |>
+    bind_cols()
+) |>
+  unnest(
+    prob,
+    names_sep = '_'
+  ) |>
+  rename_with(str_remove, pattern = '\\.{3}')
+
+logit_curve_data_long <- logit_curve_data |>
+  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
+
+logit_curve_data_long |>
+  ggplot(
+    mapping = aes(
+      latent_trait, prob,
+      color = item,
+      group = after_scale(color)
+    )
+  )                                        +
+  geom_line(linewidth = LINE_WIDTH)        +
+  labs(x = "*&theta;*", y = "P(Y~i~ = 1)") +
+  scale_y_continuous(
+    breaks       = 0:1,
+    minor_breaks = NULL,
+    limits       = 0:1,
+    expand       = expansion(mult = c(0, .05))
+  )                                        +
+  scale_x_continuous(
+    breaks       = X_BREAKS,
+    minor_breaks = NULL,
+    limits = range(PROB_AXIS),
+    expand = expansion()
+  )                                        +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  theme(
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(angle = 90),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
+  )
+```
+
+<!-- TODO: Como ecuaciones (sin punto y coma?) -->
+:::
+:::::::::
+
+::: {.r-stack .fragment fragment-index="1"}
+$a_i$ = discrimination
+:::
+
+::: {.r-stack .fragment fragment-index="1"}
+$b_i$ = difficulty (location)
+:::
+
+::: notes
+The 2-parameter logistic model is the most simple model in item response theory (IRT).
+
+It is a prototypical model that allows us to study the item properties according to a parametric conception...
+
+\[NEXT\]
+
+... that assumes a discrimination ($a_i$) and a difficulty (or location, $b_i$) for each item.
+:::
+
+## Multidimensional IRT {.smaller}
+
+-   Multidimensional 2PL (M2PL) model [@mckinley_extension_1983]:
+
+:::::: columns
+:::: column
+$$
+  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
+$$
+
+::: {.fragment fragment-index="1"}
 ```{r mirt-model-irs-3d}
@@
 :::
-
-::: column
+::::
+
+::: {.column .fragment fragment-index="1"}
+<br>
+
 ```{r mirt-model-irs-contour}
@@
 :::
-:::::
+::::::
 
 ::: notes
-Esto se puede extender, de manera bastante directa, asumiendo que la respuesta se modela como una "regresión logística multivariada".
-Aquí aparece otra vez la discriminación pero ahora en forma vectorial, que se pre-multiplica (recordad, producto escalar) por el vector de rasgo latente de la persona, y un parámetro de intersección, como antes (pero no puede haber un parámetro de dificultad simplemente, o uno por cada dimensión, porque serían indistinguibles entre sí).
-Esta es una de las extensiones multidimensionales (más antiguas, y probablemente más obvias) del modelo logístico de 2 parámetros: El modelo logístico de 2 parámetros multidimensional compensatorio.
-Pertenece a una familia de modelos llamados "compensatorios" en TRI, ya que un nivel bajo en una puntuación latente se compensa con un nivel alto en otra, dando lugar a la misma probabilidad de respuesta.
+The Multidimensional 2PL model is one of many extensions of the 2PL model, which in this case is compensatory.
+It assumes that the probability of a correct response to an item is a function of the linear combination of the latent traits weighted by item discrimination parameters.
+
+\[NEXT\]
+
+This results in a response probability function where equal probability points form a straight line in the latent space.
 :::
@@
 
-$$
-  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
-$$
+-   Unidimensional 2PL model: $P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))$
+
+    -   $a_i$ = discrimination
+
+    -   $b_i$ = difficulty (location)
+
+-   Multidimensional 2PL model: $P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)$
+
+    -   $\mathbf{a}_i$ = ?
+
+    -   $d_i$ = ?
 
 ::: fragment
-$$
-  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
-$$
-:::
-
-::: fragment
-@reckase_difficulty_1985:
+@reckase_difficulty_1985 [p. 402]:
 
@@
 ::: notes
-A pesar de este paralelismo que os he mostrado, y de la utilidad de los modelos de TRI, hay un problema con la formulación del modelo que os he planteado, y es que como habéis podido ver los parámetros no son fácilmente interpretables.
-
-La intersección del modelo por ejemplo no tiene una interpretación clara.
-Si recordáis en la TRI unidimensional el parámetro de posición permitía comparar ítems entre sí en cuanto a su dificultad (en el caso de items cognitivos).
-Esto en AF no es muy habitual (prestamos poca atención a la "intersección" o "umbral"), porque no nos interesa tanto cómo de probable sea dar una respuesta positiva o negativa a un ítem, pero podría ayudarnos por ejemplo a diagnosticar ítems que estén mal diseñados, ya que nos da una idea de si va a haber muchas o pocas respuestas positivas a un ítem.
-
-Por eso Reckase dice, ya en 1985, que se necesita una forma de describir las características de los ítems que tenga en cuenta la dimensionalidad de las habilidades (los rasgos latentes en general).
-Daos cuenta de que entonces aún no se sabe nada de ese paralelismo con el AF.
-Además menciona que de esa manera se puede intentar comparar ítems que miden diferentes combinaciones de habilidades (fijaos en la importancia que tiene el dónde están los ítems en el continuo, en TRI, en contraste con AF, como comentábamos al principio).
+The problem with the M2PL, and other multidimensional extensions, is that its parameters do not have a clear interpretation in IRT terms, and they cannot be reparameterized.
+
+\[NEXT\]
+
+That is why Reckase & McKinley propose to find a way of describing the characteristics of multidimensional items, taking their dimensionality into account.
+
+They do it in this 1985 paper, and a subsequent one in 1991.
 :::

### 2025-07-09 output/sandbox_slideshow.qmd
Update sandbox slideshow to use with IMPS slides

@@
 library(scales)
+library(fontawesome)
 library(mvtnorm)
+library(patchwork)
 ```
@@
 source("R/Mirt_toolbox.R",              encoding = 'UTF-8')
+source("src/Graphical_example_paper.R", encoding = 'UTF-8')
 ```
 
-```{r color-palettes}
-MARGINS <- c(12, 1.3, 8, 1.3) # margins for the sem path plots
+```{r graphical-constants}
+# Color palettes:
 
@@
 
+# Factor axis colors:
+F1_COLOR <- unname(PALETTE_BLUE["normal"])
+F2_COLOR <- unname(PALETTE_RASPBERRY["normal"])
+
+# Grid breaks:
+X_BREAKS <- seq(-3, 3, length.out = 5)
+
+# Probability axes:
+PROB_AXIS <- seq(-3.3, 3.3, by = .01)
+PROB_AXIS_2D <- seq(-3, 3, by = 0.05) # Axis for the probability contour plots
+```
+
+```{r graphical-output-conf}
+#| cache: false
+
+# Graphical output constants:
 GRAPH_FONT   <- "Arial"  # Replace default font to comply with UNED identity
@@
 VECTOR_WIDTH <- 1.5      # Replace default vector width for slideshow
-```
 
-```{r graphical-output-conf}
-#| cache: false
 theme_set( # `ggplot` output configuration
@@
 )
-```
 
-```{r pre}
-PROB_AXIS    <- seq(-3, 3, by = .01)
-ERROR_POINTS <- seq(-3, 3, length.out = 5)
-PROB_AXIS_2D <- seq(-3, 3, by = .1)
-F1_COLOR <- unname(PALETTE_BLUE["normal"])
-
-logit <- function(x) 1 / (1 + exp(-x))
-irf   <- function(a, b, theta) logit(a * (theta - b))
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1,    0.5,  -0.5
-)
-
-mirt_item_probs <- expand_grid(
-  trait_2 = PROB_AXIS_2D,
-  trait_1 = PROB_AXIS_2D
-) |>
-  mutate(
-    mirt_item |>
-      select(-item) |>
-      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
-      bind_cols()
-  ) |>
-    rename(prob = `...1`)
-
-ITEMS_PALETTE <- c(
-  PALETTE_BLUE["normal"],
-  PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"],
-  PALETTE_TANGERINE["normal"],
-  PALETTE_UNED["normal"]
-) |>
-  unname() |>
-  setNames(1:5)
+MARGINS <- c(12, 1.3, 8, 1.3) # margins for the sem path plots
 ```

### 2025-07-09 output/IMPS-2025-slideshow.qmd
Add slide about multidimensional parameters

@@
 
-## Multidimensional IRT parameters {.smaller}
+## Multidimensional parameters {.smaller}
 
@@
 
-They do it in this 1985 paper, and a subsequent one in 1991.
+They do it in this 1985 paper, and a subsequent one in 1991...
+:::
+
+## Multidimensional parameters {.smaller}
+
+<br>
+
+::::: columns
+::: column
+-   @reckase_discriminating_1991:
+
+$$
+  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
+$$
+
+<br>
+
+-   @reckase_difficulty_1985:
+
+$$
+  MID_i = \begin{Bmatrix}
+    D_i &= \frac{-d_i}{MDISC_i} \\
+    \cos \mathbf{\alpha}_i &= \frac{\mathbf{a}_i}{MDISC_i}
+  \end{Bmatrix}
+$$
+:::
+
+::: {.column .fragment}
+-   @ackerman_graphical_1996:
+
+```{r m2pl-representation}
+#| fig-align:  center
+#| fig-height: 6.2
+#| fig-width:  6.2
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    0.5,  -0.5
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+d_i_spike <- mirt_item_coords |>
+  select(item, trait_1 = origin_1, trait_2 = origin_2) |>
+  bind_rows(tibble(item = 1, trait_1 = 0, trait_2 = 0)) |>
+  add_column(type = c("end", "start")) |>
+  pivot_wider(names_from = type, values_from = starts_with("trait"))
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data    = d_i_spike,
+    mapping = aes(
+      trait_1_start, trait_2_start,
+      xend = trait_1_end,
+      yend = trait_2_end
+    ),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = PALETTE_BLUE["light"]
+  ) +
+  geom_point(
+    data = mirt_item_coords,
+    mapping = aes(origin_1, origin_2),
+    inherit.aes = FALSE,
+    size = I(4)
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    geom = "richtext",
+    x = 0.16, y = 0.20,
+    label = "*D~i~*",
+    size = 9,
+    angle = mirt_item_params |> pull(deg_1),
+    label.colour  = NA,
+    fill = NA
+  ) +
+  annotate(
+    "richtext",
+    x = mirt_item_coords |> pull(origin_1) + 0.40,
+    y = mirt_item_coords |> pull(origin_2) + 0.32,
+    label = "*MDISC~i~*",
+    size  = 9,
+    angle = mirt_item_params |> pull(deg_1),
+    label.colour  = NA,
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::
+
+::: notes
+...where they define the item discrimination, which can in turn be used to define the item difficulty.
+
+Note that the difficulty has two components: a "signed distance" from the origin, and a direction.
+
+\[NEXT\]
+
+It is easier to appreciate this with the graphical representation, that was proposed by Ackerman a few years later, where we can see that the item vector is applied at a point given by the difficulty, and has a length equal to the discrimination.
+Note that the item direction always goes through the origin, and that the sign of the distance (positive in this case) will determine whether the vector is displaced forward or backward along that direction.
 :::

### 2025-07-09 output/IMPS-2025-slideshow.qmd
Drop details about multidimensional parameters

@@
 
-## Multidimensional difficulty {.smaller}
-
-@reckase_difficulty_1985:
-
-> \[...\] the most reasonable point to use in defining the \[multidimensional difficulty\] for an item in the multidimensional space is the point where the item is most discriminating.
-
----(p. 402)
-
-::: fragment
-$$
-  D_i = \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
-$$
-:::
-
-::: fragment
-> The distance can be interpreted much like a $b$ parameter from unidimensional IRT
-
----(p. 405)
-:::
-
-::: notes
-Lo que hace Reckase entonces es intentar definir un parámetro de "dificultad" (lo llama, para nosotros sería de posición), que sea de algún modo equivalente a la dificultad en el modelo unidimensional, es decir, el lugar donde el ítem es lo más discriminativo posible.
-Donde la curva de probabilidad de respuesta pasa de ser cada vez más abrupta a cada vez más plana.
-
-\[NEXT\]
-
-Y llega a esta definición, que consiste simplemente en el parámetro de intersección (con signo negativo), dividido entre la raíz del producto escalar (vuelve a aparecer el producto escalar), en este caso, del vector de parámetros de discriminación consigo mismo.
-Toda esta expresión se denomina "norma" del vector.
-
-\[NEXT\]
-
-Este parámetro, de "discriminación" o "dificultad", se puede interpretar de igual manera a la dificultad unidimensional.
-:::
-
-## Multidimensional difficulty
-
-::::::: columns
-::: column
-```{r mid-representation}
-#| fig-height: 8
-#| fig-width:  8
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-d_i_spike <- mirt_item_coords |>
-  select(item, trait_1 = origin_1, trait_2 = origin_2) |>
-  bind_rows(tibble(item = 1, trait_1 = 0, trait_2 = 0)) |>
-  add_column(type = c("end", "start")) |>
-  pivot_wider(names_from = type, values_from = starts_with("trait"))
-
-mirt_item_probs |>
-  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_contour2(
-    color = F1_COLOR,
-    linewidth = LINE_WIDTH,
-    skip = 0,
-    breaks = 1:9 / 10,
-    label.placer = label_placer_fraction(frac = .90, rot_adjuster = 0)
-  ) +
-  geom_segment(
-    data    = d_i_spike,
-    mapping = aes(
-      trait_1_start, trait_2_start,
-      xend = trait_1_end,
-      yend = trait_2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = PALETTE_BLUE["light"]
-  ) +
-  geom_point(
-    data = mirt_item_coords,
-    mapping = aes(origin_1, origin_2),
-    inherit.aes = FALSE,
-    size = I(4)
-  ) +
-  scale_x_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  annotate(
-    geom = "richtext",
-    x = 0.15, y = 0.15,
-    label = "*D~i~*",
-    size = 9,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  coord_fixed() +
-  theme(
-    axis.ticks       = element_blank(),
-    axis.line        = element_blank(),
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )
-```
-:::
-
-::::: column
-::: fragment
-$$
-  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
-$$
-:::
-
-<br>
-
-::: fragment
-$$
-  MID_i = \begin{Bmatrix}
-    \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}} \\
-    \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
-  \end{Bmatrix}
-$$
-:::
-:::::
-:::::::
-
-::: notes
-Geométricamente consiste en una "distancia **con signo**" (puede ser hacia cualquiera de ambos sentidos) desde el origen hasta ese punto donde está la posición del ítem, sobre la recta de inflexión (donde la probabilidad es igual a 0.5).
-
-\[NEXT\]
-
-A lo largo de toda esa recta podría ser cualquier punto, pero en realidad es el punto más cercano al origen, donde la recta que lo une al origen es perpendicular a esa misma recta de inflexión).
-Para ello claro, no basta con la "distancia (con signo)"; hace falta saber en qué dirección mide ese ítem también.
-Eso viene dado por los llamados "cosenos directores".
-Fijáos que vuelve a aparecer la norma de este vector en el denominador.
-
-\[NEXT\]
-
-Por lo tanto, la "dificultad (posición) multidimensional" está compuesta por dos elementos, un número real que representa cuánto se desplaza la recta de inflexión del ítem respecto del origen y en qué sentido, y un vector de cosenos que determina la dirección en la que se desplaza.
-Así se puede interpretar un ítem, o varios, de manera relativamente sencilla: Todos los participantes cuyas coordenadas estén en la "recta de inflexión" (la recta que pasa por la posición del ítem, perpendicular a su dirección) tendrán una probabilidad de dar una respuesta positiva y negativa de .5 (dificultad media; igual probabilidad de acertar o fallar).
-Aquellos que estén pasada la recta, encontrarán el ítem "relativamente fácil" (tendrán una alta probabilidad de dar una respuesta positiva mayor a .5; lo encontrarán "más fácil que difícil").
-Lo contrario pasa para los que estén al otro lado de esa recta, tendrán más probabilidad de responder negativamente (mal), por lo que lo encontrarán difícil.
-
-De lo que este parámetro no informa es de "cómo de fácil o difícil" va a ser el ítem para aquellos participantes que están muy cerca o muy lejos de esta recta.
-Es decir, una vez estoy pasada la recta, aunque esté cerca, ¿es el ítem muy fácil, o más bien tengo que estar "muy lejos" (tener alta puntuación en uno u otro rasgo) para que sea más probable que de una respuesta positiva?
-:::
-
-## Multidimensional discrimination {.smaller}
-
-@reckase_discriminating_1991 [p. 362]:
-
-> the discriminating power of an item indicates how quickly the transition takes place from low probability to high probability \[...\].
-> A highly discriminating item divides the regions clearly---having a narrow region \[...\] where the probabilities are intermediate in magnitude.
-
-::::: columns
-::: {.column .fragment}
-```{r low-disc-contour}
-#| fig-width:  4.7
-#| fig-height: 4.7
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     0.5,  0.25,  -0.125
-)
-
-mirt_item_probs <- expand_grid(
-  trait_2 = PROB_AXIS_2D,
-  trait_1 = PROB_AXIS_2D
-) |>
-  mutate(
-    mirt_item |>
-      select(-item) |>
-      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
-      bind_cols()
-  ) |>
-    rename(prob = `...1`)
-
-mirt_item_probs |>
-  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_contour2(
-    color = F1_COLOR,
-    linewidth = LINE_WIDTH,
-    breaks = 1:9 / 10,
-    skip = 0,
-    label.placer = label_placer_fraction(frac = .80, rot_adjuster = 0)
-  ) +
-  scale_x_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  coord_fixed() +
-  theme(
-    axis.ticks       = element_blank(),
-    axis.line        = element_blank(),
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )
-```
-:::
-
-::: {.column .fragment}
-```{r high-disc-contour}
-#| fig-width:  4.7
-#| fig-height: 4.7
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1.5,  0.75,  -0.375
-)
-
-mirt_item_probs <- expand_grid(
-  trait_2 = PROB_AXIS_2D,
-  trait_1 = PROB_AXIS_2D
-) |>
-  mutate(
-    mirt_item |>
-      select(-item) |>
-      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
-      bind_cols()
-  ) |>
-    rename(prob = `...1`)
-
-mirt_item_probs |>
-  ggplot(mapping = aes(trait_1, trait_2, z = prob, label = after_stat(level))) +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_contour2(
-    color = F1_COLOR,
-    linewidth = LINE_WIDTH,
-    breaks = 1:9 / 10,
-    # skip = 0,
-    label.placer = label_placer_fraction(frac = .90, rot_adjuster = 0)
-  ) +
-  scale_x_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    breaks       = -1:1,
-    minor_breaks = NULL,
-    limits       = c(-1.5, 1.5),
-    expand       = expansion(),
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  coord_fixed() +
-  theme(
-    axis.ticks       = element_blank(),
-    axis.line        = element_blank(),
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )
-```
-:::
-:::::
-
-::: notes
-Esta característica viene representada por el parámetro de "discriminación", que es una medida de la "calidad general" de un ítem.
-Los ítems "malos", o poco discriminantivos, tendrán una superficie de respuesta bastante "plana", con una transición suave entre la región de baja y la de alta probabilidad de respuesta, mientras que los muy discriminativos distinguirán bien entre la gente que está en la región de alta o baja probabilidad de respuesta, con zona de transición más estrecha, más abrupta.
-:::
-
-## Multidimensional discrimination {.smaller}
-
-```{r mdisc-items}
-mdisc_items <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     4,    0.1,     0,
-  2,     0.01, 0.02,  0,
-)
-
-mirt_item_probs <- expand_grid(
-  trait_1 = PROB_AXIS_2D,
-  trait_2 = PROB_AXIS_2D
-) |>
-  mutate(
-    mdisc_items |>
-      select(-item) |>
-      pmap(irs2d, theta_1 = trait_1, theta_2 = trait_2) |>
-      bind_cols()
-  ) |>
-    rename_with(str_replace, pattern = '\\.{3}', replacement = "prob_")
-```
-
-::::: columns
-::: column
-$$
-  \mathbf{a}_1 = \begin{bmatrix}
-    0.1 \\
-    0.02
-  \end{bmatrix};\quad
-  d_1 = 0
-$$
-
-```{r low-mdisc-irs-contour}
-#| fig-height: 5
-#| fig-width:  5
-mirt_prob_surface <- mirt_item_probs |>
-  select(starts_with("trait"), prob_2) |>
-  pivot_wider(names_from = trait_1, values_from = prob_2) |>
-    select(-trait_2) |>
-    as.matrix()
-
-plot_ly(
-  x = ~PROB_AXIS_2D,
-  y = ~PROB_AXIS_2D,
-  z = ~mirt_prob_surface,
-  colorscale = list(c(0, 1), PALETTE_BLUE["normal"] |> rep(2)),
-  type = 'surface',
-  hovertemplate = 'P(I<sub>1</sub> = 1) = %{z:.2f}<extra></extra>',
-  showscale = FALSE,
-  hoverinfo = 'Z'
-) |>
-  layout(
-    scene = list(
-      xaxis = list(title = "<i>θ</i><sub>1</sub>"),
-      yaxis = list(title = "<i>θ</i><sub>2</sub>"),
-      zaxis = list(
-        range = 0:1,
-        title = "P(I<sub>1</sub> = 1)"
-      ),
-      camera = list(
-        center = list(x = -.1, z = -.2),
-        eye = list(x = -.12, y = -1.8, z = 1)
-      )
-    )
-  ) |>
-  config(displayModeBar = FALSE, scrollZoom = FALSE)
-```
-:::
-
-::: column
-$$
-  \mathbf{a}_2 = \begin{bmatrix}
-    4 \\
-    0.1
-  \end{bmatrix};\quad
-  d_2 = 0
-$$
-
-```{r high-mdisc-irs-contour}
-#| fig-height: 5
-#| fig-width:  5
-mirt_prob_surface <- mirt_item_probs |>
-  select(trait_1:prob_1) |>
-  pivot_wider(names_from = trait_1, values_from = prob_1) |>
-    select(-trait_2) |>
-    as.matrix()
-
-plot_ly(
-  x = ~PROB_AXIS_2D,
-  y = ~PROB_AXIS_2D,
-  z = ~mirt_prob_surface,
-  colorscale = list(c(0, 1), PALETTE_BLUE["normal"] |> rep(2)),
-  type = 'surface',
-  hovertemplate = 'P(I<sub>2</sub> = 1) = %{z:.2f}<extra></extra>',
-  showscale = FALSE,
-  hoverinfo = 'Z'
-) |>
-  layout(
-    scene = list(
-      xaxis = list(title = "<i>θ</i><sub>1</sub>"),
-      yaxis = list(title = "<i>θ</i><sub>2</sub>"),
-      zaxis = list(
-        range = 0:1,
-        title = "P(I<sub>2</sub> = 1)"
-      ),
-      camera = list(
-        center = list(x = -.1, z = -.2),
-        eye = list(x = -.12, y = -1.8, z = 1)
-      )
-    )
-  ) |>
-  config(displayModeBar = FALSE, scrollZoom = FALSE)
-```
-:::
-:::::
-
-::: notes
-La discriminación se puede entender con una especie de "dardo" que atraviesa de arriba abajo esa superficie perpendicularmente por esa "recta de inflexión" de la que hablábamos antes, y luego cruza el plano origen.
-Si la superficie es "totalmente plana" el dardo entra por un punto y sale por otro que está justo debajo, en las mismas coordenadas de rasgo latente.
-<!-- TODO: Añadir los vectores perpendiculares en las gráficas? -->.
-La "proyección" (si ponemos un foco justo encima, la sombra que forma en el plano) es un punto.
-Es decir, la "discriminación" es cero (el vector del ítem tiene longitud "cero").
-Si la pendiente es muy pronunciada (distingue bien entre niveles de rasgo) el dardo atravesará el plano origen muy lejos.
-La sombra será muy larga (es decir, la discriminación muy alta).
-El caso "extremo" es una superficie "vertical" en la recta de inflexión (un escalón; cosa imposible en ítems reales).
-En este caso, el dardo nunca cortaría al plano origen, y la discriminación sería infinita.
-
-Pero cuidado, que el ítem "discimine mucho" no significa que sea "bueno" midiendo todas las dimensiones, ya que eso va a depender también de "en qué dirección" mida el ítem, y eso viene dado por el componente de dirección que hemos visto antes.
-:::
-
-## Multidimensional discrimination
-
-::::::: columns
-::: column
-```{r mdisc-item}
-#| fig-width:  8
-#| fig-height: 8
-
-item_vector <- mirt_item_coords |> mutate(
-  across(ends_with("1"), ~ . - origin_1),
-  across(ends_with("2"), ~ . - origin_2)
-)
-
-item_spikes <- item_vector             |>
-  slice(1 |> rep(2))                   |>
-  add_column(dim = 'F' |> paste0(1:2)) |>
-  mutate(
-    origin_1 = if_else(dim == 'F1', end_1, origin_1),
-    origin_2 = if_else(dim == 'F2', end_2, origin_2),
-  )
-
-item_vector |>
-  ggplot(
-    aes(
-      0, 0,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data = item_spikes,
-    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color     = PALETTE_BLUE["light"]
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    "richtext",
-    x = 0.32, y = 0.25,
-    label = "*MDISC~i~*",
-    size  = 9,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  annotate(
-    "richtext",
-    x = c(0.5, -0.13), y = c(-0.12, 0.25),
-    label         = c("*a*~1*i*~", "*a*~2*i*~"),
-    size          = 8,
-    label.colour  = NA,
-    label.padding = unit(0, "lines"),
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-
-::::: column
-::: fragment
-@reckase_discriminating_1991:
-
-<br>
-
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-:::
-
-::: fragment
-$$
-  MID_i = \begin{Bmatrix}
-    \frac{-d_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}} \\
-    \frac{\mathbf{a}_i}{\sqrt{\mathbf{a}_i^T\mathbf{a}_i}}
-  \end{Bmatrix}
-$$
-:::
-:::::
-:::::::
-
-::: notes
-@reckase_discriminating_1991 hacen la deducción de la discriminación multidimensional, y el resultado es simplemente la "longitud" del vector de parámetros de discriminación.
-Esto sería, en cierto modo, equivalente a la "representación factorial" del ítem; aquí la punta del vector está donde estarían las "bolitas" que usábamos antes para representar los pesos factoriales en el espacio de factores comunes, excepto que, para un mismo ítem, estaría escalado por la varianza única del ítem y el factor aquel de 1.7 que salía en la fórmula, pero por lo demás es una representación equivalente a la del modelo factorial.
-
-El cálculo puede hacerse simplemente aplicando el teorema de Pitágoras.
-El vector es un triángulo rectángulo, $a_1$ sería un cateto, el otro cateto (SEÑALAR "SPIKE") sería igual a $a_2$, y la hipotenusa es la longitud (las dos discriminaciones al cuadrado, sumadas, y luego la raíz cuadrada de eso).
-Por ejemplo, aquí $a_1$ es 1, al cuadrado 1, más 0.5 ($a_2$) al cuadrado que es 0,25, luego la raíz cuadrada de 1,25 (sale 1,12 aproximadamente).
-
-\[NEXT\]
-
-En forma matricial, esto no es más que el producto escalar del vector consigo mismo, es decir, aparece otra vez la norma del vector, nuestra vieja amiga.
-Y esta fórmula por cierto vale para cualquier número de dimensiones.
-Aquí estamos representando sólo 2 porque es más fácil, pero podemos calcular el producto escalar de este vector de parámetros de discriminación consigo mismo (sea cual sea su dimensión), y calcular su raíz cuadrada, y eso nos daría igualmente la discriminación multidimensional (la longitud de ese vector de $\mathbf{a}$ del ítem).
-
-Si recordáis además, esta norma aparecía en el denominador de la dificultad multidimensional (en ambos componentes).
-:::
-
-## Multidimensional parameters
-
-<br>
-
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-
-<br>
-
-$$
-  MID_i = \begin{Bmatrix}
-    \frac{-d_i}{MDISC_i} \\
-    \frac{\mathbf{a}_i}{MDISC_i}
-  \end{Bmatrix}
-$$
-
-::: notes
-Luego esta expresión de la discriminación multidimensional se puede usar para expresar de manera más sencilla la dificultad.
-Y así, esta sería la expresión completa de los parámetros multidimensionales de un ítem según el modelo de TRI logístico de 2 parámetros.
-:::
-
-## Graphical representation
-
-@ackerman_graphical_1996:
-
-::: r-center
-```{r m2pl-representation}
-#| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  7.2
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data    = d_i_spike,
-    mapping = aes(
-      trait_1_start, trait_2_start,
-      xend = trait_1_end,
-      yend = trait_2_end
-    ),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = PALETTE_BLUE["light"]
-  ) +
-  geom_point(
-    data = mirt_item_coords,
-    mapping = aes(origin_1, origin_2),
-    inherit.aes = FALSE,
-    size = I(4)
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    geom = "richtext",
-    x = 0.16, y = 0.20,
-    label = "*D~i~*",
-    size = 9,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  annotate(
-    "richtext",
-    x = mirt_item_coords |> pull(origin_1) + 0.40,
-    y = mirt_item_coords |> pull(origin_2) + 0.32,
-    label = "*MDISC~i~*",
-    size  = 9,
-    angle = mirt_item_params |> pull(deg_1),
-    label.colour  = NA,
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-
-::: notes
-La propuesta de representación gráfica completa (y consistente) no llega hasta 1996.
-Ackerman propone que el ítem se represente como un vector de longitud dada por la discriminación multidimensional, aplicado en la posición del ítem (y en la dirección dada por esa posición; es decir, todos los ítems "apuntan" siempre en la dirección que pasa por el origen).
-:::
-
-## Graphical representation
-
-@ackerman_graphical_1996:
-
-::: r-center
-```{r m2pl-representation-final}
-#| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  7.2
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-
-::: notes
-La representación como tal consiste solamente en el vector, aplicado en ese punto, obviamente.
-Y lo interesante que tiene esta representación es que nos permite visualizar con claridad las propiedades de estos parámetros multidimensionales.
-:::
-
-## Properties: *MDISC* {.smaller}
-
-:::::: columns
-:::: column
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-
-::: {.fragment .r-stack style="color:#5c6eb1;" fragment-index="1"}
-$$
-  \begin{align}
-    \mathbf{a}_1 &= \begin{bmatrix}
-      1 \\
-      1
-    \end{bmatrix} \\
-    MDISC_1 &\approx 1,41
-  \end{align}
-$$
-:::
-::::
-
-::: {.column .fragment fragment-index="1"}
-```{r mdisc-comparison-1}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-mdisc_diff_items <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1,    1,    0,
-  2,     2,    2,    0,
-)
-
-mdisc_diff_params <- mdisc_diff_items |>
-  compute_mirt_params(d, starts_with('a'))
-
-mdisc_diff_coords <- mdisc_diff_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mdisc_diff_coords |>
-  filter(item == 1) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1/12),
-    axis.title.y.right = element_markdown(vjust = 1/12),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-::::::
-
-::: notes
-Por ejemplo, la discriminación multidimensional, vemos que depende solamente del parámetro de discriminación,
-
-\[NEXT\]
-
-que en este ejemplo tiene dos componentes.
-:::
-
-## Properties: *MDISC* {.smaller}
-
-::::::: columns
-::::: column
-$$
-  MDISC_i = \sqrt{\mathbf{a}_i^T \mathbf{a}_i}
-$$
-
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{align}
-    \mathbf{a}_1 &= \begin{bmatrix}
-      1 \\
-      1
-    \end{bmatrix} \\
-    MDISC_1 &\approx 1,41
-  \end{align}
-$$
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$$
-  \begin{align}
-    \mathbf{a}_2 &= \begin{bmatrix}
-      2 \\
-      2
-    \end{bmatrix} \\
-    MDISC_2 &\approx 2,83
-  \end{align}
-$$
-:::
-:::::
-
-::: column
-```{r mdisc-comparison-2}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-ITEMS_PALETTE <- c(
-  PALETTE_BLUE["normal"],
-  PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"],
-  PALETTE_TANGERINE["normal"],
-  PALETTE_UNED["medium"]
-) |>
-  unname() |>
-  setNames(1:5)
-
-# TODO: Cambiar grosor de los ítems
-mdisc_diff_plot <- mdisc_diff_coords |>
-  arrange(item |> desc()) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1/12),
-    axis.title.y.right = element_markdown(vjust = 1/12),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-
-mdisc_diff_plot
-```
-:::
-:::::::
-
-::: notes
-Cuanto más grandes sean los componentes del parámetro de discriminación, mayor será también la discriminación multidimensional.
-(Aunque en este ejemplo sean iguales entre sí, no tienen por qué serlo).
-Hasta aquí, todo es sencillo (creo).
-:::
-
-## Properties: *D* {.smaller}
-
-:::::: columns
-:::: column
-$$
-  D_i = \frac{-d_i}{MDISC_i}
-$$
-
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{matrix}
-    \begin{align}
-      d_1     &= 2 \\
-      MDISC_1 &= \sqrt{2}
-    \end{align} &
-    D_1 \approx -1,41
-  \end{matrix}
-$$
-:::
-::::
-
-::: column
-```{r d-comparison-1}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-mdisc_diff_items <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1,    1,    2,
-  2,     1,    1,    1,
-  3,     1,    1,    -1,
-  4,     2,    2,    -1,
-)
-
-mdisc_diff_params <- mdisc_diff_items |>
-  compute_mirt_params(d, starts_with('a'))
-
-mdisc_diff_coords <- mdisc_diff_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mdisc_diff_coords |>
-  filter(item == 1) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1.1/3.5),
-    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-::::::
-
-::: notes
-En cuanto a la distancia con signo, obviamente es proporcional a la intersección del modelo, eso creo que también es muy evidente
-:::
-
-## Properties: *D* {.smaller}
-
-::::::: columns
-::::: column
-$$
-  D_i = \frac{-d_i}{MDISC_i}
-$$
-
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{matrix}
-    \begin{align}
-      d_1     &= 2 \\
-      MDISC_1 &= \sqrt{2}
-    \end{align} &
-    D_1 \approx -1,41
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$$
-  \begin{matrix}
-    \begin{align}
-      d_2     &= 1 \\
-      MDISC_2 &= \sqrt{2}
-    \end{align} &
-    D_2 \approx -0,71
-  \end{matrix}
-$$
-:::
-:::::
-
-::: column
-```{r d-comparison-2}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-mdisc_diff_coords |>
-  filter(item %in% 1:2) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1.1/3.5),
-    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::::
-
-::: notes
-Aquí vemos un ítem con la mitad del valor de la intersección que el primero.
-Por lo tanto, su distancia al origen es la mitad.
-(Los ítems están un poco solapados, pero empieza aquí; espero que se vea bien).
-:::
-
-## Properties: *D* {.smaller}
-
-:::::::: columns
-:::::: column
-$$
-  D_i = \frac{-d_i}{MDISC_i}
-$$
-
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{matrix}
-    \begin{align}
-      d_1     &= 2 \\
-      MDISC_1 &= \sqrt{2}
-    \end{align} &
-    D_1 \approx -1,41
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$$
-  \begin{matrix}
-    \begin{align}
-      d_2     &= 1 \\
-      MDISC_2 &= \sqrt{2}
-    \end{align} &
-    D_2 \approx -0,71
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#749f4c;"}
-<!-- TODO: Align D_4 value (add space with equal width as minus signs) -->
-
-$$
-  \begin{matrix}
-    \begin{align}
-      d_3     &= -1 \\
-      MDISC_3 &= \sqrt{2}
-    \end{align} &
-    D_3 \approx 0,71
-  \end{matrix}
-$$
-:::
-::::::
-
-::: column
-```{r d-comparison-3}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-mdisc_diff_coords |>
-  # TODO: Item 3 on top
-  filter(item %in% 1:3) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1.1/3.5),
-    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-::::::::
-
-::: notes
-Y además, el desplazamiento respecto del origen es en sentido opuesto al signo de la intersección.
-(Si consideramos el sentido positivo hacia donde el ítem apunta, que es el ascendente para ambos rasgos latentes, o sea, hacia arriba a la derecha).
-Este tercer ítem se encuentra aquí, justo en la punta del segundo.
-:::
-
-## Properties: *D* {.smaller}
-
-::::::: columns
-::::: column
-$$
-  D_i = \frac{-d_i}{MDISC_i}
-$$
-
-::: {.r-stack style="color:#749f4c;"}
-<!-- TODO: Align D_4 value with prevous slide -->
-
-$$
-  \begin{matrix}
-    \begin{align}
-      d_3     &= -1 \\
-      MDISC_3 &= \sqrt{2}
-    \end{align} &
-    D_3 \approx 0,71
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#d76f47;"}
-$$
-  \begin{matrix}
-    \begin{align}
-      d_4     &= -1 \\
-      MDISC_4 &= \sqrt{8}
-    \end{align} &
-    D_4 \approx 0,35
-  \end{matrix}
-$$
-:::
-:::::
-
-::: column
-```{r d-comparison-4}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-# TODO: Cambiar grosor de los ítems
-mdisc_diff_coords |>
-  slice(4:3) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1.1, 2.4),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1.1/3.5),
-    axis.title.y.right = element_markdown(vjust = 1.1/3.5),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::::
-
-::: notes
-Y además, el desplazamiento respecto del origen es en sentido opuesto al signo de la intersección.
-(Si consideramos el sentido positivo hacia donde el ítem apunta, que es el ascendente para ambos rasgos latentes, o sea, hacia arriba a la derecha).
-Este tercer ítem se encuentra aquí, justo en la punta del segundo.
-:::
-
-## Properties: $\cos \, \mathbf{\alpha}$ {.smaller}
-
-::::::: columns
-::::: column
-$$
-  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
-$$
-
-<br>
-
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_1 \approx \begin{bmatrix}
-        1 \\
-        1
-      \end{bmatrix} &
-      \mathbf{\alpha}_1 =  \begin{bmatrix}
-        45º \\
-        45º
-      \end{bmatrix}
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_2 \approx \begin{bmatrix}
-        2 \\
-        2
-      \end{bmatrix} &
-      \mathbf{\alpha}_2 =  \begin{bmatrix}
-        45º \\
-        45º
-      \end{bmatrix}
-  \end{matrix}
-$$
-:::
-:::::
-
-::: column
-```{r dir-comparison-1}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-mdisc_diff_plot
-```
-:::
-:::::::
-
-::: notes
-Por último, en cuanto a la dirección del vector: Vemos que, mientras que los componentes del parámetro de discriminación se mantengan proporcionales, la dirección no va a cambiar, a pesar de que la discriminación multidimensional sí cambia.
-:::
-
-## Properties: $\cos \, \mathbf{\alpha}$ {.smaller}
-
-::::::: columns
-::::: column
-$$
-  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
-$$
-
-<br>
-
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_1 \approx \begin{bmatrix}
-        1 \\
-        1
-      \end{bmatrix} &
-      \mathbf{\alpha}_1 =  \begin{bmatrix}
-        45º \\
-        45º
-      \end{bmatrix}
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#749f4c;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_3 \approx \begin{bmatrix}
-        1,329 \\
-        0,484
-      \end{bmatrix} &
-      \mathbf{\alpha}_3 =  \begin{bmatrix}
-        20º \\
-        70º
-      \end{bmatrix}
-  \end{matrix}
-$$
-:::
-:::::
-
-::: column
-```{r dir-comparison-2}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-cos_diff_items <- tribble(
-  ~item, ~a_1,                       ~a_2,                       ~d,
-  1,     1,                          1,                          0,
-  2,     2,                          2,                          0,
-  3,     cos((20/180*pi)) * sqrt(2), cos((70/180*pi)) * sqrt(2), 0,
-  4,     0,                          2,                          0,
-  5,     2,                          0,                          0
-)
-
-cos_diff_params <- cos_diff_items |>
-  compute_mirt_params(d, starts_with('a'))
-
-cos_diff_coords <- cos_diff_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-cos_diff_coords |>
-  filter(item %in% c(1, 3)) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1/12),
-    axis.title.y.right = element_markdown(vjust = 1/12),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::::
-
-::: notes
-Ahora bien, si cambia la proporción entre ellos, la dirección sí va a cambiar; y eso va a ser independiente de que la disciminación multidimensional se mantenga constante, como vemos en este ejemplo: misma longitud del vector, distinta dirección.
-Es decir, que por ejemplo un ítem puede medir "igual de bien" dos constructos, y otro, que mida más bien uno de ellos, estaría más alineado con esa dirección, a pesar de tener la misma discriminación multidimensional, de ser igual de discriminante "en términos globales", diríamos.
-:::
-
-## Properties: $\cos \, \mathbf{\alpha}$ {.smaller}
-
-::::::: columns
-::::: column
-$$
-  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
-$$
-
-<br>
-
-::: {.r-stack style="color:#d76f47;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_4 \approx \begin{bmatrix}
-        0 \\
-        2
-      \end{bmatrix} &
-      \mathbf{\alpha}_4 =  \begin{bmatrix}
-        90º \\
-         0º
-      \end{bmatrix}
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#427562;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_5 \approx \begin{bmatrix}
-        2 \\
-        0
-      \end{bmatrix} &
-      \mathbf{\alpha}_5 =  \begin{bmatrix}
-         0º \\
-        90º
-      \end{bmatrix}
-  \end{matrix}
-$$
-:::
-:::::
-
-::: column
-```{r dir-comparison-3}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-cos_diff_coords |>
-  filter(item %in% 4:5) |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.2, 2.2),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1/12),
-    axis.title.y.right = element_markdown(vjust = 1/12),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::::
-
-::: notes
-Y por último, vemos que, cuando uno de los valores de la discriminación es cero, el vector es perpendicular al eje correspondiente (además de ser paralelo al otro eje, pero esto es una simplificación por tener dos dimensones solamente).
-Esto es muy importante, ya que nos viene a indicar que, cuando un ítem no mide en absoluto alguna de las dimensiones del instrumento que estamos considerando, entonces no proporciona absolutamente ninguna información del nivel de rasgo en esa dimensión de las personas que están respondiendo al instrumento.
-:::
-
 ## Assumptions: Increasing monotony {.smaller}

### 2025-07-09 output/IMPS-2025-slideshow.qmd
Add "assumptions" slide

@@
 
-That is why Reckase & McKinley propose to find a way of describing the characteristics of multidimensional items, taking their dimensionality into account.
-
-They do it in this 1985 paper, and a subsequent one in 1991...
+That is why Reckase proposes to find a way of describing the characteristics of multidimensional items, taking their dimensionality into account.
+
+He do it in this 1985 paper, and in a subsequent one in 1991...
 :::
@@
 ::: notes
-...where they define the item discrimination, which can in turn be used to define the item difficulty.
-
-Note that the difficulty has two components: a "signed distance" from the origin, and a direction.
+...Reckase & McKinley define the item discrimination, which is an extension of the Pythagorean theorem.
+
+Note also that the discrimination can in turn be used to define the item difficulty, as happens in the unidimensional model.
+
+And also, that the difficulty has two components: a "signed distance" from the origin, and a direction.
 
@@
 Note that the item direction always goes through the origin, and that the sign of the distance (positive in this case) will determine whether the vector is displaced forward or backward along that direction.
+:::
+
+## Assumptions {.smaller}
+
+<br>
+
+@reckase_difficulty_1985:
+
+-   Increasing monotony:
+
+> the probability of answering an item correctly increases monotonically with an increase in each dimension being measured
+
+---(p. 402)
+
+-   Orthogonality
+
+> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
+
+---(p. 404)
+
+::: notes
+However, there are two assumptions that Reckase & McKinley make when obtaining the definition of the multidimensional parameters.
+
+The first is that the probability of answering an item correctly always increases monotonically with respect to each dimension.
+
+The second, that constrain that you see there, implies that the item is represented in an orthogonal basis.
+That is the reason why the discrimination is defined by a "general Pythagorean theorem".
 :::

### 2025-07-09 output/IMPS-2025-slideshow.qmd
Add "null probability" argument slide

@@
 
-The second, that constrain that you see there, implies that the item is represented in an orthogonal basis.
-That is the reason why the discrimination is defined by a "general Pythagorean theorem".
-:::
-
-## Assumptions: Increasing monotony {.smaller}
-
-> the probability of answering an item correctly increases monotonically with an increase in each dimension being measured
-
-<!-- TODO: Put in column and make figure in 2nd column taller -->
-
----@reckase_difficulty_1985 [p. 402]
-
-::::: {.columns .fragment}
+The second, that constrain that you see there, implies that the item is represented in an orthogonal basis, which is why the discrimination is defined by a "general Pythagorean theorem".
+
+So, our purpose in this work is **to relax these two assumptions**.
+:::
+
+## Assumptions: Increasing monotony
+
+<br>
+
+::::: columns
 ::: column
@@
 $$
+:::
+
+::: column
+```{r m2pl-representation-unidimensional}
+#| fig-align:  center
+#| fig-height: 7.2
+#| fig-width:  7.2
+
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1.5,  0,    0.75
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-1, 1.5),
+    breaks       = -1:2,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-1, 1),
+    breaks       = NULL,
+    minor_breaks = NULL,
+    name         = NULL,
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = .4),
+    axis.title.y.right = element_markdown(vjust = .4),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    ),
+  )
+```
+:::
+:::::
+
+::: notes
+In the first case, one may argue that we may deal with "undimensional items" or, more generally, items that have null discrimination components.
+
+This would imply that the probability of a correct response would not change with respect to that dimension; i.e., that trait is not being measured by the item, so the isoprobability lines would be parallel to the $\theta_2$ axis (in this example.
+
+Therefore, the probability does not "increase monotonically".
+
+Nevertheless, with this caveat about null discrimination components, this assumption would be realistic in a "cognitive testing context": The more you know, the more likely you are to give a correct answer.
 :::

### 2025-07-09 output/IMPS-2025-slideshow.qmd
Add item color palette to the preface

@@
 
+# Item colors:
+ITEMS_PALETTE <- c(
+  PALETTE_BLUE["normal"],
+  PALETTE_RASPBERRY["normal"],
+  PALETTE_APPLE["normal"],
+  PALETTE_TANGERINE["normal"],
+  PALETTE_UNED["medium"]
+) |>
+  unname() |>
+  setNames(1:5)
+
 # Grid breaks:
@@
 
-ITEMS_PALETTE <- c(
-  PALETTE_BLUE["normal"],
-  PALETTE_RASPBERRY["normal"],
-  PALETTE_APPLE["normal"]
-) |>
-  unname() |>
-  setNames("prob" |> paste(1:3, sep = '_'))
-
 ITEMS <- tribble(
@@
 )
-
-logit <- function(x) 1 / (1 + exp(-x))
-irf   <- function(a, b, theta) logit(a * (theta - b))
 
@@
 logit_curve_data_long <- logit_curve_data |>
-  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
+  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob") |>
+  mutate(item = item |> str_remove("prob_"))
 

### 2025-07-09 output/IMPS-2025-slideshow.qmd
Add slide about inverse non-cognitive items

@@
 
-::: column
-```{r m2pl-representation-unidimensional}
-#| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  6
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1.5,  0,    0.75
-)
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1),
-    breaks       = NULL,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::
-
-::: notes
-Ahora bien, como hemos visto, la TRI originalmente sólo trataba con "ítems cognitivos" (lo hemos podido ver en el propio lenguaje que utiliza Reckase: "dificultad", "habilidad", "respuesta correcta").
-Uno de los supuestos que hace es este.
-Esto tiene todo el sentido, de acuerdo, ya que si la probabilidad de un ítem "decrece" al aumentar el nivel de habilidad, es que algo está mal en el ítem: No puede ser que alguien más inteligente (pongamos como ejemplo) tenga menos probabilidad de acertar un ítem que alguien menos inteligente.
-
-\[NEXT\]
-
-Pero cuidado, si el ítem es "puro" de una dimensión (o no mide "todas" las dimensiones del test), la probabilidad en la otra (u otras) dimensiones sel test sería constante (porque su discriminación sería 0); es decir, la probabilidad puede ser "plana" (es una dimensión que no "está siendo medida por el ítem").
-En este ejemplo, las rectas de isoprobabilidad serían paralelas al eje $\theta_2$.
-:::
-
-## Generalization: Inverse items {.smaller}
+## Assumptions: Increasing monotony {.smaller}
 
@@
 
-::::: columns
-::: {.column .fragment .r-center}
+:::::: {.columns .fragment}
+::: {.column .r-center}
 ```{r inverse-item-curve}
@@
 
-::: {.column .fragment .r-center}
-<br> <br>
-
+:::: {.column .r-center}
 $$
@@
 $$
-:::
-:::::
+
+<br>
+
+::: r-stack
+[$a_i < 0$]{style="color:#5c6eb1;"}
+:::
+::::
+::::::
 
 ::: notes
-Pero además, si hablamos de ítems "no cognitivos" (p.ej. personalidad) en los que la persona evaluada puede estar "de acuerdo" o "en desacuerdo", sí podemos tener ítems con probabilidad "decreciente".
-
-\[NEXT\]
-
-Formalmente, esto vendría no sería más que un parámetro de discriminación negativo.
-
-\[NEXT\]
-
+However, if we wanted to apply this model to "non-cognitive" items (e.g., personality traits) where the person evaluated can be "agreeing" or "disagreeing", we can have items with "decreasing" probability.
+
+\[NEXT\]
+
+Formally, what this would imply is a negative discrimination parameter.
+:::
+
+## Generalization: Inverse items {.smaller}
+
+::: notes
 Si recordáis la "función característica del ítem", esto consiste simplemente en un valor negativo en el parámetro $a$; eso hace que la probabilidad sea decreciente.

### 2025-07-10 output/IMPS-2025-slideshow.qmd
Add "Constant monotony" section

@@
 Formally, what this would imply is a negative discrimination parameter.
-:::
-
-## Generalization: Inverse items {.smaller}
-
-::: notes
-Si recordáis la "función característica del ítem", esto consiste simplemente en un valor negativo en el parámetro $a$; eso hace que la probabilidad sea decreciente.
-
-Llevado al caso multidimensional, lo que nos podemos encontrar es con ítems que por ejemplo tengan todos los valores de discriminación negativos, o algunos negativos y otros positivos.
-Pensad por ejemplo que tenemos el problema de los "sesgos de respuesta".
-Vamos a imaginar que alguien estuviera respondiendo con un sesgo de "deseabilidad social" (intenta dar una buena imagen de sí mismo o de sí misma, de manera consciente o no).
-A un ítem como este, tendería a responder que no, dado ese sesgo de deseabilidad social, al margen de cómo autoevalúe su verdadero nivel de responsabilidad.
-Por lo tanto, tendríamos un ítem con dos valores negativos de discriminación, uno en la dimensión de responsabilidad, y otro en la dimensión de deseabilidad social.
-Sin embargo, consideremos ahora que estamos modelando un sesgo de "aquiescencia", que hace que la gente tienda a estar de acuerdo con todo, o a decir que sí a todo sin un criterio muy reflexivo ("¿Descuidas tus obligaciones?
-Mmmmm...
-sí"; "¿Atiendes tus obligaciones con diligencia?
-Mmmmm...
-sí").
-Cuanto más aquiescente sea una persona (a igual responsabilidad) mayor será su tendencia o probabilidad a responder afirmativamente al ítem.
-Sin embargo, cuanto más responsable sea una persona (a igual aquiescencia) menor será su tendencia (o probabilidad) a responder afirmativamente.
-En ese caso, tendríamos un ítem con un valor negativo de discriminación en la dimensión de responsabilidad, y uno positivo en la de aquiescencia.
-:::
-
-## Generalization: Inverse items
-
-<br>
-
+
+In the multidimensional case, this may imply a mix of positive, negative, and null discrimination parameters.
+:::
+
+## Constant monotony {.smaller}
+
+<br>
+
+::::::: columns
+::::: column
 $$
@@
 
-::: fragment
+::: {.fragment fragment-index="2"}
 $$
@@
 :::
-
-::: notes
-Si vemos la expresión de los parámetros multidimensionales, podemos prevér cómo van a ser en estos casos.
-Aquí he preferido representar la discriminación como suma de cuadrados, en lugar de como el producto matricial, pero es lo mismo, ¿os acordáis?
-El primer elemento de un vector por el primero del otro, y asi sucesivamente...
-todos sumados.
-Sólo que en este caso el primer y el segundo vector son el mismo, así que es la suma de los elementos del vector al cuadrado.
-Por lo tanto, la discriminación multidimensional siempre va a ser positiva.
-¿Lo veis claro?
-
-Por otro lado, la distancia (con signo), al depender sólo de la intersección y la discriminación multidimensional, nos da igual que los valores de la discriminación sean negativos.
-Siempre va a cumplir que su desplazamiento es "inverso" al valor indicado por la intersección, como indicábamos antes.
-
-La dirección sin embargo tiene una peculiaridad, y es que vemos que cada uno de los cosenos puede tener un signo distinto, que va a ser el mismo que el del valor de la discriminación para esa dimensión.
-Eso lo que va a hacer es que el sentido del item en cada una de esas dimensiones venga determinado por el signo de ese valor.
-:::
-
-## Generalization: Inverse items {.smaller}
-
-::::::::: columns
-::::::: column
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1 \\
-        1
-      \end{bmatrix} &
-    d_1 = 0
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_2 &=
-      \begin{bmatrix}
-         1 \\
-        -1
-      \end{bmatrix} &
-    d_2 = 0
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#749f4c;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_3 &=
-      \begin{bmatrix}
-        -1 \\
-        -1
-      \end{bmatrix} &
-    d_3 = 0
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#d76f47;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_4 &=
-      \begin{bmatrix}
-        -1 \\
-         1
-      \end{bmatrix} &
-    d_4 = 0
-  \end{matrix}
-$$
-:::
-:::::::
-
-::: column
+:::::
+
+::: {.column .fragment fragment-index="3"}
 ```{r inverse-item-representation}
@@
 :::
-:::::::::
+:::::::
 
 ::: notes
-Veamos un ejemplo: En el ítem 1 ambos cosenos son positivos, luego ambos ánculos son menores de 90º.
-El ítem está en el primer cuadrante.
-
-El segundo ítem va a tener un coseno positivo (ángulo menor que 90º) y uno negativo, por lo que será mayor que 90º; por tanto, apunta en el sentido negativo de ese eje.
-El ángulo es 180º menos el ángulo que tendría ese mismo coseno si fuera positivo.
-
-En el tercer ítem pasa lo mismo, pero con los dos cosenos negativos (y por tanto, ángulos mayores que 90º), y en el cuarto como el segundo pero al revés.
-
-Es decir, las propiedades algebraicas de la discriminación determinan las propiedades geométricas; y vemos cómo se puede visualizar gráficamente con facilidad.
-:::
-
-## Generalization: Inverse items {.smaller}
+The generalization is straightforward: The MDISC implies squared values of the discrimination components, so it is always positive and not affected by the sign of the discrimination components.
+
+\[NEXT\]
+
+The signed distance is also the same,...
+
+\[NEXT\]
+
+...and the direction only needs to take into account the sign of each component when computing the corresponding cosine.
+
+\[NEXT\]
+
+Graphically, we can see that each item points in the direction given by its discrimination components.
+In these case, all of them have an absolute value of 1, so the direction is given by the sign of the two components.
+:::
+
+## Constant monotony {.smaller}
+
+<br>
 
@@
 ::::::: column
-::: {.r-stack style="color:#5c6eb1;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1 \\
-        1
-      \end{bmatrix} &
-    d_1 = 0
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_2 &=
-      \begin{bmatrix}
-         1 \\
-        -1
-      \end{bmatrix} &
-    d_2 = -0.5
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#749f4c;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_3 &=
-      \begin{bmatrix}
-        -1 \\
-        -1
-      \end{bmatrix} &
-    d_3 = -0.5
-  \end{matrix}
-$$
-:::
-
-::: {.r-stack style="color:#d76f47;"}
-$$
-  \begin{matrix}
-    \mathbf{a}_4 &=
-      \begin{bmatrix}
-        -1 \\
-         1
-      \end{bmatrix} &
-    d_4 = 0.25
-  \end{matrix}
-$$
+$$
+  MDISC_i = \sqrt{a_{1i}^2 + ... + a_{ni}^2}
+$$
+
+$$
+  D_i = \frac{-d_i}{MDISC_i}
+$$
+
+$$
+  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
+$$
+
+<br>
+
+::: {style="color:#5c6eb1;"}
+$d_1 = 0$
+:::
+
+::: {style="color:#90214a;"}
+$d_2 = -0.5$
+:::
+
+::: {style="color:#749f4c;"}
+$d_3 = -0.25$
+:::
+
+::: {style="color:#d76f47;"}
+$d_4 = 0.25$
 :::
@@
   2,      1,   -1,   -0.5,
-  3,     -1,   -1,   -0.5,
+  3,     -1,   -1,   -0.25,
   4,     -1,    1,    0.25
@@
 ::: notes
-Si la intersección es distinta de cero, como decíamos antes: El desplazamiento del ítem es opuesto al sentido en el que apunta el ítem.
-
-Si la intersección es negativa, como en los ítems 2 y 3, el vector se desplaza "hacia adelante" en el sentido de la cabeza de la flecha.
-Es decir, el ítem es "más difícil en general, o menos atractivo", y la región donde es relativamente fácil acertar el ítem (o estar de acuerdo con él) disminuye.
-
-Si la intersección es negativa, como en el ítem 1, el vector se desplaza "hacia atrás"; o sea, el ítem es "en general más fácil o atractivo": La región donde es fácil acertar o estar de acuerdo con el ítem se amplía.
-
-Y da igual en qué dirección apunten los vectores; esto se cumple en todos los casos.
-
-Esta generalización es bastante evidente, y se puede aplicar sin problema a ítems no cognitivos.
-Sin embargo, hay que tener en cuenta que Reckase explicitaba como condición que los ítems tenían que ser "monotonamente crecientes".
+Also, with non-null intercepts, the distance is still proportional to its absolute value, and the direction is again given by its sign and the direction of the vector (forward for negative intercepts, backward for positive intercepts).
 :::

### 2025-07-10 output/IMPS-2025-slideshow.qmd
Update content styles to fit slides better

@@
 :::::: {.fragment fragment-index="1"}
-::: r-stack
-[$a_1$ = 1.0; $b_1$ = 0.0]{style="color:#5c6eb1;"}
-:::
-
-::: r-stack
-[$a_2$ = 1.0; $b_2$ = 1.5]{style="color:#90214a;"}
-:::
-
-::: r-stack
-[$a_3$ = 2.5; $b_3$ = 0.0]{style="color:#749f4c;"}
+::: {.r-stack style="color:#5c6eb1;"}
+$a_1$ = 1.0; $b_1$ = 0.0
+:::
+
+::: {.r-stack style="color:#90214a;"}
+$a_2$ = 1.0; $b_2$ = 1.5
+:::
+
+::: {.r-stack style="color:#749f4c;"}
+$a_3$ = 2.5; $b_3$ = 0.0
 :::
@@
 
-... that assumes a discrimination ($a_i$) and a difficulty (or location, $b_i$) for each item.
+...
+that assumes a discrimination ($a_i$) and a difficulty (or location, $b_i$) for each item.
 :::
@@
 
--   Increasing monotony:
+-   Increasing monotony (p. 402):
 
@@
 
----(p. 402)
-
--   Orthogonality
+-   Orthogonality (p. 404):
 
 > \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
-
----(p. 404)
 
@@
 
-## Assumptions: Increasing monotony
+## Increasing monotony
 
@@
 
-## Assumptions: Increasing monotony {.smaller}
+## Increasing monotony {.smaller}
 
@@
 
+::::: columns
+::: column
 > "*I neglect my duties*"
@@
 -- [International Personality Item Pool](https://ipip.ori.org/newAB5CKey.htm#Conscientiousness)
-
-<br>
-
-:::::: {.columns .fragment}
-::: {.column .r-center}
+:::
+
+::: {.column .fragment fragment-index="1"}
 ```{r inverse-item-curve}
@@
 :::
-
-:::: {.column .r-center}
+:::::
+
+::::: {.columns .fragment fragment-index="1"}
+::: column
 $$
@@
 $$
-
-<br>
-
-::: r-stack
-[$a_i < 0$]{style="color:#5c6eb1;"}
-:::
-::::
-::::::
+:::
+
+::: {.column style="color:#5c6eb1;"}
+$$
+  a_i < 0
+$$
+:::
+:::::
 

### 2025-07-10 output/IMPS-2025-slideshow.qmd
Add additional explanation to M2Pl model intro
Align item parameters closer to corresponding plot
Add "inverse items in the literature" slide

@@
 The Multidimensional 2PL model is one of many extensions of the 2PL model, which in this case is compensatory.
+
 It assumes that the probability of a correct response to an item is a function of the linear combination of the latent traits weighted by item discrimination parameters.
+(This here is a vector inner product; the "T" means that the discrimination vector is transposed).
+
+And it has this additional "intercept parameter"; so it is basically a logistic multivariate regression with latent predictors.
 
@@
 
+::: {.fragment .r-center}
+### Generalization: "Constant monotony"
+:::
+
 ::: notes
@@
 
-<br>
-
-::: {style="color:#5c6eb1;"}
+::: {style="color:#5c6eb1; padding-left:300px;"}
 $d_1 = 0$
@@
 
-::: {style="color:#90214a;"}
+::: {style="color:#90214a; padding-left:300px;"}
 $d_2 = -0.5$
@@
 
-::: {style="color:#749f4c;"}
+::: {style="color:#749f4c; padding-left:300px;"}
 $d_3 = -0.25$
@@
 
-::: {style="color:#d76f47;"}
+::: {style="color:#d76f47; padding-left:300px;"}
 $d_4 = 0.25$
@@
 Also, with non-null intercepts, the distance is still proportional to its absolute value, and the direction is again given by its sign and the direction of the vector (forward for negative intercepts, backward for positive intercepts).
-:::
-
-## "Inverse" items in the literature {.smaller}
+
+So, we can see that this generalization is quite straightforward.
+However, remember that Reckase explicitly constrained the parameter definitions to items that were "monotonically increasing", not simply "with constant monotony".k
+:::
+
+## Constant monotony {.smaller}
 
@@
 
-::: fragment
 > This result occurs because $a_1$ for the item is negative, and thus as an examinee's $\theta_1$ ability increases, the chance of answering the item correctly actually decreases.
@@
 ---@ackerman_multidimensional_2005-1 [p. 17]
-:::
 ::::
@@
 
-::: {.fragment .r-center}
-### Generalization: "Constant monotony"
-:::
-
 ::: aside
@@
 ::: notes
-Pero en la literatura nos encontramos casos como estos, que no lo son, y a los autores no se les ocurre decir que "hay que generalizar esas definiciones".
-Más bien, hacen comentarios como estos, restándole importancia al hecho de que los parámetros de discriminación sean negativos ("particularmente misterioso"),...
-
-\[NEXT\]
-
-o afirmando lo obvio (el parámetro de discriminación del ítem es negativo para una dimensión, por lo que la probabilidad de respuesta correcta disminuye al incrementarse su habilidad)
-
-\[NEXT\]
-
-En lugar de eso, proponemos que los supuestos de Reckase se "relajen" para considerar ítems con probabilidad "con monotonía constante" (es decir, que no tengan un "pico" u óptimo de rasgo latente para dar una respuesta positiva y luego decrezca), sino que cuanto más extremo sea el nivel de rasgo, más extrema es también la probabilidad (positiva o negativa).
-Este caso general incluye también la probabilidad constante (ya que es monotonía "nula constante", es decir, "el cambio es nulo de manera constante").
-Este tipo de ítems, positivos, negativos, o nulos, se pueden acomodar y modelar fácilmente por un modelo como el logístico de 2 parámetros multidimensional, como hemos visto.
-Ítems que estarían excluidos, por no tener monotonía constante, podrían ser formulaciones como, por ejemplo, "Soy tan extrovertido como la mayoría de mis amigos" (alguien puede decir, "No, soy más introvertido que todos ellos", respuesta negativa con rasgo bajo, o "Qué va, soy el más extrovertido con diferencia", respuesta negativa con rasgo alto).
-
-Bien, esto por lo que respecta a los ítems que denominamos "inversos" (o que tienen parámetros de discrimiinación / pesos factoriales inversos).
-Como veis, esta generalización es necesaria cuando hablamos de medición en el dominio "no cognitivos"; en el cognitivo, ya vimos que no tenía sentido, y de ahí que hiciera falta esta matización.
+However, it is not uncommon to find cases like these in the literature, where the items are not monotonically increasing, and the authors do not suggest to "generalize these definitions".
+
+On the contrary, what we find are these kind of downplaying comments: "particularly puzzling", or "as an examinee’s ability increases, the chance of answering the item correctly actually decreases" (asserting the obvious).
+
+That is why we propose to relax the "increasing monotony assumption", considering items with "constant monotony" instead.
+This includes all items that do not have a latent trait "optimum" to give a positive response, and then decrease in both directions, but rather that the more extreme the trait level, the more extreme the probability (positive or negative) as well, in either direction.
+And this general case also includes the "constant probability" case (as "constant probability" implies no change, so "constantly null monotony").
+
+This regarding so-called "inverse" items.
+Quite straightforward.
+Now, to something also necessary, but not so straightforward: The orthogonality assumption.
 :::

### 2025-07-10 output/IMPS-2025-slideshow.qmd
Drop "inverse items in literature" slide
It is not necessary and there will be no time for it.

@@
 
-## Constant monotony {.smaller}
-
-:::::: columns
-:::: {.column width="62%"}
-> \[...\] the negative relation of Item 15 with geometry achievement is particularly puzzling.
-
----@mcdonald_basis_2000 [p. 109]
-
-> This result occurs because $a_1$ for the item is negative, and thus as an examinee's $\theta_1$ ability increases, the chance of answering the item correctly actually decreases.
-
----@ackerman_multidimensional_2005-1 [p. 17]
-::::
-
-::: {.column width="38%"}
-<!-- TODO: Annotate figure? -->
-
-!["ACT Mathematics Usage Test" items](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="left"}
-:::
-::::::
-
-::: aside
-Figure taken from @ackerman_multidimensional_2005-1
-:::
-
-::: notes
-However, it is not uncommon to find cases like these in the literature, where the items are not monotonically increasing, and the authors do not suggest to "generalize these definitions".
-
-On the contrary, what we find are these kind of downplaying comments: "particularly puzzling", or "as an examinee’s ability increases, the chance of answering the item correctly actually decreases" (asserting the obvious).
-
-That is why we propose to relax the "increasing monotony assumption", considering items with "constant monotony" instead.
-This includes all items that do not have a latent trait "optimum" to give a positive response, and then decrease in both directions, but rather that the more extreme the trait level, the more extreme the probability (positive or negative) as well, in either direction.
-And this general case also includes the "constant probability" case (as "constant probability" implies no change, so "constantly null monotony").
-
-This regarding so-called "inverse" items.
-Quite straightforward.
-Now, to something also necessary, but not so straightforward: The orthogonality assumption.
-:::
-
 ## Assumptions: Orthogonality {.smaller}

### 2025-07-10 output/IMPS-2025-slideshow.qmd
Improve speaker notes to cut down clutter

@@
 ::: notes
-The Multidimensional 2PL model is one of many extensions of the 2PL model, which in this case is compensatory.
+The Multidimensional 2PL model is one of many extensions of the 2PL model.
 
@@
 ::: notes
-The problem with the M2PL, and other multidimensional extensions, is that its parameters do not have a clear interpretation in IRT terms, and they cannot be reparameterized.
+The problem with the M2PL is that its parameters do not have a clear interpretation in IRT terms.
 
@@
 
-He do it in this 1985 paper, and in a subsequent one in 1991...
+He does it in this 1985 paper, and in a subsequent one in 1991...
 :::
@@
 
-Note also that the discrimination can in turn be used to define the item difficulty, as happens in the unidimensional model.
-
-And also, that the difficulty has two components: a "signed distance" from the origin, and a direction.
-
-\[NEXT\]
-
-It is easier to appreciate this with the graphical representation, that was proposed by Ackerman a few years later, where we can see that the item vector is applied at a point given by the difficulty, and has a length equal to the discrimination.
-Note that the item direction always goes through the origin, and that the sign of the distance (positive in this case) will determine whether the vector is displaced forward or backward along that direction.
+This discrimination can be used to define the item difficulty, finding a parameterization that parallels the unidimensional 2PL model.
+
+Note though that, in this case, the difficulty has two components: a "signed distance" from the origin, and a direction.
+
+\[NEXT\]
+
+It is easier to appreciate this with the graphical representation, that was proposed by Ackerman a few years later, where we can see that the item vector is applied at a point given by the difficulty (with its distance and direction), and has a length equal to the discrimination.
+The vector direction always goes through the origin, and the sign of the distance (positive in this case), which is the opposite of the intercept sign, will determine whether the vector is displaced forward or backward along that direction.
 :::
@@
 ::: notes
-However, there are two assumptions that Reckase & McKinley make when obtaining the definition of the multidimensional parameters.
-
-The first is that the probability of answering an item correctly always increases monotonically with respect to each dimension.
-
-The second, that constrain that you see there, implies that the item is represented in an orthogonal basis, which is why the discrimination is defined by a "general Pythagorean theorem".
-
-So, our purpose in this work is **to relax these two assumptions**.
+However, there are two assumptions for defining the multidimensional parameters.
+
+First, that the probability of a correct answer increases monotonically with each dimension.
+
+And second, this constrain you see there, which implies that the item is represented in an orthogonal basis. This is necessary for applying the "general Pythagorean theorem" as we saw before.
+
 :::
@@
 Nevertheless, with this caveat about null discrimination components, this assumption would be realistic in a "cognitive testing context": The more you know, the more likely you are to give a correct answer.
+These assumptions derive from the traditional application of the model to the cognitive domain, but we want to apply multidimensional IRT to the non-cognitive domain as well, e.g.
+to personality measures.
+So, our purpose is **to relax these two assumptions**.
 :::

### 2025-07-10 output/IMPS-2025-slideshow.qmd
Drop "unidimensional items" slide
Not necessary to justify the "inverse items" generalization

@@
 
-:::
-
-## Increasing monotony
-
-<br>
-
-::::: columns
-::: column
-<br>
-
-<!-- TODO: Same color as item? -->
-
-$$
-  \begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1.5 \\
-        0
-      \end{bmatrix} \\
-      \\
-      \\
-    d_1 &= 0.75
-  \end{align}
-$$
-:::
-
-::: column
-```{r m2pl-representation-unidimensional}
-#| fig-align:  center
-#| fig-height: 7.2
-#| fig-width:  7.2
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1.5,  0,    0.75
-)
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1, 1),
-    breaks       = NULL,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .4),
-    axis.title.y.right = element_markdown(vjust = .4),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::
-
-::: notes
-In the first case, one may argue that we may deal with "undimensional items" or, more generally, items that have null discrimination components.
-
-This would imply that the probability of a correct response would not change with respect to that dimension; i.e., that trait is not being measured by the item, so the isoprobability lines would be parallel to the $\theta_2$ axis (in this example.
-
-Therefore, the probability does not "increase monotonically".
-
-Nevertheless, with this caveat about null discrimination components, this assumption would be realistic in a "cognitive testing context": The more you know, the more likely you are to give a correct answer.
 These assumptions derive from the traditional application of the model to the cognitive domain, but we want to apply multidimensional IRT to the non-cognitive domain as well, e.g.

### 2025-07-10 output/IMPS-2025-slideshow.qmd
Cutt down (more) clutter in speaker notes

@@
 
-These assumptions derive from the traditional application of the model to the cognitive domain, but we want to apply multidimensional IRT to the non-cognitive domain as well, e.g.
-to personality measures.
 So, our purpose is **to relax these two assumptions**.
@@
 
-::::: {.columns .fragment fragment-index="1"}
+::::: {.columns .fragment fragment-index="2"}
 ::: column
@@
 ::: notes
-However, if we wanted to apply this model to "non-cognitive" items (e.g., personality traits) where the person evaluated can be "agreeing" or "disagreeing", we can have items with "decreasing" probability.
+
+Because they derive from the traditional application of the model to the cognitive domain, but we want to apply multidimensional IRT to non-cognitive items as well, e.g. to personality measures.
+
+\[NEXT\]
+
+Here, we can have items with "decreasing" probability.
 
@@
 
-In the multidimensional case, this may imply a mix of positive, negative, and null discrimination parameters.
+In the multidimensional case, this may imply a mix of positive and negative discrimination parameters, even in the same item.
 :::
@@
 
-::: {.fragment fragment-index="1"}
+::: fragment
 $$
@@
 
-::: {.fragment fragment-index="2"}
+::: fragment
 $$
@@
 
-::: {.column .fragment fragment-index="3"}
+::: {.column .fragment}
 ```{r inverse-item-representation}
@@
 ::: notes
-The generalization is straightforward: The MDISC implies squared values of the discrimination components, so it is always positive and not affected by the sign of the discrimination components.
+The generalization is straightforward: The MDISC implies squared values of the discrimination components (this is the inner product expanded), so it is always positive regardless of the sign of the discrimination components.
 
@@
 
-...and the direction only needs to take into account the sign of each component when computing the corresponding cosine.
-
-\[NEXT\]
-
-Graphically, we can see that each item points in the direction given by its discrimination components.
-In these case, all of them have an absolute value of 1, so the direction is given by the sign of the two components.
+...and the direction only needs to take into account the sign of each component when computing that corresponding cosine.
+
+\[NEXT\]
+
+Graphically, we can see that each item points in a direction given by its discrimination components, including their signs.
 :::
@@
 ::: notes
-Also, with non-null intercepts, the distance is still proportional to its absolute value, and the direction is again given by its sign and the direction of the vector (forward for negative intercepts, backward for positive intercepts).
-
-So, we can see that this generalization is quite straightforward.
-However, remember that Reckase explicitly constrained the parameter definitions to items that were "monotonically increasing", not simply "with constant monotony".k
+And the displacement from the origin is again given by the sign of the signed distance (so, opposite to the intercept sign).
 :::

### 2025-07-11 output/IMPS-2025-slideshow.qmd
Drop introductory unidimensional IRT slide
Need to cut down slides to save time

@@
 
-## Unidimensional IRT {.smaller}
-
--   2-parameter logistic (2PL) model [@birnbaum_latent_1968]:
-
-::::::::: columns
-::::::: {.column width="40%"}
-$$
-  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
-$$
-
-$$
-  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
-$$
-
-<br>
-
-:::::: {.fragment fragment-index="1"}
-::: {.r-stack style="color:#5c6eb1;"}
-$a_1$ = 1.0; $b_1$ = 0.0
-:::
-
-::: {.r-stack style="color:#90214a;"}
-$a_2$ = 1.0; $b_2$ = 1.5
-:::
-
-::: {.r-stack style="color:#749f4c;"}
-$a_3$ = 2.5; $b_3$ = 0.0
-:::
-::::::
-:::::::
-
-::: {.column .fragment width="60%" fragment-index="1"}
-```{r irt-curves}
-#| fig-asp:   0.6
-#| fig-width: 8
-
-ITEMS <- tribble(
-  ~item, ~a,   ~b,
-   1,     1,    0,
-   3,     1,    1.5,
-   2,     2.5,  0
-)
-
-logit_curve_data <- tibble(
-  latent_trait = PROB_AXIS,
-  prob         = ITEMS |>
-    select(-item) |>
-    pmap(irf, theta = latent_trait) |>
-    bind_cols()
-) |>
-  unnest(
-    prob,
-    names_sep = '_'
-  ) |>
-  rename_with(str_remove, pattern = '\\.{3}')
-
-logit_curve_data_long <- logit_curve_data |>
-  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob") |>
-  mutate(item = item |> str_remove("prob_"))
-
-logit_curve_data_long |>
-  ggplot(
-    mapping = aes(
-      latent_trait, prob,
-      color = item,
-      group = after_scale(color)
-    )
-  )                                        +
-  geom_line(linewidth = LINE_WIDTH)        +
-  labs(x = "*&theta;*", y = "P(Y~i~ = 1)") +
-  scale_y_continuous(
-    breaks       = 0:1,
-    minor_breaks = NULL,
-    limits       = 0:1,
-    expand       = expansion(mult = c(0, .05))
-  )                                        +
-  scale_x_continuous(
-    breaks       = X_BREAKS,
-    minor_breaks = NULL,
-    limits = range(PROB_AXIS),
-    expand = expansion()
-  )                                        +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  theme(
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(angle = 90),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
-  )
-```
-
-<!-- TODO: Como ecuaciones (sin punto y coma?) -->
-:::
-:::::::::
-
-::: {.r-stack .fragment fragment-index="1"}
-$a_i$ = discrimination
-:::
-
-::: {.r-stack .fragment fragment-index="1"}
-$b_i$ = difficulty (location)
-:::
-
-::: notes
-The 2-parameter logistic model is the most simple model in item response theory (IRT).
-
-It is a prototypical model that allows us to study the item properties according to a parametric conception...
-
-\[NEXT\]
-
-...
-that assumes a discrimination ($a_i$) and a difficulty (or location, $b_i$) for each item.
-:::
-
 ## Multidimensional IRT {.smaller}

### 2025-07-11 output/IMPS-2025-slideshow.qmd
Compress "constant monotony" in 1 slide

@@
 
-And second, this constrain you see there, which implies that the item is represented in an orthogonal basis. This is necessary for applying the "general Pythagorean theorem" as we saw before.
+And second, this constrain you see there, which implies that the item is represented in an orthogonal basis.
+This is necessary for applying the "general Pythagorean theorem" as we saw before.
 
@@
 ::: notes
-
 Because they derive from the traditional application of the model to the cognitive domain, but we want to apply multidimensional IRT to non-cognitive items as well, e.g. to personality measures.
@@
 
-::::::: columns
-::::: column
+:::::::::: columns
+:::::::: column
 $$
@@
 
-::: fragment
 $$
@@
 $$
-:::
-
-::: fragment
+
 $$
@@
 $$
-:::
-:::::
-
-::: {.column .fragment}
-```{r inverse-item-representation}
+
+::::::: {.fragment fragment-index="1"}
+::: {style="color:#5c6eb1; padding-left:300px;"}
+$d_1 = 0$
+:::
+
+::: {style="color:#90214a; padding-left:300px;"}
+$d_2 = -0.5$
+:::
+
+::: {style="color:#749f4c; padding-left:300px;"}
+$d_3 = -0.25$
+:::
+
+::: {style="color:#d76f47; padding-left:300px;"}
+$d_4 = 0.25$
+:::
+:::::::
+::::::::
+
+::: {.column .fragment fragment-index="1"}
+```{r inverse-item-representation-distance}
 #| fig-align:  center
@@
   1,      1,    1,    0,
-  2,      1,   -1,   0,
-  3,     -1,   -1,   0,
-  4,     -1,    1,   0
+  2,      1,   -1,   -0.5,
+  3,     -1,   -1,   -0.25,
+  4,     -1,    1,    0.25
 )
@@
 :::
-:::::::
+::::::::::
 
@@
 
-\[NEXT\]
-
-The signed distance is also the same,...
-
-\[NEXT\]
-
-...and the direction only needs to take into account the sign of each component when computing that corresponding cosine.
-
-\[NEXT\]
-
-Graphically, we can see that each item points in a direction given by its discrimination components, including their signs.
-:::
-
-## Constant monotony {.smaller}
-
-<br>
-
-::::::::: columns
-::::::: column
-$$
-  MDISC_i = \sqrt{a_{1i}^2 + ... + a_{ni}^2}
-$$
-
-$$
-  D_i = \frac{-d_i}{MDISC_i}
-$$
-
-$$
-  \cos \, \mathbf{\alpha}_i = \frac{\mathbf{a}_i}{MDISC_i}
-$$
-
-::: {style="color:#5c6eb1; padding-left:300px;"}
-$d_1 = 0$
-:::
-
-::: {style="color:#90214a; padding-left:300px;"}
-$d_2 = -0.5$
-:::
-
-::: {style="color:#749f4c; padding-left:300px;"}
-$d_3 = -0.25$
-:::
-
-::: {style="color:#d76f47; padding-left:300px;"}
-$d_4 = 0.25$
-:::
-:::::::
-
-::: column
-```{r inverse-item-representation-distance}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,      1,    1,    0,
-  2,      1,   -1,   -0.5,
-  3,     -1,   -1,   -0.25,
-  4,     -1,    1,    0.25
-)
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-  rename_with(str_remove, pattern = "transf_")
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-1.5, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-1.5, 1.5),
-    breaks       = -1:2,
-    minor_breaks = NULL,
-    name         = NULL,
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title        = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = .5),
-    axis.title.y.right = element_markdown(vjust = .5),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    ),
-  )
-```
-:::
-:::::::::
-
-::: notes
-And the displacement from the origin is again given by the sign of the signed distance (so, opposite to the intercept sign).
+The signed distance is also the same, and the direction only needs to take into account the sign of each component when computing that corresponding cosine.
+
+\[NEXT\]
+
+Graphically, we can see that each item points in a direction given by its discrimination components, including their signs (so, 1 1, 1 -1, etc.).
+
+And the displacement from the origin is again with respect to the pointing direction.
 :::

### 2025-07-11 output/IMPS-2025-slideshow.qmd
Add final annotations to "constant monotony" part

@@
   markdown: 
-    wrap:      sentence # TODO: Restore
+    wrap:      sentence
     canonical: true # TODO: Restore
@@
 And the displacement from the origin is again with respect to the pointing direction.
+
+This seems obvious enough, but we must explicitly state that this definitions are valid in these cases, and that they have the same interpretation.
+
+\[PAUSE\]
+
+Now, to the "not so obvious"...
 :::

### 2025-07-11 output/sandbox_slideshow.qmd
Add item color palette to the preface
Add "orthogonality" assumption slides (WiP)

@@
 
+## Orthogonality {.smaller}
+
+<br>
+
+:::::: columns
+::: column
+```{r item-oblique-axes-1}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     0.5,  1,    0
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+item_spikes <- mirt_item_coords        |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*^*^~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:2)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*^*^~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(0.5, -0.1), y = c(-0.1, 1),
+    label         = c("*a*^*^~11~", "*a*^*^~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1/3),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+:::: column
+$$
+  \begin{align}
+    \mathbf{a}^*_1 &=
+      \begin{bmatrix}
+        0.5 \\
+        1
+      \end{bmatrix}\\
+      d^*_1 &= 0
+  \end{align}
+$$
+
+<br>
+
+::: fragment
+$$
+  \begin{align}
+    MDISC^*_1 &= \sqrt{
+        [0.5, 1]
+        \begin{bmatrix}
+          0.5 \\
+          1
+        \end{bmatrix}
+      } \\
+      &= \sqrt{ 0.5 · 0.5 + 1 · 1 } \\
+      &= \sqrt{ 0.25 + 1 } \\
+      &= \sqrt{ 1.25 } \approx 1.12
+  \end{align}
+$$
+:::
+::::
+::::::
+
+::: notes
+You'll recall that we needed orthogonal axes to apply Reckase's formulas.
+
+So let's say we have these "theta star" axes, ...
+
+\[NEXT\]
+
+...
+where we have this item, with discrimination (.5, 1), and we can compute its MDISC.
+
+We may want to transform these axes to another coordinate system, for example, applying an oblique rotation...
+:::
+
+## Orthogonality {.smaller}
+
+<br>
+
+::::: columns
+::: {.column width="60%"}
+```{r item-oblique-axes-2}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+BASIS_VEC_1 <- c(1, 0)
+BASIS_VEC_2 <- c(-.5, 1)
+
+axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
+transf_matrix <- cbind(BASIS_VEC_1, BASIS_VEC_2) |> unname()
+
+item_coords_transf <- mirt_item_coords |>
+  pivot_longer(-item) |>
+  separate(name, into = c("coord", "dim")) |>
+  group_by(coord) |>
+  mutate(
+    dim   = paste0('F', dim),
+    value = solve(transf_matrix) %*% value |> drop()
+  ) |>
+  ungroup() |>
+  pivot_wider(names_from = dim)
+
+item_spikes_transf <- item_coords_transf         |>
+  select(-item)                                  |>
+  filter(coord == "end")                         |>
+  bind_rows(tibble(F1 = 0, F2 = 0))              |>
+  complete(F1, F2, fill = list(coord = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+item_spikes_transf <- item_spikes_transf               |>
+  bind_rows(item_spikes_transf |> slice(3))            |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+  pivot_wider(names_from = coord, values_from = F1:F2)
+
+item_spikes_transf_ob <- item_spikes_transf |>
+  pivot_longer(starts_with("F")) |>
+  separate(name, into = c("dim", "coord")) |>
+  pivot_wider(names_from = coord, values_from = value) |>
+  group_by(n) |>
+  mutate(
+    start = (transf_matrix %*% start |> drop()),
+    end   = (transf_matrix %*% end   |> drop())
+  ) |>
+  pivot_wider(names_from = dim, values_from = start:end)
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  ) +
+  geom_abline(
+    intercept = -2:3,
+    slope     = -2,
+    linewidth = GRID_WIDTH,
+    linetype  = "17",
+    color     = AXIS_COLOR
+  ) +
+  geom_abline(slope = -2, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0,  linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes_transf_ob,
+    mapping = aes(start_F1, start_F2, xend = end_F1, yend = end_F2),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-2:2)/2 + .375,
+    labels       = ~number(. - .375, .1),
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:2)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(1, -0.59), y = c(-0.07, 1),
+    label         = c("*a*~11~", "*a*~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 0),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major.y = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+::: {.column .fragment width="40%"}
+<br> <br>
+
+> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
+
+--- @reckase_difficulty_1985 [p. 404]
+:::
+:::::
+
+::: notes
+...to obtain these "theta axes" which, let's say, align with our measured personality traits.
+We have now different discrimination components than in the "theta star" axes.
+
+\[NEXT\]
+
+If we applied here Reckase's restriction...
+:::
+
+## Orthogonality {.smaller}
+
+<br>
+
+::::: columns
+::: column
+```{r item-oblique-axes-4}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+mirt_item_transf <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    1,    0
+)
+
+mirt_item_params <- mirt_item_transf |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+item_spikes <- mirt_item_coords        |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:2)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(1, -0.1), y = c(-0.1, 1),
+    label         = c("*a*~11~", "*a*~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1/3),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+::: column
+$$
+  \begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix}\\
+      d_1 &= 0
+  \end{align}
+$$
+
+<br>
+
+$$
+  \begin{align}
+    MDISC_1 &= \sqrt{
+        [1, 1]
+        \begin{bmatrix}
+          1 \\
+          1
+        \end{bmatrix}
+      } \\
+      &= \sqrt{ 1 · 1 + 1 · 1 } \\
+      &= \sqrt{ 1 + 1 } \\
+      &= \sqrt{ 2 } \approx 1.41
+  \end{align}
+$$
+:::
+:::::
+
+::: notes
+...we would be assuming that the axes are actually orthogonal, which would imply that the MDISC is not invariant.
+Also, you can see intuitively that it does not seem right to assume that both coordinate systems are orthogonal, ¿right?
+:::
+
+## Orthogonality {.smaller}
+
+<br>
+
+::::: columns
+::: column
+```{r item-oblique-axes-2}
+```
+:::
+
+::: column
+<br> <br>
+
+> ~~\[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]~~
+
+--- @reckase_difficulty_1985 [p. 404]
+:::
+:::::
+
+::: notes
+:::
+
 ## References

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add "orthogonality" assumption slides

@@
 
+## Orthogonality {.smaller}
+
+<br>
+
+:::::: columns
+::: column
+```{r item-oblique-axes-1}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+mirt_item <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     0.5,  1,    0
+)
+
+mirt_item_params <- mirt_item |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords <- mirt_item_params |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+item_spikes <- mirt_item_coords        |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*^*^~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:2)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*^*^~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(0.5, -0.1), y = c(-0.1, 1),
+    label         = c("*a*^*^~11~", "*a*^*^~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1/3),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+:::: column
+$$
+  \begin{align}
+    \mathbf{a}^*_1 &=
+      \begin{bmatrix}
+        0.5 \\
+        1
+      \end{bmatrix}\\
+      d^*_1 &= 0
+  \end{align}
+$$
+
+<br>
+
+::: fragment
+$$
+  \begin{align}
+    MDISC^*_1 &= \sqrt{
+        [0.5, 1]
+        \begin{bmatrix}
+          0.5 \\
+          1
+        \end{bmatrix}
+      } \\
+      &= \sqrt{ 0.5 · 0.5 + 1 · 1 } \\
+      &= \sqrt{ 0.25 + 1 } \\
+      &= \sqrt{ 1.25 } \approx 1.12
+  \end{align}
+$$
+:::
+::::
+::::::
+
+::: notes
+You'll recall that we needed orthogonal axes to apply Reckase's formulas.
+
+So let's say we have these "theta star" axes,...
+
+\[NEXT\]
+
+...where we have this item, with discrimination (.5, 1) in this coordinate system, where we may compute its MDISC.
+
+We can transform these axes to another coordinate system, for example, applying an oblique rotation...
+:::
+
+## Orthogonality {.smaller}
+
+<br>
+
+::::: columns
+::: column
+```{r item-oblique-axes-2}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+BASIS_VEC_1 <- c(1, 0)
+BASIS_VEC_2 <- c(-.5, 1)
+
+axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
+transf_matrix <- cbind(BASIS_VEC_1, BASIS_VEC_2) |> unname()
+
+item_coords_transf <- mirt_item_coords |>
+  pivot_longer(-item) |>
+  separate(name, into = c("coord", "dim")) |>
+  group_by(coord) |>
+  mutate(
+    dim   = paste0('F', dim),
+    value = solve(transf_matrix) %*% value |> drop()
+  ) |>
+  ungroup() |>
+  pivot_wider(names_from = dim)
+
+item_spikes_transf <- item_coords_transf         |>
+  select(-item)                                  |>
+  filter(coord == "end")                         |>
+  bind_rows(tibble(F1 = 0, F2 = 0))              |>
+  complete(F1, F2, fill = list(coord = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+item_spikes_transf <- item_spikes_transf               |>
+  bind_rows(item_spikes_transf |> slice(3))            |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+  pivot_wider(names_from = coord, values_from = F1:F2)
+
+item_spikes_transf_ob <- item_spikes_transf |>
+  pivot_longer(starts_with("F")) |>
+  separate(name, into = c("dim", "coord")) |>
+  pivot_wider(names_from = coord, values_from = value) |>
+  group_by(n) |>
+  mutate(
+    start = (transf_matrix %*% start |> drop()),
+    end   = (transf_matrix %*% end   |> drop())
+  ) |>
+  pivot_wider(names_from = dim, values_from = start:end)
+
+mirt_item_coords |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  ) +
+  geom_abline(
+    intercept = -2:3,
+    slope     = -2,
+    linewidth = GRID_WIDTH,
+    linetype  = "17",
+    color     = AXIS_COLOR
+  ) +
+  geom_abline(slope = -2, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0,  linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes_transf_ob,
+    mapping = aes(start_F1, start_F2, xend = end_F1, yend = end_F2),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-2:2)/2 + .375,
+    labels       = ~number(. - .375, .1),
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:2)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(1, -0.59), y = c(-0.07, 1),
+    label         = c("*a*~11~", "*a*~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 0),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major.y = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+::: {.column .fragment}
+<br> <br>
+
+> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
+
+--- @reckase_difficulty_1985 [p. 404]
+:::
+:::::
+
+::: notes
+...to obtain these "theta axes" (let's say, because they align with personality traits of our interest).
+We have now different discrimination components than in the "theta star" axes.
+
+\[NEXT\]
+
+If we applied here Reckase's restriction...
+:::
+
+## Orthogonality {.smaller}
+
+<br>
+
+::::: columns
+::: column
+```{r item-oblique-axes-4}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+mirt_item_transf <- tribble(
+  ~item, ~a_1, ~a_2, ~d,
+  1,     1,    1,    0
+)
+
+mirt_item_params_transf <- mirt_item_transf |>
+  compute_mirt_params(d, starts_with('a'))
+
+mirt_item_coords_transf <- mirt_item_params_transf |>
+  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
+    rename_with(str_remove, pattern = "transf_")
+
+item_spikes <- mirt_item_coords_transf |>
+  slice(1 |> rep(2))                   |>
+  add_column(dim = 'F' |> paste0(1:2)) |>
+  mutate(
+    origin_1 = if_else(dim == 'F1', end_1, origin_1),
+    origin_2 = if_else(dim == 'F2', end_2, origin_2),
+  )
+
+mirt_item_coords_transf |>
+  ggplot(
+    aes(
+      origin_1, origin_2,
+      xend = end_1,
+      yend = end_2,
+      color = factor(item), fill = factor(item)
+    ),
+  )                                                  +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_segment(
+    data = item_spikes,
+    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color     = PALETTE_BLUE["light"]
+  ) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH
+  ) +
+  scale_x_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:3)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    limits       = c(-0.75, 1.5),
+    breaks       = (-1:2)/2,
+    minor_breaks = NULL,
+    name         = NULL,
+    expand       = expansion(),
+    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
+  annotate(
+    "richtext",
+    x = c(1, -0.1), y = c(-0.1, 1),
+    label         = c("*a*~11~", "*a*~21~"),
+    size          = 8,
+    label.colour  = NA,
+    label.padding = unit(0, "lines"),
+    fill = NA
+  ) +
+  coord_fixed(clip = "on") +
+  theme(
+    axis.line          = element_blank(),
+    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
+    axis.title.x       = element_markdown(hjust = 1/3),
+    axis.title.y.right = element_markdown(vjust = 1/3),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+
+::: column
+$$
+  \begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        1
+      \end{bmatrix}\\
+      d_1 &= 0
+  \end{align}
+$$
+
+<br>
+
+$$
+  \begin{align}
+    MDISC_1 &= \sqrt{
+        [1, 1]
+        \begin{bmatrix}
+          1 \\
+          1
+        \end{bmatrix}
+      } \\
+      &= \sqrt{ 1 · 1 + 1 · 1 } \\
+      &= \sqrt{ 1 + 1 } \\
+      &= \sqrt{ 2 } \approx 1.41
+  \end{align}
+$$
+:::
+:::::
+
+::: notes
+...we would be assuming that the axes are actually orthogonal, implying that the MDISC is not invariant.
+
+As you can see, does not seem intuitively right to assume that both coordinate systems are orthogonal, ¿right?
+:::
+
+## Orthogonality {.smaller}
+
+<br>
+
+::::: columns
+::: column
+```{r item-oblique-axes-2}
+```
+:::
+
+::: column
+<br> <br>
+
+> ~~\[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]~~
+
+--- @reckase_difficulty_1985 [p. 404]
+:::
+:::::
+
+::: notes
+What we actually need is to compute the multidimensional parameters in this general space, which is not necessarily orthogonal (or, we should say more generally, not necessarily orthonormal).
+:::
+
 ## Assumptions: Orthogonality {.smaller}

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Delete old "orthogonality assumption" slides

@@
 
-## Assumptions: Orthogonality {.smaller}
-
-> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
-
----@reckase_difficulty_1985 [p. 404]
-
-::::: columns
-::: {.column .fragment}
-```{r oblique-rotation-axis-1}
-LIM_INF_X <- -2.7
-LIM_INF_Y <- -0.5
-LIM_SUP   <-  2.7
-
-axis_lims_x <- c(LIM_INF_X, LIM_SUP)
-axis_lims_y <- c(LIM_INF_Y, LIM_SUP)
-
-point <- tibble(F1 = .1, F2 = .3)
-
-point_spikes <- point      |>
-  add_column(type = "end")                      |> 
-  bind_rows(tibble(F1 = 0, F2 = 0))             |>
-  complete(F1, F2, fill = list(type = "start")) |>
-  filter(F1 != 0 | F2 != 0)
-
-point_spikes <- point_spikes                           |> 
-  bind_rows(point_spikes |> slice(3))                  |>
-  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
-  pivot_wider(names_from = type, values_from = F1:F2)
-
-point |>
-  ggplot(mapping = aes(F1, F2))            +
-  geom_segment(
-    data    = point_spikes,
-    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-  )                                        +
-  geom_point(size = I(4))                  +
-  scale_x_continuous(
-    limits = axis_lims_x,
-    breaks = 0,
-    labels = NULL,
-    name   = NULL
-  )                                        +
-  scale_y_continuous(
-    limits = axis_lims_y,
-    breaks = 0,
-    labels = NULL,
-    name   = NULL
-  )                                        +
-  coord_fixed(expand = FALSE, clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    )
-  )
-```
-:::
-
-::: column
-:::
-:::::
-
-::: notes
-Ahora, la TRI multidimensional también dijimos que se interesaba sobre todo por medir (y ordenar) a las personas, no tanto por interpretar sustantivamente esas dimensiones de rasgo latente.
-De ahí que estimar un modelo con suficientes dimensiones para explicar la variabilidad entre los sujetos fuese importante, pero no era tan importante cuáles fueran esas dimensiones.
-Por eso, hay un supuesto, implícito muchas veces en TRI, de que las dimensiones no están correlacionadas: Se estima el modelo con N dimensiones, las que haga falta, añadiendo restricciones a los ítems para identificarlas, y eso da lugar a N dimensiones independientes entre sí (con correlaciones de 0), representadas en N ejes ortogonales (ángulos rectos todos ellos).
-
-\[NEXT\]
-
-Que esos ejes representen competencia o habilidad en esto o lo otro, en principio, nos da igual.
-Con asegurarnos de que hay suficientes dimensiones para "capturar" toda la variabilidad debida a esos rasgos nos basta.
-
-Pues bien, ese es el otro supuesto que hace Reckase para calcular los parámetros multidimensionales: Los cosenos directores del ítem al cuadrado suman 1.
-Es decir, está aplicando el teorema de Pitágoras.
-Esta expresión implica, de manera implícita, que todos los ejes del espacio latente son "ortogonales" entre sí.
-Como digo, en TRI originalmente no importaba mucho que los ejes se pudieran interpretar como rasgos latentes.
-No obstante, en 1999 llega McDonald, y dice que la TRI y el AF son fundamentalmente la misma cosa.
-:::
-
-## Assumptions: Orthogonality {.smaller}
-
-<!-- TODO: Equation as LaTeX -->
-
-> ~~\[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]~~
-
----@reckase_difficulty_1985 [p. 404]
-
-::::: columns
-::: column
-<!-- TODO: Animate? -->
-
-```{r oblique-rotation-axis-2}
-SEC_COLOR <- "grey60"
-
-BASIS_VEC_1 <- c(1, 0)
-BASIS_VEC_2 <- c(-8, 1)
-BASIS_VEC_2 <- BASIS_VEC_2 / sqrt(sum(BASIS_VEC_2^2))
-
-axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
-transf_matrix <- cbind(BASIS_VEC_1, BASIS_VEC_2) |> unname()
-
-point_transf <- point |>
-  pivot_longer(everything()) |>
-  mutate(value = solve(transf_matrix) %*% value |> drop()) |>
-  pivot_wider()
-
-point_spikes_transf <- point_transf             |>
-  add_column(type = "end")                      |>
-  bind_rows(tibble(F1 = 0, F2 = 0))             |>
-  complete(F1, F2, fill = list(type = "start")) |>
-  filter(F1 != 0 | F2 != 0)
-
-point_spikes_transf <- point_spikes_transf             |>
-  bind_rows(point_spikes_transf |> slice(3))           |>
-  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
-  pivot_wider(names_from = type, values_from = F1:F2)
-
-point_spikes_transf_ob <- point_spikes_transf |>
-  pivot_longer(starts_with("F")) |>
-  separate(name, into = c("dim", "coord")) |>
-  pivot_wider(names_from = coord, values_from = value) |>
-  group_by(n) |>
-  mutate(
-    start = (transf_matrix %*% start |> drop()),
-    end   = (transf_matrix %*% end   |> drop())
-  ) |>
-  pivot_wider(names_from = dim, values_from = start:end)
-
-point |>
-  ggplot(mapping = aes(F1, F2)) +
-  geom_abline(slope = axis_slope, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data    = point_spikes,
-    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color = SEC_COLOR
-  ) +
-  geom_segment(
-    data    = point_spikes_transf_ob,
-    mapping = aes(start_F1, start_F2, xend = end_F1, yend = end_F2),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43'
-  ) +
-  geom_point(size = I(4)) +
-  scale_x_continuous(
-    limits = axis_lims_x,
-    breaks = 0,
-    labels = NULL,
-    name   = NULL
-  ) +
-  scale_y_continuous(
-    limits = axis_lims_y,
-    breaks = 0,
-    labels = NULL,
-    name   = NULL
-  ) +
-  coord_fixed(expand = FALSE, clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    # panel.grid.major   = element_line(linewidth = GRID_WIDTH),
-    panel.grid.major.x = element_line(
-      linewidth = GRID_WIDTH,
-      color = SEC_COLOR
-    ),
-    panel.grid.major.y = element_line(linewidth = GRID_WIDTH, color = "black")
-  )
-```
-:::
-
-::: column
-:::
-:::::
-
-::: notes
-Pero en AF tenemos dimensiones correlacionadas la mayoría de las veces, y eso como hemos visto implica hacer rotaciones oblicuas, que dan lugar a ejes no ortogonales.
-
-Aquí vemos un ejemplo de ejes oblicuos (muy exagerado, eso sí).
-Imaginad que esto que estamos representando aquí es un punto sobre La Tierra, y sus coordenadas rectangulares son la longitud y latitud.
-
-Es posible que por algún motivo (no sé cuál, pero imaginémoslo) nos interesen las coordenadas en estos "ejes oblicuos": Este-Oeste, y uno que podría ser algo así como "Este-Sureste este este - Oeste-Noroeste oeste oeste".
-¿Se entiende claramente?
-Proyectando cada coordenada, de forma paralela a los ejes, obtendríamos las coordenadas en este nuevo sistema de "coordenadas geodésicas oblicuas".
-
-Luego esta restricción que propone Reckase parece que en principio no se podría aplicar.
-:::
-
-## (But, an aside...) {.smaller}
-
-::::: columns
-::: column
-![A red cube in isometric perspective](../www/slideshow-assets/isometric-cube.jpg){fig-align="center"}
-:::
-
-::: {.column .fragment}
-<br>
-
-> It has been suggested by some researchers that the angle between the axes represents the degree of correlation.
-> However, for sake of clarity, an orthogonal axes system is used in which distance measures and vectors can be easily calculated, understood, and interpreted.
-
----@ackerman_multidimensional_2005-1 [p. 16]
-:::
-:::::
-
-::: aside
-Figure taken from [Vecteezy](https://www.vecteezy.com/vector-art/3709640-isometric-cube-on-a-white-background)
-:::
-
-::: notes
-Sin embargo, mirad este ejemplo:
-
-¿Qué veis?
-¿Un cubo, verdad?
-¿Son sus lados todos iguales?
-¿Y sus ángulos?
-Son todos ángulos rectos, se entiende.
-Pero si cojo un transportador de ángulos y mido, voy a ver que este mide 120 grados, y este 60.
-
-Este cubo está representado en un plano en perspectiva isométrica (Creo que todo el mundo sabe lo que es la perspectiva isométrica, si no de dibujo en bachillerato, de jugar a los SIMS, aunque sea).
-
-Bien, si ahora os pregunto por las diagonales de una cara, ¿diríais que son iguales?
-Y sin embargo, si mido (en este plano) las dos diagonales, vemos que son distintas.
-¿Estáis de acuerdo?
-
-Esta representación no es errónea: De hecho, se necesita algún sistema de perspectiva para representar un cuerpo sólido en dos dimensiones; la isométrica es solamente una de ellas.
-
-Es decir, que nada me impide, si lo necesito, representar unos ejes con un ángulo cualquiera (recto en este caso) como otro ángulo (120 grados).
-:::
-
-## Assumptions: Orthogonality {.smaller}
-
-> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
-
----@reckase_difficulty_1985 [p. 404]
-
-::::: columns
-::: column
-```{r oblique-rotation-axis-2}
-```
-:::
-
-::: {.column .fragment}
-<!-- TODO: Animate? -->
-
-```{r oblique-rotation-rect-coords}
-point_transf |>
-  ggplot(mapping = aes(F1, F2))            +
-  geom_segment(
-    data    = point_spikes_transf,
-    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-  )                                        +
-  geom_point(size = I(4))                  +
-  scale_x_continuous(
-    limits = axis_lims_x,
-    breaks = 0,
-    labels = NULL,
-    name   = NULL
-  )                                        +
-  scale_y_continuous(
-    limits = axis_lims_y,
-    breaks = 0,
-    labels = NULL,
-    name   = NULL
-  )                                        +
-  coord_fixed(expand = FALSE, clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      color     = "black",
-      linewidth = GRID_WIDTH
-    )
-  )
-```
-:::
-:::::
-
-::: notes
-Lo mismo pasa al representar los ejes de una rotación oblicua: Nada me impide representar las coordenadas resultantes en ejes rectangulares (las llamadas "coordenadas Cartesianas rectangulares").
-Son coordenadas, nada más; el espacio en el que yo las represente en realidad da igual.
-Y de hecho, esto lo habréis visto muchas veces; por ejemplo, SPSS representa los pesos factoriales siempre en coordenadas rectangulares, incluso después de aplicar una rotación oblicua.
-
-Así que si represento un ítem en ejes rectangulares, nada me impide aplicar las fórmulas de Reckase y obtener sus parámetros multidimensionales.
-
-Pero algo parece fallar cuando generalizamos los parámetros multidimensionales de TRI que hemos mostrado antes a esos espacios de coordenadas no ortogonales.
-Si rotamos los ejes de coordenadas, pero los ítems siguen siendo los mismos, es de esperar que la discriminación multidimensional de un ítem (i.e., la "longitud" de su vector) no cambie.
-Esta propiedad se llama invarianza.
-Veámoslo con un ejemplo.
-:::
-
-## Assumptions: Orthogonality {.smaller}
-
-::::::: columns
-::: column
-```{r item-oblique-axes-1}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     0.5,  1,    0
-)
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-item_spikes <- mirt_item_coords        |>
-  slice(1 |> rep(2))                   |>
-  add_column(dim = 'F' |> paste0(1:2)) |>
-  mutate(
-    origin_1 = if_else(dim == 'F1', end_1, origin_1),
-    origin_2 = if_else(dim == 'F2', end_2, origin_2),
-  )
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data = item_spikes,
-    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color     = PALETTE_BLUE["light"]
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.75, 1.5),
-    breaks       = (-1:3)/2,
-    minor_breaks = NULL,
-    name         = NULL,
-    expand       = expansion(),
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.75, 1.5),
-    breaks       = (-1:2)/2,
-    minor_breaks = NULL,
-    name         = NULL,
-    expand       = expansion(),
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    "richtext",
-    x = c(0.5, -0.1), y = c(-0.1, 1),
-    label         = c("*a*^*^~11~", "*a*^*^~21~"),
-    size          = 8,
-    label.colour  = NA,
-    label.padding = unit(0, "lines"),
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1/3),
-    axis.title.y.right = element_markdown(vjust = 1/3),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    )
-  )
-```
-:::
-
-::::: column
-::: fragment
-$$
-  \begin{align}
-    \mathbf{a}^*_1 &=
-      \begin{bmatrix}
-        0.5 \\
-        1
-      \end{bmatrix}\\
-      d^*_1 &= 0
-  \end{align}
-$$
-:::
-
-<br>
-
-::: fragment
-$$
-  \begin{align}
-    MDISC^*_1 &= \sqrt{
-        [0.5, 1]
-        \begin{bmatrix}
-          0.5 \\
-          1
-        \end{bmatrix}
-      } \\
-      &= \sqrt{ 0.5 · 0.5 + 1 · 1 } \\
-      &= \sqrt{ 0.25 + 1 } \\
-      &= \sqrt{ 1.25 } \approx 1.12
-  \end{align}
-$$
-:::
-:::::
-:::::::
-
-::: notes
-Supongamos un ítem, en coordenadas ortogonales (de una solución sin rotar).
-(Con vuestro permiso, le voy a llamar "a asterico", para representar las coordenadas "originales" de una solución ortogonal, antes de una rotación).
-El vector muestra los componentes de su parámetrosde discriminación en las dos dimensiones de los ejes representados.
-Esto sería la "solución sin rotar" que resulta típicamente de ajustar un modelo de TRI multdimensional, o un análisis factorial exploratorio (si lo convertimos a métrica de TRI).
-La discriminación multidimensional de este vector es (explicar paso a paso).
-
-Supongamos que aplicamos una rotación oblicua.
-Imaginad, por ejemplo, que la solución incluye más ítems aparte de este, y cada uno de los ejes se alinea con un grupo de ítems después de hacer la rotación, de forma que da una solución próxima a la estructura simple.
-Esa rotación como digo es oblicua, da lugar a dos ejes no ortogonales, y cada eje se identifica con un rasgo latente sustantivo.
-Como son oblicuos, la solución representa que esos dos rasgos latentes están correlacionados entre sí, con una correlación "la que sea", no nos importa cuál en este momento.
-:::
-
-## Assumptions: Orthogonality {.smaller}
-
-:::::::: columns
-::: column
-```{r item-oblique-axes-2}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-BASIS_VEC_1 <- c(1, 0)
-BASIS_VEC_2 <- c(-.5, 1)
-
-axis_slope <- BASIS_VEC_2[2] / BASIS_VEC_2[1]
-transf_matrix <- cbind(BASIS_VEC_1, BASIS_VEC_2) |> unname()
-
-item_coords_transf <- mirt_item_coords |>
-  pivot_longer(-item) |>
-  separate(name, into = c("coord", "dim")) |>
-  group_by(coord) |>
-  mutate(
-    dim   = paste0('F', dim),
-    value = solve(transf_matrix) %*% value |> drop()
-  ) |>
-  ungroup() |>
-  pivot_wider(names_from = dim)
-
-item_spikes_transf <- item_coords_transf         |>
-  select(-item)                                  |>
-  filter(coord == "end")                         |>
-  bind_rows(tibble(F1 = 0, F2 = 0))              |>
-  complete(F1, F2, fill = list(coord = "start")) |>
-  filter(F1 != 0 | F2 != 0)
-
-item_spikes_transf <- item_spikes_transf               |>
-  bind_rows(item_spikes_transf |> slice(3))            |>
-  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
-  pivot_wider(names_from = coord, values_from = F1:F2)
-
-item_spikes_transf_ob <- item_spikes_transf |>
-  pivot_longer(starts_with("F")) |>
-  separate(name, into = c("dim", "coord")) |>
-  pivot_wider(names_from = coord, values_from = value) |>
-  group_by(n) |>
-  mutate(
-    start = (transf_matrix %*% start |> drop()),
-    end   = (transf_matrix %*% end   |> drop())
-  ) |>
-  pivot_wider(names_from = dim, values_from = start:end)
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  ) +
-  geom_abline(
-    intercept = -2:3,
-    slope     = -2,
-    linewidth = GRID_WIDTH,
-    linetype  = "17",
-    color     = AXIS_COLOR
-  ) +
-  geom_abline(slope = -2, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0,  linewidth = GRID_WIDTH) +
-  geom_segment(
-    data = item_spikes_transf_ob,
-    mapping = aes(start_F1, start_F2, xend = end_F1, yend = end_F2),
-    inherit.aes = FALSE,
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color     = PALETTE_BLUE["light"]
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.75, 1.5),
-    breaks       = (-2:2)/2 + .375,
-    labels       = ~number(. - .375, .1),
-    minor_breaks = NULL,
-    name         = NULL,
-    expand       = expansion(),
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.75, 1.5),
-    breaks       = (-1:2)/2,
-    minor_breaks = NULL,
-    name         = NULL,
-    expand       = expansion(),
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    "richtext",
-    x = c(1, -0.59), y = c(-0.07, 1),
-    label         = c("*a*~11~", "*a*~21~"),
-    size          = 8,
-    label.colour  = NA,
-    label.padding = unit(0, "lines"),
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 0),
-    axis.title.y.right = element_markdown(vjust = 1/3),
-    axis.ticks         = element_blank(),
-    panel.grid.major.y = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    )
-  )
-```
-:::
-
-:::::: column
-::: fragment
-$$
-  \mathbf{Q} = \begin{bmatrix}
-      1 & -0.5 \\
-      0 & 1
-    \end{bmatrix}; \quad
-  \mathbf{Q}^{-1} = \begin{bmatrix}
-      1 & 0.5 \\
-      0 & 1
-    \end{bmatrix}
-$$
-:::
-
-::: fragment
-$$
-  \begin{align}
-    \mathbf{a}_1 =
-    \mathbf{Q}^{-1} \mathbf{a}^*_1 &= \begin{bmatrix}
-      1 & 0.5 \\
-      0 & 1
-    \end{bmatrix}
-    \begin{bmatrix}
-      0.5 \\
-      1
-    \end{bmatrix} \\
-    &= \begin{bmatrix}
-      1 · 0.5 + 0.5 · 1 \\
-      0 · 0.5 + 1 · 1
-    \end{bmatrix} \\
-    &= \begin{bmatrix}
-      0.5 + 0.5 \\
-      0   + 1
-    \end{bmatrix}
-    = \begin{bmatrix}
-      1 \\
-      1
-    \end{bmatrix}
-  \end{align}
-$$
-:::
-
-::: fragment
-$$
-  \begin{align}
-    \mathbf{a}_1 &=
-      \begin{bmatrix}
-        1 \\
-        1
-      \end{bmatrix}\\
-      d_1 &= d^*_1 = 0
-  \end{align}
-$$
-:::
-::::::
-::::::::
-
-::: notes
-Los ejes resultantes son los que se muestran en este gráfico.
-No es exactamente una rotación, ya que el eje oblicuo \[señalar\] es un poco más largo (si lo rotáramos solamente acabaría por aquí \[señalar\]), pero para el caso es una transformación de los ejes, y cualquier transformación lineal nos sirve para ejemplificar, y así simplificamos el ejemplo.
-Los nuevos valores del parámetros de discriminación del ítem, en este caso ya sí lo llamo "a", porque esta es la solución rotada, la que me interesa, la que es interpretable según mi teoría, están representados cada uno por la "proyección" de ese vector sobre cada uno de los ejes (proyección paralela al otro eje, u otros, si hubiera más de dos).
-
-Hacer una rotación supone "multiplicar" las coordenadas de este vector por la matriz de rotación que representa ese nuevo sistema de coordenadas (la transformación del ortogonal a este nuevo, oblicuo).
-
-\[NEXT\]
-
-En este caso, la matriz de rotación es esta (la llamamos matriz Q).
-Esta es la matriz que convierte las coordenadas del espacio ortogonal original a las coordenadas transformadas (coordenadas "rotadas").
-
-\[NEXT\]
-
-Como los vectores se representan en columnas, se "pre-multiplica" por la matriz de rotación.
-Esto es parecido a multiplicar por un vector; es como si cada fila de la matriz fuese un vector, y los resultados se ponen cada uno en una fila.
-\[Explicar paso a paso el producto matricial\]
-
-\[NEXT\]
-
-El resultado es, como vemos en el gráfico, (1, 1).
-El parámetro de discriminación de este ítem, en su solución rotada, interpretable, es este, (1, 1).
-El valor de la intersección del modelo no cambia.
-:::
-
-## Assumptions: Orthogonality {.smaller}
-
-:::::::: columns
-::: column
-```{r item-oblique-axes-3}
-#| fig-align:  center
-#| fig-height: 8
-#| fig-width:  8
-mirt_item <- tribble(
-  ~item, ~a_1, ~a_2, ~d,
-  1,     1,    1,    0
-)
-
-mirt_item_params <- mirt_item |>
-  compute_mirt_params(d, starts_with('a'))
-
-mirt_item_coords <- mirt_item_params |>
-  compute_mirt_coords(D, MDISC, starts_with("cos"), original_coords = FALSE) |>
-    rename_with(str_remove, pattern = "transf_")
-
-item_spikes <- mirt_item_coords        |>
-  slice(1 |> rep(2))                   |>
-  add_column(dim = 'F' |> paste0(1:2)) |>
-  mutate(
-    origin_1 = if_else(dim == 'F1', end_1, origin_1),
-    origin_2 = if_else(dim == 'F2', end_2, origin_2),
-  )
-
-mirt_item_coords |>
-  ggplot(
-    aes(
-      origin_1, origin_2,
-      xend = end_1,
-      yend = end_2,
-      color = factor(item), fill = factor(item)
-    ),
-  )                                                  +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_segment(
-    data = item_spikes,
-    mapping = aes(origin_1, origin_2, xend = end_1, yend = end_2),
-    linewidth = LINE_WIDTH,
-    linetype  = '43',
-    color     = PALETTE_BLUE["light"]
-  ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH
-  ) +
-  scale_x_continuous(
-    limits       = c(-0.75, 1.5),
-    breaks       = (-1:3)/2,
-    minor_breaks = NULL,
-    name         = NULL,
-    expand       = expansion(),
-    sec.axis     = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    limits       = c(-0.75, 1.5),
-    breaks       = (-1:2)/2,
-    minor_breaks = NULL,
-    name         = NULL,
-    expand       = expansion(),
-    sec.axis     = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  scale_color_manual(values = unname(PALETTE_BLUE["normal"]), guide = NULL) +
-  annotate(
-    "richtext",
-    x = c(1, -0.1), y = c(-0.1, 1),
-    label         = c("*a*~11~", "*a*~21~"),
-    size          = 8,
-    label.colour  = NA,
-    label.padding = unit(0, "lines"),
-    fill = NA
-  ) +
-  coord_fixed(clip = "on") +
-  theme(
-    axis.line          = element_blank(),
-    axis.title         = element_markdown(size = 20, padding = unit(0, "lines")),
-    axis.title.x       = element_markdown(hjust = 1/3),
-    axis.title.y.right = element_markdown(vjust = 1/3),
-    axis.ticks         = element_blank(),
-    panel.grid.major   = element_line(
-      linewidth = GRID_WIDTH,
-      linetype  = "17"
-    )
-  )
-```
-:::
-
-:::::: column
-::: fragment
-$$
-  \mathbf{a}_1 = \begin{bmatrix}
-    1 \\
-    1
-  \end{bmatrix};\quad
-  d_1 = 0
-$$
-:::
-
-::: fragment
-> Using orthogonal axes, discrimination corresponds to the length of the item response vector
-
----@ackerman_multidimensional_2005 [p. 6]
-:::
-
-::: fragment
-$$
-  \begin{align}
-    MDISC_1 &= \sqrt{
-        [1, 1]
-        \begin{bmatrix}
-          1 \\
-          1
-        \end{bmatrix}
-      } \\
-      &= \sqrt{ 1 · 1 + 1 · 1 } \\
-      &= \sqrt{ 1 + 1 } \\
-      &= \sqrt{ 2 } \approx 1.41
-  \end{align}
-$$
-:::
-::::::
-::::::::
-
-::: notes
-Ahora bien, al igual que vimos antes, nadie me impide representar este ítem en coordenadas rectangulares.
-
-\[NEXT\]
-
-Y si el ítem está representado en coordenadas rectangulares, puedo aplicar, como vimos, las fórmulas de Reckase.
-Pero si hago eso, me encuentro con que el resultado de la discriminación es este, raíz de 2.
-Antes, recordaréis que era raíz de 1,25, aproximadamente 1,12.
-
-Es decir, ¿que el ítem discrimina mejor o peor en función de qué ejes quiera yo considerar?
-El ítem es exactamente el mismo, y por lo tanto su discriminación (multidimensional al menos, ya que no depende de la "dirección" a considerar, como los valores del vector de discriminación) debería ser la misma.
-Esta propiedad se llama "invarianza" del ítem.
-:::
-
-## Assumptions: Orthogonality
-
-<!-- TODO: ACT en McDonald, 1999? -->
-
-::: r-center
-!["ACT Mathematics Usage Test" items](../www/Fig_1-6_Contemporary_Psychometrics.png){fig-align="center" height="400px"}
-:::
-
-::: aside
-Figure taken from @ackerman_multidimensional_2005-1
-:::
-
-::: notes
-No tiene sentido que cambie su discriminación multidimensional, y sin embargo la literatura está llena de ejemplos que calculan este parámetro con la fórmula de Reckase (en un espacio de "rasgos correlacionados", que en seguida veremos por qué eso es importante), y que no se preocupan por "los ejes" que representan esos rasgos de manera fidedigna.
-
-Para la representación gráfica esto puede valer, como dice Ackerman aquí.
-Pero a la hora de calcular los parámetros (y Ackerman lo hace en este capítulo de libro), solamente considerar el caso ortogonal, la información que obtenemos resulta deficiente; obtenemos parámetros de discriminación multidimensional "distorsionados".
-:::
-
 ## Generalization: Oblique axes {.smaller}

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Drop old "oblique axes" generalization slides

@@
 
+::: {.fragment .r-center}
+### Generalization: "Oblique axes"
+:::
+
 ::: notes
-What we actually need is to compute the multidimensional parameters in this general space, which is not necessarily orthogonal (or, we should say more generally, not necessarily orthonormal).
-:::
-
-## Generalization: Oblique axes {.smaller}
-
-> \[...\] before performing the differentiation, the constraint that Σ cos^2^**α** = 1 is added to the expression for the slope \[...\]
-
----@reckase_difficulty_1985 [p. 404]
-
-:::::::::::: columns
-:::::: column
-::::: r-stack
-::: {.fragment fragment-index="1"}
-```{r corr-traits-example}
-#| fig-width:  5.5
-#| fig-height: 5.5
-
-set.seed(7117)
-
-corr_matrix <- matrix(c(1, .5, .5, 1), nrow = 2)
-
-thetas <- rmvnorm(10^4, sigma = corr_matrix) |>
-  as_tibble(.name_repair = ~paste0("trait_", 1:2)) |>
-    rownames_to_column("i")
-
-thetas |>
-  ggplot(mapping = aes(trait_1, trait_2)) +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
-  scale_x_continuous(
-    minor_breaks = NULL,
-    limits = c(-4, 4),
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    minor_breaks = NULL,
-    limits = c(-4, 4),
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
-  ) +
-  coord_fixed() +
-  theme(
-    axis.ticks       = element_blank(),
-    axis.line        = element_blank(),
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )
-```
-:::
-
-::: {.fragment .fade-in fragment-index="4"}
-```{r corr-traits-transf-example}
-#| fig-width:  5.5
-#| fig-height: 5.5
-
-# This is a square root of the correlation matrix, then inverted and transposed:
-orth_matrix <- c(1, 0, .5, sqrt(.75)) |> matrix(nrow = 2) |> solve() |> t()
-
-thetas_transf <- thetas |>
-  pivot_longer(starts_with("trait")) |>
-  group_by(i) |>
-  mutate(value = orth_matrix %*% value |> drop()) |>
-  pivot_wider() |>
-  ungroup()
-
-thetas_transf |>
-  ggplot(mapping = aes(trait_1, trait_2)) +
-  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
-  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
-  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
-  scale_x_continuous(
-    minor_breaks = NULL,
-    limits = c(-4, 4),
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*^*^~2~", labels = NULL)
-  ) +
-  scale_y_continuous(
-    minor_breaks = NULL,
-    limits = c(-4, 4),
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "*&theta;*^*^~1~", labels = NULL)
-  ) +
-  coord_fixed() +
-  theme(
-    axis.ticks       = element_blank(),
-    axis.line        = element_blank(),
-    axis.title       = element_markdown(),
-    axis.title.y     = element_markdown(),
-    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
-  )
-```
-:::
-:::::
-::::::
-
-::::::: column
-::: {.fragment fragment-index="1"}
-$$
-  cor (\mathbf{\theta}) = \mathbf{R}
-$$
-:::
-
-::: {.fragment fragment-index="2"}
-$$
-  \mathbf{R} = \begin{bmatrix}
-      1 & 0.5 \\
-      0.5 & 1
-    \end{bmatrix}
-$$
-:::
-
-::: {.fragment fragment-index="3"}
-$$
-  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
-$$
-:::
-
-::: {.fragment fragment-index="4"}
-$$
-  cor (\mathbf{\theta}^*) =
-    \mathbf{R}^* =
-    \begin{bmatrix}
-      1 & 0 \\
-      0 & 1
-    \end{bmatrix}
-$$
-:::
-:::::::
-::::::::::::
-
-::: notes
-Hacer esta generalización no es algo tan directo como ocurría con la probabilidad "monótonamente decreciente".
-Si os acordáis, Reckase aplicaba esta restricción de que la suma de cosenos directores era 1 (que se deriva del teorema de Pitágoras).
-Pero para aplicarla hace falta asumir que esos cosenos directores son todos ortogonales entre sí, como ya hemos visto.
-Si no tenemos ángulos rectos, esa condición no se cumple.
-
-\[NEXT\]
-
-Y lo que nosotros estamos asumiendo es que tenemos una distribución de vectores de rasgo latente ($\theta$) con una matriz de correlaciones $R$ (la que sea).
-
-\[NEXT\]
-
-(En este ejemplo, vamos a usar una distribución normal estandarizada con correlación de 0,5).
-
-Normalmente representamos esos vectores en coordenadas rectangulares (ejes ortogonales), pero para obtener los parámetros multidimensionales generalizados a ejes oblicuos tenemos que representados en una base no ortogonal (aquí también hablaría de "unidades no estandarizadas", es decir, que la distancia entre el 0 y el 1 pueda ser "distinta de la base estándar, pero digamos solamente "no ortogonal" por simplificar).
-
-¿Cómo podríamos solucionar esto?
-
-\[NEXT\]
-
-Lo que podemos hacer es aplicar una "transformación" a los rasgos latentes de forma que pasemos de esa base oblicua (la que sea, aún no sabemos cuál es) a una base ortogonal.
-Esto es una transformación de las coordenadas, como la que aplicamos antes para calcular las coordenadas del ítem tras aplicar la rotación oblicua, pero aplicada en este caso al vector de rasgo latente.
-
-Ahora bien, como recordaréis, queremos representar los ítems en unos ejes que sean, por los ángulos relativos entre ellos, una representación fidedigna, de alguna manera, de las correlaciones entre las dimensiones de rasgo latente.
-
-Estas nuevas coordenadas (transformadas), que ahora están en una base ortogonal, queremos que representen una distribución en la que las correlaciones sean nulas, es decir, que la matriz de correlaciones transformada ($R^*$ la vamos a llamar) sea una matriz identidad (unos en la diagonal, ceros en el resto).
-Pero hay que tener en cuenta que esto es un "artefacto matemático".
-Estamos haciendo esto por conveniencia, pero las coordenadas que realmente nos interesan como investigadores, porque son las que verdaderamente representan algo que nos interesa investigar, son esas coordenadas originales, que pueden representar cualquier fenómeno que sea de nuestro interés (rasgos de personalidad, síntomas, actitudes, inteligencia, conocimientos...
-lo que estemos investigando).
-
-Llegados a este punto, la incógnita es: ¿Qué matriz es $\mathbf{P}$, es decir, la matriz que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$?
-Esto es equivalente a decir, ¿cuáles son los ejes originales en los que tenemos que representar $\mathbf{\theta}$?
-(Recordad, para que $\mathbf{R^*}$ sea una matriz identidad, 1's en la diagonal y 0's fuera de ella).
-:::
-
-## Generalization: Oblique axes {.smaller}
-
-<center>
-
-<br>
-
-[bit.ly/vote-coords](https://bit.ly/vote-coords) ([slido.com](https://slido.com) : #1904926)
-
-<br>
-
-![](../www/slideshow-assets/bit.ly_vote-coords.svg){width="40%" height="40%"}
-
-</center>
-
-::: notes
-En base a lo planteado anteriormente, me gustaría hacer una prueba con vosotras y vosotros.
-Aquí tenéis un enlace y QR para participar en una breve "encuesta".
-:::
-
-## Generalization: Oblique axes {.smaller}
-
-<br>
-
-<!-- # TODO: Faltan dos primeros paneles -->
-
-:::::: columns
-::: {.column width="33%"}
-```{r}
-#| fig-width:  5.5
-#| fig-height: 5.5
-thetas_transf |>
-  ggplot(mapping = aes(trait_1, trait_2)) +
-  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
-  scale_x_continuous(
-    minor_breaks = NULL,
-    labels = \(breaks) '' |> rep(length(breaks)),
-    limits = c(-4, 4),
-    expand = expansion(),
-    name   = NULL,
-    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~}", labels = NULL)
-  ) +
-  scale_y_continuous(
-    minor_breaks = NULL,
-    labels = \(breaks) '' |> rep(length(breaks)),
-    limits = c(-4, 4),
-    expand = expansion(),
-    name   = NULL
-  ) +
-  coord_fixed() +
-  theme(
-    axis.ticks   = element_blank(),
-    axis.line    = element_blank(),
-    axis.title   = element_markdown(),
-    axis.title.y = element_markdown(),
-    panel.grid   = element_blank()
-  )
-```
-:::
-
-::: {.column .fragment width="33%"}
-```{r question-mark}
-fa(
-  "question",
-  fill           = "#929acb",
-  height         = "200px",
-  vertical_align = "-10em",
-  margin_left    = "4em"
-)
-```
-:::
-
-::: {.column .fragment width="33%"}
-```{r corr-traits-transf-example}
-#| fig-width:  5.5
-#| fig-height: 5.5
-```
-:::
-::::::
-
-¿What is the coordinate system of *θ* to make *θ\** "spherical"?
-
-::: notes
-Supongamos una distribución bivariada de rasgos latentes, con una correlación de .5, como la mostrada anteriormente.
-Queremos saber cuál es la matriz $\mathbf{P}$ que nos permite pasar de las coordenadas originales, $\mathbf{\theta}$, a las transformadas $\mathbf{\theta^*}$.
-
-Esto es equivalente a preguntar:
-
-\[NEXT\]
-
-¿Cuál es el sistema de coordenadas que representaría adecuadamente los ejes de una distribución bivariada,...
-
-\[NEXT\]
-
-de forma que al transformarlos a coordenadas rectangulares (mediante la matriz $\mathbf{P}$) obtuviésemos una distribución con correlaciones de 0 y varianzas de 1, es decir, una distribución "esférica"?
-:::
-
-## Generalization: Oblique axes
-
-```{r create-grid-function}
-create_grid <- function(basis,
-                        x_limits = c(-4, 4),
-                        y_limits = c(-4, 4),
-                        break_step = 2) {
-  # Argument checking and formatting: ----
-
-  ## Basis vectors tangents and norms
-  vec_tan <- basis[2, ] / basis[1, ]
-
-  # Main: ----
-
-  # Define grid limits as a function of the axes limits (in canonical metric):
-  grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
-    x_lim = if_else(x == x_limits[1], "inf", "sup"),
-    y_lim = if_else(y == y_limits[1], "inf", "sup")
-  )
-
-  # Select delimiting grid vertices, based on the basis vector directions
-  grid_bounds <- bind_rows(
-    h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
-    v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
-    .id = "grid"
-  ) |>
-    filter(xor(slope > 0, x_lim == y_lim)) |>
-    mutate(limit_intercept = y - slope * x) # Compute their intercept
-
-  # Transformed (1, 1), expressed in the canonical basis
-  unit_point <- basis %*% c(1, 1) |> drop() |>
-    as_tibble(.name_repair = "minimal") |>
-    add_column(name = c('x', 'y') |> paste0('_unit')) |>
-    pivot_wider(values_from = value)
-
-  # Grid specifications, in canonical basis
-  grid_specs <- grid_bounds |>
-    bind_cols(unit_point) |>
-    mutate(unit_intercept = y_unit - slope * x_unit) |>
-    select(-ends_with("unit")) |>
-    mutate(
-      n_units  = trunc(limit_intercept / unit_intercept),
-      n_breaks = n_units %/% break_step * break_step
-    )
-
-  # Create vertical grid data, if the second basis vector is completely vertical
-  if (is.infinite(vec_tan[2])) {
-
-    # Create vertical grid data
-    v_grid <- grid_bounds |>
-      filter(grid == "v") |>
-      mutate(
-        unit     = basis[1, 1],
-        n_units  = trunc(x / unit),
-        n_breaks = n_units %/% break_step * break_step
-      ) |>
-      select(grid, limit, n_breaks, unit) |>
-      pivot_wider(names_from = "limit", values_from = n_breaks) |>
-      reframe(coord = seq(from = inf, to = sup, by = break_step) * unit) |>
-      mutate(
-        grid      = 'v',
-        slope     = Inf,
-        intercept = NA_real_,
-        label_pos = coord,
-        ref       = as.character(!coord)
-      )
-
-    # Filter out non-valid grid values
-    grid_specs <- grid_specs |> filter(!is.infinite(slope))
-  }
-
-  # Create grid data
-  grid <- grid_specs |>
-    select(grid, limit, slope, unit_intercept, n_breaks) |>
-    pivot_wider(names_from = limit, values_from = n_breaks) |>
-    group_by(grid) |>
-    reframe(
-      slope,
-      coord = seq(from = inf, to = sup, by = break_step),
-      intercept = coord * unit_intercept
-    ) |>
-    mutate(
-      label_pos = if_else(
-        grid == "h",
-         intercept   + x_limits[1] * vec_tan[1],
-        (y_limits[1] - intercept)  / vec_tan[2]
-      ),
-      ref = as.character(!intercept)
-    )
-
-    if (is.infinite(vec_tan[2])) return(grid |> bind_rows(v_grid))
-
-    grid
-}
-
-plot_grid <- function(grid,
-                      x_limits = c(-4, 4),
-                      y_limits = c(-4, 4),
-                      break_step = 2,
-                      linetype_axis = 'solid',
-                      linetype_grid = 'dashed') {
-
-  # Argument checking and formatting: ----
-
-  ## Grid plot linetypes
-  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
-
-  ## Basis vectors tangents and norms
-  vec_tan <- grid |> distinct(grid, slope) |> deframe()
-
-  ## Axis title positions
-  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
-    (x_limits[2] - x_limits[1])
-  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
-    (y_limits[2] - y_limits[1])
-
-  # Main: ----
-
-  # Assign x axis label values and coordinates (to account for the vertical
-  #   special case)
-  x_labels <- grid |> filter(grid == "v") |> select(coord, label_pos)
-
-  # Create vertical grid geometry if the second vector is completely vertical
-  if (is.infinite(vec_tan[2])) {
-
-    geom_vgrid <- geom_vline(
-      mapping = aes(xintercept = coord, linetype = ref),
-      data = grid |> filter(grid == "v")
-    )
-
-    grid <- grid |> filter(grid == "h")
-  }
-
-  grid_plot <- grid |>
-    ggplot() +
-    geom_abline(
-      mapping = aes(
-        intercept = intercept,
-        slope     = slope,
-        linetype  = ref
-      )
-    )
-
-  if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid
-
-  grid_plot +
-    scale_x_continuous(
-      minor_breaks = NULL,
-      breaks = x_labels |> pull(label_pos),
-      labels = x_labels |> pull(coord),
-      limits = x_limits,
-      expand = expansion(),
-      name   = NULL,
-      sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
-    ) +
-    scale_y_continuous(
-      minor_breaks = NULL,
-      breaks = grid |> filter(grid == "h") |> pull(label_pos),
-      labels = grid |> filter(grid == "h") |> pull(coord),
-      limits = y_limits,
-      expand = expansion(),
-      name   = NULL,
-      sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
-    ) +
-    scale_linetype_manual(values = linetypes, guide = NULL) +
-    coord_fixed() +
-    theme(
-      axis.ticks   = element_blank(),
-      axis.line    = element_blank(),
-      axis.title.x       = element_markdown(hjust = y_axis_titpos),
-      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
-    )
-}
-
-transform_grid <- function(basis,
-                           x_limits = c(-4, 4),
-                           y_limits = c(-4, 4),
-                           break_step = 2,
-                           linetype_axis = 'solid',
-                           linetype_grid = 'dashed') {
-
-  # Argument checking and formatting: ----
-
-  # Main: ----
-  basis |>
-    create_grid(
-      x_limits   = x_limits,
-      y_limits   = y_limits,
-      break_step = break_step
-    ) |>
-    plot_grid(
-      x_limits   = x_limits,
-      y_limits   = y_limits,
-      break_step = break_step,
-      linetype_axis = 'solid',
-      linetype_grid = 'dashed'
-    )
-}
-```
-
-<br>
-
-:::::: columns
-::: {.column width="33%"}
-<center>A</center>
-
-```{r grid-orthogonal-rotation}
-#| fig-width: 5
-rot_basis <- cbind(
-  c(sqrt(.75), -sqrt(.25)),
-  c(sqrt(.25), sqrt(.75))
-)
-
-rot_grid <- transform_grid(rot_basis)
-
-rot_grid
-```
-:::
-
-::: {.column .fragment width="33%"}
-<center>B</center>
-
-```{r grid-ts-basis}
-#| fig-width: 5
-ts_basis <- cbind(
-  c(1, 0),
-  c(sqrt(.25), sqrt(.75))
-)
-
-ts_grid <- transform_grid(ts_basis)
-
-ts_grid
-```
-:::
-
-::: {.column .fragment width="33%"}
-<center>C</center>
-
-```{r grid-ls-basis}
-#| fig-width: 5
-ls_basis <- t(solve(ts_basis))
-
-ls_grid <- transform_grid(ls_basis)
-
-ls_grid
-```
-:::
-::::::
-
-::: fragment
-[Answers in 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
-:::
-
-::: notes
--   ¿Será el A?
-
-\[NEXT\]
-
--   ¿O puede ser el B?
-
-\[NEXT\]
-
--   ¿O es el C?
-
-\[NEXT\]
-
-Tomáos vuestro tiempo y en seguida vemos las respuestas.
-:::
-
-## Generalization: Oblique axes
-
-<br>
-
-:::::::: columns
-::: {.column width="33%"}
-<center>A</center>
-
-```{r scatter-orthogonal-rotation}
-#| fig-width: 5
-thetas_transf <- thetas |>
-  group_by(i) |>
-  pivot_longer(starts_with("trait")) |>
-  mutate(value = rot_basis %*% value |> drop()) |>
-  pivot_wider() |>
-  ungroup()
-
-rot_grid +
-  geom_point(
-    data    = thetas_transf,
-    mapping = aes(trait_1, trait_2),
-    alpha   = .33,
-    shape   = 16,
-    colour = I(PALETTE_BLUE["normal"])
-  )
-```
-:::
-
-::::: {.column .fragment width="33%"}
-<center>B</center>
-
-:::: r-stack
-```{r grid-ts-basis}
-#| fig-width: 5
-```
-
-::: fragment
-```{r scatter-ts-basis}
-#| fig-width: 5
-thetas_transf <- thetas |>
-  group_by(i) |>
-  pivot_longer(starts_with("trait")) |>
-  mutate(value = ts_basis %*% value |> drop()) |>
-  pivot_wider() |>
-  ungroup()
-
-ts_grid +
-  geom_point(
-    data    = thetas_transf,
-    mapping = aes(trait_1, trait_2),
-    alpha   = .33,
-    shape   = 16,
-    colour = I(PALETTE_BLUE["normal"])
-  )
-```
-:::
-::::
-:::::
-
-::: {.column .fragment width="33%"}
-<center>C</center>
-
-```{r scatter-ls-basis}
-#| fig-width: 5
-thetas_transf <- thetas |>
-  group_by(i) |>
-  pivot_longer(starts_with("trait")) |>
-  mutate(value = ls_basis %*% value |> drop()) |>
-  pivot_wider() |>
-  ungroup()
-
-ls_scatter <- ls_grid +
-  geom_point(
-    data    = thetas_transf,
-    mapping = aes(trait_1, trait_2),
-    alpha   = .33,
-    shape   = 16,
-    colour = I(PALETTE_BLUE["normal"])
-  )
-
-ls_scatter
-```
-:::
-::::::::
-
-::: notes
-El primero yo creo que estamos de acuerdo en que no, ya que solamente es una "rotación ortogonal" (como la llamaríamos en el ámbito psicométrico).
-La transformación que se aplica podría dar lugar a correlaciones nulas (es lo que haría la transformación de los ejes mediante Análisis de Componentes Principales), pero las varianzas resultantes no serían estandarizadas.
-
-\[NEXT\]
-
-El segundo caso rotaría el eje vertical de manera que ambos ejes formarían un ángulo de 60º, que tiene un coseno igual a 0.5.
-La gente que ha pensado que estos son los ejes correctos, pensáis como yo: que el coseno del ángulo entre los ejes representa la correlación entre las dimensiones.
-Por lo tanto, cuanto más alta la correlación, más alineados deben estar los ejes.
-
-\[NEXT\]
-
-Sin embargo, mirad lo que ocurre cuando "proyectamos" esa nube de puntos sobre esos ejes.
-La "nube de puntos" se achata y se alarga aún más.
-
-\[NEXT\]
-
-La respuesta correcta era la C, que "compensa" de alguna manera el achatamiento de la nube de puntos, de forma que las coordenadas transformadas que nos quedan (en los ejes ortogonales, los "normales" del gráfico) tienen una distribución "esférica" (con matriz de correlaciones "identidad").
-
-Esto, que visto así parece ya evidente, es algo que al menos a mí me ha llevado varias decenas (puede que "cientos") de horas entender, durante al menos dos años trabajando en este problema.
+What we actually need is to compute the multidimensional parameters in this general space,...
+
+\[NEXT\]
+
+...which is not necessarily orthogonal (or, we should say more generally, not necessarily orthonormal).
+
+¿Why? Because we want this space to have a meaningful interpretation; we want it to be a geometrical representation of the probabilistic properties of our sample.
 :::

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add new "oblique axes" slides

@@
 
+## Oblique axes {.smaller}
+
+<br>
+
+:::::: columns
+::: {.column width="33%"}
+```{r}
+#| fig-width:  5.5
+#| fig-height: 5.5
+thetas_transf |>
+  ggplot(mapping = aes(trait_1, trait_2)) +
+  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    labels = \(breaks) '' |> rep(length(breaks)),
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~}", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    labels = \(breaks) '' |> rep(length(breaks)),
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks   = element_blank(),
+    axis.line    = element_blank(),
+    axis.title   = element_markdown(),
+    axis.title.y = element_markdown(),
+    panel.grid   = element_blank()
+  )
+```
+:::
+
+::: {.column .fragment width="33%" fragment-index="1"}
+```{r question-mark}
+fa(
+  "question",
+  fill           = "#929acb",
+  height         = "200px",
+  vertical_align = "-10em",
+  margin_left    = "4em"
+)
+```
+:::
+
+::: {.column .fragment width="33%" fragment-index="2"}
+```{r corr-traits-transf-example}
+#| fig-width:  5.5
+#| fig-height: 5.5
+```
+:::
+::::::
+
+<center>
+
+::: {.fragment fragment-index="2"}
+Which coordinate system *θ* makes *θ\** "spherical"?
+:::
+
+</center>
+
+::: notes
+So, if the basis matrix of $\mathbf{\theta^*}$, which has an identity covariance matrix, is an identity covariance matrix, what is $\mathbf{P}$, the basis matrix of $\mathbf{\theta}$.
+
+In other words,...
+
+\[NEXT\]
+
+...what is the coordinate system of $\mathbf{\theta}$...
+
+\[NEXT\]
+
+...that makes $\mathbf{\theta^*}$ spherical?
+:::
+
+## Oblique axes
+
+```{r create-grid-function}
+create_grid <- function(basis,
+                        x_limits = c(-4, 4),
+                        y_limits = c(-4, 4),
+                        break_step = 2) {
+  # Argument checking and formatting: ----
+
+  ## Basis vectors tangents and norms
+  vec_tan <- basis[2, ] / basis[1, ]
+
+  # Main: ----
+
+  # Define grid limits as a function of the axes limits (in canonical metric):
+  grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
+    x_lim = if_else(x == x_limits[1], "inf", "sup"),
+    y_lim = if_else(y == y_limits[1], "inf", "sup")
+  )
+
+  # Select delimiting grid vertices, based on the basis vector directions
+  grid_bounds <- bind_rows(
+    h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
+    v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
+    .id = "grid"
+  ) |>
+    filter(xor(slope > 0, x_lim == y_lim)) |>
+    mutate(limit_intercept = y - slope * x) # Compute their intercept
+
+  # Transformed (1, 1), expressed in the canonical basis
+  unit_point <- basis %*% c(1, 1) |> drop() |>
+    as_tibble(.name_repair = "minimal") |>
+    add_column(name = c('x', 'y') |> paste0('_unit')) |>
+    pivot_wider(values_from = value)
+
+  # Grid specifications, in canonical basis
+  grid_specs <- grid_bounds |>
+    bind_cols(unit_point) |>
+    mutate(unit_intercept = y_unit - slope * x_unit) |>
+    select(-ends_with("unit")) |>
+    mutate(
+      n_units  = trunc(limit_intercept / unit_intercept),
+      n_breaks = n_units %/% break_step * break_step
+    )
+
+  # Create vertical grid data, if the second basis vector is completely vertical
+  if (is.infinite(vec_tan[2])) {
+
+    # Create vertical grid data
+    v_grid <- grid_bounds |>
+      filter(grid == "v") |>
+      mutate(
+        unit     = basis[1, 1],
+        n_units  = trunc(x / unit),
+        n_breaks = n_units %/% break_step * break_step
+      ) |>
+      select(grid, limit, n_breaks, unit) |>
+      pivot_wider(names_from = "limit", values_from = n_breaks) |>
+      reframe(coord = seq(from = inf, to = sup, by = break_step) * unit) |>
+      mutate(
+        grid      = 'v',
+        slope     = Inf,
+        intercept = NA_real_,
+        label_pos = coord,
+        ref       = as.character(!coord)
+      )
+
+    # Filter out non-valid grid values
+    grid_specs <- grid_specs |> filter(!is.infinite(slope))
+  }
+
+  # Create grid data
+  grid <- grid_specs |>
+    select(grid, limit, slope, unit_intercept, n_breaks) |>
+    pivot_wider(names_from = limit, values_from = n_breaks) |>
+    group_by(grid) |>
+    reframe(
+      slope,
+      coord = seq(from = inf, to = sup, by = break_step),
+      intercept = coord * unit_intercept
+    ) |>
+    mutate(
+      label_pos = if_else(
+        grid == "h",
+         intercept   + x_limits[1] * vec_tan[1],
+        (y_limits[1] - intercept)  / vec_tan[2]
+      ),
+      ref = as.character(!intercept)
+    )
+
+    if (is.infinite(vec_tan[2])) return(grid |> bind_rows(v_grid))
+
+    grid
+}
+
+plot_grid <- function(grid,
+                      x_limits = c(-4, 4),
+                      y_limits = c(-4, 4),
+                      break_step = 2,
+                      linetype_axis = 'solid',
+                      linetype_grid = 'dashed') {
+
+  # Argument checking and formatting: ----
+
+  ## Grid plot linetypes
+  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
+
+  ## Basis vectors tangents and norms
+  vec_tan <- grid |> distinct(grid, slope) |> deframe()
+
+  ## Axis title positions
+  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
+    (x_limits[2] - x_limits[1])
+  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
+    (y_limits[2] - y_limits[1])
+
+  # Main: ----
+
+  # Assign x axis label values and coordinates (to account for the vertical
+  #   special case)
+  x_labels <- grid |> filter(grid == "v") |> select(coord, label_pos)
+
+  # Create vertical grid geometry if the second vector is completely vertical
+  if (is.infinite(vec_tan[2])) {
+
+    geom_vgrid <- geom_vline(
+      mapping = aes(xintercept = coord, linetype = ref),
+      data = grid |> filter(grid == "v")
+    )
+
+    grid <- grid |> filter(grid == "h")
+  }
+
+  grid_plot <- grid |>
+    ggplot() +
+    geom_abline(
+      mapping = aes(
+        intercept = intercept,
+        slope     = slope,
+        linetype  = ref
+      )
+    )
+
+  if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid
+
+  grid_plot +
+    scale_x_continuous(
+      minor_breaks = NULL,
+      breaks = x_labels |> pull(label_pos),
+      labels = x_labels |> pull(coord),
+      limits = x_limits,
+      expand = expansion(),
+      name   = NULL,
+      sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
+    ) +
+    scale_y_continuous(
+      minor_breaks = NULL,
+      breaks = grid |> filter(grid == "h") |> pull(label_pos),
+      labels = grid |> filter(grid == "h") |> pull(coord),
+      limits = y_limits,
+      expand = expansion(),
+      name   = NULL,
+      sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
+    ) +
+    scale_linetype_manual(values = linetypes, guide = NULL) +
+    coord_fixed() +
+    theme(
+      axis.ticks   = element_blank(),
+      axis.line    = element_blank(),
+      axis.title.x       = element_markdown(hjust = y_axis_titpos),
+      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
+    )
+}
+
+transform_grid <- function(basis,
+                           x_limits = c(-4, 4),
+                           y_limits = c(-4, 4),
+                           break_step = 2,
+                           linetype_axis = 'solid',
+                           linetype_grid = 'dashed') {
+
+  # Argument checking and formatting: ----
+
+  # Main: ----
+  basis |>
+    create_grid(
+      x_limits   = x_limits,
+      y_limits   = y_limits,
+      break_step = break_step
+    ) |>
+    plot_grid(
+      x_limits   = x_limits,
+      y_limits   = y_limits,
+      break_step = break_step,
+      linetype_axis = 'solid',
+      linetype_grid = 'dashed'
+    )
+}
+```
+
+<br>
+
+:::::: columns
+::: {.column width="33%"}
+<center>A</center>
+
+```{r grid-orthogonal-rotation}
+#| fig-width: 5
+rot_basis <- cbind(
+  c(sqrt(.75), -sqrt(.25)),
+  c(sqrt(.25), sqrt(.75))
+)
+
+rot_grid <- transform_grid(rot_basis)
+
+rot_grid
+```
+:::
+
+::: {.column width="33%"}
+<center>B</center>
+
+```{r grid-ts-basis}
+#| fig-width: 5
+ts_basis <- cbind(
+  c(1, 0),
+  c(sqrt(.25), sqrt(.75))
+)
+
+ts_grid <- transform_grid(ts_basis)
+
+ts_grid
+```
+:::
+
+::: {.column width="33%"}
+<center>C</center>
+
+```{r grid-ls-basis}
+#| fig-width: 5
+ls_basis <- t(solve(ts_basis))
+
+ls_grid <- transform_grid(ls_basis)
+
+ls_grid
+```
+:::
+::::::
+
+::: fragment
+[Answers in 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
+:::
+
+::: notes
+So, here are three possible options; one of them is the correct one.
+
+\[NEXT\]
+
+Take a few seconds to answer and we will see next which is the correct one.
+:::
+
+## Oblique axes
+
+<br>
+
+:::::::::: columns
+::: {.column width="33%"}
+<center>A</center>
+
+```{r scatter-orthogonal-rotation}
+#| fig-width: 5
+thetas_transf <- thetas |>
+  group_by(i) |>
+  pivot_longer(starts_with("trait")) |>
+  mutate(value = rot_basis %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
+rot_grid +
+  geom_point(
+    data    = thetas_transf,
+    mapping = aes(trait_1, trait_2),
+    alpha   = .33,
+    shape   = 16,
+    colour = I(PALETTE_BLUE["normal"])
+  )
+```
+:::
+
+::::: {.column width="33%"}
+<center>B</center>
+
+:::: r-stack
+```{r grid-ts-basis}
+#| fig-width: 5
+```
+
+::: fragment
+```{r scatter-ts-basis}
+#| fig-width: 5
+thetas_transf <- thetas |>
+  group_by(i) |>
+  pivot_longer(starts_with("trait")) |>
+  mutate(value = ts_basis %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
+ts_grid +
+  geom_point(
+    data    = thetas_transf,
+    mapping = aes(trait_1, trait_2),
+    alpha   = .33,
+    shape   = 16,
+    colour = I(PALETTE_BLUE["normal"])
+  )
+```
+:::
+::::
+:::::
+
+::::: {.column width="33%"}
+<center>C</center>
+
+:::: r-stack
+```{r grid-ls-basis}
+#| fig-width: 5
+```
+
+::: fragment
+```{r scatter-ls-basis}
+#| fig-width: 5
+thetas_transf <- thetas |>
+  group_by(i) |>
+  pivot_longer(starts_with("trait")) |>
+  mutate(value = ls_basis %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
+ls_scatter <- ls_grid +
+  geom_point(
+    data    = thetas_transf,
+    mapping = aes(trait_1, trait_2),
+    alpha   = .33,
+    shape   = 16,
+    colour = I(PALETTE_BLUE["normal"])
+  )
+
+ls_scatter
+```
+:::
+::::
+:::::
+::::::::::
+
+::: notes
+A is pretty clearly not the answer, as it is just an orthogonal rotation.
+
+Now, if you chose B, you probably thought that the axis units should not change (because the variances are unitary), and that the axes should form an angle which cosine is the correlation, that is, 0.5 (a 60º angle).
+
+In that case, you think exactly like I do.
+
+But if you picture this transformation in your mind, you will notice that when you make the axes closer, you will be squeezing the points in the distribution, making them look even more like a "sausage", so to speak.
+
+\[NEXT\]
+
+The resulting distribution is this; it took us hundreds of hours of work, during more than two years to realise this.
+
+So if you want to actually "stretch" the cloud of points in the distribution, you need to pull the axes apart,...
+
+\[NEXT\]
+
+...as in the coordinate axes C.
+:::
+
 ## Latent space vs. test space {.smaller}

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add missing "oblique axes" initial slide
It was omitted from the previous commit by mistake

@@
 
+:::::::::: columns
+:::::: column
+$$
+  cor (\mathbf{\theta}) = \mathbf{\Sigma}
+$$
+
+::: {.fragment fragment-index="1"}
+$$
+  \mathbf{\Sigma} = \begin{bmatrix}
+      1 & 0.5 \\
+      0.5 & 1
+    \end{bmatrix}
+$$
+:::
+
+::: {.fragment fragment-index="2"}
+$$
+  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
+$$
+:::
+
+::: {.fragment fragment-index="2"}
+$$
+  cor (\mathbf{\theta}^*) =
+    \mathbf{\Sigma}^* =
+    \begin{bmatrix}
+      1 & 0 \\
+      0 & 1
+    \end{bmatrix}
+$$
+:::
+::::::
+
+::::: column
+:::: {.r-stack .fragment .fade-in fragment-index="1"}
+```{r corr-traits-example}
+#| fig-width:  5.5
+#| fig-height: 5.5
+
+set.seed(7117)
+
+corr_matrix <- matrix(c(1, .5, .5, 1), nrow = 2)
+
+thetas <- rmvnorm(10^4, sigma = corr_matrix) |>
+  as_tibble(.name_repair = ~paste0("trait_", 1:2)) |>
+    rownames_to_column("i")
+
+thetas |>
+  ggplot(mapping = aes(trait_1, trait_2)) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks       = element_blank(),
+    axis.line        = element_blank(),
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
+  )
+```
+
+::: {.fragment .fade-in fragment-index="4"}
+```{r corr-traits-transf-example}
+#| fig-width:  5.5
+#| fig-height: 5.5
+
+# This is a square root of the correlation matrix, then inverted and transposed:
+orth_matrix <- c(1, 0, .5, sqrt(.75)) |> matrix(nrow = 2) |> solve() |> t()
+
+thetas_transf <- thetas |>
+  pivot_longer(starts_with("trait")) |>
+  group_by(i) |>
+  mutate(value = orth_matrix %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
+thetas_transf |>
+  ggplot(mapping = aes(trait_1, trait_2)) +
+  geom_vline(xintercept = 0, linewidth = GRID_WIDTH) +
+  geom_hline(yintercept = 0, linewidth = GRID_WIDTH) +
+  geom_point(alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])) +
+  scale_x_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*^*^~2~", labels = NULL)
+  ) +
+  scale_y_continuous(
+    minor_breaks = NULL,
+    limits = c(-4, 4),
+    expand = expansion(),
+    name   = NULL,
+    sec.axis = dup_axis(name = "*&theta;*^*^~1~", labels = NULL)
+  ) +
+  coord_fixed() +
+  theme(
+    axis.ticks       = element_blank(),
+    axis.line        = element_blank(),
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
+  )
+```
+:::
+::::
+:::::
+::::::::::
+
+::: notes
+So, let's say $\theta$ is multivariate normal with covariane matrix $\mathbf{\Sigma}$.
+
+\[NEXT\]
+
+For the sake of the example, let's say it has unitary variances with a 0,5 correlation.
+
+\[NEXT\]
+
+Now, we can apply a "transformation" in the latent space that results in an identity covariance matrix.
+
+\[NEXT\]
+
+And if we want the geometrical space to be a representation of the latent space, the meaningful way is to represent these coordinates in an orthonormal basis (the covariane matrix is an identity matrix, the basis also forms an identity matrix).
+:::
+
+## Oblique axes {.smaller}
+
+<br>
+
 :::::: columns
@@
 #| fig-height: 5.5
+
+ts_basis <- cbind(
+  c(1, 0),
+  c(sqrt(.25), sqrt(.75))
+)
+ls_basis <- t(solve(ts_basis))
+
+thetas_transf <- thetas |>
+  group_by(i) |>
+  pivot_longer(starts_with("trait")) |>
+  mutate(value = ls_basis %*% value |> drop()) |>
+  pivot_wider() |>
+  ungroup()
+
 thetas_transf |>
@@
 #| fig-width: 5
-ts_basis <- cbind(
-  c(1, 0),
-  c(sqrt(.25), sqrt(.75))
-)
-
 ts_grid <- transform_grid(ts_basis)
@@
 #| fig-width: 5
-ls_basis <- t(solve(ts_basis))
-
 ls_grid <- transform_grid(ls_basis)
@@
 
-::: fragment
 [Answers in 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
-:::
 
@@
 So, here are three possible options; one of them is the correct one.
-
-\[NEXT\]
 
@@
 #| fig-width: 5
-thetas_transf <- thetas |>
+thetas_transf_rot <- thetas |>
   group_by(i) |>
@@
   geom_point(
-    data    = thetas_transf,
+    data    = thetas_transf_rot,
     mapping = aes(trait_1, trait_2),
@@
 #| fig-width: 5
-thetas_transf <- thetas |>
+thetas_transf_wrong <- thetas |>
   group_by(i) |>
@@
   geom_point(
-    data    = thetas_transf,
+    data    = thetas_transf_wrong,
     mapping = aes(trait_1, trait_2),
@@
 #| fig-width: 5
-thetas_transf <- thetas |>
-  group_by(i) |>
-  pivot_longer(starts_with("trait")) |>
-  mutate(value = ls_basis %*% value |> drop()) |>
-  pivot_wider() |>
-  ungroup()
-
 ls_scatter <- ls_grid +

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add "latent space vs. test space" slides

@@
 ...as in the coordinate axes C.
+
+But how is our intuition so wrong then?
+Well, the truth is that it is not.
 :::
@@
 
-> It has been suggested by some researchers that the angle between the axes represents the degree of correlation.
-> However, for sake of clarity, an orthogonal axes system is used in which distance measures and vectors can be easily calculated, understood, and interpreted.
-
----@ackerman_multidimensional_2005-1 [p. 16]
-
-::::::::: columns
-::::: column
+<br>
+
+::::::: columns
+:::: column
+$$
+  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
+$$
+
 ::: {.fragment fragment-index="1"}
 $$
-  P(Y_i = 1) = \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
-$$
-:::
-
-::: {.fragment fragment-index="3"}
-$$
   P(Y_i = 1) = \Phi_L ({\mathbf{a}^*_i}^T \mathbf{\theta}^* + d_i)
@@
 :::
-:::::
-
-::::: column
+::::
+
+:::: column
+$$
+  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
+$$
+
+::: {.fragment fragment-index="1"}
+$$
+  \mathbf{a}^*_i = \mathbf{Q} \mathbf{a}_i \\
+$$
+:::
+::::
+:::::::
+
 ::: {.fragment fragment-index="2"}
-$$
-  \mathbf{\theta}^* = \mathbf{P} \mathbf{\theta}
-$$
-:::
-
-::: {.fragment fragment-index="4"}
-$$
-  \mathbf{a}^*_i = \mathbf{Q} \mathbf{a}_i \\
-$$
-:::
-:::::
-:::::::::
-
-::: fragment
 $$
@@
 ::: notes
-Viendo que en la literatura hay afirmaciones como esta, no me resulta totalmente descorazonador.
-Parece que no soy el único al que le ha supuesto un esfuerzo entenderlo (resultados de la "encuesta"?).
-
-Pues bien, lo que ocurre en realidad es lo siguiente:
+The model probability must be invariant to a change of basis like this.
 
@@
 
-Este es el modelo de TRI multidimensional, recordaréis (ya lo presentamos, es como un modelo de regresión multivariada, donde los predcitores, $\theta$ son variables latentes).
-El modelo tiene que ser invariante; es decir, tiene que dar igual qué sistema de coordenadas usemos para representarlo (y esto por cierto también tiene su paralelismo en análisis factorial).
+So, the item vector must be transformed as well.
 
@@
 
-Así que si transformamos los rasgos latentes de las personas a un sistema de coordenadas ortogonal, los ítems tienen que transformarse también.
-
-\[NEXT\]
-
-Pero recordaréis que llamé Q a su matriz de transformación, mientras que a la de los rasgos latentes la llamé P.
-
-\[NEXT\]
-
-Claro, para que se cumpla la invarianza, se tiene que cumplir esta igualdad: Este término debe ser el mismo, da igual el sistema de coordenadas que utilice, y por lo tanto P y Q no pueden ser la misma matriz.
-
-\[NEXT\]
-
-En realidad, para transformar los ítems hay que usar la matriz inversa (inversa traspuesta, más concretamente) de la matriz de transformación de los rasgos latentes.
+And its transformation must be the opposite of the latent trait transformation.
 :::
@@
 
-::::::: columns
-::::: column
-<center>Espacio latente</center>
-
-:::: r-stack
-```{r grid-latent-space}
-#| fig-width: 6.5
-ls_grid
-```
-
-::: fragment
+::::: columns
+::: column
+<center>Latent space</center>
+
 ```{r scatter-latent-space}
@@
 :::
-::::
-:::::
-
-::: {.column .fragment}
-<center>Espacio test</center>
+
+::: column
+<center>Test space</center>
 
@@
 :::
-:::::::
+:::::
 
 ::: notes
-Y esto nos ha permitido "descubrir" (o darnos cuenta, más bien) de que hay dos "espacios vectoriales" distintos, uno en el que se encuentran las coordenadas de rasgo latente (las puntuaciones factoriales),...
-
-\[NEXT\]
-
-...
-y otro distinto, que es el espacio en el que se encuentran las coordenadas de los ítems, y que es donde en realidad habría que representarlos.
-
-Porque en este espacio es donde los ángulos entre los ejes son los que verdaderamente son una representación geométrica de las correlaciones entre las variables latentes.
+As it turns out, this gives a space where the axes actually do represent faithfully the latent covariances.
 :::
@@
 
-@mahalanobis_reprint_2018:
-
-::::: columns
-::: {.column width="50%"}
-<center>$`r MAH_DIST_EQ`$</center>
-:::
-
-::: {.column .fragment width="50%"}
-$`r INNER_PROD_INV_COV_EQ`$
-:::
-:::::
-
-<br>
-
 ::: fragment
@@
 
----@zhang_theoretical_1999 [p. 221]
+--- @zhang_theoretical_1999 [p. 221]
 :::
@@
 ::: notes
-Esto da lugar a algo muy curioso, y es que la "norma" de un punto en el espacio latente (la longitud de un vector aplicado en el origen que termina en ese punto) viene determinada por la distancia de Mahalanobis (esto es una reimpresión de 1936, no penséis que es una cosa de anteayer).
-
-\[NEXT\]
-
-Y eso se debe a que, para que la transformación en el espacio latente se cumpla, esa matriz de transformación (la que llamababmos P) tiene que ser "raíz cuadrada" de la inversa de la matriz de covarianzas.
-
-\[NEXT\]
-
-Y, curiosamente también, aparece en la literatura, en el año 1999, en un artículo que no tiene nada que ver con esto.
-Lo menciona de soslayo y no profundiza en su deducción como hemos hecho aquí.
-Solamente lo utiliza para resolver un problema "parcial" que se encuentra en la deducción de un índice de dimensionalidad, distinto de los parámetros de los que estamos hablando.
-Pero podía haber obviado este problema y haber utilizado coordenadas ortogonales, sin importarle si había o no correlaciones entre los rasgos latentes, como ha hecho todo el mundo (antes y después que ellos).
-
-Así que, todos mis respetos por estos autores.
-
-Sin embargo claro, ellos estaban hablando de otra cosa, y sólo tocaban este tema de paso.
-Así que al final eso se quedó ahí, y nadie ha hecho caso en 25 años a lo que dijeron Zhang y Stout.
-La gente ha usado los modelos, hecho simulaciones, etc., usando distribuciones correlacionadas, y ha usado las fórmulas de Reckase para transformar entre parámetros multidimensionales y los parámetros "lineales" (podríamos llamarlos, los de la formulación original) del modelo.
-
-\[NEXT\]
-
-Y lo más importante: Esto nos demuestra efectivamente que tenemos dos espacios distintos: Uno para los parámetros de las personas (espacio latente) y otro para los ítems (espacio test).
-Y esto es así porque la operación del producto interno está definida de manera distinta en los dos espacios; si la matriz del producto interno es la inversa de las covarianzas en el espacio latente, entonces en el espacio test, donde la transformación del parámetro de discriminación es inversa (como recordaréis), pues la matriz del producto interno es la matriz de covarianzas.
+This is actually not new, it was already noted by Zhang & Stout, more than 25 years ago, to solve a different problem, but unfortunately they didn't provide a deduction as we do here.
+
+But the important thing here, is that we have two different spaces: One for the latent traits and one for the items.
+
+And knowing these...
 :::

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add "generalized parameters" slide

@@
 
-## Generalized multidimensional parameters
-
-<br>
-
+## Generalized MIRT parameters {.smaller}
+
+<br>
+
+::::: columns
+::: column
 $$
@@
 
+::: column
+```{r item-plot-out}
+#| fig-cap:   Item representation in non-orthonormal test space
+#| fig-align: center
+plot_oblique +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  theme(axis.ticks = element_blank()) ## TODO: Tema ggplot
+```
+:::
+:::::
+
 ::: notes
-Si tenemos en cuenta las correlaciones (más generalmente, covarianzas) de la distribución de puntuaciones factoriales, lo que obtenemos son estas expresiones para los parámetros multidimensionales.
-
-Muy importante, la distancia con signo y la discriminación multidimensional son iguales, no importa en qué espacio (latente, o test) queramos obtenerlas y representarlas.
-Pero los cosenos directores es necesario calcularlos en el espacio test, no en el espacio de puntuaciones latentes, como decía Reckase, porque este es el espacio que va a representar en realidad la relación entre las variables latentes.
-:::
-
-## Complete graphical representation
-
-```{r item-plot-out}
-#| fig-cap:   Orthogonal (left) and oblique (right) test space
-#| fig-align: center
-#| fig-asp:     .5
-#| fig-width: 13
-
-plot_orth <- plot_orth +
-  scale_color_manual(values = ITEMS_PALETTE) +
-  theme(
-    axis.ticks       = element_blank(),
-    legend.title     = element_blank(),
-    panel.grid.minor = element_blank()
-  )
-
-plot_oblique <- plot_oblique +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  theme(axis.ticks = element_blank())
-
-plot_orth + plot_oblique ## TODO: Tema ggplot
-```
-
-## Conclusions
-
-<br>
-
-Generalization of multidimensional parameters
-
--   To constant-monotony items
-
--   To oblique axes
-
-<br>
-
-::: fragment
-Distinction latent space - test space
-:::
-
-::: notes
-En conclusión, lo que hemos propuesto en este artículo consiste en una generalización:
-
--   A ítems con monotonía constante, creciente, decreciente o nula.
-    Esto se venía haciendo ya como hemos visto, pero creemos que hacía falta formalizarlo.
-    No basta con hacerlo y punto, si Reckase dijo que las discriminaciones tiene que ser positivas, tenemos que demostrar que también vale para discriminaciones negativas o nulas (monotonía decreciente, o probabilidad constante).
-
--   A Ejes oblicuos, las cuales consideramos que son necesarias para representar distribuciones de puntuaciones de rasgo correlacionadas, en general.
-    Hemos visto que esto ya lo mencionaron Zhang y Stout, hace 25 años, pero no proporcionaron una "deducción" de estos parámetros multidimensionales, cosa que nosotros hacemos.
-
-\[NEXT\]
-
-Por último, es imortante recalcar este "descubrimiento" (o "redescubrimiento", más bien) del espacio test, ya que tenemos que ser muy conscientes de que los ítems y las personas están en espacios vectoriales distintos.
-Aunque esto parezca solamente un "artefacto matemático" (y en el fondo lo es), hemos visto que es verdaderamente crucial tenerlo en cuenta para entender bien los modelos que estamos utilizando.
-Además, esta "distinción" se halla también en los modelos factoriales: Existe lo que se denominan "ejes primarios" y "ejes de referencia", propuestos por Thurstone hace años (pero que prácticamente no se utilizan ni tienen en consideración).
-<!-- # TODO: Hace cuántos años?
-(Thurstone) --> Sin embargo, cuidado con esto, porque no está claro que esos espacios en TRI multidimensional se puedan asimilar directamente a los ejes en un modelo de AF.
+The complete definition and representation of the parameters is as you see here.
+
+\[IF THERE IS TIME\]
+
+You can see here:
+
+- Items in the first or third quadrant are "stretched"; their MDISC become larger, while the opposite happens in the other two quadrants (these two items have components with the same absolute values)
+
+- Simple items (in one of the axis) are rotated, but their MDISC is not affected.
+
+- Also, simple iems are not necessarily orthogonal to the other axis, while complex items can be.
 :::

### 2025-07-14 output/IMPS-2025-slideshow.qmd
Add concluding slides

@@
 -   M2PL model extensions: Graded scale, nominal response, 3PL...
-:::
-
-<br>
-
-::: fragment
+
+<br>
+
 -   Other models
@@
 ::: notes
-Esta de hecho puede ser una futura línea de investigación.
-Puede parecer evidente el paralelismo (o al menos a mí me lo parece), pero si una cosa me ha enseñado el estar trabajando en este problema de los parámetros (y toparme con el problema de los dos espacios), es que no hay que fiarse de las apariencias.
-Por eso aplicamos el método científico (el método puramente deductivo, en este caso)
+This opens up other interesting questions, such as how this mirrors results from the factor analytic theory,...
 
@@
 
-El modelo multidimensional logístico de 2 parámetros se puede considerar como "anidado", o un caso más simple de varios modelos.
-Así que partiendo de este como base, lo interesante es aplicarlo a otros más complejos, pero más útiles en la práctica.
-Por cierto, que esto también lo hace mucha gente, aplica las fórmulas de Reckase a otros modelos y a correr, se quedan "tan anchos".
-
-El modelo de escala graduada, por ejemplo, sería el equivalente al modelo de AF para respuestas ordinales, con varios umbrales o "posiciones" en TRI.
-
-El de crédito parcial, por ejemplo, puede servir también para formatos con varias opciones de respuesta que miden o pesan en factores diferentes cada una de ella.
-
-El logístico de 3 parámetros es como el de 2, pero añadiendo un parámetro de pseudoazar, para modelar el que un participante acierte un ítem por pura suerte, respondiendo "a ciegas" (esto sólo valdría para ítems cognitivos).
-
-Estos modelos, que son "extensiones" de este más básico, deberían partir de estos parámetros como fundamento.
-Pero un investigador no debería dar los parámetros sin más de ese modelo, sin proporcionar una prueba formal de que se aplica esa fórmula y se cumplen los supuestos.
-
-\[NEXT\]
-
-Y por supuesto, existen muchos otros modelos psicométricos, muy útiles y aplicados en la investigación en Psicología y Educación, en los cuales no está claro aún cómo generalizar estos resultados al ámbito multidimensional.
-:::
+...and how this applies to other models, extensions of the multidimensional 2PL, or completely different ones.
+:::
+
+# The end
+
+<!-- TODO: Add contact info -->
 

### 2025-07-16 output/IMPS-2025-slideshow.qmd
Make notes and style changes
(Several channges altogether, too many!)

@@
     autodep:   true
+    cache:     true
 format:
@@
     fig-width:               6
+    footer:                  https://bit.ly/vote-coords
     incremental:             false
@@
     slide-number:            false
+    smaller:                 true
     theme:                   ../www/slideshow-assets/extra-styles.scss
@@
 
+## Participate!
+
+<center>
+
+![](../www/slideshow-assets/bit.ly_vote-coords.svg){width="400px"}
+
+<https://bit.ly/vote-coords>
+
+</center>
+
 ## Multidimensional IRT {.smaller}
@@
 ::: notes
-The Multidimensional 2PL model is one of many extensions of the 2PL model.
+The Multidimensional 2-Parameter Logistic model (M2PL) is one of many extensions of the 2PL model.
 
 It assumes that the probability of a correct response to an item is a function of the linear combination of the latent traits weighted by item discrimination parameters.
-(This here is a vector inner product; the "T" means that the discrimination vector is transposed).
-
-And it has this additional "intercept parameter"; so it is basically a logistic multivariate regression with latent predictors.
 
@@
 ## Multidimensional parameters {.smaller}
-
-<br>
 
@@
 ::: notes
-The problem with the M2PL is that its parameters do not have a clear interpretation in IRT terms.
+The M2PL has a problem though, in that its parameters do not have a clear interpretation in IRT terms.
 
@@
 $$
-  MID_i = \begin{Bmatrix}
+  MIL_i = \begin{Bmatrix}
     D_i &= \frac{-d_i}{MDISC_i} \\
@@
 
-::: {.column .fragment}
+::: column
 -   @ackerman_graphical_1996:
@@
 ::: notes
-...Reckase & McKinley define the item discrimination, which is an extension of the Pythagorean theorem.
-
-This discrimination can be used to define the item difficulty, finding a parameterization that parallels the unidimensional 2PL model.
-
-Note though that, in this case, the difficulty has two components: a "signed distance" from the origin, and a direction.
-
-\[NEXT\]
-
-It is easier to appreciate this with the graphical representation, that was proposed by Ackerman a few years later, where we can see that the item vector is applied at a point given by the difficulty (with its distance and direction), and has a length equal to the discrimination.
-The vector direction always goes through the origin, and the sign of the distance (positive in this case), which is the opposite of the intercept sign, will determine whether the vector is displaced forward or backward along that direction.
+...Reckase & McKinley define the item discrimination, which you can see is like a multidimensional version Pythagoras' theorem.
+
+This MDISC can be used to define the item difficulty (or location as we will call it here), finding a parameterization that parallels the unidimensional case.
+
+Only that, in this case, the location has two components: a "signed distance" from the origin, which is the opposite of the intercept sign, and a direction.
+
+The item vector is applied at a point given by the location (with its distance and direction), and has a length equal to the discrimination.
+The direction always goes through the origin, and the sign of the distance (positive in this case), will determine whether the vector is displaced forward or backward along that direction.
 :::
@@
 
-<br>
-
 @reckase_difficulty_1985:
@@
 ::: notes
-However, there are two assumptions for defining the multidimensional parameters.
-
-First, that the probability of a correct answer increases monotonically with each dimension.
-
-And second, this constrain you see there, which implies that the item is represented in an orthogonal basis.
-This is necessary for applying the "general Pythagorean theorem" as we saw before.
-
-So, our purpose is **to relax these two assumptions**.
+However, there are two assumptions for defining these parameters.
+
+First, that the probability of a positive response increases monotonically with each dimension.
+
+And second, this constrain you see there, which implies that the item is represented in orthogonal coodrinates.
+This is necessary for applying the "multidimensional Pythagoras' theorem" as we saw before.
+
+So, our purpose is **to relax these two assumptions**, which derive heavily from the traditional application of this model to the cognitive domain.
+
+But we want to apply multidimensional IRT to non-cognitive items as well, like personality measures, for example.
 :::
@@
 ## Increasing monotony {.smaller}
-
-<br>
 
@@
 
-::: {.column .fragment fragment-index="1"}
+::: column
 ```{r inverse-item-curve}
@@
 
-::::: {.columns .fragment fragment-index="2"}
+::::: {.columns .fragment}
 ::: column
@@
 
-::: {.fragment .r-center}
+<center>
+
+::: fragment
 ### Generalization: "Constant monotony"
@@
 
+</center>
+
 ::: notes
-Because they derive from the traditional application of the model to the cognitive domain, but we want to apply multidimensional IRT to non-cognitive items as well, e.g. to personality measures.
+In these cases, we can have items with "decreasing" probability.
 
@@
 
-Here, we can have items with "decreasing" probability.
-
-\[NEXT\]
-
-Formally, what this would imply is a negative discrimination parameter.
-
-In the multidimensional case, this may imply a mix of positive and negative discrimination parameters, even in the same item.
+Which formally implies a negative discrimination parameter.
+
+In the multidimensional case, we may have a mix of positive and negative discrimination parameters.
 :::
@@
 ## Constant monotony {.smaller}
-
-<br>
 
@@
 ::: {style="color:#5c6eb1; padding-left:300px;"}
-$d_1 = 0$
+$D_1 = 0$
 :::
@@
 ::: {style="color:#90214a; padding-left:300px;"}
-$d_2 = -0.5$
+$D_2 = 0.35$
 :::
@@
 ::: {style="color:#749f4c; padding-left:300px;"}
-$d_3 = -0.25$
+$D_3 = 0.18$
 :::
@@
 ::: {style="color:#d76f47; padding-left:300px;"}
-$d_4 = 0.25$
+$D_4 = -0.18$
 :::
@@
 ::: notes
-The generalization is straightforward: The MDISC implies squared values of the discrimination components (this is the inner product expanded), so it is always positive regardless of the sign of the discrimination components.
-
-The signed distance is also the same, and the direction only needs to take into account the sign of each component when computing that corresponding cosine.
+The generalization is straightforward: The MDISC implies squared values of the discrimination components (this is the previous inner product expanded), so it is always positive.
+
+The signed distance is the same, and the direction only needs to take into account one sign, that of the corresponding component for computing each cosine.
 
@@
 
-Graphically, we can see that each item points in a direction given by its discrimination components, including their signs (so, 1 1, 1 -1, etc.).
+We can see that each item points in a direction given by its discrimination components with their signs (so, 1 1, 1 -1, etc.).
 
@@
 
-This seems obvious enough, but we must explicitly state that this definitions are valid in these cases, and that they have the same interpretation.
+This seems obvious enough, but we must explicitly state that this definitions are valid in these cases, and that their interpretation do not change.
 
@@
 ## Orthogonality {.smaller}
-
-<br>
 
@@
 
-<br>
-
 ::::: columns
@@
 
-<br>
-
 ::::: columns
 ::: column
-```{r item-oblique-axes-4}
+```{r item-oblique-axes-3}
 #| fig-align:  center
@@
 
+<center>
+
 ::: {.fragment .r-center}
@@
 
+</center>
+
 ::: notes
@@
 
-¿Why? Because we want this space to have a meaningful interpretation; we want it to be a geometrical representation of the probabilistic properties of our sample.
+¿Why?
+Because we want this space to have a meaningful interpretation; we want it to be a geometrical representation of the probabilistic properties of our sample.
 :::
@@
 ## Oblique axes {.smaller}
-
-<br>
 
@@
 $$
-  cor (\mathbf{\theta}) = \mathbf{\Sigma}
+  cov (\mathbf{\theta}) = \mathbf{\Sigma}
 $$
@@
 $$
-  cor (\mathbf{\theta}^*) =
+  cov (\mathbf{\theta}^*) =
     \mathbf{\Sigma}^* =
@@
 ::: notes
-So, let's say $\theta$ is multivariate normal with covariane matrix $\mathbf{\Sigma}$.
+So, let's say our latent trait vector is multivariate normal with covariane matrix $\mathbf{\Sigma}$.
 
@@
 
-And if we want the geometrical space to be a representation of the latent space, the meaningful way is to represent these coordinates in an orthonormal basis (the covariane matrix is an identity matrix, the basis also forms an identity matrix).
+And if we want the geometrical space to be a representation of the latent space, the meaningful way is to represent these coordinates in an orthonormal basis (the covariance matrix is an identity matrix, the basis of this coordinate system also forms an identity matrix).
+
+But, if this is the coordinate system of $\mathbf{\theta^*}$, if we want to draw here the axes of $\mathbf{\theta}$, they cannot be these same ones, right?
+If they were, then the point cloud would be the ellipse we had before.
 :::
@@
     name   = NULL,
-    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~}", labels = NULL)
+    sec.axis = dup_axis(name = "{*&theta;*~1~, *&theta;*~2~} ?", labels = NULL)
   ) +
@@
 fa(
-  "question",
+  "arrow-right",
   fill           = "#929acb",
   height         = "200px",
-  vertical_align = "-10em",
-  margin_left    = "4em"
-)
-```
-:::
-
-::: {.column .fragment width="33%" fragment-index="2"}
+  vertical_align = "-9.5em",
+  margin_left    = "2em"
+)
+```
+:::
+
+::: {.column .fragment width="33%" fragment-index="1"}
 ```{r corr-traits-transf-example}
@@
 
+<br>
+
 <center>
 
-::: {.fragment fragment-index="2"}
-Which coordinate system *θ* makes *θ\** "spherical"?
+::: {.fragment fragment-index="1"}
+$$
+  \mathbf{\Sigma} = \begin{bmatrix}
+      1 & 0.5 \\
+      0.5 & 1
+    \end{bmatrix}
+$$
 :::
@@
 
-In other words,...
+In other words: What is the coordinate system of $\mathbf{\theta}$...
 
@@
 
-...what is the coordinate system of $\mathbf{\theta}$...
-
-\[NEXT\]
-
 ...that makes $\mathbf{\theta^*}$ spherical?
@@
 
-## Oblique axes
+## Oblique axes {.smaller}
 
@@
 
-<br>
-
 :::::: columns
@@
 
-[Answers in 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
+<center>$$
+  \mathbf{\Sigma} = \begin{bmatrix}
+      1 & 0.5 \\
+      0.5 & 1
+    \end{bmatrix}
+$$</center>
 
@@
 
-Take a few seconds to answer and we will see next which is the correct one.
+Remember that in this example, the covariance matrix has unitary variance and .5 correlation.
 :::
@@
 
+<center>
+
+![](../www/slideshow-assets/bit.ly_vote-coords.svg){width="400px"}
+
+<https://bit.ly/vote-coords>
+
+</center>
+
+## Oblique axes {.smaller}
+
+:::::::::::::::::::::::::::::::::::::::: columns
+::::::::::::::::::::::::::::::::::::::: {.column width="33%"}
+<center>
+
+A?`<`{=html}
+`/center>`{=html}
+
+```{r grid-orthogonal-rotation-spherical-option}
+#| fig-width: 5
+rot_basis <- cbind(
+  c(sqrt(.75), -sqrt(.25)),
+  c(sqrt(.25), sqrt(.75))
+)
+
+rot_grid <- transform_grid(rot_basis)
+
+rot_grid +
+    geom_point(
+    mapping = aes(trait_1, trait_2),
+    data    = thetas_transf,
+    alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])
+  )
+```
+
+:::
+
+:::::::::::::::::::::::::::::::::::::: {.column width="33%"}
+<center>
+
+B?`<`{=html}
+`/center>`{=html}
+
+```{r grid-ts-basis-spherical-option}
+#| fig-width: 5
+ts_grid <- transform_grid(ts_basis)
+
+ts_grid +
+    geom_point(
+    mapping = aes(trait_1, trait_2),
+    data    = thetas_transf,
+    alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])
+  )
+```
+
+:::
+
+::::::::::::::::::::::::::::::::::::: {.column width="33%"}
+<center>
+
+C?`<`{=html}
+`/center>`{=html}
+
+```{r grid-ls-basis-spherical-option}
+#| fig-width: 5
+ls_grid <- transform_grid(ls_basis)
+
+ls_grid +
+    geom_point(
+    mapping = aes(trait_1, trait_2),
+    data    = thetas_transf,
+    alpha = .33, shape = 16, colour = I(PALETTE_BLUE["normal"])
+  )
+```
+
+:::
+
+::::::
+
 <br>
+
+[Answers in 3, 2, 1...](https://wall.sli.do/event/dPunB4BMx3S8hfSjaFSuFh?section=ae77dd00-34fb-4600-9677-337a65b04b83)
+
+::: notes
+\[CLICK LINK AFTER A WHILE\]
+:::
+
+## Oblique axes {.smaller}
 
@@
 
-<br>
-
 ::::::: columns
@@
 
+<br>
+
 ::: {.fragment fragment-index="1"}
@@
 :::::::
+
+<br>
 
@@
 As it turns out, this gives a space where the axes actually do represent faithfully the latent covariances.
+
+But it is not the latent space, where the latent traits are represented.
+It is a specific space for the items, and we call it "test space".
+And we know it, because the inner products in both spaces have to be different from each other, for the model to be invariant to changes of bases, like the one you have just seen.
 :::
@@
 
+> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by $`r INNER_PROD_EQ`$.
+
+--- @zhang_theoretical_1999 [p. 221]
+
 <br>
-
-::: fragment
-> \[...\] the latent space should be regarded as a general Euclidean space with its *inner product* defined by $`r INNER_PROD_EQ`$.
-
---- @zhang_theoretical_1999 [p. 221]
-:::
 
@@
 ::: notes
-This is actually not new, it was already noted by Zhang & Stout, more than 25 years ago, to solve a different problem, but unfortunately they didn't provide a deduction as we do here.
-
-But the important thing here, is that we have two different spaces: One for the latent traits and one for the items.
+This is actually not new, it was already noted by Zhang & Stout, more than 25 years ago, to solve a different problem, but unfortunately they didn't provide a mathematical proof as we do here.
+
+As you see here, the latent space Gram matrix is the inverse covariance matrix (which results from the transformation matrix P, pre-multiplied by itself, P\^T P).
+The invariant distance then, in the latent space, is the Mahalanobis distance, which is the square root of the inner product of the latent trait vector with itself.
+And in the test space, the Gram matrix is the covariance distance.
+
+But the important thing here, is that we have two different spaces, and that the item representation must be in their own space, the test space.
 
@@
 
-## Generalized MIRT parameters {.smaller}
-
-<br>
+## Generalized M2PL parameters {.smaller}
 
@@
 $$
-:::
-
-::: column
+
+<br>
+
+$$
+  \mathbf{\Sigma} = \mathbf{S} \mathbf{R} \mathbf{S}
+$$
+:::
+
+::: {.column .fragment}
 ```{r item-plot-out}
@@
 
-\[IF THERE IS TIME\]
+The MDISC is the square root of the inner product of the discrimination vector; the signed distance does not change.
+
+And the direction cosines incorporate this term: It pre-multiplies the discrimination vector by the covariance matrix, and also has this "normalizing term" that is a diagonal matrix of standard deviations..
+
+\[IF THERE IS TIME; NEXT\]
 
@@
 ## To be continued...
-
-<br>
 
@@
 
-# The end
-
-<!-- TODO: Add contact info -->
-
 ## References {.smaller}
+
+::: {#refs}
+:::
+
+## The end
+
+<center>
+
+### Thanks for your attention!
+
+</center>
+
+::::: columns
+::: column
+![](../www/slideshow-assets/foto_daniel.jpg){width="200px" fig-align="center"}
+
+<center>Daniel Morillo, Ph.D.</center>
+:::
+
+::: column
+![](../www/slideshow-assets/foto_mario.jpg){width="200px" fig-align="center"}
+
+<center>Mario Luzardo, Ph.D.</center>
+:::
+:::::
+
+<br>
+
+<center>Contact: [danielmorillo.ac\@gmail.com](mailto:danielmorillo.ac@gmail.com){.email}</center>
+:::::::::::::::::::::::::::::::::::::
+::::::::::::::::::::::::::::::::::::::
+:::::::::::::::::::::::::::::::::::::::
+::::::::::::::::::::::::::::::::::::::::

### 2025-07-16 output/IMPS-2025-slideshow.qmd
Adjust title slide logos

@@
   <table>
-    <!-- # TODO: Fix vertical alignment in logos-->
     <tr>
-      <td style="width: 400px">
+      <td style="width: 290px; vertical-align: middle; text-align: center">
         [![](../www/slideshow-assets/logos_fac_psicologia.svg){height="60"}](https://www.uned.es/universidad/facultades/psicologia.html)
       </td>
-      <td style="width: 100px">
+      <td style="width: 270px; vertical-align: middle; text-align: center">
         [![](../www/slideshow-assets/logo_udelar.svg){height="120"}](https://udelar.edu.uy)

### 2025-07-16 output/IMPS-2025-slideshow.qmd
Correct html tags in latent space options

@@
 
-:::::::::::::::::::::::::::::::::::::::: columns
-::::::::::::::::::::::::::::::::::::::: {.column width="33%"}
-<center>
-
-A?`<`{=html}
-`/center>`{=html}
+:::::: columns
+::: {.column width="33%"}
+<center>A?</center>
 
@@
 ```
-
-:::
-
-:::::::::::::::::::::::::::::::::::::: {.column width="33%"}
-<center>
-
-B?`<`{=html}
-`/center>`{=html}
+:::
+
+::: {.column width="33%"}
+<center>B?</center>
 
@@
 ```
-
-:::
-
-::::::::::::::::::::::::::::::::::::: {.column width="33%"}
-<center>
-
-C?`<`{=html}
-`/center>`{=html}
+:::
+
+::: {.column width="33%"}
+<center>C?</center>
 
@@
 ```
-
-:::
-
+:::
 ::::::
@@
 <center>Contact: [danielmorillo.ac\@gmail.com](mailto:danielmorillo.ac@gmail.com){.email}</center>
-:::::::::::::::::::::::::::::::::::::
-::::::::::::::::::::::::::::::::::::::
-:::::::::::::::::::::::::::::::::::::::
-::::::::::::::::::::::::::::::::::::::::

### 2025-07-16 output/IMPS-2025-slideshow.qmd
Align contents in several slides and formulae
Fix oblique space plots
(The last one, repeated, is still not fixed, for unknown reasons)

@@
 
+<br>
+
 <center>
@@
 ::: {style="color:#5c6eb1; padding-left:300px;"}
-$D_1 = 0$
+$D_1 = \quad 0$
 :::
@@
 ::: {style="color:#90214a; padding-left:300px;"}
-$D_2 = 0.35$
+$D_2 = \quad 0.35$
 :::
@@
 ::: {style="color:#749f4c; padding-left:300px;"}
-$D_3 = 0.18$
+$D_3 = \quad 0.18$
 :::
@@
 ::: {style="color:#d76f47; padding-left:300px;"}
-$D_4 = -0.18$
+$D_4 = \; -0.18$
 :::
@@
 
-mirt_item_coords |>
+item_oblique_space <- mirt_item_coords |>
   ggplot(
@@
   )
+
+item_oblique_space
 ```
@@
 ::: column
-```{r item-oblique-axes-2}
+```{r item-oblique-axes-2-rep}
+item_oblique_space
 ```

### 2025-07-16 output/IMPS-2025-slideshow.qmd
Fix complete item representation plot

@@
 #| fig-align: center
-plot_oblique +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
-  theme(axis.ticks = element_blank()) ## TODO: Tema ggplot
+transform_grid(
+  ts_basis,
+  x_limits   = c(-2, 3),
+  y_limits   = c(-2, 2.8),
+  break_step = 1
+) +
+  geom_segment(
+    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
+    linejoin  = "mitre",
+    linewidth = VECTOR_WIDTH,
+    data      = items_oblique |> arrange(desc(item)),
+    mapping   = aes(
+      origin_transf_1, origin_transf_2,
+      xend  = end_transf_1, yend = end_transf_2,
+      color = item, fill = item
+    )
+  ) +
+  scale_color_manual(values = PALETTE, guide = NULL) +
+  coord_fixed(expand = FALSE, clip = "on")
 ```

### 2025-07-17 output/IMPS-2025-slideshow.qmd
Correct symbol in item response surface

@@
   type = 'surface',
-  hovertemplate = 'P(I<sub>i</sub> = 1) = %{z:.2f}<extra></extra>',
+  hovertemplate = 'P(Y<sub>i</sub> = 1) = %{z:.2f}<extra></extra>',
   showscale = FALSE,

### 2025-07-24 output/IMPS-2025-slideshow.qmd
Final notes tweaked
Clarify that the "transorm matrix" is an example

@@
 $$
+  \text{example:} \qquad
   \mathbf{\Sigma} = \begin{bmatrix}
@@
 The invariant distance then, in the latent space, is the Mahalanobis distance, which is the square root of the inner product of the latent trait vector with itself.
-And in the test space, the Gram matrix is the covariance distance.
+
+\[NEXT\]
+
+And in the test space, the Gram matrix is the covariance matrix.
 
@@
 ::: notes
-The complete definition and representation of the parameters is as you see here.
+The complete definition of the parameters is as you see here.
 
@@
 ## The end
-
-<br>
 
@@
 
-<br>
-
-<center>Contact: [danielmorillo.ac\@gmail.com](mailto:danielmorillo.ac@gmail.com){.email}</center>
+Contact: [danielmorillo.ac\@gmail.com](mailto:danielmorillo.ac@gmail.com){.email}
+
+Comments & questions welcome!
+
+Read the preprint: <https://bit.ly/mirt-preprint>

### 2025-09-08 output/Main_text.Rmd
Make Reviewer 4's grammar & style corrections

@@
 We apply Reckase's maximum discrimination point method to define them in an
-algebraic sense in an arbitrary basis, and then defining such basis to be a
+arbitrary algebraic basis.
+Then, we define such basis to be a
 geometrical representation of the measured construct.
@@
 [see @reckase_multidimensional_2009 for a comprehensive treatment of the topic].
-This fact is probably due to the different spotlights
+This fact is probably due to the different aspects
 these two theoretical programs put the focus on;
@@
 the latent covariance structure, and thus focuses on finding
-a structure matrix as simple as possible. MIRT literature, on the other hand,
+a structure matrix as simple as possible.
+The MIRT literature, on the other hand,
 intends to model responses to certain stimuli and estimate person scores,
-being the model interpretation secondary in relevance.
+with model interpretation secondary in relevance.
 Nevertheless, the confluence of the two theoretical approaches into a common
@@
 because the two dimensions are *perfectly aligned* and thus
-compound up in the second group,
+simply sum up in the second group,
 the item `r MDISC_SYM` should be the algebraic sum of both components,
@@
 
-On an unrelated basis, negative loadings/discrimination parameters
+Negative loadings/discrimination parameters
 are often irrelevant in MIRT applications in the cognitive testing domain.
@@
 that can be soundly applied to the correlated trait case; on the other, to
-items that yield monotonous response probabilities, but in the broader sense of
-unchanging monotony. Additionally, we aim to provide a method for
+items that yield monotonic response probabilities, but in the broader sense of
+unchanging monotonicity. Additionally, we aim to provide a method for
 graphically representing these items in the fashion of @ackerman_creating_1994
@@
 the other one, an application in marketing research [@tezza_modelo_2018].
-We round up with a discussion about
+We conclude with a discussion about
 the applicability and generalization of our results.
@@
 
-We must stop here and make a few remarks. First, as we are not constraining
+We pause here for a few remarks. First, as we are not constraining
 ourselves to cognitive traits,
@@
 
-Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
+Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as its
 Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
@@
 Reckase's [-@reckase_difficulty_1985] Equation 3 (p. 403)
-already assumes the orthonormality of the latent space, as we stated before.
+assumes the orthonormality of the latent space, as we stated before.
 
@@
 To compute the direction of the maximum slope,
-Reckase leverages on the property that the squared direction cosines sum up to 1
-(see Equation 7 therein).
+Reckase leverages the property that
+the sum of the squared direction cosines equals 1 (see Equation 7 therein).
 However, this assertion implicitly assumes orthogonal coordinates,
@@
 Before presenting the results about the parameters,
-we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
+we make a few remarks regarding the vector space $`r TEST_SPACE`$.
 As shown before, this space is an entity by itself,
@@
 
-We must make a few observations here.
+We make a few observations here.
 First, the transformation to the orthonormal bases of the two spaces allows
@@
 It is important to note though that this transformation does not have
-an also desirable property,
+another desirable property,
 which we have used to justify the covariance-based version:
@@
 
-However, formulating them in terms of the multidimensional parameters allows us
+However, formulating them in terms of the multidimensional parameters allows
 studying the items in terms of their multidimensional properties.
@@
 
+```{r compose-example-items-table}
+```
+
+```{r set-flextable-wd}
+#| cache: false
+
+# Necessary for officedown to find the template file:
+opts_knit$set(root.dir = DOC_DIR)
+```
+
+<br>
+
+```{r example-items-table-out}
+#| cache:   false
+#| tab_id:  example-items-table-out
+#| tab.cap: Item Parameters for the Graphical Representation Example
+item_params_output
+```
+
+<br>
+
 Figure \@ref(fig:item-plot-out) illustrates the effect of
@@
 
-```{r compose-example-items-table}
-```
-
-```{r set-flextable-wd}
+```{r compose-oblique-plot}
+```
+
+```{r compose-orthogonal-plot}
+```
+
+<br>
+
+```{r item-plot-out}
 #| cache:          false
-
-# Necessary for officedown to find the template file:
-opts_knit$set(root.dir = DOC_DIR)
-```
-
-<br>
-
-```{r example-items-table-out}
-#| cache:   false
-#| tab_id:  example-items-table-out
-#| tab.cap: Item Parameters for the Graphical Representation Example
-item_params_output
-```
-
-<br>
+#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
+#| fig.topcaption: true
+#| fig.height:     3.1
+#| fig.width:      6.65
+plot_orth + plot_oblique
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"))
+```
 
@@
 
-```{r compose-oblique-plot}
-```
-
-```{r compose-orthogonal-plot}
-```
-
-<br>
-
-```{r item-plot-out}
-#| cache:          false
-#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
-#| fig.topcaption: true
-#| fig.height:     3.1
-#| fig.width:      6.65
-plot_orth + plot_oblique
-
-ggsave(file.path(DOC_DIR, "Figure_1.svg"))
-```
-
 # Application to Examples From the Literature
@@
 actual empirical data, or they are a fictional example,
-created ad-hoc for simulation purposes),
+created ad hoc for simulation purposes),
 so all the discrimination parameters are positive.
@@
 items with  negative discrimination parameters.
-This is very convenient for exemplifying our results with
+This is convenient for exemplifying our results with
 items with opposite-sign discrimination parameters.
@@
 
+```{r set-cov-matrix-Tezza2018}
+# See Tezza et al., 2018, p. 927:
+# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
+# > foi de aproximadamente 0,4.
+COV_MATRIX_VALUE <- diag(4)
+DIMS_CORR        <- .4
+COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
+```
+
+```{r reset-root-wd}
+#| cache: false
+```
+
+```{r read-items-Tezza2018}
+items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
+```
+
+```{r compute-orthogonal-params}
+```
+
+```{r compare-orthogonal-params}
+#| results: hide
+```
+
+<!-- Computed item parameters in the orthonormal space match the ones in given
+in Tezza 2018, table 5 -->
+
+```{r compute-oblique-params}
+```
+
+```{r format-covariance-matrix}
+```
+
+```{r create-params-table}
+```
+
+```{r compose-output-table}
+```
+
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Tezza2018-out}
+#| cache:   false
+#| tab.id:  empirical-table-Tezza2018-out
+#| tab.cap: |
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Tezza et al. (2018)
+item_params_output |>
+  fontsize(size = 11, part = "body")
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
+
 The latter effect can be better noticed in another case, in which Reckase
@@
 
-```{r set-cov-matrix-Tezza2018}
-# See Tezza et al., 2018, p. 927:
-# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
-# > foi de aproximadamente 0,4.
-COV_MATRIX_VALUE <- diag(4)
-DIMS_CORR        <- .4
-COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
-```
-
-```{r reset-root-wd}
-#| cache: false
-```
-
-```{r read-items-Tezza2018}
-items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
-```
-
-```{r compute-orthogonal-params}
-```
-
-```{r compare-orthogonal-params}
-#| results: hide
-```
-
-<!-- Computed item parameters in the orthonormal space match the ones in given
-in Tezza 2018, table 5 -->
-
-```{r compute-oblique-params}
-```
-
-```{r format-covariance-matrix}
-```
-
-```{r create-params-table}
-```
-
-```{r compose-output-table}
-```
-
-@tezza_modelo_2018 provide us with another example that allows showcasing
+@tezza_modelo_2018 provide us with another example that showcases
 the effect of correlated dimensions on negative discrimination parameters.
@@
 `r DIMS_CORR`.
-With such latent space structure, items that strongly discriminate in any or
+With this latent space structure, items that strongly discriminate in either or
 both of these two dimensions change drastically their multidimensional
@@
 thus the decrease in their $`r MDISC_ITEM`$ parameter.
-The effect on the direction is also patent in the angles these items make with
+The effect on the direction is also apparent in the angles these items make with
 these dimensions, which change up to almost 20º, compared with at most 7º in the
@@
 
-<!---BLOCK_LANDSCAPE_START--->
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Tezza2018-out}
-#| cache:   false
-#| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: |
-#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
-#|   Tezza et al. (2018)
-item_params_output |>
-  fontsize(size = 11, part = "body")
-```
-
-<!---BLOCK_LANDSCAPE_STOP--->
-
-On the other hand, in cases such as items 12 or 59, which have
+On the other hand, cases such as items 12 or 59, which have
 relatively low discrimination parameters on those two dimensions
@@
 with that dimension.
-In this case, being the discrimination in dimension 4 so low, the situation is
+In this case, because the discrimination in dimension 4 so low, the situation is
 very similar to item 4 in Table \@ref(tab:example-items-table-out)
@@
 increase monotonically with an increase in ability on any dimension" (p. 411).
-However, we have discussed that this definition implicitly assumes
+However, this definition implicitly assumes
 the orthonormality of the latent space.
@@
 (1) to a non-orthonormal space that accounts for the covariance structure of
-the multivariate latent variable, and (2) to any case of unchanging monotony.
+the multivariate latent variable, and
+(2) to any case of unchanging monotonicity.
 The latter may seem superfluous, given that several examples in the literature
@@
 which is convenient for fidelity when representing the items.
-However, as it has been defined ad-hoc to fulfill this specific property,
+However, as it has been defined ad hoc to fulfill this specific property,
 it lacks the other desirable ones the covariance-based version has

### 2025-09-08 output/Response_to_action_editor_reviewers-2.Rmd
Add Review 2 response letter draft file

---
output:
  officedown::rdocx_document:
    mapstyles:
      First Paragraph: Body Text
    base_format: bookdown::word_document2
    reference_docx: ../www/Template.docx
    number_sections: no
    keep_md:         no
bibliography: ../www/Multidimensional-parameters-MCLM.bib
csl:          ../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| include: false

# Includes:

library(here)
library(knitr)
library(patchwork)
library(ggplot2)
library(rstudioapi)
library(svglite) # Used by {knitr} and {ggplot2} to render the figure to SVG;
                 #   added to prevent {renv} from reporting inconsistent state.
library(officer)

ROOT_DIR <- here()
DOC_DIR  <- file.path(ROOT_DIR, "output")

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      = 1200L,
  dev.args = list(type = "cairo-png"),
  tab.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE),
  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
)

# Output configuration options:
options(digits = 3)
```

```{r sources}
#| cache:   false
#| include: false
source("R/Formulae.R",  encoding = 'UTF-8')
source("R/Constants.R", encoding = 'UTF-8')
```

# Responses to the action editor:

###### 1) Minor grammatical errors/phrasing, as pointed out by Thissen’s (who is Reviewer 4 according to ScholarOne) Review 1 (that is attached) need to be addressed.

###### 2) I would like to see the graphics improved as suggested by Thissen's Review 2.  Labeling of axes is key for clarity.  And providing contour plots of the factor score ( latent variable) distributions will also help the reader’s understanding, especially for those readers coming from the IRT perspective.

###### 3) I would like to see additional comments in their closing paragraph on how testing practitioners should use or what insights can be obtained by reviewing the different representations of the latent ability space and the item space. I see the mention of multigroup equating/linking, but I would like to see more discussion about how the test results and subsequent uses could be affected.

# Responses to Reviewer 4:

###### One part of the factor model that is understated in this ms. is the population distribution of the factor scores (not estimates, but the latent variable values). That could be ameliorated by adding a few contours of the bivariate normal (0, 1, rho=0) (that is, circles :) to the left panel graphic. That population distribution is implicit in EFA, and the original Reckase work.

###### Then one could make another version of the left panel with contours for the rho=0.5 population distribution. Comparison of that with the previously-described version would show that the arrows representing discrimination do not have the same numerical meaning with respect to the variance (standard deviation) of the population distribution *in the direction of the arrows*, which is part of the point of all of this.

###### Two graphics with orthogonal axes but different population distributions are the way orthogonal vs. oblique solutions are represented in MIRT analyses. The ms. misses an opportunity to clarify that. Different population distributions on orthogonal axes are the way MIRT folks have thought about factor correlation; factor analysts have thought about 'rotation' and oblique axes (the left panel of Figure 1); it's two sides of the same coin.

###### This graphic is potentially confusing. In addition to the absence of axis labels mentioned above, it has oddly placed gray dotted grid lines. I would think it would have vertical gray dotted grid lines showing placement on the horizontal axis (x), and oblique gray dotted grid lines (orthogonal to y') showing placement on the (oblique) axis y'. With axis values -2 ... 0 ... +2 at the margins of *those* gridlines. Instead, it has the opposite: labeled horizontal grid lines that don't correspond to any axis values, and oblique grid lines labeled 0-1-2-3-4 that similarly don't correspond to any axis, but rather to how far away a point is from y'. Basically, the graphic would be easier to understand with labels on the axes, and two sets of grid lines orthogonal to those that are currently there (omitting those that are currently there).

###### That said, one could also add elliptical contours representing the population distribution to the left panel of Figure 1. That would be a little complicated to compute; it's a "squeezed in" ellipsoid due to the rotation of the y axis to y'. but when compared to a version of the right panel with ellipses for a correlation of 0.5, the arrows would represent the same numerical values of discrimination relative to the standard deviation of the population in the arrows' directions.

# References

### 2025-09-12 output/Main_text.Rmd
Correct introduction after reviewer suggestion
The dimensions do not "simply sum up" (that's what the discrimination components do)

@@
 one can intuitively say that,
-because the two dimensions are *perfectly aligned* and thus
-simply sum up in the second group,
+because the two dimensions are *perfectly aligned* in the second group,
 the item `r MDISC_SYM` should be the algebraic sum of both components,

### 2025-09-12 output/Response_to_action_editor_reviewers-2.Rmd
Add response to editor's comment 1

@@
 
+We have made most of the corrections suggested by Thissen (with only two exceptions, see below), which no doubt will improve the quality of our manuscript.
+We would also like to thank this reviewer for having the courtesy of performing this thorough grammar review; being the authors non-native English speakers, we have made the mistake of being overconfident on our language skills and automatic correction tools.
+This review has also helped setting our condifence to a more realistic level for future submissions.
+
+The only suggestions not accepted are:
+
+In p.
+4 (line 2), we think that "the two dimensions [...] _combine_ in the second group" does not reflect accurately the fact that the two dimensions are collapsed into a single one when the correlation is 1.
+We have therefore decided to just drop this clause, as it is redundant with the previous one which states that "the two dimensions are _perfectly aligned_".
+This sentence now reads:
+
+> [...] because the two dimensions are *perfectly aligned* in the second group,
+the item `r MDISC_SYM` should be the algebraic sum of both components, [...]
+
+In p. 13 (line 2), dropping, "the squared direction cosines sum ~up~ to 1" did not sound well, so instead we reformulated the sentence as:
+
+> To compute the direction of the maximum slope,
+Reckase leverages the property that
+the sum of the squared direction cosines equals 1 (see Equation 7 therein).
+[...]
+
+Finally, the sentence that was found difficult to understand in the abstract (p. 1, lines 5-7) has been rewritten as follows:
+
+>  [...]
+We apply Reckase's maximum discrimination point method to define them in an
+arbitrary algebraic basis.
+Then, we define such basis to be a
+geometrical representation of the measured construct. [...]
+
 ###### 2) I would like to see the graphics improved as suggested by Thissen's Review 2.

### 2025-09-12 output/Response_to_action_editor_reviewers-2.Rmd
Corrects grammar and errata
Drafts initial responses to Reviewer 4

@@
 We would also like to thank this reviewer for having the courtesy of performing this thorough grammar review; being the authors non-native English speakers, we have made the mistake of being overconfident on our language skills and automatic correction tools.
-This review has also helped setting our condifence to a more realistic level for future submissions.
+This review has also helped to set our confidence to a more realistic level for future submissions.
 
@@
 
+###### It is always best, many would say required, to label the axes.
+The lack of axis labels in the graphics of Figure 1 make the rest of these comments difficult to type, so for convenience I imagine the axes are labeled conventionally x and y in the left panel, and x and y' in the right panel, indicating that y' is a rotated representation of y.
+(In the ms., it would be best to tie the labeling notation to the algebra used.)
+
+We had overlooked labelling the axes and have done so in this new version.
+We thank once again the reviewer for pointing this out.
+
 ###### One part of the factor model that is understated in this ms.
@@
 That population distribution is implicit in EFA, and the original Reckase work.
+
+Our understanding of Reckase's original work is the same as the reviewer's.
+We think that the orthogonality (actually, orthonormality) assumption was implicit at the time of @reckase_difficulty_1985.
+Because of that, an orthonormal (not just orthogonal) basis was a sound assumption.
+However, as Multidimensional IRT theory, applied researchers started applying these and Reckase and McKinley's -@reckase_discriminating_1991 formulae without noticing they were applying them to a more general case (this is part of another work-in-progress manuscript).
 
@@
 
+We understand this graphic may be misleading, but the grid lines are properly placed in our version of the figure:
+Each point in the test space is intended to represent a coordinate vector, which is the displacement from the origin given by a linear combination of the basis vector, with the vector components being the coefficient of each basis vector.
+For example, the tip of the vector representing item 1 is at coordinate (+2, +2).
+This means the point is displaced twice the length of the first vector basis along its direction (i.e., along the horizontal axis), that is, two units along the horizontal axis, and then another _two units parallel to the direction of the oblique axis_.
+These displacements, parallel to the direction of the axess, is what the grid lines intend to represent.
+This representation is somewhat equivalent to a _pattern matrix_ representation in the factor model, while the one proposed by the reviewer would stand for the correlation of the item with each latent dimension, being somewhat equivalent to a _structure matrix_ representation.
+This can be more easily understood by comparing figures <!-- # TODO: complete --> and <!-- # TODO: complete --> in @harman_modern_1976 <!-- # TODO: complete -->.
+
 ###### That said, one could also add elliptical contours representing the population distribution to the left panel of Figure 1.

### 2025-09-12 output/Title_page.Rmd

@@
 
+# Data Availability Statement
+
+All the materials and code of this paper are available as
+a Software Heritage repository, under the following SWHID persistent identifier:
+
+[swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af](https://archive.softwareheritage.org/swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af)
+
 # Financial Support

### 2025-09-16 output/Title_page.Rmd
Add Review 2 acknowledgements

@@
 
-The authors would like to thank Patricia Recio-Saboya, Ph.D.
-(ORCID: [0000-0002-0530-4404](https://orcid.org/0000-0002-0530-4404))
+The authors would like to thank:
+[Patricia Recio-Saboya, Ph.D.](https://orcid.org/0000-0002-0530-4404),
 for her valuable guidance regarding multidimensional estimation
-in IRT software packages.
+in IRT software packages;
+[Jay Verkuilen, Ph.D.,](https://orcid.org/0000-0003-3546-3411),
+for his comments and suggestions regarding covariance whitening;
+David Thissen and three additional anonymous reviewers for
+their valuable comments and suggestions of early drafts of this manuscript.
 

### 2025-09-17 output/Response_to_action_editor_reviewers-2.Rmd
Complete explanation of grid lines

@@
 
-We understand this graphic may be misleading, but the grid lines are properly placed in our version of the figure:
-Each point in the test space is intended to represent a coordinate vector, which is the displacement from the origin given by a linear combination of the basis vector, with the vector components being the coefficient of each basis vector.
+We understand this graphic may be hard to interpret, but the grid lines are properly placed; the rationale is this:
+Each point in the test space is intended to represent a coordinate vector, which is the displacement from the origin given by a linear combination of the basis vectors, with the components of the coordinate vector being the coefficients of those basis vectors (please refer to the expression $`r DISCR_COORDS_ALT_EQ`$ in p.
+8 of the previous version of our manuscript, or p.
+<!-- # TODO: Complete --> in the new submission).
 For example, the tip of the vector representing item 1 is at coordinate (+2, +2).
-This means the point is displaced twice the length of the first vector basis along its direction (i.e., along the horizontal axis), that is, two units along the horizontal axis, and then another _two units parallel to the direction of the oblique axis_.
-These displacements, parallel to the direction of the axess, is what the grid lines intend to represent.
-This representation is somewhat equivalent to a _pattern matrix_ representation in the factor model, while the one proposed by the reviewer would stand for the correlation of the item with each latent dimension, being somewhat equivalent to a _structure matrix_ representation.
-This can be more easily understood by comparing figures <!-- # TODO: complete --> and <!-- # TODO: complete --> in @harman_modern_1976 <!-- # TODO: complete -->.
+This means the point is displaced twice the length of the first vector basis along its direction (i.e., along the horizontal axis), that is, two units along the horizontal axis, and then another two units parallel to the direction of the oblique axis.
+These displacements, parallel to the direction of the axes, is what the grid lines intend to represent.
+This representation is somewhat equivalent to a _factor pattern_ (i.e., description in linear combination terms, see @harman_modern_1976, p.
+267) representation in the factor model, while the one proposed by the reviewer would be somewhat equivalent to the _factor structure_ (i.e., description in terms of correlations with the latent dimensions) representation.
+Please see @harman_modern_1976 (p.
+265 and p.
+267, respectively) for the two representations.
+This can be more easily understood by comparing figures 4.2 (p. 61, a representation equivalent to the one we propose) and 12.8 (p. 266) in @harman_modern_1976.
 

### 2025-09-17 output/Letter_to_editor-2.Rmd
Add Review 2 "letter to editor"

--- 2025-05-28 output/Letter_to_editor.Rmd
@@
 
-June 2, 2025. <!-- TODO: Update date -->
+October XX <!-- # TODO: Complete -->, 2025.
 
@@
 
-Please find attached the updated version of our manuscript
+Please find attached our second review of the manuscript
 *"A generalized definition of multidimensional item response theory
 parameters"*.
-We hope that this new version lives up to the expectations of
-the action editor and the reviewers;
-we have put as much effort as we have been able to address all the comments and
-suggestions, which we believe have helped improve the quality of our paper.
-However, we acknowledge we may not have resolved satisfactorily all the action
-editor's and reviewers' concerns.
-Therefore, we would kindly ask you to forward to them our good will
-and predisposition to address any further comments or suggestions,
-or clarify futher on the previous ones.
+we have carefully taken into account Thissen's and your own
+review and suggestions, addressing all the comments and suggestions diligently.
+Where we have decided not to follow a certain suggestion,
+we have striven to explain our reasons in the clearest possible way.
+Also, besides addressing the comments,
+we have added an additional clarification
+about the "covariance matrix whitening",
+a terminology previously unknown to us.
 
-We thank you in advance for considering our reviewed manuscript.
-Please feel free to contact us again at my aforementioned email address
-(dvmorillo@psi.uned.es).
-We will be happy to address any further comments or suggestions
-the action editor or the reviewers may deem necessary.
+Once again, we thank you for considering our reviewed manuscript.
+Please feel free to contact us again regarding any further concerns at
+[dvmorillo@psi.uned.es](mailto:dvmorillo@psi.uned.es).
 

### 2025-09-17 output/IMPS-2025-slideshow.qmd
Externalize grid transformation code
Grid transformation code was repeated in the two slideshows. To use it across the project, I externalize the code to the script with the R "Output" functions

@@
 ```{r sources}
+#| cache:   false
+#| include: false
+
 source("R/Formulae.R",                  encoding = 'UTF-8')
@@
 
-```{r create-grid-function}
-create_grid <- function(basis,
-                        x_limits = c(-4, 4),
-                        y_limits = c(-4, 4),
-                        break_step = 2) {
-  # Argument checking and formatting: ----
-
-  ## Basis vectors tangents and norms
-  vec_tan <- basis[2, ] / basis[1, ]
-
-  # Main: ----
-
-  # Define grid limits as a function of the axes limits (in canonical metric):
-  grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
-    x_lim = if_else(x == x_limits[1], "inf", "sup"),
-    y_lim = if_else(y == y_limits[1], "inf", "sup")
-  )
-
-  # Select delimiting grid vertices, based on the basis vector directions
-  grid_bounds <- bind_rows(
-    h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
-    v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
-    .id = "grid"
-  ) |>
-    filter(xor(slope > 0, x_lim == y_lim)) |>
-    mutate(limit_intercept = y - slope * x) # Compute their intercept
-
-  # Transformed (1, 1), expressed in the canonical basis
-  unit_point <- basis %*% c(1, 1) |> drop() |>
-    as_tibble(.name_repair = "minimal") |>
-    add_column(name = c('x', 'y') |> paste0('_unit')) |>
-    pivot_wider(values_from = value)
-
-  # Grid specifications, in canonical basis
-  grid_specs <- grid_bounds |>
-    bind_cols(unit_point) |>
-    mutate(unit_intercept = y_unit - slope * x_unit) |>
-    select(-ends_with("unit")) |>
-    mutate(
-      n_units  = trunc(limit_intercept / unit_intercept),
-      n_breaks = n_units %/% break_step * break_step
-    )
-
-  # Create vertical grid data, if the second basis vector is completely vertical
-  if (is.infinite(vec_tan[2])) {
-
-    # Create vertical grid data
-    v_grid <- grid_bounds |>
-      filter(grid == "v") |>
-      mutate(
-        unit     = basis[1, 1],
-        n_units  = trunc(x / unit),
-        n_breaks = n_units %/% break_step * break_step
-      ) |>
-      select(grid, limit, n_breaks, unit) |>
-      pivot_wider(names_from = "limit", values_from = n_breaks) |>
-      reframe(coord = seq(from = inf, to = sup, by = break_step) * unit) |>
-      mutate(
-        grid      = 'v',
-        slope     = Inf,
-        intercept = NA_real_,
-        label_pos = coord,
-        ref       = as.character(!coord)
-      )
-
-    # Filter out non-valid grid values
-    grid_specs <- grid_specs |> filter(!is.infinite(slope))
-  }
-
-  # Create grid data
-  grid <- grid_specs |>
-    select(grid, limit, slope, unit_intercept, n_breaks) |>
-    pivot_wider(names_from = limit, values_from = n_breaks) |>
-    group_by(grid) |>
-    reframe(
-      slope,
-      coord = seq(from = inf, to = sup, by = break_step),
-      intercept = coord * unit_intercept
-    ) |>
-    mutate(
-      label_pos = if_else(
-        grid == "h",
-         intercept   + x_limits[1] * vec_tan[1],
-        (y_limits[1] - intercept)  / vec_tan[2]
-      ),
-      ref = as.character(!intercept)
-    )
-
-    if (is.infinite(vec_tan[2])) return(grid |> bind_rows(v_grid))
-
-    grid
-}
-
-plot_grid <- function(grid,
-                      x_limits = c(-4, 4),
-                      y_limits = c(-4, 4),
-                      break_step = 2,
-                      linetype_axis = 'solid',
-                      linetype_grid = 'dashed') {
-
-  # Argument checking and formatting: ----
-
-  ## Grid plot linetypes
-  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
-
-  ## Basis vectors tangents and norms
-  vec_tan <- grid |> distinct(grid, slope) |> deframe()
-
-  ## Axis title positions
-  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
-    (x_limits[2] - x_limits[1])
-  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
-    (y_limits[2] - y_limits[1])
-
-  # Main: ----
-
-  # Assign x axis label values and coordinates (to account for the vertical
-  #   special case)
-  x_labels <- grid |> filter(grid == "v") |> select(coord, label_pos)
-
-  # Create vertical grid geometry if the second vector is completely vertical
-  if (is.infinite(vec_tan[2])) {
-
-    geom_vgrid <- geom_vline(
-      mapping = aes(xintercept = coord, linetype = ref),
-      data = grid |> filter(grid == "v")
-    )
-
-    grid <- grid |> filter(grid == "h")
-  }
-
-  grid_plot <- grid |>
-    ggplot() +
-    geom_abline(
-      mapping = aes(
-        intercept = intercept,
-        slope     = slope,
-        linetype  = ref
-      )
-    )
-
-  if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid
-
-  grid_plot +
-    scale_x_continuous(
-      minor_breaks = NULL,
-      breaks = x_labels |> pull(label_pos),
-      labels = x_labels |> pull(coord),
-      limits = x_limits,
-      expand = expansion(),
-      name   = NULL,
-      sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
-    ) +
-    scale_y_continuous(
-      minor_breaks = NULL,
-      breaks = grid |> filter(grid == "h") |> pull(label_pos),
-      labels = grid |> filter(grid == "h") |> pull(coord),
-      limits = y_limits,
-      expand = expansion(),
-      name   = NULL,
-      sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
-    ) +
-    scale_linetype_manual(values = linetypes, guide = NULL) +
-    coord_fixed() +
-    theme(
-      axis.ticks   = element_blank(),
-      axis.line    = element_blank(),
-      axis.title.x       = element_markdown(hjust = y_axis_titpos),
-      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
-    )
-}
-
-transform_grid <- function(basis,
-                           x_limits = c(-4, 4),
-                           y_limits = c(-4, 4),
-                           break_step = 2,
-                           linetype_axis = 'solid',
-                           linetype_grid = 'dashed') {
-
-  # Argument checking and formatting: ----
-
-  # Main: ----
-  basis |>
-    create_grid(
-      x_limits   = x_limits,
-      y_limits   = y_limits,
-      break_step = break_step
-    ) |>
-    plot_grid(
-      x_limits   = x_limits,
-      y_limits   = y_limits,
-      break_step = break_step,
-      linetype_axis = 'solid',
-      linetype_grid = 'dashed'
-    )
-}
-```
-
 :::::: columns

### 2025-09-17 output/seminar-slideshow.qmd
Externalize grid transformation code
Grid transformation code was repeated in the two slideshows. To use it across the project, I externalize the code to the script with the R "Output" functions

@@
 )
-
-# Factor axis colors:
-F1_COLOR <- unname(PALETTE_BLUE["normal"])
-F2_COLOR <- unname(PALETTE_RASPBERRY["normal"])
 
@@
 ```
+
+::: notes
+-   Nombre de la presentación: Cero atractivo
+
+-   Nuria me pide un título más llamativo
+
+-   Lo mejor que puedo conseguir
+
+-   ¿Por qué es relevante?
+
+-   TRI muy usada en evaluación e investigación educativa
+
+-   Poco en investigación psicológica (muestras "pequeñas" en comparación)
+
+-   Pero hay un "paralelismo" entre ambas teorías (isomórficas, ante determinados supuestos)
+:::
+
+<!-- Slides "Diferencias AF - TRI" -->
+
+```{r irt-fa-table}
+af_tri_differences <- tribble(
+  ~AF,                            ~TRI,
+  "Relaciones entre variables",   "Comparación entre participantes",
+  "Cuestionarios",                "Pruebas (rendimiento)",
+  "Explicación",                  "Medición",
+  "Puntuación compuesta",         "Puntuación latente",
+  "Estructura simple",            "Diferencias individuales"
+)
+
+section_title <- "## Diferencias AF - TRI"
+
+for (slide in af_tri_differences |> nrow() |> seq_len()) {
+
+  cat( "\n", section_title, sep = '')
+
+  cat( "\n", "<br>", sep = '')
+
+  af_tri_differences |> slice_head(n = slide) |> kable() |> print()
+}
+```
+
+## AF: Estructura simple
+
+<br>
+
+```{r simple-complex-structure}
+ITEM_BASE_LOADING <- 1
+ITEM_DIST         <- 0.2
+
+# Generate data for simple and complex factorial structures
+simple_item_loadings <- tribble(
+  ~item, ~factor, ~loading,
+  'I1',  'F1',    ITEM_BASE_LOADING,
+  'I2',  'F1',    ITEM_BASE_LOADING + ITEM_DIST,
+  'I3',  'F1',    ITEM_BASE_LOADING + 3 * ITEM_DIST,
+  'I4',  'F2',    ITEM_BASE_LOADING - ITEM_DIST,
+  'I5',  'F2',    ITEM_BASE_LOADING + ITEM_DIST,
+  'I6',  'F2',    ITEM_BASE_LOADING + 2 * ITEM_DIST
+)
+complex_item_loadings <- simple_item_loadings |> bind_rows(
+  tibble(
+    item    = 'I3',
+    factor  = 'F2',
+    loading = ITEM_BASE_LOADING - 2 * ITEM_DIST # Secondary loading smaller
+  )
+)
+
+simple_item_loadings <- simple_item_loadings |>
+  complete(item, factor, fill = list(loading = 0))
+
+complex_item_loadings <- complex_item_loadings |>
+  complete(item, factor, fill = list(loading = 0))
+```
+
+::::: columns
+::: column
+```{r path-simple-structure}
+# TODO: Syntax from tibble
+simple_spec <- "
+  F1 =~ I1 + I2 + I3
+  F2 =~ I4 + I5 + I6
+
+  F1 ~~ F2
+"
+
+F1_COLOR <- unname(PALETTE_BLUE["normal"])
+F2_COLOR <- unname(PALETTE_RASPBERRY["normal"])
+
+item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
+factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
+
+simple_spec      |>
+  semPlotModel() |>
+  semPaths(
+    border.color = c(
+      item_colors,       # Item colors
+      F1_COLOR, F2_COLOR # Factor colors
+    ),
+    edge.color   = c(
+      item_colors, # Loading colors
+      "black"      # Correlation color
+    ),
+    residuals    = FALSE,
+    thresholds   = FALSE,
+    intercepts   = FALSE,
+    fixedStyle   = 1,
+    curve        = 3,
+    label.cex    = 1.6,
+    sizeMan      = 10,
+    sizeLat      = 14,
+    border.width = 3,
+    edge.width   = 3,
+    mar          = MARGINS
+  )
+```
+:::
+
+::: {.column .fragment}
+```{r scatter-simple-structure}
+LIM_INF <- -1.5
+LIM_SUP <-  2
+
+axis_lims <- c(LIM_INF, LIM_SUP)
+lim_range <- LIM_SUP - LIM_INF
+axis_pos  <- (-LIM_INF) / lim_range
+
+item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
+
+simple_item_loadings_wide <- simple_item_loadings |>
+  group_by(item)                                  |>
+  arrange(desc(loading))                          |>
+  mutate(main_factor = first(factor))             |>
+  pivot_wider(names_from = factor, values_from = loading)
+
+simple_item_loadings_wide |>
+  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
+  geom_point(size = I(4))                                          +
+  geom_text(
+    nudge_x =  .1,
+    nudge_y = -.08,
+    color   = I("black"),
+    size    = I(5)
+  )                                                                +
+  scale_x_continuous(
+    limits   = axis_lims,
+    breaks   = 0,
+    labels   = NULL,
+    name     = "F2",
+    position = "top"
+  )                                                                +
+  scale_y_continuous(
+    limits   = axis_lims,
+    breaks   = 0,
+    labels   = NULL,
+    name     = "F1",
+    position = "right"
+  )                                                                +
+  scale_color_manual(values = factor_colors, guide = NULL)         +
+  coord_fixed(expand = FALSE, clip = "on")                         +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_text(hjust = axis_pos),
+    axis.title.y.right = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = GRID_WIDTH
+    ),
+    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
+    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+  )
+```
+:::
+:::::
+
+::: notes
+Para darle más sentido al modelo factorial es común buscar la "estructura simple".
+Es decir, asumimos que la varianza de cada ítem está explicada solamente por un factor latente.
+Los pesos factoriales de un ítem en los demás factores son nulos.
+
+Si representamos los pesos factoriales como coordenadas, el ítem estará alineado con el eje correspondiente al factor en el que pesa.
+:::
+
+## AF: Estructura simple
+
+::: r-center
+```{r simple-structure-rotated}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+AXIS1_DEGREES <- -25
+AXIS2_DEGREES <-  30
+
+axis1_radians <- AXIS1_DEGREES |> as_units("degrees") |> set_units("radians")
+axis2_radians <- AXIS2_DEGREES |> as_units("degrees") |> set_units("radians")
+
+rot_matrix <- c(
+  cos(axis1_radians), cos(axis2_radians),
+  sin(axis1_radians), sin(axis2_radians)
+) |>
+  matrix(nrow = 2, byrow = TRUE)
+
+simple_item_loadings_rot <- simple_item_loadings |>
+  group_by(item) |>
+  mutate(
+    loading = rot_matrix %*% loading |> drop()
+  ) |>
+  left_join(simple_item_loadings_wide |> select(item, main_factor), by = "item")
+
+simple_item_loadings_rot_plot <- simple_item_loadings_rot |>
+  group_by(item)                                          |>
+  pivot_wider(names_from = factor, values_from = loading) |>
+  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
+  geom_text(
+    nudge_x = -.12,
+    color   = I("black"),
+    size    = I(5)
+  )                                                                +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "",
+  )                                                                +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = ""
+  )                                                                +
+  scale_color_manual(values = factor_colors, guide = NULL)         +
+  coord_fixed(expand = FALSE, clip = "on")                         +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = GRID_WIDTH
+    )
+  )
+
+simple_item_loadings_rot_plot + geom_point(size = I(4))
+```
+:::
+
+::: notes
+Ahora bien, la mayoría de procedimientos factoriales, exploratorios al menos, arrojan soluciones de este tipo: un primer factor que maximiza toda la varianza posible, el segundo (ortogonal al primeroo) maximiza la varianza restante, y si hay más, hacen lo mismo, dando ejes siempre ortogonales a los anteriores.
+:::
+
+## AF: Estructura simple
+
+<!-- # TODO: Añadir ángulos? -->
+
+::: r-center
+```{r simple-structure-rotated-axes}
+#| fig-align:  center
+#| fig-height: 8
+#| fig-width:  8
+
+rotated_axes <- tibble(
+  slope  = c(axis1_radians, axis2_radians) |> tan() |> drop_units(),
+  factor = c('F1', 'F2')
+)
+
+simple_item_loadings_rot_plot +
+  geom_abline(
+    data    = rotated_axes,
+    mapping = aes(slope = slope, intercept = 0),
+    color   = c(PALETTE_BLUE["light"], PALETTE_RASPBERRY["light"]),
+    linewidth = LINE_WIDTH
+  ) +
+  geom_point(size = I(4)) +
+  annotate(
+    "text",
+    x = -1.35, y = c(-.62, .78),
+    label = rotated_axes |> pull(factor),
+    size  = 9
+  )
+```
+:::
+
+::: notes
+Para obtener esa estructura simple, muchas veces necesitamos usar una rotación oblicua de los ejes.
+Cualquier procedimiento de rotación busca alinearse lo más posible con al menos algunos de los ítems, como vemos aquí.
+Pero esto es una situación ideal.
+Nunca van a quedar, en un modelo exploratorio, los ítems perfectamente alineados con un eje, dando cero exacto en los pesos de los factores que no le corresponden.
+En un modelo confirmatorio sí, forzamos que sea así, pero a costa de perder ajuste claro (y normalmente permitiendo también que los factores correlacionen).
+:::
+
+## AF: Estructura compleja
+
+<br>
+
+::::: columns
+::: column
+```{r path-complex-structure}
+# TODO: Syntax from tibble
+complex_spec <- "
+  F1 =~ I1 + I2 + I3
+  F2 =~ I3 + I4 + I5 + I6
+
+  F1 ~~ F2
+"
+
+edge_colors <- c(
+  F1_COLOR |> rep(3),
+  PALETTE_RASPBERRY["light"], # Cross-loading
+  F2_COLOR |> rep(3)
+)
+
+complex_spec     |>
+  semPlotModel() |>
+  semPaths(
+    border.color = c(
+      item_colors,       # Item colors
+      F1_COLOR, F2_COLOR # Factor colors
+    ),
+    edge.color   = c(
+      edge_colors, # Loading colors
+      "black"      # Correlation color
+    ),
+    residuals    = FALSE,
+    thresholds   = FALSE,
+    intercepts   = FALSE,
+    fixedStyle   = 1,
+    curve        = 3,
+    label.cex    = 1.6,
+    sizeMan      = 10,
+    sizeLat      = 14,
+    border.width = 3,
+    edge.width   = 3,
+    mar          = MARGINS
+  )
+```
+:::
+
+::: {.column .fragment}
+```{r scatter-complex-structure}
+LIM_INF <- -1.5
+LIM_SUP <-  2
+
+axis_lims <- c(LIM_INF, LIM_SUP)
+lim_range <- LIM_SUP - LIM_INF
+axis_pos  <- (-LIM_INF) / lim_range
+
+item_colors <- item_colors |> setNames(c('F1' |> rep(3), 'F2' |> rep(3)))
+
+complex_item_loadings_wide <- complex_item_loadings |>
+  group_by(item)                                    |>
+  arrange(desc(loading))                            |>
+  mutate(main_factor = first(factor))               |>
+  ungroup()                                         |>
+  pivot_wider(names_from = factor, values_from = loading)
+
+item3_spikes <- complex_item_loadings_wide      |>
+  filter(item == 'I3')                          |>
+  select(starts_with('F'))                      |>
+  add_column(type = "end")                      |> 
+  bind_rows(tibble(F1 = 0, F2 = 0))             |>
+  complete(F1, F2, fill = list(type = "start")) |>
+  filter(F1 != 0 | F2 != 0)
+
+item3_spikes <- item3_spikes                           |> 
+  bind_rows(item3_spikes |> slice(3))                  |>
+  mutate(n = paste0('F', (row_number() + 1) %% 2 + 1)) |>
+  pivot_wider(names_from = type, values_from = F1:F2)
+
+spike_colors = c(
+  'F1' = unname(PALETTE_BLUE['light']),
+  'F2' = unname(PALETTE_RASPBERRY['light'])
+)
+
+complex_item_loadings_wide |>
+  ggplot(mapping = aes(F1, F2, color = main_factor, label = item)) +
+  geom_segment(
+    data    = item3_spikes,
+    mapping = aes(F1_start, F2_start, xend = F1_end, yend = F2_end),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = spike_colors
+  )                                                                +
+  geom_point(size = I(4))                                          +
+  geom_text(
+    nudge_x =  .1,
+    nudge_y = -.08,
+    color   = I("black"),
+    size    = I(5)
+  )                                                                +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F2",
+    position = "top"
+  )                                                                +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F1",
+    position = "right"
+  )                                                                +
+  scale_color_manual(values = factor_colors, guide = NULL)         +
+  coord_fixed(expand = FALSE, clip = "on")                         +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_text(hjust = axis_pos),
+    axis.title.y.right = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = GRID_WIDTH
+    ),
+    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
+    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+  )
+```
+:::
+:::::
+
+::: notes
+Sin embargo, en ocasiones no es posible encontrar la estructura simple.
+Es cuando aparecen lo que llamamos "pesos cruzados".
+
+Esto da lugar a una "estructura compleja", que en realidad es más común que la estructura simple, en general.
+:::
+
+## AF: Estructura compleja
+
+<br>
+
+::::: columns
+::: column
+```{r path-method-factor}
+# TODO: Syntax from tibble
+method_spec <- "
+  F1 =~ I1 + I2 + I3 + I4 + I5 + I6
+  M1 =~ I4 + I5 + I6
+"
+
+edge_colors <- c(
+  F1_COLOR |> rep(6),
+  F2_COLOR |> rep(3)
+)
+
+method_spec      |>
+  semPlotModel() |>
+  semPaths(
+    border.color = c(
+      F1_COLOR |> rep(6), # Item colors
+      F1_COLOR, F2_COLOR  # Factor colors
+    ),
+    edge.color   = edge_colors, # Loading colors
+    residuals    = FALSE,
+    thresholds   = FALSE,
+    intercepts   = FALSE,
+    fixedStyle   = 1,
+    curve        = 3,
+    label.cex    = 1.6,
+    sizeMan      = 10,
+    sizeLat      = 14,
+    border.width = 3,
+    edge.width   = 3,
+    mar          = MARGINS
+  )
+```
+:::
+
+::: column
+```{r scatter-method-factor}
+method_item_loadings <- tribble(
+  ~item, ~factor, ~loading,
+  'I1',  'F1',     ITEM_BASE_LOADING,
+  'I1',  'M1',     0,
+  'I2',  'F1',     ITEM_BASE_LOADING + 2 * ITEM_DIST,
+  'I2',  'M1',     0,
+  'I3',  'F1',     ITEM_BASE_LOADING + 3 * ITEM_DIST,
+  'I3',  'M1',     0,
+  'I4',  'F1',    -ITEM_BASE_LOADING,
+  'I4',  'M1',     ITEM_BASE_LOADING - 2 * ITEM_DIST,
+  'I5',  'F1',     ITEM_BASE_LOADING + ITEM_DIST,
+  'I5',  'M1',    -ITEM_BASE_LOADING + 2 * ITEM_DIST,
+  'I6',  'F1',    -ITEM_BASE_LOADING - ITEM_DIST,
+  'I6',  'M1',    -ITEM_BASE_LOADING + 3 * ITEM_DIST
+)
+
+method_item_loadings                                      |>
+  pivot_wider(names_from = factor, values_from = loading) |>
+  ggplot(mapping = aes(F1, M1, label = item))              +
+  geom_point(size = I(4), color = F1_COLOR)                +
+  geom_text(
+    nudge_y = -.12,
+    color   = I("black"),
+    size    = I(5)
+  )                                                        +
+  scale_x_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "M1",
+    position = "top"
+  )                                                        +
+  scale_y_continuous(
+    limits = axis_lims,
+    breaks = 0,
+    labels = NULL,
+    name   = "F1",
+    position = "right"
+  )                                                        +
+  scale_color_manual(values = factor_colors, guide = NULL) +
+  coord_fixed(expand = FALSE, clip = "on")                 +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_text(hjust = axis_pos),
+    axis.title.y.right = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      color     = "black",
+      linewidth = GRID_WIDTH
+    ),
+    panel.grid.major.x = element_line(color = PALETTE_RASPBERRY["light"]),
+    panel.grid.major.y = element_line(color = PALETTE_BLUE["light"])
+  )
+```
+:::
+:::::
+
+::: notes
+Por ejemplo, si tenemos ítems directos e inversos, podemos modelar lo que llamamos "factores de método": Una variable latente para cada factor sustantivo (sólo uno en este caso), y por ejemplo uno para los ítems formulados en negativo.
+:::
+
+## Modelo factorial clásico
+
+<br>
+
+$$
+Y_1 = \tau_1 + \lambda_{11} F_1^p + \lambda_{21} F_2^p + ... + \lambda_{n1} F_n^p + U_1^p \\
+\vdots                                                                            \\
+Y_i = \tau_i + \lambda_{1i} F_1^p + \lambda_{2i} F_2^p + ... + \lambda_{ni} F_n^p + U_i^p
+$$
+
+<br>
+
+::: fragment
+$$
+Y_i = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
+$$
+:::
+
+::: notes
+Un modelo factorial no es más que una serie de regresiones lineales (una por cada variable observable) con variables latentes.
+Las variables "respuesta" o "criterio" es la variable observable, y su varianza está explicada o "predicha" por las variables latentes, que son los factores y los términos de error de cada ítem.
+Lo que ocurre es que aquí a "coeficientes de regresión" los llamamos "pesos factoriales" y al término error, "unicidad" o "varianza única".
+
+El producto de los pesos por las puntuaciones factoriales se puede expresar en "forma vectorial", de esta manera.
+Este término significa que el vector de puntuaciones factoriales está pre-multiplicado por el vector de pesos factoriales.
+Esta operación es lo que se llama un "producto escalar", que consiste simplemente en multiplicar dos a dos los términos de cada vector y sumar todos.
+:::
+
+## Modelo factorial clásico {.smaller}
+
+::::::: columns
+::::: column
+$$
+  Y_1 = \tau_1 + \mathbf{\lambda}_{1}^T \mathbf{F} + U_1
+$$
+
+$$
+  \mathbf{F}^1 =
+    \begin{bmatrix}
+      0.5 \\
+      1
+    \end{bmatrix}
+$$
+
+::: {.fragment fragment-index="1"}
+$$
+  \mathbf{\lambda}_{1} =
+    \begin{bmatrix}
+      -1 \\
+       0.5
+    \end{bmatrix}
+$$
+:::
+
+::: fragment
+$$
+  \begin{align}
+    \tau_1 &=  1 \\
+    U_1^1  &= -0.5
+  \end{align}
+$$
+:::
+:::::
+
+::: {.column .fragment fragment-index="2"}
+```{r fa_example_plot}
+#| cache: false
+AXIS_GRID <- -1:1
+
+# Generate data for simple and complex factorial structures
+fa_example <- tribble(
+  ~object, ~factor, ~loading,
+  'I~1~',  'F1',    -1,
+  'I~1~',  'F2',     0.5,
+  'F^1^',  'F1',     0.5,
+  'F^1^',  'F2',     1
+)
+
+fa_example_wide <- fa_example |>
+  pivot_wider(names_from = factor, values_from = loading)
+
+fa_example_spikes <- fa_example_wide |>
+  add_column(type = "end")           |> 
+  bind_rows(
+    tribble(
+      ~object, ~F1,  ~F2, ~type,
+      "I~1~",   0,   0.5, "start",
+      "F^1^",   0,   1,   "start",
+      "I~1~",  -1,   0,   "start",
+      "F^1^",   0.5, 0,   "start",
+    )
+  )
+
+fa_example_spikes <- fa_example_spikes |> slice(1:2)        |>
+  bind_rows(fa_example_spikes)                              |>
+  add_column(factor = paste0('F', 1:2 |> rep(2, each = 2))) |>
+  pivot_wider(names_from = type, values_from = F1:F2)
+
+fa_example_wide |>
+  ggplot(mapping = aes(F1, F2, label = object))             +
+  geom_hline(
+    yintercept = 0,
+    color      = PALETTE_BLUE["light"],
+    linewidth  = GRID_WIDTH
+  )                                                         +
+  geom_vline(
+    xintercept = 0,
+    color      = PALETTE_RASPBERRY["light"],
+    linewidth  = GRID_WIDTH
+  )                                                         +
+  geom_segment(
+    data        = fa_example_spikes,
+    mapping     = aes(
+      F1_start, F2_start,
+      xend  = F1_end, yend = F2_end
+    ),
+    inherit.aes = FALSE,
+    linewidth   = LINE_WIDTH,
+    linetype    = '43',
+    color       = spike_colors |> rep(each = 2)
+  )                                                        +
+  geom_point(size = I(4))                                  +
+  geom_richtext(
+    nudge_x = .15,
+    nudge_y = .10,
+    color   = I("black"),
+    size    = I(5),
+    lineheight = 1.2,
+    fill = NA,
+    label.color = NA
+  )                                                        +
+  scale_x_continuous(
+    limits       = axis_lims,
+    breaks       = AXIS_GRID,
+    name         = NULL,
+    minor_breaks = NULL,
+    sec.axis     = dup_axis(name = "F~2~", labels = NULL)
+  )                                                        +
+  scale_y_continuous(
+    limits       = axis_lims,
+    breaks       = AXIS_GRID,
+    name         = NULL,
+    minor_breaks = NULL,
+    sec.axis     = dup_axis(name = "F~1~", labels = NULL)
+  )                                                        +
+  scale_color_manual(values = factor_colors, guide = NULL) +
+  coord_fixed(expand = FALSE, clip = "on")                 +
+  theme(
+    axis.line          = element_blank(),
+    axis.ticks         = element_blank(),
+    axis.title.x       = element_markdown(hjust = axis_pos),
+    axis.title.y       = element_markdown(vjust = axis_pos),
+    axis.title.y.right = element_text(vjust = axis_pos),
+    panel.grid.major   = element_line(
+      linewidth = GRID_WIDTH,
+      linetype  = "17"
+    )
+  )
+```
+:::
+:::::::
+
+::: fragment
+$$
+  Y_1 = 1 + \left[ -1, 0.5 \right]
+    \begin{bmatrix}
+      0.5 \\
+      1
+    \end{bmatrix}
+    - 0.5 = 1 + (-0.5 + 0.5) - 0.5 = 0.5
+$$
+:::
+
+::: notes
+Veámoslo con un ejemplo.
+Tenemos el participante 1, en nuestro estudio, que tiene esas puntuaciones factoriales; usamos el superíndice para denotar al participante.
+Por ejemplo, en el primer rasgo, tiene un 0.5 (nos da igual la distribución pero supongamos que está estandarizada, luego está media desviación típica por encima de la media).
+En el segundo rasgo, tiene una puntuación que está una desviación típica por encima de la media.
+
+Los pesos factoriales del ítem 1 son estos: -1 para el primer rasgo (por lo cual, cuanto mayor sea la puntuación factorial en el primer factor, menor va a ser la respuesta observada en el ítem 1), y 0.5 para el segundo.
+
+Aquí vemos representados el ítem y la persona, en las mismas coordenadas cartesianas.
+Y luego el ítem tiene una intersección de 1, y la persona 1 tiene un -0.5 en el "error" de esa variable $I_1$.
+
+\[NEXT\]
+
+Su respuesta se calcularía así: Primero tenemos el parámetro de intersección que vale 1.
+La "T" del vector de pesos factoriales significa que está "traspuesto", es decir, que se pone horizontalmente (esto es una convención en álgebra matricial).
+El producto del vector de pesos por el de puntuaciones factoriales es (...).
+Y por último sumamos la puntuación de error de ese participante en la variable.
+Luego esta persona (el particpante 1) tiene la respuesta o valor observado 0.5 en esta variable observada.
+
+Lo mismo podríamos hacer con el resto de variables y de participantes, pero obviamente nosotros lo que tenemos son estas respuestas observadas, y lo que hacemos con el AF es estimar todos estos parámetros del modelo, en lugar de como lo hemos hecho ahora.
+
+Ahora bien, esta respuesta observada que hemos calculado aquí es cuantitativa.
+Tenemos costumbre de ver el análisis factorial en ese contexto, de variables cuantitativas (o que asumimos que lo son, por ejemplo en ítems Likert).
+Sin embargo, el llamado "AF de ítems" se aplica también a variables discretas ordinales (e idealmente se debería aplicar así a ítems tipo Likert, aunque a veces no se haga, por simplificar).
+:::
+
+## Análisis factorial de ítems {.smaller}
+
+::::::::: columns
+::::: {.column width="45%"}
+$$
+  y_i^* = \tau_i + \mathbf{\lambda}_i^T \mathbf{F} + U_i
+$$
+
+<br>
+
+::: {.fragment fragment-index="1"}
+$$
+  Y_i = \left\{ \begin{matrix}
+      1 &si &y_i^* \geq 0\\
+      0 &si &y_i^* <    0
+    \end{matrix} \right.
+$$
+:::
+
+<br>
+
+::: {.fragment fragment-index="2"}
+$$
+  \begin{align}
+    P(Y_i = 1) &= \int_0^{-\infty} \mathcal{N}(y_i^*) \delta y_i^* \\
+               &= \Phi_N \left(
+                 \frac{ \tau_i + \mathbf{\lambda}_i^T \mathbf{F} }
+                   { \sqrt{var(U_i)} }
+               \right)
+  \end{align}
+$$
+:::
+:::::
+
+::::: {.column width="55%"}
+::: {.fragment fragment-index="1"}
+```{r ifa-density}
+#| fig-asp:   0.5
+#| fig-width: 8
+
+PROB_AXIS      <- seq(-3.3, 3.3, by = .01)
+ERROR_POINTS   <- seq(-3, 3, length.out = 5)
+ITEM_LOADING   <-  0.45
+ITEM_INTERCEPT <- -0.3
+
+item_unique_var <-  1 - sqrt(ITEM_LOADING)
+
+latent_response <- tibble(
+  point    = ERROR_POINTS |> rep(500000),
+  response = rnorm(
+    length(point),
+    mean = ITEM_INTERCEPT + ITEM_LOADING * point,
+    sd   = sqrt(item_unique_var)
+  )
+)
+
+# Plot: Upper Panel - Linear regression with Gaussian error distributions
+latent_response |>
+  ggplot(
+    aes(x = response, y = point, fill = after_stat(x) > 0, group = point)
+  )                            +
+  geom_density_ridges_gradient(
+    scale          =  .2,
+    rel_min_height = 0.001,
+    color          = PALETTE_BLUE["normal"]
+  )                              +
+  labs(x = "y~i~^*^", y = NULL)  +
+  scale_y_continuous(
+    breaks       = ERROR_POINTS,
+    minor_breaks = NULL,
+    labels       = NULL,
+    limits       = range(PROB_AXIS),
+    expand       = expansion()
+  )           +
+  scale_x_continuous(breaks = 0) +
+  scale_fill_manual(
+    values = c(
+      `FALSE` = unname(PALETTE_BLUE["white"]),
+      `TRUE`  = unname(PALETTE_BLUE["normal"])
+    )
+  )                              +
+  theme(
+    axis.ticks.x = element_blank(),
+    axis.title.y = element_markdown(angle = 90),
+    legend.position = "none",
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH)
+  )                              +
+  coord_flip()
+```
+:::
+
+::: {.fragment fragment-index="2"}
+```{r ifa-curve}
+#| fig-asp:   0.5
+#| fig-width: 8
+
+probit_curve_data <- tibble(
+  score    = PROB_AXIS,
+  response = ITEM_INTERCEPT + ITEM_LOADING * score,
+  prob     = pnorm(response, sd = sqrt(item_unique_var))
+)
+
+# Plot: Lower Panel - Probit curve
+probit_curve_data |>
+  ggplot(mapping = aes(x = score, y = prob))                        +
+  geom_line(color = PALETTE_BLUE["normal"], linewidth = LINE_WIDTH) +
+  labs(x = "E(*y~i~^*^*)", y = "P(Y~i~ = 1)")                       +
+  scale_y_continuous(
+    breaks       = 0:1,
+    minor_breaks = NULL,
+    limits       = 0:1,
+    expand       = expansion(mult = c(0, .05)),
+    oob = oob_keep
+  )  +
+  scale_x_continuous(
+    breaks       = ERROR_POINTS,
+    minor_breaks = NULL,
+    limits = range(PROB_AXIS),
+    expand = expansion()
+  ) +
+  theme(
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(angle = 90),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
+  )
+```
+:::
+:::::
+:::::::::
+
+::: notes
+Lo que hace el análisis factorial de ítems es aplicar ese "modelo lineal" a variables ordinales, haciendo una generalización.
+(En este caso vamos a simplificarlo a variables dicotómicas; p.ej. ante un ítem como "Soy muy responsable" las opciones de respuesta son sólo V/F).
+Esto es una simplificación que rara vez se utiliza en cuestionarios no cognitivos, pero es interesante estudiarlo porque otros modelos pueden considerarse extensiones de este modelo.
+El AF de ítems asume una nueva "variable latente" lineal ($I_1^*$), con la misma fórmula anterior
+
+Después se aplica una "función de umbral": La respuesta es correcta o afirmativa, codificada como 1, si la variable latente es mayor (o igual) que 0, o incorrecta o negativa, codificada como 0, si es menor.
+Eso lo que da, en lugar de un respuesta observada (cuantitativa), es una "probabilidad" de responder correcta o "afirmativamente" a un ítem.
+:::
+
+## AF: Puntuaciones factoriales
+
+<br>
+
+:::::: columns
+:::: column
+Métodos:
+
+-   Regresión
+
+-   Bartlett
+
+::: {.fragment fragment-index="2"}
+-   **Máxima Verosimilitud**
+
+-   **Empirical Modal Bayes**
+:::
+::::
+
+::: {.column .fragment fragment-index="1"}
+```{r path-regression-scores}
+# item_colors   <- c(F1_COLOR |> rep(3), F2_COLOR |> rep(3))
+# factor_colors <- c('F1' = F1_COLOR, 'F2' = F2_COLOR)
+
+# NOTE: Manual call to qgraph:
+edge_list <- tribble(
+  ~rhs, ~lhs,
+  1,    7,
+  2,    7,
+  3,    7,
+  4,    8,
+  5,    8,
+  6,    8,
+  7,    8,
+  8,    7
+)
+vSize <- c(10 |> rep(6), 14 |> rep(2))
+
+qgraph(
+  edge_list,
+  labels = c("I1", "I2", "I3", "I4", "I5", "I6", "F1", "F2") |>
+    setNames(nm = _),
+  label.cex = 1.6,
+  shape = "square" |> rep(8),
+  vsize = vSize, 
+  vsize2 = vSize,
+  color = "background" |> rep(8),
+  border.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), F1_COLOR, F2_COLOR),
+  border.width = 3,
+  bidirectional = c(FALSE |> rep(6), TRUE, TRUE),
+  directed = TRUE |> rep(8),
+  curve = c(NA |> rep(6), 3, 3),
+  edge.color = c(F1_COLOR |> rep(3), F2_COLOR |> rep(3), "black"),
+  edge.width = 3,
+  lty = 1,
+  layout = matrix(
+    c(
+      -1, -0.6, -0.2, 0.2, 0.6, 1, -0.6, 0.6,
+      -0.5 |> rep(6), 0.5 |> rep(2)
+    ),
+    ncol = 2
+  ),
+  mar = c(12, 1.3, 8, 1.3)
+)
+```
+:::
+::::::
+
+<!--# TODO: Diagrama de CFA y de "CFA inverso" -->
+
+::: notes
+Una vez tenemos los parámetros del modelo factorial estimado, entre otras cosas, podemos aplicarlo para estimar las puntuaciones factoriales de los participantes.
+Para hacer esto hay varios métodos.
+
+El método de regresión y el de Bartlett se aplican en AF lineal y están implementados en SPSS.
+Consisten a grandes rasgos en calcular la puntuación factorial como una suma ponderada de las variables observables.
+Es decir, algo así como "darle la vuelta a las flechas" de causalidad.
+Lo que tiene más sentido, desde el punto de vista estadístico, es "estimar" qué puntuaciones factoriales hacen más probable para una participante dar las respuestas que ha dado.
+
+Eso es lo que hacen los dos métodos de abajo (SPSS no los implementa creo), Máxima Verosimilitud, o esimación Modal Bayesiana.
+Si habéis hecho esto con variables categóricas, entonces habéis aplicado la Teoría de Respuesta al Ítem, y a continuación veremos por qué.
+:::
+
+## Paralelismo AF - TRI
+
+<br>
+
+::::: columns
+::: column
+![](../www/slideshow-assets/cover_McDonald-1999.jpg){height="300px"}
+
+[@mcdonald_test_1999]
+:::
+
+::: column
+<!--# TODO: Citation by McDonald -->
+:::
+:::::
+
+::: notes
+En este libro, que se publica en 1999 (fijaos en el subtítulo de "unified treatment"), McDonald indica que estos modelos, factoriales, TRI, son a fin de cuentas "isomóficos": Sin ser exactamente iguales ni partir de los mismos supuestos, las predicciones que hacen uno y otro son indistinguibles en la práctica.
+:::
+
+## Teoría de respuesta al ítem {.smaller}
+
+Modelo logístico de 2 parámetros [@birnbaum_latent_1968]:
+
+::::::::::: columns
+:::::: {.column width="60%"}
+```{r irt-curves}
+#| fig-asp:   0.6
+#| fig-width: 8
+
+ITEMS_PALETTE <- c(
+  PALETTE_BLUE["normal"],
+  PALETTE_RASPBERRY["normal"],
+  PALETTE_APPLE["normal"]
+) |>
+  unname() |>
+  setNames("prob" |> paste(1:3, sep = '_'))
+
+ITEMS <- tribble(
+  ~item, ~a,   ~b,
+   1,     1,    0,
+   3,     1,    1.5,
+   2,     2.5,  0
+)
+
+logit <- function(x) 1 / (1 + exp(-x))
+irf   <- function(a, b, theta) logit(a * (theta - b))
+
+logit_curve_data <- tibble(
+  latent_trait = PROB_AXIS,
+  prob         = ITEMS |>
+    select(-item) |>
+    pmap(irf, theta = latent_trait) |>
+    bind_cols()
+) |>
+  unnest(
+    prob,
+    names_sep = '_'
+  ) |>
+  rename_with(str_remove, pattern = '\\.{3}')
+
+logit_curve_data_long <- logit_curve_data |>
+  pivot_longer(starts_with("prob"), names_to = "item", values_to = "prob")
+
+logit_curve_data_long |>
+  ggplot(
+    mapping = aes(
+      latent_trait, prob,
+      color = item,
+      group = after_scale(color)
+    )
+  )                                        +
+  geom_line(linewidth = LINE_WIDTH)        +
+  labs(x = "*&theta;*", y = "P(Y~i~ = 1)") +
+  scale_y_continuous(
+    breaks       = 0:1,
+    minor_breaks = NULL,
+    limits       = 0:1,
+    expand       = expansion(mult = c(0, .05))
+  )                                        +
+  scale_x_continuous(
+    breaks       = ERROR_POINTS,
+    minor_breaks = NULL,
+    limits = range(PROB_AXIS),
+    expand = expansion()
+  )                                        +
+  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+  theme(
+    axis.title       = element_markdown(),
+    axis.title.y     = element_markdown(angle = 90),
+    panel.grid.major = element_line(linetype = "17", linewidth = GRID_WIDTH),
+  )
+```
+
+<!-- TODO: Como ecuaciones (sin punto y coma?) -->
+
+::: r-stack
+[$a_1$ = 1.0; $b_1$ = 0.0]{style="color:#5c6eb1;"}
+:::
+
+::: r-stack
+[$a_2$ = 1.0; $b_2$ = 1.5]{style="color:#90214a;"}
+:::
+
+::: r-stack
+[$a_3$ = 2.5; $b_3$ = 0.0]{style="color:#749f4c;"}
+:::
+::::::
+
+:::::: {.column width="40%"}
+::: {.fragment fragment-index="1"}
+$$
+  P(Y_i = 1) = \Phi_L (a_i (\theta - b_i))
+$$
+:::
+
+<br>
+
+::: {.fragment fragment-index="2"}
+$$
+  \begin{align}
+    d_i &= -a_i b_i \\
+    P(Y_i = 1) &= \Phi_L (a_i \theta + d_i))
+  \end{align}
+$$
+:::
+
+<br>
+
+::: {.fragment fragment-index="3"}
+$$
+  \Phi_L (x) =  \frac{1}{1 + e^{-x}}
+$$
+:::
+
+<br>
+::::::
+:::::::::::
+
+::: notes
+La TRI, por su lado, asume que hay cada persona tiene una puntuación latente, y que su respuesta depende de esa puntuación, con un parámetro de discriminación que determina cómo de bien o mal mide el ítem, y uno de dificultad (aunque aquí prefiero llamarlo "de posición") que determina lo fácil o difícil que será dar una respuesta positiva (afirmativa, correcta, codificada como 1).
+Esa posición representa el nivel de rasgo latente que da lugar a una probabilidad de .5 de dar la respuesta positiva (en este modelo; en general representa el punto de inflexión de la curva, es decir, donde pasa de hacerse cada vez más pronunciada a empezar a aplanarse).
+
+Ahora bien, la posición se puede reformular como un parámetro de intersección.
+
+Como se puede ver, esto también se trata de un "modelo lineal general".
+En este caso, en lugar de una función probit (normal acumulada), la función de enlace es la función "logit".
+Es decir, la respuesta a cada ítem se modela como una regresión logistica donde los predictores son las variables latentes.
+Esto es lo que se llama "Función Característica del Ítem".
+:::
 
@@
 Pertenece a una familia de modelos llamados "compensatorios" en TRI, ya que un nivel bajo en una puntuación latente se compensa con un nivel alto en otra, dando lugar a la misma probabilidad de respuesta.
+:::
+
+## TRI multidimensional: Ejemplo {.smaller}
+
+:::::: columns
+::: column
+$$
+  P(Y_1 = 1) = \Phi_L (\mathbf{a}_1^T \mathbf{\theta} + d_1)
+$$
+
+```{r mirt-item-response-example}
+mirt_example <- tribble(
+  ~theta_1, ~theta_2,
+  1,        -1
+)
+
+mirt_example_spikes <- mirt_example |>
+  add_column(type = "end")          |> 
+  bind_rows(
+    tribble(
+      ~theta_1, ~theta_2, ~type,
+      0,        -1,        "start",
+      1,         0,        "start"
+    )
+  )
+
+mirt_example_spikes <- mirt_example_spikes |>
+  slice(1) |>
+  bind_rows(mirt_example_spikes) |>
+  add_column(factor = paste0('F', 1:2 |> rep(2))) |>
+  pivot_wider(names_from = type, values_from = starts_with("theta"))
+
+contour_plot +
+  geom_segment(
+    data    = mirt_example_spikes,
+    mapping = aes(
+      theta_1_start, theta_2_start,
+      xend = theta_1_end,
+      yend = theta_2_end
+    ),
+    inherit.aes = FALSE,
+    linewidth = LINE_WIDTH,
+    linetype  = '43',
+    color = PALETTE_BLUE["light"]
+  ) +
+  geom_point(
+    data = mirt_example,
+    mapping = aes(theta_1, theta_2),
+    inherit.aes = FALSE,
+    size = I(4)
+  )
+```
+:::
+
+:::: column
+\begin{matrix}
+  \mathbf{\theta} =
+    \begin{bmatrix}
+      1 \\
+      -1
+    \end{bmatrix}
+
+  &\begin{align}
+    \mathbf{a}_1 &=
+      \begin{bmatrix}
+        1 \\
+        0.5
+      \end{bmatrix} \\
+    d_1 &= -0.5
+  \end{align}
+\end{matrix}
+
+<br>
+
+::: fragment
+$$
+  \begin{align}
+    P(Y_1 = 1) &= \Phi_L \left([1, 0.5]
+      \begin{bmatrix}
+         1 \\
+        -1
+      \end{bmatrix}
+      -0.5 \right) \\
+      &= \Phi_L \left(1 · 1 + 0.5 · (-1) -0.5 \right) \\
+      &= \Phi_L \left(1 -0.5 -0.5 \right) = \Phi_L \left(0 \right) \\
+      &= \frac{1}{1 + e^{-0}} = \frac{1}{1 + 1} = \frac{1}{2} \\
+      &= 0.5
+  \end{align}
+$$
+:::
+::::
+::::::
+
+::: notes
+Veamos un ejemplo de cuál sería la probabilidad de respuesta de una persona con puntuaciones de rasgo de 1 y -1.
+
+(Hacer cálculos matriciales)
+:::
+
+## AF en métrica TRI {.smaller}
+
+<br>
+
+$$
+  \begin{align}
+    P(Y_i = 1) &= \Phi_N \left(
+      \frac{\mathbf{\lambda}_i^T \mathbf{F} + \tau_i}{\sqrt{var(U_i)}}
+    \right) =
+      \Phi_N \left(
+        \frac{1}{\sqrt{var(U_i)}} \mathbf{\lambda}_i^T \mathbf{F} +
+          \frac{1}{\sqrt{var(U_i)}} \tau_i
+      \right)\\
+    P(Y_i = 1) &= \Phi_L (\mathbf{a}_i^T \mathbf{\theta} + d_i)
+  \end{align}
+$$
+
+<br>
+
+::: fragment
+@takane_relationship_1987:
+
+$$
+  \begin{align}
+    \mathbf{a}_i &= \frac{1.7}{\sqrt{var(U_i)}} \mathbf{\lambda}_i \\
+    d_i &= \frac{1.7}{\sqrt{var(U_i)}} \tau_i
+  \end{align}
+$$
+:::
+
+::: notes
+Claro, si ponemos las ecuaciones del AF y de TRI lado a lado, es fácil ver que existe un paralelismo entre ambos modelos.
+
+Pues resulta que estas dos curvas, la normal y la logística, son muy parecidas; tanto, que es posible transformar los parámetros de la métrica del AF a métrica de TRI, obteniendo modelos indistinguibles en la práctica.
+Las dos curvas se diferencian por este factor de 1.7, pero aplicándo ese factor y escalando por la desviación estándar de la unicidad del ítem, obtenemos los mismos parámetros en métrica de TRI mulidimensional con función de respuesta logística.
+Y de hecho, esto es necesario y se hace así para hacer la estimación máximo-verosimil de la que hablábamos antes.
 :::
@@
 
-```{r create-grid-function}
-create_grid <- function(basis,
-                        x_limits = c(-4, 4),
-                        y_limits = c(-4, 4),
-                        break_step = 2) {
-  # Argument checking and formatting: ----
-
-  ## Basis vectors tangents and norms
-  vec_tan <- basis[2, ] / basis[1, ]
-
-  # Main: ----
-
-  # Define grid limits as a function of the axes limits (in canonical metric):
-  grid_box <- expand_grid(x = x_limits, y = y_limits) |> mutate(
-    x_lim = if_else(x == x_limits[1], "inf", "sup"),
-    y_lim = if_else(y == y_limits[1], "inf", "sup")
-  )
-
-  # Select delimiting grid vertices, based on the basis vector directions
-  grid_bounds <- bind_rows(
-    h = grid_box |> mutate(slope = vec_tan[1], limit = y_lim),
-    v = grid_box |> mutate(slope = vec_tan[2], limit = x_lim),
-    .id = "grid"
-  ) |>
-    filter(xor(slope > 0, x_lim == y_lim)) |>
-    mutate(limit_intercept = y - slope * x) # Compute their intercept
-
-  # Transformed (1, 1), expressed in the canonical basis
-  unit_point <- basis %*% c(1, 1) |> drop() |>
-    as_tibble(.name_repair = "minimal") |>
-    add_column(name = c('x', 'y') |> paste0('_unit')) |>
-    pivot_wider(values_from = value)
-
-  # Grid specifications, in canonical basis
-  grid_specs <- grid_bounds |>
-    bind_cols(unit_point) |>
-    mutate(unit_intercept = y_unit - slope * x_unit) |>
-    select(-ends_with("unit")) |>
-    mutate(
-      n_units  = trunc(limit_intercept / unit_intercept),
-      n_breaks = n_units %/% break_step * break_step
-    )
-
-  # Create vertical grid data, if the second basis vector is completely vertical
-  if (is.infinite(vec_tan[2])) {
-
-    # Create vertical grid data
-    v_grid <- grid_bounds |>
-      filter(grid == "v") |>
-      mutate(
-        unit     = basis[1, 1],
-        n_units  = trunc(x / unit),
-        n_breaks = n_units %/% break_step * break_step
-      ) |>
-      select(grid, limit, n_breaks, unit) |>
-      pivot_wider(names_from = "limit", values_from = n_breaks) |>
-      reframe(coord = seq(from = inf, to = sup, by = break_step) * unit) |>
-      mutate(
-        grid      = 'v',
-        slope     = Inf,
-        intercept = NA_real_,
-        label_pos = coord,
-        ref       = as.character(!coord)
-      )
-
-    # Filter out non-valid grid values
-    grid_specs <- grid_specs |> filter(!is.infinite(slope))
-  }
-
-  # Create grid data
-  grid <- grid_specs |>
-    select(grid, limit, slope, unit_intercept, n_breaks) |>
-    pivot_wider(names_from = limit, values_from = n_breaks) |>
-    group_by(grid) |>
-    reframe(
-      slope,
-      coord = seq(from = inf, to = sup, by = break_step),
-      intercept = coord * unit_intercept
-    ) |>
-    mutate(
-      label_pos = if_else(
-        grid == "h",
-         intercept   + x_limits[1] * vec_tan[1],
-        (y_limits[1] - intercept)  / vec_tan[2]
-      ),
-      ref = as.character(!intercept)
-    )
-
-    if (is.infinite(vec_tan[2])) return(grid |> bind_rows(v_grid))
-
-    grid
-}
-
-plot_grid <- function(grid,
-                      x_limits = c(-4, 4),
-                      y_limits = c(-4, 4),
-                      break_step = 2,
-                      linetype_axis = 'solid',
-                      linetype_grid = 'dashed') {
-
-  # Argument checking and formatting: ----
-
-  ## Grid plot linetypes
-  linetypes <- c(`TRUE` = linetype_axis, `FALSE` = linetype_grid)
-
-  ## Basis vectors tangents and norms
-  vec_tan <- grid |> distinct(grid, slope) |> deframe()
-
-  ## Axis title positions
-  y_axis_titpos <- (y_limits[2] / vec_tan[2] - x_limits[1]) /
-    (x_limits[2] - x_limits[1])
-  x_axis_titpos <- (x_limits[2] * vec_tan[1] - y_limits[1]) /
-    (y_limits[2] - y_limits[1])
-
-  # Main: ----
-
-  # Assign x axis label values and coordinates (to account for the vertical
-  #   special case)
-  x_labels <- grid |> filter(grid == "v") |> select(coord, label_pos)
-
-  # Create vertical grid geometry if the second vector is completely vertical
-  if (is.infinite(vec_tan[2])) {
-
-    geom_vgrid <- geom_vline(
-      mapping = aes(xintercept = coord, linetype = ref),
-      data = grid |> filter(grid == "v")
-    )
-
-    grid <- grid |> filter(grid == "h")
-  }
-
-  grid_plot <- grid |>
-    ggplot() +
-    geom_abline(
-      mapping = aes(
-        intercept = intercept,
-        slope     = slope,
-        linetype  = ref
-      )
-    )
-
-  if (is.infinite(vec_tan[2])) grid_plot <- grid_plot + geom_vgrid
-
-  grid_plot +
-    scale_x_continuous(
-      minor_breaks = NULL,
-      breaks = x_labels |> pull(label_pos),
-      labels = x_labels |> pull(coord),
-      limits = x_limits,
-      expand = expansion(),
-      name   = NULL,
-      sec.axis = dup_axis(name = "*&theta;*~2~", labels = NULL)
-    ) +
-    scale_y_continuous(
-      minor_breaks = NULL,
-      breaks = grid |> filter(grid == "h") |> pull(label_pos),
-      labels = grid |> filter(grid == "h") |> pull(coord),
-      limits = y_limits,
-      expand = expansion(),
-      name   = NULL,
-      sec.axis = dup_axis(name = "*&theta;*~1~", labels = NULL)
-    ) +
-    scale_linetype_manual(values = linetypes, guide = NULL) +
-    coord_fixed() +
-    theme(
-      axis.ticks   = element_blank(),
-      axis.line    = element_blank(),
-      axis.title.x       = element_markdown(hjust = y_axis_titpos),
-      axis.title.y.right = element_markdown(vjust = x_axis_titpos)
-    )
-}
-
-transform_grid <- function(basis,
-                           x_limits = c(-4, 4),
-                           y_limits = c(-4, 4),
-                           break_step = 2,
-                           linetype_axis = 'solid',
-                           linetype_grid = 'dashed') {
-
-  # Argument checking and formatting: ----
-
-  # Main: ----
-  basis |>
-    create_grid(
-      x_limits   = x_limits,
-      y_limits   = y_limits,
-      break_step = break_step
-    ) |>
-    plot_grid(
-      x_limits   = x_limits,
-      y_limits   = y_limits,
-      break_step = break_step,
-      linetype_axis = 'solid',
-      linetype_grid = 'dashed'
-    )
-}
-```
-
 <br>

### 2025-09-24 output/Main_text.Rmd
Fine tune Figure 1 after labelling axes

@@
 #| fig.topcaption: true
-#| fig.height:     3.1
+#| fig.height:     3
 #| fig.width:      6.65
+#| warning:        false
 plot_orth + plot_oblique

### 2025-09-24 R/Output.R

@@ -181,7 +181,7 @@ plot_grid <- function(grid,
                       linetype_grid = 'dashed',
                       linewidth     = .5,
                       axis_ticks    = c("margin", "axis", "both", "off"),
-                      axis_lab_disp = c(.1, -.15)) {
+                      axis_lab_disp = c(.005, -.15)) {
   
   # Argument checking and formatting: ----
   
@@ -297,7 +297,7 @@ transform_grid <- function(basis,
                            linetype_grid = 'dashed',
                            linewidth = .5,
                            axis_ticks = c("margin", "axis", "both", "off"),
-                           axis_lab_disp = c(.1, -.15)) {
+                           axis_lab_disp = c(.005, -.15)) {
   
   # Argument checking and formatting: ----
   axis_ticks <- match.arg(axis_ticks)

### 2025-09-24 src/Graphical_example_paper.R

@@ -262,7 +262,7 @@ item_params_output <- item_params                          |>
 
 plot_oblique <- transform_grid(
   transform_matrix_inv_transp,
-  x_limits = c(-2,    3),
+  x_limits = c(-1.7,    3.3),
   y_limits = c(-1.99, 2.81),
   break_step = 1,
   linetype_grid = "17",
@@ -292,7 +292,7 @@ plot_orth <- transform_grid(
   linetype_grid = "17",
   linewidth = LINE_WIDTH,
   axis_ticks = "axis",
-  axis_lab_disp = c(.15, -.15)
+  axis_lab_disp = c(-.03, -.15)
 ) +
   geom_segment(
     mapping   = aes(
@@ -306,7 +306,3 @@ plot_orth <- transform_grid(
     linewidth = VECTOR_WIDTH
   ) +
   scale_color_manual(values = PALETTE)
-
-# TODO:
-# - Change style of labels on axes
-# - Check correct behavior in slideshows

### 2025-10-01 output/Response_to_action_editor_reviewers-2.Rmd
Add (preliminary) response to editor's comment 2
Add TODO to editor's comment 3
Add pending task to editor's comment 3
Complete & clarify Reviewer 4 responses

--- 2025-09-12 output/Response_to_action_editor_reviewers-2.Rmd
@@
 
+Please see our responses to Thissen's suggestions below.
+<!-- # TODO: Review upon completion. -->
+
 ###### 3) I would like to see additional comments in their closing paragraph on how testing practitioners should use or what insights can be obtained by reviewing the different representations of the latent ability space and the item space.
@@
 
+<!-- # TODO: Discuss complex items, and the proper representation in the test space. -->
+<!-- # TODO: Discuss "latent space" transformation is an artifact, not something to "explore". -->
+
 # Responses to Reviewer 4:
+
+###### It is always best, many would say required, to label the axes.
+The lack of axis labels in the graphics of Figure 1 make the rest of these comments difficult to type, so for convenience I imagine the axes are labeled conventionally x and y in the left panel, and x and y' in the right panel, indicating that y' is a rotated representation of y.
+(In the ms., it would be best to tie the labeling notation to the algebra used.)
+
+We had overlooked labelling the axes and have done so in this new version.
+We thank the reviewer once again for pointing this out.
 
@@
 
+Our understanding of Reckase's original work is the same as the reviewer's:
+That the orthogonality (actually, orthonormality) assumption was implicit at the time of @reckase_difficulty_1985.
+Because of that, an orthonormal (not just orthogonal) basis was a sound assumption.
+However, as Multidimensional IRT theory became mainstream, some applied researchers started applying these and Reckase and McKinley's -@reckase_discriminating_1991 formulae without noticing they were applying them to a more general case with non-orthonormal dimensions (this is part of another work-in-progress manuscript).
+
 ###### Then one could make another version of the left panel with contours for the rho=0.5 population distribution.
 Comparison of that with the previously-described version would show that the arrows representing discrimination do not have the same numerical meaning with respect to the variance (standard deviation) of the population distribution *in the direction of the arrows*, which is part of the point of all of this.
+
+We have taken into account this suggestion and added the contours plots accordingly.
+<!-- # TODO: Explain the new panel labelling -->
 
@@
 Different population distributions on orthogonal axes are the way MIRT folks have thought about factor correlation; factor analysts have thought about 'rotation' and oblique axes (the left panel of Figure 1); it's two sides of the same coin.
+
+This is indeed the message we want to convey with our manuscript, and are thankful to the reviewer for the suggestion.
+<!-- # TODO: Clarify in paper and complete response. -->
 
@@
 
+We understand this graphic may be hard to interpret, but our interpretation is that the grid lines are properly placed.
+The rationale is as follows:
+Each point in the test space is intended to represent a coordinate vector, which is the displacement from the origin given by a linear combination of the basis vectors, with the components of the coordinate vector being the coefficients of those basis vectors (please refer to the expression $`r DISCR_COORDS_ALT_EQ`$ in p.
+8 of the previous version of our manuscript, or p.
+<!-- # TODO: Complete --> in the new submission).
+For example, the tip of the vector representing item 1 is at coordinate (+2, +2).
+This means the point is displaced twice the length of the first vector basis along its direction (i.e., two units along the horizontal axis) and then twice the length of the first vector basis along its direction (i.e., another two units parallel to the direction of the oblique axis).
+These displacements, parallel to the direction of the axes, are what the grid lines intend to represent.
+This representation is somewhat equivalent to a _factor pattern_ (i.e., description in linear combination terms, see @harman_modern_1976, p.
+267) representation in the factor model, while the one proposed by the reviewer would be paralleled by the _factor structure_ representation (i.e., description in terms of correlations with the latent dimensions).
+Please see @harman_modern_1976 (p. 265 and p. 267, respectively) for the two representations.
+This can be more easily understood by comparing figures 4.2 (p. 61, a representation equivalent to the one we propose) and 12.8 (p. 266) in @harman_modern_1976.
+
 ###### That said, one could also add elliptical contours representing the population distribution to the left panel of Figure 1.

### 2025-10-01 notebooks/interp_example.qmd
Try `interp::interp()` example

---
title: "Untitled"
format: html
editor: visual
---

```{r}
library(interp)

data(franke)
## x-y irregular grid points:
oldseed <- set.seed(42)
ni <- 64
xi <- runif(ni,0,1)
yi <- runif(ni,0,1)
xyi <- cbind(xi,yi)
## linear interpolation
fi <- franke.fn(xi,yi,1)
IL <- interp(xi,yi,fi,nx=80,ny=80,method="linear")
## prepare breaks and colors that match for image and contour:
breaks <- pretty(seq(min(IL$z,na.rm=TRUE),max(IL$z,na.rm=TRUE),length=11))
db <- breaks[2]-breaks[1]
nb <- length(breaks)
breaks <- c(breaks[1]-db,breaks,breaks[nb]+db)
colors <- terrain.colors(length(breaks)-1)
image(IL,breaks=breaks,col=colors,main="Franke function 1",
      sub=paste("linear interpolation, ", ni,"points"))
contour(IL,add=TRUE,levels=breaks)
points(xi,yi)
## spline interpolation
fi <- franke.fn(xi,yi,1)
IS <- interp(xi,yi,fi,method="akima",
             kernel="gaussian",solver="QR")
## prepare breaks and colors that match for image and contour:
breaks <- pretty(seq(min(IS$z,na.rm=TRUE),max(IS$z,na.rm=TRUE),length=11))
db <- breaks[2]-breaks[1]
nb <- length(breaks)
breaks <- c(breaks[1]-db,breaks,breaks[nb]+db)
colors <- terrain.colors(length(breaks)-1)
image(IS,breaks=breaks,col=colors,main="Franke function 1",
      sub=paste("spline interpolation, ", ni,"points"))
contour(IS,add=TRUE,levels=breaks)
        points(xi,yi)
```

### 2025-10-01 notebooks/interp_example.qmd
Try "deldir" triangulation method

@@
 
-```{r}
+```{r setup}
+#| include: false
 library(interp)
 
-data(franke)
-## x-y irregular grid points:
-oldseed <- set.seed(42)
-ni <- 64
-xi <- runif(ni,0,1)
-yi <- runif(ni,0,1)
-xyi <- cbind(xi,yi)
-## linear interpolation
-fi <- franke.fn(xi,yi,1)
-IL <- interp(xi,yi,fi,nx=80,ny=80,method="linear")
-## prepare breaks and colors that match for image and contour:
-breaks <- pretty(seq(min(IL$z,na.rm=TRUE),max(IL$z,na.rm=TRUE),length=11))
-db <- breaks[2]-breaks[1]
-nb <- length(breaks)
-breaks <- c(breaks[1]-db,breaks,breaks[nb]+db)
-colors <- terrain.colors(length(breaks)-1)
-image(IL,breaks=breaks,col=colors,main="Franke function 1",
-      sub=paste("linear interpolation, ", ni,"points"))
-contour(IL,add=TRUE,levels=breaks)
-points(xi,yi)
-## spline interpolation
-fi <- franke.fn(xi,yi,1)
-IS <- interp(xi,yi,fi,method="akima",
-             kernel="gaussian",solver="QR")
-## prepare breaks and colors that match for image and contour:
-breaks <- pretty(seq(min(IS$z,na.rm=TRUE),max(IS$z,na.rm=TRUE),length=11))
-db <- breaks[2]-breaks[1]
-nb <- length(breaks)
-breaks <- c(breaks[1]-db,breaks,breaks[nb]+db)
-colors <- terrain.colors(length(breaks)-1)
-image(IS,breaks=breaks,col=colors,main="Franke function 1",
-      sub=paste("spline interpolation, ", ni,"points"))
-contour(IS,add=TRUE,levels=breaks)
-        points(xi,yi)
+ROOT_DIR <- here::here()
+knitr::opts_knit$set(root.dir = ROOT_DIR)
+
 ```
+
+```{r sources}
+#| include: false
+source("src/Graphical_example_paper.R", encoding = "UTF-8")
+```
+
+```{r}
+mvn_densities <- mvn_densities |>
+  bind_cols(latent_space_grid_rot |> rename_with(~paste0(., '_rot')))
+
+interp(
+  mvn_densities$trait_1_rot,
+  mvn_densities$trait_2_rot,
+  mvn_densities$corr_norm,
+  deltri = "deldir"
+)
+```

### 2025-10-01 notebooks/interp_example.qmd
Revert attempt to try "deldir" triangulation
This method causes the R environment to crash
Plot interpolated oblique contour plots
Computed by rotating the correlated distribution, and then interpolating.

@@
 library(interp)
+library(magrittr)
 
@@
 
-interp(
+mvn_dens_obl_interp <- interp(
   mvn_densities$trait_1_rot,
   mvn_densities$trait_2_rot,
-  mvn_densities$corr_norm,
-  deltri = "deldir"
+  mvn_densities$corr_norm
 )
+
+mvn_dens_obl <- mvn_dens_obl_interp %$%
+  expand_grid(trait_1 = x, trait_2 = y) |>
+  add_column(dens = mvn_dens_obl_interp$z |> as.vector())
+
+contour_obl <- mvn_dens_obl |> geom_contour(
+  mapping     = aes(trait_1, trait_2, z = dens),
+  color        = CONTOUR_COLOR,
+  alpha        = .3,
+  linewidth    = VECTOR_WIDTH,
+  breaks       = CONTOUR_BREAKS
+)
+
+grid_oblique + contour_obl
 ```
+
+These contour plots do not seem to be correct.
+Do they actually represent a standard normal distribution with the intended covariance structure, in the test space?

### 2025-10-02 output/Main_text.Rmd
Add contours to item plots
Reorder also the plots in the order in which they are used.
Output plot with contour plots in manuscript

--- 2025-09-12 output/Main_text.Rmd
@@
 
+```{r density-contours}
+```
+
 ```{r compose-oblique-plot}
@@
 #| fig.topcaption: true
-#| fig.height:     3.1
+#| fig.height:     3
 #| fig.width:      6.65
-plot_orth + plot_oblique
+#| warning:        false
+plot_orth_uncorr_contours + plot_oblique
 

### 2025-10-02 src/Graphical_example_paper.R

@@ -314,7 +314,7 @@ contour_obl <- mvn_densities |> geom_contour(
 # Orthogonal grid:
 grid_orth <- transform_grid(
   diag(2),
-  x_limits = c(-2.8,  2.7),
+  x_limits = c(-3.2,  3.2),
   y_limits = c(-2.65, 2.65),
   break_step = 1,
   linetype_grid = "17",

### 2025-10-02 output/Main_text.Rmd
Rename "oblique" plot to "test space" plot

--- 2025-09-12 output/Main_text.Rmd
@@
 
-```{r compose-oblique-plot}
+```{r density-contours}
 ```
@@
 ```{r compose-orthogonal-plot}
+```
+
+```{r compose-test-space-oblique-plot}
 ```
@@
 #| fig.topcaption: true
-#| fig.height:     3.1
+#| fig.height:     3
 #| fig.width:      6.65
-plot_orth + plot_oblique
+#| warning:        false
+plot_orth_uncorr_contours + plot_oblique_ts
 

### 2025-10-02 src/Graphical_example_paper.R

@@ -340,12 +340,12 @@ items_geom_orth <- geom_segment(
 plot_orth_uncorr_contours <- grid_orth + contour_orth + items_geom_orth
 plot_orth_corr_contours   <- grid_orth + contour_corr + items_geom_orth
 
-## ----compose-oblique-plot----
+## ----compose-test-space-oblique-plot----
 
 # Oblique grid:
-grid_oblique <- transform_grid(
+grid_ts <- transform_grid(
   transform_matrix_inv_transp,
-  x_limits = c(-2.3,  3.2),
+  x_limits = c(-3.2,  3.2),
   y_limits = c(-2.65, 2.65),
   break_step = 1,
   linetype_grid = "17",
@@ -367,4 +367,4 @@ items_geom_oblique <- geom_segment(
   linewidth = VECTOR_WIDTH
 )
 
-plot_oblique <- grid_oblique + contour_obl + items_geom_oblique
+plot_oblique_ts <- grid_ts + contour_obl + items_geom_oblique

### 2025-10-02 output/Main_text.Rmd
Add latent space and "agnostic" plot

@@
 
+```{r compose-test-space-orthogonal-plot}
+```
+
+```{r compose-latent-space-plot}
+```
+
 <br>
@@
 #| fig.topcaption: true
-#| fig.height:     3
+#| fig.height:     5
 #| fig.width:      6.65
 #| warning:        false
-plot_orth_uncorr_contours + plot_oblique_ts
-
-ggsave(file.path(DOC_DIR, "Figure_1.svg"))
+(plot_orth_uncorr_contours + plot_orth_corr) /
+  (plot_oblique_ts + plot_oblique_ls)
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"), width = 6.65, height = 5)
 ```

### 2025-10-02 src/Graphical_example_paper.R

@@ -137,7 +137,7 @@ items_oblique_params <- items_M2PL |> compute_mirt_params(
 items_oblique_coords <- items_oblique_params |> compute_mirt_coords(
   D, MDISC, starts_with(COSINE_DIRTYPE),
   transform       = transform_matrix_inv_transp,
-  original_coords = FALSE
+  original_coords = TRUE
 )
 items_oblique        <- full_join(
   items_oblique_params, items_oblique_coords,
@@ -152,6 +152,17 @@ item_params <- full_join(
 )
 
 item_params <- full_join(items_M2PL, item_params, by = ITEM_COLKEY)
+
+# Compute items in latent space (needs to be computed manually, because the
+#   covariance matrix is computed from the transform matrix in
+#   `compute_mirt_coords()` and the inverse matrix would be used instead):
+items_ls <- items_oblique_coords |>
+  pivot_longer(origin_1:end_2) |>
+  separate(name, into = c('end', 'coord'), sep = "_", remove = FALSE) |>
+  group_by(item, end) |>
+  mutate(value = transform_matrix %*% value |> drop()) |>
+  ungroup() |>
+  pivot_wider(id_cols = item:end_transf_2, values_from = value)
 
 
 ## ----compose-example-items-table----
@@ -358,8 +369,7 @@ grid_ts <- transform_grid(
   linetype_grid = "17",
   linewidth = LINE_WIDTH,
   axis_ticks = "axis"
-) +
-  scale_color_manual(values = PALETTE, guide = NULL)
+)
 
 # Oblique items:
 items_geom_oblique <- geom_segment(
@@ -377,5 +387,42 @@ items_geom_oblique <- geom_segment(
 plot_oblique_ts <- grid_ts +
   contour_obl +
   items_geom_oblique +
+  colorscale_legend
+
+## ----compose-test-space-orthogonal-plot----
+
+plot_orth_corr <- grid_orth +
+  contour_corr +
+  items_geom_orth +
   colorscale_no_legend
 
+## ----compose-latent-space-plot----
+
+# Oblique grid:
+grid_ls <- transform_grid(
+  transform_matrix,
+  x_limits = c(-3.2,  3.2),
+  y_limits = c(-2.65, 2.65),
+  break_step = 1,
+  linetype_grid = "17",
+  linewidth = LINE_WIDTH,
+  axis_ticks = "axis"
+)
+
+# Oblique items:
[…]

### 2025-10-03 output/Main_text.Rmd
Fine tune figure 1 output
Set axis lines and limits properly.

@@
 #| fig.topcaption: true
-#| fig.height:     5
+#| fig.height:     5.2
 #| fig.width:      6.65
@@
 (plot_orth_uncorr_contours + plot_orth_corr) /
-  (plot_oblique_ts + plot_oblique_ls)
-
-ggsave(file.path(DOC_DIR, "Figure_1.svg"), width = 6.65, height = 5)
+  (plot_oblique_ts + plot_oblique_ls) +
+  plot_layout(guides = "collect")
+## TODO: Review and reformat axis labels
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"), width = 6.65, height = 5.2)
 ```

### 2025-10-03 R/Output.R

@@ -181,7 +181,7 @@ plot_grid <- function(grid,
                       linetype_grid = 'dashed',
                       linewidth     = .5,
                       axis_ticks    = c("margin", "axis", "both", "off"),
-                      axis_lab_disp = c(.005, -.15)) {
+                      axis_lab_disp = c(.005, -.2)) {
   
   # Argument checking and formatting: ----
   
@@ -297,7 +297,7 @@ transform_grid <- function(basis,
                            linetype_grid = 'dashed',
                            linewidth = .5,
                            axis_ticks = c("margin", "axis", "both", "off"),
-                           axis_lab_disp = c(.005, -.15)) {
+                           axis_lab_disp = c(.005, -.2)) {
   
   # Argument checking and formatting: ----
   axis_ticks <- match.arg(axis_ticks)

### 2025-10-03 src/Graphical_example_paper.R

@@ -42,6 +42,9 @@ LINE_WIDTH   <- .1
 VECTOR_WIDTH <- .5
 PALETTE      <- c("darkred", "darkgoldenrod3", "green3", "cyan3", "blue3")
 
+X_RANGE <- c(-2.97, 3.23)
+Y_RANGE <- c(-2.59, 2.61)
+
 # Latent space grid for density contours:
 PROB_AXIS_2D   <- seq(-3, 3, by = 0.05) # Axis for the density contour plots
 CONTOUR_BREAKS <- c(.1, .5, .9)         # Breaks for the density contour plots
@@ -331,13 +334,13 @@ contour_obl <- mvn_densities |> geom_contour(
 # Orthogonal grid:
 grid_orth <- transform_grid(
   diag(2),
-  x_limits = c(-3.2,  3.2),
-  y_limits = c(-2.65, 2.65),
+  x_limits = X_RANGE,
+  y_limits = Y_RANGE,
   break_step = 1,
   linetype_grid = "17",
   linewidth = LINE_WIDTH,
   axis_ticks = "axis",
-  axis_lab_disp = c(-.03, -.15)
+  axis_lab_disp = c(-.03, -.2)
 )
 
 # Orthogonal items:
@@ -363,8 +366,8 @@ plot_orth_uncorr_contours <- grid_orth +
 # Oblique grid:
 grid_ts <- transform_grid(
   transform_matrix_inv_transp,
-  x_limits = c(-3.2,  3.2),
-  y_limits = c(-2.65, 2.65),
+  x_limits = X_RANGE,
+  y_limits = Y_RANGE,
   break_step = 1,
   linetype_grid = "17",
   linewidth = LINE_WIDTH,
@@ -401,12 +404,13 @@ plot_orth_corr <- grid_orth +
 # Oblique grid:
 grid_ls <- transform_grid(
   transform_matrix,
-  x_limits = c(-3.2,  3.2),
-  y_limits = c(-2.65, 2.65),
+  x_limits = X_RANGE,
+  y_limits = Y_RANGE,
   break_step = 1,
   linetype_grid = "17",
   linewidth = LINE_WIDTH,
-  axis_ticks = "axis"
+  axis_ticks = "axis",
+  axis_lab_disp = c(-.03, -.15)
 )
 
 # Oblique items:

### 2025-10-06 output/Main_text.Rmd
Use rectangular coordinates for coordinate plot
Instead of transforming the items to the latent space grid, represent their coordinates on a rectangular grid (i.e. the canonical basis)

--- 2025-10-02 output/Main_text.Rmd
@@
 
+```{r compose-test-space-orthogonal-plot}
+```
+
+```{r compose-rectangular-cov-based-plot}
+```
+
 <br>
@@
 #| fig.topcaption: true
-#| fig.height:     3
+#| fig.height:     4.7
 #| fig.width:      6.65
 #| warning:        false
-plot_orth_uncorr_contours + plot_oblique_ts
-
-ggsave(file.path(DOC_DIR, "Figure_1.svg"))
+(plot_orth_uncorr_contours + plot_orth_corr) /
+  (plot_oblique_ts + plot_cov_based_orth) +
+  plot_layout(guides = "collect")
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"), width = 6.65, height = 4.7)
 ```

### 2025-10-06 src/Graphical_example_paper.R

@@ -43,7 +43,7 @@ VECTOR_WIDTH <- .5
 PALETTE      <- c("darkred", "darkgoldenrod3", "green3", "cyan3", "blue3")
 
 X_RANGE <- c(-2.97, 3.23)
-Y_RANGE <- c(-2.59, 2.61)
+Y_RANGE <- c(-2.35,  2.2)
 
 # Latent space grid for density contours:
 PROB_AXIS_2D   <- seq(-3, 3, by = 0.05) # Axis for the density contour plots
@@ -155,17 +155,6 @@ item_params <- full_join(
 )
 
 item_params <- full_join(items_M2PL, item_params, by = ITEM_COLKEY)
-
-# Compute items in latent space (needs to be computed manually, because the
-#   covariance matrix is computed from the transform matrix in
-#   `compute_mirt_coords()` and the inverse matrix would be used instead):
-items_ls <- items_oblique_coords |>
-  pivot_longer(origin_1:end_2) |>
-  separate(name, into = c('end', 'coord'), sep = "_", remove = FALSE) |>
-  group_by(item, end) |>
-  mutate(value = transform_matrix %*% value |> drop()) |>
-  ungroup() |>
-  pivot_wider(id_cols = item:end_transf_2, values_from = value)
 
 
 ## ----compose-example-items-table----
@@ -390,43 +379,33 @@ items_geom_oblique <- geom_segment(
 plot_oblique_ts <- grid_ts +
   contour_obl +
   items_geom_oblique +
-  colorscale_legend
+  colorscale +
+  annotate("text", -2.5, 2, label = "B", family = GRAPH_FONT, size = 6)
 
 ## ----compose-test-space-orthogonal-plot----
 
 plot_orth_corr <- grid_orth +
   contour_corr +
   items_geom_orth +
-  colorscale_no_legend
+  colorscale +
+  annotate("text", -2.5, 2, label = "C", family = GRAPH_FONT, size = 6)
 
-## ----compose-latent-space-plot----
+## ----compose-rectangular-cov-based-plot----
 
-# Oblique grid:
-grid_ls <- transform_grid(
-  transform_matrix,
-  x_limits = X_RANGE,
-  y_limits = Y_RANGE,
-  break_step = 1,
-  linetype_grid = "17",
-  linewidth = LINE_WIDTH,
-  axis_ticks = "axis",
-  axis_lab_disp = c(-.03, -.15)
-)
-
-# Oblique items:
-items_geom_ls_obl <- geom_segment(
[…]

### 2025-10-06 output/Main_text.Rmd
Update figure 1 referencing
Add panel labels, update figure title, and add footnote.

@@
 #| cache:          false
-#| fig.cap:   Item Plots in an Orthogonal (Left) and Oblique (Right) Test Space
+#| fig.cap:        |
+#|   Item vector plots with uncorrelated (a) and correlated latent dimensions.
+#|   The latter coordinates are computed with
+#|   (b) the agnostic multidimensional parameters, or
+#|   with the covariance/correlation-based ones, which are then plot either
+#|   (c) in an oblique basis, appropriate for the covariance structure, or
+#|   (d) in the canonical basis.
 #| fig.topcaption: true
@@
 ```
+
+```{r format-contour-plot-props}
+contour_breaks_out <- CONTOUR_BREAKS |>
+  percent() |>
+  glue_collapse(sep = ', ', last = ', and ')
+```
+
+::: {custom-style="FigureNote"}
+*Note.* The items are represented along with
+a standard bivariate normal distribution, with null correlation (a),
+or with correlation $`r CORR_OBL_OUT`$ (b-d).
+The contour plots represent the `r contour_breaks_out` of the maximum density,
+from outer to inner, respectively.
+:::
 

### 2025-10-06 src/Graphical_example_paper.R

@@ -45,6 +45,10 @@ PALETTE      <- c("darkred", "darkgoldenrod3", "green3", "cyan3", "blue3")
 X_RANGE <- c(-2.97, 3.23)
 Y_RANGE <- c(-2.35,  2.2)
 
+# Plot panel label position:
+PANEL_LABEL_X <- -2.5
+PANEL_LABEL_Y <-  2
+
 # Latent space grid for density contours:
 PROB_AXIS_2D   <- seq(-3, 3, by = 0.05) # Axis for the density contour plots
 CONTOUR_BREAKS <- c(.1, .5, .9)         # Breaks for the density contour plots
@@ -348,7 +352,15 @@ plot_orth_uncorr_contours <- grid_orth +
   contour_orth +
   items_geom_orth +
   colorscale +
-  annotate("text", -2.5, 2, label = "A", family = GRAPH_FONT, size = 6)
+  annotate(
+    "text",
+    x        = PANEL_LABEL_X,
+    y        = PANEL_LABEL_Y,
+    label    = "(a)",
+    family   = GRAPH_FONT,
+    size     = 6,
+    fontface = "bold"
+  )
 
 ## ----compose-test-space-oblique-plot----
 
@@ -380,7 +392,15 @@ plot_oblique_ts <- grid_ts +
   contour_obl +
   items_geom_oblique +
   colorscale +
-  annotate("text", -2.5, 2, label = "B", family = GRAPH_FONT, size = 6)
+  annotate(
+    "text",
+    x        = PANEL_LABEL_X,
+    y        = PANEL_LABEL_Y,
+    label    = "(b)",
+    family   = GRAPH_FONT,
+    size     = 6,
+    fontface = "bold"
+  )
 
 ## ----compose-test-space-orthogonal-plot----
 
@@ -388,7 +408,15 @@ plot_orth_corr <- grid_orth +
   contour_corr +
   items_geom_orth +
   colorscale +
-  annotate("text", -2.5, 2, label = "C", family = GRAPH_FONT, size = 6)
+  annotate(
+    "text",
+    x        = PANEL_LABEL_X,
+    y        = PANEL_LABEL_Y,
+    label    = "(c)",
+    family   = GRAPH_FONT,
+    size     = 6,
+    fontface = "bold"
+  )
 
 ## ----compose-rectangular-cov-based-plot----
 
@@ -409,4 +437,12 @@ plot_cov_based_orth <- grid_orth +
   contour_corr +
[…]

### 2025-10-06 src/Empirical_example.R
Match table footnote style to figure footnote's

@@ -272,7 +272,7 @@ item_params_output <- item_params                          |>
   )                                                        |>
   style(
     pr_p = fp_par(
-      line_spacing   = 1.25,
+      line_spacing   = 2,
       padding.top    = 2,
       padding.bottom = 0
     ),

### 2025-10-06 src/Graphical_example_paper.R

@@ -276,7 +276,7 @@ item_params_output <- item_params                          |>
   )                                                        |>
   style(
     pr_p = fp_par(
-      line_spacing   = 1.25,
+      line_spacing   = 2,
       padding.bottom = 0
     ),
     part = "footer",

### 2025-10-06 output/Main_text.Rmd
Match table footnote style to figure footnote's
Define latent-space test transformation

@@
 mapping the $`r MODEL_ACRONYM`$ parameters interchangeably in the two spaces,
-by projecting one space onto the other.
+by projecting one space onto the other;
+that is, we can set
+
+\begin{equation}
+  `r DISCR_COORDS_TRANSF_EQ`                  (\#eq:trait-coords-orth-spaces-eq)
+\end{equation}
+
+for any point, $`r TRAIT_VECTOR`$ or $`r DISCR_VECTOR`$.
 This allows representing the coordinates of the discrimination vectors
 in the latent space and, conversely,
-the coordinates of the latent trait vectors in the test space.
+the coordinates of the latent trait vectors in the test space:
+By \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
+and  \@ref(eq:trait-coords-orth-spaces-eq)
+any coordinate vector $`r DISCR_COORDS`$ in $`r TEST_SPACE_BASIS`$
+can be expressed in
+$`r LS_BASIS`$ as $`r DISCR_COORDS_TS_EQ`$, and vice versa.
 As we can see, the invariance of the model holds,
-no matter in which space we represent the coordinates.
+regardless of the space in which we represent the coordinates.
+
 Second, and most importantly,
@@
 or with correlation $`r CORR_OBL_OUT`$ (b-d).
-The contour plots represent the `r contour_breaks_out` of the maximum density,
-from outer to inner, respectively.
+The contour plots represent (from outer to inner)
+the `r contour_breaks_out` of the maximum density.
 :::

### 2025-10-06 R/Formulae.R

@@ -838,6 +838,18 @@ DISCR_STD_COORDS_MOD_EQ <- latex_eq(
 
 ### Relationship between test space and latent space: ----
 
+# Orthonormalized coordinates equivalence:
+DISCR_COORDS_TRANSF_LS   <- latex_raised_to(DISCR_VECTOR, exp = LS_ORTH_BASIS)
+DISCR_COORDS_TRANSF_EQ   <- latex_eq(DISCR_STD_COORDS, DISCR_COORDS_TRANSF_LS)
+DISCR_COORDS_LS          <- latex(DISCR_VECTOR, LS_BASIS)
+DISCR_COORDS_TRANSF_DEF  <- latex(TRANSFORM_MATRIX_INV, DISCR_STD_COORDS_DEF)
+DISCR_COORDS_GRAM_TR_DEF <- latex(INNER_PROD_MATRIX_INV, DISCR_COORDS)
+DISCR_COORDS_TS_EQ       <- latex_eq(
+  DISCR_COORDS_TRANSF_LS,
+  DISCR_COORDS_TRANSF_DEF,
+  DISCR_COORDS_GRAM_TR_DEF
+)
+
 # Item direction cosines (in test space):
 ANGLE_TS                          <- latex("\\alpha")
 ANGLE_VEC_TS                      <- latex_bf(ANGLE_TS)

### 2025-10-07 output/Main_text.Rmd
State covariance version as result of whitening

@@
 Therefore, an orthonormal basis should represent independent,
-standard coordinates;
-formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
-$`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$.
-Applying Equation \@ref(eq:change-basis), $`r COV_MATRIX_TRANSF_EQ`$, and thus,
+standard coordinates.
+Formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
+$`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$,
+which implies that Equation \@ref(eq:change-basis) actually represents
+a *whitening transformation* [@fukunaga_introduction_2013].
+Given this, $`r COV_MATRIX_TRANSF_EQ`$, and thus
 

### 2025-10-07 output/Main_text.Rmd
Use latent trait for space transform formulation
It is used later on for expressing the contour plot transformation

@@
 \begin{equation}
-  `r DISCR_COORDS_TRANSF_EQ`                  (\#eq:trait-coords-orth-spaces-eq)
+  `r TRAIT_COORDS_ORTH_EQ`                    (\#eq:trait-coords-orth-spaces-eq)
 \end{equation}
@@
 for any point, $`r TRAIT_VECTOR`$ or $`r DISCR_VECTOR`$.
-This allows representing the coordinates of the discrimination vectors
-in the latent space and, conversely,
-the coordinates of the latent trait vectors in the test space:
+This allows representing the coordinates of the latent space vectors
+in the test space and, conversely,
+the coordinates of the discrimination vectors in the latent space:
 By \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
 and  \@ref(eq:trait-coords-orth-spaces-eq)
-any coordinate vector $`r DISCR_COORDS`$ in $`r TEST_SPACE_BASIS`$
-can be expressed in
-$`r LS_BASIS`$ as $`r DISCR_COORDS_TS_EQ`$, and vice versa.
+any coordinate vector $`r TRAIT_COORDS`$ in $`r LS_BASIS`$
+can be expressed in $`r TEST_SPACE_BASIS`$ as
+
+\begin{equation}
+  `r TRAIT_COORDS_TS_EQ`,                   (\#eq:trait-coords-spaces-transform)
+\end{equation}
+
+and vice versa.
 As we can see, the invariance of the model holds,

### 2025-10-07 R/Formulae.R

@@ -839,15 +839,24 @@ DISCR_STD_COORDS_MOD_EQ <- latex_eq(
 ### Relationship between test space and latent space: ----
 
 # Orthonormalized coordinates equivalence:
-DISCR_COORDS_TRANSF_LS   <- latex_raised_to(DISCR_VECTOR, exp = LS_ORTH_BASIS)
-DISCR_COORDS_TRANSF_EQ   <- latex_eq(DISCR_STD_COORDS, DISCR_COORDS_TRANSF_LS)
-DISCR_COORDS_LS          <- latex(DISCR_VECTOR, LS_BASIS)
-DISCR_COORDS_TRANSF_DEF  <- latex(TRANSFORM_MATRIX_INV, DISCR_STD_COORDS_DEF)
-DISCR_COORDS_GRAM_TR_DEF <- latex(INNER_PROD_MATRIX_INV, DISCR_COORDS)
-DISCR_COORDS_TS_EQ       <- latex_eq(
-  DISCR_COORDS_TRANSF_LS,
-  DISCR_COORDS_TRANSF_DEF,
-  DISCR_COORDS_GRAM_TR_DEF
+TRAIT_COORDS_TRANSF_ORTH_TS <- latex_raised_to(
+  TRAIT_VECTOR,
+  exp = TEST_SPACE_STD_BASIS
+)
+TRAIT_COORDS_ORTH_EQ        <- latex_eq(
+  TRAIT_ORTH_COORDS,
+  TRAIT_COORDS_TRANSF_ORTH_TS
+)
+TRAIT_COORDS_TRANSF_TS      <- latex_raised_to(
+  TRAIT_VECTOR,
+  exp = TEST_SPACE_BASIS
+)
+TRAIT_COORDS_TRANSF_DEF     <- latex(TRANSFORM_MATRIX_TRANSP, TRAIT_TRANSFORM)
+TRAIT_COORDS_GRAM_TR_DEF    <- latex(INNER_PROD_MATRIX, TRAIT_COORDS)
+TRAIT_COORDS_TS_EQ          <- latex_eq(
+  TRAIT_COORDS_TRANSF_TS,
+  TRAIT_COORDS_TRANSF_DEF,
+  TRAIT_COORDS_GRAM_TR_DEF
 )
 
 # Item direction cosines (in test space):

### 2025-10-07 output/Main_text.Rmd
Explain new figure panels

@@
 of our general, possibly non-rectangular coordinates.
-This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
-by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
-the equivalent, orthonormalized coordinates.
+Coordinates in a rectangular Cartesian system are commonly assumed
+to be simply represented in the canonical basis.
+To represent an item in the test space properly, however,
+we must use its basis $`r TEST_SPACE_BASIS`$,
+which implies pre-multiplying its coordinates
+by $`r TRANSFORM_MATRIX_TRANSP_INV`$
+to find the equivalent coordinates in the canonical basis.
 Hence, we will need a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
@@
 ```{r compute-example-items}
+```
+
+```{r format-contour-plot-props}
+contour_breaks_out <- CONTOUR_BREAKS |>
+  percent() |>
+  glue_collapse(sep = ', ', last = ', and ')
 ```
@@
 a set of items in two different bidimensional spaces.
-In the first case, the two latent dimensions are independent,
-with $`r TRANSFORM_MATRIX_TRANSP_INV`$ being an identity matrix.
-In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
+Their parameters are shown in Table \@ref(tab:example-items-table-out),
+both in the original `r MODEL_ACRONYM` metric, and in the multidimensional
+metric in two different cases, with different latent covariance structures.
+In the first case (columns $`r CORR_ORTH_OUT`$),
+the two latent dimensions are independent;
+in the second one (columns $`r CORR_ORTH_OUT`$),
+the correlation between the dimensions is .5.
+Note that, although the `r MODEL_ACRONYM` parameters are the same in both cases,
+their (correlation-based) multidimensional parameters change when different
+correlations are used.
+
+Panel (a) of Figure \@ref(fig:item-plot-out) shows
+the item representation in the uncorrelated case,
+while the correlated case is shown in panels (b) to (d)
+under different conditions.
+In all the four panels, the contours of a bivariate standard normal density
+with the corresponding correlation are plot along with the items
+(from outer to inner, the contours are the `r contour_breaks_out` of
+the density at its mode).
+Note that the stretched ellipses in panel (c) are the result of transforming
+the distribution in the latent space to the test space,
+according to \@ref(eq:trait-coords-spaces-transform).
+In panel (b), the item coordinates are computed from
+the agnostic version of the parameters.
+These coordinates are the same ones as in the first case, but
+their relative position with respect to the latent trait distribution is not
+---This is the effect of disregarding the covariance structure.
+Panel (c) shows the items with their coordinates computed using
+the correlation-based version of the parameters.
 To keep the horizontal axis invariant,
-we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
+we have defined $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
 it only rotates the vertical axis [@harman_modern_1976]:
@@
 
-The sample items have the same `r MODEL_ACRONYM` parameters
-in both test spaces.
-However, their multidimensional parameters differ due to the correlation among
-the latent dimensions.
-The parameters in both spaces,
-along with their corresponding `r MODEL_ACRONYM` parameters,
-are shown in Table \@ref(tab:example-items-table-out).
+Comparing it to panel (b), we can see that the item coordinates are in general
+not the same as the ones computed with the agnostic version of the parameters;
+this effect is especially salient in items 2 and 3,
+which are close to the 90% contour in the agnostic version,
+but are almost along the 50% contour in the correlation-based one.
+Finally, panel (d) illustrates the effect of plotting the coordinates directly
+on the canonical basis; the relative positions and distances are correctly
+represented (which allow a more direct comparison with panel (b)),
+but the angles of the items with the axes are distorted:
+For example, item 1 forms 45-degree angles with both axes,
+instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).
 
@@
 
-Figure \@ref(fig:item-plot-out) illustrates the effect of
+A comparison of panels (a) and (c) allows understanding the effect of
 a positive correlation on the $`r MDISC_CORR_PARAM`$
@@
 
-```{r format-contour-plot-props}
-contour_breaks_out <- CONTOUR_BREAKS |>
-  percent() |>
-  glue_collapse(sep = ', ', last = ', and ')
-```
-
 ::: {custom-style="FigureNote"}
 *Note.* The items are represented along with
-a standard bivariate normal distribution, with null correlation (a),
+a bivariate standard normal distribution, with null correlation (a),
 or with correlation $`r CORR_OBL_OUT`$ (b-d).

### 2025-10-07 output/Main_text.Rmd
Explain axis labeling in Figure 1

@@
 a set of items in two different bidimensional spaces.
+(The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
+for simplicity, but note that they represent the coordinates of
+a latent trait point transformed to the test space,
+that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
 Their parameters are shown in Table \@ref(tab:example-items-table-out),
-both in the original `r MODEL_ACRONYM` metric, and in the multidimensional
-metric in two different cases, with different latent covariance structures.
+both in the original `r MODEL_ACRONYM` metric and
+in the multidimensional metric,
+in two different cases with different latent covariance structures.
 In the first case (columns $`r CORR_ORTH_OUT`$),
@@
 the correlation between the dimensions is .5.
-Note that, although the `r MODEL_ACRONYM` parameters are the same in both cases,
-their (correlation-based) multidimensional parameters change when different
+The same `r MODEL_ACRONYM` parameters are used in both cases,
+but their (correlation-based) multidimensional parameters change when different
 correlations are used.

### 2025-10-07 R/Formulae.R

@@ -1250,6 +1250,11 @@ TRANSF_MATRIX_INV_SQ_CORR_EQ <- latex_eq(INNER_PROD_MATRIX_INV_DEF, CORR_MATRIX)
 
 #### Graphical representation example: ----
 
+TRAIT_COMP_1  <- latex_sub(TRAIT_SYMBOL, 1)
+TRAIT_COMP_2  <- latex_sub(TRAIT_SYMBOL, 2)
+TRAIT_COORD_1 <- latex_raised_to(TRAIT_COMP_1, exp = TEST_SPACE_BASIS)
+TRAIT_COORD_2 <- latex_raised_to(TRAIT_COMP_2, exp = TEST_SPACE_BASIS)
+
 CORR <- latex("\\rho")
 
 TRANSFORM_MATRIX_EXAMPLE    <- latex(

### 2025-10-08 output/Title_page.Rmd
Correct acknowledgement statement

--- 2025-09-12 output/Title_page.Rmd
@@
 
-The authors would like to thank Patricia Recio-Saboya, Ph.D.
-(ORCID: [0000-0002-0530-4404](https://orcid.org/0000-0002-0530-4404))
-for her valuable guidance regarding multidimensional estimation
-in IRT software packages.
+The authors would like to thank:
+[Patricia Recio-Saboya, Ph.D.](https://orcid.org/0000-0002-0530-4404),
+for her guidance regarding multidimensional estimation
+in IRT software packages;
+[Jay Verkuilen, Ph.D.](https://orcid.org/0000-0003-3546-3411),
+for his comments and suggestions regarding covariance whitening;
+David Thissen and three additional anonymous reviewers for
+their valuable comments on early drafts of this manuscript.
 

### 2025-10-08 output/Main_text.Rmd
Explain computation of density contours in test space

@@
 the distribution in the latent space to the test space,
-according to \@ref(eq:trait-coords-spaces-transform).
+according to Equation \@ref(eq:trait-coords-spaces-transform).
+This implies that the covariance matrix of the distribution projected into
+the test space is transformed inversely as in the latent space, that is,
+$`r COV_MATRIX_TRANSF_TS_EQ`$.
+
 In panel (b), the item coordinates are computed from

### 2025-10-08 output/Response_to_action_editor_reviewers-2.Rmd
Update responses to editor

@@
 
-In p. 13 (line 2), dropping, "the squared direction cosines sum ~up~ to 1" did not sound well, so instead we reformulated the sentence as:
+In p. 13 (line 2), dropping "up", i.e., "the squared direction cosines sum ~up~ to 1", did not sound well to us, so instead we reformulated the sentence as:
 
@@
 
-Please see our responses to Thissen's suggestions below.
+Please see our detailed responses to Thissen's suggestions below.
+These changes have required a quite high effort to make sure they represent the algebra adequately.
+Also, we have deemed it necessary to rewrite some parts of the manuscript to improve the clarity of the plots.
 <!-- # TODO: Review upon completion. -->

### 2025-10-08 output/Response_to_action_editor_reviewers-2.Rmd
Explain changes to plots in responses to reviewer

@@
 
-We had overlooked labelling the axes and have done so in this new version.
-We thank the reviewer once again for pointing this out.
+We had overlooked labelling the axes and have done so in this new version;
+we thank the reviewer once again for pointing this out.
+$`r TRAIT_COMP_1`$/$`r TRAIT_COMP_2`$ have been used as labels for simplicity
+(both in the code for plotting them, and for the reader's understanding),
+but note that they represent the coordinates of
+a latent trait point transformed to the test space,
+that is, $`r TRAIT_COORD_1`$/$`r TRAIT_COORD_2`$.
+However, there is a direct correspondence between the algebraic components and
+the coordinates in either space, so this labeling should be clear enough,
+while relying only upon the algebraic interpretation of the model,
+thus avoiding the geometrical details,
+which are only necessary for deriving the results.
 
@@
 Because of that, an orthonormal (not just orthogonal) basis was a sound assumption.
-However, as Multidimensional IRT theory became mainstream, some applied researchers started applying these and Reckase and McKinley's -@reckase_discriminating_1991 formulae without noticing they were applying them to a more general case with non-orthonormal dimensions (this is part of another work-in-progress manuscript).
+However, as Multidimensional IRT theory became mainstream, some applied researchers started applying these and Reckase and McKinley's -@reckase_discriminating_1991 formulae without noticing they were applying them to a more general case with non-orthonormal dimensions, finding inconsistencies in their estimates (this topic is part of another work-in-progress manuscript).
+
+We have added the contour plots to the left panel of Figure 1 (panel (a) in the new version), as suggested.
 
@@
 We have taken into account this suggestion and added the contours plots accordingly.
-<!-- # TODO: Explain the new panel labelling -->
+This is represented as panel (b) in the new version.
+The text also explains the different interpretation of the coordinates relative to the distribution.
+<!-- # TODO: Explain changes in text -->
 
@@
 
-This is indeed the message we want to convey with our manuscript, and are thankful to the reviewer for the suggestion.
-<!-- # TODO: Clarify in paper and complete response. -->
+This is indeed the message we want to convey with our manuscript, and are thankful to the reviewer for the comments.
+As can be seen, we have made changes to the text in our manuscript to clarify this message.
+<!-- # TODO: Explain changes in text -->
 
@@
 
+As suggested, we have added these elliptical contours.
+The computation is also explained in the text; to clarify it, we have had to add some additional results in previous parts of the manuscript.
+The changes we have made in this new version are the following:
+
+<!-- # TODO: Explain changes in text -->
+
+Finally, we have added a fourth panel (panel (d)) that illustrates how the items would be plot if we omitted using the proper test space basis (and just using the canonical basis instead).
+Incidentally, this also serves the purpose of allowing an easier comparison of the coordinates computed from the two versions (the agnostic, and the correlation-based one) of the multidimensional parameters.
+We have briefly explained this, adding in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->:
+
+<!-- # TODO: Explain changes in text -->
+
 # References

### 2025-10-08 output/Main_text.Rmd
Correct coordinate equality in orthonormal bases
This equation was overwriting a previous one, by using the same symbol for the new object.

@@
 \begin{equation}
-  `r TRAIT_COORDS_ORTH_EQ`                    (\#eq:trait-coords-orth-spaces-eq)
+  `r TRAIT_COORDS_ORTH_SPACES_EQ`             (\#eq:trait-coords-orth-spaces-eq)
 \end{equation}
@@
 the coordinates of the discrimination vectors in the latent space:
-By \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
+By Equations \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
 and  \@ref(eq:trait-coords-orth-spaces-eq)

### 2025-10-08 R/Formulae.R

@@ -843,7 +843,7 @@ TRAIT_COORDS_TRANSF_ORTH_TS <- latex_raised_to(
   TRAIT_VECTOR,
   exp = TEST_SPACE_STD_BASIS
 )
-TRAIT_COORDS_ORTH_EQ        <- latex_eq(
+TRAIT_COORDS_ORTH_SPACES_EQ <- latex_eq(
   TRAIT_ORTH_COORDS,
   TRAIT_COORDS_TRANSF_ORTH_TS
 )

### 2025-10-08 output/Response_to_action_editor_reviewers-2.Rmd
Complete responses to Reviwer 4

@@
 
-Please see our detailed responses to Thissen's suggestions below.
-These changes have required a quite high effort to make sure they represent the algebra adequately.
-Also, we have deemed it necessary to rewrite some parts of the manuscript to improve the clarity of the plots.
-<!-- # TODO: Review upon completion. -->
+We have made the changes proposed, updated the figure caption accordingly, and added an appropriate footer.
+These changes have required a rather high effort to make sure they represent the algebra adequately, as well as some edits to improve the clarity of the plots and justify the computations.
+Please see our detailed responses to Thissen's comments below.
 
@@
 which are only necessary for deriving the results.
+This choice of labeling is explained in the text in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->:
+
+> (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
+> for simplicity, but note that they represent the coordinates of
+> a latent trait point transformed to the test space,
+> that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
 
@@
 
-We have taken into account this suggestion and added the contours plots accordingly.
-This is represented as panel (b) in the new version.
-The text also explains the different interpretation of the coordinates relative to the distribution.
-<!-- # TODO: Explain changes in text -->
-
 ###### Two graphics with orthogonal axes but different population distributions are the way orthogonal vs.
@@
 This is indeed the message we want to convey with our manuscript, and are thankful to the reviewer for the comments.
-As can be seen, we have made changes to the text in our manuscript to clarify this message.
-<!-- # TODO: Explain changes in text -->
+We have taken into account the suggestion and added the contours plots accordingly.
+This is represented as panel (b) in the new version.
+The text also explains the different interpretation of the coordinates relative to the distribution now.
+Please see the new paragraphs added in third and fourth place in section "Graphical Representation Example" (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
+
+> In panel (b), the item coordinates are computed from
+> the agnostic version of the parameters.
+> These coordinates are the same ones as in the first case, but
+> their relative position with respect to the latent trait distribution is not
+> ---This is the effect of disregarding the covariance structure.
+> Panel (c) shows the items with their coordinates computed using
+> the correlation-based version of the parameters. [...]
+>
+> Comparing it to panel (b), we can see that the item coordinates are in general
+> not the same as the ones computed with the agnostic version of the parameters;
+> this effect is especially salient in items 2 and 3,
+> which are close to the 90% contour in the agnostic version,
+> but are almost along the 50% contour in the correlation-based one. [...]
 
@@
 
-<!-- # TODO: Explain changes in text -->
+- The second paragraph of section "Relationship Between Test Space and Latent Space" explains how coordinates can be converted from the latent trait to the test space, and vice versa.
+For explaining this, we have needed to add and reference two new equations.
+This paragraph (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads as follows:
+
+> [...]
+> First, the transformation to the orthonormal bases of the two spaces allows
+> mapping the $`r MODEL_ACRONYM`$ parameters interchangeably in the two spaces,
+> by projecting one space onto the other;
+> that is, we can set
+> 
+> \begin{equation}
+>   `r TRAIT_COORDS_ORTH_SPACES_EQ` \t (23)
+> \end{equation}
+> 
+> 
+> for any point, $`r TRAIT_VECTOR`$ or $`r DISCR_VECTOR`$.
+> This allows representing the coordinates of the latent space vectors
+> in the test space and, conversely,
+> the coordinates of the discrimination vectors in the latent space:
+> By Equations (2), (5), and (23)
+> any coordinate vector $`r TRAIT_COORDS`$ in $`r LS_BASIS`$
+> can be expressed in $`r TEST_SPACE_BASIS`$ as
+> 
+> \begin{equation}
+>   `r TRAIT_COORDS_TS_EQ`, \t (24)
+> \end{equation}
+>   
+> and vice versa. [...]
+
+- The second paragraph in section "Graphical Representation" now clarifies better that the representation in the test space must use its basis instead of the canonical basis, to make the plot be a faithful representation of the actual directions of the items with respect to the latent trait dimensions.
+This paragraph now reads (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
+
+> [...] Coordinates in a rectangular Cartesian system are commonly assumed
+> to be simply represented in the canonical basis.
+> To represent an item in the test space properly, however,
+> we must use its basis $`r TEST_SPACE_BASIS`$,
+> which implies pre-multiplying its coordinates
+> by $`r TRANSFORM_MATRIX_TRANSP_INV`$
+> to find the equivalent coordinates in the canonical basis. [...]
+
+- The section "Graphical Representation Example" has been expanded to explain better both Table 1 and Figure 1.
+Starting in the first paragraph (p. <!-- # TODO: Complete -->, l. <!-- # TODO: Complete -->), this section now reads:
+
+> Their parameters are shown in Table 1,
+> both in the original `r MODEL_ACRONYM` metric and
+> in the multidimensional metric,
+> in two different cases with different latent covariance structures.
+> In the first case (columns $`r CORR_ORTH_OUT`$),
+> the two latent dimensions are independent;
+> in the second one (columns $`r CORR_ORTH_OUT`$),
+> the correlation between the dimensions is .5.
+> The same `r MODEL_ACRONYM` parameters are used in both cases,
+> but their (correlation-based) multidimensional parameters change when different
+> correlations are used.
+> 
+> Panel (a) of Figure 1 shows
+> the item representation in the uncorrelated case,
+> while the correlated case is shown in panels (b) to (d)
+> under different conditions.
+> In all the four panels, the contours of a bivariate standard normal density
+> with the corresponding correlation are plot along with the items
+> (from outer to inner, the contours are the `r contour_breaks_out` of
+> the density at its mode).
+> Note that the stretched ellipses in panel (c) are the result of transforming
+> the distribution in the latent space to the test space,
+> according to Equation (24).
+> This implies that the covariance matrix of the distribution projected into
+> the test space is transformed inversely as in the latent space, that is,
+> $`r COV_MATRIX_TRANSF_TS_EQ`$.
+> 
+> In panel (b), the item coordinates are computed from
+> the agnostic version of the parameters.
+> These coordinates are the same ones as in the first case, but
+> their relative position with respect to the latent trait distribution is not
+> ---This is the effect of disregarding the covariance structure.
+> Panel (c) shows the items with their coordinates computed using
+> the correlation-based version of the parameters. [...]
+> 
+> Comparing it to panel (b), we can see that the item coordinates are in general
+> not the same as the ones computed with the agnostic version of the parameters;
+> this effect is especially salient in items 2 and 3,
+> which are close to the 90% contour in the agnostic version,
+> but are almost along the 50% contour in the correlation-based one. [...]
+> 
+> A comparison of panels (a) and (c) allows understanding the effect of
+> a positive correlation on the $`r MDISC_CORR_PARAM`$
+> (and consequently the $`r MIL_CORR_PARAM`$) value:
 
@@
 
-<!-- # TODO: Explain changes in text -->
+> Finally, panel (d) illustrates the effect of plotting the coordinates directly
+> on the canonical basis; the relative positions and distances are correctly
+> represented (which allow a more direct comparison with panel (b)),
+> but the angles of the items with the axes are distorted:
+> For example, item 1 forms 45-degree angles with both axes,
+> instead of the 30-degree ones shown in Table 1.
 

### 2025-10-08 output/Response_to_action_editor_reviewers-2.Rmd
Explain inclusion of "matrix whitening" to editor

@@
 <!-- # TODO: Discuss "latent space" transformation is an artifact, not something to "explore". -->
+
+In addition to the previous changes, we have made a minor addition in the "Covariance-based version" section, to refer to the transformation that turns the latent trait distribution into a standard normal by its usual name in the literature, i.e., a whitening transformation.
+The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:
+
+> Therefore, an orthonormal basis should represent independent,
+> standard coordinates.
+> Formally, if $`r COV_MATRIX_ORTH`$ is the covariance matrix of
+> $`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$,
+> which implies that Equation \@ref(eq:change-basis) actually represents
+> a *whitening transformation* [@fukunaga_introduction_2013].
+> Given this, $`r COV_MATRIX_TRANSF_EQ`$, and thus
+
+(Please also note the update in the title page to acknowledge the colleague who brought this previous literature to our attention, as well as the valuable contributions of the reviewers.)
 

### 2025-10-09 output/Main_text.Rmd
Reorder "Graphical Representation Example" content

@@
 the distance to the origin depends on the $`r MDISC_SYM`$.
-Also, in the covariance-based,
+Also, in the covariance-based version,
 the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
@@
 The example in Figure \@ref(fig:item-plot-out) showcases a representation of
-a set of items in two different bidimensional spaces.
+a set of items with two different latent trait distributions.
 (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
@@
 both in the original `r MODEL_ACRONYM` metric and
-in the multidimensional metric,
-in two different cases with different latent covariance structures.
+in the multidimensional metric under the two cases,
+which differ in the latent trait correlation.
 In the first case (columns $`r CORR_ORTH_OUT`$),
 the two latent dimensions are independent;
-in the second one (columns $`r CORR_ORTH_OUT`$),
-the correlation between the dimensions is .5.
+in the second one ($`r CORR_OBL_OUT`$),
+the correlation between them is .5.
 The same `r MODEL_ACRONYM` parameters are used in both cases,
-but their (correlation-based) multidimensional parameters change when different
-correlations are used.
+but their (correlation-based) multidimensional parameters change
+with the correlation.
+If we used the agnostic version instead, the multidimensional parameters,
+given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.
 
@@
 the item representation in the uncorrelated case,
-while the correlated case is shown in panels (b) to (d)
-under different conditions.
-In all the four panels, the contours of a bivariate standard normal density
+while the $`r CORR_OBL_OUT`$ case is shown in panels (b) to (d)
+under different conditions:
+with item coordinates computed from the agnostic version of the parameters
+(panel (b)), or from the correlation-based version,
+either using an appropriate oblique basis (panel (c)),
+or plotting them directly in the canonical basis (panel (d)).
+Keeping the horizontal axis invariant in panel (c)
+allows for a better comparison with the other plots;
+therefore, we have chosen a $`r TRANSFORM_MATRIX_TRANSP_INV`$ that
+only rotates the vertical axis [@harman_modern_1976]:
+
+\begin{equation}
+  `r TRANSFORM_MATRIX_EXAMPLE_EQ`.
+\end{equation}
+
+In all four panels, the contours of a bivariate standard normal density
 with the corresponding correlation are plot along with the items
@@
 the density at its mode).
-Note that the stretched ellipses in panel (c) are the result of transforming
+The stretched ellipses in panel (c) are the result of transforming
 the distribution in the latent space to the test space,
@@
 This implies that the covariance matrix of the distribution projected into
-the test space is transformed inversely as in the latent space, that is,
+the test space is transformed inversely as in the latent space, i.e.,
 $`r COV_MATRIX_TRANSF_TS_EQ`$.
 
-In panel (b), the item coordinates are computed from
-the agnostic version of the parameters.
-These coordinates are the same ones as in the first case, but
+Panels (a) and (b) show that,
+although the agnostic coordinates are the same in both cases,
 their relative position with respect to the latent trait distribution is not
 ---This is the effect of disregarding the covariance structure.
-Panel (c) shows the items with their coordinates computed using
-the correlation-based version of the parameters.
-To keep the horizontal axis invariant,
-we have defined $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
-it only rotates the vertical axis [@harman_modern_1976]:
-
-\begin{equation}
-  `r TRANSFORM_MATRIX_EXAMPLE_EQ`.
-\end{equation}
-
-Comparing it to panel (b), we can see that the item coordinates are in general
-not the same as the ones computed with the agnostic version of the parameters;
-this effect is especially salient in items 2 and 3,
+The correlation-based coordinates, plot in panels (c) and (d),
+are also different in general from their agnostic counterpart (panel (b)).
+This difference is especially salient in the location of items 2 and 3,
 which are close to the 90% contour in the agnostic version,
 but are almost along the 50% contour in the correlation-based one.
-Finally, panel (d) illustrates the effect of plotting the coordinates directly
-on the canonical basis; the relative positions and distances are correctly
-represented (which allow a more direct comparison with panel (b)),
-but the angles of the items with the axes are distorted:
+Panel (d) also illustrates the effect of plotting the coordinates directly
+on the canonical basis;
+the item locations and relative positions to the distribution
+are correctly represented (allowing a more direct comparison with panel (b)),
+but their angles with the axes and discrimination parameters are distorted:
 For example, item 1 forms 45-degree angles with both axes,
-instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).
+instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out);
+items 4 and 5 seem to be equally discriminating,
+when Table \@ref(tab:example-items-table-out) panel (c) show they are not.
 
@@
 :::
+
+<br>
 

### 2025-10-09 output/Main_text.Rmd
Update "Graphical Representation Example" text
Make minor clarifications in the text.

@@
 their relative position with respect to the latent trait distribution is not
----This is the effect of disregarding the covariance structure.
+---this is the effect of disregarding the covariance structure.
 The correlation-based coordinates, plot in panels (c) and (d),
 are also different in general from their agnostic counterpart (panel (b)).
-This difference is especially salient in the location of items 2 and 3,
-which are close to the 90% contour in the agnostic version,
-but are almost along the 50% contour in the correlation-based one.
+Although they do not vary in some cases (items 1 and 4),
+the difference is especially salient in the location of items 2 and 3,
+which are close to the 90% contour in the agnostic version
+but almost on the 50% contour in the correlation-based one.
 Panel (d) also illustrates the effect of plotting the coordinates directly
 on the canonical basis;
-the item locations and relative positions to the distribution
-are correctly represented (allowing a more direct comparison with panel (b)),
-but their angles with the axes and discrimination parameters are distorted:
-For example, item 1 forms 45-degree angles with both axes,
-instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out);
-items 4 and 5 seem to be equally discriminating,
-when Table \@ref(tab:example-items-table-out) panel (c) show they are not.
+the item locations and their positions relative to the distribution
+are correctly represented (which makes comparing with panel (b) easier),
+but their discrimination parameters and angles with the axes are distorted:
+For example, items 4 and 5 seem to be equally discriminating,
+when Table \@ref(tab:example-items-table-out) panel (c) show they are not;
+and item 1 forms 45-degree angles with both axes,
+instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).
 
@@
 the distance to the origin:
-The item vector is applied at the origin when its value is null (item 1),
+An item vector is applied at the origin when its value is null (item 1),
 and tends to be shifted
@@
 as each of them is orthogonal to one of the axes in the oblique space,
-even when they are not parallel to the other axis.
+even when they are not parallel to the other one
 

### 2025-10-09 output/Response_to_action_editor_reviewers-2.Rmd
Update response to Reviewer 4
Include the latest changes in the main text in the response to the reviewer.

@@
 > for simplicity, but note that they represent the coordinates of
-> a latent trait point transformed to the test space,
+> a latent trait point expressed in test space coordinates,
 > that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
@@
 The text also explains the different interpretation of the coordinates relative to the distribution now.
-Please see the new paragraphs added in third and fourth place in section "Graphical Representation Example" (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
+Please see the new paragraph added in third place in section "Graphical Representation Example" (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
 
-> In panel (b), the item coordinates are computed from
-> the agnostic version of the parameters.
-> These coordinates are the same ones as in the first case, but
+> Panels (a) and (b) show that,
+> even if the agnostic coordinates are the same in both cases,
 > their relative position with respect to the latent trait distribution is not
-> ---This is the effect of disregarding the covariance structure.
-> Panel (c) shows the items with their coordinates computed using
-> the correlation-based version of the parameters. [...]
->
-> Comparing it to panel (b), we can see that the item coordinates are in general
-> not the same as the ones computed with the agnostic version of the parameters;
-> this effect is especially salient in items 2 and 3,
-> which are close to the 90% contour in the agnostic version,
-> but are almost along the 50% contour in the correlation-based one. [...]
+> ---this is the effect of disregarding the covariance structure.
+> The correlation-based coordinates, plot in panels (c) and (d),
+> are also different in general from their agnostic counterpart (panel (b)).
+> Although they do not vary in some cases (items 1 and 4),
+> the difference is especially salient in the location of items 2 and 3,
+> which are close to the 90% contour in the agnostic version
+> but almost on the 50% contour in the correlation-based one. [...]
 
@@
 
+> [...] The example in Figure 1 showcases a representation of
+> a set of items with two different latent trait distributions.
+> (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
+> for simplicity, but note that they represent the coordinates of
+> a latent trait point expressed in test space coordinates,
+> that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
 > Their parameters are shown in Table 1,
 > both in the original `r MODEL_ACRONYM` metric and
-> in the multidimensional metric,
-> in two different cases with different latent covariance structures.
+> in the multidimensional metric under the two cases,
+> which differ in the latent trait correlation.
 > In the first case (columns $`r CORR_ORTH_OUT`$),
 > the two latent dimensions are independent;
-> in the second one (columns $`r CORR_ORTH_OUT`$),
-> the correlation between the dimensions is .5.
+> in the second one ($`r CORR_OBL_OUT`$),
+> the correlation between them is .5.
 > The same `r MODEL_ACRONYM` parameters are used in both cases,
-> but their (correlation-based) multidimensional parameters change when different
-> correlations are used.
+> but their (correlation-based) multidimensional parameters change
+> with the correlation.
+> If we used the agnostic version instead, the multidimensional parameters,
+> given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.
 > 
@@
 > the item representation in the uncorrelated case,
-> while the correlated case is shown in panels (b) to (d)
-> under different conditions.
-> In all the four panels, the contours of a bivariate standard normal density
+> while the $`r CORR_OBL_OUT`$ case is shown in panels (b) to (d)
+> under different conditions:
+> with item coordinates computed from the agnostic version of the parameters
+> (panel (b)), or from the correlation-based version,
+> either using an appropriate oblique basis (panel (c)),
+> or plotting them directly in the canonical basis (panel (d)).
+> Keeping the horizontal axis invariant in panel (c)
+> allows for a better comparison with the other plots;
+> therefore, we have chosen a $`r TRANSFORM_MATRIX_TRANSP_INV`$ that
+> only rotates the vertical axis [@harman_modern_1976]:
+> 
+> \begin{equation}
+>   `r TRANSFORM_MATRIX_EXAMPLE_EQ`.
+> \end{equation}
+> 
+> In all four panels, the contours of a bivariate standard normal density
 > with the corresponding correlation are plot along with the items
@@
 > the density at its mode).
-> Note that the stretched ellipses in panel (c) are the result of transforming
+> The stretched ellipses in panel (c) are the result of transforming
 > the distribution in the latent space to the test space,
-> according to Equation (24).
+> according to Equation \@ref(eq:trait-coords-spaces-transform).
 > This implies that the covariance matrix of the distribution projected into
-> the test space is transformed inversely as in the latent space, that is,
+> the test space is transformed inversely as in the latent space, i.e.,
 > $`r COV_MATRIX_TRANSF_TS_EQ`$.
 > 
-> In panel (b), the item coordinates are computed from
-> the agnostic version of the parameters.
-> These coordinates are the same ones as in the first case, but
+> Panels (a) and (b) show that,
+> even if the agnostic coordinates are the same in both cases,
 > their relative position with respect to the latent trait distribution is not
-> ---This is the effect of disregarding the covariance structure.
-> Panel (c) shows the items with their coordinates computed using
-> the correlation-based version of the parameters. [...]
-> 
-> Comparing it to panel (b), we can see that the item coordinates are in general
-> not the same as the ones computed with the agnostic version of the parameters;
-> this effect is especially salient in items 2 and 3,
-> which are close to the 90% contour in the agnostic version,
-> but are almost along the 50% contour in the correlation-based one. [...]
+> ---this is the effect of disregarding the covariance structure.
+> The correlation-based coordinates, plot in panels (c) and (d),
+> are also different in general from their agnostic counterpart (panel (b)).
+> Although they do not vary in some cases (items 1 and 4),
+> the difference is especially salient in the location of items 2 and 3,
+> which are close to the 90% contour in the agnostic version
+> but almost on the 50% contour in the correlation-based one.
+> Panel (d) also illustrates the effect of plotting the coordinates directly
+> on the canonical basis;
+> the item locations and their positions relative to the distribution
+> are correctly represented (which makes comparing with panel (b) easier),
+> but their discrimination parameters and angles with the axes are distorted:
+> For example, items 4 and 5 seem to be equally discriminating,
+> when Table 1 panel (c) show they are not;
+> and item 1 forms 45-degree angles with both axes,
+> instead of the 30-degree ones shown in Table 1.
 > 
@@
 > a positive correlation on the $`r MDISC_CORR_PARAM`$
-> (and consequently the $`r MIL_CORR_PARAM`$) value:
+> (and consequently the $`r MIL_CORR_PARAM`$) value: [...]
 
@@
 Incidentally, this also serves the purpose of allowing an easier comparison of the coordinates computed from the two versions (the agnostic, and the correlation-based one) of the multidimensional parameters.
-We have briefly explained this, adding in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->:
-
-> Finally, panel (d) illustrates the effect of plotting the coordinates directly
-> on the canonical basis; the relative positions and distances are correctly
-> represented (which allow a more direct comparison with panel (b)),
-> but the angles of the items with the axes are distorted:
-> For example, item 1 forms 45-degree angles with both axes,
-> instead of the 30-degree ones shown in Table 1.
+We have briefly explained this in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete --> (please refer to the changes explained above).
 

### 2025-10-09 output/Main_text.Rmd
Update response to Reviewer 4
Include the latest changes in the main text in the response to the reviewer.

@@
 for simplicity, but note that they represent the coordinates of
-a latent trait point transformed to the test space,
+a latent trait point expressed in test space coordinates,
 that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
@@
 Panels (a) and (b) show that,
-although the agnostic coordinates are the same in both cases,
+even if the agnostic coordinates are the same in both cases,
 their relative position with respect to the latent trait distribution is not
@@
 as each of them is orthogonal to one of the axes in the oblique space,
-even when they are not parallel to the other one
+even when they are not parallel to the other one.
 

### 2025-10-09 output/Response_to_action_editor_reviewers-2.Rmd
Correct the grammar of "plot" in simple past
Should be "plotted" instead

@@
 > ---this is the effect of disregarding the covariance structure.
-> The correlation-based coordinates, plot in panels (c) and (d),
+> The correlation-based coordinates, plotted in panels (c) and (d),
 > are also different in general from their agnostic counterpart (panel (b)).
@@
 > In all four panels, the contours of a bivariate standard normal density
-> with the corresponding correlation are plot along with the items
+> with the corresponding correlation are plotted along with the items
 > (from outer to inner, the contours are the `r contour_breaks_out` of
@@
 
-Finally, we have added a fourth panel (panel (d)) that illustrates how the items would be plot if we omitted using the proper test space basis (and just using the canonical basis instead).
+Finally, we have added a fourth panel (panel (d)) that illustrates how the items would be plotted if we omitted using the proper test space basis (and just using the canonical basis instead).
 Incidentally, this also serves the purpose of allowing an easier comparison of the coordinates computed from the two versions (the agnostic, and the correlation-based one) of the multidimensional parameters.

### 2025-10-09 output/Main_text.Rmd
Correct the grammar of "plot" in simple past
Should be "plotted" instead

@@
 In all four panels, the contours of a bivariate standard normal density
-with the corresponding correlation are plot along with the items
+with the corresponding correlation are plotted along with the items
 (from outer to inner, the contours are the `r contour_breaks_out` of
@@
 ---this is the effect of disregarding the covariance structure.
-The correlation-based coordinates, plot in panels (c) and (d),
+The correlation-based coordinates, plotted in panels (c) and (d),
 are also different in general from their agnostic counterpart (panel (b)).
@@
 #|   (b) the agnostic multidimensional parameters, or
-#|   with the covariance/correlation-based ones, which are then plot either
+#|   with the correlation-based ones, which are then plotted either
 #|   (c) in an oblique basis, appropriate for the covariance structure, or

### 2025-10-09 output/Main_text.Rmd
Attempt to respond to editor's comment 3

@@
 the item psychometric properties are paralleled by their geometrical properties,
-and how this holds for any case of unchanging monotony
+and how this holds for any case of unchanging monotonicity
 of the item response function.
@@
 but for a more general, semi-parametric formulation
-(although constrained to non-decreasing monotony).
+(although constrained to non-decreasing monotonicity).
 However, they provided no formal proof for this result either.
@@
 case of the `r MODEL_ACRONYM` model.
-
-As we have seen, taking the latent space structure into account may have a
-substantial effect in the resulting values and interpretation of the parameters,
-something not contemplated by the original
-(hereby referred to as agnostic) version.
-Moreover, in applications where the latent space structure is critical
-(e.g., multi-group IRT, equating, and linking),
-the agnostic version may overlook relevant differences in
-the values of the multidimensional parameters.
-Different covariance structures across groups may significantly affect
-the value of these parameters,
-due to differences in either the variances or the correlation structure.
-Therefore, even when the model parameters are equal across groups,
-the multidimensional parameters may not be the same
-if the covariance structure differs across groups.
-Conversely, the same multidimensional parameters may correspond
-to different parameter models.
-The subsequent invariance of either the M2PL parameters
-or the multidimensional ones may lead to different interpretations of
-the same model when applied to different groups.
-Although this is an empirical question that requires further research,
-it could have significant implications for future theoretical developments
-in the understanding of MIRT theory.
 
@@
 given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
-However, the computations in the oblique space allow us to obtain
-the direction cosines relative to the original latent dimensions,
-which may give relevant information if these are actual, substantive traits.
+However, the representation in the test space allows us to obtain
+the direction cosines relative to the original latent trait axes,
+which may be useful if they represent actual, substantive traits.
 Computing the parameters in the orthonormalized (or just orthogonalized) space,
@@
 the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
+In fact, we have only defined the latent space as a geometrical artifact
+for obtaining the multidimensional parameters.
+Interpreting the items in the latent space
+could only mislead researchers to wrong conclusions.
+<!-- #TODO: Iem plots in latent space in Supplement? -->
 We hope our results help shed a clearer light on
 this often misunderstood topic.
-
 It is worth noting that a similar relationship exists in
@@
 Making a distinction between the two spaces can also have a significant impact
-on our understanding of MIRT applications that involve several latent spaces,
-such as multi-group equating and linking.
+on our understanding of MIRT applications that involve several latent spaces
+(e.g., multi-group IRT, equating, and linking).
 Although the usual transformations among coordinate systems still apply
@@
 
+As we have seen, taking the latent space structure into account may have a
+substantial effect in the resulting values and interpretation of the parameters,
+something not contemplated by the original
+(hereby referred to as agnostic) version.
+Interpreting the item properties correctly requires a solid understanding of
+their relationship with the latent covariance structure,
+and we have seen that this can only be achieved
+with a proper geometrical representation.
+Although simple items are not affected much by this issue,
+as MIRT evolves and converges with the common factor psychometric tradition,
+we find MIRT models applied more often to new problems,
+involving complex multidimensional item structures
+(even including items with opposite-signed loadings).
+Our results will be especially relevant in such cases,
+where the item interpretation may be more affected.
+For example, the relative importance of an item in scoring a given trait
+could be downplayed if it is strongly aligned with that dimension
+but this alignment is overlooked.
+Moreover, in multi-group applications,
+where differences in the latent space structure is critical,
+the agnostic version may overlook relevant differences in
+the values of the multidimensional parameters.
+Different covariance structures across groups may significantly affect
+the value of these parameters,
+due to differences in either the variances or the correlation structure.
+Therefore, even when the model parameters are equal across groups,
+the multidimensional parameters may not be the same
+if the covariance structures differ.
+Conversely, the same multidimensional parameters may correspond
+to different parameter models.
+The subsequent invariance of either the M2PL parameters
+or the multidimensional ones may lead to different interpretations of
+the same model when applied to different groups.
+Although this is an empirical question that requires further research,
+it could have significant implications for future theoretical developments
+in the understanding of MIRT theory.
+
 Finally, we must also warn against generalizing these results

### 2025-10-10 output/Response_to_action_editor_reviewers-2.Rmd
Put contour breaks text output in example script
This makes it visible also to the response to the reviewers document

@@
 source("R/Constants.R",                 encoding = 'UTF-8')
+source("src/Graphical_example_paper.R", encoding = 'UTF-8')
 ```
@@
 
-In addition to the previous changes, we have made a minor addition in the "Covariance-based version" section, to refer to the transformation that turns the latent trait distribution into a standard normal by its usual name in the literature, i.e., a whitening transformation.
+In addition to the previous changes, we have made a minor addition in the "Covariance-based version" section, to refer to the transformation that turns the latent trait distribution into a standard normal by its usual name in the literature, i.e., a "whitening transformation".
 The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:
@@
 
-(Please also note the update in the title page to acknowledge the colleague who brought this previous literature to our attention, as well as the valuable contributions of the reviewers.)
+(Please also note the update in the title page to acknowledge the colleague who brought this previous literature to our attention, as well as the contributions of the reviewers.)
 

### 2025-10-10 src/Graphical_example_paper.R

@@ -58,9 +58,12 @@ PANEL_LABEL_X <- -2.5
 PANEL_LABEL_Y <-  2
 
 # Latent space grid for density contours:
-PROB_AXIS_2D   <- seq(-3, 3, by = 0.05) # Axis for the density contour plots
-CONTOUR_BREAKS <- c(.1, .5, .9)         # Breaks for the density contour plots
-CONTOUR_COLOR  <- "#5c6eb1" # Color for the probability contour lines
+PROB_AXIS_2D       <- seq(-3, 3, by = 0.05) # Axis for density contour plots
+CONTOUR_BREAKS     <- c(.1, .5, .9)         # Breaks for density contour plots
+CONTOUR_COLOR      <- "#5c6eb1"             # Color for density contour lines
+contour_breaks_out <- CONTOUR_BREAKS |>     # Text output for contour breaks
+  percent() |>
+  glue_collapse(sep = ', ', last = ', and ')
 
 latent_space_grid <- expand_grid( # Grid for the density contour plots
   trait_1 = PROB_AXIS_2D,

### 2025-10-10 output/Main_text.Rmd
Put contour breaks text output in example script
This makes it visible also to the response to the reviewers document
Correct the appearances of multi-group MIRT
As it appears several times, it must be referred to accordingly (using appropriate connectors when repeating).
Also, minor style changes are made.

@@
 
-```{r format-contour-plot-props}
-contour_breaks_out <- CONTOUR_BREAKS |>
-  percent() |>
-  glue_collapse(sep = ', ', last = ', and ')
-```
-
 Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
@@
 changes of latent-space basis.
-Once again, this is relevant in applications where
-potentially different bases are involved.
+This is especially relevant in applications where
+potentially different bases are involved
+(e.g., multi-group IRT, equating, and linking).
 Furthermore, we have found that using this version yields a norm in
@@
 the latent covariance structure into account,
-but also makes the latent space norm invariant to changes of basis.
+but makes the latent space norm also invariant to changes of basis.
 Given these considerations,
@@
 the latent covariance (or correlation) matrix as its inner product matrix.
-On the other hand, the latent space,
-where the person parameters are represented,
+The latent space, where the person parameters are represented,
 has the inverse of the covariance (correlation) matrix as
-its inner product matrix.
+its inner product matrix though.
 [Note that in the original derivations by @reckase_difficulty_1985;
@@
 Interpreting the items in the latent space
-could only mislead researchers to wrong conclusions.
+could only lead researchers to wrong conclusions.
 <!-- #TODO: Iem plots in latent space in Supplement? -->
@@
 this often misunderstood topic.
+
 It is worth noting that a similar relationship exists in
@@
 Making a distinction between the two spaces can also have a significant impact
-on our understanding of MIRT applications that involve several latent spaces
-(e.g., multi-group IRT, equating, and linking).
+on our understanding of multi-group MIRT applications,
+which involve several latent spaces.
 Although the usual transformations among coordinate systems still apply
@@
 However, as they are mutually dependent,
-any transformation in any of the two spaces will imply
+any transformation in any of them will imply
 the inverse transformation in the other one,
-thus preserving the relationship between the two spaces.
+thus preserving their relationship.
 Nevertheless, using non-orthogonal rotation matrices for aligning the items
@@
 
-As we have seen, taking the latent space structure into account may have a
+We have seen that taking the latent space structure into account may have a
 substantial effect in the resulting values and interpretation of the parameters,
@@
 their relationship with the latent covariance structure,
-and we have seen that this can only be achieved
+and we have shown that this can only be achieved
 with a proper geometrical representation.
-Although simple items are not affected much by this issue,
+Simple items are not affected much by this issue but,
 as MIRT evolves and converges with the common factor psychometric tradition,
@@
 (even including items with opposite-signed loadings).
-Our results will be especially relevant in such cases,
-where the item interpretation may be more affected.
+Our results will be especially relevant in those cases,
+where the interpretation of the items may be affected more drastically.
 For example, the relative importance of an item in scoring a given trait
-could be downplayed if it is strongly aligned with that dimension
+could be downplayed if it is strongly aligned with that trait dimension
 but this alignment is overlooked.
-Moreover, in multi-group applications,
+Once again, in multi-group applications,
 where differences in the latent space structure is critical,
@@
 the values of the multidimensional parameters.
-Different covariance structures across groups may significantly affect
-the value of these parameters,
+Different covariance structures across groups may significantly affect them
 due to differences in either the variances or the correlation structure.

### 2025-10-10 output/Response_to_action_editor_reviewers-2.Rmd
Correct equation references in response document
Equation references need to be explicit in the response document.

@@
 > $`r TRAIT_ORTH_COORDS`$, then $`r COV_MATRIX_ORTH_EQ`$,
-> which implies that Equation \@ref(eq:change-basis) actually represents
+> which implies that Equation (2) actually represents
 > a *whitening transformation* [@fukunaga_introduction_2013].
@@
 > the distribution in the latent space to the test space,
-> according to Equation \@ref(eq:trait-coords-spaces-transform).
+> according to Equation (24).
 > This implies that the covariance matrix of the distribution projected into

### 2025-10-10 output/Response_to_action_editor_reviewers-2.Rmd
Add response to editor's third comment

@@
 
-<!-- # TODO: Discuss complex items, and the proper representation in the test space. -->
-<!-- # TODO: Discuss "latent space" transformation is an artifact, not something to "explore". -->
+We have tried our best to give a satisfactory response to this comment.
+Although we acknowledge that we are not completely sure we have succeeded, we have carefully thought about the broader implications to our results, and made changes to our "Discussion" section to address this comment.
+Besides some minor changes to adapt our discussion (please refer to the tracked changes in the main text of the submission), we have made two major changes.
+The first one is the inclusion of additional comments to discuss the implications in terms of the test results and how their uses could be affected.
+They have been included in what was previously the second paragraph;
+we consider this text was misplaced in our previous version and have moved it accordingly, nearly to the end of the section.
+After the changes, this paragraph (now in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) reads:
 
-In addition to the previous changes, we have made a minor addition in the "Covariance-based version" section, to refer to the transformation that turns the latent trait distribution into a standard normal by its usual name in the literature, i.e., a "whitening transformation".
+> We have seen that taking the latent space structure into account may have a
+> substantial effect in the resulting values and
+> interpretation of the parameters, something not contemplated by the original
+> (hereby referred to as agnostic) version.
+> Interpreting the item properties correctly requires a solid understanding of
+> their relationship with the latent covariance structure,
+> and we have shown that this can only be achieved
+> with a proper geometrical representation.
+> Simple items are not affected much by this issue but,
+> as MIRT evolves and converges with the common factor psychometric tradition,
+> we find MIRT models applied more often to new problems,
+> involving complex multidimensional item structures
+> (even including items with opposite-signed loadings).
+> Our results will be especially relevant in those cases,
+> where the interpretation of the items may be affected more drastically.
+> For example, the relative importance of an item in scoring a given trait
+> could be downplayed if it is strongly aligned with that trait dimension
+> but this alignment is overlooked.
+> Once again, in multi-group applications,
+> where differences in the latent space structure is critical,
+> the agnostic version may overlook relevant differences in
+> the values of the multidimensional parameters.
+> Different covariance structures across groups may significantly affect them
+> due to differences in either the variances or the correlation structure.
+> Therefore, even when the model parameters are equal across groups,
+> the multidimensional parameters may not be the same
+> if the covariance structures differ.
+> Conversely, the same multidimensional parameters may correspond
+> to different parameter models.
+> The subsequent invariance of either the M2PL parameters
+> or the multidimensional ones may lead to different interpretations of
+> the same model when applied to different groups.
+> Although this is an empirical question that requires further research,
+> it could have significant implications for future theoretical developments
+> in the understanding of MIRT theory.
+
+Regarding the insights that can be obtained from the different representations,
+we hope that the changes to Figure 1 and the subsequent additions
+should account for how the test space representation should be interpreted.
+Regarding the latent space, our latest version already stated that "the axes defined by the test space basis, and not the ones defined by the latent space basis, are a geometrical representation of the covariance structure.
+Indeed, the cosines of the angles among the test space axes are the geometric equivalent of the correlations.
+This result is both surprising and counterintuitive [...]".
+The manuscript implicitly showed that the latent space transformation is as an "geometric artifact", necessary for arriving at the proper representation of the items in the test space.
+We acknowledge though that this might not be necessarily clear in the previous submission.
+The second change in the discussion consists of stating this explicitly now at the end of the fourth paragraph (previously the fifth one) of the "Discussion" section, in p.
+<!-- # TODO: Complete -->, ll.
+<!-- # TODO: Complete -->:
+
+> [...] In fact, we have only defined the latent space as a geometrical artifact
+> for obtaining the multidimensional parameters.
+> Interpreting the items in the latent space
+> could only lead researchers to wrong conclusions. [...]
+
+Besides the aforementioned changes, we would like to highlight a minor addition to the "Covariance-based version" section, to refer to the transformation of the latent trait distribution into another one with an identity covariance matrix by its usual name in the literature, i.e., a "whitening transformation".
 The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:
@@
 
-(Please also note the update in the title page to acknowledge the colleague who brought this previous literature to our attention, as well as the contributions of the reviewers.)
+(Please note also the update in the title page to acknowledge the colleague who brought this previous literature to our attention, as well as the contributions of the reviewers.)
 

### 2025-10-17 notebooks/ESM_latent-space-representation.Rmd
Draft "latent space plotting" supplementary file

---
output:
  officedown::rdocx_document:
    tables:
      style: APA_Like_style
      caption:
        style: Image Caption
        pre: "Table S."
        sep: ' '
    mapstyles:
      Body Text: First Paragraph
    base_format: bookdown::word_document2
    reference_docx: ../www/Template.docx
    number_sections: no
    keep_md:         no
    fig_width:       4
    fig_height:      4
bibliography: [../www/Multidimensional-parameters-MCLM.bib, ../www/packages.bib]
csl:          ../www/apa-old-doi-prefix.csl
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| include: false

# Includes:

library(here)
library(knitr)
library(patchwork)
library(officer)

ROOT_DIR <- here()

opts_knit$set(root.dir = ROOT_DIR)
opts_chunk$set(
  echo     = FALSE,
  message  = FALSE,
  cache    = FALSE,
  dev      = "png",
  dpi      = 1200L,
  dev.args = list(type = "cairo-png"),
  fig.cap.fp_text = fp_text_lite(italic = FALSE, bold = TRUE)
)

# Output configuration options:
options(digits = 3)
```

```{r reset-root-wd}
#| cache: false

# Necessary for knitr to find the source file:
opts_knit$set(root.dir = ROOT_DIR)
```

```{r sources}
source("src/Graphical_example_paper.R", encoding = 'UTF-8')
```

# Representation in the latent space

```{r}
# Compute items in latent space (needs to be computed manually, because the
#   covariance matrix is computed from the transform matrix in
#   `compute_mirt_coords()` and the inverse matrix would be used instead):
items_ls <- items_oblique_coords |>
  pivot_longer(origin_1:end_2) |>
  separate(name, into = c('end', 'coord'), sep = "_", remove = FALSE) |>
  group_by(item, end) |>
  mutate(value = transform_matrix %*% value |> drop()) |>
  ungroup() |>
  pivot_wider(id_cols = item:end_transf_2, values_from = value)

## Compose latent space plot:

# Oblique grid:
grid_ls <- transform_grid(
  transform_matrix,
  x_limits = c(-3.2,  3.2),
  y_limits = c(-2.65, 2.65),
  break_step = 1,
  linetype_grid = "17",
  linewidth = LINE_WIDTH,
  axis_ticks = "axis"
)

# Oblique items:
items_geom_ls_obl <- geom_segment(
  mapping   = aes(
    origin_1, origin_2,
    xend  = end_1, yend = end_2,
    color = item
  ),
  data      = items_ls |> arrange(desc(item)), # Plot them in reverse order
  arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
  linejoin  = "mitre",
  linewidth = VECTOR_WIDTH
)

plot_oblique_ls <- grid_ls +
  contour_orth +
  items_geom_ls_obl +
  colorscale
```

```{r items-latent-space-plot-out}
#| cache:          false
#| fig.topcaption: true
#| fig.height:     4.7
#| fig.width:      6.65
#| warning:        false
plot_oblique_ls
```

### 2025-10-22 notebooks/ESM_latent-space-representation.Rmd
Output the latent space plot in SVG format

@@
 ROOT_DIR <- here()
+DOC_DIR  <- file.path(ROOT_DIR, "output")
 
@@
 plot_oblique_ls
+
+ggsave(file.path(DOC_DIR, "Figure_ESM.svg"), width = 6.65, height = 4.7)
 ```

### 2025-10-22 output/Main_text.Rmd
Correct singed distance as function of intercept

@@
 $`r MDISC_SYM`$ is strictly positive for non-null vectors.
-Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
+Therefore, the sign of $`r DISTANCE_PARAM`$ is the opposite of
 the sign of $`r INTERCEPT_PARAM`$.
-The implication is that the item is displaced along
-its measurement direction forward or backward with respect to the origin.
+The implication is that the item is displaced along its measurement direction
+forward or backward with respect to the origin, depending on
+whether $`r INTERCEPT_PARAM`$ is negative or positive, respectively.
 As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,

### 2025-10-22 output/Main_text.Rmd
Discard latent space representation supplement
This idea is discarded (for now) for the resubmission.
Apply Grammarly corrections

@@
 arbitrary algebraic basis.
-Then, we define such basis to be a
+Then, we define that basis to be a
 geometrical representation of the measured construct.
@@
 By Equations \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
-and  \@ref(eq:trait-coords-orth-spaces-eq)
+and  \@ref(eq:trait-coords-orth-spaces-eq),
 any coordinate vector $`r TRAIT_COORDS`$ in $`r LS_BASIS`$
@@
 a set of items with two different latent trait distributions.
-(The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
+(The axes are labeled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
 for simplicity, but note that they represent the coordinates of
@@
 the $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
-on the contrary, when the discrimination parameters have opposite signs
-they tend to cancel each other out and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
+on the contrary, when the discrimination parameters have opposite signs,
+they tend to cancel each other out, and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
 as happens with items 2 and 3.
 This effect can be especially noticed by comparing
-the $`r MDISC_CORR_PARAM_ITEM`$ of items 3 and 5:
+the $`r MDISC_CORR_PARAM_ITEM`$ values of items 3 and 5:
 Their discrimination parameters are equal in absolute value,
@@
 With this latent space structure, items that strongly discriminate in either or
-both of these two dimensions change drastically their multidimensional
+both of these two dimensions drastically change their multidimensional
 parameters from the agnostic to the covariance-based version
@@
 relatively low discrimination parameters on those two dimensions
-(compared with the ones in dimension 2 and 3, respectively) are barely affected.
+(compared with the ones in dimensions 2 and 3, respectively)
+are barely affected.
 The most extreme change in these two examples occurs in
@@
 with that dimension.
-In this case, because the discrimination in dimension 4 so low, the situation is
-very similar to item 4 in Table \@ref(tab:example-items-table-out)
-and Figure \@ref(fig:item-plot-out).
+In this case, because the discrimination in dimension 4 is so low,
+the situation is very similar to item 4
+in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).
 
@@
 the orthonormality of the latent space.
-Applying the general procedure outlined by @reckase_difficulty_1985 and later
-extended by @reckase_discriminating_1991
+Applying the general procedure outlined by @reckase_difficulty_1985,
+and later extended by @reckase_discriminating_1991,
 we have obtained a set of multidimensional parameters
@@
 the latent covariance structure into account,
-but makes the latent space norm also invariant to changes of basis.
+but also makes the latent space norm invariant to changes of basis.
 Given these considerations,
@@
 Computing the parameters in the orthonormalized (or just orthogonalized) space,
-by contrast, would require transforming the cosines to
-the original, non-orthonormal basis afterwards.
+by contrast, would require transforming the cosines back to
+the original, non-orthonormal basis.
 Nevertheless, most available MIRT software solutions,
@@
 initially estimate an uncorrelated solution (at least in the exploratory case)
-and apply an oblique rotation afterwards.
+and apply an oblique rotation afterward.
 Taking advantage of the invariance property,
@@
 has the inverse of the covariance (correlation) matrix as
-its inner product matrix though.
+its inner product matrix, though.
 [Note that in the original derivations by @reckase_difficulty_1985;
@@
 could only lead researchers to wrong conclusions.
-<!-- #TODO: Iem plots in latent space in Supplement? -->
 We hope our results help shed a clearer light on
@@
 We have seen that taking the latent space structure into account may have a
-substantial effect in the resulting values and interpretation of the parameters,
+substantial effect on the resulting values and interpretation of the parameters,
 something not contemplated by the original
@@
 Once again, in multi-group applications,
-where differences in the latent space structure is critical,
+where differences in the latent space structure are critical,
 the agnostic version may overlook relevant differences in

### 2025-10-22 output/Main_text.Rmd
Reformulate orthogonalty in Reckase's (1985) Eq. 3
It states clearly that orthogonality is assumed "as early as Eq. 3"
Correct reference to equations
After including new equations 22 and 23, these ones cannot be considered contiguous.

@@
 involved in the expression of the model in polar coordinates.
-Importantly, this shows that
-Reckase's [-@reckase_difficulty_1985] Equation 3 (p. 403)
-assumes the orthonormality of the latent space, as we stated before.
+Importantly, this shows that the orthogonality of the latent space in
+@reckase_difficulty_1985 is assumed as early on as in Equation 3 (p. 403),
+as we stated before.
 
@@
 Hence we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
-Equations \@ref(eq:distance-param) through \@ref(eq:dir-cos-vec-test-space).
+Equations \@ref(eq:distance-param), \@ref(eq:slope-param),
+and \@ref(eq:dir-cos-vec-test-space).
 However, as they depend on the inner product matrix $`r INNER_PROD_MATRIX`$,

### 2025-10-24 output/Response_to_action_editor_reviewers-2.Rmd
Explain rejection of simply dropping "already"
In response to editor. Explanation of why a suggestion by reviewer 4 was rejected, and a different change was implemented instead.

@@
 the item `r MDISC_SYM` should be the algebraic sum of both components, [...]
+
+In p. 12 (line 8), simply dropping "already" made the sentence lose the intended
+meaning that the implicit assumption of orthogonality in @reckase_difficulty_1985 is already present in Equation 3, and not just from Equation 7 (p. 404) on, as one might think in the first place.
+We deem this point important because it took us quite some effort to notice that generalizing the transformation to polar coordinates to non-orthonormal bases implied reformulating this equation.
+In an attempt to keep the intended meaning while improving the sentence structure, we have restated it as:
+
+> [...] this shows that the orthogonality of the latent space in @reckase_difficulty_1985 is assumed as early on as in Equation 3 (p. 403) [...]
 

### 2025-10-24 output/Main_text.Rmd
Update explanation of axis lableing in Figure 1
Implements the suggestion by @mluzardov after discussion.

@@
 we limit ourselves to the bidimensional case here.
-The example in Figure \@ref(fig:item-plot-out) showcases a representation of
-a set of items with two different latent trait distributions.
-(The axes are labeled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
-for simplicity, but note that they represent the coordinates of
-a latent trait point expressed in test space coordinates,
-that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
-Their parameters are shown in Table \@ref(tab:example-items-table-out),
+The example in Figure \@ref(fig:item-plot-out) showcases
+the test space representation of a set of items
+under two different latent trait distributions.
+In formal terms, the proper axis labeling of this figure
+should correspond to the vectors that make up basis $`r TEST_SPACE_BASIS`$,
+in this case, $`r ITEM_BASIS_VECTOR_FIRST`$ and $`r ITEM_BASIS_VECTOR_2`$;
+the coordinates represent the coefficients that multiply each basis vector to
+obtain the corresponding coordinate in the test space.
+However, these coefficients correspond to the values of
+a vector in the test space, be it $`r DISCR_COORDS`$ for $`r DISCR_VECTOR`$,
+or $`r TRAIT_COORDS_TRANSF_TS`$ for $`r TRAIT_VECTOR`$
+(as given by Equation \@ref(eq:trait-coords-spaces-transform)).
+For the sake of simplicity,
+we label the axes as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$,
+highlighting the comparison of the items with the latent trait vectors.
+However, note that these coordinates actually correspond to
+the components of $`r TRAIT_COORDS_TRANSF_TS`$, i.e.,
+$`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.
+
+The parameters of the items in Figure \@ref(fig:item-plot-out) are shown
+in Table \@ref(tab:example-items-table-out),
 both in the original `r MODEL_ACRONYM` metric and
@@
 with the correlation.
-If we used the agnostic version instead, the multidimensional parameters,
-given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.
+(If we used the agnostic version instead, the multidimensional parameters,
+given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.)
 

### 2025-10-24 R/Formulae.R

@@ -1250,6 +1250,8 @@ TRANSF_MATRIX_INV_SQ_CORR_EQ <- latex_eq(INNER_PROD_MATRIX_INV_DEF, CORR_MATRIX)
 
 #### Graphical representation example: ----
 
+ITEM_BASIS_VECTOR_2 <- latex_sub(ITEM_BASIS_VECTOR, 2)
+
 TRAIT_COMP_1  <- latex_sub(TRAIT_SYMBOL, 1)
 TRAIT_COMP_2  <- latex_sub(TRAIT_SYMBOL, 2)
 TRAIT_COORD_1 <- latex_raised_to(TRAIT_COMP_1, exp = TEST_SPACE_BASIS)

### 2025-10-24 output/Main_text.Rmd
Correct use of "item" subindex explaining Figure 1
The {MDISC}_R references are to certain items' values, so the "item" subindex should not be used.

@@
 a positive correlation on the $`r MDISC_CORR_PARAM`$
-(and consequently the $`r MIL_CORR_PARAM`$) value:
+(and consequently the $`r MIL_CORR_PARAM`$) value of an item:
 The components of the discrimination parameters tend to
@@
 Thus, if the discrimination parameters have the same sign, as in items 1 and 5,
-the $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
+the $`r MDISC_CORR_PARAM`$ value tends to increase;
 on the contrary, when the discrimination parameters have opposite signs,
-they tend to cancel each other out, and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
+they tend to cancel each other out, and $`r MDISC_CORR_PARAM`$ decreases,
 as happens with items 2 and 3.
 This effect can be especially noticed by comparing
-the $`r MDISC_CORR_PARAM_ITEM`$ values of items 3 and 5:
+the $`r MDISC_CORR_PARAM`$ values of items 3 and 5:
 Their discrimination parameters are equal in absolute value,
-so their $`r MDISC_CORR_PARAM_ITEM`$ values are equal in the orthogonal space;
+so their $`r MDISC_CORR_PARAM`$ values are equal in the orthogonal space;
 however, the effect of the correlation shrinks the former and
@@
 Finally, note that item 4 has one discrimination parameter equal to 0,
-so its $`r MDISC_CORR_PARAM_ITEM`$ is unaffected by the correlation.
+so its $`r MDISC_CORR_PARAM`$ is unaffected by the correlation.
 

### 2025-10-24 output/Main_text.Rmd
Correct interpretation of parameters
Stating it may affect the test scoring was incorrect; now it refers to the information an item provides about each trait and its quality assessing it.

@@
 where the interpretation of the items may be affected more drastically.
-For example, the relative importance of an item in scoring a given trait
-could be downplayed if it is strongly aligned with that trait dimension
-but this alignment is overlooked.
-Once again, in multi-group applications,
-where differences in the latent space structure are critical,
-the agnostic version may overlook relevant differences in
+For example, using the agnostic version of the parameters and/or an orthogonal plot,
+the alignment of an item with the trait axes could be easily misinterpreted.
+Then, the information it provides about each of those traits
+(and, subsequently, its quality for assessing them)
+could be either downplayed or exaggerated.
+Once again, this is paramount in multi-group applications,
+where differences in the latent space structure are critical and
+the agnostic version could overlook relevant differences in
 the values of the multidimensional parameters.

### 2025-10-29 output/Response_to_action_editor_reviewers-2.Rmd
Update response to editor and reviewer
Update response letter with latest changes in manuscript

@@
 
-Finally, the sentence that was found difficult to understand in the abstract (p. 1, lines 5-7) has been rewritten as follows:
+Finally, the sentence that was found difficult to understand in the abstract (p. 1, ll. 5-7) has been rewritten as follows (p. 1, ll. <!-- # TODO: Complete --> in the current version):
 
@@
 arbitrary algebraic basis.
-Then, we define such basis to be a
+Then, we define that basis to be a
 geometrical representation of the measured construct. [...]
@@
 > We have seen that taking the latent space structure into account may have a
-> substantial effect in the resulting values and
+> substantial effect on the resulting values and
 > interpretation of the parameters, something not contemplated by the original
@@
 > where the interpretation of the items may be affected more drastically.
-> For example, the relative importance of an item in scoring a given trait
-> could be downplayed if it is strongly aligned with that trait dimension
-> but this alignment is overlooked.
-> Once again, in multi-group applications,
-> where differences in the latent space structure is critical,
-> the agnostic version may overlook relevant differences in
+> For example, using the agnostic version of the parameters
+> and/or an orthogonal plot,
+> the alignment of an item with the trait axes could be easily misinterpreted.
+> Then, the information it provides about each of those traits
+> (and, subsequently, its quality for assessing them)
+> could be either downplayed or exaggerated.
+> Once again, this is paramount in multi-group applications,
+> where differences in the latent space structure are critical and
+> the agnostic version could overlook relevant differences in
 > the values of the multidimensional parameters.
@@
 
-Besides the aforementioned changes, we would like to highlight a minor addition to the "Covariance-based version" section, to refer to the transformation of the latent trait distribution into another one with an identity covariance matrix by its usual name in the literature, i.e., a "whitening transformation".
+Besides the aforementioned changes and some minor corrections, we would like to highlight two additional changes: a correction, and a minor addition to the "Covariance-based version" section.
+The first one corrects ll. <!-- # TODO: Complete --> in p. <!-- # TODO: Complete --> in the previous version, which stated:
+
+> the sign of $`r DISTANCE_PARAM`$ is equal to
+> the sign of $`r INTERCEPT_PARAM`$.
+> The implication is that the item is displaced along
+> its measurement direction forward or backward with respect to the origin.
+
+This was an error, as the signs of $`r DISTANCE_PARAM`$ and $`r INTERCEPT_PARAM`$ are opposite to each other.
+With this correction applied, these lines now read (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
+
+> Therefore, the sign of $`r DISTANCE_PARAM`$ is the opposite of
+> the sign of $`r INTERCEPT_PARAM`$.
+> The implication is that the item is displaced along its measurement direction
+> forward or backward with respect to the origin, depending on
+> whether $`r INTERCEPT_PARAM`$ is negative or positive, respectively.
+
+In the "Covariance-based version" section, we have referred to the transformation of the latent trait distribution into one with an identity covariance matrix as  a "whitening transformation", its usual name in the literature (unknown to us before).
 The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:
@@
 > the coordinates of the discrimination vectors in the latent space:
-> By Equations (2), (5), and (23)
+> By Equations (2), (5), and (23),
 > any coordinate vector $`r TRAIT_COORDS`$ in $`r LS_BASIS`$
@@
 > and vice versa. [...]
+
+Subsequently, we have updated ll. <!-- # TODO: Complete --> in p. <!-- # TODO: Complete --> (now p. <!-- # TODO: Complete -->, ll. p. <!-- # TODO: Complete -->) as follows:
+
+> [...] Hence we may define the $`r MIL_PARAM`$ and $`r MDISC_SYM`$ using
+> Equations (20), (21), and (25). [...]
 
@@
 
-> [...] The example in Figure 1 showcases a representation of
-> a set of items with two different latent trait distributions.
-> (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
-> for simplicity, but note that they represent the coordinates of
-> a latent trait point expressed in test space coordinates,
-> that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
-> Their parameters are shown in Table 1,
+> [...] The example in Figure 1 showcases
+> the test space representation of a set of items
+> under two different latent trait distributions.
+> In formal terms, the proper axis labeling of this figure
+> should correspond to the vectors that make up basis $`r TEST_SPACE_BASIS`$,
+> in this case, $`r ITEM_BASIS_VECTOR_FIRST`$ and $`r ITEM_BASIS_VECTOR_2`$;
+> the coordinates represent the coefficients that multiply each basis vector to
+> obtain the corresponding coordinate in the test space.
+> However, these coefficients correspond to the values of
+> a vector in the test space, be it $`r DISCR_COORDS`$ for $`r DISCR_VECTOR`$,
+> or $`r TRAIT_COORDS_TRANSF_TS`$ for $`r TRAIT_VECTOR`$
+> (as given by Equation (24)).
+> For the sake of simplicity,
+> we label the axes as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$,
+> highlighting the comparison of the items with the latent trait vectors.
+> However, note that these coordinates actually correspond to
+> the components of $`r TRAIT_COORDS_TRANSF_TS`$, i.e.,
+> $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.
+> 
+> The parameters of the items in Figure 1 are shown in Table 1,
 > both in the original `r MODEL_ACRONYM` metric and
@@
 > with the correlation.
-> If we used the agnostic version instead, the multidimensional parameters,
-> given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.
+> (If we used the agnostic version instead, the multidimensional parameters,
+> given by the values under $`r CORR_ORTH_OUT`$,
+> would be the same in both cases.)
 > 

### 2025-10-29 output/Letter_to_editor-2.Rmd
Update letter to editor with latest changes

@@
 we have striven to explain our reasons in the clearest possible way.
-Also, besides addressing the comments,
-we have added an additional clarification
+Besides addressing the comments,
+we have thoroughly reviewed the grammar and style,
+in the hope of avoiding these kind of errors in our new submission,
+and have also added an additional clarification
 about the "covariance matrix whitening",

### 2025-10-29 output/Response_to_action_editor_reviewers-2.Rmd
Add line and page references in response letter

@@
 
-Finally, the sentence that was found difficult to understand in the abstract (p. 1, ll. 5-7) has been rewritten as follows (p. 1, ll. <!-- # TODO: Complete --> in the current version):
+Finally, the sentence that was found difficult to understand in the abstract (p. 1, ll. 5-7) has been rewritten as follows (p. 1, ll. 4-7 in the current version):
 
@@
 we consider this text was misplaced in our previous version and have moved it accordingly, nearly to the end of the section.
-After the changes, this paragraph (now in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) reads:
+After the changes, this paragraph (now in p. 38, l. 19 to p. 39, l. 18) reads:
 
@@
 We acknowledge though that this might not be necessarily clear in the previous submission.
-The second change in the discussion consists of stating this explicitly now at the end of the fourth paragraph (previously the fifth one) of the "Discussion" section, in p.
-<!-- # TODO: Complete -->, ll.
-<!-- # TODO: Complete -->:
+The second change in the discussion consists of stating this explicitly now at the end of the fourth paragraph (previously the fifth one) of the "Discussion" section, in p. 37, ll. 16-18:
 
@@
 Besides the aforementioned changes and some minor corrections, we would like to highlight two additional changes: a correction, and a minor addition to the "Covariance-based version" section.
-The first one corrects ll. <!-- # TODO: Complete --> in p. <!-- # TODO: Complete --> in the previous version, which stated:
+The first one corrects ll. 16-18 in p. 19 in the previous version, which stated:
 
@@
 This was an error, as the signs of $`r DISTANCE_PARAM`$ and $`r INTERCEPT_PARAM`$ are opposite to each other.
-With this correction applied, these lines now read (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
+With this correction applied, these lines now read (p. 20, ll. 16-18):
 
@@
 In the "Covariance-based version" section, we have referred to the transformation of the latent trait distribution into one with an identity covariance matrix as  a "whitening transformation", its usual name in the literature (unknown to us before).
-The second paragraph of this section (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads:
+The second paragraph of this section (p. 16, ll. 18-20) now reads:
 
@@
 
-We had overlooked labelling the axes and have done so in this new version;
+We had overlooked labeling the axes and have done so in this new version;
 we thank the reviewer once again for pointing this out.
@@
 which are only necessary for deriving the results.
-This choice of labeling is explained in the text in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->:
+This choice of labeling is explained in the text in p. 23, ll. 1-8 (see also our response to the last comment below):
 
-> (The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
-> for simplicity, but note that they represent the coordinates of
-> a latent trait point expressed in test space coordinates,
-> that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
+> In formal terms, the proper axis labeling of this figure
+> should correspond to the vectors that make up basis $`r TEST_SPACE_BASIS`$,
+> in this case, $`r ITEM_BASIS_VECTOR_FIRST`$ and $`r ITEM_BASIS_VECTOR_2`$;
+> the coordinates represent the coefficients that multiply each basis vector to
+> obtain the corresponding coordinate in the test space.
+> However, these coefficients correspond to the values of
+> a vector in the test space, be it $`r DISCR_COORDS`$ for $`r DISCR_VECTOR`$,
+> or $`r TRAIT_COORDS_TRANSF_TS`$ for $`r TRAIT_VECTOR`$
+> (as given by Equation \@ref(eq:trait-coords-spaces-transform)).
+> For the sake of simplicity,
+> we label the axes as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$,
+> highlighting the comparison of the items with the latent trait vectors.
+> However, note that these coordinates actually correspond to
+> the components of $`r TRAIT_COORDS_TRANSF_TS`$, i.e.,
+> $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.
 
@@
 The text also explains the different interpretation of the coordinates relative to the distribution now.
-Please see the new paragraph added in third place in section "Graphical Representation Example" (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
+Please see the new paragraph added in third place in section "Graphical Representation Example" (p. 24, ll. 11-17):
 
@@
 8 of the previous version of our manuscript, or p.
-<!-- # TODO: Complete --> in the new submission).
+9, l.
+1 in the new submission).
 For example, the tip of the vector representing item 1 is at coordinate (+2, +2).
@@
 For explaining this, we have needed to add and reference two new equations.
-This paragraph (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->) now reads as follows:
+This paragraph (p. 14, l. 16 to p. 15, l. 4) now reads as follows:
 
@@
 
-Subsequently, we have updated ll. <!-- # TODO: Complete --> in p. <!-- # TODO: Complete --> (now p. <!-- # TODO: Complete -->, ll. p. <!-- # TODO: Complete -->) as follows:
+Subsequently, we have updated ll. 14-15 in p. 15 (now ll. p. 18-19) as follows:
 
@@
 - The second paragraph in section "Graphical Representation" now clarifies better that the representation in the test space must use its basis instead of the canonical basis, to make the plot be a faithful representation of the actual directions of the items with respect to the latent trait dimensions.
-This paragraph now reads (p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete -->):
+This paragraph now reads (p. 21, l. 25 to p. 22, l. 4):
 
@@
 - The section "Graphical Representation Example" has been expanded to explain better both Table 1 and Figure 1.
-Starting in the first paragraph (p. <!-- # TODO: Complete -->, l. <!-- # TODO: Complete -->), this section now reads:
+Starting in the first paragraph (p. 22, l. 24 to p. 25, l. 6), this section now reads:
 
@@
 > a positive correlation on the $`r MDISC_CORR_PARAM`$
-> (and consequently the $`r MIL_CORR_PARAM`$) value: [...]
+> (and consequently the $`r MIL_CORR_PARAM`$) value of an item: [...]
 
@@
 Incidentally, this also serves the purpose of allowing an easier comparison of the coordinates computed from the two versions (the agnostic, and the correlation-based one) of the multidimensional parameters.
-We have briefly explained this in p. <!-- # TODO: Complete -->, ll. <!-- # TODO: Complete --> (please refer to the changes explained above).
+We have briefly explained this in p. 24, ll. 17-25 (please refer to the changes explained above).
 

### 2025-11-04 output/Response_to_action_editor_reviewers-2.Rmd

@@
 > or $`r TRAIT_COORDS_TRANSF_TS`$ for $`r TRAIT_VECTOR`$
-> (as given by Equation \@ref(eq:trait-coords-spaces-transform)).
+> (as given by Equation (24)).
 > For the sake of simplicity,

### 2025-11-04 output/IMPS-2025-slideshow.qmd

@@
 ) +
-  geom_segment(
-    arrow     = arrow(angle = 20, length = unit(10, "points"), type = "closed"),
-    linejoin  = "mitre",
-    linewidth = VECTOR_WIDTH,
-    data      = items_oblique |> arrange(desc(item)),
-    mapping   = aes(
-      origin_transf_1, origin_transf_2,
-      xend  = end_transf_1, yend = end_transf_2,
-      color = item, fill = item
-    )
-  ) +
-  scale_color_manual(values = PALETTE, guide = NULL) +
+  items_geom_oblique +
+  colorscale_no_legend +
   coord_fixed(expand = FALSE, clip = "on")

### 2025-11-04 output/Title_page.Rmd

@@
 
-[swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af](https://archive.softwareheritage.org/swh:1:dir:3914435e67a2b34847e36a712a181d7d9d76f35e;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:9069b256471ea08d5748e8bb021b7ee876aee724;anchor=swh:1:rev:fe49f438bf940606ae7e3c86c3c55f202e6920af)
+[swh:1:dir:6195936c6e16023c4a0b6a4500b81a34469d64f1;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:410cf1a6fb4dd4c10bdf0ce7f12f6890f560549d;anchor=swh:1:rev:1f1a708a7477dd61c0ba5c82dcf1d8523ab3d8ff](https://archive.softwareheritage.org/swh:1:dir:6195936c6e16023c4a0b6a4500b81a34469d64f1;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:410cf1a6fb4dd4c10bdf0ce7f12f6890f560549d;anchor=swh:1:rev:1f1a708a7477dd61c0ba5c82dcf1d8523ab3d8ff)
 

### 2025-11-04 output/seminar-slideshow.qmd

@@
 
-plot_orth <- plot_orth +
-  scale_color_manual(values = ITEMS_PALETTE) +
+plot_orth <- plot_orth_uncorr_contours +
   theme(
@@
 
-plot_oblique <- plot_oblique +
-  scale_color_manual(values = ITEMS_PALETTE, guide = NULL) +
+plot_oblique <- plot_oblique_ts +
   theme(axis.ticks = element_blank())

### 2025-11-04 output/Response_to_action_editor_reviewers-2.Rmd
Correct page and line number references
Some references in the response to the action editor and reviewer were outdated

@@
 We deem this point important because it took us quite some effort to notice that generalizing the transformation to polar coordinates to non-orthonormal bases implied reformulating this equation.
-In an attempt to keep the intended meaning while improving the sentence structure, we have restated it as:
+In an attempt to keep the intended meaning while improving the sentence structure, we have restated it as (p. 12, ll. 3-4):
 
@@
 
-In p. 13 (line 2), dropping "up", i.e., "the squared direction cosines sum ~up~ to 1", did not sound well to us, so instead we reformulated the sentence as:
+In p. 13 (line 2), dropping "up", i.e., "the squared direction cosines sum ~up~ to 1", did not sound well to us, so instead we reformulated the sentence as (p. 12, ll. 20-21):
 
@@
 we consider this text was misplaced in our previous version and have moved it accordingly, nearly to the end of the section.
-After the changes, this paragraph (now in p. 38, l. 19 to p. 39, l. 18) reads:
+After the changes, this paragraph (now in p. 41, l. 4 to p. 41, l. 3) reads:
 
@@
 We acknowledge though that this might not be necessarily clear in the previous submission.
-The second change in the discussion consists of stating this explicitly now at the end of the fourth paragraph (previously the fifth one) of the "Discussion" section, in p. 37, ll. 16-18:
+The second change in the discussion consists of stating this explicitly now at the end of the fourth paragraph (previously the fifth one) of the "Discussion" section, in p. 40, ll. 1-3:
 
@@
 In the "Covariance-based version" section, we have referred to the transformation of the latent trait distribution into one with an identity covariance matrix as  a "whitening transformation", its usual name in the literature (unknown to us before).
-The second paragraph of this section (p. 16, ll. 18-20) now reads:
+The second paragraph of this section (p. 16, ll. 16-19) now reads:
 
@@
 which are only necessary for deriving the results.
-This choice of labeling is explained in the text in p. 23, ll. 1-8 (see also our response to the last comment below):
+This choice of labeling is explained in the text in p. 23, ll. 4-11 (see also our response to the last comment below):
 
@@
 
-Subsequently, we have updated ll. 14-15 in p. 15 (now ll. p. 18-19) as follows:
+Subsequently, we have updated ll. 14-15 in p. 15 (now ll. 17-18) as follows:
 

### 2025-11-11 output/Main_text.Rmd
Update code repository link

@@
 indexed Software Heritage repository for this paper, specifically:
-[URL obfuscated for blind peer-review]
+[swh:1:dir:f0a8de9085b7b72cd04ffb4a9a9c245a84ab5179;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:076d4a764ef2bbd1d961cccff56b76c59f8aa81e;anchor=swh:1:rev:11084a02d8365a878fe2135f46a459ce7044b4b2](https://archive.softwareheritage.org/swh:1:dir:6195936c6e16023c4a0b6a4500b81a34469d64f1;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:410cf1a6fb4dd4c10bdf0ce7f12f6890f560549d;anchor=swh:1:rev:1f1a708a7477dd61c0ba5c82dcf1d8523ab3d8ff)
 
@@
 
-```{r set-flextable-wd}
-#| cache: false
-```
-
-\newpage
-
-```{r empirical-table-Reckase2009-out}
-#| cache:   false
-#| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: |
-#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
-#|   Reckase (2009), With a (Rank-Complete) Three-Dimensional Covariance Matrix
-item_params_output
-```
-
-<br>
-
-```{r set-cov-matrix-Tezza2018}
-# See Tezza et al., 2018, p. 927:
-# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
-# > foi de aproximadamente 0,4.
-COV_MATRIX_VALUE <- diag(4)
-DIMS_CORR        <- .4
-COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
-```
-
-```{r reset-root-wd}
-#| cache: false
-```
-
-```{r read-items-Tezza2018}
-items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
-```
-
-```{r compute-orthogonal-params}
-```
-
-```{r compare-orthogonal-params}
-#| results: hide
-```
-
-<!-- Computed item parameters in the orthonormal space match the ones in given
-in Tezza 2018, table 5 -->
-
-```{r compute-oblique-params}
-```
-
-```{r format-covariance-matrix}
-```
-
-```{r create-params-table}
-```
-
-```{r compose-output-table}
-```
-
-<!---BLOCK_LANDSCAPE_START--->
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Tezza2018-out}
-#| cache:   false
-#| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: |
-#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
-#|   Tezza et al. (2018)
-item_params_output |>
-  fontsize(size = 11, part = "body")
-```
-
-<!---BLOCK_LANDSCAPE_STOP--->
-
 The latter effect can be better noticed in another case, in which Reckase
@@
 in the table in Online Resource 1.
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+\newpage
+
+```{r empirical-table-Reckase2009-out}
+#| cache:   false
+#| tab.id:  empirical-table-Reckase2009-out
+#| tab.cap: |
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Reckase (2009), With a (Rank-Complete) Three-Dimensional Covariance Matrix
+item_params_output
+```
+
+<br>
+
+```{r set-cov-matrix-Tezza2018}
+# See Tezza et al., 2018, p. 927:
+# > No presente estudo, a correlação entre a dimensão 1 e a dimensão 4
+# > foi de aproximadamente 0,4.
+COV_MATRIX_VALUE <- diag(4)
+DIMS_CORR        <- .4
+COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
+```
+
+```{r reset-root-wd}
+#| cache: false
+```
+
+```{r read-items-Tezza2018}
+items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
+```
+
+```{r compute-orthogonal-params}
+```
+
+```{r compare-orthogonal-params}
+#| results: hide
+```
+
+<!-- Computed item parameters in the orthonormal space match the ones in given
+in Tezza 2018, table 5 -->
+
+```{r compute-oblique-params}
+```
+
+```{r format-covariance-matrix}
+```
+
+```{r create-params-table}
+```
+
+```{r compose-output-table}
+```
+
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Tezza2018-out}
+#| cache:   false
+#| tab.id:  empirical-table-Tezza2018-out
+#| tab.cap: |
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Tezza et al. (2018)
+item_params_output |>
+  fontsize(size = 11, part = "body")
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
 

### 2025-11-11 output/Title_page.Rmd
Update code repository link

@@
 
-[swh:1:dir:6195936c6e16023c4a0b6a4500b81a34469d64f1;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:410cf1a6fb4dd4c10bdf0ce7f12f6890f560549d;anchor=swh:1:rev:1f1a708a7477dd61c0ba5c82dcf1d8523ab3d8ff](https://archive.softwareheritage.org/swh:1:dir:6195936c6e16023c4a0b6a4500b81a34469d64f1;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:410cf1a6fb4dd4c10bdf0ce7f12f6890f560549d;anchor=swh:1:rev:1f1a708a7477dd61c0ba5c82dcf1d8523ab3d8ff)
+[swh:1:dir:f0a8de9085b7b72cd04ffb4a9a9c245a84ab5179;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:076d4a764ef2bbd1d961cccff56b76c59f8aa81e;anchor=swh:1:rev:11084a02d8365a878fe2135f46a459ce7044b4b2](https://archive.softwareheritage.org/swh:1:dir:6195936c6e16023c4a0b6a4500b81a34469d64f1;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:410cf1a6fb4dd4c10bdf0ce7f12f6890f560549d;anchor=swh:1:rev:1f1a708a7477dd61c0ba5c82dcf1d8523ab3d8ff)
 

### 2025-11-11 src/Graphical_example_paper.R
Re-position figure and tables to avoid blanks

@@ -280,7 +280,7 @@ item_params_output <- item_params                          |>
   )                                                        |>
   style(
     pr_p = fp_par(
-      line_spacing   = 2,
+      line_spacing   = 1.5,
       padding.top    = 2,
       padding.bottom = 0
     ),

### 2025-11-11 output/AM_Main.Rmd
Collapse main and title files in "AM_Main" file
As requested by the journal final submission

--- 2025-11-11 output/Main_text.Rmd
@@
         sep: ' '
-    mapstyles:
-      Body Text: ['First Paragraph', 'Abstract']
     base_format: bookdown::word_document2
@@
 ```
+
+**Article title:**
+A Generalized Definition of Multidimensional Item Response Theory Parameters
+
+::: {custom-style="First Paragraph"}
+**Running title:** Generalized MIRT parameters
+:::
+
+# List of authors
+
+- Daniel Morillo-Cuadrado. Department of Behavioral Science Methodology,
+  Faculty of Psychology, Universidad Nacional de Educación a Distancia (UNED),
+  Spain.
+
+- Mario Luzardo-Verde. Department of Mathematics and Applications,
+  Universidad de La República, Uruguay.
+
+## Corresponding author:
+
+Daniel Morillo-Cuadrado \
+Faculty of Psychology, Office 2.74. \
+C. de Juan del Rosal 10, 28040 Madrid
+
+Phone nº: +34 91398 6584 \
+Email address: [danielmorillo.ac@gmail.com](mailto:danielmorillo.ac@gmail.com)
+
+# Acknowledgements
+
+The authors would like to thank:
+[Patricia Recio-Saboya, Ph.D.](https://orcid.org/0000-0002-0530-4404),
+for her guidance regarding multidimensional estimation
+in IRT software packages;
+[Jay Verkuilen, Ph.D.](https://orcid.org/0000-0003-3546-3411),
+for his comments and suggestions regarding covariance whitening;
+David Thissen and three additional anonymous reviewers for
+their valuable comments on early drafts of this manuscript.
+
+# Data Availability Statement
+
+All the materials and code of this paper are available as
+a Software Heritage repository, under the following SWHID persistent identifier:
+
+[swh:1:dir:f0a8de9085b7b72cd04ffb4a9a9c245a84ab5179;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:076d4a764ef2bbd1d961cccff56b76c59f8aa81e;anchor=swh:1:rev:11084a02d8365a878fe2135f46a459ce7044b4b2](https://archive.softwareheritage.org/swh:1:dir:6195936c6e16023c4a0b6a4500b81a34469d64f1;origin=https://github.com/DaniMori/mirt-parameters;visit=swh:1:snp:410cf1a6fb4dd4c10bdf0ce7f12f6890f560549d;anchor=swh:1:rev:1f1a708a7477dd61c0ba5c82dcf1d8523ab3d8ff)
+
+# Financial Support
+
+This research received no specific grant funding form any funding agency,
+commercial or not-for-profit sectors.
+
+# Competing Interests Statement
+
+Competing interests: The authors declare none.
+
+\newpage
 
@@
 
+<br>
+
+::: {custom-style="First Paragraph"}
 *Keywords:* Multidimensional 2-Parameter Logistic model,
@@
 item vector representation, test space
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Multidimensional item response theory (MIRT) models have been widely used since
@@
 it is sometimes referred to as a *compensatory* model.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Consider a basis $`r LS_BASIS_EQ`$,
@@
 a positive response to item $`r ITEM_INDEX`$ is
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
@@
 in Equation \@ref(eq:M2PL-formulation) we identify $`r TRAIT_GEOM_ALGEBR_EQ`$.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Let $`r TEST_SPACE`$ be the vector space made up by
@@
 defined by
+:::
 
@@
 
+::: {custom-style="Body Text"}
 The cosine between two vectors
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Let's consider now the vector $`r DIR_COS_VEC_ORG`$, made up by the $`r N_DIMS`$
@@
 the basis vectors of $`r LS_BASIS`$.
+:::
 
@@
 
+::: {custom-style="First Paragraph"}
 **Proposition 1** Under the previously defined conditions, for all
 $`r TRAIT_VEC_IN_LS`$,
+:::
 
@@
 
+::: {custom-style="First Paragraph"}
 **Proposition 2** Let's consider another basis $`r ALT_BASIS_EQ`$ in
@@
 the norm of the space invariant.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Given i,
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Applying Proposition 1 to
@@
 the `r MODEL_ACRONYM` model is expressed in polar coordinates as
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Following the procedure outlined by @reckase_difficulty_1985
@@
 the second derivative with respect to $`r TRAIT_NORM`$.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 The slope $`r SLOPE_MAX_PARAM`$ at the point defined by Equations
@@
 [@reckase_discriminating_1991],
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Before presenting the results about the parameters,
@@
 they are still two different spaces.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 The item location $`r MIL_PARAM`$ is defined as the distance and direction
@@
 the parameters.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Up to this point,
@@
 we shall refer to them as the *agnostic* version of the parameters:
+:::
 
@@
 
+::: {custom-style="Body Text"}
 On the other hand, we can let the geometrical representation of
@@
 with $`r VAR_COV_EQ`$ (i.e., the variances).
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Therefore, we may define $`r TRANSFORM_MATRIX`$ such that
@@
 resulting in the direction cosines defined by
+:::
 
@@
 
+::: {custom-style="Body Text"}
 According to @reckase_discriminating_1991, the multidimensional parameters must
@@
 of the (unidimensional) IRT parameters:
+:::
 
@@
 
+::: {custom-style="Body Text"}
 The second property derives straightforwardly from
@@
 the Mahalanobis distance, i.e., $`r MAH_DIST_EQ`$ [@mahalanobis_reprint_2018].
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
@@
 Then, we have that
+:::
 
@@
 
+::: {custom-style="Body Text"}
 As we can see, this version also fulfills the scale invariance property,
@@
 as we will see later.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 According to @ackerman_creating_1994, the most appropriate way of representing
@@
 denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
@@
 we can express the coordinates in terms of the model parameters as
+:::
 
@@
 
+::: {custom-style="Body Text"}
 We have seen how the two new versions of the parameters fulfill
@@
 the distance to the origin, and the measurement direction.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 The geometric properties are easier to apprehend with
@@
 the correlation-based version will thus be preferred.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Instead of steps 1 to 3,
@@
 and thus this matrix is necessary in any case to compute the coordinates.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
@@
 $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 In all four panels, the contours of a bivariate standard normal density
@@
 $`r COV_MATRIX_TRANSF_TS_EQ`$.
+:::
+
+```{r density-contours}
+```
+
+```{r compose-orthogonal-plot}
+```
+
+```{r compose-test-space-oblique-plot}
+```
+
+```{r compose-test-space-orthogonal-plot}
+```
+
+```{r compose-rectangular-cov-based-plot}
+```
+
+<br>
+
+```{r item-plot-out}
+#| cache:          false
+#| fig.cap:        |
+#|   Item vector plots with uncorrelated (a) and correlated latent dimensions.
+#|   The latter coordinates are computed with
+#|   (b) the agnostic multidimensional parameters, or
+#|   with the correlation-based ones, which are then plotted either
+#|   (c) in an oblique basis, appropriate for the covariance structure, or
+#|   (d) in the canonical basis.
+#| fig.topcaption: true
+#| fig.height:     4.7
+#| fig.width:      6.65
+#| warning:        false
+(plot_orth_uncorr_contours + plot_orth_corr) /
+  (plot_oblique_ts + plot_cov_based_orth) +
+  plot_layout(guides = "collect")
+
+ggsave(file.path(DOC_DIR, "Figure_1.svg"), width = 6.65, height = 4.7)
+```
+
+::: {custom-style="FigureNote"}
+*Note.* The items are represented along with
+a bivariate standard normal distribution, with null correlation (a),
+or with correlation $`r CORR_OBL_OUT`$ (b-d).
+The contour plots represent (from outer to inner)
+the `r contour_breaks_out` of the maximum density.
+:::
+
+```{r compose-example-items-table}
+```
+
+```{r set-flextable-wd}
+#| cache: false
+
+# Necessary for officedown to find the template file:
+opts_knit$set(root.dir = DOC_DIR)
+```
+
+```{r example-items-table-out}
+#| cache:   false
+#| tab_id:  example-items-table-out
+#| tab.cap: Item Parameters for the Graphical Representation Example
+item_params_output
+```
+
+<br>
 
@@
 
-```{r compose-example-items-table}
-```
-
-```{r set-flextable-wd}
-#| cache: false
-
-# Necessary for officedown to find the template file:
-opts_knit$set(root.dir = DOC_DIR)
-```
-
-<br>
-
-```{r example-items-table-out}
-#| cache:   false
-#| tab_id:  example-items-table-out
-#| tab.cap: Item Parameters for the Graphical Representation Example
-item_params_output
-```
-
-<br>
-
 A comparison of panels (a) and (c) allows understanding the effect of
@@
 
-```{r density-contours}
-```
-
-```{r compose-orthogonal-plot}
-```
-
-```{r compose-test-space-oblique-plot}
-```
-
-```{r compose-test-space-orthogonal-plot}
-```
-
-```{r compose-rectangular-cov-based-plot}
-```
-
-<br>
-
-```{r item-plot-out}
-#| cache:          false
-#| fig.cap:        |
-#|   Item vector plots with uncorrelated (a) and correlated latent dimensions.
-#|   The latter coordinates are computed with
-#|   (b) the agnostic multidimensional parameters, or
-#|   with the correlation-based ones, which are then plotted either
-#|   (c) in an oblique basis, appropriate for the covariance structure, or
-#|   (d) in the canonical basis.
-#| fig.topcaption: true
-#| fig.height:     4.7
-#| fig.width:      6.65
-#| warning:        false
-(plot_orth_uncorr_contours + plot_orth_corr) /
-  (plot_oblique_ts + plot_cov_based_orth) +
-  plot_layout(guides = "collect")
-
-ggsave(file.path(DOC_DIR, "Figure_1.svg"), width = 6.65, height = 4.7)
-```
-
-::: {custom-style="FigureNote"}
-*Note.* The items are represented along with
-a bivariate standard normal distribution, with null correlation (a),
-or with correlation $`r CORR_OBL_OUT`$ (b-d).
-The contour plots represent (from outer to inner)
-the `r contour_breaks_out` of the maximum density.
-:::
-
-<br>
-
 We can also see how the sign of $`r INTERCEPT_PARAM`$ affects
@@
 
+::: {custom-style="Body Text"}
 In the following, we apply our results to two instances taken from
@@
 items with opposite-sign discrimination parameters.
+:::
 
@@
 
+::: {custom-style="Body Text"}
 Table \@ref(tab:empirical-table-Reckase2009-out) shows the parameters of the
@@
 is standard with null correlations; however, the second one has
+:::
 
@@
 
+::: {custom-style="Body Text"}
 As we can see in Table \@ref(tab:empirical-table-Reckase2009-out),
@@
 Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out)).
+:::
 
@@
 
-```{r set-flextable-wd}
-#| cache: false
-```
-
-\newpage
-
-```{r empirical-table-Reckase2009-out}
-#| cache:   false
-#| tab.id:  empirical-table-Reckase2009-out
-#| tab.cap: |
-#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
-#|   Reckase (2009), With a (Rank-Complete) Three-Dimensional Covariance Matrix
-item_params_output
-```
-
-<br>
+## @tezza_modelo_2018
 
@@
 
-```{r reset-root-wd}
-#| cache: false
-```
-
-```{r read-items-Tezza2018}
-items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
-```
-
-```{r compute-orthogonal-params}
-```
-
-```{r compare-orthogonal-params}
-#| results: hide
-```
-
-<!-- Computed item parameters in the orthonormal space match the ones in given
-in Tezza 2018, table 5 -->
-
-```{r compute-oblique-params}
-```
-
-```{r format-covariance-matrix}
-```
-
-```{r create-params-table}
-```
-
-```{r compose-output-table}
-```
-
-<!---BLOCK_LANDSCAPE_START--->
-
-```{r set-flextable-wd}
-#| cache: false
-```
-
-```{r empirical-table-Tezza2018-out}
-#| cache:   false
-#| tab.id:  empirical-table-Tezza2018-out
-#| tab.cap: |
-#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
-#|   Tezza et al. (2018)
-item_params_output |>
-  fontsize(size = 11, part = "body")
-```
-
-<!---BLOCK_LANDSCAPE_STOP--->
-
-## @tezza_modelo_2018
-
+::: {custom-style="Body Text"}
 @tezza_modelo_2018 provide us with another example that showcases
@@
 other two dimensions, and around 1º in most cases.
+:::
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Reckase2009-out}
+#| cache:   false
+#| tab.id:  empirical-table-Reckase2009-out
+#| tab.cap: |
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Reckase (2009), With a (Rank-Complete) Three-Dimensional Covariance Matrix
+item_params_output
+```
+
+\newpage
+
+```{r reset-root-wd}
+#| cache: false
+```
+
+```{r read-items-Tezza2018}
+items <- read_csv2(ITEM_PARAMS_TEZZA_2018_FILEPATH)
+```
+
+```{r compute-orthogonal-params}
+```
+
+```{r compare-orthogonal-params}
+#| results: hide
+```
+
+<!-- Computed item parameters in the orthonormal space match the ones in given
+in Tezza 2018, table 5 -->
+
+```{r compute-oblique-params}
+```
+
+```{r format-covariance-matrix}
+```
+
+```{r create-params-table}
+```
+
+```{r compose-output-table}
+```
+
+<!---BLOCK_LANDSCAPE_START--->
+
+```{r set-flextable-wd}
+#| cache: false
+```
+
+```{r empirical-table-Tezza2018-out}
+#| cache:   false
+#| tab.id:  empirical-table-Tezza2018-out
+#| tab.cap: |
+#|   Agnostic and Covariance-Based Multidimensional Item Parameters in
+#|   Tezza et al. (2018)
+item_params_output |>
+  fontsize(size = 11, part = "body")
+```
+
+<!---BLOCK_LANDSCAPE_STOP--->
 
@@
 
+::: {custom-style="Body Text"}
 @reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
@@
 case of the `r MODEL_ACRONYM` model.
+:::
 

### 2025-12-01 output/AM_Main.Rmd
Add changes in typesetting proof

@@
 with item coordinates computed from the agnostic version of the parameters
-(panel (b)), or from the correlation-based version,
-either using an appropriate oblique basis (panel (c)),
-or plotting them directly in the canonical basis (panel (d)).
+(panel [b]), or from the correlation-based version,
+either using an appropriate oblique basis (panel [c]),
+or plotting them directly in the canonical basis (panel [d]).
 Keeping the horizontal axis invariant in panel (c)
@@
 The correlation-based coordinates, plotted in panels (c) and (d),
-are also different in general from their agnostic counterpart (panel (b)).
+are also different in general from their agnostic counterpart (panel [b]).
 Although they do not vary in some cases (items 1 and 4),
@@
 the item locations and their positions relative to the distribution
-are correctly represented (which makes comparing with panel (b) easier),
-but their discrimination parameters and angles with the axes are distorted:
+are correctly represented (which makes comparing with panel [b] easier),
+but their discrimination parameters and angles with the axes are distorted.
 For example, items 4 and 5 seem to be equally discriminating,

### 2025-12-01 output/AM_Main.Rmd
Add corrections in proof.

@@
 
-- Daniel Morillo-Cuadrado. Department of Behavioral Science Methodology,
-  Faculty of Psychology, Universidad Nacional de Educación a Distancia (UNED),
-  Spain.
+- Daniel Morillo-Cuadrado. Statistical and Computational Methods in Psychology
+  Group, Department of Behavioral Science Methodology, School of Psychology,
+  Universidad Nacional de Educación a Distancia (UNED), Spain.
 
@@
 difficulty (MID) parameters in the Multidimensional 2-Parameter Logistic model
-to account for non-identity latent covariances and negatively-keyed items.
+to account for nonidentity latent covariances and negatively keyed items.
 We apply Reckase's maximum discrimination point method to define them in an
@@
 geometrical representation of the measured construct.
-This results in three different versions of the parameters: The original one,
+This results in three different versions of the parameters: the original one,
 based on the item parameters solely, one that incorporates the covariance
@@
 
-::: {custom-style="Body Text"}
 Multidimensional item response theory (MIRT) models have been widely used since
@@
 it is sometimes referred to as a *compensatory* model.
-:::
 
@@
 This parameter consists of two parts:
-The first one is a scalar value representing the distance from
+the first one is a scalar value representing the distance from
 the origin of the multidimensional latent space to
@@
 axes is prevalent in the common factor literature
-[see, e.g., @harman_modern_1976]:
+[see, e.g., @harman_modern_1976].
 Indeed, the cosine between any two axes (or, more generally, vectors)
@@
 Take the following as an example:
-A hypothetical complex item measures two latent traits with equal validity
----i.e., its discrimination parameters are equal with value $`r DISCR_SYM`$.
+a hypothetical complex item measures two latent traits with equal validity
+---that is, its discrimination parameters are equal with value $`r DISCR_SYM`$.
 Applying Reckase & McKinley's [-@reckase_discriminating_1991] formula,
 we would find that $`r AG_MDISC_EXAMPLE`$.
-Assume we apply this item in two groups where, taken to the extreme,
+Assume that we apply this item in two groups where, taken to the extreme,
 the two measured dimensions are orthogonal in the first group
@@
 [@ackerman_multidimensional_2005-1, p. 16, see footnote 1]
-and (possibly) non-standard units,
+and (possibly) nonstandard units,
 often referred to as *general Cartesian coordinates*
@@
 The general case of MIRT model estimation implies estimating possibly
-non-standard latent dimensions and non-null covariances among them
+nonstandard latent dimensions and nonnull covariances among them
 (a more realistic case than an identity matrix).
 Assuming orthonormality among traits is usually unrealistic, be it in the
-cognitive [e.g., @reckase_multidimensional_2009] or the non-cognitive domain
+cognitive [e.g., @reckase_multidimensional_2009] or the noncognitive domain
 [e.g., @thielmann_comparability_2022].
 Moreover, certain applications impose the estimation of
-non-identity latent covariance matrices (e.g., multi-group equating) even when
+nonidentity latent covariance matrices (e.g., multi-group equating) even when
 interpreting the latent dimensions as substantive constructs is not a goal
@@
 However, when such instances of negative parameters appear,
-they are often dismissed as "particularly puzzling" (p. 109),
-or implicitly assuming item malfunctioning without further discussion
+literature often dismisses them as "particularly puzzling" (p. 109)
+or implicitly assumes item malfunctioning without further discussion
 [e.g., @ackerman_multidimensional_2005-1].
-In the non-cognitive domain though, the use of so-called negatively-keyed items
+In the noncognitive domain though, the use of so-called negatively keyed items
 ---items that tap the negative pole of a certain trait, and thus
@@
 of the Big-Five Factor Markers, @noauthor_international_nodate].
-Instruments that combine positively- and negatively-keyed items
+Instruments that combine positively and negatively keyed items
 are relatively common, and
@@
 a monotonically decreasing probability for a certain dimension
-is given by a non-positive discrimination parameter
+is given by a nonpositive discrimination parameter
 (and an unchanging probability by a null discrimination parameter,
@@
 However, the multidimensional item parameters can also be defined for
-items with monotonically non-increasing probabilities, as we will see.
+items with monotonically nonincreasing probabilities, as we will see.
 
 The main objective of this paper is thus to provide a generalization of
-the MIRT parameters in two aspects: on one hand, in an oblique coordinate system
-that can be soundly applied to the correlated trait case; on the other, to
+the MIRT parameters in two aspects: first, in an oblique coordinate system
+that can be soundly applied to the correlated trait case, and second, to
 items that yield monotonic response probabilities, but in the broader sense of
@@
 and present two examples of application of our results.
-These examples, from the cognitive and the non-cognitive domain, respectively,
+These examples, from the cognitive and the noncognitive domain, respectively,
 are taken from the MIRT literature:
-One of them is a classic dataset, often used to exemplify
+one of them is a classic dataset, often used to exemplify
 the use of the `r MODEL_ACRONYM` model [@reckase_multidimensional_2009];
@@
 
-::: {custom-style="Body Text"}
 Consider a basis $`r LS_BASIS_EQ`$,
@@
 a positive response to item $`r ITEM_INDEX`$ is
-:::
 
@@
 But it may mean endorsing an item (or a certain response option)
-in a non-cognitive domain.
+in a noncognitive domain.
 Second, as responses are not necessarily interpreted as correct or incorrect,
@@
 Third, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
-non-negative real numbers in @reckase_difficulty_1985
+nonnegative real numbers in @reckase_difficulty_1985
 ---a decrease in the probability of a correct response
@@
 of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
-the item is negatively-keyed for latent dimension $`r DIM_INDEX`$.
+the item is negatively keyed for latent dimension $`r DIM_INDEX`$.
 Fourth, $`r DISCR_VECTOR`$ is expressed in some algebraic basis,
@@
 which will be present whenever more than one component of $`r DISCR_VECTOR`$
-is non-null.
+is nonnull.
 However, the model can also account for *between-item multidimensionality*
-if the $`r DISCR_VECTOR`$ parameter only has one non-null component.
+if the $`r DISCR_VECTOR`$ parameter only has one nonnull component.
 
@@
 
-::: {custom-style="Body Text"}
-Let's consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in $`r LATENT_SPACE`$
-with the dot product as inner product.
+Let us consider an orthonormal basis $`r LS_ORTH_BASIS_EQ`$ in
+$`r LATENT_SPACE`$ with the dot product as inner product.
 Let $`r TRAIT_COORDS_EQ`$ and $`r TRAIT_COORDS_ORTH_EQ`$ be the coordinates of
@@
 in Equation \@ref(eq:M2PL-formulation) we identify $`r TRAIT_GEOM_ALGEBR_EQ`$.
-:::
 
@@
 
-::: {custom-style="Body Text"}
 Let $`r TEST_SPACE`$ be the vector space made up by
@@
 defined by
-:::
 
@@
 
-::: {custom-style="Body Text"}
 The cosine between two vectors
-:::
 
@@
 ::: {custom-style="Body Text"}
-Let's consider now the vector $`r DIR_COS_VEC_ORG`$, made up by the $`r N_DIMS`$
-direction cosines of $`r TRAIT_VECTOR`$ with
+Let us consider now the vector $`r DIR_COS_VEC_ORG`$, made up by
+the $`r N_DIMS`$ direction cosines of $`r TRAIT_VECTOR`$ with
 the basis vectors of $`r LS_BASIS`$.
@@
 ::: {custom-style="First Paragraph"}
-**Proposition 1** Under the previously defined conditions, for all
-$`r TRAIT_VEC_IN_LS`$,
+**Proposition 1** *Under the previously defined conditions, for all
+$`r TRAIT_VEC_IN_LS`$*,
 :::
@@
 
-with
+*with*
 
@@
 
-Therefore,
+*Therefore,*
 
@@
 
-Therefore, $`r DIR_COS_VEC_EQ`$ and $`r TRAIT_VECTOR_POLAR_EQ`$.
-$`r END_OF_PROOF`$
+Therefore, $`r DIR_COS_VEC_EQ`$ and
+$`r TRAIT_VECTOR_POLAR_EQ`$.$`r END_OF_PROOF`$
 :::
@@
 ::: {custom-style="First Paragraph"}
-**Proposition 2** Let's consider another basis $`r ALT_BASIS_EQ`$ in
+**Proposition 2** *Let us consider another basis $`r ALT_BASIS_EQ`$ in
 $`r LATENT_SPACE`$ and the inner product $`r ALT_BASIS_INNER_PROD`$ that keeps
-the norm of the space invariant.
+the norm of the space invariant.*
 :::
 
-Let $`r DIR_COS_VEC_STD`$ and $`r DIR_COS_VEC_ALT`$ be the vectors of
+*Let $`r DIR_COS_VEC_STD`$ and $`r DIR_COS_VEC_ALT`$ be the vectors of
 direction cosines of $`r TRAIT_VECTOR`$ in $`r LS_ORTH_BASIS`$ and
-$`r ALT_BASIS`$, respectively.
-
-Then,
+$`r ALT_BASIS`$, respectively.*
+
+*Then,*
 
@@
 
-where
+*where*
 
@@
 
-and $`r TRANSFORM_MATRIX_ALT_EQ`$.
+*and $`r TRANSFORM_MATRIX_ALT_EQ`$.*
 
@@
 
-Let's call $`r TRAIT_COORDS_ALT_EQ`$ and let $`r TRANSFORM_MATRIX_ALT_EQ`$
+Let us call $`r TRAIT_COORDS_ALT_EQ`$ and let $`r TRANSFORM_MATRIX_ALT_EQ`$
 be the change of basis matrix, such that $`r TRAIT_TRANSFORM_ALT_EQ`$.
@@
 
-::: {custom-style="Body Text"}
 Applying Proposition 1 to
@@
 the `r MODEL_ACRONYM` model is expressed in polar coordinates as
-:::
 
@@
 
-::: {custom-style="Body Text"}
 Following the procedure outlined by @reckase_difficulty_1985
@@
 the second derivative with respect to $`r TRAIT_NORM`$.
-:::
 
@@
 To compute the direction of the maximum slope,
-Reckase leverages the property that
-the sum of the squared direction cosines equals 1 (see Equation 7 therein).
+Reckase leverages the property that the sum of the squared direction cosines
+equals 1 (see Equation 7 therein, p. 404).
 However, this assertion implicitly assumes orthogonal coordinates,
@@
 
-::: {custom-style="Body Text"}
 Before presenting the results about the parameters,
@@
 they are still two different spaces.
-:::
 
@@
 the coordinates of the discrimination vectors in the latent space:
-By Equations \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
+by Equations \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
 and  \@ref(eq:trait-coords-orth-spaces-eq),
@@
 given that the test space is defined as the set of item parameters
-(see section *Test Space Definition*),
+(see Section 2.1),
 then they should be represented in this space, and not in the latent space.
@@
 
-::: {custom-style="Body Text"}
-The item location $`r MIL_PARAM`$ is defined as the distance and direction
+The multidimensional item location ($`r MIL_PARAM`$) is defined as
+the distance and direction
 from the origin to the point of maximum slope [@reckase_difficulty_1985].
@@
 the parameters.
-:::
 
@@
 
-::: {custom-style="Body Text"}
 Up to this point,
@@
 we shall refer to them as the *agnostic* version of the parameters:
-:::
 
@@
 
-::: {custom-style="Body Text"}
 On the other hand, we can let the geometrical representation of
@@
 with $`r VAR_COV_EQ`$ (i.e., the variances).
-:::
 
@@
 
-::: {custom-style="Body Text"}
 Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
@@
 while still fulfilling this property.
-To do this, let's assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
+To do this, let us assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
 $`r TRANSFORM_MATRIX_AUX_RESULT`$.
 Then, we have that
-:::
 
@@
 which we have used to justify the covariance-based version:
-The $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
+the $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
 $`r TRAIT_COORDS`$ to the orthonormal basis $`r LS_ORTH_BASIS`$
@@
 
-::: {custom-style="Body Text"}
 According to @ackerman_creating_1994, the most appropriate way of representing
@@
 denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.
-:::
 
@@
 
-::: {custom-style="Body Text"}
 We have seen how the two new versions of the parameters fulfill
@@
 the distance to the origin, and the measurement direction.
-:::
 
@@
 given by the first component of $`r MIL_PARAM`$, has an interesting property:
-By the definition of the inner product,
-$`r MDISC_SYM`$ is strictly positive for non-null vectors.
+by the definition of the inner product,
+$`r MDISC_SYM`$ is strictly positive for nonnull vectors.
 Therefore, the sign of $`r DISTANCE_PARAM`$ is the opposite of
@@
 This property leads directly to generalizing
-the parameters to the monotonically non-increasing case:
-The measurement direction relative to dimension $`r DIM_INDEX`$
+the parameters to the monotonically nonincreasing case:
+the measurement direction relative to dimension $`r DIM_INDEX`$
 will be negative when $`r IRF_ABBR`$ is monotonically decreasing
@@
 
-::: {custom-style="Body Text"}
 The geometric properties are easier to apprehend with
@@
 the correlation-based version will thus be preferred.
-:::
 
@@
 Therefore, we must compute the equivalent rectangular coordinates
-of our general, possibly non-rectangular coordinates.
+of our general, possibly nonrectangular coordinates.
 Coordinates in a rectangular Cartesian system are commonly assumed
@@
 
-::: {custom-style="Body Text"}
 Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
@@
 $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.
-:::
 
@@
 For example, items 4 and 5 seem to be equally discriminating,
-when Table \@ref(tab:example-items-table-out) panel (c) show they are not;
-and item 1 forms 45-degree angles with both axes,
+when Table \@ref(tab:example-items-table-out) and panel (c) show they are not,
+and item 1 forms 45-degree angles with both axes
 instead of the 30-degree ones shown in Table \@ref(tab:example-items-table-out).
@@
 (and consequently the $`r MIL_CORR_PARAM`$) value of an item:
-The components of the discrimination parameters tend to
+the components of the discrimination parameters tend to
 *sum up* in the same direction as they get *more aligned*.
@@
 the $`r MDISC_CORR_PARAM`$ values of items 3 and 5:
-Their discrimination parameters are equal in absolute value,
+their discrimination parameters are equal in absolute value,
 so their $`r MDISC_CORR_PARAM`$ values are equal in the orthogonal space;
@@
 the distance to the origin:
-An item vector is applied at the origin when its value is null (item 1),
+an item vector is applied at the origin when its value is null (item 1),
 and tends to be shifted
@@
 of the two parameters, but also by the correlation:
-Items 2 and 3 especially illustrate this effect,
+items 2 and 3 especially illustrate this effect,
 as each of them is orthogonal to one of the axes in the oblique space,
@@
 
-::: {custom-style="Body Text"}
 In the following, we apply our results to two instances taken from
@@
 Besides being an empirical example, applied to actual test data,
-it pertains to the non-cognitive domain and contains
+it pertains to the noncognitive domain and contains
 items with  negative discrimination parameters.
@@
 items with opposite-sign discrimination parameters.
-:::
 
@@
 In the first case, the direction angles are also coincident with the values
-provided in the original [@tezza_modelo_2018, do not provide these results].
+provided in the original [@tezza_modelo_2018 do not provide these results].
 In order to facilitate the comparison,
@@
 
-::: {custom-style="Body Text"}
 Table \@ref(tab:empirical-table-Reckase2009-out) shows the parameters of the
 30-item test in the @reckase_multidimensional_2009 example:
-There are three ten-item blocks assumed to approximate
-a simple structure: Items 1-10 measure dimension 1, items 11-20 dimension 3,
+there are three ten-item blocks assumed to approximate
+a simple structure: items 1-10 measure dimension 1, items 11-20 dimension 3,
 and items 21-30 dimension 2
@@
 In his example, Reckase tests the estimation of these items
-on two simulated datasets: In the first one, the latent trait distribution
+on two simulated datasets: in the first one, the latent trait distribution
 is standard with null correlations; however, the second one has
-:::
 
@@
 COV_MATRIX_VALUE[1, 4] <- COV_MATRIX_VALUE[4, 1] <- DIMS_CORR
-```
-
-::: {custom-style="Body Text"}
+
+dims_corr_out <- DIMS_CORR |> format_prop_like(sig = 1)
+```
+
 @tezza_modelo_2018 provide us with another example that showcases
@@
 In their model, dimensions 1 and 4 are estimated to have a correlation of
-`r DIMS_CORR`.
+`r dims_corr_out`.
 With this latent space structure, items that strongly discriminate in either or
@@
 other two dimensions, and around 1º in most cases.
-:::
 
@@
 
-::: {custom-style="Body Text"}
 @reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
@@
 but for a more general, semi-parametric formulation
-(although constrained to non-decreasing monotonicity).
+(although constrained to nondecreasing monotonicity).
 However, they provided no formal proof for this result either.
@@
 case of the `r MODEL_ACRONYM` model.
-:::
 
 With that purpose in mind, we have defined two new versions of these parameters:
-One that takes into account the whole covariance structure,
+one that takes into account the whole covariance structure,
 and another one that only considers the correlations among latent dimensions.
@@
 we propose adopting the covariance-based version of
-the multidimensional parameters as the most general definition,
+the multidimensional parameters as the most general definition
 and refer to them simply as
@@
 Therefore, we recommend its use for graphical representation purposes only,
-although in many applications both versions will be coincident, fortunately.
+although in many applications both versions will fortunately be coincident.
 As for the agnostic version, we recommend it only in
-the (unlikely) case of having no information about the latent space structure,
+the (unlikely) case of having no information about the latent space structure
 or when such structure is irrelevant
@@
 then compute the parameters in the transformed space using
-Reckase's [-@reckase_difficulty_1985] and
-Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
+the procedure of @reckase_difficulty_1985 and @reckase_discriminating_1991.
 After all, interpreting the axes of this space as substantive latent traits
@@
 we may compute the parameters in the orthogonal space
-(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
+(in fact, this is exactly what {mirt} functions MDISC() and MDIFF() do).
 Then, we may compute the direction cosines in the rotated solution by applying
@@
 the orthonormalized basis has provided us with an important insight:
-The distinction between the latent space and the test space.
+the distinction between the latent space and the test space.
 Our derivations show that, to properly represent the items,
@@
 the geometric equivalent of the correlations.
-This result is both surprising and counterintuitive
----Several instances in the MIRT literature suggest that
+This result is both surprising and counterintuitive;
+several instances in the MIRT literature suggest that
 "the angle between the axes [of the latent ability space] represents
@@
 First, we have only considered the case of dichotomous items,
-whereas non-cognitive applications often (and sometimes cognitive ones as well)
+whereas noncognitive applications often (and sometimes cognitive ones as well)
 require modeling several response categories.
